\section*{Introduction}

% \emph{Faylì'u a plltxe san Zola'u nìprrte' sìk zìya'u ye'rìn …}

Since its first inception in December of \on{2009}, the \LINK{https://files.learnnavi.org/dicts/NaviDictionary.pdf}{Taronyu} dictionary, compiled by Taronyu (\emph{alu} Richard Littauer), has been---and still remains---a vital part of the Na'vi community. Through its many iterations it is now taken care of by Markì (\emph{alu} Mark Miller) and still provides the first look at the main body of core vocabulary for the Na'vi language.

My activities as a \emph{karyu}, `teacher' in the community soon showed me that beginners would often take a word from the dictionary and apply it with the little grammatical knowledge that they had to form their first sentences. As helpful as this can be without examples to live by how a word is specifically used in Na'vi this could only end in confusion. 

Venturing into other languages and dictionaries I soon came to the realisation that I was dissatisfied with the information that the dictionary provided. In my opinion it could do so much more. My dictionaries of e.g. English, French, Swedish, and Irish all had at least a few example sentences to show the user how a word was in use, whether there were special cases, proverbs or idioms that this word was a part of. Especially with a conlang that was still in development and where there are no native speakers, I had the strong feeling that in order to provide good Na'vi one should lead with good examples and provide the necessary sources to examplify.

As late as \on{2011} I had also begun to collect all official example sentences by \emph{Karyu Pawl}, Dr. Paul Frommer who by then had started the \LINK{http://naviteri.org/}{Na'viteri} blog. These resources gave me a new spark to attempt another kind of dictionary.

% =====================

% Speaking a language is not only knowing the vocabulary and be familiar with the grammar. It also entails knowing which words go together. Not only proverbs and idioms but also certain parts of speech only go with certain verbs.

% It's not always clear how these go together and a look in a common dictionary will only show you the word. If you stack up the words you will get a hotch\hyp{}potch of Na'vi that more resembles English (or your respective native language) than cohesive Na'vi. This is especially true with prepositions. Anybody who has ventured into learning a new language will know that one word doesn't always translate in your language.

% The best way to learn is\slash know is by example. 

% Over the years since \emph{Avatar} has been released and the Learn Na'vi Community first formed there have been many attempts at a Na'vi dictionary--from \emph{Taronyu} to \emph{The \on{500} Most Commonly Used Na'vi Words}. Most of them only concentrate on the words rather than their interaction with each other. And sometimes the meaning of a word is obscured by the fact that the translation itself can have different meanings.

% This work tries to fill that gap. To lead (and teach) by example. 

% It has been attempted to find an official example of every given word in use and give it in the appropriate entry. `Official' is defined as coming directly from the source, i.e. Dr. Frommer himself (most often nowadays via his blog naviteri.org). This Dictionary also tries to be a point of reference. Therefore a link to every example is given. It may not always be the first mention of a word but sometimes subsequent information can help and be of value to interested parties.



