\newpage
\pdfbookmark[0]{Apostrophe}{'}
\begin{center}
\section*{' \I{[P]} (\emph{tìftang}, Apostrophe, Glottal Stop)}
\end{center}

\begin{multicols}{2}

% 'a'aw
\hi 
\HT{'a'aw}{\B{\cp{'a}'aw}} \I{["Pa.Paw]} \emph{adj.} (with singular of the noun; only with countable nouns) a few, several. 
\B{Lu po\-ru 'ey\-lan a\-'a\-'aw.} He has several friends.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Oel tse\-'a 'a\-'aw\-a tu\-tet.} I see several people.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\ding{43}~\HL{hol}{\B{hol}})

% 'ak 
\hi
\HT{'ak}{\B{'ak}} \I{[Pak\textcorner]} \emph{intj.} ow, ouch.~\LINK{https://forum.learnnavi.org/language-updates/correspondence-with-karyu-pawl/}{ln}

% 'akra
\hi 
\HT{'akra}{\B{\cp{'ak}ra}} \I{["Pak\textcorner.Ra]} \emph{n.} (fertile) soil (in which plants can grow). 
\B{'ak\-ra a\-pay\-nga'}, moist soil; 
\B{'ak\-ra le\-pay}, watery, saturated soil.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}

%'aku
\hi 
\HT{'aku}{\B{\cp{'a}ku}} \I{["Pa.ku]} \emph{vtr.} (\emph{a.~of clothing}) take off, remove.  
\B{Ru\-txe me\-hawn\-ven\-it 'i\-va\-ku.} Please take off your shoes.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
(\emph{b.~fig.~with} \B{ta}) remove, take away.  \B{Pot 'a\-ku fì\-tseng\-ta!} Get him out of here!~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
(\emph{c.~vulg.}) \B{'ä\-pi\-va\-ku}, get lost!~\LINK{http://forum.learnnavi.org/naviteri/new-vocabulary-clothing-19454/msg482699/\#msg482699}{ln} (\ding{214}~\HL{yemstokx}{\B{yem\-stokx}}, \HL{ioi si}{\B{\mbox{ioi} si}})

% 'al
\hi 
\HT{'al}{\B{'al}} \I{[Pal]} \emph{vtr.} waste.
\B{Rä\-'ä 'i\-val syu\-vet!} Don't waste food.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'al}{\B{tì\-'al}}, wastefulness. 
\on{2}.~\HL{le'al}{\B{le\-'al}}, wasteful (not for people). 
\on{3}. \HL{nI'al}{\B{nì\-'al}}, wastefully.

%'ali'ä
\hi
\HT{'ali'A}{\B{'a\cp{li}'ä}} \I{[Pa."li.P\ae]} \emph{n.} choker, collar. 
\B{'Ali\-'ät vul\-sìn key\-kur za\-'u fì\-tseng.} Hang the choker on the branch and come here.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% 'ampi
\hi
\HT{'ampi}{\B{\cp{'am}pi}} \I{["Pam.pi]} \emph{vtr.} (\texttt{+}\emph{control}) touch.   
\B{Oe\-ti 'am\-pi rä\-'ä, ma \mbox{skxawng}!} Don't touch me, you moron!~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}  
\B{Ze\-ne fko 'ivam\-pi prrnenit nì\-flrr fra\-krr.} One must always touch a baby gently.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\emph{a. idiom}) \B{(Na) lo\-re\-yu 'aw\-nam\-pi}, extreme shyness, [lit.: (like) a touched helicoradian] \B{Lu por mok\-ri a\-mik\-lor, slä lo\-re\-yu 'aw\-nam\-pi lu. Ke tsun ri\-vol eo su\-te.} She has a beautiful voice, but she's extremely shy. She can't sing in front of people.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}. \HL{'ampirikx}{\B{'am\-pi\-rikx}}, leaf pitcher plant.

% 'ampirikx
\hi 
\HT{'ampirikx}{\B{\cp{'am}pirikx}} \I{["Pam.pi.Rik']} \emph{n.}, \ding{96} leaf pitcher plant, \emph{pseudocenia simplex}. 
(\ding{43}~\HL{'ampi}{\B{'am\-pi}}, \HL{rikx}{\B{rikx}})

% 'ana
\hi
\HT{'ana}{\B{\cp{'a}na}} \I{["Pa.na]} \emph{n.}, \ding{96} hanging vine.
(\emph{a. proverb}) \B{Spä skxawng sìn 'a\-na a\-flì.} ``A fool jumps onto a thin vine,'' i.e. \emph{don't engage in an unpromising and\slash or potentially risky cause}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Tsa\-ye\-rik te\-rul ne 'awkx. Spä skxawng sìn 'a\-na a\-flì.} That hexapede is running toward the cliff. Only a fool jumps onto a thin vine.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{syep'an}{\B{syep\-'an}}, lift vine.

% 'anla
\hi
\HT{'anla}{\B{\cp{'an}la}} \I{["Pan.la]} \emph{vtr.} yearn for. 
\B{Tsa\-ye\-rik\-tsyìp\-ìl li 'an\-la sa'\-nok\-it.} That little hexapede is already yearning for its mother.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{sA'anla}{\B{sä\-'an\-la}}, yearning.

% 'angtsìk
\hi 
\HT{'angtsIk}{\B{\cp{'ang}tsìk}} \I{["PaN.\t{ts}Ik\textcorner]} \emph{n., fauna} hammerhead, \emph{rhinoquadruculus hammercephali}. 
\B{'Ang\-tsìk a kxap si lu le\-hrr\-ap, ma 'itan.} A threatening hammerhead is dangerous, son.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Weyn\-flit\-it 'ang\-tsìk\-ìl sro\-lu' tspang.} Wainfleet was crushed and killed by a hammerhead.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Äo fì\-ut\-ral lu tsmìm 'ang\-tsìk\-ä.} There's a hammerhead track under this tree.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}

% 'ango
\hi
\HT{'ango}{\B{\cp{'a}ngo}} \I{["Pa.No]} \emph{adj.} (\emph{of sound}) soft. 
\B{Zawr ye\-rik\-ä lu 'a\-ngo.} The call of the hexapede is quiet.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{214}~\HL{wok}{\B{wok}})

% 'are
\hi
\HT{'are}{\B{\cp{'a}re}} \I{["Pa.RE]} \emph{n.} poncho, cape, shawl. 
\B{Mo\-'at \mbox{alu} Tsa\-hìk lu O\-ma\-ti\-ka\-ya\-ä le\-'awa ha\-pxì\-tu a \mbox{ioi} sä\-pi fa 'a\-re.} Mo'at, the Tsahik, is the only member of the Omatikaya who wears a poncho.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}

% 'asap / si
\hi 
\HT{'asap}{\B{\cp{'a}sap}} \I{["Pa.sap\textcorner]} \textsc{i}. \emph{n.} sudden shock. 
\B{Fwa tse\-'a pe\-yä tì\-fkey\-tok\-it le\-fkrr lo\-lä\-ngu oer 'a\-sap nì\-ngay.} It was a real shock to me to see him in his current condition. \\
\textsc{ii}. \B{\cp{'a}sap si} \I{["Pa.sap\textcorner{} si]} \emph{vin.} be shocked, be startled. 
\B{Oe 'a\-sap so\-li krr\-a tsa\-fmawn\-it stawm.} I was startled when I heard the news.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 

% 'avun
\hi 
\HT{'avun}{\B{\cp{'a}vun}} \I{["Pa.vun]} \emph{or} \I{[.vUn]} (\textsc{rn}: \B{\cp{'a}vùn} \I{["Pa.vUn]}) \emph{vtr.} save (\emph{time, food, etc.}). 
\B{Nga\-ri txo fì\-kem si\-vi fì\-fya, krr\-ti 'a\-vun.} If you do it like this, you'll save time.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
(\ding{43}~\HL{tarep}{\B{tarep}}; \HL{zong}{\B{zong}})

% 'aw
\hi
\HT{'aw}{\B{'aw}} \I{[Paw]} \emph{num., adj.} one. (\emph{base}: \ding{43}~\B{'aw-}, \emph{rest}: \ding{43}~\HL{-aw}{\B{\mbox{-aw}}})   
\B{'aw\-a tì\-pawm}, one question. \B{Ke tsun tu\-te a\-'aw tsa\-tskxe\-ti a\-ku'\-up kxi\-vel\-tek nì\-'aw\-tu.} One person alone can't lift that heavy rock.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\ 
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'aw\-ve}{\B{'aw\-ve}}, first.   
\on{2}.~\HL{le'aw}{\B{le\-'aw}}, only. 
\on{3}.~\HL{'awlie}{\B{'aw\-lie}}, once (in the past). 
\on{4}.~\HL{'awlo}{\B{'aw\-lo}}, once, one time. 
\on{5}.~\HL{'awpo}{\B{'aw\-po}}, one (individual). 
\on{6}.~\HL{'awsiteng}{\B{'aw\-si\-teng}}, together. 
\on{7}.~\HL{'awstengyem}{\B{'aw\-steng\-yem}}, join (s.t. together).
\on{8}.~\HL{vospxI'aw}{\B{Vo\-spxì\-'aw}}, January.

% 'awkx
\hi
\HT{'awkx}{\B{'awkx}} \I{[Pawk']} \emph{n.} cliff. 
\B{Ik\-ran ya\-wo\-lo ftu txew 'awkx\-ä.} The banshee took to the air from the edge of a cliff.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
\B{Ke tsun oe fì\-'awkx\-it tsyi\-vìl.} I can't scale this cliff.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b} 

% 'awlie
\hi
\HT{'awlie}{\B{'aw\cp{li}e}} \I{[Paw."li.E]} \emph{adv.} (\emph{a.}) once (in the past) (\emph{marks an experience of something}). 
\B{Oel yo\-lom 'aw\-li\-e tey\-lut.} I once ate \emph{teylu}.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\emph{b. in questions}) ever. \B{Sra\-ke ko\-lä nga 'aw\-li\-e ne Nu Yor\-kì?} Have you ever been to New York?~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{'aw}{\B{'aw}}, \HL{lie}{\B{lie}})

% 'awlo
\hi
\HT{'awlo}{\B{\cp{'aw}lo}} \I{["Paw.lo]} \emph{adv.} once, one time (\emph{marks a single instance of something}).  
\B{Oel yo\-lom tey\-lut 'aw\-lo nì\-'aw.} I ate \emph{teylu} only once.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\ding{43}~\HL{'aw}{\B{'aw}}, \HL{alo}{\B{alo}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{kawlo}{\B{kaw\-lo}}, not once.

% 'awm
\hi
\HT{'awm}{\B{'awm}} \I{[Pawm]} \emph{n.} camp.
\B{Sra\-ne, 'awm\-it oel ta\-syok.} Yes, I intend to be at the camp.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/\#comment-2363}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}. \HL{tI'awm}{\B{tì\-'awm}}, camping.

% 'awnìm
\hi
\HT{'awnIm}{\B{\cp{'aw}nìm}} \I{["P$\cdot$aw.nIm]} \emph{vtr., inf.}\on{11} avoid. \B{Ngal 'e\-raw\-nìm oe\-ti srak?} Are you avoiding me?~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{'I'awn}{\B{'ì\-'awn}}, \HL{alIm}{\B{a\-lìm}})

% 'awpo
\hi
\HT{'awpo}{\B{\cp{'aw}po}} \I{["Paw.po]} \emph{pron.} one (individual\slash person). \B{Ha ftxey 'aw\-pot a ay\-nga\-kip.} So choose one among you.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b} (\ding{43}~\HL{'aw}{\B{'aw}}, \HL{po}{\B{po}}, \HL{fko}{\B{fko}})

% 'awsiteng
\hi
\HT{'awsiteng}{\B{'awsi\cp{teng}}} \I{[Paw.si."tEN]} \emph{adv.} together.   
\B{Ze\-ne oe 'aw\-si\-teng tì\-kang\-kem si\-vi fo\-hu.} I must work together with them.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ko\-ren a\-'aw\-ve tì\-ru\-sey\-ä 'aw\-si\-teng}, The First Rule for Living Together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} 
(\emph{a. idiom}) \B{'Aw\-si\-teng lu me\-fo la\-nay'\-ka.} \emph{praising how well two people work together and complement each other} [lit.: they are a slinger (together)].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{To\-la\-ron me\-fol me\-sa\-li\-o\-a\-ngit! Tew\-ti, 'aw\-si\-teng lu me\-fo la\-nay'\-ka.} They hunted two sturmbeest? Wow, they work very well together.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{43}~\HL{'aw}{\B{'aw}}, \HL{teng}{\B{teng}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}. \HL{'awstengyem}{\B{'aw\-steng\-yem}}, join.

% 'awstengyem
\hi
\HT{'awstengyem}{\B{'awsteng\cp{yem}}} \I{[Paw.stEN."j$\cdot$Em]}~\emph{vtr., inf.}\on{33} join (two or more things together).   
\B{Tseng\-a 'aw\-steng\-yä\-pem fì\-me\-kem\-yo lu mek\-tseng.} Where these two walls come together there's a gap.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{'aw\-steng\-yä\-po\-le\-i\-em}, (we) have happily united ourselves.~\LINK{http://forum.learnnavi.org/2012-meetup/vote-on-your-favorite-t-shirt-design/msg547581/\#msg547581}{ln} 
(\ding{43}~\HL{'awsiteng}{\B{'aw\-si\-teng}}, \HL{yem}{\B{yem}})

% 'awve
\hi
\HT{'awve}{\B{\cp{'aw}ve}} \I{["Paw.vE]} \emph{ord.~num.} first. \B{Ko\-ren a\-'aw\-ve tì\-ru\-sey\-ä 'aw\-si\-teng}, The First Rule for Living Together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} \B{'aw\-ve\-a po\-stì zì\-sìt\-ä a\-mip}, first post of the New Year.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\ding{43}~\HL{'aw}{\B{'aw}}, \HL{-ve}{\B{-ve}})

% 'ä'
\hi
\HT{'A'}{\B{'ä'}} \I{[P\ae{}P]} \emph{intj.} whoops! oops! 
\B{'Ä'! \mbox{Oel} tsngal\-it tì\-mung\-zup.} Whoops! I just dropped the cup.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}  

% 'ä'o
\hi 
\HT{'A'o}{\B{\cp{'ä}'o}} \I{["P\ae.Po]} \emph{n.}, \ding{96} pitcher plant, \emph{plantapurpurea cordata}.

% 'älek
\hi 
\HT{'Alek}{\B{\cp{'ä}lek}} \I{["P\ae.lEk\textcorner]} \emph{adj.} determined. 
\B{Tì\-flä\-ri lo\-lu po 'ä\-lek.} She was determined to succeed.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'Alek}{\B{tì\-'ä\-lek}}, determination. 
\on{2}.~\HL{nI'Alek}{\B{nì\-'ä\-lek}}, determinedly, with determination.

% 'änsyem
\hi
\HT{'Ansyem}{\B{'än\cp{syem}}} \I{[P\ae{}n."{}sjEm]} \emph{adj.} complete. 

% 'ärìp
\hi
\HT{'ArIp}{\B{\cp{'ä}rìp}} \I{["P\ae.RIp\textcorner]} \emph{vtr.} move (s.t.). \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{kA'ArIp}{\B{kä\-'ä\-rìp}}, push.  
\on{2}.~\HL{za'ArIp}{\B{za\-'ä\-rìp}}, pull.

% 'e'al
\hi
\HT{'e'al}{\B{\cp{'e}'al}} \I{["PE.Pal]} \emph{adj.} worst. 
\B{Ru\-txe, ay\-nga\-ri fì\-kem fe\-yä 'e\-'a\-la to\-pur ka\-ngay sì\-yi nì\-'aw.} Please, this will only confirm their worst fears about you.~\LINK{http://forum.learnnavi.org/language-updates/two-words-in-use-eal-and-kangay-si/msg564284/\#msg564284}{ln}  
(\ding{214}~\HL{swey}{\B{swey}})

% 'e'in
\hi
\HT{'e'in}{\B{'e\cp{'in}}} \I{[PE."Pin]} \emph{n.} pod, gourd. \\
\textsc{Derivations}: \ding{43} 
\on{1}. \HL{'e'insey}{\B{'e\-'in\-sey}}, drinking gourd.

% 'e'insey
\hi
\HT{'e'insey}{\B{'e\cp{'in}sey}} \I{[PE."Pin.sEj]} \emph{n.} drinking gourd.
(\ding{43}~\HL{'e'in}{\B{'e\-'in}}, \HL{sey}{\B{sey}})

% 'efu
\hi
\HT{'efu}{\B{\cp{'e}fu}} \I{["PE.fu]} (\emph{a. vtr.}) feel, sense, perceive. \B{Lì'\-fya\-ri le\-Na'\-vi \mbox{oel} 'e\-fu ay\-nge\-yä tì\-yawn\-it.} I feel your love for the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\emph{b. vin.}) \emph{with predicate adj. for inner feelings} 
\B{'efu} \ding{43} \HL{nitram}{\B{nit\-ram}}, \HL{keftxo}{\B{ke\-ftxo}}, \HL{som}{\B{som}}, \HL{wew}{\B{wew}}, \HL{ohakx}{\B{o\-hakx}}, \HL{vAng}{\B{väng}}, \HL{ye}{\B{ye}}, \HL{yehakx}{\B{ye\-hakx}}, \HL{yevAng}{\B{ye\-väng}}, \HL{am'ake}{\B{am\-'a\-ke}}, \HL{tunu}{\B{tu\-nu}}, \HL{kxAl}{\B{kxäl}}, \HL{rayl}{\B{rayl}} be\slash feel happy, upset, hot, cold, hungry, thirsty, satisfied, full, quenched, confident\slash sure, romantic, apathetic, sorry (for). 
\B{Me\-nga 'efu väng srak?} Are you two thirsty?~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
\B{Ngey\-ä tì\-kang\-kem\-ì\-ri 'e\-fe\-i\-u oe ye nì\-txan.} I'm very satisfied with your work.~\LINK{https://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} 
\B{Fu\-la ho\-ren sì ay\-lì'\-fya\-vi a\-mip le\-sar sa\-yi me\-ngar oe\-ti nit\-ram 'ey\-ke\-fu nì\-txan.} It makes me very happy that the new rules and words are useful to the two of you.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/comment-page-1/\#comment-1515}{b}\\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'efu}{\B{tì\-'e\-fu}}, feeling.
\on{2}.~\HL{txukxefu}{\B{txu\-kxe\-fu}}, care, be concerned about.

% 'eko
\hi
\HT{'eko}{\B{\cp{'e}ko}} \I{["{}E.ko]} \emph{vtr.} (\emph{a.} \ding{246}) attack. \B{Tsawl\-a pa\-lu\-lu\-kan\-ìl oe\-ti 'ìl\-me\-ko.} The big thanator has just attacked me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 
(\emph{b. of the weather}) \emph{severe precipitation}. \B{Fì\-re\-won tom\-pa\-meyp zar\-mup, slä set 'ì\-me\-ko nì\-txan nang.} It was drizzling this morning, but it's really started coming down now!~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}

% 'ekong
\hi
\HT{'ekong}{\B{\cp{'e}kong}} \I{["PE.koN]} \emph{n.} (\emph{pl.} \B{me\-kong}, \B{pxe\-kong}, \B{ay\-e\-kong\slash e\-kong}) (rhythmic) beat. \B{'e\-kong te'\-lan\-ä}, the beat of the hearts.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{lI'kong}{\B{lì'\-kong}}, syllable. 
\on{2}.~\HL{txe'lankong}{\B{txe'\-lan\-kong}}, heartbeat.

% 'ekxin
\hi
\HT{'ekxin}{\B{'e\cp{kxin}}} \I{[PE."k'in]} \emph{adj.} tight. \B{Fì\-ra\-spu' 'e\-kxin lu nì\-hawng.} These leggings are too tight.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b} (\ding{214}~\HL{um}{\B{um}}) 

% 'ekxinum
\hi
\HT{'ekxinum}{\B{'e\cp{kxi}num}} \I{[PE."k'i.num]} \emph{or} \I{[-nUm]} (\textsc{rn}: \B{'e\-\cp{kxi}\-nùm} \I{[PE."gi.nUm]}) \textsc{i.} \emph{n.} tightness\slash looseness. \\ 
\textsc{ii.} % 'ekxinumpe pekxinum
\B{'e\cp{kxi}numpe} \I{[PE."k'i.num.pE]} \emph{or} \B{pe\-\cp{kxi}\-num} \I{[pE."k'i.num]} \emph{inter.} how tight\slash loose? 
\B{Ngal mo\-lay' pxaw\-pxun\-it Lo\-ak\-ä a krr pe\-kxi\-num?} When you tried on Loak's armband, how tight was it?~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{'ekxin}{\B{'e\-kxin}}, \HL{um}{\B{um}}) 

% 'ele'wll
\hi 
\HT{'ele'wll}{\B{\cp{'e}le'wll}} \I{["PE.lEP.w\s{l}]} \emph{n.}, \ding{96} thorny paw (\emph{cactus-like Pandoran plant}), \emph{sigmoidia barbata}. %~\LINK{}{b}
(\ding{43}~\HL{'ewll}{\B{'e\-wll}}) 

% 'em
\hi
\HT{'em}{\B{'em}} \I{[PEm]} \emph{vtr.} cook. 
\B{Ke new oe mi\-vay' tsngan\-ti a 'o\-lem Ri\-nil.} I don't want to taste the meat that Rini cooked.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Pey\-ral\-ìl zet wur\-a wu\-tsot a\-'aw\-nem pxel sngel.} Peyral treats a cool cooked meal like garbage.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'emyu}{\B{'e\-myu}}, cook. 
\on{2}.~\HL{tI'em}{\B{tì\-'em}}, the art of cooking.

% 'emyu
\hi
\HT{'emyu}{\B{\cp{'em}yu}} \I{["PEm.ju]} \emph{n.}~(\emph{pl.} \B{mem\-yu}, \B{pxem\-yu}, \B{ay\-em\-yu\slash emyu}) cook, someone who cooks. \B{Oe lu 'em\-yu.} I am a cook.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
(\ding{43}~\HL{'em}{\B{'em}})

% 'en / si
\hi
\HT{'en}{\B{'en}} \I{[PEn]} \textsc{i.} \emph{n.} (\emph{pl.} \B{men}, \B{pxen}, \B{ay\-en\slash en}) (informed) guess, hunch, intuition. 
\B{Pe\-lun fì\-tseng\-ne nga zo\-la\-'u fte ti\-va\-ron? -- Lo\-lu 'en.} Why did you come here to hunt? -- It was a guess (hunch).~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Fì\-u\-van\-ì\-ri lu nga\-ru pxen nì\-'aw.} You only get three guesses in this game.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\ 
\textsc{ii.} \B{\cp{'en} si} \I{["PEn si]} \emph{vin.} make an informed guess.   
\B{'En si oe, Saw\-tu\-te\-ol tìl\-mok fì\-tseng\-it.} I guess some Skypeople were just here.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{le'en}{\B{le\-'en}}, speculative, intuitive.  
\on{2}.~\HL{nI'en}{\B{nì\-'en}}, making an informed guess.

% 'eng
\hi 
\HT{'eng}{\B{'eng}} \I{[PEN]} \emph{n., fauna} (\emph{pl.} \B{meng}, \B{pxeng}, \B{ay\-eng\slash eng}) beak of a bird or animal. 

% 'engeng
\hi
\HT{'engeng}{\B{\cp{'eng}eng}} \I{["PEN.EN]} \emph{adj.} level. \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{nI'eng}{\B{nì\-'eng}}, levelly, equally.

% 'eoio
\hi
\HT{'eoio}{\B{\cp{'e}oio}} \I{["PE.o.i.o]} \emph{adj.} ceremonious.   \B{Nì\-lun ay\-\mbox{ioi} a\-\mbox{'eoio} ay\-eyk\-tan\-ä lu lor fra\-to.} Of course the ceremonial wardrobes of the leaders are the most beautiful.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{sA'eoio}{\B{sä\-'e\-oio}}, ritual, rite.   
\on{2}.~\HL{nI'eoio}{\B{nì\-'e\-oio}}, ceremoniously.

% 'etnaw
\hi
\HT{'etnaw}{\B{\cp{'et}naw}} \I{["PEt\textcorner.naw]} \emph{n.} (\emph{pl.} \B{met\-naw}, \B{pxet\-naw}, \B{ay\-et\-naw\slash et\-naw}) shoulder. 
\B{Lo\-lu po\-ru txukx\-a skxir a ftu 'et\-naw a\-ski\-en ne me\-pun a\-ftär.} It had a deep wound from its right shoulder to its two left arms.~\LINK{http://naviteri.org/2020/04/awa-tskxekengtsyip-a-mikyunfpi-niul-one-more-little-listening-exercise/}{b}

% 'evan
\hi
\HT{'evan}{\B{\cp{'e}van}} \I{["PE.van]} \emph{n., coll.} (\emph{pl.} \B{me\-van}, \B{pxe\-van}, \B{ay\-e\-van\slash e\-van}) boy.    \B{'E\-van\-ìl \mbox{alo} a\-'aw\-ve nì\-'aw\-tu na'\-rìng\-it tar\-mok.} The boy was alone in the forest for the first time.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\ding{43}~\HL{'evengan}{\B{'e\-veng\-an}})

% 'eve
\hi
\HT{'eve}{\B{\cp{'e}ve}} \I{["PE.vE]} \emph{n., coll.} (\emph{pl.} \B{me\-ve}, \B{pxe\-ve}, \B{ay\-e\-ve\slash e\-ve}) girl. \B{Se\-vin\-a tsa\-'e\-ve\-ru a\-hì\-'i sa'\-sem \mbox{ioi} so\-li fa mik\-tsang.} The parents put earrings on that pretty little girl.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
(\ding{43}~\HL{'evenge}{\B{'e\-veng\-e}})

% 'eveng
\hi
\HT{'eveng}{\B{\cp{'e}veng}} \I{["PE.vEN]} \emph{n.} (\emph{pl.} \B{me\-veng}, \B{pxe\-veng}, \B{ay\-e\-veng/e\-veng}) child. 
\B{Nì\-ran lu Lo\-ak mi 'e\-veng slä tsun ti\-va\-ron nì\-teng\-fya na fyeyn\-tu.} Loak is still really just a boy but he can hunt the same as an adult.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}  \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'evengan}{\B{'e\-veng\-an}}, boy. 
\on{2}.~\HL{'evenge}{\B{'e\-veng\-e}}, girl. 
\on{3}.~\HL{'evi}{\B{'e\-vi}}, kid. 
\on{4}.~\HL{eywa'eveng}{\B{Ey\-wa\-'eveng}}, (the moon) Pandora. 
\on{5}.~\HL{kalweyaveng}{\B{kal\-wey\-a\-veng}}, `son of a bitch'.

% 'evengan
\hi
\HT{'evengan}{\B{\cp{'e}vengan}} \I{["PE.vEN.an]} \emph{n.} (\emph{pl.} \B{me\-ve\-ngan}, \B{pxe\-ve\-ngan}, \B{ay\-e\-ve\-ngan\slash e\-ve\-ngan}) boy. \B{Txe\-wì lu 'e\-veng\-an a so\-la\-lew zì\-sìt a\-vo\-mun.} Txewì is a ten\hyp{}year\hyp{}old boy.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
(\ding{43}~\HL{'eveng}{\B{'e\-veng}}, \HL{'evan}{\B{'e\-van}}) 

% 'evenge
\hi
\HT{'evenge}{\B{\cp{'e}venge}} \I{["PE.vEN.E]} \emph{n.} (\emph{pl.} \B{me\-veng\-e}, \B{pxe\-veng\-e}, \B{ay\-e\-veng\-e\slash e\-veng\-e}) girl. 
(\ding{43}~\HL{'eveng}{\B{'e\-veng}}, \HL{'eve}{\B{'e\-ve}})  

% 'evi
\hi
\HT{'evi}{\B{\cp{'e}vi}} \I{["PE.vi]} \emph{n.} (\emph{pl.} \B{me\-vi}, \B{pxe\-vi}, \B{ay\-e\-vi\slash e\-vi}) kid (\emph{affectionate form of `child'}). 
\B{Slä ay\-e\-vi nì\-wotx tso\-lun tsli\-vam (pot).} But all children could understand (her).~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
(\ding{43}~\HL{'eveng}{\B{'e\-veng}})

% 'ewan
\hi
\HT{'ewan}{\B{\cp{'e}wan}} \I{["PE.wan]} \emph{adj.} young.   
\B{ay\-lì'\-u\-fa aw\-ngey\-ä 'ey\-lan\-ä a\-'e\-wan}, in the words of our young friend.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ay\-hem\-ì\-ri 'e\-wan\-a tsa\-nan\-tang\-ur a\-hì\-'i tìng na\-ri.} Look at what that little young viperwolf is doing.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
(\ding{214}~\HL{koak}{\B{ko\-ak}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{kewan}{\B{ke\-wan}}, age.
\on{2}.~\HL{tI'ewan}{\B{tì\-'e\-wan}}, youth.

% 'ewll
\hi
\HT{'ewll}{\B{\cp{'e}wll}} \I{["PE.w\s{l}]} \emph{n.} \ding{96} (\emph{pl.} \B{me\-wll}, \B{pxe\-wll}, \B{ay\-e\-wll\slash e\-wll}) plant. 
\B{\mbox{Oel} vewng fu\-ta ay\-e\-veng ni\-vume te\-ri ay\-ewll na'\-rìng\-ä.} I see to it that the children learn about the forest plants.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
\B{Lo\-ak\-ìl pä\-nu\-to\-lìng fu\-ta kar oe\-ru fya\-'ot a 'ìp fko nem\-fa \mbox{ewll}.} Loak promised he’d teach me how to vanish into the bushes.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{prrwll}{\B{prr\-wll}}, moss. 
\on{2}.~\HL{paywll}{\B{pay\-wll}}, water plant. 
\on{3}.~\HL{pxiwll}{\B{pxi\-wll}}, hermit bud plant. 
\on{4}.~\HL{vAfewll}{\B{vä\-fe\-wll}}, Centipede. 
\on{5}.~\HL{fwAkIwll}{\B{fwä\-kì\-wll}}, mantis orchid. 
\on{6}.~\HL{'ele'wll}{\B{'e\-le'\-wll}}, thorny paw. 
\on{7}.~\HL{pa'liwll}{\B{pa'\-li\-wll}}, direhorse pitcher plant. 
\on{8}.~\HL{tompawll}{\B{tom\-pa\-wll}}, geode.
\on{9}.~\HL{hIrumwll}{\B{hì\-rum\-wll}}, puffer plant. 
\on{10}.~\HL{fkxakewll}{\B{fkxa\-ke\-wll}}, itch plant. 
\on{11}.~\HL{kllpxiwll}{\B{kll\-pxi\-wll}}, lionberry. 
\on{12}.~\HL{fngapsutxwll}{\B{fngap\-sutx\-wll}}, metalfollowing plant.
\on{13}.~\HL{yawrwll}{\B{yawr\-wll}}, feather blade.

% 'ewrang
\hi
\HT{'ewrang}{\B{\cp{'ew}rang}} \I{["PEw.RaN]} \emph{n., text.} (\emph{pl.} \B{mew\-rang}, \B{pxew\-rang}, \B{ay\-ew\-rang\slash ew\-rang}) loom. \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{sa'ewrang}{\B{sa\-'ew\-rang}}, great loom, mother loom.

% 'eylan
\hi
\HT{'eylan}{\B{\cp{'ey}lan}} \I{["PEj.lan]} \emph{n.} (\emph{pl.} \B{mey\-lan}, 
\B{pxey\-lan}, \B{ay\-ey\-lan\slash ey\-lan}) friend.  
\B{ay\-ey\-lan\-ur oe\-yä sì ey\-lan\-ur lì'\-fya\-yä le\-Na'\-vi nìwotx}, to my friends and all friends of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Me\-nga lu oe\-yä 'ey\-lan.} You two are my friends.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b} 
% \B{aw\-nge\-yä 'ey\-lan\-hu mll\-te (oe)}, (I) agree with our friend.~\LINK{http://masempul.org/2010/07/\%E2\%80\%99ite-polaheiem/}{sempul} 
\B{Sra\-ke smon ngar oe\-yä mey\-lan alu En\-tu sì Ka\-mun?} Do you know my friends En\-tu and Ka\-mun?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} 
\B{ay\-nga\-ru oe\-yä pxey\-lan\-ti}, let me introduce to you my three friends.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} 
(\ding{214}~\HL{kxutu}{\B{kxu\-tu}}, \HL{wAtu}{\B{wä\-tu}}) \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'eylan}{\B{tì\-'ey\-lan}}, friendship.
\on{2}.~\HL{'eylanay}{\B{'ey\-la\-nay}}, acquaintance.
\on{3}.~\HL{'eylyong}{\B{'eyl\-yong}}, pet.
\on{4}.~\HL{let'eylan}{\B{let\-'ey\-lan}}, friendly (ofp.).
\on{5}.~\HL{txeylan}{\B{txey\-lan}}, best friend.

% 'eylanay
\hi
\HT{'eylanay}{\B{'eyla\cp{nay}}} \I{[PEj.la."{}naj]} \emph{n.} (\emph{pl.} \B{mey\-la\-nay}, \B{pxey\-la\-nay}, \B{ay\-ey\-la\-nay\slash ey\-la\-nay}) acquaintance (with the potential of becoming a friend).%~\LINK{}{b}  
(\ding{43}~\HL{'eylan}{\B{'ey\-lan}})

% 'eylyong
\hi 
\HT{'eylyong}{\B{\cp{'eyl}yong}} \I{["PEjl.joN]} \emph{n.} pet. 
\B{Tì\-kan fì\-pay\-oang\-ä ke lu fwa tsun fko pot yi\-vom; lu oe\-yä 'eyl\-yong.} This fish is not meant to be eaten; it's my pet.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b}
(\ding{43}~\HL{'eylan}{\B{'ey\-lan}}, \HL{ioang}{\B{ioang}})

% 'eyng
\hi
\HT{'eyng}{\B{'eyng}} \I{[PEjN]} \emph{vin.} answer, respond. 
\B{'Aw\-a tì\-pawm\-ì\-ri 'i\-veyng oe set;  ay\-la\-ri zu\-saw\-krr 'ay\-eyng.} Let me answer one of the questions now; the others (I'll) answer in the future.~\LINK{http://forum.learnnavi.org/language-updates/email-from-frommer-re-numbers-\%28or-the-full-number-chart!\%29/}{ln}
\B{'Eyng oe\-ru set!} Answer me now!~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b}
(\ding{214}~\HL{pawm}{\B{pawm}}, \HL{vin}{\B{vin}}) \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'eyng}{\B{tì\-'eyng}}, answer, respond.   
\on{2}.~\HL{nI'eyng}{\B{nì\-'eyng}}, in response.

% 'eyt
\hi
\HT{'eyt}{\B{'eyt}} \I{[PEjt\textcorner]} \emph{num.}, \textsc{lw} eight, (the symbol) \on{8} (\emph{only as Human decimal figure}). 
(\ding{43}~\HL{nayn}{\B{nayn}})

% 'i'a
\hi
\HT{'i'a}{\B{\cp{'i'}a}} \I{["PiP.a]} \emph{vin.} end, conclude. 
\B{ye'\-rìn 'ì\-yi\-'a sä\-nu\-me a tsa\-ri kll\-fro' oe}, soon the teaching I am responsible for will end.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Ney\-ti\-ril nan\-tang\-it tso\-lukx fte pe\-yä tì\-sraw\-ti 'ey\-ki\-vi\-'a.} Neytiri stabbed the viperwolf to end its pain.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
(\ding{214}~\HL{sngA'i}{\B{sngä\-'i}})  \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'i'a}{\B{tì\-'i'\-a}}, ending, conclusion.  
\on{2}.~\HL{nI'i'a}{\B{nì\-'i'\-a}}, finally.

% 'ia
\hi
\HT{'ia}{\B{\cp{'i}a}} \I{["Pi.a]} \emph{vin.} lose oneself (\emph{spiritual sense}). \B{Eo Ey\-wa oe \mbox{'ia}.} I lose myself before Eywa.~\LINK{http://forum.learnnavi.org/language-updates/sieyng-a-ftu-naring-3-proverbs-what-we-asked-for-and-some-vocabulary/msg327513/\#msg327513}{a\slash ln}

% 'ipu
\hi
\HT{'ipu}{\B{\cp{'i}pu}} \I{["Pi.pu]} \emph{adj.} humorous, funny, amusing.   \B{Kaw\-krr ke lu peyä ay\-vur 'ipu kaw\-'it.} His stories are never a bit amusing.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'ipu}{\B{tì\-'i\-pu}}, humor. 
\on{2}.~\HL{sA'ipu}{\B{sä\-'i\-pu}}, s.t.~humorous.

% 'it
\hi
\HT{'it}{\B{'it}} \I{[Pit\textcorner]} \emph{n.} bit, small amount.\\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{nI'it}{\B{nì\-'it}}, a bit. 
\on{2}.~\HL{kaw'it}{\B{kaw\-'it}}, (not) at all.

% 'itan
\hi
\HT{'itan}{\B{\cp{'i}tan}} \I{["Pi.tan]} \emph{n.} son. \B{Fra\-lo a ta\-ron, oe\-yä 'itan txo\-pu si nì\-nä\-nän.} Each time he hunts, my son becomes less and less afraid.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} \B{Peyä 'itan\-ìri lu ho\-na nì\-txan a fì\-'u law lu fra\-por.} It's clear to everyone that his son is very cute.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{maitan}{\B{mai\-tan}}, my son (address).

% 'ite
\hi
\HT{'ite}{\B{\cp{'i}te}} \I{["Pi.tE]} \emph{n.} daughter. \B{Za\-'u fì\-tseng, ma 'ite\-tsyìp.} Come here, little daughter.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} \B{Nga\-ri leiu oe\-ru nrra nì\-txan, ma 'ite.} I'm very proud of you, daughter.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{maite}{\B{mai\-te}}, my daughter (address).

% 'ì'awn
\hi
\HT{'I'awn}{\B{'ì\cp{'awn}}} \I{[PI."Pawn]} \emph{vin.} remain, stay. 
\B{Nga\-ri hu Ey\-wa sa\-lew ti\-re\-a, tokx 'ì'awn slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body remains to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{a\slash wiki} 
\B{'Ì'awn alìm!} Stay back!~\LINK{http://forum.learnnavi.org/language-updates/near-distant-and-irregular-adverbs/}{a\slash ln}
\B{Ke\-tsran tu\-te ni\-vew hi\-vum, po\-ru pll\-txe san ru\-txe 'i\-vì\-'awn.} No matter who wants to leave, tell them to please stay.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}. \HL{'awnIm}{\B{'aw\-nìm}}, avoid.

% 'ìheyu
\hi
\HT{'Iheyu}{\B{'ì\cp{he}yu}} \I{[PI."hE.ju]} \emph{n.} spiral.   
\B{'ì\-he\-yu sì\-rey\-ä}, \emph{the spiral of the lives}.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki} 
\B{Ay\-wayl yìm ki\-fkey\-ä | 'Ì\-he\-yut a\-vo\-mrr}, \emph{The songs bind} | \emph{the thirteen spirals of the solid world.}~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{anIheyu}{\B{anì\-he\-yu}}, fibonacci.

% 'ìn
\hi
\HT{'In}{\B{'ìn}} \I{[PIn]} \emph{vin.} be busy, be occupied.   
\B{'Ìn nga fya\-pe nì\-fkrr?} What's been keeping you busy lately?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tIn}{\B{tìn}}, activity. 
\on{2}.~\HL{tIk'In}{\B{tìk\-'ìn}}, free time. 
\on{3}.~\HL{sulIn}{\B{su\-lìn}}, be busy (positive). 
\on{4}.~\HL{vrrIn}{\B{vrr\-ìn}}, be busy (negative). 
\on{5}.~\HL{kan'In}{\B{kan\-'ìn}}, focus on, specialize in.

% 'ìnglìsì
\hi
\HT{'InglIsI}{\B{\cp{'Ìng}lìsì}} \I{["PIN.lI.sI]} \emph{n.} English language. (\emph{mostly used in its derivations} \ding{43}~\B{le\-'Ìng\-lì\-sì} \emph{and} \ding{43}~\B{nì\-'Ìng\-lì\-sì}) 
\B{Krro lu 'Ìng\-lì\-sì txur nì\-hawng mì re\-'o.} Sometimes English is too strong in (my) head.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b}
\B{Fu\-ri\-a fì\-lì'\-fya\-vit fkol ta 'Ìng\-lì\-sì mo\-lu\-nge, sìl\-pey oe tsnì ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ke stì\-ye\-vi.} That this expression is brought from English, will hopefully not anger the members of the language group.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{le'InglIsI}{\B{le\-'Ìng\-lì\-sì}}, English, regarding English.

% 'ìp
\hi
\HT{'Ip}{\B{'ìp}} \I{[PIp\textcorner]} \emph{vin.} disappear, vanish, recede from view. \B{Lo\-ak\-ìl pä\-nu\-to\-lìng fu\-ta kar oe\-ru fya\-'ot a 'ìp fko nem\-fa \mbox{ewll}.} Loak promised he'd teach me how to vanish into the bushes.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}  
\B{Oe\-yä tska\-lep 'o\-lìp ning\-yen.} My crossbow has mysteriously disappeared.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\emph{a. proverb}) \B{Nì\-win 'ìp tì\-'e\-wan, nì\-win pä\-hem tì\-ko\-ak.} Youth vanishes quickly, old age arrives quickly.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} 
(\ding{214}~\HL{srer}{\B{srer}})

% 'llngo
\hi
\HT{'llngo}{\B{\cp{'ll}ngo}} \I{["P\s{l}.No]} \emph{n.} (\emph{pl.} \B{me\-'ll\-ngo}, \B{pxe\-'ll\-ngo}, \B{ay\-'ll\-ngo}) hip. 
\B{Fu\-r\-i\-a 'll\-ngo nge\-yä zo\-slu, 'e\-fu oe nit\-ram nì\-'aw.} I'm just happy that your hip is getting better.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/comment-page-1/\#comment-2484}{b}

% 'o'
\hi
\HT{'o'}{\B{'o'}} \I{[PoP]} \emph{adj.} (bringing) fun, exciting.   
\B{Ay\-u\-van le\-tokx 'o' lu nì\-txan.} Sports are great fun.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
\B{'O'a Ftxozä Hälowinä, ma frapo!} Happy Halloween, everyone!~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'o'}{\B{tì\-'o'}}, fun, excitement. 
\on{2}.~\HL{nI'o'}{\B{nì\-'o'}}, enjoyably, fun\hyp{}ly.

% 'ok
\hi
\HT{'ok}{\B{'ok}} \I{[Pok\textcorner]} \emph{n.} memory, remembrance. \B{Teng\-krr pa\-lu\-lu\-kan moe\-ne kxll sar\-mi, pol\-txe Ney\-ti\-ril ay\-lì\-'ut a fra\-krr 'ok seyä la\-yu oer.} As the thanator was charging towards the two of us, Neytiri said something I will always remember.~\LINK{http://thereadinginpublicproject.blogspot.de/2010/05/paul-frommer.html}{pr} 
\B{'Ok oe\-yä lu ay\-ngar srak?} Do you all remember me?~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
(\ding{43}~\HL{zerok}{\B{ze\-rok}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'okrol}{\B{'ok\-rol}}, (ancient) history. 
\on{2}.~\HL{'okvur}{\B{'ok\-vur}}, (recent) history. 
\on{3}.~\HL{'oktrr}{\B{'ok\-trr}}, day of commemoration.

% 'okrol
\hi
\HT{'okrol}{\B{'ok\cp{rol}}} \I{[Pok\textcorner."Rol]} \emph{n.} (ancient) history. 
(\ding{43}~\HL{'ok}{\B{'ok}}, \HL{rol}{\B{rol}}; \HL{'okvur}{\B{'ok\-vur}})

% 'oktrr
\hi 
\HT{'oktrr}{\B{\cp{'ok}trr}} \I{["Pok\textcorner.t\s{r}]} \emph{n.} day of commemoration. 
(\ding{43}~\HL{'ok}{\B{'ok}}, \HL{trr}{\B{trr}})

% 'okvur
\hi
\HT{'okvur}{\B{'ok\cp{vur}}} \I{[Pok\textcorner."vuR]} \emph{or} \I{[."vUR]} \emph{n.} (non\hyp{}ancient) history. \B{oe\-yä so\-ai\-ä 'ok\-vur}, my family's history. \B{'ok\-vur Saw\-tu\-te\-yä mì Ey\-wa\-'e\-veng}, the history of the Sky People on Pandora.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\ding{214}~\HL{'ok}{\B{'ok}}, \HL{vur}{\B{vur}}; \HL{'okrol}{\B{'ok\-rol}})

% 'om
\hi
\HT{'om}{\B{'om}} \I{[Pom]} \emph{adj.} violet to purple to magenta.
\B{'om na mik\-yun}, the purplish color on the inside of a Na'vi ear.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'ompin}{\B{'om\-pin}}, purple color.

% 'ompin
\hi
\HT{'ompin}{\B{\cp{'om}pin}} \I{["Pom.pin]} \emph{n.} the color \ding{43}~\HL{'om}{\B{'om}}. 
(\ding{43}~\HL{'opin}{\B{'o\-pin}})

% 'on / 'on si
\hi
\HT{'on}{\B{'on}} \I{[Pon]} \textsc{i}. \emph{n.} shape, form.
\B{Tsun fko ay\-on\-ti fì\-wopx\-ä ni\-vìn fte ya\-fkeyk\-it sre\-si\-ve\-'a.} The shapes of clouds can be used to predict the weather.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} 
\B{'On\-ìri tsko\-ä lu tì\-kin sä\-la\-tem\-ä a\-hì\-'i.} The form of the bow requires a small change.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b} \\
\textsc{ii}. \B{\cp{'on} si} \I{["Pon si]} \emph{vin.} shape; give shape (to s.t.) (\emph{physical shaping or metaphorically}).
\B{Oe 'on si tskxe\-ru fte na ik\-ran li\-vam.} I shape a rock to look like an ik\-ran.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
\B{Olo'\-ìri po\-an zu\-saw\-krr\-ur 'on so\-li.} He shaped the tribe's future.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{ko'on}{\B{ko\-'on}}, ring, oval. 
\on{2}.~\HL{'ongop}{\B{'o\-ngop}}, design (\emph{for the larger aspects of design}).

% 'ong
\hi
\HT{'ong}{\B{'ong}} \I{[PoN]} \emph{vin.} unfold, blossom. (\emph{a.}) \B{Na se\-ze | A 'ong ne tsaw\-ke.} Like the blue\hyp{}flower | Which unfolds to the sun.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} 
\B{Lì'\-fya\-ri le\-Na'\-vi 'Rr\-ta\-mì, vay set 'al\-mong a fra\-'u ze\-ra\-'u ta ngrr\-po\-ngu.} Everything that has gone on with (blossomed regarding) Na'vi until now on Earth has come from a grassroots movement.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} 
(\emph{b. idiom})  \B{'Ivong Na'\-vi!} Let Na'vi bloom!~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{'ivong nìk\-'ong}, slow is fine. \B{Ke ze\-ne win sä\-pi\-vi. 'Ivong nìk\-'ong.} Take your time; don't rush. Slow is fine.~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'ong}{\B{tì\-'ong}}, unfolding, blooming. 
\on{2}.~\HL{'onglawn}{\B{'ong\-lawn}}, \emph{exhiliration of first bonding}.

% 'onglawn
\hi 
\HT{'onglawn}{\B{\cp{'ong}lawn}} \I{["PoN.lawn]} \emph{n.} \emph{exhiliration of first bonding} (\emph{used with} \ding{43}~\HL{'efu}{\B{'efu}}). 
\B{Kaw\-krr ke tswa\-ya' oel krr\-it a 'e\-fu 'ong\-lawn\-it.} I'll never forget the time I experienced \emph{'ong\-lawn}.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
\B{Lu 'ong\-lawn tì\-'e\-fu a\-ko\-sman fra\-to mì hi\-fkey.} \emph{'Ong\-lawn} is the most wonderful feeling in the world.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{'ong}{\B{'ong}}, \HL{lawnol}{\B{law\-nol}})

% 'ongokx
\hi
\HT{'ongokx}{\B{\cp{'o}ngokx}} \I{["Po.Nok']} \emph{vin.} be born.   
\B{oe 'o\-lo\-ngokx mì sray a txam\-pay\-ì\-ri sim}, I was born in a town near the ocean~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\ding{43}~\HL{'ong}{\B{'ong}}, \HL{nokx}{\B{nokx}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI'ongokx}{\B{tì\-'o\-ngokx}}, birth.

% 'ongop
\hi 
\HT{'ongop}{\B{\cp{'o}ngop}} \I{["Po.N$\cdot$op\textcorner]} \emph{vtr., inf.}\on{22} design (\emph{for the larger aspects of design; shape or structure}).
\B{Fì\-tsko\-ti a\-fyo\-lup 'o\-ngo\-lop oe\-yä sem\-pul\-ìl.} This exquisite bow was designed by my father.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\ding{43}~\HL{rengop}{\B{re\-ngop}}; \HL{'on}{\B{'on}}, \HL{ngop}{\B{ngop}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'ongopyu}{\B{'o\-ngop\-yu}}, designer.

% 'ongopyu
\hi 
\HT{'ongopyu}{\B{\cp{'o}ngopyu}} \I{["Po.Nop\textcorner.ju]} \emph{n.} designer 
(\ding{43}~\HL{'ongop}{\B{'o\-ngop}})

% 'opin
\hi
\HT{'opin}{\B{\cp{'o}pin}} \I{["Po.pin]} \emph{n.} color. \B{Tsa\-'o\-pin hek nì\-'it, slä su\-nu oer, ha ha'.} That color is bit odd, but I like it, so it’s a good fit for me.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'ompin}{\B{'om\-pin}}, the color violet\slash purple\slash megenta.
\on{2}.~\HL{eampin}{\B{eam\-pin}}, the color blue\slash green.
\on{3}.~\HL{kllvawmpin}{\B{kll\-vawm\-pin}}, the color brown.
\on{4}.~\HL{layompin}{\B{la\-yom\-pin}}, the color black.
\on{5}.~\HL{neympin}{\B{neym\-pin}}, the color value `light'.
\on{6}.~\HL{ngulpin}{\B{ngul\-pin}}, the color grey.
\on{7}.~\HL{rimpin}{\B{rim\-pin}}, the color yellow.
\on{8}.~\HL{teyrpin}{\B{teyr\-pin}}, the color white.
\on{9}.~\HL{tumpin}{\B{tum\-pin}}, the color red\slash orange.
\on{10}.~\HL{vawmpin}{\B{vawm\-pin}}, the color value `dark'.
\on{11}.~\HL{pxayopin}{\B{pxay\-opin}}, colorful. 
\on{12}.~\HL{'opinvultsyIp}{\B{'opin\-vul\-tsyìp}}, crayon. 
\on{13}.~\HL{'opinsung}{\B{'opin\-sung}}, color, color in.

% 'opinsung
\hi
\HT{'opinsung}{\B{\cp{'o}pinsung}} \I{["Po.pin.s$\cdot$uN]} \emph{or} \I{[.s$\cdot$UN]} (\textsc{rn}: \B{\cp{'o}pinsùng} \I{["Po.pin.s$\cdot$UN]}) \emph{inf.}\on{33} (\emph{a. vtr.}) color, color in; paint, apply paints\slash dyes.
\B{'Eveng\-ìl fì\-rel\-it 'opin\-so\-lung fa pin\-vul a\-rey\-pay\-tun.} The child colored (in) this picture with a red crayon.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/\#comment-28355}{b}
(\emph{b. vin.}) \B{'Eveng 'opin\-sar\-mung mì fuk.} The child was coloring in a book.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/\#comment-28355}{b}
(\ding{43}~\HL{'opin}{\B{'opin}}, \HL{sung}{\B{sung}})

% 'opinvultsyìp
\hi 
\HT{'opinvultsyIp}{\B{\cp{'o}pinvultsyìp}} \I{["Po.pin.vul.\t{ts}jIp\textcorner]} \emph{or} \I{[.vUl.]} \emph{n.} crayon (\emph{often shortened to} \ding{43}~\HL{pinvul}{\B{pinvul}}). 
(\ding{43}~\HL{'opin}{\B{'opin}}, \HL{vultsyIp}{\B{vul\-tsyìp}})

% 'ora
\hi
\HT{'ora}{\B{\cp{'o}ra}} \I{["Po.Ra]} \emph{n.} lake. \B{Fo kel\-ku si few 'o\-ra.} They live on the opposite side of the lake.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'oratsyIp}{\B{'o\-ra\-tsyìp}}, pond, pool.

% 'oratsyìp
\hi
\HT{'oratsyIp}{\B{\cp{'o}ratsyìp}} \I{["Po.Ra.\t{ts}jIp\textcorner]} \emph{n.} pond, pool. \B{Ne 'o\-ra\-tsyìp po\-lä\-hem, yak si nì\-ftär.} When you arrive at the pond, turn to the left.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{'ora}{\B{'o\-ra}})

% 'otxang
\hi
\HT{'otxang}{\B{'o\cp{txang}}} \I{[Po."t'aN]} \emph{n.} musical instrument (\emph{generic term}).   \B{Pam fì\-'o\-txang\-ä su\-nu oer nì\-ngay.} I really like the sound of this instrument.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 

% 'rrko
\hi
\HT{'rrko}{\B{\cp{'rr}ko}} \I{["P\s{r}.ko]} (\textsc{imperf} \B{'rr\-ko}) \emph{vin.} roll.
\B{Rum 'o\-lrr\-ko oe\-ne kll\-te\-sìn.} The ball rolled towards me on the ground.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{'Eveng\-ìl skxe\-vit 'ey\-krr\-ko sko uvan.} Children roll pebbles as a game.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Tseyk 'ä\-pey\-ka\-mrr\-ko äo ut\-ral a zo\-lup fte hi\-vi\-fwo ftu ay\-sre' pa\-lu\-lu\-kan\-ä.} Jake rolled under the fallen tree to escape from the thanator's teeth.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}

% 'rrpxom
\hi
\HT{'rrpxom}{\B{\cp{'rr}pxom}} \I{["P\s{r}.p'om]} \emph{n.} thunder.%~\LINK{}{b}

% 'rrta
\hi
\HT{'Rrta}{\B{\cp{'Rr}ta}} \I{["P\s{r}.ta]} \emph{n.}, {\textsc{lw}} (\emph{planet}) Earth. \B{mì 'Rrta}, on Earth. 
\B{(Pol) Inan puk\-ot a te\-ri ay\-sam a 'Rr\-ta\-mì.} He's reading a book about the wars on Earth.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{ln} 
\B{Trr 'Rrtayä}, Earth Day.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Tsamsiyu a mì Saw 'Rrtayä}, \emph{Tsamsiyu} in the skies of Earth.~\LINK{http://naviteri.org/2013/03/tsamsiyu-a-mi-saw-rrtaya-warrior-in-our-sky/}{b} 

% 'u 
\hi
\HT{'u}{\B{'u}} \I{[Pu]} (\textsc{rn}: \B{'u} \I{[Pu]}) \textsc{i}.  \emph{n.} (\emph{no short plural, always:} \B{ayu}) thing (\emph{object, fact, abstraction}). 
\B{Kxawm set 'u a\-ngä\-zìk fra\-to lu la\-'a a\-yll.} Maybe the most difficult thing now is social distance.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} 
\B{Krr\-ka tì\-rey a\-yol, pol mo\-lu'\-ni pxay\-a ay\-ut a tsran\-ten.} During her short life, she accomplished many important things.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}  \\
\textsc{ii}. \HL{'upe}{\B{\cp{'u}pe}} \I{["Pu.pE]} \emph{inter.} what (thing) (\emph{alternative form:} \B{pe\cp{u}}). \B{Tsa\-lì\-'u\-ri alu X, ral lu 'u\-pe?} What does X mean?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
\B{Nga new yi\-vom 'u\-pet fì\-txon?} What do you want to eat tonight?~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b} 
(\ding{43}~\HL{zum}{\B{zum}}). \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{'uo}{\B{'uo}}, something. 
\on{2}.~\HL{fI'u}{\B{fì\-'u}}, this (thing). 
\on{3}.~\HL{tsa'u}{\B{tsa\-'u}}, that (thing). 
\on{4}.~\HL{fra'u}{\B{fra\-'u}}, everything. 
\on{5}.~\HL{ke'u}{\B{ke\-'u}}, nothing.
% (\emph{b}.) \B{tsa\cp{'u}} \I{[\t{ts}a."Pu]} \emph{pn., n.} (\emph{with case endings or enclitic adp.s often shortened to:} \B{tsa-} \emph{or} \B{tsaw-}, \emph{i.e.} \B{tsal, tsat(i), tsar(u), tseyä, tsari}) that (thing). \B{Ngal new a tsa\-'ut rä\-'ä wi\-vo, ma 'e\-vi.} Don't reach for what you want, child.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \B{Fko zene sivung tsat tsatseng.} One can add that there.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/comment-page-1/\#comment-2476}{b} \B{Fra\-'uä ran ngä\-pop fa fra\-syon tseyä.} The \emph{ran} of each thing arises from the totality of its attributes.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \B{Tsa\-'u\-ri po pll\-ngay.} He admits it.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}

% 'ul
\hi
\HT{'ul}{\B{'ul}} \I{[Pul]} \emph{or} \I{[PUl]} (\emph{a}.) \emph{vin.} increase. (\emph{b. to form correlative comparisons}) (i.) \B{'ul … 'ul}  the more … the more. \B{'Ul tskxe\-keng si, 'ul fnan.} The more you practice, the better you'll get.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
\B{'Ul tu\-te, 'ul tì\-ngä\-zìk.} The more people, the more problems.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
(ii.) \B{'ul … nän} the more … the less. (\ding{214}~\HL{nAn}{\B{nän}})\\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tI.'ul}{\B{tì\-'ul}}, increase. 
\on{2}.~\HL{tsan'ul}{\B{tsan\-'ul}}, improve, get better. 
\on{3}.~\HL{fe'ul}{\B{fe\-'ul}}, worsen. 

% 'umtsa
\hi
\HT{'umtsa}{\B{\cp{'um}tsa}} \I{["{}um.\t{ts}a]} \emph{or} \I{["{}Um.]} (\textsc{rn}: \B{\cp{'ùm}tsa} \I{["PUm.\t{ts}a]}) \emph{n.} medicine (\HL{wA}{\B{wä}}, against).
\B{Ra\-lul ngo\-lop 'um\-tsat fa ay\-syu\-lang fwä\-kì\-wll\-ä.} Ralu made medicine from (\emph{or}: using the) flowers of the Mantis orchid.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
\B{Fì\-sä\-spxin\-ìri nge\-yä ke lä\-ngu kea 'um\-tsa.} Unfortunately there is no medicine for this disease of yours.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
\B{Wä fì\-sä\-spxin a 'um\-tsa le\-iu set tsu\-ka\-nom.} Medicine against this disease is happily now available.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 

% 'uo
\hi
\HT{'uo}{\B{\cp{'u}o}} \I{["Pu.o]} \emph{pn., n.} something. \B{'Uo a fpi rey\-'eng Ey\-wa\-'e\-veng\-mì, 'Rr\-ta\-mì tsran\-ten nì\-txan aw\-nga\-ru nì\-wotx.} Something that matters a lot to all of us for the sake of The Balance of Life on both Pandora and Earth.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{'Uol ik\-ran\-it txo\-pu sley\-ko\-la\-tsu, ta\-lun\-a po tsìk ya\-wo.} Something must have frightened the banshee, because it suddenly took to the air.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Fì\-na\-er\-ìri ngal ew\-ku 'u\-ot a\-stxong srak?} Do you taste something strange in this drink?~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\ding{43}~\HL{'u}{\B{'u}})

% 'upe
\hi
\B{\cp{'u}pe}~:~\HL{'u}{\B{'u}}. 

% 'upxare
\hi
\HT{'upxare}{\B{'u\cp{pxa}re}} \I{[Pu."p'a.RE]} (\textsc{rn}: \B{'ù\cp{pxa}re} \I{[PU.ba.RE]}) \emph{n.} message.   
\B{Lu pì\-lok\-ur pxe\-sì\-kan sì pxe\-fne\-'upxa\-re.} The blog has three aims and three kinds of messages.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Fpo\-le' ay\-ngal oer a 'upxa\-ret sto\-lawm oel.} I have heard the message you have sent me.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Spä\-ngaw oel fu\-ta fì\-'upxa\-re\-yä tì\-ral\-peng ke lu eyawr.} Unfortunately, I don't believe the translation of this message is correct.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}

% 'ur
\hi
\HT{'ur}{\B{'ur}}\textsuperscript{\on{1}} \I{[PuR]} \emph{or} \I{[PUR]} (\textsc{rn}: \B{'ùr} \I{[PUR]}) \emph{n.} (\emph{a. sensation}) sight, look, appearance. 
(\emph{b. to form the middle voice}) \B{Nik\-re\-ri Ri\-ni\-yä 'ur fkan lor.} Rini's hair looks beautiful.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}

\hi
\B{'ur}\textsuperscript{\on{2}}~:~\HL{'u}{\B{'u}}.

\end{multicols}