\pdfbookmark[0]{A}{A}
\begin{center}
\section*{A \I{[a]}}
\end{center}

\begin{multicols}{2}

% a
\hi
\HT{a}{\B{a}} \I{[a]} \emph{part.} \emph{used to subordinate an attributive sentence to a noun (relative clause)}. 
\B{Fpo\-le' ay\-ngal oer fì\-txan nì\-ftxa\-vang a 'u\-pxa\-ret sto\-lawm oel.} I have heard the message that you have sent me so passionately.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{ke rey a tu\-te}, person who doesn't live.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln} 
\B{fì\-txon yom sat a lu syä\-'ä nì\-'aw}, tonight (we) eat only those [vegetables] that are bitter.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
\B{Tsap\-'a\-lu\-te, mì u\-pxa\-re a fpo\-le' oel ay\-nga\-ru trr\-am lu kxey\-ey}, Apologies, in the message that I sent you yesterday was a mistake.~\LINK{http://forum.learnnavi.org/language-updates/another-email-from-frommer/msg68133/\#msg68133}{ln} 
(with \emph{adp.s}) \B{Ma oeyä ey\-lan a\-yaw\-ne a lì'\-fya\-o\-lo'\-mì}, My beloved friends in the language clan.~\LINK{http://forum.learnnavi.org/news-announcements/happy-birthday-to-karyu-pawl/msg556899/\#msg556899}{ln} 
\B{Txan\-tsan\-a ul\-txa a mì Siätll so\-lu\-ne\-i\-u oer nì\-txan.} I liked the fantastic meeting in Seattle very much.~\LINK{http://naviteri.org/2011/07/attributive-\%E2\%80\%9Ca\%E2\%80\%9D-and-truncated-style/}{b} 
(with \emph{vin}.) \B{Kun\-sìp\-ì\-ri txan\-a tì\-meyp lu tsyal a mìn.} The gunship's main weakness is the rotor system. [lit.: the wing that turns.]~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{a\slash ln} 
\B{Lu fmawn a el\-tur tì\-txen si nì\-txan, ke\-fyak?} These are very interesting news, aren't they? [lit.: news that are interesting]~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
(\emph{to form conj.s}): \ding{43}~\B{\HL{alu}{\B{a\-lu}}, \HL{krra}{krr\-a}, \HL{takrra}{ta\-krr\-a}, \HL{mawkrra}{maw\-krr\-a}, \HL{srekrra}{sre\-krr\-a}, \HL{fwa}{fwa}, \HL{fula}{fu\-la}, \HL{futa}{fu\-ta}, \HL{fura}{\B{fu\-ra}}, \HL{furia}{fu\-ri\-a}, \HL{fayluta}{fay\-lut\-a}, \HL{fmawnta}{fmawn\-ta}, \HL{kuma}{kum\-a}, \HL{teynga}{tey\-nga}, \HL{teyngla}{teyng\-la}, \HL{teyngta}{teyng\-ta}, \HL{teyngra}{\B{teyng\-ra}}, \HL{teyngA}{\B{tey\-ngä}}, \HL{teyngria}{\B{teyng\-ri\-a}}, \HL{tsenga}{tse\-nga}, \HL{tsonta}{tson\-ta}} 

% -a-
\hi
\HT{a-}{\B{a-}} \emph{or} \B{-a} \I{[a]} \emph{pref., suf.} (\emph{a. used to attribute an adj. to a noun} (\emph{attributive adj.})) 
\B{tsa\-krr pa\-ye\-'un swey\-a fya\-'ot}, then (I) will decide on the best way.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ay\-lì\-'u\-fa aw\-nge\-yä 'ey\-lan\-ä a\-'e\-wan}, in the words of our young friend.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\emph{b. used to attribute active and passive participles to a noun}) 
\B{Pa\-lu\-lu\-kan a\-tu\-sa\-ron lu le\-hrr\-ap.} A hunting thanator is dangerous.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln} 
\B{lì'\-fya a\-paw\-nll\-txe}, a spoken language.~\LINK{http://forum.learnnavi.org/language-updates/a-new-participle-infix-awn/msg140685/\#msg140685}{ln} 

% a fì'u
\hi
\B{a fì\cp{'u}}~:~\HL{fwa}{\B{fwa}}

% a fì'ul
\hi
\B{a fì\cp{'ul}}~:~\HL{fula}{\B{fula}}

% a fì'ur
\hi
\B{a fì\cp{'ur}}~:~\HL{fura}{\B{fura}}

% a fì'uri
\hi
\B{a fì\cp{'u}ri}~:~\HL{furia}{\B{furia}}

% a fì'ut
\hi
\B{a fì\cp{'ut}}~:~\HL{futa}{\B{futa}}

% a krr
\hi
\HT{a krr}{\B{a \cp{krr}}} \emph{or} \HL{krra}{\B{krr a}}, \HL{krra}{\B{\cp{krr}a}} \I{[a "k\s{r}]} \emph{or} \I{["k\s{r}.a]} \emph{conj.} when (\emph{to conclude a clause of time}). 
\B{tì\-'eyng\-it oel to\-lel a krr, ay\-nga\-ru pa\-yeng}, when I have recieved an answer, (I) will tell you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\ding{43}~\HL{krr}{\B{krr}})

% a tsa'u
\hi
\B{a tsa\cp{'u}}~:~\HL{tsawa}{\B{tsawa}}

% a tsa'ul
\hi
\B{a tsa\cp{'ul}}~:~\HL{tsala}{\B{tsala}}

% a tsa'ur
\hi
\B{a tsa\cp{'ur}}~:~\HL{tsara}{\B{tsara}}

% a tsa'uri
\hi
\B{a tsa\cp{'u}ri}~:~\HL{tsaria}{\B{tsaria}}

% a tsa'ut
\hi
\B{a tsa\cp{'ut}}~:~\HL{tsata}{\B{tsata}}

% afpawng / si
\hi 
\HT{afpawng}{\B{a\cp{fpawng}}} \I{[a."fpawN]} \textsc{i}.~\emph{n.} grief. 
\B{Maw \mbox{kxitx} sem\-pul\-ä lar\-mä\-ngu Pey\-ral\-ä a\-fpawng txew\-luke.} After (her) father's death, Pey\-ral's grief was endless.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{ii}. \B{a\cp{fpawng} si} \I{[a."fpawN si]} \emph{vin.} grieve. \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nafpawng}{\B{na\-fpawng}}, grievingly, with grief.

% Aha'ri
\hi
\B{A\cp{ha'}ri} \I{[a."haP.Ri]} \emph{n.} \emph{female name}.

% aho
\hi 
\HT{aho}{\B{a\cp{ho}}} \I{[a."ho]} \emph{vin.} pray. 
\B{Ey\-wa\-ru a\-ho, ma 'i\-tan, fte Nawm\-a Sa'\-nok\-ìl tì\-ye\-vìng ngar tì\-txur\-it.} Pray to Eywa, my son, that Great Mother will give you strength.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIaho}{\B{tì\-aho}}, prayer (abstract). 
\on{2}.~\HL{saho}{\B{sa\-ho}}, prayer.

% akrrmaw
\hi
\B{a\cp{krr}maw}~:~\HL{mawkrra}{\B{mawkrra}}

% akrrta
\hi
\B{a\cp{krr}ta}~:~\HL{takrra}{\B{takrra}}

% akum
\hi
\B{a\cp{kum}}~:~\HL{kuma}{\B{kuma}}

% Akwey
\hi
\B{Ak\cp{wey}} \I{[ak\textcorner."wEj]} \emph{n.} \emph{male name}.
\B{Trr\-am\-ä ay\-u\-van\-ìri mak\-to Ak\-wey nì\-yew\-la ha snaytx.} Akwey rode disappointingly in yesterday's games so he lost.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% alaksi
\hi
\HT{alaksi}{\B{a\cp{lak}si}} \I{[a."lak\textcorner.si]} \emph{adj.} ready. 
\B{Alaksi srak, ma frapo?} Are you ready, everyone?~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{Ra\-lu rì\-kxi krr\-a srew, rì\-'ìr si pa\-lu\-kan\-ur a lu a\-lak\-si fte spi\-vä.} Ra\-lu does a shake while dancing, imitating a thanator that's ready to leap.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}

% alìm
\hi
\HT{alIm}{\B{a\cp{lìm}}} \I{[a."lIm]} \emph{adv.} far away, at a distance. 
\B{'Ì'awn alìm!} Stay back!~\LINK{http://forum.learnnavi.org/language-updates/near-distant-and-irregular-adverbs/}{a\slash ln} 
(\ding{43}~\HL{lIm}{\B{lìm}}; \ding{214} \HL{asim}{\B{asim}} ) \\
\textsc{Derivatives}:\ding{43} \on{1}. \HL{'awnIm}{\B{'awnìm}}, avoid.

% ‹alm›
\hi
‹\HT{alm}{\B{alm}}› \I{[al.m]} \emph{inf. in} ‹1› \emph{to form the perfective past}. 
\B{na'\-rìng a ke lal\-mu tsar kea rìk}, the forest that had no leaves.~\LINK{http://forum.learnnavi.org/language-updates/for-\%28duration\%29-jesus-loan-word/}{ln} 
\B{Fkxi\-let a tsaw\-fa poe ioi sä\-pal\-mi ngo\-lop Va'\-rul.} Va'ru is the one who created the necklace she was wearing.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
\B{Lì'\-fya\-ri le\-Na'\-vi 'Rr\-ta\-mì, vay set 'al\-mong a fra\-'u ze\-ra\-'u ta ngrr\-po\-ngu.} Everything that has gone on with (blossomed regarding) Na'vi until now on Earth has come from a grassroots movement.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}

% alo
\hi
\HT{alo}{\B{\cp{a}lo}} \I{["{}a.lo]} \emph{n., adv.} time, turn, instance, \emph{one of a number of repeated or recurring actions}. 
\B{Ayupxareri angim nìhawng lu alo oeyä!} Now it's my turn for a message that's too long!~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Fmi pi\-vll\-txe alo avol nì\-win!} Try to say (that) eight times fast!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
\B{nì\-sìl\-pey alo ahay}, hopefully, next time.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b} \\  
\textsc{Derivatives}:\ding{43}
\on{1}.~\HL{'awlo}{\B{'aw\-lo}}, once.  
\on{2}.~\HL{melo}{\B{me\-lo}}, twice. 
\on{3}.~\HL{pxelo}{\B{pxe\-lo}}, three times. 
\on{4}.~\HL{hayalo}{\B{ha\-ya\-lo}}, next time. 
\on{5}.~\HL{hamalo}{\B{ha\-ma\-lo}}, last time.

% alu
\hi
\HT{alu}{\B{\cp{a}lu}} \I{["{}a.lu]} \emph{conj.} (\emph{a. used for apposition}) that is, in other words \B{Tska\-lep\-it oel to\-lìng oe\-yä tsmu\-kan\-ur alu Ìstaw.} I gave the crossbow to my brother Istaw.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Txoa li\-vu, yaw\-ne lu oer So\-rewn … alu … ke tsun oeng mun\-txa sli\-vu} Sorry, but I love Sorewn \dots in other words, you and I cannot marry.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Srane, lì\-'ul alu mì fra\-krr ley\-ka\-tem lì\-'ut a\-hay tsa\-fya.} Yes, the word \emph{mì} always changes the next word like that.~\LINK{http://forum.learnnavi.org/language-updates/fmawno/msg293595/\#msg293595}{ln} 
(\emph{b. in contrastive comparisons}) \B{Fì\-kar\-yu alu fì\-po lu tsul\-fä\-tu; tsa\-kar\-yu alu tsa\-po lu \mbox{skxawng}.} \emph{This} teacher is a master; \emph{that} teacher is a fool.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
(\emph{c. to mean}) ``as, in the capacity of''. \B{Oe alu Tsa\-hìk ke tsun mì\-ftxe\-le tsngi\-vaw\-vìk; oe alu sa'\-nok tsun.} As Tsahik, I cannot weep over this matter; as a mother, I can.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} 
(\ding{43}~\HL{sko}{\B{sko}})

% alunta
\hi
\B{a\cp{lun}ta}~:~\HL{talun1}{\B{taluna}}

% ‹aly›
\hi
‹\HT{aly}{\B{aly}}› \I{[al.j]} \emph{inf. in} ‹1› \emph{to form the perfective future}.

% ‹am›
\hi
‹\HT{am}{\B{am}}› \I{[a.m]} \emph{inf. in} ‹1› \emph{to form the past tense}. 
\B{Ay\-lì'\-fya yaw\-ne le\-ru oer ta\-krr\-a 'e\-veng la\-mu.} I've loved languages since I was a child.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}

% am'a
\hi
\HT{am'a}{\B{am\cp{'a}}} \I{[am."Pa]} \emph{n.} doubt. 
\B{Ke lu oer am'a.} I have no doubt.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}. \HL{am'ake}{\B{am'ake}}, sure, confident. 
\on{2}.~\HL{am'aluke}{\B{am'aluke}}, without a doubt.

% am'ake
\hi
\HT{am'ake}{\B{am\cp{'a}ke}} \I{[am."Pa.kE]} \emph{adj.} sure, confident. 
\B{Tsa\-ri\-a pol aw\-nga\-ti ke txa\-yìng oe 'e\-fu am\-'a\-ke nì\-wotx.} I'm entirely confident that he won't abandon us.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{am'a}{\B{am\-'a}})

% am'aluke
\hi
\HT{am'aluke}{\B{am\cp{'a}luke}} \I{[am."Pa.lu.kE]} \emph{adv.} without a doubt. 
\B{Am\-'a\-lu\-ke sna\-yaytx Saw\-tu\-te, ya\-yo\-ra' Na'\-vi.} Without a doubt the Sky People will lose and the People will win.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{am'a}{\B{am\-'a}}, \HL{luke}{\B{lu\-ke}})

% amay
\hi 
\HT{amay}{\B{a\cp{may}}} \I{[a."maj]} \emph{n.} brook. 

% Amerika
\hi
\B{Amerika} \I{[a."mE.Ri.ka]} \emph{n.}, \textsc{lw} America, North America.
\B{Trr le\-fpom, ma Su\-te Ame\-ri\-ka\-yä.} Good morning, people of America.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}

% Amhul
\hi
\B{Am\cp{hul}} \I{[am."hul]} \emph{or} \I{[."hUl]} \emph{n.} \emph{female name}.

% anìheyu
\hi
\HT{anIheyu}{\B{anì\cp{he}yu}} \I{[a.nI."hE.ju]} \emph{n.}~\ding{96} fibonacci, ``blue spiral plant'', \emph{orbis caeruleus}.
(\ding{43}~\HL{ean}{\B{e\-an}}, \HL{'Iheyu}{\B{'ì\-he\-yu}})

% Anuk
\hi
\B{Anuk} \I{[a.nuk\textcorner]} \emph{or} \I{[a.nUk\textcorner]} (\textsc{rn}: \B{A\-nùk} \I{[a.nUk\textcorner]})  \emph{n.} \emph{given name}.~\LINK{}{dh}

% Anurai
\hi
\B{\cp{A}nurai} \I{["{}a.nu.Ra.i]} \emph{n.} \emph{clan name, specialise in artisinal crafts}.
\B{Oe lu Anu\-rai\-yä syen\-a ha\-pxì a rey.} I am the last living member of the Anurai.~\LINK{http://naviteri.org/2015/05/cirque-du-soleils-toruk-the-first-flight-teaser-video-online/}{b}

% Aonung
\hi 
\HT{Aonung}{\B{\cp{A}onung}} \I{["{}a.o.nuN]} \emph{or} \I{[.nUN]} \emph{n. male name}. 

% apxa
\hi
\HT{apxa}{\B{a\cp{pxa}}} \I{[a."p'a]} \emph{adj} large. 
\B{Oeyk tsa\-tì\-snaytx\-ä lu a\-pxa sä\-nu\-i tì\-eyk\-tan\-ä.} That loss was caused by a great failure of leadership.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{Mì sray apxa kelku si soaiahu.} (Txewì) lives in a large village with his family.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}  
\B{skxawng apxa}, prodigious moron.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} \\
\textsc{Derivatives}:\ding{43} \on{1}. \HL{apxangrr}{\B{apxangrr}}, delta tree.

% apxangrr
\hi
\HT{apxangrr}{\B{a\cp{pxa}ngrr}} \I{[a."p'a.N\s{r}]} \emph{n.} \ding{96} delta tree, ``large root'', \emph{magelum triangulare}.
(\ding{43}~\HL{apxa}{\B{a\-pxa}}, \HL{ngrr}{\B{ngrr}})

% ‹arm›
\hi
‹\HT{arm}{\B{arm}}› \I{[aR.m]} \emph{inf. in} ‹1› \emph{to form the imperfective past}. 
\B{Nga mal lar\-mu oer!} I trusted you!~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
\B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok, tso\-le\-'a syep\-tu\-tet a\-tsawl fra\-to mì sì\-rey.} Yesterday, while I was with Txewì in the forest, we saw the biggest Trapper I've ever seen.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{ny\slash wiki}
\B{Po\-el oe\-ti lar\-mawk.} She was talking about me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} 
\B{Tì\-kang\-kem\-ì\-ri var\-mrr\-ìn oe nì\-wotx.} I was completely overwhelmed at work.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
\B{\mbox{kxawm} oe har\-ma\-hä\-ngaw}, maybe I was sleeping.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}

% Artsut
\hi
\B{Artsut} \I{[aR.\t{ts}ut\textcorner]} \emph{or} \I{[.\t{ts}Ut\textcorner]} (\textsc{rn}: \B{Ar\-tsùt} \I{[aR.\t{ts}Ut\textcorner]}) \emph{n.} \emph{female name}.~\LINK{}{dh}

% Arvok
\hi
\B{Arvok} \I{[aR.vok\textcorner]} \emph{n.} \emph{male name}.~\LINK{}{dh}

% ‹ary›
\hi
‹\HT{ary}{\B{ary}}› \I{[aR.j]} \emph{inf.~in} ‹1› \emph{to form the imperfective future}.

% asim
\hi
\HT{asim}{\B{a\cp{sim}}} \I{[a."sim]} \emph{adv.} nearby, at close range. 
\B{New oe ri\-vun a\-sim tì\-fnu\-nga'\-a tseng\-it}, I want to find a quiet place nearby.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{sim}{\B{sim}}, \ding{214}~\HL{alIm}{\B{alìm}})

% asip
\hi 
\HT{asip}{\B{\cp{a}sip}} \I{["{}a.sip\textcorner]} \emph{n.} (a.) tall thin mass or pile of s.t. (b.) tower. 
\B{Tu\-te\-ol a\-sip\-it ay\-tä\-re\-mä txo\-lu\-la mì na'\-rìng.} Someone built a tower of bones in the forest.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}  

% ‹asy›
\hi
‹\HT{asy}{\B{asy}}› \I{[a.sj]} \emph{inf. in} ‹1› \emph{to form the intentional future}. 
\B{Ay\-oe ke wa\-syem.} We will not fight.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{g\slash ln} 
\B{Ke za\-syup lì\-'Ona ne kxu\-tu a mì\-fa fu a wrr\-pa.} The l'Ona will not perish to the enemy within or the enemy without.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{g\slash ln}

% atan 1
\hi
\HT{atan}{\B{a\cp{tan}}}\textsuperscript{\on{1}} \I{[a."tan]} \emph{n.} light. 
\B{Fra\-tseng lu atan, fnu fra\-po.} Everywhere is light, everyone is quiet.~\LINK{http://forum.learnnavi.org/news-announcements/christmas-song-ninavi-on-youtube/}{ln} \\ 
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{atanvi}{\B{atanvi}}, ray (of light). \\  
\on{2}.~\HL{atanzaw}{\B{atanzaw}}, forked lightning. \\  
\on{3}.~\HL{syuratan}{\B{syuratan}}, bioluminescence. \\
\on{4}.~\HL{txanatan}{\B{txanatan}}, bright, vivid.

% Atan 2
\hi
\B{A\cp{tan}}\textsuperscript{\on{2}} \I{[a."tan]} \emph{n.} \emph{male name}.~\LINK{}{dh}

% atanvi
\hi
\HT{atanvi}{\B{a\cp{tan}vi}} \I{[a."tan.vi]} \emph{n.} ray (of light).
(\ding{43}~\HL{atan}{\B{a\-tan}})

% atanzaw
\hi
\HT{atanzaw}{\B{a\cp{tan}zaw}} \I{[a."tan.zaw]} \emph{n.} forked lightning.
(\ding{43}~\HL{atan}{\B{a\-tan}}, \HL{swizaw}{\B{swi\-zaw}}; \HL{rawm}{\B{rawm}})

% Ateyo
\hi
\B{A\cp{te}yo} \I{[a."tE.jo]} \emph{n.} \emph{male name}.
\B{Tsay\-sam\-si\-yu lu su\-pe? -- Fo lu Ka\-mun, Ra\-lu, Ìstaw, sì Ate\-yo.} They're Ka\-mun, Ra\-lu, Ìstaw, and Ate\-yo.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}

% atokirina'
\hi
\HT{atokirina'}{\B{atoki\cp{ri}na'}} \I{[a.to.ki."Ri.naP]} \emph{n.}, \ding{96} \emph{atokirina}, seeds of the great tree, woodsprite. 
\B{Pxay\-a a\-to\-ki\-ri\-na' ne fì\-ke\-tu\-wong zì\-ma\-'u.} Many \emph{atokirina} just came to this alien.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/}{a\textsuperscript{c}\slash ln}
(\ding{43}~\HL{rina'}{\B{ri\-na'}})

% atxar
\hi
\HT{atxar}{\B{a\cp{txar}}} \I{[a."t'aR]} \emph{adj.} smell of living animals (\emph{as found around a watering hole or animal nest}). 

% atxkxe
\hi
\HT{atxkxe}{\B{atx\cp{kxe}}} \I{[at'."k'E]} \emph{n} land. 
\B{Tsun aw\-nga kel\-ku si\-vi nì\-'eng Saw\-tu\-te\-hu mì atx\-kxe aw\-nge\-yä.} We can share our land with the Skypeople.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
\B{Fì\-atx\-kxe\-ti ke ra\-ya\-'un ay\-oel kaw\-krr!} We will never give up this land.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{atxkxerel}{\B{atx\-kxe\-rel}}, map. 
\on{2}.~\HL{txanlokxe}{\B{txan\-lo\-kxe}}, tribal or clan domain; country.

% atxkxerel
\hi
\HT{atxkxerel}{\B{atx\cp{kxe}rel}} \I{[at'."k'E.REl]} \emph{n.} map.
(\ding{43}~\HL{atxkxe}{\B{atx\-kxe}}, \HL{rel}{\B{rel}})

% ‹ats›
\hi
‹\HT{ats}{\B{ats}}› \I{[a.\t{ts}]} \emph{inf. in} ‹2› \emph{to form the evidential, i.e. uncertainty or indirect knowledge}. 
\B{Fpìr\-mìl oel fu\-ta ay\-nga na\-tsew tsi\-ve\-'a fì\-'ut.} I was just thinking that you might want to see this.~\LINK{http://forum.learnnavi.org/language-updates/the-evidential-at-last/msg106055/\#msg106055}{ln} 
\B{Srung sa\-tsi fra\-por.} (That) might help everyone.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/comment-page-1/\#comment-1062}{b}
\B{Po\-an yaw\-ne la\-tsu poe\-ru nì\-lam.} Apparently she loves him.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}

% au
\hi
\HT{au}{\B{\cp{a}u}} \I{["{}a.u]} \emph{n.} drum (made of skin). 
\B{Po pam\-tse\-o si fa au nìl\-tsan nì\-txan.} She plays the drums very well.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}  \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{wokau}{\B{wokau}}, pendulum drum.

% au
\hi
\HT{au2}{\B{a\cp{u}}} \I{[a."{}u]} \emph{intj.} \emph{exclamation of consternation}.~\LINK{https://forum.learnnavi.org/intermediate/my-dictionary/msg263252/\#msg263252}{ln} 

% aungia
\hi
\HT{aungia}{\B{a\cp{u}ngia}} \I{[a."{}u.Ni.a]} \emph{n.} sign, omen. 
\B{Tsa\-krr za\-'u a\-u\-ngi\-a ta Ey\-wa.} Then came a sign from Eywa.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-296}{a\slash b} 
\B{Ze\-ne Sa'\-nok fì\-a\-u\-ngi\-a\-ti ral\-pi\-veng.} Mother has to interpret this omen.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/}{ln\slash a\textsuperscript{c}}

% awaiei
\hi
\HT{awaiei}{\B{awai\cp{e}i}} \I{[a.wa.i."{}E.i]} \emph{n.} \ding{96} banshee of paradise, \emph{bansheba terrestre}.

% ‹ay›
\hi
‹\HT{ay1}{\B{ay}}› \I{[a.j]} \emph{inf. in} ‹1› (to form the future tense). 
\B{Pol\-txe Ney\-ti\-ril ay\-lì\-'ut a fra\-krr 'ok seyä la\-yu oer.} Neytiri said something that I will always remember.~\LINK{http://thereadinginpublicproject.blogspot.de/2010/05/paul-frommer.html}{pr}. 
\B{Tì\-'eyng\-it oel to\-lel a krr, ay\-nga\-ru pa\-yeng, tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} When I've received an answer, I will tell you and then decide on the best way to bring you the words and rules of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}

\end{multicols}
