\pdfbookmark[0]{AW}{AW}
\begin{center}
\section*{AW \I{[aw]}}
\end{center}

\begin{multicols}{2}

% -aw
\hi
\HT{-aw}{\B{-aw}} \I{[.aw]} \emph{num., suf.} one (\emph{rest of} \ding{43}~\HL{'aw}{\B{'aw}} \emph{to form higher numbers}). \B{volaw}, nine (decimal), \on{11} (octal). \B{mevolaw}, \on{17} (decimal), \on{21} (octal); etc.

% ‹awn›
\hi
‹\HT{awn}{\B{awn}}› \I{[aw.n]} \emph{inf. in} ‹1› (to form the passive participle) 
(\emph{a. always used attributively}) \B{Ay\-lì\-'u a\-paw\-nll\-txe nìl\-tsan!} Words well-spoken.~\LINK{http://forum.learnnavi.org/language-updates/a-new-participle-infix-awn/msg140685/\#msg140685}{ln} 
\B{Ru\-txe oeyä tì\-o\-eyk\-tìng\-it a\-saw\-nung ni\-vìn.} Please look at my added explanation.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-581}{b} 
\B{Saw\-no'\-ha ioit ko\-la\-ne\-i\-om oel u\-van\-fa!} I got (won) the prized piece of jewelry in the game!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{b. shortened in military jargon}) \B{Tì\-kan taw\-na\-tep!} Target lost!~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{g\slash b}
(\emph{c. shortened in proverbial expression}) \B{(Na) lo\-re\-yu 'aw\-nam\-pi}, extreme shyness, [lit.: (like) a touched helicoradian] \B{Lu por mok\-ri a\-mik\-lor, slä lo\-re\-yu 'aw\-nam\-pi lu. Ke tsun ri\-vol eo su\-te.} She has a beautiful voice, but she's extremely shy. She can't sing in front of people.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\emph{d. in special adv. use}) \B{nìaw\cp{no}mum}, as you know, as is known. \B{nì\cp{zaw}nong}, safely.

% awnga
\hi
\HT{awnga}{\B{aw\cp{nga}}} \I{[aw."{}Na]} \emph{pn.} (\emph{shortened form of} \ding{43}~\HL{ayoeng}{\B{ay\-oeng}}) we all (\emph{inclusive}). 
\B{Aw\-nga sngi\-vä'i ko!} Let us begin!~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
\B{aw\-ngal yom wu\-tsot teng\-krr he\-reyn nì\-pxim}, we eat dinner while sitting upright.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
\B{Za\-ya\-'u Saw\-tu\-te fte aw\-nga\-ti ski\-va\-'a kay zì\-sìt a\-pxey!} The Sky People will come to destroy us three years from now!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
\B{Po aw\-nga\-ru ka\-vuk so\-li.} He betrayed us.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/comment-page-1/\#comment-189}{b}  
\B{Slo\-lu po tsul\-fä\-tu lì'\-fya\-yä aw\-nge\-yä.} He's become a master of our language.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}

\end{multicols}
