\pdfbookmark[0]{AY}{AY}
\begin{center}
\section*{AY \I{[aj]}}
\end{center}

\begin{multicols}{2}

% ay+
\hi
\HT{ay}{\B{ay}}\texttt{+} \I{[aj.]} \emph{pref. to form the plural of a noun}. 
\B{ay\-nu\-me\-yu}, students; 
\B{ay\-zì\-sìt}, years; 
\B{ay\-sä\-o\-mum}, information; 
\B{ay\-ya\-yo}, birds. (can be left out when the noun undergoes lenition) \B{ay\-sa\-ron\-yu\slash sa\-ron\-yu}, hunters; 
\B{ay\-har\-yu\slash har\-yu}, teachers. \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fay}{\B{fay}}\texttt{+}, those. 
\on{2}.~\HL{pay1}{\B{pay}}\texttt{+}, which (of them)? 
\on{3}.~\HL{fray}{\B{fray}}\texttt{+}, all (of them).

% ‹ay›
\hi
‹\B{ay}› \I{[a.j]} \emph{inf. in} ‹1› (to form the future tense). 
\B{Pol\-txe Ney\-ti\-ril ay\-lì\-'ut a fra\-krr 'ok seyä la\-yu oer.} Neytiri said something that I will always remember.~\LINK{http://thereadinginpublicproject.blogspot.de/2010/05/paul-frommer.html}{pr}. 
\B{Tì\-'eyng\-it oel to\-lel a krr, ay\-nga\-ru pa\-yeng, tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} When I've received an answer, I will tell you and then decide on the best way to bring you the words and rules of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}

% ayfeyä
\hi
\B{ayfeyä}~:~\HL{ayfo}{\B{ayfo}}

% ayfo
\hi
\HT{ayfo}{\B{ay\cp{fo}}}, \emph{also} \HL{fo}{\B{fo}} \I{[aj."fo]} \emph{also} \I{[fo]} \emph{pn.} (\emph{gen.} \B{ay\-fe\-yä\slash feyä}) they (3\textsuperscript{rd} person plural). 
\B{Ay\-fo so\-lop ìlä hil\-van fa u\-ran.} They traveled along the river by boat.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
\B{Ay\-oengal fte Fay\-saw\-tu\-tet ki\-vu\-rakx, ze\-ne nì\-'aw\-ve ay\-fot tsli\-vam.} We must understand these Sky People if we are to drive them out.~\LINK{http://forum.learnnavi.org/language-updates/confirmation-on-kurakx-eytukan-line-and-toitslan/}{ln} 
\B{Fo\-ru 'u\-pxa\-ret oel fpo\-le'.} I have sent them a message.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} \B{hol\-pxay ay\-zek\-wä\-yä fe\-yä}, the number of their fingers.~\LINK{http://forum.learnnavi.org/language-updates/talun-taweyk-tafral-lun-oeyk-holpxay-hol/}{ln} 
\B{Fo\-ri maw\-krr\-a fa ren\-ten ioi sä\-po\-li ho\-lum.} After they put on their goggles, they left.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
(\ding{43}~\HL{po}{\B{po}}, \HL{ay}{\B{ay}}\texttt{+})

% ayla
\hi
\HT{ayla}{\B{ay\cp{la}}} \I{[aj."la]} : \HL{aylahe}{\B{aylahe}}. 

% aylahe
\hi
\HT{aylahe}{\B{ay\cp{la}he}} \I{[aj."la.hE]} \emph{pn.} the others, others (\emph{forms:}) \B{aylahe}, \B{aylahel}, \B{aylahet(i)}, \B{aylaheyä}, \B{aylaher(u)}, \B{aylaheri}). 
\B{Kem\-ìri a nga\-ru prr\-te' ke lu, tsa\-kem rä\-'ä si\-vi ay\-la\-he\-ru.} As for any action that you don't like, don't act that way to others.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} 
(\emph{shortened forms:} \B{ayla}, \B{aylal}, \B{aylat(i)}, \B{ayleyä}, \HL{aylaru}{\B{ay\-lar(u)}}, \B{aylari})
\B{Fì\-po\-ti oel tspì\-yang, fte tì\-ke\-nong li\-ye\-vu ay\-la\-ru.} I will kill this one as a lesson to the others.~\LINK{http://forum.learnnavi.org/language-updates/capturing-avatar/}{b}
\B{'Aw\-a tì\-pawm\-ì\-ri 'i\-veyng oe set;  ay\-la\-ri zu\-saw\-krr 'ay\-eyng.} Let me answer one of the questions now; the others (I'll) answer in the future.~\LINK{http://forum.learnnavi.org/language-updates/email-from-frommer-re-numbers-\%28or-the-full-number-chart!\%29/}{ln}
\ding{43}~\HL{lahe}{\B{la\-he}})

% aylal
\hi
\HT{aylal}{\B{ay\cp{lal}}} \I{[aj."lal]} : \HL{aylahe}{\B{aylahe}}. 

% aylari
\hi
\HT{aylari}{\B{ay\cp{la}ri}} \I{[aj."la.Ri]} : \HL{aylahe}{\B{aylahe}}. 

% aylaru
\hi
\HT{aylaru}{\B{ay\cp{la}ru}} \I{[aj."la.Ru]} : \HL{aylahe}{\B{aylahe}}. 

% aylat
\hi
\HT{aylat}{\B{ay\cp{lat}}} \I{[aj."lat\textcorner]} : \HL{aylahe}{\B{aylahe}}. 

% ayleyä
\hi
\HT{ayleyA}{\B{ay\cp{le}yä}} \I{[aj."lE.j\ae]} : \HL{aylahe}{\B{aylahe}}. 

% aynga
\hi
\HT{aynga}{\B{ay\cp{nga}}} \I{[aj."Na]} \emph{pn.} (\emph{gen.} \B{ay\-\cp{nge}\-yä}) you (all) (2\textsuperscript{nd} person plural). 
\B{Le\-i\-u ay\-nga tstu\-nwi nì\-txan nì\-wotx.} You are all very kind.~\LINK{http://forum.learnnavi.org/news-announcements/happy-birthday-to-karyu-pawl/msg556899/\#msg556899}{ln} 
\B{Ay\-nga ne\-to ri\-vikx!} Step back! [lit.: move away you all]~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki} 
\B{Txo tsi\-ve\-'a ay\-ngal key\-eyt, ru\-txe oe\-ru pi\-veng.} If you see mistakes, please tell me.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
\B{Oel ay\-nga\-ti ka\-me\-i\-e, ma oe\-yä ey\-lan.} I \textsc{see} you all, my friends.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Sìl\-pey oe, la\-yu oe\-ru ye'\-rìn sìl\-tsan\-a fmawn a tsun oe ay\-nga\-ru ti\-vìng.} I hope, soon I will have good news that I can give to you all.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Si\-vu\-nu ay\-ngar fì\-tskxe\-keng\-tsyìp!} May you all enjoy this little excercise!~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Ay\-nge\-yä tì\-ftu\-si\-a 'o' li\-vu nì\-'aw!} May your studying be fun!~\LINK{http://naviteri.org/2012/10/sound-files-added-to-previous-post/}{b} \B{Sìl\-pey oe, ay\-nga\-ri fì\-tì\-päng\-kxo\-tsyìp el\-tur tì\-txen sil\-vi.} I hope, this little discussion has been interesting for you.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} 
\B{Te\-rì\-ran ay\-oe ay\-nga\-ne.} We are walking to you.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki} 
(\ding{43}~\HL{nga}{\B{nga}})

% ayngenga
\hi
\HT{ayngenga}{\B{aynge\cp{nga}}} \I{[aj.NE."Na]} \emph{pn.} (\emph{gen.} \B{ay\-nge\cp{nge}\-yä}) you all (2\textsuperscript{nd} person plural; deferential or ceremonial form of \B{ay\-nga}). 
\B{Ay\-nge\-nga\-ru o\-he\-yä tsmuk\-it alu Ne\-wey te Tska\-ha So\-rewn\-'i\-te.} Allow me to introduce my sister, Newey te Tskaha Sorewn'ite.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{nga}{\B{nga}})

% ayoe
\hi
\HT{ayoe}{\B{ay\cp{o}e}} \I{[aj."{}o.e]} (\emph{with case endings or enclitic adp.s always pronounced as:} \I{[aj."wE-]}) \emph{pn.} we (1\textsuperscript{st} person plural, exclusive). 
\B{Te\-rì\-ran ay\-oe ay\-nga\-ne.} We are walking to you.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki} 
\B{Nì\-aw\-no\-mum ngo\-lop ay\-oel fì\-nu\-mul\-txa\-ti fpi sngä\-'i\-yu.} As you know, we've created this class for beginner.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} 
\B{ul\-te kaw\-tu ke tsun ay\-oer tì\-ftang si\-vi}, and no one can stop us.~\LINK{http://forum.learnnavi.org/language-updates/stop!/}{ln} 
(\ding{43}~\HL{oe}{\B{oe}})

% ayohe
\hi
\HT{ayohe}{\B{ay\cp{o}he}} \I{[aj."{}o.hE]} \emph{pn.} (\emph{deferential or ceremonial form of}) we (1\textsuperscript{st} person plural exclusive).  The inclusive form is \B{ohe ay\-nge\-nga\-sì}, I and you, or \B{ohe pxe\-nge\-nga\-sì}, I and (the three of) you. 
(\ding{43}~\HL{ayoe}{\B{ayoe}})

% ayoeng
\hi
\HT{ayoeng}{\B{ay\cp{oeng}}} \I{[aj."wEN]} \emph{or} \HL{awnga}{\B{aw\cp{nga}}} \I{[aw."Na]} \emph{pn.} we (1\textsuperscript{st} person plural inclusive) (\emph{with case endings, the forms are:} \B{ayoe\cp{ngal}, ayoe\cp{ngat}(i), ayoe\cp{ngar}(u), ayoe\cp{nge}yä, ayoe\cp{nga}ri} \emph{with a stress shift to the} \B{nga}\emph{-part}) \I{[aj.wE."Nal]} etc. 
\B{Ay\-oengal fte Fay\-saw\-tu\-tet ki\-vu\-rakx, ze\-ne nì\-'aw\-ve ay\-fot tsli\-vam.} We must understand these Sky People if we are to drive them out.~\LINK{http://forum.learnnavi.org/language-updates/confirmation-on-kurakx-eytukan-line-and-toitslan/}{ln} 
(\ding{43}~\HL{awnga}{\B{aw\-nga}}, \HL{oeng}{\B{oeng}})

% ayram alusìng
\hi
\HT{ayram alusIng}{\B{Ay\cp{ram} Alu\cp{sìng}}} \emph{also} \B{ayRam aLusìng} \I{[aj."Ram a.lu."{}sIN]} \emph{n.} the Floating Mountains, the Hallelujah Mountains. 
\B{Aw\-nga kel\-ku si nu\-ä ay\-ram a\-lu\-sìng.} We live beyond the flying mountains.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
\B{Fwa tsyìl Ay\-ram\-it A\-lu\-sìng lu sä\-'e\-o\-i\-o a ze\-ne fra\-po si\-vi fte sl\-ivu ta\-ron\-yu.} Climbing Iknimaya is a ritual that everyone has to perform to become a hunter.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{ram}{\B{ram}}, \HL{lIng}{\B{lìng}})

% ayu
\hi
\B{ay\cp{u}} \I{[aj."{}u]}~:~\HL{'u}{\B{'u}}

% ayvitrayä ramunong
\hi
\HT{ayvitrayA ramunong}{\B{Ayvit\cp{ra}yä Ra\cp{mu}nong}} \I{[aj.vit\textcorner."Ra.j\ae{} Ra."mu.noN]} \emph{n.} the Well of Souls (\emph{the closest connection to Eywa on Pandora, a point of extreme spiritual significance to the Na'vi}).
\B{Oe tsway\-on io Ay\-vit\-ra\-yä Ra\-mu\-nong.} I fly over the Tree of Souls.~\LINK{http://forum.learnnavi.org/language-updates/over-under-and-quickly/}{ln}
(\ding{43}~\HL{vitra}{\B{vit\-ra}}, \HL{ramunong}{\B{ra\-mu\-nong}})

\end{multicols}