\pdfbookmark[0]{AE}{AE}
\begin{center}
\section*{Ä \I{[\ae]}}
\end{center}

\begin{multicols}{2}

% -ä
\hi
\HT{-A}{\B{-ä}} \I{[.\ae]} \emph{suf. to mark the genitive case of a n. ending in a consonant, pseudovowel or u, o}. (\ding{43}~\HL{-yA}{\B{-yä}})

% äie
\hi
\HT{Aie}{\B{ä\cp{i}e}} \I{[\ae."{}i.E]} \emph{n.} vision (\emph{spiritual sense}). 
\B{Fì\-ta\-ron\-yur a\-pxan te\-swo\-ti\-vìng äiet a\-ngay.} Grant this worthy warrior true vision.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Becoming_one_of_the_Na.27vi}{a\textsuperscript{c}}

% Änsìt
\hi
\B{\cp{Än}sìt} \I{["{}\ae{}n.sIt\textcorner]} \emph{n. male name}.

% ‹äng›
\hi
‹\HT{Ang}{\B{äng}}› \I{[\ae.N]} \emph{or} ‹\B{eng}› \I{[E.N]} (\emph{before i, ì}) \emph{inf. in} ‹2› (to indicate negative feelings of the speaker). 
\B{Slä vay set, ke pa\-mä\-hä\-ngem kea tì\-'eyng.} But unfortunately until now, no answer arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}. 
\B{Fì\-skxawng\-ì\-ri tsap\-'a\-lu\-te se\-ngi oe.} I apologize for this moron.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki} (\ding{214}~‹\HL{ei}{\B{ei}}›)

% äo
\hi
\HT{Ao}{\B{\cp{ä}o}} \I{["{}\ae.o]} \emph{adp.} under, below. 
\B{Pol sä\-'o\-ti wo\-lan äo ay\-rìk.} He hid his tool under the leaves.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} 
\B{Tsa\-txon ay\-ut\-ral\-äo krr a pol oe\-ti no\-lìn nì\-wa\-we.} That night beneath the trees when she looked at me meaningfully.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} (\ding{214}~\HL{io}{\B{io}})

% ‹äp› ‹äpeyk›
\hi
‹\HT{Ap}{\B{äp}}› \I{[\ae.p]} \emph{inf. in} ‹pre-1› \textsc{i}. \on{1}. (to form a reflexive statement; only possible with \emph{vtr.} and certain \B{si}-verbs). \B{Oe tsä\-pe\-'a.} I see myself.~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln} \B{oe ngar mu\-wä\-pi\-vìn\-txu}, I introduce myself to you.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} \B{Ke ze\-ne win sä\-pi\-vi.} Take your time; don’t rush.~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b} \on{2}. (together with \B{fìtsap} to convey `each other') \B{Zì\-sìt\-o a\-vol ke tsä\-po\-le\-'a fo fì\-tsap.} They haven't seen each other in eight years.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b} \\
\textsc{ii}. ‹\B{äpeyk}› \I{[\ae.pEj.k]} \emph{inf. in} ‹pre-1› (to form a causative reflexive, i.e. `cause oneself to do s.t.') \B{po tä\-pey\-kì\-ye\-ver\-ke\-i\-up nì\-näk}, I am jazzed that he is apparently about to drink himself to death.~\LINK{http://forum.learnnavi.org/language-updates/sieyng-a-ftu-naring-8-participles-causatives-and-reflexives-oh-my!/}{ln}

% ätxäle / si
\hi
\HT{AtxAle}{\B{ä\cp{txä}le}} \I{[\ae{}."t'\ae{}.lE]} \textsc{i}. \emph{n.} request. \\
\textsc{ii}. \HT{AtxAle si}{\B{ä\cp{txä}le si}} \I{[\ae{}."t'\ae{}.lE si]} \emph{vin.} request (\emph{uses} \B{tsnì} \emph{to introduce the request}). 
\B{Ä\-txä\-le si tsnì li\-vu o\-he\-ru U\-nil\-ta\-ron.} I respectfully request the Dream Hunt.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{wiki} 
(\emph{a}.) \B{ä\-tx\-äle si oe pi\-vawm}, may I ask. 
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
(\emph{b. proverb}) \B{Ä\-txä\-le si pa\-lu\-lu\-kan\-ur tsnì smar\-it li\-vo\-nu.} \emph{or} \B{Ä\-txä\-le pa\-(lu)\-lu\-kan\-ur.} \emph{a futile gesture, an attempt to achieve something that might be desirable but will clearly not happen} [lit.: Ask a thanator to release its prey.]~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\emph{c. idiom}) \B{Tì\-ka\-ra lu ä\-txä\-le pa\-lu\-kan\-ur.} Resistance is futile [lit.: `reistance is a request to the thanator'].~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tsantxAl}{\B{tsan\-txäl}}, invitation.

% äzan / si
\hi 
\HT{Azan}{\B{ä\cp{zan}}} \I{[\ae."zan]} \textsc{i}. \emph{n.} force, compulsion. \\
\textsc{ii}. \B{ä\cp{zan} si} \I{[\ae."zan si]} \emph{vin.} force, compel (\emph{used with} \ding{43}~\HL{tsnI}{\B{tsnì}})
\B{Fo ä\-zan so\-li oer tsnì tsa\-kem si\-vi.} They forced me to do it.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{Azanluke}{\B{ä\-zan\-lu\-ke}}, voluntarily, without force.
\on{2}.~\HL{Azantu}{\B{ä\-zan\-tu}}, domineering person.

% äzanluke
\hi 
\HT{Azanluke}{\B{ä\cp{zan}luke}} \I{[\ae."zan.lu.kE]} \emph{adv.} voluntarily, without force or compulsion. 
\B{Vin oel äzanluke futa nga kivä poehu.} I request, without compulsion, that you go with her (i.e., I'm asking you to go with her, but I won't force you.)~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
\B{Vin oel futa nga kivä poehu äzanluke.} I request that you go with her voluntarily.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
(\ding{43}~\HL{Azan}{\B{äzan}}, \HL{luke}{\B{lu\-ke}})

% äzantu
\hi 
\HT{Azantu}{\B{ä\cp{zan}tu}} \I{[\ae."zan.tu]} \emph{n.} domineering person; one who is bossy, authoritarian, or dictatorial. 
(\ding{43}~\HL{Azan}{\B{äzan}}, \HL{-tu}{\B{-tu}})

\end{multicols}