\textsc{Derivatives}: \pdfbookmark[0]{E}{E}
\begin{center}
\section*{E \I{[E]}}
\end{center}

\begin{multicols}{2}

% e
\hi
\HT{e}{\B{e}} \I{[E]} \emph{intj.} ah, oh (i.e., `ah, I understand').~\LINK{https://forum.learnnavi.org/language-updates/navi-equivalent-for-ah/msg698116/\#msg698116}{ln}

% eampin
\hi
\HT{eampin}{\B{\cp{e}ampin}} \I{["{}E.am.pin]} \emph{n.} the color \ding{43}~\HL{ean}{\B{ean}}. 
(\ding{43}~\HL{'opin}{\B{'o\-pin}})

% ean
\hi
\HT{ean}{\B{\cp{e}an}} \I{["{}E.an]} \emph{adj.} blue, green. 
\B{Fì\-syu\-lang lu ean na ta'\-leng.} This flower is skin\hyp{}blue.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} 
\B{Fì\-syu\-lang a\-ean\hyp{}na\hyp{}ta'\-leng lor lu nì\-txan.\slash Ta'\-leng\-na\hyp{}ean\-a fì\-syu\-lang lor lu nì\-txan.} This skin\hyp{}blue flower is very beautiful.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{eampin}{\B{eam\-pin}}, (the color) blue\slash green. 
\on{2}.~\HL{rIkean}{\B{rì\-ke\-an}}, green. 
\on{3}.~\HL{ta'lengean}{\B{ta'\-le\-nge\-an}}, blue. 
\on{4}.~\HL{eanean}{\B{ean\-ean}}, cheadle.
\on{5}.~\HL{eanfwopx}{\B{ean\-fwopx}}, mist bloom.

% eanean
\hi
\HT{eanean}{\B{\cp{e}an\cp{e}an}} \I{["{}E.an."{}E.an]} \emph{n.} \ding{96} cheadle, herbaceous plant, \emph{thylakoidia spiralis}. 
(\ding{43}~\HL{ean}{\B{ean}})

% eanfwopx
\hi 
\HT{eanfwopx}{\B{\cp{e}anfwopx}} \I{["{}E.an.fwop']} \emph{n.} \ding{96} mist bloom [lit.: `blue dust cloud']. 
(\ding{43}~\HL{ean}{\B{ean}}, \HL{fwopx}{\B{fwopx}})

% ehetx / si
\hi 
\HT{ehetx}{\B{e\cp{hetx}}} \I{[E."hEt']} \textsc{i}. \emph{n.} (\emph{pl.} \B{me\-hetx}, \B{pxe\-hetx}, \B{ay\-e\-hetx}) excuse. \\
\textsc{ii}. \B{e\cp{hetx} si} \I{[E."hEt' si]} \emph{vin.} make an excuse\slash make excuses.
\B{Fu\-ria nga ke tsan\-'ul, var nga e\-hetx si\-vi nì\-'aw.} Regarding your lack of improvement, you only keep making excuses.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}

% ‹ei›
\hi
‹\HT{ei}{\B{ei}}› \I{[E.i]} \emph{or} ‹\B{eiy}› \I{[E.i.j]} (\emph{before i, ì, ll, rr})  \emph{inf. in} ‹2› (to indicate positive feelings of the speaker). 
\B{Pe\-yä tì\-kang\-kem\-ìl mi vey\-krr\-e\-i\-yìn pot nì\-wotx.} I'm glad, his work is still completely overwhelming him.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
(\emph{a. idiom}) \B{oel nga\-ti ka\-me\-i\-e}, I \textsc{see} you.~\LINK{http://www.vanityfair.com/online/oscars/2009/12/brushing-up-on-navi-the-language-of-avatar}{vf} 
(\emph{b. idiom}) \B{sre\-fe\-re\-i\-ey nì\-prr\-te'}, looking forward. \B{Tsa\-ri\-a nga\-hu ye'\-rìn ul\-txa si nì\-mun, oe sre\-fe\-re\-i\-ey nì\-prr\-te'.} I'm looking forward to getting together with you again soon.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
(\ding{214}~‹\HL{Ang}{\B{äng}}›)

% ekxan / ekxan si
\hi
\HT{ekxan}{\B{e\cp{kxan}}} \I{[E."k'an]} \textsc{i}. \emph{n.} (\emph{pl.} \B{me\-kxan}, \B{pxe\-kxan}, \B{ay\-e\-kxan}) barricade, obstruction. 
\B{Rä\-'ä fmi\-vi li\-vok fu em\-ki\-vä ay\-ekxan\-it a fkol ngo\-lop fpi sì\-kxu\-ke ay\-frr\-tuä sì ay\-ioang\-ä.} Don't try to approach or cross the barricades that were created for the sake of the visitors' and animals' safety.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b} \\
\textsc{ii}. \B{e\cp{kxan} si} \I{[E."k'an si]} \emph{vin.} exclude, keep out, bar.
\B{Sra\-ke fì\-kxem\-yo tsun tsay\-ioang\-ur le\-hrr\-ap ekxan si\-vi?} Can this wall keep out those dangerous animals?~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{kawkxan}{\B{kaw\-kxan}}, free, unblocked, unobstructed, clear.
\on{2}.~\HL{lekxan}{\B{le\-kxan}}, blocked, obstructed; frustrated.

% ekxtxu
\hi
\HT{ekxtxu}{\B{ekx\cp{txu}}} \I{[Ek'."t'u]} \emph{adj.} rough.   
\B{Ta'\-leng prr\-nen\-ä lu faoi, pum ko\-ak\-tu\-ä ekx\-txu.} A baby's skin is smooth, an old person's is rough.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
\B{Kll\-te lu ekx\-txu.} The ground is rough.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\ding{214}~\HL{faoi}{\B{faoi}})

% eltu
\hi
\HT{eltu}{\B{\cp{el}tu}} \I{["{}El.tu]} \textsc{i}. \emph{n.}~(\emph{pl.} \B{mel\-tu}, \B{pxel\-tu}, \B{ay\-el\-tu}) 
(\emph{a.}) brain. 
(\emph{b.}) \B{\cp{el}tu le\cp{fngap}} \I{["{}El.tu lE."fNap\textcorner]} computer, PC [lit.: metallic brain]. \\
\textsc{ii}. \B{\cp{el}tu si} \I{["{}El.tu si]} \emph{vin.} pay attention, quit goofing off. 
\B{El\-tu si! Tsa\-tstal a\-fwem lu litx nì\-txan.} Pay attention! That blunt knife is very sharp.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{iii}. \B{\cp{el}tur tì\cp{txen} si} \I{["{}El.tuR tI."t'En si]} \emph{vin.} be interesting, intriguing.   
\B{Tsa\-'u el\-tur tì\-txen si.} That's interesting.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{Tsa\-vur oe\-yä el\-tur tì\-txen sol\-i.} That story intrigued me.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{Tsa\-vur fe\-yä ay\-el\-tur tì\-txen so\-li.} They found that story interesting.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \B{Pxoe\-ri tsa\-vur el\-tur tì\-txen so\-li.} The three of us were interested in that story.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \\
\textsc{iv}. \B{\cp{el}tut hey\cp{ka}haw} \I{["{}El.tut\textcorner hEj."ka.haw]} \emph{or} \I{[.tUt\textcorner]} \emph{ph.} be boring. 
\B{Tsa\-sä\-ftxu\-lì\-'ul pe\-yä el\-tut hey\-ko\-la\-haw nì\-txan.} That speech of his was very boring.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{skeykx}{\B{skeykx}}) \\ 
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{eltungawng}{\B{el\-tu\-ngawng}}, brainworm.

% eltungawng
\hi
\HT{eltungawng}{\B{\cp{el}tungawng}} \I{["{}El.tu.NawN]} \emph{n., fauna}~(\emph{pl.} \B{mel\-tu\-ngawng}, \B{pxel\-tu\-ngawng}, \B{ay\-el\-tu\-ngawng}) brainworm. 
(\ding{43}~\HL{eltu}{\B{el\-tu}}, \HL{ngawng}{\B{ngawng}})

% emkä
\hi
\HT{emkA}{\B{em\cp{kä}}} \I{[Em."k$\cdot$\ae]} \emph{vtr.}\on{22} cross (s.t.). 
\B{Rä\-'ä fmi\-vi li\-vok fu em\-ki\-vä ay\-ekxan\-it a fkol ngo\-lop fpi sì\-kxuke ay\-frr\-tuä sì ay\-ioang\-ä.} Do not attempt to approach or cross any barriers designed for Guest and animal safety.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b}
(\ding{43}~\HL{kA}{\B{kä}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{semkA}{\B{sem\-kä}}, bridge.  
\on{2}.~\HL{emkAfya}{\B{em\-kä\-fya}}, ford, crossing. 

% emkäfya
\hi 
\HT{emkAfya}{\B{em\cp{kä}fya}} \I{[Em."k\ae.fja]} \emph{n.} ford, crossing. 
\B{Fì\-tseng pay\-fya vi\-rä ka ngip a\-reng, ha tsun aw\-nga tsat si\-var sko em\-kä\-fya.} Here the stream spreads over a shallow area, so we can use it as a ford.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{emkA}{\B{em\-kä}}, \HL{fya'o}{\B{fya\-'o}}; \HL{semkA}{\B{sem\-kä}})

% emrey
\hi
\HT{emrey}{\B{em\cp{rey}}} \I{[Em."R$\cdot$Ej]} \emph{vin., inf.}\on{22} survive (a life-threatening situation). 
\B{Tsawl\-a pa\-lu\-lu\-kan\-ìl oe\-ti 'ìl\-me\-ko a krr, Nawm\-a Sa'\-nok lrr\-tok sei\-yi ul\-te oe em\-ro\-ley.} The Great Mother was looking after me when the big thanator attacked and I've survived.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{rey}{\B{rey}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lemrey}{\B{lem\-rey}}, surviving. 
\on{2}.~\HL{temrey}{\B{tem\-rey}}, survival. 
\on{3}.~\HL{nemrey}{\B{nem\-rey}}, as if one's life were at stake.

% emza'u
\hi
\HT{emza'u}{\B{em\cp{za}'u}} \I{[Em."z$\cdot$a.P$\cdot$u]} \emph{vtr., inf.}\on{23} pass (a test). \B{Sì\-fme\-tok\-it em\-zo\-la\-'u ohel.} I have passed the tests.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{wiki} 
\B{Nän ftia, nän lu skxom a em\-za\-'u.} The less you study, the less chance you have of passing.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{za'u}{\B{za\-'u}})

% Entu
\hi
\B{\cp{En}tu} \I{["{}En.tu]} \emph{n. male name}. 
\B{Ma En\-tu, fì\-kem rä\-'ä si\-vi; lu ke\-fpom\-ro\-nga'.} Entu, don't do this; it's not healthy (mentally).~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{En\-tul kan\-'ìn tì\-wu\-sem\-it.} Entu specializes in fighting.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}

% ‹eng›
\hi
‹\HT{eng}{\B{eng}}›~:~‹\HL{Ang}{\B{äng}}›

% eo
\hi
\HT{eo}{\B{\cp{e}o}} \I{["{}E.o]} \emph{adp.} before, in front of.
\B{Eo ay\-oeng lu txan\-a tì\-kawng.} A great evil is upon us.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{wiki} 
\B{Eo Ey\-wa oe \mbox{'ia}.} I lose myself before Eywa.~\LINK{http://forum.learnnavi.org/language-updates/sieyng-a-ftu-naring-3-proverbs-what-we-asked-for-and-some-vocabulary/msg327513/\#msg327513}{a\slash ln} 
\B{Ut\-ral a lu few pay\-fya a eo kel\-ku oeyä tsawl lu nì\-txan.} The tree on the other side of the stream in front of my house is very tall.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
(\emph{a. idiom}) \B{tokx eo tokx}, face to face, in person.~\LINK{http://forum.learnnavi.org/language-updates/fmawno/msg293595/\#msg293595}{ln} 
(\emph{b. idiom}) \B{na fkxen eo fkio}, to go to waste [lit.: `like vegetable food before a tetrapteron'] 
\B{Nge\-yä fì\-tì\-kang\-kem\-vi a\-txan\-tsan ke sla\-yu na fkxen eo fkio.} This excellent work of yours will not go to waste.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{214}~\HL{uo}{\B{uo}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{eolI'uvi}{\B{eo\-lì\-'u\-vi}}, prefix.

% eolì'uvi
\hi 
\HT{eolI'uvi}{\B{\cp{e}olì'uvi}} \I{["{}E.o.lI.Pu.vi]} \emph{n.} prefix. 
\B{Lu tsa\-lì\-'ur alu fì\-haw\-re'\-ti me\-lì\-'u\-vi alu 'aw\-a eo\-lì\-'u\-vi sì 'aw\-a uo\-lì\-'u\-vi.} The word \emph{fì\-haw\-re'\-ti} has two affixes--one prefix and one suffix.~\LINK{http://naviteri.org/2017/10/more-language-for-talking-about-language/}{b}
(\ding{43}~\HL{eo}{\B{eo}}, \HL{lI'uvi}{\B{lì\-'uvi}})

% ep'ang
\hi
\HT{ep'ang}{\B{ep\cp{'ang}}} \I{[Ep\textcorner."PaN]} \emph{adj.} complex. \B{Ul\-te ke stì\-yawm ay\-ngal fì\-trr kea ho\-ren\-it a\-ep\-'ang.} And you will not hear any complex rules today.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} 
(\ding{214}~\HL{fyin}{\B{fyin}})\\
\textsc{Derivatives}: \ding{43} 
\on{1}. \HL{fyinep'ang}{\B{fyi\-nep\-'ang}}, complexity.

% epxang
\hi
\HT{epxang}{\B{e\cp{pxang}}} \I{[E."p'aN]} \emph{n.}~(\emph{pl.} \B{me\-pxang}, \B{pxe\-pxang}, \B{ay\-e\-pxang}) stone jar (\emph{to hold small toxic arachnid}).
\B{Epxang\-mì lu 'u\-pe?--Tì\-lo\-ho.} What's in the stone jar?--It's a surprise.~\LINK{http://http://naviteri.org/2019/06/50a-liu-amip-40-new-words}{b}

% ‹er›
\hi
‹\HT{er}{\B{er}}› \I{[E.R]} \emph{inf. in} ‹1› (to form imperfective aspect, on-going action). \B{Ney\-ti\-ri he\-ra\-haw.} Neytiri is sleeping.~\LINK{http://wiki.learnnavi.org/Corpus\#MSNBC}{wiki} \on{1}. (required with \B{tengkrr}) \B{aw\-ngal yom wu\-tsot teng\-krr he\-reyn nì\-pxim}, we eat dinner while sitting upright.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} \on{2}. (can combine with tense infixes to form a frame narrative) \B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok, tso\-le'a syep\-tu\-tet a\-tsawl fra\-to mì sì\-rey.} Yesterday I was with Txewì in the forest and we saw the biggest Trapper I've ever seen.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki} 
(\ding{43}~‹\HL{arm}{\B{arm}}›, ‹\HL{ary}{\B{ary}}›, ‹\HL{Irm}{\B{ìrm}}›, ‹\HL{Iry}{\B{ìry}}›; \ding{214}~‹\HL{ol}{\B{ol}}›)

% etrìp
\hi
\HT{etrIp}{\B{\cp{et}rìp}} \I{["{}Et\textcorner.RIp\textcorner]} \emph{adj.} favorable, auspicious. 
(\emph{a}) (idiom) 
\B{Et\-rìp\-a syay\-vi!} Good luck!~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\ding{214}~\HL{ketrIp}{\B{ket\-rìp}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{netrIp}{\B{net\-rìp}}, luckily, happily.
\on{2}.~\HL{ketrIp}{\B{ket\-rìp}}, unfavorable, inauspicious.

% Europa
\hi
\B{Eu\cp{ro}pa} \I{[E.u."Ro.pa]} \emph{n.}, \textsc{lw} Europe. 

% eyawr
\hi
\HT{eyawr}{\B{e\cp{yawr}}} \I{[E."jawR]} \emph{adj.} correct, right. 
\B{Sä\-pll\-txe\-vi\-ri a\-la\-he, lu lì'\-fya\-vi nge\-yä e\-yawr, lu pum oe\-yä e\-yawr nì\-teng.} Concerning the other comment, your expression is correct, mine is correct as well.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1903}{b} 
\B{Slä fma\-syi ri\-vun e\-yawr\-a tì\-'eyng\-it.} But I'll try to find the correct answer.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-920}{b} 
(\ding{214}~\HL{keyawr}{\B{ke\-yawr}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIyawr}{\B{tì\-yawr}}, correctness. 
\on{2}.~\HL{nIyawr}{\B{nì\-yawr}}, correctly. 
\on{3}.~\HL{keyawr}{\B{ke\-yawr}}, incorrect. 
\on{4}.~\HL{eyawrfya}{\B{e\-yawr\-fya}}, righ way, correct path.

% eyawrfya
\hi 
\HT{eyawrfya}{\B{e\cp{yawr}fya}} \I{[E."jawR.fja]} \emph{n.} right way (of doing s.t.), correct path. 
\B{Ney\-ti\-ril Tsyeyk\-ur ko\-lar eyawr\-fyat a fyep tsko\-ti.} Neytiri taught Jake the right way to hold a bow.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
\B{Eyawr\-fya\-ri ze\-ne tsli\-vam fya\-'ot a mìn ki\-fkey.} To know the right way, you have to understand how the world turns.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{eyawr}{\B{eyawr}}, \HL{fya'o}{\B{fya\-'o}})

% eyaye
\hi
\B{e\cp{ya}ye} \I{[E."ja.jE]} \emph{n.} \ding{96}~(\emph{pl.} \B{me\-ya\-ye}, \B{pxe\-ya\-ye}, \B{ay\-e\-ya\-ye}) warbonnet, \emph{bellicum pennatum}.  

\end{multicols}