\pdfbookmark[0]{EY}{EY}
\begin{center}
\section*{EY \I{[Ej]}}
\end{center}

\begin{multicols}{2}

% eyk
\hi
\HT{eyk}{\B{eyk}} \I{[Ejk\textcorner]} \emph{vtr.} lead. \B{Tsam\-po\-ngut Tsu'\-tey\-ìl i\-veyk.} Tsu'tey will lead the war party.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Preparing_the_war_party}{a\slash wiki} \B{Eyk Ka\-mun a fra\-lo lä\-ngu tsa\-sä\-ta\-ron vel\-ke nì\-wotx.} Every time Kamun is in charge, the hunt is a mess.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{eyktan}{\B{eyktan}}, leader. 
\on{2}.~\HL{eykyu}{\B{eykyu}}, leader (of a small group).

% ‹eyk›
\hi
‹\HT{eyk2}{\B{eyk}}› \I{[Ej.k]} \emph{inf. in} pre-‹1› (\emph{a. to make a causative statement, to make s.o. do s.t.}) 
\B{Ey\-tu\-kan\-ìl Ney\-ti\-rir ye\-rik\-it tey\-ko\-la\-ron.} Eytukan made Neytiri hunt a hexapede.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#More_extracts_from_various_emails}{wiki} 
(\emph{b. alternatively with} \HL{fa}{\B{fa}} \emph{to introduce the caused person}) \B{Ey\-tu\-kan\-ìl fa Ney\-ti\-ri ye\-ri\-kit tey\-ko\-la\-ron.} Eytukan had a hexapede hunted by Neytiri.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
(\emph{c. to make a \emph{vin.} into a \emph{vtr.}}) \B{A\-pxa tì\-väng\-ìl po\-ti stey\-ko\-li.} (His) great thirst made him angry [lit.: `great thirst caused him to be angry'].~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
\B{oe\-ri U\-nil\-tì\-ran\-tokx\-ìl tì\-rey\-ti ley\-ko\-la\-te\-i\-em nì\-txan}, \emph{Avatar} has greatly changed my life for the better.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/msg350756/\#msg350756}{ln} 
\B{Po\-el tì\-kang\-kem\-it sngey\-ko\-lä\-'i.} She began the work.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{d. with vtr. modals}) \B{Fì\-pam\-tseo\-l oe\-ru ney\-kew fu\-ta srew.} This music makes me want to dance.~\LINK{https://naviteri.org/2020/12/mrra-tipangkxotsyip-five-little-discussions/}{b}

% eykil
\hi
\HT{eykil}{\B{ey\cp{kil}}} \I{[Ej."k$\cdot$il]} \emph{vtr., inf.}\on{22} bend s.t., pulling things together.
\B{Me\-tew\-it fì\-swek\-ä ey\-ki\-vil.} Pull the two ends of this bar together.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\ding{43}~\HL{il}{\B{il}})

% eyktan
\hi
\HT{eyktan}{\B{\cp{eyk}tan}} \I{["{}Ejk\textcorner.tan]} \emph{n.}~(\emph{pl.} \B{meyk\-tan}, \B{pxeyk\-tan}, \B{ay\-eyk\-tan}) leader. \B{Na'viru set lu nawma eyktan amip a larmu Tawtute.} The Na'vi have a great leader now who was a Skyperson.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \B{Nì\-lun ay\-ioi a\-'eoio ay\-eyk\-tan\-ä lu lor fra\-to.} Of course the ceremonial wardrobes of the leaders are the most beautiful.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
(\ding{43}~\HL{eyk}{\B{eyk}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIeyktan}{\B{tìeyktan}}, leadership. 
\on{2}.~\HL{eyktanay}{\B{eyktanay}}, deputy, general.

% eyktanay
\hi
\HT{eyktanay}{\B{eykta\cp{nay}}} \I{[Ejk\textcorner.ta."naj]} \emph{n.}~(\emph{pl.} \B{meyk\-ta\-nay}, \B{pxeyk\-ta\-nay}, \B{ay\-eyk\-ta\-nay}) deputy, general (s.o.~who is one step down in rank from leader). 
\B{Ul\-te ftxo\-ley pol ay\-eyk\-ta\-nay\-ti a lu stum nì\-ftxan kawng na po.} And he has chosen deputies who are almost as bad as he is.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{eyktan}{\B{eyk\-tan}})

% eykyu
\hi
\HT{eykyu}{\B{\cp{eyk}yu}} \I{["{}Ejk\textcorner.ju]} \emph{n.}~(\emph{pl.} \B{meyk\-yu}, \B{pxeyk\-yu}, \B{ay\-eyk\-yu}) leader (typically of a small group). 
\B{Vll eyk\-yu ne\-fä fte po\-ngu fä\-ki\-vä.} The leader signals the party to ascend by pointing upwards.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
\B{Mì sa\-ngek a sä\-vll\-it ngo\-lop eyk\-yul tar\-po\-ngu\-ä.}
The sign on the tree trunk was made by the leader of the hunting party.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
(\ding{43}~\HL{eyk}{\B{eyk}})

% Eytukan
\hi
\B{\cp{Ey}tukan} \I{["{}Ej.tu.kan]} \emph{n. male name}. 
\B{Pol\-txe Ey\-tu\-kan san oe ka\-yä sìk, slä oel pot ke spaw.} Eytukan said he would come but I don't believe him.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175491/\#msg175491}{ln}
\B{Spaw oel fu\-ta Mo\-'at\-ìl tso\-le\-'a Ney\-ti\-rit. -- Ke\-he. Tso\-le\-'a Ney\-ti\-rit Ey\-tu\-kan\-ìl.} I believe Mo\-'at saw Ney\-ti\-ri. -- No. It's Ey\-tu\-kan who saw Ney\-ti\-ri.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/}{b}
\B{New Va'\-ru tskot Ey\-tu\-kan\-ä za\-sri\-vìn.} Va'ru wants to borrow Eytukan's bow.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}

% Eywa
\hi
\HT{eywa}{\B{\cp{Ey}wa}} \I{["{}Ej.wa]} \emph{n.} Eywa, `the Great Mother' (the deity of the Na'vi who protects the balance of life). 
\B{Ey\-wal zey\-ki\-vo ngat nì\-win.} May Eywa heal you quickly.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Spaw Na'\-vil fu\-ta tì\-yawn Ey\-wa\-yä lu txew\-lu\-ke.} The Na'vi believe that Eywa's love is boundless.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\emph{a. idiom}) \B{Ey\-wa nga\-hu} Goodbye! [lit.: Eywa (be) with you!]~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ey\-wa va\-yun.} Ey\-wa will provide.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}. \HL{eywa'eveng}{\B{Eywa'eveng}}, (the moon) Pandora.

% Eywa'eveng
\hi
\HT{eywa'eveng}{\B{Eywa\cp{'e}veng}} \I{[""Ej.wa."PE.vEN]} \emph{n.} (the moon) Pandora [lit.: child of Eywa]. \B{Pol sla'\-tsu ay\-i\-o\-ang\-it a tse\-'a fkol mì Ey\-wa\-'e\-veng.} She describes animals seen on Pandora.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
(\ding{43}~\HL{eywa}{\B{Ey\-wa}}, \HL{'eveng}{\B{'e\-veng}}, \HL{eyweveng}{\B{Ey\-we\-veng}})

% Eyweveng
\hi
\HT{eyweveng}{\B{Ey\cp{we}veng}} \I{[Ej."wE.vEN]} \emph{n.}~(\emph{short form of}) \ding{43}~\HL{eywa'eveng}{\B{Ey\-wa\-'e\-veng}}. 
\B{Oel ron\-srel\-ngop fu\-ta Ey\-we\-veng\-it tok.} I imagine that I'm on Pandora.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
\B{Txon\-krr lu syu\-ra\-tan na'\-rìng\-ä Ey\-we\-veng\-ä lor nì\-txan.} At night, the bioluminescence of the Pandoran forest is very beautiful.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
(\ding{43}~\HL{eywa}{\B{Ey\-wa}}, \HL{'eveng}{\B{'e\-veng}})

\end{multicols}