\pdfbookmark[0]{F}{F}
\begin{center}
\section*{F \I{[f]} (\emph{Fä})}
\end{center}

\begin{multicols}{2}

% fa
\hi
\HT{fa}{\B{fa}} \I{[fa]} \emph{adp.} with, by means of. 
\B{Ay\-lì\-'u\-fa aw\-nge\-yä 'ey\-lan\-ä a\-'e\-wan.} In the words of our young friend.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{O\-lo\-me\-i\-um oel fa key\-rel fu\-ta lu yaw\-ne oe po\-ru.} I knew by the expression on her face that she loves me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Tsun fko tsa\-tse\-nge\-ne ki\-vä nì\-sle\-le fu fa fwa ik\-ran\-it mak\-to nì\-'aw.} You can only get there by swimming or riding an ikran.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
(\emph{a. with certain verbs}) \ding{43}~\HL{ioi}{\B{ioi s(äp)i fa}}, adorn (oneself) with; \HL{pamtseo}{\B{pam\-tseo si fa}}, play an instrument; \HL{rawn}{\B{rawn fa}}, replace s.t.~with s.t.~else; \HL{tswayon}{\B{tsway\-on fa}}, fly with s.t.; \HL{pxek}{\B{pxek ki\-nam\-fa\slash pxun\-fa}}, kick, shove. \HL{rumtsyokx}{\B{ta\-kuk fa rum\-tsyokx}}, punch s.o.; \HL{rIktsyokx}{\B{ta\-kuk fa rìk\-tsyokx}}, slap s.o.
(\emph{b. alternative to inroduce the caused person in causative sentences}) \B{Ey\-tu\-kan\-ìl fa Ney\-ti\-ri ye\-ri\-kit tey\-ko\-la\-ron.} Eytukan had a hexapede hunted by Neytiri.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
(\emph{c. proverb}) \B{Tì\-tok slu tìk\-tok fa pam\-tsyìp a\-'aw.} Presence becomes absence by means of one little sound. [i.e. `Small things can make a big difference.']~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}

% fay+
\hi
\HT{fay}{\B{fay}}\texttt{+} \I{[faj.]} \emph{prod.~pref.~on nouns} (\emph{short form of} \HL{fI}{\B{fì}}\texttt{-} \emph{and} \HL{ay}{\B{ay}}\texttt{+}) these (\dots~here) (\emph{proximal plural demonstrative prefix}). 
\B{fay\-vrr\-tep fì\-tse\-nge lu kxa\-nì}, these demons are forbidden here.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki} 
\B{Sra\-ke tsun oe fay\-u\-pxa\-ret tsl\-ivam nì\-ftue?} Can I understand these messages easily?~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln}
(\ding{43} \HL{tsay}{\B{tsay}\texttt{+}})

% fayluta
\hi
\HT{fayluta}{\B{fay\cp{lu}ta}} \I{[faj."lu.ta]} \emph{conj.} (\emph{short form of} \B{fay\-lì\-'ut a}) these words that\dots (to form reported speech) \B{Pol\-txe pol fay\-lu\-ta oe new ki\-vä.} She said, `I want to go'. \emph{or} She said she wanted to go.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} 

% fäkä
\hi
\HT{fAkA}{\B{fä\cp{kä}}} \I{[f\ae."k$\cdot$\ae]} \emph{vin., inf.}\on{22} go up, ascend. \B{Vll eyk\-yu ne\-fä fte po\-ngu fä\-ki\-vä.} The leader signals the party to ascend by pointing upwards.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{fApa}{\B{fä\-pa}}, \HL{kA}{\B{kä}}; \ding{214}~\HL{kllkA}{\B{kll\-kä}})

% fäpa
\hi
\HT{fApa}{\B{\cp{fä}pa}} \I{["f\ae.pa]} \emph{n.} top. \B{ro fäpa}, at the top. \LINK{http://forum.learnnavi.org/language-updates/all-adpositions/msg137792/\#msg137792}{ln}
(\ding{43}~\HL{pa'o}{\B{pa\-'o}}; \ding{214}~\HL{kllpa}{\B{kll\-pa}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fApyo}{\B{fäpyo}}, roof.

% fäpyo
\hi 
\HT{fApyo}{\B{\cp{fäp}yo}} \I{["f\ae{}p\textcorner.jo]} \emph{n.} roof. 
\B{Tom\-pa ze\-rup ul\-te pay lipx kxam\-lä fäp\-yo.} It's raining and water is dripping through the roof.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b} 
(\ding{43}~\HL{fApa}{\B{fäpa}}, \HL{yo}{\B{yo}})

% fäza'u
\hi
\HT{fAza'u}{\B{fä\cp{za}'u}} \I{[f\ae."z$\cdot$a.P$\cdot$u]} \emph{vin., inf.}\on{23} come up, ascend. (\emph{a.})  \B{Fä\-zi\-va\-'u ne tse\-nge a oel tok!} Come up to where I am!~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
(\emph{b. of astronomical bodies rising}) \B{Fä\-za'u tsaw\-ke krr\-pe?} When will the sun come up?~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
(\ding{43}~\HL{fApa}{\B{fä\-pa}}, \HL{za'u}{\B{za\-'u}}; \HL{fpxAkIm}{\B{fpxä\-kìm}}; \ding{214}~\HL{kllkA}{\B{kll\-kä}}, \HL{kllza'u}{\B{kll\-za\-'u}})

% fahew
\hi
\HT{fahew}{\B{fa\cp{hew}}} \I{[fa."hEw]} \emph{n.} (\emph{a. sensation}) smell. 
\B{Oe\-ri ta pe\-yä fa\-hew a\-ke\-wong on\-tu te\-ya lä\-ngu.} My nose is full of his alien smell.~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll\slash a} 
\B{Fì\-syu\-lang\-it syam. Fa\-hew lor lu nì\-txan, ke\-fyak?} Smell this flower. Its fragrance is beautiful, isn't it?~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
(\emph{b. to form the middle voice}) \B{Ra\-lu\-ri fa\-hew fkan oe\-ru na ye\-rik.} Ralu smells like a hexapede to me.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}

% faoi
\hi
\HT{faoi}{\B{\cp{fa}oi}} \I{["fa.o.i]} \emph{adj.} smooth. 
\B{Ta'\-leng prr\-nen\-ä lu faoi, pum ko\-ak\-tu\-ä ekx\-txu.} A baby's skin is smooth, an old person's is rough.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
(\ding{214}~\HL{ekxtxu}{\B{ekx\-txu}})

% fe'
\hi
\HT{fe'}{\B{fe'}} \I{[fEP]} \emph{adj., nfp.} bad. (\emph{a.}) \B{Peyä tsa\-tì\-pe\-'un a swey\-lu txo wi\-vem ayoeng Oma\-ti\-ka\-ya\-wä lu fe'.} His decision to fight against the Omaticaya was a bad one.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} \B{tì\-len afe'}, bad event, something bad that happened.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
(\emph{b.~proverb}) \B{Kem a\-muiä, kum a\-fe'.} Proper action, bad result. (\emph{something that should have turned out well didn't}).~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
(\ding{214}~\HL{sIltsan}{\B{sìl\-tsan}}) \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{nIfe'}{\B{nì\-fe'}}, badly, in a bad way. 
\on{2}.~\HL{fe'ul}{\B{fe\-'ul}}, worsen, get worse.
\on{3}.~\HL{fekem}{\B{fe\-kem}}, accident.
\on{4}.~\HL{fe'ran}{\B{fe'\-ran}}, flawed nature.
\on{5}.~\HL{fe'lup}{\B{fe'\-lup}}, tacky, in poor taste. 
\on{6}.~\HL{fekum}{\B{fe\-kum}}, disadvantage, drawback, downside.
\on{7}.~\HL{fe'pey}{\B{fe'\-pey}}, dread, expect something bad to happen.

% fe'lup
\hi
\HT{fe'lup}{\B{\cp{fe'}lup}} \I{["fEP.lup\textcorner]} \emph{or} \I{[.lUp\textcorner]} (\textsc{rn}: \B{\cp{fe'}lùp} \I{["fEP.lUp\textcorner]}) \emph{adj.} tacky, in poor taste.
(\ding{43}~\HL{fe'}{\B{fe'}}, \HL{lupra}{\B{lup\-ra}})

% fe'pey
\hi 
\HT{fe'pey}{\B{fe'\cp{pey}}} \I{[fE."p$\cdot$Ej]} \emph{vin., inf.}\on{22} dread, expect something bad to happen. (\emph{used with} \ding{43} \HL{tsnI}{\B{tsnì}})
\B{Krr\-a pä\-hem Saw\-tu\-te, pxay\-a Na'\-vi fe'\-par\-mey.} When the Sky People arrived, many Na'vi felt dread.
\B{Po fe'\-po\-ley tsnì 'i\-tan sne\-yä tì\-fme\-tok\-it ke em\-zì\-ye\-va\-'u.} He feared his son might not pass the test.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
(\ding{43}~\HL{fe'}{\B{fe'}}, \HL{pey}{\B{pey}}; \ding{214}~\HL{sIlpey}{\B{sìl\-pey}})

% fe'ran
\hi
\HT{fe'ran}{\B{\cp{fe'}ran}} \I{["fEP.Ran]} \emph{n.} flawed nature, something ill-conceived or inherently defective. \B{Fì\-tì\-hawl\-ì\-ri fe'\-ran law lä\-ngu fra\-por.} Unfortunately the flawed nature of this plan is obvious to everyone.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\ding{43}~\HL{fe'}{\B{fe'}}, \HL{ran}{\B{ran}}; \ding{214}~\HL{loran}{\B{lo\-ran}}) \\
\textsc{Derivations:} \ding{43} \on{1}. \HL{fe'ranvi}{\B{fe'ranvi}}, blemish, flawed feature.

% fe'ranvi
\hi
\HT{}{\B{\cp{fe'}ranvi}} \I{["fEP.Ran.vi]} \emph{n.} blemish, deformity, stain, flawed feature.  \B{Hu\-fwa lu fil\-ur Va'\-ru\-ä fne\-fe'\-ran\-vi, tsal\-su\-ngay fpìl fu\-ta say\-rìp lu nì\-txan.} Although Va'ru's facial stripes are rather uneven, I still think he's very handsome.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{fe'ran}{\B{fe'\-ran}})

% fe'ul
\hi
\HT{fe'ul}{\B{\cp{fe}'ul}} \I{["fE.P$\cdot$ul]} \emph{vin., inf.}\on{22} worsen, get worse. \B{Ke tsun oe tsli\-vam teyng\-ta tì\-ru\-sol peyä lum\-pe fe\-'ul krra oe tìng mik\-yun.} I can't understand why her singing gets worse when I listen.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Nge\-yä tsay\-lì\-'ul tì\-fkey\-tok\-it fe\-'ey\-ko\-lul nì\-'aw.} Those words of yours have only made the situation worse.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b} 
(\ding{214}~\HL{tsan'ul}{\B{tsan\-'ul}}, \ding{43}~\HL{fe'}{\B{fe'}}, \HL{'ul}{\B{'ul}})  \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIfe'ul}{\B{tìfe'ul}}, worsening (abstract) 
\on{2}.~\HL{sAfe'ul}{\B{säfe'ul}}, worsening (specific)

% few
\hi
\HT{few}{\B{few}} \I{[fEw]} \emph{adp.} across, aiming for the other side. \B{Po spä few pay\-fya fte smar\-it si\-vutx.} He jumped across the stream to track his prey.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fewtusok}{\B{fewtusok}}, opposite (side).

% fewtusok
\hi
\HT{fewtusok}{\B{\cp{few}tusok}} \I{["fEw.tu.sok\textcorner]} \emph{or coll.} \I{["fEw.\t{ts}ok\textcorner]} (\emph{in rapid speech}) \emph{adj.} opposite, on the opposite side. 
\B{Oe kaw\-krr ne few\-tu\-sok\-a pa\-'o kil\-van\-ä ke ka\-mä.} I never went to the opposite side of the river.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{few}{\B{few}}, \HL{tok}{\B{tok}})

% fekem
\hi
\HT{fekem}{\B{\cp{fe}kem}} \I{["fE.kEm]} \emph{n.} accident. \B{Na\-ri si fte kea fe\-kem ke li\-ven ngar!} Be careful you don't have an accident!~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{fe'}{\B{fe'}}, \HL{kem}{\B{kem}})

% fekum
\hi
\HT{fekum}{\B{\cp{fe}kum}} \I{["fE.kum]} \emph{or} \I{[.kUm]} \emph{n.} disadvantage, drawback, downside. 
\B{Tì\-ta\-ron\-ìri lä\-ngu tì\-kak\-pam fe\-kum. Kin fkol fra\-inan\-fyat.} I'm sorry to say that deafness is a disadvantage for hunting. You need all your senses.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
(\ding{43}~\HL{fe'}{\B{fe'}}, \HL{kum}{\B{kum}}, \ding{214}~\HL{tsankum}{\B{tsan\-kum}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fekumnga'}{\B{fekumnga'}}, disadvantageous.

% fekumnga'
\hi
\HT{fekumnga'}{\B{\cp{fe}kumnga'}} \I{["fE.kum.NaP]} \emph{or} \I{[.kUm.]} \emph{adj.} disadvantageous. 
(\ding{43}~\HL{fekum}{\B{fe\-kum}}, \ding{214}~\HL{tsankumnga'}{\B{tsan\-kum\-nga'}})

% fewi
\hi
\HT{fewi}{\B{\cp{fe}wi}} \I{["fE.wi]} \emph{vtr.} chase. \B{Rum\-it a no\-lu\-i rä\-'ä fe\-wi.} Don't chase after a foul ball.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% feyä
\hi
\B{feyä}~:~\HL{ayfo}{\B{ayfo}}

% fil
\hi 
\HT{fil}{\B{fil}}\textsuperscript{\on{1}} \I{[fil]} \emph{n.} child's toy, plaything.

% 
\hi 
\B{fil}\textsuperscript{\on{2}} : \HL{pil}{\B{pil}}

% fì-
\hi
\HT{fI}{\B{fì-}} \I{[fI.]} \emph{prod. pref. on nouns} (\emph{a.}) this (here) (\emph{proximal singular demonstrative prefix}). \B{Fì\-lì\-'u\-ä ral lu tì\-me\-'em.} This word means harmony.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/ }{ln} 
(\emph{b. with words describing time}) present, this \B{fì\-trr}, today; \B{fì\-kaym}, this evening; \B{fì\-txon}, tonight; \B{fì\-kin\-trr}, this week; etc. 
(\emph{c. with words describing time and} \B{-o}) duration \B{fì\-trr\-o}, all day (long). 
(\ding{43}~\HL{tsa}{\B{tsa-}}, \HL{fay}{\B{fay}}\texttt{+})

% fì'u
\hi
\HT{fI'u}{\B{fì\cp{'u}}} \I{[fI."Pu]} \emph{pn., n.} this (thing). 
\B{fay\-sä\-num\-vi\-ri ru\-txe fì\-'ut tsli\-vam}, concerning these lessons, please understand this.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
\B{Fì\-'u\-ti ni\-vìn, ma smuk.} Look at this, siblings.~\LINK{http://naviteri.org/2013/03/tsamsiyu-a-mi-saw-rrtaya-warrior-in-our-sky/}{b} 
(\emph{a. to form contrastive demonstratives}, !stress change!) \B{Fì\-fkxen alu \cp{fì}\-'u lu ftxì\-lor; tsa\-fkxen alu \cp{tsa}\-'u nga\-ti tspang.} \emph{This} vegetable is delicious; \emph{that} one will kill you.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
(\ding{43}~\HL{'u}{\B{'u}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fwa}{\B{fwa}}, that (\emph{subjective}). 
\on{2}.~\HL{fula}{\B{fula}}, that (\emph{agentive}). 
\on{3}.~\HL{futa}{\B{futa}}, that (\emph{patientive}). 
\on{4}.~\HL{furia}{\B{furia}}, that (\emph{topic}). 

% fìfya 
\hi
\HT{fIfya}{\B{fì\cp{fya}}} \I{[fI."fja]} \emph{adv.} this way, like this. 
\B{Oel nga\-ti stum ke stä\-ngawm krra nga fì\-fya tse\-rì\-syì.} I can barely hear you when you're whispering like this.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\emph{a. idiom}) \B{fì\-fya tsa\-fya}, one way or another. 
\B{Slä fì\-fya tsa\-fya, nga\-ri ti\-re\-al tsa\-tse\-nget ta\-yok.} But one way or the other your spirit will be there.~\LINK{http://naviteri.org/2013/04/sin-asok-sin-zusawkrra-recent-and-upcoming-activities/comment-page-1/\#comment-2316}{b}
(\ding{43}~\HL{fya'o}{\B{fya\-'o}})

% fìkem
\hi
\HT{fIkem}{\B{fì\cp{kem}}} \I{[fI."kEm]} \emph{pn., n.} \textsc{i}.~this (action). 
\B{Ay\-nga\-ri fì\-kem fe\-yä 'e\-'al\-a to\-pur ka\-ngay sì\-yi nì\-'aw.} This will only confirm their worst fears about you.~\LINK{http://forum.learnnavi.org/language-updates/two-words-in-use-eal-and-kangay-si/msg564284/\#msg564284}{ln} \\
\textsc{ii}.~\HT{fIkem si}{\B{fìkem si}} \I{[fI."kEm si]} \emph{vin.} do this (action). \B{Swey lu fwa nga fì\-kem ke si\-vi nì\-sngum.} It's best that you not do this fretfully.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
(\ding{43}~\HL{kem}{\B{kem}})

% fìkintrr
\hi
\HT{fIkintrr}{\B{fì\cp{kin}trr}} \I{[fI."kin.t\s{r}]} \emph{adv.} this (present) week. 
(\ding{43}~\HL{kintrr}{\B{kin\-trr}})

% fìme+
\hi
\HT{fIme}{\B{fìme}}\texttt{+} \I{[fI.mE]} \emph{prod. pref. on n.s} these two (here) (\emph{proximal singular demonstrative prefix}).
\B{Tse\-nga 'aw\-steng\-yä\-pem fì\-me\-kem\-yo lu mek\-tseng a tsun fpxi\-vä\-kìm hì\-'ang tsaw\-ìlä.} Where these two walls come together there's a gap through which insects can get in.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% fìmuntrr
\hi
\HT{fImuntrr}{\B{fì\cp{mun}trr}} \I{[fI."mun.t\s{r}]}~\emph{or}~\I{[."mUn.]} \emph{adv.} this (present) weekend. 
\B{Sra\-ne, fì\-mun\-trr lo\-lu oer ma\-wey.} Yes, I had a quiet weekend.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/\#comment-33458}{b}
(\ding{43}~\HL{muntrr}{\B{mun\-trr}})

% fìpo
\hi
\HT{fIpo}{\B{fì\cp{po}}} \I{[fI."po]} (\emph{gen.}~\B{fìpeyä}) \emph{pn.} this one (person or thing). 
\B{Fì\-po\-ti oel tspì\-yang, fte tì\-ke\-nong li\-ye\-vu ay\-la\-ru.} I will kill this one as a lesson to the others.~\LINK{http://forum.learnnavi.org/language-updates/capturing-avatar/}{b} 
\B{Fì\-por syaw fko Ìstaw.} This is Ìstaw.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} (\emph{a. in contrastive demonstratives}, !stress change!) 
\B{Fì\-kar\-yu alu \cp{fì}\-po lu tsul\-fä\-tu; tsa\-kar\-yu alu \cp{tsa}\-po lu \mbox{skxawng}.} This teacher is a master; that teacher is a fool.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b} 
(\ding{43}~\HL{po}{\B{po}})

% fìpxe+
\hi
\HT{fIpxe}{\B{fìpxe}}\texttt{+} \I{[fI.p'E]} \emph{prod. pref. on n.s} these three (here) (\emph{proximal singular demonstrative prefix}).

% fìrewon
\hi
\HT{fIrewon}{\B{fì\cp{re}won}} \I{[fI."RE.won]} \emph{n.; adv.} this (present) morning. 
(\ding{43}~\HL{rewon}{\B{re\-won}})

% fìtrr
\hi
\HT{fItrr}{\B{fì\cp{trr}}} \I{[fI."t\s{r}]} \emph{n.; adv.} this (present) day, today. 
\B{fì\-trr\-mì le\-tsran\-ten}, on this important day.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Ul\-te fì\-trr\-ä ftxo\-zä\-ri, sìl\-pey oe, ay\-nga\-ru prr\-te' li\-vu.} And I hope you my have pleasure with today's clebration.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
(\ding{43}~\HL{trr}{\B{trr}})

% fìtxan
\hi
\HT{fItxan}{\B{fì\cp{txan}}} \I{[fI."t'an]} \emph{adv.} so, to such an extent. 
(\emph{a.}) \B{Po ä\-txä\-le so\-li nì\-ho\-na fì\-txan, ke tsun oe sti\-vo.} She asked so sweetly that I couldn't refuse.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
(\emph{b. in result clauses}) \B{Lu poe se\-vin fì\-txan kum\-a yaw\-ne slo\-lu oer.} She was so beautiful that I fell in love with her.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{txan}{\B{txan}})

% fìtxon
\hi
\HT{fItxon}{\B{fì\cp{txon}}} \I{[fI."t'on]} \emph{n., adv.} tonight, this (present) night. 
\B{Fì\-txon na ton a\-la\-he nì\-wotx pe\-lun ke lu teng?} Why is this night not the same like all other nights?~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
(\ding{43}~\HL{txon}{\B{txon}})

% fìtsap
\hi
\HT{fItsap}{\B{fì\cp{tsap}}} \I{[fI."\t{ts}ap\textcorner]} \emph{adv.} each other. (\emph{a. with vtr.}, ‹\B{äp}› \emph{is required}) 
\B{Me\-fo fì\-tsap mä\-po\-le\-yam teng\-krr tsngaw\-vìk.} The two of them hugged each other and wept.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
(\emph{b. with vin.}) \B{Tse\-nu sì Lo\-ak fì\-tsap ke ha' kaw\-'it.} Tsenu and Loak are a terrible match for each other. (\emph{neither one is to blame for the mismatch})~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Moe smon (moe\-ru) fì\-tsap.} We know each other.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b} 
\B{Me\-fo yaw\-ne lu (snor) fì\-tsap.} They love each other.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}

% fìtseng
\hi
\HT{fItseng}{\B{fì\cp{tseng}}} \I{[fI."\t{ts}EN]} \emph{or} \B{fì\cp{tse}nge} \I{[fI."\t{ts}E.NE]} \emph{n.; adv.} here, this place. 
\B{Fay\-vrr\-tep fì\-tse\-nge lu kxa\-nì.} These demons are forbidden here.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki\slash a} 
\B{Nga za\-'u fì\-tseng pxìm srak?} Do you come here often?~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{wiki\slash ld} 
\B{Nì\-'i'a tok oel nì\-mun fì\-tse\-nget!} Finally, I'm here again.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
\B{'En si oe, Saw\-tu\-te\-ol tìl\-mok fì\-tseng\-it.} I guess, some Skypeople were just here.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
(\ding{43}~\HL{tsenge}{\B{tse\-nge}})

% fìvospxì
\hi
\HT{fIvospxI}{\B{fìvo\cp{spxì}}} \I{[fI.vo."{}sp'I]} \emph{n.; adv.} this (present) month. 
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}})

% fkan
\hi
\HT{fkan}{\B{fkan}} \I{[fkan]} \emph{vin. resemble in a sensory modality, come to the senses as}, (used to form the `middle voice'). 
\B{Nik\-re\-ri Ri\-ni\-yä 'ur fkan lor.} Rini's hair looks beautiful.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Nik\-re\-ri Ri\-ni\-yä fkan lor.} Rini's hair is pleasant to the senses.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Ra\-lu\-ri fa\-hew fkan oe\-ru na ye\-rik.} Ralu smells like a hexapede to me.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Ra\-lu\-ri fkan na ye\-rik.} Ralu smells (looks\slash sounds) like a hexapede.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} \B{Fì\-na\-er\-ì\-ri sur fkan lor.} This drink tastes good.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Fì\-na\-er\-ì\-ri fkan ftxì\-lor.} This drink tastes good.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Syu\-ve\-ri ay\-ftxo\-zä\-yä ay\-nga\-ru fki\-van on\-lor ftxì\-lor\-sì nì\-wotx!} May the food of your holidays taste and smell good.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}

% fkarut
\hi
\HT{fkarut}{\B{\cp{fka}rut}} \I{["fka.Rut\textcorner]} \emph{or} \I{[.RUt\textcorner]} (\textsc{rn}: \B{\cp{fka}rùt} \I{["fka.RUt\textcorner]}) \emph{vtr.} peel. 

% fkay
\hi
\HT{fkay}{\B{fkay}} \I{[fkaj]} \emph{adj.} hateful. 

% fkew
\hi
\HT{fkew}{\B{fkew}} \I{[fkEw]} \emph{adj.} mighty. 
\B{na ay\-sa\-ngek a\-fkew}, like mighty trunks (of trees).~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}

% -fkeyk
\hi
\HT{-fkeyk}{\B{-fkeyk}} \I{[.fkEjk\textcorner]} \emph{prod. suf. on nouns to mean} situation, state, status. 
\B{Kil\-van\-fkeyk lu fya\-pe fì\-trr?} What's the condition of the river today?~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b} 
\B{Saw\-tu\-te\-ri ron\-sem\-fkeyk\-it ke tsun kaw\-tu tsli\-vam.} No one can understand the state of mind of the Sky People.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b} 
(\emph{a. conv. phrase}) \B{Nga\-fkeyk pe\-fya\slash fya\-pe?} `What’s your status?\slash How ya doin'?'~\LINK{http://naviteri.org/2018/02/negative-questions-in-navi/\#comment-28201}{b}
(\ding{43}~\HL{tIfkeytok}{\B{tì\-fkey\-tok}})

% fkeytok
\hi
\HT{fkeytok}{\B{\cp{fkey}tok}} \I{["fkEj.t$\cdot$ok\textcorner]} \emph{vin., inf.}\on{22} exist. \B{Ngal fwe\-rew a tu\-te ke fkey\-tok.} The person you're looking for doesn't exist.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b} 
(\ding{43}~\HL{kifkey}{\B{ki\-fkey}}, \HL{tok}{\B{tok}}) \\
\textsc{Derivations:} \ding{43} \on{1}. \HL{tIfkeytok}{\B{tìfkeytok}}, state, condition, situation.

% fkeyä
\hi
\B{\cp{fke}yä}~:~\HL{fko}{\B{fko}}

% fkio
\hi
\HT{fkio}{\B{\cp{fki}o}} \I{["fki.o]} \emph{n., fauna} tetrapteron. 

% fkip
\hi
\HT{fkip}{\B{fkip}} \I{[fkip\textcorner]} \emph{adp.} up among. 
(\emph{a. idiom}) \B{(na) fwam\-pop fkip fì\-wopx}, ``Fish out of water.'' [lit.: (like a) tapirus in the clouds].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{43}~\HL{kip}{\B{kip}}, \HL{takip}{\B{ta\-kip}}, \HL{tafkip}{\B{ta\-fkip}}) 

% fko
\hi
\HT{fko}{\B{fko}} \I{[fko]} (\emph{gen.}~\B{fkeyä}) \emph{pn.} (\emph{in the general sense}) one, you, they; (\emph{also used to convey passive voice}).  
\B{Oe\-ru syaw fko Ney\-ti\-ri.} My name is Neytiri [lit.: one calls (to) me Neytiri].~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln} 
\B{Ney\-ti\-ri\-ti fkol pä\-nu\-to\-lìng oe\-ru!} Neytiri was promised to me.~\LINK{http://forum.learnnavi.org/language-updates/panuting/}{ln\slash a\textsuperscript{c}} 
\B{Pa\-lu\-lu\-kan new fkot yi\-vom.} Thanator wants to eat you.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Hem nge\-yä zen\-ke fko\-ru li\-vu, Tsay\-hem a nga\-ru prr\-te' ke lu.} As for any action you don't like, don't do that action to others.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
(\emph{a. proverb}) \B{Ke kur fko fa kxe\-tse.} ``One can't hang by a tail,'' i.e. \emph{don't rely on s.t.\slash s.o. untrustworthy or useless, just as a Na'vi tail can't be relied on to bear weight}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% fkxake
\hi
\HT{fkxake}{\B{\cp{fkxa}ke}} \I{["fk'a.kE]} \emph{vin.} itch. 
\B{Sran, sran, oe\-ri pì\-tìk tsa\-tse\-nget a mì tal! Fkxa\-ke nì\-ftxan kum\-a ter\-kup.} Yeah, yeah, scratch that place on my back! It itches like crazy!~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{214}~\HL{pItIk}{\B{pì\-tìk}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fkxakewll}{\B{fkxakewll}}, itch plant.

% fkxakewll
\hi
\B{\cp{fkxa}kewll} \I{["fk'a.kE.w\s{l}]} \emph{n.}, \ding{96} itch plant, \emph{cynaroidia glauca}. 
(\ding{43}~\HL{fkxake}{\B{fkxa\-ke}}, \HL{'ewll}{\B{'e\-wll}})

% fkxara
\hi 
\HT{fkxara}{\B{\cp{fkxa}ra}} \I{["fk'a.Ra]} \emph{n.} stress (\emph{mental or emotional feeling}). 
\B{Krra oe ftxu\-lì\-'u, pxìm 'efu fkxa\-rat nì\-txan.} When I give a speech, I often feel a lot of stress.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fkxaranga'}{\B{fkxaranga'}}, stressful.

% fkxaranga'
\hi 
\HT{fkxaranga'}{\B{\cp{fkxa}ranga'}} \I{["fk'a.Ra.NaP]} \emph{adj.} stressful. 
\B{Nge\-yä fpom\-tokx\-ìri fì\-tì\-fkey\-tok a\-fkxa\-ra\-nga' lu le\-hrr\-ap.} This stressful situation is dangerous to your health.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
(\ding{43}~\HL{fkxara}{\B{fkxa\-ra}})

% fkxen
\hi
\HT{fkxen}{\B{fkxen}} \I{[fk'En]} \emph{n.}~\ding{96} food of vegetable (as opposed to animal) origin. 
\B{aw\-ngal yom fkxen\-ti le\-rìk nì\-wotx}, we eat leafy vegetable-food.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
\B{Fkxen\-ti pxìm yom fkol nì\-yrr.} Vegetables are often eaten raw.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}

% fkxile
\hi
\HT{fkxile}{\B{\cp{fkxi}le}} \I{["fk'i.lE]} \emph{n.} bib necklace. 
\B{Fkxi\-le pewn\-ta tu\-té\-yä kur.} The bib necklace hangs from the woman's neck.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Pll\-txe nì\-tsleng! Tsa\-fkxi\-let ke to\-lìng ngar En\-tul!}Liar! En\-tu didn't give you that necklace!~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}
\B{Poe\-ri sä\-fmong lor\-a tsa\-fkxi\-le\-yä lo\-lu na ay\-skxe mì te'\-lan.} For her, the theft of that beautiful necklace was like stones in her heart.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 

% flawkx
\hi
\HT{flawkx}{\B{flawkx}} \I{[flawk']} \emph{n.} leather.

% flawm
\hi
\HT{flawm}{\B{flawm}} \I{[flawm]} \emph{n.} cheek.

% flä
\hi
\HT{flA}{\B{flä}} \I{[fl\ae]} \emph{vin.} succeed. 
\B{Tì\-ste\-ftxaw nge\-yä flo\-lä nìl\-tsan!} Your examination succeeded well!~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{Tì\-tu\-sa\-ron\-ì\-ri fte fli\-vä, zene fko si\-vutx smar\-it ni\-nan nì\-no.} To succeed at hunting, you have to track your prey by reading (the forest) with attention to detail.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} 
(\emph{a. proverb}) \B{Flä ke flä, ley sä\-fmi.} Whether you succeed or not, the attempt has value.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b} 
(\ding{214}~\HL{nui}{\B{nui}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{sAflA}{\B{sä\-flä}}, (instance of) success.
\on{2}.~\HL{tIflA}{\B{tì\-flä}}, success (in general). 
\on{3}.~\HL{nIflA}{\B{nì\-flä}}, successfully.

% flew
\hi
\HT{flew}{\B{flew}} \I{[flEw]} \emph{n.} throat. 

% flefle
\hi
\HT{flefle}{\B{\cp{fle}fle}} \I{["flE.flE]} \emph{n.}, \ding{96} Sol's delight, magnetotrophic plant, \emph{calamariphyllum elegans}. 

% flel
\hi 
\HT{flel}{\B{flel}} \I{[flEl]} \emph{vtr.} trick s.o., fool s.o. 
\B{En\-tul Pey\-ral\-it fì\-txan flo\-lel ku\-ma fpìl poe san oe yaw\-ne lu por.} En\-tu fooled Pey\-ral so mucht that she thought he loved her.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b} \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{tIflel}{\B{tìflel}}, trickery (abstract).
\on{2}.~\HL{sAflel}{\B{sä\-flel}}, trick, hoax.
\on{3}.~\HL{fleltu}{\B{flel\-tu}}, fool, sucker.
\on{4}.~\HL{fleltrr}{\B{flel\-trr}}, April Fool's Day.

% fleltrr
\hi 
\HT{fleltrr}{\B{\cp{flel}trr}} \I{["flEl.t\s{r}]} \emph{n.} April Fool's Day. 
\B{Flel\-tu slu rä\-'ä! Fì\-trr lu Flel\-trr!} Don't be fooled! Today is April Fool's Day!.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}
(\ding{43}~\HL{flel}{\B{flel}}, \HL{trr}{\B{trr}}) 

% fleltu
\hi 
\HT{fleltu}{\B{\cp{flel}tu}} \I{["flEl.tu]} \emph{n.} fool, sucker, mark, s.o. easily tricked. 
\B{Flel\-tu slu rä\-'ä! Fì\-trr lu Flel\-trr!} Don't be fooled! Today is April Fool's Day!~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}
(\ding{43}~\HL{flel}{\B{flel}})

% flì
\hi
\HT{flI}{\B{flì}} \I{[flI]} \emph{adj., nfp.} thin. 
\B{Krro krro, flì\-a vul a\-ru\-sey to nutx\-a pum a\-ke\-ru\-sey lu txur.} A thin living branch is sometimes stronger than a thick dead one.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 
\B{Tskxe\-pay\-ri lu frir a\-flì sìn 'o\-ra.} There's a thin layer of ice on the lake.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
(\emph{a. proverb}) \B{Spä skxawng sìn 'a\-na a\-flì.} ``A fool jumps onto a thin vine,'' i.e. \emph{don't engage in an unpromising and\slash or potentially risky cause}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Tsa\-ye\-rik te\-rul ne 'awkx. Spä skxawng sìn 'a\-na a\-flì.} That hexapede is running toward the cliff. Only a fool jumps onto a thin vine.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{214}~\HL{nutx}{\B{nutx}}; \ding{43}~\HL{litsi}{\B{li\-tsi}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{flInutx}{\B{flìnutx}}, thickness.

% flìnutx
\hi
\HT{flInutx}{\B{flì\cp{nutx}}} \I{[flI."nut']} \emph{n.} thickness.   
\B{Sre fwa sìn tskxe\-pay tì\-ran, ze\-ne fko flì\-nutx\-it sti\-ve\-ftxaw.} It's necessary to check the thickness of the ice before walking on it.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\ding{43}~\HL{flI}{\B{flì}}, \HL{nutx}{\B{nutx}})

% flrr
\hi
\HT{flrr}{\B{flrr}} \I{[fl\s{r}]} \emph{adj.} gentle, mild, tender (\emph{for both people and things}). 
\B{Keng tsam\-si\-yu ze\-ne flrr li\-vu ay\-e\-veng\-hu.} Even a warrior must be gentle with children.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 
\B{Flrr\-a tom\-pa ze\-rup.} A gentle rain is falling.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 
(\ding{214}~\HL{kakan}{\B{ka\-kan}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{nIflrr}{\B{nìflrr}}, tenderly, mildly, gently. 
\on{2}.~\HL{tIflrr}{\B{tìflrr}}, gentleness, tenderness.
\on{3}.~\HL{flrrtsawl}{\B{flrr\-tsawl}}, sailfin goliath. 

% flrrtsawl
\hi 
\HT{flrrtsawl}{\B{\cp{flrr}tsawl}} \I{["fl\s{r}.\t{ts}awl]} \emph{n., fauna} sailfin goliath. 
(\ding{43}~\HL{flrr}{\B{flrr}}, \HL{tsawl}{\B{tsawl}})

% fmal
\hi
\HT{fmal}{\B{fmal}} \I{[fmal]} \emph{vtr.} sustain. \B{Te\-ri fwa fmal fì\-lì'\-fya\-ti a\-yaw\-ne.} On keeping this beloved language alive.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b} \B{Lu\-ke pay, ke tsun ay\-oe tì\-rey\-ti fmi\-val.} Without water we cannot sustain life.~\LINK{http://forum.learnnavi.org/language-updates/for-\%28duration\%29-jesus-loan-word/msg372740/\#msg372740}{ln\slash g}

% fmawn
\hi
\HT{fmawn}{\B{fmawn}} \I{[fmawn]} \emph{n.} news, something to report.  
\B{Sìl\-pey oe, la\-yu oe\-ru ye'\-rìn sìl\-tsan\-a fmawn.} I hope that I will have good news soon.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Slä lu oe\-ru fmawn\-o a\-sìl\-tsan.} But I have some good news.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Nì\-sngä\-'i fmawn\-it fo nar\-mew wi\-van, slä nì\-'i'\-a fra\-por lo\-lo\-nu.} They originally wanted to hide the news, but in the end they revealed it everyone.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
(\emph{a. idiom}) \B{Pe\-fmawn?} `What's the report?\slash What's the news?\slash What's new?'~\LINK{http://naviteri.org/2018/02/negative-questions-in-navi/\#comment-28201}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fmawnta}{\B{fmawnta}}, (the news) that.

% fmawnta
\hi
\HT{fmawnta}{\B{\cp{fmawn}ta}} \I{["fmawn.ta]} \emph{conj.} (\emph{short form for} \B{fmawnit a}) \emph{to introduce reported speech of the verbs} \ding{43}~\HL{peng}{\B{peng}} \emph{and} \HL{stawm}{\B{stawm}}, (the news) that. 
\B{Sto\-lawm oel fmawn\-ta fo new hi\-vum.} I heard they want to leave.\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} 
\B{Ngal po\-leng oer fmawn\-ta po to\-ler\-kup.} You told me that he died.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} 
(\ding{43}~\HL{fmawn}{\B{fmawn}})

% fmetok
\hi
\HT{fmetok}{\B{\cp{fme}tok}} \I{["fmE.tok\textcorner]} \emph{vtr.} test (s.o.).
\B{Ney\-ti\-ril fma\-ye\-tok fì\-tsam\-si\-yut.} Neytiri will test this `warrior'.~\LINK{https://forum.learnnavi.org/language-updates/about-fmetok-and-pofo/}{a\textsuperscript{c}\slash ln} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIfmetok}{\B{tìfmetok}}, test.

% fmi
\hi
\HT{fmi}{\B{fmi}} \I{[fmi]} \emph{vtr., modal} try, attempt. 
\B{Fmi pi\-vll\-txe alo avol nì\-win!} Try to say that eight times fast!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
\B{Oe fmo\-li nu\-ä\-ngi.} I tried, but unfortunately I failed.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{Fma\-yi oe 'i\-veyng ye'\-rìn.} I will try to answer soon.~\LINK{http://forum.learnnavi.org/language-updates/another-email-ta-frommer-no-answers-as-of-yet-but-a-few-interesting-points/msg196441/\#msg196441}{ln} (\ding{43}~\B{may'})\\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIfmi}{\B{tìfmi}}, attempt. 
\on{2}.~\HL{sAfmi}{\B{säfmi}}, attempt.

% fmokx
\hi
\HT{fmokx}{\B{fmokx}} \I{[fmok']} \emph{n.} jealousy, envy (\emph{neutral connotation unless otherwise specified with} ‹\B{äng}› \emph{or} ‹\B{ei}›). 
\B{lu oe\-ru fmokx}, I'm jealous.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
\B{Fu\-ri\-a fì\-txan fnä\-ngan Ul\-rey\-ìl tsko swi\-zaw\-it, lu oe\-ru fmokx.} I'm jealous of the fact that Ul\-rey is such an excellent archer.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
\B{Fko pll\-txe\-i\-e san me\-nga mun\-txa slo\-lu sìk!~Sey\-kxel sì ni\-tram!~Slä lu oe\-ru fmokx nì\-'it.} I'm so happy to hear you got married! Congratulations! I am a little envious, though.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{nIfmokx}{\B{nìfmokx}}, jealously, enviously.

% fmong
\hi 
\HT{fmong}{\B{fmong}} \I{[fmoN]} \emph{vtr.} steal, rob. 
\B{Fo fmong, fo nekx, fo ska'a.} They steal, they burn, they destroy.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIfmong}{\B{tì\-fmong}}, theft.  
\on{2}.~\HL{fmongyu}{\B{fmong\-yu}}, thief.
\on{3}.~\HL{sAfmong}{\B{sä\-fmong}}, theft (\emph{particular instance}).

% fmongyu
\hi 
\HT{fmongyu}{\B{\cp{fmong}yu}} \I{["fmoN.ju]} \emph{n.} thief. 
(\ding{43}~\HL{fmong}{\B{fmong}}) 

% fnan
\hi
\HT{fnan}{\B{fnan}} \I{[fnan]} \emph{vtr.} be good at. 
\B{Sra\-ke fnan ngal lì'\-fya\-ti le\-Na'\-vi?} Are you good at Na'vi?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
\B{Fol fnan fu\-ta 'em tey\-lut a fì\-'u sìn pe\-yì?} How good are they at cooking \emph{teylu}?~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{214}~\HL{wAtx}{\B{wätx}})

% fnawe'
\hi
\HT{fnawe'}{\B{fna\cp{we'}}} \I{[fna."wEP]} \emph{adj.} cowardly. \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fnawe'tu}{\B{fna\-we'\-tu}}, coward.
\on{2}.~\HL{tIfnawe'}{\B{tì\-fna\-we'}}, cowardice.

% fnawe'tu
\hi
\HT{fnawe'tu}{\B{fna\cp{we'}tu}} \I{[fna."wEP.tu]} \emph{n.} coward. 
\B{Txan\-tstew sä\-ro\-'a si, fna\-we'\-tu ke si.} A hero does great deeds, a coward does not.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Mu\-nge fna\-we'\-tul lewng\-it soaia\-ru sne\-yä.} A coward brings shame to his\slash her family.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
(\emph{a. proverb}) \B{Kxìm utu\-ftu fna\-we'\-tu.} ``A coward commands from the canopy,'' i.e. \emph{a real leader will have `boots on the ground' and will help out, whereas a coward will only tell people (from afar) what to do}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{43}~\HL{fnawe'}{\B{fna\-we'}})

% fne-
\hi
\HT{fne}{\B{fne-}} \I{[fnE]} \emph{pref. on nouns} kind (of), type (of). 
\B{Kel\-ut\-ral lu fne\-ut\-ral a\-zey.} Hometree is a special kind of tree.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
\B{la\-ye\-i\-u oer krr nì\-'ul fte ngi\-vop ay\-lì\-'ut sì tsay\-fne\-sä\-num\-vit a tsun fra\-por srung si\-vi}, I will have more time to create words and those kinds of lessons that will help everyone.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Lu pì\-lok\-ur pxe\-sì\-kan sì pxe\-fne\-'u\-pxa\-re.} The blog has three aims and three kinds of messages.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
\B{Por war\-mo\-u fra\-fne\-i\-o\-ang nì\-tut, slä nì\-pxi pxe\-syì\-rä\-fì.} He just couldn't get over all the animals, but was especially taken with the 3 giraffe.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
(\ding{43}~\HL{fnel}{\B{fnel}})

% fnel
\hi
\HT{fnel}{\B{fnel}} \I{[fnEl]} \emph{n.} kind, type. 
\B{tsa\-fnel syu\-lang\-ä}, that kind of flower.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
(\ding{43}~\HL{fne}{\B{fne-}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fnelan}{\B{fnelan}}, male. 
\on{2}.~\HL{fnele}{\B{fnele}}, female.
\on{3}.~\HL{lI'fyafnel}{\B{lì'\-fya\-fnel}}, dialect, variety of a language.

% fnelan
\hi 
\HT{fnelan}{\B{\cp{fne}lan}} \I{["fnE.lan]} \emph{n.} male.
(\ding{43}~\HL{fnel}{\B{fnel}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{lefnelan}{\B{lefnelan}}, male.

% fnele
\hi 
\HT{fnele}{\B{\cp{fne}le}} \I{["fnE.lE]} \emph{n.} female.
(\ding{43}~\HL{fnel}{\B{fnel}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{lefnele}{\B{lefnele}}, female.

% fnepe
\hi
\HT{fnepe}{\B{\cp{fne}pe}} \I{["fnE.pE]} \emph{inter.} which kind. 
(\ding{43}~\HL{fnel}{\B{fnel}}, \HL{pefnel}{\B{pe\-fnel}})

% fnetxum
\hi
\HT{fnetxum}{\B{fne\cp{txum}}} \I{[fnE."t'um]} \emph{or} \I{[."t'Um]} (\textsc{rn}: \B{fne\cp{txùm}} \I{[fne."dUm]}) \emph{idiom} 
\B{Oe\-ri lu swo\-a fne\-txum.} I'm allergic to alcohol [lit.: alcohol is a kind of poison for me].~\LINK{http://forum.learnnavi.org/language-updates/allergies/}{ln}
(\ding{43}~\HL{fne}{\B{fne-}}, \HL{txum}{\B{txum}})

% fnu 
\hi
\HT{fnu}{\B{fnu}} \I{[fnu]} \emph{vin.} be quiet, be still. 
\B{Fnu, ma smuk, fnu! Oel he\-fi ye\-rik\-it!} Quiet, everyone! I smell a hexapede!~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Oe ä\-txä\-le so\-li tsnì fra\-po fni\-vu.} I request that everyone be quiet.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/comment-page-1/\#comment-1060}{b}\\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIfnu}{\B{tìfnu}}, quiet, silence. 
\on{2}.~\HL{nIfnu}{\B{nìfnu}}, quietly.

% fngap
\hi
\HT{fngap}{\B{fngap}} \I{[fNap\textcorner]} \emph{n.} metal. \B{Fu\-ri\-a txu\-la tsa\-lew\-ti lu srä sìl\-tsan to fngap.} For constructing that cover, woven cloth is better than metal.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}\\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{lefngap}{\B{lefngap}}, metallic. 
\on{2}.~\HL{fngapsutxwll}{\B{fngapsutxwll}}, metalfollowing plant.

% fngapsutxwll
\hi
\HT{fngapsutxwll}{\B{\cp{fngap}sutxwll}} \I{["fNap\textcorner.sut'.w\s{l}]} \emph{n.}, \ding{96} metalfollowing plant, \emph{pennanemone ciliare}.
(\ding{43}~\HL{fngap}{\B{fngap}}, \HL{sutx}{\B{sutx}}, \HL{'ewll}{\B{'e\-wll}})

% fngä' 
\hi
\HT{fngA'}{\B{fngä'}} \I{[fN\ae{}P]} \emph{vin.} relieve oneself. 
\B{Fko tsun fngi\-vä' pe\-seng?} Where is the bathroom? [lit.: Where can one relieve oneself?]~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% fngä'tseng
\hi 
\HT{fngA'tseng}{\B{\cp{fngä'}tseng}} \I{["fN\ae{}P.\t{ts}EN]} \emph{n.} restroom (\emph{general term, any `place of elimination'}). 
(\ding{43}~\HL{fngA'}{\B{fngä'}}, \HL{tseng}{\B{tseng(e)}}; \HL{moafngA'}{\B{mo a fngä'}})

% fngo'
\hi
\HT{fngo'}{\B{fngo'}} \I{[fNoP]} \emph{vtr.} require, demand. 
\B{Fol fte ay\-spe\-'e\-tut li\-vo\-nu fngo' 'u\-pet?} What are they demanding in order for them to release the captives?~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
\B{Fì\-fne\-tì\-fkey\-tok\-ìl fngo' fu\-ta kem si\-vi fko pxi\-ye'\-rìn.} This kind of situation requires immediate action.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
\B{Kar\-yul fngo\-lo' fu\-ta ay\-nu\-me\-yu pi\-va\-te ye'\-krr.} The teacher required the students to arrive early.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\  
\textsc{Derivations:} \ding{43} \on{1}.~\HL{sAfngo'}{\B{säfngo'}}, requirement, demand.

% fo
\hi
\HT{fo}{\B{fo}} : \HL{ayfo}{\B{ayfo}} 

% fpak
\hi
\HT{fpak}{\B{fpak}} \I{[fpak\textcorner]} \emph{vin.} hold off, suspend an action (that is already in progress). 

% fpap
\hi 
\HT{fpap}{\B{fpap}} \I{[fpap\textcorner]} \emph{vtr.} pound (\emph{striking heavily and repeatedly}). 
\B{Krr\-a sti nì\-txan, pol fyan\-yot fpap fa me\-syokx.} When he's angry, he pounds the table with his hand.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}

% fpe'
\hi
\HT{fpe'}{\B{fpe'}} \I{[fpEP]} \emph{vtr.} send. 
\B{Fpo\-le' ay\-ngal oer a 'u\-pxa\-re.} The message you have sent me.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}

% fpeio
\hi
\HT{fpeio}{\B{fpe\cp{i}o}} \I{[fpE."{}i.o]} \emph{n.} (ceremonial) challenge. 

% fpi
\hi
\HT{fpi}{\B{fpi}} \I{[fpi]} \emph{adp.}\texttt{+} for (the benefit or sake of). \B{Fay\-u\-pxa\-re la\-yu ay\-sngä\-'i\-yu\-fpi.} These messages will be for beginners.~\LINK{http://naviteri.org/2010/06/first-post/}{b} \B{Stxe\-lit fpo\-le' oel nga\-fpi.} I sent the gift for you (\emph{i.e. I sent it to someone else for your sake})~\LINK{http://forum.learnnavi.org/language-updates/i-have-something-to-say/}{ln} \B{Oe\-ri tì\-rey\-ti oel stxe\-nu\-to\-lìng fpi o\-lo' aw\-nge\-yä.} I offered my life for the sake of our clan.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}

% fpivìl
\hi 
\HT{fpivIl}{\B{fpi\cp{vìl}}} \I{[fpi."vIl]} \emph{intj.} ``hmm, let's see, let me think''. 
\B{Fpi\-vìl … Kxawm nga\-ru tì\-yawr.} Hmm . . . Perhaps you're right.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{fpIl}{\B{fpìl}})

% fpìl
\hi
\HT{fpIl}{\B{fpìl}} \I{[fpIl]} \emph{vtr.} think. 
\B{Fwa fpìl ngal fu\-ta fay\-lì'\-fya\-vi li\-vu le\-sar oe\-ru prr\-te' lu.} That you think that these expressions are useful pleases me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-293}{b} 
\B{Fra\-krr po fpìl nì\-tsim nì\-wotx.} Her thinking is always completely original.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
\B{Fpìr\-mìl oel fu\-ta ay\-nga na\-tsew tsi\-ve\-'a fì\-'ut.} I was just thinking that you might want to see this.~\LINK{http://forum.learnnavi.org/language-updates/the-evidential-at-last/msg106055/\#msg106055}{ln} 
(\emph{a. with} \B{san … sìk}) \B{Lu su\-te a fpìl san oe\-ri tì\-rey tì\-sraw si, ha tì\-rey\-ti nge\-yä oel tì\-sraw sey\-ka\-syi nì\-teng … sìk.} There are people who think, `My life is hurting, so I will make your life hurt as well ….'~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} 
(\emph{b. conv.}) \B{fpìl oe …} `I think …' \B{Fpìl oe, fì\-vur za\-yaw\-prr\-te' ay\-nga\-ne!} I think, this story will be enjoyable for you all!~\LINK{http://naviteri.org/2020/04/keltrrtrra-tieylan-an-unusual-friendship/}{b} \\  
\textsc{Derivations:} \ding{43} \on{1}.~\HL{sAfpIl}{\B{säfpìl}}, idea, thought.
\on{2}.~\HL{fpIlfya}{\B{fpìlfya}}, thought pattern, way of thinking.
\on{3}.~\HL{kanfpIl}{\B{kanfpìl}}, concentrate, focus one's attention. 
\on{4}.~\HL{fpivIl}{\B{fpi\-vìl}}, ``hmmm \ldots, let me think''.
\on{5}.~\HL{txakrrfpIl}{\B{txa\-krr\-fpìl}}, consider, ponder.

% fpìlfya
\hi
\HT{fpIlfya}{\B{\cp{fpìl}fya}} \I{["fpIl.fja]} \emph{n.} thought pattern, way of thinking. \B{Slä sa'\-nok\-ì\-ri fpìl\-fya lu ke\-teng.} But mother's way of thinking is different.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
(\ding{43}~\HL{fpIl}{\B{fpìl}}, \HL{fya'o}{\B{fya\-'o}}) \\ 
\textsc{Derivations:} \ding{43} \on{1}.~\HL{snafpIlfya}{\B{snafpìlfya}}, philosophy.

% fpom
\hi
\HT{fpom}{\B{fpom}} \I{[fpom]} \emph{n.} well\hyp{}being, happiness; peace.
\B{Pxìm lu tì\-new le\-hawng kxu\-tu fpom\-ä.} Excessive desire is often the enemy of peace.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Oey fpom\-it sweyn!} Leave me alone! [lit.: do not disturb my peace]~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\emph{a. idiom}) \B{Nga\-ru lu fpom srak?} How are you? [lit.: Do you have well\hyp{}being?]~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Vanity_Fair}{wiki} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{lefpom}{\B{lefpom}}, happy. 
\on{2}.~\HL{fpomtokx}{\B{fpomtokx}}, (physical) health. 
\on{3}.~\HL{fpomron}{\B{fpomron}}, (mental) health.

% fpomron
\hi
\HT{fpomron}{\B{fpom\cp{ron}}} \I{[fpom."Ron]} \emph{n.} (mental) health, well\hyp{}being.
\B{Po\-ri fpom\-tokx sì fpom\-ron yo'.} His physical and mental health are perfect.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\B{fpom}, \B{ron\-sem}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{lefpomron}{\B{lefpomron}}, (mentally) healthy. 
\on{2}.~\HL{kelfpomron}{\B{kelfpomron}}, (mentally) unhealthy. 
\on{3}.~\HL{fpomronga'}{\B{fpomronga'}}, (mentally) healthful. 
\on{4}.~\HL{kefpomronga'}{\B{kefpomronga'}}, (mentally) unhealthful. 

% fpomronga'
\hi
\HT{fpomronga'}{\B{fpom\cp{ro}nga'}} \I{[fpom."Ro.NaP]} \emph{adj., nfp.} (mentally) healthful.
\B{Fwa lawk ay\-sì\-'efu\-ti ay\-ey\-lan\-kip lu fpom\-ro\-nga'.} It's healthy among friends to discuss feelings.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{fpomron}{\B{fpom\-ron}}, \ding{214}~\HL{kefpomronga'}{\B{ke\-fpom\-ro\-nga'}})

% fpomtokx
\hi
\HT{fpomtokx}{\B{fpom\cp{tokx}}} \I{[fpom."tok']} \emph{n.} (physical) health. 
\B{Po\-ri fpom\-tokx sì fpom\-ron yo'.} His physical and mental health are perfect.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Sìl\-pey oe tsnì fì\-zì\-sìt\-ìl ay\-nga\-ru za\-mì\-ye\-vu\-nge txan\-a fpom\-it sì fpom\-tokx\-it.} I hope this year will bring you much happiness and health.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/}{b}
\B{Nge\-yä fpom\-tokx\-ìri fì\-tì\-fkey\-tok a\-fkxa\-ra\-nga' lu le\-hrr\-ap.} This stressful situation is dangerous to your health.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\B{fpom}, \B{tokx}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{lefpomtokx}{\B{lefpomtokx}}, (physically) healthy. 
\on{2}.~\HL{kelfpomtokx}{\B{kelfpomtokx}}, (physically) unhealthy. 
\on{3}.~\HL{fpomtokxnga'}{\B{fpomtokxnga'}}, (physically) healthful.  
\on{4}.~\HL{kefpomtokxnga'}{\B{kefpomtokxnga'}}, (physically) unhealthful.

% fpomtokxnga'
\hi
\HT{fpomtokxnga'}{\B{fpom\cp{tokx}nga'}} \I{[fpom."tok'.NaP]} \emph{adj., nfp.} (physically) healthful.
\B{Tsat rä\-'ä yi\-vom! Ke lu fpom\-tokx\-nga'.} Don't eat that. It's not healthful  (i.e., it will make you unhealthy).~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{fpomtokx}{\B{fpom\-tokx}}, \ding{214}~\HL{kefpomtokxnga'}{\B{ke\-fpom\-tokx\-nga'}})

% fpxafaw
\hi
\HT{fpxafaw}{\B{\cp{fpxa}faw}} \I{["fp'a.faw]} \emph{n., fauna} medusa, \emph{electromedusa aerae}. 

% fpxamo
\hi 
\HT{fpxamo}{\B{\cp{fpxa}mo}} \I{["fp'a.mo]} \emph{adj.} terrible, horrible, awful. 
\B{Maw\-krr\-a fko lie so\-li tì\-len\-ur a\-fpxa\-mo fì\-txan, tì\-rey ke lu teng kaw\-krr.} After experiencing such a terrible event, life is never the same.~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b} 
\B{Maw fpxa\-mo\-a kin\-trr a\-fpxa\-mo mì tan\-lo\-kxe oe\-yä …} After a truly horrible week in my country … \LINK{http://naviteri.org/2022/05/results-of-the-tttc/}{b}
(\ding{214}~\HL{kosman}{\B{kosman}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIfpxamo}{\B{tìfpxamo}}, horror. 
\on{2}.~\HL{nIfpxamo}{\B{nìfpxamo}}, horribly, terribly, awfully.

% fpxäkìm
\hi
\HT{fpxAkIm}{\B{\cp{fpxä}kìm}} \I{["fp'\ae.kIm]} \emph{vin.} (\emph{a.}) enter (\B{ìlä}, through).   
\B{Lu mek\-tseng a tsun fpxi\-vä\-kìm hì\-'ang tsaw\-ìlä.}  There's a gap through which insects can get in.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{Fpxo\-lä\-kìm fo ìlä rawng a\-hì\-'i.} They entered through a small doorway.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\emph{b. of astronomical bodies rising}) \B{Tsaw\-ke fpxe\-rä\-kìm nem\-fa taw.} \emph{or} \B{Tsaw\-ke fpxe\-rä\-kìm.} The sun is rising.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} (\ding{214}~\HL{hum}{\B{hum}})

% fra-
\hi
\HT{fra}{\B{fra-}} \I{[fRa.]} \emph{pref. on nouns} every, all. 
\B{Po\-ri wem\-tswo fra\-tsam\-si\-yur ro\-lo\-'a nì\-txan.} His ability to fight greatly impressed all the warriors.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} 
\B{Fra\-syu\-ra\-ti fkol za\-sro\-lìn nì\-'aw ul\-te trr\-o ze\-ne tey\-ki\-vä\-txaw.} All energy is only borrowed, and one day it will have to be given back.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}

% fra'u
\hi
\HT{fra'u}{\B{\cp{fra}'u}} \I{["fRa.Pu]} (\textsc{rn}: \B{frau} \I{["fRa.u]}) \emph{n.} everything. 
\B{Tse \dots za\-'u fra\-'u ne tu\-te lem\-wey\-pey, ke\-fyak?} Well, everything comes to the patient people, doesn't it?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} 
\B{Sra\-ke tso\-lun fra\-'ut tsli\-vam kxey\-ey\-lu\-ke?} Could you understand everything without error?~\LINK{http://naviteri.org/2012/10/txon-eywaevenga-text-and-translation/}{b} 
\B{Fra\-'uä ran ngä\-pop fa fra\-syon tseyä.} The \emph{ran} of each thing arises from the totality of its attributes.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\emph{a. coll. shortened to} \B{fraw}) \B{'U\-pe ke zo?--Fraw mì la\-'ang.} What's the matter?--Everything is screwed up.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{43}~\HL{'u}{\B{'u}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{frawzo}{\B{frawzo}}, all is well.

% frakrr
\hi
\HT{frakrr}{\B{\cp{fra}krr}} \I{["fRa.k\s{r}]} \textsc{i}. \emph{n.} every time. \\
\textsc{ii}. \emph{adv.} always. \B{Pol\-txe Ney\-ti\-ril ay\-lì\-'ut a fra\-krr 'ok se\-yä la\-yu oer.} Neytiri said something that I will always remember.~\LINK{http://thereadinginpublicproject.blogspot.de/2010/05/paul-frommer.html}{pr}
(\ding{43}~\HL{krr}{\B{krr}})

% fralo
\hi
\HT{fralo}{\B{\cp{fra}lo}} \I{["fRa.lo]} \emph{adv.} every time, each time. 
\B{Fra\-lo poe pol\-txe san ke\-he.} Each time she said `no'.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\ding{43}~\HL{alo}{\B{alo}}) 

% franse
\hi
\HT{franse}{\B{\cp{Fran}se}} \I{["fRan.sE]} \emph{n.}, \textsc{lw} French, the French language. \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{nIfranse}{\B{nìFranse}}, in French.

% frapo
\hi
\HT{frapo}{\B{\cp{fra}po}} \I{["fRa.po]} (\emph{gen.}~\B{frapeyä}) \emph{pn.} everyone, everybody. 
\B{Fra\-po ne mì\-fa!} Everbody inside!~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b} 
\B{tslam fra\-pol fu\-ta slu po O\-lo'\-eyk\-tan a\-mip}, everyone understood that he had become the new Clan Leader.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
\B{Po\-leng Ney\-ti\-ril hang\-vur\-it a fra\-pot hey\-kang\-ham.} Neytiri told a joke that made everyone laugh.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
\B{Fì\-tì\-hawl\-ì\-ri fe'\-ran law lä\-ngu fra\-por.} Unfortunately the flawed nature of this plan is obvious to everyone.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\ding{43}~\HL{po}{\B{po}})

% frato
\hi
\HT{frato}{\B{\cp{fra}to}} \I{["fRa.to]} \emph{part.} (\emph{superlative marker}) than all. 
(\emph{a. with (scalable) adjectives and adverbs}) 
\B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok, tso\-le\-'a syep\-tu\-tet a\-tsawl fra\-to mì sì\-rey.} Yesterday I was with Txewì in the forest and we saw the biggest Trapper I've ever seen.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki} 
\B{New oe ti\-vìng ay\-ngar lì\-'ut a tì\-'e\-fu\-mì oe\-yä lu lor fra\-to mì lì'\-fya le\-Na'\-vi.} I want to give you the word that, in my opinion, is the most beautiful in the Na'vi language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Pol ye\-rik\-it ta\-ron nìl\-tsan fra\-to.} He hunts yerik the best of all.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln} 
(\emph{b. with verbs}) \B{Tsran\-ten fra\-to a syon tsam\-si\-yu\-ä lu tì\-tstew.} The most important characteristic of a warrior is bravery.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Tì\-'o'\-ì\-ri pe\-u sunu ngar fra\-to?} What is your favorite way to have fun?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
\B{Nga\-ru le\-i\-o\-a\-e si oe fra\-to, ma 'ey\-lan.} I respect you the most of all, friend.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
\B{pi\-za\-yu Ney\-ti\-ri\-yä txan\-rar\-mo\-'a fra\-to}, Ney\-ti\-ri's ancestor was the most famous.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}

% fratrr
\hi
\HT{fratrr}{\B{fra\cp{trr}}} \I{[fRa."t\s{r}]} \emph{adv.} every day.
\B{Lì'\-fya\-ri le\-Na'\-vi nga tsan\-'e\-re\-i\-ul fra\-trr.} I'm delighted that your Na'vi is improving every day.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\ding{43}~\HL{trr}{\B{trr}})

% fratxon
\hi
\HT{fratxon}{\B{fra\cp{txon}}} \I{[fRa."t'on]} \emph{adv.} every night. 
(\ding{43}~\HL{txon}{\B{txon}})

% fratseng
\hi
\HT{fratseng}{\B{\cp{fra}tseng}} \I{["fRa.\t{ts}EN]} \emph{adv.} everywhere. 
\B{Tsun Txil\-te pam\-rel\-it ivi\-nan; ta\-fral puk\-ot a\-nutx mu\-nge fra\-tseng.} Txilte knows how to read; therefore she brings a thick book wherever she goes.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\ding{43}~\HL{tsenge}{\B{tse\-nge}})


% fraw
\hi
\HT{fraw}{\B{fraw}} \I{[fRaw]}~:~\HL{fra'u}{\B{fra'u}}

% frawzo
\hi
\HT{frawzo}{\B{fraw\cp{zo}}} \I{[fRaw."zo]} (\emph{idiom}) all is well, everything is fine, everything is \textsc{ok}. 
\B{Sìl\-pey Txe\-wì tsnì tsi\-vun set fko pi\-vll\-txe san fraw\-zo.} Txewì hopes that now one can say, `everything is fine'.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Nge\-yä tsa\-sä\-ngä\-'än\-tsyìp\-ì\-ri set fra\-wzo srak?} Have you recovered from being down for a while?~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{fra'u}{\B{fra\-'u}}, \HL{zo}{\B{zo}})

% fray+
\hi
\HT{fray}{\B{fray}}\texttt{+} \I{[fRaj.]} (\emph{pref. to nouns}) every, all of \LINK{https://forum.learnnavi.org/language-updates/navi-details-from-avatarmeet-2013/}{ln}. 
\B{Kay\-mo zo\-la\-'u fray\-ul\-txa\-tu ne kel\-ku moe\-yä.} One evening all the members of the meeting came to our house.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b} 
(\ding{43}~\HL{fra}{\B{fra-}}, \HL{ay}{\B{ay}}\texttt{+})

% frir
\hi 
\HT{frir}{\B{frir}} \I{[fRiR]} \emph{n.} layer. 
\B{Tskxe\-pay\-ri lu frir a\-flì sìn 'o\-ra.} There's a thin layer of ice on the lake.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
\B{Fay\-frir le\-tskxe lor lu nì\-txan.} These stone layers are very beautiful.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{lefrir}{\B{lefrir}}, layered. 
\on{2}.~\HL{nIfrir}{\B{nìfrir}}, in layers.

% frìp
\hi
\HT{frIp}{\B{frìp}} \I{[fRIp\textcorner]} \emph{vtr.} bite. \B{Frìp, frìp, ma Evi\-tsyìp, To\-ruk\-ä tsawl\-a kxa!} The big mouth of the Great Leonopteryx bites, bites, children.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{sAfrIp}{\B{säfrìp}}, a bite.

% frrfen
\hi
\HT{frrfen}{\B{\cp{frr}fen}} \I{["f\s{r}.fEn]} (\emph{with} ‹\B{er}›: \B{frr\-fen}) \emph{vtr.} visit. 
\B{Oe\-ri so\-la\-lew wum zì\-sìt °a14 a krr, fo\-lrr\-fen spo\-not alo a\-'aw\-ve.} When I was about 12 years old I visited an island for the first time.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{frrtu}{\B{frrtu}}, visitor.
\on{2}.~\HL{sAfrrfen}{\B{sä\-frr\-fen}}, a visit.

% frrtu
\hi
\HT{frrtu}{\B{\cp{frr}tu}} \I{["f\s{r}.tu]} (\textsc{rn}: \B{frrtu} \I{["f\s{r}.tu]}) \emph{n.} visitor, guest. 
\B{Sko frr\-tu, fra\-krr mì hel\-ku nge\-yä lu oe\-ru ho\-an nì\-txan, ma tsmuk.} I always feel very comfortable as a guest in your home, brother.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
\B{Rä\-'ä fmi\-vi li\-vok fu em\-ki\-vä ay\-ekxan\-it a fkol ngo\-lop fpi sì\-kxu\-ke ay\-frr\-tu\-ä sì ay\-ioang\-ä.} Do not attempt to approach or cross any barriers designed for Guest and animal safety.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b}
(\ding{43}~\HL{frrfen}{\B{frr\-fen}})

% fta
\hi
\HT{fta}{\B{fta}}\textsuperscript{\on{1}} \I{[fta]} \textsc{i}.~\emph{n.} knot. \\
\textsc{ii}.~\HT{fta si}{\B{fta si}} \I{[fta si]} \emph{vin.} knot, make or tie a knot. \B{Txur\-tel\-mì fo fta so\-li fte\-ke ka tsyokx fwi\-vi.} They tied a knot in the rope so it wouldn't slip through their hands.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% fta
\hi
\B{fta}\textsuperscript{\on{2}} : \HL{futa}{\B{futa}}

% ftang
\hi
\HT{ftang}{\B{ftang}} \I{[ftang]} \emph{vin., modal} stop (to do s.t.\slash an action). \B{Ik\-ran\-ì\-ri krra hu tu\-te tsa\-heyl si, ftang li\-vu yrr.} When an ikran bonds with a person, it ceases to be wild.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \B{Hay\-a yak\-ro fti\-vang.~Sa\-lew rä\-'ä.} Stop at the next fork.~Do not proceed further.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIftang}{\B{tìftang}}, stopping, glottal stop. 
\on{2}.~\HL{nIlkeftang}{\B{nìlkeftang}}, continuously, incessantly. 
\on{3}.~\HL{ftanglen}{\B{ftanglen}}, prevent.

% ftanglen
\hi 
\HT{ftanglen}{\B{ftang\cp{len}}} \I{[ftaN."lEn]} \emph{vtr.} prevent. 
\B{Tsran\-ten nì\-txan fwa ftang\-len aw\-ngal fu\-ta fì\-sä\-spxin vi\-vi\-rä.} It's very important that we prevent this disease from spreading.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b} 
(\ding{43}~\HL{ftang}{\B{ftang}}, \HL{len}{\B{len}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIftanglen}{\B{tìftanglen}}, prevention.

% ftawnemkrr
\hi
\HT{ftawnemkrr}{\B{ftaw\cp{nem}krr}} \I{[ftaw."nEm.k\s{r}]} \emph{n.} past. 
\B{txo tsa\-kem li\-ven mì ftaw\-nem\-krr}, if it happens in the past.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1910}{b} 
\B{Ul\-te ftaw\-nem\-krr\-ìri, ‹imv› tam nì\-teng.} And concerning the past ‹imv› suffices as well.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1910}{b}
(\ding{43}~\HL{ftem}{\B{ftem}}, \HL{krr}{\B{krr}})

% ftär
\hi
\HT{ftAr}{\B{ftär}} \I{[ft\ae{}R]} \emph{adj.} left (\emph{not right}). 
\B{Lo\-lu po\-ru txukx\-a skxir a ftu 'et\-naw a\-ski\-en ne me\-pun a\-ftär.} It had a deep wound from its right shoulder to its two left arms.~\LINK{http://naviteri.org/2020/04/awa-tskxekengtsyip-a-mikyunfpi-niul-one-more-little-listening-exercise/}{b}
(\ding{214}~\HL{skien}{\B{ski\-en}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{ftArpa}{\B{ftärpa}}, left side.   
\on{2}.~\HL{nIftAr}{\B{nìftär}}, on the left.

% ftärpa
\hi
\HT{ftArpa}{\B{\cp{ftär}pa}} \I{["ft\ae{}R.pa]} \emph{n.} left side. \B{Mìn ne ftär\-pa.} Turn to the left.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{b} 
(\ding{214}~\HL{skiempa}{\B{ski\-em\-pa}}, \ding{43}~\HL{ftAr}{\B{ftär}}, \HL{nIftAr}{\B{nì\-ftär}})

% fte
\hi
\HT{fte}{\B{fte}} \I{[ftE]} \emph{conj.} (\emph{requires} ‹\B{iv}›) so that, (in order) to. 
\B{Aw\-nga rol fte ki\-vame}, we sing to \textsc{see}.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}  
\B{He\-tu\-wong\-ìl aw\-nge\-yä swo\-tut ska\-'a, fte kll\-ki\-vu\-lat ke\-ru\-sey\-a tskxet.} The aliens destroy our sacred place to dig up dead rock.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{{ln\slash a}\textsuperscript{c}} 
\B{New oel fu\-ta u\-pxa\-re oe\-yä lu\-ke key\-ey li\-vu nì\-wotx, fte eyawr\-a sì\-ke\-nong\-it ti\-vìng su\-te\-ru.} I want my messages to be without errors, in order to give correct examples to people.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln} 
(\ding{214}~\HL{fteke}{\B{fte\-ke}})

% fteke
\hi
\HT{fteke}{\B{\cp{fte}ke}} \I{["ftE.kE]} \emph{conj.} (\emph{requires} ‹\B{iv}›) so that not, lest. 
\B{Nari so\-li ay\-oe fte\-ke nì\-hawng li\-vok.} We were careful not to get too close.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki} 
\B{Tì\-ran nì\-fnu txan\-txew\-vay fte\-ke ay\-ye\-rik\-ìl aw\-nga\-ti sti\-vawm.} Walk as quietly as possible so the hexapedes won't hear us.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} 
(\ding{214}~\HL{fte}{\B{fte}})

% ftem
\hi
\HT{ftem}{\B{ftem}} \I{[ftEm]} \emph{vtr.} pass by (s.t.). \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{ftawnemkrr}{\B{ftawnemkrr}}, (the) past.

% ftia
\hi
\HT{ftia}{\B{fti\cp{a}}} \I{[fti."{}a]} \emph{vtr.} study. 
\B{Zì\-sìt\-o a\-mrr fto\-li\-a o\-he, slä ze\-ne fko ni\-vume nì\-txan.} I studied for five years but there is much to learn.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Grace_and_Norm_in_the_Link_room}{wiki\slash a} 
\B{Fti\-a oel lì'\-fya\-ti le\-Na'\-vi nì\-'o' nì\-wotx!} Studying Na'vi is a ton of fun for me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIftia}{\B{tì\-ftia}}, studying.

% ftu
\hi
\HT{ftu}{\B{ftu}} \I{[ftu]} (\textsc{rn}: \B{ftu} \I{[ftu]}) \emph{adp.} from (\emph{direction}). \B{Ftu oe ne\-to rikx, ma \mbox{skxawng}!} Get away from me, moron!~\LINK{http://www.youtube.com/watch?v=ojr_BE8ZmVE}{yt} 
\B{Ftu fì\-tseng ze\-ne hi\-vum.} We gotta get out of here.~\LINK{http://forum.learnnavi.org/language-updates/txing-and-hum/}{ln} 
\B{Nga za\-'u pe\-senge\-ftu\slash tseng\-pe\-ftu?} Where are you from?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
(\ding{214}~\HL{ne}{\B{ne}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{ftumfa}{\B{ftum\-fa}}, out of, from inside. 
\on{2}.~\HL{ftuopa}{\B{ftuo\-pa}}, from behind.

% ftue
\hi
\HT{ftue}{\B{\cp{ftu}e}} \I{["ftu.E]} \emph{adj.} easy, simple.   
\B{Sì\-pawm nge\-yä ke lu ftu\-e}, your questions aren't easy.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#More_extracts_from_various_emails}{wiki} 
\B{Fì\-pam\-tseo\-tu\-ri ke la\-yu ftue fwa run fkol sä\-rawn\-it a tam.} It won't be easy to find a satisfactory replacement for this musician.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{214}~\HL{ngAzIk}{\B{ngä\-zìk}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{nIftue}{\B{nìftue}}, easily.

% ftumfa
\hi
\HT{ftumfa}{\B{\cp{ftum}fa}} \I{["ftum.fa]} \emph{adp.} out of, from inside. 
\B{Ri\-ti tswo\-lay\-on ftum\-fa slär.} The stingbat flew out of the cave.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{Rey\-pay skxir\-ftum\-fa he\-rum.} Blood is coming out of the wound.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\ding{214}~\HL{nemfa}{\B{nem\-fa}}, \ding{43}~\HL{ftu}{\B{ftu}}, \HL{mIfa}{\B{mì\-fa}})

% ftuopa
\hi 
\HT{ftuopa}{\B{\cp{ftu}opa}} \I{["ftu.o.pa]} \emph{adp.} from behind. 
\B{Sro\-ler fwä\-kì ftu\-opa tskxe.} A mantis appeared from behind a rock.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{ftu}{\B{ftu}}, \HL{uo}{\B{uo}}, \HL{pa'o}{\B{pa'o}})

% ftxavang
\hi
\HT{ftxavang}{\B{\cp{ftxa}vang}} \I{["ft'a.vaN]} \emph{adj.} passionate. \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIftxavang}{\B{tìftxavang}}, passion. 
\on{2}.~\HL{nIftxavang}{\B{nìftxavang}}, passionately. 

% ftxey
\hi
\HT{ftxey}{\B{ftxey}} \I{[ft'Ej]} \textsc{i}. \emph{vtr.} choose. 
\B{ftxey 'aw\-pot a ay\-nga\-kip} choose one among you.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki} 
\B{Ul\-te ftxo\-ley pol ay\-eyk\-ta\-nay\-ti a lu stum nì\-ftxan kawng na po.} And he has chosen deputies who are almost as bad as he is.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}  \\
\textsc{ii}. \emph{conj.} (\emph{a.}) whether (or). 
\B{Fra\-po, ftxey sngä\-'i\-yu ftxey tsul\-fä\-tu, tsì\-ye\-vun fì\-tse\-nge ri\-vun 'u\-ot le\-sar.} Everyone, whether beginner or master, can find something useful here.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
(\emph{b.}) whether (or not). \B{Pi\-veng oer ftxey nga new ri\-vey fu\-ke.} Tell me whether you want to live (or not)?~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/msg156629/\#msg156629}{ln} 
\B{Ftxey nga za\-'u fu\-ke?} Are you coming (or not)?~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/msg156629/\#msg156629}{ln} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{tIftxey}{\B{tìftxey}}, choice.

% ftxì
\hi
\HT{ftxI}{\B{ftxì}} \I{[ft'I]} \textsc{i}.~\emph{n.} tongue.
\B{Nga\-ri ftxì vi\-var le\-fpom\-tokx li\-vu.} May your tongue remain healthy.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/comment-page-1/\#comment-10624}{b}  \\  
\textsc{ii}.~\B{tìng ftxì} \I{[tIN ft'I]} \emph{vin.} taste, take a taste. 
(\ding{43}~\HL{syam}{\B{syam}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{ftxIlor}{\B{ftxìlor}}, delicious. 
\on{2}.~\HL{ftxIvA'}{\B{ftxìvä'}}, bad-tasting.

% ftxìlor 
\hi
\HT{ftxIlor}{\B{ftxì\cp{lor}}} \I{[ft'I."loR]} \emph{adj.} delicious, good\hyp{}tasting. 
\B{Tey\-lu a oel ye\-rom lu ftxì\-lor.} The teylu I'm eating is delicious.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
\B{Fì\-na\-er\-ìri fkan ftxì\-lor.} This drink tastes good.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
(\ding{214}~\HL{ftxIvA'}{\B{ftxì\-vä'}})

% ftxìvä'
\hi
\HT{ftxIvA'}{\B{ftxì\cp{vä'}}} \I{[ft'I."v\ae{}P]} \emph{adj.} bad-tasting. 
\B{Fì\-na\-er ftxì\-vä' lu nì\-hawng, ha swey\-lu txo ngal tsat kxi\-vukx nì\-win.} This drink tastes horrible, so you'd better swallow it quickly.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 
(\ding{214}~\HL{ftxIlor}{\B{ftxì\-lor}})

% ftxozä
\hi
\HT{ftxozA}{\B{ftxo\cp{zä}}} \I{[ft'o."z\ae]} \textsc{i}. \emph{n.} holiday, celebration, happy occasion.  
\B{Ay\-ftxo\-zä Le\-fpom!} Happy Holidays!~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Out_of_Office_AutoReply}{wiki} 
\B{Ftxo\-zä\-ri ay\-lrr\-tok ngar.} Smiles to you on your celebration.~\LINK{http://forum.learnnavi.org/language-updates/good-luck-with-this-one!/}{ln} \\
\textsc{ii}. \HT{ftxozA si}{\B{ftxo\cp{zä} si}} \I{[ft'o."z\ae{} si]} \emph{vin.} celebrate. 
\B{Teng\-krr ftxo\-zä se\-re\-i\-yi aw\-nga, ke tswi\-va' ay\-lom\-tu\-ti ko!} A toast: To absent friends. [lit.: While we are celebrating, let us not forget those who we wish could also be here (but can't).]~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{leftxozA}{\B{leftxozä}}, celebratory.

% ftxulì'u
\hi
\HT{ftxulI'u}{\B{ftxu\cp{lì}'u}} \I{[ft'$\cdot$u."lI.Pu]} \emph{vin., inf.}\on{11} orate, give a speech. 
\B{Krra oe ftxu\-lì\-'u, pxìm 'efu fkxa\-rat nì\-txan.} When I give a speech, I often feel a lot of stress.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
\B{Maw\-krra Tsyeyk ftxo\-lu\-lì\-'u, tslam fra\-pol fu\-ta slu po O\-lo'\-eyk\-tan a\-mip.} After Jake spoke, everyone understood that he had become the new Clan Leader.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{ftxulI'uyu}{\B{ftxulì'uyu}}, orator.   
\on{2}.~\HL{tIftxulI'u}{\B{tìftxulì'u}}, speech-making, public speaking. 
\on{3}.~\HL{sAftxulI'u}{\B{säftxulì'u}}, speech, oration.

% ftxulì'uyu
\hi
\HT{ftxulI'uyu}{\B{ftxu\cp{lì}'uyu}} \I{[ft'u."lI.Pu.ju]} \emph{n.} orator, (public) speaker.
(\ding{43}~\HL{ftxulI'u}{\B{ftxu\-lì\-'u}})

% fu
\hi
\HT{fu1}{\B{fu}} \I{[fu]} \emph{conj.} (\emph{a.}) or. 
\B{Tsun fko tsa\-tse\-nge\-ne ki\-vä nì\-sle\-le fu fa fwa ik\-ran\-it mak\-to nì\-'aw.} You can only get there by swimming or riding an ikran.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} 
\B{Ke za\-syup lì\-'O\-na ne kxu\-tu a mì\-fa fu a wrr\-pa.} The l'Ona will not perish to the enemy within or the enemy without.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g} 
(\emph{b. choice statement}) 
\B{Nul\-new oel fì\-'ut fu tsa\-'ut.} I want this or that.~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/}{b}
(\emph{c. choice question}) 
\B{Nul\-new ngal fu fì\-'ut fu tsa\-'ut?} Do you want this or that?~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/}{b} 
\B{New nga fu yi\-vom fu ni\-väk?} Do you want to eat or drink?~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/\#comment-30389}{b}
(\emph{d. non\hyp{}choice question}) 
\B{Sra\-ke nul\-new ngal fì\-'ut fu tsa\-'ut?} Do you want this or that?~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fuke}{\B{fuke}}, or not.

% -fu
\hi
\HT{fu2}{\B{-fu}} \I{[.fu]} \emph{num., suf.}~six (\emph{rest form of} \ding{43}~\HL{pukap}{\B{pu\-kap}} \emph{to form higher numbers}). \B{vofu}, \on{14} (decimal), \on{16} (octal). \B{mevofu}, \on{22} (decimal), \on{26} (octal); etc.

% fuke
\hi
\HT{fuke}{\B{fu\cp{ke}}} \I {[fu."kE]} \emph{conj.} or not. \B{Oeng swey\-lu txo ki\-vä fu\-ke?} What do you think? Should we go or not?~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \B{Ftxey nga za\-'u fu\-ke?} Are you coming (or not)?~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/msg156629/\#msg156629}{ln} 
(\ding{43}~\HL{ftxey}{\B{ftxey}}, \HL{fu}{\B{fu}})

% fula
\hi
\HT{fula}{\B{\cp{fu}la}}, \B{fì\-\cp{'ul} a}, \emph{or} \B{a fì\-\cp{'ul}} \I{["fu.la]} \emph{conj.} that (\emph{to nominalize an agentive subordinate clause}) 
\B{Fu\-la ho\-ren sì ay\-lì'\-fya\-vi a\-mip le\-sar sa\-yi me\-ngar oe\-ti nit\-ram 'ey\-ke\-fu nì\-txan.} That the rules and new expressions are useful to you two makes me feel very happy.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/comment-page-1/\#comment-1515}{b} 
\B{Vì\-ming\-kap oe\-ti fu\-la poe ke li ke pol\-txe san oe za\-sya\-'u.} It just occurred to me that she hasn't yet said she's coming.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} 
\B{Fay\-lì\-'u su\-nu ay\-ngar a fì\-'ul oe\-ti nit\-ram sley\-ke\-i\-u nì\-txan.} That you all like these words makes me very happy.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/comment-page-1/\#comment-1443}{b} 
(\ding{43}~\HL{fI'u}{\B{fì\-'u}}, \emph{nominalizer of} \ding{43}~\HL{tsa'u}{\B{tsa\-'u}}, \emph{i.e.} \HL{tsala}{\B{tsa\-la}} \emph{exists but is rarely in use})

% fura
\hi
\HT{fura}{\B{\cp{fu}ra}}, \B{fì\-\cp{'ur} a}, \emph{or} \B{a fì\-\cp{'ur}} \I{["fu.Ra]} \emph{conj.} that (\emph{to introduce a dative subordinate clause}) 
\B{Fì\-'u\-pxa\-re ka\-ngay si fu\-ra po za\-ya\-'u.} This message confirms that he will come.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
(\ding{43}~\HL{fI'u}{\B{fì\-'u}}, \HL{tsara}{\B{tsa\-ra}} \emph{exists but is rarely in use})

% furia
\hi
\HT{furia}{\B{\cp{fu}ria}}, \B{fì\-\cp{'u}\-ri a}, \emph{or} \B{a fì\-\cp{'u}\-ri} \I{["fu.Ri.a]} \emph{conj.} that (\emph{to nominalize a topic subordinate clause}) 
\B{Fu\-ri\-a nì\-'Ìng\-lì\-sì pam\-rel si\-vi, oe\-ru txoa li\-vu.} Forgive me that I write in English.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg139677/\#msg139677}{ln}
\B{Ngal oe\-yä 'u\-pxa\-ret ay\-su\-te\-ru fpo\-le' a fì\-'u\-ri, nga\-ru ir\-ayo sei\-yi oe nì\-txan!} Concerning the fact, that you sent my message to the people, I thank you very much!~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
(\ding{43}~\HL{fI'u}{\B{fì\-'u}}, \emph{nominalizer of} \ding{43}~\HL{tsa'u}{\B{tsa\-'u}}, \emph{i.e.} \HL{tsaria}{\B{tsa\-ri\-a}} \emph{exists but is rarely in use})

% futa
\hi
\HT{futa}{\B{\cp{fu}ta}}, \B{fì\-\cp{'ut} a}, \emph{or} \B{a fì\-\cp{'ut}} \I{["fu.ta]} \emph{or coll.} \I{[fta]} \emph{conj.} that (\emph{to nominalize a patientive subordinate clause}) 
(\emph{a. with vtr.}) \B{Ul\-te omum oel fu\-ta tì\-fya\-wìn\-txu\-ri oe\-yä pe\-rey ay\-nga nì\-wotx.} And I know that you all are waiting for my guidance.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\emph{b. with transitive modals}, requires ‹\B{iv}› in the dependent verb) \B{\mbox{Oel} new fu\-ta ta\-ron\-yu ki\-vä.} I want the hunter to go.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-email/}{b} 
\B{Po\-el sto\-la\-tso fu\-ta me\-fo ti\-va\-ron tsa\-ha'\-ngir.} She must have refused to hunt that afternoon.~\LINK{http://forum.learnnavi.org/language-updates/sto-has-the-same-syntax-as-new/msg567034/\#msg567034}{b} 
\B{(oel) nul\-new fu\-ta si\-vop oeng nì\-txi\-lu\-ke}, (I) prefer to travel leisurely.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{c. rarely written as} \B{fta}) \B{Ro\-lun oel olo'\-ti a 'e\-fu fta oe man tsa\-tseng\-mì.} I've found a community where I feel I belong.~\LINK{http://naviteri.org/2022/06/solalew-maul-zisita-half-the-year-is-over/}{b} 
(\ding{43}~\HL{fI'u}{\B{fì\-'u}}, \emph{nominalizer of} \ding{43}~\HL{tsa'u}{\B{tsa\-'u}}, \emph{i.e.} \HL{tsata}{\B{tsa\-ta}} \emph{exists but is rarely in use})

% fwa
\hi
\HT{fwa}{\B{fwa}}, \B{fì\-\cp{'u} a}, \emph{or} \B{a fì\-\cp{'u}} \I{[fwa]} \emph{conj.} that (\emph{to nominalize a subordinate clause}). (\emph{a. with adj.}) \B{Tsun oe nga\-hu nì\-Na'\-vi pi\-väng\-kxo a fì\-'u oe\-ru prr\-te' lu.} It's a pleasure to be able to chat with you in Na'vi.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki} 
\B{Swey lu fwa nga fì\-kem ke si\-vi nì\-sngum.} It's best that you not do this fretfully.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
(\emph{b. with vin.}) \B{Su\-nu oe\-ru fwa fì\-tseng\-it te\-rok oel nì\-mun.} I like that I'm here again.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/}{b} 
\B{Lam oer fwa po lu ka\-nu nì\-txan.} It seems to me that he is very clever.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{Toruk Makto po\-lä\-hem a fì\-'u ro\-lo\-'a nì\-txan Oma\-ti\-ka\-ya\-ru.} The arrival of \emph{Toruk Makto} made a great impression on the Omatikaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{c. to introduce verb phrases}) \B{Oe\-yä txin\-tìn lu fwa stä'\-nì fay\-o\-ang\-it.} My central societal occupation is to catch fish.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\emph{d. with} \B{to} \emph{to compare verb phrases}) \B{Ftu\-e lu fwa ta\-ron ngong\-a i\-o\-ang\-it to fwa ta\-ron pum\-it a lu wa\-lak sì win.} It's easier to hunt lethargic animals than to hunt perky, speedy ones.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
(\emph{e. with adp.s to introduce a verb phrase}) \B{Sre fwa sìn tskxe\-pay tì\-ran, zene fko flì\-nutx\-it sti\-ve\-ftxaw.} It's necessary to check the thickness of the ice before walking on it.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 
(\emph{f. with} \B{sweylu} \emph{to mean}) `should have' \B{Swey\-lu fwa nga ko\-lä.} You should have gone.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
(\ding{43}~\HL{fI'u}{\B{fì\-'u}}, \emph{nominalizer of} \ding{43}~\HL{tsa'u}{\B{tsa\-'u}}, \emph{i.e.} \HL{tsawa}{\B{tsawa}} \emph{exists but is rarely in use})

% fwal
\hi 
\HT{fwal}{\B{fwal}} \I{[fwal]} \emph{vtr.} wipe. 
\B{Yo\-ti fwal ru\-txe. Mei slo\-lu maw tom\-pa.} Please wipe the table. It's gotten wet after the rain.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}

% fwampop
\hi
\HT{fwampop}{\B{\cp{fwam}pop}} \I{["fwam.pop\textcorner]} \emph{n., fauna} tapirus. \B{Oel tso\-le\-'a fwam\-pop\-it a\-win fra\-to.} I saw the fastest Tapirus.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Fwam\-pop\-ä tem\-rey tsa\-tseng\-mì ngä\-zìk lu nì\-wotx.} It's very hard for a tapirus to survive there.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\emph{a. idiom}) \B{(na) fwam\-pop fkip fì\-wopx}, ``Fish out of water.'' [lit.: (like a) tapirus in the clouds].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% fwang
\hi
\HT{fwang}{\B{fwang}} \I{[fwaN]} \emph{adj.} (\emph{of taste and smell}) savory, umami, rich.

% fwäkì
\hi
\HT{fwAkI}{\B{\cp{fwä}kì}} \I{["fw\ae.kI]} \emph{n., fauna} mantis.
\B{Sro\-ler fwä\-kì ftu\-opa tskxe.} A mantis appeared from behind a rock.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\emph{a. proverb}) \B{Fwä\-kì ke fwe\-fwi.} It's against s.o.'s\slash s.t.'s nature [lit.: a mantis doesn't whistle].~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/comment-page-1/\#comment-1143}{b} \\
\textsc{Derivations:} \ding{43} \on{1}. \HL{fwAkIwll}{\B{fwäkìwll}}, mantic orchid.

% fwäkìwll
\hi
\HT{fwAkIwll}{\B{\cp{fwä}kìwll}} \I{["fw\ae.kI.w\s{l}]} \emph{n.}~\ding{96} mantis orchid, \emph{pandorchidus insectoralis}. 
\B{Ra\-lul ngo\-lop 'um\-tsat fa ay\-syu\-lang fwä\-kì\-wll\-ä.} Ralu made medicine from flowers of the Mantis orchid.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{fwAkI}{\B{fwä\-kì}}, \HL{'ewll}{\B{'e\-wll}})

% fwefwi
\hi
\HT{fwefwi}{\B{\cp{fwe}fwi}} \I{["fwE.fwi]} \emph{vin.} whistle.   
\B{Nga fwe\-fwi nìl\-ke\-ftang a fì\-'u lu sä\-srätx a\-txan oe\-ru.} It's a major annoyance to me that you whistle continuously.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{nIfwefwi}{\B{nìfwefwi}}, by whistling.

% fwew
\hi
\HT{fwew}{\B{fwew}} \I{[fwEw]} \emph{vtr.} look for, seek. \B{Ngal fwe\-rew a tu\-te ke fkey\-tok.} The person you're looking for doesn't exist.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b} 
\B{Tsa\-'u\-pxa\-ret fwo\-lew oel slä ke tsä\-ngun ri\-vun.} I looked for that message but I couldn't find it.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b}
(\ding{214}~\HL{run}{\B{run}})

% fwel
\hi
\HT{fwel}{\B{fwel}} \I{[fwEl]} \emph{adj.} broken. 

% fwem
\hi
\HT{fwem}{\B{fwem}} \I{[fwEm]} \emph{adj.} (\emph{of a point}) dull, blunt. 
\B{Ko\-ak\-tan\-ä ay\-sre' lä\-ngu fwem.} An old man's teeth are dull.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
\B{Tsa\-tstal a\-fwem lu litx nì\-txan.} That blunt knife is very sharp.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
(\ding{214}~\HL{pxi}{\B{pxi}}, \ding{43}~\HL{tete}{\B{te\-te}})

% fwep
\hi 
\HT{fwep}{\B{fwep}} \I{[fwEp\textcorner]} \emph{n.} dust (\emph{on a surface}).
\B{Slär\-ìl nga' fwep\-it a\-txan.} The cave is very dusty.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Txe\-wìl 'o\-la\-ku fwep\-it ftum\-fa kel\-ku sne\-yä.} Txewì removed the dust from inside his house.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fwopx}{\B{fwopx}}, dust cloud, dust (in the air).

% fwi
\hi 
\HT{fwi}{\B{fwi}} \I{[fwi]} \emph{vin.} slip, slide.
\B{Na\-ri si! Kll\-te\-sìn lu pay a\-txan. Fwi rä\-'ä!} Be careful! There’s a lot of water on the ground. Don't slip!~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Txur\-tel\-mì fo fta so\-li fte\-ke ka tsyokx fwi\-vi.} They tied a knot in the rope so it wouldn't slip through their hands.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Po\-ti fwey\-ko\-li ay\-oel ìlä rong.} We let him slide through the tunnel.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% fwìng / si
\hi 
\HT{fwIng}{\B{fwìng}} \I{[fwIN]} \textsc{i}. \emph{n.} humiliation, embarrassment, loss of face. 
\B{Ra\-lu\-ri fwa tì\-fme\-tok\-it ke em\-zo\-la\-'u lä\-ngu fwìng a\-txan.} Ra\-lu's not passing the test was a great humiliation (to him).~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
\B{Peyä hem\-ìl a\-ke\-muia\-nga' za\-mo\-lu\-nge fwìng\-it ay\-oer.} His dishonorable behavior (lit.: actions) humiliated us (lit.: brought humiliation to us).~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b} \\
\textsc{ii}. \B{fwìng si} \I{[fwIN si]} \emph{vin.} humiliate. \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{txanfwIngtu}{\B{txanfwìngtu}}, bastard, loser (\emph{insult}).

% fwopx
\hi 
\HT{fwopx}{\B{fwopx}} \I{[fwop']} \emph{n.} dust (\emph{in the air}).
\B{Tsaw\-ke slo\-lu vawm ta\-lun fwopx.} The sun became dark due to dust in the air.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\ding{43}~\HL{fwep}{\B{fwep}}, \HL{pIwopx}{\B{pì\-wopx}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{eanfwopx}{\B{ean\-fwopx}}, mist bloom.

% fwum
\hi 
\HT{fwum}{\B{fwum}} \I{[fwum]} \emph{or} \I{[fwUm]} \emph{vin.} float (\emph{on the surface of a liquid}). 
\B{Me\-rìk a\-lor pay\-sìn fwar\-mum.} Two beautiful leaves were floating on the water.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
\B{Ke omum teyng\-ta fì\-u\-ran a\-ku'\-up fu fwa\-yum fu wa\-yapx.} I don't know (or: It's not known) whether this heavy boat will float or sink.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}

% fya'o
\hi
\HT{fya'o}{\B{\cp{fya}'o}} \I{["fja.Po]} \emph{n.} path, way; method, manner. \B{Fol lì'\-fya\-ti aw\-nge\-yä sar a fya\-'o lu ko\-sman.} It's wonderful the way they use our language.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \B{pa\-ye\-'un swey\-a fya\-'ot} I will decide the best way.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} \\  
\textsc{Derivations}: \ding{43} \on{1}.~\HL{fyape}{\B{fyape}\slash \B{pefya}}, how?   
\on{2}.~\HL{nIfya'o}{\B{nìfya'o}}, in a manner (that). 
\on{3}.~\HL{eyawrfya}{\B{eyawrfya}}, righ way, correct path. 
\on{4}.~\HL{fyalI'u}{\B{fya\-lì\-'u}}, adverb. 
\on{5}.~\HL{yo'kofya}{\B{yo'\-ko\-fya}}, cycle.

% fyalì'u
\hi 
\HT{fyalI'u}{\B{\cp{fya}lì'u}} \I{["fja.lI.Pu]} \emph{n.} adverb. 
(\ding{43}~\HL{fya'o}{\B{fya\-'o}}, \HL{lI'u}{\B{lì\-'u}})

% fyan
\hi
\HT{fyan}{\B{fyan}} \I{[fjan]} \emph{n.} \emph{constructed device for keeping s.t. off the ground and clean}. \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fyanyo}{\B{fyanyo}}, table. 
\on{2}.~\HL{fyanyI}{\B{fyanyì}}, shelf. 
\on{3}.~\HL{kurfyan}{\B{kurfyan}}, hamper, suspended rack.

% fyanyì
\hi
\HT{fyanyI}{\B{\cp{fyan}yì}} \I{["fjan.jI]} \emph{n.} shelf.
(\ding{43}~\HL{fyan}{\B{fyan}}, \HL{yI}{\B{yì}})

% fyanyo
\hi
\HT{fyanyo}{\B{\cp{fyan}yo}} \I{["fjan.jo]} \emph{n.} table, elevated utilitarian surface; \emph{coll.}: \ding{43}~\HL{yo}{\B{yo}}.
\B{Oe\-yä tstal\-ìl tok pe\-se\-nget? -- Lu yo\-sìn.} Where's my knife? -- It's on the table.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\ding{43}~\HL{fyan}{\B{fyan}}, \HL{yo}{\B{yo}})

% fyape
\hi
\HT{fyape}{\B{\cp{fya}pe}} \emph{or} \B{pe\cp{fya}} \I{["fja.pE]} \emph{or} \I{[pE."fja]} \emph{inter.} how?   
\B{Fya\-pe fko syaw ngar?} What's your name?~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln} 
\B{'Ìn nga fya\-pe nì\-fkrr?} What's been keeping you busy lately?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
\B{Ya\-fkeyk za\-'u fya\-pe?} How's the weather?~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Mak\-to fya\-pe?} ``How are things? How's everything going? How are you doing?'
\B{A: Mak\-to fya\-pe?---B: Zong. Nga\-ri tut?---A: Nìk\-sran. Oe\-ru lu fpom, slä oey 'i\-tan lu spxin.} A: How's everything?---B: Good. You?---A: So\hyp{}so. I'm fine, but my son is sick.~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b} 

% fyawìntxu
\hi
\HT{fyawIntxu}{\B{fyawìn\cp{txu}}} \I{[fja.w$\cdot$In."t'$\cdot$u]} \emph{vin., inf.}\on{23} guide. \B{oe new nì\-txan ay\-nga\-ru fya\-wi\-vìn\-txu}, I want to guide you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\ding{43}~\HL{fya'o}{\B{fya\-'o}}, \HL{wIntxu}{\B{wìn\-txu}}) 
\on{1}.~\HL{tIfyawIntxu}{\B{tìfyawìntxu}}, guidance. 
\on{2}.~\HL{fyawIntxuyu}{\B{fyawìntxuyu}}, guide.

% fyawìntxuyu
\hi 
\HT{fyawIntxuyu}{\B{fyawìn\cp{txu}yu}} \I{[fja.wIn."t'u.ju]} \emph{n.} guide (\emph{of a person}). 
(\ding{43}~\HL{fyawIntxu}{\B{fya\-wìn\-txu}}, \HL{-yu}{\B{-yu}})

% fyel
\hi
\HT{fyel}{\B{fyel}} \I{[fjEl]} \emph{vtr.} seal, cement, make impervious (\B{wä}, against) (\emph{making something tight and secure, so that it is unlikely (and not intended) to be broken}). 
\B{Txo fkol ke fyi\-vel u\-ran\-it pay\-wä, ze\-ne fko sli\-ve\-le.} If one does not seal a boat against water, one must swim.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}

% fyeng
\hi 
\HT{fyeng}{\B{fyeng}} \I{[fjEN]} \emph{adj.} steep. 
\B{Na\-ri si! Fay\-ram\-tsyìp lu fyeng.} Be careful! These hills are steep.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}

% fyengkekin
\hi 
\HT{fyengkekin}{\B{\cp{fyeng}kekin}} \I{["fjEN.kE.kin]} \emph{conv.} `no need to use honorific language' (\emph{from} \B{lì'fyeng kelkin}). 
(\ding{43}~\HL{lI'fyeng}{\B{lì'\-fyeng}}, \HL{kelkin}{\B{kel\-kin}}) 

% fyep
\hi
\HT{fyep}{\B{fyep}} \I{[fjEp\textcorner]} \emph{vtr.} (\emph{a.}) hold in the hand, grasp, grip. \B{Ngä\-zìk lu fwa var tsko\-ti fyi\-vep teng\-krr ut\-ral\-it tsye\-rìl.} It's hard to keep holding a bow while climbing a tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
(\emph{b. general holding}) \B{Oel tstal\-it fyo\-lep fa ay\-sre'.} I held the knife in my teeth.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{sAfyep}{\B{säfyep}}, handle, grip.

% fyeyn
\hi
\HT{fyeyn}{\B{fyeyn}} \I{[fjEjn]} \emph{adj.} ripe, mature, adult. \\  
\on{1}. \HL{fyeyntu}{\B{fyeyntu}}, adult person.

% fyeyntu
\hi
\HT{fyeyntu}{\B{\cp{fyeyn}tu}} \I{[fjEjn.tu]} \emph{n.} adult person. 
\B{Le\-hrr\-ap lu fwa evi\-tsyìp sle\-le mì hil\-van lu\-ke fwa fyeyn\-tu te\-rìng nari.} It's dangerous for tiny ones to swim in the river without an adult watching.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} 
\B{Nì\-ran lu Lo\-ak mi 'e\-veng slä tsun ti\-va\-ron nì\-teng\-fya na fyeyn\-tu.} Loak is still really just a boy but he can hunt the same as an adult.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\ding{43}~\HL{fyeyn}{\B{fyeyn}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{nIfyeyntu}{\B{nìfyeyntu}}, maturely.

% fyin
\hi
\HT{fyin}{\B{fyin}} \I{[fjin]} \emph{adj.} simple. 
\B{Ta\-fral kì\-syar a ay\-sä\-num\-vi sì ay\-lì'\-fya\-vi lu fyin.} That's why I'll teach lessons and expressions that are simple.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} 
(\ding{214}~\HL{ep'ang}{\B{ep\-'ang}}) \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fyinep'ang}{\B{fyinep'ang}}, complexity.

% fyinep'ang / fyinep'angpe
\hi
\HT{fyinep'ang}{\B{fyinep\cp{'ang}}} \I{[fjin.Ep\textcorner."PaN]} \textsc{i}. \emph{n.} degree of complexity. \\  
\textsc{ii}. \B{fyinep\cp{'ang}pe} \I{[fjin.Ep\textcorner."PaN]} \emph{inter.} how complex? 
\B{Ngal ke tslä\-ngam teyng\-ta fì\-tì\-ngä\-zìk\-ì\-ri pe\-fyin\-ep\-'ang.} Unfortunately you don't understand how complex this problem is.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{fyin}{\B{fyin}}, \HL{ep'ang}{\B{ep\-'ang}})

% fyìp
\hi 
\HT{fyIp}{\B{fyìp}} \I{[fjIp\textcorner]} \emph{n.} tendril, tentacle. \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fyIpmaut}{\B{fyìpmaut}}, squid fruit tree.

% fyìpmaut
\hi 
\HT{fyIpmaut}{\B{fyìpmaut}} \I{[fjIp\textcorner.ma.ut\textcorner]} \emph{n.} \ding{96} squid fruit tree, \emph{octocrus folliculus}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln} 
(\ding{43}~\HL{fyIp}{\B{fyìp}}, \HL{mauti}{\B{mau\-ti}}, \HL{utral}{\B{ut\-ral}})

% fyole
\hi
\HT{fyole}{\B{\cp{fyo}le}} \I{["fjo.lE]} \emph{adj.} sublime, beyond perfection. 
\B{Ran tì\-ru\-sol\-ä pe\-yä lu fyo\-le.} The \emph{ran} of her singing is sublime.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Nge\-yä tì\-kang\-kem a\-fyo\-le me\-u\-i\-a le\-i\-u lì'\-fya\-o\-lo'\-ru, ma tsmu\-kan.} Your sublime work honors the language clan, brother.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} \\
\textsc{Derivations:} \ding{43} \on{1}.~\HL{fyolup}{\B{fyolup}}, exquisite, sublime in style.

% fyolup
\hi
\HT{fyolup}{\B{\cp{fyo}lup}} \I{["fjo.lup\textcorner]} \emph{or} \I{[.lUp\textcorner]} (\textsc{rn}: \B{\cp{fyo}lùp} \I{["fjo.lUp\textcorner]}) \emph{adj.} exquisite, sublime in style.
\B{Fì\-tsko\-ti a\-fyo\-lup 'o\-ngo\-lop oe\-yä sem\-pul\-ìl.} This exquisite bow was designed by my father.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Fyo\-lup\-a ay\-sä\-ngop ay\-nge\-yä oe\-ru te\-ya si nì\-ngay.} Your exquisite creations fill me greatly with joy.~\LINK{http://naviteri.org/2014/12/ayngaru-tsinga-yoratu-may-i-present-the-four-winners/}{b}
(\ding{43}~\HL{lupra}{\B{lup\-ra}}, \HL{fyole}{\B{fyo\-le}}) 

% fyuatx
\hi 
\HT{fyuatx}{\B{fyu\cp{atx}}} \I{[fju."{}at']} \emph{n., fauna} anemonoid.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln} 

% fyufye
\hi 
\HT{fyufye}{\B{\cp{fyu}fye}} \I{["fju.fjE]} (\textsc{rn}: \B{\cp{fyù}fye} \I{["fjU.fjE]}) \emph{vin.} splash. 
\B{Ra\-nu kil\-van\-mì fyar\-mu\-fye na tsawl\-a pa\-yo\-ang a\-pì\-saw.} Ra\-nu was splashing in the river like a big clumsy fish.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}

\end{multicols}