\pdfbookmark[0]{H}{H}
\begin{center}
\section*{H \I{[h]} \emph{Hä}}
\end{center}

\begin{multicols}{2}

% ha'
\hi
\HT{ha'}{\B{ha'}} \I{[haP]} \emph{vin.} fit, suit, complement, inherently enhance (two entities (things, people, situations, \dots) fit or suit each other; they ``go together'' well).  
\B{Tse\-nu sì Lo\-ak fì\-tsap ke ha' kaw\-'it.} Tse\-nu and Lo\-ak are a terrible match for each other.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Tse\-nu Lo\-ak\-ur ke ha'.} Tse\-nu is a bad match for Lo\-ak.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Tse\-nur Lo\-ak ke hä\-nga'.} Lo\-ak isn't good for Tse\-nu.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Hu\-fwa nge\-yä tì\-hawl\-ìri ke lu kea kxey\-ey, tsal\-su\-ngay oe\-ru ke ha' nì\-tam.} Although there's nothing wrong with your plan, it just doesn't suit me.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Ngay. Tsa\-'o\-pin hek nì\-'it, slä su\-nu oer, ha ha'.} True. That color is bit odd, but I like it, so it's a good fit for me.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivation}: \on{1}.~\HL{leha'}{\B{le\-ha'}}, suitable, appropriate, fitting.

% ha'ngir
\hi
\HT{ha'ngir}{\B{ha'\cp{ngir}}} \I{[haP."NiR]} \emph{n., adv.} afternoon. \B{Po\-el sto\-la\-tso fu\-ta me\-fo ti\-va\-ron tsa\-ha'\-ngir.} She must have refused their request to hunt that afternoon.~\LINK{http://forum.learnnavi.org/language-updates/sto-has-the-same-syntax-as-new/msg567034/\#msg567034}{ln} \\
(I.) \B{ha'ngi\cp{ram}} \I{[haP.Ni."Ram]} \emph{n., adv.} yesterday afternoon. \\ 
(II.) \B{ha'ngi\cp{ray}} \I{[haP.Ni."Raj]} \emph{n., adv.} tomorrow afternoon. 

% ha
\hi
\HT{ha}{\B{ha}} \I{[ha]} \emph{conj.} so, in that case. 
\B{Tom\-pa 'e\-ko nì\-hawng, ha ze\-ne aw\-nga kll\-wi\-vo.} The rain is too strong, so we must land.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}

% hafyon
\hi 
\HT{hafyon}{\B{ha\cp{fyon}}} \I{[ha."fjon]} \emph{n.} wisdom.
\B{Nge\-nge\-yä ha\-fyon\-ìri ira\-yo, ma le\-i\-o\-ko\-ak\-te.} Thank you for your wisdom, respected female elder.~\LINK{https://naviteri.org/2025/03/conlang-adventure-and-more/}{b} 
(\ding{43}~\HL{tItxantslusam}{\B{tìtxantslusam}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lafyon}{\B{la\-fyon}}, wise. 
\on{2}.~\HL{hafyonga'}{\B{ha\-fyo\-nga'}}, wise.

% hafyonga'
\hi 
\HT{hafyonga'}{\B{ha\cp{fyo}nga'}} \I{[ha."fjo.NaP]} \emph{adj., nfp.} wise. 
\B{ay\-lì\-'u a\-ha\-fyo\-nga'}, wise words.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\HL{hafyon}{\B{hafyon}}, \HL{lafyon}{\B{lafyon}})

% hahaw
\hi
\HT{hahaw}{\B{\cp{ha}haw}} \I{["ha.haw]} \emph{vin.} sleep. 
(\emph{a.}) \B{Ney\-ti\-ri he\-ra\-haw.} Ney\-ti\-ri is sleeping.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#MSNBC}{wiki}  
\B{Sem\-pul lì'\-fya\-yä har\-ma\-haw.} The father of the language was sleeping.~\LINK{http://forum.learnnavi.org/language-updates/a-long-answer-to-a-quick-question/msg155104/\#msg155104}{ln} 
\B{Lu na'\-rìng tsuk\-ha\-haw.} It's possible to sleep in the forest.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b} 
(\emph{b. idiom}) \B{Hi\-va\-haw nìm\-wey.} Goodnight (\emph{when it's bedtime}). 
(\emph{c. proverb}) \B{Ha\-haw nì\-'aw txo pa\-lu\-kan smi\-von ngar.} ``Only sleep if you are familiar with the Thanator,'' i.e. \emph{don't think you're safe unless you're aware of the danger}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{hawtsyIp}{\B{haw\-tsyìp}}, nap. 
\on{2}.~\HL{tsukhahaw}{\B{tsuk\-ha\-haw}}, `sleepable', able to sleep in. 
\on{3}.~\HL{snomo}{\B{sno\-mo a ha\-haw}}, bedroom, area to sleep.
\on{4}.~\HL{haway}{\B{ha\-way}}, lullaby. 
\on{5}.~\HL{yehaw}{\B{ye\-haw}}, well\hyp{}rested.

% ham
\hi
\HT{ham}{\B{ham}} \I{[ham]} \emph{adj.} last, previous. 
(\ding{214}~\HL{hay}{\B{hay}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hamalo}{\B{ha\-ma\-lo}}, last time.

% hamalo
\hi
\HT{hamalo}{\B{ha\cp{ma}lo}} \I{[ha."ma.lo]} \emph{adv.} last time. 
(\ding{214}~\HL{hayalo}{\B{hayalo}}, \ding{43}~\HL{}{\B{ham}}, \HL{alo}{\B{alo}}) 

% hametsì
% \hi
% \B{ha\cp{me}tsì} \I{[ha."mE.\t{ts}I]} \emph{n.}, \textsc{lw} Chametz (\emph{leavened bread}). \B{aw\-ngal yom ha\-me\-tsìt, yom ma\-tsat, ke tsran\-ten}, it doesn't matter whether we eat Chametz or Matzah.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{this night}

% han
\hi 
\HT{han}{\B{han}} \I{[han]} \emph{vtr.} lose (\emph{not having s.t. you once had}). \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIhah}{\B{tì\-han}}, loss.

% hangham
\hi
\HT{hangham}{\B{\cp{hang}ham}} \I{["haN.ham]} \emph{vin.} laugh. 
\B{Oe ho\-lang\-ham nì\-teng.} I laughed too.~\LINK{http://masempul.org/2010/07/\%E2\%80\%99ite-polaheiem/}{sempul} 
\B{Po\-leng Ney\-ti\-ril hang\-vur\-it a fra\-pot hey\-kang\-ham.} Ney\-ti\-ri told a joke that made everyone laugh.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}\\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{hangvur}{\B{hang\-vur}}, joke, funny story.
\on{2}.~\HL{tIhangham}{\B{tì\-hang\-ham}}, laughter.
\on{3}.~\HL{lehangham}{\B{le\-hang\-ham}}, laughable, ridiculous, absurd.

% hangvur
\hi
\HT{hangvur}{\B{\cp{hang}vur}} \I{["haN.vuR]} \emph{or} \I{[.vUR]} \emph{n.} joke, funny story. 
\B{Po\-leng Ney\-ti\-ril hang\-vur\-it a fra\-pot hey\-kang\-ham.} Neytiri told a joke that made everyone laugh.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{hangham}{\B{hang\-ham}}, \HL{vur}{\B{vur}})

% hapxì
\hi
\HT{hapxI}{\B{ha\cp{pxì}}} \I{[ha."p'I]} \emph{n.} part. 
\B{Nga\-ri hu Ey\-wa sa\-lew ti\-re\-a, tokx 'ì'awn slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body remains to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{a\slash wiki} 
\B{Pì\-lok\-ä fì\-ha\-pxì\-yä tì\-kan lu law.} The aim of this part of the blog is clear.~\LINK{http://naviteri.org/2010/06/first-post/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hapxItu}{\B{ha\-pxì\-tu}}, member.   
\on{2}.~\HL{kApxI}{\B{kä\-pxì}}, rear (part\slash section).   
\on{3}.~\HL{zapxI}{\B{za\-pxì}}, front (part\slash section).   
\on{4}.~\HL{hImpxI}{\B{hìm\-pxì}}, minority. 
\on{5}.~\HL{txampxI}{\B{txam\-pxì}}, majority.   
\on{6}.~\HL{vospxI}{\B{vo\-spxì}}, month. 
\on{7}.~\HL{pxI}{\B{\mbox{-pxì}}}, \emph{to form fractions}. 
\on{8}.~\HL{pxImun'i}{\B{pxì\-mun\-'i}}, devide.
\on{9}.~\HL{swapxI}{\B{swa\-pxì}}, family member.
\on{10}.~\HL{napxI}{\B{na\-pxì}}, partially, in part.

% hapxìtu
\hi
\HT{hapxItu}{\B{ha\cp{pxì}tu}} \I{[ha."p'I.tu]} \emph{n.} member. 
\B{sìl\-pey oe tsnì ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ke stì\-ye\-vi.} I hope that the members of the language clan won't be angry.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{Me\-ha\-pxì\-tul lì'\-fya\-o\-lo'\-ä pxe\-stxe\-lit a\-lor to\-lìng aw\-nga\-ru.} Two members of the language clan have given us three beautiful gifts.~\LINK{http://naviteri.org/2010/12/beautiful-christmas-carols-sung-in-navi/}{b} 
\B{Ul\-te pi\-veng ru\-txe ay\-ha\-pxì\-tur a\-la\-he san se\-i\-yi i\-ra\-yo sìk nì\-teng.} And please tell the other members `thank you', too. 
\B{Va'\-rul zä\-nget ik\-ran\-it sn\-eyä pxel ha\-pxì\-tu so\-a\-i\-ä.} Va'\-ru treats his ikran like a member of the family.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\ding{43}~\HL{hapxI}{\B{hapxì}})

% hasa'
\hi 
\HT{hasa'}{\B{ha\cp{sa'}}} \I{[ha."{}saP]} \emph{n.} orphan. 
\B{Krr\-ka tsam slä\-ngu pxay\-a 'e\-veng ha\-sa'.} During war, sad to say, many children become orphans.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 

% hasey / si
\hi
\HT{hasey}{\B{ha\cp{sey}}} \I{[ha."{}sEj]} \textsc{i}.~\emph{adj.} done, finished. \B{Hu Saw\-tu\-te a tsam lu ha\-sey.} The war with the Skypeople is finished.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \\
\textsc{ii}. \HT{hasey si}{\B{ha\cp{sey} si}} \I{[ha."{}sEj si]} \emph{vin.} accomplish, bring to a conclusion. 
\B{Nì\-'i'\-a po tsa\-tì\-kang\-kem\-vir ha\-sey so\-li.} She finally completed the project.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} 
\B{Sä\-tsway\-on a\-'aw\-ve tsa\-heyl\-ur ha\-sey si; ke tsun nga pi\-vey.} First flight seals the bond; you cannot wait.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Ha\-sey si fu\-ra yom!} Finish eating!~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\emph{a. idiom}) \B{Sey\-so\-nìl\-\cp{tsan}!} \I{[sEj.so.nIl."\t{ts}an]} Well done! (\emph{from} \B{ha\-sey so\-li nìl\-tsan})~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} (\ding{43}~\HL{mu'ni}{\B{mu'\-ni}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{seysonIltsan}{\B{sey\-so\-nìl\-tsan}}, well done, bravo.

% haway
\hi 
\HT{haway}{\B{\cp{ha}way}} \I{["ha.waj]} \emph{n.} lullaby. 
(\ding{43}~\HL{hahaw}{\B{ha\-haw}}, \HL{way}{\B{way}})

% hawl
\hi
\HT{hawl}{\B{hawl}} \I{[hawl]} \emph{vtr.} prepare. 
\B{Ta\-fral ho\-lawl ay\-oel ay\-nga\-fpi 'a\-'aw\-a tì\-päng\-kxo\-tsyìp\-it.} That's why we've prepared several little dialogs for you.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIhawl}{\B{tì\-hawl}}, preparation, plan.

% hawmpam
\hi
\HT{hawmpam}{\B{\cp{hawm}pam}} \I{["hawm.pam]} \emph{n.} noise (\emph{sound that is excessive, unnecessary, inappropriate, unexpected, or startling})
\B{Fì\-hawm\-pam pe\-lun, ma 'itan? Fni\-vu set!} Why all this noise, son? Be quiet now!~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{vApam}{\B{vä\-pam}}, \HL{hawng}{\B{hawng}}, \HL{pam}{\B{pam}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lehawmpam}{\B{le\-hawm\-pam}}, noisy. 
\on{2}.~\HL{nIhawmpam}{\B{nì\-hawm\-pam}}, noisily.

% hawntu
\hi 
\HT{hawntu}{\B{\cp{hawn}tu}} \I{["hawn.tu]} \emph{n.} one under s.o.'s protection, protected person. 
\B{Oey yawn\-tu lu oey hawn\-tu.} My beloved is under my protection.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}
(\ding{43}~\HL{hawnu}{\B{hawnu}})

% hawntsyokx
\hi
\HT{hawntsyokx}{\B{hawn\cp{tsyokx}}} \I{[hawn."\t{ts}jok']} \emph{n.} glove. (\emph{plurals in all but very careful or ceremonial conversation usually pronounced} \B{me\-hawn\-\cp{tsyokx}} \I{[mawn."\t{ts}jok']}, \B{pxe\-hawn\-\cp{tsyokx}} \I{[p'awn."\t{ts}jok']}, \B{ay\-hawn\-\cp{tsyokx}} \I{[aj.awn."\t{ts}jok']})
(\ding{43}~\HL{hawnu}{\B{haw\-nu}}, \HL{tsyokx}{\B{tsyokx}})

% hawnu
\hi
\HT{hawnu}{\B{\cp{haw}nu}} \I{["haw.nu]} \emph{vtr.} protect, shelter (\B{wä}, from). 
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{tIhawnu}{\B{tì\-haw\-nu si}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIhawnu}{\B{tì\-haw\-nu}}, protection. 
\on{2}.~\HL{hawntsyokx}{\B{hawn\-tsyokx}}, glove. 
\on{3}.~\HL{hawnven}{\B{hawn\-ven}}, shoe. 
\on{4}.~\HL{hawre'}{\B{haw\-re'}}, hat.
\on{5}.~\HL{hawntu}{\B{hawn\-tu}}, protected person.

% hawnven
\hi
\HT{hawnven}{\B{hawn\cp{ven}}} \I{[hawn."vEn]} \emph{n.} shoe. (\emph{plurals in all but very careful or ceremonial conversation usually pronounced} \B{me\-hawn\-\cp{ven}} \I{[mawn."vEn]}, \B{pxe\-hawn\-\cp{ven}} \I{[p'awn."vEn]}, \B{ay\-hawn\-\cp{ven}} \I{[aj.awn."vEn]}) 
\B{Rutxe me\-hawn\-ven\-it 'i\-va\-ku.} Please take off your shoes.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
\B{Hawn\-ven\-ì\-ri lu oe\-ru mun\-sna a\-mrr.} I have five pairs of shoes.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{hawnu}{\B{haw\-nu}}, \HL{venu}{\B{ve\-nu}})

% hawng
\hi
\HT{hawng}{\B{hawng}} \I{[hawN]} \emph{n.} overabundance. \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIhawng}{\B{nìhawng}}, too, excessively. 
\on{2}.~\HL{lehawng}{\B{lehawng}}, excessive. 
\on{3}.~\HL{hawngkrr}{\B{hawngkrr}}, late (\emph{adv.}). 
\on{4}.~\HL{hawmpam}{\B{hawmpam}}, noise.

% hawngkrr
\hi
\HT{hawngkrr}{\B{\cp{hawng}krr}} \I{["hawN.k\s{r}]} \emph{adv.} late.   
\B{Hawng\-krr rä\-'ä zi\-va\-'u!} Don't come late!~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\ding{214}~\HL{ye'krr}{\B{ye'krr}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lehawngkrr}{\B{lehawngkrr}}, late (\emph{adj.}).

% hawre'
\hi
\HT{hawre'}{\B{haw\cp{re'}}} \I{[haw."REP]} \emph{n.} hat. (\emph{plurals in all but very careful or ceremonial conversation usually pronounced} \B{me\-haw\-\cp{re'}} \I{[maw."REP]}, \B{pxe\-haw\-\cp{re'}} \I{[p'aw."REP]}, \B{ay\-haw\-\cp{re'}} \I{[aj.aw."REP]}) 
\B{Su\-nu oer haw\-re' a ngal yo\-lem\-stokx.} I like the hat you’re wearing.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
\B{Fì\-re\-won ngal lum\-pe kea ha\-wre'\-it ke yo\-lem\-stokx?} Why didn’t you put on a hat this morning?~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
(\ding{43}~\HL{hawnu}{\B{haw\-nu}}, \HL{re'o}{\B{re\-'o}})

% hawtsyìp
\hi
\HT{hawtsyIp}{\B{\cp{haw}tsyìp}} \I{["haw.\t{ts}jIp\textcorner]} \emph{n.} nap. 
\B{Oel new fu\-ta li\-vu oer set haw\-tsyìp.} I want to take a nap now.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
\B{Sa'\-nur le\-ru haw\-tsyìp.} Mommy is taking a nap.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{hahaw}{\B{ha\-haw}})

% hay
\hi
\HT{hay}{\B{hay}} \I{[haj]} \emph{adj.} next, following. 
\B{Sìl\-pey oe, tsun oeng ul\-txa si\-vi ro hay\-a ul\-txa!} I hope, we can meet at the next meeting.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/comment-page-1/\#comment-844}{b}   
\B{Alo a\-hay oe za\-sya\-'u mì re\-won}, next time I'll come in the morning.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
(\ding{214}~\HL{ham}{\B{ham}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hayalo}{\B{ha\-ya\-lo}}, next time.

% hayalo
\hi
\HT{hayalo}{\B{ha\cp{ya}lo}} \I{[ha."ja.lo]} \emph{adv.} next time. (\emph{a. idiom}) \B{ha\-ya\-lo ta oe} \emph{or} \B{hay\-a\-lo oe\-ta}~\I{[ha."ja.lo "wE.ta]}. You're welcome [lit.: next time from me] (\emph{phrase for responding to thanks}).~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} 
(\emph{b. idiom}) \B{slä ha\-ya\-lo a\-la\-he} ``\ldots but that is for another time'' (\emph{set phrase in storytelling})~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}  \\
(i.) \B{ha\-\cp{ya}\-lo\-vay} \I{[ha."ja.lo.vaj]}, until next time.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} 
(\ding{214}~\HL{hamalo}{\B{ha\-ma\-lo}}, \ding{43}~\HL{hay}{\B{hay}}, \HL{alo}{\B{alo}})

% hän
\hi
\HT{hAn}{\B{hän}} \I{[h\ae{}n]} \emph{n.} net; web.
\B{Tsun fko si\-var hän\-it fte pay\-oang\-it sti\-vä'\-nì.} One can use a net to catch a fish.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Fì\-hì\-'ang\-ìl txu\-la hän\-ti fte smar\-it syi\-vep.} This insect constructs a web to trap its prey.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}

% he'a
\hi 
\HT{he'a}{\B{\cp{he}'a}} \I{["hE.Pa]} \emph{vin.} cough. 
\B{Nga\-ri krr\-a he\-'a, swey\-lu txo kxa\-ru lew si\-vi.} When you cough, it's best to cover your mouth.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{sAhe'a}{\B{sä\-he\-'a}}, a cough.

% hefi
\hi
\HT{hefi}{\B{\cp{he}fi}} \I{["hE.fi]} \emph{vtr.} smell (-\emph{control}). 
\B{Oel he\-fi ye\-rik\-it!} I smell a hexapede!~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
(\ding{43}~\HL{syam}{\B{syam}}, \HL{ontu}{\B{tìng on\-tu}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hefitswo}{\B{he\-fi\-tswo}}, sense of smell.

% hefitswo
\hi
\HT{hefitswo}{\B{\cp{he}fitswo}} \I{["hE.fi.\t{ts}wo]} \emph{n.} sense of smell. 
(\ding{43}~\HL{hefi}{\B{he\-fi}})

% hek
\hi
\HT{hek}{\B{hek}} \I{[hEk\textcorner]} \emph{vin.} be curious, odd, strange, unexpected. 
\B{Nge\-yä sä\-fpìl Saw\-tu\-te\-teri he\-i\-ek oer nì\-txan.} Your idea about the Sky People is very interesting to me (because it seems unusual).~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Fo nì\-Na'\-vi pll\-txe nì\-fya\-'o a hek.} They speak Na'vi strangely.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{hu\-fwa lì\-'u\-pam hek nì\-'it}, although the pronunciation is a bit strange.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Ay\-ron\-srel pe\-yä hä\-ngek nì\-txan.} His imaginings are (unpleasantly) weird.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIhek}{\B{nì\-hek}}, strangely, oddly.

% hena
\hi
\HT{hena}{\B{\cp{he}na}} \I{["hE.na]} \emph{vtr.} carry.
\B{Ru\-txe hi\-ve\-na fì\-e\-pxang\-it fpi oe.} Please carry this stone jar for me.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{sAhena}{\B{sä\-he\-na}}, container, vessel, carrier. 
\on{2}.~\HL{kahena}{\B{ka\-he\-na}}, transport.

% henga / si
\hi 
\HT{henga}{\B{\cp{he}nga}} \I{["hE.Na]} \textsc{i}. \emph{n.} formally polite behavior. 
\B{He\-nga kel\-kin.} Formality isn't necessary.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b} \\
\textsc{ii}. \B{\cp{he}nga si} \I{["hE.Na si]} \emph{vin.} act in a formally polite or honorific way.
\B{Krr\-a ul\-txa si nga tsa\-txan\-ro'\-tu\-hu, ze\-rok fu\-ta ze\-ne he\-nga si\-vi, ma 'e\-veng.} When you meet that famous person, remember that you have to be formally polite, child.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b} 
\B{He\-nga rä\-'ä si, ma tsmuk.} Don't be so formal, bro\slash sis.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{leheng}{\B{leheng}}, formally polite. 

% hermeyp
\hi
\HT{hermeyp}{\B{her\cp{meyp}}} \I{[hER."mEjp\textcorner]} \emph{n.} snow flurries. 
(\ding{43}~\HL{herwI}{\B{her\-wì}}, \HL{meyp}{\B{meyp}})

% hertxayo
\hi
\HT{hertxayo}{\B{\cp{her}txayo}} \I{["hER.t'a.jo]} \emph{n.} field of snow. 
(\ding{43}~\HL{herwI}{\B{her\-wì}}, \HL{txayo}{\B{txa\-yo}})

% herwì
\hi
\HT{herwI}{\B{\cp{her}wì}} \I{["hER.wI]} \emph{n.} snow. \B{Her\-wì ze\-re\-i\-up fì\-trr\-o nì\-wotx!} It's been snowing all day!~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{herwIva}{\B{her\-wì\-va}}, snowflake.  
\on{2}.~\HL{hermeyp}{\B{her\-meyp}}, snow flurries. 
\on{3}.~\HL{hertxayo}{\B{her\-txa\-yo}}, field of snow. 
\on{4}.~\HL{txanfwerwI}{\B{txan\-fwer\-wì}}, blizzard. 

% herwìva
\hi
\HT{herwIva}{\B{\cp{her}wìva}} \I{["hER.wI.va]} \emph{n.} snowflake.
(\ding{43}~\HL{herwI}{\B{her\-wì}}, \HL{Ilva}{\B{ìl\-va}}) 

% hewne
\hi
\HT{hewne}{\B{\cp{hew}ne}} \I{["hEw.nE]} \emph{adj.} soft (\emph{of an object}). 
(\ding{214}~\HL{txa'}{\B{txa'}}) 

% heyn
\hi
\HT{heyn}{\B{heyn}} \I{[hEjn]} \emph{vin.} sit. 
(\emph{a.}) \B{aw\-ngal yom wu\-tsot teng\-krr he\-reyn nì\-pxim}, we eat dinner while sitting upright.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
\B{Lä\-ngu fì\-seyn kel\-ho\-an nì\-ngay. Tsun oe hi\-veyn tsaw\-sìn 'a\-'aw\-a swaw\-tsyìp nì\-'aw.} This chair is really uncomfortable. I can only sit on it a few seconds.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\emph{b. proverb}) \B{Txìm a\-'aw ke tsun hi\-veyn mì tal me\-fa'\-li\-yä.} You can't sit on the fence; you need to decide [lit.: One butt can't sit on the backs of two direhorses].~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{seyn}{\B{seyn}}, chair, stool, bench.
\on{2}.~\HL{heynyI}{\B{heyn\-yì}}, lap.

% heynyì
\hi 
\HT{heynyI}{\B{\cp{heyn}yì}} \I{["hEjn.jI]} \emph{n.} lap. 
\B{Za\-'u heyn sìn oey heyn\-yì, ma 'e\-vi.} Come sit on my lap, my child.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\ding{43}~\HL{heyn}{\B{heyn}}, \HL{yI}{\B{yì}}) 

% heyr
\hi
\HT{heyr}{\B{heyr}} \I{[hEjR]} \emph{n.} chest (\emph{of human or animal body}), \emph{the area between the stomach and throat}. 
\B{Oe\-ri heyr tì\-sraw sä\-ngi ta\-lu\-na zi\-ze'\-ìl oet sngo\-lap tsa\-tseng.} My chest hurts because a hellfire wasp stung me there.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Po\-an\-it tso\-lukx po\-el fa tstal nì\-txukx nem\-fa heyr.} She stabbed him deeply in the chest with a knife.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}

% hifwo
\hi
\HT{hifwo}{\B{\cp{hi}fwo}} \I{["hi.fwo]} \emph{vin.} flee, escape (\B{ftu}, from). \B{Pa\-lu\-kan\-it tso\-le\-'a, ye\-rik lopx hi\-fwo kxam\-lä ze\-swa.} Spotting a thanator, the hexapede panicked and escaped through the grass.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Tseyk 'ä\-pey\-ka\-mrr\-ko äo ut\-ral a zo\-lup fte hi\-vi\-fwo ftu ay\-sre' pa\-lu\-lu\-kan\-ä.} Jake rolled under the fallen tree to escape from the thanator's teeth.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\emph{a. idiom}) \B{Hì\-pey ta\-ron\-yu, hi\-fwo ye\-rik.} ``He who hesitates is lost.'' [lit.: The hunter hesitates and the hexapede escapes]~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}

% hilu
\hi 
\HT{hilu}{\B{\cp{hi}lu}} \I{["hi.lu]} \emph{n.} nipple. 

% -hin
\hi
\HT{hin}{\B{-hin}} \I{[.hin]} \emph{num., suf.} seven (\emph{rest of} \ding{43}~\HL{kinA}{\B{ki\-nä}} \emph{to form higher numbers}). \B{vo\-hin}, \on{15} (decimal), \on{17} (octal). \B{me\-vo\-hin}, \on{23} (decimal), \on{27} (octal); etc.

% hipx
\hi 
\HT{hipx}{\B{hipx}} \I{[hip']} \emph{vtr.} control. 
\B{Kar\-yu a\-sìl\-tsan ze\-ne tsi\-vun ay\-nu\-me\-yut hi\-vipx mì num\-tseng\-vi.} A good teacher has to be able to control (his/her) students in the classroom.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIhipx}{\B{tì\-hipx}}, control.

% hiup
\hi 
\HT{hiup}{\B{\cp{hi}up}} \I{["hi.up\textcorner]} \emph{or} \I{[.Up\textcorner]} (\textsc{rn}: \B{\cp{hi}ùp} \I{["hi.Up\textcorner]}) (\emph{a. vin.}) spit.
\B{Tsa\-kem rä\-'ä si, ma 'itan. Fwa hi\-up fì\-tseng ke lu muiä.} Don't do that, son. It's not proper to spit here.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
(\emph{b. vtr.}) spit out. \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hiupwopx}{\B{hi\-up\-wopx}}, cloud spitter.

% hiupwopx
\hi 
\HT{hiupwopx}{\B{\cp{hi}upwopx}} \I{["hi.up\textcorner.wop']} (\textsc{rn}: \B{\cp{hi}ùpwopx} \I{["hi.Up\textcorner.wop']}) \emph{n.} \ding{96} cloud spitter. 
(\ding{43}~\HL{hiup}{\B{hiup}}, \HL{pIwopx}{\B{pì\-wopx}})

% hiyìk
\hi
\HT{hiyIk}{\B{\cp{hi}yìk}} \I{["hi.jIk\textcorner]} \emph{adj.} funny, strange. 

% hì'ang
\hi
\HT{hI'ang}{\B{\cp{hì}'ang}} \I{["hI.PaN]} \emph{n.} insect. 
\B{Lu mek\-tseng a tsun fpxi\-vä\-kìm hì\-'ang tsaw\-ìlä.} There's a gap through which insects can get in.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{hI'i}{\B{hì\-'i}}, \HL{ioang}{\B{ioang}})

% hì'i
\hi
\HT{hI'i}{\B{\cp{hì}'i}} \I{["hI.Pi]} \emph{adj.} little, small (\emph{in size}). 
\B{Fì\-syu\-lang a\-rim lu hì\-'i fra\-to.} This yellow flower is the smallest of all.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} 
\B{su\-te tsun oe\-yä hì\-'i\-a tì\-ngop\-it si\-var}, people can use my strange/funny little creation.~\LINK{http://forum.learnnavi.org/language-updates/score!-\%281\%29-the-verb-to-use-\%282\%29-more-detail-on-teya-si/}{ln} 
\B{win\-a sä\-wä\-sul\-tsyìp a\-hì\-'i}, a quick little contest.~\LINK{http://naviteri.org/2012/10/wina-sawasultsyip-ahii-a-quick-little-contest/}{b} 
(\ding{214}~\HL{tsawl}{\B{tsawl}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tsawlhI'}{\B{tsawl\-hì'}}, size. 
\on{2}.~\HL{hI'ang}{\B{hì\-'ang}}, insect.   
\on{3}.~\HL{hIkrr}{\B{hì\-krr}}, second, very short time. 
\on{4}.~\HL{hItxoa}{\B{hì\-txoa}}, excuse me. 
\on{5}.~\HL{tanhI}{\B{tan\-hì}}, star. 
\on{6}.~\HL{hIno}{\B{hì\-no}}, fine, detailed. 
\on{7}.~\HL{hIpey}{\B{hì\-pey}}, hesitate, hold back. 
\on{8}.~\HL{hIrumwll}{\B{hì\-rum\-wll}}, puffer plant.

% hìkrr
\hi
\HT{hIkrr}{\B{\cp{hì}krr}} \I{["hI.k\s{r}]} \textsc{i}. \emph{n.} second, a very short time, a moment. 
\B{Maw hì\-krr ay\-oe tì\-yä\-txaw.} We'll return after a short time.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} \\
\textsc{ii}.~\emph{adv.} for a moment, for a second. 
\B{Tsun mi\-vä\-kxu hì\-krr srak?} May I interrupt for a moment?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} 
(\ding{43}~\HL{hI'i}{\B{hì\-'i}}, \HL{krr}{\B{krr}})

% hìm
\hi
\HT{hIm}{\B{hìm}} \I{[hIm]} \emph{adj.} small (\emph{in quantity}), a little. 
(\ding{214}~\HL{txan}{\B{txan}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hImtxan}{\B{hìm\-txan}}, amount. 
\on{2}.~\HL{hImpxI}{\B{hìm\-pxì}}, minority. 
\on{3}.~\HL{hImtxew}{\B{hìm\-txew}}, minimum. 

% hìmpxì
\hi
\HT{hImpxI}{\B{hìm\cp{pxì}}} \I{[hIm."p'I]} \emph{n.} minority, least, small part. 
\B{Hìm\-pxì Saw\-tu\-te\-yä lu tstu\-nwi}, A minority of the Sky People are kind.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 
(\ding{214}~\HL{txampxI}{\B{txam\-pxì}}, \ding{43}~\HL{hIm}{\B{hìm}}, \HL{hapxI}{\B{ha\-pxì}})

% hìmtxampe
\hi
\HT{hImtxampe}{\B{hìm\cp{txam}pe}} \I{[hIm."t'am.pE]} \emph{or} \B{pìm\cp{txan}} \I{[pIm.\-"t'an]} \emph{inter.} how much? 
\B{Fì\-nik\-roi ley pìm\-txan?} How valuable is this hair ornament?~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\ding{43}~\HL{hImtxan}{\B{hìm\-txan}})

% hìmtxan
\hi
\HT{hImtxan}{\B{hìm\cp{txan}}} \I{[hIm."t'an]} \emph{n.} amount. \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hImtxampe}{\B{hìm\-txam\-pe}\slash \B{pìm\-txan}}, how much? 

% hìmtxew
\hi
\HT{hImtxew}{\B{hìm\cp{txew}}} \I{[hIm."t'Ew]} \emph{n.} minimum. 
(\ding{214}~\HL{txantxew}{\B{txan\-txew}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hItxewvay}{\B{hìm\-txew\-vay}}, minimally.

% hìmtxewvay
\hi
\HT{hImtxewvay}{\B{hìm\cp{txew}vay}} \I{[hIm."t'Ew.vaj]} \emph{adv.} minimally. 
(\ding{43}~\HL{hImtxew}{\B{hìm\-txew}})

% hìno
\hi
\HT{hIno}{\B{hì\cp{no}}} \I{[hI."no]} \emph{adj.} fine, detailed, precise (\emph{of things}). 
\B{Pol ngop fra\-krr sì\-ke\-nong\-it a hì\-no lu nì\-hawng.} He always creates excessively detailed examples.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
\B{Fì\-way\-ri hì\-no\-a re\-nut ngam\-pam\-ä ke tsä\-ngun oe tsl\-ivam.} I'm afraid I can't understand the intricate rhyme scheme of this poem.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
(\ding{43}~\HL{leno}{\B{le\-no}}, \HL{nIno}{\B{nì\-no}})

% hìpey
\hi
\HT{hIpey}{\B{\cp{hì}pey}} \I{["hI.p$\cdot$Ej]} \emph{vin., inf.}\on{22} hesitate, hold back for a short time; \emph{defer the start of an action, reluctance to begin or accomplish an action, for whatever reason}.
\B{Fu\-ria peng fmawn\-it Ey\-tu\-kan\-ur po hì\-po\-ley.} He hesitated to tell Eytukan the news.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\emph{a. idiom}) \B{Hì\-pey ta\-ron\-yu, hi\-fwo ye\-rik.} ``He who hesitates is lost.'' [lit.: The hunter hesitates and the hexapede escapes]~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b} 
(\emph{b. proverb}) \B{'Uo\-ri hì\-pey, kxawm nga\-ru ke ley.} ``If you hesitate doing something, it might not be important to you,'' i.e. \emph{more meant\slash used as a motivation for s.o. hesitating, or even as a teasing to get s.o. into action}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{43}~\HL{hI'i}{\B{hì\-'i}}, \HL{pey}{\B{pey}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIhIpey}{\B{tì\-hì\-pey}}, hesitation. 
\on{2}.~\HL{lehIpey}{\B{le\-hì\-pey}}, hesitant, in a state of hesitation. 
\on{3}.~\HL{nIhIpey}{\B{nì\-hì\-pey}}, hesitantly.

% hìrumwll
\hi
\HT{hIrumwll}{\B{hì\cp{rum}wll}} \I{[hI."Rum.w\s{l}]} \emph{or} \I{[."RUm.]} \emph{n.}~\ding{96} puffer plant. 
(\ding{43}~\HL{hI'i}{\B{hì\-'i}}, \HL{rum}{\B{rum}}, \HL{'ewll}{\B{'e\-wll}})

% hìtxoa
\hi
\HT{hItxoa}{\B{hì\cp{txo}a}} \I{[hI."t'o.a]} \emph{idiom} excuse me, pardon me. \B{Hì\-txo\-a, ke tslo\-lam.} Sorry, I didn't get that.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} (\ding{43}~\HL{hI'i}{\B{hì\-'i}}, \HL{txoa}{\B{txoa}})

% hoan
\hi
\HT{hoan}{\B{\cp{ho}an}} \I{["ho.an]} \emph{n.} comfort.
\B{Sko frr\-tu, fra\-krr mì hel\-ku nge\-yä lu oe\-ru ho\-an nì\-txan, ma tsmuk.} I always feel very comfortable as a guest in your home, brother.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lehoan}{\B{le\-ho\-an}}, comfortable. 
\on{2}.~\HL{nIhoan}{\B{nì\-ho\-an}}, comfortably.

% hoet
\hi
\HT{hoet}{\B{\cp{ho}et}} \I{["ho.Et\textcorner]} \emph{adj.} vast, broad, expansive. \B{'Rr\-ta\-mì a tam\-pay lu ho\-et.} The oceans on Earth are vast.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
\B{Lì'\-fya le\-Na'\-vi lu teng\-wotx, ken ay\-reym\-sä\-tsa le\-ke\-teng\-syon Ey\-wa\-'e\-veng\-ä, ken ho\-et\-a ta\-yo a mì\-kam ay\-olo'.} The language of the Na'vi is consistent, despite the exo\hyp{}moon's varied landscapes and the vast expanse between clans.~\LINK{https://naviteri.org/2026/06/tskxekeng-a-mikyunfpi-pamrel-listening-exercise-text/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIhoet}{\B{nì\-ho\-et}}, widely, pervasively.

% hol
\hi
\HT{hol}{\B{hol}} \I{[hol]} \emph{adj.} few. 
\B{Tsay\-lì\-'u lu hol nì\-'aw.} Those words are few.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1117}{b} 
\B{Oel tse\-'a hol\-a tu\-tet.} I see only a few people.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\ding{214}~\HL{pxay}{\B{pxay}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{holpxay}{\B{hol\-pxay}}, number.

% holpxay
\hi
\HT{holpxay}{\B{hol\cp{pxay}}} \I{[hol."p'aj]} \emph{n.} number. 
\B{hol\-pxay ay\-zek\-wä\-yä feyä}, the number of their fingers.~\LINK{http://forum.learnnavi.org/language-updates/talun-taweyk-tafral-lun-oeyk-holpxay-hol/}{ln} 
\B{Ko\-ren Hol\-pxay\-ä}, Number Principle.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b} 
(\ding{43}~\HL{hol}{\B{hol}}, \HL{pxay}{\B{pxay}})

% holpxaype
\hi
\HT{holpxaype}{\B{hol\cp{pxay}pe}} \I{[hol."p'aj.pE]}~\emph{or}~\B{pol\cp{pxay}}~\I{[pol."p'aj]} \emph{inter.} how many? (behaves like an ordinary \emph{adj.}) \B{Nga\-ri so\-la\-lew pol\-pxay\-a zì\-sìt?} How old are you?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
(\ding{43}~\HL{holpxay}{\B{hol\-pxay}})

% hona
\hi
\HT{hona}{\B{\cp{ho}na}} \I{["ho.na]} \emph{adj.} endearing, adorable, cute. 
\B{Pe\-yä 'i\-tan\-ìri lu ho\-na nì\-txan.} His son is very cute.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
\B{ho\-na pa\-lu\-kan\-tsyìp a\-hì\-'i}, sweet little cat.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
\B{Prr\-nen lrr\-tok si a krr, srer me\-song\-tsyìp a\-ho\-na.} When the baby smiles, two adorable dimples appear.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIhona}{\B{nì\-ho\-na}}, sweetly, endearingly.  
\on{2}.~\HL{tIhona}{\B{tì\-ho\-na}}, cuteness, adorableness.

% horenleykekyu
\hi 
\HT{horenleykekyu}{\B{ho\cp{ren}ley\cp{kek}yu}} \I{[ho.""ren.lEj."kEk\textcorner.ju]} \emph{n.} law enforcer, (\emph{on Earth}) police officer. (\emph{often shortened to} \ding{43}~\HL{leykekyu}{\B{ley\-kek\-yu}}) 
\B{Kin\-trr\-am ay\-oe tar\-mìng na\-ri teng\-krr 'aw\-a ho\-ren\-ley\-kek\-yul tu\-tan\-it a\-'aw tspe\-rang--lun\-lu\-ke nì\-wotx, nìk\-'ong, nì\-ze\-vakx.} Last week we watched while one law enforcer killed a man--without any reason, slowly, cruelly.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} 
(\ding{43}~\HL{leykekyu}{\B{leykekyu}}, \HL{koren}{\B{ko\-ren}})

% hrrap
\hi
\HT{hrrap}{\B{\cp{hrr}ap}} \I{["h\s{r}.ap\textcorner]} \emph{n.} danger. \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lehrrap}{\B{le\-hrr\-ap}}, dangerous.     
\on{2}.~\HL{yrrap}{\B{yrr\-ap}}, storm. 
\on{3}.~\HL{penghrrap}{\B{peng\-hrr\-ap}}, binary sunshine. 
\on{4}.~\HL{penghrr}{\B{peng\-hrr}}, warn.

% hu
\hi
\HT{hu}{\B{hu}} \I{[hu]} \emph{adp.} with (\emph{together with s.o., accompaniment}). \B{Nga\-ri hu Ey\-wa sa\-lew ti\-re\-a, tokx 'ì\-'awn slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body remains to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{a\slash wiki} \B{Lar\-mu tsa\-tsam\-si\-yu\-hu tì\-vawm a lu stxong ay\-oer.} That warrior carried with him a darkness unknown to us.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\emph{a. idiom}) \B{Ey\-wa nga\-hu.} Ey\-wa (be) with you. 
(\emph{b. with verbs}) \HL{pAngkxo}{\B{päng\-kxo hu}}, speak\slash chat with; 
\HL{kAteng}{\B{kä\-teng hu}}, spend time with; 
\HL{mllte}{\B{mll\-te hu}}, agree with; 
\HL{wem}{\B{wem hu}}, fight (together) with (against s.o.~else); 
\HL{ultxa}{\B{ul\-txa si hu}}, meet (intentionally) with; 
\HL{tsaheylu}{\B{tsa\-heyl si hu}}, bond with, establish the neural connection with; 
\HL{uvan}{\B{u\-van si hu}}, play with s.o., 
\HL{nawang}{\B{na\-wang hu}}, merge, become one with.
\HL{nalsteng}{\B{nal\-steng hu}}, feel empthay\slash compassion for s.o.\\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{huslew}{\B{hu\-slew}}, accompany. 

% hufwa
\hi
\HT{hufwa}{\B{hu\cp{fwa}}} \I{[hu."fwa]} \emph{conj.} although. \B{hu\-fwa lì\-'u\-pam hek nì\-'it}, although the pronunciation is a bit strange.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \B{Fì\-ut\-ral\-ìri ta\-ngek\-ä zir fkan vawt, hu\-fwa ke rey.} The trunk of the tree feels solid, although it's dead.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} \B{Hu\-fwa ro\-lun oel 'a\-'aw\-a kxey\-ey\-ti, fì\-tì\-kang\-kem\-vi lu txan\-tsan ka wotx.} Although I found a few errors, this piece of work is generally excellent.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}

% hufwe
\hi
\HT{hufwe}{\B{hu\cp{fwe}}} \I{[hu."fwE]} \emph{n.} (\emph{a.}) wind. 
\B{ay\-nga\-ti spi\-vu\-le hu\-fwel}, the wind may propel you.~\LINK{http://forum.learnnavi.org/language-updates/naviya-the-other-vocative/}{a\slash ln} \B{Tì\-ran hu\-fwe.} It's breezy (but pleasantly so). 
\B{Tul hu\-fwe nì\-win.} It's very windy.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\emph{b. expression}) \B{nì\-win na hu\-fwe}, ``as fast as the wind'' (\emph{any situation to express rapidity}). \B{na hu\-fwe}, fluent (not just rapid) use of language.
\B{Fte\-ria oel lì'\-fya\-ti le\-Na'\-vi, slä mi ke tsä\-ngun pi\-vll\-txe na hu\-fwe.} I'm studying Na'vi, but I'm afraid I still can't speak it fluently.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{hufwetsyIp}{\B{hu\-fwe\-tsyìp}}, breeze, light wind. 
\on{2}.~\HL{txanfwerwI}{\B{txan\-fwer\-wì}}, blizzard. 
\on{3}.~\HL{nawfwe}{\B{naw\-fwe}}, fluent (for speech).

% hufwetsyìp
\hi
\HT{hufwetsyIp}{\B{hu\cp{fwe}tsyìp}} \I{[hu."fwE.\t{ts}jIp\textcorner]} \emph{n.} breeze, light wind.
\B{Hu\-fwe\-tsyìp le\-fpom tar\-mì\-ran.} A pleasant little breeze was blowing.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b} 
\B{Rìk a\-ukxo mì hu\-fwe\-tsyìp ka\-fi sar\-mi.} The dry leaf was sailing in the breeze.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b}
(\ding{43}~\HL{hufwe}{\B{hu\-fwe}})

% hultstxem
\hi
\HT{hultstxem}{\B{hul\cp{tstxem}}} \I{[hul."\t{ts}t'Em]} \emph{or} \I{[hUl.]} (\textsc{rn}: \B{hùl\cp{tstxem}} \I{[hUl."\t{ts}t'Em]}) \emph{vtr.} (\emph{usually negative connotation}) hinder, be an obstacle (to a person or thing). \B{Ke new oel fu\-ta fì\-tì\-päng\-kxot ay\-nge\-yä hi\-vul\-tstxem, slä tsun mi\-vä\-kxu hì\-krr nì\-'aw srak?.} I don't want to derail your chat, but can I interrupt for just a moment?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{mAkxu}{\B{mä\-kxu}})

% hum
\hi
\HT{hum}{\B{hum}}\textsuperscript{\on{1}} \I{[hum]} \emph{or} \I{[hUm]} (\textsc{rn}: \B{hùm} \I{[hUm]}) \emph{vin.} (\emph{a.}) exit, leave, depart. 
\B{Hum ne wrr\-pa!} Get outside!~\LINK{http://forum.learnnavi.org/language-updates/txing-and-hum/}{ln} 
\B{Ftu fì\-tseng ze\-ne hi\-vum.} We gotta get outta here.~\LINK{http://forum.learnnavi.org/language-updates/txing-and-hum/}{ln} 
\B{Po\-lä\-hem Saw\-tu\-te kam zì\-sìt a\-mrr, hum me\-zì\-sìt maw\-krr.} The Sky People arrived five years ago and left two years later.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Rey\-pay skxir\-ftum\-fa he\-rum.} Blood is coming out of the wound.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\emph{b. for astronomical bodies setting}) \B{Tsay\-san\-hì ha\-yum ye'\-rìn taw\-ftu.}~\emph{or} \B{Tsay\-san\-hì ha\-yum ye'\-rìn.} Those stars will soon set.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
(\ding{214}~\HL{fpxAkIm}{\B{fpxä\-kìm}})

\hi
\B{hum}\textsuperscript{\on{2}} : \HL{kum}{\B{kum}}

% hupx
\hi
\HT{hupx}{\B{hupx}} \I{[hup']} \emph{or} \I{[hUp']} \emph{vtr.} miss, not hit a target. 
\B{Txe\-wìl ye\-rik\-it ko\-lan slä hä\-ngupx.} Txe\-wì aimed at the hexapede but unfortunately missed.~\LINK{http://naviteri.org/2019/12/tengkrr-zisit-leratem-as-the-year-changes/}{b}
(\ding{214}~\HL{takuk}{\B{ta\-kuk}})

% huru
\hi
\HT{huru}{\B{\cp{hu}ru}} \I{["hu.Ru]} \emph{n.} cooking pot. 

% huslew
\hi 
\HT{huslew}{\B{\cp{hu}slew}} \I{["hu.sl$\cdot$Ew]} \emph{vtr., inf.}\on{22} accompany. 
\B{Sra\-ke ni\-vew nga hu\-sli\-vew oe\-ti ftxo\-zä\-ne?} Would you like to accompany me to the celebration?~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b}
(\ding{43}~\HL{hu}{\B{hu}}, \HL{salew}{\B{sa\-lew}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tIhuslew}{\B{tì\-hu\-slew}}, accompaniment.
\on{2}.~\HL{huslewtu}{\B{hu\-slew\-tu}}, accompanier, escort, companion.

% huslewtu
\hi 
\HT{huslewtu}{\B{\cp{hu}slewtu}} \I{["hu.slew.tu]} \emph{n.} accompanier, escort, companion. 
(\ding{43}~\HL{huslew}{\B{hu\-slew}}) 

% huta
\hi 
\HT{huta}{\B{\cp{hu}ta}} \I{["hu.ta]} \emph{adj.} unexpected (usually for positive outcomes). 
\B{tì\-'o\-ngokx a\-hu\-ta}, an unexpected birth.~\LINK{http://naviteri.org/2022/12/neytiriya-waytelem-neytiris-song-cord/}{b\slash a\on{2}} 

\end{multicols}