\pdfbookmark[0]{I}{I}
\begin{center}
\section*{I \I{[i]}}
\end{center}

\begin{multicols}{2}

% i'en
\hi
\HT{i'en}{\B{\cp{i}'en}} \I{["{}i.PEn]} \emph{n.} \emph{stringed Pandoran instrument}. 
\B{Pe\-o\-txang\-fa nga pam\-tseo si?--Pam\-tse\-o si fa i\-'en.} What instrument do you play?--I play the \emph{i'en}.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 
\B{Tew\-ti, nga lu tsul\-fä\-tu i\-'en\-ä.} Wow, you are a master on the \emph{i'en}.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}

% iknimaya
\hi
\HT{iknimaya}{\B{ikni\cp{ma}ya}} \I{[ik\textcorner.ni."ma.ja]} \emph{n.} Stairway to Heaven (\emph{a rite of passage of the Na'vi}). 
\B{Tsyìl Ik\-ni\-ma\-yat ul\-te tsa\-heyl si ik\-ran\-hu a fì\-'u lu tì\-fme\-tok a ze\-ne fra\-ta\-ron\-yu a\-'e\-wan em\-zi\-va\-'u.} Scaling Iknimaya and bonding with a banshee is a test that every young hunter must pass.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}

% ikran
\hi
\HT{ikran}{\B{\cp{ik}ran}} \I{["{}ik\textcorner.Ran]} \emph{n., fauna} ikran, mountain banshee. 
\B{Oe\-yä ik\-ran sli\-vu nga, tsa\-krr oeng 'aw\-si\-teng mi\-vak\-to.} Be my \emph{ikran} and let's ride together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{wiki} 
\B{Nge\-yä ik\-ran\-ìl ngip\-it le\-tam kin fte tsi\-vun kll\-pi\-vä.} Your ikran needs enough open space to be able to land.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Va'\-rul zä\-nget ik\-ran\-it sn\-eyä pxel ha\-pxì\-tu so\-a\-i\-ä.} Va'ru treats his \emph{ikran} like a member of the family.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{nan\-tang ke na\-mew fpi\-ve' ik\-ran\-ur upxa\-ret fa ngu\-way}, the viperwolf did not want to give messages to the banshee by howling.~\LINK{http://forum.learnnavi.org/language-updates/about-subordination/msg156378/\#msg156378}{ln} 
\B{Ik\-ran\-ìri krr\-a hu tu\-te tsa\-heyl si, ftang li\-vu yrr.} When an \emph{ikran} bonds with a person, it ceases to be wild.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
\B{Tsyeyk tswa\-mayon fa ik\-ran.} Jake flew with an \emph{ikran}.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\emph{a.~special usage}) \B{\cp{ik}\-ran \cp{mak}\-to}, the act of riding an ikran.  \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{ikranay}{\B{ik\-ra\-nay}}, forest banshee.
\on{2}.~\HL{kxetsikran}{\B{kxe\-tsik\-ran}}, banshee's tail.

% ikranay
\hi
\HT{ikranay}{\B{ikra\cp{nay}}} \I{[ik\textcorner.Ra."naj]} \emph{n., fauna} forest banshee, lesser banshee (\emph{smaller cousin to the mountain banshee}). 
(\ding{43}~\HL{ikran}{\B{ik\-ran}})

% ikut
\hi
\HT{ikut}{\B{\cp{i}kut}} \I{["{}i.kut\textcorner]} \emph{or} \I{[.kUt\textcorner]} (\textsc{rn}: \B{\cp{i}kùt} \I{["{}i.kUt\textcorner]}) \emph{n.} large pestle (\emph{grinding tool}); meal\hyp{}mashing pole. 

% il
\hi
\HT{il}{\B{il}} \I{[il]} \emph{vin.} bend, \emph{s.t.~straight that bends or hinges at a joint}. 
\B{Txo vul i\-vil nì\-hawng kxì\-ye\-vakx.} If a branch bends too much, it might break.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{Me\-tew\-it fì\-swek\-ä ey\-ki\-vil.} Pull the two ends of this bar together.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{til}{\B{til}}, joint, hinge. \\  
\on{2}.~\HL{eykil}{\B{ey\-kil}}, bend s.t., pulling things together.

% ilu
\hi 
\HT{ilu}{\B{\cp{i}lu}} \I{["{}i.lu]} \emph{n., fauna} ilu (\emph{a large plesiosaur\hyp{}like sea creature}). 

% ‹ilv›
\hi
‹\HT{ilv}{\B{ilv}}› \I{[il.v]} \emph{inf. in} ‹1› \emph{to form the perfective subjunctive}.
\B{Ul\-te sìl\-pey oe, fì\-'upxa\-ret inan a fì\-'u sil\-vu\-nu ay\-nga\-ru nì\-wotx!} And I hope, you all liked to read this message.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}

% ‹imv›
\hi
‹\HT{imv}{\B{imv}}› \I{[im.v]} \emph{inf. in} ‹1› \emph{to form the past subjunctive}.

% inan
\hi
\HT{inan}{\B{i\cp{nan}}} \I{[i."nan]} \emph{vtr.} (\emph{a.}) read (e.g. the forest), gain knowledge from sensory input. (\emph{with inf. stress changes to first syllable, e.g.}~\B{i\cp{vi}nan} \I{[i."vi.nan]}) 
\B{Tì\-o\-mum\-mì oe\-yä, pol na'\-rìng\-it i\-nan nìl\-tsan.} As far as I know, he reads the forest well.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} 
(\emph{b. colloquially}) `know what someone is about, know someone's ``deal'''.
\B{Pol ze\-ret oe\-ti pxel tu\-te a ke inan pot.} He's treating me like I don't know him.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tinan}{\B{ti\-nan}}, reading. 
\on{2}.~\HL{ninan}{\B{ni\-nan}}, by reading. 
\on{3}.~\HL{inanfya}{\B{inan\-fya}}, sense (means of perception).

% inanfya
\hi
\HT{inanfya}{\B{i\cp{nan}fya}} \I{[i."nan.fja]} \emph{n.} sense (\emph{means of perception}).
\B{Tì\-ta\-ron\-ìri lä\-ngu tì\-kak\-pam fe\-kum. Kin fkol fra\-inan\-fyat.} I'm sorry to say that deafness is a disadvantage for hunting. You need all your senses.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
\B{re\-nu ay\-i\-nan\-fya\-yä}, sense paradigm.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
(\ding{43}~\HL{inan}{\B{inan}}, \HL{fya'o}{\B{fya\-'o}})

% ingyen
\hi
\HT{ingyen}{\B{\cp{ing}yen}} \I{["{}iN.jEn]} \emph{n.} feeling of mystery, feeling of noncomprehension.
\B{Lu oer ing\-yen a Ì\-staw nim lu fì\-txan kum\-a pxìm wä\-pan.} It's a mystery to me why Ìstaw is so shy that he frequently hides.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{ningyen}{\B{ning\-yen}}, mysteriously. 
\on{2}.~\HL{ingyentsim}{\B{ing\-yen\-tsim}}, mystery, riddle. 
\on{3}.~\HL{ingyentsyIp}{\B{ing\-yen\-tsyìp}}, trick, clever methodology. 
\on{4}.~\HL{ingyenga'}{\B{ing\-ye\-nga'}}, mysterious, puzzling. 

% ingyentsim
\hi
\HT{ingyentsim}{\B{\cp{ing}yentsim}} \I{["{}iN.jEn.\t{ts}im]} \emph{n.} mystery, riddle, enigma, conundrum.
\B{Fwa po ke zo\-la\-'u lu ing\-yen\-tsim.} It's a mystery that he didn't come.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
\B{Sra\-ke tsun nga fì\-ing\-yen\-tsim\-ur tì\-ke\-zin si\-vi?} Can you solve this riddle?~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} 
(\ding{43}~\HL{ingyen}{\B{ing\-yen}}, \HL{tsim}{\B{tsim}})

% ingyentsyìp
\hi
\HT{ingyentsyIp}{\B{\cp{ing}yentsyìp}} \I{["{}iN.jEn.\t{ts}jIp\textcorner]} \emph{n.} trick, slight of hand, clever methodology, special methodology.
\B{Pol\-txe po san lu ing\-yen\-tsyìp a\-zey.} He said there's a special trick to it.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\ding{43}~\HL{ingyen}{\B{ing\-yen}})

% ingyenga'
\hi
\HT{ingenga'}{\B{\cp{ing}yenga'}} \I{["{}iN.jE.NaP]} \emph{adj.} mysterious, puzzling, enigmatic.
\B{Pe\-yä ay\-lì\-'u a\-ing\-ye\-nga' lo\-lu sngum\-tsim ay\-oe\-ru nì\-wotx.} His mysterious words worried us all.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\ding{43}~\HL{ingyen}{\B{ing\-yen}})

% io
\hi
\HT{io}{\B{\cp{i}o}} \I{["{}i.o]} \emph{adp.} above, over. 
\B{Oe tsway\-on io Ay\-vit\-ra\-yä Ra\-mu\-nong.} I fly over the Tree of Souls.~\LINK{http://forum.learnnavi.org/language-updates/over-under-and-quickly/}{ln} 
(\ding{214}~\HL{Ao}{\B{äo}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{ionar}{\B{io\-nar}}, banshee rider visor.

% ionar
\hi
\HT{ionar}{\B{\cp{i}onar}} \I{["{}i.o.naR]} \emph{n.} banshee rider visor. 
(\ding{43}~\HL{io}{\B{io}}, \HL{nari}{\B{na\-ri}}; \HL{renten}{\B{ren\-ten}})

% ioang
\hi
\HT{ioang}{\B{i\cp{o}ang}} \I{[i."{}o.aN]} \emph{n., fauna} animal, beast. 
\B{Fì\-i\-o\-ang lu tsuk\-yom.} This animal is edible.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b} 
\B{Na'\-rìng\-ìl nga' pxay\-a ioang\-it.} The forest contains many animals.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b} 
\B{Ioang\-ìri, tu\-te\-ri, tsun fko si\-var lì\-'ut alu tsmìm.} One can use the word `track' for animals or people.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1118}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{payoang}{\B{pay\-oang}}, fish.   
\on{2}.~\HL{hI'ang}{\B{hì\-'ang}}, insect.   
\on{3}.~\HL{yayo}{\B{ya\-yo}}, bird.  
\on{4}.~\HL{tarnioang}{\B{tar\-ni\-oang}}, predator animal. 
\on{5}.~\HL{talioang}{\B{ta\-li\-oang}}, sturmbeest. 
\on{6}.~\HL{yomioang}{\B{yom\-ioang}}, chalice plant. 
\on{7}.~\HL{yomhI'ang}{\B{yom\-hì\-'ang}}, dakteron. 
\on{8}.~\HL{tireaioang}{\B{Ti\-re\-a\-ioang}}, Spirit Animal.
\on{9}.~\HL{'eylyong}{\B{'eyl\-yong}}, pet.

% ioi / ioi si / ioi säpi
\hi
\HT{ioi}{\B{i\cp{o}i}} \I{[i."{}o.i]} \textsc{i}. \emph{n.} (item of) adornment or ceremonial apparel.
\B{Nì\-lun ay\-ioi a\-'eoio ay\-eyk\-tan\-ä lu lor fra\-to.} Of course the ceremonial wardrobes of the leaders are the most beautiful.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} 
\B{Saw\-no'\-ha ioit ko\-la\-ne\-i\-om oel u\-van\-fa!} I got (won) the prized piece of jewelry in the game!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} (\ding{43}~\HL{pxen}{\B{pxen}}) \\
\textsc{ii}. \HT{ioi si}{\B{i\cp{o}i si}} \I{[i."{}o.i si]} \emph{vin.} adorn s.o. (with, \B{fa}), put on (adornment). 
\B{Se\-vin\-a tsa\-'e\-ve\-ru a\-hì\-'i me\-sa'\-sem ioi so\-li fa mik\-tsang.} The parents adorned that pretty little girl with earrings.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} \\
\textsc{iii}. \HT{ioi sApi}{\B{i\cp{o}i sä\cp{pi}}} \I{[i."{}o.i s\ae."pi]} \emph{or coll.} \I{[spi]} \emph{vin.} adorn oneself (\HL{fa}{\B{fa}}, with), put on (adornment). 
\B{Fo\-ri maw\-krr\-a fa ren\-ten ioi sä\-po\-li ho\-lum.} After they put on their goggles, they left.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
(\ding{214}~\HL{'aku}{\B{'a\-ku}}, \ding{43}~\HL{yemstokx}{\B{yem\-stokx}})

% irayo / irayo si
\hi
\HT{irayo}{\B{i\cp{ra}yo}} \I{[i."Ra.jo]} \textsc{i}.~\emph{n.} thank; (\emph{a. idiom, intj.}) `Thanks,' `Thank you.' 
\B{Txa\-na ira\-yo.} Many thanks.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b} \\
\textsc{ii}. \HT{irayo si}{\B{i\cp{ra}yo si}} \I{[i."Ra.jo si]} \emph{vin.} thank s.o. (\emph{in the dative}) for s.t. (\emph{with the topic}). 
(\emph{a.}) \B{Tsa\-krr\-vay, ay\-nge\-yä tìm\-wey\-pey\-ri ira\-yo sei\-yi oe.} Until then, I thank you all for your patience.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Pxe\-fe\-yä tì\-tstun\-wi\-ri aw\-nga ira\-yo si\-vi ko!} Let us all thank those three for their kindness.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{Fu\-ri\-a ngal oe\-yä 'u\-pxa\-ret ay\-su\-te\-ru fpo\-le', nga\-ru ira\-yo sei\-yi oe nì\-txan!} I thank you very much for sending my message to people.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
(\emph{b. word order can be broken}) \B{Oel ay\-nga\-ti ka\-me\-i\-e, ma oe\-yä ey\-lan, ul\-te ay\-nga\-ru sei\-yi ira\-yo.} I \textsc{see} you all, my friends, and thank you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}

% ‹irv›
\hi
‹\HT{irv}{\B{irv}}› \I{[iR.v]} \emph{inf. in} ‹1› \emph{to form the imperfective subjunctive}.

% -it
\hi
\HT{-it}{\B{-it}} \I{[.it\textcorner]} \emph{suf. to mark the patientive case of a n. ending in a consonant, pseudovowel or diphthong}. 
(\ding{43}~\HL{-t}{\B{-t}}, \HL{-ti}{\B{-ti}})

% ‹iv›
\hi
‹\HT{iv}{\B{iv}}› \I{[i.v]} \emph{inf. in} ‹1›. (\emph{a.}) \emph{to form the subjunctive}. 
(\emph{b.}) \emph{required after} \B{fte}, \B{fteke} \emph{and all modal verbs}.
(\emph{c.}) \emph{optionally to indicate the imperative}.

% ‹iyev›
\hi
‹\HT{iyev}{\B{iyev}}› \I{[i.jE.v]} \emph{inf. in} ‹1› \emph{to form the (approximate) future subjunctive}. 
(\ding{43}~\HL{Iyev}{‹\B{ìyev}›})

\end{multicols}