\pdfbookmark[0]{SHORT I}{SHORT I}
\begin{center}
\section*{Ì \I{[I]}}
\end{center}

\begin{multicols}{2}

% ìì
\hi
\HT{II}{\B{ìì}} \I{[I:]} \emph{intj.} um, er (\emph{hesitation marker}). \B{Lu oe\-ru … ìì … tì\-ngä\-zìk a\-hì\-'i.} I have … um … a slight problem.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}

% -ìl
\hi
\HT{Il}{\B{-ìl}} \I{[.Il]} \emph{suf. to mark the agentive case of a n. ending in a consonant, pseudovowel or diphthong}. 
(\ding{43}~\HL{l}{\B{-l}})

% ìlä
\hi
\HT{IlA}{\B{ì\cp{lä}}} \I{[I."l\ae]} \emph{or} \B{\cp{ì}lä} \I{["{}I.l\ae]} \emph{adp.}\texttt{+} 
(\emph{a.}) by, via, following; in. \B{Ay\-fo so\-lop ìlä hil\-van fa uran.} They traveled along the river by boat.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Po\-ti fwey\-ko\-li ay\-oel ìlä rong.} We let him slide through the tunnel.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Na'\-vi ìlä ho\-'on kll\-kxo\-lem teng\-krr re\-rol.} The People were standing in a circle, singing.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\emph{b.}) according to. \B{Ìlä Fey\-ral, mun\-txa so\-li Ra\-lu sì Ne\-wey nì\-wan me\-srr\-am.} According to Peyral, Ralu and Newey were secretly married the day before yesterday.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b} (\ding{214}~\HL{wA}{\B{wä}})
(\emph{c. with} \HL{latem}{\B{la\-tem}}) depend on. \B{Tì\-flä la\-tem ìlä seyng\-a ftxey fkol sä\-nu\-met li\-vek fu\-ke.} Success depends on whether or not one follows instructions.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kxamlA}{\B{kxam\-lä}}, through, via the middle of.

% ìley
\hi
\HT{Iley}{\B{ì\cp{ley}}} \I{[I."lEj]} \emph{intj. sound of a war cry}.

% ‹ìlm›
\hi
‹\HT{Ilm}{\B{ìlm}}› \I{[Il.m]} \emph{inf.~in} ‹1› \emph{to indicate perfective approximate past}.

% ìlva
\hi
\HT{Ilva}{\B{\cp{ìl}va}} \I{["{}Il.va]} \emph{n.} flake, drop, chip. \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{herwIva}{\B{her\-wì\-va}}, snowflake. 
\on{2}.~\HL{payIva}{\B{pay\-ìva}}, drop of water. 
\on{3}.~\HL{txepIva}{\B{txe\-pì\-va}}, ash, cinder. 
\on{4}.~\HL{tompIva}{\B{tom\-pì\-va}}, raindrop.

% ‹ìly›
\hi
‹\HT{Ily}{\B{ìly}}› \I{[Il.j]} \emph{inf.~in} ‹1› \emph{to indicate perfective approximate future}.

% ‹ìm›
\hi
‹\HT{Im}{\B{ìm}}› \I{[I.m]} \emph{inf.~in} ‹1› \emph{to indicate approximate past}.

% ìpxa
\hi
\HT{Ipxa}{\B{ì\cp{pxa}}} \I{[I."p'a]} \emph{n.}~\ding{96} fern. 
\B{Fnu, fnu, wä\-pan wä ku | Uo tsa\-ìpxa.} Quiet, quiet, hide yourself from harm | Behind that fern.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}

% -ìri
\hi
\HT{-Iri}{\B{-ìri}} \I{[.I.ri]} \emph{suf. to mark the topical case of a n. ending in a consonant or pseudovowel}.
(\ding{43}~\HL{-ri}{\B{\mbox{-ri}}})

% ‹ìrm›
\hi
‹\HT{Irm}{\B{ìrm}}› \I{[IR.m]} \emph{inf.~in} ‹1› \emph{to indicate imperfective approximate past}.

% ‹ìry›
\hi
‹\HT{Iry}{\B{ìry}}› \I{[IR.m]} \emph{inf.~in} ‹1› \emph{to indicate imperfective approximate future}.

% Ìstaw
\hi
\B{Ì\cp{staw}} \I{[I."staw]} \emph{n. male name}. 
\B{Fì\-po lu Ìstaw.} \emph{or} \B{Fì\-por syaw fko Ìstaw.} This is Ìstaw.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Wo\-leyn Ìstawl yey\-fyat mì hll\-te.} Ìstaw drew a line on the ground.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}

% ‹ìsy›
\hi
‹\HT{Isy}{\B{ìsy}}› \I{[I.sj]} \emph{inf.~in} ‹1› \emph{to indicate determinational approximate future}.
\B{Pì\-syeng oe ngar.} I intend to tell you.~\LINK{http://forum.learnnavi.org/language-updates/a-collection/}{ln}
\B{Ta\-fral ke lì\-syek oel nge\-yä ke\-ye\-'ung\-it.} Therefore I will not heed your insanity.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{g\slash ln}

% ‹ìy›
\hi
‹\HT{Iy}{\B{ìy}}› \I{[I.j]} \emph{inf.~in} ‹1› \emph{to indicate approximate future}. 
\B{Maw hì\-krr ay\-oe tì\-yä\-txaw.} We'll be right back after this break.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
\B{ye'\-rìn 'ì\-yi\-'a sä\-nu\-me a tsa\-ri kll\-fro' oe}, soon the teaching I am responsible for will end.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}

% ‹ìyev›
\hi
‹\HT{Iyev}{\B{ìyev}}› \I{[I.jE.v]} \emph{inf.~in} ‹1› \emph{to indicate subjunctive future}.
\B{Sìl\-pey oe, fra\-po tsì\-ye\-vun fì\-tse\-nge ri\-vun 'uot le\-sar.} I hope that everyone may find something useful here.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Lu oe\-ru sngum a sa\-ron\-yu ke tì\-ye\-vä\-txaw.} I'm worried that the hunters will not return (soon, as expected).~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\emph{a. idiom}) \B{Kì\-ye\-va\-me.} \textsc{See} you soon.
(\ding{43}~\HL{iyev}{‹\B{iyev}›})

\end{multicols}