\pdfbookmark[0]{K}{K}
\begin{center}
\section*{K \I{[k]} (\emph{KeK})}
\end{center}

\begin{multicols}{2}

% ka
\hi
\HT{ka}{\B{ka}}\textsuperscript{\on{1}} \I{[ka]} \emph{adp.} (\emph{a. locative}) across. 
\B{Txur\-tel\-mì fo fta so\-li fte\-ke ka tsyokx fwi\-vi.} They tied a knot in the rope so it wouldn't slip through their hands.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{b. temporal}) 
\B{Nge\-yä tì\-eyk\-tan\-ìri, tì\-slan\-ìri sì tsran\-ten fra\-to a tì\-'ey\-lan\-ìri a ka ay\-zì\-sìt nì\-wotx, sei\-yi oe ira\-yo nì\-txan.} Thank you so much for your leadership, your support, and most importantly your friendship throughout the years.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{to disambiguate durative interpretation together with indefinite} \ding{43}~\B{-o}) 
\B{Pol lì'\-fya\-ti le\-Na'\-vi fto\-lia ka trr\-o (a\-'aw).} He studied Na'vi for a day.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/comment-page-1/\#comment-10619}{b}
(\emph{c. idiom}) \B{ka wotx}, generally, for the most part. 
\B{Hu\-fwa ro\-lun oel 'a\-'aw\-a kxey\-ey\-ti, fì\-tì\-kang\-kem\-vi lu txan\-tsan ka wotx.} Although I found a few errors, this piece of work is generally excellent.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{krrka}{\B{krr\-ka}}, during.
\on{2}.~\HL{katIng}{\B{ka\-tìng}}, distribute.
\on{3}.~\HL{kahena}{\B{ka\-he\-na}}, transport.

\hi
\B{ka}\textsuperscript{\on{2}} : \HL{kxa}{\B{kxa}}

% Ka'ani
\hi
\B{Ka'ani} \I{[ka.Pa.ni]} \emph{n.} \emph{male name}.~\LINK{}{dh}

% kafi / si
\hi 
\HT{kafi}{\B{\cp{ka}fi}} \I{["ka.fi]} \textsc{i}. \emph{n.} sail. \\
\textsc{ii}. \B{kafi si} \I{["ka.fi si]} \emph{vin.} to sail, move by means of a sail (\emph{lit. and fig.}).
\B{Rìk a\-ukxo mì hu\-fwe\-tsyìp ka\-fi sar\-mi.} The dry leaf was sailing in the breeze.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kaifuran}{\B{ka\-fi\-u\-ran}}, sailboat. 

% kafiuran
\hi 
\HT{kafiuran}{\B{\cp{ka}fiuran}} \I{["ka.fi.u.Ran]} \emph{n.} sailboat. 
(\ding{43}~\HL{kafi}{\B{kafi}}, \HL{uran}{\B{uran}})

% kahena
\hi 
\HT{kahena}{\B{ka\cp{he}na}} \I{[ka."h$\cdot$E.n$\cdot$a]} \emph{vtr., inf.}\on{23} transport. 
\B{Fwa ka\-he\-na fì\-uran\-it a\-tsawl ftu tsray oe\-yä ne pum nge\-yä la\-yu ngä\-zìk.} It's going to be difficult to transport this large boat from my village to yours.~\LINK{}{} 
(\ding{43}~\HL{ka}{\B{ka}}, \HL{hena}{\B{he\-na}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkanhena}{\B{tì\-ka\-hena}}, transportation (abstract). 
\on{2}.~\HL{sAkahena}{\B{sä\-ka\-he\-na}}, means of transport, vehicle.

% kakan
\hi 
\HT{kakan}{\B{\cp{ka}kan}} \I{["ka.kan]} \emph{adj.} rough (\emph{behavior; used for people and things}). 
\B{Ka\-kan\-a ay\-lì\-'u\-ri a pol\-txe oel nì\-sti, tsap\-'a\-lu\-te.} I apologize for the rough words that I spoke in anger.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{214}~\HL{flrr}{\B{flrr}}; \ding{43}~\HL{ekxtxu}{\B{ekx\-txu}}) \\
\textsc{Derivatives}: \on{1}.~\HL{nIkakan}{\B{nì\-ka\-kan}}, roughly.

% kakmokri
\hi 
\HT{kakmokri}{\B{kak\cp{mok}ri}} \I{[kak\textcorner."mok\textcorner.Ri]} \emph{adj.} mute.
\B{Txo po lu kak\-mok\-ri, fya\-pe sä\-fpìl\-yewn?} If he's mute, how does he communicate?~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{mokri}{\B{mokri}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkakmokri}{\B{tì\-kak\-mok\-ri}}, muteness.
\on{2}.~\HL{nIkakmokri}{\B{nì\-kak\-mok\-ri}}, mutely.

% kakpam
\hi
\HT{kakpam}{\B{kak\cp{pam}}} \I{[kak\textcorner."pam]} \emph{adj.} deaf.
\B{Hì\-krr\-o me\-fo kak\-pam lar\-mu maw\-krr\-a pxo\-lor kun\-sìp.} The two of them were deaf for a short time after the gunship exploded.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} 
(\ding{43}~\HL{pam}{\B{pam}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkakpam}{\B{tì\-kak\-pam}}, deafness.

% kakrel
\hi
\HT{kakrel}{\B{kak\cp{rel}}} \I{[kak\textcorner."REl]} \emph{adj.} blind. 
(\ding{43}~\HL{rel}{\B{rel}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkakrel}{\B{tì\-kak\-rel}}, blindness.

% kakzir
\hi 
\HT{kakzir}{\B{kak\cp{zir}}} \I{[kak\textcorner."ziR]} \emph{adj.} (\emph{a. physically}) numb. 
\B{Po\-ri pxun kak\-zir la\-tsu.} His arm must be\slash appears to be numb.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b}
(\emph{b. emotionally}) numb. 
\B{Te\-ri kxitx pe\-yä stawm a krr, oe kak\-zir slo\-lu nì\-wotx.} When I heard about her death, I became completely numb.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b}
(\ding{43}~\HL{zir}{\B{zir}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tIkakzir}{\B{tì\-kak\-zir}}, numbness.

% kali'weya
\hi
\HT{kali'weya}{\B{kali'\cp{wey}a}} \I{[ka.liP."wEj.a]} \emph{n., fauna} species of poisonous arachnid, \emph{scorpiosista virosae}. \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kalweyaveng}{\B{kal\-we\-ya\-veng}}, `son of a bitch'.

% kalin
\hi
\HT{kalin}{\B{ka\cp{lin}}} \I{[ka."lin]} \emph{adj.} sweet. 
\B{Fì\-na\-er\-ìri sur fkan oe\-ru ka\-lin.} This drink tastes sweet to me.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kalintu}{\B{ka\-lin\-tu}}, sweet person.

% kalintu
\hi 
\HT{kalintu}{\B{ka\cp{lin}tu}} \I{[ka."lin.tu]} \emph{n.} sweet person. 
\B{Lu nga ka\-lin\-tu nì\-ngay.} You're really a sweet person.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{kalin}{\B{ka\-lin}}) 

% kaltxì
\hi
\HT{kaltxI}{\B{kal\cp{txì}}} \I{[kal."t'I]} \textsc{i}. \emph{idiom} Hello! Greetings! 
\B{Kal\-txì!} Hello!~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Vanity_Fair}{wiki} 
\B{Kal\-txì ay\-nga\-ru}, I greet you all.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b} 
\B{Kal\-txì ta Ko\-pen\-han}, Hi from Copenhagen.~\LINK{http://naviteri.org/2010/09/kaltxi-ta-kopenhan-hi-from-copenhagen/}{b} \\
\textsc{ii}. \HT{kaltxI si}{\B{kal\cp{txì} si}} \I{[kal."t'I si]} \emph{vin.} say hello, greet. 
\B{Kal\-txì si\-vi, Ìstaw.} Say hello, Ìstaw.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} 
\B{Za\-'u kal\-txì si ko!} Come and say hello!~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
\B{Lo\-ak kal\-txì so\-li oer.} Loak greeted me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-294}{b}
(\ding{43}~\HL{kxI}{\B{kxì}})

% kalweyaveng
\hi 
\HT{kalweyaveng}{\B{kal\cp{wey}aveng}} \I{[kal."wEj.a.vEN]} \emph{n.} `son of a bitch', \emph{lit.}: `child of a poisonous arachnid' (\emph{insult}). 
\B{Txo sran, tsa\-krr fwa syaw por ``kal\-wey\-a\-veng'' lu mu\-iä nì\-wotx!} If yes, then to call them `son of a bitch' is entirely justified.~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/\#comment-30388}{b}
(\ding{43}~\HL{kali'weya}{\B{ka\-li'\-wey\-a}}, \HL{'eveng}{\B{'e\-veng}})

% kam 
\hi
\HT{kam}{\B{kam}}\textsuperscript{\on{1}} \I{[kam]} \emph{adp.} ago. 
\B{Tskot sngo\-lä\-'i po si\-var 'a\-'aw\-a trr\-kam.} He started to use the bow several days ago.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
\B{Po\-lä\-hem Saw\-tu\-te kam zì\-sìt a\-mrr, hum me\-zì\-sìt maw\-krr.} The Sky People arrived five years ago and left two years later.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{214}~\HL{kay}{\B{kay}}) 

\hi
\HT{kam}{\B{kam}}\textsuperscript{\on{2}} : \HL{kxam}{\B{kxam}}

% kame
\hi
\HT{kame}{\B{\cp{ka}me}} \I{["ka.mE]} \emph{vtr.} see (\emph{spiritual sense}), acknowledge, see into, understand, connect. (\emph{a}) \B{Oel nga\-ti ka\-me\-i\-e.} I \textsc{see} you.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog}{wiki} 
(\emph{b, idiom}) \B{Kì\-ye\-vame.} Good-bye. \textsc{See} you soon.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
(\ding{43}~\HL{tse'a}{\B{tse\-'a}})

% Kamun
\hi
\B{\cp{Ka}mun} \I{["ka.mun]}~\emph{or}~\I{[.mUn]} \emph{n. male name}. \B{Pol\-txe Ka\-mun san oe za\-sya\-'u sìk.} Kamun said, ``I will come.''~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/comment-page-1/\#comment-191}{b}
(\emph{shortened with} \B{-tsyìp}) \B{Kam\-tsyìp\-ìl wu\-tsot ye\-rom.} Little Kamun is having dinner.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}

% kan
\hi
\HT{kan}{\B{kan}} \I{[kan]} (\emph{a. vtr.}) aim (s.t. at\slash against s.t.\slash s.o.; \emph{used with} \ding{43} \HL{ne}{\B{ne}} \emph{or} \HL{wA}{\B{w\"a}}).
\B{Ney\-ti\-ril tsko\-ti ke\-ran ne ye\-rik.} Neytiri is aiming her bow at a hexapede.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/comment-page-1/\#comment-22792}{g\slash b}
\B{Pol tsko\-ti ko\-lan wä ku\-tu.} He raised his bow against the enemy.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/comment-page-1/\#comment-22792}{g\slash b}
(\emph{b. vtr.}) aim (at). 
\B{Ta\-ron\-yul le\-hì\-pey kan smar\-it nìl\-ke\-ftang slä ke ta\-kuk kaw\-krr.} A hesitant hunter will aim at a prey forever but never hit it.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\emph{c. vtr., modal}) intend (\emph{either with} ‹\B{iv}› \emph{or} \B{futa} \emph{and} ‹\B{iv}›). 
\B{Oe kan ki\-vä.} I intend to go.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/comment-page-1/\#comment-22792}{b} 
\B{Oel kan fu\-ta po ki\-vä.} I intend him to go.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/comment-page-1/\#comment-22792}{b} 
(\emph{d. conv., self\hyp{}correction as} \B{ko\cp{lan}}) ``I mean …, rather''.
\B{Oe\-ri tsyokx tì\-sraw si … ko\-lan ze\-kwä.} My hand hurts--I mean, my finger.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
\B{Fo ko\-lä tsa\-tseng fte ti\-va\-ron ye\-ri … ìì … ko\-lan ta\-li\-o\-ang\-it.} They went there to hunt hexa … um … I mean sturmbeast.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} 
(\emph{e. proverb}) \B{Fwa kan ke tam; ze\-ne swi\-zaw\-it li\-vo\-nu.} Intent is not enough; it's action that counts. [lit.: To aim is not enough; one must release the arrow.]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkan}{\B{tì\-kan}}, aim, goal, purpose, target. 
\on{2}.~\HL{kan'In}{\B{kan\-'ìn}}, focus on, specialize in.  
\on{3}.~\HL{kanfpIl}{\B{kan\-fpìl}}, concentrate. 
\on{4}.~\HL{kantseng}{\B{kan\-tseng}}, destination.

% kan'ìn
\hi
\HT{kan'In}{\B{\cp{kan}'ìn}} \I{["kan.PIn]} \emph{vtr.} focus on, specialize in, be particularly interested in.  \B{En\-tul kan\-'ìn tì\-wu\-sem\-it.} Entu specializes in fighting.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{kan}{\B{kan}}, \HL{'In}{\B{'ìn}})

% kanfpìl
\hi
\HT{kanfpIl}{\B{\cp{kan}fpìl}} \I{["kan.fpIl]} \emph{vin.} concentrate, focus one's attention. \B{Fu\-ri\-a sne\-yä tsko\-ti ngop po kan\-fpìl.} He's concentrating on making his bow.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Txo new nga tsli\-vam, ze\-ne ki\-van\-fpìl.} If you want to understand, you have to concentrate.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{kan}{\B{kan}}, \HL{fpIl}{\B{fpìl}})

% kanom
\hi
\HT{kanom}{\B{\cp{ka}nom}} \I{["ka.nom]} \emph{vtr.} acquire, get.
\B{Oe\-yä tsmu\-kan\-ìl mip\-a tsko\-ti kì\-ma\-ne\-i\-om.} My brother just got a new bow, I'm happy to say.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{sAkanom}{\B{sä\-ka\-nom}}, s.t.~acquired, an acquisition, a possession.
\on{2}.~\HL{tsukanom}{tsu\-ka\-nom}, available, obtainable.

% kantseng
\hi 
\HT{kantseng}{\B{\cp{kan}tseng}} \I{["kan.\t{ts}EN]} \emph{n.} destination. 
\B{Nge\-yä fì\-tì\-sop\-ìri pe\-han\-tseng?} This journey of yours--what's its destination?~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{kan}{\B{kan}}, \HL{tseng}{\B{tseng(e)}})

% kanu
\hi
\HT{kanu}{\B{\cp{ka}nu}} \I{["ka.nu]} \emph{adj.} smart, intelligent (\emph{of a person}). 
\B{Lam oer fwa po lu ka\-nu nì\-txan.} It seems to me that he is very smart.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkanu}{\B{tì\-ka\-nu}}, intelligence.  
\on{2}.~\HL{nIkanu}{\B{nì\-ka\-nu}}, intelligently, in a smart way. 

% kangay / si
\hi
\HT{kangay}{\B{ka\cp{ngay}}} \I{[ka."Naj]} \textsc{i}. \emph{adj.} valid. \\
\textsc{ii}. \HT{kangay si}{\B{ka\cp{ngay} si}} \I{[ka."Naj si]} \emph{vin.} confirm, validate. 
\B{Ru\-txe, ay\-nga\-ri fì\-kem fe\-yä 'e\-'a\-la to\-pur ka\-ngay sì\-yi nì\-'aw.} Please, this will only confirm their worst fears about you.~\LINK{http://forum.learnnavi.org/language-updates/two-words-in-use-eal-and-kangay-si/msg564284/\#msg564284}{ln}
\B{Fì\-'u\-pxa\-re ka\-ngay si fu\-ra po za\-ya\-'u.} This message confirms that he will come.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}

% kangkem
\hi 
\HT{kangkem}{\B{\cp{kang}kem}} \I{["kaN.kEm]} \emph{n.} work (\emph{coll. of} \ding{43}~\HL{tIkangkem}{\B{tì\-kang\-kem}}). 
\B{Kang\-kem val!} Work hard!~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 
(\emph{a. idiom}) \B{Po\-ri kang\-kem lu kawng\-kem.} He's really lazy [lit.: `for him, work is a crime'].~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}

% Kapsavan
\hi 
\HT{kapsavan}{\B{Kapsavan}} \I{[kap\textcorner.sa.van]} \emph{n.} the river in the Valley of Mo\-'a\-ra along which the Shaman of Songs resides, the Na'\-vi River Journey river.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln} 

% kar
\hi
\HT{kar}{\B{kar}} \I{[kaR]} \emph{vtr.} teach. 
\B{Lo\-ak\-ìl pä\-nu\-to\-lìng fu\-ta kar oe\-ru fya\-'ot a 'ìp fko nem\-fa ewll.} Loak promised he'd teach me how to vanish into the bushes.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Stxe\-nu\-to\-lìng oel fu\-ta lì'\-fyat le\-Na'\-vi po\-e\-ru ki\-var.} I offered to teach her Na'vi.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Ta\-fral kì\-syar a ay\-sä\-num\-vi sì ay\-lì'\-fya\-vi lu fyin.} That's why I'll teach lessons and expressions that are simple.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} 
\B{Fo\-ri tsa\-fne\-tì\-ku\-sar ke lu le\-ha'.} That kind of teaching isn't appropriate for them.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}  \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{karyu}{\B{ka\-ryu}}, teacher. 

% kara
\hi 
\HT{kara}{\B{ka\cp{ra}}} \I{[ka."Ra]} (\emph{a. vin.}) resist. 
\B{Fol nga\-ti spo\-le\-'e a krr, nga lum\-pe ke ka\-ra?} When they captured you, why didn't you resist?~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(\emph{b. vtr.}) resist s.t.\slash s.o.
\B{Ay\-sä\-la\-tem\-it a za\-mo\-lu\-nge Saw\-tu\-tel nga fmi ki\-va\-ra. Lä\-ngu kel\-tsun.} You try to resist the changes brought by the Sky People. Sadly, that's impossible.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkara}{\B{tì\-ka\-ra}}, resistance. 

% karyu
\hi
\HT{karyu}{\B{\cp{kar}yu}} \I{["kaR.ju]} \emph{n.} teacher.
\B{Lo\-lu tok\-tor Kì\-rey\-sì kar\-yu a\-sìl\-tsan.} Doctor Grace was a good teacher.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Kar\-yul fngo\-lo' fu\-ta ay\-nu\-me\-yu pi\-va\-te ye'\-krr.} The teacher required the students to arrive early.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Nga ze\-ne ay\-lì\-'ut kar\-yu\-ä yi\-vu\-ne, ma 'evi.} You must listen to your teacher, my son.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Fì\-sä\-tsan\-'ul\-ìri nge\-yä lu oe\-ru sko har\-yu nrra.} As your teacher, I'm proud of your improvement.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\ding{43}~\HL{kar}{\B{kar}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{karyunay}{\B{kar\-yu\-nay}}, apprentice teacher.

% karyunay
\hi
\HT{karyunay}{\B{karyu\cp{nay}}} \I{[kaR.ju."naj]} \emph{n.} apprendice teacher. 
(\ding{43}~\HL{karyu}{\B{kar\-yu}})

% katir
\hi 
\HT{katir}{\B{\cp{ka}tir}} \I{["ka.tiR]} \emph{n.} rainbow. 

% katìng
\hi 
\HT{katIng}{\B{\cp{ka}tìng}} \I{["ka.t$\cdot$IN]} \emph{vtr., inf.}\on{22} distribute. 
\B{Eyk\-yul ay\-swi\-zaw\-ti ka\-to\-lìng ay\-ha\-pxì\-tur tsam\-po\-ngu\-ä.} The leader distributed the arrows to the members of the war party.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{ka}{\B{ka}}, \HL{tIng}{\B{tìng}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkatIng}{\B{tì\-ka\-tìng}}, distribution. 

% kato
\hi
\HT{kato}{\B{\cp{ka}to}} \I{["ka.to]} \emph{n.} rhythm. 
\B{Tom\-pa\-yä ka\-to | Tsaw\-ke\-yä ka\-to | Ka\-tot tä\-ftxu oel | Nì\-ean nì\-rim.} The rhythm of the rain | The rhythm of the sun | I weave the rhythm | In blue and yellow.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}

% kavan
\hi 
\HT{kavan}{\B{\cp{ka}van}} \I{["ka.van]} \emph{vtr.} support (\emph{physically}). 
\B{Fol kar\-ma\-van ko\-ak\-tet teng\-krr fme\-ri po ti\-vì\-ran.} They supported the old woman as she was trying to walk.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
(\ding{43}~\HL{slan}{\B{slan}})

% kavuk
\hi
\HT{kavuk}{\B{ka\cp{vuk}}} \I{[ka."vuk\textcorner]} \emph{or} \I{[."vUk\textcorner]} (\textsc{rn}: \B{ka\cp{vùk}} \I{[ka."vUk\textcorner]}) \textsc{i}.~\emph{n.} treachery, betrayal. 
\B{Lo\-lu ka\-vuk, slä Tse\-nul tì\-ngay\-it ko\-lu\-lat.} There was treachery, but Tsenu revealed the truth.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{ii}.~\HL{kavuk si}{\B{kavuk si}} \I{[ka."vuk\textcorner{} si]} \emph{vin.} betray. 
\B{Po aw\-nga\-ru ka\-vuk so\-li.} He betrayed us.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/comment-page-1/\#comment-189}{b}

% kaw'it
\hi
\HT{kaw'it}{\B{kaw\cp{'it}}} \I{[kaw."Pit\textcorner]} \emph{adv.} (\emph{used with} \ding{43}~\HL{ke}{\B{ke}} \emph{to mean}) (not) a bit, (not) at all.
\B{Txan\-krr ngal ke lawk pot kaw\-'it.} You haven't mentioned a thing about him in a long time.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} 
\B{Ke lu po tsam\-si\-yu kaw\-'it!} There's not a warrior's bone in his whole body!~\LINK{http://forum.learnnavi.org/language-updates/ke-kawit/msg174572/\#msg174572}{ln}
(\emph{a. conv.}) \B{Ke srätx kaw\-'it!\slash Kea sä\-srätx kaw\-'it!\slash Ke\-he kaw\-'it!} `Not at all!' [lit.: (it) doesn't bother (me) at all\slash no bother at all]~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{'it}{\B{'it}})

% kawkrr
\hi
\HT{kawkrr}{\B{\cp{kaw}krr}} \I{["kaw.k\s{r}]} \emph{n., adv.} no time; never (\emph{requires negation of the verb}). 
\B{Kaw\-krr ke sla\-yu nga Na'\-vi\-yä ha\-pxì!} You will never become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake.2C_honest_about_betrayal}{a\slash wiki}
\B{Fay\-lì\-'u a\-lor oe\-ru te\-ya si nì\-txan, ul\-te ke tswa\-ya' oel sat kaw\-krr.} These beautiful words fill me with joy, and I will never forget them.~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
(\ding{43}~\HL{krr}{\B{krr}})

% kawkxan
\hi 
\HT{kawkxan}{\B{kaw\cp{kxan}}} \I{[kaw."k'an]} \emph{adj.} free, unblocked, unobstructed, clear. 
\B{Nga tsun ki\-vä set. Fya\-'o lu kaw\-kxan.} You can go now. The way is clear.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\HL{ekxan}{\B{e\-kxan}})

% kawl
\hi 
\HT{kawl}{\B{kawl}} \I{[kawl]} \emph{adv.} hard, diligently. 
\B{Mak\-to kawl, ma sam\-si\-yu, fte tsi\-vun pi\-vä\-hem nì\-win!} Ride hard, warriors, so you can get there fast!~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{43}~\HL{val}{\B{val}}) 

% kawlo
\hi 
\HT{kawlo}{\B{\cp{kaw}lo}} \I{["kaw.lo]} \emph{adv.} not once (\emph{requires negation of the verb}). 
\B{Oel keng kawlo ke solar kea räptulì'fyat ngahu!} I have never even once used vulgar language with you!~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
\B{Oe kaw\-lo ke ko\-lä tsa\-tseng\-ne.} I have never once gone there.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/\#comment-50641}{b}
(\ding{43}~\HL{ke}{\B{ke}}, \HL{'awlo}{\B{'aw\-lo}})

% kawnomum
\hi 
\HT{kawnomum}{\B{kaw\cp{no}mum}} \I{[kaw."no.mum]} \emph{or} \I{[.mUm]} (\textsc{rn}: \B{kaw\cp{no}nùm} \I{[kaw."no.nUm]}) \emph{adj.} unknown. 
(\ding{43}~\HL{ke}{\B{ke}}, \HL{awn}{‹\B{awn}›}, \HL{omum}{\B{omum}})

% kawng
\hi
\HT{kawng}{\B{kawng}} \I{[kawN]} \emph{adj.} bad, evil. 
\B{su\-te\-kip nì\-wotx, ftxey Na'\-vi ftxey Saw\-tu\-te, lu sìl\-tsan lu kawng}, among all people, whether Na'vi or Skypeople, there are good and bad (ones).~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Lam set fwa Saw\-tu\-te a\-kawng ho\-lum.} Now it seems that the evil Skypeople have left.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Kawng\-a hem a\-na\-fì'u lo\-lä\-ngen alo a\-pxay mì ok\-vur ay\-oe\-yä.} Many such bad actions have happened many times in our history.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} 
(\ding{214}~\HL{sIltsan}{\B{sìl\-tsan}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkawng}{\B{tì\-kawng}}, evil. 
\on{2}.~\HL{kawnglan}{\B{kawng\-lan}}, malicious.   
\on{3}.~\HL{kawngsar}{\B{kawng\-sar}}, exploit.
\on{4}.~\HL{kawngtu}{\B{kawng\-tu}}, bad person.
\on{5}.~\HL{kawngkem}{\B{kawng\-kem}}, evil deed, crime.

% kawnglan
\hi
\HT{kawnglan}{\B{\cp{kawng}lan}} \I{["kawN.lan]} \emph{adj.} malicious, bad\hyp{}hearted. 
\B{Hìm\-pxì Saw\-tu\-te\-yä lu tstun\-wi, slä fe\-yä txam\-pxì lä\-ngu kawng\-lan.} A minority of the Sky People are kind, but the majority are malicious.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\ding{43}~\HL{kawng}{\B{kawng}}, \HL{txe'lan}{\B{txe'\-lan}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tstunlan}{\B{tstun\-lan}}, kind\hyp{}hearted.

% kawngkem
\hi 
\HT{kawngkem}{\B{\cp{kawng}kem}} \I{["kawN.kEm]} \emph{n.} evil deed, crime. 
\B{Tsa\-kawng\-kem\-ìri lu oe layl!} I am innocent of that crime!~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\emph{a. idiom}) \B{Po\-ri kang\-kem lu kawng\-kem.} He's really lazy [lit.: `for him, work is a crime'].~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{kawng}{\B{kawng}}, \HL{kem}{\B{kem}}) 

% kawngsar
\hi
\HT{kawngsar}{\B{\cp{kawng}sar}} \I{["kawN.s$\cdot$aR]} \emph{vtr., inf.}\on{22} exploit. 
\B{Ay\-nge\-yä ki\-fkey\-ti fol kawng\-sar nì\-tut, fì\-'u\-ri ke\-kem ke si ay\-nga.} They continuously exploit your world and you do nothing about it.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{kawng}{\B{kawng}}, \HL{sar}{\B{sar}})

% kawngtu
\hi 
\HT{kawngtu}{\B{\cp{kawng}tu}} \I{["kawN.tu]} \emph{n.} bad person, `bad guy'. 
\B{Lal\-a tsa\-rel\-mì a\-ru\-sikx, yem\-stokx tsan\-tul haw\-re'\-ti a\-teyr, kawng\-tul pum\-it a\-la\-yon.} In that old movie, the good guy wears a white hat, the bad guy a black one.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\ding{214}~\HL{tsantu}{\B{tsantu}}; \ding{43}~\HL{kawng}{\B{kawng}}, \HL{tute1}{\B{tute}})

% kawtu
\hi
\HT{kawtu}{\B{\cp{kaw}tu}} \I{["kaw.tu]} \emph{pn.} no one, nobody (\emph{requires double negation}). 
\B{kaw\-tu ke tsun ay\-oer tì\-ftang si\-vi}, no one can stop us.~\LINK{http://forum.learnnavi.org/language-updates/stop!/}{ln} 
\B{En\-tu lu tu\-vom\-a ta\-ron\-yu. Kaw\-tut na po ke tso\-le\-'a oel mì sì\-rey.} Entu is an incredible hunter. I've never seen anyone like him before.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Pe\-yä sä\-ftxu\-lì\-'u lo\-lä\-ngu ke\-sran ul\-te kaw\-tur slan\-ti\-re ke si.} Unfortunately his speech was only so-so and inspired no one.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{tute1}{\B{tu\-te}})

% kawtseng
\hi 
\HT{kawtseng}{\B{\cp{kaw}tseng}} \I{["kaw.\t{ts}EN]} \emph{adv.} nowhere (\emph{requires negation of the verb}). 
\B{Nga ke tsun wä\-pi\-van; nga\-ri ke lu ke\-a u\-tu\-ru kaw\-tseng.} You cannot hide; there is no sanctuary for you anywhere.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 
(\ding{43}~\HL{tseng}{\B{tse\-nge}})

% kay
\hi
\HT{kay}{\B{kay}} \I{[kaj]} \emph{adp.} in (\emph{the future}), from now. 
\B{Za\-ya\-'u Saw\-tu\-te fte aw\-nga\-ti ski\-va\-'a kay zì\-sìt a\-pxey!} The Sky People will come to destroy us three years from now!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}  
\B{Oe pä\-hem kay\slash maw trr\-pxì\-vi a\-mrr.} I'll be there in five minutes.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} 
(\ding{214}~\HL{kam}{\B{kam}}) 

% kaym / kaymam / kaymay
\hi
\HT{kaym}{\B{kaym}} \I{[kajm]} \emph{n.} evening; \emph{adv.} in the evening. 
\B{Nän yom kxam\-trr, 'ul 'e\-fu o\-hakx kaym.} The less you eat at noon, the hungrier you'll feel in the evening.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
\B{Kaym\-krr\-ka to\-lel mo\-el ta ay\-ul\-txa\-tu stxe\-lit a\-ko\-sman.} During the evening we received a wonderful gift from the meeting members.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b} \\  
(i.) \B{kay\cp{mam}} \I{[kaj."mam]} \emph{n., adv.} last, yesterday evening. \\
(ii.) \B{kay\cp{may}} \I{[kaj."maj]} \emph{n., adv.} next, tomorrow evening. 

% kä-
\hi
\HT{kA-}{\B{kä-}} \I{[k\ae.]} \emph{non\hyp{}productive part., pref. to mean} away from, out, e.g. in \ding{43} \HL{kA'ArIp}{\B{kä\-'ä\-rìp}}, push; \HL{kAmakto}{\B{kä\-mak\-to}}, ride out; \HL{kApxI}{\B{kä\-pxì}}, rear; \HL{kAsrIn}{\B{kä\-srìn}}, lend; \HL{kAsatseng}{\B{kä\-sa\-tseng}}, out there; etc.

% kä
\hi
\HT{kA}{\B{kä}} \I{[k\ae]} \emph{vin.} go. 
\B{Tsa\-tse\-nge le\-hrr\-ap lu fì\-txan kum\-a tsa\-ne ke kä aw\-nga kaw\-krr.} That place is so dangerous we never go there.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}  
\B{Ne\-to rä\-'ä ki\-vä!} Don't go away.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} 
\B{Oel new fu\-ta ta\-ron\-yu ki\-vä.} I want the hunter to go.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-email/}{ln}
\B{Sra\-ke ko\-lä nga 'aw\-li\-e ne Nu Yor\-kì?} Have you ever been to New York?~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
\B{Oe kaw\-krr ne few\-tu\-so\-ka pa\-'o kil\-van\-ä ke ka\-mä.} I never went to the opposite side of the river.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Lam ngay oer, fo ka\-yä ìlä hil\-van tup na'\-rìng.} I bet they'll take the river rather than the forest.~\LINK{http://forum.learnnavi.org/language-updates/concerning-tup/}{ln} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kAteng} {\B{kä\-teng}}, spend time with.     
\on{2}.~\HL{kllkA}{\B{kll\-kä}}, go down, descend. 
\on{3}.~\HL{fAkA}{\B{fä\-kä}}, go up, ascend. 
\on{4}.~\HL{emkA}{\B{em\-kä}}, cross (s.t.).

% kä'ärìp
\hi
\HT{kA'ArIp}{\B{kä\cp{'ä}rìp}} \I{[k\ae."P$\cdot$\ae.R$\cdot$Ip\textcorner]} \emph{vtr., inf.}\on{23} push (\emph{coll. also} \ding{43}~\HL{kArIp}{\B{kä\-rìp}}). 
(\ding{43}~\HL{'ArIp}{\B{'ärìp}}, \HL{kA-}{\B{kä-}}; \ding{214}~\HL{za'ArIp}{\B{za\-'ä\-rìp}}) 

% kämakto
\hi
\HT{kAmakto}{\B{kä\cp{mak}to}} \I{[k\ae."m$\cdot$ak\textcorner.t$\cdot$o]} \emph{vin., inf.}\on{23} ride out. 
\B{kä\-mak\-to nì\-win}, ride out quickly.~\LINK{http://forum.learnnavi.org/language-updates/naviya-the-other-vocative/}{a\slash ln}
(\ding{43}~\HL{makto}{\B{mak\-to}}, \HL{kA-}{\B{kä-}})

% kämunge
\hi 
\HT{kAmunge}{\B{kä\cp{mu}nge}} \I{[k\ae."m$\cdot$u.N$\cdot$E]} (\textsc{rn}: \B{kä\cp{mù}nge} \I{[k\ae."mU.NE]}) \emph{vtr., inf.}\on{23} take (\emph{away from the speaker}). 
(\ding{43}~\HL{munge}{\B{mu\-nge}}, \HL{kA-}{\B{kä-}}; \ding{214}~\HL{zamunge}{\B{za\-mu\-nge}})

% käpxì
\hi
\HT{kApxI}{\B{kä\cp{pxì}}} \I{[k\ae."p'I]} \emph{n.} rear (part or section). 
\B{Zo\-la\-'u ftu kä\-pxì num\-tseng\-vi\-yä sä\-mewn a\-wok.} From the back of the classroom came a loud sigh.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
(\ding{214}~\HL{zapxI}{\B{za\-pxì}}) 

% kärìp
\hi 
\HT{kArIp}{\B{\cp{kä}rìp}} \I{["k\ae.RIp\textcorner]} \emph{vtr.} (\emph{coll. form of}) push. 
(\ding{43}~\HL{kA'ArIp}{\B{kä\-'ä\-rìp}}) 

% käsrìn
\hi
\HT{kAsrIn}{\B{kä\cp{srìn}}} \I{[k\ae."sR$\cdot$In]} \emph{vtr.}\on{22} lend.
\B{Sne\-yä ma\-sat\-it pol kä\-sro\-lìn oer.} He lent me his breastplate.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{srIn}{\B{srìn}}, \HL{kA-}{\B{kä-}}; \ding{214}~\HL{zasrIn}{\B{za\-srìn}}) 

% käteng
\hi
\HT{kAteng}{\B{\cp{kä}teng}} \I{["k$\cdot$\ae.tEN]} \emph{vin., inf.}\on{11} spend time, hang out (with, \HL{hu}{\B{hu}}). 
\B{Nga pe\-su\-hu kä\-teng nì\-trr\-trr.} Who do you typically spend time with?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
\B{Kä\-teng oe hu ey\-lan Per\-lin\-mì a mrr\-trr lu tsyeym a ke tsun tswi\-va' kaw\-krr.} The week I spent with my friends in Berlin is a treasure that I'll never forget.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}
\B{Kaym\-o zo\-la\-'u fray\-ul\-txa\-tu ne kel\-ku moe\-yä, fte ki\-ve\-i\-ä\-teng nì\-'o'.} All the members came to our house for an evening to hang out in a fun way.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}
(\ding{43}~\HL{kA-}{\B{kä-}}, \HL{teng}{\B{teng}})

% käsatseng 
\hi
\HT{kAsatseng}{\B{kä\cp{sa}tseng}} \I{[k\ae."{}sa.\t{ts}EN]} \emph{adv.} out there. 
(\ding{43}~\HL{tsatseng}{\B{tsa\-tseng}}, \HL{kA-}{\B{kä-}})

% ke
\hi
\HT{ke}{\B{ke}} \I{[kE]} \emph{adv.} not.
(\emph{a. with verbs}) 
\B{Ke lu oer set krr atxan}, I don't have much time now.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg139677/\#msg139677}{ln}
\B{oel pot ke spaw} I don't believe him.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{ke rey a tu\-te}, person who doesn't live\slash isn't alive.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Participles}{wiki}
(\emph{b. with} \ding{43}~\HL{kea}{\B{kea}}, \HL{ke'u}{\B{ke\-'u}}, \HL{kawtu}{\B{kaw\-tu}}, \HL{kawkrr}{\B{kaw\-krr}}, \HL{kawtseng}{\B{kaw\-tseng}}, \HL{kaw'it}{\B{kaw\-'it}}, \HL{kekem}{\B{ke\-kem}} \emph{to form}) double negation.
\B{Vay set ke pa\-mä\-hä\-ngem kea tì\-'eyng.} Up to now no answer has arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ke\-'u ke lu ngay.} Nothing is true.~\LINK{https://forum.learnnavi.org/language-updates/double-negatives-not-optional/}{ln}
\B{Ul\-te kaw\-tu ke tsun ay\-oer tì\-ftang si\-vi.} And no one can stop us.~\LINK{http://forum.learnnavi.org/language-updates/stop!/}{ln}
\B{Kaw\-krr ke sla\-yu nga Na'\-vi\-yä ha\-pxì!} You will never be one of The People!~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake.2C_honest_about_betrayal}{a\slash wiki}
\B{Ke lu po tsam\-si\-yu kaw\-'it!} He is not a worrior at all!~\LINK{http://forum.learnnavi.org/language-updates/ke-kawit/msg174572}{ln}
\B{Ay\-nge\-yä ki\-fkey\-ti fol kawng\-sar nì\-tut, fì\-'u\-ri ke\-kem ke si ay\-nga.} They continuously exploit your world and you do nothing about it.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\emph{c. with adj.s; idiom}) \B{Ke pxan.} `Thank you, I don't deserve your praise.' (\emph{in response to a praise or compliment}) [lit.: not worthy].
(\emph{d. with} \ding{43}~\HL{ki1}{\B{ki}} \emph{to mean}) `not … but rather'. 
\B{Nga pll\-txe ke nì\-fyeyn\-tu ki nì\-'e\-veng.} You speak not like an adult but a child.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\emph{e. with adp.s}) \B{Kll\-kxa\-yem fì\-tì\-kang\-kem oe\-yä ro\-fa--ke io--pum fe\-yä.} My work will stand beside--not above--theirs.~\LINK{http://naviteri.org/2010/06/first-post/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{ke'aw}{\B{ke\-'aw}}, divided, torn apart. 
\on{2}.~\HL{ke'u}{\B{ke\-'u}}, nothing. 
\on{3}.~\HL{kefyak}{\B{ke\-fyak}}, … right? … isn't it? 
\on{4}.~\HL{kezemplltxe}{\B{ke\-zem\-pll\-txe}}, needless to say. 
\on{5}.~\HL{zenke}{\B{zen\-ke}}, must not. 
\on{6}.~\HL{kekem}{\B{ke\-kem}}, nothing (action).
\on{7}.~\HL{keve'o}{\B{ke\-ve\-'o}}, chaos.
\on{8}.~\HL{kawlo}{\B{kaw\-lo}}, not once.

% ke-
\hi
\HT{ke2}{\B{ke-}} \I{[kE.]} \emph{productive pref. on adj.s to form the opposite, unless there is a separate lexical item}.~\LINK{https://forum.learnnavi.org/language-updates/how-to-form-negativeopposite-adjectivesadverbs/}{ln} 

% ke'aw
\hi
\HT{ke'aw}{\B{ke\cp{'aw}}} \I{[kE."Paw]} \emph{adj.} divided, torn apart, strife\hyp{}ridden. 
(\ding{43}~\HL{ke}{\B{ke}}, \HL{'aw}{\B{'aw}})

% ke'u
\hi
\HT{ke'u}{\B{\cp{ke}'u}} \I{["kE.Pu]} \emph{pn., n.} nothing (\emph{requires negation of the verb}). 
(\emph{a.}) \B{Ke lä\-ngu fì\-po\-stì\-mì ke\-'u a lu mip.} I'm afraid there's nothing new in this post.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b} 
\B{Fì\-ta\-ron\-yu\-tsyìp ke tsun ke\-'ut sti\-vä'\-nì.} This little hunter can't catch anything.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} 
(\emph{b. idiom with the dative}, !stress change!) \B{Oeru ke\cp{'u}.} I don't care. [lit.: it's nothing to me]~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Nga new yi\-vom 'u\-pet fì\-txon? -- Ke\-tsran. Oe\-ru ke\-'u.} What do you want to eat tonight?  Whatever. I don't care.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Oe tì\-kang\-kem si trr\-txon nì\-wotx, nga\-ru ke\-'u!} I work all day and all night, and you don't give a damn!~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
(\ding{43}~\HL{'u}{\B{'u}}, \HL{ke}{\B{ke}})

% kea
\hi
\HT{}{\B{\cp{ke}a}} \I{["kE.a]} \emph{adj.} no (… \emph{noun}) (\emph{requires negation of the verb}). 
\B{slä vay set ke pa\-mä\-hä\-ngem kea tì\-'eyng}, but until now no answer arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ke lu po\-ru kea smu\-ke.} He doesn't have any sisters.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}

% kefpomronga'
\hi
\HT{kefpomronga'}{\B{kefpom\cp{ro}nga'}} \I{[kE.fpom."Ro.NaP]} \emph{adj., nfp.} (mentally) unhealthful.
\B{Ma En\-tu, fì\-kem rä\-'ä si\-vi; lu ke\-fpom\-ro\-nga'.} Entu, don't do this; it's not healthy (mentally).~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}, \ding{214}~\HL{fpomronga'}{\B{fpom\-ro\-nga'}})

% kefpomtokxnga'
\hi
\HT{kefpomtokxnga'}{\B{kefpom\cp{tokx}nga'}} \I{[kE.fpom."tok'.NaP]} \emph{adj., nfp.} (physcially) unhealthful.
(\ding{43}, \ding{214}~\HL{fpomtokxnga'}{\B{fpom\-tokx\-nga'}})

% keftxo
\hi
\HT{keftxo}{\B{ke\cp{ftxo}}} \I{[kE."ft'o]} \emph{adj.} unhappy, upset; (\emph{for inner feelings used with} \B{'efu}) \B{Krr a sto\-lawm Txe\-wì te\-ri hem a poe\-ru lo\-len, 'o\-le\-fu ke\-ftxo nì\-txan.} When Txewì heard about what happened to her, he was very upset.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Po ho\-lum a\-krr\-maw, oe 'e\-fu ke\-ftxo.} After he left, I was unhappy.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-986}{b} 
(\emph{a. idiom}) \B{Ke\-ftxo.} `How sad!' \B{Tìng na\-ri. Tsa\-ye\-rik\-tsyìp\-ìl li 'an\-la sa'\-nok\-it. Ke\-ftxo!} Look! That little hexapede is already yearning for its mother. How sad!~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\emph{b. idiom}) \B{Ngari keftxo (fwa) ke tok.} ``We missed you.\slash Sorry you couldn't make it.\slash Too bad you couldn't be there.''~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkeftxo}{\B{tì\-ke\-ftxo}}, sadness. 
\on{2}.~\HL{nIkeftxo}{\B{nì\-ke\-ftxo}}, unfortunately, sadly.

% kefyak
\hi
\HT{kefyak}{\B{ke\cp{fyak}}} \I{[kE."fjak\textcorner]} \emph{idiom, tag question} … right? … isn't it? (\emph{from} \B{ke fì\-fya srak})
\B{Nga lu So\-rewn, ke\-fyak?} You're Sorewn, right?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}

% kehe
\hi
\HT{kehe}{\B{\cp{ke}he}} \I{["kE.hE]} \emph{intj.} no. 
\B{Alo a\-mrr po\-an po\-lawm, slä fra\-lo poe pol\-txe san ke\-he.} He asked five times, but each time she said, `no.'~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Sra\-ke fnan ngal lì'\-fya\-ti le\-Na'\-vi?~Ke\-he.} Are you good at Na'vi?~No.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\emph{a. conv.}) \B{Ke\-he kaw\-'it!} `Not at all!'~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{214}~\HL{sran}{\B{sra\-ne}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kesran}{\B{ke\-sran}}, mediocre. 
\on{2}.~\HL{srankehe}{\B{sran\-ke\-he}}, more or less.

% kekem
\hi
\HT{kekem}{\B{\cp{ke}kem}} \I{["kE.kEm]} \textsc{i}. \emph{adv.} no action or activity, nothing. \\
\textsc{ii}. \HT{kekem si}{\B{\cp{ke}kem ke si}} \I{["kE.kEm kE si]} \emph{vin.} do nothing.
\B{Ay\-nge\-yä ki\-fkey\-ti fol kawng\-sar nì\-tut, fì\-'u\-ri ke\-kem ke si ay\-nga.} They continuously exploit your world and you do nothing about it.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{ke}{\B{ke}}, \HL{kem}{\B{kem}})

% Kekunan
\hi
\B{\cp{Ke}kunan} \I{["kE.ku.nan]} \emph{n.} \emph{clan name, specialise in banshee riding}.

% kelemweypey
\hi
\HT{kelemweypey}{\B{kelem\cp{wey}pey}} \I{[kE.lEm."wEj.pEj]} \emph{adj.} impatient.
(\ding{214}, \ding{43}~\HL{lemweypey}{\B{lem\-wey\-pey}})

% kelfpomron
\hi
\HT{kelfpomron}{\B{kelfpom\cp{ron}}} \I{[kEl.fpom."Ron]} \emph{adj., ofp.} (mentally) unhealthy.
\B{Ke tsun nga tì\-kxìm si\-vi oer. Lu nga kel\-fpom\-ron!} You can't order me around. You're mentally unsound!~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}, \ding{214} \HL{lefpomron}{\B{le\-fpom\-ron}})

% kelfpomtokx
\hi
\HT{kelfpomtokx}{\B{kelfpom\cp{tokx}}} \I{[kEl.fpom."tok']} \emph{adj.} (physcially) unhealthy.
(\ding{43}, \ding{214} \HL{lefpomtokx}{\B{le\-fpom\-tokx}})

% kelha'
\hi
\HT{kelha'}{\B{kel\cp{ha'}}} \I{[kEl."haP]} \emph{adj.} inappropriate, unsuitable, unfitting.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/\#comment-31844}{b}
(\ding{43}~\HL{leha'}{\B{le\-ha'}})

% kelhoan
\hi
\HT{kelhoan}{\B{kel\cp{ho}an}} \I{[kEl."ho.an]} \emph{adj.} uncomfortable.
\B{Lä\-ngu fì\-seyn kel\-ho\-an nì\-ngay. Tsun oe hi\-veyn tsaw\-sìn 'a\-'aw\-a swaw\-tsyìp nì\-'aw.} This chair is really uncomfortable. I can only sit on it a few seconds.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\ding{43}~\HL{hoan}{\B{ho\-an}}, \ding{214}~\HL{lehoan}{\B{le\-ho\-an}})

% kelkin
\hi
\HT{kelkin}{\B{kel\cp{kin}}} \I{[kEl."kin]} \emph{adj.} unnecessary.  (\ding{43}, \ding{214}~\HL{lekin}{\B{le\-kin}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIkelkin}{\B{nì\-kel\-kin}}, unnecessarily.

% kelku
\hi
\HT{kelku}{\B{\cp{kel}ku}} \I{["kEl.ku]} \textsc{i}. \emph{n.} home, house. 
\B{Lu me\-nge\-yä kel\-ku na'\-rìng\-ä lu\-ke kxu a\-txan a fì\-'u fmawn a\-sìl\-tsan lu nì\-ngay.} It's truly good news that your forest house is without much harm.\LINK{http://forum.learnnavi.org/language-updates/luke-kxu-without-harm/}{ln} 
\B{Aw\-ngal tok kel\-kut.} We're home.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
\B{Ay\-nga\-ri li\-vu hel\-ku sang ul\-te te'\-lan le\-fpom.} May your homes be warm and your hearts be happy.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} 
\B{Ut\-ral a lu few pay\-fya a eo kel\-ku oe\-yä tsawl lu nì\-txan.} The tree on the other side of the stream in front of my house is very tall.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\
\textsc{ii}. \HT{kelku si}{\B{\cp{kel}ku si}} \emph{vin.} live, dwell. 
\B{Mì sray apxa kel\-ku si so\-a\-i\-a\-hu.} (He) lives with his family in a large village.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Tsun aw\-nga kel\-ku si\-vi nì\-'eng Saw\-tu\-te\-hu mì atx\-kxe aw\-nge\-yä.} We can share our land with the Skypeople.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{kelutral}{\B{Ke\-lut\-ral}}, Hometree.  
\on{2}.~\HL{prrku}{\B{prr\-ku}}, womb.

% kelpxìmrun
\hi 
\HT{kelpxImrun}{\B{kel\cp{pxìm}run}} \I{[kEl."p'Im.Run]} \emph{or} \I{[.RUn]} \emph{adj.} rare. 
(\ding{43}~\HL{lepxImrun}{\B{le\-pxìm\-run}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIkelpxImrun}{\B{tì\-kel\-pxìm\-run}}, rarity, rareness. 

% kelsar
\hi
\HT{kelsar}{\B{kel\cp{sar}}} \I{[kEl."saR]} \emph{adj.} useless, in vain. 
(\ding{214}~\HL{lesar}{\B{le\-sar}})

% keltrrtrr
\hi 
\HT{keltrrtrr}{\B{kel\cp{trr}trr}} \I{[kEl."t\s{r}.t\s{r}]} \emph{adj.} unusual.  
\B{Kel\-trr\-trr\-a tì'eylan.} An unusual friendship.~\LINK{http://naviteri.org/2020/04/keltrrtrra-tieylan-an-unusual-friendship/}{b}
(\ding{43}~\HL{ke}{\B{ke}}, \HL{letrrtrr}{\B{le\-trr\-trr}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{txankeltrrtrr}{\B{txan\-kel\-trr\-trr}}, extraordinary.

% keltsun
\hi
\HT{keltsun}{\B{kel\cp{tsun}}} \I{[kEl."\t{ts}un]} \emph{or} \I{[."\t{ts}Un]} (\textsc{rn}: \B{kel\cp{tsùn}} \I{[kEl."\t{ts}Un]}) \emph{adj.} impossible. 
\B{Fwa fpìl nì\-fya\-'o a\-ve\-nga' fì\-tì\-zin\-mì lu kel\-tsun.} It's impossible to think clearly (i.e., in an organized manner) in this muddled situation.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/\#comment-33483}{b}
(\ding{214}~\HL{letsunslu}{\B{le\-tsun\-slu}}) 

% Kelutral
\hi
\HT{kelutral}{\B{\cp{Kel}utral}} \I{["kEl.ut\textcorner.Ral]} \emph{n.} Hometree (\emph{home of the Omaticaya clan}). 
\B{Kel\-ut\-ral lu fne\-ut\-ral a\-zey.} Hometree is a special kind of tree.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Saw\-tu\-te ze\-ra\-'u fte fol Kel\-ut\-ral\-ti ski\-va\-'a.} The Skypeople are coming to destroy Hometree.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{wiki} 
\B{mì hil\-van a lok Kel\-ut\-ral}, in the river near Hometree.~\LINK{http://naviteri.org/2010/07/thoughts-on-ambiguity/}{b}
(\ding{43}~\HL{kelku}{\B{kel\-ku}}, \HL{utral}{\B{ut\-ral}})

% kem
\hi
\HT{kem}{\B{kem}} \I{[kEm]} \textsc{i}. \emph{n.} action, deed. 
\B{Kea kem le\-yew\-la rä\-'ä si, ru\-txe.} Please, don't let me down.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Krr a sto\-lawm Txe\-wì te\-ri hem a poe\-ru lo\-len, 'o\-le\-fu ke\-ftxo nì\-txan.} When Txewì heard about what happened to her, he was very upset.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Hem le\-'en tsun le\-hrr\-ap li\-vu.} Speculative moves can be dangerous.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Krr\-o len ay\-hem a\-na\-fì\-'u.} Sometimes such things happen.~\LINK{http://forum.learnnavi.org/language-updates/just-another-few-words/}{ln} 
\B{Saw\-tu\-te\-yä hem\-ìri lu aw\-nga\-ru ya\-yayr.} The Skypeople's actions confuse us.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\emph{idiom}) \B{Kem\-ìri a nga\-ru prr\-te' ke lu, tsa\-kem rä\-'ä si\-vi ay\-la\-he\-ru.} As for any action that you don't like, don't act that way to others.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} 
(\emph{a. proverb}) \B{Hem a\-sru\-nga' nì\-'ul, hum a\-sngu\-nga' nì\-nän.}
More helpful actions lead to less troubling outcomes.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\
\textsc{ii}. \B{\cp{kem} si} \I{["kEm si]} \emph{vin.} do (an action).   \B{Kem si li trr\-ay\-sre.} Do it by tomorrow.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Oel tse\-'a kem\-it a tsa\-tse\-nge nga so\-li.} I see what you did there.~\LINK{http://forum.learnnavi.org/language-updates/just-another-few-words/}{ln} \\
% \textsc{iii}. \B{fì\cp{kem}} \I{[fI."kEm]} \emph{pn., n.} this (action), it.
% \B{Ru\-txe, ay\-nga\-ri fì\-kem fe\-yä 'e\-'a\-la to\-pur ka\-ngay sì\-yi nì\-'aw.} Please, this will only confirm their worst fears about you.~\LINK{http://forum.learnnavi.org/language-updates/two-words-in-use-eal-and-kangay-si/msg564284/\#msg564284}{ln} \\
% \textsc{iv}. \B{\cp{tsa}kem} \I{["\t{ts}a.kEm]} \emph{pn., n.} that (action), it.
% \B{Tsa\-kem kop krr\-nekx.} That also takes time.~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln} \\
% \textsc{v}. \B{\cp{kem}pe} \I{["kEm.pE]} \emph{inter.} what (action)?
% \B{Kem\-pe le\-ren?} What's happening?~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} \\
% \textsc{vi}. \B{\cp{kem}pe si} \I{["kEm.pE si]} \emph{inter.} do what (action)? 
% \B{Kem\-pe si nga, ma sa'\-nu\-tsyìp?} What are you doing, little mommy?~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{kempe}{\B{kem\-pe\slash pe\-hem}}, what (action). 
\on{2}.~\HL{fIkem}{\B{fì\-kem}}, this (action).
\on{3}.~\HL{tsakem}{\B{tsa\-kem}}, that (action).
\on{4}.~\HL{fekem}{\B{fe\-kem}}, accident. 
\on{5}.~\HL{tIkangkem}{\B{tì\-kang\-kem}}, work.
\on{6}.~\HL{kemlI'u}{\B{kem\-lì\-'u}}, verb.
\on{7}.~\HL{kawngkem}{\B{kawng\-kem}}, evil deed, crime.

% kemlì'u 
\hi
\HT{kemlI'u}{\B{\cp{kem}lì'u}} \I{["kEm.lI.Pu]} \emph{n.} verb. 
\B{Kem\-lì\-'ut alu var fkol sar pxel pum alu new.} You can use the verb `var' like (the verb) `new'.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-582}{b} 
(\ding{43}~\HL{kem}{\B{kem}}, \HL{lI'u}{\B{lì\-'u}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kemlI'uvi}{\B{kem\-lì\-'u\-vi}}, infix.

% kemlì'uvi
\hi
\HT{kemlI'uvi}{\B{\cp{kem}lì'uvi}} \I{["kEm.lI.Pu.vi]} \emph{n.} verb infix. 
\B{Tsa\-lì\-'u\-ri alu zey\-ko lu kem\-lì\-'u\-vi (lì\-'u a\-txan\-tsan nang!) mì kam\-tseng.} Concerning the verb `zeyko', there is an infix (excellent word!) in the middle.~\LINK{http://naviteri.org/2010/07/vocabulary-update/comment-page-1/\#comment-95}{b}
(\ding{43}~\HL{kemlI'u}{\B{kem\-lì\-'u}})

% kempe / si
\hi
\HT{kempe}{\B{\cp{kem}pe}} \emph{or} \B{pe\cp{hem}} \I{["kEm.pE]} \emph{or} \I{[pE."hEm]} \textsc{i}. \emph{inter.} what (action)?
\B{Kem\-pe le\-ren?} What's happening?~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} \\
\textsc{ii}. \B{\cp{kem}pe si} \emph{or} \B{pe\cp{hem} si} \I{["kEm.pE si]} \emph{or} \I{[pE."hEm si]} \emph{vin., inter.} do what (action)? 
\B{Kem\-pe si nga, ma sa'\-nu\-tsyìp?} What are you doing, little mommy?~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Pol po\-le\-'un fu\-ta pe\-hem si nì\-'en.} He decided what to do on a hunch.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}

% kemu'nitkan
\hi 
\HT{kemu'nitkan}{\B{ke\cp{mu'}nitkan}} \I{[kE."muP.nit\textcorner.kan]} \emph{or} \I{[."mUP.]} \emph{adj.} ineffective. 
(\ding{214}~\HL{mu'nitkan}{\B{mu'\-nit\-kan}}) 

% kemuia / si
\hi 
\HT{kemuia}{\B{ke\cp{mu}ia}} \I{[kE."mu.i.a]} \textsc{i}. \emph{n.} dishonor. \\
\textsc{ii}. \B{ke\cp{mu}ia si} \I{[kE."mu.i.a si]} \emph{vin.} dishonor.
\B{Nge\-yä tì\-fna\-we' ke\-muia so\-li fì\-soaia\-ru.} Your cowardice has dishonored this family.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
(\ding{214}~\HL{meuia}{\B{meuia}}; \ding{43}~\HL{meuia}{\B{meuia}}, \HL{ke}{\B{ke}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kemuianga'}{\B{ke\-mu\-ia\-nga'}}, dishonorable.

% kemuianga'
\hi 
\HT{kemuianga'}{\B{ke\cp{mu}ianga'}} \I{[kE."mu.i.a.NaP]} \emph{adj.} dishonorable. 
\B{Peyä hem\-ìl a\-ke\-muia\-nga' za\-mo\-lu\-nge fwìng\-it ay\-oer.} His dishonorable behavior (lit.: actions) humiliated us (lit.: brought humiliation to us).~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
(\ding{43}~\HL{kemuia}{\B{ke\-mu\-ia}}, \HL{-nga'}{\B{-nga'}})

% kemum
\hi 
\HT{kemum}{\B{ke\cp{mum}}} \I{[kE."mum]} \emph{or} \I{[."mUm]} (\textsc{rn}: \B{ke\cp{mùm}} \I{[kE."mUm]})~:~\HL{omum}{\B{omum}}

% kemwiä
\hi 
\HT{kemwiA}{\B{kem\cp{wi}ä}} \I{[kEm."wi.\ae]} \emph{adj.} improper, unfair, wrong, unjustified. 
\B{Lu kem\-wi\-ä!} It's unfair.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{214}~\HL{muiA}{\B{muiä}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkemwiA}{\B{tì\-kem\-wiä}}, unfairness, injustice.
\on{2}.~\HL{leymkem}{\B{leym\-kem}}, protest.

% ken
\hi 
\HT{ken}{\B{ken}}\textsuperscript{\on{1}} \I{[kEn]} \emph{adp.} despite, in spite of. 
\B{Ken tì\-naw\-ri pe\-yä, ke flo\-lä.} Despite her talent, she didn't succeed.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
\B{ken fwa lu por 'aw\-a na\-ri nì\-'aw, lu Ma\-ti ta\-ron\-yu a\-swey.} In spite of having only one eye, Ma\-ti is the best hunter.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}

% ken
\hi 
\HT{ken2}{\B{ken}}\textsuperscript{\on{2}} \I{[kEn]} \emph{vin.} behave assertively and with confidence. 
\B{Eyk\-tan a\-sìl\-tsan, ken ay\-skxe mì te'\-lan, zene ki\-ven fra\-krr.} A good leader, despite the stones in his heart, must always behave confidently.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b} 
(\emph{a. idiom}) \B{Kll\-kxem ki\-ven!} `Stand up straight!' [lit.: ``stand straight and act with confidence'']~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b}\slash \LINK{https://naviteri.org/2026/01/follow-up-to-txep-si-txeva/}{a\on{3}}

% ken'aw
\hi
\HT{ken'aw}{\B{ken\cp{'aw}}} \I{[kEn."Paw]} \emph{adv.} not only (\emph{together with} \ding{43}~\HL{slAkop}{\B{slä\-kop}}).
\B{Nge\-yä tsmu\-ke lu ken\-'aw lor slä\-kop ka\-nu.} Your sister is not only beautiful but also intelligent.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Ay\-nga\-ri sìl\-pey oe tsnì ken\-'aw fpom\-tokx slä\-kop fpom\-ron yi\-vo'.} I hope that not only your physical but also your mental health will be perfect.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{ke}{\B{ke}}, \HL{nI'aw}{\B{nì\-'aw}})

% kenong
\hi
\HT{kenong}{\B{\cp{ke}nong}} \I{["kE.noN]} \emph{vtr.} model, represent, exemplify.
\B{Fay\-hem\-ìl pe\-yä ke ke\-nong tì\-sayt a fkol fngo' po\-ta.} These actions of his do not represent the loyalty that is required of him.~\LINK{http://naviteri.org/2022/01/aawa-tipangkxotsyip-a-teri-horen-lifyaya-a-few-little-discussions-about-grammar/}{b}
 \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkenong}{\B{tì\-ke\-nong}}, example.

% kenten
\hi
\HT{kenten}{\B{\cp{ken}ten}} \I{["kEn.tEn]} \emph{n., fauna} fan lizard, \emph{fanisaurus pennatus}.
\B{Mìn, mìn, ken\-ten mìn. | Pe\-lun ke 'efu spxin?} Turn, turn, the fanlizard turns. | Why doesn't it feel sick?~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}
(\emph{a. idiom}) \B{(Na) ken\-ten mì kum\-pay}, being in an environment where one is prevented from acting naturally or doing what one wants to do, [lit.: (like) a fan lizard in gel] \B{Nar\-mew oe fo\-ru na'\-rìng\-ä tì\-lor\-it wi\-vìn\-txu, slä ke tsä\-ngun fo tsli\-vam. 'Efu oe na ken\-ten mì kum\-pay.} I wanted to show them the beauty of the forest, but sadly, they weren't able to understand. I felt completely stymied.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}

% kenzen
\hi
\HT{kenzen}{\B{ken\cp{zen}}} \I{[kEn."zEn]} \emph{adv.} not necessarily.
\B{Fay\-sä\-fpìl fay\-sì\-'e\-fu\-sì lu pum oe\-yä nì\-'aw, ken\-zen pum su\-te\-yä a za\-mo\-lu\-nge aw\-ngar rel\-it a\-ru\-sikx alu Unil\-tì\-ran\-tokx.} These thoughts and feelings are only mine, not necessarily of the people who brought us the movie \emph{Avatar}.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{nIzen}{\B{nì\-zen}})

% keng
\hi
\HT{keng}{\B{keng}} \I{[kEN]} \emph{adv.} even.
\B{Yom tey\-lut keng oel!} Even \emph{I} eat teylu!~\LINK{http://forum.learnnavi.org/language-updates/even-keng/}{ln} 
\B{Ngay\-txoa, krro tì\-kxey si keng kar\-yu.} Sorry, sometimes even the teacher messes up.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Ke su\-nu oer keng nì\-'it.} I don't like (them) one bit.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}

% keri
\hi 
\HT{keri}{\B{\cp{ke}ri}} \I{["kE.Ri]} \emph{adj.} fierce (\emph{of people or things}). 
\B{tsam\-si\-yu a\-ke\-ri}, fierce warrior or things.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b} 
\B{ay\-lì\-'u a\-ke\-ri}, fierce words.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b} 
(\emph{a. idiom}) \B{Ay\-lì\-'u a\-ke\-ri | Le\-kxu to tì\-tse\-ri.} Fierce words can be more harmful than you think.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b} (\emph{longer form:}) \B{Ay\-lì\-'u a\-ke\-ri tsun le\-kxu li\-vu nì\-'ul to tì\-tse\-ri.}

% kerìsmìsì
\hi 
\HT{kerIsmIsI}{\B{Ke\cp{rì}smìsì}} \I{[kE."RI.smI.sI]} (\emph{gen.} \B{Ke\-rì\-smì\-sä}) \textsc{lw}, \emph{n.} Christmas.
\B{Ftxo\-zä Ke\-rì\-smì\-sä Le\-fpom.} Merry Christmas.~\LINK{http://naviteri.org/2022/01/aawa-tipangkxotsyip-a-teri-horen-lifyaya-a-few-little-discussions-about-grammar/}{b}

% kerusey
\hi
\HT{kerusey}{\B{\cp{ke}rusey}} \I{["kE.Ru.sEj]} \emph{adj.} dead.
\B{He\-tu\-wong\-ìl aw\-nge\-yä swo\-tut ska\-'a, fte kll\-ki\-vu\-lat ke\-ru\-sey\-a tskxet.} The aliens destroy our sacred place to dig up dead rock.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{Krro krro, flì\-a vul a\-ru\-sey to nutx\-a pum a\-ke\-ru\-sey lu txur.} A thin living branch is sometimes stronger than a thick dead one.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 
(\ding{214}~\HL{rusey}{\B{ru\-sey}}) 

% kesätarenga'
\hi 
\HT{kesAtarenga'}{\B{kesä\cp{ta}renga'}} \I{[kE.s\ae."ta.RE.NaP]} \emph{or coll.} \I{[kE."{}sta.RE.NaP]} \emph{adj.} irrelevant. 
(\ding{214}~\HL{sAtarenga'}{\B{sä\-ta\-re\-nga'}})

% kesran
\hi
\HT{kesran}{\B{ke\cp{sran}}} \I{[kE."{}sRan]} \emph{adj.} so\hyp{}so, mediocre (in quality).
\B{Pe\-yä sä\-ftxu\-lì\-'u lo\-lä\-ngu ke\-sran ul\-te kaw\-tur slan\-ti\-re ke si.} Unfortunately his speech was only so-so and inspired no one.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Fu\-ri\-a tä\-ftxu ngal tok yì\-pet?~Tok yìt a\-ke\-sran.} How's your weaving?~I'm so-so.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{kehe}{\B{ke\-he}}, \HL{sran}{\B{sra\-ne}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIksran}{\B{nìk\-sran}}, in a mediocre manner.

% ketartu
\hi 
\HT{ketartu}{\B{ke\cp{tar}tu}} \I{[kE."taR.tu]} \emph{n.} outcast. 
(\ding{43}~\HL{tare}{\B{ta\-re}}) 

% keteng
\hi
\HT{keteng}{\B{\cp{ke}teng}} \I{["kE.tEN]} \emph{adj.} different. 
\B{Slä sa'\-nok\-ìri fpìl\-fya lu ke\-teng.} But mother's way of thinking is different.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
(\ding{214}~\HL{teng}{\B{teng}})  \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIketeng}{\B{tì\-ke\-teng}}, difference.
\on{2}.~\HL{ketengsyon}{\B{ke\-teng\-syon}}, variety, diversity.

% ketengsyon
\hi 
\HT{ketengsyon}{\B{\cp{ke}tengsyon}} \I{["kE.tEN.sjon]} \emph{n.} variety, diversity.
\B{Naw\-num\-tseng\-ä ay\-nu\-me\-yu\-ri ley ay\-nga\-ru tì\-teng\-wotx, ay\-oe\-ru ke\-teng\-syon.} Regarding university students, you value uniformity, we value diversity.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
(\ding{43}~\HL{keteng}{\B{keteng}}, \HL{syon}{\B{syon}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{leketengsyon}{\B{le\-ke\-teng\-syon}}, diverse, various.

% ketrìp
\hi 
\HT{ketrIp}{\B{\cp{ket}rìp}} \I{["kEt\textcorner.RIp\textcorner]} \emph{adj.} unfavorable, inauspicious. 
\B{Nga ket\-rìp\-a krr zo\-la\-'u; Ra\-lu set sti ul\-te ke new nga\-hu pi\-väng\-kxo.} You came at the wrong time; Ralu is angry and won't speak with you.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
(\ding{214}~\HL{etrIp}{\B{et\-rìp}})

% ketsran
\hi
\HT{ketsran}{\B{ke\cp{tsran}}} \I{[kE."\t{ts}Ran]} \textsc{i}. \emph{adj.} (\emph{to form compound relative pronouns, i.e.}) whatever, whoever, however. 
\B{Ke\-tsran\-a tu\-te a ni\-vew hi\-vum tsun tsa\-kem si\-vi.} Whoever wants to leave can do so.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{'U a\-ke\-tsran tsun ti\-vam.} Anything at all will be fine.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b} \\
\textsc{ii}. \emph{conj.} 
\B{Ke\-tsran tseng\-ne nga ki\-vä, kä oe tsa\-tseng nì\-teng.} Wherever you go, I'll go there too.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Ke\-tsran fya\-'o si\-vu\-nu ngar, kem si.} Do it however you'd like.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b} \\
\textsc{iii}. (\emph{idiom}) \B{Nga new yi\-vom 'u\-pet fì\-txon?--Ke\-tsran. Oe\-ru ke\-'u.} What do you want to eat tonight?--Whatever. I don't care.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
(\ding{43}~\HL{tsranten}{\B{tsran\-ten}})

% ketstun
\hi 
\HT{ketstun}{\B{ke\cp{tstun}}} \I{[kE."\t{ts}tun]} \emph{or} \I{[."\t{ts}tUn]} \emph{adj., ofp.} unkind, emotionally cold. 
(\emph{a. proverb}) \B{Tu\-te a\-ke\-tstun sli\-vu 'ey\-lan ke tsun.} An unkind person cannot be a friend.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b}  
(\ding{43}~\HL{tstunwi}{\B{tstun\-wi}}, \HL{tIktstunga'}{\B{tìk\-tstu\-nga'}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIktstun}{\B{tìk\-tstun}}, unkindness, emotional coldness.

% ketsuk-
\hi
\HT{ketsuk}{\B{ketsuk-}} \I{[kE.\t{ts}uk\textcorner.]} \emph{or} \I{[.\t{ts}Uk\textcorner.]} (\textsc{rn}: \B{ketsùk-} \I{[kE.\t{ts}Uk\textcorner.]}) \emph{prod. pref. on vtr. to form receptive ability, that s.t. is not capable of receiving the action of the verb}, ``un-…-able''. \B{ke\-tsuk\-tswa'}, unforgettable.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b} 
(\ding{214}~\HL{tsuk}{\B{tsuk-}})

% ketsukanom
\hi 
\HT{ketsukanom}{\B{ketsu\cp{ka}nom}} \I{[kE.\t{ts}u."ka.nom]} (\textsc{rn}: \B{ketsù\cp{ka}nom} \I{[kE.\t{ts}U."ka.nom]}) \emph{adj.} unavailable, unobtainable. 
(\ding{43}~\HL{ketsuk}{\B{ketsuk-}}, \HL{kanom}{\B{ka\-nom}}; \ding{214}~\HL{tsukanom}{\B{tsu\-ka\-nom}})

% ketsuklewn
\hi 
\HT{ketsuklewn}{\B{ketsuk\cp{lewn}}} \I{[kE.\t{ts}uk\textcorner."lEwn]} \emph{or} \I{[.\t{ts}Uk\textcorner.]} (\textsc{rn}: \B{ketsùk\cp{lewn}} \I{[kE.\t{ts}Uk\textcorner."lEwn]}) \emph{adj.} intolerable, unaccaptable, unbearable. 
\B{Tsa\-fne\-vo\-ìk lu ke\-tsuk\-lewn.} That kind of behavior is intolerable.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(\ding{43}~\HL{ketsuk}{\B{ke\-tsuk-}}, \HL{lewn}{\B{lewn}}) 

% ketsuktiam
\hi
\HT{ketsuktiam}{\B{ketsuk\cp{ti}am}} \I{[kE.\t{ts}uk\textcorner."ti.am]} \emph{or} \I{[.\t{ts}Uk\textcorner.]} (\textsc{rn}: \B{ketsùk\cp{ti}am} \I{[kE.\t{ts}Uk\textcorner]."ti.am}) \emph{adj.} uncountable (\emph{for countable nouns}).
\B{Hol\-pxay san\-hì\-yä a mì saw lu ke\-tsuk\-ti\-am; keng ke tsun fko tsi\-ve\-'a sat nì\-wotx.} The number of stars in the sky is infinite; it's not even possible to see them all.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{tiam}{\B{ti\-am}}, \HL{txewluke}{\B{txew\-lu\-ke}})

% ketsuktswa'
\hi
\HT{ketsuktswa'}{\B{ketsuk\cp{tswa'}}} \I{[kE.\t{ts}uk\textcorner."\t{ts}waP]} \emph{or} \I{[.\t{ts}Uk\textcorner.]} (\textsc{rn}: \B{ketsùk\cp{tswa'}} \I{[kE.\t{ts}Uk\textcorner."\t{ts}waP]}) \emph{adj.} unforgettable.
(\ding{43}~\HL{tswa'}{\B{tswa'}})

% ketuwong
\hi
\HT{ketuwong}{\B{\cp{ke}tuwong}} \I{["kE.tu.woN]} \emph{n.} alien (person, creature).
\B{Txo\-pu rä\-'ä si, lu ke\-tu\-wong\-o nì\-'aw.} Don't be afraid, it's only an alien.~\LINK{http://forum.learnnavi.org/language-updates/a-collection/}{ln\slash a\textsuperscript{c}}
\B{He\-tu\-wong\-ìl aw\-nge\-yä swo\-tut ska\-'a, fte kll\-ki\-vu\-lat ke\-ru\-sey\-a tskxet.} The aliens destroy our sacred place to dig up dead rock.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{Tsa\-swi\-rä\-ti lo\-nu! Ay\-nga ne\-to ri\-vikx! Fì\-ke\-tu\-wong\-ti oel stì\-ye\-ftxaw.} Release this creature! Step back! I will look at this alien.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{wiki\slash a}
(\ding{43}~\HL{kewong}{\B{ke\-wong}})

% kew
\hi
\HT{kew}{\B{kew}} \I{[kEw]} \emph{num.} zero, 0. 

% key
\hi
\HT{key}{\B{key}} \I{[kEj]} \emph{n.} face. 
\B{Oe\-ri key tun slo\-lu nì\-txan.} My face became very red.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-694}{b}
\B{Nga\-ri key ne\-re\-i\-yrr, ma 'ey\-lan.} Your face is glowing, friend.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{keyrel}{\B{key\-rel}}, facial expression.

% keykur
\hi
\HT{keykur}{\B{key\cp{kur}}} \I{[kEj."k$\cdot$uR]} \emph{or} \I{[."kUR]} (\textsc{rn}: \B{key\cp{kùr}} \I{[kEj."kUR]}) \emph{vtr., inf.}\on{22} hang (on, onto, \ding{43}~\HL{sIn}{\B{sìn}}), let hang.
\B{'A\-li\-'ät vul\-sìn key\-kur za\-'u fì\-tseng.} Hang the choker on the branch and come here.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{kur}{\B{kur}})

% keyn
\hi 
\HT{keyn}{\B{keyn}} \I{[kEjn]} \emph{vtr.} put down. 
\B{Ngey tsko\-ti ki\-veyn. Li ye\-rik ho\-li\-fwo.} Put down your bow. The hexapede has already run away.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
(\ding{214}~\HL{kxeltek}{\B{kxeltek}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{keynven}{\B{keyn\-ven}}, step.

% keynven
\hi 
\HT{keynven}{\B{keyn\cp{ven}}} \I{[k$\cdot$Ejn."vEn]} \emph{vin., inf.}\on{11} step. 
\B{Na\-ri si teng\-krr ke\-reyn\-ven fì\-tseng. Lu kll\-te ekx\-txu.} Step carefully here. The ground is rough.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b}
\B{Na\-ri si keyn\-ven!} Step carefully!~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b}
(\emph{a. idiom}) \B{Po keyn\-ven sìn ketse.} He is socially awkward. (\emph{lit.}: `He steps on tails.')~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{keyn}{\B{keyn}}, \HL{venu}{\B{ve\-nu}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sAkeynven}{\B{sä\-keyn\-ven}}, step.

% keyrel
\hi
\HT{keyrel}{\B{\cp{key}rel}} \I{["kEj.rEl]} \emph{n.} facial expression.
\B{O\-lo\-me\-i\-um oel fa key\-rel fu\-ta lu yaw\-ne oe po\-ru.} I knew by the expression on her face that she loves me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{key}{\B{key}}, \HL{rel}{\B{rel}})

% keve'o
\hi 
\HT{keve'vo}{\B{ke\cp{ve}'o}} \I{[kE."vE.Po]} \emph{n.} chaos, disorder. 
\B{Nge\-yä ay\-sä\-fpìl\-mì tse\-'ä\-nga oel ke\-ve\-'ot nì\-'aw.} In your thinking, I see only chaos.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{ve'o}{\B{ve\-'o}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lekve'o}{\B{lek\-ve\-'o}}, chaotic. 
\on{2}.~\HL{kekve'otu}{\B{ke\-ve\-'o\-tu}}, chaotic. 

% keve'otu
\hi 
\HT{keve'otu}{\B{ke\cp{ve}'otu}} \I{[kE."vE.Po.tu]} \emph{n.} troublemaker, \emph{one who habitually creates chaos and disorder}. 
\B{Ay\-sä\-leym\-fe'\-it pe\-yä yu\-ne rä\-'ä. Lu po ke\-ve\-'o\-tu nì\-'aw.} Don't listen to his complaints. He's just a troublemaker.~\LINK{https://naviteri.org/2025/03/conlang-adventure-and-more/}{b} 
(\ding{43}~\HL{keve'o}{\B{ke\-ve\-'o}})

% kewan
\hi 
\HT{kewan}{\B{\cp{ke}wan}} \I{["kE.wan]} \emph{n.} age. 
\B{Nge\-yä ke\-wan pìm\-txan?} How old are you? [formal and stiff]~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\emph{a. proverb}) \B{Koakturi kewanti keyìl ke wan.} Some things can't be covered up. [lit.: `An old person's face doesn't hide their age.']~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{koak}{\B{ko\-ak}}, \HL{'ewan}{\B{'e\-wan}})

% kewong
\hi
\HT{kewong}{\B{\cp{ke}wong}} \I{["kE.woN]} \emph{adj.} alien, unknown.
\B{Oe\-ri ta pe\-yä fa\-hew a\-ke\-wong on\-tu te\-ya lä\-ngu.} My nose is full of his alien smell.~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{ketuwong}{\B{ke\-tu\-wong}}, alien.

% keyawr
\hi
\HT{keyawr}{\B{ke\cp{yawr}}} \I{[kE."jawR]} \emph{adj.} wrong, incorrect. 
\B{Ul\-te tsa\-lì'\-fya\-vi nge\-yä ke lu ke\-yawr kaw\-'it.} And your expression isn't wrong at all.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-995}{b}
(\ding{214}~\HL{eyawr}{\B{eyawr}})

% keye'ung
\hi
\HT{keye'ung}{\B{ke\cp{ye}'ung}} \I{[kE."jE.PuN]} \emph{or} \I{[.PUN]} (\textsc{rn}: \B{ke\cp{ye}ùng} \I{[kE."jE.UN]}) \emph{n.} insanity. 
\B{Ta\-fral ke lì\-syek oel nge\-yä ke\-ye\-'ung\-it.} Therefore I will not heed your insanity.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g}

% kezemplltxe
\hi
\HT{kezemplltxe}{\B{kezempll\cp{txe}}} \I{[kE.zEm.p\s{l}."t'E]} \emph{adv.} of course, needless to say. 
\B{New Va'\-ru tskot Ey\-tu\-kan\-ä za\-sri\-vìn.~Ke\-zem\-pll\-txe pa\-yll\-txe san ke\-he.} Va'ru wants to borrow Eytukan's bow.~Of course he'll say no.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{ke}{\B{ke}}, \HL{zene}{\B{ze\-ne}}, \HL{plltxe}{\B{pll\-txe}})

% kezin
\hi 
\HT{kezin}{\B{\cp{ke}zin}} \I{["kE.zin]} \emph{adj.} untangled. 
(\ding{214}~\HL{zin}{\B{zin}}, \HL{ke}{\B{ke}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIkezin}{\B{tì\-ke\-zin}}, s.t. in an untangled state; solution.

% ki
\hi
\HT{ki1}{\B{ki}} \I{[ki]} \emph{conj. with} \ding{43}~\HL{ke}{\B{ke}} \emph{to mean} not … but rather, not … but instead. 
\B{Nga pll\-txe ke nì\-fyeyn\-tu ki nì\-'e\-veng.} You speak not like an adult but a child.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Tsyokx\-ìri ke lu zek\-wä a\-tsìng ki a\-mrr.} The hand doesn't have four but five digits.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}

% ki-
\hi
\HT{ki2}{\B{ki-}} \I{[ki.]} \emph{num., pref.} seven (\emph{base to form higher numbers or fractions}). 
\B{\cp{ki}vol}, \on{56} (decimal), \on{70} (octal); 
\B{ki\cp{pxì}}, one-seventh, $\frac{1}{7}$; 
\B{ki\-pxì a\-tsìng}, four-seventh, $\frac{4}{7}$.
(\ding{43}~\HL{kinA}{\B{ki\-nä}})

% kian
\hi 
\HT{kian}{\B{ki\cp{an}}} \I{[ki."{}an]} \emph{vtr.} blame. 
\B{Oe\-ti kian rä\-'ä! Ke no\-lui oe!} Don't blame me! It wasn't my fault.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} \\
\textsc{Derivatives}: \on{1}.~\HL{kiantu}{\B{ki\-an\-tu}}, blameworthy person.

% kiantu
\hi 
\HT{kiantu}{\B{ki\cp{an}tu}} \I{[ki."{}an.tu]} \emph{n.} blameworthy person. 
\B{Fì\-tì\-snaytx\-ìri lu nge\-yä tsmu\-kan ki\-an\-tu.} Your brother is to blame for this loss.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b}
(\ding{43}~\HL{kian}{\B{ki\-an}})

% kifkey
\hi
\HT{kifkey}{\B{ki\cp{fkey}}} \I{[ki."fkEj]} \emph{n.} the (physical, solid) world. 
\B{Pll\-txe Saw\-tu\-te san ki\-fkey\-ìl ay\-oe\-yä tok txan\-a ngip\-it a san\-hì\-kip.} The Sky People say that their world is in the great space among the stars.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{tì\-kang\-kem si\-vi fo\-hu a Unil\-tì\-ran\-tokx\-it sì ki\-fkey\-ti Ey\-wa\-'e\-veng\-ä za\-mo\-lu\-nge aw\-ngar}, working with those that brought us \emph{Avatar} and the World of Pandora.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ay\-wayl yìm ki\-fkey\-ä, 'ì\-he\-yut a\-vo\-mrr}, The Songs bind the thirteen spirals of the solid world.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Oe new si\-vop ne tsa\-ki\-fkey le\-ron\-srel.} I want to journey to that imaginary world.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
(i.) \B{tì\-fti\-a ki\-fkey\-ä}, science [lit.: the study of the physical world].~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Science_Magazine}{wiki} 
(ii.) \B{tì\-fti\-a\-tu ki\-fkey\-ä}, scientist.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{fkeytok}{\B{fkey\-tok}}, exist.

% kilvan
\hi
\HT{kilvan}{\B{kil\cp{van}}} \I{[kil."van]} \emph{n.} river. 
\B{kil\-van a\-reng}, shallow river; \B{kil\-van a\-txukx}, deep river; \B{ngima kilvan}, long river. 
\B{Na\-ri si, ma 'itan. Kil\-van tswe\-sya si nì\-win nì\-txan.} Be careful, son. The river is flowing very swiftly.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
\B{Kil\-van\-ä tì\-fkey\-tok lu fya\-pe fì\-trr?} What's the condition of the river today?~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Kil\-van\-ìri tsa\-tseng slo\-snep\-pe?} How wide is the river there?~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}

% kin
\hi
\HT{kin}{\B{kin}} \I{[kin]} \emph{vtr.} need. 
\B{Tì\-ta\-ron\-ìri lä\-ngu tì\-kak\-pam fe\-kum. Kin fkol fra\-inan\-fyat.} I'm sorry to say that deafness is a disadvantage for hunting. You need all your senses.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
\B{Po 'e\-fu ngeyn ul\-te kin tì\-rawn\-it nì\-txan.} He is tired and very much needs to be replaced.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIkin}{\B{tì\-kin}}, (the) need. 
\on{2}.~\HL{lekin}{\B{le\-kin}}, necessary. 

% kinam
\hi
\HT{kinam}{\B{ki\cp{nam}}} \I{[ki."nam]} \emph{n.} leg.
\B{Ut\-ral\-ä a\-nawm | Ay\-ri\-na' lu ay\-oeng | A pe\-yä tì\-txur mì hi\-nam aw\-nge\-yä.} We are all seeds | Of the Great Tree | Whose strength is in our legs.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kinamtil}{\B{ki\-nam\-til}}, knee.

% kinamtil
\hi
\HT{kinamtil}{\B{ki\cp{nam}til}} \I{[ki."nam.til]} \emph{n.} knee. (\ding{43}~\HL{kinam}{\B{kinam}}, \HL{til}{\B{til}})

% kinä
\hi
\HT{kinA}{\B{\cp{ki}nä}} \I{["ki.n\ae]} \emph{num.} seven. (\emph{base}: \ding{43}~\HL{ki2}{\B{ki-}}, \emph{rest}: \ding{43}~\HL{hin}{\B{\mbox{-hin}}}) 
\B{Maria kxam\-lä na'\-rìng kä a zì\-sìt\-o a\-ki\-nä ke lal\-mu tsar kea rìk.} Maria went through the forest that had no leaf for seven years.~\LINK{http://forum.learnnavi.org/language-updates/for-\%28duration\%29-jesus-loan-word/}{ln} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kintrr}{\B{kin\-trr}}, (\on{7}-day) week. 
\on{2}.~\HL{vospxIkin}{\B{Vo\-spxì\-kin}}, July.

% kintrr
\hi
\HT{kintrr}{\B{\cp{kin}trr}} \I{[kin.t\s{r}]} \emph{n., adv.} (\on{7}-day) week.
\B{Maw fpxa\-mo\-a kin\-trr a\-fpxa\-mo mì tan\-lo\-kxe oe\-yä …} After a truly horrible week in my country … \LINK{http://naviteri.org/2022/05/results-of-the-tttc/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\B{kintrr\cp{am}} \emph{n., adv.}~last week. 
\B{Kin\-trr\-am fay\-fam\-rel\-si\-yu\-hu oe ul\-txa so\-le\-i\-yi.} Last week I met with those writers.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
\B{Kin\-trr\-am\-ä sä\-tsyìl lu le\-hrr\-ap slä 'o' nì\-txan.} Last week's climb was dangerous but very exciting.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\  
\on{2}.~\B{kintrr\cp{ay}}  \emph{n., adv.}~next week.
\B{Zun nga srung sil\-vi oer, zel ke kì\-ye\-vä oe ne Wa\-syìng\-ton kin\-trr\-ay.} If you had helped me, I wouldn’t be going to Washington next week.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
(\ding{43}~\HL{kinA}{\B{ki\-nä}}, \HL{trr}{\B{trr}})

% kinglor
\hi 
\HT{kinglor}{\B{\cp{king}lor}} \I{["kiN.loR]} \emph{n., fauna} kinglor. 
(\ding{43}~\HL{kIng}{\B{kìng}}, \HL{lor}{\B{lor}}) 

% kip
\hi
\HT{kip}{\B{kip}} \I{[kip\textcorner]} \emph{adp.} among.
\B{pi\-za\-yu Ney\-ti\-ri\-yä txan\-rar\-mo\-'a fra\-to kip ay\-ha\-pxì\-tu O\-ma\-ti\-ka\-ya\-ä}, Neytiri's ancestor was the most famous among the members of the Omaticaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}   
\B{Tì\-'e\-fu\-mì oe\-yä su\-te\-kip nì\-wotx, ftxey Na'\-vi ftxey Saw\-tu\-te, lu sìl\-tsan lu kawng.} In my opinion among all people, whether Na'vi or Skypeople, there are good and bad.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{fkip}{\B{fkip}}, up among. 
\on{2}.~\HL{takip}{\B{ta\-kip}}, from among 

% kipxì
\hi
\HT{kipxI}{\B{ki\cp{pxì}}} \I{[ki."p'I]} \emph{num., n.} one\hyp{}seventh, $\frac{1}{7}$. 
(\ding{43}~\HL{ki2}{\B{ki-}}, \HL{pxI}{\B{-pxì}})

% kiri
\hi 
\HT{kiri}{\B{\cp{Ki}ri}} \I{["ki.Ri]} \emph{n. female name}.  

% kive
\hi
\HT{kive}{\B{\cp{ki}ve}} \I{["ki.vE]} \emph{ord. num., adj.} seventh. 

% kivol
\hi
\HT{kivol}{\B{\cp{ki}vol}} \I{["ki.vol]} \emph{num., adj.} \on{56} (decimal), \on{70} (octal). 
(\ding{43}~\HL{ki2}{\B{ki-}}, \HL{vol}{\B{vol}})

% kì'ong
\hi
\HT{kI'ong}{\B{kì\cp{'ong}}} \I{[kI."PoN]} \emph{adj.} slow. 
(\emph{a. idiom}) \B{Kxe\-tse kì\-'ong!} ``Don't get angry.\slash Calm down.'' [lit.: slow tail] (\emph{short for} \B{Nga\-ri kxe\-tse kì\-'ong li\-vu.}).~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{214}~\HL{win}{\B{win}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIk'ong}{\B{nìk\-'ong}}, slowly.

% kìm
\hi
\HT{kIm}{\B{kìm}} \I{[kIm]} \emph{vtr.} spin. 
\B{Pol rum\-it ko\-lìm.} He spun the ball.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln}

% kìmar
\hi
\HT{kImar}{\B{kì\cp{mar}}} \I{[kI."maR]} \emph{adj.} (\emph{a.~of foods, vegetable or animal}) in season. 
\B{Tey\-lu kì\-mar lì\-yu a fì\-'u oe\-ru te\-ya si.} It fills me with joy that \emph{teylu} is about to be in season.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{b.~of events and situations that occur at a convenient or appropriate time}) \B{Kì\-mar lu fwa fo ke päng\-kxo oe\-hu.} It was opportune that they didn't chat with me.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIkmar}{\B{nìk\-mar}}, convenient, in time.

% kìng
\hi
\HT{kIng}{\B{kìng}} \I{[kIN]} \emph{n.} thread. 
\B{na way\-te\-lem\-ä hìng}, like the threads of a song cord.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} 
\B{Lä\-ngu fay\-hìng zin nì\-wotx; ke tsun sat si\-var.} Unfortunately, these threads are all tangled up; they can't be used.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lI'ukIng}{\B{lì\-'u\-kìng}}, sentence.

% kìte'e
\hi
\HT{kIte'e}{\B{kìte\cp{'e}}} \I{[kI.tE."PE]} \textsc{i}. \emph{n.} service. \\
\textsc{ii}. \B{kìte\cp{'e} si} \I{[kI.tE."PE si]} \emph{vin.} serve, be of service. \B{Eo nge\-nga kll\-kxem ohe, a\-lak\-si (fte) Oma\-ti\-ka\-ya\-ru kì\-te\-'e si\-vi.} I stand before you, ready to serve the Omaticaya people.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie}{wiki\slash a}

% kìyevame
\hi
\HT{kIyevame}{\B{kìye\cp{va}me}} \I{[kI.jE."va.mE]} \emph{idiom} \textsc{See} you again soon! Goodbye! 
\B{Kì\-ye\-va\-me ul\-te Ey\-wa nga\-hu.} Goodbye for now, and may Eywa be with you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\ding{43}~\HL{kame}{\B{ka\-me}})

% kllfro'
\hi
\HT{kllfro'}{\B{kll\cp{fro'}}} \I{[k\s{l}."fR$\cdot$oP]} \emph{vin., inf.}\on{22} be responsible for (\emph{with the topic}), having s.o. or s.t. as a job, duty, or area of concern. 
\B{Po\-ri ze\-ne kll\-fri\-vo' nga.} He is your responsibility [lit.: you are responsible for him].~\LINK{http://forum.learnnavi.org/language-updates/misc-answers/}{ln\slash a\textsuperscript{c}}
\B{Ye'\-rìn 'ì\-yi'\-a sä\-nu\-me a tsa\-ri kll\-fro' oe.} Soon the teaching that I am responsible for will end.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}

% kllkä
\hi
\HT{kllkA}{\B{kll\cp{kä}}} \I{[k\s{l}."k$\cdot$\ae]} \emph{vin., inf.}\on{22} (\emph{a.}) go down, descend. 
(\emph{b. with} \HL{ftu}{\B{ftu}}) get off, dismount.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
(\emph{c. for astronomical bodies setting})
(\ding{214}~\HL{fAkA}{\B{fä\-kä}}, \ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{kA}{\B{kä}}, \HL{hum}{\B{hum}}) 

% kllkulat
\hi
\HT{kllkulat}{\B{kll\cp{ku}lat}} \I{[k\s{l}."k$\cdot$u.l$\cdot$at\textcorner]} \emph{vtr., inf.}\on{23} dig up.  
\B{He\-tu\-wong\-ìl aw\-nge\-yä swo\-tut ska\-'a, fte kll\-ki\-vu\-lat ke\-ru\-sey\-a tskxet.} The aliens destroy our sacred place to dig up dead rock.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln\slash a\textsuperscript{c}}
(\ding{214}~\HL{kllyem}{\B{kll\-yem}}, \ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{kulat}{\B{ku\-lat}}) 

% kllkxem
\hi
\HT{kllkxem}{\B{kll\cp{kxem}}} \I{[k\s{l}."k'$\cdot$Em]} \emph{vin., inf.}\on{22} stand (\HL{mI}{\B{mì}}, in; \HL{sIn}{\B{sìn}}, on; \HL{IlA}{\B{ìlä}}, in shape\slash form of). 
(\emph{a. locally}) \B{Tsa\-ye\-rik kll\-kxem sìn yì a\-kxayl.} That \emph{yerik} is standing on a high ledge.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Kll\-kxa\-yem fì\-tì\-kang\-kem oe\-yä ro\-fa--ke io--pum fe\-yä.} My work will stand beside--not above--theirs.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{oe\-yä ay\-syu\-lang set mì fay kll\-kxe\-rem}, my flowers are standing in water now.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-296}{b}
\B{Na'\-vi ìlä ho\-'on kll\-kxo\-lem teng\-krr re\-rol.} The People were standing in a circle, singing.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\emph{b. metaphorically with} \B{sìn yì} \emph{to describe anything scalable, e.g. water level, temperature, talent, anger, etc.}) \B{Lì'\-fya\-ri po kll\-kxem sìn pe\-yì?} How good is her Na'vi?~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\emph{c. idiom}) \B{Kll\-kxem ki\-ven!} `Stand up straight!' [lit.: ``stand straight and act with confidence'']~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b}
(\ding{43}~\HL{kxem}{\B{kxem}})

% kllpa
\hi
\HT{kllpa}{\B{\cp{kll}pa}} \I{["k\s{l}.pa]} \emph{n.} bottom. 
(\ding{214}~\HL{fApa}{\B{fä\-pa}}, \ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{pa'o}{\B{pa\-'o}}) 

% kllpä 
\hi
\HT{kllpA}{\B{kll\cp{pä}}} \I{[k\s{l}."p$\cdot$\ae]} \emph{vin., inf.}\on{22} land, arrive at the ground. 
\B{Maw sä\-tsway\-on a\-yol ay\-oe kll\-po\-lä mì ta\-yo a lu ro\-fa kil\-van.} After a short flight we landed in a field beside the river.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\B{kllwo}, \HL{kllte}{\B{kll\-te}}, \HL{pAhem}{\B{pä\-hem}})

% kllpxiwll
\hi
\HT{kllpxiwll}{\B{kll\cp{pxi}wll}} \I{[k\s{l}."p'i.w\s{l}]} \emph{n.}~\ding{96} lionberry, \emph{cynaroidia decumbens}.
(\ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{pxi}{\B{pxi}}, \HL{'ewll}{\B{'e\-wll}})

% kllpxìltu
\hi
\HT{kllpxIltu}{\B{kll\cp{pxìl}tu}} \I{[k\s{l}."p'Il.tu]} \emph{n.} territory. 
\B{ul\-te kop su\-teo a\-la\-he a tok fra\-kll\-pxìl\-tut ki\-fkey\-ä} and (there are) also some others who are in all territories of the world.~\LINK{http://forum.learnnavi.org/language-updates/ke-kawit/msg174572/\#msg174572}{ln}

% kllrikx
\hi
\HT{kllrikx}{\B{kll\cp{rikx}}} \I{[k\s{l}."Rik']} \emph{n.} earthquake. 
\B{Txe\-wì pll\-txe san kll\-rikx txewm la\-mu sìk.} Txewì says that the earthquake was frightening.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{rikx}{\B{rikx}})

% kllte
\hi
\HT{kllte}{\B{\cp{kll}te}} \I{["k\s{l}.tE]} \emph{n.} ground.  
\B{Kll\-te lu ekx\-txu.} The ground is rough.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Fwa ya\-wo ftu kll\-te to fwa tsway\-on ftu 'awkx lu ngä\-zìk.} Taking off from the ground is harder than flying off a cliff.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{a. idiom}) \B{ne kllte}, get down! \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nekll}{\B{ne\-kll}}, see below.
\on{2}.~\HL{kllrikx}{\B{kll\-rikx}}, earthquake.
\on{3}.~\HL{kllkA}{\B{kll\-kä}}, go down.
\on{4}.~\HL{kllkulat}{\B{kll\-ku\-lat}}, dig up.
\on{5}.~\HL{kllkxem}{\B{kll\-kxem}}, stand.
\on{6}.~\HL{kllpa}{\B{kll\-pa}}, bottom.
\on{7}.~\HL{kllpA}{\B{kll\-pä}}, land, arrive at the ground.
\on{8}.~\HL{kllpxiwll}{\B{kll\-pxi\-wll}}, lionberry.
\on{9}.~\HL{klltxay}{\B{kll\-txay}}, lie on the ground.
\on{10}.~\HL{klltseng}{\B{kll\-tseng}}, position.
\on{11}.~\HL{kllvawm}{\B{kll\-vawm}}, brown.
\on{12}.~\HL{kllwo}{\B{kll\-wo}}, alight, land.
\on{13}.~\HL{kllyem}{\B{kll\-yem}}, bury.
\on{14}.~\HL{kllza'au}{\B{kll\-za\-'u}}, come down, descend.

% klltxay
\hi
\HT{klltxay}{\B{kll\cp{txay}}} \I{[k\s{l}."t'$\cdot$aj]} \emph{vin., inf.}\on{22} lie on the ground.
(\ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{txay}{\B{txay}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\B{kll\-txey\-kay}, lay (s.t.) on the ground.

% klltxeykay
\hi
\HT{klltxeykay}{\B{klltxey\cp{kay}}} \I{[k\s{l}.t'Ej."k$\cdot$aj]} \emph{vtr., inf.}\on{33} lay (s.t.) on the ground. 
(\ding{43}~\HL{klltxay}{\B{kll\-txay}})

% klltseng
\hi
\HT{klltseng}{\B{\cp{kll}tseng}} \I{["k\s{l}.\t{ts}EN]} \emph{n.} position. 
\B{Sra\-ne, tsun fko si\-var hem\-lì\-'u\-vit mì lì\-'u alu swey\-lu, ul\-te sra\-ne, kll\-tseng lu \on{2},\on{2}.} Yes, you can use infixes in the word \emph{sweylu}, and (their) position is \on{2},\on{2}.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-677}{b}
(\ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{tsenge}{\B{tse\-nge}})

% kllvawm
\hi
\HT{kllvawm}{\B{\cp{kll}vawm}} \I{["k\s{l}.vawm]} \emph{adj.} brown. 
\B{Tsa\-tu\-tan\-ur a fkol tspo\-lang lo\-lu ta'\-leng a\-kll\-vawm; tspang\-yur lu pum a\-teyr.} That person that was killed had brown skin, the killer had white.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{vawm}{\B{vawm}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kllvawmpin}{\B{kll\-vawm\-pin}}, (the) brown (color).

% kllvawmpin
\hi
\HT{kllvawmpin}{\B{\cp{kll}vawmpin}} \I{["k\s{l}.vawm.pin]} \emph{n.} the color \ding{43}~\HL{kllvawm}{\B{kll\-vawm}}.
(\ding{43}~\HL{'opin}{\B{'o\-pin}})

% kllwo 
\hi
\HT{kllwo}{\B{kll\cp{wo}}} \I{[k\s{l}."w$\cdot$o]} \emph{vin., inf.}\on{22} alight, land (\emph{process}), reach for the ground.  
\B{Tom\-pa 'e\-ko nì\-hawng, ha ze\-ne aw\-nga kll\-wi\-vo.} The rain is too strong, so we must land.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{214}~\HL{yawo}{\B{ya\-wo}}, \ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{wo}{\B{wo}}, \HL{kllpA}{\B{kll\-pä}}) 

% kllyem
\hi
\HT{kllyem}{\B{kll\cp{yem}}} \I{[k\s{l}."j$\cdot$Em]} \emph{vtr., inf.}\on{22} bury. 
\B{Po\-ti kll\-yo\-lem ay\-oel äo ut\-ral\-o a\-lor a ro\-fa kil\-van.} We buried her under a beautiful tree beside the river.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{214}~\HL{kllkulat}{\B{kll\-ku\-lat}}, \ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{yem}{\B{yem}}) 

% kllza'u
\hi
\HT{kllza'u}{\B{kll\cp{za}'u}} \I{[k\s{l}."z$\cdot$a.P$\cdot$u]} \emph{vin., inf.}\on{23} come down, descend. 
\B{Kll\-zo\-la\-'u Mo\-'at fa sna\-yì teng\-krr pe\-rll\-txe san Ay\-nga ne\-to ri\-vikx!} Mo'at came down the stairway saying, ``Get back, all of you!''~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{214}~\HL{fAza'u}{\B{fä\-za\-'u}}, \ding{43}~\HL{kllte}{\B{kll\-te}}, \HL{za'u}{\B{za\-'u}}) 

% ko 
\hi
\HT{ko}{\B{ko}} \I{[ko]} \emph{sentence\hyp{}final part. for solicit agreement.} (\emph{a.}) ``Don't you think so?'' ``Wouldn't you agree?'' ``Let's \dots'' ``Why don't you \dots'' ``I'll do X, OK?''
\B{Aw\-nga sngi\-vä\-'i ko!} Let us begin!~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Var ni\-vu\-me ko!} Let's keep learning!~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln}
\B{Za\-'u kal\-txì si ko!} Come and say hello!~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Teng\-krr ftxo\-zä se\-re\-i\-yi aw\-nga, ke tswi\-va’ ay\-lom\-tu\-ti ko!} A toast:~To absent friends. [lit.: While we are celebrating, let us not forget those who we wish could also be here]~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay.--Ye\-i\-o'!~Tsa\-krr\-vay ko!} Let's meet on the hill tomorrow at nightfall.--Perfect!~See you then.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Nga lä\-pi\-vawk nì\-'it nì\-'ul ko.} Tell me a bit more about yourself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Lu fo le\-hrr\-ap.~Tsun tu\-tet tspi\-vang ko.} Those things are dangerous. They can kill a person, you know.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
\B{Fnu, ma 'e\-vi.~Sa'\-nur le\-ru haw\-tsyìp.~Tsi\-vu\-rokx ko.}
Quiet, young one.~Mommy is taking a nap.~Let her rest.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\emph{b. idiom}) \B{Li ko.} Well, get to it, then.\slash Let's get on it.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{May' ko.} Give it a try!\slash Have a go!~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}

% ko'on
\hi
\HT{ko'on}{\B{\cp{ko}'on}} \I{["ko.Pon]} \emph{n.} ring, oval, closed shaped (\emph{roughly circular}). 
\B{Na'\-vi ìlä ho\-'on kll\-kxo\-lem teng\-krr re\-rol.} The People were standing in a circle, singing.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{koum}{\B{ko\-um}}, \HL{'on}{\B{'on}}; \HL{yo'ko}{\B{yo'\-ko}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{ko'onspul}{\B{ko\-'on\-spul}}, sunflower gigantus.

% ko'onspul
\hi 
\HT{ko'onspul}{\B{\cp{ko}'onspul}} \I{["ko.Pon.spul]} \emph{or} \I{[.spUl]} \emph{n.} \ding{96} sunflower gigantus. 
(\ding{43}~\HL{ko'on}{\B{ko'on}}, \HL{spule}{\B{spule}})

% koak
\hi
\HT{koak}{\B{\cp{ko}ak}} \I{["ko.ak\textcorner]} \emph{adj.} old, aged (\emph{for living things}). 
\B{Ko\-ak nì\-ngay, ke\-fyak?} Really old, right?~\LINK{http://forum.learnnavi.org/news-announcements/happy-birthday-to-karyu-pawl/msg556899/\#msg556899}{ln}
(\ding{214}~\HL{'ewan}{\B{'e\-wan}}, \ding{43}~\HL{spuwin}{\B{spu\-win}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{koaktu}{\B{ko\-ak\-tu}}, old person. 
\on{2}.~\HL{koaktan}{\B{ko\-ak\-tan}}, old man.   
\on{3}.~\HL{koakte}{\B{ko\-ak\-te}}, old woman. 
\on{4}.~\HL{kewan}{\B{ke\-wan}}, age.
\on{5}.~\HL{tIkoak}{\B{tì\-ko\-ak}}, old age.

% koaktan
\hi
\HT{koaktan}{\B{\cp{ko}aktan}} \I{["ko.ak\textcorner.tan]} \emph{n.} old man.  
\B{Ay\-sre' lom lu tsa\-ko\-ak\-tan\-ur.} The old man misses his teeth.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Ko\-ak\-tan\-ä ay\-sre' lä\-ngu fwem.} An old man's teeth are dull.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{koak}{\B{ko\-ak}})

% koakte
\hi
\HT{koakte}{\B{\cp{ko}akte}} \I{["ko.ak\textcorner.tE]} \emph{n.} old woman.
\B{Po\-leng ay\-oe\-ru ko\-ak\-tel vur\-it a\-txan\-lal.} The old woman told us an ancient story.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}    
\B{Tsa\-ko\-ak\-te\-ri stawm\-tswo lu fe'.} That old woman's hearing is poor.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\ding{43}~\HL{koak}{\B{ko\-ak}})

% koaktu
\hi
\HT{koaktu}{\B{\cp{ko}aktu}} \I{["ko.ak\textcorner.tu]} \emph{n.} old person.
\B{Ul\-te txo smi\-von ngar ay\-ho\-ak\-tu fo\-ti pa\-lang fte tsi\-vun ivo\-mum teyng\-ta ftxey lu fo\-ru fpom fu\-ke.} And if you know elders contact them to know whether they are alright or not.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b}
\B{Ta'\-leng prr\-nen\-ä lu fa\-o\-i, pum ko\-ak\-tu\-ä ekx\-txu.} A baby's skin is smooth, an old person's is rough.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{a. proverb}) \B{Koakturi kewanti keyìl ke wan.} Some things can't be covered up. [lit.: `An old person's face doesn't hide their age.']~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
(\ding{43}~\HL{koak}{\B{ko\-ak}}, \HL{koaktan}{\B{ko\-ak\-tan}}, \HL{koakte}{\B{ko\-ak\-te}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{koaktutral}{\B{ko\-ak\-tut\-ral}}, goblin thistle.

% koaktutral
\hi 
\HT{koaktutral}{\B{ko\cp{ak}tutral}} \I{[ko."{}ak\textcorner.tut\textcorner.Ral]} \emph{or} \I{[.tUt\textcorner.]} \emph{n.} \ding{96} goblin thistle, \emph{cobalus carduus}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln}  
(\ding{43}~\HL{koaktu}{\B{ko\-ak\-tu}}, \HL{utral}{\B{ut\-ral}})

% kolan
\hi
\HT{kolan}{\B{ko\cp{lan}}} \I{[ko."lan]} \emph{conv.} ``I mean \ldots, rather''.
(\ding{43}~\HL{kan}{\B{kan}})

% kom
\hi 
\HT{kom}{\B{kom}} \I{[kom]} \emph{vin., modal} dare. 
\B{Oe ke kom ki\-vä.} I don't dare to go.
\B{Nga kom pi\-vll\-txe oe\-hu tsa\-fya srak?} You dare to speak to me like that?~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}

% komum
\hi
\HT{komum}{\B{ko\cp{mum}}} \I{[ko."mum]} \emph{or} \I{[."mUm]} (\textsc{rn}: \B{ko\cp{mùm}} \I{[ko."mUm]})~:~\HL{omum}{\B{omum}}

% kop
\hi
\HT{kop}{\B{kop}} \I{[kop\textcorner]} \textsc{i}. \emph{adv.} also, too, as well, additionally. 
\B{Tsa\-kem kop krr\-nekx.} That also takes time.~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln}
\B{Kop oe\-ru lo\-lä\-ngu lie a ha\-pxì\-tu soaiä ter\-kup.} Sadly, I too have had the experience of a family member dying.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{a. idiom}) \B{Kxe\-tse sì mik\-yun kop pll\-txe.} Body language is important [lit.: The tail and ears also speak].~\LINK{http://forum.learnnavi.org/language-updates/grid-tidbits/}{ln}  \\
\textsc{ii}. \emph{conj.} and also.
\B{Pe\-ioang tsun mì tam\-pay kop mì reym ri\-vey?} What animal can live both in the sea and on the land?~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{slAkop}{\B{slä\-kop}}, but also.

% koren
\hi
\HT{koren}{\B{ko\cp{ren}}} \I{[ko."REn]} \emph{n.} rule, guideline.  
\B{Ko\-ren A\-'aw\-ve Tì\-ru\-sey\-ä 'Aw\-si\-teng}, the First Rule of Living Together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
\B{Oe\-ri tsa\-ko\-ren\-it pxìm tswä\-nga'.} I often forget that rule.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b}
\B{za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi}, I will bring you the words and rules of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{ke tsun fko ni\-vu\-me fte nì\-lì'\-fya a\-mip pi\-vll\-txe nìl\-tsan fa fwa tìng mik\-yun ho\-ren\-ur nì\-'aw}, you can't learn to speak a new language well just by listening rules.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} \\
(i.) \B{koren ayll} law, societal rule.
\B{Pol aw\-nga\-ru ho\-ren\-it a\-yll ley\-kek.} He makes us obey the laws.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lekoren}{\B{le\-ko\-ren}}, lawlike, regarding rules. 
\on{2}.~\HL{horenleykekyu}{\B{ho\-ren\-ley\-kek\-yu}}, law enforcer.

% kosman
\hi
\HT{kosman}{\B{ko\cp{sman}}} \I{[ko."{}sman]} \emph{adj.} wonderful, terrific, fantastic.
\B{Fol lì'\-fya\-ti aw\-nge\-yä sar a fya\-'o lu ko\-sman.} It's wonderful the way they use our language.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
\B{Nge\-yä sä\-mok\-ìri a\-ko\-sman se\-i\-yi oe ira\-yo.} Thanks for that excellent suggestion (of yours).~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIksman}{\B{nìk\-sman}}, wonderfully.

% kot
\hi
\HT{kot}{\B{kot}} \I{[kot\textcorner]} \emph{n.}, \textsc{lw} coat.

% koum
\hi
\HT{koum}{\B{\cp{ko}um}} \I{["ko.um]} \emph{or} \I{[.Um]} (\textsc{rn}: \B{\cp{ko}ùm} \I{["ko.Um]}) \emph{adj.} rounded, curved.
\B{Fì\-tskxe\-ri fa\-'o lu yey; ke lu ko\-um.} This rock has straight sides; it's not rounded.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{ko'on}{\B{ko\-'on}}, ring, oval, closed shape roughly circular. 
\on{2}.~\HL{yo'ko}{\B{yo'\-ko}}, circle, perfectly circular ring.

% krr
\hi
\HT{krr}{\B{krr}} \I{[k\s{r}]} \emph{uncountable n.} time. 
\B{maw\-krr la\-ye\-i\-u oer krr nì\-'ul}, after that I will have more time.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Ke lu oer set krr a\-txan.} I don't have much time now.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg139677/\#msg139677}{ln} 
\B{Nga\-ri txo fì\-kem si\-vi fì\-fya, krr\-ti 'a\-vun.} If you do it like this, you'll save time.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
\B{Slä mì tam\-pxì krr\-ä, ke sar fkol kea hem\-lì\-'u\-vit fì\-lì\-'u\-mì.} But most of the time, one doesn't use any infixes in this word.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-677}{b} 
(\emph{a. idiom, poetic}) \B{tì\-'i'\-a\-vay krr\-ä}, forever, 'til the end of time. 
\B{To\-ler\-kup tu\-te; pam\-tseo pe\-yä tì\-'i'\-a\-vay krr\-ä ra\-yey.} The man has died; his music will live forever.~\LINK{http://naviteri.org/2015/06/na-ayskxe-mi-telan-sad-news/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{frakrr}{\B{fra\-krr}}, always. 
\on{2}.~\HL{ftawnemkrr}{\B{ftaw\-nem\-krr}}, past. 
\on{3}.~\HL{hIkrr}{\B{hì\-krr}}, short time, second. 
\on{4}.~\HL{kawkrr}{\B{kaw\-krr}}, never. 
\on{5}.~\HL{hawngkrr}{\B{hawng\-krr}}, late. 
\on{6}.~\HL{krra}{\B{krra}}\slash \HL{a krr}{\B{a krr}}, when, as. 
\on{7}.~\HL{krrnekx}{\B{krr\-nekx}}, take\slash consume time. 
\on{8}.~\HL{krro}{\B{krro}}, sometimes. 
\on{9}.~\HL{krro krro}{\B{krro krro}}, at times, on occasion. 
\on{10}.~\HL{lefkrr}{\B{le\-fkrr}}, current. 
\on{11}.~\HL{mawkrr}{\B{maw\-krr}}, after. 
\on{12}.~\HL{nIfkrr}{\B{nì\-fkrr}}, currently, lately. 
\on{13}.~\HL{nulkrr}{\B{nul\-krr}}, longer. 
\on{14}.~\HL{sekrr}{\B{se\-krr}}, present, now. 
\on{15}.~\HL{sngA'ikrr}{\B{sngä\-'i\-krr}}, beginning. 
\on{16}.~\HL{srekrr}{\B{sre\-krr}}, before. 
\on{17}.~\HL{tengkrr}{\B{teng\-krr}}, while, at the same time. 
\on{18}.~\HL{takrra}{\B{ta\-krra\slash akrr\-ta}}, since. 
\on{19}.~\HL{tsakrr}{\B{tsakrr}}, then, at that time. 
\on{20}.~\HL{txankrr}{\B{txan\-krr}}, (for a) long time. 
\on{21}.~\HL{txonkrr}{\B{txon\-krr}}, at night. 
\on{22}.~\HL{vaykrr}{\B{vay\-krr}}, until. 
\on{23}.~\HL{ye'krr}{\B{ye'\-krr}}, early. 
\on{24}.~\HL{zIsIkrr}{\B{zì\-sì\-krr}}, season. 
\on{25}.~\HL{zusawkrr}{\B{zu\-saw\-krr}}, future.  
\on{26}.~\HL{krrka}{\B{krr\-ka}}, during.
\on{27}.~\HL{txakrrfpIl}{\B{txa\-krr\-fpìl}}, consider, ponder.
\on{28}.~\HL{krrlik}{\B{krr\-lik}}, point in time.

% krra
\hi
\HT{krra}{\B{\cp{krr}a}}, \B{krr a} \emph{or} \HL{a krr}{\B{a krr}} \I{["k\s{r}.a]} \emph{conj.} when, at the time that. 
\B{Oel tsko\-ti nga\-ru ta\-syìng krra oeng ul\-txa si.} I'll give you the bow when we meet.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{pam\-rel sa\-yi trr\-ay krr a skxom la\-tsu oe\-ru}, I'll write tomorrow when I have the chance.~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln}
\B{Tì\-'eyng\-it oel to\-lel a krr, ay\-nga\-ru pa\-yeng.} When I've received an answer, I will tell you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Oel fo\-ru fì\-ay\-lì\-'ut to\-lìng a krr, kxawm oe har\-ma\-hä\-ngaw.} When I gave them these words I was probably sleeping.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
(\ding{43}~\HL{krr}{\B{krr}})

% krrka
\hi
\HT{krrka}{\B{\cp{krr}ka}} \I{["k\s{r}.ka]} \emph{adp.} during.
\B{Krr\-ka tsawl\-ul\-txa vo\-spxì\-am.} During the great meeting last month.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}
\B{Kaym\-krr\-ka to\-lel mo\-el ta ay\-ul\-txa\-tu stxe\-lit a\-ko\-sman.} During the evening we received a wonderful gift from the meeting members.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}
(\ding{43}~\HL{krr}{\B{krr}}, \HL{ka}{\B{ka}})

% krrlik
\hi 
\HT{krrlik}{\B{\cp{krr}lik}} \I{["k\s{r}.lik\textcorner]} \textsc{i}. \emph{n.} point in time. 
\B{Krr\-lik lu trr\-pxì a\-mrr\-ve.} The time is five o'clock.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} \\
\textsc{ii}. \B{\cp{krr}likpe} \I{["k\s{r}.lik\textcorner.pE]} \emph{or} \B{pe\cp{hrr}lik} \I{[pE."h\s{r}.lik\textcorner]} \emph{inter.} what point in time? what hour?
(\ding{43}~\HL{krr}{\B{krr}}, \HL{lik}{\B{lik}}) 

% krro
\hi
\HT{krro}{\B{\cp{krr}o}} \I{["k\s{r}.o]} \emph{adv.} sometimes.
\B{Krro len ay\-hem a\-na\-fì\-'u.} Sometimes such things happen.~\LINK{http://forum.learnnavi.org/language-updates/just-another-few-words/}{ln}
\B{Krro lu 'Ìng\-lì\-sì txur nì\-hawng mì re\-'o.} Sometimes English is too strong in my head.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b}
\B{Ngay\-txoa, krro tì\-kxey si keng kar\-yu.} I'm sorry, sometimes even the teacher messes up.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{krr}{\B{krr}})

% krro krro
\hi
\HT{krro krro}{\B{\cp{krr}o \cp{krr}o}} \I{["k\s{r}.o "k\s{r}.o]} \emph{adv.} at times, on occasion.
\B{Krro krro fì\-tse\-nge oe tì\-kxey sa\-yi.} At times, I will make mistakes here.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Krro krro, flì\-a vul a\-ru\-sey to nutx\-a pum a\-ke\-ru\-sey lu txur.} A thin living branch is sometimes stronger than a thick dead one.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\ding{43}~\HL{krr}{\B{krr}})

% krrnekx
\hi
\HT{krrnekx}{\B{krr\cp{nekx}}} \I{[k\s{r}."n$\cdot$Ek']} \emph{vin., inf.}\on{22} take\slash consume time. 
\B{Tsa\-kem kop krr\-nekx.} That also takes time.~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln}
\B{fwa oel fì\-tì\-kang\-kem\-vit sngey\-ki\-vä\-'i krr\-no\-lekx nì\-txan}, me starting this project took a long time.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\ding{43}~\HL{krr}{\B{krr}}, \HL{nekx}{\B{nekx}})

% krrpe
\hi
\HT{krrpe}{\B{\cp{krr}pe}} \emph{or} \HL{pehrr}{\B{pe\cp{hrr}}} \I{["k\s{r}.pE]} \emph{or} \I{[pE."h\s{r}]} \emph{inter.} when?
\B{Fä\-za\-'u tsaw\-ke krr\-pe?} When will the sun come up?~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Nga\-ri sno\-mot krr\-pe ve\-zey\-ko, ma 'itan?} When are you going to organize your room, son?~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{krr}{\B{krr}})

% ku'up
\hi
\HT{ku'up}{\B{\cp{ku'}up}} \I{["kuP.up\textcorner]} \emph{or} \I{["kUP.Up\textcorner]} (\textsc{rn}: \B{\cp{ku}ùp} \I{["ku.Up\textcorner]}) \emph{adj.} heavy (\emph{physical weight}). 
\B{Ay\-skxe a mì sa\-sna\-'o ku'\-up lu nì\-txan.} The rocks in that pile are very heavy.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Ke tsun tu\-te a\-'aw tsa\-tskxe\-ti a\-ku'\-up kxi\-vel\-tek nì\-'aw\-tu.} One person alone can't lift that heavy rock.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{a. idiom}) \B{Ngä\-zìk fwa fkol rum\-it a\-ku\-'up pey\-ku\-wup.} \emph{it's difficult to make s.o. stubborn or inept do what you want them to} [lit.: it's hard to bounce a heavy ball].~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b}
(\ding{214}~\HL{syo}{\B{syo}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{syokup}{\B{syo\-kup}}, weight.

% kulat
\hi
\HT{kulat}{\B{\cp{ku}lat}} \I{["ku.lat\textcorner]} \emph{vtr.} reveal, bring forth, uncover (\emph{literally and metaphorically}). 
\B{Maw txan\-tom\-pa, pxay\-a rìk\-äo la\-mu tska\-lep pe\-yä, ha tsat ku\-lat ay\-oel.} After the rainstorm, his crossbow was under a lot of leaves, so we uncovered it.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Lo\-lu ka\-vuk, slä Tse\-nul tì\-ngay\-it ko\-lu\-lat.} There was treachery, but Tsenu revealed the truth.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kllkulat}{\B{kll\-ku\-lat}}, dig out.

% kum
\hi
\HT{kum}{\B{kum}} \I{[kum]}~\emph{or}~\I{[kUm]} \emph{n.} result.
\B{Po\-ri tì\-kak\-rel tì\-kak\-pam\-sì kum tsam\-ä lu.} His blindness and deafness are a result of the war.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
\B{Mllte oe: am'aluke zamayunge tsatìpuseyìl humit apxan.} I agree: without a doubt the waiting will bring worthy results.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/comment-page-1/\#comment-2478}{b}
(\emph{a. proverb}) \B{Kem a\-muiä, kum a\-fe'.} Proper action, bad result. (\emph{something that should have turned out well didn't}).~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Hem a\-sru\-nga' nì\-'ul, hum a\-sngu\-nga' nì\-nän.}
More helpful actions lead to less troubling outcomes.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kuma}{\B{ku\-ma\slash a\-kum}}, that (\emph{as a result}). 
\on{2}.~\HL{tsankum}{\B{tsan\-kum}}, advantage, benefit, upside, gain. 
\on{3}.~\HL{fekum}{\B{fe\-kum}}, disadvantage, drawback, downside.

% kuma
\hi
\HT{kuma}{\B{\cp{kum}a}} \emph{or} \B{a\cp{kum}} \I{["kum.a]} \emph{or} \I{[a."kum]} \emph{conj. with} \HL{fItxan}{\B{fì\-txan}} \emph{or} \HL{nIftxan}{\B{nì\-ftxan}} \emph{to form result clauses.} that (\emph{as a result}). 
\B{Lu poe se\-vin nì\-ftxan kum\-a yaw\-ne slo\-lu oer.}~\emph{or} \B{Poe yaw\-ne slo\-lu oer a\-kum, nì\-ftxan lu se\-vin.} She was so beautiful that I fell in love with her.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Lu oer ing\-yen a Ìstaw nim lu fì\-txan kum\-a pxìm wä\-pan.} It's a mystery to me why Ìstaw is so shy that he frequently hides.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{without} \B{fì\-txan} \emph{or} \B{nì\-ftxan}) \B{Pxe\-fo\-ru oe srung so\-li, kum\-a oe\-ru set pxe\-fo srung se\-ri.} I helped the three of them, so they're now helping me.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{a. coll. idiom}) \B{[verb] nì\-ftxan kum\-a ter\-kup}, \emph{commenting on an action that is so intense that it results in dying}; ``like crazy, \ldots{} to death''. 
\B{Sran, sran, oe\-ri pì\-tìk tsa\-tse\-nget a mì tal! Fkxa\-ke nì\-ftxan kum\-a ter\-kup.} Yeah, yeah, scratch that place on my back! It itches like crazy!~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{43}~\HL{kum}{\B{kum}})

% kunsìp
\hi
\HT{kunsIp}{\B{\cp{kun}sìp}} \I{["kun.sIp\textcorner]} \emph{or} \I{["kUn.]} (\textsc{rn}: \B{\cp{kùn}sìp} \I{["kUn.sIp\textcorner]}) \emph{n.} \textsc{lw} gunship. 
\B{Kun\-sìp\-ìri txan\-a tì\-meyp lu tsyal a mìn.} The gunship's main weakness is the rotor system.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln\slash a\textsuperscript{c}}

% kung
\hi 
\HT{kung}{\B{kung}} \I{[kuN]} \emph{or} \I{[kUN]} (\textsc{rn}: \B{kùng} \I{[kUN]}) \emph{adj.} putrid, fetid, rotten (\emph{referred to rotten meat or a pile of dead and rotting animal matter}). 
\B{Tsa\-fa\-hew a\-on\-vä' ftu kung\-a ioang za\-'u.} That stinking smell comes from a rotten animal.~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/}{b}
(\ding{43}~\HL{snAm}{\B{snäm}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kurkung}{\B{kur\-kung}}, asshole (\emph{insult}).

% kur
\hi
\HT{kur}{\B{kur}} \I{[kuR]} \emph{or} \I{[kUR]} (\textsc{rn}: \B{kùr} \I{[kUR]}) \emph{vin.} hang (\HL{ta}{\B{ta}}, from).
\B{Fkxi\-le pewn\-ta tu\-té\-yä kur.} The bib necklace hangs from the woman's neck.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{a. proverb}) \B{Ke kur fko fa kxe\-tse.} ``One can't hang by a tail,'' i.e. \emph{don't rely on s.t.\slash s.o. untrustworthy or useless, just as a Na'vi tail can't be relied on to bear weight}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{keykur}{\B{key\-kur}}, hang (s.t.), let hang. 
\on{2}.~\HL{kurfyan}{\B{kur\-fyan}}, hamper, suspended rack.

% kurakx
\hi
\HT{kurakx}{\B{ku\cp{rakx}}} \I{[ku."Rak']} \emph{vtr.} drive out. 
\B{Ay\-oe\-ngal fte Fay\-saw\-tu\-tet ki\-vu\-rakx, ze\-ne nì\-'aw\-ve ay\-fot tsli\-vam.} We must understand these Sky People if we are to drive them out.~\LINK{http://forum.learnnavi.org/language-updates/confirmation-on-kurakx-eytukan-line-and-toitslan/}{ln\slash a\textsuperscript{c}}

% kurfyan
\hi
\HT{kurfyan}{\B{\cp{kur}fyan}} \I{["kuR.fjan]} \emph{or} \I{["kUR.]} (\textsc{rn}: \B{\cp{kùr}fya} \I{["kUR.fja]}) \emph{n.} hamper, suspended rack.
(\ding{43}~\HL{kur}{\B{kur}}, \HL{fyan}{\B{fyan}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{snokfyan}{\B{snok\-fyan}}, personal belongings rack. 
\on{2}.~\HL{kurfyavi}{\B{kur\-fya\-vi}}, hook.

% kurfyavi
\hi
\HT{kurfyavi}{\B{\cp{kur}fyavi}} \I{["kuR.fja.vi]} \emph{or} \I{["kUR.]} (\textsc{rn}: \B{\cp{kùr}fyavi} \I{["kUR.fja.vi]}) \emph{n.} hook (\emph{for hanging or suspending an item}).
(\ding{43}~\HL{kurfyan}{\B{kur\-fyan}}) 

% kurkung
\hi 
\HT{kurkung}{\B{\cp{kur}kung}} \I{["kuR.kuN]} \emph{or} \I{["kUR.kUN]} (\textsc{rn}: \B{\cp{kur}kùng} \I{["kuR.kUN]}) \emph{n., vulg.} asshole (\emph{strong insult}). 
(\ding{43}~\HL{kuru}{\B{ku\-ru}}, \HL{kung}{\B{kung}})

% kuru
\hi
\HT{kuru}{\B{kuru}} \I{["ku.Ru]} \emph{n.} neural queue.
\ding{43}~\HL{tswin}{\B{tswin}} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{kurkung}{\B{kur\-kung}}, asshole (\emph{strong insult}). 

\end{multicols}