\pdfbookmark[0]{KX}{KX}
\begin{center}
\section*{KX \I{[k']} (\emph{KxeKx})}
\end{center}

\begin{multicols}{2}

% kxa
\hi
\HT{kxa}{\B{kxa}} \I{[k'a]} \emph{n.} mouth. 
\B{Kxa\-ri tstu\-tswo tsran\-ten krra ke lu kea sä\-fpìl le\-sar.} When one has no useful thoughts, the ability to close one's mouth is important.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}

% kxakx
\hi
\HT{kxakx}{\B{kxakx}} \I{[k'ak']} \emph{vin.} break, snap in two.
\B{Txo vul ivil nì\-hawng kxì\-ye\-vakx.} If a branch bends too much, it might break.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% kxam
\hi
\HT{kxam}{\B{kxam}} \I{[k'am]} \emph{n.} middle, midpoint. \\ 
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxamlA}{\B{kxam\-lä}}, through.  
\on{2}.~\HL{kxamtrr}{\B{kxam\-trr}}, midday, noon. 
\on{3}.~\HL{kxamtxon}{\B{kxam\-txon}}, midnight.
\on{4}.~\HL{kxamtseng}{\B{kxam\-tseng}}, center.
\on{5}.~\HL{kxamyI}{\B{kxam\-yì}}, intermediate level. 
\on{6}.~\HL{mIkam}{\B{mì\-kam}}, between. 

% kxamlä
\hi
\HT{kxamlA}{\B{\cp{kxam}lä}} \I{["k'am.l\ae]} \emph{adp.} through, via the middle of. 
\B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{kxam}{\B{kxam}}, \HL{IlA}{\B{ìlä}})

% kxamtrr
\hi
\HT{kxamtrr}{\B{\cp{kxam}trr}} \I{["k'am.t\s{r}]} \emph{n.}; \emph{adv.} midday, noon.
\B{Nän yom kxam\-trr, 'ul 'e\-fu o\-hakx kaym.} The less you eat at noon, the hungrier you'll feel in the evening.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Kxam\-trr lam fwa san\-hì a mì saw 'o\-lìp nì\-wotx slä tsa\-krr ke tsun fko sat tsi\-ve\-'a nì\-'aw.} At mid-day it seems that the stars in the sky have all vanished but they just can't be seen then.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{kxam}{\B{kxam}}, \HL{trr}{\B{trr}}) \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxamtrrmaw}{\B{kxam\-trr\-maw}}, after midday. 
\on{2}.~\HL{srekamtrr}{\B{sre\-kam\-trr}}, before midday.

% kxamtrrmaw
\hi
\HT{kxamtrrmaw}{\B{\cp{kxam}trrmaw}} \I{["k'am.t\s{r}.maw]} \emph{n.}; \emph{adv.} after midday, after noon.
(\ding{43}~\HL{kxamtrr}{\B{kxam\-trr}}, \HL{maw}{\B{maw}})

% kxamtxomaw
\hi
\HT{kxamtxomaw}{\B{\cp{kxam}txomaw}} \I{["k'am.t'o.maw]} \emph{n.}; \emph{adv.} after the middle of the night.
(\ding{43}~\HL{kxamtxon}{\B{kxam\-txon}}, \HL{maw}{\B{maw}})

% kxamtxon
\hi
\HT{kxamtxon}{\B{\cp{kxam}txon}} \I{["k'am.t'on]} \emph{n.}; \emph{adv.} middle of the night, midnight.
(\ding{43}~\HL{kxam}{\B{kxam}}, \HL{txon}{\B{txon}}) \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxamtxomaw}{\B{kxam\-txo\-maw}}, after midnight. 
\on{2}.~\HL{srekamtxon}{\B{sre\-kam\-txon}}, before midnight.

% kxamtseng
\hi
\HT{kxamtseng}{\B{\cp{kxam}tseng}} \I{["k'am.\t{ts}EN]} \emph{n.} center, place in the middle. 
\B{Re\-rol teng\-krr ke\-rä | Ìlä fya\-'o a\-vol | Ne kxam\-tseng.} We sing our way, down the eight paths, to the center.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
(\ding{43}~\HL{kxam}{\B{kxam}}, \HL{tseng}{\B{tse\-nge}})

% kxamyì
\hi
\HT{kxamyI}{\B{\cp{kxam}yì}} \I{["k'am.jI]} \emph{n.} intermediate level.
\B{Lì'\-fya\-ri tsa\-taw\-tu\-te pe\-yì?--Kxam\-yì.~Pll\-txe nìk\-sran, slä tsun fko pe\-yä ay\-lì\-'ut tsli\-vam.} How is that Sky Person's Na'vi?--It's intermediate.~His speech is mediocre, but you can understand him.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{kxam}{\B{kxam}}, \HL{yI}{\B{yì}})

% kxanì
\hi
\HT{kxanI}{\B{\cp{kxa}nì}} \I{["k'a.nI]} \emph{adj.} forbidden.  
\B{Fay\-vrr\-tep fì\-tse\-nge lu kxa\-nì.} These demons are forbidden here.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}

% kxangangang
\hi
\HT{kxangangang}{\B{\cp{kxang}angang}} \I{["k'aN.aN.aN]} \emph{onom.} boom. 

% kxange
\hi 
\HT{kxange}{\B{\cp{kxa}nge}} \I{["k'a.NE]} \emph{vin.} yawn (\emph{as a result of fatigue or boredom}). 
\B{Oe kxì\-ma\-nge ta\-lu\-na 'e\-fu ngeyn.} I just yawned because I feel tired.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
\B{Keng krra sä\-num\-vi el\-tur tì\-txen ke si, nga swey\-lu txo ke kxi\-va\-nge mì num\-tseng.} Even when the lesson isn't interesting, you shouldn't yawn in school.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b} \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{sAkxange}{\B{sä\-kxa\-nge}}, a yawn.

% kxap / si
\hi
\HT{kxap}{\B{kxap}} \I{[k'ap\textcorner]} \textsc{i}. \emph{n.} threat.
\B{Tì\-pä\-hem Saw\-tu\-te\-yä kxap a\-txan lar\-mu Na'\-vi\-ru.} The arrival of the Sky People was a great threat to the Na'\-vi.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Kap sì ay\-u\-nil say\-la\-he.} Threats, dreams, and other things.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\emph{a. with} \B{wä}) a threat to s.t.\slash s.o., threaten.
\B{Krr\-a kxap lar\-mu sì\-rey\-wä fe\-yä nì\-wotx.} When their lives were threatened.~\LINK{http://naviteri.org/2015/05/cirque-du-soleils-toruk-the-first-flight-teaser-video-online/}{b} \\
\textsc{ii}. \HT{kxap si}{\B{kxap si}} \I{[k'ap\textcorner{} si]} \emph{vin.} (\emph{a.}) threaten (\ding{43}~\HL{fko}{\B{fkor}}, s.o.). 
\B{Sra\-ke nga kxap si oer, ma \mbox{skxawng}?} Are you threatening me, you moron?~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\emph{b. of animals and people}) threatening. \B{'Ang\-tsìk a kxap si lu le\-hrr\-ap, ma 'itan.} A threatening hammerhead is dangerous, son.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxapnga'}{\B{kxap\-nga'}}, threatening. 
\on{2}.~\HL{nIkxap}{\B{nì\-kxap}}, threateningly.

% kxapnga'
\hi
\HT{kxapnga'}{\B{\cp{kxap}nga'}} \I{[k'ap\textcorner.NaP]} \emph{adj.} threatening (\emph{not for people}).
\B{Nge\-yä ay\-lì\-'ul a\-kxap\-nga' txo\-pu ke sley\-ku oet kaw\-'it.} Your threatening words don't scare me one bit.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{kxap si}{\B{kxap si}} (\emph{b.}), \HL{kxap}{\B{kxap}})

% kxawm
\hi
\HT{kxawm}{\B{kxawm}} \I{[k'awm]} \emph{adv.} perhaps, maybe.
\B{Oel fo\-ru fì\-ay\-lì\-'ut to\-lìng a krr, kxawm oe har\-ma\-hä\-ngaw.} When I gave them these words I was probably sleeping.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{Kxawm swo\-at nì\-hawng nil\-vä\-ngäk oel txon\-am.} Maybe I drank too much last night.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/comment-page-1/\#comment-2585}{b}

% kxayl
\hi
\HT{kxayl}{\B{kxayl}} \I{[k'ajl]} \emph{adj.} high (\emph{scalable height}). 
\B{Ney\-ti\-ri\-ri lu yì tì\-sti\-yä kxayl nì\-ngay.} Neytiri is really angry [lit.: Neytiri's level of anger is truly high].~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Tsa\-ye\-rik kll\-kxem sìn yì a\-kxayl.} That \emph{yerik} is standing on a high ledge.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b} \\ 
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxaylyI}{\B{kxayl\-yì}}, high level.

% kxaylte
\hi
\HT{kxaylte}{\B{\cp{kxayl}te}} \I{["k'ajl.tE]} \emph{n.}~\ding{96} cillaphant, \emph{clawenia purpurea}.

% kxaylyì
\hi
\HT{kxaylyI}{\B{\cp{kxayl}yì}} \I{["k'ajl.jI]} \emph{n.} high level.
\B{Lì'\-fya\-ri tsa\-taw\-tu\-te pe\-yì?--Kxayl\-yì. Ke tsun oe spi\-vaw. Slo\-lu po tsul\-fä\-tu lì'\-fya\-yä aw\-nge\-yä.} How is that Sky Person's Na'vi?--It's excellent. I can't believe it. He's become a master of our language.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{kxayl}{\B{kxayl}}, \HL{yI}{\B{yì}})

% kxänäng
\hi
\HT{kxAnAng}{\B{\cp{kxä}näng}} \I{["k\ae.n\ae{}N]} \emph{adj.} putrid, \emph{smell of decaying animal\slash flesh}.
\B{Fì\-tsngan\-ur a sno\-läm lä\-ngu fa\-hew a\-kxä\-näng.} This rotten meat has a putrid smell.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}

% kxeltek
\hi
\HT{kxeltek}{\B{\cp{kxel}tek}} \I{["k'El.tEk\textcorner]} \emph{vtr.} pick up, lift.   
\B{Pxi\-set nge\-yä tska\-lep\-it kxel\-tek!} Pick up your crossbow right now!~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
\B{Ke tsun tu\-te a\-'aw tsa\-tskxe\-ti a\-ku'\-up kxi\-vel\-tek nì\-'aw\-tu.} One person alone can't lift that heavy rock.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{214}~\HL{keyn}{\B{keyn}})

% kxem
\hi
\HT{kxem}{\B{kxem}} \I{[k'Em]} \emph{vin.} be vertical.   
\B{Fì\-ru\-mut lum\-pe ke kxem?} Why isn't this puffball tree vertical?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\  
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kllkxem}{\B{kllkxem}}, stand. 
\on{2}.~\HL{kxemyo}{\B{kxem\-yo}}, wall. 
\on{3}.~\HL{nIkxem}{\B{nì\-kxem}}, vertically.

% kxemyo
\hi
\HT{kxemyo}{\B{\cp{kxem}yo}} \I{["k'Em.jo]} \emph{n.} wall, vertical surface. 
\B{Rìk a\-'aw syar\-mar sìn kxem\-yo.} A leaf was sticking to the wall.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
\B{Tseng\-a 'aw\-steng\-yä\-pem fì\-me\-kem\-yo lu mek\-tseng.} Where these two walls come together there's a gap.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/} {b}
(\ding{43}~\HL{kxem}{\B{kxem}}, \HL{yo}{\B{yo}})

% kxener
\hi
\HT{kxener}{\B{\cp{kxe}ner}} \I{["k'E.nER]} \emph{n.} smoke. 
\B{Kxe\-ner lu sä\-vll txep\-ä.} Smoke is a sign\slash an indication that there is fire.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b} \\
(i.) \B{tswìk \cp{kxe}nerit} \I{[\t{ts}wIk\textcorner{} "k'E.nE.Rit\textcorner]} \emph{vin.} smoke (\emph{a cigarette})
\B{Ke tung Na'\-vil fu\-ta tswìk kxe\-ner\-it Mo\-'a\-ra\-ka nì\-wotx.} The Na'vi have banned smoking in Mo'ara.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b}

% kxetse
\hi
\HT{kxetse}{\B{\cp{kxe}tse}} \I{["k'E.\t{ts}E]} \emph{n.} tail. 
(\emph{a.}) \B{Nge\-yä kxe\-tse lor lu nì\-txan.} Your tail is very beautiful.~\LINK{http://www.youtube.com/watch?v=ojr_BE8ZmVE}{yt}
\B{Ut\-ral\-ti a tsa\-uo Lo\-ak wä\-par\-man pol vol fa kxe\-tse.} She used her tail to point out the tree Loak was hiding behind.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{b. idiom}) \B{Kxe\-tse sì mik\-yun kop pll\-txe.} Body language is important [lit.: The tail and ears also speak].~\LINK{http://forum.learnnavi.org/language-updates/grid-tidbits/}{ln} 
(\emph{c. idiom}) \B{Po keyn\-ven sìn ke\-tse.} He is socially awkward. (\emph{lit.}: `He steps on tails.')~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\emph{d. proverb}) \B{Ke kur fko fa kxe\-tse.} ``One can't hang by a tail,'' i.e. \emph{don't rely on s.t.\slash s.o. untrustworthy or useless, just as a Na'vi tail can't be relied on to bear weight}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\emph{e. idiom}) \B{Kxe\-tse kì\-'ong!} ``Don't get angry.\slash Calm down.'' [lit.: slow tail] (\emph{short for} \B{Nga\-ri kxe\-tse kì\-'ong li\-vu.}).~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxetsikran}{\B{kxe\-tsik\-ran}}, banshee's tail.

% kxetsikran
\hi 
\HT{kxetsikran}{\B{kxe\cp{tsik}ran}} \I{[k'E."\t{ts}ik\textcorner.Ran]} \emph{n.} \ding{96} banshee's tail. 
(\ding{43}~\HL{kxetse}{\B{kxetse}}, \HL{ikran}{\B{ikran}})

% kxeyey
\hi
\HT{kxeyey}{\B{\cp{kxey}ey}} \I{["k'Ej.Ej]} \emph{n.} error, mistake.
\B{Tsa\-kxey\-ey li zo\-slo\-lu.} That mistake has already been fixed.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/comment-page-1/\#comment-497}{b}
\B{Txo kxey\-ey\-ti ay\-ngal tsi\-ve\-'a, ru\-txe oe\-ru pi\-veng.} If you see a mistake, please tell me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Ke ro\-lun oel kea ay\-key\-eyt!} I have found no mistakes!~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-584}{b}
\B{Kxey\-ey\-ri tsap\-'a\-lu\-te se\-ngi oe.} I apologize for the mistake.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-email/msg56711/\#msg56711}{ln}
\B{Slä krr a nì\-Na'\-vi pam\-rel si oe, new oel fu\-ta upxa\-re oe\-yä lu\-ke key\-ey li\-vu nì\-wotx, fte eyawr\-a sì\-ke\-nong\-it ti\-vìng su\-te\-ru.} But when I write Na'vi, I want my messages to be without any mistakes to give people correct examples.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln} \\  
\textsc{Derivations}: \ding{43} \on{1}.~\HL{tIkxey}{\B{tì\-kxey}}, incorrectness.

% kxitx
\hi
\HT{kxitx}{\B{kxitx}} \I{[k'it']} \emph{n.} death (\emph{the event of an individual's death}).
\B{Sa'\-sem\-ìri lu 'eveng\-ä \mbox{kxitx} yeng\-wal a\-tu\-vom.} For a parent, the greatest sorrow is the death of a child.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Maw \mbox{kxitx} sem\-pul\-ä lar\-mä\-ngu Pey\-ral\-ä a\-fpawng txew\-luke.} After (her) father's death, Pey\-ral's grief was endless.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\HL{tIterkup}{\B{tì\-ter\-kup}})

% kxì
\hi
\HT{kxI}{\B{kxì}} \I{[k'I]} \emph{intj.} hi, hiya (\emph{casual greeting}).
\B{Kxì, ma 'ey\-lan! Kem\-pe le\-ren?} Hey dude! What's happenin'?.~\LINK{http://naviteri.org/2018/02/negative-questions-in-navi/}{b}
(\ding{43}~\HL{kaltxI}{\B{kal\-txì}})

% kxìm
\hi
\HT{kxIm}{\B{kxìm}} \I{[k'Im]} \emph{vtr.} command, order, assign a task (\emph{with} \ding{43}~\HL{txonta}{\B{tson\-ta}}).
\B{Ay\-evengur kxo\-lìm sa'\-nok\-ìl fì\-tson\-it.} Mother imposed this task on the children.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Ay\-eveng\-ur kxo\-lìm sa'\-nok\-ìl tson\-ta pay\-it za\-mu\-nge.} The children were told by their mother to fetch water.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\emph{a. proverb}) \B{Kxìm utu\-ftu fna\-we'\-tu.} ``A coward commands from the canopy,'' i.e. \emph{a real leader will have `boots on the ground' and will help out, whereas a coward will only tell people (from afar) what to do}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxImyu}{\B{kxìm\-yu}}, commander, one with authority over another. 
\on{2}.~\HL{tIkxIm}{\B{tì\-kxìm}}, commanding, ordering, assigning tasks.

% kxìmyu
\hi
\HT{kxImyu}{\B{\cp{kxìm}yu}} \I{["k'Im.ju]} \emph{n.} commander, one with authority over another.
\B{Nge\-yä kxìm\-yu pe\-su?} Who's your boss?~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\emph{a. idiom, politeness formula})
\B{Kxìm\-yu nga.} ``Please! Go ahead. You first.'' [lit.: (my) commander (is) you]~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{kxIm}{\B{kxìm}})

% kxll / kxll si
\hi
\HT{kxll}{\B{kxll}} \I{[k'\s{l}]} \textsc{i}.~\emph{n.} charge. \\
\textsc{ii}.~\HT{kxll si}{\B{kxll si}} \I{[k'\s{l} si]} \emph{vin.} charge (toward, \B{ne}).   
\B{Teng\-krr pa\-lu\-lu\-kan moe\-ne kxll sar\-mi, pol\-txe Ney\-ti\-ril ay\-lì\-'ut a fra\-krr 'ok se\-yä la\-yu oer.} As the thanator was charging towards the two of us, Neytiri said something I will always remember.~\LINK{http://thereadinginpublicproject.blogspot.de/2010/05/paul-frommer.html}{pr}

% kxon
\hi 
\HT{kxon}{\B{kxon}} \I{[k'on]} \emph{n.} internal organ. 
\B{Lu tokx\-ur tu\-te\-yä pol\-pxay\-a kxon?} How many internal organs does a person's body have?~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}

% kxor
\hi
\HT{kxor}{\B{kxor}} \I{[k'oR]} \emph{rarely countable n.} a wall or bank of powerful waterfalls noted for its deafening roar and deadly force. 

% kxu / kxu si
\hi
\HT{kxu}{\B{kxu}} \I{[k'u]} \textsc{i}. \emph{n.} harm.
\B{Lu me\-nge\-yä kel\-ku na'\-rìng\-ä lu\-ke kxu a\-txan a fì\-'u fmawn a\-sìl\-tsan lu nì\-ngay.} It's truly good news that your forest house is without much harm.\LINK{http://forum.learnnavi.org/language-updates/luke-kxu-without-harm/}{ln}
\B{Fnu, fnu, wä\-pan wä ku | Uo tsa\-ìpxa.} Quiet, quiet, hide yourself from harm | Behind that fern.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}  \\  
\textsc{ii}. \B{kxu si} \I{[k'u si]} \emph{vin.} harm.
\B{Po\-eru kxu rä\-'ä si!} Do not harm her!~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b} 
\B{Rä\-'ä fmi\-vi kxu si\-vi ay\-ioang\-ur fu hel\-kur fe\-yä} Do not attempt to harm the animals or their habitats.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b} \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxutu}{\B{kxu\-tu}}, enemy. 
\on{2}.~\HL{kxuke}{\B{kxu\-ke}}, safe. 
\on{3}.~\HL{mAkxu}{\B{mä\-kxu}}, interrupt. 
\on{4}.~\HL{kxutslu}{\B{kxu\-tslu}}, risk.

% kxuke
\hi
\HT{kxuke}{\B{\cp{kxu}ke}} \I{["k'u.kE]} \emph{adj.} safe. 
(\ding{43}~\HL{kxu}{\B{kxu}}, \HL{luke}{\B{lu\-ke}}) \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{tIkxuke}{\B{tì\-kxu\-ke}}, safety.

% kxukx
\hi
\HT{kxukx}{\B{kxukx}} \I{[k'uk']} \emph{or} \I{[k'Uk']} \emph{vtr.} swallow.   
\B{Fì\-na\-er ftxì\-vä' lu nì\-hawng, ha swey\-lu txo ngal tsat kxi\-vukx nì\-win.} This drink tastes horrible, so you'd better swallow it quickly.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 

% kxum
\hi
\HT{kxum}{\B{kxum}} \I{[k'um]} \emph{or} \I{[k'Um]} \emph{adj.} viscous, gelatinous, thick. \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{kxumpay}{\B{kxum\-pay}}, viscous liquid, gel.

% kxumpay
\hi
\HT{kxumpay}{\B{\cp{kxum}pay}} \I{["k'um.paj]} \emph{or} \I{["k'Um.]} (\textsc{rn}: \B{\cp{kxum}\-pay} \I{["gum.paj]}) \emph{n.} viscous liquid, gel. 
(\emph{a. idiom}) \B{(Na) ken\-ten mì kum\-pay}, being in an environment where one is prevented from acting naturally or doing what one wants to do, [lit.: (like) a fan lizard in gel] \B{Nar\-mew oe fo\-ru na'\-rìng\-ä tì\-lor\-it wi\-vìn\-txu, slä ke tsä\-ngun fo tsli\-vam. 'Efu oe na ken\-ten mì kum\-pay.} I wanted to show them the beauty of the forest, but sadly, they weren't able to understand. I felt completely stymied.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{kxum}{\B{kxum}}, \HL{pay}{\B{pay}}) \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{txekxumpay}{\B{txe\-kxum\-pay}}, magma, lava.
\on{2}.~\HL{kxumpaysyar}{\B{kxum\-pay\-syar}}, glue.

% kxumpaysyar
\hi 
\HT{kxumpaysyar}{\B{\cp{kxum}paysyar}} \I{["k'um.paj.sjaR]} \emph{or} \I{["k'Um.]} \emph{n.} glue. 
\B{Pol kxum\-pay\-syar\-it so\-lar syey\-kar rìk\-it sìn kxem\-yo.} She stuck the leaf onto the wall with glue. [lit.: ``She used glue (and then) stuck the leaf onto the wall.'']~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{kxumpay}{\B{kxum\-pay}}, \HL{syar}{\B{syar}})

% kxutu
\hi
\HT{kxutu}{\B{\cp{kxu}tu}} \I{["k'u.tu]} (\textsc{rn}: \B{\cp{kxu}tu} \I{["gu.tu]}) \emph{n.} enemy. 
\B{Ke za\-syup lì\-'O\-na ne kxu\-tu a mì\-fa fu a wrr\-pa.} The l'Ona will not perish to the enemy within or the enemy without.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g}
\B{Ay\-ha\-pxì\-tu po\-ngu\-ä txo\-pu si nì\-nän ta\-krr\-a Va'\-rul pxe\-ku\-tut lä\-txayn.} The members of the group are less afraid since Va’ru defeated three of the enemy.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Nì\-trr\-trr pxì\-mun\-'i sam\-si\-yul ay\-swi\-zaw\-it ku\-tu\-ä a\-law\-nä\-txayn sno\-kip nì\-'eng.} Warriors typically share the arrows of their defeated enemies among themselves.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{kxu}{\B{kxu}})

% kxutslu
\hi 
\HT{kxutslu}{\B{\cp{kxu}tslu}} \I{["k'u.\t{ts}lu]} \emph{n.} risk. 
(\ding{43}~\HL{kxu}{\B{kxu}}, \HL{tItsunslu}{\B{tìtsunslu}})

\end{multicols}