\pdfbookmark[0]{L}{L}
\begin{center}
\section*{L \I{[l]} (\emph{LeL})}
\end{center}

\begin{multicols}{2}

% -l
\hi
\HT{l}{\B{-l}} \I{[l]} \emph{suf. to mark the agentive case of a n. ending in a vowel or diphthong}. 
(\ding{43}~\HL{Il}{\B{-ìl}})

% la'a
\hi 
\HT{la'a}{\B{\cp{la}'a}} \I{["la.Pa]} \emph{n.} physical separation, distance between two places or objects. 
\B{Kxawm set 'u a\-ngä\-zìk fra\-to lu la\-'a a\-yll.} Maybe the most difficult thing now is social distance.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{pela'a}{\B{pe\-la\-'a}}, how near, how far, what distance.
\on{2}.~\HL{la'ape}{\B{la\-'a\-pe}}, how near, how far, what distance.

% la'ang
\hi 
\HT{la'ang}{\B{\cp{la}'ang}} \I{["la.Pang]} \emph{n.} pile of stinking, rotting animal matter. 
\B{'U\-pe ke zo?--Fraw mì la\-'ang.} What's the matter?--Everything is screwed up.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\emph{a. expression}) \B{Pe\-la\-'ang?} `What the hell? What's going on?' [lit.: what pile of stinking, rotting animal matter?]~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/\#comment-31420}{b}

% la'ape
\hi 
\HT{la'ape}{\B{\cp{la}'ape}} \I{["la.Pa.pE]} \emph{inter.}  how near? how far? what distance?
(\ding{43}~\HL{la'a}{\B{la'a}}, \HL{pe}{\B{pe}\texttt{+}}; \HL{pelImsim}{\B{pelìmsim\slash lìmsimpe}})

% la'um
\hi 
\HT{la'um}{\B{\cp{la}'um}} \I{["la.Pum]} \emph{or} \I{[.PUm]} (\textsc{rn}: \B{\cp{la}\-ùm} \I{["la.Um]}) \emph{vin.} pretend (\emph{used with} \ding{43}~\HL{tsnI}{\B{tsnì}}). 
\B{Pll\-txe po san nga yaw\-ne lu oer sìk, sl\"a la\-'um nì\-'aw.} He says he loves you, but he’s only pretending.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
\B{Lum\-pe nga la\-'um tsnì ke tsun sri\-vew?} Why are you pretending (that) you can't dance?~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIla'um}{\B{tì\-la\-'um}}, pretence.

% lafyon
\hi 
\HT{lafyon}{\B{la\cp{fyon}}} \I{[la."fjon]} \emph{adj., ofp.} wise. 
\B{eyk\-tan la\-fyon}, a wise leader.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\HL{hafyon}{\B{hafyon}}, \HL{hafyonga'}{\B{hafyonga'}})

% lahe
\hi
\HT{lahe}{\B{\cp{la}he}} \I{["la.hE]} \textsc{i}.~\emph{adj.} other, else. 
\B{Fì\-txon na ton a\-la\-he nì\-wotx pe\-lun ke lu teng?} Why isn't this night the same like all the others?~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
\B{Lu nga sì la\-he\-a ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ay\-aw\-po a ma\-wey\-pe\-rey.} It's you and the other members of the sprachbund who are being patient.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Lu law 'uo a\-la\-he, ma ey\-lan.} Something else is clear to me, my friends.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{c. idiom}) \B{slä ha\-ya\-lo a\-la\-he} ``\ldots but that is for another time'' (\emph{set phrase in storytelling})~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b} \\
\textsc{ii}.~\emph{pn.} 
\B{tsa\-kem rä\-'ä si\-vi ay\-la\-he\-ru}, don't do that action to others.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} 
\B{Oel po\-ti tspì\-yang fte tì\-ke\-nong lì\-ye\-vu ay\-la\-he\-ru\slash aylaru.} I will kill him as an example to the others.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/}{ln} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lapo}{\B{la\-po}}, another (one). 
\on{2}.~\HL{saylahe}{\B{say\-la\-he}}, et cetera.

% lak
\hi 
\HT{lak}{\B{lak}} \I{[lak\textcorner]} \emph{n.} shell, \emph{hard covering of a plant or animal}. 
(\ding{43}~\HL{sum}{\B{sum}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tslikxyutsawlak}{\B{tslikxyu tsawlak}}, scarab crawler.

% lal
\hi
\HT{lal}{\B{lal}} \I{[lal]}  \emph{adj.} old (\emph{not for people}).
\B{lal\-a sä\-fpìl}, an old idea.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
\B{Hum zì\-sìt a\-lal, pä\-hem pum a\-mip. Yo'\-ko\-fya a\-tì\-'i\-lu\-ke.} The old year leaves, the new one arrives. An endless cycle.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
(\ding{214}~\HL{mip}{\B{mip}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{txanlal}{\B{txan\-lal}}, ancient, very old.

% lam
\hi
\HT{lam}{\B{lam}} \I{[lam]}  \emph{vin.} seem, appear. (\emph{a.}) 
\B{fay\-he\-tu\-wong fmi Na'\-vi\-na li\-vam}, these aliens try to look like Na'vi.~\LINK{http://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie}{wiki\slash a\textsuperscript{c}}
\B{Poe\-ri unil\-tì\-ran\-tokx\-it tar\-mok a krr lam stum nì\-ay\-fo.} When she was in her avatar body she almost appear like them.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\emph{b. idiom}) \B{lam oer (fwa)}, it seems\slash appears to me (that), seemingly.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{lam fwa}, it seems that. \B{Lam ngay oer, fo ka\-yä ìlä hil\-van tup na'\-rìng.} I bet they'll take the river rather than the forest.~\LINK{http://forum.learnnavi.org/language-updates/concerning-tup/}{ln} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIlam}{\B{tì\-lam}}, appearance.   
\on{2}.~\HL{nIlam}{\B{nì\-lam}}, apparently. 

% lamaytxa
\hi 
\HT{lamaytxa}{\B{lamay\cp{txa}}} \I{[la.maj."t'a]} \emph{n.} flood, powerful gathering of water. 
\B{La\-may\-txal a\-txan pxay\-a kel\-kut sko\-la\-'ä\-nga.} The great flood sadly destroyed many homes.~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b}

% lan
\hi 
\HT{lan}{\B{lan}} \I{[lan]} \emph{n.} resin. \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lanutral}{\B{la\-nut\-ral}}, dandetiger.

% lanay'ka
\hi 
\HT{lanay'ka}{\B{la\cp{nay'}ka}} \I{[la."najP.ka]} \emph{n., fauna} slinger, \emph{acediacutus xenoterribili}. 
(\emph{a. idiom}) \B{'Aw\-si\-teng lu me\-fo la\-nay'\-ka.} \emph{praising how well two people work together and complement each other} [lit.: they are a slinger (together)].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{To\-la\-ron me\-fol me\-sa\-li\-o\-a\-ngit! Tew\-ti, 'aw\-si\-teng lu me\-fo la\-nay'\-ka.} They hunted two sturmbeest? Wow, they work very well together.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\emph{b. idiom}) \B{(na) la\-nay'\-ka lu\-ke re\-'o}, ``completely lost,'' [lit.: Like a slinger without a head].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Po maw kxitx mun\-txa\-tu\-ä 'a\-me\-fu na la\-nay'\-ka re'\-o\-lu\-ke.} After the death of his spouse, he felt completely lost.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% lante
\hi 
\HT{lante}{\B{\cp{lan}te}} \I{["lan.tE]} \emph{vin.} wander. 
\B{Fo ka na'\-rìng le\-ran\-te teng\-krr syu\-vet fwe\-rew.} They're wandering across the forest looking for food.~\LINK{http://naviteri.org/2023/05/reef-navi-part-2-morphology-syntax-lexicon-and-more/}{b}

% lanutral
\hi 
\HT{lanutral}{\B{\cp{la}nutral}} \I{["la.nut\textcorner.Ral]} \emph{or} \I{[.nUt\textcorner.]} \emph{n.} \ding{96} dandetiger, \emph{candea inflata}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln} 
(\ding{43}~\HL{lan}{\B{lan}}, \HL{utral}{\B{ut\-ral}})

% lang
\hi
\HT{lang}{\B{lang}} \I{[laN]} \emph{vtr.} investigate, explore. 
\B{Lum\-pe le\-rang Kel\-ut\-ral\-it Saw\-tu\-tel?} Why are the humans exploring Hometree?~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIlang}{\B{tì\-lang}}, exploration. 
\on{2}.~\HL{sAlang}{\B{sä\-lang}}, exploration, investigation.

% lapo
\hi
\HT{lapo}{\B{\cp{la}po}} \I{["la.po]} (\emph{gen.}~\B{lapeyä}) \emph{pn.} another one (person or thing). 
(\ding{43}~\HL{lahe}{\B{la\-he}}, \HL{po}{\B{po}}) 

% lapx
\hi 
\HT{lapx}{\B{lapx}} \I{[lap']} \emph{vtr.} regret s.t. 
\B{Kem\-it a oe so\-li oel lä\-ngapx.} I regret what I did.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}  \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIlapx}{\B{tì\-lapx}}, regret. 

% lare
\hi 
\HT{lare}{\B{\cp{la}re}} \I{["la.RE]} \emph{vin.} be watchful, alert (\emph{staying alert and watchful, keenly aware of your surroundings, particularly in dangerous circumstances}). 
\B{Tsun na'\-rìng le\-hrr\-ap li\-vu, ma 'e\-veng. Ze\-ne li\-va\-re fra\-krr.} The forest can be dangerous, child. You need to be alert at all times.~\LINK{https://naviteri.org/2025/03/conlang-adventure-and-more/}{b} 
(\ding{43}~\HL{nari}{\B{na\-ri si}}) 

% laro / si
\hi
\HT{laro}{\B{\cp{la}ro}} \I{["la.Ro]} \textsc{i}. \emph{adj.} clean. \\ 
\textsc{ii}. \B{\cp{la}ro si} \I{["la.Ro si]} \emph{vin.} clean, make free of dirt.
\B{Txo ke li\-vu pay, tsun me\-syokx\-ur la\-ro si\-vi fa srä.} If water isn't available, you can clean your hands with a cloth.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{214}~\HL{tsewtx}{\B{tsewtx}}; \ding{43}~\HL{yur}{\B{yur}})

% latem
\hi
\HT{latem}{\B{\cp{la}tem}} \I{["la.tEm]}  \emph{vin.} (\emph{a.}) change, i.e. s.t. changes.
\B{fra\-'u la\-tem}, everything changes.
\B{Ke\-zem\-pll\-txe, ta\-lun tì\-vi\-rä fì\-sä\-spxin\-ä, lo\-la\-tä\-ngem ki\-fkey, lo\-la\-tä\-ngem tì\-rey.} Needless to say, because of the spread of this illness, the world and life changes.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b}
(\emph{b.~with} \HL{IlA}{\B{ìlä}}) depend on. \B{Tì\-flä la\-tem ìlä seyng\-a ftxey fkol sä\-nu\-met li\-vek fu\-ke.} Success depends on whether or not one follows instructions.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{leykatem}{\B{ley\-ka\-tem}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIlatem}{\B{tì\-la\-tem}}, change (abstract concept). 
\on{2}.~\HL{sAlatem}{\B{sä\-la\-tem}}, (instance of) change.

% latsi
\hi
\HT{latsi}{\B{la\cp{tsi}}} \I{[la."\t{ts}i]}  \emph{vin.} keep up with s.o. 
\B{Slä pll\-txe rä\-'ä san Oe ke tsun li\-va\-tsi sìk!} But don't say, `I can't keep up!'~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/\#comment-31193}{b}

% law / si
\hi
\HT{law}{\B{law}} \I{[law]} \textsc{i}.~\emph{adj.} clear, certain. (\emph{a.}) \B{Pì\-lok\-ä fì\-ha\-pxì\-yä tì\-kan lu law.} The aim of this part of the blog is clear.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Poe pol\-txe nì\-fya\-'o a\-law.} She spoke clearly [lit.: in a clear manner].~\LINK{http://forum.learnnavi.org/language-updates/why-is-this-night/msg154875/\#msg154875}{ln}
(\emph{b.~with the dative}) be clear to one. \B{Pe\-yä 'i\-tan\-ìri lu ho\-na nì\-txan a fì\-'u law lu fra\-por.} It's clear to everyone that his son is very cute.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
\B{Nge\-yä Ti\-re\-a\-i\-o\-ang zo\-la\-'u a krr, law la\-yu nga\-ru.} When your Spirit Animal comes, you will know [lit.: it will be clear to you].~\LINK{http://forum.learnnavi.org/language-updates/auxilary-verb-si-possessive-dative-krr/}{ln} \\
\textsc{ii}.~\HT{law si}{\B{law si}} \I{[law si]}  \emph{vin.} make clear (\emph{allows for double dative}).   
\B{Tsun nga law si\-vi nì\-'it srak?} Could you make that a bit clearer?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
\B{Ra\-lur law so\-li fo oe\-ru.} They made the meaning clear to me.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIlaw}{\B{nì\-law}}, clearly.

% lawk
\hi
\HT{lawk}{\B{lawk}} \I{[lawk\textcorner]}  \emph{vtr.} discourse on, talk about, say s.t. concerning.   
\B{Txan\-krr ngal ke lawk pot kaw\-'it.} You haven't mentioned a thing about him in a long time.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Po\-el oe\-ti lar\-mawk.} She was talking about me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\emph{b. in the reflexive} \ding{43}~\HL{lApawk}{\B{lä\-pawk}}) introduce oneself. 

% lawnol
\hi
\HT{lawnol}{\B{\cp{law}nol}} \I{["law.nol]} \emph{n.} great joy.
\B{Ay\-zì\-sìt a\-zu\-sa\-'u ta law\-nol te\-ya li\-vu nì\-wotx.} May the coming years be full of great joy.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{'onglawn}{\B{'ong\-lawn}}, \emph{exhiliration of first bonding}.
\on{2}.~\HL{lawnoltsim}{\B{law\-nol\-tsim}}, source of (great) joy.

% lawnoltsim
\hi 
\HT{lawnoltsim}{\B{\cp{law}noltsim}} \I{["law.nol.\t{ts}im]} \emph{or coll.} \I{["law.no.\t{ts}im]} \emph{n.} source of (great) joy. 
\B{Oe\-ri leiu fì\-lì'\-fya\-olo' law\-nol\-tsim.} This language community is a source of great joy to me.~\LINK{naviteri.org/2022/05/results-of-the-tttc/}{b} 
(\ding{43}~\HL{lawnol}{\B{law\-nol}}, \HL{tsim}{\B{tsim}})

% lawr
\hi
\HT{lawr}{\B{lawr}} \I{[lawR]} \textsc{i}.~\emph{n.} tune, melody. \\
\textsc{ii}.~\HT{tIng lawr}{\B{tìng lawr}} \I{[tIN lawR]} \emph{vin.} sing wordlessly, give out a tune or melody.
\B{'E\-van\-ìl alo a\-'aw\-ve nì\-'aw\-tu na'\-rìng\-it tar\-mok, ha to\-lìng lawr nì\-fwe\-fwi fte\-ke txo\-pu si\-vi.} The boy was alone in the forest for the first time, so he whistled a tune to calm his fears.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}

% layl
\hi 
\HT{layl}{\B{layl}} \I{[lajl]} \emph{adj.} innocent. 
\B{Tsa\-kawng\-kem\-ìri lu oe layl!} I am innocent of that crime!~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(\ding{214}~\HL{tokat}{\B{to\-kat}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIlayl}{\B{tì\-layl}}, innocence. 
\on{2}.~\HL{nIlayl}{\B{nì\-layl}}, innocently. 

% layompin
\hi
\HT{layompin}{\B{la\cp{yom}pin}} \I{[la."jom.pin]}  \emph{n.} the color \ding{43}~\HL{layon}{\B{la\-yon}}. 
(\ding{43}~\HL{'opin}{\B{'opin}})

% layon
\hi
\HT{layon}{\B{la\cp{yon}}} \I{[la."jon]}  \emph{adj.} black.
\B{mrr\-a ya\-yo a\-tsawl sì la\-yon\slash mrr\-a ya\-yo a lu tsawl sì la\-yon}, five big, black birds.~\LINK{http://naviteri.org/2021/02/aysipawm-si-aysieyng-questions-and-answers/\#comment-32798}{b}
\B{tsawl\-a ya\-yo a\-la\-yon}, a big, black bird.~\LINK{http://naviteri.org/2021/02/aysipawm-si-aysieyng-questions-and-answers/\#comment-32798}{b}
(\ding{214}~\HL{teyr}{\B{teyr}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{layompin}{\B{la\-yom\-pin}}, (the color) black.

% layro
\hi 
\HT{layro}{\B{\cp{lay}ro}} \I{["laj.Ro]} \emph{adj.} free (\emph{from oppression}). \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIlayro}{\B{tì\-lay\-ro}}, freedom. 

% läpawk
\hi
\HT{lApawk}{\B{lä\cp{pawk}}} \I{[l\ae."p$\cdot$awk\textcorner]} \emph{vin., inf.}\on{22} introduce oneself, talk about oneself.
\B{Ru\-txe lä\-pi\-vawk nì\-'it.} Please talk a little bit about yourself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Po lä\-po\-lawk.} He talked about himself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Nga lä\-pi\-vawk nì\-no ko.} Tell me all about yourself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\emph{reflexive of} \ding{43}~\HL{lawk}{\B{lawk}})

% lätxayn
\hi
\HT{lAtxayn}{\B{lä\cp{txayn}}} \I{[l\ae."t'ajn]}  \emph{vtr.} defeat in battle, conquer.
\B{Fwa ay\-ioang a\-pxay fì\-txan Na'\-vi\-ru srung so\-li fte Saw\-tu\-tet li\-vä\-txayn lu pa\-rul nì\-ngay.} That so many animals helped the Na’vi defeat the Sky People was a genuine miracle.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
\B{Nì\-trr\-trr pxì\-mun\-'i sam\-si\-yul ay\-swi\-zaw\-it ku\-tu\-ä a\-law\-nä\-txayn sno\-kip nì\-'eng.} Warriors typically share the arrows of their defeated enemies among themselves.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}  \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{sAlAtxayn}{\B{sä\-lä\-txayn}}, defeat.

% le-
\hi
\HT{le}{\B{le-}} \I{[lE.]} \emph{unprod. pref. to form adj.s with unpredictable meaning}.

% le'aw
\hi
\HT{le'aw}{\B{le\cp{'aw}}} \I{[lE."Paw]}  \emph{adj.} only, sole.   
\B{Lu tsa\-tsam\-si\-yu le\-'aw\-a ha\-pxì\-tu tsam\-po\-ngu\-ä a mal lu moer.} That warrior is the only member of the war party that we both trust.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{'aw}{\B{'aw}})

% le'awtu
\hi 
\HT{le'awtu}{\B{le\cp{'aw}tu}} \I{[lE."Paw.tu]} \emph{adj.} (\emph{a. neutral connotation}) alone, on one's own, lone, by oneself, solitary. 
\B{Oe 'e\-fu le\-'aw\-tu.} I feel alone.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
\B{Oe lu le\-'aw\-a tu\-te a tsun srung si\-vi, ul\-te 'e\-fu le\-'aw\-tu nì\-ngay.} I'm the only one who can help, and I feel really alone.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
\B{Le\-'aw\-tu\-a ta\-li\-oang\-ìri lu ki\-fkey tse\-nge le\-hrr\-ap.} The world is a dangerous place for a lone sturmbeest.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
(\emph{b. often with} \ding{43} ‹\HL{Ang}{\B{äng}}›) lonely.
\B{Oe 'e\-fä\-ngu le\-'aw\-tu.} I feel lonely.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
\B{Oe lu le\-'aw\-a tu\-te a tsun srung si\-vi, ul\-te 'e\-fu\slash 'e\-fä\-ngu le\-'aw\-tu nì\-ngay.} I'm the only one who can help, and I feel really lonely.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 

% le'al
\hi 
\HT{le'al}{\B{le\cp{'al}}} \I{[lE."Pal]} \emph{n.} wasteful (\emph{not for people}).
\B{Fwa sar pay\-it fì\-txan lu le\-'al.} Using this much water is wasteful.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} 
(\ding{43}~\HL{'al}{\B{'al}})

% le'en
\hi
\HT{le'en}{\B{le\cp{'en}}} \I{[lE."PEn]}  \emph{adj.} speculative, intuitive (\emph{of an action, not a person}).
\B{Hem le\-'en tsun le\-hrrap li\-vu.} Speculative moves can be dangerous.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{'en}{\B{'en}})

% le'Ìnglìsì
\hi
\HT{le'InglIsI}{\B{le\cp{'Ìng}lìsì}} \I{[lE."PIN.lI.sI]} \emph{adj.}~\textsc{lw} English, regarding English. 
\B{Kxawm tsa\-lì\-'u le\-'Ìng\-lì\-sì alu ``No\"el'' mì ron\-sem lar\-mu.} Maybe I had the English word ``No\"el'' in my mind.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/comment-page-1/\#comment-499}{b}
(\ding{43}~\HL{'InglIsI}{\B{'Ìng\-lì\-sì}})

% lefkeytongay
\hi 
\HT{lefkeytongay}{\B{lefkeyto\cp{ngay}}} \I{[lE.fkEj.to."Naj]} \emph{adj.} real. 
\B{Yu\-ne oet! Ke lu fì\-vrr\-tep tu\-te le\-fkey\-to\-ngay!} Listen to me! This demon is not a real person!~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{tIfkeytongay}{\B{tì\-fkey\-to\-ngay}})

% lefkrr
\hi
\HT{lefkrr}{\B{le\cp{fkrr}}} \I{[lE."fk\s{r}]}  \emph{adj.} current. 
\B{Tì\-fkey\-tok le\-fkrr le\-hrr\-ap lu nì\-txan.} The current situation is very dangerous.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{… fu\-la tsun le\-fkrr\-a sì\-len\-it tswi\-va' hì\-krr (…) oe\-ti 'ey\-ke\-fu nit\-ram nì\-txan.} … that we can forget current events for a short time, makes me very happy.~\LINK{http://naviteri.org/2022/05/results-of-the-tttc/}{b}
(\ding{43}~\HL{krr}{\B{krr}})

% lefnelan
\hi 
\HT{lefnelan}{\B{le\cp{fne}lan}} \I{[lE."fnE.lan]} \emph{adj.} male. 
(\ding{43}~\HL{fnelan}{\B{fne\-lan}})

% lefnele
\hi 
\HT{lefnele}{\B{le\cp{fne}le}} \I{[lE."fnE.lE]} \emph{adj.} female. 
(\ding{43}~\HL{fnele}{\B{fne\-le}})

% lefngap
\hi
\HT{lefngap}{\B{le\cp{fngap}}} \I{[lE."fNap\textcorner]}  \emph{adj.} metallic. \B{el\-tu le\-fngap}, computer, \textsc{pc} [lit.: metallic brain].~\LINK{http://forum.learnnavi.org/language-updates/frommerian-words-from-nytimes-6442/}{ln}
\B{Kì\-ma\-nom oel mip\-a el\-tut le\-fngap yoa ew\-ro.} I just bought a new computer.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
(\ding{43}~\HL{fngap}{\B{fngap}})

% lefpom
\hi
\HT{lefpom}{\B{le\cp{fpom}}} \I{[lE."fpom]}  \emph{adj., nfp.} happy, peaceful, joyous, pleasant. 
\B{trr le\-fpom}, good day, good morning (cf.~French \emph{bon jour}).
\B{Trr le\-fpom, ma Su\-te Ame\-ri\-ka\-yä.} Good morning, America.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
\B{Txon le\-fpom (li\-vu ngar).} (May you have a) good night. 
\B{Ay\-ftxo\-zä Le\-fpom!}~Happy Holidays!
\B{Mipa Zìsìt Lefpom, ma frapo!}~Happy New Year, y'all!~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Txep\-ìl tìng le\-fpom\-a sä\-nrr\-ti.} The fire gives a pleasant glow.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{fpom}{\B{fpom}}; \HL{nitram}{\B{nit\-ram}})

% lefpomron
\hi
\HT{lefpomron}{\B{lefpom\cp{ron}}} \I{[lE.fpom."Ron]} \emph{adj., ofp.} (mentally) healthy.
(\ding{43}~\HL{fpomron}{\B{fpom\-ron}}, \ding{214}~\HL{kelfpomron}{\B{kel\-fpom\-ron}})

% lefpomtokx
\hi
\HT{lefpomtokx}{\B{lefpom\cp{tokx}}} \I{[lE.fpom."tok']}  \emph{adj., ofp.} (physically) healthy.
\B{Nga\-ri ftxì vi\-var le\-fpom\-tokx li\-vu, ma tsmuk!} May your tongue remain healthy, sibling!~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/comment-page-1/\#comment-10624}{b}
(\ding{43}~\HL{fpomtokx}{\B{fpom\-tokx}}, \ding{214}~\HL{kelfpomtokx}{\B{kel\-fpom\-tokx}})

% lefrir
\hi 
\HT{lefrir}{\B{le\cp{frir}}} \I{[lE."fRiR]} \emph{adj.} layered. 
(\ding{43}~\HL{frir}{\B{frir}})

% leftxozä
\hi 
\HT{leftxozA}{\B{leftxo\cp{zä}}} \I{[lE.ft'o."z\ae]} \emph{adj.} celebratory. 
\B{Ay\-nge\-yä lor\-a fì\-'u\-pxa\-re le\-ftxo\-zä oe\-ru te\-ya so\-li nì\-ngay.} Your beautiful, celebratory message truly filled me with joy.~\LINK{https://forum.learnnavi.org/language-updates/new-words-leftxoza-(celebratory)-poston-(boston)-kelni-(koln-cologne)/}{ln} 
(\ding{43}~\HL{ftxozA}{\B{ftxo\-zä}})

% leha'
\hi 
\HT{leha'}{\B{le\cp{ha'}}} \I{[lE."haP]} \emph{adj.} appropriate, suitable, fitting. 
\B{Fo\-ri tsa\-fne\-tì\-ku\-sar ke lu le\-ha'.} That kind of teaching isn't appropriate for them.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} 
(\ding{214}~\HL{kelha'}{\B{kel\-ha'}} \ding{43}~\HL{ha'}{\B{ha'}})

% lehangham
\hi 
\HT{lehangham}{\B{le\cp{hang}ham}} \I{[lE."haN.ham]} \emph{adj.} laughable, ridiculous, absurd. 
\B{Fo\-le\-nge fra\-pol pe\-yä tì\-hawl\-it le\-hang\-ham.} Everyone ridiculed his ridiculous plan.~\LINK{naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b}  
(\ding{43}~\HL{hangham}{\B{hang\-ham}})

% lehawmpam
\hi
\HT{lehawmpam}{\B{le\cp{hawm}pam}} \I{[lE."hawm.pam]}  \emph{adj.} noisy. 
\B{Ta\-ron\-yul le\-hawm\-pam ska\-'a sä\-ta\-ron\-it.} A noisy hunter destroys the hunt.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{hawmpam}{\B{hawmpam}})

% lehawng
\hi
\HT{lehawng}{\B{le\cp{hawng}}} \I{[lE."hawN]}  \emph{adj.} excessive. 
\B{Pxìm lu tì\-new le\-hawng kxu\-tu fpom\-ä.} Excessive desire is often the enemy of peace.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{hawng}{\B{hawng}})

% lehawngkrr
\hi
\HT{lehawngkrr}{\B{le\cp{hawng}krr}} \I{[lE."hawN.k\s{r}]}  \emph{adj.} late. 
\B{tì\-pä\-hem le\-hawng\-krr}, a late arrival.
(\ding{43}~\HL{hawngkrr}{\B{hawng\-krr}}, \ding{214}~\HL{leye'krr}{\B{le\-ye'\-krr}})

% leheng
\hi 
\HT{leheng}{\B{le\cp{heng}}} \I{[lE."hEN]} \emph{adj.} formally polite. \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lI'fyeng}{\B{lì'\-fyeng}}, honorific language. 

% lehìpey
\hi
\HT{lehIpey}{\B{le\cp{hì}pey}} \I{[lE."hI.pEj]} \emph{adj.} hesitant, in a state of hesitation.
\B{Ta\-ron\-yul le\-hì\-pey kan smar\-it nìl\-ke\-ftang slä ke ta\-kuk kaw\-krr.} A hesitant hunter will aim at a prey forever but never hit it.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{hIpey}{\B{hì\-pey}})

% lehoan
\hi
\HT{lehoan}{\B{le\cp{ho}an}} \I{[lE."ho.an]} \emph{adj.} comfortable.
(\ding{43}~\HL{hoan}{\B{ho\-an}}, \ding{214}~\HL{kelhoan}{\B{kel\-ho\-an}})

% lehrrap
\hi
\HT{lehrrap}{\B{le\cp{hrr}ap}} \I{[lE."h\s{r}.ap\textcorner]}  \emph{adj.} dangerous. 
\B{Lu fo le\-hrr\-ap.} They are dangerous.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
(\ding{43}~\HL{hrrap}{\B{hrrap}})

% leioae / si
\hi
\HT{leioae}{\B{leio\cp{a}e}} \I{[lE.i.o."{}a.E]} \textsc{i}.~\emph{n.} respect.   
\B{Lu\-ke le\-i\-o\-a\-e o\-lo'\-ä ke tsun kea eyk\-tan fli\-vä.} Without the respect of the clan, no leader can succeed.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{le\-i\-o\-a\-e a\-mek,} feigned respect.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} \\
\textsc{ii}.~\HT{leioae si}{\B{leio\cp{a}e si}} \I{[lE.i.o."{}a.E si]}  \emph{vin.} to respect (s.o.~with the dative). 
\B{Nga\-ru le\-i\-o\-a\-e si oe fra\-to, ma 'ey\-lan.} I respect you the most of all, friend.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{leiokoaktu}{\B{le\-io\-ko\-ak\-tan}}, respected elder.

% leiokoaktan
\hi 
\HT{leiokoaktan}{\B{leio\cp{ko}aktan}} \I{[lE.i.o."ko.ak\textcorner.tan]} \emph{n} respected male elder. 
(\ding{214}~\HL{leiokoakte}{\B{le\-i\-o\-ko\-ak\-te}}, \ding{43}~\HL{leiokoaktu}{\B{le\-i\-o\-ko\-ak\-tu}})

% leiokoakte
\hi 
\HT{leiokoakte}{\B{leio\cp{ko}akte}} \I{[lE.i.o."ko.ak\textcorner.tE]} \emph{n.} respected female elder. 
\B{Nge\-nge\-yä ha\-fyon\-ìri ira\-yo, ma le\-i\-o\-ko\-ak\-te.} Thank you for your wisdom, respected female elder.~\LINK{https://naviteri.org/2025/03/conlang-adventure-and-more/}{b} 
(\ding{214}~\HL{leiokoaktan}{\B{le\-i\-o\-ko\-ak\-tan}}, \ding{43}~\HL{leiokoaktu}{\B{le\-i\-o\-ko\-ak\-tu}}) 

% leiokoaktu
\hi 
\HT{leiokoaktu}{\B{leio\cp{ko}aktu}} \I{[lE.i.o."ko.ak\textcorner.tu]} \emph{n.} respected elder. 
(\ding{43}~\HL{leioae}{\B{le\-i\-o\-a\-e}}, \HL{koaktu}{\B{ko\-ak\-tu}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{leiokoaktan}{\B{le\-io\-ko\-ak\-tan}}, respected male elder.
\on{2}.~\HL{leiokoakte}{\B{le\-i\-o\-ko\-ak\-te}}, respected female elder.

% lek
\hi
\HT{lek}{\B{lek}} \I{[lEk\textcorner]}  \emph{vtr.} heed, obey, follow (command, order, suggestion etc.). 
\B{Tì\-flä la\-tem ìlä seyng\-a ftxey fkol sä\-nu\-met li\-vek fu\-ke.} Success depends on whether or not one follows instructions.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Ta\-fral ke lì\-syek oel nge\-yä ke\-ye\-'ung\-it.} Therefore I will not heed your insanity.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lek}{\B{ley\-kek}}, enforce.

% leketengsyon
\hi 
\HT{leketengsyon}{\B{le\cp{ke}tengsyon}} \I{[lE."kE.tEN.sjon]} \emph{adj.} diverse, various. 
\B{Pll\-txe Saw\-tu\-te fa ay\-lì'\-fya le\-ke\-teng\-syon.} Humans speak diverse (OR: a variety of) languages.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
\B{Ay\-u le\-ke\-teng\-syon.} Various things.~\LINK{https://naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b}
(\ding{43}~\HL{ketengsyon}{\B{ke\-teng\-syon}}) 

% lekin
\hi
\HT{lekin}{\B{le\cp{kin}}} \I{[lE."kin]}  \emph{adj.} necessary.
\B{Le\-kin lu tì\-txur, lu tì\-tstew.~Slä le\-tsran\-ten fra\-to lu tì\-mal.} Strength and courage are necessary.~But most important of all is trustworthiness.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{kin}{\B{kin}}; \ding{214}~\HL{kelkin}{\B{kel\-kin}})

% lekoren
\hi
\HT{lekoren}{\B{leko\cp{ren}}} \I{[lE.ko."REn]}  \emph{adj.} lawlike, regarding rules.
\B{päng\-kxo\-yu le\-ko\-ren}, lawyer, one who discusses rules.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-words-from-nytimes-6442/}{ln}
(\ding{43}~\HL{koren}{\B{ko\-ren}})

% lekve'o
\hi 
\HT{lekve'o}{\B{lek\cp{ve}'o}} \I{[lEk\textcorner."vE.Po]} \emph{adj.} chaotic. 
\B{Lam oe\-ru ki\-fkey lek\-ve\-'o.} The world seems chaotic to me.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{ve'o}{\B{ve\-'o}}, \HL{velke}{\B{vel\-ke}}) 

% lekye'ung
\hi
\HT{lekye'ung}{\B{lek\cp{ye}'ung}} \I{[lEk\textcorner."jE.PuN]} \emph{or} \I{[.PUN]} (\textsc{rn}: \B{lek\cp{ye}ùng} \I{[lEk\textcorner.UN]}) \emph{adj.} insane, crazy. 
\B{Fì\-sä\-'an\-lal Ne\-we\-yä nga\-ti sley\-ka\-yu lek\-ye\-'ung!} This yearning for Newey is going to drive you crazy.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Krr\-ka lek\-ye\-'ung\-a fay\-srr, sìl\-pey oe, li\-vu ay\-nga\-ru fpom nì\-wotx.} During these crazy days, I hope you are all well.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{keye'ung}{\B{ke\-ye\-'ung}})

% lekxan
\hi 
\HT{lekxan}{\B{le\cp{kxan}}} \I{[lE."k'an]} \emph{adj.} (\emph{a. literally}) blocked, obstructed.
\B{Lu fì\-fya\-'o le\-kxan; ke tsun aw\-nga si\-va\-lew.} The path is obstructed; we can't proceed further.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
(\emph{b. metaphorically, with} \ding{43}~\HL{'efu}{\B{'efu}}) frustrated. 
\B{Oe pll\-txe, po ke tìng mik\-yun. 'E\-fu oe le\-kxan nì\-txan.} I talk, but he doesn't listen. I feel very frustrated.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b} 
(\ding{43}~\HL{ekxan}{\B{ekxan}}; \HL{tIkankxang'}{\B{tì\-kan\-kxa\-nga'}})

% lekxutslu
\hi 
\HT{lekxutslu}{\B{le\cp{kxu}tslu}} \I{[lE."k'u.\t{ts}lu]} \emph{adj.} risky. 
\B{Aw\-nga zen\-ke fì\-kem si\-vi. Lu le\-kxu\-tslu nì\-hawng.} We mustn't do this. It's too risky.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}  
(\ding{43}~\HL{kxutslu}{\B{kxutslu}})

% lelewng
\hi 
\HT{lelewng}{\B{le\cp{lewng}}} \I{[lE."lEwN]} \emph{adj., ofp.} shameful. 
\B{tu\-te le\-lewng}, a shameful person.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
(\ding{43}~\HL{lewng}{\B{lewng}}) 

% lelì'fya
\hi
\HT{lelI'fya}{\B{le\cp{lì'}fya}} \I{[lE."lIP.fja]}  \emph{adj.} concerning language.   
\B{tì\-päng\-kxo le\-lì'\-fya}, language discussion.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{sul\-fä\-tu le\-lì'\-fya}, language masters.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/}{b}
(\ding{43}~\HL{lI'fya}{\B{lì'\-fya}})

% lemrey
\hi
\HT{lemrey}{\B{lem\cp{rey}}} \I{[lEm."Rej]}  \emph{adj.} surviving (e.g., of entities from a group some of whom have died). 
\B{Maw fwa fkol Kel\-ut\-ral\-it sko\-la\-'a, lem\-rey\-a ha\-pxì\-tul tsa\-soai\-ä txo\-lu\-la mip\-a kel\-ku\-ti mì ta\-yo.} After Hometree was destroyed, the surviving members of that family built a new home on the plains.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{emrey}{\B{em\-rey}})

% lemweypey
\hi
\HT{lemweypey}{\B{lem\cp{wey}pey}} \I{[lEm."wEj.pEj]}  \emph{adj.} patient.
\B{Za\-'u fra\-'u ne tu\-te lem\-wey\-pey, ke\-fyak?} Everything comes to the patient people, right?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
(\ding{43}~\HL{maweypey}{\B{ma\-wey\-pey}})

% len
\hi
\HT{len}{\B{len}} \I{[lEn]}  \emph{vin.} occur, happen. 
\B{Krro len ay\-hem a\-na\-fì\-'u.} Sometimes such actions happen.~\LINK{http://forum.learnnavi.org/language-updates/just-another-few-words/}{ln} 
\B{Krr a sto\-lawm Txe\-wì te\-ri hem a poe\-ru lo\-len, 'o\-le\-fu ke\-ftxo nì\-txan.} When Txewì heard about what happened to her, he felt very upset.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Nari si fte kea fe\-kem ke li\-ven ngar!} Be careful you don't have an accident!~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
(\emph{a. idiom}) \B{Kem\-pe le\-ren?} `What's happenin'?' 
\B{Kxì, ma 'ey\-lan! Kem\-pe le\-ren?} Hey dude! What's happenin'?.~\LINK{http://naviteri.org/2018/02/negative-questions-in-navi/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIlen}{\B{tì\-len}}, event. 
\on{2}.~\HL{ftanglen}{\B{ftang\-len}}, prevent.

% leNa'vi
\hi
\HT{lena'vi}{\B{le\cp{Na'}vi}} \I{[lE."naP.vi]} \emph{adj.} having to do with the Na'vi, regarding Na'vi. 
\B{Ay\-ey\-lan\-ur oe\-yä sì ey\-lan\-ur lì'\-fya\-yä le\-Na'\-vi nì\-wotx.} To my friends and all the frieands of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Sna\-fpìl\-fya\-ri le\-Na'\-vi krra smar\-it fkol tspang, tsran\-ten nì\-txan fwa po ke ngä\-'än nì\-kel\-kin.} It's important in Na'vi philosophy that the prey not suffer unnecessarily when it's killed.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Kin aw\-ngal le\-Na'\-vi\-a lì\-'ut a ral lu ``typo.''} We need a Na'vi word for `typo.'~\LINK{http://naviteri.org/2015/06/na-ayskxe-mi-telan-sad-news/comment-page-1/\#comment-20585}{b} 
(\ding{43}~\HL{na'vi}{\B{Na'vi}})

% lenalsteng
\hi 
\HT{lenalsteng}{\B{le\cp{nal}steng}} \I{[lE."nal.stEN]} \emph{adj., ofp.} compassionate. 
(\ding{43}~\HL{nalsteng}{\B{nal\-steng}}; \HL{tInalstenga'}{\B{tì\-nal\-ste\-nga'}})

% leno
\hi
\HT{leno}{\B{le\cp{no}}} \I{[lE."no]}  \emph{adj.} thorough, detail\hyp{}oriented (\emph{of a person}).
\B{Le\-no lu Lo\-ak nì\-txan.} Loak is very thorough.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}

% lenomum
\hi
\HT{lenomum}{\B{le\cp{no}mum}} \I{[lE."no.mum]} \emph{or} \I{[.mUm]} (\textsc{rn}: \B{le\cp{no}mùm} \I{[lE."no.mUm]}) \emph{adj.} curious.
\B{Txo\-pu rä\-'ä si. Nga\-ti ke nìn pol nì\-kxap. Lu le\-no\-mum nì\-'aw.} Don't be scared. He's not looking at you threateningly. He's just curious.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{newomum}{\B{new\-omum}})

% lenrra
\hi
\HT{lenrra}{\B{le\cp{nrr}a}} \I{[lE."n\s{r}.a]}  \emph{adj.} proud, \emph{feeling of pride one has in others}.
\B{Sa'\-nok le\-nrr\-a lrr\-tok so\-li krra prr\-nen alo a\-'aw\-ve pol\-txe.} The proud mother smiled when the baby spoke for the first time.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{nrra}{\B{nrra}})

% lepay
\hi
\HT{lepay}{\B{le\cp{pay}}} \I{[lE."paj]}  \emph{adj.} watery, saturated. 
\B{'ak\-ra le\-pay}, watery, saturated soil.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{pay}{\B{pay}}, \HL{paynga'}{\B{pay\-nga'}})

% lepwopx
\hi
\HT{lepwopx}{\B{lep\cp{wopx}}} \I{[lEp\textcorner."wop']}  \emph{adj.} cloudy.
\B{lep\-wopx nì\-hol}, lightly cloudy, just a few clouds.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{lep\-wopx nì\-pxay}, heavily cloud-covered, many clouds.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{pIwopx}{\B{pì\-wopx}})

% lepxìmrun
\hi 
\HT{lepxImrun}{\B{le\cp{pxìm}run}} \I{[lE."p'Im.Run]} \emph{adj.} often\hyp{}found, common.
(\ding{43}~\HL{pxIm}{\B{pxìm}}, \HL{run}{\B{run}}; \ding{214}~\HL{kelpxImrun}{\B{kel\-pxìm\-run}})

% ler
\hi 
\HT{ler}{\B{ler}} \I{[lER]} \emph{adj.} steady, smooth (of motion).  \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIler}{\B{nì\-ler}}, steadily.

% lereyfya
\hi
\HT{lereyfya}{\B{le\cp{rey}fya}} \I{[lE."REj.fja]}  \emph{adj.} cultural. 
\B{Ay\-lì'\-fya\-vi Le\-rey\-fya} Cultural Terms.~\LINK{http://naviteri.org/2015/07/aylifyavi-lereyfya-1-cultural-terms-1/}{b} 
(\ding{43}~\HL{reyfya}{\B{rey\-fya}}) 

% lerìk
\hi
\HT{lerIk}{\B{le\cp{rìk}}} \I{[lE."RIk\textcorner]}  \emph{adj.} leafy. 
\B{Aw\-ngal yom fkxen\-ti le\-rìk nì\-wotx.} We eat leafy vegetable.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
(\ding{43}~\HL{rIk}{\B{rìk}}) \\
\textsc{Derivatives}: \ding{43} \on{1}. \HL{yomyolerIk}{\B{yom\-yo le\-rìk}}, leaf plate.

% lerìn
\hi 
\HT{lerIn}{\B{le\cp{rìn}}} \I{[lE."RIn]} \emph{adj.} wooden, of wood. 
\B{tuk\-ru le\-rìn}, a spear made of wood.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{43}~\HL{rIn}{\B{rìn}})

% leronsrel
\hi
\HT{leronsrel}{\B{le\cp{ron}srel}} \I{[lE."Ron.sREl]} \emph{adj.} imaginary.
\B{Oe new si\-vop ne tsa\-ki\-fkey le\-ron\-srel.} I want to journey to that imaginary world.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\ding{43}~\HL{ronsrel}{\B{ron\-srel}})

% lertu
\hi
\HT{lertu}{\B{\cp{ler}tu}} \I{["lER.tu]}  \emph{n.} colleague, co-worker.
\B{Nga\-ru oe\-yä ler\-tut.} \emph{or} \B{Oel mu\-wi\-vìn\-txu nga\-ru oe\-yä ler\-tut.} (\emph{formal}) Allow me to introduce my colleague.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}

% lesar / si
\hi
\HT{lesar}{\B{le\cp{sar}}} \I{[lE."{}saR]} \textsc{i}.~\emph{adj.} useful.
\B{Tsuk\-yom\-a ioang lu le\-sar.} An edible animal is useful.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}
\B{Kxa\-ri tstu\-tswo tsran\-ten krra ke lu kea sä\-fpìl le\-sar.} When one has no useful thoughts, the ability to close one's mouth is important.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} \\
\textsc{ii}.~\B{le\cp{sar} si} \I{[lE."{}saR si]} \emph{vin.} be useful, be of use, come in handy.
\B{Fu\-la ho\-ren sì ay\-lì'\-fya\-vi a\-mip le\-sar sa\-yi me\-ngar oe\-ti nit\-ram 'ey\-ke\-fu nì\-txan.} It makes me very happy that the rules and new expressions will be useful to you both.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/comment-page-1/\#comment-1515}{b}
(\ding{43}~\HL{sar}{\B{sar}})

% leskxir
\hi
\HT{leskxir}{\B{le\cp{skxir}}} \I{[lE."sk'iR]}  \emph{adj.} wounded.
\B{Oe\-ri ski\-en\-a tsyokx lu le\-skxir.} My right hand is wounded.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{skxir}{\B{skxir}})

% lesnonrra
\hi
\HT{lesnonrra}{\B{lesno\cp{nrr}a}} \I{[lE.sno."n\s{r}.a]}  \emph{adj.} full of self-pride, arrogant.
\B{Kea tsam\-si\-yu le\-sno\-nrr\-a ke tsun Na'\-vit iveyk.} No warrior full of self-pride can lead the People.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Le\-sno\-nrr\-a tu\-te a swä\-pe\-ri fra\-krr slu eyk\-tan a\-fpxa\-mo.} An arrogant individual who always praises himself makes a terrible leader.~\LINK{naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b} 
(\ding{43}~\HL{snonrra}{\B{sno\-nrr\-a}})

% lesngä'i
\hi
\HT{lesngA'i}{\B{le\cp{sngä}'i}} \I{[lE."{}sN\ae.Pi]}  \emph{adj.} original, existing at\slash{}from the start, first in a series.
\B{Swey\-lu txo ngal ke txi\-vìng sä\-fpìl\-it le\-sngä\-'i.} You shouldn't abandon your original idea.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{sngA'i}{\B{sngä\-'i}}, \HL{letsim}{\B{le\-tsim}})

% leso'ha
\hi
\HT{leso'ha}{\B{le\cp{so'}ha}} \I{[lE."{}soP.ha]}  \emph{adj.} enthusiastic, keen (\emph{of a person}).
\B{Ro\-le\-i\-un fu\-ta lu fo ka\-nu sì le\-so'\-ha nì\-wotx.} I found that they are intelligent and completely enthusiastic.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
(\ding{43}~\HL{so'ha}{\B{so'ha}})

% let'eylan
\hi 
\HT{let'eylan}{\B{let\cp{'ey}lan}} \I{[lEt\textcorner."PEj.lan]} \emph{adj., ofp.} friendly. 
\B{Ay\-ha\-pxì\-tu soaiä nge\-yä lu oe\-ru let\-'ey\-lan nì\-wotx.} Your family members are all friendly to me.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{tI'eylanga'}{\B{tì\-'ey\-la\-nga'}}; \HL{tI'eylan}{\B{tì\-'ey\-lan}}) 

% letam
\hi
\HT{letam}{\B{le\cp{tam}}} \I{[lE."tam]}  \emph{adj.} sufficient, enough.
\B{Ngay\-txoa, nì\-aw\-no\-mum ke lo\-lu oer nì\-ke\-ftxo mì sok\-a srr ay\-skxom le\-tam fte lì'\-fya\-ri aw\-nge\-yä tì\-kang\-kem si\-vi.} My apologies: As you know, in recent days I have not had sufficient opportunity to work on our language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
(\ding{43}~\HL{tam}{\B{tam}})

% letokx
\hi
\HT{letokx}{\B{le\cp{tokx}}} \I{[lE."tok']}  \emph{adj.} bodily, physical.   
\B{Ay\-u\-van le\-tokx 'o' lu nì\-txan.} Sports are great fun.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Tì\-sraw le\-tokx sì tì\-ngu\-sä\-'än pxìm tä\-pa\-re fì\-tsap.} Physical pain and mental suffering are often interrelated.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{tokx}{\B{tokx}})

% letrr
\hi 
\HT{letrr}{\B{le\cp{trr}}} \I{[lE."t\s{r}]}  \emph{adj.} daily. 
\B{Oe new pi\-vll\-txe nì\-Na'\-vi mì oe\-yä le\-trr\-a tì\-rey.} I want to speak Na'vi in my daily life.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-email/}{ln}
(\ding{43}~\HL{trr}{\B{trr}})

% letrrtrr
\hi
\HT{letrrtrr}{\B{le\cp{trr}trr}} \I{[lE."t\s{r}.t\s{r}]}  \emph{adj.} ordinary. 
\B{Ton\-ìri a\-la\-he, aw\-ngal yom wu\-tsot nì\-fya\-'o le\-trr\-trr; fì\-txon yom nì\-'eoio.} As for other nights, we eat dinner in an ordinary way; tonight we eat it ceremoniously.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
\B{Ta\-fral ho\-lawl ay\-oel ay\-nga\-fpi 'a\-'aw\-a tì\-päng\-kxo\-tsyìp\-it a tsun fko si\-var mì sì\-rey le\-trr\-trr.} That's why we prepared several little dialogues for you that one can use in ordinary life.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{mo letrrtrr}, living room.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{trr}{\B{trr}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{keltrrtrr}{\B{kel\-trr\-trr}}, unusual.

% letut
\hi 
\HT{letut}{\B{le\cp{tut}}} \I{[lE."tut\textcorner]} \emph{or} \I{[.tUt\textcorner]} (\textsc{rn}: \B{le\cp{tùt}} \I{[lE."tUt\textcorner]}) \emph{adj.} constant, continual. 
\B{Pe\-yä tì\-pu\-sll\-txel le\-tut oe\-ti srätx.} His constant talking annoys me.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}  
(\ding{43}~\HL{tut}{\B{tut}}, \HL{lukftang}{\B{luk\-ftang}})

% letwan
\hi
\HT{letwan}{\B{let\cp{wan}}} \I{[lEt\textcorner."wan]} \emph{adj.} dodgy, sneaky (\emph{of a person}).
(\ding{43}~\HL{tIwan}{\B{tì\-wan}})

% letsaktap
\hi 
\HT{letsaktap}{\B{le\cp{tsak}tap}} \I{[lE."\t{ts}ak\textcorner.tap\textcorner]} \emph{adj.} violent. 
(\ding{43}~\HL{tsaktap}{\B{tsak\-tap}})

% letsim
\hi
\HT{letsim}{\B{le\cp{tsim}}} \I{[lE."\t{ts}im]}  \emph{adj.} original, unique, not derived from another source.
\B{Swey\-lu txo ngal ke txi\-vìng sä\-fpìl\-it le\-tsim.} You shouldn't abandon your original idea.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{tsim}{\B{tsim}}, \HL{lesngA'i}{\B{le\-sngä\-'i}})

% letskxe
\hi 
\HT{letskxe}{\B{le\cp{tskxe}}} \I{[lE."\t{ts}k'E]} \emph{adj.} stony, of stone. 
\B{Fay\-frir le\-tskxe lor lu nì\-txan.} These stone layers are very beautiful.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
(\ding{43}~\HL{tskxe}{\B{tskxe}})

% letsranten
\hi
\HT{letsranten}{\B{le\cp{tsran}ten}} \I{[lE."\t{ts}Ran.tEn]}  \emph{adj.} important.
\B{Fì\-trr\-mì le\-tsran\-ten--Trr 'Rr\-ta\-yä--new oe pi\-vll\-txe ay\-nga\-ru san kal\-txì sìk.} On this important day--Earth Day--I want to say hello to you.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Mì\-ftxe\-le po\-ri lu oer le\-tsran\-ten\-a fmawn a new pi\-veng ngar.} By the way, I have some important news I want to tell you about him.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Le\-kin lu tì\-txur, lu tì\-tstew. Slä le\-tsran\-ten fra\-to lu tì\-mal.} Strength and courage are necessary. But most important of all is trustworthiness.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{tsranten}{\B{tsran\-ten}})

% letsunslu
\hi
\HT{letsunslu}{\B{le\cp{tsun}slu}} \I{[lE."\t{ts}un.slu]} \emph{or} \I{[."\t{ts}Un.]} (\textsc{rn}: \B{le\cp{tsùn}slu} \I{[lE."\t{ts}Un.slu]})  \emph{adj.} possible. 
(\ding{43}~\HL{tsunslu}{\B{tsun\-slu}}, \ding{214}~\HL{keltsun}{\B{kel\-tsun}})

% letswal
\hi 
\HT{letswal}{\B{le\cp{tswal}}} \I{[lE."\t{ts}wal]} \emph{adj., ofp.} powerful . 
(\ding{43}~\HL{tswal}{\B{tswal}}, \HL{tswalnga'}{\B{tswal\-nga'}})

% letxi
\hi
\HT{letxi}{\B{le\cp{txi}}} \I{[lE."t'i]}  \emph{adj.} hurried, frenzied.
(\ding{43}~\HL{txi}{\B{txi}}, \ding{214}~\HL{letxiluke}{\B{le\-txi\-lu\-ke}})

% letxiluke
\hi
\HT{letxiluke}{\B{le\cp{txi}luke}} \I{[lE."t'i.lu.kE]}  \emph{adj.} unhurried.
(\ding{43}~\HL{txi}{\B{txi}}, \HL{luke}{\B{lu\-ke}}; \ding{214}~\HL{letxi}{\B{le\-txi}})

% lew / si / säpi
\hi
\HT{lew}{\B{lew}} \I{[lEw]} \textsc{i}.~\emph{n.} cover, lid.
\B{Fu\-ri\-a txu\-la tsa\-lew\-ti lu srä sìl\-tsan to fngap.} For constructing that cover, woven cloth is better than metal.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b} \\
\textsc{ii}.~\HT{lew si}{\B{lew si}} \I{[lEw si]}  \emph{vin.} cover. 
\B{Ne\-ni lew si fì\-txa\-yor vay txam\-pay nì\-wotx.} Sand covers this expanse all the way to the ocean.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} \\
\textsc{iii}.~\HT{lew sApi}{\B{lew sä\cp{pi}}} \I{[lEw s\ae."p$\cdot$i]} \emph{vin.} cover oneself.
\B{Oe\-ru ke tsun li\-vam ke\-'u lor to Ey\-wa\-'e\-veng\-ä na'\-rìng a lew sä\-po\-li fa prr\-wll.} Nothing could be more beautiful to me than a Pandoran forest that has covered in moss.~\LINK{http://forum.learnnavi.org/language-updates/prrwll-moss-the-comparative-to-construct/}{ln}
\B{Kxawm mung\-wrr fì\-ki\-fkey a lew sä\-pì\-ye\-vi fa fpom sì lì'\-fya le\-Na'\-vi.} Except perhaps this world soon covering itself in peaceful well-being and the Na'vi language.~\LINK{http://forum.learnnavi.org/language-updates/prrwll-moss-the-comparative-to-construct/}{ln}

% lewn
\hi 
\HT{lewn}{\B{lewn}} \I{[lEwn]} \emph{vtr.} endure, stand, tolerate, bear. 
\B{Pe\-yä tì\-ru\-sol\-it ke tsun oe li\-vewn.} I can't stand her singing.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
\B{Hu\-fwa tì\-sraw lu txan, tsun ay\-oe tsat li\-vewn.} Although the pain is great, we can endure it.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \on{1}.~\HL{ketsuklewn}{\B{ke\-tsuk\-lewn}}, intolerable, unacceptable, unbearable.

% lewng
\hi 
\HT{lewng}{\B{lewng}} \I{[lEwN]} \emph{n.} shame. 
\B{Mu\-nge fna\-we'\-tul lewng\-it soaia\-ru sne\-yä.} A coward brings shame to his\slash her family.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
(\ding{214}~\HL{nrra}{\B{nrr\-a}}) \\
\textsc{Derivatives}: \on{1}.~\HL{lelewng}{\B{le\-lewng}}, shameful. \\
\on{2}.~\HL{lewnga'}{\B{lew\-nga'}}, shameful.

% lewnga'
\hi 
\HT{lewnga'}{\B{\cp{lew}nga'}} \I{["lEw.NaP]} \emph{adj., nfp.} shameful. 
\B{vo\-ìk a\-lew\-nga'}, shameful behavior.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{lewng}{\B{lewng}})

% ley
\hi
\HT{ley}{\B{ley}} \I{[lEj]} \emph{vin.} be of value, have some positive value, be worth s.t. 
\B{Fì\-'u ley.} This thing is valuable.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b} 
\B{Fì\-'u ke ley.} This thing is worthless.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\emph{a.~in comparisons}) \B{Oe\-ri tsaw ke ley fì\-'u\-to.} I don't value that over this.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
\B{Ma\-sat oeyä ley nì\-ftxan na pum nge\-yä.} My breastplate is as valuable as yours.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\emph{b.~with the dative}) \B{Sìl\-pey oe, fì\-po\-stì lil\-vey ay\-nga\-ru nì\-wotx!} I hope this post has been valuable to you.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\emph{c. proverb}) \B{Flä ke flä, ley sä\-fmi.} Whether you succeed or not, the attempt has value.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b} 
(\emph{d. proverb}) \B{'Uo\-ri hì\-pey, kxawm nga\-ru ke ley.} ``If you hesitate doing something, it might not be important to you,'' i.e. \emph{more meant\slash used as a motivation for s.o. hesitating, or even as a teasing to get s.o. into action}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivation}: \on{1}.~\HL{leytslam}{\B{ley\-tslam}}, appreciate.

% leykatem
\hi
\HT{leykatem}{\B{ley\cp{ka}tem}} \I{[lEj."k$\cdot$a.t$\cdot$Em]} \emph{vtr., inf.}\on{23} change s.t.
\B{Sra\-ne, lì\-'ul alu mì fra\-krr ley\-ka\-tem lì\-'ut a\-hay tsa\-fya.} Yes, the word \emph{in} always changes the following word like that.~\LINK{http://forum.learnnavi.org/language-updates/fmawno/msg293595/\#msg293595}{ln}
\B{Oe\-ri Unil\-tì\-ran\-tokx\-ìl tì\-rey\-ti ley\-ko\-la\-te\-i\-em nì\-txan.} \emph{Avatar} has greatly changed my life for the better.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/msg350756/\#msg350756}{ln}
(\ding{43}~\HL{latem}{\B{la\-tem}})

% leykek
\hi 
\HT{leykek}{\B{ley\cp{kek}}} \I{[lEj."kEk\textcorner]} \emph{vtr.} enforce. 
\B{Pol ho\-ren\-it a\-yll ley\-kek.} He enforces the laws.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} 
(\ding{43}~\HL{lek}{\B{lek}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{horenleykekyu}{\B{ho\-ren\-ley\-kek\-yu}}, law enforcer, police officer.

% leykekyu
\hi 
\HT{leykekyu}{\B{ley\cp{kek}yu}} \I{[lEj."kEk\textcorner.ju]} \emph{n.} law enforcer, (\emph{on Earth}) police officer. (\emph{shortened form of} \ding{43}~\HL{horenleykekyu}{\B{ho\-ren\-ley\-kek\-yu}}).
(\ding{43}~\HL{leykek}{\B{ley\-kek}}) 

% leym
\hi 
\HT{leym}{\B{leym}} \I{[lEjm]}  \emph{vin.} call out, cry out, exclaim.
\B{Zun Ey\-wa\-'e\-veng\-it oel ti\-vok, zel leym san Srung si ay\-oe\-ru, ma Ey\-wa sìk!} If I were on Pandora, I would call out, ``Help us, Eywa!''~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIleym}{\B{tì\-leym}}, call.
\on{2}.~\HL{leymfe'}{\B{leym\-fe'}}, complain.
\on{3}.~\HL{leymkem}{\B{leym\-kem}}, protest.

% leymfe'
\hi 
\HT{leymfe'}{\B{leym\cp{fe'}}} \I{[l$\cdot$Ejm."fEP]} \emph{vin., inf.}\on{11} complain. 
\B{Fo le\-reym\-fe' tsnì syu\-ve lu wew.} They're complaining that the food is cold.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{43}~\HL{leym}{\B{leym}}, \HL{fe'}{\B{fe'}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIleymfe'}{\B{tì\-leym\-fe'}}, complaining. 
\on{2}.~\HL{sAleymfe'}{\B{sä\-leym\-fe'}}, complaint.

% leymkem
\hi 
\HT{leymkem}{\B{leym\cp{kem}}} \I{[l$\cdot$Ejm."kEm]}, \emph{coll. pronounced as} \I{[lejN."kEm]} \emph{vin., inf.}\on{11} protest (\B{tsnì}, that). 
\B{Oe leym\-kem! Fì\-tsam\-ìl Na'\-vit tì\-sraw sey\-ka\-yi nì\-'aw ul\-te ku\-tut ke lä\-txayn.} I protest! This war will only harm the People and not defeat the enemy.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}
\B{Tsay\-hem\-ìri\slash Tsay\-hem\-te\-ri po lo\-leym\-kem.} She protested those actions.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}
\B{Lo\-leym\-kem po tsnì fwa Ak\-wey slu olo'\-eyk\-tan lu kem\-wi\-ä.} He protested that it was unfair for Ak\-wey to become clan leader.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
\B{Fpxa\-mo\-a fì\-kem\-ìl a\-fpxa\-mo ay\-oe\-ti tsngey\-ko\-law\-vìk, tsa\-krr ley\-ko\-leym\-kem.} This terrible, terrible action has made us cry, then protest.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{leym}{\B{leym}}, \HL{kemwiA}{\B{kem\-wi\-ä}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIleymkem}{\B{tì\-leym\-kem}}, protesting, protest. 
\on{2}.~\HL{sAleymkem}{\B{sä\-leym\-kem}}, protest, instance of protesting.

% leyn
\hi 
\HT{leyn}{\B{leyn}} \I{[lEjn]}  \emph{vtr.} repeat, do again.
\B{Ru\-txe li\-veyn.} Please repeat (it).~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Rä\-'ä li\-veyn!} Don't do that again!~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIleyn}{\B{nì\-leyn}}, repeatedly.

% leyr
\hi 
\HT{leyr}{\B{leyr}} \I{[lEjR]} \textsc{i}. \emph{adj.} frozen. 
\B{Ke tsun ioang ri\-vun syu\-vet mì hll\-te a\-leyr.} Animals can't find food in the frozen ground.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b} \\
\textsc{ii}. \B{leyr si} \I{[lEjR si]} \emph{vin.} freeze (s.t.).
\B{Txo awnga fì\-tsngan\-ur leyr si\-vi, tsun tsat yi\-vom kin\-trr\-ay.} If we freeze this meat, we can eat it next week.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b} \\
\textsc{iii}. \B{slu leyr} \I{[slu lEjR]} \emph{vin.} s.t. freezes.
\B{Mì zì\-sì\-krr a\-txa\-wew slu ay\-ora leyr.} In the very cold season, the lakes freeze.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}

% leytslam
\hi 
\HT{leytslam}{\B{\cp{ley}tslam}} \I{["lEj.\t{ts}l$\cdot$am]} \emph{vtr., inf.}\on{22} appreciate. 
\B{Nge\-yä fay\-lì\-'ut a\-tì\-tstun\-wi\-nga' oel ley\-tslam, ma 'ey\-lan.} I appreciate your kind words, friend.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{ley}{\B{ley}}, \HL{tslam}{\B{tslam}})

% leweopx
\hi 
\HT{leweopx}{\B{lewe\cp{opx}}} \I{[lE.wE."{}op']} \emph{adj.} wave\hyp{}like. 
\B{Tsay\-re\-nur le\-we\-opx a sìn ne\-ni tìng na\-ri.} Look at those wave\hyp{}like patterns in the sand.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{43}~\HL{weopx}{\B{we\-opx}})

% lewäte
\hi
\HT{lewAte}{\B{lewä\cp{te}}} \I{[lE.w\ae."tE]}  \emph{adj.} disagreeable, argumentative (\emph{of an agent}).
(\ding{43}~\HL{wAte}{\B{wä\-te}})

% leyapay
\hi
\HT{leyapay}{\B{le\cp{ya}pay}} \I{[lE."ja.paj]}  \emph{adj.} foggy, misty. 
(\ding{43}~\HL{yapay}{\B{ya\-pay}})

% leyayl
\hi 
\HT{leyayl}{\B{le\cp{yayl}}} \I{[lE."jajl]} \emph{adj.} nonsensical. 
(\ding{43}~\HL{yayl}{\B{yayl}})

% leye'krr
\hi
\HT{leye'krr}{\B{le\cp{ye'}krr}} \I{[lE."jEP.k\s{r}]}  \emph{adj.} early. 
\B{tì\-pä\-hem le\-ye'\-krr}, an early arrival.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{ye'krr}{\B{ye'\-krr}}, \ding{214}~\HL{lehawngkrr}{\B{le\-hawng\-krr}})

% leyewla 
\hi
\HT{leyewla}{\B{le\cp{yew}la}} \I{[lE."jEw.la]}  \emph{adj.} disappointing. 
\B{Pe\-yä sä\-ngä\-'än\-ìri lu sä\-fe\-'ul le\-yew\-la nì\-txan.} The worsening of his depression is very disappointing.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Kea kem le\-yew\-la rä\-'ä si, ru\-txe.} Please don't let me down.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{yewla}{\B{yew\-la}})

% lezeswa
\hi
\HT{lezeswa}{\B{le\cp{ze}swa}} \I{[lE."zE.swa]}  \emph{adj.} grassy.
(\ding{43}~\HL{zeswa}{\B{zeswa}}) 

% li
\hi
\HT{li}{\B{li}} \I{[li]}  \emph{adv.} (\emph{a.}) already.   
\B{Tì\-kang\-kem li ha\-sey lu.} The work is already finished.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\emph{b. negated}) not yet.
\B{Vì\-ming\-kap oe\-ti fu\-la poe ke li ke pol\-txe san oe za\-sya\-'u.} It just occurred to me that she hasn't yet said she's coming.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Fo li po\-lä\-hem srak?--Ke li.} Have they already arrived?--Not yet.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\emph{c. in imperatives to convey strong urgency})
\B{Ngal mi fì\-tseng\-it te\-rok srak? Li \cp{kä}!} You're still here? Get going!~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\emph{d. set phrase})
\B{\cp{Li} ko.} Well, get to it, then.\slash Let's get on it.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
(\emph{e. hesitant or weak `yes'}) 
\B{ Nga mll\-te srak?--Li, slä\ldots} Do you agree?--Well, yes, I guess so, but\ldots{}~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
(\emph{f. with} \ding{43}~\HL{sre}{\B{sre}}) by; before or up to, but not after. 
\B{Kem si li trr\-ay\-sre.} Do it by tomorrow.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} (\emph{also} \ding{43}~\HL{lisre}{\B{li\-sre}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIli}{\B{nì\-li}}, in advance. \\
\on{2}.~\HL{lisre}{\B{li\-sre}}, by; before but not after.

% lie / si
\hi
\HT{lie}{\B{\cp{li}e}} \I{["li.E]} \textsc{i}. \emph{n.} experience. 
\B{Kop oe\-ru lo\-lä\-ngu lie a ha\-pxì\-tu soaiä ter\-kup.} Sadly, I too have had the experience of a family member dying.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Lu oe\-ru lie a yom tey\-lut.} I once ate \emph{teylu}.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \\
\textsc{ii}. \B{\cp{li}e si} \I{["li.E si]} \emph{vin.} experience (\emph{an event, a feeling, a person}). 
\B{Tu\-te a ke\-ftxo fra\-to lu tsa\-po a tì\-yawn\-ur lie ke so\-li kaw\-krr.} The saddest person of all is the one who has never experienced love.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} \\ 
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{'awlie}{\B{'aw\-lie}}, once (\emph{experiential}).

% lik
\hi 
\HT{lik}{\B{lik}} \I{[lik\textcorner]} \emph{n.} point, spot, particular place or position in some area. 
\B{Tsko\-ti fyep fì\-lik\-ro.} Grasp the bow at this spot.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
\B{Ro sa\-lik a mì sä\-ftxu\-lì\-'u a\-tì\-'i\-lu\-ke pe\-yä, oe ho\-lum.} At that point in his endless speech, I left.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{krrlik}{\B{krr\-lik}}, point in time.

% limang
\hi
\HT{limang}{\B{li\cp{mang}}} \I{[li."maN]}  \emph{n.} moot, celebratory gathering or event. 
\B{A\-ha'\-ri wrr\-ko\-lä fte li\-mang si\-vu\-nu por.} Aha'ri is out enjoying the moot.~\LINK{https://forum.learnnavi.org/language-updates/limangmoot-from-afop-dialogue/}{ln\slash fop}

% lini
\hi
\HT{lini}{\B{\cp{li}ni}} \I{["li.ni]}  \emph{countable n.} young (\emph{of an animal, bird, fish, insect}). 
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nga'lini}{\B{nga'\-li\-ni}}, be pregnant (for animals).

% lipx
\hi 
\HT{lipx}{\B{lipx}} \I{[lip']} \emph{vin.} drip. 
\B{Tom\-pa ze\-rup ul\-te pay lipx kxam\-lä fäp\-yo.} It's raining and water is dripping through the roof.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
(\emph{a. idiom}) \B{Pay\-ìl a lipx tskxe\-ti ripx.} Dripping water pierces a stone. [\emph{persistent effort can accomplish unexpected and amazing things}]~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}  \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{sAlipx}{\B{sä\-lipx}}, drip (of liquid).

% lisre
\hi
\HT{lisre}{\B{\cp{li}sre}} \I{["li.sRE]} \emph{adp.}\texttt{+} by, before, up to but not after.
\B{Kem si li\-sre srr\-ay.} Do it by tomorrow.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\ding{43}~\HL{li}{\B{li}}, \HL{sre}{\B{sre}})

% liswa / si
\hi 
\HT{liswa}{\B{li\cp{swa}}} \I{[li."{}swa]} \textsc{i}. \emph{n.} nourishment. \\
\textsc{ii}. \B{li\cp{swa} si} \I{[li."{}swa si]} \emph{vin.} nourish, provide nourishment.
\B{Fì\-'e\-wll li\-swa si Na'\-vi\-ru.} This plant provides nourishment to the People.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 

% litx
\hi
\HT{litx}{\B{litx}} \I{[lit']}  \emph{adj.} sharp (as a blade).
\B{El\-tu si! Tsa\-tstal a\-fwem lu litx nì\-txan.} Pay attention! That blunt knife is very sharp.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{214}~\HL{tete}{\B{te\-te}})

% litsi
\hi 
\HT{litsi}{\B{\cp{li}tsi}} \I{["li.\t{ts}i]} \emph{adj., ofp.} thin, lean, lithe. 
(\ding{214}~\HL{ompu}{\B{om\-pu}}; \ding{43}~\HL{flI}{\B{flì}}) 

% lì'fya
\hi
\HT{lI'fya}{\B{\cp{lì'}fya}} \I{["lIP.fja]}  \emph{n.} 
(\emph{a.}) language. 
\B{Ay\-lì'\-fya yaw\-ne le\-ru oer ta\-krr\-a 'e\-veng la\-mu.} I've loved languages since I was a child.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Fol lì'\-fya\-ti aw\-nge\-yä sar a fya\-'o lu ko\-sman.} It's wonderful the way they use our language.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\emph{b.}) the language spoken on Pandora.
\B{Lì'\-fya\-ri pol tok pe\-yìt?} \emph{or} \B{Lì'\-fya\-ri po kll\-kxem sìn pe\-yì?}\slash \B{Lì'\-fya\-ri po sìn pe\-yì?}\slash \B{Lì'\-fya\-ri po pe\-yì?} How good is her Na'vi?~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\emph{c.}) \B{lì'\-fya le\-Na'\-vi}, the Na'vi language.
\B{Nge\-yä lì'\-fya le\-Na'\-vi txan\-tsan lu nì\-ngay.} Your Na'vi language is truly excellent.~\LINK{http://forum.learnnavi.org/language-updates/score!-\%281\%29-the-verb-to-use-\%282\%29-more-detail-on-teya-si/}{ln}
\B{Sra\-ke fnan ngal lì'\-fya\-ti le\-Na'\-vi?} Are you good at Na'vi?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Unil\-tì\-ran\-tokx\-ìri lì'\-fya\-yä le\-Na'\-vi wa\-wet fol tslam.} They understand the importance of the Na'vi language for \emph{Avatar}.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
\B{Lì'\-fya\-ri le\-Na'\-vi nga tsan\-'e\-re\-i\-ul fra\-trr.} I'm delighted that your Na'vi is improving every day.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lI'fyavi}{\B{lì'\-fya\-vi}}, expression.   
\on{2}.~\HL{lI'fyaolo'}{\B{lì'\-fya\-o\-lo'}}, language group.
\on{3}.~\HL{lI'fyeng}{\B{lì'\-fyeng}}, honorific language.
\on{4}.~\HL{lI'fyafnel}{\B{lì'\-fya\-fnel}}, dialect, variety of a language.
\on{5}.~\HL{rAptulI'fya}{\B{räp\-tu\-lì'\-fya}}, coarse, vulgar language.

% lì'fyafnel
\hi 
\HT{lI'fyafnel}{\B{\cp{lì'}fyafnel}} \I{["lIP.fja.fnEl]} \emph{n.} dialect, variety of a language. 
\B{Lu mì `Fya\-'o Pay\-ä' mip\-a lì'\-fya\-fnel lì'\-fya\-yä le\-Na'\-vi.} There is a new dialect of the Na'vi language in \emph{Avatar: The Way of Water}.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 
(\ding{43}~\HL{lI'fya}{\B{lì'fya}}, \HL{fnel}{\B{fnel}})

% lì'fyaolo'
\hi
\HT{lI'fyaolo'}{\B{\cp{lì'}fyaolo'}} \I{["lIP.fja.o.loP]}  \emph{n.} language group, \emph{Sprachbund}.
\B{Nge\-yä tì\-kang\-kem a\-fyo\-le meuia leiu lì'\-fya\-o\-lo'\-ru, ma tsmu\-kan.} Your sublime work honors the language group, brother.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Fu\-ri\-a fì\-lì'\-fya\-vit fkol ta 'Ìng\-lì\-sì mo\-lu\-nge, sìl\-pey oe tsnì ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ke stì\-ye\-vi.} That this expression is brought from English, will hopefully not anger the members of the language group.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\ding{43}~\HL{lI'fya}{\B{lì'\-fya}}, \HL{olo'}{\B{o\-lo'}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lI'fyaolo'tu}{\B{lì'\-fya\-o\-lo'\-tu}}, member of the language community.

% lì'fyaolo'tu
\hi
\HT{lI'fyaolo'}{\B{\cp{lì'}fyaolo'tu}} \I{["lIP.fja.o.loP.tu]}  \emph{n.} member of the language group\slash \emph{Sprachbund}.~\LINK{https://forum.learnnavi.org/language-updates/a-collection-of-questions-answered/msg679687/\#msg679687}{ln}
(\ding{43}~\HL{lI'fyaolo'}{\B{lì'\-fya\-olo'}}, \HL{-tu}{\B{-tu}})

% lì'fyavi
\hi
\HT{lI'fyavi}{\B{\cp{lì'}fyavi}} \I{["lIP.fja.vi]}  \emph{n.} expression, bit of language.
\B{Lì'\-fya\-vi le\-sar nì\-txan nang!} The expression is very useful.~\LINK{http://forum.learnnavi.org/language-updates/just-another-few-words/}{ln}
\B{\ldots slä nì\-sìl\-pey pum a nga' ay\-lì'\-fya\-vit le\-sar.} \ldots but hopefully one that contains useful expressions.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Tsa\-lì'\-fya\-vi\-ri fko ze\-ne si\-var lì\-'ut alu fì\-'u me\-lo.} One must use the word \emph{fì'u} twice in that expression.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-677}{b}
(\ding{43}~\HL{lI'fya}{\B{lì'\-fya}}, \HL{tIyewn}{\B{tì\-yewn}})

% lì'fyeng / si
\hi 
\HT{lI'fyeng}{\B{lì'\cp{fyeng}}} \I{[lIP."fjEN]} \textsc{i}. \emph{n.} honorific language. 
\B{Lì'\-fyeng kel\-kin.} No need to speak so formally.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b}  
\B{Lì'\-fyeng kel\-kin ko.} Let's not speak so formally with each other.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b} \\
\textsc{ii}. \B{lì'\cp{fyeng} si} \I{[lIP."fjEN si]} \emph{vin.} use honorific language.
\B{Lì'fyeng rä'ä si.} Don't use honorific language.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b} 
\B{Fwa lì'fyeng si lu kelkin.} It's not necessary to use honorific language.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b}
(\ding{43}~\HL{lI'fya}{\B{lì'\-fya}}, \HL{leheng}{\B{le\-heng}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{fyengkekin}{\B{fyeng\-ke\-kin}}, no need to use honorific language.

% lì'kong
\hi 
\HT{lI'kong}{\B{\cp{lì'}kong}} \I{["lIP.koN]} \emph{n.} syllable. 
\B{Lu tsa\-lì\-'ur alu fì\-haw\-re'\-ti tsìng\-a lì'\-kong.} The word \emph{fì\-haw\-re'\-ti} has four syllables.~\LINK{http://naviteri.org/2017/10/more-language-for-talking-about-language/}{b} 
(\ding{43}~\HL{lI'u}{\B{lì\-'u}}, \HL{'ekong}{\B{'e\-kong}})

% Lì'ona
\hi
\HT{lI'ona}{\B{Lì\cp{'o}na}} \I{[lI."Po.na]}  \emph{n. name of a clan}. l'Ona clan.
\B{Ke za\-syup lì\-'O\-na ne kxu\-tu a mì\-fa fu a wrr\-pa.} The l'Ona will not perish to the enemy within or the enemy without.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{g\slash ln}

% lì'u
\hi
\HT{lI'u}{\B{\cp{lì}'u}} \I{["lI.Pu]}  \emph{n.} word.
\B{Ma Ey\-tu\-kan, lu oe\-ru ay\-lì\-'u fra\-por.} Eytukan, I have something to say, to everyone.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{a\slash wiki}
\B{Nge\-yä tsay\-lì\-'ul tì\-fkey\-tok\-it fe\-'ey\-ko\-lul nì\-'aw.} Those words of yours have only made the situation worse.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Sra\-ne, lì\-'ul alu mì fra\-krr ley\-ka\-tem lì\-'ut a\-hay tsa\-fya.} Yes, the word \emph{in} always changes the following word like that.~\LINK{http://forum.learnnavi.org/language-updates/fmawno/msg293595/\#msg293595}{ln}
\B{Ro sngä\-'i\-tseng tsa\-lì\-'uä alu 'ey\-lan lu tì\-ftang.} At the beginning of the word \emph{'eylan} there's a glottal stop.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Nge\-yä fay\-lì\-'u\-ri sei\-yi oe i\-ra\-yo.} I thank you for these words of yours.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-995}{b} \\ 
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lI'fya}{\B{lì'\-fya}}, language. 
\on{2}.~\HL{lI'upam}{\B{lì\-'u\-pam}}, pronunciation. 
\on{3}.~\HL{lI'upuk}{\B{lì\-'u\-puk}}, dictionary. 
\on{4}.~\HL{lI'uvan}{\B{lì\-'u\-van}}, pun, word play. 
\on{5}.~\HL{fayluta}{\B{fay\-lu\-ta}}, that (\emph{reported speech}). 
\on{6}.~\HL{ftxulI'u}{\B{ftxu\-lì\-'u}}, orate, give a speech. 
\on{7}.~\HL{kemlI'u}{\B{kem\-lì\-'u}}, verb. 
\on{8}.~\HL{tstxolI'u}{\B{tstxo\-lì\-'u}}, noun.
\on{9}.~\HL{lI'upe}{\B{lì\-'u\-pe\slash pe\-lì\-'u}}, what, which (word, utterance)?
\on{10}.~\HL{lI'ukIng}{\B{lì\-'u\-kìng}}, sentence.
\on{11}.~\HL{lI'kong}{\B{lì'\-kong}}, syllable.
\on{12}.~\HL{lI'uvi}{\B{lì\-'u\-vi}}, affix.
\on{13}.~\HL{syonlI'u}{\B{syon\-lì\-'u}}, adjective.
\on{14}.~\HL{fyalI'u}{\B{fya\-lì\-'u}}, adverb.
\on{15}.~\HL{starlI'u}{\B{star\-lì\-'u}}, adposition.
\on{16}.~\HL{tilI'u}{\B{ti\-lì\-'u}}, conjunction.
\on{17}.~\HL{rAptulI'u}{\B{räp\-tu\-lì\-'u}}, coarse or swear word.
\on{18}.~\HL{rawnlI'u}{\B{rawn\-lì\-'u}}, pronoun.
\on{19}.~\HL{tengralI'u}{\B{teng\-ra\-lì\-'u}}, synonym.
\on{20}.~\HL{wAralI'u}{\B{wä\-ra\-lì\-'u}}, antonym.

% lì'ukìng
\hi 
\HT{lI'ukIng}{\B{\cp{lì}'ukìng}} \I{["lI.Pu.kIN]} \emph{n.} sentence (of words), \emph{lit.}: word thread. 
(\ding{43}~\HL{lI'u}{\B{lì\-'u}}, \HL{kIng}{\B{kìng}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lI'ukIngvi}{\B{lì\-'u\-kìng\-vi}}, phrase.

% lì'ukìngvi
\hi 
\HT{lI'ukIngvi}{\B{\cp{lì}'ukìngvi}} \I{["lI.Pu.kIN.vi]} \emph{n.} phrase. 
(\ding{43}~\HL{lI'ukIng}{\B{lì\-'u\-kìng}})

% lì'upam
\hi
\HT{lI'upam}{\B{\cp{lì}'upam}} \I{["lI.Pu.pam]}  \emph{n.} pronunciation. 
\B{Poe tso\-lun nì\-Na'\-vi pi\-vll\-txe, hu\-fwa lì\-'u\-pam hek nì\-'it.} She can speak Na'vi although her pronunciation is a bit strange.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Slä lì\-'u\-pam\-ìri pll\-txe fko san tì\-o\-mu\-mì sìk.} But as for pronunciation one says \emph{tìomumì}.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b}
(\ding{43}~\HL{lI'u}{\B{lì'u}}, \HL{pam}{\B{pam}})

% lì'upe
\hi
\HT{lI'upe}{\B{\cp{lì}'upe}} \I{["lI.Pu.pE]} \emph{or} \B{pe\cp{lì}'u} \I{[pE."lI.Pu]} \emph{inter.} what, which (word, utterance)?
(\ding{43}~\HL{lI'u}{\B{lì\-'u}})

% lì'upuk
\hi
\HT{lI'upuk}{\B{\cp{lì}'upuk}} \I{["lI.Pu.puk\textcorner]} \emph{or} \I{[.pUk\textcorner]}  \emph{adj.} dictionary.
(\ding{43}~\HL{lI'u}{\B{lì'u}}, \HL{puk}{\B{puk}}) 

% lì'uvan
\hi
\HT{lI'uvan}{\B{\cp{lì}'uvan}} \I{["lI.Pu.van]}  \emph{n.} pun, word\hyp{}play.   
\B{Ay\-lì\-'u\-van a\-swey lu 'i\-pu, lu sìl\-ron\-sem.} The best puns are both funny and clever.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{lI'u}{\B{lì\-'u}}, \HL{uvan}{\B{u\-van}})

% lì'uvi
\hi 
\HT{lI'uvi}{\B{\cp{lì}'uvi}} \I{["lI.Pu.vi]} \emph{n.} affix. 
\B{Lu tsa\-lì\-'ur alu fì\-haw\-re'\-ti me\-lì\-'u\-vi alu 'aw\-a eo\-lì\-'u\-vi sì 'aw\-a uo\-lì\-'u\-vi.} The word \emph{fì\-haw\-re'\-ti} has two affixes--one prefix and one suffix.~\LINK{http://naviteri.org/2017/10/more-language-for-talking-about-language/}{b}
(\ding{43}~\HL{lI'u}{\B{lì\-'u}}) \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{eolI'uvi}{\B{eo\-lì\-'u\-vi}}, prefix. \\
\on{2}.~\HL{uolI'uvi}{\B{uo\-lì\-'u\-vi}}, suffix.

% lìktap
\hi
\HT{lIktap}{\B{\cp{lìk}tap}} \I{["lIk\textcorner.tap\textcorner]}  \emph{adj.} crooked.
\B{Ke tsun fko fì\-swi\-zaw\-ti si\-var--lu lìk\-tap.} This arrow can't be used--it's crooked.~\LINK{http://naviteri.org/2019/12/tengkrr-zisit-leratem-as-the-year-changes/}{b}
(\ding{214}~\HL{yey}{\B{yey}})

% lìlì
\hi 
\HT{lIlI}{\B{\cp{lì}lì}} \I{["lI.lI]} \emph{n.} breast. 

% lìm
\hi
\HT{lIm}{\B{lìm}} \I{[lIm]}  \emph{vin.} far, be far.
\B{'Ì\-'awn nì\-fya\-'o a lìm.} Stay away [lit.: remain in a way that is far].~\LINK{http://forum.learnnavi.org/language-updates/near-distant-and-irregular-adverbs/}{ln}
(\ding{43}~\HL{alIm}{\B{alìm}}; \ding{214}~\HL{sim}{\B{sim}})

% lìmsimpe 
\hi 
\HT{lImsimpe}{\B{\cp{lìm}simpe}} \I{["lIm.sim.pE]} \emph{inter.} how near? how far? what distance? (\emph{less common than} \ding{43}~\HL{la'ape}{\B{la'ape}}). 
(\ding{43}~\HL{pelImsim}{\B{pelìmsim}})

% lìng
\hi
\HT{lIng}{\B{lìng}} \I{[lIN]}  \emph{vin.} float in the air or water, hover. 
\ding{43}~\HL{ayram alusIng}{\B{Ay\-ram A\-lu\-sìng}}, the Floating Mountains. \\
\textsc{Derivatives}: \on{1}.~\HL{lIngtskxe}{\B{lìng\-tskxe}}, unobtanium.

% lìngtskxe
\hi 
\HT{lIngtskxe}{\B{\cp{lìng}tskxe}} \I{["lIN.\t{ts}k'E]} \emph{n.} unobtanium. 
(\ding{43}~\HL{lIng}{\B{lìng}}, \HL{tskxe}{\B{tskxe}})

% lo'a
\hi 
\HT{lo'a}{\B{\cp{lo}'a}} \I{["lo.Pa]} \emph{n.} totem; amulet. \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lo'akur}{\B{lo'akur}}, Toruk Makto Amulet.

% lo'ak
\hi 
\HT{lo'ak}{\B{\cp{Lo}'ak}} \I{["lo.Pak\textcorner]} \emph{n. male name}. 

% lo'akur
\hi 
\HT{lo'akur}{\B{lo'a\cp{kur}}} \I{[lo.Pa."kur]} \emph{or} \I{[."kUR]} (\textsc{rn}: \B{loa\cp{kùr}} \I{[lo.a."kUR]}) \emph{n.} Toruk Makto Amulet.
(\ding{43}~\HL{lo'a}{\B{lo'a}}, \HL{kur}{\B{kur}})

% Loak
\hi
\B{\cp{Lo}ak} \I{["lo.ak\textcorner]}  \emph{n., male name}. 
\B{Lo\-ak kal\-txì so\-li oer.} Loak greeted me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-294}{b}
\B{Lo\-ak\-ìl pä\-nu\-to\-lìng fu\-ta kar oe\-ru fya\-'ot a 'ìp fko nem\-fa ewll.} Loak promised he'd teach me how to vanish into the bushes.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tse\-nu Lo\-ak\-ur ke ha'.} Tsenu is a bad match for Loak.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Ngal mo\-lay' pxaw\-pxun\-it Lo\-ak\-ä a krr pe\-kxin\-um?} When you tried on Loak's armband, how tight was it?~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}

% loho
\hi 
\HT{loho}{\B{\cp{lo}ho}} \I{["lo.ho]} \emph{vin.} be surprising. 
\B{Tä\-ftxu\-tswo Ri\-ni\-yä lo\-ho oer nì\-txan.} Rini's ability to weave surprises me a lot.
\B{Fo tsìk sro\-ler a fì\-'u lo\-lo\-ho po\-an\-ur.\slash Lo\-lo\-ho po\-an\-ur fwa fo tsìk sro\-ler.} It surprised him that they suddenly appeared.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tIloho}{\B{tì\-lo\-ho}}, surprise. \\
\on{2}.~\HL{nIloho}{\B{nì\-lo\-ho}}, surprisingly. 

% loi
\hi
\HT{loi}{\B{\cp{lo}i}} \I{["lo.i]}  \emph{n.} egg.
\B{Ro\-lun ay\-oel tsrul\-mì hì\-'i\-a pxe\-lo\-it a\-teyr.} We found three little white eggs in the nest.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}

% lok
\hi
\HT{lok1}{\B{lok}}\textsuperscript{\on{1}} \I{[lok\textcorner]}  \emph{adp.} close to. 
\B{(fwa sle\-le) mì hil\-van a lok Kel\-ut\-ral}, to swim in a river near Hometree.~\LINK{http://naviteri.org/2010/07/thoughts-on-ambiguity/}{b}

% lok
\hi
\HT{lok2}{\B{lok}}\textsuperscript{\on{2}} \I{[lok\textcorner]}  \emph{vtr.} get close to, approach.   
\B{Na\-ri so\-li ay\-oe fte\-ke nì\-hawng li\-vok.} We were careful not to get too close.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
\B{Fra\-po ne mì\-fa! Le\-rok yrr\-ap apxa!} Everybody inside! A big storm is approaching!~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Sra\-ke pol la\-yok txew\-ti na'\-rìng\-ä?} Will he approach the edge of the forest?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Ke tsun aw\-nga pi\-vey nul\-krr -- txew lok.} We can't wait any longer -- time is almost up.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}

% lom
\hi
\HT{lom}{\B{lom}} \I{[lom]}  \emph{adj.} missing, missed (\emph{as an absent person who is longed for; covers only s.t.~one once had but no longer does; a sense of emotional loss; not s.t.~one lacks but never had in the first place}).
\B{Nga lom lu oer.} I miss you.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Ay\-sre' lom lu tsa\-ko\-ak\-tan\-ur.} The old man misses his teeth.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lomtu}{\B{lom\-tu}}, missed person.

% lomtu
\hi
\HT{lomtu}{\B{\cp{lom}tu}} \I{["lom.tu]}  \emph{n.} (\emph{reserved for special circumstances}) missed person.
(\emph{a. idiom}) \B{Teng\-krr ftxo\-zä se\-rei\-yi aw\-nga, ke tswi\-va' ay\-lom\-tu\-ti ko!} A toast: To absent friends. [lit.: While we are celebrating, let us not forget those who we wish could also be here (but can't)]~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{lom}{\B{lom}}) 

% lonu
\hi
\HT{lonu}{\B{lo\cp{nu}}} \I{[lo."nu]}  \emph{vtr.} release, let go. 
\B{Pot lo\-nu.} Release him.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki}
\B{Nì\-sngä\-'i fmawn\-it fo nar\-mew wi\-van, slä nì\-'i'\-a fra\-por lo\-lo\-nu.} They originally wanted to hide the news, but in the end they revealed it everyone.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{a. proverb}) \B{Fwa kan ke tam; ze\-ne swi\-zaw\-it li\-vo\-nu.} Intent is not enough; it's action that counts. [lit.: To aim is not enough; one must release the arrow.]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIklonu}{\B{nìk\-lo\-nu}}, firmly, steadfastly, faithfully. \\
\on{2}.~\HL{stxenu}{\B{stxe\-nu}}, offer. \\
\on{3}.~\HL{lonusye}{\B{lo\-nu\-sye}}, exhale, blow.

% lonusye
\hi 
\HT{lonusye}{\B{lo\cp{nu}sye}} \I{[l$\cdot$o."n$\cdot$u.sjE]} \emph{vin., inf.}\on{12} exhale; blow (\ding{43}~\HL{ne}{\B{ne}}, on). 
\B{Txo syu\-ve som li\-vu nì\-hawng, lo\-nu\-sye tsa\-ne.} If the food is too hot, blow on it.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
\B{Lo\-nu\-sye tong txep\-tsyìp\-it.} Blow out the flame.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{43}~\HL{lonu}{\B{lo\-nu}}, \HL{syeha}{\B{sye\-ha}}; \ding{214}~\HL{mungsye}{\B{mung\-sye}}) 

% lopx
\hi
\HT{lopx}{\B{lopx}} \I{[lop']}  \emph{vin.} panic.
\B{Pa\-lu\-kan\-it tso\-le\-'a, ye\-rik lopx hi\-fwo kxam\-lä ze\-swa.} Spotting a thanator, the hexapede panicked and escaped through the grass.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}

% lor
\hi
\HT{lor}{\B{lor}} \I{[loR]}  \emph{adj.} beautiful, pleasant to the senses.
\B{New oe ti\-vìng ay\-ngar lì\-'ut a tì\-'e\-fu\-mì oe\-yä lu lor fra\-to mì lì'\-fya le\-Na'\-vi.} I want to present to you the word that, in my opinion, is the most beautiful in the Na'vi language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Fay\-lì\-'u a\-lor oe\-ru te\-ya si nì\-txan, ul\-te ke tswa\-ya' oel sat kaw\-krr.} These beautiful words fill me with joy and I will never forget them.~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln} 
\B{Lor\-a fì\-tì\-kang\-kem me\-nge\-yä meuia lu\-yu aw\-nga\-ru nì\-wotx.} This beautiful work of theirs honors us very much.~\LINK{http://naviteri.org/2010/12/beautiful-christmas-carols-sung-in-navi/}{b}
\B{Lu po lor\-a tu\-té a\-lor.} She's an extremely beautiful woman. \emph{or} \B{Tsa\-tu\-té lu lor\-a pum a\-lor.} That woman is an extremely beautiful one.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{214}~\HL{vA'}{\B{vä'}}) \\  
\textsc{Derivatives}: \on{1}.~\HL{tIlor}{\B{tì\-lor}}, beauty.   
\on{2}.~\HL{ftxIlor}{\B{ftxì\-lor}}, delicious. 
\on{3}.~\HL{onlor}{\B{on\-lor}}, good-smelling. 
\on{4}.~\HL{miklor}{\B{mik\-lor}}, pleasant sounding. 
\on{5}.~\HL{loran}{\B{lo\-ran}}, elegance. 
\on{6}.~\HL{loreyu}{\B{lo\-re\-yu}}, helicordian.
\on{7}.~\HL{lortsyal}{\B{lor\-tsyal}}, shimmyfly.
\on{8}.~\HL{kinglor}{\B{king\-lor}}, kinglor. 
\on{9}.~\HL{zeklor}{\B{zek\-lor}}, pleasant to the touch.
\on{10}.~\HL{lornusrr}{\B{lor\-nu\-srr}}, radiant.

% loran
\hi
\HT{loran}{\B{\cp{lo}ran}} \I{["lo.Ran]}  \emph{n.} elegance, grace.   
\B{Ya\-mì tsun fko tsi\-ve\-'a lo\-ran\-it re\-nu\-ä kil\-van\-ä slä kll\-te\-sìn wä\-pan.} From the air you can see the grace of the river's form but from the ground it's hidden.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\ding{43}~\HL{lor}{\B{lor}}, \HL{ran}{\B{ran}})

% loreyu
\hi
\HT{loreyu}{\B{lo\cp{re}yu}} \I{[lo."RE.ju]} \emph{n.} \ding{96} helicoradian, ``beautiful spiral'', \emph{helicoradium spirale}. 
\B{Mì na'\-rìng Tsyeyk\-ìl lo\-re\-yu\-ti 'o\-lam\-pi a tsa\-swaw wou oer nì\-wotx!} That instant when Jake touched the \emph{Helicoradiam Spirale} in the forest I was totally amazed.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\emph{a. idiom}) \B{(Na) lo\-re\-yu 'aw\-nam\-pi}, extreme shyness, [lit.: (like) a touched helicoradian] \B{Lu por mok\-ri a\-mik\-lor, slä lo\-re\-yu 'aw\-nam\-pi lu. Ke tsun ri\-vol eo su\-te.} She has a beautiful voice, but she's extremely shy. She can't sing in front of people.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{lor}{\B{lor}}, \HL{'Iheyu}{\B{'ì\-he\-yu}})

% lornusrr
\hi 
\HT{lornusrr}{\B{\cp{lor}nusrr}} \I{["loR.nu.s\s{r}]} \emph{adj.} radiant. 
\B{Ey\-we\-veng lu lor\-nu\-srr.} Pandora is radiant.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
(\ding{43}~\HL{lor}{\B{lor}}, \HL{nrr}{\B{nrr}}) 

% lortsyal
\hi 
\HT{lortsyal}{\B{\cp{lor}tsyal}} \I{["loR.\t{ts}jal]} \emph{n., fauna} shimmyfly.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln}
(\ding{43}~\HL{lor}{\B{lor}}, \HL{tsyal}{\B{tsyal}})

% lrrtok / si
\hi
\HT{lrrtok}{\B{\cp{lrr}tok}} \I{["l\s{r}.tok\textcorner]} \textsc{i}.~\emph{n.} smile. 
\B{Nge\-yä lrr\-tok lor lu nì\-txan.} Your smile is very beautiful.
\B{Ay\-nge\-yä ay\-u\-pxa\-re\-ri sì ay\-lrr\-tok\-ìri sei\-yi oe ira\-yo.} Thank you for all your messages and smiles.~\LINK{http://forum.learnnavi.org/news-announcements/happy-birthday-to-karyu-pawl/msg556899/\#msg556899}{ln}
(\emph{a.~idiom}) \B{Lrr\-tok ngar!\slash Ay\-lrr\-tok ngar (li\-vu)!} Good luck.\slash Good luck to you.~\LINK{http://forum.learnnavi.org/language-updates/good-luck-with-this-one!/}{ln} 
\B{Zì\-tsal\-trr\-ìri tì\-mun\-txa\-yä ay\-lrr\-tok!} Happy anniversary (of your marriage)!~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b} \\
\textsc{ii}.~\HT{lrrtok si}{\B{\cp{lrr}tok si}} \I{["l\s{r}.tok\textcorner{} si]}  \emph{vin.}~(\emph{a.}) smile. 
\B{Prr\-nen lrr\-tok si a krr, srer me\-song\-tsyìp a\-ho\-na.} When the baby smiles, two adorable dimples appear.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Nì\-law Nawm\-a Sa'\-nok lrr\-tok so\-li me\-ngar.} Clearly the Great Mother smiled on you both.~\LINK{http://forum.learnnavi.org/language-updates/luke-kxu-without-harm/}{ln} 
(\emph{b. idiom}) shine (of the sun) \B{Tsaw\-ke lrr\-tok si.} The sun shines.~\LINK{http://forum.learnnavi.org/language-updates/some-news-from-berlin/}{ln}

% lu
\hi
\HT{lu}{\B{lu}} \I{[lu]}  \emph{vin.} \on{1}.~(\emph{copula}) be (am, is, are, was, were etc.).
\B{Rel\-tse\-o\-tu a\-txan\-tsan lu nga!} You are an excellent artist!~\LINK{http://forum.learnnavi.org/language-updates/art-related-vocab/}{ln}
\B{Taw\-sìp nge\-yä lu sngel\-tseng.} Your ship is a garbage scow.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog_.28Twitter_Questions.29}{wiki\slash ugo}
\B{Lu nga win sì txur \slash{} Lu nga txan\-tslu\-sam}, You are fast and strong \slash{} You are wise.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Fì\-tskxe\-ri fa\-'o lu yey; ke lu ko\-um.} This rock has straight sides; it's not rounded.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} 
\on{2}.~(\emph{a. to indicate existence}) it is, it's. \B{Nga\-ru tì\-yawr, lu kxey\-ey.} You are right, it's a mistake.~\LINK{http://forum.learnnavi.org/language-updates/a-long-answer-to-a-quick-question/msg155104/\#msg155104}{ln} 
(\emph{b.}) there is\slash are.
\B{Äo fì\-ut\-ral lu tsmìm 'ang\-tsìk\-ä.} There's a hammerhead track under this tree.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Lo\-lu txan\-tsan\-a ul\-txa mì Si\-ä\-tll.} There was a great meeting in Seattle.~\LINK{http://naviteri.org/2011/07/attributive-\%E2\%80\%9Ca\%E2\%80\%9D-and-truncated-style/}{b}
\B{Lo\-lu ka\-vuk, slä Tse\-nul tì\-ngay\-it ko\-lu\-lat.} There was treachery, but Tsenu revealed the truth.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Tskxe\-keng\-lu\-ke ke lu kea tì\-tsan\-'ul.} Without practice there is no improvement.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\emph{c. as parallel structure}) \B{Fra\-u\-van\-ìri lu yo\-ra'\-tu, lu snay\-tu.} For every game, there's a winner and a loser.~\LINK{http://naviteri.org/2011/12/a-note-on-the-word-\%E2\%80\%9Cyora\%E2\%80\%99tu\%E2\%80\%9D/}{b}
\B{Su\-te\-kip nì\-wotx, ftxey Na'\-vi, ftxey Saw\-tu\-te, lu sìl\-tsan, lu kawng.} There are good and bad among all people, whether Na'vi or Skypeople.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\on{3}.~(\emph{in combination with the dative to indicate possession}) have. (`have' constructions usually begin with the verb)
\B{Nga\-ru lu fpom srak?} How are you? [lit.: Do you have well-being?]~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{La\-yu oe\-ru ye'\-rìn sìl\-tsan\-a fmawn.} Soon, I will have good news.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ke lu oer set krr a\-txan, ul\-te ke new tì\-kxey si\-vi.} I don't have much time now, and I don't want to make a mistake.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg139677/\#msg139677}{ln}
(\emph{a. allows double dative})
\B{Lu oe\-ru ay\-lì\-'u fra\-por.} I have something to say to everyone.~\LINK{http://forum.learnnavi.org/language-updates/i-have-something-to-say/}{ln}
\B{Lu fra\-por ay\-lì\-'u oe\-ru.} Everyone has something to say to me.~\LINK{http://forum.learnnavi.org/language-updates/i-have-something-to-say/}{ln}
(\emph{b. idiomatic uses})
\B{lu fko\-ru ya\-yayr}, to be confused. \B{Saw\-tu\-te\-yä hem\-ìri lu aw\-nga\-ru ya\-yayr.} The Skypeople's actions confuse us.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Slä 'en si oe, tsa\-me\-lì\-'u oe\-ru ya\-yayr lo\-la\-tsu.} But I guess, those two words would have confused me.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-677}{b}
\B{lu fko\-ru sngum}, to be worried. \B{Lu oe\-ru sngum a sa\-ron\-yu ke tì\-ye\-vä\-txaw.} I'm worried that the hunters will not return.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{lu fko\-ru ing\-yen}, to be puzzled. \B{Lu oer ing\-yen a Ìstaw nim lu fì\-txan kum\-a pxìm wä\-pan.} I'm puzzled that Ìstaw is so shy that he frequently hides.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{lu fko\-ru tì\-kin a} \texttt{+} \emph{verb}, need to do s.t. \B{Po\-ri aw\-nga\-ru lu tì\-kin a nu\-me nì\-'ul.} We need to learn more about him.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki}
\B{lu fko\-ru haw\-tsyìp}, to take a nap. 
\B{Oel new fu\-ta li\-vu oer set haw\-tsyìp.} I want to take a nap now.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
\B{Sa'\-nur le\-ru haw\-tsyìp.} Mommy is taking a nap.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{lu fko\-ru txoa (a ta tu\-teo)}, to forgive s.o. \B{Oe\-ru txoa li\-vu.} Forgive me. [lit.: may I have forgiveness]~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg139677/\#msg139677}{ln} \B{Säm\-yam\-ìl po\-ru wa\-yìn\-txu fu\-ta nga\-ta lo\-lu li txoa.} A hug will show him that you've already forgiven him.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{lu fko\-ru fmokx}, to be jealous (\emph{neutral unless otherwise specified with} ‹\B{äng}› \emph{or} ‹\B{ei}›). \B{Fu\-ria fì\-txan fnä\-ngan Ul\-rey\-ìl tsko swi\-zaw\-it, lu oe\-ru fmokx.} I'm jealous of the fact that Ulrey is such an excellent archer.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{lu fko\-ru nrra}, to be proud of s.o. (with the topic). \B{Nga\-ri le\-iu oe\-ru nrra nì\-txan, ma 'ite.} I'm very proud of you, daughter.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{lu fko\-ru yew\-la}, to be disappointed. \B{Oer lu txan\-a yew\-la a ke tsun nga oe\-hu ki\-vä\-teng me\-srr\-ay.} I'm very disappointed you can't hang out with me the day after tomorrow.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{lu fko\-ru tì\-yawr}, to be right. \B{Nga\-ru tì\-yawr, lu kxey\-ey.} You are right, it's a mistake.~\LINK{http://forum.learnnavi.org/language-updates/a-long-answer-to-a-quick-question/msg155104/\#msg155104}{ln}
\B{lu fko\-ru tì\-kxey}, to be wrong. \B{Po ke tsun pi\-vll\-ngay san oe\-ru tì\-kxey.} He can't admit he's wrong.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{lu fko\-ru ho\-an}, to feel comfortable.
\B{Sko frr\-tu, fra\-krr mì hel\-ku nge\-yä lu oe\-ru ho\-an nì\-txan, ma tsmuk.} I always feel very comfortable as a guest in your home, brother.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
\B{lu fko\-ru tì\-hì\-pey}, to be hesitant.
\B{Sar tsa\-lì\-'ut a fì\-'u\-ri lu oe\-ru tì\-hì\-pey nì\-'it.} I'm a bit hesitant about using that word.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
\on{4}.~(\emph{can be omitted in coll. conversation}) 
\B{Nga (lu) pe\-su?} Who are you?~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{luke}{\B{lu\-ke}}, without. 
\on{2}.~\HL{tIkelu}{\B{tì\-ke\-lu}}, lack.

% luan
\hi 
\HT{luan}{\B{\cp{lu}an}} \I{["lu.an]} (\textsc{rn}: \B{\cp{lu}an} \I{["lu.an]}) \emph{vtr.} owe (\emph{moral obligation to give s.t. to s.o.}). 
\B{Fol nge\-yä tsmu\-ke\-ru lu\-an tsko\-ti a\-mip} They owe your sister a new bow.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
\B{Oey vo\-ìk\-ìri a\-lew\-nga' lu\-an oel ngar tì\-oeyk\-tìng\-it.} I owe you an explanation for my shameful behavior.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}

% luke
\hi
\HT{luke}{\B{\cp{lu}ke}} \I{["lu.kE]}  \emph{adp.} without.
\B{Lu me\-nge\-yä kel\-ku na'\-rìng\-ä lu\-ke kxu a\-txan a fì\-'u fmawn a\-sìl\-tsan lu nì\-ngay.} It's truly good news that your forest house is without great harm.~\LINK{http://forum.learnnavi.org/language-updates/luke-kxu-without-harm/}{ln} 
\B{Ke tsun oe tsaw\-te\-ri fpi\-vìl lu\-ke fwa sngä\-'i tsngi\-vaw\-vìk.} I can't think about it without starting to cry.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Sra\-ke tso\-lun fra\-'ut tsli\-vam kxey\-ey\-lu\-ke?} Could you understand everything without a mistake?~\LINK{http://naviteri.org/2012/10/txon-eywaevenga-text-and-translation/}{b} 
(\emph{a. idiom}) \B{(na) la\-nay'\-ka lu\-ke re\-'o}, ``completely lost,'' [lit.: Like a slinger without a head].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Po maw kxitx mun\-txa\-tu\-ä 'a\-me\-fu na la\-nay'\-ka re'\-o\-lu\-ke.} After the death of his spouse, he felt completely lost.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lukpen}{\B{luk\-pen}}, naked. 
\on{2}.~\HL{kxuke}{\B{kxu\-ke}}, safe. 
\on{3}.~\HL{velke}{\B{vel\-ke}}, chaotic. 
\on{4}.~\HL{am'aluke}{\B{am\-a'a\-lu\-ke}}, without a doubt. 
\on{5}.~\HL{letxiluke}{\B{le\-txi\-lu\-ke}}, unhurried. 
\on{6}.~\HL{nItxiluke}{\B{nì\-txi\-lu\-ke}}, unhurriedly, leisurely. 
\on{7}.~\HL{nItkanluke}{\B{nìt\-kan\-lu\-ke}}, accidentally. 
\on{8}.~\HL{renulke}{\B{re\-nul\-ke}}, irregular. 
\on{9}.~\HL{am'ake}{\B{am\-'a\-ke}}, sure, confident.
\on{10}.~\HL{txewluke}{\B{txew\-lu\-ke}}, endless, boundless.
\on{11}.~\HL{tI'iluke}{\B{tì\-'i\-lu\-ke}}, endless, never\hyp{}ending.
\on{12}.~\HL{lukftang}{\B{luk\-ftang}}, constant, continual.

% lukftang
\hi 
\HT{lukftang}{\B{luk\cp{ftang}}} \I{[luk\textcorner."ftaN]} \emph{or} \I{[lUk\textcorner.]} \emph{adj.} constant, continual. 
\B{Pe\-yä tì\-pu\-sll\-txel a\-luk\-ftang oe\-ti srätx.} His constant talking annoys me.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(\ding{43}~\HL{luke}{\B{lu\-ke}}, \HL{tIftang}{\B{tì\-ftang}}; \HL{letut}{\B{le\-tut}})

% lukpen
\hi
\HT{lukpen}{\B{luk\cp{pen}}} \I{[luk\textcorner."pEn]} \emph{or} \I{[lUk\textcorner.]}  \emph{adj.} naked, without clothing. 
(\ding{43}~\HL{luke}{\B{lu\-ke}}, \HL{pxen}{\B{pxen}})

% lumpe
\hi
\HT{lumpe}{\B{\cp{lum}pe}} \I{["lum.pE]} \emph{or} \I{["lUm.]} \emph{or} \HL{pelun}{\B{pe\cp{lun}}} \I{[pE."lun]} \emph{or} \I{[."lUn]} \emph{inter.} why? what reason?
\B{Oel lum\-pe tsat so\-lung mì upxa\-re?} Why did I add that in the message?~\LINK{http://naviteri.org/2010/07/vocabulary-update/comment-page-1/\#comment-95}{b}
\B{Tey\-nga lum\-pe fo ho\-lum ke lu law.} It's not clear why they left.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Nge\-yä tsngal lum\-pe lu mek?} Why is your cup empty?~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{lun}{\B{lun}})

% lun
\hi
\HT{lun}{\B{lun}} \I{[lun]} \emph{or} \I{[lUn]} \emph{n.} reason. 
(\ding{43}~\HL{oeyk}{\B{oeyk}}) \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIlun}{\B{nì\-lun}}, of course, logically.   
\on{2}.~\HL{talun1}{\B{ta\-lu\-na\slash a\-lun\-ta}}, because. 
\on{3}.~\HL{talun2}{\B{ta\-lun}}, because of. 
\on{4}.~\HL{lumpe}{\B{pe\-lun\slash lum\-pe}}, why? 
\on{5}.~\HL{velun}{\B{ve\-lun}}, logic.

% lupra
\hi
\HT{lupra}{\B{\cp{lup}ra}} \I{["lup\textcorner.Ra]} \emph{or} \I{["lUp\textcorner.]} (\textsc{rn}: \B{\cp{lùp}ra} \I{["lUp\textcorner].Ra}) \emph{n.} style.
\B{Pll\-txe fra\-po san fì\-fkxi\-le lor lu nì\-ngay sìk, slä oe\-ri ke su\-nu oer lup\-ra kaw\-'it.} Everybody says this necklace is really beautiful, but me, I don't like the style one bit.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{fyolup}{\B{fyo\-lup}}, exquisite, sublime in taste. 
\on{2}.~\HL{fe'lup}{\B{fe'\-lup}}, tacky, in poor taste. 
\on{3}.~\HL{snolup}{\B{sno\-lup}}, personal style, aesthetic, presence.

\end{multicols}