\pdfbookmark[0]{M}{M}
\begin{center}
\section*{M \I{[m]} (\emph{MeM})}
\end{center}

\begin{multicols}{2}

% ma
\hi
\HT{ma}{\B{ma}} \I{[ma]} \emph{part., vocative marker} `O, \emph{name}', `dear \emph{name}' (\emph{but usually not translated}).
\B{Tew\-ti, ma Prr\-ton!} Wow, Prrton!~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll}
\B{Oel ay\-nga\-ti ka\-meie, ma oe\-yä ey\-lan.} I see you, my friends.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ma smu\-kan, ma smu\-ke}, Brothers and sisters!~\LINK{https://forum.learnnavi.org/language-updates/naviya-the-other-vocative/}{ln\slash a}
(\ding{43}~\HL{-ya}{\B{-ya}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{maitan}{\B{ma\-i\-tan}}, my son (address). 
\on{2}.~\HL{maite}{\B{ma\-i\-te}}, my daughter (address).
\on{3}.~\HL{manawmtu}{\B{ma\-nawm\-tu}}, ``excuse me''. 
\on{4}.~\HL{matu}{\B{ma\-tu}}, ``excuse me, hey''.
\on{5}.~\HL{manga}{\B{ma\-nga}}, ``hey you''. 

% maitan
\hi 
\HT{maitan}{\B{ma\cp{i}tan}} \I{[ma."{}i.tan]} \emph{ph.} my son (\emph{form of address}). 
\B{Ma\-i\-tan za\-'u fì\-tseng.} Come here, son.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{'itan}{\B{'itan}})

% maite
\hi 
\HT{maite}{\B{ma\cp{i}te}} \I{[ma."{}i.tE]} \emph{ph.} my daughter (\emph{form of address}).  
(\ding{43}~\HL{'ite}{\B{'ite}})

% makto
\hi
\HT{makto}{\B{\cp{mak}to}} \I{["mak\textcorner.to]} \emph{vtr.} ride.   
\B{Trr\-am\-ä ay\-u\-van\-ìri mak\-to Ak\-wey nì\-yew\-la ha snaytx.} Akwey rode disappointingly in yesterday's games so he lost.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Tsun fko tsa\-tse\-nge\-ne ki\-vä nì\-sle\-le fu fa fwa ik\-ran\-it mak\-to nì\-'aw.} You can only get there by swimming or riding an ikran.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
\B{Oe\-yä ik\-ran sli\-vu nga, tsa\-krr oeng 'aw\-si\-teng mi\-vak\-to.} Be my ikran and let's ride together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{wiki} 
(\emph{a.~idiom}) \B{Mak\-to fya\-pe?} ``How are things? How's everything going? How are you doing?'
\B{A: Mak\-to fya\-pe?---B: Zong. Nga\-ri tut?---A: Nìk\-sran. Oe\-ru lu fpom, slä oey 'i\-tan lu spxin.} A: How's everything?---B: Good. You?---A: So\hyp{}so. I'm fine, but my son is sick.~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b} 
(\emph{b.~idiom}) \B{Mak\-to zong.\slash Mi\-vak\-to nì\-zaw\-nong.} Take care.\slash Travel safely.~\LINK{http://forum.learnnavi.org/language-updates/makto-zong!/}{ln} 
(\emph{c.~special usage}) \B{Tor\-uk Mak\-to}, Rider of the Last Shadow, \emph{legendary hero of the Na'vi}. 
\B{Tor\-uk Mak\-to po\-lä\-hem a fì\-'u ro\-lo\-'a nì\-txan Oma\-ti\-ka\-ya\-ru.} The arrival of Toruk Makto made a great impression on the Omatikaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
(\emph{d.~special usage}) \B{\cp{ik}\-ran \cp{mak}\-to}, the act of riding an ikran.~\textsc{\cb{dh}} 
(\emph{e.~special usage}) \B{\cp{pa'}\-li \cp{mak}\-to}, the act of riding a pa'\-li.~\textsc{\cb{dh}}  \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{kAmakto}{\B{kämakto}}, ride out.
\on{2}.~\B{Toruk Makto}, Rider of the Last Shadow.

% mal 
\hi
\HT{mal}{\B{mal}} \I{[mal]} \emph{adj.} trustworthy, trust\hyp{}inspiring.  
\B{Fì\-tì\-kang\-kem\-vi\-ri le\-tsran\-ten ke new oe hu Ra\-lu tì\-kang\-kem si\-vi. Po ke lä\-ngu mal.} I don't want to work with Ralu on this important project. He's not trustworthy, unfortunately.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Nga mal lar\-mu oer!} I trusted you!~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
\B{Lu tsa\-tsam\-si\-yu le\-'aw\-a ha\-pxì\-tu tsam\-po\-ngu\-ä a mal lu moer.} That warrior is the only member of the war party that we both trust.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImal}{\B{tìmal}}, trustworthiness.  
\on{2}.~\HL{nImal}{\B{nìmal}}, trustingly, without hesitation.

% mam
\hi 
\HT{mam}{\B{mam}} \I{[mam]} \emph{vtr.} wrap. 
\B{Fì\-srä\-ti pxaw sey mi\-vam fte tsat hi\-vaw\-nu.} Wrap this cloth around the bowl to protect it.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}

% man
\hi 
\HT{man}{\B{man}} \I{[man]} \emph{vin.} belong (\ding{43}~\HL{hu}{\B{hu}}, with; \HL{mI}{\B{mì}}, in) (\emph{fitting in; feeling comfortable as part of a group; being in a place, position, or relationship where one belongs}). 
\B{Man po fì\-po\-ngu\-mì nì\-law.} He clearly belongs in this group.~\LINK{http://naviteri.org/2022/06/solalew-maul-zisita-half-the-year-is-over/}{b}
\B{Nga man oe\-hu, ma pa\-ska\-lin.} You belong with me, honey.~\LINK{http://naviteri.org/2022/06/solalew-maul-zisita-half-the-year-is-over/}{b}
\B{Krro krro fpìl oel fu\-ta ke man oe kaw\-tseng.} Sometimes I think I don't belong anywhere.~\LINK{http://naviteri.org/2022/06/solalew-maul-zisita-half-the-year-is-over/}{b}
\B{Ro\-lun oel olo'\-ti a 'e\-fu fta oe man tsa\-tseng\-mì.} I've found a community where I feel I belong.~\LINK{http://naviteri.org/2022/06/solalew-maul-zisita-half-the-year-is-over/}{b} 

% manawmtu
\hi 
\HT{manawmtu}{\B{ma\cp{nawm}tu}} \I{[ma."nawm.tu]} \emph{intj.} ``excuse me sir, excuse me madam'' (\emph{attract s.o.'s attention politely\slash honorificly}). 
\B{Ma\-nawm\-tu, sra\-ke lu\-yu nge\-nga eyk\-tan fì\-olo'\-ä?} Excuse me, sir, are you the leader of this clan?~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
(\ding{43}~\HL{ma}{\B{ma}}, \HL{nawmtu}{\B{nawm\-tu}}; \HL{matu}{\B{ma\-tu}}, \HL{manga}{\B{ma\-nga}})

% mankuan
\hi 
\HT{mankuan}{\B{\cp{Man}kuan}} \I{["man.ku.an]} \emph{n., clan name} the Ash People.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b}

% manga
\hi 
\HT{manga}{\B{ma\cp{nga}}} \I{[ma."Na]} \emph{intj.} ``hey, hey you'' (\emph{attract s.o.'s attention one is close\slash superior to}). 
\B{Ma\-nga! Kem\-pe si?} Hey! What are you doing?~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{ma}{\B{ma}}, \HL{nga}{\B{nga}}; \HL{manawmtu}{\B{ma\-nawm\-tu}}, \HL{matu}{\B{ma\-tu}})

% Maru
\hi
\B{Maru} \I{[ma.Ru]} \emph{n. female name}. \LINK{}{dh}
\B{Tì\-mun\-txa me\-fe\-yä zo\-law\-prr\-te' Ma\-ru\-ne nì\-layl. Ke lu por kea fmokx kaw\-'it.} Their marriage brought Ma\-ru pleasure. She felt no jealousy at all.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}

% masat
\hi
\HT{masat}{\B{\cp{ma}sat}} \I{["ma.sat\textcorner]} \emph{n.}~\ding{246} breastplate (\emph{armor}).
\B{Ma\-sat oe\-yä ley nì\-ftxan na pum nge\-yä.} My breastplate is as valuable as yours.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
\B{Sne\-yä ma\-sat\-it pol kä\-sro\-lìn oer.} He lent me his breastplate.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}

% Mati
\hi 
\B{Mati} \I{[ma.ti]} \emph{name}.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 

% matu
\hi 
\HT{matu}{\B{ma\cp{tu}}} \I{[ma."tu]} \emph{intj.} ``excuse me, hey'' (\emph{attract s.o.'s attention in neutral language, not overly polite\slash familiar}). 
\B{Ma\-tu, ngal hawn\-tsyokx\-it tì\-mung\-zup.} Excuse me, you just dropped your glove.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
(\ding{43}~\HL{ma}{\B{ma}}, \HL{tute1}{\B{tu\-te}}; \HL{manawmtu}{\B{ma\-nawm\-tu}}, \HL{manga}{\B{ma\-nga}}) 

% matsa
% \hi
% \B{\cp{ma}tsa} \I{["ma.\t{ts}a]} \emph{n.}, \textsc{lw} Matzah (\emph{unleavened bread}). \B{aw\-ngal yom ha\-me\-tsìt, yom ma\-tsat, ke tsran\-ten}, it doesn't matter whether we eat Chametz or Matzah.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{this night}

% maw
\hi
\HT{maw}{\B{maw}} \I{[maw]} \emph{adp.} after (\emph{temporal}).
\B{Maw hì\-krr ay\-oe tì\-yä\-txaw.} We'll be right back after this break.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} 
\B{Maw txan\-tom\-pa, pxay\-a rìk\-äo la\-mu tska\-lep pe\-yä, ha tsat ku\-lat ay\-oel.} After the rainstorm, his crossbow was under a lot of leaves, so we uncovered it.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Maw fwa fkol Kel\-ut\-ral\-it sko\-la\-'a, lem\-rey\-a ha\-pxì\-tul tsa\-soai\-ä txo\-lu\-la mip\-a kel\-ku\-ti mì ta\-yo.} After Hometree was destroyed, the surviving members of that family built a new home on the plains.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\emph{a. of telling time}) in.  
\B{Oe pä\-hem maw\slash kay trr\-pxì\-vi a\-mrr.} I'll be there in five minutes.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{mawkrr}{\B{maw\-krr}}, after, afterwards. 
\on{2}.~\HL{mawkrra}{\B{maw\-krr\-a\slash a\-krr\-maw}}, after (\emph{conj.}). 
\on{3}.~\HL{pximaw}{\B{pxi\-maw}}, right after. 
\on{4}.~\HL{txon'ongmaw}{\B{txon\-'ong\-maw}}, twilight, after sunset.  
\on{5}.~\HL{trr'ongmaw}{\B{trr\-'ong\-maw}}, dawn, time after sunrise.
\on{6}.~\HL{kxamtxomaw}{\B{kxam\-txo\-maw}}, after midnight.
\on{7}.~\HL{kxamtrrmaw}{\B{kxam\-trr\-maw}}, after noon.
\on{8}.~\HL{mawfwa}{\B{maw\-fwa}}, after (\emph{conj.}).
\on{9}.~\HL{mawftxele}{\B{maw\-ftxe\-le}}, belatedly, in hindsight.

% mawftxele
\hi 
\HT{mawftxele}{\B{maw\cp{ftxe}le}} \I{[maw."ft'E.lE]} \emph{adv.} belatedly, in hindsight, after the fact, as an afterthought. 
\B{Oel pe\-yä ftxo\-zä\-ti tswo\-lä\-nga', ha pol\-txe por san ftxo\-zä\-ri ay\-lrr\-tok nga\-ru sìk maw\-ftxe\-le.} Unfortunately I forgot his birthday, so I said ``Happy Birthday'' belatedly.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}  
(\ding{43}~\HL{maw}{\B{maw}}, \HL{txele}{\B{txe\-le}})

% mawfwa
\hi 
\HT{mawfwa}{\B{\cp{maw}fwa}} \I{["maw.fwa]} \emph{conj.} after.
\B{Maw\-fwa zup tom\-pa, lu ngo\-a a\-txan fì\-ha\-pxì\-mì na'\-rìng\-ä.} After rain, there's a lot of mud in this part of the forest.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\ding{43}~\HL{maw}{\B{maw}}, \HL{fwa}{\B{fwa}})

% mawkrr
\hi
\HT{mawkrr}{\B{\cp{maw}krr}} \I{["maw.k\s{r}]} \emph{adv.} after (\emph{temporal}), afterwards, later.
\B{Ye'\-rìn 'ì\-yi'\-a sä\-nu\-me a tsa\-ri kll\-fro' oe; maw\-krr la\-ye\-i\-u oer krr nì\-'ul.} My teaching responsibilities will soon end; after that I will have more time.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Nì\-'aw\-ve oeng yo\-lom wu\-tsot; maw\-krr u\-van si.} First we had dinner; afterwards we played (a game).~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b} 
\B{Po\-lä\-hem Saw\-tu\-te kam zì\-sìt a\-mrr, hum me\-zì\-sìt maw\-krr.} The Sky People arrived five years ago and left two years later.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{maw}{\B{maw}}, \HL{krr}{\B{krr}}, \HL{mawkrra}{\B{maw\-krr\-a}}; \ding{214}~\HL{srekrr}{\B{sre\-krr}})

% mawkrra
\hi
\HT{mawkrra}{\B{maw\cp{krr}a}} \I{[maw."k\s{r}.a]} \emph{or} \HL{mawkrra}{\B{a\-\cp{\-krr}\-maw}} \I{[a."k\s{r}. maw]} \emph{conj.} after (\emph{temporal}).
\B{Fo\-ri maw\-krr\-a fa ren\-ten ioi sä\-po\-li ho\-lum.} After they put on their goggles, they left.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Maw\-krr\-a po ho\-lum, oe 'e\-fu ke\-ftxo.} I feel sad after he left.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-986}{b}
(\ding{43}~\HL{mawkrr}{\B{maw\-krr}}, \HL{mawfwa}{\B{maw\-fwa}})

% mawl
\hi
\HT{mawl}{\B{mawl}} \I{[mawl]} \emph{n., num.} half, one\hyp{}half, $\frac{1}{2}$.
\B{Po\-ri lu sno\-lup räp\-tum, ul\-te mawl ay\-lì\-'uä ke lu ngay.} His personal style is coarse and half of the words are not true.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
\B{Tsu'\-tey\-ìl to\-lìng oer mawl\-it smar\-ä.} Tsu'tey gave me a half of the prey.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\emph{a. to tell time}) \B{trr\-pxì a\-mrr\-ve sì mawl} \emph{or coll.} \B{mrr\-ve sì mawl}, (it's) \on{5}:\on{30}.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}

% mawey
\hi
\HT{mawey}{\B{ma\cp{wey}}} \I{[ma."wEj]} \emph{adj.} calm.
\B{Tok oel lo\-ra tsa\-mo\-ti a mì na'\-rìng a krr, 'e\-fu ma\-wey sì nit\-ram.} When I'm in that beautiful hollow in the forest, I feel calm and happy.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\emph{a. idiom}) \B{Nga\-ri txe'\-lan ma\-wey li\-vu.} Don't worry (about it).~\LINK{http://forum.learnnavi.org/language-updates/fmawno/msg293595/\#msg293595}{ln} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{maweypey}{\B{ma\-wey\-pey}}, be patient. 

% maweypey
\hi
\HT{maweypey}{\B{ma\cp{wey}pey}} \I{[ma."wEj.p$\cdot$Ej]} \emph{vin., inf.}\on{33} be patient. 
\B{Lu nga sì la\-he\-a ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ay\-aw\-po a ma\-wey\-pe\-rey.} You and the other members of our \emph{Sprachbund} are the ones who are being patient.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Ze\-ne ma\-wey\-pi\-vey, ma ey\-lan.} You must be patient, friends.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} 
(\ding{43}~\HL{mawey}{\B{ma\-wey}}, \HL{pey}{\B{pey}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImweypey}{\B{tìm\-wey\-pey}}, patience.   
\on{2}.~\HL{lemweypey}{\B{lem\-wey\-pey}}, patient.   
\on{3}.~\HL{nImweypey}{\B{nìm\-wey\-pey}}, patiently.  
\on{4}.~\HL{maweypeyyu}{\B{ma\-wey\-pey\-yu}}, patient person. 

% maweypeyyu 
\hi
\HT{maweypeyyu}{\B{ma\cp{wey}peyyu}} \I{[ma."wEj.pEj.ju]} \emph{n.} one who is patient, patient person. 
\B{Ay\-ma\-wey\-pey\-yu lä\-ngu nga sì la\-he\-a ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä.} The one's who're being patient are you and the other \emph{Sprachbund} members.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\ding{43}~\HL{maweypey}{\B{ma\-wey\-pey}})

% mawup
\hi
\HT{mawup}{\B{\cp{ma}wup}} \I{["ma.wup\textcorner]} \emph{or} \I{[.wUp\textcorner]} \emph{n., fauna} turtapede. \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{mawuptsyIp}{\B{ma\-wup\-tsyìp}}, turtle.

% mawuptsyìp
\hi 
\HT{mawuptsyIp}{\B{\cp{ma}wuptsyìp}} \I{["ma.wup\textcorner.\t{ts}jIp\textcorner]} \emph{or} \I{[.wUp\textcorner.]} \emph{n., fauna} turtle (Earth animal). 
(\ding{43}~\HL{mawup}{\B{ma\-wup}}) 

% may'
\hi
\HT{may'}{\B{may'}} \I{[majP]} \emph{vtr.} (\emph{a}.~\texttt{+}\emph{control}) taste, check out something by tasting. 
\B{Ke new oe mi\-vay' tsngan\-ti a 'o\-lem Ri\-nil.} I don't want to taste the meat that Rini cooked.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\emph{b}.) try, sample, evaluate, check out, test\hyp{}drive.
\B{Ngal mo\-lay' pxaw\-pxun\-it Lo\-ak\-ä a krr pe\-kxin\-um?} When you tried on Loak’s armband, how tight was it?~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\emph{idiom}) \B{Mi\-vay'.\slash May' ko.} Give it a try.\slash Have a go.
(\emph{modal use}) \B{May' mi\-vak\-to pa'\-lit!} Try riding a direhorse!~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}

% mäkxu
\hi
\HT{mAkxu}{\B{mä\cp{kxu}}} \I{[m$\cdot$\ae."k'u]} \emph{vtr., inf.}\on{11} to interrupt, throw out of harmonious balance (\emph{only  for activities or established conditions})
\B{Tsun (oe) mi\-vä\-kxu hì\-krr srak?} May I interrupt a moment?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Pol mo\-lä\-kxu ul\-txa\-ti.} He interrupted the meeting.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}

% mauti
\hi
\HT{mauti}{\B{\cp{ma}uti}} \I{["ma.u.ti]} \emph{n.}, \ding{96} fruit.
\B{Pe\-fne\-ma\-u\-ti su\-nu ngar fra\-to?} What's your favorite fruit?~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\emph{a. idiom}) 
\B{Rey\-kì\-kxi ut\-ral\-ti, zup ma\-u\-ti.} If you shake the tree, the fruit will fall (i.e, ``actions have consequences'').~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{utumauti}{\B{u\-tu\-ma\-u\-ti}}, banana fruit. 
\on{2}.~\HL{paymaut}{\B{pay\-ma\-ut}}, fountain tree.
\on{3}.~\HL{tskxemauti}{\B{tskxe\-ma\-u\-ti}}, nut.

% me+
\hi
\HT{me+}{\B{me}\texttt{+}} \I{[mE.]} \emph{dual pref.} two (of s.t.).

% me'em
\hi
\HT{me'em}{\B{me\cp{'em}}} \I{[mE."PEm]} \emph{adj.} harmonious. \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIme'em}{\B{tì\-me\-'em}}, (general) harmony.

% mewn
\hi 
\HT{mewn}{\B{mewn}} \I{[mEwn]} \emph{vin.} sigh. 
\B{Nga me\-rewn pe\-lun, ma pa\-ska\-lin?} Why are you sighing, sweetie?~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{sAmewn}{\B{sä\-mewn}}, sigh, intance of sighing.

% mefan
\hi
\HT{mefan}{\B{me\cp{fan}}} \I{[mE."fan]} \emph{num., n.} two\hyp{}thirds, $\frac{2}{3}$.
(\ding{43}~\HL{pan}{\B{pan}})

% mefo
\hi
\HT{mefo}{\B{me\cp{fo}}} \I{[mE."fo]} (\emph{gen.} \B{me\-\cp{fe}\-yä}) \emph{pn.} they (two), the two of them (3\textsuperscript{rd} person dual).
\B{Me\-fo fì\-tsap mä\-po\-le\-yam teng\-krr tsngaw\-vìk.} The two of them hugged each other and wept.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{'Aw\-po a me\-fo\-kip tsul\-fä\-tu le\-iu lì'\-fya\-yä le\-Na'\-vi.} One of the two is a master of the Na'vi language.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/comment-page-1/\#comment-2619}{b}
(\emph{a. idiom}) \B{'Aw\-si\-teng lu me\-fo la\-nay'\-ka.} \emph{praising how well two people work together and complement each other} [lit.: they are a slinger (together)].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{To\-la\-ron me\-fol me\-sa\-li\-o\-a\-ngit! Tew\-ti, 'aw\-si\-teng lu me\-fo la\-nay'\-ka.} They hunted two sturmbeest? Wow, they work very well together.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{43}~\HL{po}{\B{po}})

% mei
\hi 
\HT{mei}{\B{\cp{me}i}} \I{["mE.i]} \emph{adj.} wet. 
\B{Kll\-te lu mei a krr, fwa fwi lu ftue.} When the ground is wet, it's easy to slip.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
(\ding{214}~\HL{ukxo}{\B{ukxo}}, \ding{43}~\HL{lepay}{\B{le\-pay}}, \HL{paynga'}{\B{pay\-nga'}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{meitayo}{\B{mei\-ta\-yo}}, wetlands.

% meitayo
\hi 
\HT{meitayo}{\B{mei\cp{ta}yo}} \I{[me.i."ta.jo]} \emph{or coll.} \I{[mej."ta.jo]} \emph{n.} wetlands. 
(\ding{43}~\HL{mei}{\B{mei}}, \HL{txayo}{\B{txa\-yo}})

% mek
\hi
\HT{mek}{\B{mek}} \I{[mEk\textcorner]} \emph{adj.}~(\emph{a}.) empty.   
\B{Nge\-yä tsngal lum\-pe lu mek?} Why is your cup empty?~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\emph{b.~fig.}) `empty', having no valuable content. 
\B{mek\-a sä\-fpìl}, an empty\slash dumb idea~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}; 
\B{mek\-a sä\-pll\-txe\-vi}, an insipid\slash thoughtless comment~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}; 
\B{sä\-mok a\-mek}, a useless suggestion~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b};
\B{\mbox{leioae} a\-mek}, feigned respect.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
(\ding{214}~\HL{teya}{\B{te\-ya}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{mektseng}{\B{mek\-tseng}}, gap, breach.  
\on{2}.~\HL{momek}{\B{mo\-mek}}, hollow (not solid).

% mekong
\hi
\B{\cp{me}kong}~:~\HL{'ekong}{\B{'ekong}}

% mekre
\hi
\HT{mekre}{\B{\cp{mek}re}} \I{["mEk\textcorner.RE]} \emph{n.} supplies. 

% mektseng
\hi
\HT{mektseng}{\B{\cp{mek}tseng}} \I{["mEk\textcorner.\t{ts}EN]} \emph{n.} gap, breach.
\B{Lu mek\-tseng a tsun fpxi\-vä\-kìm hì\-'ang tsaw\-ìlä.}  There's a gap through which insects can get in.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Fì\-po\-stì mek\-tseng\-ur te\-ya si.} This post fills a gap (in our understanding of Na'vi syntax).~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
(\ding{43}~\HL{mek}{\B{mek}}, \HL{tseng}{\B{tseng}})

% mele'wll
\hi
\B{mele'wll}~:~\HL{'ele'wll}{\B{'ele'wll}}

% melo
\hi
\HT{melo}{\B{\cp{me}lo}} \I{["mE.lo]} \emph{adv.} twice.
\B{Tsa\-lì'\-fya\-vi\-ri fko ze\-ne si\-var lì\-'ut alu fì\-'u me\-lo.} One must use the word \emph{fì'u} twice in that expression.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-677}{b}
(\ding{43}~\HL{alo}{\B{alo}})

% memyu
\hi
\B{\cp{mem}yu}~:~\HL{'emyu}{\B{'emyu}}

% men
\hi
\B{men}~:~\HL{'en}{\B{'en}}

% menga
\hi
\HT{menga}{\B{me\cp{nga}}} \I{[mE."Na]} \emph{pn.} (\emph{gen.} \B{me\-\cp{nge}\-yä}) you (two), the two of you. (2\textsuperscript{nd} person dual). 
\B{Me\-nga 'e\-fu väng srak?} Are you two thirsty?~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Me\-nga lu kar\-yu.} You two are teachers.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
\B{Ay\-oel me\-nga\-ti ka\-me\-i\-e.} We \textsc{see} you both.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Me\-nga\-ru tì\-yawr.} You are both right.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/comment-page-1/\#comment-347}{b}
\B{Lor\-a fì\-tì\-kang\-kem me\-nge\-yä meuia lu\-yu aw\-nga\-ru nì\-wotx.} This beautiful work of theirs honors us very much.~\LINK{http://naviteri.org/2010/12/beautiful-christmas-carols-sung-in-navi/}{b}
(\ding{43}~\HL{nga}{\B{nga}})

% mengenga
\hi
\HT{mengenga}{\B{menge\cp{nga}}} \I{[mE.NE."Na]} \emph{pn.} (\emph{gen.} \B{me\-nge\-\cp{nge}\-yä}) you (two), the two of you. (2\textsuperscript{nd} person dual; deferential\slash ceremonial form).
(\ding{43}~\HL{ngenga}{\B{nge\-nga}})

% meoauniaea
\hi
\HT{meoauniaea}{\B{meoaunia\cp{e}a}} \I{[""mE.o.a.u.ni.a."{}E.a]} \emph{n.} harmony, living at one with nature.
\B{New oe ti\-vìng ay\-ngar lì\-'ut a tì\-'e\-fu\-mì oe\-yä lu lor fra\-to mì lì'\-fya le\-Na'\-vi: me\-o\-a\-u\-ni\-a\-e\-a.} I want to present to you the word that, in my opinion, is the most beautiful in the Na’vi language: \emph{me\-o\-a\-u\-ni\-a\-e\-a}.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}

% merki
\hi
\HT{merki}{\B{\cp{mer}ki}} \I{["mER.ki]} \emph{n.} ground rack (\emph{for smoking meats}).

% mesrram
\hi
\HT{mesrram}{\B{mesrr\cp{am}}} \I{[mE.s\s{r}."am]} \emph{n.; adv.} the day before yesterday, two days ago. 
\B{Ìlä Fey\-ral, mun\-txa so\-li Ra\-lu sì Ne\-wey nì\-wan me\-srr\-am.} According to Peyral, Ralu and Newey were secretly married the day before yesterday.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{trram}{\B{trr\-am}})

% mesrray
\hi
\HT{mesrray}{\B{mesrr\cp{ay}}} \I{[mE.s\s{r}."aj]} \emph{n.; adv.} the day after tomorrow, two days from now.
\B{Oer lu txan\-a yew\-la a ke tsun nga oe\-hu ki\-vä\-teng me\-srr\-ay.} I'm very disappointed you can't hang out with me the day after tomorrow.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{trray}{\B{trr\-ay}})

% metkayina
\hi
\B{Metka\cp{yi}na} \I{[mEt\textcorner.ka."ji.na]} \emph{n. clan name}, one clan of the reef people.

% metnaw
\hi
\B{\cp{met}naw}~:~\HL{'etnaw}{\B{'etnaw}}

% meuia / si
\hi
\HT{meuia}{\B{me\cp{u}ia}} \I{[mE."{}u.i.a]} \textsc{i}. \emph{n.} honor.
\B{Fì\-me\-u\-i\-a\-ri sei\-yi moe ira\-yo nì\-txan, ma smuk.} We both thank you very much for this honor, siblings.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}
(\emph{a.}) \B{Fay\-sul\-fä\-tu\-ä tì\-kang\-kem o\-he\-ru me\-u\-i\-a lu\-yu nì\-ngay.} The work of these experts truly honors me.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{b.~idiom}) \B{Oeru me\-u\-i\-a.} ``You're welcome.\slash The honor is mine.'' (in response to thanks) \\  
\textsc{ii}. \B{me\cp{u}ia si} \I{[me."{}u.i.a si]} \emph{vin.} honor. 
\B{Nge\-yä fay\-lì\-'u a\-tì\-tstun\-wi\-nga' oe\-ru meuia so\-li nì\-ngay.} These kind words of yours have honored me greatly.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
(\ding{214}~\HL{kemuia}{\B{ke\-mu\-ia}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{meuianga'}{\B{me\-uia\-nga'}}, honorable. 
\on{2}.~\HL{kemuia}{\B{ke\-mu\-ia}}, dishonor.

% meuianga'
\hi
\HT{meuianga'}{\B{me\cp{u}ianga'}} \I{[mE."{}u.i.a.NaP]} \emph{adj.} honorable. 
(\ding{43}~\HL{meuia}{\B{me\-uia}})

% mevan
\hi
\B{\cp{me}van}~:~\HL{'evan}{\B{'evan}}

% meve
\hi
\B{\cp{me}ve}~:~\HL{'eve}{\B{'eve}}

% meveng
\hi
\B{\cp{me}veng}~:~\HL{'eveng}{\B{'eveng}}

% mevi
\hi
\B{\cp{me}vi}~:~\HL{'evi}{\B{'evi}}

% mevol
\hi
\HT{mevol}{\B{\cp{me}vol}} \I{["mE.vol]} \emph{num., adj.} \on{20} (octal), \on{16} (decimal).
\B{Po\-el haw\-re'\-tsyìp\-it ko\-la\-nom yoa txo\-lar a\-me\-vol.} She bought a little cap for \$\on{16}.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}

% meyam
\hi
\HT{meyam}{\B{me\cp{yam}}} \I{[mE."jam]} \emph{vtr.} hug, embrace, hold in one's arms.   
\B{Ma sa'\-nu, oe txo\-pu si. Me\-yam oe\-ti!} Mommy, I'm scared. Hold me!~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} 
\B{Me\-fo fì\-tsap mä\-po\-le\-yam teng\-krr tsngaw\-vìk.} The two of them hugged each other and wept.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{sAmyam}{\B{säm\-yam}}, hug, embrace.

% meylan
\hi
\B{\cp{mey}lan}~:~\HL{'eylan}{\B{'eylan}}

% meylanay
\hi
\B{meyla\cp{nay}}~:~\HL{'eylanay}{\B{'eylanay}}

% meylyong
\hi
\B{\cp{meyl}yong}~:~\HL{'eylyong}{\B{'eylyong}}

% meyp
\hi
\HT{meyp}{\B{meyp}} \I{[mEjp\textcorner]} \emph{adj.} (\emph{a.~of strength}) weak. 
(\emph{b.~fig.}) \B{Tsa\-sä\-pll\-txe\-ri sä\-wì\-ngay a to\-lìng ngal lu meyp.} The proof you gave of that statement is weak.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImeyp}{\B{tì\-meyp}}, weakness. 
\on{2}.~\HL{nImeyp}{\B{nì\-meyp}}, weakly, loosely. 
\on{3}.~\HL{tompameyp}{\B{tom\-pa\-meyp}}, drizzle. 
\on{4}.~\HL{hermeyp}{\B{her\-meyp}}, snow flurry. 
\on{5}.~\HL{meyptu}{\B{meyp\-tu}}, weakling.

% meyptu
\hi 
\HT{meyptu}{\B{\cp{meyp}tu}} \I{["mEjp\textcorner.tu]} \emph{n.} weakling, \emph{a person of weak physique or character}. 
(\ding{43}~\HL{meyp}{\B{meyp}}, \HL{-tu}{\B{-tu}}, \ding{214}~\HL{txurtu}{\B{txur\-tu}})

% mi
\hi
\HT{mi}{\B{mi}} \I{[mi]} \emph{adv.} yet, still, as before.
\B{Pol trr\-am ye\-rik\-it tar\-ma\-ron.--Fì\-trr mi te\-ra\-ron.} He was hunting hexapede yesterday.--He's still hunting today.~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln}
\B{Tsa\-tsko lu spu\-win ul\-te ke lu mi txur, slä oe\-ri txan\-wa\-we le\-iu.} That bow is old and no longer strong, but it means a lot to me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 

% miklor
\hi
\HT{miklor}{\B{mik\cp{lor}}} \I{[mik\textcorner."loR]}  \emph{adj.} pleasant sounding, beautiful sounding. 
\B{Lu por mok\-ri a\-mik\-lor, slä lo\-re\-yu 'aw\-nam\-pi lu.} She has a beautiful voice, but she's extremely shy.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{214}~\HL{mikvA'}{\B{mik\-vä'}}; \ding{43}~\HL{mikyun}{\B{mik\-yun}}, \HL{lor}{\B{lor}})

% miktsang
\hi
\HT{miktsang}{\B{\cp{mik}tsang}} \I{["mik\textcorner.\t{ts}aN]} \emph{n.} earring.
\B{Pey\-ral\-ä mik\-tsang\-it a\-mip ngal tso\-le\-ri srak?} Did you notice Pey\-ral's new earring?~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Se\-vin\-a tsa\-'e\-ve\-ru a\-hì\-'i me\-sa'\-sem ioi so\-li fa mik\-tsang.} The parents put earrings on that pretty little girl.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
(\ding{43}~\HL{mikyun}{\B{mik\-yun}}, \HL{tsang}{\B{tsang}})

% mikvä'
\hi
\HT{mikvA'}{\B{mik\cp{vä'}}} \I{[mik\textcorner."v\ae{}P]}  \emph{adj.} bad\hyp{}sounding.
(\ding{214}~\HL{miklor}{\B{mik\-lor}}; \ding{43}~\HL{mikyun}{\B{mik\-yun}}, \HL{vA'}{\B{vä'}})

% mikyun
\hi
\HT{mikyun}{\B{\cp{mik}yun}} \I{["mik\textcorner.jun]} \emph{or} \I{[.jUn]} \textsc{i}.~\emph{n.} ear. 
\B{Lu Ney\-ti\-ri\-ru 'aw\-a mik\-yun a\-raw\-nipx nì\-'aw.} Neytiri has only one pierced ear.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
\B{'om na mik\-yun}, the purplish color on the inside of a Na'vi ear.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Tskxe\-keng\-tsyìp a Mik\-yun\-fpi}, A Little Listening Exercise.~\LINK{http://naviteri.org/2015/09/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise-2/}{b}
(\emph{idiom}) \B{Kxe\-tse sì mik\-yun kop pll\-txe.} Body language is important [lit.: the tail and ear also speak].~\LINK{http://forum.learnnavi.org/language-updates/grid-tidbits/}{ln} \\
\textsc{ii}. \HT{tIng mikyun}{\B{tìng \cp{mik}yun}} \I{[tIN "mik\textcorner.jun]} \emph{or coll.} \I{[tI."mik\textcorner.jun]} \emph{vin.} (\texttt{+}\emph{control}) listen, listen to. 
\B{Tìng mik\-yun fte tsli\-vam.} Listen in order to understand.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Oe\-ru tìng mik\-yun, ma Ra\-lu.} Listen to me, Ralu.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Po pa\-mll\-txe a krr, fra\-po tar\-mìng mik\-yun nì\-pxi, ta\-lu\-na mok\-ri lu sä\-tsì\-syì\-tsyìp.} When he spoke, everyone listened intently, because his voice was a tiny whisper.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{miktsang}{\B{mik\-tsang}}, earring. 
\on{2}.~\HL{miklor}{\B{mik\-lor}}, pleasant\slash beautiful sounding. 
\on{3}.~\HL{mikvA'}{\B{mik\-vä'}}, bad\hyp{}sounding.

% mimu
\hi 
\HT{mimu}{\B{\cp{mi}mu}} \I{["mi.mu]} \emph{vtr.} develop. 
\B{Ze\-ne aw\-nga mi\-vi\-mu tì\-hawl\-it a lä\-txayn kxu\-tut.} We have to develop a plan to defeat the enemy.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tImimu}{\B{tì\-mi\-mu}}, development.

% mip
\hi
\HT{mip}{\B{mip}} \I{[mip\textcorner]} \emph{adj.} new.
\B{Ke lä\-ngu fì\-po\-stì\-mì ke\-'u a lu mip.} Unfortunately, there is nothing in this post that is new.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
\B{Mip\-a ay\-lì'\-fya\-vi, mip\-a ho\-ren}, New expressions, new rules.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Zi\-va\-'u nì\-mun ye'\-rìn \ldots tsa\-krr ra\-yun ay\-ngal ay\-u\-pxa\-ret a\-mip.} Come again soon \ldots then you'll find new messages.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{a. idiom}) \B{Mip\-a zì\-sìt le\-fpom nga\-ru!} Happy New Year to you!~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{214}~\HL{lal}{\B{lal}})

% mì
\hi
\HT{mI}{\B{mì}} \I{[mI]} \emph{adp.}\texttt{+} (\emph{a.~local})  in, on. 
\B{Mì sray a\-pxa kel\-ku si soaia\-hu.} (He) lives in a great village with his family.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Nge\-yä tì\-rol\-tsyìp\-ìri, oe\-yä ay\-syu\-lang set mì fay kll\-kxe\-rem.} As for your little song, my flowers are standing in water now.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-296}{b}
\B{Ul\-txa a mì na'\-rìng}, Meeting in the forest.~\LINK{http://naviteri.org/2010/12/catching-up/}{b}
\B{Pam\-tse\-ol ngop ay\-re\-nut | Mì ron\-sem\-ä tì\-fnu.} The music creates patterns | In the silence of the mind.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Ay\-lì\-'u na ay\-skxe mì te'\-lan.} The words (are) like stones in (my) heart.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{wiki}
\B{Krro lu 'Ìng\-lì\-sì txur nì\-hawng mì re\-'o.} Sometimes English is too strong in my head.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b}
\B{Wo\-leyn Ìstawl yey\-fyat mì hll\-te fte oeyk\-ti\-vìng fra\-po\-ru tì\-hawl\-te\-ri sne\-yä.} Ìstaw drew a line on the ground to explain his plan to everyone.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Oe\-ri skxir a mì syokx tì\-sraw se\-ngi.} The wound on my hand hurts.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Txur\-tel\-mì fo fta so\-li fte\-ke ka tsyokx fwi\-vi.} They tied a knot in the rope so it wouldn't slip through their hands.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{b. local, fig.}) in. 
\B{Law lu oe\-ru fwa nga mì rel\-tseo no\-lu\-me nì\-txan!} I'm sure that you've learned a lot in music!~\LINK{http://forum.learnnavi.org/language-updates/art-related-vocab/}{ln}
\B{Ta\-fral ho\-lawl ay\-oel ay\-nga\-fpi 'a\-'aw\-a tì\-päng\-kxo\-tsyìp\-it a tsun fko si\-var mì sì\-rey le\-trr\-trr.} That's why we prepared several little dialogues for you that one can use in ordinary life.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{Sìl\-pey oe, ay\-nga\-ri fra\-'u a mì sì\-rey vi\-var tsan\-'i\-vul fra\-fya, ma ey\-lan.} I hope, everything in your life continues to improve for you in every way, friends.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{New oe ti\-vìng ay\-ngar lì\-'ut a tì\-'e\-fu\-mì oe\-yä lu lor fra\-to mì lì'\-fya le\-Na'\-vi: me\-o\-a\-u\-ni\-a\-e\-a.} I want to present to you the word that, in my opinion, is the most beautiful in the Na’vi language: \emph{me\-o\-a\-u\-ni\-a\-e\-a}.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Sra\-ke tsun nga ri\-vun fì\-tì\-fkey\-tok\-mì a tì\-'i\-put?} Can you find the humor in this situation?~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{c.~local, planetary surface}) on. 
\B{mì 'Rr\-ta\slash 'Rr\-ta\-mì}, on Earth.
\B{mì Ey\-wa\-'e\-veng\slash Ey\-wa\-'e\-veng\-mì}, on Pandora.
(\emph{d.~temporal, for range of time}) 
\B{Ke lo\-lu oer nì\-ke\-ftxo mì sok\-a srr ay\-skxom le\-tam fte lì'\-fya\-ri aw\-nge\-yä tì\-kang\-kem si\-vi.} In recent days I have not had sufficient opportunity to work on our language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Fì\-trr\-mì le\-tsran\-ten}, on this important day.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Alo a\-hay oe za\-sya\-'u mì re\-won.} Next time, I'll come in the morning.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{mì ftaw\-nem\-krr\slash zu\-saw\-krr}, in the past\slash future.
(\emph{e.~idiom})
\B{tì\-'e\-fu\-mì oe\-yä}, in my opinion.
\B{tì\-o\-mum\-mì oe\-yä} \I{[tI."{}o.mu.mI "wE.j\ae]}, to my knowledge, as far as I know.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Ay\-te\-le a nge\-yä ha\-pxì\-mì ki\-fkey\-ä lu fya\-pe?} What's new in your part of the world?~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
(\emph{f.~proverb}) 
\B{Txìm a\-'aw ke tsun hi\-veyn mì tal me\-fa'\-li\-yä.} You can't sit on the fence; you need to decide [lit.: One butt can't sit on the backs of two direhorses].~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{mIfa}{\B{mìfa}}, inside.   
\on{2}.~\HL{mIftxele}{\B{mì\-ftxe\-le}}, by the way.   
\on{3}.~\HL{mIkam}{\B{mì\-kam}}, between. 
\on{4}.~\HL{mIso}{\B{mì\-so}}, away.

% mìfa
\hi
\HT{mIfa}{\B{\cp{mì}fa}} \I{["mI.fa]} \emph{or} \B{mì\cp{fa}} \I{[mI."fa]} \emph{n., adv.} inside.
\B{Fra\-po ne mì\-fa! Le\-rok yrr\-ap a\-pxa!} Everybody inside! A big storm is approaching!~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Ke za\-syup lì\-'O\-na ne kxu\-tu a mì\-fa fu a wrr\-pa.} The l'Ona will not perish to the enemy within or the enemy without.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{g\slash ln} 
(\ding{43}~\HL{mI}{\B{mì}}, \HL{pa'o}{\B{pa'o}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nemfa}{\B{nem\-fa}}, into. 
\on{2}.~\HL{ftumfa}{\B{ftum\-fa}}, out of. 
\on{3}.~\HL{tamIfa}{\B{ta\-mì\-fa}}, internal. 

% mìftxele
\hi
\HT{mIftxele}{\B{mì\cp{ftxe}le}} \I{[mI."ft'E.lE]} \emph{adv.} by the way, in this regard, related to this matter.
\B{Mì\-ftxe\-le po\-ri lu oer le\-tsran\-ten\-a fmawn a new pi\-veng ngar.} By the way, I have some important news I want to tell you about him.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Sko Sa\-hìk ke tsun oe mì\-ftxe\-le tsngi\-vaw\-vìk.} As Tsahik, I cannot weep over this matter.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
(\ding{43}~\HL{nIvingkap}{\B{nì\-ving\-kap}})

% mìkam
\hi
\HT{mIkam}{\B{mì\cp{kam}}} \I{[mI."kam]} \emph{adp.} between. 
\B{Ta\-lu\-na fì\-tì\-'ey\-lan lu pum a mì\-kam tu\-tan sì syak\-syuk.} Because this friendship is between a man and a Prolemuris.~\LINK{http://naviteri.org/2020/04/awa-tskxekengtsyip-a-mikyunfpi-niul-one-more-little-listening-exercise/}{b}
(\ding{43}~\HL{mI}{\B{mì}}, \HL{kxam}{\B{kxam}}) 

% mìn
\hi
\HT{mIn}{\B{mìn}} \I{[mIn]} \emph{vin.} turn.
\B{Mìn ne ftär\-pa.} Turn to the left.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln}
\B{Kun\-sìp\-ìri txan\-a tì\-meyp lu tsyal a mìn.} The gunship's main weakness is the rotor system.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln}  \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{mInyu}{\B{mìn\-yu}}, turner; twisted lily.

% mìnyu
\hi
\HT{mInyu}{\B{\cp{mìn}yu}} \I{["mIn.ju]} \emph{n.} turner; \ding{96} twisted lily, \emph{pseudolilium contortum}.
(\ding{43}~\HL{mIn}{\B{mìn}})

% mìso
\hi
\HT{mIso}{\B{mì\cp{so}}} \I{[mI."{}so]} \emph{adv.} away (\emph{positional}). 
% \B{'Ì\-'awn mì\-so.} Stay away.
(\ding{43}~\HL{neto}{\B{ne\-to}})

% mll'an
\hi
\HT{mll'an}{\B{mll\cp{'an}}} \I{[m\s{l}."Pan]} (\textsc{perf.}~\B{mol'an}) \emph{vtr.} accept.   
\B{Mun\-txa\-tu\-ri So\-rewn\-ti ke tsun oe mi\-vll\-'an.} I can't accept Sorewn as my spouse.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Stxe\-nu\-to\-lìng oel fu\-ta lì'\-fyat le\-Na'\-vi poe\-ru ki\-var. Mol\-'an nì\-prr\-te'.} I offered to teach her Na'vi. She accepted gladly.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Fu\-ta nga\-ta tel pxen\-yoa srät, mll\-'ei\-an oel.} I'm happy to take cloth from you for finished garments.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b} 
(\ding{214}~\HL{tsyAr}{\B{tsyär}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImll'an}{\B{tì\-mll\-'an}}, acceptance.

% mllte
\hi
\HT{mllte}{\B{mll\cp{te}}} \I{[m\s{l}."tE]} (\textsc{perf.}~\B{mol\-te}) \emph{vin.} agree (\HL{hu}{\B{hu}}, with; \emph{with the thing agreed upon in the topic}). 
\B{Nga\-hu mll\-te oe.} I agree with you.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Txam\-pay lìm nì\-hawng a tsa\-ri mll\-te oe hu sem\-pul.} I agree with father that the ocean is too far away.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}

% mo
\hi
\HT{mo}{\B{mo}} \I{[mo]} \emph{n.} space, hollow, enclosed open area.
\B{Tok oel lora tsamoti a mì na'rìng a krr, 'efu mawey sì nitram.} When I'm in that beautiful hollow in the forest, I feel calm and happy.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
(i.) \HT{mo a hahaw}{\B{mo a \cp{ha}haw}}, bedroom, sleeping area. \\
(ii.) \HT{mo a yom}{\B{mo a yom}}, dining room, dining area. \\
(iii.) \HT{mo letrrtrr}{\B{mo le\cp{trr}trr}}, living room. \\  
(iv.) \HT{mo a fngA'}{\B{mo a fngä'}}, restroom (\emph{an enclosed private structure or room}). \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{snomo}{\B{sno\-mo}}, private space that one can retreat to.  
\on{2}.~\HL{momek}{\B{mo\-mek}}, hollow, not solid.

% mo'ara
\hi
\HT{mo'ara}{\B{Mo\cp{'a}ra}} \I{[mo."Pa.Ra]} (\emph{a}.) \emph{n.} gathering place (\emph{generally for some specific purpose}).
(\emph{b}.) \emph{name of the Disney theme park}.

% mo'at
\hi
\HT{mo'at}{\B{\cp{Mo}'at}} \I{["mo.Pat\textcorner]} \emph{n., female name}.
\B{Mo\-'at \mbox{alu} Tsa\-hìk lu O\-ma\-ti\-ka\-ya\-ä le\-'awa ha\-pxì\-tu a \mbox{ioi} sä\-pi fa 'a\-re.} Mo'at, the Tsahik, is the only member of the Omatikaya who wears a poncho.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Spaw oel fu\-ta Mo\-'at\-ìl tso\-le\-'a Ney\-ti\-rit.} I believe, Mo'at saw Neytiri.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/}{b}

% moe
\hi
\HT{moe}{\B{\cp{mo}e}} \I{["mo.E]} \emph{pn.} we (two, not you), the two of us (1\textsuperscript{st} person dual exclusive).
\B{Moe sìl\-pey nì\-teng tsnì pxo\-eng tsä\-pì\-ye\-ve\-'a fì\-tsap ye'\-rìn nì\-mun.} We also hope that we will see each other again soon.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/comment-page-1/\#comment-2483}{b}
\B{Moe smon (moe\-ru) fì\-tsap.} We know each other.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Teng\-krr pa\-lu\-lu\-kan moe\-ne kxll sar\-mi, pol\-txe Ney\-ti\-ril ay\-lì\-'ut a fra\-krr 'ok se\-yä la\-yu oer.} As the thanator was charging towards the two of us, Neytiri said something I will always remember.~\LINK{http://thereadinginpublicproject.blogspot.de/2010/05/paul-frommer.html}{pr}
(\ding{43}~\HL{oe}{\B{oe}}, \HL{me+}{\B{me}\texttt{+}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{mohe}{mo\-he}, we two (deferential).

% mohe
\hi
\HT{mohe}{\B{\cp{mo}he}} \I{["mo.hE]} \emph{pn.} we (two, not you), the two of us (1\textsuperscript{st} person dual exclusive; deferential\slash ceremonial form). 
(\ding{43}~\HL{ohe}{\B{ohe}}, \HL{me+}{\B{me}\texttt{+}})

% mok
\hi
\HT{mok}{\B{mok}} \I{[mok\textcorner]} \emph{vtr.} suggest.
\B{Nge\-yä tì\-pawm\-ìri mok oel fu\-ta ngal si\-var kem\-lì\-'u\-vit alu ‹irv›.} As for your question, I suggest that you use the infix ‹irv›.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1951}{b}
\B{Tsa\-tsmu\-kel alu Ri\-ni mo\-lok fu\-ta oe ngar mu\-wä\-pi\-vìn\-txu.} Sister Rini over there suggested that I introduce myself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImok}{\B{tì\-mok}}, suggestion, suggesting.   
\on{2}.~\HL{sAmok}{\B{sä\-mok}}, suggestion. 

% mokri
\hi
\HT{mokri}{\B{\cp{mok}ri}} \I{["mok\textcorner.Ri]} \emph{n.} voice. 
\B{Po pa\-mll\-txe a krr, fra\-po tar\-mìng mik\-yun nì\-pxi, ta\-lu\-na mok\-ri lu sä\-tsì\-syì\-tsyìp.} When he spoke, everyone listened intently, because his voice was a tiny whisper.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Fu\-la nga\-ri oel mok\-rit sta\-yawm, oe\-ti nit\-ram sley\-ku nì\-txan.} It makes me very happy that I will hear your voice.~\LINK{http://forum.learnnavi.org/language-updates/fula-short-form-of-fiul-a-and-tsal-oeti-steyki-nitram/}{ln}
(i.) \B{Ut\-ral Ay\-mok\-ri\-yä}, Tree of Voices (\emph{important spiritual site}).
\B{Ke lu kaw\-tu a nul\-ni\-vew oe po\-hu ti\-re\-a\-pi\-väng\-kxo äo Ut\-ral Ay\-mok\-ri\-yä.} There's nobody I'd rather commune with under the Tree of Voices.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{wiki} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{utral aymokriyA}{\B{Ut\-ral Ay\-mok\-ri\-yä}}, Tree of Voices. 
\on{2}.~\HL{spulmokri}{\B{spul\-mok\-ri}}, telephone. 
\on{3}.~\HL{kakmokri}{\B{kak\-mok\-ri}}, mute.

% momek
\hi
\HT{momek}{\B{\cp{mo}mek}} \I{["mo.mEk\textcorner]} \emph{adj.} hollow (\emph{not solid}).
\B{Tsa\-ta\-ngek\-ìri pam fkan mo\-mek.} That treetrunk sounds hollow.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{mo}{\B{mo}}, \HL{mek}{\B{mek}})

% mong
\hi
\HT{mong}{\B{mong}} \I{[moN]} \emph{vtr.} depend on, rely on, trust for protection.
\B{Mong prr\-nen\-ìl fu\-ta sa'\-nul ve\-rewng nì\-tut.} The infant depends on her Mommy to look after her constantly.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Tsun nga oet mi\-vong.} You can rely on me.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tsukmong}{\B{tsukmong}}, reliable, dependable.

% mowan
\hi
\HT{mowan}{\B{mo\cp{wan}}} \I{[mo."wan]} \emph{adj.} (\emph{a}.) pleasing, enjoyable (\emph{implies physical or sensual pleasure, often has a sexual connotation}).
\B{Pll\-txe fko san nga\-ru lu mo\-wan Txil\-te ul\-te po\-ru nga.} I hear you like Txilte and vice versa.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Tì\-ru\-sol lu oe\-ru mo\-wan.} Singing is enjoyable to me.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\emph{b}.~\emph{slang}) like. \B{Tì\-tu\-sa\-ron mo\-wan lu oer nì\-ngay.} Hunting really turns me on.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}

% mowar / si
\hi
\HT{mowar}{\B{mo\cp{war}}} \I{[mo."waR]} \textsc{i}. \emph{n.} advice, bit or piece of advice.
\B{Nge\-yä mo\-war txan\-ley oer.} Your advice is invaluable to me.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b}
\B{Ma Ney\-ti\-ri, ay\-oel kin mo\-war\-it nge\-yä.} Neytiri, we need your advice.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} \\
\textsc{ii}. \B{mo\cp{war si}} \I{[mo."waR si]} \emph{vin.} advise, give advice; (\B{tsnì}) advise s.o. to do s.t.
\B{Tsun oe mo\-war si\-vi ngar, slä ke tsun fya\-wi\-vìn\-txu.} I can advise you, but I can't guide you.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
\B{Poe mo\-war so\-li poan\-ur tsnì hi\-vum.} She advised him to leave.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{mowarsiyu}{\B{mo\-war\-si\-yu}}, advisor.

% mowarsiyu
\hi
\HT{mowarsiyu}{\B{mo\cp{war}siyu}} \I{[mo."waR.si.ju]} \emph{n.} advisor.
\B{Lu fra\-eyk\-tan\-ur a\-sìl\-tsan txan\-tslu\-sam\-a ay\-mo\-war\-si\-yu.} Every good leader has wise advisors.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
\B{Ay\-sä\-mok\-ìri a\-txan\-tsan, oe\-yä ay\-mo\-war\-si\-yu\-ru ira\-yo nì\-txan!} I thank my advisors very much for the excellent suggestions!~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{mowar}{\B{mo\-war}})

% mrr
\hi
\HT{mrr}{\B{mrr}} \I{[m\s{r}]} \emph{num., adj.} five.
\B{Alo a\-mrr po\-an po\-lawm, slä fra\-lo poe pol\-txe san ke\-he.} He asked five times but each time she said `no'.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{mrra riti}, five stingbats.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{mrrtrr}{\B{mrr\-trr}}, (\on{5}\hyp{}day) week. 
\on{2}.~\HL{vospxImrr}{\B{Vo\-spxì\-mrr}}, May.

% mrr-
\hi
\HT{mrr-}{\B{mrr-}} \I{[m\s{r}.]} \emph{num.~pref.} five (\emph{base to form higher numbers or fractions}). \B{mrr\-vol}, \on{40} (decimal), \on{50} (octal); \B{mrr\-pxì}, one-fifth, $\frac{1}{5}$; \B{mrr\-pxì a\-tsìng}, four-fifth, $\frac{4}{5}$.

% mrrpxì
\hi
\HT{mrrpxI}{\B{mrr\cp{pxì}}} \I{[m\s{r}."p'I]} \emph{n.} one-fifth, $\frac{1}{5}$.
(\ding{43}~\HL{mrr-}{\B{mrr-}}, \HL{hapxI}{\B{ha\-pxì}})

% mrrtrr / mrrtrram / mrrtrray
\hi
\HT{mrrtrr}{\B{\cp{mrr}trr}} \I{["m\s{r}.t\s{r}]} \emph{n., adv.} (\on{5}\hyp{}day) workweek.
\B{Kä\-teng oe hu ey\-lan Per\-lin\-mì a mrr\-trr lu tsyeym a ke tsun tswi\-va' kaw\-krr.} The (\on{5}-day) week I spent with my friends in Berlin is a treasure that I’ll never forget.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b} \\ 
(i.)~\B{mrr\-trr\-\cp{am}} \I{[m\s{r}.t\s{r}."{}am]} \emph{n., adv.} last (\on{5}\hyp{}day) workweek. \\
(ii.)~\B{mrr\-trr\-\cp{ay}} \I{[m\s{r}.t\s{r}"{}aj]} \emph{n., adv.} next (\on{5}\hyp{}day) workweek.

% mrrve
\hi
\HT{mrrve}{\B{\cp{mrr}ve}} \I{["m\s{r}.vE]} \emph{ord.~num., adj.} fifth. 

% mrrvol
\hi
\HT{mrrvol}{\B{\cp{mrr}vol}} \I{["m\s{r}.vol]} \emph{num., adj.} \on{40} (decimal), \on{50} (octal). 

% mu'ni
\hi 
\HT{mu'ni}{\B{\cp{mu'}ni}} \I{["muP.ni]} \emph{vtr.} accomplish, achieve (\emph{for achievements that are in some way significant}). 
\B{Krr\-ka tì\-rey a\-yol, pol mo\-lu'\-ni pxay\-a ay\-ut a tsran\-ten.} During her short life, she accomplished many important things.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} 
(\ding{43}~\HL{hasey si}{\B{ha\-sey si}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImu'ni}{\B{tì\-mu'\-ni}}, achievement, accomplishment.
\on{2}.~\HL{mu'nitkan}{\B{mu'\-nit\-kan}}, effective.

% mu'nitkan
\hi 
\HT{mu'nitkan}{\B{\cp{mu'}nitkan}} \I{["muP.nit\textcorner.kan]} \emph{or} \I{["mUP.]} \emph{adj.} effective. 
\B{Tsa\-ye\-rik\-ìl mu'\-nit\-kan\-a wan\-tseng\-it ro\-lun; ke tsun fko pot tsi\-ve\-'a kaw\-'it.} That hexapede has found an effective hiding place. He can't be seen at all.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
(\ding{43}~\HL{mu'ni}{\B{mu'\-ni}}, \HL{tIkan}{\B{tì\-kan}}; \ding{214}~\HL{kemu'nitkan}{\B{ke\-mu'\-nit\-kan}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImu'nitkan}{\B{tì\-mu'\-nit\-kan}}, effectiveness.
\on{2}.~\HL{nImu'nitkan}{\B{nì\-mu'\-nit\-kan}}, effectively.

% muiä
\hi
\HT{muiA}{\B{mu\cp{i}ä}} \I{[mu."{}i.\ae]} \emph{adj.} (\emph{a}.) proper, fair, right, justified. 
\B{Tsay\-lì\-'u ke lu muiä!} Those words are improper!~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
\B{Fra\-trr 'e\-rul a fay\-sä\-leym\-kem a\-yll muiä lu nì\-wotx.} The communal protests that are increasing every day are completely fair.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
\B{Nga sìl\-mi a tsa\-kem ke lu vo\-ìk a\-mu\-iä!} What you just did was not proper behavior!~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} 
(\emph{b.~proverb}) \B{Kem a\-muiä, kum a\-fe'.} Proper action, bad result.\slash Well, their heart was in the right place (\emph{something that should have turned out well didn't}).~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImwiA}{\B{tìm\-wiä}}, fairness, justice.

% mulpxar
\hi
\HT{mulpxar}{\B{mul\cp{pxar}}} \I{[mul."p'aR]} \emph{or} \I{[mUl.]} \emph{n.}, \ding{96} roosterhead plant, \emph{alectophyllum molle}.

% -mun
\hi
\HT{-mun}{\B{-mun}} \I{[.mun]} \emph{or} \I{[.mUn]} \emph{num. suf.} two (\emph{to form higher numbers}). \B{vo\-mun}, \on{10} (decimal), \on{12} (octal). \B{me\-vo\-mun}, \on{18} (decimal), \on{22} (octal); etc. 
(\ding{43}~\HL{mune}{\B{mu\-ne}})

% mun'i
\hi
\HT{mun'i}{\B{mun\cp{'i}}} \I{[mun."Pi]} \emph{or} \I{[mUn.]} (\textsc{rn}: \B{mùn\cp{'i}} \I{[mUn."Pi]}) \emph{vtr.} cut. \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{pxImun'i}{\B{pxì\-mun\-'i}}, divide, cut into parts. 
\on{2}.~\HL{syImawnun'i}{\B{syì\-maw\-nun\-'i}}, cut tie.

% mune
\hi
\HT{mune}{\B{\cp{mu}ne}} \I{["mu.nE]} \emph{num., adj.} two. (\emph{base}: \ding{43}~\B{me}\texttt{+}, \emph{rest}: \ding{43}~\HL{-mun}{\B{-mun}}, \emph{ordinal}: \ding{43}~\HL{muve}{\B{mu\-ve}})  \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nImun}{\B{nì\-mun}}, again. 
\on{2}.~\HL{muntrr}{\B{mun\-trr}}, weekend. 
\on{3}.~\HL{munsna}{\B{mun\-sna}}, pair.

% munsna
\hi
\HT{munsna}{\B{\cp{mun}sna}} \I{["mun.sna]} \emph{or} \I{["mUn.]} \emph{n.} pair.
\B{Hawn\-ven\-ìri lu oe\-ru mun\-sna a\-mrr.} I have five pairs of shoes.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Hu\-fwa me\-fo le\-ru mun\-txa\-tu txan\-krr, mi lu mun\-snar ho\-na tì\-flrr a na pum me\-yaw\-ne\-tu\-ä a\-mip nì\-wotx.} Although the two of them have been mates a long time, they still have all the adorable tenderness of new sweethearts.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{mune}{\B{mu\-ne}}, \HL{sna'o}{\B{sna'o}})

% munsna-
\hi
\HT{munsna-}{\B{munsna-}} \I{[mun.sna.]} \emph{or} \I{[mUn.]} \emph{pref.~to mean} pair, couple. 
\B{Tì\-kang\-kem si fa mun\-sna\-tu\-te.} Work in pairs.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Ta\-yel Tse\-nul pxe\-swi\-zaw\-ti yoa mun\-sna\-hawn\-ven.} Tsenu will receive three arrows in exchange for a pair of shoes.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}

% muntrr / muntrram / muntrray
\hi
\HT{muntrr}{\B{\cp{mun}trr}} \I{["mun.t\s{r}]} \emph{or} \I{["mUn.]} \emph{n., adv.} weekend.  \\
(i.) \B{muntrr\cp{am}} \I{[mun.t\s{r}."{}am]} \emph{or} \I{[mUn.]} \emph{n., adv.} last weekend. 
\B{Mun\-trr\-am oe ko\-lä\-teng hu Ra\-lu ul\-te su\-nu oer tsa\-krr.} I spent last weekend with Ralu and had a good time [lit.: that time was pleasant to me].~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} \\
(ii.) \B{muntrr\cp{ay}} \I{[mun.t\s{r}."{}aj]} \emph{or} \I{[mUn.]} \emph{n., adv.} next weekend. 
(\ding{43}~\HL{mune}{\B{mu\-ne}}, \HL{trr}{\B{trr}})

% muntxa / si / slu
\hi
\HT{muntxa}{\B{mun\cp{txa}}} \I{[mun."t'a]} \emph{or} \I{[mUn.]} \textsc{i}.~\emph{adj.} mated. \\
\textsc{ii}.a.~\B{mun\cp{txa} si} \I{[mun."t'a si]} \emph{or} \I{[mUn.]} \emph{vin.} mate (\ding{43}~\HL{hu}{\B{hu}}, with s.o.), marry s.o.
\B{Ìlä Fey\-ral, mun\-txa so\-li Ra\-lu sì Ne\-wey nì\-wan me\-srr\-am.} According to Peyral, Ralu and Newey were secretly married the day before yesterday.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}  
\B{Ne\-wey yaw\-ne lu oer ul\-te new oe mun\-txa si\-vi poe\-hu.} I love Ne\-wey and want to marry her.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{ii}.b.~\B{mun\cp{txa} slu} I{[mun."t'a slu]} \emph{vin.} get married.
\B{Fko pll\-txe\-i\-e san me\-nga mun\-txa slo\-lu sìk!} I'm so happy to hear you got married!~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{muntxatu}{\B{mun\-txa\-tu}}, spouse.  
\on{2}.~\HL{tImuntxa}{\B{tì\-mun\-txa}}, marriage. 

% muntxatan
\hi
\HT{muntxatan}{\B{mun\cp{txa}tan}} \I{[mun."t'a.tan]} \emph{or} \I{[mUn.]} \emph{n.} husband, male spouse. 
(\ding{43}~\HL{muntxatu}{\B{mun\-txa\-tu}})

% muntxate
\hi
\HT{muntxate}{\B{mun\cp{txa}te}} \I{[mun."t'a.tE]} \emph{or} \I{[mUn.]} \emph{n.} wife, female spouse. 
(\ding{43}~\HL{muntxatu}{\B{mun\-txa\-tu}})

% muntxatu
\hi
\HT{muntxatu}{\B{mun\cp{txa}tu}} \I{[mun."t'a.tu]} \emph{or} \I{[mUn.]} \emph{n.} spouse. 
\B{Ma mun\-txa\-tu, oeng yaw\-ne lu (oe\-nga\-ru) fì\-tsap, ke\-fyak?} We love each other, don't we, my spouse?~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Mip\-a me\-mun\-txa\-tu\-ru sey\-kxel sì ni\-tram.} Congratulations to the two newly-weds.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Mun\-txa\-tu\-ri So\-rewn\-ti ke tsun oe mi\-vll\-'an.} I can't accept Sorewn as my spouse.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{muntxa}{\B{mun\-txa}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{muntxatan}{\B{mun\-txa\-tan}}, husband.   
\on{2}.~\HL{muntxate}{\B{mun\-txa\-te}}, wife.

% munge
\hi
\HT{munge}{\B{\cp{mu}nge}} \I{["mu.NE]} (\textsc{rn}: \B{\cp{mù}nge} \I{["mU.NE]}) \emph{vtr.} take, bring. 
\B{Tsun Txil\-te pam\-rel\-it ivi\-nan; ta\-fral puk\-ot a\-nutx mu\-nge fra\-tseng.} Txilte knows how to read; therefore she brings a thick book wherever she goes.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Fu\-ri\-a fì\-lì'\-fya\-vit fkol ta 'Ìng\-lì\-sì mo\-lu\-nge, sìl\-pey oe tsnì ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ke stì\-ye\-vi.} That this expression is brought from English, will hopefully not anger the members of the language group.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{sAmunge}{\B{sä\-mu\-nge}}, transportation device. 
\on{2}.~\HL{zamunge}{\B{za\-mu\-nge}}, bring. 
\on{3}.~\HL{mungwrr}{\B{mung\-wrr}}, except. 
\on{4}.~\HL{mungsye}{\B{mung\-sye}}, inhale.

% mungsye
\hi 
\HT{mungsye}{\B{\cp{mung}sye}} \I{["m$\cdot$uN.sjE]} \emph{or} \I{["m$\cdot$UN.]} (\textsc{rn}: \B{\cp{mùng}sye} \I{["m$\cdot$UN.SE]}) \emph{vin., inf.}\on{11} inhale. 
(\ding{43}~\HL{munge}{\B{mu\-nge}}, \HL{syeha}{\B{sye\-ha}}; \ding{214}~\HL{lonusye}{\B{lo\-nu\-sye}})

% mungwrr
\hi
\HT{mungwrr}{\B{mung\cp{wrr}}} \I{[muN."w\s{r}]} \emph{or} \I{[mUN.]} (\textsc{rn}: \B{mùng\cp{wrr}} \I{[mUN."w\s{r}]}) \emph{adp.} except.
\B{Me\-lì\-'u alu mung\-wrr sì nì\-fkrr ngam\-pam si.} The words \emph{mungwrr} and \emph{nìfkrr} rhyme.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Oe\-ru ke tsun li\-vam ke\-'u lor to Ey\-wa\-'e\-veng\-ä na'\-rìng a lew sä\-po\-li fa prr\-wll, kxawm mung\-wrr fì\-ki\-fkey a lew sä\-pì\-ye\-vi fa fpom sì lì'\-fya le\-Na'\-vi.} Nothing could be more beautiful to me than a Pandoran forest that has covered in moss, except perhaps this world soon covering itself in peaceful well-being and the Na'vi language.~\LINK{http://forum.learnnavi.org/language-updates/prrwll-moss-the-comparative-to-construct/}{ln} 
(\emph{a. with} \HL{fwa}{\B{fwa}}) except that. 
\B{Po\-ru ke po\-leng oel ke\-'ut mung\-wrr fwa Ra\-lul ke tsa\-tse\-nget.} I told her nothing except that Ralu wasn't there.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
(\ding{43}~\HL{munge}{\B{mu\-nge}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tImungwrr}{\B{tì\-mung\-wrr}}, exception. 
\on{2}.~\HL{mungwrrtxo}{\B{mung\-wrr\-txo}}, unless, except if.
\on{3}.~\HL{lemungwrr}{\B{le\-mung\-wrr}}, exceptional.

% mungwrrtxo
\hi 
\HT{mungwrrtxo}{\B{mung\cp{wrr}txo}} \I{[muN."w\s{r}.t'o]} \emph{or} \I{[mUN.]} \emph{or coll.} \I{[muN."w\s{r}.to]} (\textsc{rn}: \B{mùng\cp{wrr}txo} \I{[mUN."w\s{r}.do]}) \emph{conj.} unless, except if. 
\B{Tsak\-tap rä\-'ä si kaw\-krr mung\-wrr\-txo ke li\-vu kea fya\-'o a\-la\-he.} Never use violence unless there is no other way. ~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
(\ding{43}~\HL{mungwrr}{\B{mung\-wrr}}, \HL{txo}{\B{txo}})

% muve
\hi
\HT{muve}{\B{\cp{mu}ve}} \I{["mu.vE]} \emph{ord. num., adj.} second. 
\B{Lì\-'ut a\-mu\-ve oel tswo\-la'.} I forgot the second word.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-677}{b}
\B{Ul\-te mu\-ve\-a sä\-pll\-txe le\-iu eyawr nì\-teng.} And the second statement is also correct, I'm happy to say.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/comment-page-1/\#comment-2614}{b}
(\ding{43}~\HL{mune}{\B{mu\-ne}})

% muwìntxu
\hi
\HT{muwIntxu}{\B{muwìn\cp{txu}}} \I{[mu.w$\cdot$In."t'$\cdot$u]} \emph{vtr., inf.}\on{23} introduce, present (\emph{people or things}).   
\B{Oel mu\-wi\-vìn\-txu nga\-ru oe\-yä ler\-tut.} Let me present my colleague to you (\emph{formal}).\slash \B{Nga\-ru oe\-yä ler\-tut.} This is my colleague (\emph{coll.}).~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-297}{b}
\B{Tsa\-tsmu\-kel alu Ri\-ni mo\-lok fu\-ta oe ngar mu\-wä\-pi\-vìn\-txu.} Sister Rini over there suggested that I introduce myself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{wIntxu}{\B{wìn\-txu}}) \\

\end{multicols}

