\pdfbookmark[0]{N}{N}
\begin{center}
\section*{N \I{[n]} (\emph{NeN})}
\end{center}

\begin{multicols}{2}

% na
\hi
\HT{na}{\B{na}} \I{[na]} \emph{adp.} (\emph{a.}) like, as.   
\B{Ay\-lì\-'u (lu) na ay\-skxe mì te'\-lan.} The words are like stones in my heart.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{a\slash wiki}
\B{Fì\-txon na ton a\-la\-he nì\-wotx pe\-lun ke lu teng?} Why is this night not the same like all other nights?~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
(\emph{b. with colors to further subdivide the spectrum and name colors more specifically}) 
\B{Fì\-syu\-lang lu ta'\-leng\-na ean.} This flower is skin\hyp{}blue.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Fì\-syu\-lang a\-ean\hyp{}na\hyp{}ta'leng lor lu nì\-txan.} This skin\hyp{}blue flower is very beautiful.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
(\emph{c. in comparisons of equality}) as \emph{adj.}\slash \emph{adv.} as \emph{n.}\slash \emph{pn.} \B{Oe lu nì\-ftxan sìl\-tsan na nga.} I am as good as you.~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln}
\B{Oe tul nì\-ftxan nì\-win na nga.} I run as fast as you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Oel ye\-rik\-it ta\-ron nì\-ftxan nìl\-tsan na nga.} I hunt \emph{yerik} as well as you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
(\emph{d. idiom}) \B{nì\-win na hu\-fwe}, `as fast as the wind' (\emph{any situation to express rapidity}). \B{na hu\-fwe}, fluent (not just rapid) use of language.
\B{Fte\-ria oel lì'\-fya\-ti le\-Na'\-vi, slä mi ke tsä\-ngun pi\-vll\-txe na hu\-fwe.} I'm studying Na'vi, but I'm afraid I still can't speak it fluently.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\emph{e. idiom}) \B{na fkxen eo fkio}, to go to waste [lit.: `like vegetable food before a tetrapteron'] 
\B{Nge\-yä fì\-tì\-kang\-kem\-vi a\-txan\-tsan ke sla\-yu na fkxen eo fkio.} This excellent work of yours will not go to waste.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{pxel}{\B{pxel}})

% na'rìng
\hi
\HT{na'rIng}{\B{\cp{na'}rìng}} \I{["naP.RIN]} \emph{n.} forest. 
\B{Na'\-rìng sngä\-'i ni\-vrr txon\-krr syu\-ra\-tan\-fa.} The forest begins to glow at night with bioluminescence.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Lu na'\-rìng tsuk\-ha\-haw.} One can sleep in the forest [lit.: the forest ist `sleepable'].~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}
\B{Na'\-rìng\-ìl nga' pxay\-a io\-ang\-it.} The forest contains many animals.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok.} Yesterday I was with Txewì in the forest.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
\B{Tì\-o\-mum\-mì oe\-yä, pol na'\-rìng\-it inan nìl\-tsan.} As far as I know, he reads the forest well.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Sra\-ke pol la\-yok txew\-ti na'\-rìng\-ä?} Will he approach the edge of the forest?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{a. idiom}) \B{zawr (a) mì na'\-rìng}, ``Old news.'' [lit.: an animal cry in the forest].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Sra\-ke ngal sto\-lawm fu\-ta Tse\-nu Ra\-lur mo\-wan lu nì\-txan? -- Zawr mì na'\-rìng, ma tsmuk! Tsat o\-mum oel kin\-trr\-o.} Have you heard that Ralu has the hots for Tsenu? -- Old news, brother! I've known it for a week.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% Na'vi / nìNa'vi / leNa'vi
\hi
\HT{na'vi}{\B{\cp{Na'}vi}} \I{["naP.vi]} \emph{plural n.} (\emph{a.}) the Na'vi, the People, indigenous Pandorean sentient humanoids.  
\B{Tì\-'e\-fu\-mì oe\-yä su\-te\-kip nì\-wotx, ftxey Na'\-vi ftxey Saw\-tu\-te, lu sìl\-tsan lu kawng.} In my opinion, among all people, whether Na'vi or Skypeople, there are good and bad.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Nì\-trr\-trr yom Na'\-vil wu\-tsot 'aw\-si\-teng pxaw yll\-txep.} The Na'vi regularly eat dinner together around a communal fire.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Kea tsam\-si\-yu le\-sno\-nrr\-a ke tsun Na'\-vit iveyk.} No warrior full of self-pride can lead the People.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Tsa\-sä\-lä\-txayn Na'\-vi\-ru srung so\-li nì\-'aw fte sli\-vu txur nì\-'ul.} That defeat only helped the People become stronger.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Nga\-ri hu Ey\-wa sa\-lew ti\-re\-a, tokx 'ì\-'awn slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body remains to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{a\slash wiki}
\B{Na'\-vi\-ri txin\-a sä\-'o tì\-tu\-sa\-ron\-ä lu tsko swi\-zaw.} For the Na'vi the bow and arrow is the main hunting tool.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\emph{b.}) \emph{shortened for} the Na'vi language. \B{'I\-vong Na'\-vi.} Let Na'vi bloom.\slash May the Na'vi language bloom.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{lena'vi}{\B{le\-Na'\-vi}}, having to do with the Na'\-vi. 
\on{2}.~\HL{nIna'vi}{\B{nì\-Na'\-vi}}, like the Na'\-vi, Na'\-vi\-ly.

% naer
\hi
\HT{naer}{\B{na\cp{er}}} \I{[na."{}ER]} \emph{countable n.} drink, beverage.
\B{Fì\-na\-er ftxì\-vä' lu nì\-hawng, ha swey\-lu txo ngal tsat kxi\-vukx nì\-win.} This drink tastes horrible, so you'd better swallow it quickly.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Fì\-na\-er\-ìri ngal ew\-ku 'u\-ot a\-stxong srak?} Do you taste something strange in this drink?~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Fì\-na\-er\-ìri sur fkan oe\-ru ka\-lin.} This drink tastes sweet to me.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}

% nafì'u
\hi
\HT{nafI'u}{\B{nafì\cp{'u}}} \I{[na.fI."Pu]} \emph{adj.} such.
\B{Na\-fì\-'u\-a mok\-ri lu sä\-kel\-pxìm\-run.} Such a voice is a rarity.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b}
\B{Krr\-o len ay\-hem a\-na\-fì\-'u.} Sometimes these things happen.~\LINK{http://forum.learnnavi.org/language-updates/just-another-few-words/}{ln}
(\ding{43}~\HL{na}{\B{na}}, \HL{fI'u}{\B{fì\-'u}})

% nafpawng
\hi 
\HT{nafpawng}{\B{na\cp{fpawng}}} \I{[na."fpawN]} \emph{adv.} grievingly, with grief. 
(\ding{43}~\HL{afpawng}{\B{afpawng}})

% nal
\hi 
\HT{nal}{\B{nal}} \I{[nal]} \emph{vin.} suffer. 
\B{Krr\-a tse\-'a oel tì\-fkey\-tok\-it txan\-lo\-kxe\-yä oey, nal nì\-txan mì te'\-lan.} When I see the condition of my country, I suffer greatly in my heart.~\LINK{naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b} 
(\ding{43}~\HL{ngA'An}{\B{ngä\-'än}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nalsteng}{\B{nal\-steng}}, feel empathy\slash compassion.

% nalsteng
\hi 
\HT{nalsteng}{\B{\cp{nal}steng}} \I{["n$\cdot$al.stEN]} \emph{vin., inf.}\on{11} feel empathy, feel compassion (for, \ding{43}~\HL{hu}{\B{hu}} \emph{or the topic}). 
\B{Po\-hu\slash Po\-ri  oe nal\-steng.} I feel compassion for him.~\LINK{naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b} 
(\ding{43}~\HL{nal}{\B{nal}}, \HL{'awsiteng}{\B{'aw\-si\-teng}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tInalsteng}{\B{tì\-nal\-steng}}, compassion, empathy.
\on{2}.~\HL{lenalsteng}{\B{le\-nal\-steng}}, compassionate (\emph{ofp.}).

% nalutsa
\hi 
\HT{nalutsa}{\B{na\cp{lu}tsa}} \I{[na."lu.\t{ts}a]} \emph{n., fauna} nalutsa.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln} 

% nam'ake
\hi
\HT{nam'ake}{\B{nam\cp{'a}ke}} \I{[nam."Pa.kE]} \emph{adv.} confidently. 
\B{Tu\-ke\-ru pol\-txe Ak\-wey nam\-'a\-ke, omum fu\-ta ke tsun poe sti\-vo.} Akwey spoke to Tuke confidently, knowing that she couldn't refuse.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{am'ake}{\B{am\-'a\-ke}})

% nantang
\hi
\HT{nantang}{\B{\cp{nan}tang}} \I{["nan.taN]} \emph{n., fauna} viperwolf, striped armored wolf, \emph{caniferratus costatus}.
\B{ay\-nan\-tang sì ay\-ri\-ti}, viperwolves and stingbats.~\LINK{http://naviteri.org/2010/07/thoughts-on-ambiguity/}{b}
\B{Nan\-tang\-ìl yom ye\-rik\-it.} Viperwolves eat hexapedes.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
\B{Ay\-hem\-ìri 'e\-wan\-a tsa\-nan\-tang\-ur a\-hì\-'i tìng na\-ri.} Look at what that little young viperwolf is doing.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nantangtsyIp}{\B{nantangtsyìp}}, dog.

% nantangtsyìp
\hi 
\HT{nantangtsyIp}{\B{\cp{nan}tangtsyìp}} \I{["nan.taN.\t{ts}jIp\textcorner]} \emph{n.} dog (Earth animal). 
\B{Oey nan\-tang\-tsyìp o\-lu\-e' ta\-lu\-na yom nì\-hawng.} My dog vomited because it ate too much.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{nantang}{\B{nan\-tang}}, \HL{-tsyIp}{\B{-tsyìp}})

% nang
\hi
\HT{nang}{\B{nang}} \I{[naN]} \emph{part. for surprise, exclamation, encouragement (mirative); always sentence\hyp{}final; appears with a form of} \ding{43}~\HL{txan}{\B{txan}}.
\B{Lì'\-fya\-vi le\-sar nì\-txan nang!} A very useful expression, indeed!~\LINK{http://forum.learnnavi.org/language-updates/just-another-few-words/}{ln}
\B{kem\-lì\-'u\-vi (lì\-'u a\-txan\-tsan nang!)}, infix (a truly excellent word!)~\LINK{http://naviteri.org/2010/07/vocabulary-update/comment-page-1/\#comment-95}{b}

% napxì
\hi 
\HT{napxI}{\B{na\cp{pxì}}} \I{[na."p'I]} \emph{adv.} partially, in part. 
\B{Fì\-tì\-kang\-kem\-vi ha\-sey lu na\-pxì nì\-'aw.} This project is only partially complete.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
(\ding{43}~\HL{hapxI}{\B{ha\-pxì}}) 

% naranawm
\hi 
\HT{naranawm}{\B{Nara\cp{nawm}}} \I{[naR.a."nawm]} \emph{n., astr.} Polyphemus [lit.: Great Eye]. 
\B{Na\-ra\-nawm\-ä mawl mi lu uo taw\-txew.} Half of Polyphemus is still behind the horizon.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
(\ding{43}~\HL{nari}{\B{na\-ri}}, \HL{nawm}{\B{nawm}})

% nari / nari si
\hi
\HT{nari}{\B{\cp{na}ri}} \I{["na.Ri]} \textsc{i}.~\emph{n.} eye.
\B{Nge\-yä me\-na\-ri lor lu nì\-txan.} Your eyes are very beautiful.~\LINK{https://www.youtube.com/watch?v=ojr_BE8ZmVE}{yt}
\B{Oe\-ri me\-na\-ri\-ru pi\-ak si.} I open my eyes.~\LINK{http://forum.learnnavi.org/language-updates/open-close/}{ln}
\B{Po\-ri me\-na\-ri pi\-ak sä\-po\-li.} His eyes opened.~\LINK{http://forum.learnnavi.org/language-updates/open-close/}{ln} \\
\textsc{ii}.a.~\B{\cp{na}ri si} \I{["na.Ri si]} \emph{vin.} watch out, be careful. \B{Na\-ri so\-li ay\-oe fte\-ke nì\-hawng li\-vok.} We were careful not to get too close.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
\B{Na\-ri si fte kea fe\-kem ke li\-ven ngar!} Be careful you don't have an accident!~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Fì\-txe\-le\-ri ze\-ne oe na\-ri si\-vi!} I have to be careful about this matter!~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1093}{b}
\B{Na\-ri si txo\-ke\-fyaw tì\-ran nì\-nu.} Be careful or you'll trip.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\emph{to express `do X carefully'}) \B{nari si} \texttt{+} \emph{root verb} 
\B{Na\-ri si keyn\-ven!} Step carefully!~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b}
\B{Na\-ri si lo\-nu swi\-zaw\-it.} Release the arrow carefully.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{lare}{\B{la\-re}}) \\
\textsc{ii}.b.~\B{tìng \cp{na}ri} \I{[tIN."na.Ri]} \emph{or coll.} \I{[tI."na.Ri]} \emph{vin.} (\texttt{+}\emph{control}) look, look out, look at.
\B{Tìng na\-ri ne\-kll.} Look below.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
\B{Po\-ru tìng na\-ri!} Look at him.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Le\-hrr\-ap lu fwa evi\-tsyìp sle\-le mì hil\-van lu\-ke fwa fyeyn\-tu te\-rìng na\-ri.} It's dangerous for tiny ones to swim in the river without an adult watching.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{narlor}{\B{nar\-lor}}, (visually) beautiful. 
\on{2}.~\HL{narvA'}{\B{nar\-vä'}}, ugly, unsightly.
\on{3}.~\HL{naranawm}{\B{Na\-ra\-nawm}}, Polyphemus.
\on{4}.~\HL{naritxim}{\B{na\-ri\-txim}}, eyethorn.

% naritxim
\hi 
\HT{naritxim}{\B{\cp{na}ritxim}} \I{["na.Ri.t'im]} \emph{n.} \ding{96} eyethorn. 
(\ding{43}~\HL{nari}{\B{nari}}, \HL{txim}{\B{txim}}) 

% narlor
\hi
\HT{narlor}{\B{nar\cp{lor}}} \I{[naR."loR]}  \emph{adj.} beautiful visually. 
(\ding{214}~\HL{narvA'}{\B{nar\-vä'}}; \ding{43}~\HL{nari}{\B{na\-ri}}, \HL{lor}{\B{lor}})

% narvä'
\hi
\HT{narvA'}{\B{nar\cp{vä'}}} \I{[naR."v\ae{}P]}  \emph{adj.} ugly, unsightly. 
(\ding{214}~\HL{narlor}{\B{nar\-lor}}; \ding{43}~\HL{nari}{\B{na\-ri}}, \HL{vA'}{\B{vä'}})

% natkenong 
\hi
\HT{natkenong}{\B{nat\cp{ke}nong}} \I{[nat\textcorner."kE.noN]} \emph{conj.\slash adv.} for example.
\B{Slä lu 'a\-'aw\-a tì\-ke\-teng, nat\-ke\-nong, tsyokx\-ìri ke lu zek\-wä a\-tsìng ki a\-mrr.} But there are several differences, for example, the hand doesn't have four but five digits.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{na}{\B{na}}, \HL{tIkenong}{\B{tì\-ke\-nong}})

% natxu
\hi 
\HT{natxu}{\B{na\cp{txu}}} \I{[na."t'u]} \emph{vtr.} disapprove of. 
\B{Oel nge\-yä tì\-hawl\-it na\-txu ulte tsa\-wä wa\-syem.} I disapprove of your plan and will oppose (fight against) it.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b} 
(\ding{214}~\HL{smaw}{\B{smaw}}) \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tInatxu}{\B{tì\-na\-txu}}, disapproval.

% nawang
\hi 
\HT{nawang}{\B{\cp{na}wang}} \I{["na.waN]} \emph{vin.} merge, become one with (\HL{hu}{\B{hu}}, with). 
\B{Tì\-mun\-txa\-maw lam Ni\-nat\-ur fwa vit\-ra sne\-yä no\-la\-wang hu pum mun\-txa\-tu\-ä.} After her marriage, it seemed to Ninat that her soul had merged with that of her mate.~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b} 

% nawfwe
\hi
\HT{nawfwe}{\B{naw\cp{fwe}}} \I{[naw."fwE]} \emph{adj.} fluent (for speech). (\emph{from:} \B{(nì\-win) na hu\-fwe})
\B{Toi\-tsye\-ri lu poe pll\-txe\-yu a\-naw\-fwe nì\-txan.} She's a very fluent speaker of German.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}

% nawkx 
\hi 
\HT{nawkx}{\B{nawkx}} \I{[nawk']} \emph{n., fauna} bone helm rhino.

% nawm
\hi
\HT{nawm}{\B{nawm}} \I{[nawm]} \emph{adj.} great, noble.
\B{Na'\-vi\-ru set lu nawm\-a eyk\-tan a\-mip a lar\-mu Taw\-tu\-te.} The Na'vi have a great new leader now who had been a Skyperson.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Ut\-ral\-ä a\-nawm | Ay\-ri\-na' lu ay\-oeng.} We are all seeds | Of the Great Tree.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} \\
(i.) \B{Nawm\-a Sa'\-nok}, the Great Mother, Eywa. 
\B{Tsawl\-a pa\-lu\-lu\-kan\-ìl oe\-ti 'ìl\-me\-ko a krr, Nawm\-a Sa'\-nok lrr\-tok sei\-yi ul\-te oe em\-ro\-ley.} The Great Mother was looking after me when the big thanator attacked and I've survived.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nawmtu}{\B{nawm\-tu}}, great, noble person. 
\on{2}.~\HL{naranawm}{\B{Na\-ra\-nawm}}, Polyphemus. 
\on{3}.~\HL{nawmtoruktek}{\B{nawm\-to\-ruk\-tek}}, Toruk Makto Totem.
\on{4}.~\HL{manawmtu}{\B{ma\-nawm\-tu}}, ``excuse me''.
\on{5}.~\HL{nawnumtseng}{\B{naw\-num\-tseng}}, university.

% nawmtoruktek
\hi
\HT{nawmtoruktek}{\B{nawm\cp{to}ruktek}} \I{[nawm."to.Ruk\textcorner.tek\textcorner]} \emph{or} \I{[.rUk.]} (\textsc{rn}: \B{nawm\cp{to}rùktek} \I{[nawm."to.RUk\textcorner.tEk\textcorner]}) \emph{n.} Toruk Makto Totem.
(\ding{43}~\HL{nawm}{\B{nawm}}, \HL{toruk}{\B{to\-ruk}}, \HL{tekre}{\B{tek\-re}})

% nawmtu
\hi
\HT{nawmtu}{\B{\cp{nawm}tu}} \I{["nawm.tu]} \emph{n.} great person. 
(\ding{43}~\HL{nawm}{\B{nawm}})
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{manawmtu}{\B{ma\-nawm\-tu}}, ``excuse me''.

% nawnumtseng
\hi 
\HT{nawnumtseng}{\B{naw\cp{num}tseng}} \I{[naw."num.\t{ts}EN]} \emph{or} \I{["nUm.]} \emph{n.} university. 
\B{Naw\-num\-tseng\-ä ay\-nu\-me\-yu\-ri ley ay\-nga\-ru tì\-teng\-wotx, ay\-oe\-ru ke\-teng\-syon.} Regarding university students, you value uniformity, we value diversity.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
(\ding{43}~\HL{nawm}{\B{nawm}}, \HL{numtseng}{\B{num\-tseng}})

% nawri
\hi
\HT{nawri}{\B{\cp{naw}ri}} \I{["naw.ri]} \emph{adj.} talented. 
\B{Nga lu rol\-yu a\-naw\-ri slä Ni\-nat lu pum a\-swey.} You're a talented singer but Ninat is the best (one).~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tInawri}{\B{tì\-naw\-ri}}, talent.

% nayn
\hi
\HT{nayn}{\B{nayn}} \I{[najn]} \emph{num.}, \textsc{lw} nine, (the symbol) \on{9} (\emph{only as Human decimal figure}). 
(\ding{43}~\HL{'eyt}{\B{'eyt}})

% näk 
\hi
\HT{nAk}{\B{näk}} \I{[n\ae{}k\textcorner]} \emph{vtr.} drink.
\B{Nge\-yä tsngal lum\-pe lu mek? Näk nì\-'ul ko!} Why is your cup empty? Drink up!~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Ngay\-txoa, ke tsun oe swo\-at ni\-väk.} Sorry, but I can't drink spirits.~\LINK{http://forum.learnnavi.org/language-updates/new-words-from-the-2012-avatarmeet-navi-class/msg551479/\#msg551479}{ln}
\B{Kxawm swo\-at nì\-hawng nil\-vä\-ngäk oel txo\-nam.} Perhaps I've had too much to drink last night.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/comment-page-1/\#comment-2585}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nInAk}{\B{nì\-näk}}, by drinking, in a liquid way.

% nän
\hi
\HT{nAn}{\B{nän}} \I{[n\ae{}n]} \textsc{i}. \emph{vin.} decrease.
\B{Fra\-zì\-sì\-krr pay kil\-van\-ä nän nì\-'ul\-'ul.} Every season the river dries up more and more. \\
\textsc{ii}. \emph{conj. to form correlative comparisons} (\emph{a.}) \B{nän … nän}, the less … the less.   
\B{Nän ftia, nän lu skxom a em\-za\-'u.} The less you study, the less chance you have of passing.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
(\emph{b.}) \B{nän … 'ul}, the less … the more. 
\B{Nän yom kxam\-trr, 'ul 'e\-fu ohakx kaym.} The less you eat at noon, the hungrier you'll feel in the evening.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{214}~\HL{'ul}{\B{'ul}}) \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nInAn}{\B{nì\-nän}}, less.

% ne 
\hi
\HT{ne}{\B{ne}} \I{[nE]} \emph{adp.} to, towards (\emph{direction}).   
\B{Re\-rol teng\-krr ke\-rä | Ìlä fya\-'o a\-vol | Ne kxam\-tseng.} We sing our way | Down the eight paths | To the center.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Te\-rì\-ran ay\-oe ay\-nga\-ne.} We are walking your way.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Mìn ne ftär\-pa.} Turn to the left.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln}
\B{Tswìl\-may\-on oe ftu Wa\-syìng\-ton ne kel\-ku.} I've just flown home from Washington.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
(\emph{a. may be used to disambiguate the predicate of the verb} \ding{43}~\HL{slu}{\B{slu}}) \B{Ta\-ron\-yu slu ne tsam\-si\-yu.} The hunter becomes a warrior.~\LINK{http://wiki.learnnavi.org/Canon/2010/UltxaAyharyu\%C3\%A4\#Word_Order_with_Slu}{wiki}
(\emph{b.1 with} \ding{43}~\HL{za'u}{\B{za\-'u}}) 
\B{Zo\-la\-'u nì\-prr\-te' ne pì\-lok.} Welcome to the blog.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
(\emph{b.2 and} \ding{43}~\HL{nI'eng}{nì\-'eng}) \B{za\-'u nì\-'eng}, share an interest in common. 
\B{Tì\-ru\-sol za\-'u ne fo nì\-'eng.} They share an interest in singing.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\emph{b.3, an adv. and a sense organ}) `s.o. looks\slash sounds\slash tastes\slash smells good or bad. 
\B{Nga za\-'u nìl\-tsan ne na\-ri.} You look good.~\LINK{https://naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b}
\B{Fì\-syu\-ve za\-'u nì\-fe' ne ftxì.} This food tastes bad.~\LINK{https://naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b} 
\B{Nga za\-'u nì\-fpxa\-mo ne mik\-yun. Mok\-ri\-ri kem\-pe lo\-len?} You sound terrible. What happened to your voice?~\LINK{https://naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b}
(\emph{c.}) \B{Ne kll\-te!} ``Get down!'' [lit.: to the ground] 
(\emph{d. with} \ding{43}~\HL{zup}{\B{zup}}) perish. \B{Ke za\-syup Lì\-'o\-na ne kxu\-tu.} The l'Ona will not perish to the enemy. [lit.: will not fall to the enemy]~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g} 
(\emph{e. with} \ding{43}~\HL{kan}{\B{kan}}) aim s.t. at s.t.\slash s.o.
\B{Ney\-ti\-ril tsko\-ti ke\-ran ne ye\-rik.} Neytiri is aiming her bow at a hexapede.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/comment-page-1/\#comment-22792}{g\slash b}
(\emph{f. with} \ding{43}~\HL{lonusye}{\B{lo\-nu\-sye}}) blow on.
\B{Txo syu\-ve som li\-vu nì\-hawng, lo\-nu\-sye tsa\-ne.} If the food is too hot, blow on it.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{214}~\HL{ftu}{\B{ftu}}) \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nefA}{\B{ne\-fä}}, up, upwards.   
\on{2}.~\HL{nekll}{\B{ne\-kll}}, down, below.   
\on{3}.~\HL{nemfa}{\B{nem\-fa}}, into, inside.   
\on{4}.~\HL{neto}{\B{ne\-to}}, away.

% ne'ìm
\hi
\HT{ne'Im}{\B{ne\cp{'ìm}}} \I{[nE."PIm]} \emph{adv.} back (\emph{direction}). 

% nefä
\hi
\HT{nefA}{\B{ne\cp{fä}}} \I{[nE."f\ae]} \emph{adv.} up (\emph{direction}), upwards. 
\B{Vll eyk\-yu ne\-fä fte po\-ngu fä\-ki\-vä.} The leader signals the party to ascend by pointing upwards.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
(\ding{43}~\HL{ne}{\B{ne}}, \HL{fApa}{\B{fä\-pa}})

% Nekawn
\hi
\HT{nekawn}{\B{Nekawn}} \I{[nE.kawn]} \emph{n. female name}.~\LINK{}{dh}

% nekll
\hi
\HT{nekll}{\B{ne\cp{kll}}} \I{[nE."k\s{l}]} \emph{adv.} down (\emph{direction}).
\B{Tìng na\-ri ne\-kll.} Look below.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
\B{Po pxun\-fa pxek tsa\-krr me\-fo zup ne\-kll!} He shoved the two of them and they fell down.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} 
(\ding{43}~\HL{ne}{\B{ne}}, \HL{kllte}{\B{kll\-te}})

% nekx
\hi
\HT{nekx}{\B{nekx}} \I{[nEk']} \emph{vtr.} burn, consume.
\B{Tsa\-srä a\-naw\-nekx ley nì\-'it nì\-'aw.} That burnt cloth is of little value.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b} 
(\ding{43}~\HL{palon}{\B{pa\-lon}}) \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{krrnekx}{\B{krr\-nekx}}, take time.

% nemfa
\hi
\HT{nemfa}{\B{\cp{nem}fa}} \I{["nEm.fa]} \emph{adp.} (\emph{a.}) into. 
\B{Lo\-ak\-ìl pä\-nu\-to\-lìng fu\-ta kar oe\-ru fya\-'ot a 'ìp fko nem\-fa ewll.} Loak promised he'd teach me how to vanish into the bushes.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{b. idiomatic use}) (i.) with \HL{fpxAkIm}{\B{fpxä\-kìm}} to mean ``born''.
\B{Fì\-tu\-te a\-hì\-'i a nem\-fa ki\-fkey fpxì\-mä\-kìm.} This little person that just entered into the world.~\LINK{http://masempul.org/2010/07/\%E2\%80\%99ite-polaheiem/}{sempul} (ii.) (for astronomical bodies rising) \B{Tsaw\-ke fpxe\-rä\-kìm nem\-fa taw.} The sun is rising.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{214}~\HL{ftumfa}{\B{ftum\-fa}}; \ding{43}~\HL{ne}{\B{ne}}, \HL{mIfa}{\B{mì\-fa}})

% nemrey
\hi
\HT{nemrey}{\B{nem\cp{rey}}} \I{[nEm."REj]} \emph{adv.} in a fashion as if one's life were at stake. 
\B{Kä li!--Slä\dots{}--Nem\-rey!} Get going!--But\dots{}--Run for your life!~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{emrey}{\B{em\-rey}})

% neni
\hi
\HT{neni}{\B{\cp{ne}ni}} \I{["nE.ni]} \emph{n.} sand.
\B{Ne\-ni lew si fì\-txa\-yor vay txam\-pay nì\-wotx.} Sand covers this expanse all the way to the ocean.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Tsay\-re\-nur le\-we\-opx a sìn ne\-ni tìng na\-ri.} Look at those wave\hyp{}like patterns in the sand.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 

% neteyam
\hi 
\HT{neteyam}{\B{Ne\cp{te}yam}} \I{[nE."tE.jam]} \emph{n. male name}. 

% neto
\hi
\HT{neto}{\B{ne\cp{to}}} \I{[nE."to]} \emph{adv.} away (\emph{direction}). 
\B{Ay\-nga ne\-to ri\-vikx!} Step back! [lit.: move away you all]~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki}
\B{Ne\-to rä\-'ä ki\-vä!} Don't go away!~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
\B{Ftu oe ne\-to rikx!} Get away from me!~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-994}{b}
(\ding{43}~\HL{mIso}{\B{mì\-so}})

% netrìp
\hi
\HT{netrIp}{\B{\cp{net}rìp}} \I{["nEt\textcorner.RIp\textcorner]} \emph{adv.} luckily, happily.
(\ding{43}~\HL{etrIp}{\B{et\-rìp}})

% new
\hi
\HT{new}{\B{new}} \I{[nEw]} \emph{vtr.} (\emph{a.}) want (s.t.).
\B{Nga\-tsyìp\-ìl new pe\-ut ta oe?} What does little you want from me?~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Ngal new a tsa\-'ut rä\-'ä wi\-vo, ma 'e\-vi.} Don't reach for what you want, child.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{b. modal}) (\emph{either with} ‹\B{iv}› \emph{or} \B{futa} \emph{and} ‹\B{iv}›).
\B{Txo new nga ri\-vey, oe\-hu!} Come with me, if you want to live.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog_.28Twitter_Questions.29}{wiki}
\B{Oe new nì\-txan ay\-nga\-ru fya\-wi\-vìn\-txu.} I want to guide you very much.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Oe new yi\-vom tey\-lut.} I want to eat \emph{teylu}.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
\B{Oe new ki\-vä.}\slash \B{Oel new fu\-ta ki\-vä.} I want to go.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-email/}{ln}
\B{Oel new fu\-ta Ta\-ron\-yu ki\-vä.} I want Taronyu to go.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-email/}{ln} 
\B{Fì\-pam\-tseo\-l oe\-ru ney\-kew fu\-ta srew.} This music makes me want to dance.~\LINK{https://naviteri.org/2020/12/mrra-tipangkxotsyip-five-little-discussions/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nulnew}{\B{nul\-new}}, prefer.   
\on{2}.~\HL{newomum}{\B{new\-omum}}, be curious.  
\on{3}.~\HL{txanew}{\B{txa\-new}}, greedy. 
\on{4}.~\HL{nInew}{\B{nì\-new}}, voluntarily, willingly, by desire.

% Newey
\hi
\HT{newey}{\B{Ne\cp{wey}}} \I{[nE."wEj]} \emph{n. female name}.
\B{Ìlä Fey\-ral, mun\-txa so\-li Ra\-lu sì Ne\-wey nì\-wan me\-srr\-am.} According to Pey\-ral, Ra\-lu and Ne\-wey were secretly married the day before yesterday.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Fì\-sä\-'an\-lal Ne\-wey\-ä nga\-ti sley\-ka\-yu lek\-ye\-'ung!} This yearning for Newey is going to drive you crazy.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}

% newomum
\hi
\HT{newomum}{\B{new\cp{o}mum}} \I{[n$\cdot$Ew."{}o.mum]} \emph{or} \I{[.mUm]} (\textsc{rn}: \B{new\cp{o}mùm} \I{[n$\cdot$Ew."{}o.mUm]}) \emph{vin.}\on{11} be curious. 
(\ding{43}~\HL{new}{\B{new}}, \HL{omum}{\B{omum}}) \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tInomum}{\B{tì\-no\-mum}}, curiosity.   
\on{2}.~\HL{lenomum}{\B{le\-no\-mum}}, curious.

% neympin
\hi
\HT{neympin}{\B{\cp{neym}pin}} \I{["nEjm.pin]} \emph{n.} the color \ding{43}~\HL{neyn}{\B{neyn}}.
(\ding{43}~\HL{'opin}{\B{'o\-pin}})

% neyn
\hi
\HT{neyn}{\B{neyn}} \I{[nEjn]} \emph{adj.} light colored, ``shades of white''. 
\B{neyn na txä\-rem}, the light color of bone.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} 
\B{neyn na ya\-pay}, the light, nondescript color of mist or fog.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{neympin}{\B{neym\-pin}}, (the color) \B{neyn}.

% Neytep
\hi
\B{Neytep} \I{[nEj.tEp\textcorner]} \emph{n.} \emph{given name}.~\LINK{}{dh}

% Neytiri
\hi
\HT{neytiri}{\B{Ney\cp{ti}ri}} \I{[nEj."ti.ri]} \emph{n. female name}.
\B{Ney\-ti\-ri he\-ra\-haw.} Neytiri is sleeping.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#MSNBC}{wiki}
\B{Ney\-ti\-ril ye\-rik\-it to\-la\-ron.} Neytiri hunted a hexapede.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#MSNBC}{wiki}
\B{Ney\-ti\-ri\-ti fkol pä\-nu\-to\-lìng oe\-ru!} Neytiri was promised to me.~\LINK{http://forum.learnnavi.org/language-updates/panuting/}{ln\slash a}\textsuperscript{c}
\B{To\-ruk Mak\-to alu pi\-za\-yu Ney\-ti\-ri\-yä txan\-rar\-mo\-'a fra\-to kip ay\-ha\-pxì\-tu Oma\-ti\-ka\-yaä.} Neytiri's ancestor was the most famous Toruk Makto among the Omatikaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Ney\-ti\-ri\-ri lu yì tì\-sti\-yä kxayl nì\-ngay.} Neytiri is really angry.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}

% niä
\hi
\HT{niA}{\B{ni\cp{ä}}} \I{[ni"{}\ae]} \emph{vtr.} grab. 
\B{Oe mìn, to\-ruk, stum niä pol oe\-ti.} I turn, Toruk almost grabs me.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Iknimaya}{a\textsuperscript{c}}

% nik
\hi
\HT{nik}{\B{nik}} \I{[nik\textcorner]} \emph{adj.} convenient (\emph{usable without much expenditure of effort}). \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{niktsyey}{\B{nik\-tsyey}}, food wrap.

% nikre 
\hi
\HT{nikre}{\B{\cp{nik}re}} \I{["nik\textcorner.RE]} \emph{n.} hair.
\B{vawm na nik\-re}, the dark color of Na'vi hair.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Ri\-ni\-ri nik\-re yä\-ngo' nì\-tut.} Rini's hair is always perfect.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Nik\-re\-ri Ri\-ni\-yä 'ur fkan lor.} Rini's hair looks beautiful.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Nik\-re\-ri Ri\-ni\-yä fkan lor.} Rini's hair is pleasant to the senses.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{New sa'\-nok sli\-vayk nik\-ret 'eveng\-ä.} The mother wants to brush the child's hair.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nikroi}{\B{nik\-roi}}, hair ornament.

% nikroi 
\hi
\HT{nikroi}{\B{nik\cp{ro}i}} \I{[nik\textcorner."Ro.i]} \emph{n.} hair ornament. 
\B{Fì\-nik\-roi ley pìm\-txan?} How valuable is this hair ornament?~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
\B{Kä\-sro\-lìn oel nik\-roit Pey\-ral\-ur yoa fwa po rol oer.} I loaned Peyral a hair ornament in exchange for her singing to me.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
(\ding{43}~\HL{nikre}{\B{nik\-re}}, \HL{ioi}{\B{ioi}})

% niktsyey
\hi
\HT{niktsyey}{\B{\cp{nik}tsyey}} \I{["nik\textcorner.\t{ts}jEj]} \emph{n.} food wrap (\emph{food items wrapped in edible leaves and vines}). 
(\ding{43}~\HL{nik}{\B{nik}}, \HL{tsyey}{\B{tsyey}})

% nim
\hi
\HT{nim}{\B{nim}} \I{[nim]} \emph{adj.} timid, shy.   
\B{Ye\-rik lu swi\-rä a\-nim nì\-txan.} The hexapede is a very timid creature.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Nim rä\-'ä lu!} Don't be shy!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}

% ninan 
\hi
\HT{ninan}{\B{\cp{ni}nan}} \I{["ni.nan]} \emph{adv.} by reading. 
\B{Tì\-tu\-sa\-ron\-ìri fte fli\-vä, ze\-ne fko si\-vutx smar\-it ni\-nan nì\-no.} To succeed at hunting, you have to track your prey by reading (the forest) with attention to detail.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
(\ding{43}~\HL{inan}{\B{inan}})

% Ninat
\hi
\B{Ni\cp{nat}} \I{[ni."nat\textcorner]} \emph{n. female name}.
\B{Nga lu rol\-yu a\-naw\-ri slä Ni\-nat lu pum a\-swey.} You're a talented singer but Ninat is the best (one).~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Ni\-nat\-ìri tì\-ru\-sol Txe\-wì\-yä lu vä\-pam.} To Ninat, Txewì's singing is noise.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}

% ningyen 
\hi
\HT{ningyen}{\B{\cp{ning}yen}} \I{["niN.jEn]} \emph{adv.} mysteriously, in a puzzling fashion.
\B{Oe\-yä tska\-lep 'o\-lìp ning\-yen.} My crossbow has mysteriously disappeared.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{ingyen}{\B{ing\-yen}})

% nip
\hi
\HT{nip}{\B{nip}} \I{[nip\textcorner]} \emph{vin.} become stuck, get caught in s.t. 
\B{Ri\-ni fmar\-mi hi\-vi\-fwo slä ve\-nu no\-lip äo tskxe.} Ri\-ni tried to escape but her foot got caught under a rock.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{nip nìk\-lo\-nu\slash nì\-syep\slash nì\-meyp}, stuck tightly\slash stuck irremovably\slash weakly, loosely stuck.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 

% nitram
\hi
\HT{nitram}{\B{nit\cp{ram}}} \I{[nit\textcorner."Ram]} \emph{adj.} happy (\emph{used with} \HL{'efu}{\B{'efu}}). 
(\emph{a. only for people})
\B{Saw\-tu\-te zo\-la\-'u a\-krr\-ta po ke 'ole\-fu nit\-ram.} Since the Skypeople came, she had not been happy.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Fu\-la tsa\-yun oeng pi\-väng\-kxo ye'\-rìn oe\-ti nit\-ram sley\-ku nì\-txan.} It makes me very happy that we'll be able to chat soon.~\LINK{http://forum.learnnavi.org/language-updates/fula-short-form-of-fiul-a-and-tsal-oeti-steyki-nitram/}{ln}
\B{Fu\-la ho\-ren sì ay\-lì'\-fya\-vi a\-mip le\-sar sa\-yi me\-ngar oe\-ti nit\-ram 'ey\-ke\-fu nì\-txan.} It makes me feel happy that the rules and new expressions are useful to you both.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/comment-page-1/\#comment-1515}{b}
(\emph{b. idioms}) \B{Sey\-kxel sì nit\-ram!} Congratulations!~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Nit\-ram nì\-'aw!} Cheers!\slash Toast!~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\ding{43}~\HL{tInitramnga'}{\B{tìnitramnga'}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tInitram}{\B{tì\-nit\-ram}}, happiness.

% nivi
\hi
\HT{nivi}{\B{\cp{ni}vi}} \I{["ni.vi]} \emph{n.} sleeping hammock (\emph{general term}).
\B{Tsa\-ni\-vi a\-txaw\-nu\-la nì\-fe' hu\-fwe\-mì pa\-lo\-la\-te.} That poorly constructed hammock fell apart in the wind.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{swaynivi}{\B{sway\-ni\-vi}}, family hammock. 
\on{2}.~\HL{snonivi}{\B{sno\-ni\-vi}}, single\hyp{}person hammock. 

% nì-
\hi
\B{nì}- \I{[nI.]} (\emph{a. prod. pref. to form adv.s from adj.s or ord. num.s}) 
\B{Ey\-wal zey\-ki\-vo ngat nì\-win.} May Eywa heal you quickly.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Nì\-'aw\-ve oeng yo\-lom wu\-tsot; maw\-krr uvan si.} First we had dinner; afterwards we played.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
(\emph{b. prod. pref. on pron.s to mean}) like, as (s.o. does).
\B{Nì\-ay\-nga oe pe\-rey nì\-teng.} Like you I'm waiting too.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
(\emph{c. unprod. pref. to form adv.s from other word classes}) e.g. \B{nì\-li}, in advance; \B{nì\-'en}, making an informed guess; \B{nì\-no}, in detail.

% nì'aw
\hi
\HT{nI'aw}{\B{nì\cp{'aw}}} \I{[nI."Paw]} \emph{adv.} only.
\B{Ru\-txe, ay\-nga\-ri fì\-kem fe\-yä 'e\-'a\-la to\-pur ka\-ngay sì\-yi nì\-'aw.} Please, this will only confirm their worst fears about you.~\LINK{http://forum.learnnavi.org/language-updates/two-words-in-use-eal-and-kangay-si/msg564284/\#msg564284}{ln}
\B{Soaia lu hì\-'i; nì\-'aw sem\-pul, sa'\-nok, tsmu\-kan sì Txe\-wì.} The family is small; only father, mother, a brother and Txewì.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{'aw}{\B{'aw}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{le'awnI'aw}{\B{le\-'aw\-nì\-'aw}}, unique.

% nì'awtu 
\hi
\HT{nI'awtu}{\B{nì\cp{'aw}tu}} \I{[nI."Paw.tu]} \emph{adv.} alone (\emph{as one person}). 
\B{Ke tsun tu\-te a\-'aw tsa\-tskxe\-ti a\-ku'\-up kxi\-vel\-tek nì\-'aw\-tu.} One person alone can't lift that heavy rock.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{'aw}{\B{'aw}}, \HL{tute1}{\B{tu\-te}})

% nì'awve
\hi
\HT{nI'awve}{\B{nì\cp{'aw}ve}} \I{[nI."Paw.vE]} \emph{adv.} first, firstly. 
\B{Nì\-'aw\-ve oeng yo\-lom wu\-tsot; maw\-krr uvan si.} First we had dinner; afterwards we played.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
\B{Txo ni\-vew fko sä\-ro\-'a si\-vi, ze\-ne nì\-'aw\-ve ve\-nga' li\-vu.} If you want to accomplish great things, you first have to be organized.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{'awve}{\B{'aw\-ve}})

% nì'älek
\hi 
\HT{nI'Alek}{\B{nì\cp{'ä}lek}} \I{[nI."P\ae.lEk\textcorner]} \emph{adv.} determinedly, with determination. 
(\ding{43}~\HL{'Alek}{\B{'ä\-lek}})

% nì'al
\hi 
\HT{nI'al}{\B{nì\cp{'al}}} \I{[tI."Pal]} \emph{adv.} wastefully. 
(\ding{43}~\HL{'al}{\B{'al}})

% nì'en
\hi
\HT{nI'en}{\B{nì\cp{'en}}} \I{[nI."PEn]} \emph{adv.} making an informed guess, acting on intuition.
\B{Pol po\-le\-'un fu\-ta pe\-hem si nì\-'en.} He decided what to do on a hunch.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{'en}{\B{'en}})

% nì'eng
\hi
\HT{nI'eng}{\B{nì\cp{'eng}}} \I{[nI."PEN]} \emph{adv.} (\emph{a.}) levelly, equally.   
(\emph{b. idiom}) \B{za'u nì'eng}, share an interest in common (\HL{ne}{\B{ne}}, with). 
\B{Tì\-ru\-sol za\-'u ne fo nì\-'eng.} They share an interest in singing.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\emph{c. to express equal sharing}) \B{Fol tsngan\-it pxì\-mo\-lun\-'i nì\-'eng.} They shared the meat.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Tsun aw\-nga kel\-ku si\-vi nì\-'eng Saw\-tu\-te\-hu mì atx\-kxe aw\-nge\-yä.} We can share our land with the Skypeople.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\emph{d. to express value and worth; \emph{\B{nì'eng}} can be omitted in informal context}) \B{Ma\-sat oeyä ley nì\-ftxan na pum nge\-yä.} My breastplate is as valuable as yours.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b} 
(\emph{also monetary value on Earth}) \B{Oe\-yä el\-tu le\-fngap ley nì\-'eng na ew\-ro a\-za\-fu.} My computer is worth \on{70} euros.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\ding{43}~\HL{'engeng}{\B{'eng\-eng}})

% nì'eoio
\hi
\HT{nI'eoio}{\B{nì\cp{'e}oio}} \I{[nI."PE.o.i.o]} \emph{adv.} ceremoniously. 
\B{Aw\-ngal yom wu\-tsot nì\-fya\-'o le\-trr\-trr; fì\-txon yom nì\-'eoio.} We eat dinner in an ordinary manner; tonight we eat ceremoniously.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
(\ding{43}~\HL{'eoio}{\B{'eoio}})

% nì'eveng
\hi
\HT{nI'eveng}{\B{nì\cp{'e}veng}} \I{[nI."PE.vEN]} \emph{adv.} like a child, immaturely.
\B{Nga pll\-txe ke nì\-fyeyn\-tu ki nì\-'e\-veng.} You speak not like an adult but a child.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{'eveng}{\B{'eveng}})

% nì'eyng
\hi
\HT{nI'eyng}{\B{nì\cp{'eyng}}} \I{[nI."EjN]} \emph{adv.} back, in response, in answer.
\B{Oel tel 'u\-pxa\-ret le\-Na'\-vi a krr, new oe nì\-teng\-fya pam\-rel si\-vi nì\-'eyng.} When I receive a Na'vi message, I want to write the same way in response.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln}
(\ding{43}~\HL{'eyng}{\B{'eyng}}) 

% nì'i'a
\hi
\HT{nI'i'a}{\B{nì\cp{'i'}a}} \I{[nI."PiP.a]} \emph{adv.} finally, at last. 
\B{Slä nì\-'i'a tsun oe pi\-vll\-txe san Zo\-la\-'u nì\-prr\-te' ne pì\-lok Na'\-vi\-te\-ri sìk!} But at last I can say ``Welcome to the blog Na'viteri!''~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Oe\-ri nì\-'i'a tsyokx zo\-slo\-lu.} My hand is finally healed.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{'i'a}{\B{'i'a}})

% nì'it
\hi
\HT{nI'it}{\B{nì\cp{'it}}} \I{[nI."Pit\textcorner]} \emph{adv.} small amount, a bit. 
\B{Ay\-lì\-'u ngi\-an nì\-'it ske\-pek lu.} However the words are a bit formal.~\LINK{http://forum.learnnavi.org/language-updates/more-grammatical-tidbits/}{ln\slash a}
\B{Ru\-txe pi\-vll\-txe nì\-wok nì\-'it.} Please speak up a bit.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Tsa\-srä a\-naw\-nekx ley nì\-'it nì\-'aw.} That burnt cloth is of little value.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\ding{43}~\HL{'it}{\B{'it}})

% nì'Ìnglìsì
\hi
\HT{nI'InglIsI}{\B{nì\cp{'Ìng}lìsì}} \I{[nI."PIN.lI.sI]} \emph{adv.} English, in English.
\B{X-(ì)ri pe\-ral nì\-'Ìng\-lì\-sì?} What does X mean in English?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tsun Txe\-wì kop pi\-vll\-txe nì\-'it nì\-'Ìng\-lì\-sì a no\-lu\-me mì num\-tseng tok\-tor Kì\-rey\-sä.} Txewì can also speak a bit of English that he has learned in doctor Grace's school.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{'InglIsI}{\B{'Ìng\-lì\-sì}})

% nì'o'
\hi
\HT{nI'o'}{\B{nì\cp{'o'}}} \I{[nI."PoP]} \emph{adv.} in a manner that is very enjoyable,  `funly', enjoyably. 
\B{Ftia oel lì'\-fya\-ti le\-Na'\-vi nì\-'o' nì\-wotx!} Studying Na'vi is a ton of fun for me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Tìng mik\-yun nì\-'o'!} Have fun listening!~\LINK{http://naviteri.org/2012/10/ting-mikyun-listen/}{b}
(\ding{43}~\HL{'o'}{\B{'o'}})

% nì'ul
\hi
\HT{nI'ul}{\B{nì\cp{'ul}}} \I{[nI."Pul]} \emph{or} \I{[."PUl]} \emph{adv.} more. (\emph{a.}) \B{Maw\-krr la\-ye\-iu oer krr nì\-'ul.} Afterwards I will have more time.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Nga lä\-pi\-vawk nì\-'it nì\-'ul ko.} Tell me a bit more about yourself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Nì\-'ul ye'\-rìn …} More soon …~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\emph{b. to form comparatives}) \B{Oe nga\-to yom nì\-'ul.} I eat more than you.~\LINK{http://forum.learnnavi.org/language-updates/prefer-x-to-y-24474/}{ln}
\B{Tsa\-sä\-lä\-txayn Na'\-vi\-ru srung so\-li nì\-'aw fte sli\-vu txur nì\-'ul.} That defeat only helped the People become stronger.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{'ul}{\B{'ul}}) \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nI'ul'ul}{\B{nì\-'ul\-'ul}}, more and more.

% nì'ul'ul
\hi
\HT{nI'ul'ul}{\B{nì\cp{'ul}'ul}} \I{[nI."Pul.Pul]} \emph{or} \I{[."PUl.PUl]} \emph{adv.} increasingly, more and more.
\B{Fra\-zì\-sì\-krr pay kil\-van\-ä nän nì\-'ul\-'ul.} Every season the river dries up more and more.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Sìn yì sngä\-'i\-yu\-ä, slä tsye\-rìl (hay\-a yì\-ne) nì\-'ul\-'ul.} They're at a beginner's level, but they're getting better and better.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{nI'ul}{\B{nì\-'ul}})

% nìawnomum
\hi
\HT{nIawnomum}{\B{nìaw\cp{no}mum}} \I{[nI.aw."no.mum]} \emph{or} \I{[.mUm]}, \emph{in fast speech often} \I{[naw."no.mum]} (\textsc{rn}: \B{nìaw\cp{no}mùm} \I{[nI.aw."no.mUm]}) \emph{adv.} as you know, as is known, as everyone knows.
\B{Slä nì\-aw\-no\-mum, ze\-ne oe 'aw\-si\-teng tì\-kang\-kem si\-vi fo\-hu.} But as you know, I must work together with them.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Nga nì\-aw\-no\-mum to oe\-tsyìp lu txur nì\-txan.} As everyone knows, you're a lot stronger than little old me.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\ding{43}~\HL{awn}{‹\B{awn}›}, \HL{omum}{\B{omum}})

% nìayoeng
\hi
\HT{nIayoeng}{\B{nìay\cp{oeng}}} \I{[naj."wEN]} \emph{adv.} like us, as we do. 
\B{Ma 'ite, aw\-nge\-yä fya\-'o\-ri ze\-ne nga sä\-nume si\-vi po\-ru fte tsi\-vun pil\-vll\-txe sì ti\-vì\-ran nì\-ay\-oeng.} My daughter, you will teach him our way to speak and walk as we do.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki}
(\ding{43}~\HL{ayoeng}{\B{ay\-oeng}})

% nìfe'
\hi
\HT{nIfe'}{\B{nì\cp{fe'}}} \I{[nI."fEP]} \emph{adv.} badly.
\B{Oe pll\-ngay san mo\-lak\-to oe nì\-fe', ta\-fral sno\-laytx; wä\-tu lu oe\-to txur.} I acknowledge that I rode badly, so I lost; my opponent was stronger than I was.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{fe'}{\B{fe'}})

% nìfkeytongay
\hi
\HT{nIfkeytongay}{\B{nìfkeyto\cp{ngay}}} \I{[nI.fkEj.to."Naj]} \emph{adv.} actually, as a matter of fact.
\B{Tsat ngal tswo\-lä\-nga' nì\-lam. -- Nì\-fkey\-to\-ngay ke tswo\-la' kaw\-'it.} Apparently you've forgotten it. -- As a matter of fact, I haven't.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
(\ding{43}~\HL{tIfkeytongay}{\B{tì\-fkey\-to\-ngay}})

% nìfkrr
\hi
\HT{nIfkrr}{\B{nì\cp{fkrr}}} \I{[nI."fk\s{r}]} \emph{adv.} lately. 
\B{'Ìn nga fya\-pe nì\-fkrr?} What's been keeping you busy lately?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{krr}{\B{krr}})

% nìflä
\hi 
\HT{nIflA}{\B{nì\cp{flä}}} \I{[nI."fl\ae]} \emph{adv.} successfully. 
\B{So\-leia! Ngal tì\-fme\-tok\-it em\-zo\-la\-'u nì\-flä! Sey\-kxel sì nit\-ram!} You rose to the challenge! You passed the test successfully! Congratulations!~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
(\ding{43}~\HL{flA}{\B{flä}})

% nìflrr
\hi
\HT{nIflrr}{\B{nì\cp{flrr}}} \I{[nI."fl\s{r}]} \emph{adv.} gently, tenderly.
\B{Ze\-ne fko 'i\-vam\-pi prr\-nen\-it nì\-flrr fra\-krr.} One must always touch a baby gently.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{flrr}{\B{flrr}})

% nìfmokx
\hi
\HT{nIfmokx}{\B{nì\cp{fmokx}}} \I{[nI."fmok']} \emph{adv.} jealously, enviously.   
\B{Txe\-wìl tuk\-rut Lo\-ak\-ä nar\-mìn nì\-fmokx.} Txe\-wì was eyeing Lo\-ak's spear enviously.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{fmokx}{\B{fmokx}})

% nìfnu
\hi
\HT{nIfnu}{\B{nì\cp{fnu}}} \I{[nI."fnu]} \emph{adv.} silently.
\B{Tì\-ran nì\-fnu txan\-txew\-vay fte\-ke ay\-ye\-rik\-ìl aw\-nga\-ti sti\-vawm.} Walk as quietly as possible so the hexapedes won't hear us.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b}
(\ding{43}~\HL{fnu}{\B{fnu}})

% nìfpxamo
\hi 
\HT{nIfpxamo}{\B{nì\cp{fpxa}mo}} \I{[nI."fp'a.mo]} \emph{adv.} horribly, terribly, awfully. 
\B{Fpä\-ngìl oe, txo\-nam oe ro\-lol nì\-fpxa\-mo.} Sadly, I think I sang terribly last night.~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b} 
(\ding{43}~\HL{fpxamo}{\B{fpxa\-mo}})

% nìfrakrr
\hi
\HT{nIfrakrr}{\B{nì\cp{fra}krr}} \I{[nI."fRa.k\s{r}]} \emph{adv.} as always.
\B{Nì\-fra\-krr fol 'o\-lem a wu\-tso ftxì\-vä' lu nì\-ngay.} As always, the dinner they cooked tasted really terrible.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{frakrr}{\B{fra\-krr}})

% nìFranse
\hi
\HT{nIfranse}{\B{nì\cp{Fran}se}} \I{[nI."fRan.sE]} \emph{adv.} French, in French.
\B{X-(ì)ri pe\-ral nì\-Fran\-se?} What does X mean in French?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}

% nìfrir
\hi 
\HT{nIfrir}{\B{nì\cp{frir}}} \I{[nI."fRiR]} \emph{adv.} in layers. 
(\ding{43}~\HL{frir}{\B{frir}})

% nìftär
\hi
\HT{nIftAr}{\B{nì\cp{ftär}}} \I{[nE."ft\ae{}R]} \emph{adv.} to the left.
\B{Ne 'o\-ra\-tsyìp po\-lä\-hem, yak si nì\-ftär.} When you arrive at the pond, turn to the left.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{ftAr}{\B{ftär}})

% nìftue
\hi
\HT{nIftue}{\B{nì\cp{ftu}e}} \I{[nI."ftu.E]} \emph{adv.} easily. 
\B{Sra\-ke tsun oe fay\-upxa\-ret tsli\-vam nì\-ftue?} Can I understand these messages easily?~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln}
\B{Fay\-upxa\-re la\-yu ay\-sngä\-'i\-yu\-fpi, fte lì'\-fya\-ri aw\-nge\-yä fo tsì\-ye\-vun nì\-ftue nìl\-tsan\-sì ni\-vume.} These messages will be for beginners, so that they can learn about our language easily and well.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
(\ding{43}~\HL{ftue}{\B{ftue}})

% nìftxan
\hi
\HT{nIftxan}{\B{nì\cp{ftxan}}} \I{[nI."ft'an]} \emph{adv.} (\emph{a.}) so, to such an extent. 
\B{Fwa ay\-ngal fì\-tse\-nget tok nì\-pxay nì\-ftxan oe\-ru te\-ya si nì\-ngay.} It fills me with joy that you're here in such great number.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\emph{b. with adj. or adv. and} \HL{na}{\B{na}}\slash \HL{pxel}{\B{pxel}} \emph{to mean}) as \emph{adj.\slash adv.} as.
\B{Oe lu nì\-ftxan sìl\-tsan na nga.} I am as good as you.~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln} 
(\emph{A pronoun or definite point of comparison may take the topical}) 
\B{Nga\-ri lu oe nì\-ftxan sìl\-tsan.} I am as good as you.~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln}
\B{Oe tul nì\-ftxan nì\-win pxel nga.} I run as fast as you.~\LINK{https://forum.learnnavi.org/language-updates/eapressions-of-\%27like-we\%27/}{ln}
\B{Oel ye\-rik\-it ta\-ron nì\-ftxan nìl\-tsan na nga.} I hunt yerik as well as you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
(\emph{c. to form result clauses with} \HL{kuma}{\B{kuma}}) so.
\B{Lu poe se\-vin nì\-ftxan kuma yaw\-ne slo\-lu oer.} She was so beautiful that I fell in love with her.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{d. coll. idiom}) \B{[verb] nì\-ftxan kum\-a ter\-kup}, \emph{commenting on an action that is so intense that it results in dying}; ``like crazy, \ldots{} to death''. 
\B{Sran, sran, oe\-ri pì\-tìk tsa\-tse\-nget a mì tal! Fkxa\-ke nì\-ftxan kum\-a ter\-kup.} Yeah, yeah, scratch that place on my back! It itches like crazy!~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{43}~\HL{fItxan}{\B{fì\-txan}})

% nìftxavang
\hi
\HT{nIftxavang}{\B{nì\cp{ftxa}vang}} \I{[nI."ft'a.vaN]} \emph{adv.} passionately, with all heart.
\B{Fpo\-le' ay\-ngal oer fì\-txan nì\-ftxa\-vang a 'upxa\-ret sto\-lawm oel.} I have heard the message you have sent me so passionately.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Po\-ri mok\-ri lor lu nì\-txan ul\-te rol po nì\-ftxa\-vang.} Her voice is very beautiful and she sings passionately.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
(\ding{43}~\HL{ftxavang}{\B{ftxa\-vang}})

% nìfwefwi
\hi
\HT{nIfwefwi}{\B{nì\cp{fwe}fwi}} \I{[nI."fwE.fwi]} \emph{adv.} by whistling, in a whistling manner.
\B{'E\-van\-ìl alo a\-'aw\-ve nì\-'aw\-tu na'\-rìng\-it tar\-mok, ha to\-lìng lawr nì\-fwe\-fwi fte\-ke txo\-pu si\-vi.} The boy was alone in the forest for the first time, so he whistled a tune to calm his fears.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{fwefwi}{\B{fwe\-fwi}})

% nìfya'o
\hi
\HT{nIfya'o}{\B{nì\cp{fya}'o}} \I{[nI."fja.Po]} \emph{adv.} in a manner. 
\B{Aw\-ngal yom wu\-tsot nì\-fya\-'o le\-trr\-trr; fì\-txon yom nì\-'eoio.} We eat dinner in an ordinary manner; tonight we eat ceremoniously.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
\B{Poe pol\-txe nì\-fya'o a\-law.} She spoke clearly.~\LINK{http://forum.learnnavi.org/language-updates/why-is-this-night/msg154875/\#msg154875}{ln}
\B{Fo nì\-Na'\-vi pll\-txe nì\-fya\-'o a hek.} They speak Na'vi strangely.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{fya'o}{\B{fya\-'o}})

% nìfyeyntu
\hi
\HT{nIfyeyntu}{\B{nì\cp{fyeyn}tu}} \I{[nI."fjEjn.tu]} \emph{adv.} like an adult, maturely. 
\B{Nga pll\-txe ke nì\-fyeyn\-tu ki nì\-'e\-veng.} You speak not like an adult but a child.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{fyeyntu}{\B{fyeyn\-tu}})

% nìhawmpam
\hi
\HT{nIhawmpam}{\B{nì\cp{hawm}pam}} \I{[nI."hawm.pam]} \emph{adj.} noisily. 
\B{Txo fko ti\-vul mì na'\-rìng nì\-hawm\-pam, stawm ay\-ioang.} If you run noisily in the forest, the animals will hear.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{hawmpam}{\B{hawm\-pam}})

% nìhawng
\hi
\HT{nIhawng}{\B{nì\cp{hawng}}} \I{[nI."hawN]} \emph{adv.} too, excessively. 
\B{Na\-ri so\-li ay\-oe fte\-ke nì\-hawng li\-vok.} We were careful not to get too close.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
\B{Ay\-upxa\-re\-ri a\-ngim nì\-hawng lu alo oe\-yä!} Now it's my turn for a message that's too long!~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Fì\-na\-er ftxì\-vä' lu nì\-hawng, ha swey\-lu txo ngal tsat kxi\-vukx nì\-win.} This drink tastes horrible, so you'd better swallow it quickly.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\ding{43}~\HL{hawng}{\B{hawng}})

% nìhay
\hi
\HT{nIhay}{\B{nì\cp{hay}}} \I{[nI."haj]} \emph{adv.} next (in line). 
\B{Nga kä pe\-seng nì\-hay?} Where are you going next?~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{43}~\HL{hay}{\B{hay}})

% nìhek
\hi
\HT{nIhek}{\B{nì\cp{hek}}} \I{[nI."hEk\textcorner]} \emph{adv.} oddly, strangely   (\emph{sentence adverbial only}). 
\B{Nì\-hek fo nì\-Na'\-vi pll\-txe.} Strangely, they speak Na'\-vi.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{hek}{\B{hek}})

% nìhìpey
\hi
\HT{nIhIpey}{\B{nì\cp{hì}pey}} \I{[nI."hI.pEj]} \emph{adv.} hesitantly.
(\ding{43}~\HL{hIpey}{\B{hì\-pey}})

% nìhoan
\hi
\HT{nIhoan}{\B{nì\cp{ho}an}} \I{[nI."ho.an]} \emph{adv.} comfortably.
(\ding{43}~\HL{hoan}{\B{ho\-an}})

% nìhoet
\hi
\HT{nIhoet}{\B{nì\cp{ho}et}} \I{[nI."ho.Et\textcorner]} \emph{adv.} widely, pervasively.   
\B{Run fkol tey\-lut nì\-ho\-et.} You find \emph{teylu} everywhere.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{hoet}{\B{ho\-et}})

% nìhol
\hi
\HT{nIhol}{\B{nì\cp{hol}}} \I{[nI."hol]} \emph{adv.} lightly, just a few.
\B{lep\-wopx nì\-hol}, lightly cloudy, just a few clouds.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{hol}{\B{hol}})

% nìhona
\hi
\HT{nIhona}{\B{nì\cp{hon}a}} \I{[nI."ho.na]} \emph{adv.} endearingly, sweetly.
\B{Po ätxä\-le so\-li nì\-ho\-na fì\-txan, ke tsun oe sti\-vo.} She asked so sweetly that I couldn't refuse.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{hona}{\B{ho\-na}})

% nìk'ärìp
\hi
\HT{nIk'ArIp}{\B{nìk\cp{'ä}rìp}} \I{[nIk\textcorner."P\ae.RIp\textcorner]} \emph{adv.} steadily (\emph{without letting it move}).
\B{fyep nìk\-'ä\-rìp}, hold steadily.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{'ArIp}{\B{'ä\-rìp}})

% nìk'ong
\hi
\HT{nIk'ong}{\B{nìk\cp{'ong}}} \I{[nIk\textcorner."PoN]} \emph{adv.} slowly. 
(\emph{a. idiom}) \B{Ke ze\-ne win sä\-pi\-vi. 'Ivong nìk\-'ong.} Take your time; don't rush. Slow is fine.~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b}
(\ding{43}~\HL{kI'ong}{\B{kì\-'ong}})

% nìkakan
\hi 
\HT{nIkakan}{\B{nì\cp{ka}kan}} \I{[nI."ka.kan]} \emph{adv.} roughly (\emph{for behavior, people and things}). 
\B{Sem\-pul\-ìl a\-sìl\-tsan sney eveng\-it ke txaw nì\-ka\-kan.} A good father doesn't punish his children roughly.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{kakan}{\B{ka\-kan}})

% nìkakmokri
\hi 
\HT{nIkakmokri}{\B{nìkak\cp{mok}ri}} \I{[nI.kak\textcorner."mok\textcorner.ri]} \emph{adv.} mutely. 
\B{Kll\-kxo\-lem fo nì\-kak\-mok\-ri lu\-ke fwa 'aw\-a lì\-'uti pll\-txe.} They stood there mutely without saying a word.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{kakmokri}{\B{kak\-mok\-ri}}) 

% nìkanu
\hi
\HT{nIkanu}{\B{nì\cp{ka}nu}} \I{[nI."ka.nu]} \emph{adv.} intelligently, in a smart way. 
(\ding{43}~\HL{kanu}{\B{ka\-nu}})

% nìkeftxo
\hi
\HT{nIkeftxo}{\B{nìke\cp{ftxo}}} \I{[nI.kE."ft'o]} \emph{adv.} unfortunately, sadly. 
\B{Nì\-ke\-ftxo lu fì\-'e\-veng\-ur 'aw\-a sa'\-sem nì\-'aw.} Unfortunately this child has only one parent.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
\B{Nì\-ke\-ftxo oe\-ri ke tam.} Unfortunately, it doesn't work for me.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1116}{b}
(\ding{43}~\HL{keftxo}{\B{ke\-ftxo}})

% nìkelkin
\hi
\HT{nIkelkin}{\B{nìkel\cp{kin}}} \I{[nI.kEl."kin]} \emph{adv.} unnecessarily.
\B{Sna\-fpìl\-fya\-ri le\-Na'\-vi krra smar\-it fkol tspang, tsran\-ten nì\-txan fwa po ke ngä\-'än nì\-kel\-kin.} It's important in Na'vi philosophy that the prey not suffer unnecessarily when it's killed.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{kelkin}{\B{kel\-kin}})

% nìkemweypey
\hi
\HT{nIkemweypey}{\B{nìkem\cp{wey}pey}} \I{[nI.kEm."wEj.pEj]} \emph{adv.} impatiently.
(\ding{43}~\HL{maweypey}{\B{ma\-wey\-pey}}, \HL{ke}{\B{ke}})

% nìklonu
\hi
\HT{nIklonu}{\B{nìklo\cp{nu}}} \I{[nIk\textcorner.lo."nu]} \emph{adv.} firmly, steadfastly, faithfully (\emph{without releasing it}).
\B{fyep nìk\-lo\-nu}, hold firmly.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{nip nìk\-lo\-nu}, stuck tightly.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 
(\ding{43}~\HL{lonu}{\B{lo\-nu}})

% nìkmar
\hi
\HT{nIkmar}{\B{nìk\cp{mar}}} \I{[nIk\textcorner."maR]} \emph{adv.} in the right season, opportunely.
\B{Po tsap\-'a\-lu\-te so\-li nìk\-mar.} He apologized at the right time.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Nìk\-mar zup tom\-pa set.} Now is a good time for it to be raining.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Fo ke pe\-räng\-kxo oe\-hu nìk\-mar.} They were chatting with me at a bad time.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{kImar}{\B{kì\-mar}}) 

% nìksman
\hi
\HT{nIksman}{\B{nìk\cp{sman}}} \I{[nIk\textcorner."sman]} \emph{adv.} wonderfully.
\B{Fol lì'\-fya\-ti aw\-nge\-yä sar nìk\-sman.} They use our language wonderfully.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{kosman}{\B{ko\-sman}})

% nìksran
\hi
\HT{nIksran}{\B{nìk\cp{sran}}} \I{[nIk\textcorner."sRan]} \emph{adv.} in a mediocre manner.
\B{Pll\-txe nìk\-sran, slä tsun fko pe\-yä ay\-lì\-'ut tsli\-vam.} His speech is mediocre, but you can understand him.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{kesran}{\B{ke\-sran}})

% nìktungzup
\hi
\HT{nIktungzup}{\B{nìktung\cp{zup}}} \I{[nIk\textcorner.tuN."zup\textcorner]} \emph{or} \I{[.tUN."zUp\textcorner]} (\textsc{rn}: \B{nìktùng\cp{zup}} \I{[nIk\textcorner.tUN."zup\textcorner]}) \emph{adv.} carefully, firmly (\emph{without letting it fall}).
\B{fyep nìk\-tung\-zup}, hold firmly.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{tungzup}{\B{tung\-zup}})

% nìkx
\hi
\HT{nIkx}{\B{nìkx}} \I{[nIk']} \emph{n.} gravel.

% nìkxap
\hi
\HT{nIkxap}{\B{nì\cp{kxap}}} \I{[nI."k'ap\textcorner]} \emph{adv.} threateningly.
\B{Txo\-pu rä\-'ä si. Nga\-ti ke nìn pol nì\-kxap. Lu le\-no\-mum nì\-'aw.} Don't be scared. He's not looking at you threateningly. He's just curious.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{kxap}{\B{kxap}})

% nìkxäl
\hi 
\HT{nIkxAl}{\B{nì\cp{kxäl}}} \I{[nI."k'\ae{}l]} \emph{adv.} apathetically. 
(\ding{43}~\HL{kxAl}{\B{kxäl}}) 

% nìkxem
\hi
\HT{nIkxem}{\B{nì\cp{kxem}}} \I{[nI."k'Em]} \emph{adv.} vertically.  
(\ding{43}~\HL{kxem}{\B{kxem}})

% nìlam
\hi
\HT{nIlam}{\B{nì\cp{lam}}} \I{[nI."lam]} \emph{adv.} apparently.
\B{Po\-an yaw\-ne la\-tsu poe\-ru nì\-lam.} Apparently she loves him.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{Tsat ngal tswo\-lä\-nga' nì\-lam.} Apparently you've forgotten it.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
(\ding{43}~\HL{lam}{\B{lam}})

% nìlaw
\hi
\HT{nIlaw}{\B{nì\cp{law}}} \I{[nI."law]} \emph{adv.} clearly (\emph{sentence and manner adverbial}). 
\B{Poe pol\-txe nì\-law.} She spoke clearly.\slash Clearly, she spoke.~\LINK{http://forum.learnnavi.org/language-updates/why-is-this-night/msg154875/\#msg154875}{ln}
\B{Nì\-law po yaw\-ne lu ngar.} It's clear you love her.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{law}{\B{law}})

% nìlayl
\hi 
\HT{nIlayl}{\B{nì\cp{layl}}} \I{[nI."lajl]} \emph{adv.} innocently. 
\B{zaw\-prr\-te' nì\-layl fko\-ne}, be pleasurable to one in an innocent way. 
\B{Tì\-mun\-txa me\-fe\-yä zo\-law\-prr\-te' Ma\-ru\-ne nì\-layl. Ke lu por kea fmokx kaw\-'it.} Their marriage brought Ma\-ru pleasure. She felt no jealousy at all.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(\ding{43}~\HL{layl}{\B{layl}}; \ding{214}~\HL{nItokat}{\B{nì\-to\-kat}}) 

% nìleyn
\hi 
\HT{nIleyn}{\B{nì\cp{leyn}}} \I{[nI."lEjn]} \emph{adv.} repeatedly. 
(\ding{43}~\HL{leyn}{\B{leyn}}) 

% nìler
\hi 
\HT{nIler}{\B{nì\cp{ler}}} \I{[nI."lER]} \emph{adv.} (\emph{a. of motion}) steadily. 
\B{Nì\-sngä\-'i ke tsun Tsyeyk tswi\-vay\-on nì\-ler.} At first, Jake couldn't fly steadily.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
(\emph{b. metaphorically of smooth, unbroken action}) 
\B{tì\-kang\-kem si nì\-ler}, to work steadily (without stopping).~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{ler}{\B{ler}})

% nìli
\hi
\HT{nIli}{\B{nì\cp{li}}} \I{[nI."li]} \emph{adv.} in advance. 
\B{Nge\-yä stxe\-li\-ri a\-lor oe new nga\-ru pi\-vll\-txe san irayo sìk nì\-li.} I want to thank you in advance for your beautiful gift.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{li}{\B{li}})

% nìlkeftang
\hi
\HT{nIlkeftang}{\B{nìlke\cp{ftang}}} \I{[nIl.kE."ftaN]} \emph{adv.} continuously, without stopping. 
\B{Nga fwe\-fwi nìl\-ke\-ftang a fì\-'u lu sä\-srätx a\-txan oe\-ru.} It's a major annoyance to me that you whistle continuously.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{ftang}{\B{ftang}}, \HL{luke}{\B{lu\-ke}})

% nìloho
\hi 
\HT{nIloho}{\B{nì\cp{lo}ho}} \I{[nI."lo.ho]} \emph{adv.} surprisingly. 
\B{Pol\-txe po nì\-lo\-ho san oe za\-sya\-'u.} Surprisingly, he said he would come.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{loho}{\B{lo\-ho}})

% nìltsan
\hi
\HT{nIltsan}{\B{nìl\cp{tsan}}} \I{[nIl."\t{ts}an]} \emph{adv.} well.
\B{Pll\-txe nga nìl\-tsan!} You speak well!~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll}
\B{Tì\-ste\-ftxaw nge\-yä flo\-lä nìl\-tsan!} Your examination succeeded well!~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
\B{Ay\-lì\-'u a\-paw\-nll\-txe nìl\-tsan!} Words well spoken!~\LINK{http://forum.learnnavi.org/language-updates/a-new-participle-infix-awn}{ln}
\B{Pol ye\-rik\-it ta\-ron nìl\-tsan fra\-to.} He hunts yerik better than anyone.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
(i.) \B{Sey\-so\-nìl\-\cp{tsan}!} \I{[sEj.so.nIl."\t{ts}an]} Well done! (\emph{from} \B{ha\-sey so\-li nìl\-tsan})~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
(\ding{43}~\HL{sIltsan}{\B{sìl\-tsan}})

% nìlun
\hi
\HT{nIlun}{\B{nì\cp{lun}}} \I{[nI."lun]} \emph{or} \I{[."lUn]} \emph{adv.} of course, logically, following common sense. 
\B{Nì\-lun ay\-ioi a\-'eoio ay\-eyk\-tan\-ä lu lor fra\-to.} Of course the ceremonial wardrobes of the leaders are the most beautiful.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
(\ding{43}~\HL{lun}{\B{lun}})  

% nìmal
\hi
\HT{nImal}{\B{nì\cp{mal}}} \I{[nI."mal]} \emph{adv.} trustingly, without hesitation.  
\B{Ri\-ni tsa\-po\-hu ho\-lum nì\-mal nì\-wotx.} Rini left with that guy without thinking twice about it.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{mal}{\B{mal}})

% nìmeyp
\hi
\HT{nImeyp}{\B{nì\cp{meyp}}} \I{[nI."mEjp\textcorner]} \emph{adv.} weakly, loosely.
\B{fyep nì\-meyp}, hold loosely.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{nip nì\-meyp}, weakly, loosely stuck.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 
(\ding{43}~\HL{meyp}{\B{meyp}})

% nìmu'nitkan
\hi 
\HT{nImu'nitkan}{\B{nì\cp{mu'}nitkan}} \I{[nI."muP.nit\textcorner.kan]} \emph{or} \I{[."mUP.]} \emph{adv.} effectively, in such a way as to reach a target. 
(\ding{43}~\HL{mu'nitkan}{\B{mu'\-nit\-kan}}) 

% nìmun
\hi
\HT{nImun}{\B{nì\cp{mun}}} \I{[nI."mun]} \emph{or} \I{[."mUn]} \emph{adv.} again.
\B{Nga\-ru irayo sei\-yi oe nì\-mun.} I thank you again.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{Zi\-va\-'u nì\-mun ye'\-rìn.} Come again soon.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Kal\-txì nì\-mun, ma oe\-yä ey\-lan.} Hello again, my friends.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{mune}{\B{mu\-ne}})

% nìmwey
\hi
\HT{nImwey}{\B{nìm\cp{wey}}} \I{[nIm."wEj]} \emph{adv.} calmly, peacefully. 
(\emph{a. idiom}) \B{hi\-va\-haw nìm\-wey} \I{[hi."va.haw nIm."wEj]} ``goodnight''.   (when it's bedtime).~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\ding{43}~\HL{mawey}{\B{ma\-wey}})

% nìmweypey
\hi
\HT{nImweypey}{\B{nìm\cp{wey}pey}} \I{[nIm."wEj.pEj]} \emph{adv.} patiently.
(\ding{43}~\HL{maweypey}{\B{ma\-wey\-pey}})

% nìn
\hi
\HT{nIn}{\B{nìn}} \I{[nIn]} \emph{vtr.} look at (\texttt{+}\emph{control}). 
\B{Po\-ti nìn!} Look at him!~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Nìn tsat!} Look at that!~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Ru\-txe oe\-yä tì\-oeyk\-tìng\-it a\-saw\-nung ni\-vìn.} Please look at my added explanation.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-581}{b}
\B{Txe\-wìl tuk\-rut Lo\-ak\-ä nar\-mìn nì\-fmokx.} Txewì was eyeing Loak's spear enviously.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{tIng}{\B{tìng na\-ri}})

% nìNa'vi
\hi
\HT{nIna'vi}{\B{nì\cp{Na'}vi}} \I{[nI."naP.vi]} \emph{adv.} like the Na'\-vi. 
\B{Tsun oe nga\-hu nì\-Na'\-vi pi\-väng\-kxo a fì\-'u oe\-ru prr\-te' lu.} It's a pleasure to be able to chat with you in Na'vi.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Nga nì\-Na'\-vi pll\-txe nìl\-tsan.} You speak Na'\-vi well.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\emph{a. idiom}) \B{X nì\-Na'\-vi slu lì\-'u\-pe\slash pe\-lì\-'u?} How do you say X in Na'\-vi?~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b}
(\ding{43}~\HL{na'vi}{\B{Na'\-vi}})

% nìnäk
\hi
\HT{nInAk}{\B{nì\cp{näk}}} \I{[nI."n\ae{}k\textcorner]} \emph{adv.} by drinking, in a liquid way. 
\B{Po tä\-pey\-kì\-ye\-ver\-kei\-up nì\-näk.} I'm so jazzed that he may be about to drink himself to death.~\LINK{http://naviteri.org/2010/12/catching-up/}{b}
(\ding{43}~\HL{nAk}{\B{näk}}) 

% nìnän
\hi
\HT{nInAn}{\B{nì\cp{nän}}} \I{[nI."n\ae{}n]} \emph{adv.} less.   
\B{Ru\-txe wi\-vem nì\-nän.} Please fight less.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
\B{Ay\-ha\-pxì\-tu po\-ngu\-ä txo\-pu si nì\-nän ta\-krr\-a Va'\-rul pxe\-ku\-tut lä\-txayn.} The members of the group are less afraid since Va'ru defeated three of the enemy.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{nAn}{\B{nän}}, \ding{214}~\HL{nI'ul}{\B{nì\-'ul}})

% nìnänän
\hi
\HT{nInAnAn}{\B{nì\cp{nä}nän}} \I{[nI."n\ae.n\ae{}n]} \emph{adv.} decreasingly, less and less.
\B{Fra\-lo a ta\-ron, oe\-yä 'i\-tan txo\-pu si nì\-nä\-nän.} Each time he hunts, my son becomes less and less afraid.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{nInAn}{\B{nì\-nän}}, \ding{214}~\HL{nI'ul'ul}{\B{nì\-'ul\-'ul}})

% nìnew
\hi 
\HT{nInew}{\B{nì\cp{new}}} \I{[nI."nEw]} \emph{adv.} voluntarily, willingly, by desire. 
\B{Nga tsa\-kem so\-li nì\-new srak?!} You did that without being asked to?!~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
\B{Tsa\-sä\-num\-vit oel po\-ru ka\-yei\-ar nì\-new!} I'm happy to teach him that lesson!~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
\B{Oel pe\-lun ftxal\-mey nì\-new fu\-ta srung si skxawng\-ur a\-na\-fì\-'u?} Why did I choose, of my own free will, to help such a fool?~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
(\ding{43}~\HL{new}{\B{new}})

% nìno
\hi
\HT{nIno}{\B{nì\cp{no}}} \I{[nI."no]} \emph{adv.} in detail, thoroughly, with attention to detail. 
\B{Nga lä\-pi\-vawk nì\-no ko.} Tell me all about yourself. [lit.: introduce yourself in detail]~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tì\-tu\-sa\-ron\-ìri fte fli\-vä, ze\-ne fko si\-vutx smar\-it ni\-nan nì\-no.} To succeed at hunting, you have to track your prey by reading (the forest) with attention to detail.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
(\ding{43}~\HL{leno}{\B{le\-no}}, \HL{hIno}{\B{hì\-no}}, \HL{tIno}{\B{tì\-no}})

% nìnrra
\hi
\HT{nInrra}{\B{nì\cp{nrr}a}} \I{[nI."n\s{r}.a]} \emph{adv.} proudly, with pride.   
\B{Sa'\-nok nì\-nrr\-a lrr\-tok so\-li krr\-a prr\-nen a\-lo a\-'aw\-ve pol\-txe.} The mother smiled with pride when the baby spoke for the first time.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{nrra}{\B{nrra}})

% nìnu
\hi
\HT{nInu}{\B{nì\cp{nu}}} \I{[nI."nu]} \emph{adv.} failingly, falteringly, in vain, fruitlessly; \emph{not achieving the desired or expected end}.
\B{Oe\-ru txoa li\-vu. Pol\-txä\-nge nì\-nu.} Forgive me. I misspoke.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Kll\-te lu ekx\-txu. Na\-ri si txo\-ke\-fyaw tì\-ran nì\-nu.} The ground is rough. Be careful or you'll trip.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{nui}{\B{nui}})

% nìngay
\hi
\HT{nIngay}{\B{nì\cp{ngay}}} \I{[nI."Naj]} \emph{adv.} truly.   
\B{Oe\-yä swi\-zaw nì\-ngay ti\-va\-kuk.} Let my arrow strike true.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Ay\-lì\-'u nge\-yä lor lu nì\-ngay.} Your words are truly beautiful.~\LINK{http://forum.learnnavi.org/language-updates/combined-answers-1-\%28feb-16\%29/}{ln}
\B{Nì\-ngay pll\-txe nga.} You speak truth.~\LINK{http://forum.learnnavi.org/language-updates/confirmation-on-kurakx-eytukan-line-and-toitslan/}{ln\slash a\textsuperscript{c}}
\B{Zì\-sìt a\on{104}°! Ko\-ak nì\-ngay, ke\-fyak?} \on{68} years! Really old, right?~\LINK{http://forum.learnnavi.org/news-announcements/happy-birthday-to-karyu-pawl/msg556899/\#msg556899}{ln}
(\ding{43}~\HL{ngay}{\B{ngay}})

% nìngong
\hi
\HT{nIngong}{\B{nì\cp{ngong}}} \I{[nI."NoN]} \emph{adv.} lethargically, lazily.
(\ding{43}~\HL{ngong}{\B{ngong}})

% nìolo'
\hi
\HT{nIolo'}{\B{nìol\cp{o'}}} \I{[nI.o."loP]} \emph{adv.} (together) as members of a clan.
\B{Txil\-te Ri\-ni\-sì tä\-pa\-re fì\-tsap nì\-soaia, slä tsal\-su\-ngay ke nì\-olo' ta\-krr\-a Ri\-ni mun\-txa slo\-lu.} Txilte and Rini are related by blood, but nevertheless not by clan since the time Rini got married.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{olo'}{\B{olo'}}) 

% nìprrte'
\hi
\HT{nIprrte'}{\B{nì\cp{prr}te'}} \I{[nI."p\s{r}.tEP]} \emph{adv.} gladly, with pleasure.
\B{Fì\-'u\-pxa\-ret a\-syi\-nan oel nì\-prr\-te'.} I intend to read this message gladly.
(\emph{a. idiom}) \B{Zo\-la\-'u nì\-prr\-te'.} ``Welcome.'' (\HL{ne}{\B{ne}}, to)~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{b.}) \B{Smon nì\-prr\-te'.} ``Nice to know you.''~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b}
(\emph{c.}) \B{To\-lä\-txaw nì\-prr\-te'.} ``Welcome back.''~\LINK{http://forum.learnnavi.org/language-updates/some-news-from-berlin/msg585072/\#msg585072}{ln}  
(\emph{d. responding to thanks}) \B{Nì\-prr\-te'.} ``You're welcome.''~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\emph{e.}) \B{sre\-fe\-re\-i\-ey nì\-prr\-te'}, looking forward (to that). \B{Tsa\-ria nga\-hu ye'\-rìn ul\-txa si nì\-mun, oe sre\-fe\-re\-i\-ey nì\-prr\-te'.} I'm looking forward to getting together with you again soon.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\ding{43}~\HL{prrte'}{\B{prr\-te'}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{zawprrte'}{\B{zaw\-prr\-te'}}, be enjoyable.

% nìpxay
\hi
\HT{nIpxay}{\B{nì\cp{pxay}}} \I{[nI."p'aj]} \emph{adv.} many, heavily.
\B{Fwa ay\-ngal fì\-tse\-nget tok nì\-pxay nì\-ftxan oe\-ru te\-ya si nì\-ngay.} It fills me with joy that you are here in such a great number.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\emph{a. of weather}) \B{lep\-wopx nì\-pxay}, heavily cloud\hyp{}covered, many clouds.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{pxay}{\B{pxay}})

% nìpxi
\hi
\HT{nIpxi}{\B{nì\cp{pxi}}} \I{[nI."p'i]} \emph{adv.} especially, pointedly, unambiguously. 
\B{Por war\-mou fra\-fne\-i\-o\-ang nì\-tut, slä nì\-pxi pxe\-syì\-rä\-fì.} He just couldn't get over all the animals, but was especially taken with the three giraffe.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Po pa\-mll\-txe a krr, fra\-po tar\-mìng mik\-yun nì\-pxi, ta\-lu\-na mok\-ri lu sä\-tsì\-syì\-tsyìp.} When he spoke, everyone listened intently, because his voice was a tiny whisper.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{pxi}{\B{pxi}})

% nìpxim
\hi
\HT{nIpxim}{\B{nì\cp{pxim}}} \I{[nI."p'im]} \emph{adv.} erectly, uprightly. 
\B{Ton\-ìri a\-la\-he, aw\-ngal yom wu\-tsot teng\-krr he\-reyn nì\-pxim.} As for other nights, we eat dinner while sitting erect.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
(\ding{43}~\HL{pxim}{\B{pxim}})

% nìpxul
\hi 
\HT{nIpxul}{\B{nì\cp{pxul}}} \I{[nI."p'ul]} \emph{or} \I{[.p'Ul]} \emph{adv.} formidably, imposingly.~\LINK{http://naviteri.org/2020/08/hiia-vur-a-teri-mefalukantsyip-a-little-story-about-two-cats/}{b}  
(\ding{43}~\HL{pxul}{\B{pxul}})

% nìran
\hi
\HT{nIran}{\B{nì\cp{ran}}} \I{[nI."Ran]} \emph{adv.} basically, fundamentally, in essence.
\B{Nì\-ran lu Lo\-ak mi 'e\-veng slä tsun ti\-va\-ron nì\-teng\-fya na fyeyn\-tu.} Loak is still really just a boy but he can hunt the same as an adult.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{ran}{\B{ran}}) 

% nìrangal
\hi
\HT{nIrangal}{\B{nì\cp{rang}al}} \I{[nI."RaN.al]} \emph{adv.} ``I wish …'', ``oh that …'' (\emph{indicating counterfactuals}) 
(\emph{a. with the imperfective subjunctive} ‹\B{irv}› \emph{for a present counterfactual})
\B{Po\-el nì\-rang\-al tir\-vok fì\-tseng\-it.} I wish she were here. (But I know she's not.)~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{b. with the perfective subjunctive} ‹\B{ilv}› \emph{or past subjunctive} ‹\B{imv}› \emph{for a past counterfactual})
\B{Poe nì\-rang\-al zim\-va\-'u\slash zil\-va\-'u trr\-am.} I wish she had come yesterday. (But she didn't.)~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{c. with the future subjunctive} ‹\B{iyev}›\slash ‹\B{ìyev}› \emph{for a future counterfactual})
\B{Poe nì\-rang\-al zì\-ye\-va\-'u trr\-ay.} I wish she were coming tomorrow. (But I know she isn't.)~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{rangal}{\B{rang\-al}}, \HL{nIsIltsan}{\B{nì\-sìl\-pey}})

% nìräptum
\hi 
\HT{nIrAptum}{\B{nìräp\cp{tum}}} \I{[nI.R\ae{}p\textcorner."tum]} \emph{or} \I{[."tUm]} (\textsc{rn}: \B{nì\-räp\-tùm} \I{[nI.R\ae{}p\textcorner."tUm]}) \emph{adv.} coarsly, vulgarly. 
\B{Fya\-pe yaw\-ne lu fko\-ru tu\-te a fra\-krr vo\-ìk si fì\-txan nì\-räp\-tum?} How does one love a person who always behaves so coarsely?~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{rAptum}{\B{räp\-tum}})

% nìrim
\hi 
\HT{nIrim}{\B{nì\cp{rim}}} \I{[nI."Rim]} \emph{adv.} in yellow, `yellow-ly'.
\B{Tom\-pa\-yä ka\-to | Tsaw\-ke\-yä ka\-to | Ka\-tot tä\-ftxu oel | Nì\-ean nì\-rim.} The rhythm of the rain | The rhythm of the sun | I weave the rhythm | In blue and yellow.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
(\ding{43}~\HL{rim}{\B{rim}})

% nìrìkxi
\hi 
\HT{nIrIkxi}{\B{nìrì\cp{kxi}}} \I{[nI.RI."k'i]} \emph{adv.} shakily, tremblingly.
(\ding{43}~\HL{rIkxi}{\B{rì\-kxi}})

% nìronsrel
\hi
\HT{nIronsrel}{\B{nì\cp{ron}srel}} \I{[nI."Ron.sREl]} \emph{adv.} in\slash by imagination.
\B{Oe pxìm päng\-kxo nga\-hu nì\-ron\-srel.} I often talk with you in my imagination.\slash I often imagine I'm talking with you.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\ding{43}~\HL{ronsrel}{\B{ron\-srel}})

% nìsay
\hi 
\HT{nIsay}{\B{nì\cp{say}}} \I{[nI."{}saj]} \emph{adv.} loyally.  
(\ding{43}~\HL{say}{\B{say}})

% nìsìlpey
\hi
\HT{nIsIlpey}{\B{nìsìl\cp{pey}}} \I{[nI.sIl."pEj]} \textsc{i}. \emph{sentence adv.} hopefully (\emph{expresses a hope that s.t. is true}) 
(\emph{a. with the subjunctive} ‹\B{iv}› \emph{for a present wish})
\B{Po\-el nì\-sìl\-pey ti\-vok fì\-tseng\-it.} I hope she's here.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{b. with the perfective subjunctive} ‹\B{ilv}› \emph{or past subjunctive} ‹\B{imv}› \emph{for a past wish})
\B{Poe nì\-sìl\-pey zim\-va\-'u\slash zil\-va\-'u trr\-am.} I hope she came yesterday.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{c. with the future subjunctive} ‹\B{iyev}›\slash ‹\B{ìyev}› \emph{for a future wish})
\B{Poe nì\-sìl\-pey zì\-ye\-va\-'u trr\-ay.} I hope she'll come tomorrow.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \\
\textsc{ii}. \emph{manner adv.} in a hopeful way.
\B{Tsyeyk ä\-txä\-le so\-li nì\-sìl\-pey tsnì li\-vu por Unil\-ta\-ron.} Jake hopefully requested the Dream Hunt.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{sIlpey}{\B{sìl\-pey}}, \HL{nIrangal}{\B{nì\-rang\-al}})

% nìskien
\hi
\HT{nIskien}{\B{nì\cp{ski}en}} \I{[nI."{}ski.En]} \emph{adv.} to the right.
\B{Yak si nì\-ski\-en.} Turn to the right.~\ding{43}~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{skien}{\B{ski\-en}}, \ding{214}~\HL{nIftAr}{\B{nì\-ftär}})

% nìslele
\hi
\HT{nIslele}{\B{nì\cp{sle}le}} \I{[nI."{}slE.lE]} \emph{adv.} by swimming.
\B{Tsun fko tsa\-tse\-nge\-ne ki\-vä nì\-sle\-le fu fa fwa ik\-ran\-it mak\-to nì\-'aw.} You can only get there by swimming or riding an ikran.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
(\ding{43}~\HL{slele}{\B{sle\-le}})

% nìsngä'i
\hi
\HT{nIsngA'i}{\B{nì\cp{sngä}'i}} \I{[nI."{}sN\ae.Pi]} \emph{adv.} originally, at first.  
\B{Nì\-sngä\-'i fmawn\-it fo nar\-mew wi\-van, slä nì\-'i'\-a fra\-por lo\-lo\-nu.} They originally wanted to hide the news, but in the end they revealed it everyone.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{sngA'i}{\B{sngä\-'i}}, \HL{nItsim}{\B{nì\-tsim}})

% nìsngum 
\hi
\HT{nIsngum}{\B{nì\cp{sngum}}} \I{[nI."{}sNum]} \emph{or} \I{[."{}sNUm]} \emph{adv.} worryingly, fretfully. 
\B{Swey lu fwa nga fì\-kem ke si\-vi nì\-sngum.} It's best that you not do this fretfully.\slash It's best that you don't freak out about doing this.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{sngum}{\B{sngum}})

% nìso'ha
\hi
\HT{nIso'ha}{\B{nì\cp{so'}ha}} \I{[nI."{}soP.ha]} \emph{adv.} enthusiastically.
(\ding{43}~\HL{so'ha}{\B{so'\-ha}})

% nìsoaia
\hi
\HT{nIsoaia}{\B{nìso\cp{a}ia}} \I{[nI.so."{}a.i.a]} \emph{adv.} (together) as members of a family. 
\B{Txil\-te Ri\-ni\-sì tä\-pa\-re fì\-tsap nì\-soaia, slä tsal\-su\-ngay ke nì\-olo' ta\-krr\-a Ri\-ni mun\-txa slo\-lu.} Txilte and Rini are related by blood, but nevertheless not by clan since the time Rini got married.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{soaia}{\B{soaia}})

% nìsok
\hi 
\HT{nIsok}{\B{nì\cp{sok}}} \I{[nI."{}sok\textcorner]} \emph{adv.} recently. 
(\ding{43}~\HL{sok}{\B{sok}})

% nìsteng
\hi
\HT{nIsteng}{\B{nì\cp{steng}}} \I{[nI."{}stEN]} \emph{adv.} similarly, in a similar fashion.
\B{Tseyk tswa\-may\-on fa ik\-ran sre\-krr; ta\-fral fmo\-li fì\-kem si\-vi fa to\-ruk nì\-steng.} Jake flew with an ikran before; therefore he tried to do it with a toruk in a similar fashion.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\ding{43}~\HL{steng}{\B{steng}})

% nìsti
\hi
\HT{nIsti}{\B{nì\cp{sti}}} \I{[nI."{}sti]} \emph{adv.} angrily. 
(\ding{43}~\HL{sti}{\B{sti}})

% nìsung
\hi
\HT{nIsung}{\B{nì\cp{sung}}} \I{[nI."{}suN]} \emph{or} \I{[."{}sUN]} (\textsc{rn}: \B{nì\cp{sùng}} \I{[nI."{}sUN]}) \emph{adv.} besides, additionally, furthermore.
\B{Nì\-sung, txo tsa\-kem li\-ven mì zu\-saw\-krr, ze\-ne fko si\-var kem\-lì\-'u\-vit a\-lu ‹ìyev›.} Furthermore, if that happens in the future, one must use the infix ‹\B{ìyev}›.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1910}{b}
(\ding{43}~\HL{sung}{\B{sung}})

% nìswey
\hi
\HT{nIswey}{\B{nì\cp{swey}}} \I{[nI."{}swEj]} \emph{adv.} optimally, best.
\B{Maw\-krr la\-ye\-iu oer krr nì\-'ul fte ngi\-vop ay\-lì\-'ut sì tsay\-fne\-sä\-num\-vit a tsun fra\-por srung si\-vi fte ni\-vu\-me sì zi\-ve\-rok nì\-swey.} After that I will have more time to create words and the kinds of lessons that can help everyone best learn and remember.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
(\ding{43}~\HL{swey}{\B{swey}})

% nìsyayvi
\hi
\HT{nIsyayvi}{\B{nì\cp{syay}vi}} \I{[nI."{}sjaj.vi]} \emph{adv.} by chance, by coincidence, as luck would have it.
\B{Et\-rìp\-a syay\-vi!} Good luck!~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}.
(\ding{43}~\HL{syayvi}{\B{syay\-vi}})

% nìsyen
\hi
\HT{nIsyen}{\B{nì\cp{syen}}} \I{[nI."{}sjEn]} \emph{adv.} lastly, finally.
\B{Ul\-te nì\-syen …}, And lastly …~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{syen}{\B{syen}}, \HL{nI'i'a}{\B{nì'i'a}})

% nìsyep
\hi
\HT{nIsyep}{\B{nì\cp{syep}}} \I{[nI."{}sjEp\textcorner]} \emph{adv.} tightly, in an iron grip.
\B{fyep nì\-syep}, hold tightly.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{nip nì\-syep}, stuck irremovably.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 
(\ding{43}~\HL{syep}{\B{syep}})

% nìt'iluke
\hi 
\HT{nIt'iluke}{\B{nìt\cp{'i}luke}} \I{[nIt\textcorner."{}i.lu.kE]} \emph{adv.} never\hyp{}endingly, forever. 
(\ding{43}~\HL{tI'iluke}{\B{tì\-'i\-lu\-ke}})

% nìtam
\hi
\HT{nItam}{\B{nì\cp{tam}}} \I{[nI."tam]} \emph{adv.} enough.
\B{Hu\-fwa nge\-yä tì\-hawl\-ìri ke lu kea kxey\-ey, tsal\-su\-ngay oe\-ru ke ha' nì\-tam.} Although there's nothing wrong with your plan, it just doesn't suit me.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{tam}{\B{tam}})

% nìteng
\hi
\HT{nIteng}{\B{nì\cp{teng}}} \I{[nI."tEN]} \emph{adv.} too, likewise, as well.   
\B{Nì\-ay\-nga oe pe\-rey nì\-teng.} Like you, I too am waiting.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Lu poe lor, lu yu\-ey nì\-teng.} She's beautiful on the outside \emph{and} the inside.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{teng}{\B{teng}}, \HL{kop}{\B{kop}})

% nìtengfya
\hi
\HT{nItengfya}{\B{nì\cp{teng}fya}} \I{[nI."tEN.fja]} \emph{adv.} in the same way. 
\B{Oel tel 'u\-pxa\-ret le\-Na'\-vi a krr, new oe nì\-teng\-fya pam\-rel si\-vi nì\-'eyng.} When I receive a Na'vi message, I want to write the same way in response.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln}
(\ding{43}~\HL{tengfya}{\B{teng\-fya}})

% nìtkan
\hi
\HT{nItkan}{\B{nìt\cp{kan}}} \I{[nIt\textcorner."kan]} \emph{adv.} purposefully, deliberately.
(\ding{43}~\HL{tIkan}{\B{tì\-kan}})

% nìtkanluke
\hi
\HT{nItkanluke}{\B{nìt\cp{kan}luke}} \I{[nIt\textcorner."kan.lu.kE]} \emph{adv.} accidentally, unintentionally.
(\ding{43}~\HL{tIkan}{\B{tì\-kan}}, \HL{luke}{\B{lu\-ke}})

% nìToitsye
\hi
\HT{nItoitsye}{\B{nì\cp{To}itsye}} \I{[nI."to.i.\t{ts}jE]} \emph{adv.} German, in German.
\B{X-(ì)ri pe\-ral nì\-To\-i\-tsye?} What does X mean in German?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\ding{43}~\HL{toitsye}{\B{Toi\-tsye}})

% nìtokat
\hi 
\HT{nItokat}{\B{nì\cp{to}kat}} \I{[nI."to.kat\textcorner]} \emph{adv.} guiltily. 
\B{zaw\-prr\-te' nì\-to\-kat fko\-ne}, be pleasurable to one in a guilty way, \emph{schadenfreude, i.e. taking pleasure in s.o. else's misfortune}.
\B{En\-tul fot ve'\-kì ul\-te sä\-nui fe\-yä zo\-law\-prr\-te' nì\-to\-kat po\-ne.} En\-tu hates them and their failure brought him pleasure.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}  
(\ding{43}~\HL{tokat}{\B{to\-kat}}; \ding{214}~\HL{nIlayl}{\B{nì\-layl}})

% nìtrea
\hi 
\HT{nItrea}{\B{nìt\cp{re}a}} \I{[nIt\textcorner."RE.a]} \emph{adv.} in spirit. 
\B{Ke tsä\-ngun Tsyìm zi\-va'\-u ftxo\-zä\-ne, slä tok nìt\-re\-a.} Sadly, Jim couldn't come to the celebration, but he was there in spirit.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 
(\ding{43}~\HL{tirea}{\B{ti\-re\-a}})

% nìtrrtrr
\hi
\HT{nItrrtrr}{\B{nì\cp{trr}trr}} \I{[nI."t\s{r}.t\s{r}]} \emph{adv.} daily, regularly, on a daily basis, typically.   
\B{Nga\-ru lu pe\-fne\-txin\-tìn nì\-trr\-trr?} What is your primary role (in society)?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Nga pe\-su\-hu kä\-teng nì\-trr\-trr?} Who do you usually hang out with?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Nì\-trr\-trr yom Na'\-vil wu\-tsot 'aw\-si\-teng pxaw yll\-txep.} The Na'vi regularly eat dinner together around a communal fire.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Nì\-trr\-trr pxì\-mun\-'i sam\-si\-yul ay\-swi\-zaw\-it ku\-tu\-ä a\-law\-nä\-txayn sno\-kip nì\-'eng.} Warriors typically share the arrows of their defeated enemies among themselves.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{letrrtrr}{\B{le\-trr\-trr}})

% nìtut
\hi
\HT{nItut}{\B{nì\cp{tut}}} \I{[nI."tut\textcorner]} \emph{or} \I{[."tUt\textcorner]} (\textsc{rn}: \B{nì\cp{tùt}} \I{[nI."tUt\textcorner]}) \emph{adv.} continually, continuously, always.
\B{Ay\-nge\-yä ki\-fkey\-ti fol kawng\-sar nì\-tut, fì\-'u\-ri ke\-kem ke si ay\-nga.} They continuously exploit your world and you do nothing about it.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Ri\-ni\-ri nik\-re yä\-ngo' nì\-tut.} Ri\-ni's hair is always perfect.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Por war\-mou fra\-fne\-i\-o\-ang nì\-tut, slä nì\-pxi pxe\-syì\-rä\-fì.} He just couldn't get over all the animals, but was especially taken with the three giraffe.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Mong prr\-nen\-ìl fu\-ta sa'\-nul ve\-rewng nì\-tut.} The infant depends on her Mommy to look after her constantly.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Pa\-lu\-lu\-kan\-ìri lu pxe\-syon a ze\-ne fko zi\-ve\-rok nì\-tut.} Three things about the thanator must always be kept in mind.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{tut}{\B{tut}})

% nìtxan
\hi
\HT{nItxan}{\B{nì\cp{txan}}} \I{[nI."t'an]} \emph{adv.} (\emph{a.}) very much, very.
\B{Oe new nì\-txan ay\-nga\-ru fya\-wi\-vìn\-txu.} I want very much to guide you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Lam oer fwa po lu ka\-nu nì\-txan.} It seems to me that he is very intelligent.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Nga nì\-aw\-no\-mum to oe\-tsyìp lu txur nì\-txan.} As everyone knows, you're a lot stronger than little old me.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Ay\-nge\-yä ay\-sä\-pll\-txe\-vi\-ri sei\-yi oe ira\-yo nì\-txan ay\-nga\-ru nì\-wotx!} I thank you all very much for the comments!~\LINK{http://naviteri.org/2011/07/attributive-\%E2\%80\%9Ca\%E2\%80\%9D-and-truncated-style/}{b}
\B{Ira\-yo nì\-txan.} Thanks very much.~\LINK{http://naviteri.org/2010/06/irayo-thank-you-and-some-miscellaneous-thoughts/}{b}
(\emph{b. of time}) long.   
\B{Fwa oel fì\-tì\-kang\-kem\-vit sngey\-ki\-vä\-'i krr\-no\-lekx nì\-txan.} Me starting this project took a long time.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\ding{43}~\HL{txan}{\B{txan}})

% nìtxankeltrrtrr
\hi 
\HT{nItxankeltrrtrr}{\B{nì\cp{txan}kel\cp{trr}trr}} \I{[nI."t'an.kEl."t\s{r}.t\s{r}]} \emph{adj.} extraordinarily. 
\B{Oey 'ey\-lan pll\-txe nì\-Na'\-vi na hu\-fwe nì\-txan\-kel\-trr\-trr.} My friend speaks Na'vi extraordinarily fluently.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\HL{txankeltrrtrr}{\B{txan\-kel\-trr\-trr}})

% nìtxay
\hi
\HT{nItxay}{\B{nì\cp{txay}}} \I{[nI."t'aj]} \emph{adv.} horizontally.
(\ding{43}~\HL{txay}{\B{txay}})

% nìtxi
\hi
\HT{nItxi}{\B{nì\cp{txi}}} \I{[nI."t'i]} \emph{adv.} hurriedly, in a frenzied way. 
(\ding{43}~\HL{txi}{\B{txi}}; \ding{214}~\HL{nItxiluke}{\B{nì\-txi\-lu\-ke}})

% nìtxiluke
\hi
\HT{nItxiluke}{\B{nì\cp{txi}luke}} \I{[nI."t'i.lu.kE]} \emph{adv.} unhurriedly, leisurely   \B{Tsun oe nga\-hu tsa\-tse\-nge\-ne ki\-vä, slä nul\-new fu\-ta si\-vop oeng nì\-txi\-lu\-ke.} I can go there with you, but I prefer to travel leisurely.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{txi}{\B{txi}}, \HL{luke}{\B{lu\-ke}}; \ding{214}~\HL{nItxi}{\B{nì\-txi}})

% nìtxukx
\hi 
\HT{nItxukx}{\B{nì\cp{txukx}}} \I{[nI."t'uk']} \emph{or} \I{[."t'Uk']} (\textsc{rn}: \B{nìtxukx} \I{[nI."duk']}) \emph{adv.} deeply (\emph{also metaphorically}). 
\B{Po\-an\-it tso\-lukx po\-el fa tstal nì\-txukx nem\-fa heyr.} She stabbed him deeply in the chest with a knife.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
\B{Fì\-tì\-pawm\-te\-ri fpar\-mìl oe nì\-txukx, slä vay set ke ro\-lä\-ngun tì\-'eyng\-it.} I've thought about this question deeply, but I'm sad to say I haven't yet found the answer.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
(\ding{43}~\HL{txukx}{\B{txukx}})

% nìtsim
\hi
\HT{nItsim}{\B{nì\cp{tsim}}} \I{[nI."\t{ts}im]} \emph{adv.} originally, in an original way, with originality.   
\B{Fra\-krr po fpìl nì\-tsim nì\-wotx.} Her thinking is always completely original.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{tsim}{\B{tsim}}, \HL{nIsngA'i}{\B{nì\-sngä\-'i}})

% nìtsìsyì
\hi
\HT{nItsIsyI}{\B{nì\cp{tsì}syì}} \I{[nI."\t{ts}I.sjI]} \emph{adv.} by whispering, in a whisper.   
\B{Pol tstxo\-ti oe\-yä pol\-txe\-i\-e nì\-tsì\-syì.} He whispered my name, I'm happy to say.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{tsIsyI}{\B{tsìsyì}})

% nìtsleng
\hi 
\HT{nItsleng}{\B{nì\cp{tsleng}}} \I{[nI."\t{ts}lEN]} \emph{adv.} falsely. 
\B{Nì\-tsleng pll\-txe nga.} You speak falsely [i.e. you are a liar].~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
\B{Pll\-txe nì\-tsleng! Tsa\-fkxi\-let ke to\-lìng ngar En\-tul!}Liar! En\-tu didn't give you that necklace!~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}  
(\ding{43}~\HL{tsleng}{\B{tsleng}})

% nìtson
\hi
\HT{nItson}{\B{nì\cp{tson}}} \I{[nI."\t{ts}on]} \emph{adv.} dutifully, as an obligation.
\B{Pol vewng fra\-trr ay\-eveng\-it nì\-tson.} He observes his duty to care for the kids every day.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{tson}{\B{tson}})

% nìtstew
\hi 
\HT{nItstew}{\B{nì\cp{tstew}}} \I{[nI."\t{ts}tEw]} \emph{adv.} bravely, courageously.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} 
(\ding{43}~\HL{tstew}{\B{tstew}})

% nìtsyìl
\hi
\HT{nItsyIl}{\B{nì\cp{tsyìl}}} \I{[nI."\t{ts}jIl]} \emph{adv.} by climbing.
(\ding{43}~\HL{tsyIl}{\B{tsyìl}})

% nìTsyungwen
\hi
\HT{nItsyungwen}{\B{nì\cp{Tsyung}wen}} \I{[nI."\t{ts}juN.wEn]} \emph{or} \I{[."\t{ts}jUN.]} (\textsc{rn}: \B{nì\cp{Tsyùng}wen} \I{[nI."\t{tS}UN.wEn]}) \emph{adv.} like the Chinese, in Mandarin.
\B{X-(ì)ri pe\-ral nì\-Tsyung\-wen?} What does X mean in Chinese?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\ding{43}~\HL{tsyungwen}{\B{Tsyung\-wen}})

% nìvingkap
\hi
\HT{nIvingkap}{\B{nì\cp{ving}kap}} \I{[nI."viN.kap\textcorner]} \emph{adv.} by the way, incidentally (\emph{to change the subject and introduce something new into the conversation}). 
\B{Nì\-ving\-kap nge\-yä tsmu\-kan\-ur alu Ra\-lu lu fpom srak?} Oh by the way, how's your brother Ralu?~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
(\ding{43}~\HL{vingkap}{\B{ving\-kap}}, \HL{mIftxele}{\B{mì\-ftxe\-le}})

% nìwan
\hi
\HT{nIwan}{\B{nì\cp{wan}}} \I{[nI."wan]} \emph{adv.} secretly; in hiding, by hiding.  
\B{Sam\-si\-yu pe\-rey nì\-wan.} The hunters lie in wait, prepared to ambush.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
\B{Ìlä Fey\-ral, mun\-txa so\-li Ra\-lu sì Ne\-wey nì\-wan me\-srr\-am.} According to Peyral, Ralu and Newey were secretly married the day before yesterday.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{wan}{\B{wan}})

% nìwawe
\hi
\HT{nIwawe}{\B{nìwa\cp{we}}} \I{[nI.wa."wE]} \emph{adv.} meaningfully, significantly. 
\B{Tsa\-txon ay\-ut\-ral\-äo krr a pol oe\-ti no\-lìn nì\-wa\-we, olo\-me\-i\-um oel fa key\-rel fu\-ta lu yaw\-ne oe po\-ru.} That night beneath the trees when she looked at me meaningfully, I knew by the expression on her face that she loves me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{wawe}{\B{wa\-we}})

% nìwä
\hi
\HT{nIwA}{\B{nì\cp{wä}}} \I{[nI."w\ae]} \emph{adv.} on the contrary, conversely. 
(\ding{43}~\HL{wA}{\B{wä}})

% nìwäte
\hi
\HT{nIwAte}{\B{nìwä\cp{te}}} \I{[nI.w\ae."tE]} \emph{adv.} disagreeably, begrudgingly.
(\ding{43}~\HL{wAte}{\B{wä\-te}}) 

% nìwin
\hi
\HT{nIwin}{\B{nì\cp{win}}} \I{[nI."win]} \emph{adv.} (\emph{a.}) fast, quickly. 
\B{kä\-mak\-to nì\-win}, ride out fast.~\LINK{http://forum.learnnavi.org/language-updates/naviya-the-other-vocative/}{ln\slash a}
\B{Ey\-wal zey\-ki\-vo ngat nì\-win.} May Eywa heal you quickly.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Oe tul nì\-ftxan nì\-win na nga.} I run as fast as you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Oe to nga tul nì\-win.} I run faster than you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Po tul nì\-win fra\-to.} She is the fastest.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
(\emph{b. expression}) \B{nì\-win na hu\-fwe}, ``as fast as the wind'' (\emph{any situation to express rapidity}). \B{na hu\-fwe}, fluent (not just rapid) use of language.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\ding{43}~\HL{win}{\B{win}})

% nìwok
\hi
\HT{nIwok}{\B{nì\cp{wok}}} \I{[nI."wok\textcorner]} \emph{adv.} loudly. 
\B{Ru\-txe pi\-vll\-txe nì\-wok nì\-'it.} Please speak up a bit.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{wok}{\B{wok}})

% nìwotx
\hi
\HT{nIwotx}{\B{nì\cp{wotx}}} \I{[nI."wot']} \emph{adv.} all (of), completely, \emph{in toto}.   
\B{Ay\-ftxo\-zä le\-fpom ay\-nga\-ru nì\-wotx!} Happy Holidays to You All!~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Out_of_Office_AutoReply}{wiki} 
\B{Ay\-ey\-lan\-ur oe\-yä sì ey\-lan\-ur lì'\-fya\-yä le\-Na'\-vi nì\-wotx.} To all my friends and friends of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ri\-ni tsa\-po\-hu ho\-lum nì\-mal nì\-wotx.} Rini left with that guy without thinking twice about it.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Nga\-ru tì\-yawr nì\-wotx!} You are completely right.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/comment-page-1/\#comment-191}{b}
\B{Fti\-a oel lì'\-fya\-ti le\-Na'\-vi nì\-'o' nì\-wotx!} Studying Na'vi is a ton of fun for me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{wotx}{\B{wotx}})

% nìyawr
\hi
\HT{nIyawr}{\B{nì\cp{yawr}}} \I{[nI."jawR]} \emph{adv.} correctly, rightly. 
(\ding{43}~\HL{eyawr}{\B{e\-yawr}})

% nìyewla
\hi
\HT{nIyewla}{\B{nì\cp{yew}la}} \I{[nI."jEw.la]} \emph{adv.} in a disappointing fashion, \emph{in a way failing to meet expectations}.   
\B{Trr\-am\-ä ay\-u\-van\-ìri mak\-to Ak\-wey nì\-yew\-la ha snaytx.} Akwey rode disappointingly in yesterday's games so he lost.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{yewla}{\B{yew\-la}})

% nìyey
\hi
\HT{nIyey}{\B{nì\cp{yey}}} \I{[nI."jEj]} \emph{adv.} directly, straight to the point; just.
\B{Kem si nì\-yey!} Just do it!~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\ding{43}~\HL{yey}{\B{yey}})

% nìyeyfya
\hi
\HT{nIyeyfya}{\B{nì\cp{yey}fya}} \I{[nI."jEj.fja]} \emph{adv.} straight ahead, in a straight line. 
\B{Sa\-lew nì\-yey\-fya.} Proceed straight ahead.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{yey}{\B{yey}}, \HL{fya'o}{\B{fya\-'o}})

% nìyll
\hi
\HT{nIyll}{\B{nì\cp{yll}}} \I{[nI."j\s{l}]} \emph{adv.} communally, in a communal manner (\emph{to express sharing with the community}).
\B{Fol tsngan\-it pxì\-mo\-lun\-'i nì\-yll.} They shared the meat with the entire clan.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Fì\-tey\-lu\-ri ke nar\-mew Va'\-ru yi\-vom nì\-yll.} Va'ru didn't want to share this \emph{teylu} with the Omatikaya.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{yll}{\B{yll}}; \HL{nI'eng}{\B{nì\-'eng}})

% nìyo'
\hi
\HT{nIyo'}{\B{nì\cp{yo'}}} \I{[nI."joP]} \emph{adv.} perfectly, flawlessly.
(\emph{a. proverb}) \B{Txo ke nì\-yo' tsa\-krr nì\-yol.} `If you can't be flawless, at least be brief.'~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{yo'}{\B{yo'}}) 

% nìyol
\hi
\HT{nIyol}{\B{nì\cp{yol}}} \I{[nI."jol]} \emph{adv.}  briefly, shortly (\emph{of time}). 
(\emph{a. proverb}) \B{Txo ke nì\-yo' tsa\-krr nì\-yol.} `If you can't be flawless, at least be brief.'~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{yol}{\B{yol}})

% nìyrr
\hi
\HT{nIyrr}{\B{nì\cp{yrr}}} \I{[nI."j\s{r}]} \emph{adv.} naturally, without tampering with nature, without changing the nature of something. 
\B{Fkxen\-ti pxìm yom fkol nì\-yrr.} Vegetables are often eaten raw.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{yrr}{\B{yrr}})

% nìzen
\hi
\HT{nIzen}{\B{nì\cp{zen}}} \I{[nI."zEn]} \emph{adv.} necessarily. 
(\ding{43}~\HL{zene}{\B{ze\-ne}}; \ding{214}~\HL{kenzen}{\B{ken\-zen}})

% nìzevakx
\hi 
\HT{nIzevakx}{\B{nì\cp{ze}vakx}} \I{[nI."zE.vak']} \emph{adv.} cruelly. 
\B{Kin\-trr\-am ay\-oe tar\-mìng na\-ri teng\-krr 'aw\-a ho\-ren\-ley\-kek\-yul tu\-tan\-it a\-'aw tspe\-rang--lun\-lu\-ke nì\-wotx, nìk\-'ong, nì\-ze\-vakx.} Last week we watched while one law enforcer killed a man--without any reason, slowly, cruelly.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} 
(\ding{43}~\HL{zevakx}{\B{ze\-vakx}})

% nìzawnong
\hi
\HT{nIzawnong}{\B{nì\cp{zaw}nong}} \I{[nI."zaw.noN]} \emph{adv.} safely. 
\B{Si\-vop nì\-zaw\-nong, ma ay\-sop\-yu.} Travel safely, travellers.~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b}
\B{Ti\-va\-ron nì\-zaw\-nong, ma ey\-lan!} Hunt safely, friends!~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
(\emph{a. idiom}) \B{Mi\-vak\-to nì\-zaw\-nong.} `Take care.' `Travel safely.'~\LINK{http://forum.learnnavi.org/language-updates/makto-zong!/}{b}
(\emph{often shortened to} \HL{zong}{\B{zong}}) \emph{in the answer to the question} \B{Mak\-to fya\-pe?} ``How are things? How's everything going? How are you doing?'' \B{Nì\-zaw\-nong.\slash Zong.} ``Well.''~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b}
(\ding{43}~\HL{zong}{\B{zong}})

% nokx
\hi
\HT{nokx}{\B{nokx}} \I{[nok']} \emph{vtr.} give birth to.
\B{Sa'\-nok\-ìl oe\-yä tsmu\-ket a\-mip no\-lokx trr\-am.} Mom gave birth to my new sister yesterday.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}. \HL{'ongokx}{\B{'o\-ngokx}}, be born.

% nong
\hi
\HT{nong}{\B{nong}} \I{[noN]} \emph{vtr.} follow, proceed after.
\B{Nong oet!} Follow me!~\LINK{http://naviteri.org/2022/01/aawa-tipangkxotsyip-a-teri-horen-lifyaya-a-few-little-discussions-about-grammar/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nongspe'}{\B{nong\-spe'}}, persue with an intend to capture.

% nongspe'
\hi 
\HT{nongspe'}{\B{nong\cp{spe'}}} \I{[noN."{}spEP]} \emph{vtr.} persue with an intend to capture. 
\B{Ta\-ron\-yul ye\-rik\-it nar\-mong\-spe', slä tsun ye\-rik hi\-vi\-fwo.} The hunter was pursuing a hexapede, but the hexapede was able to escape.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
(\ding{43}~\HL{nong}{\B{nong}}, \HL{spe'e}{\B{spe'e}})

% nopx
\hi
\HT{nopx}{\B{nopx}} \I{[nop']} \emph{vtr.} put away, store.
\B{Tsko swi\-zaw\-ti ni\-vopx, ma 'ite. Ke ta\-ron oeng fì\-trr.} Put away your bow and arrow, daughter. You and I are not hunting today.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}

% nrr
\hi
\HT{nrr}{\B{nrr}} \I{[n\s{r}]} \emph{vin.} glow, be luminous.   
\B{Na'\-rìng sngä\-'i ni\-vrr txon\-krr syu\-ra\-tan\-fa.} The forest begins to glow at night with bioluminescence.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Nga\-ri key ne\-re\-i\-yrr, ma 'ey\-lan. Nì\-law po yaw\-ne lu ngar.} Your face is glowing, my friend (and I'm pleased to see it). It's clear you love her.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{sAnrr}{\B{sä\-nrr}}, glow, an instance of glowing. 
\on{2}.~\HL{nrra}{\B{nrr\-a}}, pride, feeling of pride.
\on{3}.~\HL{lornusrr}{\B{lor\-nu\-srr}}, radiant.

% nrra
\hi
\HT{nrra}{\B{\cp{nrr}a}} \I{["n\s{r}.a]} \emph{n.} pride, feeling of pride.   
\B{Nga\-ri lu\slash leiu oe\-ru nrra nì\-txan, ma 'ite.} I'm very proud of you, daughter.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} 
(\ding{214}~\HL{lewng}{\B{lewng}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lenrra}{\B{le\-nrra}}, proud, feeling of pride in others. 
\on{2}.~\HL{nInrra}{\B{nì\-nrra}}, proudly, with pride. 
\on{3}.~\HL{snonrra}{\B{sno\-nrra}}, self\hyp{}pride.

% nuä
\hi
\HT{nuA}{\B{\cp{nu}ä}} \I{["nu.\ae]} \emph{adp.}\texttt{+} beyond.   
\B{Aw\-nga kel\-ku si nuä ay\-ram a\-lu\-sìng.} We live beyond the flying mountains.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Fo kel\-ku si nuä ora.} They live beyond the lake.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nuAslew}{\B{nu\-ä\-slew}}, surpass, transcend. 

% nuäslew
\hi 
\HT{nuAslew}{\B{\cp{nu}äslew}} \I{["nu.\ae.sl$\cdot$Ew]} \emph{vtr., inf.}\on{33} surpass, transcend. 
\B{Nge\-yä way\-ìl nuä\-slew pum\-it oe\-yä.} Your song surpasses mine.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b} 
(\ding{43}~\HL{nuA}{\B{nuä}}, \HL{salew}{\B{sa\-lew}}) 

% nui
\hi
\HT{nui}{\B{\cp{nu}i}} \I{["nu.i]} \emph{vin.} (\emph{a.}) fail, falter, go astray (\emph{not obtain expected or desired result})  
\B{Oe fmo\-li nu\-ä\-ngi.} I tried, but unfortunately I failed.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Rum\-it a no\-lu\-i rä\-'ä fe\-wi.} Don't chase after a foul ball.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{b.}) mess up, do wrong (\emph{used for placing blame}).
\B{No\-lu\-i \cp{nga}!} \emph{You} failed! It's \emph{your} fault! \emph{You're} the one who messed up.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{214}~\HL{flA}{\B{flä}}, \ding{43}~\HL{tIkxey}{\B{tì\-kxey si}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tInui}{\B{tì\-nui}}, failure (abstract).  
\on{2}.~\HL{sAnui}{\B{sä\-nui}}, failure (particular instance).   
\on{3}.~\HL{nInu}{\B{nì\-nu}}, failingly, falteringly, in vain. 

% nulkrr
\hi
\HT{nulkrr}{\B{nul\cp{krr}}} \I{[nul."k\s{r}]} \emph{or} \I{[nUl.]} \emph{adv.} longer (time). 
\B{Ke tsun aw\-nga pi\-vey nul\-krr--txew lok.} We can't wait any longer--time is almost up.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Ru\-txe ma\-wey\-pi\-vey nul\-krr nì\-'it, ma ey\-lan.} Please, wait a bit longer, friends.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
(\ding{43}~\HL{krr}{\B{krr}}, \HL{nI'ul}{\B{nì\-'ul}})

% nulnew
\hi
\HT{nulnew}{\B{nul\cp{new}}} \I{[nul."n$\cdot$Ew]} \emph{or} \I{[nUl.]} \emph{vtr.}\on{22} (\emph{a.}) prefer, would rather.
\B{Oel pay\-ti nul\-new to swo\-at.} I prefer water to alcohol.~\LINK{http://forum.learnnavi.org/language-updates/prefer-x-to-y-24474/}{ln}
(\emph{b. modal}) (\emph{either with} ‹\B{iv}› \emph{or} \B{futa} \emph{and} ‹\B{iv}›).
\B{Ke lu kaw\-tu a nul\-ni\-vew oe po\-hu ti\-rea\-pi\-väng\-kxo äo Ut\-ral Ay\-mok\-ri\-yä.} There's nobody I'd rather commune with under the Tree of Voices.~\LINK{http://wiki.learnnavi.org/Corpus\#Lemondrop.com}{ld\slash wiki}
\B{Tsun oe nga\-hu tsa\-tse\-nge\-ne ki\-vä, slä nul\-new fu\-ta si\-vop oeng nì\-txi\-lu\-ke.} I can go there with you, but I prefer to travel leisurely.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{new}{\B{new}}, \HL{nì'ul}{\B{nì\-'ul}})

% nume
\hi
\HT{nume}{\B{\cp{nu}me}} \I{["nu.mE]} \emph{vin.} learn, acquire knowledge or understanding (\HL{teri}{\B{teri}}, about).
\B{Tsko swi\-zaw\-fa a tì\-ta\-ron\-ìri ze\-ne fko ni\-vu\-me.} You must learn to hunt with a bow and arrow.~\LINK{http://forum.learnnavi.org/language-updates/ordinals-nume/}{ln\slash a\textsuperscript{c}}
\B{Maw\-krr la\-ye\-iu oer krr nì\-'ul fte ngi\-vop ay\-lì\-'ut sì tsay\-fne\-sä\-num\-vit a tsun fra\-por srung si\-vi fte ni\-vu\-me sì zi\-ve\-rok nì\-swey.} After that I will have more time to create words and the kinds of lessons that can help everyone best learn and remember.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Var ni\-vu\-me ko!} Let's keep learning!~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln}
\B{Oel vewng fu\-ta ay\-e\-veng ni\-vu\-me te\-ri ay\-ewll na'\-rìng\-ä.} I see to it that the children learn about the forest plants.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{numeyu}{\B{nu\-me\-yu}}, student, learner. 
\on{2}.~\HL{numtseng}{\B{num\-tseng}}, school. 
\on{3}.~\HL{numtsengvi}{\B{num\-tseng\-vi}}, classroom. 
\on{4}.~\HL{numultxa}{\B{nu\-mul\-txa}}, class (for instruction). 
\on{5}.~\HL{numultxatu}{\B{nu\-mul\-txa\-tu}}, classmate. 
\on{6}.~\HL{snanumultxa}{\B{sna\-nu\-mul\-txa}}, course (for instruction).
\on{7}.~\HL{sAnume}{\B{sä\-nu\-me}}, instruction. 
\on{8}.~\HL{sAnumvi}{\B{sä\-num\-vi}}, lesson.

% numeyu
\hi
\HT{numeyu}{\B{\cp{nu}meyu}} \I{["nu.mE.ju]} \emph{n.} student, learner. 
\B{Oe lu nu\-me\-yu.} I am a student.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Kar\-yul fngo\-lo' fu\-ta ay\-nu\-me\-yu pi\-va\-te ye'\-krr.} The teacher required the students to arrive early.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{nume}{\B{nu\-me}})

% numtseng
\hi
\HT{numtseng}{\B{\cp{num}tseng}} \I{["num.\t{ts}EN]} \emph{or} \I{["nUm.]} \emph{n.} school. 
\B{Tsun Txe\-wì kop pi\-vll\-txe nì\-'it nì\-'Ìng\-lì\-sì a no\-lu\-me mì num\-tseng tok\-tor Kì\-rey\-sä.} Txewì can also speak a bit of English that he has learned in doctor Grace's school.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{nume}{\B{nu\-me}}, \HL{tseng}{\B{tse\-nge}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{numtsengvi}{\B{num\-tseng\-vi}}, classroom.
\on{2}.~\HL{nawnumtseng}{\B{naw\-num\-tseng}}, university.

% numtsengvi
\hi
\HT{numtsengvi}{\B{\cp{num}tsengvi}} \I{["num.\t{ts}EN.vi]} \emph{or} \I{["nUm.]} \emph{n.} classroom, division of a school.
\B{num\-tseng\-vi ay\-sngä\-'i\-yu\-ä}, beginners' classroom.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Zo\-la\-'u ftu kä\-pxì num\-tseng\-vi\-yä sä\-mewn a\-wok.} From the back of the classroom came a loud sigh.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
\B{Kar\-yu a\-sìl\-tsan ze\-ne tsi\-vun ay\-nu\-me\-yut hi\-vipx mì num\-tseng\-vi.} A good teacher has to be able to control (his/her) students in the classroom.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
(\ding{43}~\HL{numtseng}{\B{num\-tseng}})

% numultxa
\hi
\HT{numultxa}{\B{numul\cp{txa}}} \I{[nu.mul."t'a]} \emph{or} \I{[.mUl.]} \emph{n.} (\textsc{rn}: \B{numùl\cp{txa}} \I{[nu.mUl."da]}) class (for instruction).   
\B{Te\-ri lì'\-fya le\-Na'\-vi a tsa\-nu\-mul\-txa lo\-le\-i\-u sä\-flä!} The class about the Na'vi language was a success.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{Nì\-aw\-no\-mum ngo\-lop ay\-oel fì\-nu\-mul\-txa\-ti fpi sngä\-'i\-yu.} As you know, we created this class for beginners.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{Mì Si\-ä\-tll a nu\-mul\-txa\-ri lì'\-fya\-yä le\-Na'\-vi, sìl\-pey oe slì\-ye\-vu nga nu\-mul\-txa\-tu!} I hope you'll participate in the upcoming Na'vi class in Seattle!~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{nume}{\B{nu\-me}}, \HL{ultxa}{\B{ul\-txa}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{numultxatu}{\B{nu\-mul\-txa\-tu}}, classmate, member of a class.

% numultxatu
\hi
\HT{numultxatu}{\B{numul\cp{txa}tu}} \I{[nu.mul."t'a.tu]} \emph{or} \I{[.mUl.]} (\textsc{rn}: \B{numùl\cp{txa}tu} \I{[nu.mUl."da.tu]}) \emph{n.} classmate, member of a class.
\B{Mì Si\-ä\-tll a nu\-mul\-txa\-ri lì'\-fya\-yä le\-Na'\-vi, sìl\-pey oe slì\-ye\-vu nga nu\-mul\-txa\-tu!} I hope you'll participate in the upcoming Na'vi class in Seattle!~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{numultxa}{\B{nu\-mul\-txa}})

% nun
\hi 
\HT{nun}{\B{nun}} \I{[nun]} \emph{or} \I{[nUn]} (\textsc{rn}: \B{nun} \I{[nun]}) \emph{adj.} long (\emph{of time}). 
\B{Su\-nu oer Ra\-lu, slä fì\-sä\-frr\-fen pe\-yä lu nun nì\-hawng.} I like Ralu, but this visit of his is too long.~\LINK{http://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{214}~\HL{yol}{\B{yol}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{zIsItnun}{\B{zì\-sìt\-nun}}, leap year. 
\on{2}.~\HL{nunyol}{\B{nun\-yol}}, length (of time).

% nunyol
\hi 
\HT{nunyol}{\B{\cp{nun}yol}} \I{["nun.yol]} \emph{or} \I{["nUn.]} (\textsc{rn}: \B{\cp{nun}yol} \I{["nun.yol]}) \textsc{i}. \emph{n.} length (of time). \\
\textsc{ii}. \B{\cp{nun}yolpe} \I{["nun.yol.pE]} \emph{or} \B{pe\cp{nun}yol} \I{[pE."nun.yol]} \emph{inter.} what length? how long (of time)?
\B{Nga har\-ma\-haw pe\-nun\-yol?} How long were you sleeping?~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
(\ding{43}~\HL{nun}{\B{nun}}, \HL{yol}{\B{yol}}; \HL{ngimpup}{\B{ngim\-pup}}) 

% nutx
\hi
\HT{nutx}{\B{nutx}} \I{[nut']} \emph{or} \I{[nUt']} \emph{adj., nfp.} thick.   
\B{Tsun Txil\-te pam\-rel\-it ivi\-nan; ta\-fral puk\-ot a\-nutx mu\-nge fra\-tseng.} Txilte knows how to read; therefore she brings a thick book wherever she goes.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Krro krro, flì\-a vul a\-ru\-sey to nutx\-a pum a\-ke\-ru\-sey lu txur.} A thin living branch is sometimes stronger than a thick dead one.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\ding{214}~\HL{flI}{\B{flì}}; \ding{43}~\HL{ompu}{\B{om\-pu}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{flInutx}{\B{flì\-nutx}}, thickness.

\end{multicols}