\pdfbookmark[0]{NG}{NG}
\begin{center}
\section*{NG \I{[N]} (\emph{NgeNg})}
\end{center}

\begin{multicols}{2}

% nga
\hi
\HT{nga}{\B{nga}} \I{[Na]} \emph{pn.} (\emph{gen.} \B{\cp{nge}yä}) you (2\textsuperscript{nd} person singular). 
\B{Lu nga win sì txur | Lu nga txan\-tslu\-sam.} You are fast and strong | You are wise.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
\B{Nga lu So\-rewn, ke\-fyak?} You're Sorewn, right?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Tew\-ti, nga lu tsul\-fä\-tu i\-'en\-ä.} Wow, you are a master on the \emph{i'en}.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Pll\-txe nga nìl\-tsan!} You speak well!~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll}
\B{Ngal fì\-'u\-pxa\-ret oe\-ru fpo\-le'.} You have sent me this message.~\LINK{http://forum.learnnavi.org/language-updates/a-few-small-things/msg299949/\#msg299949}{ln}
\B{Oel nga\-ti ka\-me\-i\-e.} I \textsc{see} you.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog}{wiki}
\B{Ey\-wal zey\-ki\-vo ngat nì\-win.} May Eywa heal you quickly.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Nga\-ru ira\-yo sei\-yi oe nì\-mun.} I (happily) thank you again.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
\B{Oel mu\-wi\-vìn\-txu nga\-ru oe\-yä ler\-tut.} Let me present my colleague to you (\emph{formal}).\slash \B{Nga\-ru oe\-yä ler\-tut.} This is my colleague (\emph{coll.}).~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-297}{b}
\B{Fya\-pe fko syaw ngar?} What's your name? [lit.: how does one call you?]~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln}
\B{Tì\-ste\-ftxaw nge\-yä flo\-lä nìl\-tsan!} Your examination succeeded well!~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{Lu nge\-yä tì\-ftxey!} It's your choice!~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{Nga\-ri txe'\-lan ma\-wey li\-vu.} ``Don't worry (about it).''~\LINK{http://forum.learnnavi.org/language-updates/fmawno/}{ln}
\B{Nga\-ri so\-la\-lew pol\-pxay\-a zì\-sìt?} How old are you?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tsun oe nga\-hu nì\-Na'\-vi pi\-väng\-kxo a fì\-'u oe\-ru prr\-te' lu.} It's a pleasure to be able to chat with you in Na'vi.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Ey\-wa nga\-hu.} Eywa be with you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Tul hu nga | Na'\-rìng\-ka nì\-'o'.} Run with you | Through the forest in a fun way.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}
(\emph{a. conv. phrase}) \B{Nga\-fkeyk pe\-fya\slash fya\-pe?} `What's your status?\slash How ya doin'?'~\LINK{http://naviteri.org/2018/02/negative-questions-in-navi/\#comment-28201}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{ngenga}{\B{nge\-nga}}, you (ceremonial).
\on{2}.~\HL{oeng}{\B{oeng}}, us two (inclusive).
\on{3}.~\HL{manga}{\B{ma\-nga}}, ``hey you''. 

% nga'
\hi
\HT{nga'}{\B{nga'}} \I{[NaP]} \emph{vtr.} (\emph{a.}) contain.   
\B{Na'\-rìng\-ìl nga' pxay\-a io\-ang\-it.} The forest contains many animals.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\emph{b.}) be pregnant with. 
\B{Pol pxey\-a prr\-nen\-it nge\-i\-a'.} I'm delighted to say she's pregnant with triplets.~\LINK{http://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
\B{Krr\-a ngal oe\-ti ngar\-ma', 'e\-fu pe\-fya?} How did you feel when you were pregnant with me?~\LINK{http://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{nga'prrnen}{\B{nga'prrnen}}, be pregnant (for people).
\on{2}.~\HL{nga'lini}{\B{nga'lini}}, be pregnant (for animals).

% -nga'
\hi
\HT{-nga'}{\B{-nga'}} \I{[.NaP]} \emph{unprod. suf. to form adj.s from n.s with the meaning} containing, -ful; e.g., \B{meuia\-nga'}, honorable, honorably. \B{ral\-nga'}, meaningful, instructive. \B{tì\-tstew\-nga'}, courageous, brave, etc.

% nga'lini
\hi 
\HT{nga'lini}{\B{nga'\cp{li}ni}} \I{[NaP."li.ni]} \emph{vin.} be pregnant (\emph{for animals}). 
(\ding{43}~\HL{nga'}{\B{nga'}}, \HL{lini}{\B{li\-ni}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tInga'lini}{\B{tì\-nga'\-li\-ni}}, pregnancy (\emph{for animals}).

% nga'prrnen
\hi 
\HT{nga'prrnen}{\B{nga'\cp{prr}nen}} \I{[N$\cdot$aP."p\s{r}.nEn]} \emph{vin., ofp., inf.}\on{11} (\emph{a}.) be pregnant. 
\B{Zun ngal oey tsmu\-ket tsi\-ve\-'a, zel am\-'a\-lu\-ke i\-vo\-mum fu\-ta poe nga'\-prr\-nen.} If you saw my sister, you'd certainly know she was pregnant.~\LINK{http://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
(\emph{b.}) be pregnant \emph{with}, use \ding{43}~\HL{nga'}{\B{nga'}}, e.g. \B{Pol pxey\-a prr\-nen\-it nge\-i\-a'.} I'm delighted to say she's pregnant with triplets.~\LINK{http://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
(\ding{43}~\HL{nga'}{\B{nga'}}, \HL{prrnen}{\B{prr\-nen}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tInga'prrnen}{\B{tì\-nga'\-prr\-nen}}, pregnancy (\emph{for people}).

% ngam
\hi
\HT{ngam}{\B{ngam}} \I{[Nam]} \emph{n.} echo.   
\B{Fì\-slär\-mì tsun fko sti\-vawm ngam\-it a\-pxay.} You can hear a lot of echoes in this cave.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{ngampam}{\B{ngam\-pam}}, rhyme.

% ngampam
\hi
\HT{ngampam}{\B{\cp{ngam}pam}} \I{["Nam.pam]} \textsc{i}. \emph{n.} rhyme.
\B{Fì\-way\-ri hì\-no\-a re\-nut ngam\-pam\-ä ke tsä\-ngun oe tsli\-vam.} I'm afraid I can't understand the intricate rhyme scheme of this poem.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
\textsc{ii}. \B{\cp{ngam}pam si} \I{["Nam.pam si]} \emph{vin.} (\emph{a.}) rhyme.   
\B{Me\-lì\-'u alu mung\-wrr sì nì\-fkrr ngam\-pam si.} The words \emph{mung\-wrr} and \emph{nì\-fkrr} rhyme.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{b. fig.}) get along, fit together well. 
\B{New Ri\-ni sì Ra\-lu mun\-txa sli\-vu, slä tì\-'e\-fu\-mì oe\-yä, ngam\-pam ke si.} Rini and Ralu want to marry, but I feel they're not compatible.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{ngam}{\B{ngam}}, \HL{pam}{\B{pam}})

% ngawng
\hi
\HT{ngawng}{\B{ngawng}} \I{[NawN]} \emph{n., fauna} worm. 
\B{Ma txan\-tslu\-sam\-a ngawng, a swok\-a ut\-ral\-ti yom.} O wise worm, who eats the Sacred Tree.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Becoming_one_of_the_Na.27vi}{a\textsuperscript{c}}  \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{eltungawng}{\B{el\-tu\-ngawng}}, brainworm.

% ngay
\hi
\HT{ngay}{\B{ngay}} \I{[Naj]} \emph{adj.} true.   
\B{Oe\-yä txe'\-lan li\-vu ngay.} Let my heart be true.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki} 
\B{Ke\-'u ke lu ngay.} Nothing is true.~\LINK{http://forum.learnnavi.org/language-updates/double-negatives-not-optional/}{ln} 
\B{Lam ngay oer, fo ka\-yä ìlä hil\-van tup na'\-rìng.} I bet they'll take the river rather than the forest.~\LINK{http://forum.learnnavi.org/language-updates/concerning-tup/}{ln\slash g} 
\B{Tsyan\-ìri sì oe\-ri lo\-le\-iu tsa\-kaym tsyeym a\-ngay.} There was a true treasure that evening for John and I.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b} \\  
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIngay}{\B{nì\-ngay}}, truly.  
\on{2}.~\HL{tIngay}{\B{tì\-ngay}}, truth.   
\on{3}.~\HL{ngaytxoa}{\B{ngay\-txoa}}, sorry, my apologies.  
\on{4}.~\HL{pllngay}{\B{pll\-ngay}}, admit. 
\on{5}.~\HL{nIfkeytongay}{\B{nì\-fkey\-to\-ngay}}, actually, as a matter of fact. 
\on{6}.~\HL{wIngay}{\B{wì\-ngay}}, prove.
\on{7}.~\HL{pe'ngay}{\B{pe'\-ngay}}, judge, conclude.

% ngaytxoa
\hi
\HT{ngaytxoa}{\B{ngay\cp{txo}a}} \I{[Naj."t'o.a]} \emph{idiom} forgive me, sorry, my apologies (\emph{acknowledgment of guilt and regret}).
\B{Ngay\-txo\-a, nì\-aw\-no\-mum ke lo\-lu oer nì\-ke\-ftxo mì sok\-a srr ay\-skxom le\-tam fte lì'\-fya\-ri aw\-nge\-yä tì\-kang\-kem si\-vi.} My apologies: As you know, in recent days I have not had sufficient opportunity to work on our language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Kxey\-ey\-tsyìp\-ìri ngay\-txo\-a.} My opologies for the little mistake.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/comment-page-1/\#comment-2108}{b}
(\ding{43}~\HL{ngay}{\B{ngay}}, \HL{txoa}{\B{txoa}})

% ngä'än 
\hi
\HT{ngA'An}{\B{ngä\cp{'än}}} \I{[N\ae."P\ae{}n]} \emph{vin.} suffer (mentally, emotionally), be miserable.
\B{Sna\-fpìl\-fya\-ri le\-Na'\-vi krra smar\-it fkol tspang, tsran\-ten nì\-txan fwa po ke ngä\-'än nì\-kel\-kin.} It's important in Na'vi philosophy that the prey not suffer unnecessarily when it's killed.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 
\B{Sra\-ne, skxir tì\-sraw si nì\-txan, slä ke nge\-rä\-'än oe kaw\-'it.} Yes, the wound is very painful, but I'm not in the least suffering emotionally.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Oe\-ri ke lu tìm\-wiä, ha nga\-ti oel ngey\-ka\-syä\-'än nì\-teng.} For me there isn't justice, so I'll make you be miserable as well.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} \\  
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAngA'An}{\B{sä\-ngä\-'än}}, bout of suffering, episode of drepression.

% ngäng
\hi
\HT{ngAng}{\B{ngäng}} \I{[N\ae{}N]} \emph{n.} stomach. 

% ngäzìk
\hi
\HT{ngAzIk}{\B{\cp{ngä}zìk}} \I{["N\ae.zIk\textcorner]} \emph{adj.} difficult, hard.
\B{Fwam\-pop\-ä tem\-rey tsa\-tseng\-mì ngä\-zìk lu nì\-wotx.} It's very hard for a tapirus to survive there.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Kxawm set 'u a\-ngä\-zìk fra\-to lu la\-'a a\-yll.} Maybe the most difficult thing now is social distance.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b}
(\ding{214}~\B{ftue}) \\  
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIngAzIk}{\B{tìngäzìk}}, difficulty, problem.

% ngenga
\hi
\HT{ngenga}{\B{nge\cp{nga}}} \I{[NE."Na]} \emph{pn.} (\emph{gen.} \B{nge\-{\color{Plum}nge}\-yä}) you. (2\textsuperscript{nd} person singular; deferential\slash ceremonial form).
\B{Kal\-txì. Nge\-nga lu tu\-pe\slash pe\-su?--Oe lu Txe\-wì.} Hello. Who are you?--I'm Txewì.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to? (\emph{overly formal, stilted Na'vi})~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Ey\-wa nge\-nga\-hu!--Ul\-te nge\-nga\-hu, ma 'E\-veng.} Eywa be with you!--And with you, 'Eveng.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/comment-page-1/\#comment-1069}{b}
(\ding{43}~\HL{nga}{\B{nga}})

% ngep
\hi
\HT{ngep}{\B{ngep}} \I{[NEp\textcorner]} \emph{n.} navel.

% nget
\hi
\HT{nget}{\B{nget}} \I{[NEt\textcorner]} \emph{adj.} smell of decaying wood and leaves.

% ngeyn
\hi
\HT{ngeyn}{\B{ngeyn}} \I{[NEjn]} \emph{adj.} tired. 
\B{Po 'e\-fu ngeyn ul\-te kin tì\-rawn\-it nì\-txan.} He is tired and very much needs to be replaced.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{214}~\HL{txen}{\B{txen}})

% ngian
\hi
\HT{ngian}{\B{ngi\cp{an}}} \I{[Ni."{}an]} \emph{adv.} however. 
\B{Ay\-lì\-'u ngi\-an nì\-'it ske\-pek lu.} However the words are a bit formal.~\LINK{http://forum.learnnavi.org/language-updates/more-grammatical-tidbits/}{ln\slash a}
\B{Fì\-kxey\-ey\-ri tsap\-'a\-lu\-te. Ngi\-an la\-he\-a lì\-'u alu Pol ke lu kxey\-ey.} Apologies for this mistake. However the other word `pol' is not a mistake.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/comment-page-1/\#comment-520}{b}

% ngim
\hi
\HT{ngim}{\B{ngim}} \I{[Nim]} \emph{adj.} long (\emph{physical length}).
\B{Ay\-upxa\-re\-ri a\-ngim nì\-hawng lu alo oe\-yä!} Now it's my turn for messages that are too long!~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\ding{214}~\B{pup}) \\  
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{ngimpup}{\B{ngim\-pup}}, length.

% ngimpup
\hi
\HT{ngimpup}{\B{ngim\cp{pup}}} \I{[Nim."pup\textcorner]} \emph{or} \I{[."pUp\textcorner]} (\textsc{rn}: \B{ngim\cp{pùp}} \I{[Nim."pUp\textcorner]}) \textsc{i}. \emph{n.} length (of physical extension). \\
\textsc{ii}. \B{\cp{ngim}puppe} \I{["Nim.pup\textcorner.pE]} \emph{or} \I{[."pUp\textcorner.]} \emph{or} \HL{pengimpup}{\B{pe\cp{ngim}pup}} \emph{or} \I{[."pUp\textcorner]} \emph{inter.} what length? how long (of physical extension)?
(\ding{43}~\HL{ngim}{\B{ngim}}, \HL{pup}{\B{pup}}; \HL{nunyol}{\B{nun\-yol}})

% ngip
\hi
\HT{ngip}{\B{ngip}} \I{[Nip\textcorner]} \emph{n.} space, open or borderless area.
\B{Nge\-yä ik\-ran\-ìl ngip\-it le\-tam kin fte tsi\-vun kll\-pi\-vä.} Your ikran needs enough open space to be able to land.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Pll\-txe Saw\-tu\-te san ki\-fkey\-ìl ay\-oe\-yä tok txan\-a ngip\-it a san\-hì\-kip.} The Sky People say that their world is in the great space among the stars.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{pxawngip}{\B{pxawngip}}, environment.

% ngoa
\hi 
\HT{ngoa}{\B{\cp{ngo}a}} \I{["No.a]} \emph{n.} mud.
\B{Maw\-fwa zup tom\-pa, lu ngo\-a a\-txan fì\-ha\-pxì\-mì na'\-rìng\-ä.} After rain, there's a lot of mud in this part of the forest.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}

% ngong
\hi
\HT{ngong}{\B{ngong}} \I{[NoN]} \emph{adj.} (\emph{a.}) lethargic, lacking sufficient energy.    
\B{Ftu\-e lu fwa ta\-ron ngong\-a io\-ang\-it to fwa ta\-ron pum\-it a lu wa\-lak sì win.} It's easier to hunt lethargic animals than to hunt perky, speedy ones.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{b. of people with pejorative force}) lazy.
\B{Fì\-trr oe 'e\-fu ngong nì\-wotx.} I'm just not motivated to do anything today.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{214}~\B{walak}) \\  
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIngong}{\B{nì\-ngong}}, lethargically, lazily. 
\on{2}.~\HL{tIngong}{\B{tì\-ngong}}, lethargy, laziness.

% ngop
\hi
\HT{ngop}{\B{ngop}} \I{[Nop\textcorner]} \emph{vtr.} create.
\B{Pam\-tse\-ol ngop ay\-re\-nut}, Music creates patterns.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Maw\-krr la\-ye\-iu oer krr nì\-'ul fte ngi\-vop ay\-lì\-'ut sì tsay\-fne\-sä\-num\-vit a tsun fra\-por srung si\-vi fte ni\-vu\-me sì zi\-ve\-rok nì\-swey.} After that I will have more time to create words and the kinds of lessons that can help everyone best learn and remember.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
\B{Nì\-aw\-no\-mum ngo\-lop ay\-oel fì\-nu\-mul\-txa\-ti fpi sngä\-'i\-yu.} As you know, we created this class for beginners.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{Fra\-'u\-ä ran ngä\-pop fa fra\-syon tse\-yä.} The \emph{ran} of each thing arises from the totality of its attributes.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\  
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{'ongop}{\B{'o\-ngop}}, design (form, shape). 
\on{2}.~\HL{ngopyu}{\B{ngop\-yu}}, creator.   
\on{3}.~\HL{tIngop}{\B{tì\-ngop}}, creation. 
\on{4}.~\HL{rengop}{\B{re\-ngop}}, design (finer detail).   
\on{5}.~\HL{ronsrelngop}{\B{ron\-srel\-ngop}}, imagine.

% ngopyu
\hi
\HT{ngopyu}{\B{\cp{ngop}yu}} \I{["Nop\textcorner.ju]} \emph{n.} creator. 
\B{Ul\-te set … mok\-ri ngop\-yu\-ä!} And now … the voice of the creator.~\LINK{http://naviteri.org/2012/10/ulte-sawasultsyipa-yoratu-leiu-and-the-winner-is/}{b} 
(\ding{43}~\B{ngop}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{pamtseongopyu}{\B{pam\-tseo\-ngop\-yu}}, music creator, composer.

% ngrr
\hi
\HT{ngrr}{\B{ngrr}} \I{[N\s{r}]} \emph{n.} \ding{96} root. \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{ngrrpongu}{\B{ngrr\-po\-ngu}}, grassroots movement.
\on{2}.~\HL{tsawlapxangrr}{\B{tsawl\-a\-pxa\-ngrr}}, unidelta tree.
\on{3}.~\HL{ngrrfpIl}{\B{ngrr\-fpìl}}, key assumption.

% ngrrfpìl 
\hi 
\HT{ngrrfpIl}{\B{\cp{ngrr}fpìl}} \I{["N\s{r}.fpIl]} \emph{n.} key assumption (\emph{a basic assumption that underlies a presentation or argument}). 
\B{Nì\-law lu pe\-yä ngrr\-fpìl fwa Saw\-tu\-te ke lu mal.} His assumption is clearly that the Skypeople can't be trusted.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(\ding{43}~\HL{ngrr}{\B{ngrr}}, \HL{sAfpIl}{\B{sä\-fpìl}}; \HL{txinfpIl}{\B{txin\-fpìl}})

% ngrrpongu
\hi
\HT{ngrrpongu}{\B{\cp{ngrr}pongu}} \I{["N\s{r}.po.Nu]} \emph{n.} grassroots movement, popular movement.
\B{Lì'\-fya\-ri le\-Na'\-vi 'Rr\-ta\-mì, vay set 'al\-mong a fra\-'u ze\-ra\-'u ta ngrr\-po\-ngu.} Everything that has gone on with (blossomed regarding) Na'vi until now on Earth has come from a grassroots movement.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
(\ding{43}~\HL{ngrr}{\B{ngrr}}, \HL{pongu}{\B{po\-ngu}})

% ngul
\hi
\HT{ngul}{\B{ngul}} \I{[Nul]} \emph{or} \I{[NUl]} \emph{adj.} drab-colored, grey. 
\B{ngul na tskxe}, the drab color of stone.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} \\ 
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{ngulpin}{\B{ngul\-pin}}, (the color) grey.

% ngulpin
\hi
\HT{ngulpin}{\B{\cp{ngul}pin}} \I{["Nul.pin]} \emph{or} \I{["NUl.]} \emph{n.} the color \ding{43}~\HL{ngul}{\B{ngul}}.
(\ding{43}~\HL{'opin}{\B{'opin}})

% ngungung
\hi 
\HT{ngungung}{\B{\cp{ngu}ngung}} \I{["Nu.NuN]} \emph{or} \I{[.NUN]} (\textsc{rn}: \B{\cp{ngù}ngùng} \I{["NU.NUN]}) \emph{vtr.} rub. 
\B{Nga\-ri pxun\-ti ngu\-ngung pe\-lun? Sra\-ke tì\-sraw si?} Why are you rubbing your arm? Does it hurt?~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}

% nguway
\hi
\HT{nguway}{\B{\cp{ngu}way}} \I{["Nu.waj]} \emph{n.} howl, viperwolf cry. \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{ronguway}{\B{ro\-ngu\-way}}, howl.

\end{multicols} 