\pdfbookmark[0]{O}{O}
\begin{center}
\section*{O \I{[o]}}
\end{center}

\begin{multicols}{2}

% -o
\hi
\HT{-o}{\B{-o}} \I{[.o]} \emph{suf.} `indefinite -o'. 
(\emph{a. on regular nouns to show indefiniteness}) 
\B{Txo\-pu rä\-'ä si, lu ke\-tu\-wong\-o nì\-'aw.} Don't be afraid, it's just an alien.~\LINK{http://forum.learnnavi.org/language-updates/a-collection/}{ln\slash a} 
(\emph{b. on words relating to time of day or calender to refer to a point in time or duration; use} \ding{43}~\B{ro} \emph{or} \ding{43}~\B{ka} \emph{to disambiguate})
\B{Kaym\-o zo\-la\-'u fray\-ul\-txa\-tu ne kel\-ku moe\-yä fte yi\-vom wu\-tsot, fti\-via nì\-'it lì'\-fya\-ti le\-Na'\-vi, ul\-te ki\-vä\-teng nì\-'o'.} All the participants came to our house one evening to have dinner, study a little Na'vi, hang out together, and have fun.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}

% oare
\hi 
\HT{oare}{\B{o\cp{a}re}} \I{[o."{}a.RE]} \emph{n., astr.} moon (\emph{generic term}). 
\B{Pol\-pxay\-a oa\-ret tse\-'a ngal mì saw pxi\-set?} How many moons do you see in the sky right now?~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b} 

% oe
\hi
\HT{oe}{\B{\cp{o}e}} \I{["{}o.E]} (\emph{with case endings or enclitic adp.s always pronounced as:} \I{["wE-]}) \emph{pn.} I (1\textsuperscript{st} person singular). 
(\emph{a. subjective}) 
\B{Li\-vu win sì txur oe ze\-ne | Ha nì\-'aw | Pxan li\-vu txo nì\-'aw oe nga\-ri | Tsa\-krr nga Na'\-vi\-ru yom\-tì\-yìng.} I must be fast and strong | So only | If I am worthy of you | Will you feed the People.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
(\emph{b. agentive}) \B{oel} \I{[wEl]}
\B{Oel nga\-ti ka\-me\-i\-e.} I \textsc{see} you.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog}{wiki}
\B{Ka\-tot tä\-ftxu oel | Nì\-ean nì\-rim}, I weave the rhythm | In yellow and blue.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki} 
\B{Pol\-txe Ey\-tu\-kan san oe ka\-yä sìk, slä oel pot ke spaw.} Eytukan said he would go, but I don't believe him.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
(\emph{c. patientive}) \B{oet(i)} \I{[wEt\textcorner \slash "wE.ti]}
\B{Spi\-vaw oe\-ti ru\-txe, ma oe\-yä ey\-lan.} Please, believe me, friends.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Fu\-la tsa\-yun oeng pi\-väng\-kxo ye'\-rìn ul\-te nga\-ri oel mok\-rit sta\-yawm, oe\-ti nit\-ram sley\-ku nì\-txan.} It makes me very happy that we'll be able to chat soon and that I'll hear your voice.~\LINK{http://forum.learnnavi.org/language-updates/fula-short-form-of-fiul-a-and-tsal-oeti-steyki-nitram/}{ln}
\B{Tsun nga oet mi\-vong.} You can rely on me.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\emph{d. genitive}) \B{\cp{oe}yä} \I{["wE.j\ae]}
\B{Oe\-yä swi\-zaw nì\-ngay ti\-va\-kuk | Oe\-yä tuk\-rul txe'\-lan\-it ti\-va\-kuk.} Let my arrow strike true | Let my spear strike the heart.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
\B{Ay\-ey\-lan\-ur oe\-yä sì ey\-lan\-ur lì'\-fya\-yä le\-Na'\-vi nì\-wotx.} To all my friends and friends of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
(\emph{idiom}) \B{tì\-'e\-fu\-mì oe\-yä}, in my opinion. \B{tì\-o\-mum\-mì oe\-yä}, as far as I know, to my knowledge.
(\emph{e. dative}) \B{oer(u)} \I{[wER \slash "wE.Ru]}
\B{Tsun oe nga\-hu nì\-Na'\-vi pi\-väng\-kxo a fì\-'u oe\-ru prr\-te' lu.} It's a pleasure to be able to chat with you in Na'vi.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Oe\-ru syaw fko Txe\-wì.} My name is Txewì. [lit.: one calls (to) me Txewì]~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Fpo\-le' ay\-ngal oer fì\-txan nì\-ftxa\-vang a 'upxa\-ret sto\-lawm oel.} I have heard the message you have sent me so passionately.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Pi\-veng oer ftxey nga new ri\-vey fu\-ke.} Tell me whether you want to live or not.~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/msg156629/\#msg156629}{ln}
(\emph{f. topic}) \B{\cp{oe}ri} \I{["wE.Ri]}
\B{Oe\-ri lu swoa fne\-txum.} I'm allergic to intoxicating beverage.~\LINK{http://forum.learnnavi.org/language-updates/new-words-from-the-2012-avatarmeet-navi-class/msg551479/\#msg551479}{b}
\B{Oe\-ri so\-la\-lew zì\-sìt a\-pxe\-vol.} I'm \on{24} years old.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\emph{for inalienable possessions}) 
\B{Oe\-ri tì\-ngay\-ìl txe'\-lan\-it ti\-va\-kuk.} Let the truth strike my heart.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b} 
\B{Oe\-ri ski\-en\-a tsyokx lu le\-skxir.} My right hand is wounded.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}

% oeng
\hi
\HT{oeng}{\B{o\cp{eng}}} \I{[o."{}EN]} (\emph{with case endings or enclitic adp.s always pronounced as:} \I{[wE-]}) \emph{pn.} we (two), you and I, the two of us (1\textsuperscript{st} person dual inclusive).
(\emph{a. subjective}) 
\B{Nì\-'aw\-ve oeng yo\-lom wu\-tsot; maw\-krr uvan si.} First we had dinner; afterwards we played.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
(\emph{b. agentive}) \B{oe\cp{ngal}} \I{[wE."Nal]}
(\emph{c. patientive}) \B{oe\cp{ngat}(i)} \I{[wE."Nat\textcorner \slash wE."Na.ti]}
(\emph{d. genitive}) \B{oe\cp{nge}yä} \I{[wE."NE.j\ae]}
(\emph{e. dative}) \B{oe\cp{ngar}(u)} \I{[wE."NaR \slash wE."Na.Ru]} 
\B{Ma mun\-txa\-tu, oeng yaw\-ne lu oe\-nga\-ru fì\-tsap, ke\-fyak?} We love each other, don't we, my spouse?~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
(\emph{f. topic}) \B{oe\cp{nga}ri} \I{[wE."Na.Ri]} 
(\ding{43}~\HL{oe}{\B{oe}}, \HL{nga}{\B{nga}})

% oeyk
\hi
\HT{oeyk}{\B{o\cp{eyk}}} \I{[o."{}Ejk\textcorner]} \emph{n.} cause.
\B{Oeyk tsa\-tì\-snaytx\-ä lu apxa sä\-nui tì\-eyk\-tan\-ä.} That loss was caused by a great failure of leadership.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} \\
\on{1}.~\HL{oeyktIng}{\B{oeyktìng}}, explain (why).

% oeyktìng
\hi
\HT{oeyktIng}{\B{o\cp{eyk}tìng}} \I{[o."{}Ejk\textcorner.t$\cdot$IN]} \emph{or} \I{["wEjk\textcorner.]} \emph{vtr.}\on{33} explain (why). 
\B{Ney\-ti\-ril Tsyey\-kur oeyk\-to\-lìng teyng\-ta fya\-pe ze\-ne vo\-ìk si\-vi tsa\-tì\-fkey\-tok\-mì.} Neytiri explained to Jake how to behave in that situation.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
\B{Wo\-leyn Ìstawl yey\-fyat mì hll\-te fte oeyk\-ti\-vìng fra\-po\-ru tì\-hawl\-te\-ri sne\-yä.} Ìstaw drew a line on the ground to explain his plan to everyone.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} 
\B{Tsa\-lì'\-u\-ri alu nì\-fkey\-to\-ngay oe oeyk\-ta\-syìng ye'\-rìn mì fo\-stì a\-hay.} Concerning the word \emph{nì\-fkey\-to\-ngay} I'll explain (that) soon in a future post.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/comment-page-1/\#comment-3841}{b}
\B{Tì\-yäkx ke lu sru\-nga', ma tsmuk. Nga txo sti, oeyk\-tìng teyng\-ta pe\-lun.} Snubbing isn't helpful, brother. If you're angry, explain why.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{oeyk}{\B{oeyk}}) \\
\on{1}.~\HL{tIoeyktIng}{\B{tìoeyktìng}}, explanation.

% ohakx
\hi
\HT{ohakx}{\B{o\cp{hakx}}} \I{[o."hak']} \emph{adj.} hungry. 
\B{Nän yom kxam\-trr, 'ul 'e\-fu ohakx kaym.} The less you eat at noon, the hungrier you'll feel in the evening.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Sra\-ke nga 'e\-fu o\-hakx?--Sra\-ne, 'e\-fu nì\-txan. Yi\-vom ko!} Are you hungry?--Yes, quite. Let's eat.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{214}~\HL{vAng}{\B{väng}}) \\
\on{1}.~\HL{tIohakx}{\B{tìohakx}}, hunger.

% ohe
\hi
\HT{ohe}{\B{\cp{o}he}} \I{["{}o.hE]} \emph{pn.} I (1\textsuperscript{st} person singular; deferential\slash ceremonial form).
(\emph{a. subjective}) 
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to? (\emph{overly formal, stilted Na'vi})~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\emph{b. agentive}) \B{\cp{o}hel} \I{["{}o.hEl]}
\B{Vu\-yin ohel Unil\-ta\-ron\-it.} I respectfully request the Dreamhunt.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
(\emph{c. patientive}) \B{\cp{o}het(i)} \I{["{}o.hEt\textcorner \slash "{}o.hE.ti]}
(\emph{d. genitive}) \B{\cp{o}heyä} \I{["{}o.hE.j\ae]}
\B{Ay\-nge\-nga\-ru o\-he\-yä tsmuk\-it alu Ne\-wey te Tska\-ha So\-rewn\-'i\-te.} Allow me to introduce my sister, Newey te Tskaha Sorewn'ite.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\emph{e. dative}) \B{\cp{o}her(u)} \I{["{}o.hER \slash "{}o.hE.Ru]} 
\B{Ätxä\-le si tsnì li\-vu ohe\-ru Unil\-ta\-ron.} I respectfully request the Dreamhunt.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Fay\-sul\-fä\-tu\-ä tì\-kang\-kem ohe\-ru meuia lu\-yu nì\-ngay.} The work of these masters truly honors me.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{f. topic}) \B{\cp{o}heri} \I{["{}o.hE.Ri]}
(\ding{43}~\HL{oe}{\B{oe}})

% oìsss
\hi
\HT{oIsss}{\B{o\cp{ìsss}}} \I{[o."{}Is:]} \textsc{i}.~\emph{intj.} watch it! (\emph{angry snarl}).  \\
\textsc{ii}.~\B{o\cp{ìsss} si} \I{[o."{}Is: si]} \emph{vin.} hiss (at s.o.)
\B{Nga lum\-pe oìsss so\-li por?} Why did you hiss at him?~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% okup
\hi
\HT{okup}{\B{\cp{o}kup}} \I{["{}o.kup\textcorner]} \emph{or} \I{[.kUp\textcorner]} \emph{n.} milk.
\B{Sa'\-nok prr\-nen\-ur yom\-tìng fa okup sne\-yä.} A mother feeds an infant with her milk.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}

% ‹ol›
\hi
‹\B{ol}› \I{[o.l]} \emph{inf. in} ‹1› \emph{to form the perfective aspect}.

% Olangi
\hi
\HT{olangi}{\B{Olangi}} \I{[o.la.Ni]} \emph{n.} \emph{clan name, clan of the plains and Direhorse riders}.

% olo'
\hi
\HT{olo'}{\B{o\cp{lo'}}} \I{[o."loP]} \emph{n.} clan, tribe; community.
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to? (\emph{overly formal, stilted Na'vi})~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tsa\-krr syay ay\-nge\-yä, syay o\-lo'\-ä oe\-yä la\-yu teng.} Then you will suffer the same fate as my clan.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Ay\-o\-lo'\-ru a\-la\-he peng zi\-va\-'u.} Tell the other clans (to) come.~\LINK{http://forum.learnnavi.org/language-updates/naviya-the-other-vocative/}{ln\slash a}
\B{Oe\-ri tì\-rey\-ti oel stxe\-nu\-to\-lìng fpi olo' aw\-nge\-yä.} I offered my life for the sake of our clan.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\  
\on{1}.~\HL{nIolo'}{\B{nìo\-lo'}}, (together) as members of a clan. 
\on{2}.~\HL{olo'eyktan}{\B{olo'\-eyk\-tan}}, clan leader. 
\on{3}.~\HL{lI'fyaolo'}{\B{lì'\-fya\-o\-lo'}}, language clan, \emph{Sprachbund}. 
\on{4}.~\HL{uniltIrantokxolo'}{\B{Unil\-tì\-ran\-tokx\-o\-lo'}}, \emph{Avatar} (fan) community.
\on{5}.~\HL{txanlokxe}{\B{txan\-lo\-kxe}}, clan or tribal domain; country.

% olo'eyktan 1
\hi
\HT{olo'eyktan}{\B{olo\cp{'eyk}tan}}\textsuperscript{\on{1}} \I{[o.lo."PEjk\textcorner.tan]} \emph{n.} clan leader (\emph{generally gender\hyp{}neutral}). 
\B{Maw\-krr\-a Tsyeyk ftxo\-lu\-lì\-'u, tslam fra\-pol fu\-ta slu po Olo\-'eyk\-tan a\-mip.} After Jake spoke, everyone understood that he had become the new Clan Leader.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Zun li\-vu oe O\-lo'\-eyk\-tan, zel oe nga\-ru srung si\-vi set.} If I were Clan Leader, then I would help you now.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
(\ding{43}~\HL{olo'}{\B{olo'}}, \HL{eyktan}{\B{eyk\-tan}}; \HL{olo'eykte}{\B{olo'eykte}})

% olo'eyktan 2
\hi
\HT{olo'eyktan2}{\B{olo'eyk\cp{tan}}}\textsuperscript{\on{2}} \I{[o.lo.PEjk\textcorner."tan]} \emph{n.} male clan leader. 

% olo'eykte
\hi 
\HT{olo'eykte}{\B{olo'eyk\cp{te}}} \I{[o.loP.Ejk."tE]} \emph{n.} female clan leader. 
(\ding{43}~\HL{olo'eyktan}{\B{olo'eyktan}})

% Omatikaya
\hi
\HT{omatikaya}{\B{Omati\cp{ka}ya}} \I{[o.ma.ti."ka.ja]} (\emph{gen.} \B{O\-ma\-ti\-ka\-ya\-ä}) \emph{n.} the Omaticaya clan of the Na'vi.
\B{Mo\-'at \mbox{alu} Tsa\-hìk lu O\-ma\-ti\-ka\-ya\-ä le\-'awa ha\-pxì\-tu a \mbox{ioi} sä\-pi fa 'a\-re.} Mo'at, the Tsahik, is the only member of the Omaticaya who wears a poncho.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{To\-ruk Mak\-to po\-lä\-hem a fì\-'u ro\-lo\-'a nì\-txan Oma\-ti\-ka\-ya\-ru.} The arrival of Toruk Makto made a great impression on the Omaticaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Pe\-yä tsa\-tì\-pe\-'un a swey\-lu txo wi\-vem ay\-oeng Oma\-ti\-ka\-ya\-wä lu fe'.} His decision to fight against the Omaticaya was a bad one.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}

% ompu
\hi 
\HT{ompu}{\B{\cp{om}pu}} \I{["{}om.pu]} \emph{adj., ofp.} fat, corpulent. 
(\ding{214}~\HL{litsi}{\B{li\-tsi}}; \ding{43}~\HL{nutx}{\B{nutx}}) 

% omum
\hi
\HT{omum}{\B{o\cp{mum}}} \I{[o."mum]} \emph{or} \I{[."mUm]} \emph{vtr.} know (a fact). (\emph{in any inflected or derived form, the stress shifts to the o, e.g.}~\B{i\cp{vo}mum} \I{[i."vo.mum]})    
\B{Ul\-te o\-mum oel fu\-ta tì\-fya\-wìn\-txu\-ri oe\-yä pe\-rey ay\-nga nì\-wotx.} And I know you are all waiting for my guidance.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Pol ke tslam stum ke\-'ut, o\-mum lì\-'ut a\-vol nì\-'aw.} He understands almost nothing and only knows eight words.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Tsa\-txon ay\-ut\-ral\-äo krr a pol oe\-ti no\-lìn nì\-wa\-we, olo\-me\-i\-um oel fa key\-rel fu\-ta lu yaw\-ne oe po\-ru.} That night beneath the trees when she looked at me meaningfully, I knew by the expression on her face that she loves me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 
\B{Ke o\-mä\-ngum oel teyng\-ta ftu pe\-seng zo\-la\-'u tsa\-pam\-rel\-vi alu t.} I'm afraid I don't know where that letter t came from.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/comment-page-1/\#comment-2612}{b}
\B{Ke o\-mum oel teyng\-ta fo kä pe\-seng\-ne.} I don't know where they're going.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
(\emph{a. coll.}) \B{ko\cp{mum}} \I{[ko."mum]} \emph{or} \B{ke\cp{mum}} \I{[kE."mUm]} `I don't know \ldots' 
\B{Ko\-mum teyng\-ta kem\-pe lo\-len, ta\-lu\-na me\-sì\-ke\-nong\-mì ke lu kea key\-ey.} I don't know what happened because in the two examples there were no mistakes.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/\#comment-33161}{b}
(\ding{43}~\HL{syawm}{\B{syawm}}) \\
\textsc{Derivations}: \ding{43} \on{1}.~\HL{tIomum}{\B{tì\-o\-mum}}, knowledge. 
\on{2}.~\HL{sAomum}{\B{sä\-o\-um}}, information. 
\on{3}.~\HL{nIawnomum}{\B{nì\-aw\-no\-mum}}, as is known. 
\on{4}.~\HL{newomum}{\B{new\-o\-mum}}, be curious.

% onlor
\hi
\HT{onlor}{\B{on\cp{lor}}} \I{[on."loR]} \emph{adj.} good-smelling. 
\B{Syu\-ve\-ri ay\-ftxo\-zä\-yä ay\-nga\-ru fki\-van on\-lor ftxì\-lor\-sì nìwotx!} May the food of the celebrations smell and taste good to you.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\ding{214}~\HL{onvA'}{\B{on\-vä'}}, \ding{43}~\HL{ontu}{\B{on\-tu}}, \HL{lor}{\B{lor}})

% ontu
\hi
\HT{ontu}{\B{\cp{on}tu}} \I{["{}on.tu]} \emph{n.} nose. 
\B{Oe\-ri ta pe\-yä fa\-hew a\-ke\-wong on\-tu te\-ya lä\-ngu.} My nose is full of his alien smell.~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll} \\  
\on{1}.~\HL{onlor}{\B{onlor}}, good-smelling. \\  
\on{2}.~\HL{onvA'}{\B{onvä'}}, bad-smelling. \\
\on{3}.~\HL{ontsang}{\B{ontsang}}, nose ring. \\
\on{4}.~\HL{tIng}{\B{tìng ontu}}, (to) smell.

% ontsang
\hi
\HT{ontsang}{\B{\cp{on}tsang}} \I{["{}on.\t{ts}aN]} \emph{n.} nose ring.
(\ding{43}~\HL{ontu}{\B{on\-tu}}, \HL{tsang}{\B{tsang}})

% onvä'
\hi
\HT{onvA'}{\B{on\cp{vä'}}} \I{[on."v\ae{}P]} \emph{adj.} bad\hyp{}smelling.
\B{Tsa\-fa\-hew a\-on\-vä' ftu kung\-a ioang za\-'u.} That stinking smell comes from a rotten animal.~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/}{b}
(\ding{214}~\HL{onlor}{\B{on\-lor}}, \ding{43}~\HL{ontu}{\B{on\-tu}}, \HL{vA'}{\B{vä'}}) \\
\on{1}.~\HL{vonvA'}{\B{vonvä'}}, asshole, dickhead.

% Otranyu
\hi
\HT{otranyu}{\B{Ot\cp{ran}yu}} \I{[ot\textcorner."Ran.ju]} \emph{n. name, probably clan}.

\end{multicols}