\pdfbookmark[0]{P}{P}
\begin{center}
\section*{P \I{[p]} (\emph{PeP})}
\end{center}

\begin{multicols}{2}

% pa'li
\hi
\HT{pa'li}{\B{\cp{pa'}li}} \I{["paP.li]} \emph{n., fauna} direhorse, \emph{equidirus hoplites}. 
\B{Fmi mi\-vak\-to pa'\-lit!} Try to ride a direhorse!~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Lu pa'\-lir sngem a\-txur.} A direhorse has strong muscles.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\emph{proverb}) \B{Txìm a\-'aw ke tsun hi\-veyn mì tal me\-fa'\-li\-yä.} You can't take both positions or sit on the fence; you need to decide. [lit.: One butt can't sit on the backs of two direhorses]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 
(\emph{a.~special usage}) \B{\cp{pa'}\-li \cp{mak}\-to}, the act of riding a pa'\-li.  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pa'liwll}{\B{pa'liwll}}, direhorse pitcher plant.

% pa'liwll
\hi
\HT{pa'liwll}{\B{\cp{pa'}liwll}} \I{["paP.li.w\s{l}]} \emph{n.}, \ding{96} direhorse pitcher plant, \emph{pseudocenia equina}. 

% pa'o
\hi
\HT{pa'o}{\B{\cp{pa}'o}} \I{["pa.Po]} \emph{n.} side. 
\B{Oe kaw\-krr ne few\-tu\-sok\-a pa\-'o kil\-van\-ä ke ka\-mä.} I never went to the opposite side of the river.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Fì\-tskxe\-ri fa\-'o lu yey; ke lu ko\-um.} This rock has straight sides; it's not rounded.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{fApa}{\B{fä\-pa}}, top, topside. 
\on{2}.~\HL{ftArpa}{\B{ftär\-pa}}, left side. 
\on{3}.~\HL{kllpa}{\B{kll\-pa}}, bottom, bottom side. 
\on{4}.~\HL{pxawpa}{\B{pxaw\-pa}}, perimeter, border. 
\on{5}.~\HL{skiempa}{\B{ski\-em\-pa}}, right side. 
\on{6}.~\HL{wrrpa}{\B{wrr\-pa}}, outside. 
\on{7}.~\HL{rofa}{\B{ro\-fa}}, beside, alongside. 
\on{8}.~\HL{ftumfa}{\B{ftum\-fa}}, out of, from inside. 
\on{9}.~\HL{mIfa}{\B{mì\-fa}}, inside, inner side. 
\on{10}.~\HL{nemfa}{\B{nem\-fa}}, into, inside.
\on{11}.~\HL{ftuopa}{\B{ftuo\-pa}}, from behind.

% pak 
\hi
\HT{pak}{\B{pak}} \I{[pak\textcorner]} \emph{part. for disparagement; always sentence\hyp{}final}. 
\B{Po\-an pak!? Ke lu po tsam\-si\-yu kaw\-'it!} Him!? There's not a warrior's bone in his whole body!~\LINK{http://forum.learnnavi.org/language-updates/ke-kawit/msg174572}{ln}
\B{Tsaw ke ley kaw\-'it pak!} That's not worth a thing!\slash That's just worthless junk!~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}

% palang
\hi 
\HT{palang}{\B{\cp{pa}lang}} \I{["pa.laN]} \emph{vtr.} contact (\emph{in a social sense}), communicate with. 
\B{Ul\-te txo smi\-von ngar ay\-ho\-ak\-tu, ftxey soaia\-mì ftxey sko ey\-lan, fo\-ti pa\-lang fte tsi\-vun ivo\-mum teyng\-ta ftxey lu fo\-ru fpom fu\-ke.} And if you know elders, whether in the family or as friends, contact them to know whether they are alright or not.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b}
(\emph{a. expression}) \B{Pa\-lang ko!} Keep in touch!~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpalang}{\B{tìpalang}}, (social) contact.

% palon
\hi 
\HT{palon}{\B{\cp{pa}lon}} \I{["pa.lon]} \emph{vin.} burn (\emph{of a fire}). 
\B{Txep a\-hì\-'i mì tep\-tseng par\-ma\-lon.} A little fire was burning in the fireplace.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
(\ding{43}~\HL{nekx}{\B{nekx}})

% palukan 
\hi
\HT{palukan}{\B{pa\cp{lu}kan}} \I{[pa."lu.kan]} \emph{n. short form of} \ding{43}~\HL{palulukan}{\B{pa\-lu\-lu\-kan}}. \emph{on Earth often used for} cat.
\B{Ho\-na pa\-lu\-kan\-tsyìp a\-hì\-'i.} A sweet little cat.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Pa\-lu\-kan\-it tso\-le\-'a, ye\-rik lopx hi\-fwo kxam\-lä ze\-swa.} Spotting a thanator, the hexapede panicked and escaped through the grass.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\emph{a. idiom}) \B{Tì\-ka\-ra lu ä\-txä\-le pa\-lu\-kan\-ur.} Resistance is futile [lit.: `reistance is a request to the thanator'].~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}

% palukantsyìp
\hi 
\HT{palukantsyIp}{\B{pa\cp{lu}kantsyìp}} \I{[pa."lu.kan.\t{ts}jIp\textcorner]} \emph{n.} cat (Earth animal). 
\B{Teng\-krr hu pa\-lu\-kan\-tsyìp uvan se\-ri zo\-lìm oel mì sa'\-leng a 'u\-ot a lu txa' sì ekx\-txu.} While playing with my cat I felt something hard and rough on his skin.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
(\ding{43}~\HL{palukan}{\B{pa\-lu\-kan}})

% palulukan
\hi
\HT{palulukan}{\B{palu\cp{lu}kan}} \I{[""pa.lu."lu.kan]} \emph{n., fauna} thanator, \emph{pestiapanthera ferox}. 
\B{Pa\-lu\-lu\-kan a\-tu\-sa\-ron lu le\-hrr\-ap.} A hunting thanator is dangerous.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{Tsawl\-a pa\-lu\-lu\-kan\-ìl oe\-ti 'ìl\-me\-ko a krr, Nawm\-a Sa'\-nok lrr\-tok sei\-yi ul\-te oe em\-ro\-ley.} The Great Mother was looking after me when the big thanator attacked and I've survived.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Oel tso\-le\-'a pa\-lu\-lu\-kan\-it a\-tsawl fra\-to mì sì\-rey.} I saw the biggest thanator I had ever seen.~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
\B{Pa\-lu\-lu\-kan\-ìri lu pxe\-syon a ze\-ne fko zi\-ve\-rok nì\-tut.} Three things about the thanator must always be kept in mind.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\emph{a. proverb}) \B{Ätxä\-le si pa\-lu\-lu\-kan\-ur tsnì smar\-it li\-vo\-nu.\slash Ätxä\-le pa\-(lu)\-lu\-kan\-ur.} \emph{a futile gesture, an attempt to achieve something that might be desirable but will clearly not happen} [lit.: Ask a thanator to release its prey.]~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\emph{b. proverb}) \B{Ha\-haw nì\-'aw txo pa\-lu\-kan smi\-von ngar.} ``Only sleep if you are familiar with the Thanator,'' i.e. \emph{don't think you're safe unless you're aware of the danger}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}\\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{palukantsyIp}{\B{palukantsyìp}}, cat.

% pam / pam si
\hi
\HT{pam}{\B{pam}} \I{[pam]} \textsc{i}. \emph{n.} sound (\emph{sensation}).
\B{Tsa\-ta\-ngek\-ìri pam fkan mo\-mek.} That (tree)\-trunk sounds hollow.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} \\
\textsc{ii}. \B{pam si} \I{[pam si]} \emph{vin.} make a sound.
\B{Fnu! Pam si rä\-'ä!} Be quiet! Don't make a sound!~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
\B{Ut\-ral pam a\-wok so\-li krr\-a zup.} The tree made a loud sound when it fell.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lI'upam}{\B{lì\-'u\-pam}}, pronunciation. 
\on{2}.~\HL{ngampam}{\B{ngam\-pam}}, rhyme. 
\on{3}.~\HL{pamrel}{\B{pam\-rel}}, writing, s.t. written. 
\on{4}.~\HL{pamtseo}{\B{pam\-tseo}}, music.   
\on{5}.~\HL{pxorpam}{\B{pxor\-pam}}, ejective. 
\on{6}.~\HL{kakpam}{\B{kak\-pam}}, deaf. 
\on{7}.~\HL{vApam}{\B{vä\-pam}}, noise. 
\on{8}.~\HL{hawmpam}{\B{hawm\-pam}}, noise.
\on{9}.~\HL{pamuvan}{\B{pam\-u\-van}}, soundplay.
\on{10}.~\HL{tIranpam}{\B{tì\-ran\-pam}}, footstep (\emph{sound}).
\on{11}.~\HL{stA'nIpam}{\B{stä'\-nì\-pam}}, recording.
\on{12}.~\HL{plltxepam}{\B{pll\-txe\-pam}}, speech sound.
\on{13}.~\HL{txampam}{\B{txam\-pam}}, soundblast colossus.
\on{14}.~\HL{pamrIrlI'u}{\B{pam\-rìr\-lì'u}}, onomatopoetic word.
\on{15}.~\HL{pamtsyIp}{\B{pam\-tsyìp}}, small\slash light sound.

% pamrel
\hi
\HT{pamrel}{\B{pam\cp{rel}}} \I{[pam."REl]} \textsc{i}.~\emph{n.} writing, s.t. written; text.
\B{Tok pe\-se\-nget pam\-rel\-ìl?} Where is the text?~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
\B{Tsun Txil\-te pam\-rel\-it ivi\-nan; ta\-fral puk\-ot a\-nutx mu\-nge fra\-tseng.} Txilte knows how to read; therefore she brings a thick book wherever she goes.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} \\
\textsc{ii}.~\HT{pamrelsi}{\B{pam\cp{rel} si}} \I{[pam."REl si]} \emph{vin.} write. (\emph{allows for double dative})
\B{Oe 'upxa\-re\-ru tsmu\-kan\-ur oe\-yä pam\-rel so\-li.} I wrote my brother a message.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}
\B{Fu\-ri\-a nì\-'Ìng\-lì\-sì pam\-rel si\-vi, oe\-ru txoa li\-vu.} I'm sorry for writing in English.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/}{ln}
\B{Tsa\-lì\-'u\-ri fko pam\-rel si fya\-pe?} How is that word written?~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/}{b}
\B{Pam\-rel sa\-yi trr\-ay krr a skxom la\-tsu oe\-ru.} I will write tomorrow when I have time.~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pamrelfya}{\B{pam\-rel\-fya}}, spelling. \\
\on{2}.~\HL{pamrelsiyu}{\B{pam\-rel\-si\-yu}}, writer, author. \\
\on{3}.~\HL{pamrelvi}{\B{pam\-rel\-vi}}, letter (\emph{phonetic symbol}). \\
\on{4}.~\HL{pamrelvul}{\B{pam\-rel\-vul}}, pen, pencil.

% pamrelfya
\hi
\HT{pamrelfya}{\B{pam\cp{rel}fya}} \I{[pam."REl.fja]} \emph{n.} spelling.
\B{Pam\-rel\-fya\-ri fya\-pe?} How do you spell it?~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/}{b}
\B{Lì\-'u\-ri alu tskxe pam\-rel fya\-pe?--Pam\-rel\-fya lu na Tsä, KxeKx, E.} How do you spell the word \emph{tskxe}?--It's spelled ts, kx, e.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/}{b} 
(\ding{43}~\HL{pamrel}{\B{pam\-rel}})

% pamrelsiyu
\hi
\HT{pamrelsiyu}{\B{pam\cp{rel}siyu}} \I{[pam."REl.si.ju]} \emph{n.} writer, author.
\B{Kin\-trr\-am fay\-fam\-rel\-si\-yu\-hu oe ul\-txa so\-lei\-yi.} Last week I've met with these writers.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
(\ding{43}~\HL{pamrelsi}{\B{pam\-rel si}})

% pamrelvi
\hi
\HT{pamrelvi}{\B{pam\cp{rel}vi}} \I{[pam."REl.vi]} \emph{n.} letter (\emph{phonetic symbol}).
\B{Ke o\-mä\-ngum oel teyng\-ta ftu pe\-seng zo\-la\-'u tsa\-pam\-rel\-vi alu t.} I'm afraid I don't know where that letter t came from.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/comment-page-1/\#comment-2612}{b}
\B{U\-van Vo\-mun\-a Pam\-rel\-vi\-yä.} The Ten Letter Game.~\LINK{http://www.pbs.org/wgbh/nova/blogs/secretlife/blogposts/learn-navi-with-avatars-paul-frommer/}{nova}
(\ding{43}~\B{pam\-rel}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{snapamrelvi}{\B{snapamrelvi}}, alphabet.

% pamrelvul
\hi
\HT{pamrelvul}{\B{pam\cp{rel}vul}} \I{[pam."REl.vul]} \emph{or} \I{[.vUl]} \emph{n.} pen, pencil, writing utensil (\emph{often shortened to} \ding{43}~\HL{}{\B{rel\-vul}}).
(i.) \B{pam\cp{rel}vul le\cp{rìn}}, pencil.
(\ding{43}~\HL{pamrel}{\B{pam\-rel}}, \HL{vul}{\B{vul}}; \HL{pensIl}{\B{pen\-sìl}})

% pamrìrlì'u
\hi 
\HT{pamrìrlI'u}{\B{pam\cp{rìr}lì'u}} \I{[pam."RIR.lI.Pu]} \emph{n.} onomatopoetic word.  
(\ding{43}~\HL{pam}{\B{pam}}, \HL{rI'Ir}{\B{rì'ìr}}, \HL{lI'u}{\B{lì'u}})

% pamtseo
\hi
\HT{pamtseo}{\B{\cp{pam}tseo}} \I{["pam.\t{ts}E.o]} \textsc{i}.~\emph{n.} music.
\B{Tsa\-fne\-pam\-tseo ke zaw\-prr\-te' oe\-ne.} I don't enjoy that kind of music.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Pam\-tse\-ol ngop ay\-re\-nut | Mì ron\-sem\-ä tì\-fnu}, Music creates patterns | In the silence of the mind.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Pam\-tseo\-ri po su\-lä\-ngìn nì\-hawng.} He is overly engrossed in his music (and I'm displeased about it).~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} \\
\textsc{ii}.~\HT{pamtseosi}{\B{\cp{pam}tseo si}} \I{["pam.\t{ts}E.o si]} \emph{vin.} play music, (\B{fa}) play an instrument. 
\B{Po pam\-tseo si fa au nìl\-tsan nì\-txan.} She plays the drums very well.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
(\ding{43}~\HL{pam}{\B{pam}}, \HL{tseo}{\B{tseo}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{pamtseotu}{\B{pamtseotu}}, musician. 
\on{2}.~\HL{pamtseowll}{\B{pamtseowll}}, cat ear (plant). 
\on{3}.~\HL{pamtseongopyu}{\B{pamtseongopyu}}, music creator, composer.
\on{4}.~\HL{pamtseovi}{\B{pamtseovi}}, musical piece.

% pamtseongopyu
\hi
\HT{pamtseongopyu}{\B{pamtseo\cp{ngop}yu}} \I{[pam.\t{ts}E.o."Nop\textcorner.ju]} \emph{n.} music creator, composer.
\B{Txo tì\-ke\-ftxo\-nga'\-a fmawn\-it ke stil\-vawm ay\-ngal, ze\-ne oe pi\-veng san Unil\-tì\-ran\-tokx\-ä pam\-tseo\-ngop\-yu a\-yaw\-ne to\-ler\-kä\-ngup.} If you haven't heard the sad news, I have to inform that the dear music creator of \emph{Avatar} has died.~\LINK{http://naviteri.org/2015/06/na-ayskxe-mi-telan-sad-news/}{b}
(\ding{43}~\HL{pamtseo}{\B{pam\-tseo}}, \HL{ngopyu}{\B{ngop\-yu}})

% pamtseotu
\hi
\HT{pamtseotu}{\B{\cp{pam}tseotu}} \I{["pam.\t{ts}E.o.tu]} \emph{n.} musician.
\B{Fì\-pam\-tseo\-tu\-ri ke la\-yu ftue fwa run fkol sä\-rawn\-it a tam.} It won't be easy to find a satisfactory replacement for this musician.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{pamtseo}{\B{pam\-tseo}}, \HL{tute1}{\B{tu\-te}})

% pamtseovi
\hi 
\HT{pamtseovi}{\B{\cp{pam}tseovi}} \I{["pam.\t{ts}E.o.vi]} \emph{n.} musical piece. 
\B{Aw\-nga tìng mik\-yun ay\-lì\-'u\-lu\-ke a pam\-tseo\-vir ko!} Let's listen to some music without words.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{pamtseo}{\B{pam\-tseo}})

% pamtseowll
\hi
\HT{pamtseowll}{\B{\cp{pam}tseowll}} \I{["pam.\t{ts}E.o.w\s{l}]} \emph{n.}, \ding{96} cat ear, ``music plant'', \emph{felinafolia ferruginea}.
(\ding{43}~\HL{pamtseo}{\B{pam\-tseo}}, \HL{'ewll}{\B{'ewll}})

% pamtsyìp
\hi 
\HT{pamtsyIp}{\B{\cp{pam}tsyìp}} \I{["pam.\t{ts}jIp\textcorner]} \emph{n.} small or light sound. 
(\emph{a. proverb}) \B{Tì\-tok slu tìk\-tok fa pam\-tsyìp a\-'aw.} Presence becomes absence by means of one little sound. [i.e. `Small things can make a big difference.']~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\ding{43}~\HL{pam}{\B{pam}}, \HL{tsyIp}{\B{-tsyìp}})

% pamuvan
\hi
\HT{pamuvan}{\B{\cp{pam}uvan}} \I{["pam.u.van]} \emph{n.} sound play (\emph{playing with and enjoying the sounds of language}).
(\ding{43}~\HL{pam}{\B{pam}}, \HL{uvan}{\B{uvan}})

% pan
\hi
\HT{pan}{\B{pan}} \I{[pan]} (\emph{dual} \ding{43}~\HL{mefan}{\B{me\cp{fan}}}) \emph{n., num.} third, one third, $\frac{1}{3}$.
\B{Tsu'\-tey\-ìl to\-lìng oer pan\-it smar\-ä.} Tsu'tey gave me a third of the prey.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}

% Pari
% \hi
% \B{Pa\cp{ri}} \I{[pa."Ri]} \emph{n.}, \textsc{lw} Paris (\emph{capital of France}).
% \B{Pa\-ri yaw\-ne lu oer nì\-ngay!} I truly love Paris.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}

% parul
\hi 
\HT{parul}{\B{pa\cp{rul}}} \I{[pa."Rul]} \emph{or} \I{[."RUl]} (\textsc{rn}: \B{pa\cp{rùl}} \I{[pa."RUl]}) \emph{n.} miracle. 
\B{Fwa ay\-ioang a\-pxay fì\-txan Na'\-vi\-ru srung so\-li fte Saw\-tu\-tet li\-vä\-txayn lu pa\-rul nì\-ngay.} That so many animals helped the Na’vi defeat the Sky People was a genuine miracle.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{parulnga'}{\B{parulnga'}}, miraculous. \\
\on{2}.~\HL{parultsyIp}{\B{parultsyìp}}, \emph{term of affection for children.}

% parulnga'
\hi 
\HT{parulnga'}{\B{pa\cp{rul}nga'}} \I{[pa."Rul.NaP]} \emph{or} \I{[."RUl.]} (\textsc{rn}: \B{pa\cp{rùl}nga'} \I{[pa."RUl.NaP]}) \emph{adj.} miraculous. 
(\ding{43}~\HL{parul}{\B{pa\-rul}})

% parultsyìp
\hi 
\HT{parultsyIp}{\B{pa\cp{rul}tsyìp}} \I{[pa."Rul.\t{ts}jIp\textcorner]} \emph{or} \I{[."RUl.]} (\textsc{rn}: \B{pa\cp{rùl}tsyìp} \I{[pa."RUl.\t{tS}Ip\textcorner]}) \emph{n. term of affection for children} (lit. `little wonder'). 
\B{Txon le\-fpom, ma pa\-rul\-tsyìp. Hi\-va\-haw nìm\-wey.} Good night, my dear little one. Sleep peacefully.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
(\ding{43}~\HL{parul}{\B{parul}})

% paskalin
\hi 
\HT{paskalin}{\B{paska\cp{lin}}} \I{[pa.ska."lin]} \emph{n.} sweet berry (term of endearment). 
\B{Hivahaw nìmwey, ma paskalin.} Sleep well, honey.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{pasuk}{\B{pa\-suk}}, \HL{kalin}{\B{ka\-lin}})

% pasuk
\hi 
\HT{pasuk}{\B{\cp{pa}suk}} \I{["pa.suk\textcorner]} \emph{or} \I{[.sUk\textcorner]} (\textsc{rn}: \B{\cp{pa}sùk} \I{["pa.sUk\textcorner]}) \emph{n.}, \ding{96} berry. 
\B{Tsay\-fa\-suk tsu\-ka\-nom lu krr\-ka fì\-zì\-sì\-krr nì\-'aw.} Those berries are available during this season only.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vozampasukut}{\B{vozampasukut}}, grinch tree. 
\on{2}.~\HL{paskalin}{\B{paskalin}}, sweet berry (term of endearment). 

% pate
\hi
\HT{pate}{\B{\cp{pa}te}} \I{["pa.tE]} \emph{vin.} get to a place, arrive. 
\B{Kar\-yul fngo\-lo' fu\-ta ay\-nu\-me\-yu pi\-va\-te ye'\-krr.} The teacher required the students to arrive early.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{pAhem}{\B{pä\-hem}})

% paw
\hi 
\HT{paw}{\B{paw}} \I{[paw]} \emph{vin.} grow, germinate and develop (of a plant). 
\B{Fì\-ut\-ral paw kil\-van\-lok nì\-'aw. Tsawl slu nì\-win nì\-txan.} This tree only grows (i.e., germinates, develops) near a river. It grows (i.e., gets big) very quickly.~\LINK{http://naviteri.org/2018/10/aysrr-ayvospxi-ayzisikrr-days-months-seasons/}{b} 
\B{Fkol yo\-lem nem\-fa kll\-te ut\-ral\-it a\-kawng ul\-te tsat pey\-ko\-law.} A bad tree was planted and was made to grow.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpaw}{\B{tìpaw}}, growth. 
\on{2}.~\HL{zIskrrmipaw}{\B{zìskrrmipaw}}, spring (season).

% pawk
\hi
\HT{pawk}{\B{pawk}} \I{[pawk\textcorner]} \emph{n.} horn, wind instrument. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txArpawk}{\B{txärpawk}}, Palulukan Bone Horn.

% pawm
\hi
\HT{pawm}{\B{pawm}} \I{[pawm]} \emph{vtr.\slash vin.} ask.
(\emph{a. transitive use}) 
\B{Pol po\-lawm teng\-a tì\-pawm\-it a li oel pal\-mawm trr\-am.} He asked the same question that I had already asked yesterday.~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b}
\B{Oel po\-ru po\-lawm fì\-'ut.} I asked him this.~\LINK{https://forum.learnnavi.org/intermediate/question-about-pawm/msg323238/\#msg323238}{ln}
(\emph{b. intransitive use})
\B{Alo a\-mrr po\-an po\-lawm, slä fra\-lo poe pol\-txe san ke\-he.} He asked five times, but each time she said, ``no.''~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\emph{c. with direct speech}) 
\B{Po\-lawm po san sra\-ke Sä\-li ho\-lum sìk.} He asked, ``Did Sally leave?''~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/}{ln}
(\emph{d. optional} \B{san … sìk}) 
\B{Po\-lawm po, Ney\-ti\-ri kä pe\-seng\-ne?} He asked where Neytiri was going.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} 
(\emph{e. with} \HL{ta}{\B{ta}} \emph{to form the indirect object})
\B{Pol po\-lawm tì\-pawm\-it oe\-ta.} He asked me a question.~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b}
(\emph{f. phrase}) \B{ätxä\-le si oe pi\-vawm …}, may I ask …
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to? (\emph{overly formal, stilted Na'vi})~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
(\ding{43}~\HL{vin}{\B{vin}}, \ding{214}~\HL{'eyng}{\B{'eyng}}) \\ 
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpawm}{\B{tì\-pawm}}, question.

% pay+
\hi
\HT{pay1}{\B{pay}}\texttt{+} \I{[paj.]} \emph{inter. pref. on nouns to mean} which (of them)? (\emph{plural})
\B{Pol\-txe pol pay\-lì\-'ut?} What did she say? [lit.: what words did she say?]~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{ay}{\B{ay}}\texttt{+})

% pay
\hi
\HT{pay2}{\B{pay}} \I{[paj]} \emph{n.} water, liquid. 
\B{Fra\-zì\-sì\-krr pay kil\-van\-ä nän nì\-'ul\-'ul.} Every season the river dries up more and more.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
\B{Na\-ri si! Kll\-te\-sìn lu pay a\-txan. Fwi rä\-'ä!} Be careful! There's a lot of water on the ground. Don't slip!~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Oe\-ri pay\-ìl tìng rì\-'ìr\-it key\-ä.} My face is reflected in the water.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Oel pay\-ti nul\-new to swo\-at.} I prefer water to alcohol.~\LINK{http://forum.learnnavi.org/language-updates/prefer-x-to-y-24474/}{ln}
\B{Fwa sar pay\-it fì\-txan lu le\-'al.} Using this much water is wasteful.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} 
\B{Pay\-ri som\-wew\-pe?} What's the temperature of the water?~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b} 
(\emph{a. idiom}) \B{Pay\-ìl a lipx tskxe\-ti ripx.} Dripping water pierces a stone. [{\em persistent effort can accomplish unexpected and amazing things}]~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{payfya}{\B{pay\-fya}}, stream. 
\on{2}.~\HL{paynga'}{\B{pay\-nga'}}, moist, humid, damp. 
\on{3}.~\HL{lepay}{\B{le\-pay}}, watery.
\on{4}.~\HL{yemfpay}{\B{yem\-fpay}}, dipping, immersion.
\on{5}.~\HL{paywll}{\B{pay\-wll}}, dapophet. 
\on{6}.~\HL{payIva}{\B{pay\-ìva}}, drop of water.   
\on{7}.~\HL{yapay}{\B{ya\-pay}}, mist. 
\on{8}.~\HL{reypay}{\B{rey\-pay}}, blood.  
\on{9}.~\HL{payoang}{\B{pay\-oang}}, fish.   
\on{10}.~\HL{txampay}{\B{txam\-pay}}, sea, ocean. 
\on{11}.~\HL{tskxepay}{\B{tskxe\-pay}}, ice. 
\on{12}.~\HL{tsngawpay}{\B{tsngaw\-pay}}, tears. 
\on{13}.~\HL{txumpaywll}{\B{txum\-pay\-wll}}, scorpion thistle. 
\on{14}.~\HL{paysena}{\B{pay\-se\-na}}, water container.
\on{15}.~\HL{kxumpay}{\B{kxum\-pay}}, gel. 
\on{16}.~\HL{paysmung}{\B{pay\-smung}}, water carrier.
\on{17}.~\HL{paytxew}{\B{pay\-txew}}, shoreline, water's edge.
\on{18}.~\HL{paysyul}{\B{pay\-syul}}, water lily.
\on{19}.~\HL{paytsim}{\B{pay\-tsim}}, water fountain.
\on{20}.~\HL{paymaut}{\B{pay\-ma\-ut}}, fountain tree.
\on{21}.~\HL{kxapay}{\B{kxapay}}, spit, saliva.

% payakan
\hi 
\HT{payakan}{\B{\cp{Pay}akan}} \I{["paj.a.kan]} \emph{n., name of a rogue tul\-kun}. 

% payfya 
\hi
\HT{payfya}{\B{\cp{pay}fya}} \I{["paj.fja]} \emph{n.} stream. 
\B{Po spä few pay\-fya fte smar\-it si\-vutx.} He jumped across the stream to track his prey.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{fya'o}{\B{fya'o}})

% payìva
\hi
\HT{payIva}{\B{\cp{pay}ìva}} \I{["paj.I.va]} \emph{n.} drop of water.
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{Ilva}{\B{ìlva}})

% paymaut
\hi 
\HT{paymaut}{\B{\cp{pay}maut}} \I{["paj.ma.ut\textcorner]} \emph{or} \I{[.Ut\textcorner]} \emph{n.} \ding{96} fountain tree [lit.: `liquid fruit']. 
(\ding{43}~\HL{pay}{\B{pay}}, \HL{mauti}{\B{mauti}}) 

% paynga'
\hi
\HT{paynga'}{\B{\cp{pay}nga'}} \I{["paj.NaP]} \emph{adj.} moist, damp, humid. 
\B{Fì\-fne\-rìn ke ley krr\-a slu pay\-nga'.} This kind of wood is worthless when it gets damp.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
\B{'ak\-ra a\-pay\-nga'}, moist soil.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{-nga'}{\B{-nga'}})

% payoang
\hi
\HT{payoang}{\B{pay\cp{o}ang}} \I{[paj."{}o.aN]} \emph{countable n., fauna} fish. 
\B{Lu tsa\-fne\-pay\-oang ftxì\-lor fra\-to krr\-a lu yrr.} That kind of fish is most tasty when eaten as \emph{sashimi}.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Lu fay\-oang a\-lor mì hil\-van Ey\-wa\-'e\-veng\-ä.} There are beautiful fish in Pandora's rivers.~\LINK{http://naviteri.org/2010/07/thoughts-on-ambiguity/}{b}
\B{Oe\-yä txin\-tìn lu fwa stä'\-nì fay\-oang\-it.} My central societal occupation is to catch fish.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tì\-kan fì\-pay\-oang\-ä ke lu fwa tsun fko pot yi\-vom; lu oe\-yä 'eyl\-yong.} This fish is not meant to be eaten; it's my pet.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b}
\B{Fol ko\-la\-nom po\-ta ay\-srok\-it fay\-oang\-yoa.} They acquired beads from him in exchange for fish.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{ioang}{\B{ioang}})

% paysena
\hi
\HT{paysena}{\B{\cp{pay}sena}} \I{["paj.sE.na]} \emph{n.} water container. 
\B{\mbox{Tsngal} lu fne\-pay\-se\-na.} A cup is a type of water container.~{\color{blue}\href{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{\textsc{b}}}
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{sAhena}{\B{sä\-he\-na}})

% paysmung
\hi
\HT{paysmung}{\B{\cp{pay}smung}} \I{["paj.smuN]} \emph{or} \I{[.smUN]} (\textsc{rn}: \B{\cp{pay}smùng} \I{["paj.smUN]}) \emph{n.} water carrier. 
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{sAmunge}{\B{sä\-mu\-nge}})

% paystan
\hi
\HT{paystan}{\B{pay\cp{stan}}} \I{[paj."{}stan]} \emph{inter.} who? what men? (\emph{alternate form:} \B{(ay)stampe})
(\ding{43}~\HL{pay1}{\B{pay}}\texttt{+}, \HL{tutan}{\B{tu\-tan}})

% payste
\hi
\HT{payste}{\B{pay\cp{ste}}} \I{[paj."{}stE]} \emph{inter.} who? what women? (\emph{alternate form:} \B{(ay)stepe})
(\ding{43}~\HL{pay1}{\B{pay}}\texttt{+}, \HL{tute2}{\B{tu\-té}})

% paysu
\hi
\HT{paysu}{\B{pay\cp{su}}} \I{[paj."{}su]} \emph{inter.} who? what people? (\emph{alternate form:} \B{(ay)supe})
(\ding{43}~\HL{pay1}{\B{pay}}\texttt{+}, \HL{tute1}{\B{tu\-te}})

% paysyul
\hi 
\HT{paysyul}{\B{\cp{pay}syul}} \I{["paj.sjul]} \emph{or} \I{[.sjUl]} \emph{n.} \ding{96} water lily, \emph{inrigo lilliam}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln}
(\ding{43}~\HL{pay}{\B{pay}}, \HL{syulang}{\B{syu\-lang}})

% paytxew
\hi
\HT{paytxew}{\B{pay\cp{txew}}} \I{[paj."t'Ew]} \emph{n.} shoreline, water's edge.
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{txew}{\B{txew}})

% paytsim
\hi 
\HT{paytsim}{\B{\cp{pay}tsim}} \I{["paj.\t{ts}im]} \emph{n.} water fountain.~\LINK{https://forum.learnnavi.org/vocab-phrases/words-from-the-disney-cast-members-training-on-the-na'vi-language/}{ln} 
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{tsim}{\B{tsim}})

% paywll
\hi
\HT{paywll}{\B{\cp{pay}wll}} \I{["paj.w\s{l}]} \emph{n.}, \ding{96}  dapophet, ``water plant'', \emph{aloeparilus succulentus}.
(\ding{43}~\HL{pay2}{\B{pay}}, \HL{'ewll}{\B{'ewll}})

% pähem
\hi
\HT{pAhem}{\B{\cp{pä}hem}} \I{["p\ae.hEm]} \emph{vin.} arrive.
\B{Fo\-ru 'upxa\-ret oel fpo\-le', slä vay set ke pa\-mä\-hä\-ngem kea tì\-'eyng.} I have sent them a message, but up to now no answer has arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Fì\-tseo\-ri ke tsun kaw\-tu pi\-vä\-hem tì\-yo'\-ne; tsran\-ten tì\-pä\-hem\-ä tì\-fmi nì\-'aw.} In this art it's impossible to arrive at perfection; the only thing that matters is the attempt to arrive there.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Ne 'o\-ra\-tsyìp po\-lä\-hem, yak si nì\-ftär.} When you arrive at the pond, turn to the left.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\emph{a. proverb}) \B{Nì\-win 'ìp tì\-'e\-wan, nì\-win pä\-hem tì\-ko\-ak.} Youth vanishes quickly, old age arrives quickly.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} 
(\ding{43}~\HL{pate}{\B{pa\-te}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpAhem}{\B{tì\-pä\-hem}}, arrival.  
\on{2}.~\HL{kllpA}{\B{kll\-pä}}, land, reach the ground.

% pänu
\hi
\HT{pAnu}{\B{\cp{pä}nu}} \I{["p\ae.nu]} \emph{n.} promise. \\ 
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{pAnutIng}{\B{pä\-nu\-tìng}}, promise.

% pänutìng
\hi
\HT{panutIng}{\B{\cp{pä}nutìng}} \I{["p\ae.nu.t$\cdot$IN]} \emph{vtr.}\on{33} promise (s.t. to s.o.). 
\B{Ney\-ti\-ri\-ti fkol pä\-nu\-to\-lìng oe\-ru!} Neytiri was promised to me!~\LINK{http://forum.learnnavi.org/language-updates/panuting/}{ln\slash a\textsuperscript{c}}
\B{Lo\-ak\-ìl pä\-nu\-to\-lìng fu\-ta kar oe\-ru fya\-'ot a 'ìp fko nem\-fa ewll.} Loak promised he'd teach me how to vanish into the bushes.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{pAnu}{\B{pä\-nu}}, \HL{tIng}{\B{tìng}})

% pängkxo
\hi
\HT{pAngkxo}{\B{päng\cp{kxo}}} \I{[p\ae{}N."k'o]} \emph{vin.} chat, converse, have a conversation (\B{hu}, with s.o.).
\B{Oe pxìm päng\-kxo nga\-hu nì\-ron\-srel.} I often talk with you in my imagination.\slash I often imagine I'm talking with you.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Tsun oe nga\-hu nì\-Na'\-vi pi\-väng\-kxo a fì\-'u oe\-ru prr\-te' lu.} It's a pleasure to be able to chat with you in Na'vi.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Fay\-upxa\-re\-mì oe pa\-yäng\-kxo te\-ri ho\-ren lì'\-fya\-yä le\-Na'\-vi.} In these messages I will chat about the rules of the Na'vi language.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Fo ke pe\-räng\-kxo oe\-hu nìk\-mar.} They were chatting with me at a bad time.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}\\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpAngkxo}{\B{tì\-päng\-kxo}}, chat, conversation. 
\on{2}.~\HL{pAngkxoyu}{\B{päng\-kxo\-yu}}, chatter, converser. 
\on{3}.~\HL{tireapAngkxo}{\B{ti\-re\-a\-päng\-kxo}}, commune.

% pängkxoyu
\hi
\HT{pAngkxoyu}{\B{päng\cp{kxo}yu}} \I{[p\ae{}N."k'o.ju]} \emph{n.} chatter, converser.
(\ding{43}~\HL{pAngkxo}{\B{päng\-kxo}})

% pängkxoyu lekoren
\hi
\HT{pAngkxoyulekoren}{\B{päng\cp{kxo}yu leko\cp{ren}}} \I{[p\ae{}N."k'o.ju lE.ko."REn]} \emph{n.} lawyer, rule chatter.

% pätsì
\hi
\HT{pAtsI}{\B{\cp{pä}tsì}} \I{["p\ae.\t{ts}I]} \emph{n.}, \textsc{lw} badge. 

% pe+
\hi
\HT{pe1}{\B{pe}\texttt{+}} \I{[pE.]} \emph{inter. pref. on nouns to mean} what? which?
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to? (\emph{overly formal, stilted Na'vi})~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
\B{Nga\-ru lu pe\-fne\-txin\-tìn nì\-trr\-trr?} What is your primary role (in society)?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Nga\-ri tstxo lu pe\-lì\-'u?} What's your name?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pay1}{\B{pay}\texttt{+}}, which (of them)? (plural). 
\on{2}.~\HL{pem}{\B{pem}\texttt{+}}, what two? 
\on{3}.~\HL{pep}{\B{pep}\texttt{+}}, what three? 
\on{4}.~\HL{pefne}{\B{pefne}\texttt{-}}, what kind?

% -pe
\hi
\HT{pe2}{\B{-pe}} \I{[.pE]} \emph{inter. suf. on nouns to mean} what? which?
\B{Ya\-ri som\-wew\-pe?} What's the temperature?~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{Kil\-van\-ìri tsa\-tseng slo\-snep\-pe?} How wide is the river there?~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}

% pe'ngay
\hi 
\HT{pe'ngay}{\B{pe'\cp{ngay}}} \I{[p$\cdot$EP."Naj]} \emph{vin., inf.}\on{11} judge, conclude (\emph{used with} \ding{43}~\HL{tsnI}{\B{tsnì}}). 
\B{Po\-ri key\-rel\-fa oe po\-le'\-ngay tsnì ke new zi\-va\-'u.} From her expression, I concluded that she didn’t want to come.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 
(\ding{43}~\HL{pe'un}{\B{pe\-'un}}, \HL{ngay}{\B{ngay}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpe'ngay}{\B{tì\-pe'\-ngay}}, conclusion.
\on{2}.~\HL{pe'ngayyu}{\B{pe'\-ngay\-yu}}, judge.

% pe'ngayyu
\hi 
\HT{pe'ngayyu}{\B{pe'\cp{ngay}yu}} \I{[pEP."Naj.ju]} \emph{n.} judge (\emph{as a person}). 
(\ding{43}~\HL{pe'ngay}{\B{pe'\-ngay}})

% pe'un
\hi
\HT{pe'un}{\B{\cp{pe}'un}} \I{["pE.Pun]} \emph{or} \I{[.PUn]} (\textsc{rn}: \B{\cp{pe}ùn} \I{["pE.Un]}) \emph{vtr.} decide.  
\B{Tì\-'eyng\-it oel to\-lel a krr, tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} When I receive an answer, I will then decide the best way to bring you the words and rules of Na'vi.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Pol po\-le\-'un fu\-ta pe\-hem si nì\-'en.} He decided what to do on a hunch.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpe'un}{\B{tì\-pe\-'un}}, decision. 
\on{2}.~\HL{pe'ngay}{\B{pe'\-ngay}}, judge, conclude.

% pefne-
\hi
\HT{pefne}{\B{pefne-}} \I{[pE.fnE.]} \emph{inter. pref. on nouns to mean} which kind?
\B{Nga\-ru lu pe\-fne\-txin\-tìn nì\-trr\-trr?} What is your primary role (in society)?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}

% pefnel
\hi
\HT{pefnel}{\B{pe\cp{fnel}}} \I{[pE."fnEl]} \emph{inter.} which kind? (\emph{alternate of} \ding{43}~\HL{fnepe}{\B{fne\-pe}})

% pefya
\hi
\HT{pefya}{\B{pe\cp{fya}}} \I{[pE."fja]} \emph{inter.} how? (\emph{alternate of} \ding{43}~\HL{fyape}{\B{fya\-pe}})
\B{Ya\-fkeyk (za\-'u) pe\-fya?} How is the weather?~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Pe\-fya nga fpìl? Oeng swey\-lu txo ki\-vä fu\-ke?} What do you think? Should we go or not?~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{pe1}{\B{pe}}\texttt{+}, \HL{fya'o}{\B{fya\-'o}})

% pefyinep'ang
\hi
\HT{pefyinep'ang}{\B{pefyinep\cp{'ang}}} \I{[pE.fjin.Ep."PaN]} \emph{inter.} how complex?
\B{Ngal ke tslä\-ngam teyng\-ta fì\-tì\-ngä\-zìk\-ìri pe\-fyin\-ep\-'ang.} Unfortunately you don't understand how complex this problem is.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{pe1}{\B{pe}}\texttt{+}, \HL{fyinep'ang}{\B{fyin\-ep\-'ang}})

% pehem
\hi
\HT{pehem}{\B{pe\cp{hem}}} \I{[pE."hEm]} \textsc{i}. \emph{inter.} what (action)? (\emph{alternate of} \ding{43}~\HL{kempe}{\B{kem\-pe}}) 
\B{Pol po\-le\-'un fu\-ta pe\-hem si nì\-'en.} He decided what to do on a hunch.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\
\textsc{ii}. \B{pe\cp{hem} si} \I{[pE."hEm si]} \emph{vin., inter.} do what (action)? 
(\emph{a. idiom}) \B{Tsun pe\-hem?} (\emph{short for} \B{Tsun fko pe\-hem si\-vi?}) `What can one do?' 
\B{Ri\-ni yaw\-ne lu oer, slä oe yaw\-ne ke lä\-ngu por kaw\-'it.---Tsun pe\-hem?} I'm in love with Rini, but she doesn't love me one bit.---What are you gonna do? That's life.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
(\ding{43}~\HL{pe1}{\B{pe}}\texttt{+}, \HL{kem}{\B{kem}})

% pehrr
\hi
\HT{pehrr}{\B{pe\cp{hrr}}} \I{[pE."h\s{r}]} \emph{inter.} when (time)? (\emph{alternate of} \ding{43}~\HL{krrpe}{\B{krr\-pe}}). 
(\ding{43}~\HL{pe}{\B{pe}\texttt{+}}, \HL{krr}{\B{krr}})

% pehrrlik
\hi 
\B{pe\cp{hrr}lik} \I{[pE."h\s{r}.lik\textcorner]} \emph{inter.} what point in time? what hour? (\emph{alternate form of} \ding{43} \HL{krrlik}{\B{krrlikpe}})

% pek
\hi 
\HT{pek}{\B{pek}} \I{[pEk\textcorner]} \emph{n.} fin (\emph{of an aquatic animal}).

% pekxinum 
\hi
\HT{pekxinum}{\B{pe\cp{kxi}num}} \I{[pE."k'i.num]} \emph{or} \I{[.nUm]} (\textsc{rn}: \B{pe\cp{kxi}nùm} \I{[pE."gi.nUm]}) \emph{inter.} how tight\slash loose? (\emph{alternate form:} \ding{43}~\HL{'ekxinum}{\B{'e\-kxin\-um\-pe}})
\B{Ngal mo\-lay' pxaw\-pxun\-it Lo\-ak\-ä a krr pe\-kxin\-um?} When you tried on Loak's armband, how tight was it?~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{pe1}{\B{pe}}\texttt{+}, \HL{'ekxinum}{\B{'e\-kxi\-num}})

% pela'a
\hi 
\HT{pela'a}{\B{pe\cp{la}'a}} \I{[pE."la.Pa]} \emph{inter.}  how near? how far? what distance?. 
\B{Ftu Kel\-ut\-ral ne Txin\-tseng Saw\-tu\-te\-yä pe\-la\-'a?} How far is Hometree from Hell's Gate?~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b}
(\ding{43}~\HL{la'a}{\B{la'a}}, \HL{pe}{\B{pe}\texttt{+}}; \HL{pelImsim}{\B{pelìmsim\slash lìmsimpe}})

% pelìmsim
\hi 
\HT{pelImsim}{\B{pe\cp{lìm}sim}} \I{[pE."lIm.sim]} \emph{inter.} how near? how far? what distance? (\emph{less common than} \ding{43}~\HL{pela'a}{\B{pela'a}})
(\ding{43}~\HL{lImsimpe}{\B{lìmsimpe}})

% pelun
\hi
\HT{pelun}{\B{pe\cp{lun}}} \I{[pE."lun]} \emph{or} \I{[."lUn]} \emph{inter.} why? (\emph{alternative form:} \ding{43}~\HL{lumpe}{\B{lum\-pe}}).   
\B{Fì\-txon na ton a\-la\-he nì\-wotx pe\-lun ke lu teng?} Why is this night not the same as all the other nights?~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
\B{Nga pe\-lun wä\-pe\-ran?} Why are you hiding?~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
(\ding{43}~\HL{pe1}{\B{pe}}\texttt{+}, \HL{lun}{\B{lun}})

% pem+
\hi
\HT{pem}{\B{pem}\texttt{+}} \I{[pEm.]} \emph{inter. pref. on nouns to mean} what two? which two?
(\ding{43}~\HL{}{\B{pe}}\texttt{+}, \HL{me}{\B{me}}\texttt{+})

% pemsu
\hi
\HT{pemsu}{\B{pem\cp{su}}} \I{[pEm."{}su]} \emph{inter.} who? what two people?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b} (\emph{alternate form:} \ding{43}~\HL{mesupe}{\B{me\-su\-pe}})
(\ding{43}~\HL{pem}{\B{pem}}\texttt{+}, \HL{tute1}{\B{tu\-te}}) 
	
% pemstan
\hi
\HT{pemstan}{\B{pem\cp{stan}}} \I{[pEm."{}stan]} \emph{inter.} who? what two men?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b} (\emph{alternate form:} \ding{43}~\HL{mestampe}{\B{me\-stam\-pe}})
(\ding{43}~\HL{pem}{\B{pem}}\texttt{+}, \HL{tutan}{\B{tu\-tan}}) 	
	
% pemste
\hi
\HT{pemste}{\B{pem\cp{ste}}} \I{[pEm."{}stE]} \emph{inter.} who? what two women?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b} (\emph{alternate form:} \ding{43}~\HL{}{\B{me\-ste\-pe}})
(\ding{43}~\HL{pem}{\B{pem}}\texttt{+}, \HL{tute2}{\B{tuté}}) 

% pensìl
\hi
\HT{pensIl}{\B{\cp{pen}sìl}} \I{["pEn.sIl]} \emph{n.}, \textsc{lw} pencil.
(\ding{43}~\HL{pamrelvul}{\B{pam\-rel\-vul}})

% peng
\hi
\HT{peng}{\B{peng}} \I{[pEN]} \emph{vtr.} tell, inform. 
\B{Lu poe tstun\-wi, ul\-te evi\-ru peng el\-tur tì\-txen si a ay\-vur\-it te\-ri 'Rr\-ta, alu tse\-nge a\-stxong nì\-ngay.} She is kind and tells the kids interesting stories about the truly strange place Earth.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Tì\-'eyng\-it oel to\-lel a krr, ay\-nga\-ru pa\-yeng, tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} When I receive an answer, I will let you know, and I will then decide the best way to bring you the words and rules of Na'vi.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Pi\-veng oer ftxey nga new ri\-vey fu\-ke.} Tell me whether you want to live (or not).~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/}{ln}
\B{New oe me\-nga\-ru pi\-veng fu\-ta tì\-kang\-kem\-ìri a\-txan\-tsan me\-nge\-yä fpi ``Trr Le\-fpom ma Ame\-ri\-ka'' oe\-ru te\-ya so\-li nì\-txan.} I want to tell you both that I am very much filled with joy for your excellent work for ``Good Morning Amerika.''~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/}{ln} 
(\emph{in indirect speech with} \B{fmawn\-ta}) 
\B{Ngal po\-leng oer fmawn\-ta po to\-ler\-kup.} You told me that he died.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{ralpeng}{\B{ralpeng}}, translate, interpret. 
\on{2}.~\HL{penghrrap}{\B{penghrrap}}, binary sunshine. 
\on{3}.~\HL{penghrr}{\B{penghrr}}, warn.

% penghrr
\hi 
\HT{penghrr}{\B{peng\cp{hrr}}} \I{[p$\cdot$EN."h\s{r}]} \emph{vin., inf.}\on{11} warn (\B{te\-ri}, about). 
\B{Tsyeyk Na'\-vi\-ru po\-leng\-hrr te\-ri Saw\-tu\-te.} Jake warned the Na'vi about the Skypeople.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{pllhrr}{\B{pll\-hrr}}; \HL{peng}{\B{peng}}, \HL{hrrap}{\B{hrr\-ap}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sApenghrr}{\B{säpenghrr}}, warning.

% penghrrap
\hi
\HT{penghrrap}{\B{peng\cp{hrr}ap}} \I{[pEN."h\s{r}.ap\textcorner]} \emph{n.}, \ding{96} binary sunshine, fringed lamp, \emph{lucinaria fibriata}.
(\ding{43}~\HL{peng}{\B{peng}}, \HL{hrrap}{\B{hrr\-ap}}) 

% pengimpup
\hi 
\HT{pengimpup}{\B{pe\cp{ngim}pup}} \I{[pE."Nim.pup\textcorner]} \emph{or} \I{[.pUp\textcorner]} (\textsc{rn}: \B{pe\cp{ngim}pùp} \I{[pE."Nim.pUp\textcorner]}) \emph{inter.} what length? how long (of physical extension)? (\emph{alternate form:} \ding{43}~\HL{ngimpup}{\B{ngim\-pup\-pe}}) 

% penunyol
\hi 
\HT{penunyol}{\B{pe\cp{nun}yol}} \I{[pE."nun.yol]} \emph{or} \I{[."nUn.]} (\textsc{rn}: \B{pe\cp{nun}yol} \I{[pE."nun.yol]})  \emph{inter.} what length? how long (of time)? (\emph{alternate form:} \ding{43}~\HL{nunyol}{\B{nun\-yol\-pe}}). 
\B{Nga har\-ma\-haw pe\-nun\-yol?} How long were you sleeping?~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}

% pep+
\hi
\HT{pep}{\B{pep}\texttt{+}} \I{[pEp\textcorner.]} \emph{inter. pref. on nouns to mean} what three? which three?
(\ding{43}~\HL{pe1}{\B{pe}}\texttt{+}, \HL{pxe}{\B{pxe}}\texttt{+})

% pepsu
\hi
\HT{pepsu}{\B{pep\cp{su}}} \I{[pEp\textcorner."{}su]} \emph{inter.} who? what three people? \emph{alternate form:} \ding{43}~\HL{pxesupe}{\B{pxe\-su\-pe}}
\B{Pep\-sul tok fì\-tse\-nget?} What three people are here?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b} 
(\ding{43}~\HL{pep}{\B{pep}}\texttt{+}, \HL{tute1}{\B{tu\-te}}) 
	
% pepstan
\hi
\HT{pepstan}{\B{pep\cp{stan}}} \I{[pEp\textcorner."{}stan]} \emph{inter.} who? what three men? \emph{alternate form:} \ding{43}~\HL{pxestampe}{\B{pxe\-stam\-pe}}
(\ding{43}~\HL{pep}{\B{pep}}\texttt{+}, \HL{tutan}{\B{tu\-tan}}) 	
	
% pepste
\hi
\HT{pepste}{\B{pep\cp{ste}}} \I{[pEp\textcorner."{}stE]} \emph{inter.} who? what three women? \emph{alternate form:} \ding{43}~\HL{pxestepe}{\B{pxe\-ste\-pe}}
(\ding{43}~\HL{pep}{\B{pep}}\texttt{+}, \HL{tute2}{\B{tu\-té}}) 

% peseng
\hi
\HT{peseng}{\B{pe\cp{seng}}} \I{[pE."{}sEN]} \emph{or} \B{pe\cp{se}nge} \I{[pE."{}sE.NE]} \emph{inter.} where? what place? (\emph{alternative form:} \ding{43}~\HL{tsengpe}{\B{tseng\-pe}})
\B{Tok pe\-se\-nget pam\-rel\-ìl?} Where is the text?~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
\B{Oe\-ta a tsa\-sä\-srìn\-ìl tok pe\-seng\-it?} Where's the thing (you) borrowed from me?~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Nga zo\-la\-'u ftu pe\-seng?} Where are you from?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Vo\-lin oel teyng\-ta Ney\-ti\-ri kä pe\-seng\-ne.} I asked where Neytiri was going.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{tseng}{\B{tse\-nge}}) 

% peslosnep
\hi
\HT{peslosnep}{\B{peslo\cp{snep}}} \I{[pE.slo."{}snEp\textcorner]} \emph{inter.} what width? how wide? (\emph{alternate form:} \ding{43}~\HL{slosneppe}{\B{slo\-snep\-pe}})
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{slosnep}{\B{slo\-snep}}) 

% pesomwew
\hi
\HT{pesomwew}{\B{pesom\cp{wew}}} \I{[pE.som."wEw]} \emph{inter.} what temperature? (\emph{alternate form:} \ding{43}~\HL{somwewpe}{\B{som\-wew\-pe}})
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{somwew}{\B{som\-wew}}) 

% pesrrpxì
\hi 
\B{pesrr\cp{pxì}} \I{[pE.s\s{r}."p'I]} \emph{inter.} what part of the day? what hour? (\emph{alternate form:} \ding{43}~\HL{trrpxI}{\B{trr\-pxì\-pe}}) 
\B{Fo pä\-hem pe\-srr\-pxì?---Sre\-kam\-trr.} What part of the day will they arrive?---Before noon.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} 
(\ding{43}~\HL{krrlik}{\B{krrlikpe\slash pe\-hrr\-lik}})

% pestan
\hi
\HT{pestan}{\B{pe\cp{stan}} \emph{or} \B{tu\cp{tam}pe}} \I{[pE."{}stan]} \emph{inter.} who? what man? \emph{alternate form:} \ding{43}~\HL{tutampe}{\B{tu\-tam\-pe}}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{tutan}{\B{tu\-tan}}) 

% peste
\hi
\HT{peste}{\B{pe\cp{ste}} \emph{or} \B{tu\cp{te}pe}} \I{[pE."{}stE]} \emph{inter.} who? what woman? \emph{alternate form:} \ding{43}~\HL{tutepe}{\B{tu\-te\-pe}}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{tute2}{\B{tu\-té}}) 

% pesu
\hi
\HT{pesu}{\B{pe\cp{su}}} \I{[pE."{}su]} \emph{inter.} who? what person? (\emph{alternate form:} \ding{43}~\HL{tupe}{\B{tu\-pe}})
\B{Nge\-nga lu pe\-su?} Who are you?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Fo lu pesu?} Who are they?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
\B{Nga pe\-su\-hu kä\-teng nì\-trr\-trr?} Who do you usually hang out with?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{tute1}{\B{tu\-te}}) 

% peu 
\hi
\HT{peu}{\B{pe\cp{u}}} \I{[pE."{}u]} \emph{inter.} what (thing)? (\emph{alternate form:} \ding{43}~\HL{'upe}{\B{'u\-pe}})
\B{Tì\-'o'\-ìri pe\-u su\-nu ngar fra\-to?} What is your favorite way to have fun?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Nga\-tsyìp\-ìl new pe\-ut ta oe?} What does little you want from me?~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{'u}{\B{'u}}) 

% pewn
\hi
\HT{pewn}{\B{pewn}} \I{[pEwn]} \emph{n.} neck. 
\B{Fkxi\-le pewn\-ta tu\-té\-yä kur.} The bib necklace hangs from the woman's neck.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{a. idiom})
\B{Ke tsun fko tspi\-vang to\-ruk\-it fa fwa pewn\-ti snew.} \emph{Proverbial expression for a method that will not work.} [lit.: You can't kill a great leonopteryx by constricting its throat]~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} 
(\emph{shortened to}) \B{pewn to\-ruk\-ä}. \B{Sra\-ke fpìl ay\-ngal fu\-ta aw\-nga tsun hu Saw\-tu\-te a tsam\-it 'i\-vaw\-nìm fa fwa päng\-kxo fo\-hu nì\-'aw, ma smuk? Pewn to\-ruk\-ä!} Do you think we can avoid a war with the Sky People solely by talking with them, my brothers? That won't work!~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/comment-page-1/\#comment-1755}{b}

% pey
\hi
\HT{pey}{\B{pey}} \I{[pEj]} \emph{vin.} wait, wait for (\emph{with the topic}).
\B{Ul\-te o\-mum oel fu\-ta tì\-fya\-wìn\-txu\-ri oe\-yä pe\-rey ay\-nga nì\-wotx.} And I know you are all waiting for my guidance.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Nì\-ay\-nga oe pe\-rey nì\-teng.} Like you, I too am waiting.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ke tsun aw\-nga pi\-vey nul\-krr--txew lok.} We can't wait any longer--time is almost up.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sIlpey}{\B{sìl\-pey}}, hope. 
\on{2}.~\HL{maweypey}{\B{ma\-wey\-pey}}, be patient. 
\on{3}.~\HL{srefey}{\B{sre\-fey}}, expect. 
\on{4}.~\HL{fe'pey}{\B{fe'\-pey}}, feel dread, fear.

\hi
\HT{-pey}{\B{-pey}} \I{[.pEj]} \emph{num., suf.} three (\emph{rest form of} \ding{43}~\HL{pxey}{\B{pxey}} \emph{to form higher numbers}). \B{vopey}, \on{11} (decimal), \on{13} (octal). \B{me\-vo\-pey}, \on{19} (decimal), \on{23} (octal); etc.

% peyä
\hi
\B{\cp{pe}yä}~:~\HL{po}{\B{po}}.

% Peyral
\hi
\B{Pey\cp{ral}} \I{[pEj."Ral]} \emph{female name}.
\B{Pll\-txe fra\-po san Pey\-ral ke tsun ri\-vol nìl\-tsan sìk.} They say that Peyral can't sing well.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
\B{Pey\-ral\-ìl zet wur\-a wu\-tsot a\-'aw\-nem pxel sngel.} Peyral treats a cool cooked meal like garbage.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Kä\-sro\-lìn oel nik\-ro\-it Pey\-ral\-ur yoa fwa po rol oer.} I loaned Peyral a hair ornament in exchange for her singing to me.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Ìlä Fey\-ral, mun\-txa so\-li Ra\-lu sì Ne\-wey nì\-wan me\-srr\-am.} According to Peyral, Ralu and Newey were secretly married the day before yesterday.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}

% piak
\hi
\HT{piak}{\B{pi\cp{ak}}} \I{[pi."{}ak\textcorner]} \textsc{i}. \emph{adj.} (\emph{a.}) open.   
(\emph{b. of the weather}) no clouds, completely clear.
\B{Ya\-fkeyk fya\-pe set?--Taw lu pi\-ak.} What's the weather like now?--The sky is clear.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} \\
\textsc{ii}. \HT{piaksi}{\B{piak si}} \I{[pi."{}ak\textcorner{} si]} \emph{vin.} open.
\B{Oe\-ri me\-na\-ri\-ru pi\-ak si.} I open my eyes.~\LINK{http://forum.learnnavi.org/language-updates/open-close/}{ln} \\
\textsc{iii}. \HT{piaksApi}{\B{piak säpi}} \I{[pi."{}ak\textcorner{} s\ae."pi]} \emph{vin.} open by itself, open on its own.
\B{Po\-ri me\-na\-ri pi\-ak sä\-po\-li.} His eyes opened.~\LINK{http://forum.learnnavi.org/language-updates/open-close/}{ln}
(\ding{214}~\HL{tstu}{\B{tstu}})

% pil
\hi
\HT{pil}{\B{pil}} \I{[pil]} \emph{n.} facial stripe.
\B{ean na pil}, facial\hyp{}stripe blue.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Hu\-fwa lu fil\-ur Va'\-ru\-ä fne\-fe'\-ran\-vi, tsal\-su\-ngay fpìl fu\-ta say\-rìp lu nì\-txan.} Although Va'ru's facial stripes are rather uneven, I still think he's very handsome.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\emph{a. proverbial expression}) \B{Po\-ri fra\-pil sì fra\-sil lor lu nì\-txan.} He\slash She is very attractive.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b} 
\B{Fra\-pil fra\-sil!} `Wow! Check him\slash her out! He\slash she is gorgeous!'~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{teylupil}{\B{tey\-lu\-pil}}, \emph{teylu}\hyp{}face (\emph{childish insult}). 

% pinvul
\hi 
\HT{pinvul}{\B{\cp{pin}vul}} \I{["pin.vul]} \emph{or} \I{[.vUl]} \emph{n.} crayon (\emph{short form of} \ding{43}~\HL{'opinvultsyIp}{\B{'opin\-vul\-tsyìp}}). 
\B{Pol rel\-it wo\-leyn fa pin\-vul.} He drew a picture with a crayon.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/\#comment-28355}{b}

% pizayu
\hi
\HT{pizayu}{\B{\cp{pi}zayu}} \I{["pi.za.ju]} \emph{n.} ancestor. 
\B{Vay fwa zo\-la\-'u Tsyeyk\-Su\-li, To\-ruk Mak\-to alu pi\-za\-yu Ney\-ti\-ri\-yä txan\-rar\-mo\-'a fra\-to kip ay\-ha\-pxì\-tu Oma\-ti\-ka\-yaä.} Until Jake Sully arrived, Neytiri's ancestor was the most famous Toruk Makto among the Omatikaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{a. in storytelling}) \emph{opening formula} \B{Ay\-fi\-za\-yu pll\-txu\-ye san trro la\-mu krr a …}, Once upon a time … [lit.: the ancestors say, some day there was a time that] 
\emph{closing formula} \B{Fì\-fya pll\-txu\-ye ay\-fi\-za\-yu.} Thus say the ancestors.~\LINK{http://forum.learnnavi.org/intermediate/translation-exercise-a-navajo-coyote-tale-in-navi/}{ln}

% pìlok
\hi
\HT{pIlok}{\B{pì\cp{lok}}} \I{[pI."lok\textcorner]} \emph{n.}, \textsc{lw} blog.
\B{Sìl\-pey oe, aw\-nge\-yä lì'\-fya\-olo'\-ìri fì\-pì\-lok lì\-ye\-vu pxan.} I hope, this blog will be worth for our language community.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Zo\-la\-'u nì\-prr\-te' ne pì\-lok Na'\-vi\-te\-ri.} Welcome to the \emph{Na'\-vi\-te\-ri blog}.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Lu pì\-lok\-ur pxe\-sì\-kan sì pxe\-fne\-'upxa\-re.} The blog has three aims and three kinds of messages.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Pì\-lok\-ä fì\-ha\-pxì\-yä tì\-kan lu law.} The aim of this part of the blog is clear.~\LINK{http://naviteri.org/2010/06/first-post/}{b}

% pìmtxan
\hi
\HT{pImtxan}{\B{pìm\cp{txan}}} \I{[pIm."t'an]} \emph{inter., adj.} how much? (\emph{alternate form:} \ding{43}~\HL{hImtxampe}{\B{hìm\-txam\-pe}})
\B{Fì\-nik\-roi ley pìm\-txan?} How valuable is this hair ornament?~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
\B{pìmtxana pay?} how much water?~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b}
\B{pìmtxana tìsom?} how hot?~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{hImtxan}{\B{hìm\-txan}}) 

% pìsaw
\hi 
\HT{pIsaw}{\B{pì\cp{saw}}} \I{[pI."{}saw]} (\emph{a. adj.}) clumsy, accident\hyp{}prone. 
\B{Po lu pì\-saw. Trr\-am tol\-tem ve\-nu\-ti sne\-yä nìt\-kan\-lu\-ke.} He is accident\hyp{}prone. Yesterday he unintentionally shot his own foot.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
\B{Lu Saw\-tu\-te wok sì pì\-saw nì\-wotx, na prr\-nen.} The skypeople are all loud and clumsy, like a baby.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
\B{Ra\-nu kil\-van\-mì fyar\-mu\-fye na tsawl\-a pa\-yo\-ang a\-pì\-saw.} Ra\-nu was splashing in the river like a big clumsy fish.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(\emph{b. intj.}) \emph{e.g. when you've acted clumsily with unintended negative results, often accompanied by a sharp intake of breath.} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpsaw}{\B{tìp\-saw}}, clumsiness.

% pìtìk
\hi 
\HT{pItIk}{\B{\cp{pì}tìk}} \I{["pI.tIk\textcorner]} \emph{vtr.} scratch non\hyp{}harmfully (\emph{as an itch}). 
\B{Sran, sran, oe\-ri pì\-tìk tsa\-tse\-nget a mì tal! Fkxa\-ke nì\-ftxan kum\-a ter\-kup.} Yeah, yeah, scratch that place on my back! It itches like crazy!~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{43}~\HL{tsupx}{\B{tsupx}})

% pìwopx
\hi
\HT{pIwopx}{\B{pì\cp{wopx}}} \I{[pI."wop']} \emph{n.} cloud.
\B{Tsun fko ay\-on\-ti fì\-wopx\-ä ni\-vìn fte ya\-fkeyk\-it sre\-si\-ve\-'a.} The shapes of clouds can be used to predict the weather.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\emph{a. idiom}) \B{(na) fwam\-pop fkip fì\-wopx}, ``Fish out of water.'' [lit.: (like a) tapirus in the clouds].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lepwopx}{\B{lepwopx}}, cloudy. 
\on{2}.~\HL{pIwopxtsyIp}{\B{pìwopxtsyìp}}, nebula. 
\on{3}.\HL{fwopx}{\B{fwopx}}, dust (in the air).
\on{4}.~\HL{hiupwopx}{\B{hi\-up\-wopx}}, cloud spitter.

% pìwopxtsyìp
\hi
\HT{pIwopxtsyIp}{\B{pì\cp{wopx}tsyìp}} \I{[pI."wop'.\t{ts}jIp\textcorner]} \emph{n.} nebula (\emph{astronomical}).
(\ding{43}~\HL{pIwopx}{\B{pì\-wopx}}, -\HL{tsyIp}{\B{tsyìp}})

% pllhrr
\hi
\HT{pllhrr}{\B{pll\cp{hrr}}} \I{[p$\cdot$\s{l}."h\s{r}]} (\textsc{perf}, \B{pol\cp{hrr}})\emph{vin.}\on{11} warn (\B{teri}, about).
\B{Tseyk Na'\-vi\-ru pol\-hrr te\-ri Saw\-tu\-te.} Jake warned the Na'vi about the Skypeople.~\LINK{http://naviteri.org/2017/01/meliuteri-alu-tung-si-pllhrr-about-tung-and-pllhrr/}{b}
(\ding{43}~\HL{penghrr}{\B{peng\-hrr}}; \HL{plltxe}{\B{pll\-txe}}, \HL{hrrap}{\B{hrr\-ap}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sApllhrr}{\B{sä\-pll\-hrr}}, warning.

% pllngay
\hi
\HT{pllngay}{\B{pll\cp{ngay}}} \I{[p$\cdot$\s{l}."Naj]} (\textsc{perf}, \B{pol\cp{ngay}}) \emph{vin.}\on{11} admit, acknowledge.
(\emph{a. with direct speech}) \B{Po ke tsun pi\-vll\-ngay san oe\-ru tì\-kxey.} He can't admit he's wrong.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Oe pll\-ngay san mo\-lak\-to oe nì\-fe', ta\-fral sno\-laytx; wä\-tu lu oe\-to txur.} I acknowledge that I rode badly, so I lost; my opponent was stronger than I was.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\emph{b. with the topic}) \B{Tsa\-'u\-ri po pll\-ngay.} He admits it.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{plltxe}{\B{pll\-txe}}, \HL{ngay}{\B{ngay}}) 

% plltxe
\hi
\HT{plltxe}{\B{pll\cp{txe}}} \I{[p\s{l}."t'E]} (\textsc{perf}, \B{pol\cp{txe}}) \textsc{i}. \emph{vin.} speak, say. 
\B{Pll\-txe nga nìl\-tsan!} You speak well!~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll}
\B{Ke\-tsran tu\-te ni\-vew hi\-vum, po\-ru pll\-txe san ru\-txe 'i\-vì\-'awn.} No matter who wants to leave, tell them to please stay.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Ma 'ite, aw\-nge\-yä fya\-'o\-ri ze\-ne nga sä\-nume si\-vi po\-ru fte tsi\-vun pil\-vll\-txe sì ti\-vì\-ran nì\-ay\-oeng.} My daughter, you will teach him our way so that (he) can speak and walk as we do.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki}
\B{New Va'\-ru tskot Ey\-tu\-kan\-ä za\-sri\-vìn. Ke\-zem\-pll\-txe pa\-yll\-txe san ke\-he.} Va'ru wants to borrow Eytukan's bow. Of course he'll say no.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Sa'\-nok nì\-nrr\-a lrr\-tok so\-li krr\-a prr\-nen alo a\-'aw\-ve pol\-txe.} The mother smiled with pride when the baby spoke for the first time.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Fko pll\-txe\-i\-e san me\-nga mun\-txa slo\-lu sìk!} I'm so happy to hear you got married!~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
\textsc{ii}. \emph{vtr.}
\B{Pol\-txe pol pay\-lì\-'ut?} What did she say? [lit.: What words did she say?]
\B{Ke pol\-txe pol tsay\-lì\-'ut.} She didn't say that.
\B{Pol\-txe pol fay\-lu\-ta oe new ki\-vä.} She said, `I want to go'. \emph{or} She said she wanted to go.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} 
(\emph{a. with} \B{nì\-nu}) misspeak, speak in vain.
\B{Oe\-ru txoa li\-vu. Pol\-txä\-nge nì\-nu.} Forgive me. I misspoke.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{b. compound}) \B{way a pll\-txe}, spoken poem.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b}
(\emph{c. idiom}) \B{Kxe\-tse sì mik\-yun kop pll\-txe.} Body language is important [lit.: the tail and ears also speak].~\LINK{http://forum.learnnavi.org/language-updates/grid-tidbits/}{ln} 
(\emph{d. idiom}) \B{pll\-txe räp\-tum}, speak vulgarly, use vulgar language, swear.\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{plltxeyu}{\B{pll\-txe\-yu}}, speaker. 
\on{2}.~\HL{sAplltxe}{\B{sä\-pll\-txe}}, statement. 
\on{3}.~\HL{kezemplltxe}{\B{ke\-zem\-pll\-txe}}, of course.
\on{4}.~\HL{pllngay}{\B{pll\-ngay}}, admit, acknowledge.
\on{5}.~\HL{pllhrr}{\B{pll\-hrr}}, warn. 
\on{6}.~\HL{wAte}{\B{wä\-te}}, argue.
\on{7}.~\HL{plltxepam}{\B{pll\-txe\-pam}}, speech sound.

% plltxepam
\hi 
\HT{plltxepam}{\B{pll\cp{txe}pam}} \I{[p\s{l}."t'E.pam]} \emph{n.} speech sound; phone, phoneme. 
\B{Slä ke\-zem\-pll\-txe, tsa\-pll\-txe\-pam alu KX ke fkey\-tok nì\-Fran\-se.} But of course, the phoneme `kx' doesn't exist in French.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/\#comment-31171}{b} 
(\ding{43}~\HL{plltxe}{\B{plltxe}}, \HL{pam}{\B{pam}})

% plltxeyu
\hi
\HT{plltxeyu}{\B{pll\cp{txe}yu}} \I{[p\s{l}."t'E.ju]} \emph{n.} speaker.
\B{Toi\-tsye\-ri lu poe pll\-txe\-yu a\-naw\-fwe nì\-txan.} She's a very fluent speaker of German.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\ding{43}~\HL{plltxe}{\B{pll\-txe}})

% po 
\hi
\HT{po}{\B{po}} \I{[po]} (\emph{gen.} \B{peyä}) \emph{pn.} he\slash she (3\textsuperscript{rd} person singular).
(\emph{a. subjective}) \B{Lam oer fwa po lu ka\-nu nì\-txan.} It seems to me that he is very smart.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Maw\-krr\-a Tsyeyk ftxo\-lu\-lì\-'u, tslam fra\-pol fu\-ta slu po Olo'\-eyk\-tan a\-mip.} After Jake spoke, everyone understood that he had become the new Clan Leader.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Lu po lor\-a tu\-té a\-lor.} She's an extremely beautiful woman.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\emph{b. agentive}) \B{pol} \I{[pol]}
\B{Li pol te\-rok fì\-tseng\-it srak?} Is she already here?~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\emph{c. patientive}) \B{pot(i)} \I{[pot\textcorner \slash "po.ti]}
\B{Apxa tì\-väng\-ìl po\-ti stey\-ko\-li.} (His) great thirst made him angry.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Pol\-txe Ey\-tu\-kan san oe ka\-yä sìk, slä oel pot ke spaw.} Eytukan said he would come but I don't believe him.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175491/\#msg175491}{ln}
(\emph{d. genitive}) \B{\cp{pe}yä} \I{["pE.j\ae]}
\B{Ut\-ral\-ä A\-nawm | Ay\-ri\-na' lu ay\-oeng, | A pe\-yä tì\-txur mì hi\-nam aw\-nge\-yä.} We are all seeds | Of the Great Tree, | Whose strength is in our legs.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} 
\B{Ay\-ron\-srel pe\-yä hä\-ngek nì\-txan.} His imaginings are (unpleasantly) weird.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Pol 'o\-lem pe\-yä wu\-tsot.} He made his (i.e., someone else's) dinner.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
(\ding{43}~\B{sne\-yä})
(\emph{e. dative}) \B{por(u)} \I{[poR \slash "po.Ru]}
\B{Ma Ra\-lu, srung si por!} Ralu, help her!~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Ke su\-nu por fwa rol eo tu\-te a\-pxay.} She doesn't like to sing in front of many people.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
(\emph{f. topic}) \B{\cp{po}ri} \I{["po.Ri]} 
\B{Pori mokri lor lu nìtxan ulte rol po nìftxavang.} Her voice is very beautiful and she sings passionately.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
\B{Po\-ri ze\-ne kll\-fri\-vo' nga.} He is your responsibility [lit.: you are responsible for him].~\LINK{http://forum.learnnavi.org/language-updates/misc-answers/}{ln\slash a\textsuperscript{c}} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pan}{\B{po\-an}}, he. 
\on{2}.~\HL{poe}{\B{po\-e}}, she. 
\on{3}.~\HL{frapo}{\B{fra\-po}}, everyone, everybody. 
\on{4}.~\HL{fIpo}{\B{fì\-po}}, this one (person). 
\on{5}.~\HL{tsapo}{\B{tsa\-po}}, that one (person). 
\on{6}.~\HL{'awpo}{\B{'aw\-po}}, one (person). 
\on{7}.~\HL{lapo}{\B{la\-po}}, another one.
\on{8}.~\HL{poho}{\B{po\-ho}}, he\slash she (deferential form).

% poan
\hi
\HT{poan}{\B{po\cp{an}}} \I{[po."{}an]} \emph{pn.} he (3\textsuperscript{rd} person singular masculine). 
\B{Po\-an pak!? Ke lu po tsam\-si\-yu kaw\-'it!} Him!? There's not a warrior's bone in his whole body!~\LINK{http://forum.learnnavi.org/language-updates/ke-kawit/msg174572}{ln}
\B{Poe mo\-war so\-li poan\-ur tsnì hi\-vum.} She advised him to leave.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{po}{\B{po}})  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pohan}{\B{po\-han}}, he (deferential).

% poe
\hi
\HT{poe}{\B{po\cp{e}}} \I{[po."{}E]} \emph{pn.} she (3\textsuperscript{rd} person singular feminine).
\B{Poe pol\-txe nì\-fya\-'o a\-law} She spoke clearly.~\LINK{http://forum.learnnavi.org/language-updates/why-is-this-night/msg154875/\#msg154875}{ln} 
\B{Po\-el oe\-ti lar\-mawk.} She was talking about me.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Krr a sto\-lawm Txe\-wì te\-ri hem a poe\-ru lo\-len, 'o\-le\-fu ke\-ftxo nì\-txan.} When Txewì heard about what happened to her, he felt very upset.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Poe\-ri unil\-tì\-ran\-tokx\-it tar\-mok a krr lam stum nì\-ay\-fo.} When she is in her Avatar body, she appears almost like them.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{po}{\B{po}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pohe}{\B{po\-he}}, she (deferential).

% pohan
\hi 
\HT{pohan}{\B{po\cp{han}}} \I{[po."han]} \emph{pn.} he (3\textsuperscript{rd} person singular; deferential\slash ceremonial form). 
(\ding{43}~\HL{poho}{\B{po\-ho}}, \HL{poan}{\B{po\-an}}) 

% pohe
\hi 
\HT{pohe}{\B{po\cp{he}}} \I{[po."hE]} \emph{pn.} she (3\textsuperscript{rd} person singular; deferential\slash ceremonial form). 
(\ding{43}~\HL{poho}{\B{po\-ho}}, \HL{poe}{\B{po\-e}}) 

% poho
\hi 
\HT{poho}{\B{\cp{po}ho}} \I{["po.ho]} \emph{pn.} he\slash she (3\textsuperscript{rd} person singular; deferential\slash ceremonial form). 
\B{Sra\-ke lu\-yu po\-ho set ro hel\-ku?} Is he\slash she at home now?~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b}
(\ding{43}~\HL{po}{\B{po}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{pohe}{\B{po\-he}}, she (deferential).
\on{2}.~\HL{pohan}{\B{po\-han}}, he (deferential).

% polpxay
\hi
\HT{polpxay}{\B{pol\cp{pxay}}} \I{[pol."p'aj]} \emph{inter., adj.} how many? (\emph{alternate form:} \ding{43}~\HL{holpxaype}{\B{hol\-pxay\-pe}})
\B{Nga\-ri so\-la\-lew pol\-pxay\-a zì\-sìt?\slash Nga\-ri so\-la\-lew zì\-sìt a\-pol\-pxay?} How old are you?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\ding{43}~\HL{pe}{\B{pe}}\texttt{+}, \HL{holpxay}{\B{hol\-pxay}}) 

% poltxe
\hi
\B{pol\cp{txe}} \I{[pol."t'E]}~:~\HL{plltxe}{\B{plltxe}}.

% pom
\hi
\HT{pom}{\B{pom}} \I{[pom]} \emph{vtr.} kiss. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sApom}{\B{säpom}}, kiss.

% pon
\hi 
\HT{pon}{\B{pon}} \I{[pon]} \emph{vtr.} balance. 
\B{Fwa pon sey\-ti sìn ki\-nam\-til lu le\-hrr\-ap, ma 'itan. Tsun ne\-kll zi\-vup tsawng.} Balancing a cup on your knee is dangerous, son. It can fall to the ground and break.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
\B{Nì\-sngä\-'i Tsyeyk lu pì\-saw ul\-te ke tsun vul\-sìn pä\-pi\-von.} At first Jake was clumsy and wasn't able to balance on a branch.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b} 

% pongu
\hi
\HT{pongu}{\B{\cp{po}ngu}} \I{["po.Nu]} \emph{n.} group or party of people. 
\B{Vll eyk\-yu ne\-fä fte po\-ngu fä\-ki\-vä.} The leader signals the party to ascend by pointing upwards.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Ay\-ha\-pxì\-tu po\-ngu\-ä txo\-pu si nì\-nän ta\-krr\-a Va'\-rul pxe\-ku\-tut lä\-txayn.} The members of the group are less afraid since Va'ru defeated three of the enemy.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{ngrrpongu}{\B{ngrr\-po\-ngu}}, grassroots movement.   
\on{2}.~\HL{tsampongu}{\B{tsam\-po\-ngu}}, war party. 
\on{3}.~\HL{wempongu}{\B{wem\-po\-ngu}}, squad, battle party. 
\on{4}.~\HL{tarpongu}{\B{tar\-po\-ngu}}, hunting party.

% postì
\hi
\HT{postI}{\B{\cp{po}stì}} \I{["po.stI]} \emph{n.}, \textsc{lw} (forum\slash blog) post.
\B{'Aw\-ve\-a po\-stì zì\-sìt\-ä a\-mip.} First post of the new year.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Mì fo\-stì a kxey\-ey\-ti oel li zo\-sley\-ko\-lu.} I already fixed the mistake in the post.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1910}{b}

% prrku
\hi
\HT{prrku}{\B{\cp{prr}ku}} \I{["p\s{r}.ku]} \emph{n.} womb.
(\ding{43}~\HL{prrnen}{\B{prr\-nen}}, \HL{kelku}{\B{kel\-ku}})

% prrnen
\hi 
\HT{prrnen}{\B{\cp{prr}nen}} \I{["p\s{r}.nEn]} \emph{n.} infant, baby.
\B{Sa'\-nok nì\-nrr\-a lrr\-tok so\-li krr\-a prr\-nen a\-lo a\-'aw\-ve pol\-txe.} The mother smiled with pride when the baby spoke for the first time.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Mong prr\-nen\-ìl fu\-ta sa'\-nul ve\-rewng nì\-tut.} The infant depends on her Mommy to look after her constantly.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Ze\-ne fko 'i\-vam\-pi prr\-nen\-it nì\-flrr fra\-krr.} One must always touch a baby gently.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Oel vewng frr\-nen\-it.} I look after the infants.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Ta'\-leng prr\-nen\-ä lu faoi, pum ko\-ak\-tu\-ä ekx\-txu.} A baby's skin is smooth, an old person's is rough.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Sa'\-nok prr\-nen\-ur yom\-tìng fa okup sne\-yä.} A mother feeds an infant with her milk.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{prrku}{\B{prrku}}, womb. 
\on{2}.~\HL{prrnesyul}{\B{prrnesyul}}, bud (of a flower). 
\on{3}.~\HL{prrwll}{\B{prrwll}}, moss. 
\on{4}.~\HL{prrsmung}{\B{prrsmung}}, baby carrier.
\on{5}.~\HL{nga'prrnen}{\B{nga'prrnen}}, be pregnant (for people).

% prrnesyul
\hi
\HT{prrnesyul}{\B{\cp{prr}nesyul}} \I{["p\s{r}.nE.sjul]} \emph{or} \I{[.sjUl]} \emph{n.} bud (of a flower). 
\B{Ay\-ut\-ral\-ur lu frr\-ne\-syul, Ey\-wa tì\-haw\-nu si\-vi!} The trees have buds, may Eywa protect (them)!~\LINK{http://forum.learnnavi.org/news-announcements/christmas-song-ninavi-on-youtube/}{ln}
(\ding{43}~\HL{prrnen}{\B{prr\-nen}}, \HL{syulang}{\B{syu\-lang}})

% prrsmung
\hi
\HT{prrsmung}{\B{\cp{prr}smung}} \I{["p\s{r}.smuN]} \emph{or} \I{[.smUN]} (\textsc{rn}: \B{\cp{prr}smùng} \I{["p\s{r}.smUN]}) \emph{n.} baby carrier. 
(\ding{43}~\HL{prrnen}{\B{prr\-nen}}, \HL{sAmunge}{\B{sä\-mu\-nge}})

% prrte'
\hi
\HT{prrte'}{\B{\cp{prr}te'}} \I{["p\s{r}.tEP]} \emph{adj.} pleasurable (of an activity).   
\B{Tsun oe nga\-hu nì\-Na'\-vi pi\-väng\-kxo a fì\-'u oe\-ru prr\-te' lu.} It's a pleasure to be able to chat with you in Na'vi.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Kem\-ìri a nga\-ru prr\-te' ke lu, tsa\-kem rä\-'ä si\-vi ay\-la\-he\-ru.} As for any action that you don't like, don't act that way to others.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
\B{Fì\-trr\-ä ftxo\-zä\-ri, sìl\-pey oe, ay\-nga\-ru prr\-te' li\-vu.} I hope you enjoy today’s celebration.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIprrte'}{\B{nìprrte'}}, with pleasure. 
\on{2}.~\HL{tIprrte'}{\B{tìprrte'}}, pleasure.

% prrwll
\hi
\HT{prrwll}{\B{\cp{prr}wll}} \I{["p\s{r}.w\s{l}]} \emph{n.} moss. 
\B{Oe\-ru ke tsun li\-vam ke\-'u lor to Ey\-wa\-'eveng\-ä na'\-rìng a lew sä\-po\-li fa prr\-wll.} Nothing could appear more beautiful to me than a Pandoran forest that has covered in moss.~\LINK{http://forum.learnnavi.org/language-updates/prrwll-moss-the-comparative-to-construct/}{ln}
(\ding{43}~\HL{prrnen}{\B{prr\-nen}}, \HL{'ewll}{\B{'e\-wll}})

% pu-
\hi
\HT{pu}{\B{pu}}- \I{[pu.]} \emph{num. pref.} six (\emph{base form of \ding{43}~\HL{pukap}{\B{pu\-kap}} to form higher numbers or fractions}). 
\B{\cp{pu}vol}, \on{48} (decimal), \on{60} (octal); 
\B{pu\cp{pxì}}, one-sixth, $\frac{1}{6}$; 
\B{pu\-pxì a\-tsìng}, four\hyp{}sixth, $\frac{4}{6}$.

% puk
\hi
\HT{puk}{\B{puk}} \I{[puk\textcorner]} \emph{or} \I{[pUk\textcorner]} (\textsc{rn}: \B{pùk} \I{[pUk\textcorner]}) \emph{n.}, \textsc{lw} book. 
\B{Lu oe\-ru puk.} I have a book.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Puk\-it a\-ke\-tsran ivi\-nan.} Read any book at all.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Tsun Txil\-te pam\-rel\-it ivi\-nan; ta\-fral puk\-ot a\-nutx mu\-nge fra\-tseng.} Txilte knows how to read; therefore she brings a thick book wherever she goes.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Kì\-rey\-sì\-ri lu tsa\-puk\-ä ay\-vur a te\-ri 'Rr\-ta txan\-wa\-we nì\-ngay, slä oe\-ri, hu\-fwa el\-tur tì\-txen si, lu ral\-nga' nì\-'aw.} For Grace, the stories in that book relating to Earth are personally meaningful, but for me, although interesting, they’re simply instructive.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{puktsyIp}{\B{puk\-tsyìp}}, booklet.   
\on{2}.~\HL{lI'upuk}{\B{lì\-'u\-puk}}, dictionary.

% puktsyìp
\hi
\HT{puktsyIp}{\B{\cp{puk}tsyìp}} \I{["puk\textcorner.\t{ts}jIp\textcorner]} \emph{or} \I{["pUk\textcorner.]} (\textsc{rn}: \B{\cp{pùk}tsyìp} \I{["pUk\textcorner.\t{tS}Ip\textcorner]}) \emph{n.} booklet, pamphlet. 
(\ding{43}~\HL{puk}{\B{puk}}, \HL{tsyIp}{\B{-tsyìp}})

% pukap
\hi
\HT{pukap}{\B{\cp{pu}kap}} \I{["pu.kap\textcorner]} \emph{num.} six. (\emph{base}: \ding{43}~\HL{pu}{\B{pu-}}, \emph{rest}: \ding{43}~\HL{fu2}{\B{\mbox{-fu}}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vospxIpuk}{\B{Vo\-spxì\-puk}}, June.

% pum
\hi
\HT{pum}{\B{pum}} \I{[pum]} \emph{or} \I{[pUm]} (\textsc{rn:} \B{pùm} \I{[pUm]}) 
\emph{n.} (``dummy noun,'' \emph{not marked for number}) 
(\emph{a.}) ``one'' (\emph{always used with a narrowing descriptor})
\B{Lam set fwa Saw\-tu\-te a\-kawng ho\-lum, pum a\-sìl\-tsan 'ì\-'awn.} It now seems that the bad Skypeople have left, the good ones remain.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Ftu\-e lu fwa ta\-ron ngong\-a io\-ang\-it to fwa ta\-ron pum\-it a lu wa\-lak sì win.} It's easier to hunt lethargic animals than to hunt perky, speedy ones.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Lu nga\-ru swi\-zaw nì\-hawng srak? Tìng oer pum\-it a\-mrr!} Do you have too many arrows? Give me five.~\LINK{https://naviteri.org/2025/01/tsaliu-alu-pum-aysaomum-asawnung-additional-information-about-pum/}{b}
\B{Tìng oer pum\-it a\-'aw!} Give me one.~\LINK{https://naviteri.org/2025/01/tsaliu-alu-pum-aysaomum-asawnung-additional-information-about-pum/}{b}
(\emph{a.1 optionally with} \B{fota} or \B{sawta}) ``of them''.
\B{Lu nga\-ru 'e\-veng a\-pu\-kap, slä smon oer fo\-ta pum a\-mrr nì\-'aw.} You have six children, but I only know five of them.~\LINK{https://naviteri.org/2025/01/tsaliu-alu-pum-aysaomum-asawnung-additional-information-about-pum/}{b}
(\emph{b. to convey emphasis})
\B{Tsa\-tu\-té lu lor\-a pum a\-lor.} That woman is an extremely beautiful one.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\emph{c. with possessive pronouns to form ``disjunctive possessives''}) ``possession, thing possessed.''
\B{Kel\-ku nge\-yä lu tsawl; pum oe\-yä lu hì\-'i.} Your house is large; mine is small.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to? (\emph{overly formal, stilted Na'vi})~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\emph{d. idiom}) \B{Pum nge\-yä.} (\emph{in response to thanks}) ``You're welcome,'' i.e. The thanks belong to you.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\emph{e. can be ommitted in coll. conversation})
\B{Fì\-tsko lu (pum) oe\-yä.} This bow is mine.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b}

% pung
\hi 
\HT{pung}{\B{pung}} \I{[puN]} \emph{or} \I{[pUN]} \emph{vtr.} hurt, injure. 
\B{Ngal pe\-rung oet fì\-fya pe\-lun?} Why are you hurting me like this?~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
\B{Te\-ya si oer fwa ngal paw\-nu\-nga ay\-ioang\-it zong.} It moves me that you save injured animals.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{tIsraw}{\B{tì\-sraw si}})

% pup
\hi
\HT{pup}{\B{pup}} \I{[pup\textcorner]} \emph{or} \I{[pUp\textcorner]} (\textsc{rn}: \B{pùp} \I{[pUp\textcorner]}) \emph{adj.} short (\emph{physical length}). 
\B{Vo\-spxì A\-yol, Po\-stì A\-pup.} Short Post for a Short Month.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 
\B{Lu srey a\-pup ta Nga\-ri zì\-sìt a\-mip li\-vu le\-fpom.} It's a short version of \emph{Ngari zìsìt amip livu lefpom}.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/\#comment-27964}{b}
(\ding{214}~\HL{ngim}{\B{ngim}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{ngimpup}{\B{ngimpup}}, length.

% pupxì
\hi
\HT{pupxI}{\B{pu\cp{pxì}}} \I{[pu."p'I]} \emph{num.} one\hyp{}sixth, $\frac{1}{6}$.

% puve
\hi
\HT{puve}{\B{\cp{pu}ve}} \I{["pu.vE]} \emph{adj., ord. num.} sixth. \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{trrpuve}{\B{trr\-pu\-ve}}, Friday.

% puvol
\hi
\HT{puvol}{\B{\cp{pu}vol}} \I{["pu.vol]} \emph{num.} \on{48} (decimal), \on{60} (octal).
(\ding{43}~\HL{pu}{\B{pu-}}, \HL{vol}{\B{vol}}) 

% puwup
\hi 
\HT{puwup}{\B{\cp{pu}wup}} \I{["pu.wup\textcorner]} \emph{or} \I{[.wUp\textcorner]} (\textsc{rn}: \B{\cp{pu}wùp} \I{["pu.wUp\textcorner]}) \emph{vin.} bounce. 
\B{Rum 'aw\-lo po\-lu\-wup 'rr\-ko ne\-to.} The ball bounced once and rolled away.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
(\emph{a. idiom}) \B{Ngä\-zìk fwa fkol rum\-it a\-ku\-'up pey\-ku\-wup.} \emph{It's difficult to make s.o. stubborn or inept do what you want them to} [lit.: it's hard to bounce a heavy ball].~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
\\
\end{multicols}



