\pdfbookmark[0]{PX}{PX}
\begin{center}
\section*{PX \I{[p']} (\emph{PxePx})}
\end{center}

\begin{multicols}{2}

% pxan
\hi
\HT{pxan}{\B{pxan}} \I{[p'an]} \emph{adj.} worthy, worthy of\slash worth for (\emph{with the topic}).   
\B{Pxan li\-vu txo nì\-'aw oe nga\-ri.} Only if I am worthy of you.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Sìl\-pey oe, aw\-nge\-yä lì'\-fya\-olo'\-ìri fì\-pì\-lok lì\-ye\-vu pxan.} I hope, this blog will be worth for our language community.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Mll\-te oe: am\-'a\-lu\-ke za\-ma\-yu\-nge tsa\-tì\-pu\-sey\-ìl hum\-it a\-pxan.} I agree: without a doubt that hope will bring worthy results.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/comment-page-1/\#comment-2478}{b}
(\emph{a. idiom}) \B{Ke pxan.} ``Don't mention it.\slash Thank you.'' (\emph{in response to a compliment}) [lit.: (I'm) not worthy].~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}

% pxasìk
\hi
\HT{pxasIk}{\B{\cp{pxa}sìk}} \I{["p'a.sIk\textcorner]} \emph{idiom, vulg.} Screw that! No way!~\LINK{http://www.lemondrop.com/2010/01/26/avatar-pick-up-lines-talk-dirty-in-Navi/}{ld}

% pxasul
\hi
\HT{pxasul}{\B{\cp{pxa}sul}} \I{["p'a.sul]} \emph{or} \I{[.sUl]} (\textsc{rn}: \B{\cp{pxa}sùl} \I{["ba.sUl]}) \emph{adj.} fresh, appealing as food.

% pxazang
\hi 
\HT{pxazang}{\B{\cp{pxa}zang}} \I{["p'a.zaN]} \emph{n., fauna} A\-ku\-la (\emph{shark\hyp{}like aquatic creature}).

% pxaw
\hi
\HT{pxaw}{\B{pxaw}} \I{[p'aw]} \emph{adp.} around. 
\B{Nì\-trr\-trr yom Na'\-vil wu\-tsot 'aw\-si\-teng pxaw yll\-txep.} The Na'vi regularly eat dinner together around a communal fire.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{pxawpa}{\B{pxaw\-pa}}, perimeter, border. 
\on{2}.~\HL{pxawpun}{\B{pxaw\-pxun}}, armband. 
\on{3}.~\HL{pxawngip}{\B{pxaw\-ngip}}, environment. 
\on{4}.~\HL{pxawtxap}{\B{pxaw\-txap}}, squeeze.
\on{5}.~\HL{pxawtok}{\B{pxaw\-tok}}, surround.

% pxawngip
\hi
\HT{pxawngip}{\B{\cp{pxaw}ngip}} \I{["p'aw.Nip\textcorner]} \emph{n.} environment. 
\B{Fwa 'e\-fu ma\-wey sì nit\-ram mì paw\-ngip a\-mip krr\-nekx.} It takes time (for them) to feel calm and happy in a new environment.~\LINK{http://naviteri.org/2015/09/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise-2/}{b}
\B{Na fra\-ve\-'o ay\-ru\-sey\-ä, far\-wun\-tu Ey\-wa\-'ev\-eng\-ä sä\-fpìl\-yewn nì\-'aw\-nì\-'aw fa ay\-fya\-'o a ta\-re fey pxaw\-ngip\-it.} As with any ecosystem, the inhabitants of Pandora communicate in ways unique to their environment.~\LINK{https://naviteri.org/2026/06/tskxekeng-a-mikyunfpi-pamrel-listening-exercise-text/}{b}
(\ding{43}~\HL{pxaw}{\B{pxaw}}, \HL{ngip}{\B{ngip}})

% pxawpa
\hi
\HT{pxawpa}{\B{\cp{pxaw}pa}} \I{["p'aw.pa]} \emph{n.} perimeter, circumference, border. 
\B{Tu\-tan a'\-ewan tar\-mìng na\-ri pxaw\-pa\-ro txan\-lo\-kxe\-yä.} A young man was watching at the border of his country.~\LINK{http://naviteri.org/2020/04/awa-tskxekengtsyip-a-mikyunfpi-niul-one-more-little-listening-exercise/}{b}
(\ding{43}~\HL{pxaw}{\B{pxaw}}, \HL{pa'o}{\B{pa'o}})

% pxawpxun
\hi
\HT{pxawpxun}{\B{\cp{pxaw}pxun}} \I{["p'aw.p'un]} \emph{or} \I{[.p'Un}] \emph{n.} armband.   
\B{Po\-ri pxun\-pxaw lu pxaw\-pxun.} Around his arm is an armband.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Ngal mo\-lay' pxaw\-pxun\-it Lo\-ak\-ä a krr pe\-kxin\-um?} When you tried on Loak's armband, how tight was it?~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{pxaw}{\B{pxaw}}, \HL{pxun}{\B{pxun}})

% pxawtok
\hi 
\HT{pxawtok}{\B{\cp{pxaw}tok}} \I{["p'aw.t$\cdot$ok]} \emph{vtr., inf.}\on{22} surround. 
\B{Pxaw\-to\-lok sna\-nan\-tang\-ìl ye\-rik\-it.} The nan\-tang pack surrounded the ye\-rik.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}  
(\ding{43}~\HL{pxaw}{\B{pxaw}}, \HL{tok}{\B{tok}})

% pxawtxap
\hi 
\HT{pxawtxap}{\B{pxaw\cp{txap}}} \I{[p'aw."t'$\cdot$ap\textcorner]} \emph{vtr., inf.}\on{22} squeeze. 
\B{Po\-el oey tsyokx\-it pxaw\-txo\-lap a krr, pol\-txe nì\-fnu san nga yaw\-ne lu oer.} When she squeezed my hand, she silently said she loved me.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}  
(\ding{43}~\HL{pxaw}{\B{pxaw}}, \HL{txap}{\B{txap}})

% pxay
\hi
\HT{pxay}{\B{pxay}} \I{[p'aj]} \emph{adj.} many (\emph{for countable nouns}). 
(\emph{a. mainly used with the singular form of the noun}) 
\B{Na'\-rìng\-ìl nga' pxay\-a io\-ang\-it.} The forest contains many animals.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{Fì\-slär\-mì tsun fko sti\-vawm ngam\-it a\-pxay.} You can hear a lot of echoes in this cave.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{b. coll. with the plural form}) \B{Fu\-ri\-a fya\-pe fkol se\-rar lì'\-fya\-ti aw\-nge\-yä, le\-iu set nì\-law pxay\-a su\-te a oe\-to lu sìl\-tsan!} Concerning the use of our language, clearly many people are now better than me, I'm happy to say.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/comment-page-1/\#comment-191}{b}
(\ding{214}~\HL{hol}{\B{hol}}, \ding{43}~\HL{'a'aw}{\B{'a'aw}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIpxay}{\B{nì\-pxay}}, many, in great number. 
\on{2}.~\HL{holpxay}{\B{hol\-pxay}}, number. 
\on{3}.~\HL{pxayopin}{\B{pxay\-opin}}, colorful, multi\hyp{}colored. 
\on{4}.~\HL{pxayzekwA}{\B{pxay\-zek\-wä}}, spiny whips.

% pxayopin
\hi
\HT{pxayopin}{\B{\cp{pxay}opin}} \I{["p'aj.o.pin]} \emph{adj.} colorful, multi\hyp{}colored, variegated. 
\B{'En si oe, lor\-a tsa\-fkxi\-le\-ri a\-pxay\-opin so\-lar Tse\-nul srok\-it a\-vo\-zam.} I would guess that Tsenu used a thousand (lit. \on{512}) beads for that beautiful multi\hyp{}colored bib necklace.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
(\ding{43}~\HL{pxay}{\B{pxay}}, \HL{'opin}{\B{'opin}})

% pxayzekwä
\hi
\HT{pxayzekwA}{\B{pxay\cp{zek}wä}} \I{[p'aj."zEk\textcorner.w\ae]} \emph{n.}, \ding{96} spiny whips, \emph{phytotinium poydactylum}. 
(\ding{43}~\HL{pxay}{\B{pxay}}, \HL{zekwA}{\B{zek\-wä}})

% pxe+
\hi
\HT{pxe}{\B{pxe}\texttt{+}} \I{[p'E.]} \emph{trial pref.} three (of something).

% pxefo
\hi
\HT{pxefo}{\B{pxe\cp{fo}}} \I{[p'E."fo]} (\emph{gen.} \B{pxe\-\cp{fe}\-yä}) \emph{pn.} they (three), the three of them, (3\textsuperscript{rd} person trial).
(\emph{a. subjective}) \B{Sra\-ke pxe\-fo li po\-lä\-ha\-tsem?} I wonder if the three of them have already arrived.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{b. genitive}) \B{Pxe\-fe\-yä tì\-tstun\-wi\-ri aw\-nga ira\-yo si\-vi ko!} Let's thank the three of them for their kindness.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\emph{c. dative}) \B{Pxe\-fo\-ru oe srung so\-li, kuma oe\-ru set pxe\-fo srung se\-ri.} I helped the three of them, so they're now helping me.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{po}{\B{po}})

% pxek
\hi
\HT{pxek}{\B{pxek}} \I{[p'Ek\textcorner]} \emph{vtr.} kick, shove (\emph{can be used with} \B{kinamfa} \emph{and} \B{pxunfa} \emph{to clear meaning}). 
\B{Ye\-rik\-ìl nan\-tang\-it pxo\-lek fte hi\-vi\-fwo.} The hexapede kicked the viperwolf in order to flee.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
\B{Po pxun\-fa pxek tsa\-krr me\-fo zup ne\-kll!} He shoved the two of them and they fell down.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}

% pxel
\hi
\HT{pxel}{\B{pxel}} \I{[p'El]} \emph{adp.} like, as. 
\B{Kem\-lì\-'ut alu var fkol sar pxel pum alu new.} One uses the verb \emph{var} like \emph{new}.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-582}{b}
(\emph{a. required with} \HL{zet}{\B{zet}}) \B{Pey\-ral\-ìl zet wur\-a wu\-tsot a\-'aw\-nem pxel sngel.} Peyral treats a cool cooked meal like garbage.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\emph{b. with} \HL{nIftxan}{\B{nìftxan}} \emph{and adj. or adv. to mean}) as \emph{adj.\slash adv.} as.
\B{Oe tul nì\-ftxan nì\-win pxel nga.} I run as fast as you.~\LINK{https://forum.learnnavi.org/language-updates/eapressions-of-\%27like-we\%27/}{ln}
(\ding{43}~\HL{na}{\B{na}})

% pxelo
\hi
\HT{pxelo}{\B{\cp{pxe}lo}} \I{["p'E.lo]} \emph{adv.} thrice, three times. 
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{alo}{\B{alo}})

% pxen
\hi
\HT{pxen}{\B{pxen}}\textsuperscript{\on{1}} \I{[p'En]} \emph{n.} (item of) functional clothing.
\B{Pen\-it yem\-stokx!} Get dressed!~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Fu\-ta nga\-ta tel pxen\-yoa srät, mll\-'e\-i\-an oel.} I'm happy to take cloth from you for finished garments.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
(\ding{43}~\HL{ioi}{\B{ioi}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lukpen}{\B{luk\-pen}}, naked.

% pxen
\hi
\B{pxen}\textsuperscript{\on{2}} : \HL{'en}{\B{'en}}

% pxenga
\hi
\HT{pxenga}{\B{pxe\cp{nga}}} \I{[p'E."Na]} \emph{pn.} you (three), the three of you (2\textsuperscript{nd} person trial).
(\emph{a. genitive}) \B{Pxe\-nge\-yä tsul\-fä lì'\-fya\-yä le\-Na'\-vi oe\-ru te\-ya si.} Their mastery of the Na'vi language fills me with joy.~\LINK{http://naviteri.org/2010/12/navi-writing-contest-the-first-place-winners/}{b}
(\emph{b. dative}) \B{Ira\-yo nì\-txan pxe\-nga\-ru!} Thank you three very much!~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{nga}{\B{nga}})

% pxengenga
\hi
\HT{pxengenga}{\B{pxenge\cp{nga}}} \I{[p'E.NE."Na]} \emph{pn.} you (three), the three of you (2\textsuperscript{nd} person trial; deferential or ceremonial form).
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{ngenga}{\B{nge\-nga}})

% pxesrram
\hi
\HT{pxesrram}{\B{pxesrr\cp{am}}} \I{[p'E.s\s{r}."{}am]} \emph{n., adv.} three days ago. 
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{trram}{\B{trr\-am}})

% pxesrray
\hi
\HT{pxesrray}{\B{pxesrr\cp{ay}}} \I{[p'E.s\s{r}."{}aj]} \emph{n., adv.} three days from now.
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{trray}{\B{trr\-ay}})

% pxevol
\hi
\HT{pxevol}{\B{\cp{pxe}vol}} \I{["p'E.vol]} \emph{num., adj.} \on{30} (octal), \on{24} (decimal).
\B{Oe\-ri so\-la\-lew zì\-sìt a\-pxe\-vol.} I'm 24 years old.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{vol}{\B{vol}})

% pxevotrrpxì
\hi 
\HT{pxevotrrpxI}{\B{pxevotrr\cp{pxì}}} \I{[p'E.vo.t\s{r}."p'I]} \emph{n.} hour (in the \on{24}\hyp{}human cycle, \emph{often shortened to} \ding{43}~\HL{trrpxI}{\B{trrpxì}}).  
(\ding{43}~\HL{pxevol}{\B{pxevol}}, \HL{trrpxI}{\B{trrpxì}})

% pxey
\hi
\HT{pxey}{\B{pxey}} \I{[p'Ej]} \emph{num., adj.} three (\emph{base:} \B{pxey-}, \emph{rest:} \HL{-pey}{\B{-pey}}). 
\B{Za\-ya\-'u Saw\-tu\-te fte aw\-nga\-ti ski\-va\-'a kay zì\-sìt a\-pxey (fu: pxey\-a zì\-sìt\-kay)!} The Sky People will come to destroy us three years from now!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vospxey}{\B{Vo\-spxey}}, March.

% pxeyve
\hi
\HT{pxeyve}{\B{\cp{pxey}ve}} \I{["p'Ej.vE]} \emph{num., adj.} third.
\B{ta\-ron\-yu a\-pxey\-ve\slash pxey\-ve\-a ta\-ron\-yu}, hunter number 3, the third hunter.~\LINK{http://forum.learnnavi.org/language-updates/update-on-counting/}{ln}

% pxi
\hi
\HT{pxi}{\B{pxi}} \I{[p'i]} \emph{adj.} sharp (\emph{as a point}). 
\B{Na\-ri si! Tsa\-ioang\-ur lu pxi\-a me\-slukx!} Careful! That animal has two sharp horns!~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{214}~\HL{fwem}{\B{fwem}}, \ding{43}~\HL{litx}{\B{litx}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIpxi}{\B{nì\-pxi}}, especially, pointedly. 
\on{2}.~\HL{pximaw}{\B{pxi\-maw}}, right after. 
\on{3}.~\HL{pxiset}{\B{pxi\-set}}, right now. 
\on{4}.~\HL{pxisre}{\B{pxi\-sre}}, right before. 
\on{5}.~\HL{pxiswawam}{\B{pxi\-swaw\-am}}, just a moment ago. 
\on{6}.~\HL{pxiswaway}{\B{pxi\-swaw\-ay}}, in just a moment from now. 
\on{7}.~\HL{pxiut}{\B{pxi\-ut}}, razor palm. 
\on{8}.~\HL{pxiwll}{\B{pxi\-wll}}, hermit bud plant. 
\on{9}.~\HL{kllpxiwll}{\B{kll\-pxi\-wll}}, lionberry. 
\on{10}.~\HL{pxiye'rIn}{\B{pxi\-ye'\-rìn}}, immediately. 

% pxim
\hi
\HT{pxim}{\B{pxim}} \I{[p'im]} \emph{adj.} erect, upright. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIpxim}{\B{nì\-pxim}}, errectly, rightly. 

% pximaw
\hi
\HT{pximaw}{\B{pxi\cp{maw}}} \I{[p'i."maw]} \emph{adp.} right after.
\B{Pxi\-maw Tsawl\-ul\-txa}, Just After the Meet\hyp{}up.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{maw}{\B{maw}})

% pxir
\hi
\HT{pxir}{\B{pxir}} \I{[p'iR]} \emph{n.}, \textsc{lw} beer. 
\B{Su\-nu nga\-ru pxir srak?} Do you like beer?~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Pxir\-it fu swo\-at näk rä\-'ä; ni\-väk pay\-it nì\-'aw.} Don't drink beer or spirits; just drink water.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}

% pxiset
\hi
\HT{pxiset}{\B{pxi\cp{set}}} \I{[p'i."sEt\textcorner]} \emph{adv.} right now.
\B{Pxi\-set nge\-yä tska\-lep\-it kxel\-tek!} Pick up your crossbow right now!~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Zun oe pxi\-set tir\-va\-ron, zel oe 'i\-ve\-fu nit\-ram.} If I were hunting right now, then I would be happy.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{set}{\B{set}})

% pxisre
\hi
\HT{pxisre}{\B{pxi\cp{sre}}} \I{[p'i."{}sRE]} \emph{adp.}\texttt{+} right before (\emph{temporal}).
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{sre}{\B{sre}}) 

% pxiswawam
\hi
\HT{pxiswawam}{\B{pxiswaw\cp{am}}} \I{[p'i.swaw."{}am]} \emph{adv.} just a moment ago.
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{swaw}{\B{swaw}})

% pxiswaway
\hi
\HT{pxiswaway}{\B{pxiswaw\cp{ay}}} \I{[p'i.swaw."{}aj]} \emph{adv.} in just a second from now.
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{swaw}{\B{swaw}}; \HL{pxiye'rIn}{\B{pxi\-ye'\-rìn}})

% pxiut
\hi
\HT{pxiut}{\B{\cp{pxi}ut}} \I{["p'i.ut\textcorner]} \emph{or} \I{[.Ut\textcorner]} \emph{n.}, \ding{96} razor palm, ``sharp tree'', \emph{saltcellar gracilis}. 
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{utral}{\B{utral}})

% pxiwll
\hi
\HT{pxiwll}{\B{\cp{pxi}wll}} \I{["p'i.w\s{l}]} \emph{n.}, \ding{96} hermit bud plant, \emph{cynaroidia discolor}. 
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{'ewll}{\B{'ewll}})

% pxiye'rìn
\hi
\HT{pxiye'rIn}{\B{pxi\cp{ye'}rìn}} \I{[p'i."jEP.RIn]} \emph{adv.} immediately (\emph{but not as soon as} \B{pxi\-swaw\-ay}).
\B{Fì\-fne\-tì\-fkey\-tok\-ìl fngo' fu\-ta kem si\-vi fko pxi\-ye'\-rìn.} This kind of situation requires immediate action.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{pxi}{\B{pxi}}, \HL{ye'rIn}{\B{ye'rìn}}; \HL{pxiswaway}{\B{pxi\-swaw\-ay}})

% -pxì
\hi
\HT{pxI}{\B{-pxì}} \I{[.p'I]} \emph{num. suf. to form fractions with the base form of numbers}. 
\B{ki\-pxì}, one\hyp{}seventh, $\frac{1}{7}$. 
(\ding{43}~\HL{hapxI}{\B{ha\-pxì}})

% pxìm
\hi
\HT{pxIm}{\B{pxìm}} \I{[p'Im]} \emph{adv.} often.   
\B{Nga za\-'u fì\-tseng pxìm srak?} Do you come here often?~\LINK{http://www.lemondrop.com/2010/01/26/avatar-pick-up-lines-talk-dirty-in-Navi/}{ld}
\B{Oe pxìm päng\-kxo nga\-hu nì\-ron\-srel.} I often talk with you in my imagination.\slash I often imagine I'm talking with you.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Sra\-ke tsun oe fay\-upxa\-ret tsli\-vam nì\-ftue? Tse … ze\-ne pi\-vll\-txe san pxìm tsa\-fya lu sìk.} Can I understand these messages easily? Well…  I must say ``Often that way''.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lepxìmrun}{\B{le\-pxìm\-run}}, common, often found.

% pxìmun'i
\hi
\HT{pxImun'i}{\B{pxìmun\cp{'i}}} \I{[p'I.m$\cdot$un."P$\cdot$i]} \emph{or} \I{[.mUn.]} (\textsc{rn}: \B{pxìmùn\cp{'i}} \I{[bI.m$\cdot$Un."P$\cdot$i]}) \emph{vtr.}\on{23} divide, cut into parts.
\B{Nì\-trr\-trr pxì\-mun\-'i sam\-si\-yul ay\-swi\-zaw\-it ku\-tu\-ä a\-law\-nä\-txayn sno\-kip nì\-'eng.} Warriors typically share the arrows of their defeated enemies among themselves.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Fol tsngan\-it pxì\-mo\-lun\-'i nì\-'eng.} They shared the meat.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Fol tsngan\-it pxì\-mo\-lun\-'i nì\-yll.} They shared the meat with the entire clan.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
(\ding{43}~\HL{hapxI}{\B{hapxì}}, \HL{mun'i}{\B{mun'i}})

% pxoe
\hi
\HT{pxoe}{\B{\cp{pxo}e}} \I{["p'o.E]} \emph{pn.} we (three), the three of us (not you) (1\textsuperscript{st} person trial, exclusive).
(\emph{a. subjective}) \B{Fwa tso\-lun nga zi\-va'u moe\-yä kel\-ku\-ne fte tsi\-vun pxoe 'aw\-si\-teng ki\-vä\-teng nì\-mun oe\-ru te\-ya so\-lei\-yi nì\-ngay.} That you could come to our house, so that we three could spend time together again, filled me truly with joy.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/comment-page-1/\#comment-7689}{b}
(\emph{b. topic}) \B{Pxoe\-ri tsa\-vur el\-tur tì\-txen so\-li.} The three of us were interested in that story.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{oe}{\B{oe}})

% pxoeng
\hi
\HT{pxoeng}{\B{pxo\cp{eng}}} \I{[p'o."{}EN]} \emph{pn.} we (three), the three of us (1\textsuperscript{st} person trial, inclusive).
(\emph{a. subjective}) \B{Moe sìl\-pey nì\-teng tsnì pxo\-eng tsä\-pì\-ye\-ve\-'a fì\-tsap ye'\-rìn nì\-mun.} We both also hope that the three of us will see each other again soon.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/comment-page-1/\#comment-2483}{b}
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{oeng}{\B{oeng}})

% pxohe
\hi
\HT{}{\B{\cp{pxo}he}} \I{["p'o.hE]} \emph{pn.} we (three), the three of us (not you) (1\textsuperscript{st} person trial, exclusive; deferential or ceremonial form)
% (\emph{a. subjective})
% (\emph{b. agentive})
% (\emph{c. patientive})
% (\emph{d. genitive})
% (\emph{e. dative})
% (\emph{f. topic})
(\ding{43}~\HL{pxe}{\B{pxe}}\texttt{+}, \HL{ohe}{\B{ohe}})

% pxor
\hi
\HT{pxor}{\B{pxor}} \I{[p'oR]} \emph{vin.} explode. 
\B{Txep\-ram pxor a krr, txan\-a txe\-kxum\-pay wrr\-za\-'u.} When a volcano erupts, a lot of lava comes out.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
\B{Hì\-krr\-o me\-fo kak\-pam lar\-mu maw\-krr\-a pxo\-lor kun\-sìp.} The two of them were deaf for a short time after the gunship exploded.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pxorpam}{\B{pxor\-pam}}, ejective. 
\on{2}.~\HL{pxorna'}{\B{pxor\-na'}}, episoth. 
\on{3}.~\HL{pxorna'lor}{\B{pxor\-na'\-lor}}, sari.
\on{4}.~\HL{sApxor}{\B{sä\-pxor}}, explosion.

% pxorna'
\hi
\HT{pxorna'}{\B{\cp{pxor}na'}} \I{["p'oR.naP]} \emph{n.}, \ding{96} episoth, ``exploding seed'', \emph{croquembouche columnare}. 
(\ding{43}~\HL{pxor}{\B{pxor}}, \HL{rina'}{\B{ri\-na'}})

% pxorna'lor
\hi
\HT{pxorna'lor}{\B{\cp{pxor}na'lor}} \I{["p'oR.naP.""loR]} \emph{n.}, \ding{96} sari, ``beautiful exploding seed'', \emph{pandargonium cyanellum}. 
(\ding{43}~\HL{pxorna'}{\B{pxor\-na'}}, \HL{lor}{\B{lor}})

% pxorpam
\hi
\HT{pxorpam}{\B{\cp{pxor}pam}} \I{["p'oR.pam]} \emph{n.} ejective consonant. 
(\ding{43}~\HL{pxor}{\B{pxor}}, \HL{pam}{\B{pam}})

% pxul
\hi 
\HT{pxul}{\B{pxul}} \I{[p'ul]} \emph{or} \I{[p'Ul]} \emph{adj.} formidable, imposing. 
\B{Lu Saw\-tu\-te wä\-tu a\-pxul.} The Skypeople are formidable opponents.~\LINK{http://naviteri.org/2020/08/hiia-vur-a-teri-mefalukantsyip-a-little-story-about-two-cats/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpxul}{\B{tì\-pxul}}, formidableness, imposingness. 
\on{2}.~\HL{nIpxul}{\B{nì\-pxul}}, formidably, imposingly.

% pxun
\hi
\HT{pxun}{\B{pxun}} \I{[p'un]} \emph{or} \I{[p'Un]} \emph{n.} arm. 
\B{A pe\-yä tì\-txur … Mì pun | Na ay\-vul a\-hu\-saw\-nu.} Whose strength is in our arms | As sheltering branches.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} 
\B{Me\-pun\-ìri oe 'efu ngeyn nì\-ngay!} My arms are really tired!~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
\B{Po pxun\-fa pxek tsa\-krr me\-fo zup ne\-kll!} He shoved the two of them and they fell down.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
\B{Po\-ri pxun\-pxaw lu pxaw\-pxun.} Around his arm is an armband.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{pxuntil}{\B{pxun\-til}}, elbow. 
\on{2}.~\HL{pxawpxun}{\B{pxaw\-pxun}}, armband.

% pxuntil
\hi
\HT{pxuntil}{\B{\cp{pxun}til}} \I{["p'un.til]} \emph{or} \I{["p'Un.]} \emph{n.} elbow. 
(\ding{43}~\HL{pxun}{\B{pxun}}, \HL{til}{\B{til}})

\end{multicols}
