\pdfbookmark[0]{R}{R}
\begin{center}
\section*{R \I{[R]} (\emph{ReR})}
\end{center}

\begin{multicols}{2}

% -r
\hi
\HT{-r}{\B{-r}} \I{[R]} \emph{suf. to mark the dative case of a n. ending in a vowel or the diphthongs aw and ew}.
(\ding{43}~\HL{-ru}{\B{-ru}}, \HL{-ur}{\B{-ur}})

% ra'un
\hi 
\HT{ra'un}{\B{\cp{ra}'un}} \I{["Ra.Pun]} \emph{or} \I{[.PUn]} (\textsc{rn}: \B{\cp{ra}ùn} \I{[Ra.Un]}) \emph{vtr.} surrender s.t., relinquish s.t., give up s.t. 
\B{Fì\-atx\-kxe\-ti ke ra\-ya\-'un ay\-oel kaw\-krr!} We will never give up this land.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{velek}{\B{ve\-lek}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIra'un}{\B{tì\-ra\-'un}}, surrender, relinquishment.

% rayl
\hi 
\HT{rayl}{\B{rayl}} \I{[Rajl]} \emph{adj.} feeling pity and sorrow for s.o.'s misfortune (\emph{used with} \ding{43}~\HL{'efu}{\B{'e\-fu}}). 
\B{Po\-ri sa'\-nok tì\-mer\-kup; 'e\-fu oe rayl.} His mother just died; I feel bad for him.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b}

% ral
\hi
\HT{ral}{\B{ral}} \I{[Ral]} \emph{n.} meaning.
\B{Fì\-lì\-'u\-ä ral lu tì\-me\-'em sì tì\-ru\-sey mì hi\-fkey na Nawm\-a Sa'\-nok\-ä ha\-pxì.} The meaning of this word is ``harmony, living in the world as part of the Great Mother.''~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Ral\-it fay\-lì\-'u\-ä oel ke tsli\-vam.} I don't understand the meaning of these words.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Ra\-lur law so\-li fo oe\-ru.} They made the meaning clear to me.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}
\B{Sra\-ke tsun nga ral\-ur 'en si\-vi?} Can you guess the meaning?~\LINK{http://naviteri.org/2014/03/value-and-worth/comment-page-1/\#comment-2669}{b}
\B{Ral\-ìri ke 'efu oel kea tì\-ke\-teng\-it.} I don't feel any difference concerning the meaning.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1903}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{ralpeng}{\B{ral\-peng}}, interpret. 
\on{2}.~\HL{tafral}{\B{ta\-fral}}, therefore. 
\on{3}.~\HL{ralnga'}{\B{ral\-nga'}}, meaningful, instructive. 
\on{4}.~\HL{ralke}{\B{ral\-ke}}, meaningless, devoid of content.
\on{5}.~\HL{tengralI'u}{\B{teng\-ra\-lì\-'u}}, synonym.
\on{6}.~\HL{wAralI'u}{\B{wä\-ra\-lì\-'u}}, antonym.

% ralke
\hi 
\HT{ralke}{\B{\cp{ral}ke}} \I{["Ral.kE]} \emph{adj.} meaningless, devoid of content. 
\B{Txe\-wì ka trro nì\-wotx ftxo\-lu\-lì\-'u, slä ay\-lì\-'u peyä lä\-ngu ral\-ke.} Txewì spoke for an entire day, but sadly, his words were meaningless.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{ral}{\B{ral}}, \HL{luke}{\B{lu\-ke}})

% ralnga'
\hi
\HT{ralnga'}{\B{\cp{ral}nga'}} \I{["Ral.NaP]} \emph{adj.} meaningful, instructive, s.t. from which a lesson can be learned.   
\B{Kì\-rey\-sì\-ri lu tsa\-puk\-ä ay\-vur a te\-ri 'Rr\-ta txan\-wa\-we nì\-ngay, slä oe\-ri, hu\-fwa el\-tur tì\-txen si, lu ral\-nga' nì\-'aw.} For Grace, the stories in that book relating to Earth are personally meaningful, but for me, although interesting, they’re simply instructive.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{ral}{\B{ral}}, \HL{-nga'}{\B{-nga'}}; \ding{214}~\HL{ralke}{\B{ral\-ke}}) 

% ralpeng
\hi
\HT{ralpeng}{\B{ral\cp{peng}}} \I{[Ral."p$\cdot$EN]} \emph{vtr.}\on{22} interpret, translate. 
\B{Wa\-we\-ti ke tsun fko ral\-pi\-veng.} One can't explain (or: put into words) \emph{wawe}.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Ze\-ne Sa'\-nok fì\-au\-ngia\-ti ral\-pi\-veng.} Mother must interpret this sign.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/}{ln\slash a\textsuperscript{c}}
(\ding{43}~\HL{ral}{\B{ral}}, \HL{peng}{\B{peng}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIralpeng}{\B{tì\-ral\-peng}}, translation, interpretation.

% Ralu
\hi
\B{\cp{Ra}lu} \I{["Ra.lu]} \emph{n.} \emph{male name}. 
\B{Ra\-lu to\-lì\-ran äo vul apxa.} Ralu walked under the large branch.~\LINK{http://forum.learnnavi.org/language-updates/over-under-and-quickly/}{ln}
\B{Ra\-lu\-ri fa\-hew fkan oe\-ru na ye\-rik.} Ralu smells like a hexapede to me.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}

% ram
\hi
\HT{ram}{\B{ram}} \I{[Ram]} \emph{n.} mountain.
\B{Ay\-ram A\-lu\-sìng}, the Floating Mountains. \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{ramtsyIp}{\B{ram\-tsyìp}}, hill. \\
\on{2}.~\HL{txepram}{\B{txep\-ram}}, volcano.

% ramtsyìp
\hi
\HT{ramtsyIp}{\B{\cp{ram}tsyìp}} \I{["Ram.\t{ts}jIp\textcorner]} \emph{n.} hill.
\B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay.} Let's meet on the hill tomorrow at nightfall.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{ram}{\B{ram}}, \HL{-tsyIp}{\B{-tsyìp}}) 

% ramu
\hi 
\HT{ramu}{\B{Ramu}} \I{[Ra.mu]} \emph{male name}. 
\B{Ra\-mu ke lu txur nì\-txan, slä u\-van\-it yo\-lo\-ra', ti\-'a.} Ramu isn't very strong, but surprisingly, he won the game.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} 

% ramunong
\hi
\HT{ramunong}{\B{ra\cp{mu}nong}} \I{[Ra."mu.noN]} \emph{n.} well.
\ding{43}~\HL{ayvitrayA ramunong}{\B{Ayvitrayä Ramunong}}, the Well of Souls.

% ran
\hi
\HT{ran}{\B{ran}} \I{[Ran]} \emph{n.} (\emph{a.}) intrinsic character or nature, essence, constitution, (\emph{basic nature of s.t. resulting from the totality of its properties}).
\B{Fra\-'u\-ä ran ngä\-pop fa fra\-syon tse\-yä.} The \emph{ran} of each thing arises from the totality of its attributes.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Ran tì\-ru\-sol\-ä pe\-yä lu fyo\-le.} The \emph{ran} of her singing is sublime.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\emph{b. for people}) personality.
\B{Mun\-txa\-tu\-ri So\-rewn\-ti ke tsun oe mi\-vll\-'an. Ran pe\-yä oe\-ru ke ha'.} I can't accept Sorewn as my spouse. Her personality doesn't suit me.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIran}{\B{nì\-ran}}, basically, fundamentally, in essence. 
\on{2}.~\HL{loran}{\B{lo\-ran}}, elegance, grace. 
\on{3}.~\HL{fe'ran}{\B{fe'\-ran}}, flawed nature.

% Ranu
\hi 
\HT{ranu}{\B{Ranu}} \I{[Ra.nu]} \emph{n. given name}.

% rangal
\hi
\HT{rangal}{\B{\cp{rang}al}} \I{["RaN.al]} \emph{vin.} wish (\emph{used with} \HL{tsnI}{\B{tsnì}}). \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIrangal}{\B{nì\-ra\-ngal}}, ``I wish that …''.
\on{2}.~\HL{sArangal}{\B{sä\-ra\-ngal}}, a wish.

% raspu'
\hi
\HT{raspu'}{\B{ra\cp{spu'}}} \I{[Ra."{}spuP]} \emph{or} \I{[."{}spUP]} \emph{n.} \ding{246} leggings (\emph{used in war}).
\B{Fì\-ra\-spu' 'e\-kxin lu nì\-hawng. Ke tsun oe yi\-vem\-stokx.} These leggings are too tight. I can't wear them.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}

% raw
\hi 
\HT{raw}{\B{raw}} \I{[Raw]} \emph{adp.} down to. 
\B{Ko\-lä oe raw kil\-van fte iva\-ho.} I went down to the river to pray.
\B{Kll\-za\-'u yì\-raw a\-mu\-ve.} Descend to the second level.
\B{Ti\-am ta vo\-mrr raw pxey.} Count down from thirteen to three. ~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{214}~\HL{vay}{\B{vay}})

% rawke
\hi
\HT{rawke}{\B{\cp{raw}ke}} \I{["Raw.kE]} \emph{intj.} alarm cry, call to defense.

% rawm
\hi
\HT{rawm}{\B{rawm}} \I{[Rawm]} \emph{n.} (\emph{of weather}) lightning (\emph{general term}).
(\ding{43}~\HL{atanzaw}{\B{atan\-zaw}})  \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{rawmpxom}{\B{rawm\-pxom}}, thunder and lightning.

% rawmpxom
\hi
\HT{rawmpxom}{\B{\cp{rawm}pxom}} \I{["Rawm.p'om]} \emph{n.} thunder and lightning.
(\ding{43}~\HL{rawm}{\B{rawm}}, \HL{'rrpxom}{\B{'rr\-pxom}}) 

% rawn
\hi
\HT{rawn}{\B{rawn}} \I{[Rawn]} \emph{vtr.} replace, substitute. 
\B{Ro\-lawn oel pa'\-lit fa ik\-ran.} I replaced my direhorse with a banshee.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIrawn}{\B{tì\-rawn}}, replacement, act of replacing. 
\on{2}.~\HL{sArawn}{\B{sä\-rawn}}, replacement, substitute.
\on{3}.~\HL{rawnlI'u}{\B{rawn\-lì\-'u}}, pronoun.

% rawnlì'u
\hi 
\HT{rawnlI'u}{\B{\cp{rawn}lì'u}} \I{["Rawn.lI.Pu]} \emph{n.} pronoun. 
(\ding{43}~\HL{rawn}{\B{rawn}}, \HL{lI'u}{\B{lì'u}})

% rawng
\hi 
\HT{rawng}{\B{rawng}} \I{[RawN]} \emph{n.} entrance, doorway. 
\B{Fpxo\-lä\-kìm fo ìlä rawng a\-hì\-'i.} They entered through a small doorway.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 

% rawp
\hi
\HT{rawp}{\B{rawp}} \I{[Rawp\textcorner]} \emph{n.}, \ding{96} bladder polyp, \emph{polypiferus brevisimus}. 

% rä'ä
\hi
\HT{rA'A}{\B{rä\cp{'ä}}} \I{[R\ae."P\ae]} \emph{part. to form a negative imperative}, ``don't (do s.t.)!''.
(\emph{a. usually before the verb})
\B{Nim rä\-'ä lu!} Don't be shy!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Hawng\-krr rä\-'ä zi\-va\-'u!} Don't come late!~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Txo\-pu rä\-'ä si, lu ke\-tu\-wong\-o nì\-'aw.} Don't be afraid, it's just an alien.~\LINK{http://forum.learnnavi.org/language-updates/a-collection/}{ln\slash a} 
(\emph{b. after the verb for special emphasis}) \B{Oe\-ti 'am\-pi rä\-'ä, ma skxawng!} Don't touch me, you moron!~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{Hay\-a yak\-ro fti\-vang. Sa\-lew rä\-'ä.} Stop at the next fork. Do not proceed further.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Na\-ri si! Kll\-te\-sìn lu pay a\-txan. Fwi rä\-'ä!} Be careful! There’s a lot of water on the ground. Don't slip!~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{c. proverb}) \B{Kem\-ìri a nga\-ru prr\-te' ke lu, tsa\-kem rä\-'ä si\-vi ay\-la\-he\-ru.} As for any action that you don't like, don't act that way to others.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}

% räptulì'fya
\hi 
\HT{rAptulI'fya}{\B{räptu\cp{lì'}fya}} \I{[R\ae{}p\textcorner.tu."lIP.fja]} (\textsc{rn}: \B{räptù\cp{lì'}fya} \I{[REp\textcorner.tU."lIP.fja]}) \emph{n.} coarse, vulgar language. 
(\ding{43}~\HL{rAptum}{\B{räp\-tum}}, \HL{lI'fya}{\B{lì'\-fya}}) 

% räptulì'u
\hi 
\HT{rAptulI'u}{\B{räptu\cp{lì}'u}} \I{[R\ae{}p\textcorner.tu."lI.Pu]} (\textsc{rn}: \B{räptù\cp{lì}u} \I{[REp\textcorner.tU."lI.u]}) \emph{n.} coarse or swear word. 
(\ding{43}~\HL{rAptum}{\B{räp\-tum}}, \HL{lI'u}{\B{lì\-'u}}) 

% räptum
\hi
\HT{rAptum}{\B{räp\cp{tum}}} \I{[R\ae{}p\textcorner."tum]} \emph{or} \I{[."tUm]} (\textsc{rn}: \B{räp\cp{tùm}} \I{[REp\textcorner."tUm]}) \emph{adj.} coarse, vulgar, socially unacceptable.
\B{Po\-ri lu sno\-lup räp\-tum, ul\-te mawl ay\-lì\-'uä ke lu ngay.} His personal style is coarse and half of the words are not true.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\emph{a. idiom}) \B{Rä\-'ä räp\-tum!} Don't be impolite.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
\B{Ngal new pe\-ut? -- Rä\-'ä räp\-tum, ma 'eveng! Tsay\-lì\-'u ke lu muiä!} What do you want? -- Don't be impolite, child! Those words are improper!~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\emph{b. idiom}) \B{pll\-txe räp\-tum}, speak vulgarly, use vulgar language, swear.\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIrAptum}{\B{nì\-räp\-tum}}, coarsly, vulgarly.  
\on{2}.~\HL{rAptulI'u}{\B{räp\-tu\-lì\-'u}}, coarse or swear word.
\on{3}.~\HL{rAptulI'fya}{\B{räp\-tu\-lì'\-fya}}, coarse, vulgar language.

% räzekx
\hi 
\HT{rAzekx}{\B{\cp{rä}zekx}} \I{["R\ae.zEk']} \emph{n.} curse. \\
(i.) \B{tìng \cp{rä}zekx} \emph{vin.}, to curse (s.o.), wishing s.o. ill.
\B{Oe nga\-ru tìng rä\-zekx, ma kal\-wey\-a\-veng!} I curse you, (OR: Damn you!), you son of a bitch.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
\B{Rä\-zekx ngar!} Damn you! [lit.: ``(I) curse you!'']~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/\#comment-54817}{b}
(\ding{214}~\HL{syawn}{\B{syawn}}) 

% re'o
\hi
\HT{re'o}{\B{\cp{re}'o}} \I{["RE.Po]} \emph{n.} head. 
\B{Oe\-ri ay\-som\-pì\-va sìn re\-'o var zi\-vup.} Raindrops keep falling on my head.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Krro lu 'Ìng\-lì\-sì txur nì\-hawng mì re\-'o.} Sometimes English is too strong in (my) head.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b} 
(\emph{a. idiom}) \B{(na) la\-nay'\-ka lu\-ke re\-'o}, ``completely lost,'' [lit.: Like a slinger without a head].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Po maw kxitx mun\-txa\-tu\-ä 'a\-me\-fu na la\-nay'\-ka re'\-o\-lu\-ke.} After the death of his spouse, he felt completely lost.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{hawre'}{\B{haw\-re'}}, hat. 
\on{2}.~\HL{zare'}{\B{za\-re'}}, forehead, brow. 

% rel
\hi
\HT{rel}{\B{rel}} \I{[REl]} \emph{n.} image, picture.
\B{Rel oe\-yä na unil\-tì\-ran\-yu lor lu nì\-ngay!} My picture as a Dreamwalker is truly beautiful.~\LINK{http://forum.learnnavi.org/language-updates/art-related-vocab/}{ln} 
\B{Lu tsay\-rel ko\-sman nì\-wotx!} Those pictures are wonderful!~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/comment-page-1/\#comment-2485}{b}
\B{Fì\-me\-rel\-it 'o\-ngo\-lop aw\-nge\-yä tsul\-fä\-tul rel\-tseo\-ä alu Älìn.} The two portraits were created by our Master of Visual Arts, Alan.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
\B{Ean\-a rel\-ìri lei\-am nga say\-rìp nì\-txan nang!} It's nice for me to see you look extremely handsome in your blue photo.~\LINK{http://forum.learnnavi.org/language-updates/eana-reliri-leiam-nga-sayrip-nitxan-nang!/}{ln}  \\ 
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{reltseo}{\B{rel\-tseo}}, visual art.  
\on{2}.~\HL{rel arusikx}{\B{rel a\-ru\-sikx}}, movie, film, video. 
\on{3}.~\HL{pamrel}{\B{pam\-rel}}, writing, s.t. written. 
\on{4}.~\HL{keyrel}{\B{key\-rel}}, facial expression. 
\on{5}.~\HL{ronsrelngop}{\B{ron\-srel\-ngop}}, imagine. 
\on{6}.~\HL{ronsrel}{\B{ron\-srel}}, s.t. imagined. 
\on{7}.~\HL{syeprel}{\B{syep\-rel}}, camera. 
\on{8}.~\HL{kakrel}{\B{kak\-rel}}, blind. 
\on{9}.~\HL{atxkxerel}{\B{atx\-kxe\-rel}}, map.
\on{10}.~\HL{relvul}{\B{rel\-vul}}, pen, pencil.

% rel arusikx
\hi
\HT{rel arusikx}{\B{rel aru\cp{sikx}}} \I{[REl a.Ru."{}sik']} \emph{n.} movie, film, video [lit.: moving picture].
\B{Am\-'a\-lu\-ke hay\-a pxe\-rel a\-ru\-sikx wa\-you!} Doubtless the next three movies will be amazing!~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
(\ding{43}~\HL{rel}{\B{rel}}, \HL{rikx}{\B{rikx}})

% reltseo
\hi
\HT{reltseo}{\B{\cp{rel}tseo}} \I{["REl.\t{ts}E.o]} \emph{n.} visual art.
\B{Law lu oe\-ru fwa nga mì rel\-tseo no\-lu\-me nì\-txan!} It's clear to me that you learned a lot in visual art.~\LINK{http://forum.learnnavi.org/language-updates/art-related-vocab/}{ln}
\B{Fì\-me\-rel\-it 'o\-ngo\-lop aw\-nge\-yä tsul\-fä\-tul rel\-tseo\-ä alu Älìn.} The two portraits were created by our Master of Visual Arts, Alan.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
(\ding{43}~\HL{rel}{\B{rel}}, \HL{tseo}{\B{tseo}}) \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{reltseotu}{\B{rel\-tseo\-tu}}, artist.

% reltseotu
\hi
\HT{reltseotu}{\B{\cp{rel}tseotu}} \I{["REl.\t{ts}E.o.tu]} \emph{n.} artist.
\B{Rel\-tseo\-tu a\-txan\-tsan lu nga!} You are an excellent artist.~\LINK{http://forum.learnnavi.org/language-updates/art-related-vocab/}{ln}
(\ding{43}~\HL{reltseo}{\B{rel\-tseo}}, \HL{tute1}{\B{tu\-te}}) 

% relvul
\hi
\HT{relvul}{\B{\cp{rel}vul}} \I{["REl.vul]} \emph{or} \I{[.vUl]} \emph{n.} pen, pencil (\emph{short form of} \ding{43}~\HL{pamrelvul}{\B{pam\-rel\-vul}}).

% rem
\hi 
\HT{rem}{\B{rem}} \I{[REm]} \emph{n.} fuel. 
\B{Na'\-vi\-ri lu fì\-ut\-ral\-ä rìn rem le\-tsran\-ten.} The wood of this tree is an important fuel for the Na'vi.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 

% renten
\hi
\HT{renten}{\B{\cp{ren}ten}} \I{["REn.tEn]} \emph{n.} goggles (\emph{made from insect wings, carved wood etc.}).
\B{Fo\-ri maw\-krr\-a fa ren\-ten ioi sä\-po\-li ho\-lum.} After they put on their goggles, they left.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}

% renu
\hi
\HT{renu}{\B{\cp{re}nu}} \I{["RE.nu]} \emph{n.} pattern; paradigm. 
\B{Renu Ayinanfyayä}, the Senses Paradigm.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Sìl\-pey oe, fì\-re\-nu a\-mip nga\-ru le\-sar lì\-ye\-vu.} I hope, this new (grammatical) pattern will be useful for you.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/comment-page-1/\#comment-2237}{b}
\B{Pam\-tse\-ol ngop ay\-re\-nut}, Music creates patterns.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Tsay\-re\-nur le\-we\-opx a sìn ne\-ni tìng na\-ri.} Look at those wave\hyp{}like patterns in the sand.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
\B{Ya\-mì tsun fko tsi\-ve\-'a lo\-ran\-it re\-nu\-ä kil\-van\-ä slä kll\-te\-sìn wä\-pan.} From the air you can see the grace of the river's form but from the ground it's hidden.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{renulke}{\B{re\-nul\-ke}}, irregular, random. 
\on{2}.~\HL{rengop}{\B{re\-ngop}}, design (finer detail). 
\on{3}.~\HL{renu ngampamA}{\B{re\-nu ngam\-pa\-mä}}, rhyme scheme.

% renu ngampamä
\hi
\HT{renu ngampamA}{\B{\cp{re}nu \cp{ngam}pamä}} \I{["RE.nu "Nam.pa.m\ae]} \emph{n.} rhyme scheme.   
\B{Fì\-way\-ri hì\-no\-a re\-nut ngam\-pam\-ä ke tsä\-ngun oe tsli\-vam.} I'm afraid I can't understand the intricate rhyme scheme of this poem.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{renu}{\B{re\-nu}}, \HL{ngampam}{\B{ngam\-pam}})

% renulke
\hi
\HT{renulke}{\B{\cp{re}nulke}} \I{["RE.nul.kE]} \emph{or} \I{[.nUl.]} \emph{adj.} irregular, random. 
\B{Eo ay\-fo a fya\-'o la\-mu ay\-skxe\-ta te\-ya sì re\-nul\-ke.} The path ahead of them was full of rocks and irregular.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{renu}{\B{re\-nu}}, \HL{luke}{\B{lu\-ke}})

% reng
\hi
\HT{reng}{\B{reng}} \I{[REN]} \emph{adj.} (\emph{a. physical}) shallow. 
\B{kil\-van a\-reng}, a shallow river.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\emph{b. sometimes applied metaphorically to thoughts, ideas, analyses, etc.}) shallow.
\B{Tì\-eyk\-tan\-ìri ay\-sä\-fpìl pe\-yä lu reng, ay\-sì\-hawl lu fe'\-ran.} Concerning leadership his thoughts are shallow, his plans are flawed.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{214}~\HL{txukx}{\B{txukx}})

% rengop
\hi
\HT{rengop}{\B{\cp{re}ngop}} \I{["RE.N$\cdot$op\textcorner]} \emph{vtr., inf.}\on{22} design (\emph{for finer details, e.g. ornament design}).
\B{Nge\-yä tsa\-fkxi\-let tu\-pel re\-ngo\-lop?} Who designed that necklace of yours?~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{renu}{\B{re\-nu}}, \HL{ngop}{\B{ngop}}; \HL{'ongop}{\B{'o\-ngop}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{sArengop}{\B{sä\-re\-ngop}}, design (particular instance). 
\on{2}.~\HL{tIrengop}{\B{tì\-re\-ngop}}, design (abstract concept).

% rewon
\hi
\HT{rewon}{\B{\cp{re}won}} \I{["RE.won]} \emph{n.; adv.} morning.
\B{Fì\-re\-won tom\-pa\-meyp zar\-mup, slä set 'ì\-me\-ko nì\-txan nang.} It was drizzling this morning, but it's really started coming down now!~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Alo a\-hay oe za\-sya\-'u mì re\-won.} Next time I'll come in the morning.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \\ 
(i.) \B{rewo\cp{nay}} \I{[RE.wo."naj]} \emph{n.; adv.} tomorrow morning.   
\B{Oeng re\-wo\-nay 'aw\-si\-teng ti\-va\-ron ko.} Let's you and I go hunting together tomorrow morning.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} \\  
(ii.) \B{rewo\cp{nam}} \I{[RE.wo."nam]} \emph{n.; adv.} yesterday morning. 

% rey
\hi
\HT{rey}{\B{rey}} \I{[REj]} \emph{vin.} live.  
\B{ke rey a tu\-te}, person who doesn't live\slash isn't alive.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{Txo new nga ri\-vey, oe\-hu!} Come with me, if you want to live!~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog_.28Twitter_Questions.29}{wiki} 
\B{Pi\-veng oer ftxey nga new ri\-vey fu\-ke.} Tell me whether you want to live (or not).~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/}{ln} 
\B{Fì\-lì\-'u\-ä ral lu tì\-me\-'em sì tì\-ru\-sey mì hi\-fkey na Nawm\-a Sa'\-nok\-ä ha\-pxì.} The meaning of this word is ``harmony, living in the world as part of the Great Mother.''~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Ko\-ren A\-'aw\-ve Tì\-ru\-sey\-ä 'Aw\-si\-teng}, the First Rule of Living Together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tIrey}{\B{tì\-rey}}, life. 
\on{2}.~\HL{rusey}{\B{ru\-sey}}, living thing; alive, living. 
\on{3}.~\HL{rey'eng}{\B{rey\-'eng}}, balance of life. 
\on{4}.~\HL{reyfya}{\B{rey\-fya}}, way of living, culture. 
\on{5}.~\HL{reypay}{\B{rey\-pay}}, blood. 
\on{6}.~\HL{emrey}{\B{em\-rey}}, survive.

% rey'eng
\hi
\HT{rey'eng}{\B{rey\cp{'eng}}} \I{[REj."PEN]} \emph{n.} balance of life.
\B{'Uo a fpi rey\-'eng Ey\-wa\-'e\-veng\-mì, 'Rr\-ta\-mì tsran\-ten nì\-txan aw\-nga\-ru nì\-wotx.} Something that matters a lot to all of us for the sake of The Balance of Life on both Pandora and Earth.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\ding{43}~\HL{rey}{\B{rey}}, \HL{'engeng}{\B{'eng\-eng}})

% Reya
\hi
\HT{Reya}{\B{Reya}} \I{[RE.ja]} \emph{n., female name} (\emph{short form of:} \ding{43}~\HL{Tsireya}{\B{Tsi\-re\-ya}}).

% reyfya
\hi
\HT{reyfya}{\B{\cp{rey}fya}} \I{["REj.fja]} \emph{n.} way of living, culture.
\B{'Rr\-ta\-mì a rey\-fya Saw\-tu\-te\-yä la\-tsu fe'\-ran nì\-ngay.} The Skypeople's culture on earth must truly be flawed.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{rey}{\B{rey}}, \HL{fya'o}{\B{fya\-'o}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lereyfya}{\B{le\-rey\-fya}}, cultural.

% reykol
\hi
\HT{reykol}{\B{rey\cp{kol}}} \I{[REj."k$\cdot$ol]} \emph{vtr., inf.}\on{22} play music (\emph{poetic}).  
\B{Tew\-ti, nga lu tsul\-fä\-tu i\-'en\-ä. Ngal tsat rey\-kì\-mol!} Wow, you are a master on the \emph{i'en}. You just made it sing!~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{rol}{\B{rol}}, ‹\HL{eyk2}{\B{eyk}}›)

% reym
\hi
\HT{reym}{\B{reym}} \I{[REjm]} \emph{n.} dry land (\emph{as distinct from water}). 
\B{Pe\-ioang tsun mì tam\-pay kop mì reym ri\-vey?} What animal can live both in the sea and on the land?~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 

% reypay
\hi
\HT{reypay}{\B{\cp{rey}pay}} \I{["REj.paj]} \emph{n.} blood. 
\B{Rey\-pay skxir\-ftum\-fa he\-rum.} Blood is coming out of the wound.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\ding{43}~\HL{rey}{\B{rey}}, \HL{pay}{\B{pay}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{reypaytun}{\B{rey\-pay\-tun}}, red. 
\on{2}.~\HL{reyptswIk}{\B{reyp\-tswìk}}, wolf tick.

% reypaytun
\hi 
\HT{reypaytun}{\B{\cp{rey}paytun}} \I{["REj.paj.tun]} \emph{or} \I{[.tUn]} \emph{adj.} red.
\B{'Eveng\-ìl fì\-rel\-it 'opin\-so\-lung fa pin\-vul a\-rey\-pay\-tun.} The child colored (in) this picture with a red crayon.~ \LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/\#comment-28355}{b}
(\ding{43}~\HL{reypay}{\B{rey\-pay}}, \HL{tun}{\B{tun}})

% reyptswìk
\hi 
\HT{reyptswIk}{\B{\cp{reyp}tswìk}} \I{["REjp\textcorner.\t{ts}wIk\textcorner]} \emph{n., fauna} wolf tick. 
(\ding{43}~\HL{reypay}{\B{rey\-pay}}, \HL{tswIk}{\B{tswìk}}) 

% -ri
\hi
\HT{-ri}{\B{-ri}} \I{[.Ri]} \emph{suf. to mark the topic case of a n. ending in a vowel or diphthong}.
(\ding{43}~\HL{-Iri}{\B{-ìri}})

% rikx
\hi
\HT{rikx}{\B{rikx}} \I{[Rik']} \emph{vin.} move, shift position. 
\B{Ftu oe ne\-to rikx!} Get away from me!~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-994}{b}
\B{Tsun kxam\-lä na'\-rìng ri\-vikx nì\-fnu nì\-wotx.} (The thanator) can move silently through the forest.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{'ampirikx}{\B{'am\-pi\-rikx}}, leaf pitcher plant. 
\on{2}.~\HL{kllrikx}{\B{kll\-rikx}}, earthquake. 
\on{3}.~\HL{rel arusikx}{\B{rel a\-ru\-sikx}}, movie, film, video.

% rim
\hi
\HT{rim}{\B{rim}} \I{[Rim]} \emph{adj.} yellow.   
\B{Fì\-syu\-lang lu rim.} This flower is yellow.\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Fì\-syu\-lang a\-rim lu hì\-'i fra\-to.} This yellow flower is the smallest of all.\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{rim na na\-ri}, the yellow of a Na'vi eye.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIrim}{\B{nì\-rim}}, in yellow. 
\on{2}.~\HL{rimpin}{\B{rim\-pin}}, the color yellow. 

% rimpin
\hi
\HT{rimpin}{\B{\cp{rim}pin}} \I{["Rim.pin]} \emph{n.} the color \ding{43}~\HL{rim}{\B{rim}}.
\B{Ke su\-nu oe\-ru rim\-pin.} I don't like the color yellow.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
(\ding{43}~\HL{'opin}{\B{'opin}})

% rina'
\hi
\HT{rina'}{\B{ri\cp{na'}}} \I{[Ri."naP]} \emph{n.}, \ding{96} seed. 
\B{Ut\-ral\-ä A\-nawm | Ay\-ri\-na' lu ay\-oeng.} We are all seeds | Of the Great Tree.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Ay\-fol za\-mo\-lu\-nge aw\-ngar ay\-ri\-na'\-it fa sä\-he\-na a\-pxa.} They brought us the seeds in a large container.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{atokirina'}{\B{ato\-ki\-ri\-na'}}, seeds of the great tree, woodsprite.  
\on{2}.~\HL{tsyorina'wll}{\B{tsyo\-ri\-na'\-wll}}, cycad.

% Rini
\hi
\B{\cp{Ri}ni} \I{["Ri.ni]} \emph{n.} \emph{female name}. 
\B{Ri\-ni tsa\-po\-hu ho\-lum nì\-mal nì\-wotx.} Rini left with that guy without thinking twice about it.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Ke new oe mi\-vay' tsngan\-ti a 'o\-lem Ri\-nil.} I don't want to taste the meat that Rini cooked.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Nik\-re\-ri Ri\-ni\-yä 'ur fkan lor.} Rini's hair looks beautiful.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Ri\-ni\-ri nik\-re yä\-ngo' nì\-tut.} Rini's hair is always perfect.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}

% ripx
\hi 
\HT{ripx}{\B{ripx}} \I{[Rip']} \emph{vtr.} pierce. 
\B{Lu Ney\-ti\-ri\-ru 'aw\-a mik\-yun a\-raw\-nipx nì\-'aw.} Neytiri has only one pierced ear.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
(\emph{a. idiom}) \B{Pay\-ìl a lipx tskxe\-ti ripx.} Dripping water pierces a stone. [{\em persistent effort can accomplish unexpected and amazing things}]~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}

% riti
\hi
\HT{riti}{\B{\cp{ri}ti}} \I{["Ri.ti]} \emph{n., fauna} stingbat, \emph{scorpiobattus volansii}. 
\B{Ri\-ti tswo\-lay\-on ftum\-fa slär.} The stingbat flew out of the cave.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{mrr\-a ri\-ti}, five stingbats.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}

% rì'ìr
\hi
\HT{rI'Ir}{\B{\cp{rì}'ìr}} \I{["RI.PIR]} \textsc{i}. \emph{n.} reflection.
\B{Oe\-ri pay\-ìl tìng rì\-'ìr\-it key\-ä.} My face is reflected in the water.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
\textsc{ii}. \HL{rI'Irsi}{\B{\cp{rì}'ìr si}} \I{["RI.PIR si]} \emph{vin.} reflect, imitate.   
\B{Rì\-'ìr rä\-'ä si\-vi tsmuk\-tur!} Don't imitate your sibling!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Por lu sno\-lup a new fra\-po rì\-'ìr si\-vi.} He\slash She has a personal style that everyone wants to emulate.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}

% rìk
\hi
\HT{rIk}{\B{rìk}} \I{[RIk\textcorner]} \emph{n.} (\emph{a.} \ding{96}) leaf.
\B{Na'\-rìng a zì\-sìt\-o a\-ki\-nä ke lal\-mu tsar kea rìk.} A forest that had no leaf for seven years.~\LINK{http://forum.learnnavi.org/language-updates/for-\%28duration\%29-jesus-loan-word/}{ln}
\B{Pol sä\-'o\-ti wo\-lan äo ay\-rìk.} He hid his tool under the leaves.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
\B{ean na rìk}, leaf\hyp{}green.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
(\emph{b. coll. form of} \ding{43}~\HL{yomyolerIk}{\B{yom\-yo le\-rìk}}) leaf plate. \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lerIk}{\B{le\-rìk}}, leafy. 
\on{2}.~\HL{rIkean}{\B{rìk\-ean}}, leaf\hyp{}green.
\on{3}.~\HL{rIktsyokx}{\B{rìk\-tsyokx}}, flat hand.

% rìkean
\hi
\HT{rIkean}{\B{\cp{rì}kean}} \I{["RI.kE.an]} \emph{adj.} leaf\hyp{}green, i.e. green.
(\ding{43}~\HL{rIk}{\B{rìk}}, \HL{ean}{\B{ean}}; \HL{ta'lengean}{\B{ta'\-leng\-ean}})

% rìktsyokx
\hi 
\HT{rIktsyokx}{\B{\cp{rìk}tsyokx}} \I{["RIk\textcorner.\t{ts}jok']} \emph{n.} flat hand. 
\B{A\-te\-yol En\-tut to\-la\-kuk fa rìk\-tsyokx.} Ateyo slapped Entu (i.e. struck him with his flat hand).~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b} 
(\ding{43}~\HL{rIk}{\B{rìk}}, \HL{tsyokx}{\B{tsyokx}}) 

% rìkxi
\hi 
\HT{rIkxi}{\B{rì\cp{kxi}}} \I{[RI."k'i]} \emph{vin.} tremble, shake, shiver. 
\B{Po\-ri me\-syokx rì\-kxi, ha ke tsa\-yun ye\-rik\-it ti\-va\-kuk.} His hands tremble, so he will not be able to hit the hexapede.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
\B{Ra\-lu rì\-kxi krr\-a srew, rì\-'ìr si pa\-lu\-kan\-ur a lu a\-lak\-si fte spi\-vä.} Ra\-lu does a shake while dancing, imitating a thanator that's ready to leap.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
(\emph{a. idiom}) 
\B{Rey\-kì\-kxi ut\-ral\-ti, zup ma\-u\-ti.} If you shake the tree, the fruit will fall (i.e, ``actions have consequences'').~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIrIkxi}{\B{nì\-rìkxi}}, shakily, tremblingly.

% rìn
\hi
\HT{rIn}{\B{rìn}} \I{[RIn]} \emph{n.} \ding{96} wood.
\B{Na'\-vi\-ri lu fì\-ut\-ral\-ä rìn rem le\-tsran\-ten.} The wood of this tree is an important fuel for the Na'vi.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
\B{Fì\-fne\-rìn ke ley krr\-a slu pay\-nga'.} This kind of wood is worthless when it gets damp.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
\B{Txep\-ìri, yrr\-a rìn\-ti rä\-'ä sar, ma skxawng.} Don't use that green wood for a fire, you fool.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{lerIn}{\B{le\-rìn}}, wooden, of wood.

% ro
\hi
\HT{ro}{\B{ro}} \I{[Ro]} \emph{adp.}\texttt{+} (\emph{a. locative}) at.   
\B{ro fä\-pa}, at the top.~\LINK{http://forum.learnnavi.org/language-updates/all-adpositions/msg137792/\#msg137792}{ln}
\B{ro hel\-ku}, at home.~\LINK{http://forum.learnnavi.org/language-updates/all-adpositions/msg137792/\#msg137792}{ln}
\B{Sìl\-pey oe, tsun oeng ul\-txa si\-vi ro hay\-a ul\-txa!} I hope we can meet at the next meeting.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/comment-page-1/\#comment-844}{b}
\B{Ro sngä\-'i\-tseng tsa\-lì\-'u\-ä alu 'ey\-lan lu tì\-ftang.} At the beginning of the word \emph{'eylan} there's a glottal stop.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Hay\-a yak\-ro fti\-vang. Sa\-lew rä\-'ä.} Stop at the next fork. Do not proceed further.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{New oe ri\-vun a\-sim tì\-fnu\-nga'\-a tseng\-it a tsa\-ro tsun syi\-vor tsi\-vu\-rokx fte spä\-pi\-veng.} I want to find a quiet place nearby where I can chill out and rest to get my head back on straight.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{b. temporal to disambiguate point\hyp{}in\hyp{}time interpretation together with indefinite} \ding{43}~\B{-o}) 
\B{Ro srr\-o Ra\-lu zo\-la\-'u fte oe\-hu ul\-txa si\-vi.} One day Ralu came to meet with me.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/comment-page-1/\#comment-10619}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{rofa}{\B{ro\-fa}}, beside, alongside.

% ro'a
\hi
\HT{ro'a}{\B{\cp{ro}'a}} \I{["Ro.Pa]} \emph{vin.} be impressive, inspire awe or respect.  
\B{To\-ruk Mak\-to po\-lä\-hem a fì\-'u ro\-lo\-'a nì\-txan Oma\-ti\-ka\-ya\-ru.} The arrival of Toruk Makto made a great impression on the Omaticaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Po\-ri wem\-tswo fra\-tsam\-si\-yur ro\-lo\-'a nì\-txan.} His ability to fight greatly impressed all the warriors.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{sAro'a}{\B{sä\-ro\-'a}}, feat, accomplishment, great deed.  
\on{2}.~\HL{txanro'a}{\B{txan\-ro\-'a}}, be famous.

% rofa
\hi
\HT{rofa}{\B{\cp{ro}fa}} \I{["Ro.fa]} \emph{adp.} beside, alongside. 
\B{Kll\-kxa\-yem fì\-tì\-kang\-kem oe\-yä ro\-fa--ke io--pum fe\-yä.} My work will stand beside--not above--theirs.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Maw sä\-tsway\-on a\-yol ay\-oe kll\-po\-lä mì ta\-yo a lu ro\-fa kil\-van.} After a short flight we landed in a field beside the river.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{ro}{\B{ro}}, \HL{pa'o}{\B{pa'o}})

% rol
\hi
\HT{rol}{\B{rol}} \I{[Rol]} \emph{vin.} sing.   
\B{Kä\-sro\-lìn oel nik\-ro\-it Pey\-ral\-ur yoa fwa po rol oer.} I loaned Peyral a hair ornament in exchange for her singing to me.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Pll\-txe fra\-po san Pey\-ral ke tsun ri\-vol nìl\-tsan sìk.} They say that Peyral can't sing well.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
\B{Re\-rol ay\-oe ay\-nga\-ne}, We are singing your way.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Na'\-vi ìlä ho\-'on kll\-kxo\-lem teng\-krr re\-rol.} The People were standing in a circle, singing.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Tì\-ru\-sol za\-'u ne fo nì\-'eng.} They share an interest in singing.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} 
\B{Tì\-ru\-sol lu oe\-ru mo\-wan.} Singing is enjoyable to me.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{rolyu}{\B{rol\-yu}}, singer.
\on{2}.~\HL{tIrol}{\B{tì\-rol}}, song. 
\on{3}.~\HL{ronguway}{\B{ro\-ngu\-way}}, howl.

% rolyu
\hi
\HT{rolyu}{\B{\cp{rol}yu}} \I{["Rol.ju]} \emph{n.} singer.
\B{Nì\-fkey\-to\-ngay lu po rol\-yu a\-txan\-tsan.} As a matter of fact, she is an excellent singer.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
(\ding{43}~\HL{rol}{\B{rol}})

% Ronal
\hi
\HT{Ronal}{\B{\cp{Ro}nal}} \I{["Ro.nal]} \emph{n., female name}.

% ronsem
\hi
\HT{ronsem}{\B{\cp{ron}sem}} \I{["Ron.sEm]} \emph{n.} mind. 
\B{Pam\-tse\-ol ngop ay\-re\-nut | Mì ron\-sem\-ä tì\-fnu}, Music creates patterns | In the silence of the mind.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Kxawm tsa\-lì\-'u le\-'Ìng\-lì\-sì alu ``No\"el'' mì ron\-sem lar\-mu.} Maybe I had the English word ``No\"el'' in my mind.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/comment-page-1/\#comment-499}{b}
\B{Saw\-tu\-te\-ri ron\-sem\-fkeyk\-it ke tsun kaw\-tu tsl\-ivam.} No one can understand the state of mind of the Sky People.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{ronsrel}{\B{ron\-srel}}, s.t. imagined. 
\on{2}.~\HL{fpomron}{\B{fpom\-ron}}, (mental) health. 
\on{3}.~\HL{sIlronsem}{\B{sìl\-ron\-sem}}, clever, smart (of things).

% ronsrel
\hi
\HT{ronsrel}{\B{\cp{ron}srel}} \I{["Ron.sREl]} \emph{n.} s.t. imagined, imagining.
\B{Ay\-ron\-srel pe\-yä hä\-ngek nì\-txan.} His imaginings are (unpleasantly) weird.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
(\ding{43}~\HL{ronsem}{\B{ron\-sem}}, \HL{rel}{\B{rel}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{ronsrelngop}{\B{ron\-srel\-gop}} 
\on{2}.~\HL{tIronsrel}{\B{tì\-ron\-srel}}, imagination. 
\on{3}.~\HL{leronsrel}{\B{le\-ron\-srel}}, imaginary. 
\on{4}.~\HL{nIronsrel}{\B{nì\-ron\-srel}}, in\slash by imagination.

% ronsrelngop
\hi
\HT{ronsrelngop}{\B{\cp{ron}srelngop}} \I{["Ron.sREl.N$\cdot$op\textcorner]} \emph{or coll.} \I{["Ron.sREw.Nop\textcorner]} \emph{vtr., inf.}\on{33} imagine, envision.
\B{Oel ron\-srel\-ngop fu\-ta Ey\-we\-veng\-it tok.} I imagine that I'm on Pandora.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Tsat ke tsun oe ron\-srel\-ngi\-vop.} I can't imagine that.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\ding{43}~\HL{ronsrel}{\B{ron\-srel}}, \HL{ngop}{\B{ngop}})

% rong
\hi
\HT{rong}{\B{rong}} \I{[RoN]} \emph{n.} tunnel.
\B{Po\-ti fwey\-ko\-li ay\-oel ìlä rong.} We let him slide through the tunnel.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% Rongloa
\hi
\B{\cp{Rong}loa} \I{["RoN.lo.a]} \emph{n.} \emph{clan name}.~\LINK{http://forum.learnnavi.org/language-updates/stxo/}{ln} 

% ronguway
\hi
\HT{ronguway}{\B{ro\cp{ngu}way}} \I{[R$\cdot$o."Nu.waj]} \emph{vin., inf.}\on{11} howl.
\B{Nan\-tang ro\-ngu\-way.} The viper wolf howls.~\LINK{http://forum.learnnavi.org/projects/kaltxi-ma-prrnen!-animal-learning-book/msg612075/\#msg612075}{ln}
(\ding{43}~\HL{rol}{\B{rol}}, \HL{nguway}{\B{ngu\-way}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{ronguwayyu}{\B{ro\-ngu\-way\-yu}}, howler.

% ronguwayyu
\hi
\HT{ronguwayyu}{\B{ro\cp{ngu}wayyu}} \I{[Ro."Nu.waj.ju]} \emph{n.} howler, one who howls.
(\ding{43}~\HL{ronguway}{\B{ro\-ngu\-way}})

% ropx
\hi
\HT{ropx}{\B{ropx}} \I{[Rop']} \emph{n.} hole (\emph{going clear through an object}). 
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang. Fì\-tì\-kan\-ìri ropx ke ha'.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators. For this purpose, a hole going right through the tree trunk isn't suitable.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tsongropx}{\B{tsong\-ropx}}, hole, cavity.

% rotxo
\hi 
\HT{rotxo}{\B{Ro\cp{txo}}} \I{[Ro."t'o]} \emph{n. male name}.

% rou
\hi
\HT{rou}{\B{\cp{ro}u}} \I{["Ro.u]} \emph{vin.} be drunk, get drunk.   
\B{Sra\-ne, fì\-txon tsun nga ni\-väk swo\-at nì\-'it, slä rä\-'ä rou!} Yes, you can have a little alcohol tonight, but don't get drunk!~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Nga tsun ni\-väk wew\-a pay\-it. Ul\-te ke ra\-you!} You can drink cold water. And not get drung!~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}

% -ru
\hi
\HT{-ru}{\B{-ru}} \I{[.Ru]} \emph{suf. to mark the dative case of a n. ending in a vowel, diphthong or glottal stop}.
(\ding{43}~\HL{-r}{\B{-r}}, \HL{-ur}{\B{-ur}})

% rum
\hi
\HT{rum}{\B{rum}} \I{[Rum]} \emph{or} \I{[RUm]} \emph{n.} ball. 
\B{Rum 'o\-lrr\-ko oe\-ne kll\-te\-sìn.} The ball rolled towards me on the ground.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Pol rum\-it ko\-lìm.} He spun the ball.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln}
\B{Rum\-it a no\-lu\-i rä\-'ä fe\-wi.} Don't chase after a foul ball.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\emph{a. idiom}) \B{Ngä\-zìk fwa fkol rum\-it a\-ku\-'up pey\-ku\-wup.} \emph{it's difficult to make s.o. stubborn or inept do what you want them to} [lit.: it's hard to bounce a heavy ball].~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{rumut}{\B{ru\-mut}}, puffball tree. 
\on{2}.~\HL{rumaut}{\B{ru\-ma\-ut}}, cannonball fruit tree.
\on{3}.~\HL{rumtsyokx}{\B{rum\-tsyokx}}, fist.

% rumaut
\hi 
\HT{rumaut}{\B{\cp{ru}maut}} \I{["Ru.ma.ut\textcorner]} \emph{or coll.} \I{["Ru.mawt\textcorner]} \emph{n.} \ding{96} cannonball fruit tree, \emph{ecdurus putamen pomus}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln}  
(\ding{43}~\HL{rum}{\B{rum}}, \HL{mauti}{\B{mauti}})

% rumtsyokx
\hi 
\HT{rumtsyokx}{\B{\cp{rum}tsyokx}} \I{["Rum.\t{ts}jok']} \emph{or} \I{["RUm.]} \emph{n.} fist. 
\B{A\-te\-yol En\-tut to\-la\-kuk fa rum\-tsyokx.} Ateyo punched Entu (i.e. struck him with his fist).~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b} 
(\ding{43}~\HL{rum}{\B{rum}}, \HL{tsyokx}{\B{tsyokx}})

% rumut
\hi
\HT{rumut}{\B{\cp{ru}mut}} \I{["Ru.mut\textcorner]} \emph{or} \I{[.mUt\textcorner]} \emph{n.}, \ding{96} puffball tree, \emph{obesus rotundus}. 
\B{Fì\-rum\-ut lum\-pe ke kxem?} Why isn't this puffball tree vertical?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{rum}{\B{rum}}, \HL{utral}{\B{ut\-ral}})  \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{vArumut}{\B{vä\-ru\-mut}}, vein pod.

% run
\hi
\HT{run}{\B{run}} \I{[Run]} \emph{or} \I{[RUn]} (\textsc{rn}: \B{run} \I{[Run]}) \emph{vtr.} (\emph{a.}) find, discover.
\B{Run fkol tey\-lut nì\-ho\-et.} You find \emph{teylu} everywhere.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}  
\B{Slä fma\-syi ri\-vun e\-yawr\-a tì\-'eyng\-it.} But I'll try to find the correct answer.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-920}{b}
\B{Ira\-yo 'ey\-lan\-ur aw\-nge\-yä a kxey\-ey\-ti ro\-lun} Thanks to our friend who found the mistake.~\LINK{http://forum.learnnavi.org/language-updates/another-email-from-frommer/msg68133/\#msg68133}{ln} 
\B{Zi\-va\-'u nì\-mun ye'\-rìn \ldots tsa\-krr ra\-yun ay\-ngal ay\-u\-pxa\-ret a\-mip.} Come again soon \ldots then you'll find new messages.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Ro\-le\-i\-un fu\-ta lu fo ka\-nu sì le\-so'\-ha nì\-wotx.} I found that they are intelligent and completely enthusiastic.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
(\emph{b. conv. idiom}) \B{Ro\cp{lun}!} ``Eureka! I found it!''~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\ding{214}~\B{fwew}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.\HL{ultxarun}{\B{ul\-txa\-run}}, encounter, meet by chance. 
\on{2}.~\HL{lepxìmrun}{\B{le\-pxìm\-run}}, common, often found.
\on{3}.~\HL{tIrun}{\B{tì\-run}}, discovery.

% rurur
\hi
\HT{rurur}{\B{ru\cp{rur}}} \I{[Ru."RuR]} \emph{or} \I{[."RUR]} (\textsc{rn}: \B{rù\cp{rùr}} \I{[RU."RUR]}) \emph{uncountable n.} kind of waterfall, \emph{water that is aerated while flowing among the rocks of a very gradually sloping stream}.

% rusey
\hi
\HT{rusey}{\B{ru\cp{sey}}} \I{[Ru."{}sEj]} \textsc{i}. \emph{adj.} alive, living.
\B{Krro krro, flì\-a vul a\-ru\-sey to nutx\-a pum a\-ke\-ru\-sey lu txur.} A thin living branch is sometimes stronger than a thick dead one.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 
(\ding{214}~\HL{kerusey}{\B{ke\-ru\-sey}}) \\ 
\textsc{ii}. \emph{n.} living thing.~\LINK{http://forum.learnnavi.org/language-updates/conceptual-adjectives/}{ln}
(\ding{43}~\HL{rey}{\B{rey}})

% rutxe
\hi
\HT{rutxe}{\B{ru\cp{txe}}} \I{[Ru."t'E]} \emph{intj.} please.
\B{Spi\-vaw oe\-ti ru\-txe, ma oe\-yä ey\-lan.} Please believe me, my friends.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ru\-txe ti\-vìng mik\-yun, ma fra\-po.} Your attention, please, everyone!~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Syaw oer ru\-txe trr\-ay fa spul\-mok\-ri.} Please phone me tomorrow.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}

\end{multicols}