\pdfbookmark[0]{S}{S}
\begin{center}
\section*{S \I{[s]} (\emph{Sä})}
\end{center}

\begin{multicols}{2}

% sa
\hi 
\HT{sa}{\B{sa}} \I{[sa]} \emph{vin.} (\emph{a.}) rise to a challenge. 
(\emph{b. idioms}) \B{Si\-\cp{va} ko!} Rise to the challenge! Courage! You can do it!~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
\B{So\-lei\cp{a}}! \I{[so.lE.i."{}a]} \emph{or} \I{[so.lEj."{}a]} Congratulations! Nice going! You met the challenge! You did it!~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
\B{Sa\cp{sya}!} I'll rise to the challenge! I can do it!~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
\B{Ätxä\-le si, oer syi\-vaw trr\-ay ha'\-ngir fa Skxayp?---Sa\-sya!} Could you call me on Skype tomorrow afternoon?---Sure will!\slash Will do!~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}

% sa'ewrang
\hi
\HT{sa'ewrang}{\B{sa\cp{'ew}rang}} \I{[sa."PEw.RaN]} \emph{n.} great loom, mother loom.
(\ding{43}~\HL{sa'nok}{\B{sa'\-nok}}, \HL{'ewrang}{\B{'ew\-rang}})

% sa'nok
\hi
\HT{sa'nok}{\B{\cp{sa'}nok}} \I{["{}saP.nok\textcorner]} \emph{n.} mother. 
\B{Trr\-am to\-ler\-kä\-ngup sa'\-nok a\-yaw\-ne.} My dear mother died yesterday.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Sa'\-nok\-ìl oe\-yä tsmu\-ket a\-mip no\-lokx trr\-am.} Mom gave birth to my new sister yesterday.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tsa\-ye\-rik\-tsyìp\-ìl li 'an\-la sa'\-nok\-it.} That little hexapede is already yearning for its mother.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Slä sa'\-nok\-ìri fpìl\-fya lu ke\-teng.} But mothers way of thinking is different.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(i.) \B{Nawm\-a Sa'\-nok}, the Great Mother, Eywa.
(\emph{a. idiom}) \B{(Tsa\-ri) Nga\-ru lrr\-tok (si\-vi (Nawm\-a Sa'\-nok))!} Good luck with that! [lit.: (concerning that) may the great mother smile upon you]~\LINK{http://forum.learnnavi.org/language-updates/good-luck-with-this-one!/}{ln}
\B{Fì\-lì\-'u\-ä ral lu tì\-me'\-em sì tì\-ru\-sey mì hi\-fkey na Nawm\-a Sa'\-nok\-ä ha\-pxì.} The meaning of this word is ``harmony, living in the world as part of the Great Mother''.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sa'nu}{\B{sa'\-nu}}, mommy. 
\on{2}.~\HL{sa'sem}{\B{sa'\-sem}}, parents, a parent. 
\on{3}.~\HL{sa'ewrang}{\B{sa'\-ewrang}}, great loom, mother loom.

% sa'nu
\hi
\HT{sa'nu}{\B{\cp{sa'}nu}} \I{["{}saP.nu]} \emph{n.} mommy. 
\B{Mong prr\-nen\-ìl fu\-ta sa'\-nul ve\-rewng nì\-tut.} The infant depends on her Mommy to look after her constantly.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Fnu, ma 'evi. Sa'\-nur le\-ru haw\-tsyìp. Tsi\-vu\-rokx ko.} Quiet, young one. Mommy is taking a nap. Let her rest.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Nì\-trr\-trr oel unil\-tsa sa'\-nu\-ä tey\-lut.} I regularly dream of my mom's \emph{teylu}.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{sa'nok}{\B{sa'\-nok}})

% sa'sem
\hi
\HT{sa'sem}{\B{\cp{sa'}sem}} \I{["{}saP.sEm]} \emph{n.} parents, one parent.
\B{Nì\-ke\-ftxo lu fì\-'e\-veng\-ur 'aw\-a sa'\-sem nì\-'aw.} Unfortunately this child has only one parent.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
\B{Se\-vin\-a tsa\-'eve\-ru a\-hì\-'i me\-sa'\-sem ioi so\-li fa mik\-tsang.} The parents put earrings on that pretty little girl.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Sa'\-sem\-ìri lu 'eveng\-ä kxitx yeng\-wal a\-tu\-vom.} For a parent, the greatest sorrow is the death of a child.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{sa'nok}{\B{sa'\-nok}}, \HL{sempul}{\B{sem\-pul}})

% sa'u
\hi
\HT{sa'u}{\B{\cp{sa}'u}} \I{["{}sa.Pu]}~:~\B{'u} (short plural of \HL{tsa'u}{\B{tsa'u}}).

% saa
\hi
\HT{saa}{\B{saa}} \I{[sa:]} \emph{intj.} \emph{threatening cry}.

% Saeyla
\hi
\B{Saeyla} \I{[sa.Ej.la]} \emph{n.} \emph{female name}.~\LINK{}{dh}

% saho
\hi 
\HT{saho}{\B{sa\cp{ho}}} \I{[sa."ho]} \emph{n.} prayer (\emph{concrete concept}). 
(\ding{43}~\HL{aho}{\B{aho}})

% sal
\hi
\HT{sal}{\B{sal}} \I{[sal]}~:~\B{'u} (\emph{shortened agentive form of the plural of} \HL{tsa'u}{\B{tsa\-'u}}).

% salew 
\hi
\HT{salew}{\B{sa\cp{lew}}} \I{[sa."lEw]} \emph{vin.} (\emph{a.}) proceed, go.
\B{Nga\-ri hu Ey\-wa sa\-lew ti\-rea, tokx 'ì\-'awn, slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body stays behind to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake.27s_final_kill}{a\slash wiki}
\B{Ay\-nge\-yä ay\-srr tì\-flä\-ta te\-ya li\-vu ul\-te si\-va\-lew nì\-'o' nì\-'aw.} May your days be full of success and proceed only in a fun way.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/}{b}
\B{Sa\-lew nì\-yey\-fya.} Proceed straight ahead.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\emph{b. used for stating age}) 
\B{Nga\-ri so\-la\-lew pol\-pxay\-a zì\-sìt?\slash Nga\-ri so\-la\-lew zì\-sìt a\-pol\-pxay?} How old are you?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
\B{Oe\-ri so\-la\-lew (so\-lew) zì\-sìt a\-pxe\-vol.} I'm 24 years old.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{salewfya}{\B{salewfya}}, direction, course. 
\on{2}.~\HL{zIsItsaltrr}{\B{zìsìtsaltrr}}, (yearly) anniversary.

% salewfya
\hi
\HT{salewfya}{\B{sa\cp{lew}fya}} \I{[sa."lEw.fja]} \emph{n.} direction, course.
\B{Swey\-lu set txo aw\-nga ki\-vä pe\-sa\-lew\-fya\slash pe\-sa\-lew\-fya\-ìlä?} What direction should we go in now?~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Aw\-nga ze\-ne vi\-var ìlä fì\-sa\-lew\-fya. Zen\-ke yak si\-vi.} We must continue in this direction. We must not go astray.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\emph{a. idiom}) \B{Set pe\-sa\-lew\-fya?} What do we do now?~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{salew}{\B{sa\-lew}}, \HL{fya'o}{\B{fya\-'o}})

% san
\hi
\HT{san}{\B{san}} \I{[san]} \emph{part. to indicate the beginning of a quote with verbs of speech} \HL{plltxe}{\B{pll\-txe}}, \HL{peng}{\B{peng}}, \HL{pllngay}{\B{pll\-ngay}}, \HL{pawm}{\B{pawm}}, \emph{thought} \HL{fpIl}{\B{fpìl}}, \emph{and writing} \HL{pamrel}{\B{pam\-rel si}}; \emph{used together with} \ding{43}~\HL{sIk}{\B{sìk}} \emph{when the speaker continues to say s.t. after the cited quote.}
\B{Pol\-txe Ey\-tu\-kan san oe ka\-yä sìk, slä oel pot ke spaw.} Eytukan said he would go but I don't believe him.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
\B{New oe pi\-vll\-txe ay\-nga\-ru san kal\-txì sìk ul\-te ti\-vìng ay\-ngar lì\-'ut a tì\-'e\-fu\-mì oe\-yä lu lor fra\-to mì lì'\-fya le\-Na'\-vi.} I want to say hello to you and present to you the word that, in my opinion, is the most beautiful in the Na'vi language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Po\-lawm po san sra\-ke Sä\-li ho\-lum sìk.} He asked, ``Did Sally leave?''~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/}{ln} 
\B{Alo a\-mrr po\-an po\-lawm, slä fra\-lo poe pol\-txe san ke\-he.} He asked five times, but each time she said, `no.'~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Po ke tsun pi\-vll\-ngay san oe\-ru tì\-kxey.} He can't admit he's wrong.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Oe pll\-ngay san mo\-lak\-to oe nì\-fe', ta\-fral sno\-laytx; wä\-tu lu oe\-to txur.} I acknowledge that I rode badly, so I lost; my opponent was stronger than I was.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Oel ho\-ren\-it 'Ìng\-lì\-sì\-yä so\-lar, nì\-teng\-fya fko pam\-rel si\-vi san \ldots} I used the rules of English, the same way as you write, ``\ldots''.~\LINK{http://forum.learnnavi.org/language-updates/fmawno/}{ln}
\B{Oe pam\-rel so\-li nga\-ru 'upxa\-rer san Oe za\-ya\-'u trr\-ay.} I wrote a message to you, saying ``I'll come tomorrow''.~\LINK{http://forum.learnnavi.org/language-updates/some-news-from-berlin/}{ln}

% sang
\hi
\HT{sang}{\B{sang}}\textsuperscript{\on{1}} \I{[saN]} \emph{adj.} warm. 
\B{Ya\-ri som\-wew\-pe (set)? Ya lu sang.} What's the temperature? It's warm.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Ay\-nga\-ri teng\-krr ya wur sle\-ru nì\-'ul, sìl\-pey oe, li\-vu hel\-ku sang ul\-te te'\-lan le\-fpom.} As for you all, while it's getting cooler, I hope your homes may be warm and your hearts happy.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Sìl\-pey oe tsnì fpom li\-vu ay\-nga\-ru nì\-wotx ul\-te snge\-rä\-'i a zì\-sì\-krr a\-sang zi\-vaw\-prr\-te' ay\-nga\-ne.} I hope that you are all well and that you are enjoying the beginning warm season.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{214}~\HL{wur}{\B{wur}})

\hi
\B{sang}\textsuperscript{\on{2}} : \HL{tsang}{\B{tsang}}.

% sar
\hi
\HT{sar}{\B{sar}}\textsuperscript{\on{1}} \I{[saR]} \emph{vtr.} use.
\B{Tsat sa\-yar oel nìl\-tsan.} I will use it well.~\LINK{http://forum.learnnavi.org/language-updates/score!-\%281\%29-the-verb-to-use-\%282\%29-more-detail-on-teya-si/}{ln}
\B{Kem\-lì\-'ut alu var fkol sar pxel pum alu new.} One uses the verb \emph{var} like \emph{new}.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-582}{b}
\B{Fol lì'\-fya\-ti aw\-nge\-yä sar a fya\-'o lu ko\-sman.} It's wonderful the way they use our language.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Txep\-ìri, yrr\-a rìn\-ti rä\-'ä sar, ma skxawng.} Don't use that green wood for a fire, you fool.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tskot sngo\-lä\-'i po si\-var 'a\-'aw\-a trr\-kam.} He started to use the bow several days ago.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{'En si oe, lor\-a tsa\-fkxi\-le\-ri a\-pxay\-opin so\-lar Tse\-nul srok\-it a\-vo\-zam.} I would guess that Tsenu used a thousand (lit. \on{512}) beads for that beautiful multi\hyp{}colored bib necklace.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}

\hi
\B{sar}\textsuperscript{\on{2}} : \B{'u} (\emph{shortened dative form of the plural of} \HL{tsa'u}{\B{tsa\-'u}}).

% sari
\hi
\B{\cp{sa}ri} \I{["{}sa.Ri]}~:~\B{'u} (\emph{shortened topic form of the plural of} \HL{tsa'u}{\B{tsa\-'u}}).

% sat
\hi
\B{sat} \I{[sat\textcorner]}~:~\B{'u} (\emph{shortened patientive form of the plural of} \HL{tsa'u}{\B{tsa\-'u}}).

% satu'li
\hi
\HT{satu'li}{\B{sa\cp{tu'}li}} \I{[sa."tuP.li]} \emph{n.} heritage.

% sau
\hi
\HT{sau}{\B{sa\cp{u}}} \I{[sa."{}u]} \emph{intj.} \emph{exclamation upon exertion}.

% sawtute
% \hi
% \HT{sawtute}{\B{\cp{saw}tute}} \I{["{}saw.tu.tE]}~:~\B{tawtute}. 

% say
\hi 
\HT{say}{\B{say}} \I{[saj]} \emph{adj.} loyal. 
\B{Le\-iu po ken\-'aw say\-rìp slä\-kop say.} He's not only handsome but also, I'm happy to say, loyal.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsay}{\B{tì\-say}}, loyalty. 
\on{2}.~\HL{nIsay}{\B{nì\-say}}, loyally.

% saylahe
\hi
\HT{saylahe}{\B{say\cp{la}he}} \I{[saj."la.hE]} \emph{adv.} et cetera (etc.). (\emph{from} \B{sì ay\-la\-he}, \emph{in writing on Earth may be shortened to} \B{sl.}) 
\B{Kap sì ay\-u\-nil say\-la\-he.} Threats, dreams, and other things.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}

% sayrìp
\hi
\HT{sayrIp}{\B{\cp{say}rìp}} \I{["{}saj.RIp\textcorner]} \emph{adj.} handsome, good\hyp{}looking (\emph{primarily of males}).
\B{Hu\-fwa lu fil\-ur Va'\-ru\-ä fne\-fe'\-ran\-vi, tsal\-su\-ngay fpìl fu\-ta say\-rìp lu nì\-txan.} Although Va'ru's facial stripes are rather uneven, I still think he's very handsome.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{sevin}{\B{se\-vin}}, \HL{lor}{\B{lor}})

% sä-
\hi
\HT{sA-}{\B{sä-}} \I{[s\ae.]} \emph{unprod. pref. to create n.s from v.s and adj.s with the meaning of (1) an instrument or tool, or the means by which s.t. is achieved, or of (2) a particular, concrete instance of a general action.}
(\ding{43}~\HL{tI-}{\B{tì-}})

% sä'anla
\hi
\HT{sA'anla}{\B{sä\cp{'an}la}} \I{[s\ae."Pan.la]} \emph{n.} yearning.
\B{Fì\-sä\-'an\-lal Ne\-we\-yä nga\-ti sley\-ka\-yu lek\-ye\-'ung!} This yearning for Newey is going to drive you crazy.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{'anla}{\B{'an\-la}})

% sä'eoio
\hi
\HT{sA'eoio}{\B{sä\cp{'e}oio}} \I{[s\ae."PE.o.i.o]} \textsc{i}. \emph{n.} ceremony, ritual, rite. \\
 \textsc{ii}. \B{sä\cp{'e}oio si} \I{[s\ae."PE.o.i.o si]} \emph{vin.} take part in a ceremony, perform a ritual.
\B{Fwa tsyìl Ay\-ram\-it A\-lu\-sìng lu sä\-'e\-o\-i\-o a ze\-ne fra\-po si\-vi fte sl\-ivu ta\-ron\-yu.} Climbing Iknimaya is a ritual that everyone has to perform to become a hunter.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{'eoio}{\B{'eoio}})

% sä'ipu
\hi
\HT{sA'ipu}{\B{sä\cp{'i}pu}} \I{[s\ae."Pi.pu]} \emph{n.} s.t. humorous, a particular instance of being humorous, e.g. a joke.
\B{Oe\-ru txoa li\-vu, ma 'ey\-lan. Rä\-'ä sti\-vi. Lu hì\-'i\-a sä\-'i\-pu nì\-'aw.} I'm sorry, friend. Don't be angry. It was just a small bit of humor.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{'ipu}{\B{'i\-pu}}, \HL{tI'ipu}{\B{tì\-'i\-pu}})

% sä'o
\hi
\HT{sA'o}{\B{\cp{sä}'o}} \I{["{}s\ae.Po]} \emph{n.} tool, utensil.   
\B{Na'\-vi\-ri txin\-a sä\-'o tì\-tu\-sa\-ron\-ä lu tsko swi\-zaw.} For the Na'vi the bow and arrow is the main hunting tool.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Pol sä\-'o\-ti wo\-lan äo ay\-rìk.} He hid his tool under the leaves.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tsamsA'o}{\B{tsam\-sä\-'o}}, weapon of war.

% säfe'ul
\hi
\HT{sAfe'ul}{\B{sä\cp{fe}'ul}} \I{[s\ae."fE.Pul]} \emph{or} \I{[.PUl]} \emph{n.} worsening (\emph{specific instance}). 
\B{Pe\-yä sä\-ngä\-'än\-ìri lu sä\-fe\-'ul le\-yew\-la nì\-txan.} The worsening of his depression is very disappointing.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\ding{43}~\HL{fe'ul}{\B{fe\-'ul}}, \HL{tIfe'ul}{\B{tì\-fe\-'ul}})

% säflel
\hi 
\HT{sAflel}{\B{sä\cp{flel}}} \I{[s\ae."flEl]} \emph{n.} trick, hoax, dishonest act or scheme. 
\B{Pot spaw rä\-'ä! Lu fì\-'u sä\-flel!} Don't believe him! It's a trick!~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}
(\ding{43}~\HL{flel}{\B{flel}}, \HL{tIflel}{\B{tì\-flel}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{sAfleltsyIp}{\B{sä\-flel\-tsyìp}}, practical joke.

% säfleltsyìp
\hi 
\HT{sAfleltsyIp}{\B{sä\cp{flel}tsyìp}} \I{[s\ae."flEl.\t{ts}jIp\textcorner]} \emph{n.} practical joke. 
(\ding{43}~\HL{sAflel}{\B{säflel}})

% säflä
\hi
\HT{sAflA}{\B{sä\cp{flä}}} \I{[s\ae."fl\ae]} \emph{n.} success, an instance of succeeding.
\B{Te\-ri lì'\-fya le\-Na'\-vi a tsa\-nu\-mul\-txa lo\-le\-i\-u sä\-flä!} The class about the Na'vi language was a success.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\ding{43}~\HL{flA}{\B{flä}}, \HL{tIflA}{\B{tì\-flä}})

% säfmi
\hi
\HT{sAfmi}{\B{sä\cp{fmi}}} \I{[s\ae."fmi]} \emph{n.} attempt. 
(\emph{a. proverb}) \B{Flä ke flä, ley sä\-fmi.} Whether you succeed or not, the attempt has value.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\ding{43}~\HL{fmi}{\B{fmi}}, \HL{tIfmi}{\B{tì\-fmi}})

% säfmong
\hi 
\HT{sAfmong}{\B{sä\cp{fmong}}} \I{[s\ae."fmoN]} \emph{n.} theft (\emph{particular instance}). 
\B{Poe\-ri sä\-fmong lor\-a tsa\-fkxi\-le\-yä lo\-lu na ay\-skxe mì te'\-lan.} For her, the theft of that beautiful necklace was like stones in her heart.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{fmong}{\B{fmong}})

% säfngo'
\hi
\HT{sAfngo'}{\B{sä\cp{fngo'}}} \I{[s\ae."fNoP]} \emph{n.} requirement, demand.   
\B{Nge\-yä fay\-sä\-fngo'\-ìl nì\-wotx stey\-ke\-rä\-ngi oe\-ti nì\-hawng.} All these demands of yours are making me exceedingly angry.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{fngo'}{\B{fngo'}})

% säfpìl
\hi
\HT{sAfpIl}{\B{sä\cp{fpìl}}} \I{[s\ae."fpIl]} \emph{n.} idea, thought.   
\B{Nge\-yä sä\-fpìl Saw\-tu\-te\-te\-ri hei\-ek oer nì\-txan.} Your idea about the Sky People is very interesting to me (because it seems unusual).~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Lor\-a ay\-lì\-'u, lor\-a ay\-sä\-fpìl.} Beautiful words and thoughts.~\LINK{http://masempul.org/2010/07/\%E2\%80\%99ite-polaheiem/}{sempul} 
\B{Saw\-tu\-te lu ay\-vrr\-tep nì\-wotx a sä\-fpìl\-it oel wä\-te.} I dispute the idea that the Skypeople are all demons.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Pol sä\-fpìl\-it ve\-rar wi\-van.} He's keeping his idea a secret.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
(\emph{a. metaphorically}) \B{ay\-sä\-fpìl a\-txukx}, deep thoughts.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{mek\-a sä\-fpìl}, an empty\slash dumb idea.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\emph{b. proverb}) \B{Sä\-fpìl a\-steng, tì\-kan a\-teng.} Great minds think alike [lit.: similar thought, same aim].~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{fpIl}{\B{fpìl}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{txinfpIl}{\B{txin\-fpìl}}, main point.
\on{2}.~\HL{ngrrfpIl}{\B{ngrr\-fpìl}}, key assumption.
\on{3}.~\HL{sAfpIlyewn}{\B{sä\-fpìl\-yewn}}, communicate.

% säfpìlyewn
\hi 
\HT{sAfpIlyewn}{\B{sä\cp{fpìl}yewn}} \I{[s\ae."fpIl.j$\cdot$Ewn]} \emph{vin., inf.}\on{33} communicate. 
\B{Txo po lu kak\-mok\-ri, fya\-pe sä\-fpìl\-yewn?} If he's mute, how does he communicate?~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{sAfpIl}{\B{sä\-fpìl}}, \HL{yewn}{\B{yewn}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tIsAfpIlyewn}{\B{tì\-sä\-fpìl\-yewn}}, communication.

% säfrìp
\hi
\HT{sAfrIp}{\B{sä\cp{frìp}}} \I{[s\ae."fRIp\textcorner]} \emph{n.} a bite.
\B{Lam fwa skxir za\-'u ta sä\-frìp nan\-tang\-ä.} It seemed the wound came from a bite of a viperwolf.~\LINK{http://naviteri.org/2020/04/awa-tskxekengtsyip-a-mikyunfpi-niul-one-more-little-listening-exercise/}{b}
(\ding{43}~\HL{frIp}{\B{frìp}})

% säfrrfen
\hi 
\HT{sAfrrfen}{\B{sä\cp{frr}fen}} \I{[s\ae."f\s{r}.fEn]} \emph{n.} a visit, an instance of visiting. 
\B{Fu\-ria nga zo\-la\-'u ira\-yo; ngey sä\-frr\-fen txa\-su\-nu oer.} Thank you for coming; I enjoyed your visit very much.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
(\ding{43}~\HL{frrfen}{\B{frr\-fen}})

% säftxulì'u
\hi
\HT{sAftxulI'u}{\B{säftxu\cp{lì}'u}} \I{[s\ae.ft'u."lI.Pu]} \emph{n.} speech, oration.   
\B{Sä\-ftxu\-lì\-'u Tsyeyk\-ä Na'\-vi\-ru slan\-ti\-re so\-li nì\-wotx.} Jake's speech inspired all the People.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Fni\-vu! Sä\-ftxu\-lì\-'u\-ri ke tsun oe sti\-vawm.} Hush! I can't hear the speech.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Su\-nu oe\-ru nì\-txan ay\-sä\-ftxu\-lì\-'u pe\-yä.} I like his speeches a lot.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{ftxulI'u}{\B{ftxu\-lì\-'u}})

% säfyep
\hi
\HT{sAfyep}{\B{sä\cp{fyep}}} \I{[s\ae."fjEp\textcorner]} \emph{n.} handle (\emph{the part of an object where one holds it, e.g. the handle of a cup}).
(\ding{43}~\HL{fyep}{\B{fyep}})

% sähe'a
\hi 
\HT{sAhe'a}{\B{sä\cp{he}'a}} \I{[s\ae."hE.Pa]} \emph{n.} a cough, an instance of coughing. 
(\ding{43}~\HL{he'a}{\B{he\-'a}})

% sähena
\hi
\HT{sAhena}{\B{sä\cp{he}na}} \I{[s\ae."hE.na]} \emph{n.} container, vessel, carrier.
\B{Ay\-fol za\-mo\-lu\-nge aw\-ngar ay\-ri\-na'\-it fa sä\-he\-na a\-pxa.} They brought us the seeds in a large container.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{hena}{\B{he\-na}}) \\  
\on{1}.~\HL{paysena}{\B{pay\-se\-na}}, water container. 
\on{2}.~\HL{tutsena}{\B{tut\-se\-na}}, stretcher. 
\on{3}.~\HL{tstalsena}{\B{tstal\-se\-na}}, knife sheath. 
\on{4}.~\HL{swizawsena}{\B{swi\-zaw\-se\-na\slash zaw\-se\-na}}, quiver.

% säkahena
\hi 
\HT{sAkahena}{\B{säka\cp{he}na}} \I{[s\ae.ka."hE.na]} \emph{or coll.} \I{[ska."hE.na]} \emph{n.} means of transport, transporation device, vehicle (\emph{typically s.t. that can move large things, including people}).  
(\ding{43}~\HL{kahena}{\B{ka\-he\-na}}, \HL{tIkahena}{\B{tì\-ka\-he\-na}}; \HL{sAmunge}{\B{sä\-mu\-nge}})

% säkanom
\hi
\HT{sAkanom}{\B{sä\cp{ka}nom}} \I{[s\ae."ka.nom]} \emph{or coll.} \I{["{}ska.nom]} \emph{n.} something acquired, an acquisition, a possession.   
\B{Tì\-'efu\-mì oe\-yä, nge\-yä fì\-sä\-ka\-nom lu le\-hrr\-ap ul\-te tsun nga\-ti tì\-sraw sey\-ki\-vi.} In my opinion, this acquisition of yours is dangerous and can hurt you.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{kanom}{\B{ka\-nom}})

% säkeynven
\hi 
\HT{sAkeynven}{\B{säkeyn\cp{ven}}} \I{[s\ae.kEjn."vEn]} \emph{or coll.} \I{[skEjn."vEn]} \emph{n.} (\emph{a. physical}) step. 
\B{Rawng tsa\-slär\-ä ftu fì\-tsenge lu vol\-a skeyn\-ven nì\-'aw.} The entrance to the cave is only eight steps from here.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\emph{b. metaphorical}) step in a procedure.
(\ding{43}~\HL{keynven}{\B{keyn\-ven}}) 

% säkxange
\hi 
\HT{sAkxange}{\B{sä\cp{kxa}nge}} \I{[s\ae."k'a.NE]} \emph{or} \I{["{}sk'a.NE]} \emph{n.} a yawn. 
(\ding{43}~\HL{kxange}{\B{kxa\-nge}})

% sälang
\hi
\HT{sAlang}{\B{sä\cp{lang}}} \I{[s\ae."laN]} \emph{or coll.} \I{[slaN]} \emph{n.} expedition, investigation.
\B{Fì\-txe\-lel fngo' sä\-lang\-it a\-ta\-wrr\-pa.} This matter requires an external investigation.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
\B{Kum sä\-lang\-ä le\-yew\-la lä\-ngu. Ke ro\-lun aw\-ngal ke\-'ut.} The result of the investigation was, I'm sorry to say, disappointing. We found nothing.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{lang}{\B{lang}}, \HL{tIlang}{\B{tì\-lang}})

% sälatem
\hi
\HT{sAlatem}{\B{sä\cp{la}tem}} \I{[s\ae."la.tEm]} \emph{or coll.} \I{["{}sla.tEm]} \emph{n.} (instance of) change, edit, modification.
\B{Ay\-sä\-la\-tem\-it a za\-mo\-lu\-nge Saw\-tu\-tel nga fmi ki\-va\-ra. Lä\-ngu kel\-tsun.} You try to resist the changes brought by the Sky People. Sadly, that's impossible.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
\B{'On\-ìri tsko\-ä lu tì\-kin sä\-la\-tem\-ä a\-hì\-'i.} The form of the bow requires a small change.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{latem}{\B{la\-tem}}, \HL{tIlatem}{\B{tì\-la\-tem}})

% sälätxayn
\hi
\HT{sAlAtxayn}{\B{sälä\cp{txayn}}} \I{[s\ae.l\ae."t'ajn]} \emph{or coll.} \I{[sl\ae."t'ajn]} \emph{n.} defeat, an instance of defeat.
\B{Tsa\-sä\-lä\-txayn Na'\-vi\-ru srung so\-li nì\-'aw fte sli\-vu txur nì\-'ul.} That defeat only helped the People become stronger.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{lAtxayn}{\B{lä\-txayn}})

% säleymfe'
\hi 
\HT{sAleymfe'}{\B{säleym\cp{fe'}}} \I{[s\ae.lEjm."fEP]} \emph{or coll.} \I{[slEjm."fEP]} \emph{n.} complaint. 
\B{Ay\-sä\-leym\-fe'\-it pe\-yä yu\-ne rä\-'ä. Lu po ke\-ve\-'o\-tu nì\-'aw.} Don't listen to his complaints. He's just a troublemaker.~\LINK{https://naviteri.org/2025/03/conlang-adventure-and-more/}{b} 
(\ding{43}~\HL{leymkem}{\B{leym\-kem}})

% säleymkem
\hi 
\HT{sAleymkem}{\B{säleym\cp{kem}}} \I{[s\ae.lEjm.\-"kEm]} \emph{or coll.} \I{[slEjm.\-"kEm]} \emph{n.} protest, instance of protesting. 
\B{Eyk\-tan\-ìl nge\-yä sä\-leym\-kem\-it sto\-lawm ul\-te pa\-ye\-'un teyng\-ta ze\-ne fko pe\-hem si\-vi.} The leader has heard your protest and will decide what must be done.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{43}~\HL{leymkem}{\B{leym\-kem}})

% sälipx
\hi 
\HT{sAlipx}{\B{sä\cp{lipx}}} \I{[s\ae."lip']} \emph{or coll.} \I{[slip']} \emph{n.} drop ({\em of a liquid}). 
(\ding{43}~\HL{lipx}{\B{lipx}})

% sämewn
\hi 
\HT{sAmewn}{\B{sä\cp{mewn}}} \I{[s\ae."mEwn]} \emph{or coll.} \I{[smEwn]} \emph{n.} sigh, an instance of sighing. 
\B{Zo\-la\-'u ftu kä\-pxì num\-tseng\-vi\-yä sä\-mewn a\-wok.} From the back of the classroom came a loud sigh.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
(\ding{43}~\HL{mewn}{\B{mewn}}) 

% sämok
\hi
\HT{sAmok}{\B{sä\cp{mok}}} \I{[s\ae."mok\textcorner]} \emph{or coll.} \I{[smok\textcorner]} \emph{n.} a suggestion, a concrete instance of suggesting.
\B{Fe\-yä ay\-sä\-mok lu fe' nì\-wotx.} All of their suggestions are bad.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Oel sä\-mok\-ti nge\-yä txa\-krr\-fpo\-lìl.} I have considered your suggestion.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
\B{Nge\-yä sä\-mok\-ìri a\-ko\-sman sei\-yi oe ira\-yo.} Thanks for that excellent suggestion (of yours).~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\emph{a. fig.}) \B{sä\-mok a\-mek}, a useless suggestion.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{mok}{\B{mok}}, \HL{tImok}{\B{tì\-mok}})

% sämunge
\hi
\HT{sAmunge}{\B{sä\cp{mu}nge}} \I{[s\ae."mu.NE]} \emph{or coll.} \I{["{}smu.NE]} (\textsc{rn}: \B{sä\cp{mù}nge} \I{[sE."mU.NE]}) \emph{n.} transportation tool or device (\emph{general term for any object used to carry or transport something else}).
\B{Sìn skxir yem 'um\-tsat a ho\-le\-na li pol mì sä\-mu\-nge.} He put medicine, which he had already carried in his pouch, on the wound.~\LINK{http://naviteri.org/2020/04/awa-tskxekengtsyip-a-mikyunfpi-niul-one-more-little-listening-exercise/}{b}
(\ding{43}~\B{mu\-nge}, \B{sä\-he\-na}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{syusmung}{\B{syu\-smung}}, tray. 
\on{2}.~\HL{paysmung}{\B{pay\-smung}}, water carrier. 
\on{3}.~\HL{prrsmung}{\B{prr\-smung}}, baby carrier.

% sämyam
\hi
\HT{sAmyam}{\B{säm\cp{yam}}} \I{[s\ae{}m."jam]} \emph{n.} hug, embrace.   
\B{Säm\-yam\-ìl po\-ru wa\-yìn\-txu fu\-ta nga\-ta lo\-lu li txo\-a.} A hug will show him that you've already forgiven him.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{meyam}{\B{me\-yam}})

% sänrr
\hi
\HT{sAnrr}{\B{sä\cp{nrr}}} \I{[s\ae."n\s{r}]} \emph{or coll.} \I{[sn\s{r}]} \emph{n.} (\emph{a.}) glow, an instance of glowing.   
\B{Txep\-ìl tìng le\-fpom\-a sä\-nrr\-ti.} The fire gives a pleasant glow.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} 
(\emph{b.}) light source; lamp.
(\ding{43}~\HL{nrr}{\B{nrr}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tsmisnrr}{\B{tsmi\-snrr}}, bladder lantern, nectar lantern. 
\on{2}.~\HL{tawsnrrtsyIp}{\B{Taw\-snrr\-tsyìp}}, Alpha Centauri C.

% sänui
\hi
\HT{sAnui}{\B{sä\cp{nu}i}} \I{[s\ae."nu.i]} \emph{or coll.} \I{["{}snu.i]} \emph{n.} failure (particular instance of failure).
\B{Oeyk tsa\-tì\-snaytx\-ä lu apxa sä\-nui tì\-eyk\-tan\-ä.} That loss was caused by a great failure of leadership.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Fì\-sä\-nu\-it ze\-ne nga tswi\-va', ma 'i\-tan.} You must forget about this failure, my son.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{nui}{\B{nui}}, \HL{tInui}{\B{tì\-nui}})

% sänume / sänume si
\hi
\HT{sAnume}{\B{sä\cp{nu}me}} \I{[s\ae."nu.mE]} \emph{or coll.} \I{["{}snu.mE]} \textsc{i}. \emph{n.} teaching, instruction. 
\B{Ye'\-rìn 'ì\-yi\-'a sä\-nu\-me a tsa\-ri kll\-fro' oe.} My teaching responsibilities will soon end.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Tì\-flä la\-tem ìlä seyng\-a ftxey fkol sä\-nu\-met li\-vek fu\-ke.} Success depends on whether or not one follows instructions.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Nge\-yä te\-ri fay\-te\-le a ay\-sä\-nu\-me\-ri ngar ira\-yo sei\-yi ay\-oe nì\-wotx.} We all thank you for your teachings about these topics.~\LINK{http://forum.learnnavi.org/language-updates/combined-answers-1-\%28feb-16\%29/}{ln} \\
\textsc{ii}. \B{sä\cp{nu}me si} \I{[s\ae{}."nu.mE si]} \emph{or coll.} \I{["{}snu.mE si]} \emph{vin.} teach.
\B{Ma 'ite, aw\-nge\-yä fya\-'o\-ri ze\-ne nga sä\-nu\-me si\-vi po\-ru fte tsi\-vun pil\-vll\-txe sì ti\-vì\-ran nì\-ay\-oeng.} My daughter, you will teach him our way to speak and walk as we do.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki}
(\ding{43}~\HL{nume}{\B{nu\-me}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAnumvi}{\B{sä\-num\-vi}}, lesson.

% sänumvi
\hi
\HT{sAnumvi}{\B{sä\cp{num}vi}} \I{[s\ae."num.vi]} \emph{or} \I{[."nUm.]} \emph{or coll.} \I{["{}snum.vi]} \emph{n.} lesson.
\B{Ta\-fral kì\-syar a ay\-sä\-num\-vi sì ay\-lì'\-fya\-vi lu fyin.} That's why I'll teach lessons and expressions that are simple.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
\B{Maw\-krr la\-ye\-iu oer krr nì\-'ul fte ngi\-vop ay\-lì\-'ut sì tsay\-fne\-sä\-num\-vit a tsun fra\-por srung si\-vi fte ni\-vu\-me sì zi\-ve\-rok nì\-swey.} After that I will have more time to create words and the kinds of lessons that can help everyone best learn and remember.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
(\ding{43}~\HL{sAnume}{\B{sä\-nu\-me}})

% sängä'än
\hi
\HT{sAngA'An}{\B{sängä\cp{'än}}} \I{[s\ae.N\ae."P\ae{}n]} \emph{or coll.} \I{[sN\ae."P\ae{}n]} \emph{n.} bout of suffering, episode of depression.
\B{Nge\-yä tsa\-sä\-ngä\-'än\-tsyìp\-ì\-ri set fra\-wzo srak?} Have you recovered from being down for a while?~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Pe\-yä sä\-ngä\-'än\-ìri lu sä\-fe\-'ul le\-yew\-la nì\-txan.} The worsening of his depression is very disappointing.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\ding{43}~\HL{ngA'An}{\B{ngä\-'än}})

% sängop
\hi
\HT{sAngop}{\B{sä\cp{ngop}}} \I{[s\ae."Nop\textcorner]} \emph{or coll.} \I{[sngop\textcorner]} \emph{n.} creation.
\B{Fyo\-lup\-a ay\-sä\-ngop ay\-nge\-yä oe\-ru te\-ya si nì\-ngay.} Your exquisite creations fill me greatly with joy.~\LINK{http://naviteri.org/2014/12/ayngaru-tsinga-yoratu-may-i-present-the-four-winners/}{b}
(\ding{43}~\HL{ngop}{\B{ngop}}, \HL{tIngop}{\B{tì\-ngop}})

% säomum
\hi
\HT{sAomum}{\B{sä\cp{o}mum}} \I{[s\ae."{}o.mum]} \emph{or} \I{[.mUm]} (\textsc{rn}: \B{sä\cp{o}mùm} \I{[sE."{}o.mUm]}) \emph{countable n.} (piece of) information.   
\B{Fì\-sä\-omum txan\-krr lo\-lu oer.} I had this information for a long time.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
\B{Ay\-sä\-omum\-ìri le\-sar sei\-yi oe nga\-ru ira\-yo nì\-txan.} Thanks very much for (all) this useful information.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\ding{43}~\HL{omum}{\B{o\-mum}})

% säpenghrr
\hi 
\HT{sApenghrr}{\B{säpeng\cp{hrr}}} \I{[s\ae.pEN."h\s{r}]} \emph{or coll.} \I{[spEN."h\s{r}]} \emph{n.} warning. 
\B{Som\-wew\-ä tì\-'ul a ka 'Rr\-ta sä\-peng\-hrr lu aw\-nga\-ru nì\-wotx.} Global warming is a warning to us all.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{penghrr}{\B{peng\-hrr}}, \HL{sApllhrr}{\B{sä\-pll\-hrr}})

% säpllhrr
\hi 
\HT{sApllhrr}{\B{säpll\cp{hrr}}} \I{[s\ae.p\s{l}."h\s{r}]} \emph{or coll.} \I{[sp\s{l}."h\s{r}]} \emph{n.} warning. 
(\ding{43}~\HL{pllhrr}{\B{pll\-hrr}}; \HL{sApenghrr}{\B{sä\-peng\-hrr}})

% säplltxe
\hi
\HT{sAplltxe}{\B{säpll\cp{txe}}} \I{[s\ae.p\s{l}."t'E]} \emph{or coll.} \I{[sp\s{l}."t'E]} \emph{n.} statement.
\B{'Aw\-ve\-a sä\-pll\-txe nge\-yä lä\-ngu e\-yawr.} Your first statement is unfortunately correct.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/comment-page-1/\#comment-2614}{b}
\B{Sä\-pll\-txel kar\-yu\-ä ke to\-la\-rä\-nge tì\-pawm\-it kaw\-'it.} The teacher's statement in no way pertained to the question. Drat!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{plltxe}{\B{pll\-txe}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAplltxevi}{\B{sä\-pll\-txe\-vi}}, comment.

% säplltxevi
\hi
\HT{sAplltxevi}{\B{säpll\cp{txe}vi}} \I{[s\ae.p\s{l}."t'E.vi]} \emph{or coll.} \I{[sp\s{l}."t'E.vi]} \emph{n.} comment.
\B{Tsa\-lì\-'u\-ri alu yo\-ra'\-tu, nge\-yä sä\-pll\-txe\-vi lu txan\-tsan.} Your comment concerning the word \emph{yora'tu} is excellent.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/comment-page-1/\#comment-1267}{b}
\B{Ay\-nge\-yä ay\-sä\-pll\-txe\-vi\-ri sei\-yi oe ira\-yo nì\-txan ay\-nga\-ru nì\-wotx!} I thank you all very much for the comments!~\LINK{http://naviteri.org/2011/07/attributive-\%E2\%80\%9Ca\%E2\%80\%9D-and-truncated-style/}{b}
(\emph{a. fig.}) \B{mek\-a sä\-pll\-txe\-vi}, an insipid\slash thoughtless comment.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{sAplltxe}{\B{sä\-pll\-txe}})

% säpom
\hi
\HT{sApom}{\B{sä\cp{pom}}} \I{[s\ae."pom]} \emph{or coll.} \I{[spom]} \emph{n.} kiss.
(\ding{43}~\HL{pom}{\B{pom}})

% säpxor
\hi 
\HT{sApxor}{\B{sä\cp{pxor}}} \I{[s\ae."p'oR]} \emph{or coll.} \I{[sp'oR]}  \emph{n.} explosion. 
\B{Stì\-mawm sä\-pxor\-it! Tul!} (I) just heard an explosion! Run!~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{pxor}{\B{pxor}}) 

% särangal
\hi 
\HT{sArangal}{\B{sä\cp{ra}ngal}} \I{[s\ae."Ra.Nal]} \emph{or coll.} \I{["{}sRa.Nal]} \emph{n.} wish. 
\B{Fay\-sä\-ra\-ngal\-ìri lu oe nga\-hu zam mì zam.} Concerning these wishes I'm a hundred percent with you.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/\#comment-39760}{b}
(\ding{43}~\HL{rangal}{\B{ra\-ngal}}) 

% särawn
\hi
\HT{sArawn}{\B{sä\cp{rawn}}} \I{[s\ae."Rawn]} \emph{or coll.} \I{[sRawn]} \emph{n.} replacement, substitute, something that replaces something else.
\B{Fì\-pam\-tseo\-tu\-ri ke la\-yu ftue fwa run fkol sä\-rawn\-it a tam.} It won't be easy to find a satisfactory replacement for this musician.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{rawn}{\B{rawn}})

% särengop
\hi
\HT{sArengop}{\B{sä\cp{re}ngop}} \I{[s\ae."RE.Nop\textcorner]} \emph{or coll.} \I{["{}sRE.Nop\textcorner]} \emph{n.} design, a particular instance of designing.
\B{Fay\-sä\-re\-ngop\-it a\-vä' oe\-ru rä\-'ä wìn\-txu nì\-mun, ru\-txe. Ke su\-nu oer keng nì\-'it.} Please don't show me these ugly designs again. I don't like them one bit.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{rengop}{\B{re\-ngop}}, \HL{tIrengop}{\B{tì\-re\-ngop}})

% säro'a / säro'a si
\hi
\HT{sAro'a}{\B{sä\cp{ro}'a}} \I{[s\ae."Ro.Pa]} \emph{or coll.} \I{["{}sRo.Pa]} \textsc{i}. \emph{n.} feat, accomplishment, great deed. \\
\textsc{ii}. \B{sä\cp{ro}'a si} \I{[s\ae."Ro.Pa si]} \emph{or coll.} \I{["{}sRo.Pa si]} \emph{vin.} do great deeds.  
\B{Txan\-tstew sä\-ro\-'a si, fna\-we'\-tu ke si.} A hero does great deeds, a coward does not.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Txo ni\-vew fko sä\-ro\-'a si\-vi, ze\-ne nì\-'aw\-ve ve\-nga' li\-vu.} If you want to accomplish great things, you first have to be organized.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{ro'a}{\B{ro\-'a}})

% säseyto
\hi
\HT{sAseyto}{\B{säsey\cp{to}}} \I{[s\ae.sEj."to]} \emph{n.} butchering tool.
(\ding{43}~\HL{seyto}{\B{sey\-to}})

% säsìlpey
\hi
\HT{sAsIlpey}{\B{säsìl\cp{pey}}} \I{[s\ae.sIl."pEj]} \emph{n.} hope, a particular instance of hoping.
\B{Krr\-a wä\-tut tse\-'a, pe\-yä sä\-sìl\-pey a yo\-ra' 'o\-lìp.} When he saw his opponent, his hope of winning vanished.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{sIlpey}{\B{sìl\-pey}}, \HL{tsIlpey}{\B{tsìl\-pey}})

% säspaw
\hi 
\HT{sAspaw}{\B{sä\cp{spaw}}} \I{[s\ae."{}spaw]} \emph{n.} belief (\emph{particular instance}). 
\B{Tsa\-sä\-spaw a\-tsleng lu le\-hrr\-ap.} That false belief is dangerous.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
(\ding{43}~\HL{spaw}{\B{spaw}}; \HL{tIspaw}{\B{tì\-spaw}}) 

% säspxin
\hi
\HT{sAspxin}{\B{sä\cp{spxin}}} \I{[s\ae."{}sp'in]} \emph{n.} disease, sickness.
\B{Fu\-ria ke tsun tì\-kang\-kem si\-vi, pe\-yä sä\-spxin lu tì\-la\-'um nì\-'aw.} As for not being able to work, his illness is only a pretence.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b} 
\B{Fì\-sä\-spxin\-ìri nge\-yä ke lä\-ngu kea 'um\-tsa.} Unfortunately there is no medicine for this disease of yours.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{spxin}{\B{spxin}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAspxintsyIp}{\B{sä\-spxin\-tsyìp}}, minor ailment.

% säspxintsyìp
\hi
\HT{sAspxintsyIp}{\B{sä\cp{spxin}tsyìp}} \I{[s\ae."{}sp'in.\t{ts}jIp\textcorner]} \emph{n.} minor ailment. 
(\ding{43}~\HL{sAspxin}{\B{sä\-spxin}})

% säsrätx
\hi
\HT{sAsrAtx}{\B{sä\cp{srätx}}} \I{[s\ae."{}sR\ae{}t']} \emph{n.} annoyance.
\B{Nga fwe\-fwi nìl\-ke\-ftang a fì\-'u lu sä\-srätx a\-txan oe\-ru.} It's a major annoyance to me that you whistle continuously.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\emph{a. expression}) \B{Sra\-ke srätx (ngat) txo \ldots?} (\emph{with} ‹iv›) `Would you mind if \ldots?'
\B{Kea sä\-srätx kaw\-'it!} `Not at all!'~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{srAtx}{\B{srätx}})

% säsrese'a
\hi 
\HT{sAsrese'a}{\B{säsrese\cp{'a}}} \I{[s\ae.sRE.sE."Pa]} \emph{n.} prediction. 
\B{Ay\-nge\-yä sä\-sre\-se\-'a lo\-lu e\-yawr!} Your prediction has been correct.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/\#comment-30279}{b} 
(\ding{43}~\HL{srese'a}{\B{sre\-se'a}})

% säsrìn
\hi
\HT{sAsrIn}{\B{sä\cp{srìn}}} \I{[s\ae."{}sRIn]} \emph{n.} lent or borrowed thing.
\B{Oe\-ta a tsa\-sä\-srìn\-ìl tok pe\-seng\-it?} Where's the thing (you) borrowed from me?~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{srIn}{\B{srìn}})

% säsrung
\hi 
\HT{sAsrung}{\B{sä\cp{srung}}} \I{[s\ae."{}sRuN]} \emph{or} \I{[."{}sRUN]} \emph{n.} helper (\emph{inanimate}), s.t. that helps. 
\B{Oe\-yä tì\-tslam\-ìri tì\-oeyk\-tìng nge\-yä lo\-lu sä\-srung.} Your explanation helped my understanding.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}
(\ding{43}~\HL{srung}{\B{srung}}, \HL{srungsiyu}{\B{srung\-si\-yu}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{txawnulsrung}{\B{txaw\-nul\-srung}}, machine.

% sästarsìm
\hi
\HT{sAstarsIm}{\B{sä\cp{star}sìm}} \I{[s\ae."{}staR.sIm]} \emph{n.} collection (put together intentionally by a person). 
(i.) \B{sä\-star\-sìm syu\-lang\-ä}, bouquet (\emph{collection of flowers selected and put together intentionally by a person}).~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
(\ding{43}~\HL{starsIm}{\B{star\-sìm}})

% säsulìn
\hi
\HT{sAsulIn}{\B{sä\cp{su}lìn}} \I{[s\ae."{}su.lIn]} \emph{n.} hobby.   
\B{Oe tskxe\-keng si sä\-su\-lìn\-ur alu tsko swi\-zaw.} I practice my hobby, which is archery.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{sulIn}{\B{su\-lìn}})

% säsyep
\hi
\HT{sAsyep}{\B{sä\cp{syep}}} \I{[s\ae."{}sjEp\textcorner]} \emph{n.} trap.
(\ding{43}~\HL{syep}{\B{syep}})

% sätare
\hi
\HT{sAtare}{\B{sä\cp{ta}re}} \I{[s\ae."ta.RE]} \emph{or coll.} \I{["{}sta.RE]} \emph{n.} connection, relationship.   
\B{Nì\-ngay le\-iu oer sì sem\-pul\-ur sä\-ta\-re a\-sìl\-tsan.} Father and I really have a good relationship; it's nice.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{tare}{\B{ta\-re}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{starlI'u}{\B{star\-lì\-'u}}, adposition.
\on{2}.~\HL{sAtarenga'}{\B{sä\-ta\-re\-nga'}}, relevant, pertinent.

% sätarenga'
\hi 
\HT{sAtarenga'}{\B{sä\cp{ta}renga'}} \I{[s\ae."ta.RE.NaP]} \emph{or coll.} \I{["{}sta.RE.NaP]} \emph{adj.} relevant, pertinent. 
\B{Tsa\-sä\-pll\-txe\-vi\-ri a\-sä\-ta\-re\-nga' ira\-yo.} Thanks for that pertinent comment.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 
(\ding{43}~\HL{sAtare}{\B{sä\-ta\-re}}, \HL{-nga'}{\B{-nga'}}; \ding{214}~\HL{kesAtarenga'}{\B{ke\-sä\-ta\-re\-nga'}})

% sätarep
\hi 
\HT{sAtarep}{\B{sä\cp{ta}rep}} \I{[s\ae."ta.REp\textcorner]} \emph{or coll.} \I{["{}sta.REp\textcorner]} \emph{n.} a rescue, an incident of rescuing.  
(\ding{43}~\HL{tarep}{\B{ta\-rep}})

% sätaron
\hi
\HT{sAtaron}{\B{sä\cp{ta}ron}} \I{[s\ae."ta.Ron]} \emph{or coll.} \I{["{}sta.Ron]} \emph{n.} hunt.   
\B{Eyk Ka\-mun a fra\-lo lä\-ngu tsa\-sä\-ta\-ron vel\-ke nì\-wotx. Ta\-ron\-yut yom smar\-ìl!} Every time Kamun is in charge, the hunt is a mess. Everything goes wrong that can!~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Ta\-ron\-yul le\-hawm\-pam ska\-'a sä\-ta\-ron\-it.} A noisy hunter destroys the hunt.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{taron}{\B{ta\-ron}})

% sätsan'ul
\hi
\HT{sAtsan'ul}{\B{sä\cp{tsan}'ul}} \I{[s\ae."\t{ts}an.Pul]} \emph{or} \I{[.PUl]} \emph{n.} improvement, a specific instance of improving.
\B{Set pll\-txe nga nìl\-tsan nì\-ngay. Fì\-sä\-tsan\-'ul\-ìri nge\-yä lu oe\-ru sko har\-yu nrr\-a.} You speak really well now. As your teacher, I'm proud of your improvement.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\ding{43}~\HL{tsan'ul}{\B{tsan\-'ul}}, \HL{tItsan'ul}{\B{tì\-tsan\-'ul}})

% sätsawn
\hi 
\HT{sAtsawn}{\B{sä\cp{tsawn}}} \I{[s\ae."\t{ts}awn]} \emph{n.} harvest (\emph{particular instance}). 
\B{Fì\-zì\-sìt\-ä sä\-tsawn txan\-tsan lei\-u.} I'm pleased to say that this year's harvest was excellent.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} 
(\ding{43}~\HL{tsawn}{\B{tsawn}}) 

% sätseri
\hi 
\HT{sAtseri}{\B{sä\cp{tse}ri}} \I{[s\ae."\t{ts}E.Ri]} \emph{n.} observation, something noticed. 
\B{Sìl\-tsan\-a sä\-tse\-ri, sìl\-tsan\-a tì\-pawm.} Good observation, good question.~\LINK{http://naviteri.org/2020/08/hiia-vur-a-teri-mefalukantsyip-a-little-story-about-two-cats/\#comment-31473}{b} 
\B{Nge\-yä tsa\-sä\-tse\-ri\-ri a el\-tur tì\-txen si ira\-yo.} Thank you for that interesting observation of yours.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} 
(\ding{43}~\HL{tseri}{\B{tse\-ri}})

% sätsìsyì
\hi
\HT{sAtsIsyI}{\B{sä\cp{tsì}syì}} \I{[s\ae."\t{ts}I.sjI]} \emph{n.} a whisper.
\B{Po pa\-mll\-txe a krr, fra\-po tar\-mìng mik\-yun nì\-pxi, ta\-lu\-na mok\-ri lu sä\-tsì\-syì\-tsyìp.} When he spoke, everyone listened intently, because his voice was a tiny whisper.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{tsIsyI}{\B{tsì\-syì}})

% sätswayon
\hi
\HT{sAtswayon}{\B{sä\cp{tsway}on}} \I{[s\ae."\t{ts}waj.on]} \emph{n.} flight, an instance of flying.
\B{Sä\-tsway\-on a\-'aw\-ve tsa\-heyl\-ur ha\-sey si; ke tsun nga pi\-vey.} First flight seals the bond; you cannot wait.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Maw sä\-tsway\-on a\-yol ay\-oe kll\-po\-lä mì ta\-yo a lu ro\-fa kil\-van.} After a short flight we landed in a field beside the river.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{tswayon}{\B{tsway\-on}})

% sätsyìl
\hi
\HT{sAtsyIl}{\B{sä\cp{tsyìl}}} \I{[s\ae."\t{ts}jIl]} \emph{n.} climbing event, a climb.   
\B{Kin\-trr\-am\-ä sä\-tsyìl lu le\-hrr\-ap slä 'o' nì\-txan.} Last week's climb was dangerous but very exciting.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} 
\B{Tsa\-sä\-tsyìl lo\-lu ngä\-zìk nì\-ngay!} That was a really hard climb!~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{tsyIl}{\B{tsyìl}})

% sävll
\hi 
\HT{sAvll}{\B{sä\cp{vll}}} \I{[s\ae."v\s{l}]} \emph{n.} sign, signal, indication. 
\B{Kxe\-ner lu sä\-vll txep\-ä.} Smoke is a sign\slash an indication that there is fire.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
\B{Mì sa\-ngek a sä\-vll\-it ngo\-lop eyk\-yul tar\-po\-ngu\-ä.} The sign on the tree trunk was made by the leader of the hunting party.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
(\ding{43}~\HL{vll}{\B{vll}})

% säwäsul
\hi
\HT{sAwAsul}{\B{sä\cp{wä}sul}} \I{[s\ae."w\ae.sul]} \emph{or} \I{[.sUl]} \emph{or coll.} \I{["{}sw\ae.sul]} \emph{n.} a competition, a particular instance of competing.
(i.) \B{sä\-wä\-sul a tul}, foot race.~\LINK{http://naviteri.org/2012/10/wina-sawasultsyip-ahii-a-quick-little-contest/}{b}
(\ding{43}~\B{wä\-sul}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAwAsultsyIp}{\B{sä\-wä\-sul\-tsyìp}}, contest.

% säwäsultsyìp
\hi
\HT{sAwAsultsyIp}{\B{sä\cp{wä}sultsyìp}} \I{[s\ae."w\ae.sul.\t{ts}jIp\textcorner]} \emph{or} \I{[.sUl.]} \emph{or coll.} \I{["{}sw\ae.sul.\t{ts}jIp\textcorner]} \emph{n.} contest.
\B{Sìl\-pey oe, fì\-sä\-wä\-sul\-tsyìp 'o' lì\-ye\-vu ay\-nga\-ru!} I hope, this little contest will be fun for you!~\LINK{http://naviteri.org/2012/10/wina-sawasultsyip-ahii-a-quick-little-contest/}{b}
\B{Win\-a Sä\-wä\-sul\-tsyìp A\-hì\-'i.} A Quick Little Contest.~\LINK{http://naviteri.org/2012/10/wina-sawasultsyip-ahii-a-quick-little-contest/}{b}
(\ding{43}~\HL{sAwAsul}{\B{sä\-wä\-sul}})

% säwäte
\hi
\HT{sAwAte}{\B{säwä\cp{te}}} \I{[s\ae.w\ae."tE]} \emph{or coll.} \I{[sw\ae."tE]} \emph{n.} point of contention, source of argument, thing disputed. 
(\ding{43}~\HL{wAte}{\B{wä\-te}}, \HL{tIwAte}{\B{tì\-wä\-te}})

% säwem
\hi 
\HT{sAwem}{\B{sä\cp{wem}}} \I{[s\ae."wEm]} \emph{or coll.} \I{[swEm]} \emph{n.} fight. 
\B{Fì\-sä\-wem\-ìri ze\-ne aw\-nga kawl hä\-pi\-vawl. Lu Saw\-tu\-te wä\-tu a\-pxul.} We must prepare diligently for this fight. The Skypeople are formidable opponents.~\LINK{http://naviteri.org/2020/08/hiia-vur-a-teri-mefalukantsyip-a-little-story-about-two-cats/}{b}
(\ding{43}~\HL{wem}{\B{wem}})

% säwìntxu
\hi
\HT{sAwIntxu}{\B{säwìn\cp{txu}}} \I{[s\ae.wIn."t'u]} \emph{or coll.} \I{[swIn."t'u]} \emph{n.} a showing, an exhibition.
(\ding{43}~\HL{wIntxu}{\B{wìn\-txu}})

% säwìngay
\hi
\HT{sAwIngay}{\B{säwì\cp{ngay}}} \I{[s\ae.wI."Naj]} \emph{or coll.} \I{[swI."Naj]} \emph{n.} proof (\emph{particular instance}).
\B{Tsa\-sä\-pll\-txe\-ri sä\-wì\-ngay a to\-lìng ngal lu meyp.} The proof you gave of that statement is weak.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{wIngay}{\B{wì\-ngay}})

% säyäkx
\hi
\HT{sAyAkx}{\B{sä\cp{yäkx}}} \I{[s\ae."j\ae{}k']} \emph{or coll.} \I{[sj\ae.k']} \emph{n.} snub.
\B{Fì\-sä\-yäkx\-it ay\-oel ke tswa\-ya'.} We will not forget this snub.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{yAkx}{\B{yäkx}})

% säyìm
\hi 
\HT{sAyIm}{\B{sä\cp{yìm}}} \I{[s\ae."jIm]} \emph{or coll.} \I{[sjIm]} \emph{n.} tie, something used for binding. 
(\ding{43}~\HL{yIm}{\B{yìm}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{syImawnun'i}{\B{syì\-maw\-nun\-'i}}, cut tie. 

% säzärìp
\hi
\HT{sAzArIp}{\B{sä\cp{zä}rìp}} \I{[s\ae."z\ae.RIp\textcorner]} \emph{n.} direhorse lead, rein.
(\ding{43}~\HL{za'ArIp}{\B{za\-'ä\-rìp}}) 

% se'ayl
\hi
\HT{se'ayl}{\B{se\cp{'ayl}}} \I{[sE."Pajl]} \emph{countable n.} kind of waterfall, \emph{an individual tall, thin waterfall that pours down a sheer high cliff or off of a floating mountain}.

% sekrr
\hi
\HT{sekrr}{\B{\cp{se}krr}} \I{["{}sE.k\s{r}]} \emph{n.} present (\emph{temporal}). 
(\ding{43}~\HL{set}{\B{set}}, \HL{krr}{\B{krr}})

% semkä
\hi 
\HT{semkA}{\B{sem\cp{kä}}} \I{[sEm."k\ae]} \emph{n.} bridge.  
(\ding{43}~\HL{sA'o}{\B{sä\-'o}}, \HL{emkA}{\B{em\-kä}}; \HL{emkAfya}{\B{em\-kä\-fya}})

% sempu
\hi
\HT{sempu}{\B{\cp{sem}pu}} \I{["{}sEm.pu]} \emph{n.} daddy. 
\B{Ma sem\-pu, oey yom\-yo tso\-lawng!} Daddy, my plate broke!~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
\B{Sem\-pu\-ti rä\-'ä srätx, ma yawn\-tu\-tsyìp. Tì\-kang\-kem se\-ri.} Don't bother daddy, little one. He's working.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}
(\ding{43}~\HL{sempul}{\B{sem\-pul}})

% sempul
\hi
\HT{sempul}{\B{\cp{sem}pul}} \I{["{}sEm.pul]} \emph{or} \I{[.pUl]} \emph{n.} father. 
\B{Ma Sem\-pul, pxay\-a a\-to\-ki\-ri\-na' ne fì\-ke\-tu\-wong zì\-ma\-'u. Ze\-ne Sa'\-nok fì\-au\-ngia\-ti ral\-pi\-veng.} Father, many \emph{atokirina} just came to this alien. Mother must interpret this sign.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/}{a\textsuperscript{c}\slash ln}
\B{Tì\-re\-ngop\-ìri ioi\-yä lu sem\-pul pe\-yä tsul\-fä\-tu.} Her father is a master designer of ceremonial adornments.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Fì\-tsko\-ti a\-fyo\-lup 'o\-ngo\-lop oe\-yä sem\-pul\-ìl.} This exquisite bow was designed by my father.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Nì\-ngay le\-iu oer sì sem\-pul\-ur sä\-ta\-re a\-sìl\-tsan.} Father and I really have a good relationship; it's nice.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Txe\-wì\-yä sem\-pul\-ìri sngä\-'i\-krr Saw\-tu\-tet ve\-re'\-kì.} In the beginning Txewì's father hated the Sky People.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Tsa\-ri\-a fkol po\-le\-'un fu\-ta Lo\-ak slu ta\-ron\-yu, sem\-pul 'efu ye.} Father is content that it's been decided Loak will be a hunter.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Na\-ri si!  Tsa\-tsko lu spu\-win ul\-te ke lu mi txur, slä oe\-ri txan\-wa\-we le\-iu. Lu stxe\-li a sem\-pul\-ta.} Careful! That bow is old and no longer strong, but it means a lot to me. It was a gift from my father.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(i.) \B{Sem\-pul Lì'\-fya\-yä}, father of the language, Dr.~Paul Frommer.~\LINK{http://forum.learnnavi.org/language-updates/a-long-answer-to-a-quick-question/msg155104/\#msg155104}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sempul}{\B{sem\-pu}}, daddy. 
\on{2}.~\HL{sa'sem}{\B{sa'\-sem}}, parent, parents.

% set
\hi
\HT{set}{\B{set}} \I{[sEt\textcorner]} \emph{adv.} now.   
\B{'Aw\-pot set ftxey ay\-ngal a lu ay\-nga\-kip.} Choose one among you now.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
\B{'Aw\-a tì\-pawm\-ì\-ri 'i\-veyng oe set;  ay\-la\-ri zu\-saw\-krr 'ay\-eyng.} Let me answer one of the questions now; the others (I'll) answer in the future.~\LINK{http://forum.learnnavi.org/language-updates/email-from-frommer-re-numbers-\%28or-the-full-number-chart!\%29/}{ln}
\B{Fo\-ru 'upxa\-ret oel fpo\-le', slä vay set ke pa\-mä\-hä\-ngem kea tì\-'eyng.} I have sent them a message, but up to now no answer has arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Lì'\-fya\-ri le\-Na'\-vi 'Rr\-ta\-mì, vay set 'al\-mong a fra\-'u ze\-ra\-'u ta ngrr\-po\-ngu.} Everything that has gone on with (blossomed regarding) Na'vi until now on Earth has come from a grassroots movement.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sekrr}{\B{se\-krr}}, present, present time. 
\on{2}.~\HL{pxiset}{\B{pxi\-set}}, right now.

% sevin
\hi
\HT{sevin}{\B{se\cp{vin}}} \I{[sE."vin]} \emph{adj.} pretty (\emph{primarily of females}).
\B{Lu poe se\-vin nì\-ftxan kum\-a yaw\-ne slo\-lu oer.} She was so beautiful that I fell in love with her.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Se\-vin\-a tsa\-'eve\-ru a\-hì\-'i me\-sa'\-sem ioi so\-li fa mik\-tsang.} The parents put earrings on that pretty little girl.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Nga\-ri tswin\-tsyìp se\-vin nì\-txan lu nang!} What a pretty little queue you have!~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\ding{43}~\HL{sayrIp}{\B{say\-rìp}}, \HL{lor}{\B{lor}})

% sey
\hi
\HT{sey}{\B{sey}} \I{[sEj]} \emph{n.} cup, bowl (\emph{minimally modified from naturally occurring resources}).
\B{Fì\-srä\-ti pxaw sey mi\-vam fte tsat hi\-vaw\-nu.} Wrap this cloth around the bowl to protect it.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
\B{Fwa pon sey\-ti sìn ki\-nam\-til lu le\-hrr\-ap, ma 'itan.} Balancing a cup on your knee is dangerous, son.~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}  \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{'e'insey}{\B{'e\-'in\-sey}}, drinking gourd. 
\on{2}.~\HL{sumsey}{\B{sum\-sey}}, drinking vessel (from shell). 
\on{3}.~\HL{swoasey}{\B{swoa\-sey}}, kava bowl.

% seykxel
\hi
\HT{seykxel}{\B{sey\cp{kxel}}} \I{[sEj."k'El]} \emph{adj.} strong: inner strength, a quiet feeling of confidence in one's own capability.   
(\emph{a. conv.}) \B{Sey\-kxel sì nit\-ram!} ``Congratulations!'' [lit.: strong and happy] (\emph{for congratulating s.o. on their good fortune}).~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}

% seyn
\hi
\HT{seyn}{\B{seyn}} \I{[sEjn]} \emph{n.} chair, stool, bench; \emph{any tool or device to facilitate sitting}. 
\B{Lä\-ngu fì\-seyn kel\-ho\-an nì\-ngay. Tsun oe hi\-veyn tsaw\-sìn 'a\-'aw\-a swaw\-tsyìp nì\-'aw.} This chair is really uncomfortable. I can only sit on it a few seconds.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
(\ding{43}~\HL{sA}{\B{sä-}}, \HL{heyn}{\B{heyn}})

% seyri
\hi
\HT{seyri}{\B{\cp{sey}ri}} \I{["{}sEj.Ri]} \emph{n.} lip. 
\B{Nge\-yä me\-sey\-ri lor lu nì\-txan.} Your lips are very beautiful.~\LINK{https://www.youtube.com/watch?v=ojr_BE8ZmVE}{ted}

% seysonìltsan
\hi
\HT{seysonIltsan}{\B{seysonìl\cp{tsan}}} \I{[sEj.so.nIl."\t{ts}an]} \emph{conv.} ``Well done!'' (\emph{from} \B{ha\-sey so\-li nìl\-tsan})
\B{Nge\-yä tì\-kang\-kem\-ìri 'e\-fe\-iu oe ye nì\-txan. Sey\-so\-nìl\-tsan!} I'm very satisfied with your work. Well done!~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
(\ding{43}~\HL{hasey}{\B{ha\-sey si}}, \HL{nIltsan}{\B{nìl\-tsan}})

% seyto
\hi
\HT{seyto}{\B{sey\cp{to}}} \I{[sEj."to]} \emph{vtr.} butcher (\emph{i.e. separating or processing the carcass of a dead animal}).
\B{Aw\-ngal fì\-ye\-rik\-it nì\-win si\-vey\-to ko.} Let's cut up this hexapede quickly.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAseyto}{\B{sä\-sey\-to}}, butchering tool.

% seyä
\hi
\HT{seyA}{\B{\cp{se}yä}} \I{["{}sE.j\ae]}~:~\HL{'u}{\B{'u}} (\emph{shortened genitive form of the plural of} \HL{tsa'u}{\B{tsa\-'u})}.

% seze
\hi
\HT{seze}{\B{\cp{se}ze}} \I{["{}sE.zE]} \emph{n.} (\emph{a. \ding{96}}) blue flower. 
\B{Na se\-ze | A 'ong ne tsaw\-ke.} Like the blue\hyp{}flower | Which unfolds to the sun.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} 
(\emph{b. proper name}) \emph{name of Neytiri's banshee}.

% si
\hi
\HT{si}{\B{si}} \I{[si]} \emph{unprod. auxiliary v. to turn a n. or adj. into a v.:} \HL{tIkangkem}{\B{tì\-kang\-kem si}}, to work. \HL{win}{\B{win si}}, to hurry. \emph{Grammatically vin., the logical direct object takes the dative case:}  
\B{Nì\-'i'\-a po tsa\-tì\-kang\-kem\-vir ha\-sey so\-li.} She finally completed the project.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} 
\emph{Follows the non\hyp{}verbal element; with} \B{ira\-yo si} \emph{word order can be broken:} 
\B{Oel ay\-nga\-ti ka\-me\-i\-e, ma oe\-yä ey\-lan, ul\-te ay\-nga\-ru sei\-yi ira\-yo.} I \textsc{see} you all, my friends, and thank you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\emph{Can stand alone when the non\hyp{}verbal element has already been mentioned and is understood from context:}
\B{Nga tsap\-'alu\-te so\-li srak? -- Soli.} Did you apologize? -- (Yes,) I have.~\LINK{http://forum.learnnavi.org/language-updates/auxilary-verb-si-possessive-dative-krr/}{ln}
\emph{Negation precedes the auxiliary v.:} 
\B{Swey lu fwa nga fì\-kem ke si\-vi nì\-sngum.} It's best that you not do this fretfully.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Txo\-pu rä\-'ä si, lu ke\-tu\-wong\-o nì\-'aw.} Don't be afraid, it's only an alien.~\LINK{http://forum.learnnavi.org/language-updates/a-collection/}{ln\slash a\textsuperscript{c}} \\
\emph{Special forms in} ‹\HL{Ap}{\B{äp}}› \emph{and} ‹\HL{eyk2}{\B{eyk}}› \emph{exist, e.g.:} \ding{43} \HL{ioi}{\B{ioi sä\-pi}}, adorn oneself. \HL{win}{\B{win sä\-pi}}, hurry oneself. \HL{tIsraw}{\B{tì\-sraw sey\-ki}}, hurt s.o. \B{sä\-pi} \emph{and its derivations is often shortened to} \I{[spi]}.

% sil
\hi 
\HT{sil}{\B{sil}}\textsuperscript{1} \I{[sil]} \emph{n.} body stripe (\emph{non\hyp{}facial}). 
(\emph{a. proverbial expression}) \B{Po\-ri fra\-pil sì fra\-sil lor lu nì\-txan.} He\slash She is very attractive.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b} 
\B{Fra\-pil fra\-sil!} `Wow! Check him\slash her out! He\slash she is gorgeous!'~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\ding{43}~\HL{pil}{\B{pil}})

\hi
\B{sil}\textsuperscript{2} \I{[sil]}~:~\HL{til}{\B{til}}

% sim
\hi
\HT{sim}{\B{sim}}\textsuperscript{\on{1}} \I{[sim]} \emph{vin.} near, be near. 
\B{Oe 'o\-long\-okx mì sray a txam\-pay\-ìri sim.} I was born in a town near the ocean.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{asim}{\B{asim}}, near by, at close range.

\hi
\B{sim}\textsuperscript{\on{2}} : \HL{tsim}{\B{tsim}}

% sì
\hi
\HT{sI}{\B{sì}} \I{[sI]} \emph{conj., suf.} and (\emph{for enumeration}).
\B{Ay\-ey\-lan\-ur oe\-yä sì ey\-lan\-ur lì'\-fya\-yä le\-Na'\-vi nì\-wotx.} To all my friends and friends of the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Lu nga win sì txur.} You are fast and strong.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Tsay\-fne\-sä\-num\-vi a tsun fra\-por srung si\-vi fte ni\-vu\-me sì zi\-ve\-rok nì\-swey.} The kinds of lessons that can help everyone best learn and remember.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
(\emph{a. can be used as an enclitic})
\B{Ta 'ey\-lan kar\-yu\-sì ay\-nge\-yä}. From your friend and teacher.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} I will then decide the best way to bring you the words and rules of Na'vi.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Fay\-upxa\-re la\-yu ay\-sngä\-'i\-yu\-fpi, fte lì'\-fya\-ri aw\-nge\-yä fo tsì\-ye\-vun nì\-ftue nìl\-tsan\-sì ni\-vume.} These messages will be for beginners, so that they can learn about our language easily and well.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
(\emph{b. often merges in coll. pronunciation}) \B{ay\-nan\-tang sì ay\-ri\-ti} \emph{pronounced:} \I{[aj."nan.taN saj."Ri.ti]}, viperwolfs and stingbats.~\LINK{http://naviteri.org/2010/07/thoughts-on-ambiguity/}{b}
(\emph{c. idiom}) \B{Kxe\-tse sì mik\-yun kop pll\-txe.} Body language is important [lit.: The tail and ears also speak].~\LINK{http://forum.learnnavi.org/language-updates/grid-tidbits/}{ln}
(\ding{43}~\HL{ulte}{\B{ul\-te}})

% sìk
\hi
\HT{sIk}{\B{sìk}} \I{[sIk\textcorner]} \emph{part. to indicate the end of a quote with verbs of speech} \HL{plltxe}{\B{pll\-txe}}, \HL{peng}{\B{peng}}, \HL{pllngay}{\B{pll\-ngay}}, \HL{pawm}{\B{pawm}}, \emph{and writing} \HL{pamrel}{\B{pam\-rel si}} \emph{when the speaker continues to say s.t. after the cited quote; used together with} \ding{43}~\HL{san}{\B{san}}.
\B{Po\-lawm po san sra\-ke Sä\-li ho\-lum sìk.} He asked, ``Did Sally leave?''~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/}{ln}
\B{Pol\-txe Ey\-tu\-kan san oe ka\-yä sìk, slä oel pot ke spaw.} Eytukan said he would go but I don't believe him.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}

% sìlpey
\hi
\HT{sIlpey}{\B{sìl\cp{pey}}} \I{[sIl."p$\cdot$Ej]} \emph{vin., inf.}\on{22} hope (\emph{used optionally with} \HL{tsnI}{\B{tsnì}}, \emph{often with} ‹\B{iv}›), \B{sìl\-pey oe}, I hope, hopefully.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Sìl\-pey oe, la\-yu oe\-ru ye'\-rìn sìl\-tsan\-a fmawn.} I hope, I will soon have good news.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Ul\-te sìl\-pey oe, fay\-sì\-oeyk\-tìng li\-vu law nì\-wotx!} And I hope, all these explanations are clear.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Fu\-ri\-a fì\-lì'\-fya\-vit fkol ta 'Ìng\-lì\-sì mo\-lu\-nge, sìl\-pey oe tsnì ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ke stì\-ye\-vi.} That this expression is brought from English, will hopefully not anger the members of the language group.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
(\ding{43}~\HL{sIltsan}{\B{sìl\-tsan}}, \HL{pey}{\B{pey}}; \ding{214}~\HL{fe'\-pey}{\B{fe'\-pey}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIsIlpey}{\B{nì\-sìl\-pey}}, hopefully. 
\on{2}.~\HL{tsIlpey}{\B{tsìl\-pey}}, hope (abstract). 
\on{3}.~\HL{sAsIlpey}{\B{sä\-sìl\-pey}}, a hope.

% sìlronsem
\hi
\HT{sIlronsem}{\B{sìl\cp{ron}sem}} \I{[sIl."Ron.sEm]} \emph{adj.} (\emph{of things}) clever, smart. 
\B{Ay\-lì\-'u\-van a\-swey lu 'i\-pu, lu sìl\-ron\-sem.} The best puns are both funny and clever.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{sIltsan}{\B{sìl\-tsan}}, \HL{ronsem}{\B{ron\-sem}})

% sìltsan
\hi
\HT{sIltsan}{\B{sìl\cp{tsan}}} \I{[sIl."\t{ts}an]} \emph{adj.} good.   
\B{Tì\-'e\-fu\-mì oe\-yä su\-te\-kip nì\-wotx, ftxey Na'\-vi ftxey Saw\-tu\-te, lu sìl\-tsan lu kawng.} In my opinion among all people, whether Na'vi or Skypeople, there are good and bad.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Sìl\-pey oe, la\-yu oe\-ru ye'\-rìn sìl\-tsan\-a fmawn.} I hope, I will soon have good news.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Slä lu oe\-ru fmawn\-o a\-sìl\-tsan.} But I have some good news.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Po to oe lu sìl\-tsan.} S\slash he is better than me.~\LINK{http://forum.learnnavi.org/language-updates/prrwll-moss-the-comparative-to-construct/}{ln} 
(\ding{43}~\HL{swey}{\B{swey}}; \ding{214}~\HL{fe'}{\B{fe'}}, \HL{kawng}{\B{kawng}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIltsan}{\B{nìl\-tsan}}, well, in a good way. 
\on{2}.~\HL{tsIltsan}{\B{tsìl\-tsan}}, good, goodness.
\on{3}.~\HL{tsantu}{\B{tsan\-tu}}, good person.
\on{4}.~\HL{tsantxAl}{\B{tsan\-txäl}}, invitation.

% sìn
\hi
\HT{sIn}{\B{sìn}}\textsuperscript{\on{1}} \I{[sIn]} \emph{adp.} (\emph{a. locally}) on, onto.
\B{Ay\-wayl yìm ki\-fkey\-ä | 'Ì\-he\-yut a\-vo\-mrr | Sìn ti\-rea\-fya\-'o a\-vol | Na way\-te\-lem\-ä hìng.} The songs bind | the thirteen spirals of the solid world | To the eight spirit paths | Like the threads of a Songcord.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Sre fwa sìn tskxe\-pay tì\-ran, ze\-ne fko flì\-nutx\-it sti\-ve\-ftxaw.} It's necessary to check the thickness of the ice before walking on it.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay.} Let's meet on the hill tomorrow at nightfall.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Ya\-mì tsun fko tsi\-ve\-'a lo\-ran\-it re\-nu\-ä kil\-van\-ä slä kll\-te\-sìn wä\-pan.} From the air you can see the grace of the river's form but from the ground it's hidden.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Ngey tsko\-ti yem tsa\-yì\-sìn tsa\-krr za\-'u fì\-tseng!} Put your bow on that ledge and come here!~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Oe\-ri ay\-som\-pì\-va sìn re\-'o var zi\-vup.} Raindrops keep falling on my head.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
(\emph{b. with} \HL{fAkA}{\B{fäkä}}) get on, mount. \B{Po sìn pa'\-li fä\-ko\-lä mak\-to ne\-to.} She got on her direhorse and rode away.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
(\emph{c. metaphorically with} \HL{kllkxem}{\B{kll\-kxem}} \emph{and} \HL{yI}{\B{yì}} \emph{to refer to the level of anything scalable: water level, temperature, talent, anger, etc.}) 
\B{Lì'\-fya\-ri po (kll\-kxem) sìn pe\-yì?} [\emph{coll.}: \I{[sIm.pE."jI]}]\slash \B{Lì'\-fyari po pe\-yì?} How good is her Na'vi?~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Sìn yì sngä\-'i\-yu\-ä, slä tsye\-rìl (hay\-a yì\-ne) nì\-'ul\-'ul.} They're at a beginner's level, but they're getting better and better.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Sìn yì a ke tsun kaw\-tu spi\-vaw, nì\-win fra\-to.} He runs at an incredible level, faster than anyone else.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}

\hi
\B{sìn}\textsuperscript{\on{2}} : \B{tìn}

% -sìng
\hi
\HT{sIng}{\B{-sìng}} \I{[.sIN]} \emph{num., suf.} four (\emph{rest form of \ding{43}~\HL{tsIng}{\B{tsìng}} to form higher numbers}). \B{vo\-sìng}, \on{12} (decimal), \on{14} (octal). \B{me\-vo\-sìng}, \on{20} (decimal), \on{24} (octal); etc.

% ska'a
\hi
\HT{ska'a}{\B{ska\cp{'a}}} \I{[ska."Pa]} \emph{vtr.} destroy.
\B{He\-tu\-wong\-ìl aw\-nge\-yä swo\-tut ska\-'a, fte kll\-ki\-vu\-lat ke\-ru\-sey\-a tskxet.} The aliens destroy our sacred place to dig up dead rock.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln\slash a\textsuperscript{c}}
\B{Zo\-la\-'u saw\-tu\-te fì\-tseng fte aw\-nga\-ti ski\-va\-'a.} The Skypeople have come to destroy us.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Maw fwa fkol Kel\-ut\-ral\-it sko\-la\-'a, lem\-rey\-a ha\-pxì\-tul tsa\-soai\-ä txo\-lu\-la mip\-a kel\-ku\-ti mì ta\-yo.} After Hometree was destroyed, the surviving members of that family built a new home on the plains.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Lu oer sngum\-tsim a pol Ame\-ri\-kat ski\-ye\-va\-'a.} My source of worry is that he will destroy America.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIska'a}{\B{tì\-ska\-'a}}, destruction.

% skepek
\hi
\HT{skepek}{\B{\cp{ske}pek}} \I{["{}skE.pEk\textcorner]} \emph{adj.} formal. 
\B{Ay\-lì\-'u ngi\-an nì\-'it ske\-pek lu.} However the words are a bit formal.~\LINK{http://forum.learnnavi.org/language-updates/more-grammatical-tidbits/}{ln\slash a}

% skeykx
\hi 
\HT{skeykx}{\B{skeykx}} \I{[skEjk']} \emph{adj.} bored. 
\B{Oe 'e\-fu skeykx ul\-te new ti\-vä\-txaw ne kel\-ku.} I'm bored and I want to go home.~\LINK{http://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
(\ding{214}~\HL{eltu}{\B{eltu si}})

% skiempa
\hi
\HT{skiempa}{\B{\cp{ski}empa}} \I{["{}ski.Em.pa]} \emph{n.} right side.
\B{Mìn ne ski\-em\-pa.} Turn to the right.~\ding{43}~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln}
(\ding{43}~\HL{skien}{\B{ski\-en}}, \HL{pa'o}{\B{pa\-'o}}; \ding{214}~\HL{ftArpa}{\B{ftär\-pa}})

% skien
\hi
\HT{skien}{\B{\cp{ski}en}} \I{["{}ski.En]} \emph{adj.} right (\emph{not left, direction}). 
\B{Oe\-ri ski\-en\-a tsyokx lu le\-skxir.} My right hand is wounded.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{214}~\B{ftär}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{skiempa}{\B{ski\-em\-pa}}, right side. 
\on{2}.~\HL{nIskien}{\B{nì\-ski\-en}}, to the right.

% sko
\hi
\HT{sko}{\B{sko}}\textsuperscript{\on{1}} \I{[sko]} \emph{adp.}\texttt{+} (\emph{a}) as, in the capacity of, in the role of.   
\B{Sko Sa\-hìk ke tsun oe mì\-ftxe\-le tsngi\-vaw\-vìk, sko sa'\-nok tsun.} As Tsahik, I cannot weep over this matter, as mother I can.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Set pll\-txe nga nìl\-tsan nì\-ngay. Fì\-sä\-tsan\-'ul\-ìri nge\-yä lu oe\-ru sko har\-yu nrr\-a.} You speak really well now. As your teacher, I'm proud of your improvement.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{'Eveng\-ìl skxe\-vit 'ey\-krr\-ko sko uvan.} Children roll pebbles as a game.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\emph{b}) \emph{to convey the meaning of} `I myself, you yourself' etc. 
\B{Nga sko nga pol\-txe san rä\-'ä ki\-vä!\slash Nga nga\-sko pol\-txe …} You yourself said don't go!~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}

% sko
\hi
\B{sko}\textsuperscript{\on{2}} : \B{tsko}.

% skuka
\hi 
\HT{skuka}{\B{\cp{sku}ka}} \I{["{}sku.ka]} \emph{n., fauna} Sagittaria, \emph{cephalopod\hyp{}like aquatic creature with a hard shell and long tentacles}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln}

% skxakep
\hi
\HT{skxakep}{\B{\cp{skxa}kep}} \I{["{}sk'a.kEp\textcorner]} \textsc{i}. \emph{adj.} probable. \\ 
\textsc{ii}. \emph{adv.} probably. 

% skxawng
\hi
\HT{skxawng}{\B{skxawng}} \I{[sk'awN]} \emph{n.} moron, idiot.
\B{Fì\-kar\-yu alu fì\-po lu tsul\-fä\-tu; tsa\-kar\-yu alu tsa\-po lu skxawng.} \emph{This} teacher is a master; \emph{that} teacher is a fool.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Oe\-ti 'am\-pi rä\-'ä, ma \mbox{skxawng}!} Don't touch me, you moron!~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Fì\-skxawng\-ìri tsap\-'a\-lu\-te se\-ngi oe.} I apologize for this moron.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Nem\-fa na'\-rìng, ma skxawng\-tsyìp! Kä li!} Into the forest, you moron! Get going! (\emph{endearing})~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\emph{a. proverb}) \B{Spä skxawng sìn 'a\-na a\-flì.} ``A fool jumps onto a thin vine,'' i.e. \emph{don't engage in an unpromising and\slash or potentially risky cause}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Tsa\-ye\-rik te\-rul ne 'awkx. Spä skxawng sìn 'a\-na a\-flì.} That hexapede is running toward the cliff. Only a fool jumps onto a thin vine.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% skxir / si
\hi
\HT{skxir}{\B{skxir}} \I{[sk'iR]} \textsc{i}. \emph{n.} wound.  
\B{Oe\-ri skxir a mì syokx tì\-sraw se\-ngi.} The wound on my hand hurts.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Ze\-ne nga yi\-vur pxìm fì\-skxir\-it.} You must clean the wound frequently.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Rey\-pay skxir\-ftum\-fa he\-rum.} Blood is coming out of the wound.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} \\
\textsc{ii}. \B{skxir si} \I{[sk'iR si]} \emph{vin.} wound.
\B{Ta\-ron\-yu ye\-rik\-ur skxir so\-li.} The hunter wounded the hexapede.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{leskxir}{\B{le\-skxir}}, wounded. \\
\on{2}.~\HL{skxirtsyIp}{\B{skxir\-tsyìp}}, cut, bruise, minor wound.

% skxirtsyìp
\hi
\HT{skxirtsyIp}{\B{\cp{skxir}tsyìp}} \I{["sk'iR.\t{ts}jIp\textcorner]} \emph{n.} cut, bruise, minor wound. 
(\ding{43}~\HL{skxir}{\B{skxir}}, \HL{tsyIp}{\B{-tsyìp}})

% skxom
\hi
\HT{skxom}{\B{skxom}} \I{[sk'om]} \emph{n.} chance, opportunity. 
\B{Ngay\-txo\-a, nì\-aw\-no\-mum ke lo\-lu oer nì\-ke\-ftxo mì sok\-a srr ay\-skxom le\-tam fte lì'\-fya\-ri aw\-nge\-yä tì\-kang\-kem si\-vi.} My apologies: As you know, in recent days I have not had sufficient opportunity to work on our language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Nän ftia, nän lu skxom a em\-za\-'u.} The less you study, the less chance you have of passing.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}

% sl.
\hi
\B{sl.}~:~\HL{saylahe}{\B{saylahe}}.

% sla'tsu
\hi
\HT{sla'tsu}{\B{\cp{sla'}tsu}} \I{["{}slaP.\t{ts}u]} \emph{vtr.} describe.   
\B{Pol sla'\-tsu ay\-ioang\-it a tse\-'a fko mì Ey\-wa\-'e\-veng.} She describes animals seen on Pandora.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsla'tsu}{\B{tì\-sla'\-tsu}}, description.

% slayk
\hi
\HT{slayk}{\B{slayk}} \I{[slajk\textcorner]} \emph{vtr.} brush, comb.
\B{New sa'\-nok sli\-vayk nik\-ret 'eveng\-ä.} The mother wants to brush the child's hair.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}

% slan
\hi
\HT{slan}{\B{slan}} \I{[slan]} \emph{vtr.} support (\emph{emotionally, socially, or personally}). 
\B{Tì\-wä\-te\-ri ngal oe\-ti pe\-lun ke slan kaw\-krr?} Why don't you ever support me in an argument?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
\B{Ul\-te wä sì\-kawng a fì\-tì\-wu\-sem\-ìri, ze\-ne aw\-nga nì\-wotx fì\-tsap slä\-pi\-van.} And in this fighting against this evil we must all support each other.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b} 
(\ding{43}~\HL{kavan}{\B{ka\-van}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIslan}{\B{tì\-slan}}, support. 
\on{2}.~\HL{slantire}{\B{slan\-ti\-re}}, inspiration.

% slantire / si
\hi
\HT{slantire}{\B{slanti\cp{re}}} \I{[slan.ti."RE]} \textsc{i}. \emph{n.} inspiration.
\B{Mll\-te oe nì\-wotx. Fì\-slan\-ti\-re\-ri irayo nì\-txan, ma Tsyi\-li!} I agree completely. Thank you for this inspiration, Tsyi\-li.~\LINK{http://naviteri.org/2019/09/choice-statements-vs-choice-questions-and-some-insults/\#comment-30391}{b} \\
\textsc{ii}. \B{slanti\cp{re} si} \I{[slan.ti."RE si]} \emph{vin.} inspire.
\B{Sä\-ftxu\-lì\-'u Tsyeyk\-ä Na'\-vi\-ru slan\-ti\-re so\-li nì\-wotx.} Jake's speech inspired all the People.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\B{slan}, \B{tirea})

% slä
\hi
\HT{slA}{\B{slä}} \I{[sl\ae]} \emph{conj.} but.   
\B{Fo\-ru 'upxa\-ret oel fpo\-le', slä vay set ke pa\-mä\-hä\-ngem kea tì\-'eyng.} I have sent them a message, but up to now no answer has arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Pol\-txe Ey\-tu\-kan san oe ka\-yä sìk, slä oel pot ke spaw.} Eytukan said he would go but I don't believe him.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
(\ding{43}~\HL{ki}{\B{ki}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{slAkop}{\B{slä\-kop}}, but also.

% släkop
\hi
\HT{slAkop}{\B{\cp{slä}kop}} \I{["{}sl\ae.kop\textcorner]} \emph{adv.} but also. 
(\emph{a. together with} \ding{43}~\HL{ken'aw}{\B{ken\-'aw}})
\B{Nge\-yä tsmu\-ke lu ken\-'aw lor slä\-kop ka\-nu.} Your sister is not only beautiful but also intelligent.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Ay\-nga\-ri sìl\-pey oe tsnì ken\-'aw fpom\-tokx slä\-kop fpom\-ron yi\-vo'.} I hope that not only your physical but also your mental health will be perfect.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\emph{b. on its own}) \B{Fra\-krr lu nge\-yä sì\-pawm ngä\-zìk slä\-kop le\-tsran\-ten.} Your questions are always difficult but also important.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{slA}{\B{slä}}, \HL{kop}{\B{kop}})

% slär
\hi
\HT{slAr}{\B{slär}} \I{[sl\ae{}R]} \emph{n.} cave.
\B{Slär\-ìl nga' fwep\-it a\-txan.} The cave is very dusty.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Fì\-slär\-mì tsun fko sti\-vawm ngam\-it a\-pxay.} You can hear a lot of echoes in this cave.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Ri\-ti tswo\-lay\-on ftum\-fa slär.} The stingbat flew out of the cave.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 

% slele
\hi
\HT{slele}{\B{\cp{sle}le}} \I{["{}slE.lE]} \emph{vin.} swim.   
\B{Le\-hrr\-ap lu fwa evi\-tsyìp sle\-le mì hil\-van lu\-ke fwa fyeyn\-tu te\-rìng nari-} It's dangerous for tiny ones to swim in the river without an adult watching.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} 
\B{Txo fkol ke fyi\-vel u\-ran\-it pay\-wä, ze\-ne fko sli\-ve\-le.} If one does not seal a boat against water, one must swim.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIslele}{\B{nì\-sle\-le}}, by swimming.

% sleyku
\hi
\HT{sleyku}{\B{sley\cp{ku}}} \I{[slEj."k$\cdot$u]} \emph{vtr., inf.}\on{22} (\emph{a.}) produce, make.
\B{Nga pam\-rel so\-li san sley\-ki\-vu ut\-ral\-ìl lì'\-fya\-yä le\-Na'\-vi pxay\-a ay\-vur\-it a\-txur sìk.} You wrote, ``May the Na'vi language tree produce many strong stories.''~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
\B{Ta\-kuk set tsaw\-ke. Sley\-ku fra\-'ut u\-kxo.} Now the sun bears down. (It) makes everything dry.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg475008/\#msg475008}{ln}
(\emph{b. for emotional states}) ``make''. 
\B{Fu\-la tsa\-yun oeng pi\-väng\-kxo ye'\-rìn oe\-ti nit\-ram sley\-ku nì\-txan.} It makes me very happy that we'll be able to chat soon.~\LINK{http://forum.learnnavi.org/language-updates/fula-short-form-of-fiul-a-and-tsal-oeti-steyki-nitram/}{ln}
\B{Fì\-sä\-'an\-lal Ne\-we\-yä nga\-ti sley\-ka\-yu lek\-ye\-'ung!} This yearning for Newey is going to drive you crazy.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(i.) \B{txo\-pu sley\-ku}, frighten, scare, produce fear. \B{'Uol ik\-ran\-it txo\-pu sley\-ko\-la\-tsu, ta\-lun\-a po tsìk ya\-wo.} Something must have frightened the banshee, because it suddenly took to the air.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{slu}{\B{slu}})

% sloa
\hi
\HT{sloa}{\B{\cp{slo}a}} \I{["{}slo.a]} \emph{adj.} wide.
\B{Tsa\-ut\-ral\-ìri ta\-ngek lu sloa nì\-txan; 'evi ke tsun tsyi\-vìl.} The trunk of that tree is very wide; the kid cannot climb it.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{214}~\HL{snep}{\B{snep}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{slosnep}{\B{slo\-snep}}, width.
\on{2}.~\HL{slotsyal}{\B{slo\-tsyal}}, stormglider.

% sloan
\hi
\HT{sloan}{\B{slo\cp{an}}} \I{[slo."{}an]} \emph{vtr.} pour (s.t.).
\B{Ru\-txe sli\-vo\-an ngal pay\-it oe\-fpi; 'efu väng nì\-txan.} Please pour me some water; I'm very thirsty.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}

% slosnep / slosneppe
\hi
\HT{slosnep}{\B{slo\cp{snep}}} \I{[slo."{}snEp\textcorner]} \textsc{i}. \emph{n.} width. \\
\textsc{ii}. \B{slo\cp{snep}pe} \I{[slo."{}snEp\textcorner.pE]} \emph{or} \B{pe\-slo\cp{snep}} \I{[pE.slo."{}snEp\textcorner]} \emph{inter.} what width? how wide? 
\B{Kil\-van\-ìri tsa\-tseng slo\-snep\-pe?} How wide is the river there?~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{sloa}{\B{sloa}}, \HL{snep}{\B{snep}})

% slotsyal
\hi 
\HT{slotsyal}{\B{\cp{slo}tsyal}} \I{["{}slo.\t{ts}jal]} \emph{n., fauna} stormglider [lit.: `wide wing']. 
(\ding{43}~\HL{sloa}{\B{sloa}}, \HL{tsyal}{\B{tsyal}}) 

% slu
\hi
\HT{slu}{\B{slu}} \I{[slu]} \emph{vin.}  become. 
(\emph{a. with n.s}) \B{Nga\-ri hu Ey\-wa sa\-lew ti\-rea, tokx 'ì\-'awn, slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body stays behind to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake.27s_final_kill}{a\slash wiki}
\B{Tsa\-ri\-a fkol po\-le\-'un fu\-ta Lo\-ak slu ta\-ron\-yu, sem\-pul 'efu ye.} Father is content that it's been decided Loak will be a hunter.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Mì Si\-ä\-tll a nu\-mul\-txa\-ri lì'\-fya\-yä le\-Na'\-vi, sìl\-pey oe slì\-ye\-vu nga nu\-mul\-txa\-tu!} I hope you'll participate in the upcoming Na'vi class in Seattle!~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Oe\-yä ik\-ran sli\-vu nga, tsa\-krr oeng 'aw\-si\-teng mi\-vak\-to.} Be my ikran and let's ride together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{ld}
(\emph{b. optionally with} \B{ne} \emph{to disambiguate the predicate of a sentence}) 
\B{Ta\-ron\-yu slu ne tsam\-si\-yu.} The hunter becomes a warrior.~\LINK{http://wiki.learnnavi.org/Canon/2010/UltxaAyharyu\%C3\%A4\#Word_Order_with_Slu}{wiki}
(\emph{c. with adj.s}) 
\B{Fì\-fne\-rìn ke ley krr\-a slu pay\-nga'.} This kind of wood is worthless when it gets damp.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b} 
\B{Ay\-nga\-ri teng\-krr ya wur sle\-ru nì\-'ul, sìl\-pey oe, li\-vu hel\-ku sang ul\-te te'\-lan le\-fpom.} As for you all, while it's getting cooler, I hope your homes may be warm and your hearts happy.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Oeri key tun slolu nìtxan.} I'm so embarrassed [lit.: my face has become very red].~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-694}{b}
(\emph{d. idiom}) \B{X nì\-Na'\-vi slu lì\-'u\-pe\slash pe\-lì\-'u?} How do you say X in Na'vi?~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b} 
(\emph{e. proverb}) \B{Tì\-tok slu tìk\-tok fa pam\-tsyìp a\-'aw.} Presence becomes absence by means of one little sound. [i.e. `Small things can make a big difference.']~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b} \\
(i.) \B{tsawl slu}, grow. \\
(ii.) \B{yawne slu}, fall in love. \B{Lu poe se\-vin nì\-ftxan kuma yaw\-ne slo\-lu oer.} She was so beautiful that I fell in love with her.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
(iii.) \B{muntxa slu}, get married. \B{Fko pll\-txe\-i\-e san me\-nga mun\-txa slo\-lu sìk!} I'm so happy to hear you got married!~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}  \\
(iv.) \B{slu leyr}, freeze. \B{Mì zì\-sì\-krr a\-txa\-wew slu ay\-ora leyr.} In the very cold season, the lakes freeze.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sleyku}{\B{sley\-ku}}, produce, make. 
\on{2}.~\HL{zoslu}{\B{zo\-slu}}, heal, become well, get fixed. 
\on{3}.~\HL{tsunslu}{\B{tsun\-slu}}, may, be possible.

% slukx
\hi 
\HT{slukx}{\B{slukx}} \I{[sluk']} \emph{or} \I{[slUk']} \emph{n.} horn (\emph{of an animal}). 
\B{Na\-ri si! Tsa\-ioang\-ur lu pxi\-a me\-slukx!} Careful! That animal has two sharp horns!~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 

% smaw
\hi 
\HT{smaw}{\B{smaw}} \I{[smaw]} \emph{vtr.} approve of. 
\B{Fay\-hemit oel smaw nì\-wotx.} I completely approve of these actions.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
(\ding{214}~\HL{natxu}{\B{na\-txu}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsmaw}{\B{tì\-smaw}}, approval.

% smaoe
\hi
\HT{smaoe}{\B{sma\cp{o}e}} \I{[sma."{}o.E]} \emph{n.}, \ding{96} phalanxia, heavy thorned plant, \emph{phalanxia ferox}. 

% smar
\hi
\HT{smar}{\B{smar}} \I{[smaR]} \emph{n.} prey, a hunted thing.
\B{Tì\-tu\-sa\-ron\-ìri fte fli\-vä, ze\-ne fko si\-vutx smar\-it ni\-nan nì\-no.} To succeed at hunting, you have to track your prey by reading (the forest) with attention to detail.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Tsu'\-tey\-ìl to\-lìng oer mawl\-it smar\-ä.} Tsu'tey gave me a half of the prey.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\emph{a. proverb}) (i.) \B{Ätxä\-le si pa\-lu\-lu\-kan\-ur tsnì smar\-it li\-vo\-nu.} \emph{a futile gesture, an attempt to achieve something that might be desirable but will clearly not happen} [lit.: ask a thanator to release its prey.]~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(ii.) \B{Ta\-ron\-yut yom smar\-ìl.} \emph{A totally botched situation, chaos} [lit.: the prey eats the hunter].
\B{Eyk Ka\-mun a fra\-lo lä\-ngu tsa\-sä\-ta\-ron vel\-ke nì\-wotx. Ta\-ron\-yut yom smar\-ìl!} Every time Kamun is in charge, the hunt is a mess. Everything goes wrong that can!~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% smon
\hi
\HT{smon}{\B{smon}} \I{[smon]} \emph{vin.} be familiar to, ``know s.o.''.
\B{Po smon oer.} I know him.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Sra\-ke smon ngar oe\-yä mey\-lan alu En\-tu sì Ka\-mun?} Do you know my friends En\-tu and Ka\-mun?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b} 
\B{Moe smon (moe\-ru) fì\-tsap.} We know each other.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
(\emph{a. conv.}) \B{Smon nì\-prr\-te'.} ``Nice to know you.''~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b}
(\emph{b. proverb}) \B{Ha\-haw nì\-'aw txo pa\-lu\-kan smi\-von ngar.} ``Only sleep if you are familiar with the Thanator,'' i.e. \emph{don't think you're safe unless you're aware of the danger}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% sna-
\hi
\HT{sna}{\B{sna-}} \I{[sna.]} \on{1}. \emph{prod. pref. to describe a naturally occurring group or collection of animals or plants.} 
\B{sna\-ta\-li\-o\-ang}, a herd of sturmbeest. 
\on{2}. \emph{unprod. pref. to describe a collection of n.s denoting non\hyp{}living things.} \B{sna\-txä\-rem}, skeleton.
(\ding{43}~\HL{sna'o}{\B{sna\-'o}})

% sna'o
\hi
\HT{sna'o}{\B{\cp{sna}'o}} \I{["{}sna.Po]} \emph{n.} set, group, pile, clump, stand.   
\B{Ay\-skxe a mì sa\-sna\-'o ku'\-up lu nì\-txan.} The rocks in that pile are very heavy.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Hay\-ku\-ri sna\-'o ay\-nge\-yä lor nì\-txan lu nang!} What a beautiful collection of haiku!~\LINK{http://naviteri.org/2012/10/ulte-sawasultsyipa-yoratu-leiu-and-the-winner-is/}{b}

% snafpìlfya
\hi
\HT{snafpIlfya}{\B{sna\cp{fpìl}fya}} \I{[sna."fpIl.fja]} \emph{n.} philosophy.
\B{Sna\-fpìl\-fya\-ri le\-Na'\-vi krra smar\-it fkol tspang, tsran\-ten nì\-txan fwa po ke ngä\-'än nì\-kel\-kin.} It's important in Na'vi philosophy that the prey not suffer unnecessarily when it's killed.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{sna}{\B{sna-}}, \HL{fpIlfya}{\B{fpìl\-fya}})

% snanumultxa
\hi
\HT{snanumltxa}{\B{snanumul\cp{txa}}} \I{[sna.nu.mul."t'a]} \emph{or} \I{[.mUl.]} (\textsc{rn}: \B{snanumùl\cp{txa}} \I{[sna.nu.mUl."da]}) \emph{n.} course (\emph{of lectures}). 
(\ding{43}~\HL{sna}{\B{sna-}}, \HL{numultxa}{\B{nu\-mul\-txa}})

% snapamrelvi
\hi
\HT{snapamrelvi}{\B{snapam\cp{rel}vi}} \I{[sna.pam."REl.vi]} \emph{n.} alphabet. 
\B{Sna\-pam\-rel\-vi sì Lì\-'u\-pam}, Alphabet and Pronunciation.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{43}~\HL{sna}{\B{sna-}}, \HL{pamrelvi}{\B{pam\-rel\-vi}})

% snatanhì
\hi
\HT{snatanhI}{\B{snatan\cp{hì}}} \I{[sna.tan."hI]} \emph{n., astr.} constellation (\emph{of stars}).
(\ding{43}~\HL{sna}{\B{sna-}}, \HL{tanhI}{\B{tan\-hì}})

% snatanhìtsyìp
\hi
\HT{snatanhItsyIp}{\B{snatan\cp{hì}tsyìp}} \I{[sna.tan."hI.\t{ts}jIp\textcorner]} \emph{n., astr.} star cluster. 
(\ding{43}~\HL{snatanhI}{\B{sna\-tan\-hì}}, \HL{tsyIp}{\B{-tsyìp}})

% snatxärem
\hi
\HT{snatxArem}{\B{sna\cp{txä}rem}} \I{[sna."t'\ae.REm]} \emph{n.} skeleton.
(\ding{43}~\HL{sna}{\mbox{\B{sna-}}}, \HL{txArem}{\B{txä\-rem}})

% snaytu
\hi
\HT{snaytu}{\B{\cp{snay}tu}} \I{["{}snaj.tu]} \emph{n.} loser.   
\B{Fra\-u\-van\-ìri lu yo\-ra'\-tu, lu snay\-tu.} For every game, there's a winner and a loser.~\LINK{http://naviteri.org/2011/12/a-note-on-the-word-\%E2\%80\%9Cyora\%E2\%80\%99tu\%E2\%80\%9D/}{b}
(\ding{43}~\HL{snaytx}{\B{snaytx}})

% snaytx
\hi
\HT{snaytx}{\B{snaytx}} \I{[snajt']} \emph{vtr.} lose (\emph{a game}).   
\B{Trr\-am\-ä ay\-u\-van\-ìri mak\-to Ak\-wey nì\-yew\-la ha snaytx.} Ak\-wey rode disappointingly in yesterday's games, so he lost.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Ay\-oe sno\-laytx fì\-trr ta\-lun\-a oel rum\-it to\-lung\-zup.} We lost today because I dropped the ball.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
\B{Oe pll\-ngay san mo\-lak\-to oe nì\-fe', ta\-fral sno\-laytx; wä\-tu lu oe\-to txur.} I acknowledge that I rode badly, so I lost; my opponent was stronger than I was.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{snaytu}{\B{snay\-tu}}, loser.

% snayì
\hi
\HT{snayI}{\B{sna\cp{yì}}} \I{[sna."jI]} \emph{n.} (natural) staircase, series of step\hyp{}like levels. 
\B{Kll\-zo\-la\-'u Mo\-'at fa sna\-yì teng\-krr pe\-rll\-txe san Ay\-nga ne\-to ri\-vikx!} Mo'at came down the stairway saying, ``Get back, all of you!''~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{sna}{\B{sna-}}, \HL{yI}{\B{yì}})

% snäm
\hi
\HT{snAm}{\B{snäm}} \I{[sn\ae{}m]} \emph{vin.} rot, decay, degrade over time (\emph{of an object or s.t. abstract}). 
\B{Fì\-tsngan\-ur a sno\-läm lä\-ngu fa\-hew a\-kxä\-näng.} This rotten meat has a putrid smell.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
\B{Ze\-ne fko tsko swi\-zaw\-it si\-var nì\-trr\-trr fte\-ke fì\-tsu\-'o sni\-väm.} One must use a bow and arrow regularly to prevent this ability degrading over time.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}

% snep
\hi
\HT{snep}{\B{snep}} \I{[snEp\textcorner]} \emph{adj.} narrow.
(\ding{214}~\HL{sloa}{\B{sloa}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{slosnep}{\B{slo\-snep}}, width.

% snew
\hi
\HT{snew}{\B{snew}} \I{[snEw]} \emph{vtr.} constrict, tighten. 
(\emph{a. idiom})
\B{Ke tsun fko tspi\-vang to\-ruk\-it fa fwa pewn\-ti snew.} \emph{Proverbial expression for a method that will not work.} [lit.: you can't kill a great leonopteryx by constricting its throat]~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} 
(\emph{shortened to}) \B{pewn to\-ruk\-ä}. \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{snewsye}{\B{snew\-sye}}, weird, spooky.

% snewsye
\hi
\HT{snewsye}{\B{\cp{snew}sye}} \I{["{}snEw.sjE]} \emph{adj.} weird, spooky.   
\B{Ay\-nga\-ri ftxo\-zä Hä\-lo\-win\-ä li\-vu 'o' sì snew\-sye txan\-txew\-vay.} May your Halloween be as fun and spooky as possible.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b}
\B{Oe\-ri lu fì\-tseng snew\-sye nì\-hawng. Hi\-vum ko.} This place is too spooky for me. Let's get out of here.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b}
(\ding{43}~\HL{snew}{\B{snew}}, \HL{syeha}{\B{sye\-ha}})

% sno
\hi
\HT{sno}{\B{sno}} \I{[sno]} (\emph{gen.} \ding{43}~\HL{sneyA}{\B{\cp{sne}yä}}) \emph{pn. 3\textsuperscript{rd} person to mean} ``his\slash her\slash its\slash their self'' or ``his\slash her\slash their own''; \emph{indifferent to number; most often used in its genitive form}.
(\emph{a. subjective})
\B{Nì\-trr\-trr pxì\-mun\-'i sam\-si\-yul ay\-swi\-zaw\-it ku\-tu\-ä a\-law\-nä\-txayn sno\-kip nì\-'eng.} Warriors typically share the arrows of their defeated enemies among themselves.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\emph{b. genitive})
\B{Pol 'o\-lem sne\-yä wu\-tsot.} He made his (own) dinner.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Sa'\-nok prr\-nen\-ur yom\-tìng fa okup sne\-yä.} A mother feeds an infant with her milk.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}
(\emph{c. dative})
\B{Po yaw\-ne lu snor.} He loves himself.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Me\-fo yaw\-ne lu (snor) fì\-tsap.} They (= those two) love each other.
\B{Fo smon (sno\-ru) fì\-tsap nì\-wotx.} They all know each other.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{snolup}{\B{sno\-lup}}, personal style. 
\on{2}.~\HL{snomo}{\B{sno\-mo}}, private space, room.
\on{3}.~\HL{snonrra}{\B{sno\-nrra}}, self-pride. 
\on{4}.~\HL{snokfyan}{\B{snok\-fyan}}, personal belongings rack.
\on{5}.~\HL{snonivi}{\B{sno\-ni\-vi}}, single\hyp{}person hammock.
\on{6}.~\HL{snotipx}{\B{sno\-tipx}}, self\hyp{}control.

% snokfyan
\hi
\HT{snokfyan}{\B{\cp{snok}fyan}} \I{["snok\textcorner.fjan]} \emph{n.} personal belongings rack.
(\ding{43}~\HL{sno}{\B{sno}}, \HL{kurfyan}{\B{kur\-fyan}})

% snolup
\hi
\HT{snolup}{\B{\cp{sno}lup}} \I{["sno.lup\textcorner]} \emph{or} \I{[.lUp\textcorner]} (\textsc{rn}: \B{\cp{sno}lùp} \I{["sno.lUp\textcorner]}) \emph{n.} personal style or aesthetic, presence.
\B{Por lu sno\-lup a new fra\-po rì\-'ìr si\-vi.} He\slash She has a personal style that everyone wants to emulate.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\ding{43}~\HL{sno}{\B{sno}}, \HL{lupra}{\B{lup\-ra}})

% snomo
\hi
\HT{snomo}{\B{\cp{sno}mo}} \I{["{}sno.mo]} \emph{n.} private space that one can retreat to. 
\B{Nga\-ri sno\-mot krr\-pe ve\-zey\-ko, ma 'itan?} When are you going to organize your room, son?~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(i.) \B{(sno)\-mo a ha\-haw}, bedroom.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{sno}{\B{sno}}, \HL{mo}{\B{mo}})

% snonivi
\hi
\HT{snonivi}{\B{\cp{sno}nivi}} \I{["{}sno.ni.vi]} \emph{n.} single\hyp{}person hammock.
\B{Ze\-ne aw\-nga tì\-ngä\-zìk\-ur sno\-ni\-vi\-yä tì\-ke\-zin si\-vi nì\-win.} We need to solve the hammock problem quickly.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/\#comment-33483}{b}
(\ding{43}~\HL{sno}{\B{sno}}, \HL{nivi}{\B{ni\-vi}})

% snonrra
\hi
\HT{snonrra}{\B{sno\cp{nrr}a}} \I{[sno."n\s{r}.a]} \emph{n.} self\hyp{}pride (\emph{negative connotation}), arrogance.
(\ding{43}~\B{sno}, \B{nrr\-a}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{lesnonrra}{\B{le\-sno\-nrra}}, full of self\hyp{}pride.

% snotipx
\hi 
\HT{snotipx}{\B{sno\cp{tipx}}} \I{[sno."tip']} \emph{n.} self\hyp{}control. 
\B{Ke fkey\-tok tì\-eyk\-tan a\-tì\-flä\-nga' lu\-ke sno\-tipx.} Successful leadership does not exist without self-control.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
(\ding{43}~\HL{sno}{\B{sno}}, \HL{tIhipx}{\B{tì\-hipx}})

% Snrrtsyìp
\hi 
\HT{snrrtsyIp}{\B{\cp{Snrr}tsyìp}} \I{["s\s{r}.\t{ts}jIp\textcorner]} \emph{n., astr., short form of} \ding{43}~\HL{tawsnrrtsyIp}{\B{Tawsnrrtsyìp}}.

% snumìna
\hi
\HT{snumIna}{\B{\cp{snu}mìna}} \I{["snu.mI.na]} (\textsc{rn}: \B{\cp{snù}mìna} \I{["snU.mI.na]}) \emph{adj.} dim (\emph{of a person}), stupid. 
\B{Oe\-ru po snu\-mìn\-a la\-tsam.} He seems dim to me.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Wutso_in_Hometree}{a\textsuperscript{c}}
\B{snu\-mì\-na skxawng\slash skxawng a\-snu\-mì\-na}, stupid moron.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}

% sngap
\hi
\HT{sngap}{\B{sngap}} \I{[sNap\textcorner]} \emph{vtr.} sting. 
(\emph{a. idiom}) \B{sre fwa sngap zi\-ze'}, ``as quickly as possible'' [lit: before the hellfire wasp stings].~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}

% sngä'i
\hi
\HT{sngA'i}{\B{\cp{sngä}'i}} \I{["{}sN\ae.Pi]} (\emph{a.}) \emph{vin.} begin, start, s.t. begins\slash starts.   
\B{Ye'\-rìn sngä\-'i tom\-pa.} Soon the rain starts.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg475008/\#msg475008}{ln}
\B{Aw\-nga sngi\-vä\-'i ko!} Let's begin!~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Tì\-kang\-kem sngo\-lä\-'i.} The work began.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{b. modal}) start, begin to do s.t. 
\B{Na'\-rìng sngä\-'i ni\-vrr txon\-krr syu\-ra\-tan\-fa.} The forest begins to glow at night with bioluminescence.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Tskot sngo\-lä\-'i po si\-var 'a\-'aw\-a trr\-kam.} He started to use the bow several days ago.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{tsyul}{\B{tsyul}}, \ding{214}~\HL{'i'a}{\B{'i'\-a}}) \\  
\textsc{Derivations}:
\on{1}.~\HL{sngA'ikrr}{\B{sngä\-'i\-krr}}, beginning (time).  
\on{2}.~\HL{sngA'itseng}{\B{sngä\.'i\-tseng}}, beginning, starting position.  
\on{3}.~\HL{sngA'iyu}{\B{sngä\-'i\-yu}}, beginner. 
\on{4}.~\HL{nIsngA'i}{\B{nì\-sngä\-'i}}, originally, at first. 
\on{5}.~\HL{lesngA'i}{\B{le\-sngä\-'i}}, original, first in a series.

% sngä'ikrr
\hi
\HT{sngA'ikrr}{\B{\cp{sngä}'ikrr}} \I{["{}sN\ae.Pi.k\s{r}]} \textsc{i}. \emph{n.} beginning, start time. \\
\textsc{ii}. \emph{adv.} in the beginning, at the start.
\B{Txe\-wì\-yä sem\-pul\-ìri sngä\-'i\-krr Saw\-tu\-tet ve\-re'\-kì.} In the beginning, Txe\-wì's father hated the Skypeople.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Tsa\-lì'\-fya\-vi\-ri, sngä\-'i\-krr lo\-lu am\-'a\-lu\-ke aw\-nge\-yä 'ey\-lan alu Tsyeyk tsa\-fne\-tu\-te.} As for that expression, our friend Jake was doubtless that kind of person in the beginning.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/\#comment-30289}{b}
(\ding{43}~\HL{sngA'i}{\B{sngä\-'i}}, \HL{krr}{\B{krr}}, \ding{214}~\HL{tI'i'a}{\B{tì\-'i'a}})

% sngä'itseng
\hi
\HT{sngA'itseng}{\B{\cp{sngä}'itseng}} \I{["{}sN\ae.Pi.\t{ts}EN]} \emph{n.} beginning, starting position, initial location.
\B{Ro sngä\-'i\-tseng tsa\-lì\-'u\-ä alu 'ey\-lan lu tì\-ftang.} At the beginning of the word \emph{'eylan} there's a glottal stop.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{sngA'i}{\B{sngä\-'i}}, \HL{tseng}{\B{tse\-nge}})

% sngä'iyu
\hi
\HT{sngA'iyu}{\B{\cp{sngä}'iyu}} \I{["{}sN\ae.Pi.ju]} \emph{n.} beginner.   
\B{Fra\-po, ftxey sngä\-'i\-yu ftxey tsul\-fä\-tu, tsì\-ye\-vun fì\-tse\-nge ri\-vun 'u\-ot le\-sar.} Everyone, whether beginner or master, can find something useful here.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
\B{num\-tseng\-vi ay\-sngä\-'i\-yu\-ä}, beginners' classroom.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Sìn yì sngä\-'i\-yu\-ä, slä tsye\-rìl (hay\-a yì\-ne) nì\-'ul\-'ul.} They're at a beginner's level, but they're getting better and better.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Nì\-aw\-no\-mum ngo\-lop ay\-oel fì\-nu\-mul\-txa\-ti fpi sngä\-'i\-yu.} As you know, we created this class for the beginner.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\ding{43}~\HL{sngA'i}{\B{sngä\-'i}}, \ding{214}~\HL{tsulfAtu}{\B{tsul\-fä\-tu}})

% sngeykä'i
\hi
\HT{sngeykA'i}{\B{sngey\cp{kä}'i}} \I{[sNEj."k$\cdot$\ae.P$\cdot$i]} \emph{vtr.} begin, start s.t.
\B{Poel tì\-kang\-kem\-it sngey\-ko\-lä\-'i.} She began the work.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
\B{Fwa oel fì\-tì\-kang\-kem\-vit sngey\-ki\-vä\-'i krr\-no\-lekx nì\-txan.} Me starting this project took a long time.~\LINK{http://naviteri.org/2010/06/first-post/}{b} 
(\ding{43}~\HL{sngA'i}{\B{sngä\-'i}}; \HL{tsyul}{\B{tsyul}})

% sngel
\hi
\HT{sngel}{\B{sngel}} \I{[sNEl]} \emph{countable n.} garbage, trash. 
\B{Oel vewng ay\-sngel\-it.} I tend to the refuse.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Ngal kaw\-krr sngel\-it ke mu\-nge wrr\-pa\-ne.} You never take the trash out.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Pey\-ral\-ìl zet wur\-a wu\-tsot a\-'aw\-nem pxel sngel.} Peyral treats a cool cooked meal like garbage.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sngeltseng}{\B{sngel\-tseng}}, garbage dump.

% sngeltseng
\hi
\HT{sngeltseng}{\B{\cp{sngel}tseng}} \I{["{}sNEl.\t{ts}EN]} \emph{n.} garbage dump.
\B{Taw\-sìp nge\-yä lu sngel\-tseng.} Your ship is a garbage scow.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog_.28Twitter_Questions.29}{ugo\slash wiki}
(\ding{43}~\HL{sngel}{\B{sngel}}, \HL{tseng}{\B{tse\-nge}})

% sngukx
\hi
\HT{sngukx}{\B{sngukx}} \I{[sNuk']} \emph{or} \I{[sNUk']} \emph{n.}, \ding{96} grub plant, \emph{limacina erecta}. 

% sngum
\hi
\HT{sngum}{\B{sngum}} \I{[sNum]} \emph{or} \I{[sNUm]} \textsc{i}. \emph{n.} worry.   
\B{Lu oe\-ru sngum a sa\-ron\-yu ke tì\-ye\-vä\-txaw.} I'm worried that the hunters will not return (soon, as expected).~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\
\textsc{ii}. \B{sngum si} \I{[sNum si]} \emph{vin.} worry.
\B{Sngum rä\-'ä si.} Don't worry.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/comment-page-1/\#comment-10619}{b}
\B{Fu\-ri\-a po ke za\-ya\-'u oe sngum si.} I'm worried that he won't come.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/comment-page-1/\#comment-10734}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sngumtsim}{\B{sngum\-tsim}}, worrisome matter. 
\on{2}.~\HL{sngunga'}{\B{sngu\-nga'}}, worrisome, troubling. 
\on{3}.~\HL{txansngum}{\B{txan\-sngum}}, desperation.
 
% sngumtsim
\hi
\HT{sngumtsim}{\B{\cp{sngum}tsim}} \I{["{}sNum.\t{ts}im]} \emph{or} \I{["{}sNUm.]} \emph{n.} worrisome matter, source of worry.
\B{Lu oer sngum\-tsim fwa po ke zo\-la\-'u.} I have a feeling of worry that he didn't come.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Pe\-yä ay\-lì\-'u a\-ing\-ye\-nga' lo\-lu sngum\-tsim ay\-oe\-ru nì\-wotx.} His mysterious words worried us all.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Syor, ma 'ey\-lan, syor. Ke lu kea sngum\-tsim.} Relax, friend, relax. There's nothing to worry about.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{sngum}{\B{sgnum}}, \HL{tsim}{\B{tsim}})

% sngunga'
\hi
\HT{sngunga'}{\B{\cp{sngu}nga'}} \I{["{}sNu.NaP]} \emph{adj.} worrisome, troubling.  
(\emph{a. proverb}) \B{Hem a\-sru\-nga' nì\-'ul, hum a\-sngu\-nga' nì\-nän.}
More helpful actions lead to less troubling outcomes.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{sngum}{\B{sngum}}, \HL{-nga'}{\B{-nga'}})
 
% so'ha
\hi
\HT{so'ha}{\B{\cp{so'}ha}} \I{["{}soP.ha]} \emph{vtr.} (\emph{a.}) be enthusiastic about, show enthusiasm for, be excited about. 
\B{Oel so'\-ha fu\-ta trr\-ay nga\-hu kä.} I'm excited about going with you tomorrow.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Tse\-nul so'\-ha tey\-lut nì\-hawng nì\-'it, ke\-fyak?} Don't you think Tsenu is a bit too into \emph{teylu}?~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Saw\-no'\-ha ioit ko\-la\-ne\-i\-om oel u\-van\-fa!} I got (won) the prized piece of jewelry in the game!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{b. conv., intj.}) \B{Le\-iam fwa Txe\-wì Ri\-ni\-sì mun\-txa slì\-yu.---So'\-ha!} It looks like Txewì and Rini are about to get married.---That's great!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIso'ha}{\B{tì\-so'\-ha}}, enthusiasm, having a good attitude. 
\on{2}.~\HL{leso'ha}{\B{le\-so'\-ha}}, enthusiastic, keen. 
\on{3}.~\HL{nIso'ha}{\B{nì\-so'\-ha}}, enthusiastically.
\on{4}.~\HL{so'hayu}{\B{so'\-ha\-yu}}, enthusiast.

% so'hayu
\hi
\HT{so'hayu}{\B{\cp{so'}hayu}} \I{["{}soP.ha.ju]} \emph{n.} enthusiast.
(\ding{43}~\HL{so'ha}{\B{so'\-ha}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{txanso'hayu}{\B{txanso'hayu}}, fan, enthusiast.

% so'yu
\hi 
\HT{so'yu}{\B{\cp{so'}yu}} \I{["{}soP.ju]} \emph{n.} fan, enthusiast; (\emph{coll. form of} \ding{43}~\HL{txanso'hayu}{\B{txan\-so'\-ha\-yu}}). 

% soaia
\hi
\HT{soaia}{\B{so\cp{a}ia}} \I{[so."{}a.i.a]} (\emph{gen.} \B{soaiä}) \emph{n.} family.
\B{Soaia lu hì\-'i -- nì\-'aw sem\-pul, sa'\-nok, tsmu\-kan sì Txe\-wì.} The family is small -- only father, mother, a brother and Txewì.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Mu\-nge fna\-we'\-tul lewng\-it soaia\-ru sne\-yä.} A coward brings shame to his\slash her family.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
\B{Va'\-rul zä\-nget ik\-ran\-it sn\-eyä pxel ha\-pxì\-tu so\-a\-i\-ä.} Va'ru treats his ikran like a member of the family.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Mì sray apxa kel\-ku si so\-a\-i\-a\-hu.} (He) lives with his family in a large village.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIsoaia}{\B{nì\-soaia}}, (together) as members of a family. 
\on{2}.~\HL{swaynivi}{\B{sway\-ni\-vi}}, family hammock.
\on{3}.~\HL{swapxI}{\B{swa\-pxì}}, family member.

% sok
\hi
\HT{sok}{\B{sok}} \I{[sok\textcorner]} \emph{adj.} recent.
\B{Ngay\-txo\-a, nì\-aw\-no\-mum ke lo\-lu oer nì\-ke\-ftxo mì sok\-a srr ay\-skxom le\-tam fte lì'\-fya\-ri aw\-nge\-yä tì\-kang\-kem si\-vi.} My apologies: As you know, in recent days I have not had sufficient opportunity to work on our language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Sìn A\-sok, Sìn Zu\-saw\-krr\-ä}, Recent and Upcoming Activities [lit.: recent activities, future's activities].~\LINK{http://naviteri.org/2013/04/sin-asok-sin-zusawkrra-recent-and-upcoming-activities/}{b}
\on{1}.~\HL{nIsok}{\B{nì\-sok}}, recently.

% solew
\hi
\HT{solew}{\B{so\cp{lew}}} \I{[so."lEw]}~:~\HL{salew}{\B{salew}}.

% som
\hi
\HT{som}{\B{som}} \I{[som]} \emph{adj.} hot. 
\B{Ya\-ri som\-wew\-pe (set)?---Ya lu som.} How's the temperature (now)?---It's hot.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{214}~\B{wew}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsom}{\B{tì\-som}}, heat. 
\on{2}.~\HL{txasom}{\B{txa\-som}}, very hot. 
\on{3}.~\HL{somwew}{\B{som\-wew}}, temperature. 
\on{4}.~\HL{somtIlor}{\B{som\-tì\-lor}}, popsicle, ``hot beauty''. 
\on{5}.~\HL{zIskrrsom}{\B{zì\-skrr\-som}}, summer.

% somtìlor
\hi
\HT{somtIlor}{\B{\cp{som}tìlor}} \I{["{}som.tI.loR]} \emph{n.}, \ding{96} popsicle, ``hot beauty'',
\emph{capsulatum virgatum}.
(\ding{43}~\HL{som}{\B{som}}, \HL{tIlor}{\B{tì\-lor}})

% somwew
\hi
\HT{somwew}{\B{som\cp{wew}}} \I{[som."wEw]} \textsc{i}. \emph{n.} temperature. 
\B{Som\-wew\-ä tì\-'ul a ka 'Rr\-ta sä\-peng\-hrr lu aw\-nga\-ru nì\-wotx.} Global warming (lit.: the increase in temperature across the Earth) is a warning to us all.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} \\
\textsc{ii}. \B{som\cp{wew}pe} \I{[som."{}wEw.pE]} \emph{or} \B{pe\-som\-\cp{wew}} \I{[pE.som."{}wEw]} \emph{inter.} what temperature? ``How cold\slash warm?''
\B{Ya\-ri som\-wew\-pe (set)?---Ya lu sang.} What's the temperature?---It's warm.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Pay\-ri som\-wew\-pe?} What's the temperature of the water?~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\ding{43}~\HL{som}{\B{som}}, \HL{wew}{\B{wew}})

% sop
\hi
\HT{sop}{\B{sop}} \I{[sop\textcorner]} \emph{vin.} travel. 
\B{Tsun oe nga\-hu tsa\-tse\-nge\-ne ki\-vä, slä nul\-new fu\-ta si\-vop oeng nì\-txi\-lu\-ke.} I can go there with you, but I prefer to travel leisurely.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Ay\-fo so\-lop ìlä hil\-van fa u\-ran.} They traveled along (up, down) the river by boat.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\emph{a. idiom}) \B{Si\-vop nì\-zaw\-nong.} Travel safely.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsop}{\B{tì\-sop}}, journey. 
\on{2}.~\HL{sopyu}{\B{sop\-yu}}, traveller.

% sopyu
\hi
\HT{sopyu}{\B{\cp{sop}yu}} \I{["sop\textcorner.ju]} \emph{n.} traveller.
\B{Si\-vop nì\-zaw\-nong, ma ay\-sop\-yu.} Travel safely, travellers.~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b}
(\ding{43}~\HL{sop}{\B{sop}})

% Sorewn
\hi
\B{So\cp{rewn}} \I{[so."REwn]} \emph{n.} \emph{female name}. 
\B{Nga lu So\-rewn, ke\-fyak?} You're Sorewn, right?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{So\-rewn\-ìl kan\-'ìn tì\-'em\-it nì\-txan ul\-te kxawm tsa\-txe\-le me\-nga\-ne za\-'a\-tsu nì\-'eng.} Sorewn is really into cooking and you two might share that in common.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Mun\-txa\-tu\-ri So\-rewn\-ti ke tsun oe mi\-vll\-'an.} I can't accept Sorewn as my spouse.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% sosul
\hi
\HT{sosul}{\B{so\cp{sul}}} \I{[so."sul]} \emph{or} \I{[."sUl]} (\textsc{rn}: \B{so\cp{sùl}} \I{[so."sUl]}) \emph{adj.} pleasant smell (\emph{of nearby running water, rain, moist vegetation}).

% spakat
\hi 
\HT{spakat}{\B{\cp{spa}kat}} \I{["spa.kat\textcorner]} \emph{vin.} drown. 
\B{Fì\-ioang ke tsun sli\-ve\-le; nem\-fa pay zup tìk spa\-kat.} This animal cannot swim; if it falls into the water, it immediately drowns.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}

% spaw
\hi
\HT{spaw}{\B{spaw}} \I{[spaw]} \emph{vtr.} believe (s.o.).
\B{Spi\-vaw oe\-ti ru\-txe, ma oe\-yä ey\-lan.} Please believe me, my friends.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Pol\-txe Ey\-tu\-kan san oe ka\-yä sìk, slä oel pot ke spaw.} Eytukan said he would come but I don't believe him.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175491/\#msg175491}{ln}
(\emph{a. conv.}) \B{spaw oe …\slash spaw oel fu\-ta …}, I believe (that).
\B{Spaw oel fu\-ta Mo\-'at\-ìl tso\-le\-'a Ney\-ti\-rit.} I believe Mo'at saw Neytiri.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/}{b}
\B{Spaw oe, fwa po ko\-lä lä\-ngu kxey\-ey.} I believe it was a mistake for him to have gone.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tIspaw}{\B{tì\-spaw}}, belief (abstract).
\on{2}.~\HL{sAspaw}{\B{sä\-spaw}}, belief (particular).

% spä
\hi
\HT{spA}{\B{spä}} \I{[sp\ae]} \emph{vin.} jump.
\B{Po spä few pay\-fya fte smar\-it si\-vutx.} He jumped across the stream to track his prey.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Maw\-krr\-a pa\-lu\-lu\-kan po\-sìn spo\-lä, ke tsä\-ngun fko pot ti\-va\-rep.} Sadly, once the thanator had jumped on her, she could not be rescued.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\emph{a. proverb}) \B{Spä skxawng sìn 'a\-na a\-flì.} ``A fool jumps onto a thin vine,'' i.e. \emph{don't engage in an unpromising and\slash or potentially risky cause}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Tsa\-ye\-rik te\-rul ne 'awkx. Spä skxawng sìn 'a\-na a\-flì.} That hexapede is running toward the cliff. Only a fool jumps onto a thin vine.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% spe'e
\hi
\HT{spe'e}{\B{spe\cp{'e}}} \I{[spE."PE]} \emph{vtr.} capture (s.o.\slash s.t.). 
\B{Fol nga\-ti spo\-le\-'e a krr, nga lum\-pe ke ka\-ra?} When they captured you, why didn't you resist?~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tIspe'e}{\B{tì\-spe\-'e}}, capture. 
\on{2}.~\HL{spe'etu}{\B{spe\-'e\-tu}}, captive. 
\on{3}.~\HL{nongspe'}{\B{nong\-spe'}}, pursue with an intent to capture.

% spe'etu
\hi
\HT{spe'etu}{\B{spe\cp{'e}tu}} \I{[spE."PE.tu]} \emph{n.} captive. 
\B{Po spe\-'e\-tu lu oe\-yä.} He is \emph{my} captive.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/}{ln\slash a\textsuperscript{c}}
\B{Fol fte ay\-spe\-'e\-tut li\-vo\-nu fngo' 'u\-pet?} What are they demanding in order for them to release the captives?~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{spe'e}{\B{spe\-'e}})

% speng
\hi
\HT{speng}{\B{speng}} \I{[spEN]} \emph{vtr.} (\emph{a.}) restore. 
(\emph{b. metaphorically as} \B{spä\-peng}) restore oneself, get one's head back on straight.
\B{New oe ri\-vun a\-sim tì\-fnu\-nga'\-a tseng\-it a tsa\-ro tsun syi\-vor tsi\-vu\-rokx fte spä\-pi\-veng.} I want to find a quiet place nearby where I can chill out and rest to get my head back on straight.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}

% spono
\hi
\HT{spono}{\B{\cp{spo}no}} \I{["{}spo.no]} \emph{n.} island.   
\B{Ay\-oel ro\-lun mip\-a spo\-not mì hil\-van.} We found a new island in the river.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Oe\-ri so\-la\-lew wum zì\-sìt °a14 a krr, fo\-lrr\-fen spo\-not alo a\-'aw\-ve.} When I was about 12 years old I visited an island for the first time.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% spule
\hi
\HT{spule}{\B{\cp{spu}le}} \I{["{}spu.lE]} \emph{vtr.} propel.   
\B{Kä\-mak\-to nì\-win, ay\-nga\-ti spi\-vu\-le hu\-fwel.} You ride out as fast as the wind can carry you [lit.: ride out fast, may the wind propel you].~\LINK{http://forum.learnnavi.org/language-updates/naviya-the-other-vocative/}{ln\slash a} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{spulmokri}{\B{spul\-mok\-ri}}, telephone. 
\on{2}.~\HL{spulyaney}{\B{spul\-ya\-ney}}, canoe paddle.
\on{3}.~\HL{yIspul}{\B{yì\-spul}}, mermaid tail.
\on{4}.~\HL{ko'onspul}{\B{ko\-'on\-spul}} sunflower gigantus. 

% spulmokri
\hi
\HT{spulmokri}{\B{spul\cp{mok}ri}} \I{[spul."mok\textcorner.Ri]} \emph{or} \I{[spUl.]} \emph{n.} telephone.
\B{Syaw oer ru\-txe trr\-ay fa spul\-mok\-ri.} Please phone me tomorrow.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{spule}{\B{spu\-le}}, \HL{mokri}{\B{mok\-ri}})

% spulyaney
\hi
\HT{spulyaney}{\B{spulya\cp{ney}}} \I{[spul.ja."nEj]} \emph{or} \I{[spUl.]} \emph{n.} canoe paddle.
(\ding{43}~\HL{spule}{\B{spu\-le}}, \HL{yaney}{\B{ya\-ney}})

% spuwin
\hi
\HT{spuwin}{\B{\cp{spu}win}} \I{["{}spu.win]} \emph{adj.} old (\emph{not for people}), former. 
\B{Tsa\-tsko lu spu\-win ul\-te ke lu mi txur, slä oe\-ri txan\-wa\-we le\-iu.} That bow is old and no longer strong, but it means a lot to me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{lal}{\B{lal}})

% spxam
\hi
\HT{spxam}{\B{spxam}} \I{[sp'am]} \emph{n., funga} fungus. 
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{torukspxam}{\B{to\-ruk\-spxam}}, octoshroom.
\on{2}.~\HL{txepvispxam}{\B{txep\-vi\-spxam}}, sparkle pod. 

% spxin
\hi
\HT{spxin}{\B{spxin}} \I{[sp'in]} \emph{adj.} sick.
\B{Mìn, mìn, ken\-ten mìn. | Pe\-lun ke 'efu spxin?} Turn, turn, the fanlizard turns. | Why doesn't it feel sick?~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}
\B{Oe\-yä 'ey\-lan ke 'e\-fu spxin. Slä ke tslä\-ngam pol fu\-ta nì\-fkey\-to\-ngay lu spxin nì\-txan.} My friend doesn't feel sick. But he doesn't understand that he is in fact very sick.~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/\#comment-31207}{b}
(\ding{214}~\B{le\-fpom\-tokx}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIspxin}{\B{tì\-spxin}}, sickness, state of being sick. 
\on{2}.~\HL{sAspxin}{\B{sä\-spxin}}, sickness, disease, illness.
\on{3}.~\HL{spxintu}{\B{spxin\-tu}}, sick or ill person.

% spxintu
\hi 
\HT{spxintu}{\B{\cp{spxin}tu}} \I{["{}sp'in.tu]} \emph{n.} sick or ill person. 
\B{Lä\-ngu zam\-a spxin\-tu mì sray oe\-yä.} I'm sorry to say there are sixty\hyp{}four sick people in my village.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
(\ding{43}~\HL{spxin}{\B{spxin}})

% srak / srake
\hi
\HT{srak}{\B{srak}} \I{[sRak\textcorner]} \emph{or} \B{\cp{sra}ke} \I{["{}sRa.kE]} \emph{part. to indicate a yes\slash no\hyp{}question.}   
(\emph{a. at the end}) \B{Nga\-ru lu fpom srak?} How are you? [lit.: do you have well\hyp{}being?]~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Vanity_Fair}{wiki}
\B{Nga zo srak?} Are you well? (\emph{e.g. have you recovered from your illness\slash fall, etc.?})~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Nga za\-'u fì\-tseng pxìm srak?} Do you come here often?~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{wiki\slash ld}
\B{Ngal 'e\-raw\-nìm oe\-ti srak?} Are you avoiding me?~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Tsun (oe) mi\-vä\-kxu hì\-krr srak?} May I interrupt a moment?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\emph{b. at the beginning}) \B{Sra\-ke nga za\-'u?} Are you coming?~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/}{ln}
\B{Sra\-ke tsun oe fay\-upxa\-ret tsli\-vam nì\-ftue?} Can I understand these messages easily?~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln}
(\ding{43}~\HL{sran}{\B{sra\-ne}}, \HL{kehe}{\B{ke\-he}})

% srakat
\hi 
\HT{srakat}{\B{\cp{sra}kat}} \I{["{}sRa.kat\textcorner]} \emph{n., fauna} dinicthoid, \emph{large piranha\hyp{}like aquatic creature that lives in Pandora's lakes and murky lowland drainages}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln}

% sran / srane
\hi
\HT{sran}{\B{sran}} \I{[sRan]} \emph{or} \B{\cp{sra}ne} \I{["{}sRa.nE]} \emph{intj.} yes, (\emph{coll.}) yeah.
\B{Sra\-ne, fì\-txon tsun nga ni\-väk swo\-at nì\-'it, slä rä\-'ä rou!} Yes, you can have a little alcohol tonight, but don't get drunk!~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Nga 'e\-fu väng srak?---Sran, 'e\-fu nì\-txan.} Are you thirsty?---Yep, I'm parched.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{214}~\HL{kehe}{\B{ke\-he}}; \ding{43}~\HL{li}{\B{li}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{kesran}{\B{ke\-sran}}, so\hyp{}so, mediocre. 
\on{2}.~\HL{srak}{\B{srak\slash sra\-ke}}, \emph{yes\slash no particle}. 
\on{3}.~\HL{srankehe}{\B{sran\-ke\-he}}, more or less.

% srankehe
\hi 
\HT{srankehe}{\B{sran\cp{ke}he}} \I{[sRan."kE.hE]} \emph{or} \I{[sRaN."kE.hE]} \emph{part., intj.} more or less, somewhat, yes and no, kind of (\emph{an equivocal response to a yes\slash no question, so as not to commit oneself}). 
\B{Sra\-ke fay\-sä\-fpìl lu pum ngey nì\-wotx?--Sran\-ke\-he.} Are all these ideas your own?--More or less.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\HL{sran}{\B{sran(e)}}, \HL{kehe}{\B{ke\-he}})

% sraw
\hi
\HT{sraw}{\B{sraw}} \I{[sRaw]} \emph{adj.} painful. \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsraw}{\B{tì\-sraw}}, pain.

% srä
\hi
\HT{srA}{\B{srä}} \I{[sR\ae]} \emph{n.} cloth (\emph{a piece of cloth woven on a loom}).
\B{Tsa\-srä a\-naw\-nekx ley nì\-'it nì\-'aw.} That burnt cloth is of little value.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
\B{Fu\-ri\-a txu\-la tsa\-lew\-ti lu srä sìl\-tsan to fngap.} For constructing that cover, woven cloth is better than metal.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
\B{Fu\-ta nga\-ta tel pxen\-yoa srät, mll\-'ei\-an oel.} I'm happy to take cloth from you for finished garments.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b} 

% srätx
\hi
\HT{srAtx}{\B{srätx}} \I{[sR\ae{}t']} \emph{vtr.} bother, annoy.   
\B{Oe\-ti rä\-'ä srätx.} Don't bother me.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\emph{a. expression}) \B{Sra\-ke srätx (ngat) txo \ldots?} (\emph{with} ‹iv›) `Would you mind if \ldots?'~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
\B{Ke srätx (oet) txo \ldots} `I don't mind if \ldots'~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}       
\B{Ke srätx kaw\-'it!\slash Kea sä\-srätx kaw\-'it!\slash Ke\-he kaw\-'it!} `Not at all!'~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAsrAtx}{\B{sä\-srätx}}, annoyance.

% sre
\hi
\HT{sre}{\B{sre}} \I{[sRE]} \emph{adp.}\texttt{+} before (\emph{temporal}). (\emph{a.}) \B{Sre fwa sìn tskxe\-pay tì\-ran, ze\-ne fko flì\-nutx\-it sti\-ve\-ftxaw.} It's necessary to check the thickness of the ice before walking on it.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\emph{b. of telling time}) \B{trr\-pxì\-vi a\-vo\-mun sre srr\-pxì a\-mrr\-ve} \emph{or coll.} \B{vo\-mun sre mrr\-ve}, (it's) \on{4}:\on{50}, or `ten to five'.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
(\emph{c. with} \B{li}) by; before or up to, but not after. 
\B{Kem si li trr\-ay\-sre.} Do it by tomorrow. (\emph{also} \ding{43}~\HL{lisre}{\B{li\-sre}})
(\emph{d. idiom}) \B{sre fwa sngap zi\-ze'}, ``as quickly as possible'' [lit: before the hellfire wasp stings].~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{214}~\HL{maw}{\B{maw}}) \\
\textsc{Derivations}:
\on{1}.~\HL{lisre}{\B{li\-sre}}, by; up to, but not after. 
\on{2}.~\HL{srefey}{\B{sre\-fey}}, expect. 
\on{3}.~\HL{srefpIl}{\B{sre\-fpìl}}, assume, presume. 
\on{4}.~\HL{srese'a}{\B{sre\-se\-'a}}, prophesize, predict. 
\on{5}.~\HL{srefwa}{\B{sre\-fwa}}, before (\emph{conj.}). 
\on{6}.~\HL{srekrr}{\B{sre\-krr}}, before (\emph{adv.}). 
\on{7}.~\HL{srekamtrr}{\B{sre\-kam\-trr}}, time before noon.
\on{8}.~\HL{srekamtxon}{\B{sre\-kam\-txon}}, time before midnight. 
\on{9}.~\HL{sresrr'ong}{\B{sre\-srr\-'ong}}, before dawn. 
\on{10}.~\HL{sreton'ong}{\B{sre\-ton\-'ong}}, dusk.

% sre'
\hi
\HT{sre'}{\B{sre'}} \I{[sREP]} \emph{n.} tooth. 
\B{Ay\-sre' lom lu tsa\-ko\-ak\-tan\-ur.} The old man misses his teeth.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Oel tstal\-it fyo\-lep fa aysre'.} I held the knife in my teeth.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}

% srefey
\hi
\HT{srefey}{\B{sre\cp{fey}}} \I{[sRE."f$\cdot$Ej]} \emph{vtr.}\slash \emph{vin.}\on{22} expect (\emph{takes} \HL{tsnI}{\B{tsnì}} \emph{when used as vin.}).
\B{Set sre\-fey oel fu\-ta tsam\-po\-ngu tä\-txaw maw txon\-'ong.\slash Set sre\-fey oe tsnì tsam\-po\-ngu …} I'm currently expecting the war party back after nightfall.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\emph{a. idiom}) \B{sre\-fe\-re\-i\-ey nì\-prr\-te'}, looking forward (to s.t.).~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} \B{Tsa\-ria nga\-hu ye'\-rìn ul\-txa si nì\-mun, oe sre\-fe\-re\-i\-ey nì\-prr\-te'.} I'm looking forward to getting together with you again soon.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\ding{43}~\HL{sre}{\B{sre}}, \HL{pey}{\B{pey}})

% srefpìl
\hi
\HT{srefpIl}{\B{sre\cp{fpìl}}} \I{[sRE."fp$\cdot$Il]} \emph{vtr.}\slash \emph{vin., inf.}\on{22} assume, presume.
(\emph{a. vtr.}) \B{Sre\-fpìl oel fu\-ta nga lu tok\-tor Lì\-vìng\-sì\-ton.} Doctor Livingstone, I presume.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\emph{a. vin. used with} \HL{tsnI}{\B{tsnì}}) \B{Sre\-fpìl Oma\-ti\-ka\-ya tsnì Tsyeyk kaw\-krr ke ta\-yä\-txaw maw ka\-vuk sne\-yä.} The Omaticaya assumed that Jake would never return after his treachery.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{sre}{\B{sre}}, \HL{fpIl}{\B{fpìl}})

% srefwa
\hi
\HT{srefwa}{\B{\cp{sre}fwa}} \I{["{}sRE.fwa]} \emph{conj.} before.
\B{Sre\-fwa oe hum, new pi\-vll\-txe.} Before I leave, I want to speak.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{sre}{\B{sre}}, \HL{fwa}{\B{fwa}}, \ding{214}~\HL{mawfwa}{\B{maw\-fwa}})

% srekamtrr
\hi
\HT{srekamtrr}{\B{sre\cp{kam}trr}} \I{[sRE."kam.t\s{r}]} \emph{n.}; \emph{adv.} before midday, before noon.
\B{Fo pä\-hem trr\-pxì\-pe?--Sre\-kam\-trr.} What part of the day will they arrive?--Before noon.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
(\ding{43}~\HL{sre}{\B{sre}}, \HL{kxamtrr}{\B{kxam\-trr}})

% srekamtxon
\hi
\HT{srekamtxon}{\B{sre\cp{kam}txon}} \I{[sRE."kam.t'on]} \emph{n.}; \emph{adv.} before midnight.
(\ding{43}~\HL{sre}{\B{sre}}, \HL{kxamtxon}{\B{kxam\-txon}})

% srekrr
\hi
\HT{srekrr}{\B{sre\cp{krr}}} \I{[sRE."k\s{r}]} \emph{adv.} before (\emph{temporal}), beforehand.  
\B{Sre\-krr 'a\-me\-fu väng, set ye\-väng.} Before, I was thirsty; now my thirst has been quenched.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Fay\-upxa\-re\-mì oe pa\-yäng\-kxo te\-ri ho\-ren lì'\-fya\-yä le\-Na'\-vi fpi su\-te a tsun sre\-krr tsat si\-var.} In these messages I will chat about the rules of the Na'vi language.for people who can already use it.~\LINK{http://naviteri.org/2010/06/irayo-thank-you-and-some-miscellaneous-thoughts/}{b}
\B{Tseyk tswa\-may\-on fa ik\-ran sre\-krr; ta\-fral fmo\-li fì\-kem si\-vi fa to\-ruk nì\-steng.} Jake flew with an ikran before; therefore he tried to do it with a \emph{toruk} in a similar fashion.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\ding{43}~\HL{sre}{\B{sre}}, \HL{krr}{\B{krr}})

% srer
\hi
\HT{srer}{\B{srer}} \I{[sRER]} \emph{vin.} appear, materialize, come into view.
\B{Prr\-nen lrr\-tok si a krr, srer me\-song\-tsyìp a\-ho\-na.} When the baby smiles, two adorable dimples appear.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{214}~\HL{'Ip}{\B{'ìp}})

% srese'a
\hi
\HT{srese'a}{\B{srese\cp{'a}}} \I{[sRE.s$\cdot$E."P$\cdot$a]} \emph{vtr., inf.}\on{23} prophesy, predict. 
\B{Tsun fko ay\-on\-ti fì\-wopx\-ä ni\-vìn fte ya\-fkeyk\-it sre\-si\-ve\-'a.} The shapes of clouds can be used to predict the weather.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{sre}{\B{sre}}, \HL{tse'a}{\B{tse\-'a}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAsrese'a}{\B{sä\-sre\-se\-'a}}, prediction.

% sresrr'ong
\hi
\HT{sresrr'ong}{\B{sresrr\cp{'ong}}} \I{[sRE.s\s{r}."PoN]} \emph{n.}; \emph{adv.} before daybreak, before dawn.
\B{Sa'\-nok 'eveng\-ur tì\-txen si sre\-srr\-'ong.} Mother woke the child in the early morning.~\LINK{http://forum.learnnavi.org/language-updates/little-confirmation-titxen-si/}{ln}
(\ding{43}~\HL{sre}{\B{sre}}, \HL{trr'ong}{\B{trr\-'ong}})

% sreton'ong
\hi
\HT{sreton'ong}{\B{sreton\cp{'ong}}} \I{[sRE.ton."oN]} \emph{n.}; \emph{adv.} before darkness falls, before night unfolds.
(\ding{43}~\HL{sre}{\B{sre}}, \HL{txon'ong}{\B{txon\-'ong}})

% srew
\hi
\HT{srew}{\B{srew}} \I{[sREw]} \emph{vin.} dance. 
\B{Zun tom\-pa zì\-ye\-vup trr\-ay, zel fo srì\-ye\-vew.\slash … zel fo srew.} If it rained tomorrow, they'd do a dance.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}

% srey
\hi
\HT{srey}{\B{srey}} \I{[sREj]} \emph{n.} version.
\B{Fì\-srey puk\-ä lu swey nì\-law.} This version of the book is clearly the best.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Lu srey a\-pup ta Nga\-ri zì\-sìt a\-mip li\-vu le\-fpom.} It's a short version of \emph{Ngari zìsìt amip livu lefpom}.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/\#comment-27964}{b}

% srìn
\hi
\HT{srIn}{\B{srìn}} \I{[sRIn]} \emph{vtr.} temporarily transfer from one to another (\emph{mainly used in its derivational forms}). \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{kAsrIn}{\B{kä\-srìn}}, lend. 
\on{2}.~\HL{zasrIn}{\B{za\-srìn}}, borrow. 
\on{3}.~\HL{sAsrIn}{\B{sä\-srìn}}, lent\slash borrowed thing.

% srok
\hi
\HT{srok}{\B{srok}} \I{[sRok\textcorner]} \emph{n.} bead (\emph{decorative}).
\B{'En si oe, lor\-a tsa\-fkxi\-le\-ri a\-pxay\-opin so\-lar Tse\-nul srok\-it a\-vo\-zam.} I would guess that Tsenu used a thousand (lit. \on{512}) beads for that beautiful multi\hyp{}colored bib necklace.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
\B{Fol ko\-la\-nom po\-ta ay\-srok\-it fay\-oang\-yoa.} They acquired beads from him in exchange for fish.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}

% sru'
\hi
\HT{sru'}{\B{sru'}} \I{[sRuP]} \emph{or} \I{[sRUP]} \emph{vtr.} crush, trample.   \B{Weyn\-flit\-it 'ang\-tsìk\-ìl sro\-lu' tspang.} Wainfleet was crushed and killed by a hammerhead.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}

% srung / srung si
\hi
\HT{srung}{\B{srung}} \I{[sRuN]} \emph{or} \I{[sRUN]} (\textsc{rn}: \B{srùng} \I{[sRUN]}) \textsc{i}. \emph{n.} help, assistance. \\
\textsc{ii}. \HT{srungsi}{\B{srung si}} \I{[sRuN si]} (\textsc{rn}: \B{srùng si} \I{[sRUN si]}) \emph{vin.} (\emph{a.}) help (s.o.). 
\B{Ma Ra\-lu, srung si por!} Ralu, help her!~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Txan\-a srung so\-li po oer.} He helped me a lot.~\LINK{http://forum.learnnavi.org/language-updates/si-verbs-modification/}{ln}
\B{Tsa\-tì\-oeyk\-tìng a\-ya\-yayr\-nga' srung ke so\-li oer fte tsli\-vam teyng\-ta kem\-pe ze\-ne si\-vi.} That confusing explanation didn't help me understand what I have to do.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{b. of people}) helpful. \B{tu\-te a srung si\slash srung si a tu\-te}, a helpful person.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\  
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{srungsiyu}{\B{srung\-si\-yu}}, assistant, helper. 
\on{2}.~\HL{srungtsyIp}{\B{srung\-tsyìp}}, helpful hint, tip. 
\on{3}.~\HL{srunga'}{\B{sru\-nga'}}, helpful.
\on{4}.~\HL{sAsrungsiyu}{\B{srung\-si\-yu}}, assistant.
\on{5}.~\HL{sAsrung}{\B{sä\-srung}}, helper (inanimate).

% srunga' 
\hi
\HT{srunga'}{\B{\cp{sru}nga'}} \I{["sRu.NaP]} (\textsc{rn}: \B{\cp{srù}nga'} \I{["sRU.NaP]}) \emph{adj., nfp.} helpful.
\B{Tì\-yäkx ke lu sru\-nga', ma tsmuk. Nga txo sti, oeyk\-tìng teyng\-ta pe\-lun.} Snubbing isn't helpful, brother. If you're angry, explain why.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\emph{a. proverb}) \B{Hem a\-sru\-nga' nì\-'ul, hum a\-sngu\-nga' nì\-nän.}
More helpful actions lead to less troubling outcomes.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\ding{43}~\HL{srung}{\B{srung}}, \HL{-nga'}{\B{-nga'}}, \HL{srung}{\B{srung si} (\emph{b. of people})})

% srungsiyu
\hi
\HT{srungsiyu}{\B{\cp{srung}siyu}} \I{["sRuN.si.ju]} \emph{or} \I{["sRUN.]} \emph{n.} assistant, helper, one who helps.
\B{Hu\-fwa lu oe fì\-trr kar\-yu a\-txin, le\-iu oer kop pxe\-srung\-si\-yu a fì\-lì'\-fya\-ri slo\-lu tsul\-fä\-tu.} Although I'm the main teacher today, I'm happy to say I also have three assistants who have become masters of the language.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\ding{43}~\HL{srung}{\B{srung si}}, \HL{-yu}{\B{-yu}})

% srungtsyìp
\hi
\HT{srungtsyIp}{\B{\cp{srung}tsyìp}} \I{["{}sRuN.\t{ts}jIp\textcorner]} (\textsc{rn}: \B{\cp{srùng}tsyìp} \I{["sRUN.\t{tS}Ip\textcorner]})  \emph{n.} helpful hint, tip.
\B{Srung\-tsyìp a\-1ve}, 1st tip.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{43}~\HL{srung}{\B{srung}}, \HL{-tsyIp}{\B{-tsyìp}})

% stang
\hi 
\HT{stang}{\B{stang}} \I{[staN]} \emph{n.} torso. 

% starlì'u
\hi 
\HT{starlI'u}{\B{\cp{star}lì'u}} \I{["{}staR.lI.Pu]} \emph{n.} adposition. 
(\ding{43}~\HL{sAtare}{\B{sä\-ta\-re}}, \HL{lI'u}{\B{lì\-'u}})

% starsìm
\hi
\HT{starsIm}{\B{\cp{star}sìm}} \I{["{}staR.sIm]} \emph{vtr.} gather, collect.
\B{Aw\-nga tsong\-ne ki\-vä fte sti\-var\-sìm tey\-lut.} Let's go to the valley to gather beetle larvae.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAstarsIm}{\B{sä\-star\-sìm}}, collection.

% stawm
\hi
\HT{stawm}{\B{stawm}} \I{[stawm]} \emph{vtr.} (\texttt{-}\emph{control}) hear (s.t.), hear about s.t. (\B{te\-ri}).
\B{Txo fko ti\-vul mì na'\-rìng nì\-hawm\-pam, stawm ay\-ioang.} If you run noisily in the forest, the animals will hear.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Fì\-slär\-mì tsun fko sti\-vawm ngam\-it a\-pxay.} You can hear a lot of echoes in this cave.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Fpo\-le' ay\-ngal oer fì\-txan nì\-ftxa\-vang a 'upxa\-ret sto\-lawm oel.} I have heard the message you have sent me so passionately.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Fol oe\-yä tì\-pawm\-it ke sto\-lä\-ngawm.} Unfortunately they didn't hear my question.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Krr a sto\-lawm Txe\-wì te\-ri hem a poe\-ru lo\-len, 'o\-le\-fu ke\-ftxo nì\-txan.} When Txewì heard about what happened to her, he felt very upset.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{yune}{\B{yu\-ne}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{stawmtswo}{\B{stawm\-tswo}}, (sense of) hearing.

% stawmtswo
\hi
\HT{stawmtswo}{\B{\cp{stawm}tswo}} \I{["{}stawm.\t{ts}wo]} \emph{n.} (sense of) hearing. 
\B{Tsa\-ko\-ak\-te\-ri stawm\-tswo lu fe'. Po\-hu a tì\-päng\-kxo ngä\-zìk lu nì\-txan.} That old woman's hearing is poor. A conversation with her is very difficult.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\ding{43}~\HL{stawm}{\B{stawm}}, \HL{-tswo}{\B{-tswo}})

% stä'nì
\hi
\HT{stA'nI}{\B{\cp{stä'}nì}} \I{["{}st\ae{}P.nI]} \emph{vtr.} catch. 
\B{Oe\-yä txin\-tìn lu fwa stä'\-nì fay\-oang\-it.} My central societal occupation is to catch fish.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Fì\-ta\-ron\-yu\-tsyìp ke tsun ke\-'ut sti\-vä'\-nì.} This little hunter can't catch anything.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{stA'nIpam}{\B{stä'\-nì\-pam}}, recording.

% stä'nìpam
\hi 
\HT{stA'nIpam}{\B{\cp{stä'}nìpam}} \I{["st\ae{}P.nI.pam]} \textsc{i}. \emph{n.} recording. 
\B{Fì\-vur\-way\-ri a\-lor sì stä'\-nì\-pam\-ìri tse\-yä ira\-yo nì\-txan, ma mesmuk!} Thank you two very much for this beautiful narravite poem and its recording!
(\ding{43}~\HL{stA'nI}{\B{stä'\-nì}}, \HL{pam}{\B{pam}}) \\
\textsc{ii}. \B{\cp{stä'}nìpam si} \I{["st\ae{}P.nI.pam si]} \emph{vin.} record, make a recording.~\LINK{http://naviteri.org/2019/10/vurway-alor-a-beautiful-narrative-poem-2/}{b}
\B{Sä\-ftxu\-lì\-'u a\-tì\-txur\-nga' nì\-txan nang! Fu\-ria tsa\-ru nga stä'\-nì\-pam so\-li, ira\-yo.} What a powerful speech! Thank you for recording it.~\LINK{http://naviteri.org/2019/10/vurway-alor-a-beautiful-narrative-poem-2/}{b}

% steftxaw
\hi
\HT{steftxaw}{\B{ste\cp{ftxaw}}} \I{[stE."ft'aw]} \emph{vtr.} examine, check. 
\B{Sre fwa sìn tskxe\-pay tì\-ran, ze\-ne fko flì\-nutx\-it sti\-ve\-ftxaw.} It's necessary to check the thickness of the ice before walking on it.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b} 
\B{Tsa\-swi\-rä\-ti lo\-nu! Ay\-nga ne\-to ri\-vikx! Fì\-ke\-tu\-wong\-ti oel stì\-ye\-ftxaw.} Release this creature! Step back! I will look at this alien.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{wiki\slash a}
(\ding{43}~\HL{lang}{\B{lang}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsteftxaw}{\B{tì\-ste\-ftxaw}}, examination.

% steng
\hi
\HT{steng}{\B{steng}} \I{[stEN]} \emph{adj.} similar.   
\B{Me\-unil\-tì\-ran\-tokx Tok\-tor Kì\-rey\-sä sì Tsyeyk\-ä ke lu teng ki steng.} The avatars of Grace and Jake aren't the same, but they are similar.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Nga\-ru tì\-yawr. Lo\-lu oer steng\-a sä\-fpìl.} You're right. I had a similar idea.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\emph{a. proverb}) \B{Sä\-fpìl a\-steng, tì\-kan a\-teng.} Great minds think alike [lit.: similar thought, same aim].~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIsteng}{\B{nì\-steng}}, similarly.

% steyki
\hi
\HT{steyki}{\B{stey\cp{ki}}} \I{[stEj."k$\cdot$i]} \emph{vtr., inf.}\on{22} anger (s.o.), make (s.o.) angry. 
\B{Ngal kaw\-krr sngel\-it ke mu\-nge wrr\-pa\-ne ul\-te nì\-tut ze\-ne ätxä\-le sir\-vä\-ngi ngar oe, stey\-ki fì\-txan.} Your never taking out the trash and my having to constantly ask you to do it makes me so angry.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Apxa tì\-väng\-ìl po\-ti stey\-ko\-li.} (His) great thirst made him angry.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Nge\-yä fay\-sä\-fngo'\-ìl nì\-wotx stey\-ke\-rä\-ngi oe\-ti nì\-hawng.} All these demands of yours are making me exceedingly angry.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{sti}{\B{sti}})

% sti
\hi
\HT{sti}{\B{sti}} \I{[sti]} \emph{vin.} angry, be angry. 
\B{Tì\-yäkx ke lu sru\-nga', ma tsmuk. Nga txo sti, oeyk\-tìng teyng\-ta pe\-lun.} Snubbing isn't helpful, brother. If you're angry, explain why.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Oe\-ru txoa li\-vu, ma 'ey\-lan. Rä\-'ä sti\-vi. Lu hì\-'i\-a sä\-'i\-pu nì\-'aw.} I'm sorry, friend. Don't be angry. It was just a small bit of humor.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Fu\-ri\-a fì\-lì'\-fya\-vit fkol ta 'Ìng\-lì\-sì mo\-lu\-nge, sìl\-pey oe tsnì ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ke stì\-ye\-vi.} That this expression is brought from English, will hopefully not anger the members of the language group.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{steyki}{\B{stey\-ki}}, make (s.o.) angry. 
\on{2}.~\HL{tIsti}{\B{tì\-sti}}, anger. 
\on{3}.~\HL{nIsti}{\B{nì\-sti}}, angrily.

% stiwi
\hi 
\HT{stiwi}{\B{\cp{sti}wi}} \I{["{}sti.wi]} \textsc{i}. \emph{n.} mischief. \\
\textsc{ii}. \B{\cp{sti}wi si} \I{["{}sti.wi si]} \emph{vin.} be naughty, do mischief.
\B{Sti\-wi rä\-'ä si, ma 'e\-veng! U\-van si mì se\-ngo a\-la\-he.} Don't be naughty, child! Play somewhere else.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{stiwinga'}{\B{sti\-wi\-nga'}}, mischievous. 
\on{2}.~\HL{stiwiyu}{\B{sti\-wi\-yu}}, mischief\hyp{}maker.

% stiwinga'
\hi 
\HT{stiwinga'}{\B{\cp{sti}winga'}} \I{["{}sti.wi.NaP]} \emph{adj.} mischievous.
(\ding{43}~\HL{stiwi}{\B{sti\-wi}})

% stiwisiyu
\hi 
\HT{stiwisiyu}{\B{\cp{sti}wisiyu}} \I{["{}sti.wi.si.ju]} \emph{n.} mieschief\hyp{}maker. 
(\ding{43}~\HL{stiwi}{\B{sti\-wi}})

% sto
\hi
\HT{sto}{\B{sto}} \I{[sto]} \emph{vtr.} (\emph{a.}) refuse (s.t.).
\B{Po ätxä\-le so\-li nì\-ho\-na fì\-txan, ke tsun oe sti\-vo.} She asked so sweetly that I couldn't refuse.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Sto\-lo oel stxe\-nut pe\-yä.} I refused his offer.~\LINK{http://forum.learnnavi.org/language-updates/sto-has-the-same-syntax-as-new/}{ln}
(\emph{b. modal; either with} ‹\B{iv}› \emph{or} \HL{futa}{\B{fu\-ta}} \emph{and} ‹\B{iv}›) refuse (to do s.t.). 
\B{Sto\-lo po hi\-vum fo\-hu.} She refused to leave with them.~\LINK{http://forum.learnnavi.org/language-updates/sto-has-the-same-syntax-as-new/}{ln}
\B{Po\-el sto\-la\-tso fu\-ta me\-fo ti\-va\-ron tsa\-ha'\-ngir.} She must have refused their request to hunt that afternoon.~\LINK{http://forum.learnnavi.org/language-updates/sto-has-the-same-syntax-as-new/}{ln}

% stum
\hi
\HT{stum}{\B{stum}} \I{[stum]} \emph{or} \I{[stUm]} (\textsc{rn}: \B{stùm} \I{[stUm]})  \emph{adv.} (\emph{a.}) almost. 
\B{Poe\-ri unil\-tì\-ran\-tokx\-it tar\-mok a krr lam stum nì\-ay\-fo.} When she is in her Avatar body, she appears almost like them.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Lì'\-fya\-ri tsa\-taw\-tu\-te pe\-yì?---Tìm\-yì. Pol ke tslam stum ke\-'ut, omum lì\-'ut a\-vol nì\-'aw.} How is that Sky Person's Na'vi?---Pretty bad. He understands almost nothing and only knows eight words.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\emph{b. negated}) hardly, barely, scarcely.
\B{Ru\-txe pi\-vll\-txe nì\-wok nì\-'it, oel nga\-ti stum ke stä\-ngawm krr\-a nga fì\-fya tsye\-rì\-syì!} Please speak up a bit, I can barely hear you when you're whispering like this!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}

% stxang
\hi 
\HT{stxang}{\B{stxang}} \I{[st'aN]} \emph{n.} axe, hatchet, tomahawk. 

% stxeli
\hi
\HT{stxeli}{\B{\cp{stxe}li}} \I{["{}st'E.li]} \emph{n.} gift, present.   
\B{Ftxo\-zä\-ri fì\-tì\-ngop a\-lor lu swey\-a stxe\-li a to\-lel oel.} This beautiful creation is the best present I have received for (my birthday) celebration.~\LINK{http://forum.learnnavi.org/community-calendar/karyu-pawls-birthday/msg312733/\#msg312733}{ln}
\B{Tsa\-tsko lu spu\-win ul\-te ke lu mi txur, slä oe\-ri txan\-wa\-we le\-iu. Lu stxe\-li a sem\-pul\-ta.} That bow is old and no longer strong, but it means a lot to me. It was a gift from my father.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Fì\-me\-stxe\-li a\-lor kur set ta kxem\-yo a mì hel\-ku moe\-yä.} These two beautiful gifts are now hanging on a wall in our home.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
\B{Stxe\-lit fpo\-le' oel nga\-ru.} I sent the gift to you.~\LINK{http://forum.learnnavi.org/language-updates/i-have-something-to-say/}{ln}
\B{Nge\-yä stxe\-li\-ri a\-lor oe new nga\-ru pi\-vll\-txe san ira\-yo sìk nì\-li.} I want to thank you in advance for your beautiful gift.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
 \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{stxenu}{\B{stxe\-nu}}, offer.

% stxenu
\hi
\HT{stxenu}{\B{\cp{stxe}nu}} \I{["{}stE.nu]} \emph{n.} offer. 
\B{Ngal lum\-pe oe\-yä stxe\-nut tsyo\-lä\-ngär?} Why did you reject my offer?~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Po pol\-txe san tì\-tstew\-nga'\-a stxe\-nu\-ri ira\-yo sei\-yi oe nga\-ru nì\-txan.} She said, ``I thank you very much for your courageous offer.''~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{stxeli}{\B{stxe\-li}}, \HL{lonu}{\B{lo\-nu}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{stxenutIng}{\B{stxe\-nu\-tìng}}, offer (s.t.).

% stxenutìng
\hi
\HT{stxenutIng}{\B{\cp{stxe}nutìng}} \I{["{}stE.nu.t$\cdot$IN]} \emph{vtr., inf.}\on{33} offer (s.t.).
\B{Oe\-ri tì\-rey\-ti oel stxe\-nu\-to\-lìng fpi olo' aw\-nge\-yä.} I offered my life for the sake of our clan.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Stxe\-nu\-to\-lìng oel fu\-ta lì'\-fyat le\-Na'\-vi poe\-ru ki\-var. Mol\-'an nì\-prr\-te'.} I offered to teach her Na’vi. She accepted gladly.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{stxenu}{\B{stxe\-nu}}, \HL{tIng}{\B{tìng}})

% stxong
\hi
\HT{stxong}{\B{stxong}} \I{[st'oN]} \emph{adj.} strange, unfamiliar, unknown (\emph{usually negative connotation, potentially threatening or dangerous}).   
\B{Lar\-mu tsa\-tsam\-si\-yu\-hu tì\-vawm a lu stxong ay\-oer.} That warrior carried with him a darkness unknown to us.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Fì\-na\-er\-ìri ngal ew\-ku 'u\-ot a\-stxong srak?} Do you taste something strange in this drink?~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}

% sulìn
\hi
\HT{sulIn}{\B{\cp{su}lìn}} \I{["{}su.lIn]} \emph{vin.} be busy (\emph{positive sense}), \emph{be engrossed in s.t. one finds especially pleasant and energizing}.
\B{Pam\-tseo\-ri po su\-lä\-ngìn nì\-hawng.} He is overly engrossed in his music (and I'm displeased about it).~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{'In}{\B{'ìn}}, \ding{214}~\HL{vrrIn}{\B{vrr\-ìn}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sAsulIn}{\B{sä\-su\-lìn}}, hobby.

% sum
\hi
\HT{sum}{\B{sum}} \I{[sum]} \emph{or} \I{[sUm]} \emph{n.} shell (\emph{from the ocean}). 
(\ding{43}~\HL{lak}{\B{lak}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{sumsey}{\B{sum\-sey}}, drinking vessel (made of shell).

% sumsey
\hi
\HT{sumsey}{\B{\cp{sum}sey}} \I{["{}sum.sEj]} \emph{or} \I{["{}sUm.]} \emph{n.} drinking vessel (\emph{made of shell}).
(\ding{43}~\HL{sum}{\B{sum}}, \HL{sey}{\B{sey}})

% sunkesun
\hi 
\HT{sunkesun}{\B{\cp{sun}kesun}} \I{["{}sun.kE.sun]} \emph{or} \I{["{}sUn.kE.sUn]} \emph{adv.} ``like it or not'' (\emph{default addressee is ``you''}). 
\B{Sun\-ke\-sun po sla\-yu olo'\-eyk\-tan.} Whether you like it or not, he's going to become chief.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{sunu}{\B{su\-nu}}, \HL{ke}{\B{ke}})

% sunu
\hi
\HT{sunu}{\B{\cp{su}nu}} \I{["{}su.nu]} \emph{vin.} be pleasing or likeable, bring enjoyment.   
\B{Su\-nu oe\-ru tey\-lu.} I like \emph{teylu}.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Fra\-por a ke su\-nu por zop\-lo si.} He insults everyone he doesn't like.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
\B{Su\-nu oe\-ru fwa fì\-tseng\-it te\-rok oel nì\-mun.} I'm happy to be here again.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/}{b}
\B{Tì\-'o'\-ìri pe\-u su\-nu ngar fra\-to?} What is your favorite way to have fun?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Ke\-tsran fya\-'o si\-vu\-nu ngar, kem si.} Do it however you'd like.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Ul\-te sìl\-pey oe, fì\-'upxa\-ret inan a fì\-'u sil\-vu\-nu ay\-nga\-ru nì\-wotx!} And I hope, you all liked to read this message.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{txasunu}{\B{txa\-su\-nu}}, love greatly, enjoy tremendously. 
\on{2}.~\HL{sunkesun}{\B{sun\-ke\-sun}}, ``like it or not''.

% sung 
\hi
\HT{sung}{\B{sung}} \I{[suN]} \emph{or} \I{[sUN]} (\textsc{rn}: \B{sùng} \I{[sUN]}) \emph{vtr.} add. 
\B{Fko ze\-ne si\-vung tsat tsa\-tseng.} One must add that there.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/comment-page-1/\#comment-2476}{b}
\B{Oel lum\-pe tsat so\-lung mì upxa\-re?} Why did I add that in the message?~\LINK{http://naviteri.org/2010/07/vocabulary-update/comment-page-1/\#comment-95}{b}
\B{Ru\-txe oe\-yä tì\-oeyk\-tìng\-it a\-saw\-nung ni\-vìn.} Please look at my added explanation.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-581}{b}
(i.) \B{Trr A\-saw\-nung}, Leap Day.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIsung}{\B{nì\-sung}}, besides, additionally, furthermore. 
\on{2}.~\HL{tIsung}{\B{tì\-sung}}, addition, post script. 
\on{3}.~\HL{'opinsung}{\B{'o\-pin\-sung}}, color, color in.

% sur
\hi
\HT{sur}{\B{sur}} \I{[suR]} \emph{or} \I{[sUR]} \emph{n.} (\emph{a. sensation}) taste, flavor. 
(\emph{b. to form the middle voice}) \B{Fì\-na\-er\-ìri sur fkan oe\-ru ka\-lin.} This drink tastes sweet to me.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Fì\-na\-er\-ìri sur fkan lor.} This drink tastes good.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}

% susyang
\hi
\HT{susyang}{\B{su\cp{syang}}} \I{[su."sjaN]} \emph{adj.} fragile, delicate. 
\B{Lu fì\-vul su\-syang nì\-txan. Txo fko ti\-vìng zek\-wä kxakx.} This branch is very fragile. If you touch it, it'll break.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Rey\-'eng lu su\-syang.} The balance of life is fragile.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}

% sutx
\hi
\HT{sutx}{\B{sutx}} \I{[sut']} \emph{or} \I{[sUt']} \emph{vtr.} track, follow. 
\B{Po spä few pay\-fya fte smar\-it si\-vutx.} He jumped across the stream to track his prey.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Tì\-tu\-sa\-ron\-ìri fte fli\-vä, ze\-ne fko si\-vutx smar\-it ni\-nan nì\-no.} To succeed at hunting, you have to track your prey by reading (the forest) with attention to detail.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{fngapsutxwll}{\B{fngap\-sutx\-wll}}, metalfollowing plant.

% swapxì
\hi 
\HT{swapxI}{\B{swa\cp{pxì}}} \I{[swa."p'I]} \emph{n.} family member. 
\B{Ay\-swapx\-ìl oe\-yä tok fì\-tseng\-it nì\-wotx.} All the members of my family are here.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
(\ding{43}~\HL{soaia}{\B{soaia}}, \HL{hapxì}{\B{ha\-pxì}}) 

% swaran
\hi 
\HT{swaran}{\B{\cp{swa}ran}} \I{["{s}wa.Ran]} \emph{adj.} humble, modest, self\hyp{}effacing. 
\B{Tsam\-si\-yu a\-sìl\-tsan lu tstew slä\-kop swa\-ran.} A good warrior is courageous but also humble.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIswaran}{\B{tì\-swa\-ran}}, humility, humbleness.

% swaw
\hi
\HT{swaw}{\B{swaw}} \I{[swaw]} \emph{n.} moment. 
\B{Mì na'\-rìng Tsyeyk\-ìl lo\-re\-yu\-ti 'o\-lam\-pi a tsa\-swaw wou oer nì\-wotx!} That instant when Jake touched the \emph{Helicoradiam Spirale} in the forest I was totally amazed.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{pxisawam}{\B{pxi\-swaw\-am}}, just a moment ago. 
\on{2}.~\HL{pxiswaway}{\B{pxi\-swaw\-ay}}, in just second (from now). 
\on{3}.~\HL{swawtsyIp}{\B{swaw\-tsyìp}}, little moment. 

% swawtsyìp 
\hi
\HT{swawtsyIp}{\B{\cp{swaw}tsyìp}} \I{["{}swaw.\t{ts}jIp\textcorner]} \emph{n.} little moment, second.
\B{'Aw\-a swaw\-tsyìp. Oe fpe\-rìl.} Just a moment. I'm thinking.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{'Aw\-a swaw\-tsyìp li\-vu oer.} May I have a tiny moment.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{swaw}{\B{swaw}}, \HL{-tsyIp}{\B{-tsyìp}})

% swaynivi
\hi
\HT{swaynivi}{\B{\cp{sway}nivi}} \I{["{}swaj.ni.vi]} \emph{n.} family hammock. 
(\ding{43}~\HL{soaia}{\B{soaia}}, \HL{nivi}{\B{ni\-vi}})

% swek
\hi
\HT{swek}{\B{swek}} \I{[swEk\textcorner]} \emph{n.} bar, rod, pole.
\B{Me\-tew\-it fì\-swek\-ä ey\-ki\-vil.} Pull the two ends of this bar together.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}

% swey
\hi
\HT{swey}{\B{swey}} \I{[swEj]} \emph{adj.} best.
\B{Fì\-srey puk\-ä lu swey nì\-law.} This version of the book is clearly the best.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Tì\-'eyng\-it oel to\-lel a krr, tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} When I receive an answer, I will then decide the best way to bring you the words and rules of Na'vi.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ay\-lì\-'u\-van a\-swey lu 'i\-pu, lu sìl\-ron\-sem.} The best puns are both funny and clever.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Swey lu fwa nga fì\-kem ke si\-vi nì\-sngum.} It's best that you not do this fretfully.\slash It's best that you don't freak out about doing this.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{214}~\HL{'e'al}{\B{'e'al}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIswey}{\B{nì\-swey}}, optimally. 
\on{2}.~\HL{sweylu}{\B{swey\-lu}}, should.

% sweylu
\hi
\HT{sweylu}{\B{\cp{swey}lu}} \I{["{}swEj.lu]} \emph{vin.} should, ``it's best (that)''.
(\emph{a. for s.t. that hasn't yet happened with} \HL{txo}{\B{txo}} \emph{and} ‹\B{iv}›)
\B{Fì\-na\-er ftxì\-vä' lu nì\-hawng, ha swey\-lu txo ngal tsat kxi\-vukx nì\-win.} This drink tastes horrible, so you'd better swallow it quickly.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Swey\-lu txo nga ki\-vä.\slash Nga swey\-lu txo ki\-vä.} You should go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Swey\-lu txo nga ke ki\-vä.\slash Nga swey\-lu txo ke ki\-vä.} You shouldn't go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Swey\-lu txo ki\-vä.} I\slash you\slash she\slash one\slash etc. should go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
(\emph{b. for s.t. that's already happened with} \HL{fwa}{\B{fwa}} \emph{or} \HL{tsawa}{\B{tsawa}} \emph{and the past or perfective indicative})
\B{Swey\-lu fwa nga ko\-lä.} You should have gone.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
(\ding{43}~\HL{swey}{\B{swey}}, \HL{lu}{\B{lu}})

% sweyn
\hi 
\HT{sweyn}{\B{sweyn}} \I{[swEjn]} \emph{vtr.} (\emph{a.}) keep, preserve. 
\B{Ay\-ngal syu\-vet sweyn pe\-seng fte\-ke ay\-ioang tsi\-vun tsat ki\-va\-nom?} Where do you keep the food so that animals can't get it?~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}  
(\emph{b.}) leave alone, not disturb.
\B{Tsa\-ya\-yo\-tsrul\-it sweyn, ma 'i\-tan.} Don't disturb that bird's nest, son. ~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
\B{Oey fpom\-it sweyn!} Leave me alone!\slash Do not disturb my peace!~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 

% Sweriye
% \hi
% \B{Sweriye} \I{[]} \emph{n.} Sweden. 

% swirä
\hi
\HT{swirA}{\B{swi\cp{rä}}} \I{[swi."R\ae]} \emph{n.} creature. 
\B{Ye\-rik lu swi\-rä a\-nim nì\-txan.} The hexapede is a very timid creature.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Tsa\-swi\-rä\-ti lo\-nu! Ay\-nga ne\-to ri\-vikx! Fì\-ke\-tu\-wong\-ti oel stì\-ye\-ftxaw.} Release this creature! Step back! I will look at this alien.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{wiki\slash a}

% swizaw
\hi
\HT{swizaw}{\B{swi\cp{zaw}}} \I{[swi."zaw]} \emph{n.} arrow.
\B{Oe\-yä swi\-zaw nì\-ngay ti\-va\-kuk.} Let my arrow strike true.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Nì\-trr\-trr pxì\-mun\-'i sam\-si\-yul ay\-swi\-zaw\-it ku\-tu\-ä a\-law\-nä\-txayn sno\-kip nì\-'eng.} Warriors typically share the arrows of their defeated enemies among themselves.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Ta\-yel Tse\-nul pxe\-swi\-zaw\-ti yoa mun\-sna\-hawn\-ven.} Tsenu will receive three arrows in exchange for a pair of shoes.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
(\emph{a. with} \HL{tsweykayon}{\B{tswey\-ka\-yon}}) shoot an arrow.
\B{Swi\-zaw\-ti tswey\-kay\-on ne\-fä ne taw, tse\-nga zup ke lu law.} I shot an arrow into the air, | It fell to earth, I know not where.~\LINK{http://naviteri.org/2017/03/titusemteri-concerning-shooting/}{b} 
(i.) \B{tsko swizaw}, bow and arrow, archery.
\B{Na'\-vi\-ri txin\-a sä\-'o tì\-tu\-sa\-ron\-ä lu tsko swi\-zaw.} For the Na'vi the bow and arrow is the main hunting tool.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Tsko swi\-zaw\-fa a tì\-ta\-ron\-ìri ze\-ne fko ni\-vu\-me.} You must learn to hunt with a bow and arrow.~\LINK{http://forum.learnnavi.org/language-updates/ordinals-nume/}{ln\slash a\textsuperscript{c}}
\B{Oe tskxe\-keng si sä\-su\-lìn\-ur alu tsko swi\-zaw.} I practice my hobby, which is archery.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\emph{a. proverb}) \B{Fwa kan ke tam; ze\-ne swi\-zaw\-it li\-vo\-nu.} Intent is not enough; it's action that counts. [lit.: To aim is not enough; one must release the arrow.]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{swizawsena}{\B{swi\-zaw\-se\-na}}, quiver. 
\on{2}.~\HL{tsko}{\B{tsko swi\-zaw}}, bow and arrow, archery. 
\on{3}.~\HL{atanzaw}{\B{atan\-zaw}}, forked lightning.
\on{4}.~\HL{winzaw}{\B{win\-zaw}}, arrow deer.

% swizawsena
\hi
\HT{swizawsena}{\B{swi\cp{zaw}sena}} \I{[swi."zaw.sE.na]} \emph{n.} quiver (\emph{attached to the ikran's saddle}), \emph{coll. often shortened to} \HL{zawsena}{\B{zaw\-se\-na}}.
(\ding{43}~\HL{swizaw}{\B{swi\-zaw}}, \HL{sAhena}{\B{sä\-he\-na}})

% swoa
\hi
\HT{swoa}{\B{\cp{swo}a}} \I{["{}swo.a]} \emph{n.} intoxicating beverage.
\B{Oe\-ri lu swoa fne\-txum.} I'm allergic to intoxicating beverage.~\LINK{http://forum.learnnavi.org/language-updates/new-words-from-the-2012-avatarmeet-navi-class/msg551479/\#msg551479}{b}   
\B{Sra\-ne, fì\-txon tsun nga ni\-väk swo\-at nì\-'it, slä rä\-'ä rou!} Yes, you can have a little alcohol tonight, but don't get drunk!~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{swoasey}{\B{swoa\-sey}}, kava bowl.

% swoasey
\hi
\HT{swoasey}{\B{\cp{swo}asey}} \I{["{}swo.a.sEj]} \emph{n.} (hand\hyp{}sized) kava bowl (\emph{constructed from seed pods, used for drinking intoxicating beverages}). \\
(i.) \B{swoasey ayll} \emph{n.} large, social kava bowl. 
(\ding{43}~\HL{swoa}{\B{swoa}}, \HL{sey}{\B{sey}})

% swok
\hi
\HT{swok}{\B{swok}} \I{[swok\textcorner]} \emph{adj.} sacred. 
\B{Ma txan\-tslu\-sam\-a ngawng, a swok\-a ut\-ral\-ti yom.} O wise worm, who eats the Sacred Tree.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Becoming_one_of_the_Na.27vi}{a\textsuperscript{c}}

% swotu
\hi
\HT{swotu}{\B{\cp{swo}tu}} \I{["{}swo.tu]} \emph{n.} sacred place.
\B{He\-tu\-wong\-ìl aw\-nge\-yä swo\-tut ska\-'a, fte kll\-ki\-vu\-lat ke\-ru\-sey\-a tskxet.} The aliens destroy our sacred place to dig up dead rock.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln\slash a\textsuperscript{c}}

% Swotu Wayä
\hi 
\HT{swotu wayA}{\B{\cp{Swo}tu \cp{Way}ä}} \I{["{}swo.tu waj.\ae]} \emph{n.} Sacred Place of Song, \emph{situated in the Valley of Mo'ara}. 
(\ding{43}~\HL{swotu}{\B{swo\-tu}}, \HL{way}{\B{way}})

% swotulu
\hi
\HT{swotulu}{\B{\cp{swo}tulu}} \I{["{}swo.tu.lu]} \emph{n.} \emph{name of a sacred river}. 

% syaw
\hi
\HT{syaw}{\B{syaw}} \I{[sjaw]} \emph{vin.} call, name. 
\B{Fya\-pe fko syaw ngar? -- Oe\-ru syaw (fko) Ney\-ti\-ri. Nga\-ru tut? -- Oe\-ru syaw Tseyk.} What's your name? -- I'm Neytiri. How about you? -- I'm Jake.~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln}
\B{Fì\-por syaw fko Ìstaw.} This is Ìstaw.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Syaw oer ru\-txe trr\-ay fa spul\-mok\-ri.} Please phone me tomorrow.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}

% syawm
\hi 
\HT{syawm}{\B{syawm}} \I{[sjawm]} \emph{vtr.}, \textsc{rn} know. 
% \B{Ayoe ke syawm.} We don't know.~\LINK{}{a\on{2}} 
(\ding{43}~\HL{omum}{\B{o\-mum}}) 

% syawn
\hi 
\HT{syawn}{\B{syawn}} \I{[sjawn]} \textsc{i}.~\emph{n.} blessing. 
\B{Ne\-wey yaw\-ne lu oer ul\-te new oe mun\-txa si\-vi poe\-hu. Ru\-txe, ma sem\-pul, tìng mo\-er nge\-yä syawn\-it.} I love Ne\-wey and want to marry her. Please, father, give us your blessing.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
\B{Eywal tivìng ngar sneyä syawnit.} May Ey\-wa bless you. \\
\textsc{ii}.~\B{tìng syawn} \I{[t$\cdot$IN syawn]} \emph{vin.} bless.
\B{Eywa ti\-vìng syawn ngar, ma 'i\-te.} May Ey\-wa bless you, my daughter.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
(\ding{214}~\HL{rAzekx}{\B{rä\-zekx}}) 

% syay
\hi
\HT{syay}{\B{syay}} \I{[sjaj]} \emph{n.} fate, destiny, the arc of one's life. 
\B{Tsa\-krr syay ay\-nge\-yä, syay o\-lo'\-ä oe\-yä la\-yu teng.} Then you will suffer the same fate as my clan.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{syayvi}{\B{syay\-vi}}, luck, chance.

% syayvi
\hi
\HT{syayvi}{\B{\cp{syay}vi}} \I{["{}sjaj.vi]} \emph{n.} luck, chance.
\B{Et\-rìp\-a syay\-vi!} Good luck!~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} \\
\on{1.} \B{nìsyayvi}, by chance, by coincidence.
(\ding{43}~\HL{syay}{\B{syay}}, \HL{-vi}{\B{-vi}})

% syaksyuk
\hi
\HT{syaksyuk}{\B{syak\cp{syuk}}} \I{[sjak\textcorner."{}sjuk\textcorner]} \emph{or} \I{[."{}sjUk\textcorner]} (\textsc{rn}: \B{syak\cp{syùk}} \I{[Sak\textcorner."SUk\textcorner]}) \emph{n., fauna} prolemuris, \emph{prolemuris noctis}.

% syam
\hi
\HT{syam}{\B{syam}} \I{[sjam]} \emph{vtr.} smell (\texttt{+}\emph{control}). 
\B{Fì\-syu\-lang\-it syam. Fa\-hew lor lu nì\-txan, ke\-fyak?} Smell this flower. Its fragrance is beautiful, isn't it?~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\ding{43}~\HL{hefi}{\B{he\-fi}})

% syanan
\hi
\HT{syanan}{\B{\cp{sya}nan}} \I{["{}sja.nan]} \emph{countable n.} a type of waterfall, \emph{a single drop or series of smaller falls occurring sequentially along a stream or series of pools}.

% syar
\hi 
\HT{syar}{\B{syar}} \I{[sjar]} \emph{vin.} stick, stick to, adhere. 
\B{Rìk a\-'aw syar\-mar sìn kxem\-yo.} A leaf was sticking to the wall.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
\B{Pol kxum\-pay\-syar\-it so\-lar syey\-kar rìk\-it sìn kxem\-yo.} She stuck the leaf onto the wall with glue. [lit.: ``She used glue (and then) stuck the leaf onto the wall.'']~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{kxumpaysyar}{\B{kxumpaysyar}}, glue.

% syä'ä
\hi
\HT{syA'A}{\B{\cp{syä}'ä}} \I{["{}sj\ae.P\ae]} \emph{adj.}(of taste) bitter. 
\B{Ton\-ìri a\-la\-he, aw\-ngal yom fkxen\-ti le\-rìk nì\-wotx; fì\-txon yom sat a lu syä\-'ä nì\-'aw.} As for other nights, we eat all leafy vegetable; this night we eat only those that are bitter.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
\B{To\-lìng ay\-oer ma\-u\-tit a\-syä\-'ä.} (It) has given us bitter fruit.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}

% sye'otxang
\hi 
\HT{sye'otxang}{\B{\cp{sye}'otxang}} \I{["{}sjE.Po.t'aN]} \emph{n.} wind instrument. 
(\ding{43}~\HL{syeha}{\B{sye\-ha}}, \HL{'otxang}{\B{'o\-txang}})

% syeha
\hi
\HT{syeha}{\B{\cp{sye}ha}} \I{["{}sjE.ha]} \textsc{i}. \emph{n.} breath.  \\
\textsc{ii}. \B{\cp{sye}ha si} \I{["{}sjE.ha si]} \emph{vin.} breathe.   
\B{Ma Ra\-lu, srung si por! Nì\-win! Sye\-ha ke si!} Ralu, help her! Quick! She's not breathing!~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{snewsye}{\B{snew\-sye}}, weird, spooky. 
\on{2}.~\HL{sye'otxang}{\B{sye\-'o\-txang}}, wind instrument.
\on{3}.~\HL{mungsye}{\B{mung\-sye}}, inhale.
\on{4}.~\HL{lonusye}{\B{lo\-nu\-sye}}, exhale, blow.

% syen
\hi
\HT{syen}{\B{syen}} \I{[sjEn]} \emph{adj.} last, final (\emph{in a series}). 
\B{tì\-pawm a\-syen}, last question.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
(\ding{43}~\HL{tor}{\B{tor}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{nIsyen}{\B{nì\-syen}}, lastly, finally.

% syep
\hi
\HT{syep}{\B{syep}} \I{[sjEp\textcorner]} \emph{vtr.} trap (s.o.\slash s.t.).
\B{Fì\-hì\-'ang\-ìl txu\-la hän\-ti fte smar\-it syi\-vep.} This insect constructs a web to trap its prey.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} \\
\textsc{Derivation}: \ding{43} \on{1}.~\HL{sAsyep}{\B{sä\-syep}}, a trap. 
\on{2}.~\HL{nIsyep}{\B{nì\-syep}}, tightly, in an iron grip. 
\on{3}.~\HL{syeptute}{\B{syep\-tu\-te}}, hyneman. 
\on{4}.~\HL{syeprel}{\B{syep\-rel}}, camera.
\on{5}.~\HL{syep'an}{\B{syep\-'an}}, lift vine.

% syep'an
\hi 
\HT{syep'an}{\B{\cp{syep}'an}} \I{["{}sjEp\textcorner.Pan]} \emph{n.} \ding{96} lift vine [lit.: `trap vine']. 
(\ding{43}~\HL{syep}{\B{syep}}, \HL{'ana}{\B{'a\-na}})

% syeprel
\hi
\HT{syeprel}{\B{syep\cp{rel}}} \I{[sjEp\textcorner."REl]} \emph{n.} camera.
\B{Lu oe\-ru syep\-rel.} I have a camera.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{43}~\HL{syep}{\B{syep}}, \HL{rel}{\B{rel}})

% syeptute
\hi
\HT{syeptute}{\B{\cp{syep}tute}} \I{["{}sjEp\textcorner.tu.tE]} \emph{n.}, \ding{96} hyneman, ``person trapper'' (\emph{type of carnivorous plant}), \emph{pandoratonia myopora}. 
\B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok, tso\-le\-'a syep\-tu\-tet a\-tsawl fra\-to mì sì\-rey.} Yesterday I was with Txewì in the forest and we saw the biggest Trapper I've ever seen.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
(\ding{43}~\HL{syep}{\B{syep}}, \HL{tute}{\B{tu\-te}})

% syewe
\hi 
\HT{syewe}{\B{\cp{sye}we}} \I{["{}sjE.wE]} \emph{n.} fat (substance in meat). \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{syewenga'}{\B{sye\-we\-nga'}}, fatty.

% syewenga'
\hi 
\HT{syewenga'}{\B{\cp{sye}wenga'}} \I{["{}sjE.wE.NaP]} \emph{adj.} fatty. 
\B{Po\-an\-ur su\-nu tsngan a\-sye\-we\-nga'; poe\-ru ke su\-nu kaw\-'it.} He likes fatty meat; she doesn't like it at all.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{syewe}{\B{sye\-we}})

% syìmawnun'i
\hi 
\HT{syImawnun'i}{\B{syìmawnun\cp{'i}}} \I{[sjI.maw.nun."Pi]} \emph{or} \I{[.nUn.]} \emph{n.} cut tie, dissolved connection. 
\B{Oe\-nga\-ri sä\-ta\-re syì\-maw\-nun\-'i slo\-lä\-ngu.} As for the two of us, I'm sorry to say our relationship is dissolved.~\LINK{http://naviteri.org/2023/07/trr-tsyimawnuniya-lefpom-happy-independence-day/}{b} 
(\ding{43}~\HL{sAyIm}{\B{sä\-yìm}}, \HL{mun'i}{\B{mun\-'i}}) \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{tIsyImawnun'i}{\B{tì\-syì\-maw\-nun\-'i}}, independence. 

% syo
\hi
\HT{syo}{\B{syo}}\textsuperscript{\on{1}} \I{[sjo]} \emph{adj.} light, lightweight.
(\ding{214}~\HL{ku'up}{\B{ku'\-up}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{syokup}{\B{syo\-kup}}, weight.

\hi
\B{syo}\textsuperscript{\on{2}} : \HL{tsyo}{\B{tsyo}}

% syokup
\hi 
\HT{syokup}{\B{syo\cp{kup}}} \I{[sjo."kup\textcorner]} \emph{or} \I{[."kUp\textcorner]} (\textsc{rn}: \B{syo\cp{kùp}} \I{[So."kUp\textcorner]}) \emph{n.} weight (physical). 
(\ding{43}~\HL{syo}{\B{syo}}, \HL{ku'up}{\B{ku'\-up}})

% syon
\hi
\HT{syon}{\B{syon}} \I{[sjon]} \emph{n.} feature, trait, attribute, characteristic, point, aspect, facet, property. 
\B{Tsran\-ten fra\-to a syon tsam\-si\-yu\-ä lu tì\-tstew.} The most important characteristic of a warrior is bravery.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}  
\B{Pa\-lu\-lu\-kan\-ìri lu pxe\-syon a ze\-ne fko zi\-ve\-rok nì\-tut.} Three things about the thanator must always be kept in mind.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Fra\-'u\-ä ran ngä\-pop fa fra\-syon tse\-yä.} The \emph{ran} of each thing arises from the totality of its attributes.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{syonlI'u}{\B{syon\-lì\-'u}}, adjective.
\on{2}.~\HL{ketengsyon}{\B{ke\-teng\-syon}}, variety, diversity.

% syonlì'u
\hi 
\HT{syonlI'u}{\B{\cp{syon}lì'u}} \I{["{}sjon.lI.Pu]} \emph{n.} adjective. 
\B{Kehe, ke tsun\-slu fwa yem me\-syon\-lì\-'ut uo tstxo\-lì\-'u.} No, it's not possible to put two adjectives after a noun.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/\#comment-27964}{b} 
(\ding{43}~\HL{syon}{\B{syon}}, \HL{lI'u}{\B{lì\-'u}})

% syor
\hi
\HT{syor}{\B{syor}} \I{[sjoR]} \emph{vin.} relax, chill out.
\B{Syor, ma 'ey\-lan, syor. Ke lu kea sngum\-tsim.} Relax, friend, relax. There's nothing to worry about.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{New oe ri\-vun a\-sim tì\-fnu\-nga'\-a tseng\-it a tsa\-ro tsun syi\-vor tsi\-vu\-rokx fte spä\-pi\-veng.} I want to find a quiet place nearby where I can chill out and rest to get my head back on straight.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Fì\-u\-van\-ìl oe\-ti syey\-kor nì\-txan.} I find this game very relaxing.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{tIsyor}{\B{tì\-syor}}, relaxation.

% syulang
\hi
\HT{syulang}{\B{\cp{syu}lang}} \I{["{}sju.lang]} \emph{n.}, \ding{96} flower. 
\B{Fì\-syu\-lang a\-rim lu hì\-'i fra\-to.} This yellow flower is the smallest of all.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Oe\-yä ay\-syu\-lang set mì fay kll\-kxe\-rem.} My flowers are standing in water now.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-296}{b}
\B{Ra\-lul ngo\-lop 'um\-tsat fa ay\-syu\-lang fwä\-kì\-wll\-ä.} Ralu made medicine from (\emph{or}: using the) flowers of the Mantis orchid.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
\B{Fì\-syu\-lang\-it syam. Fa\-hew lor lu nì\-txan, ke\-fyak?} Smell this flower. Its fragrance is beautiful, isn't it?~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(i.) \B{sä\-star\-sìm syu\-lang\-ä}, bouquet (\emph{collection of flowers selected and put together intentionally by a person}).~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{prrnesyul}{\B{prr\-ne\-syul}}, bud (of a flower).
\on{2}.~\HL{paysyul}{\B{pay\-syul}}, water lily.
\on{3}.~\HL{tsawksyul}{\B{tsawk\-syul}}, sun lily.
\on{4}.~\HL{tarsyu}{\B{tar\-syu}}, tarsyu.

% syura
\hi
\HT{syura}{\B{syu\cp{ra}}} \I{[sju."Ra]} \emph{n.} energy (\emph{physical and spiritual}).
\B{Fra\-syu\-ra\-ti fkol za\-sro\-lìn nì\-'aw ul\-te trr\-o ze\-ne tey\-ki\-vä\-txaw.} All energy is only borrowed, and one day it will have to be given back.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{syuratan}{\B{syuratan}}, bioluminescence.

% syuratan
\hi
\HT{syuratan}{\B{syu\cp{ra}tan}} \I{[sju."Ra.tan]} \emph{n.} bioluminescence.   
\B{Txon\-krr lu syu\-ra\-tan na'\-rìng\-ä Ey\-we\-veng\-ä lor nì\-txan.} At night, the bioluminescence of the Pandoran forest is very beautiful.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Na'\-rìng sngä\-'i ni\-vrr txon\-krr syu\-ra\-tan\-fa.} The forest begins to glow at night with bioluminescence.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{syura}{\B{syu\-ra}}, \HL{atan}{\B{a\-tan}})

% syusmung
\hi
\HT{syusmung}{\B{\cp{syu}smung}} \I{["{}sju.smuN]} \emph{or} \I{[.smUN]} (\textsc{rn}: \B{\cp{syu}smùng} \I{["Su.smUN]}) \emph{n.} tray. 
(\ding{43}~\HL{syuve}{\B{syu\-ve}}, \HL{sAmunge}{\B{sä\-mu\-nge}})

% syuve
\hi
\HT{syuve}{\B{\cp{syu}ve}} \I{["{}sju.vE]} \emph{uncountable n.} food. 
\B{Lu oe\-ru syu\-ve.} I have food.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Rä\-'ä 'i\-val syu\-vet!} Don't waste food.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Zun ngal tsa\-fne\-syu\-vet tim\-vìng oer, zel li\-vu oe txur fì\-trr.} If you had given me that kind of food, I would be strong today.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
\B{Syu\-ve\-ri ay\-ftxo\-zä\-yä ay\-nga\-ru fki\-van on\-lor ftxì\-lor\-sì nìwotx!} May the food of the celebrations smell and taste good to you.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{syuvekel}{\B{syu\-ve\-kel}}, famine. 
\on{2}.~\HL{syusmung}{\B{syu\-smung}}, tray. 
\on{3}.~\HL{tsyosyu}{\B{tsyo\-syu}}, food made from flour.
\on{4}.~\HL{okupsyu}{\B{o\-kup\-syu}}, dairy product.

% syuvekel
\hi
\HT{syuvekel}{\B{\cp{syu}vekel}} \I{["{}sju.vE.kEl]} \emph{n.} famine.
(\ding{43}~\HL{syuve}{\B{syu\-ve}}, \HL{tIkelu}{\B{tì\-ke\-lu}}) \\

\end{multicols} 

