\pdfbookmark[0]{T}{T}
\begin{center}
\section*{T \I{[t]} (\emph{TeT})}
\end{center}

\begin{multicols}{2}

% -t
\hi
\HT{-t}{\B{-t}} \I{[t\textcorner]} \emph{suf. to mark the patientive case of a n. ending in a vowel or diphthong}. 
(\ding{43}~\HL{-it}{\B{-it}}, \HL{-ti}{\B{-ti}})

% ta
\hi
\HT{ta}{\B{ta}} \I{[ta]} \emph{adp.} from.   
(\emph{a. origin}) from, of. 
\B{Oe\-ri ta pe\-yä fa\-hew a\-ke\-wong on\-tu te\-ya lä\-ngu.} My nose is full of his alien smell.~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll\slash a}  
\B{Ta 'ey\-lan kar\-yu\-sì ay\-nge\-yä}. [This letter comes] From your friend and teacher.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Fu\-ri\-a fì\-lì'\-fya\-vit fkol ta 'Ìng\-lì\-sì mo\-lu\-nge, sìl\-pey oe tsnì ay\-ha\-pxì\-tu lì'\-fya\-o\-lo'\-ä aw\-nge\-yä ke stì\-ye\-vi.} That this expression is brought from English, will hopefully not anger the members of the language group.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Kal\-txì ta Ko\-pen\-han.} Hi from Copenhagen.~\LINK{http://naviteri.org/2010/09/kaltxi-ta-kopenhan-hi-from-copenhagen/}{b}
\B{Pot 'a\-ku fì\-tseng\-ta!} Get him out of here!~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Säm\-yam\-ìl po\-ru wa\-yìn\-txu fu\-ta nga\-ta lo\-lu li txo\-a.} A hug will show him that you've already forgiven him.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Fkxi\-le pewn\-ta tu\-té\-yä kur.} The bib necklace hangs from the woman's neck.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Za\-'u tsa\-tson ta Ey\-wa.} That obligation comes from Eywa.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Lu Ney\-ti\-rir ta Mo\-'at a tson a kar Tsyeyk\-ur ay\-fya\-'ot Na'\-vi\-yä.} Neytiri is under obligation by\slash from Mo'at to teach Jake the ways of the Na'vi.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{tuk\-ru a txo\-lu\-la fkol ta rìn\slash tuk\-ru a ta rìn}, a spear made of wood.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}
(\emph{b. in counting}) from. 
\B{Ti\-am ta pxey vay vol.} Count from three to eight.~\LINK{https://forum.learnnavi.org/language-updates/'count-to'-wa-and-sequential-genitive-(discussion)/}{ln}
(\emph{c. temporal}) from, since. 
\B{Trr\-'ong\-ta Txon\-'ong\-vay po to\-lì\-ran.} He walked from dawn until dusk.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\emph{d. with} \HL{pawm}{\B{pawm}} \emph{to form the indirect object})
\B{Pol po\-lawm tì\-pawm\-it oe\-ta.} He asked me a question.~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b}
(\emph{e. conv. phrase for responding to thanks}) \B{ha\-ya\-lo ta oe} \emph{or} \B{hay\-a\-lo oe\-ta}~\I{[ha."ja.lo "wE.ta]}. You're welcome [lit.: next time from me].~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\ding{43}~\HL{ftu}{\B{ftu}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{ta'em}{\B{ta\-'em}}, from above. 
\on{2}.~\HL{tafkip}{\B{ta\-fkip}}, from up among. 
\on{3}.~\HL{tafral}{\B{ta\-fral}}, therefore, because of that. 
\on{4}.~\HL{takip}{\B{ta\-kip}},  from among. 
\on{5}.~\HL{takrra}{\B{ta\-krra\slash akrr\-ta}}, since. 
\on{6}.~\HL{talun1}{\B{ta\-lun(a)\slash alun\-ta}}, because, because of. 
\on{7}.~\HL{tatlam}{\B{tat\-lam}}, apparently. 
\on{8}.~\HL{taweyk}{\B{ta\-weyk(a)}}, because.
\on{9}.~\HL{tamIfa}{\B{ta\-mì\-fa}}, internal. 
\on{10}.~\HL{tawrrpa}{\B{ta\-wrr\-pa}}, external.

% ta'em
\hi
\HT{ta'em}{\B{ta\cp{'em}}} \I{[ta."PEm]} \emph{adv.} from above. 
\B{Ik\-ran\-ti mak\-to, 'eko ta\-'em.} Take the Ikran, attack from above.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Attack_on_hometree}{wiki\slash a}

% ta'leng
\hi
\HT{ta'leng}{\B{\cp{ta'}leng}} \I{["taP.lEN]} \emph{n.} skin. 
\B{Ta'\-leng prr\-nen\-ä lu faoi, pum ko\-ak\-tu\-ä ekx\-txu.} A baby's skin is smooth, an old person's is rough.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Fì\-syu\-lang lu ean na ta'\-leng.} This flower is skin\hyp{}blue.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Teng\-krr hu pa\-lu\-kan\-tsyìp uvan se\-ri zo\-lìm oel mì sa'\-leng a 'u\-ot a lu txa' sì ekx\-txu.} While playing with my cat I felt something hard and rough on his skin.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{ta'lengean}{\B{ta'\-le\-nge\-an}}, skin\hyp{}blue.

% ta'lengean
\hi
\HT{ta'lengean}{\B{\cp{ta'}lengean}} \I{["taP.lEN.E.an]} \emph{adj.} skin\hyp{}blue.
(\ding{43}~\HL{ta'leng}{\B{ta'\-leng}}, \HL{ean}{\B{ean}})

% ta'unui
\hi
\B{Ta'unui} \I{[ta.Pu.nu.i]} \emph{n. a clan name}, one clan of the reef people.

% tafkip
\hi
\HT{tafkip}{\B{ta\cp{fkip}}} \I{[ta."fkip\textcorner]} \emph{adp.} from up among. 
(\ding{43}~\HL{ta}{\B{ta}}, \HL{fkip}{\B{fkip}})

% tafral
\hi
\HT{tafral}{\B{ta\cp{fral}}} \I{[ta."fRal]} \emph{adv.} therefore, because of that.
\B{Ta\-fral ke lì\-syek oel nge\-yä ke\-ye\-'ung\-it.} Therefore I will not heed your insanity.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g}
\B{Oe pll\-ngay san mo\-lak\-to oe nì\-fe', ta\-fral sno\-laytx; wä\-tu lu oe\-to txur.} I acknowledge that I rode badly, so I lost; my opponent was stronger than I was.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
(\ding{43}~\HL{ta}{\B{ta}}, \HL{ral}{\B{ral}})

% takip
\hi
\HT{takip}{\B{ta\cp{kip}}} \I{[ta."kip\textcorner]} \emph{adp.} from among. 
(\ding{43}~\HL{ta}{\B{ta}}, \HL{kip}{\B{kip}})

% takrra
\hi
\HT{takrra}{\B{ta\cp{krr}a}} \emph{or} \B{a\cp{krr}ta} \I{[ta."k\s{r}.a]} \emph{conj.} since.   
\B{Ay\-lì'\-fya yaw\-ne le\-ru oer ta\-krr\-a 'e\-veng la\-mu.} I've loved languages since I was a child.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
\B{Ay\-ha\-pxì\-tu po\-ngu\-ä txo\-pu si nì\-nän ta\-krr\-a Va'\-rul pxe\-ku\-tut lä\-txayn.} The members of the group are less afraid since Va'ru defeated three of the enemy.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{ta}{\B{ta}}, \HL{krr}{\B{krr}}, \HL{a}{\B{a}})

% taksyokx
\hi
\HT{taksyokx}{\B{tak\cp{syokx}}} \I{[ta$\cdot$k\textcorner."{}sjok']} \emph{vin., inf.}\on{11} clap hands.  
\B{Tak\-syokx, ma fra\-po!} Let's give (him\slash her\slash them) a hand, everybody!~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b}
(\ding{43}~\HL{takuk}{\B{ta\-kuk}}, \HL{tsyokx}{\B{tsyokx}})

% takuk 1
\hi
\HT{takuk}{\B{\cp{ta}kuk}}\textsuperscript{\on{1}} \I{["ta.kuk\textcorner]} \emph{or} \I{[.kUk\textcorner]} (\textsc{rn}: \B{\cp{ta}kùk} \I{["ta.kUk\textcorner]}) \emph{vtr.} (\emph{a.}) strike, hit one's target, slap, beat.   
\B{Oe\-yä swi\-zaw nì\-ngay ti\-va\-kuk.} Let my arrow strike true.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Ta\-ron\-yul le\-hì\-pey kan smar\-it nìl\-ke\-ftang slä ke ta\-kuk kaw\-krr.} A hesitant hunter will aim at a prey forever but never hit it.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
\B{Oe\-ti ta\-kuk rä\-'ä!} Don't hit me!~\LINK{https://forum.learnnavi.org/language-updates/little-confirmation-on-takuk/}{ln}
(\emph{b. with} \HL{fa}{\B{fa}}) \HL{rumtsyokx}{\B{ta\-kuk fa rum\-tsyokx}}, punch s.o. \B{A\-te\-yol En\-tut to\-la\-kuk fa rum\-tsyokx.} Ateyo punched Entu.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}; \HL{rIktsyokx}{\B{ta\-kuk fa rìk\-tsyokx}}, slap s.o. \B{A\-te\-yol En\-tut to\-la\-kuk fa rìk\-tsyokx.} Ateyo slapped Entu.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\ding{214}~\HL{hupx}{\B{hupx}})
(\emph{c. of weather}) \B{Ta\-kuk set tsaw\-ke. Sley\-ku fra\-'ut u\-kxo.} Now the sun bears down. (It) makes everything dry.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg475008/\#msg475008}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsamkuk}{\B{tsam\-kuk}}, war drum.

% Takuk 2
\hi
\B{\cp{Ta}kuk}\textsuperscript{\on{2}} \I{["ta.kuk\textcorner]} \emph{or} \I{[.kUk\textcorner]} \emph{n.} \emph{male name}.~\LINK{}{dh}

% talioang
\hi
\HT{talioang}{\B{\cp{ta}lioang}} \I{["tal.i.""{}o.aN]} \emph{n., fauna} sturmbeest, \emph{bovindicum monocerii}. 
\B{Fo ko\-lä tsa\-tseng fte ti\-va\-ron ye\-ri … ìì … ko\-lan ta\-li\-o\-ang\-it.} They went there to hunt hexa … um … I mean sturmbeast.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} 

% talun / taluna
\hi
\HT{talun1}{\B{ta\cp{lun}}}\textsuperscript{\on{1}} \emph{or} \B{ta\cp{lu}na} \I{[ta."lun]} \emph{or} \I{[ta."lu.na]} \emph{conj.} because, since.
\B{Ay\-oe sno\-laytx fì\-trr ta\-lun\-a oel rum\-it to\-lung\-zup.} We lost today because I dropped the ball.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Tsa\-ik\-ran\-ìri ta\-lun\-a new nga\-ti tspi\-vang, law lu fwa mi ke lu zä\-fi.} Since that ikran wants to kill you, it's clear it's still not tame.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{ta}{\B{ta}}, \HL{lun}{\B{lun}}, \HL{a}{\B{a}})

\hi
\HT{talun2}{\B{ta\cp{lun}}}\textsuperscript{\on{2}} \I{[ta."lun]} \emph{or} \I{[."lUn]} \emph{adp.} because of, due to.
\B{Tì\-'i'\-a\-ri tsam\-ä ze\-ne Saw\-tu\-te vi\-ve\-lek ta\-lun tì\-txur Ey\-wa\-yä.} At the end of the war, the Sky People had to give up due to the power of Eywa.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Tsaw\-ke slo\-lu vawm ta\-lun fwopx.} The sun became dark due to dust in the air.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\emph{a. to form a causative meaning with vin. modals}) \B{Tom\-pa\-kel\-ta\-lun ze\-ne tu\-te Kä\-lì\-for\-ni\-a\-yä pay\-it si\-var nì\-nän.} The drought causes Californians to use less water.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{ta}{\B{ta}}, \HL{lun}{\B{lun}})

% tam
\hi
\HT{tam}{\B{tam}} \I{[tam]} \emph{vin.} suffice, be enough, ``do''. 
\B{Fì\-pam\-tseo\-tu\-ri ke la\-yu ftue fwa run fkol sä\-rawn\-it a tam.} It won't be easy to find a satisfactory replacement for this musician.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Nì\-ke\-ftxo oe\-ri ke tam.} Unfortunately, it doesn't work for me.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1116}{b}
\B{Pe\-yä 'itan\-ìri lu ho\-na nì\-txan a fì\-'u law lu fra\-por. Slä tì\-ho\-na nì\-'aw ke tam.} It's clear to everyone that his son is very cute. But cuteness alone isn't enough.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\emph{a. idiom, conv.}) \B{Tam.} \textsc{ok}! 
\B{Tsun ti\-vam.} It's ok.~\LINK{http://forum.learnnavi.org/language-updates/ok!-we-got-ok/}{ln} 
\B{'U a\-ke\-tsran tsun ti\-vam.} Anything at all will be fine.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
(\emph{b. idiom, conv.}) \B{Tam ke tam}. `Meh\slash so\hyp{}so.'~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/comment-page-1/\#comment-1621}{b}
\B{Nga\-ru lu fpom srak?--Tam ke tam.} How are you?--Meh, so\hyp{}so.
(\emph{c. proverb}) \B{Fwa kan ke tam; ze\-ne swi\-zaw\-it li\-vo\-nu.} Intent is not enough; it's action that counts. [lit.: To aim is not enough; one must release the arrow.]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}\\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letam}{\B{le\-tam}}, sufficient, enough. 
\on{2}.~\HL{nItam}{\B{nì\-tam}}, enough.

% tamìfa
\hi 
\HT{tamIfa}{\B{ta\cp{mì}fa}} \I{[ta."mI.fa]} \emph{adj.} internal. 
(\ding{214}~\HL{tawrrpa}{\B{ta\-wrr\-pa}}, \ding{43}~\HL{ta}{\B{ta}}, \HL{mIfa}{\B{mì\-fa}}) 

% tanhì
\hi
\HT{tanhI}{\B{tan\cp{hì}}} \I{[tan."hI]} \emph{n.} (\emph{a. astr.}) star.  
\B{Tsay\-san\-hì ha\-yum ye'\-rìn (taw\-ftu).} Those stars will soon set.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Kxam\-trr lam fwa san\-hì a mì saw 'o\-lìp nì\-wotx slä tsa\-krr ke tsun fko sat tsi\-ve\-'a nì\-'aw.} At mid-day it seems that the stars in the sky have all vanished but they just can't be seen then.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Hol\-pxay san\-hì\-yä a mì saw lu ke\-tsuk\-ti\-am; keng ke tsun fko tsi\-ve\-'a sat nì\-wotx.} The number of stars in the sky is infinite; it's not even possible to see them all.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Pll\-txe Saw\-tu\-te san ki\-fkey\-ìl ay\-oe\-yä tok txan\-a ngip\-it a san\-hì\-kip.} The Sky People say that their world is in the great space among the stars.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\emph{b. of the body}) bioluminescent freckle. 
(\ding{43}~\HL{atan}{\B{a\-tan}}, \HL{hI'i}{\B{hì\-'i}})

% tanleng
\hi 
\HT{tanleng}{\B{\cp{tan}leng}} \I{["tan.lEN]} \emph{n.}, \ding{96} bark (of a tree). 
(\ding{43}~\HL{tangek}{\B{ta\-ngek}}, \HL{ta'leng}{\B{ta'\-leng}})

% tangek
\hi
\HT{tangek}{\B{\cp{ta}ngek}} \I{["taN.Ek\textcorner]} \emph{n.} trunk (\emph{of a tree}). 
\B{Tsa\-ut\-ral\-ìri ta\-ngek lu sloa nì\-txan; 'evi ke tsun tsyi\-vìl.} The trunk of that tree is very wide; the kid cannot climb it.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Fì\-ut\-ral\-ìri ta\-ngek\-ä zir fkan vawt, hu\-fwa ke rey.} The trunk of the tree feels solid, although it's dead.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Tsa\-ta\-ngek\-ìri pam fkan mo\-mek.} That (tree)trunk sounds hollow.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Na ay\-sa\-ngek a\-fkew.} Like mighty trunks (of trees).~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tanleng}{\B{tan\-leng}}, bark.

% tare
\hi
\HT{tare}{\B{\cp{ta}re}} \I{["ta.RE]} \emph{vtr.} connect, relate to, have a relationship with.   
\B{Sä\-pll\-txel kar\-yu\-ä ke to\-la\-rä\-nge tì\-pawm\-it kaw\-'it.} The teacher's statement in no way pertained to the question. Drat!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Txil\-te Ri\-ni\-sì tä\-pa\-re fì\-tsap nì\-soaia, slä tsal\-su\-ngay ke nì\-olo' ta\-krr\-a Ri\-ni mun\-txa slo\-lu.} Txilte and Rini are related by blood, but nevertheless not by clan since the time Rini got married.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{a. conv.}) \B{Ke ta\-re.} It's irrelevant.\slash It doesn't matter.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sAtare}{\B{sä\-ta\-re}}, connection, relationship. 
\on{2}.~\HL{ketartu}{\B{ke\-tar\-tu}}, outcast.
\on{3}.~\HL{tarsyu}{\B{tar\-syu}}, tarsyu.

% tarep
\hi 
\HT{tarep}{\B{\cp{ta}rep}} \I{["ta.REp\textcorner]} \emph{vtr.} save, rescue (\emph{from a dangerous or distressing situation}). 
\B{Maw\-krr\-a pa\-lu\-lu\-kan po\-sìn spo\-lä, ke tsä\-ngun fko pot ti\-va\-rep.} Sadly, once the thanator had jumped on her, she could not be rescued.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\emph{a. idiom}) \B{Ta\-rep!} ``Help!''~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
(\ding{43}~\HL{'avun}{\B{'a\-vun}}; \HL{zong}{\B{zong}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tareptu}{\B{ta\-rep\-tu}}, rescuee, rescued person. 
\on{2}.~\HL{sAtarep}{\B{sä\-ta\-rep}}, a rescue. 
\on{3}.~\HL{tarepyu}{\B{ta\-rep\-yu}}, rescuer.

% tareptu
\hi 
\HT{tareptu}{\B{\cp{ta}reptu}} \I{["ta.REp\textcorner.tu]} \emph{n.} rescuee, s.o. who has been rescued or saved, saved person, rescued person. 
\B{Maw tsa\-fe\-kem, new ta\-rep\-tu sne\-yä ta\-rep\-yur ira\-yo si\-vi.} After the accident, the rescuee wanted to thank his rescuer.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
(\ding{43}~\HL{tarep}{\B{ta\-rep}})

% tarepyu
\hi 
\HT{tarepyu}{\B{\cp{ta}repyu}} \I{["ta.REp\textcorner.ju]} \emph{n.} rescuer. 
\B{Maw tsa\-fe\-kem, new ta\-rep\-tu sne\-yä ta\-rep\-yur ira\-yo si\-vi.} After the accident, the rescuee wanted to thank his rescuer.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
(\ding{43}~\HL{tarep}{\B{ta\-rep}})

% tarnioang
\hi
\HT{tarnioang}{\B{\cp{tar}nioang}} \I{["taR.ni.o.aN]} \emph{n.} predator animal.
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang. Fì\-tì\-kan\-ìri ropx ke ha'.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators. For this purpose, a hole going right through the tree trunk isn't suitable.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{taron}{\B{ta\-ron}}, \HL{ioang}{\B{ioang}})

% taron
\hi
\HT{taron}{\B{\cp{ta}ron}} \I{["ta.Ron]} \emph{vtr.} hunt. 
\B{Ney\-ti\-ril ye\-rik\-it to\-la\-ron.} Neytiri hunted a hexapede.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#MSNBC}{wiki}
\B{Pol ye\-rik\-it ta\-ron nìl\-tsan fra\-to.} He hunts yerik the best of all.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln} 
\B{Ftu\-e lu fwa ta\-ron ngong\-a io\-ang\-it to fwa ta\-ron pum\-it a lu wa\-lak sì win.} It's easier to hunt lethargic animals than to hunt perky, speedy ones.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Pe\-lun fì\-tseng\-ne nga zo\-la\-'u fte ti\-va\-ron? -- Lo\-lu 'en.} Why did you come here to hunt? -- It was a guess (hunch).~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Pa\-lu\-lu\-kan a te\-ra\-ron lu le\-hrr\-ap.} A thanator that's hunting is dangerous.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{Pa\-lu\-lu\-kan a\-tu\-sa\-ron lu le\-hrr\-ap.} A hunting thanator is dangerous.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{Pol trr\-am ye\-rik\-it tar\-ma\-ron. -- Fì\-trr mi te\-ra\-ron.} He was hunting hexapede yesterday. -- He's still hunting today.~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{taronyu}{\B{ta\-ron\-yu}}, hunter. 
\on{2}.~\HL{tItaron}{\B{tì\-ta\-ron}}, hunt. 
\on{3}.~\HL{sAtaron}{\B{sä\-ta\-ron}}, hunt. 
\on{4}.~\HL{tarnioang}{\B{tar\-ni\-oang}}, predator animal. 
\on{5}.~\HL{uniltaron}{\B{unil\-ta\-ron}}, dream hunt. 
\on{6}.~\HL{taronway}{\B{ta\-ron\-way}}, the Hunt Song.
\on{7}.~\HL{tarpongu}{\B{tar\-po\-ngu}}, hunting party.

% taronway
\hi
\HT{taronway}{\B{\cp{ta}ronway}} \I{["ta.Ron.waj]} \emph{n.} the Hunt Song.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
(\ding{43}~\HL{taron}{\B{ta\-ron}}, \HL{way}{\B{way}})

% taronyu
\hi
\HT{taronyu}{\B{\cp{ta}ronyu}} \I{["ta.Ron.ju]} \emph{n.} hunter. 
\B{Oe lu ta\-ron\-yu.} I am a hunter.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tsa\-ri\-a fkol po\-le\-'un fu\-ta Lo\-ak slu ta\-ron\-yu, sem\-pul 'efu ye.} Father is content that it's been decided Loak will be a hunter.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Fì\-ta\-ron\-yu\-tsyìp ke tsun ke\-'ut sti\-vä'\-nì.} This little hunter can't catch anything.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Ta\-ron\-yul le\-hawm\-pam ska\-'a sä\-ta\-ron\-it.} A noisy hunter destroys the hunt.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Lu oe\-ru sngum a sa\-ron\-yu ke tì\-ye\-vä\-txaw.} I'm worried that the hunters will not return (soon, as expected).~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\emph{a. proverb}) \B{Ta\-ron\-yut yom smar\-ìl.} \emph{A totally botched situation, chaos} [lit.: the prey eats the hunter].
\B{Eyk Ka\-mun a fra\-lo lä\-ngu tsa\-sä\-ta\-ron vel\-ke nì\-wotx. Ta\-ron\-yut yom smar\-ìl!} Every time Kamun is in charge, the hunt is a mess. Everything goes wrong that can!~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Hì\-pey ta\-ron\-yu, hi\-fwo ye\-rik.} ``He who hesitates is lost.'' [lit.: The hunter hesitates and the hexapede escapes]~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{taron}{\B{ta\-ron}}, \HL{-yu}{\B{-yu}})

% tarpongu
\hi
\HT{tarpongu}{\B{\cp{tar}pongu}} \I{["taR.po.Nu]} \emph{n.} hunting party. 
\B{Ni\-nat tì\-mung\-wrr sìl\-mi fte Ra\-lu tsi\-vun ki\-vä hu tar\-po\-ngu.} Ninat just made an exception so that Ralu can go with the hunting party.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\ding{43}~\HL{taron}{\B{ta\-ron}}, \HL{pongu}{\B{po\-ngu}})

% tarsyu
\hi 
\HT{tarsyu}{\B{\cp{tar}syu}} \I{["taR.sju]} \emph{n.} \ding{96} tarsyu. 
(\ding{43}~\HL{tare}{\B{ta\-re}}, \HL{syulang}{\B{syu\-lang}})

% tatep
\hi
\HT{tatep}{\B{\cp{ta}tep}} \I{["ta.tEp\textcorner]} \emph{vtr.} lose, lose track\slash  awareness of s.t. 
\B{Tì\-kan taw\-na\-tep!} (\emph{mil.}) Target lost!~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b\slash g}

% tatlam
\hi
\HT{tatlam}{\B{tat\cp{lam}}} \I{[tat\textcorner."lam]} \emph{adv.} apparently. 
(\ding{43}~\HL{ta}{\B{ta}}, \HL{tIlam}{\B{tì\-lam}})

% tautral
\hi
\HT{tautral}{\B{\cp{ta}utral}} \I{["ta.ut\textcorner.Ral]} \emph{n.}, \ding{96} sky tree.
(\ding{43}~\HL{taw}{\B{taw}}, \HL{utral}{\B{ut\-ral}})

% tayng
\hi 
\HT{tayng}{\B{tayng}} \I{[tajN]} \emph{n.}, \ding{96} thistle\hyp{}like plant (\emph{generic term}).  \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tompatayng}{\B{tom\-pa\-tayng}}, rain thistle. 

% Tayrangi
\hi
\HT{tayrangi}{\B{Tayrangi}} \I{[taj.Ra.Ni]} \emph{n., clan name, specialise in banshee fishing}.

% taweyk
\hi
\HT{taweyk}{\B{ta\cp{weyk}}} \emph{or} \B{ta\cp{wey}ka} \I{[ta."wEjk\textcorner]}  \I{[ta."wEj.ka]} \emph{conj.} because.
(\ding{43}~\HL{ta}{\B{ta}}, \HL{oeyk}{\B{o\-eyk}}, \HL{talun1}{\B{ta\-lun(a)}}) 

% tawrrpa
\hi 
\HT{tawrrpa}{\B{ta\cp{wrr}pa}} \I{[ta."w\s{r}.pa]} \emph{adj.} external. 
\B{Fì\-txe\-lel fngo' sä\-lang\-it a\-ta\-wrr\-pa.} This matter requires an external investigation.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\ding{214}~\HL{tamIfa}{\B{ta\-mì\-fa}}, \ding{43}~\HL{ta}{\B{ta}}, \HL{wrrpa}{\B{wrr\-pa}}) 

% taw
\hi
\HT{taw}{\B{taw}} \I{[taw]} \emph{n.} sky.
\B{Kxam\-trr lam fwa san\-hì a mì saw 'o\-lìp nì\-wotx slä tsa\-krr ke tsun fko sat tsi\-ve\-'a nì\-'aw.} At mid\hyp{}day it seems that the stars in the sky have all vanished but they just can't be seen then.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
\B{Hol\-pxay san\-hì\-yä a mì saw lu ke\-tsuk\-ti\-am; keng ke tsun fko tsi\-ve\-'a sat nì\-wotx.} The number of stars in the sky is infinite; it's not even possible to see them all.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\emph{a. of weather}) \B{Ya\-fkeyk fya\-pe set? -- Taw lu pi\-ak\slash lep\-wopx\slash tstu.} What's the weather like now? -- The sky is clear\slash cloudy\slash overcast.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} 
\B{Taw\-ri fya\-pe?} What's the sky like?~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\emph{b. for celestial bodies rising and setting})
\B{Tsaw\-ke fpxe\-rä\-kìm (nem\-fa taw).} The sun is rising.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Tsay\-san\-hì ha\-yum ye'\-rìn (taw\-ftu).} Those stars will soon set.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tawtute}{\B{taw\-tu\-te}}, human, sky\hyp{}person. 
\on{2}.~\HL{tawsyuratan}{\B{taw\-syu\-ra\-tan}}, aurora. 
\on{3}.~\HL{tawtsngal}{\B{taw\-tsngal}}, panopyra. 
\on{4}.~\HL{tautral}{\B{ta\-ut\-ral}}, sky\hyp{}tree.
\on{5}.~\HL{tawsIp}{\B{taw\-sìp}}, spaceship.
\on{6}.~\HL{tawtxew}{\B{tawtxew}}, horizon, skyline.

% Tawkami
\hi
\HT{Tawkami}{\B{Taw\cp{ka}mi}} \I{[taw."ka.mi]} \emph{n.} \emph{clan name, experts in botany}.

% tawng
\hi
\HT{tawng}{\B{tawng}} \I{[tawN]} \emph{vin.} dive, duck (\emph{into water}).
\B{tawng nem\-fa pay}, dive into water.~\LINK{http://naviteri.org/2021/02/aysipawm-si-aysieyng-questions-and-answers/}{b} 

% tawsìp
\hi
\HT{tawsIp}{\B{\cp{taw}sìp}} \I{["taw.sIp\textcorner]} \emph{n.} spaceship, \emph{compound of Na'vi} \HL{taw}{\B{taw}} ``sky'' \emph{and English} \textsc{lw} ``ship''.
\B{Taw\-sìp nge\-yä lu sngel\-tseng.} Your ship is a garbage scow.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog_.28Twitter_Questions.29}{ugo\slash wiki}

% Tawsnrrtsyìp
\hi 
\HT{tawsnrrtsyIp}{\B{Taw\cp{snrr}tsyìp}} \I{[taw."{}sn\s{r}.\t{ts}jIp\textcorner]} \emph{n., astr.} Alpha Centauri C\slash Proxima Centauri; \emph{often shortened to} \ding{43}~\HL{snrrtsyIp}{\B{Snrrtsyìp}} 
(\ding{43}~\HL{taw}{\B{taw}}, \HL{sAnrr}{\B{sä\-nrr}}, \HL{-tsyIp}{\B{-tsyìp}})

% tawsyuratan
\hi
\HT{tawsyuratan}{\B{tawsyu\cp{ra}tan}} \I{[taw.sju."Ra.tan]} \emph{n.} aurora.
(\ding{43}~\HL{taw}{\B{taw}}, \HL{syuratan}{\B{syu\-ra\-tan}}) 

% tawtute
\hi
\HT{tawtute}{\B{\cp{taw}tute}} \I{["taw.tu.tE]} \emph{n.} human, ``skyperson''.
\B{Na'\-vi\-ru set lu nawm\-a eyk\-tan a\-mip a lar\-mu Taw\-tu\-te.} The Na'vi have a great new leader now who had been a Skyperson.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Lì'\-fya\-ri tsa\-taw\-tu\-te pe\-yì?} How is that Sky Person's Na'vi?.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Tì\-'e\-fu\-mì oe\-yä su\-te\-kip nì\-wotx, ftxey Na'\-vi ftxey Saw\-tu\-te, lu sìl\-tsan lu kawng.} In my opinion among all people, whether Na'vi or Skypeople, there are good and bad.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Lum\-pe le\-rang Kel\-ut\-ral\-it Saw\-tu\-tel?} Why are the humans exploring Hometree?~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Ay\-oe\-ngal fte Fay\-saw\-tu\-tet ki\-vu\-rakx, ze\-ne nì\-'aw\-ve ay\-fot tsli\-vam.} We must understand these Sky People if we are to drive them out.~\LINK{http://forum.learnnavi.org/language-updates/confirmation-on-kurakx-eytukan-line-and-toitslan/}{ln\slash a\textsuperscript{c}}
\B{Sra\-ne, su\-nu Saw\-tu\-te\-ru tì\-lang, slä ke omum fol teyng\-ta kem\-pe ze\-ne si\-vi maw\-krr\-a 'uo\-ti ro\-lun.} Yes, the Skypeople love exploration, but they don't know what to do once they find something.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
\B{Saw\-tu\-te\-yä hem\-ìri lu aw\-nga\-ru ya\-yayr.} The Skypeople's actions confuse us.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Saw\-tu\-te\-ri tì\-fkey\-tok ke tsan\-'o\-lul kaw\-'it.} The situation with the Skypeople hasn't improved one bit.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\ding{43}~\HL{taw}{\B{taw}}, \HL{tute1}{\B{tu\-te}}) 

% tawtxew
\hi 
\HT{tawtxew}{\B{\cp{taw}txew}} \I{["taw.t'Ew]} \emph{n.} horizon, skyline (\B{sìn}\slash \B{ro}, on; \B{eo}, in front of; \B{uo}, behind). 
\B{Lu ay\-ram sìn taw\-txew.} There are mountains on the horizon.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
\B{Na\-ra\-nawm\-ä mawl mi lu uo taw\-txew.} Half of Polyphemus is still behind the horizon.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
(\ding{43}~\HL{taw}{\B{taw}}, \HL{txew}{\B{txew}})

% tawtsngal
\hi
\HT{tawtsngal}{\B{\cp{taw}tsngal}} \I{["taw.\t{ts}Nal]} \emph{n.}, \ding{96} panopyra, ``sky cup'', \emph{panopyra\slash maraca aerii}.
(\ding{43}~\HL{taw}{\B{taw}}, \HL{tsngal}{\B{tsngal}}) 

% täftxu
\hi
\HT{tAftxu}{\B{tä\cp{ftxu}}} \I{[t\ae."ft'u]} \emph{vtr.} weave.   
\B{Ka\-tot tä\-ftxu oel | Nì\-ean nì\-rim.} I weave the rhythm | In yellow and blue.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
\B{Fu\-ri\-a tä\-ftxu ngal tok yì\-pet? -- Tok yìt a\-ke\-sran.} How's your weaving? -- I'm so\hyp{}so.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tAftxuyu}{\B{tä\-ftxu\-yu}}, weaver. 
\on{2}.~\HL{ftxulI'u}{\B{ftxu\-lì\-'u}}, orate, give a speech.
\on{3}.~\HL{sAtAftxu}{\B{sä\-tä\-ftxu}}, piece of weaving.

% täftxuyu
\hi
\HT{tAftxuyu}{\B{tä\cp{ftxu}yu}} \I{[t\ae."ft'u.ju]} \emph{n.} weaver, one who weaves. 
\B{Pam\-tseol ngop ay\-re\-nut | Mì ron\-sem\-ä tì\-fnu | Teng\-fya ngop sä\-ftxu\-yul | Mì hi\-fkey.} Music creates patterns | In the silence of the mind | As weavers do | In the physical word.~\LINK{https://www.pandorapedia.com/navi/music/ritual_music.html}{pp}
(\ding{43}~\HL{tAftxu}{\B{tä\-ftxu}}, \HL{-yu}{\B{-yu}}) 

% tätxaw
\hi
\HT{tAtxaw}{\B{tä\cp{txaw}}} \I{[t\ae."t'aw]} \emph{vin.} return (\B{ne}, to).   
\B{Set sre\-fey oel fu\-ta tsam\-po\-ngu tä\-txaw maw txon\-'ong.} I'm currently expecting the war party back after nightfall.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Krr\-pe nga tä\-txaw ne kel\-ku?} When will you return home?~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Maw\-krr\-a Saw\-tu\-te\-yä txam\-pxì ho\-lum, tä\-txaw Na'\-vi\-ne nì\-'i'\-a ve\-'o.} After most of the Sky People left, order finally returned to the People.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Maw hì\-krr ay\-oe tì\-yä\-txaw.} We'll return after a short time.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
\B{Sre\-fpìl Oma\-ti\-ka\-ya tsnì Tsyeyk kaw\-krr ke ta\-yä\-txaw maw ka\-vuk sne\-yä.} The Omaticaya assumed that Jake would never return after his treachery.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\emph{a. idiom}) \B{To\-lä\-txaw nì\-prr\-te'.} ``Welcome back.''~\LINK{http://forum.learnnavi.org/language-updates/some-news-from-berlin/msg585072/\#msg585072}{ln} 

% te
\hi
\HT{te}{\B{te}} \I{[tE]} \emph{part. used in full names}. 
\B{Ney\-ti\-ri te Tska\-ha Mo\-'at\-'ite}, Neytiri of the Tskaha, daughter of Mo'at.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Wutso_in_Hometree}{wiki\slash a\textsuperscript{c}} 
\B{Ay\-nge\-nga\-ru ohe\-yä tsmuk\-it alu Ne\-wey te Tska\-ha So\-rewn\-'ite.} Allow me to introduce my sister, Ne\-wey te Tska\-ha So\-rewn\-'ite.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Tsu'\-tey te Rong\-loa, A\-te\-yo\-'itan}, Tsu'tey of the Rongloa, son of Ateyo.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Toruk_Makto}{wiki\slash a}

% tekre
\hi
\HT{tekre}{\B{\cp{tek}re}} \I{["tEk\textcorner.rE]} \emph{n.} skull. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nawmtoruktek}{\B{nawm\-to\-ruk\-tek}}, Toruk Makto Totem.

% tel
\hi
\HT{tel}{\B{tel}} \I{[tEl]} \emph{vtr.} receive (\B{ta}, from, \B{yoa}, in exchange for).   
\B{Oel tel 'u\-pxa\-ret le\-Na'\-vi a krr, new oe nì\-teng\-fya pam\-rel si\-vi nì\-'eyng.} When I receive a Na'vi message, I want to write the same way in response.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln}
\B{Fu\-ta nga\-ta tel pxen\-yoa srät, mll\-'e\-i\-an oel.} I'm happy to take cloth from you for finished garments.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Tì\-'eyng\-it oel to\-lel a krr, ay\-nga\-ru pa\-yeng.} When I receive an answer, I will let you know.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
(\emph{a. conv. idiom}) \B{To\-\cp{lel}}! ``Eureka! I got it! I understand!''~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}

% telem
\hi
\HT{telem}{\B{te\cp{lem}}} \I{[tE."lEm]} \emph{n.} cord. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txurtel}{\B{txur\-tel}}, rope. 
\on{2}.~\HL{waytelem}{\B{way\-te\-lem}}, song cord.

% telisi
\hi 
\HT{telisi}{\B{teli\cp{si}}} \I{[tE.li."{}si]} \emph{n.} whirlwind. 

% tem
\hi
\HT{tem}{\B{tem}} \I{[tEm]} \emph{vin.} shoot (\B{ne}, at).
\B{Tem rä\-'ä!} Don't shoot!~\LINK{http://naviteri.org/2017/03/titusemteri-concerning-shooting/}{b} 
\B{Oe\-ne fko te\-rem!} Someone is shooting at me!~\LINK{http://naviteri.org/2017/03/titusemteri-concerning-shooting/}{b} 
\B{Txo nga ze\-ne ti\-vem, tsa\-tì\-tu\-sem li\-vu muiä.} If you must shoot, let it be justified.~\LINK{http://naviteri.org/2017/03/titusemteri-concerning-shooting/}{b} 
(\ding{43}~\HL{toltem}{\B{tol\-tem}}; \HL{tsweykayon}{\B{tswa\-yon} (\emph{b. \ding{246}})})

% temrey
\hi
\HT{temrey}{\B{tem\cp{rey}}} \I{[tEm."REj]} \emph{n.} survival (\emph{in the face of danger}). 
\B{Fwam\-pop\-ä tem\-rey tsa\-tseng\-mì ngä\-zìk lu nì\-wotx.} It's very hard for a tapirus to survive there.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{emrey}{\B{em\-rey}})

% teng
\hi
\HT{teng}{\B{teng}} \I{[tEN]} \emph{adj.} equal, same. 
\B{Fì\-txon na ton a\-la\-he nì\-wotx pe\-lun ke lu teng?} Why is this night not the same as all the other nights?~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
\B{Tsa\-krr syay ay\-nge\-yä, syay o\-lo'\-ä oe\-yä la\-yu teng.} Then you will suffer the same fate as my clan.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Me\-unil\-tì\-ran\-tokx Tok\-tor Kì\-rey\-sä sì Tsyeyk\-ä ke lu teng ki steng.} The avatars of Grace and Jake aren't the same, but they are similar.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\emph{a. proverb}) \B{Sä\-fpìl a\-steng, tì\-kan a\-teng.} Great minds think alike [lit.: similar thought, same aim].~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{214}~\HL{keteng}{\B{ke\-teng}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{nIteng}{\B{nì\-teng}}, too, also, likewise. 
\on{2}.~\HL{tengfya}{\B{teng\-fya}}, as, same way as (conj.). 
\on{3}.~\HL{tengkrr}{\B{teng\-krr}}, while (conj.). 
\on{4}.~\HL{'awsiteng}{\B{'aw\-si\-teng}}, together. 
\on{5}.~\HL{keteng}{\B{ke\-teng}}, different. 
\on{6}.~\HL{kAteng}{\B{kä\-teng}}, spend time with, hang out with. 
\on{7}.~\HL{'awstengyem}{\B{'aw\-steng\-yem}}, join (two or more things) together.
\on{8}.~\HL{tengralI'u}{\B{teng\-ra\-lì\-'u}}, synonym.
\on{9}.~\HL{tengwotx}{\B{teng\-wotx}}, uniform, non\hyp{}diverse.

% tengfya
\hi
\HT{tengfya}{\B{\cp{teng}fya}} \I{["tEN.fja]} \emph{conj.} as. 
\B{Pam\-tseol ngop ay\-re\-nut | Mì ron\-sem\-ä tì\-fnu | Teng\-fya ngop sä\-ftxu\-yul | Mì hi\-fkey.} Music creates patterns | In the silence of the mind | As weavers do | In the physical word.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Teng\-fya tsun tsi\-ve\-'a, lup\-ra el\-tur tì\-txen si nì\-txan.} As you can see, the style is very interesting.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
(\ding{43}~\B{teng}, \B{fya\-'o}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nItengfya}{\B{nì\-teng\-fya}}, in the same way.

% tengkrr
\hi
\HT{tengkrr}{\B{teng\cp{krr}}} \I{[tEN."k\s{r}]} \emph{conj.} while (\emph{doing s.t.}). \emph{Usually with a form of} ‹\B{er}› \emph{in the following verb.}
\B{Re\-rol teng\-krr ke\-rä | Ìlä fya\-'o a\-vol | Ne kxam\-tseng.} We sing our way | Down the eight paths | To the center.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Na'\-vi ìlä ho\-'on kll\-kxo\-lem teng\-krr re\-rol.} The People were standing in a circle, singing.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Ay\-nga\-ri teng\-krr ya wur sle\-ru nì\-'ul, sìl\-pey oe, li\-vu hel\-ku sang ul\-te te'\-lan le\-fpom.} As for you all, while it's getting cooler, I hope your homes may be warm and your hearts happy.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{a. idiom}) \B{Teng\-krr ftxo\-zä se\-rei\-yi aw\-nga, ke tswi\-va' ay\-lom\-tu\-ti ko!} A toast: To absent friends. [lit.: While we are celebrating, let us not forget those who we wish could also be here (but can't)]~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
(\ding{43}~\HL{teng}{\B{teng}}, \HL{krr}{\B{krr}})

% tengralì'u
\hi 
\HT{tengralI'u}{\B{\cp{teng}ralì'u}} \I{["tEN.Ra.lI.Pu]} \emph{n.} synonym. 
\B{Me\-lì\-'u alu pxel sì na lu teng\-ra\-lì\-'u.} The two words \emph{pxel} and \emph{na} are synonyms.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{teng}{\B{teng}}, \HL{ral}{\B{ral}}, \HL{lI'u}{\B{lì\-'u}}; \ding{214}~\HL{wAralI'u}{\B{wä\-ra\-lì\-'u}}) 

% tengwotx
\hi 
\HT{tengwotx}{\B{\cp{teng}wotx}} \I{["tEN.wot']} \emph{adj.} uniform, non\hyp{}diverse.
\B{Lì'\-fya le\-Na'\-vi lu teng\-wotx, ken ay\-reym\-sä\-tsa le\-ke\-teng\-syon Ey\-wa\-'e\-veng\-ä, ken ho\-et\-a ta\-yo a mì\-kam ay\-olo'.} The language of the Na'vi is consistent, despite the exo\hyp{}moon's varied landscapes and the vast expanse between clans.~\LINK{https://naviteri.org/2026/06/tskxekeng-a-mikyunfpi-pamrel-listening-exercise-text/}{b}
(\ding{43}~\HL{teng}{\B{teng}}, \HL{nIwotx}{\B{nì\-wotx}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tItengwotx}{\B{tì\-teng\-wotx}}, uniformity.

% teri
\hi
\HT{teri}{\B{te\cp{ri}}} \I{[tE."Ri]} \emph{adp.} about, concerning.
(\emph{a.})
\B{Lu fra\-'u te\-ri tì\-ftxa\-vang.} Everything is about passion.~\LINK{http://forum.learnnavi.org/news-announcements/paul-frommer-at-grid-2010/}{ln}
\B{Nge\-yä te\-ri fay\-te\-le a ay\-sä\-nu\-me\-ri ngar ira\-yo sei\-yi ay\-oe nì\-wotx.} We all thank you for your teachings about these topics.~\LINK{http://forum.learnnavi.org/language-updates/combined-answers-1-\%28feb-16\%29/}{ln}
(\emph{b. with the verbs} \ding{43}~\B{\HL{pAngkxo}{päng\-kxo}, \HL{peng}{peng}, \HL{stawm}{stawm}, \HL{nume}{nu\-me}, \HL{oeyktIng}{oeyk\-tìng} te\-ri}) chat, tell, hear, learn, explain about.
\B{Fay\-upxa\-re\-mì oe pa\-yäng\-kxo te\-ri ho\-ren lì'\-fya\-yä le\-Na'\-vi.} In these messages I will chat about the rules of the Na'vi language.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Oe pi\-veng ay\-nga\-ru nì\-'it te\-ri Txe\-wì alu Na'\-vi\-yä 'e\-veng.} I'd like to tell you a bit about Txewì, a Na'vi child.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Krr a sto\-lawm Txe\-wì te\-ri hem a poe\-ru lo\-len, 'o\-le\-fu ke\-ftxo nì\-txan.} When Txewì heard about what happened to her, he felt very upset.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Oel vewng fu\-ta ay\-e\-veng ni\-vu\-me te\-ri ay\-ewll na'\-rìng\-ä.} I see to it that the children learn about the forest plants.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Wo\-leyn Ìstawl yey\-fyat mì hll\-te fte oeyk\-ti\-vìng fra\-po\-ru tì\-hawl\-te\-ri sne\-yä.} Ìstaw drew a line on the ground to explain his plan to everyone.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}

% terkup
\hi
\HT{terkup}{\B{\cp{ter}kup}} \I{["tER.kup\textcorner]} \emph{or} \I{[.kUp\textcorner]} (\textsc{rn}: \B{\cp{ter}kùp} \I{["tER.kUP\textcorner]}) \emph{vin.} die. 
\B{Kop oe\-ru lo\-lä\-ngu lie a ha\-pxì\-tu soaiä ter\-kup.} Sadly, I too have had the experience of a family member dying.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Ngal po\-leng oer fmawn\-ta po to\-ler\-kup.} You told me that he died.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Trr\-am to\-ler\-kä\-ngup sa'\-nok a\-yaw\-ne.} My dear mother died yesterday.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Po tä\-pey\-kì\-ye\-ver\-kei\-up nì\-näk.} I'm so jazzed that he may be about to drink himself to death.~\LINK{http://naviteri.org/2010/12/catching-up/}{b}
(\emph{a. coll. idiom}) \B{[verb] nì\-ftxan kum\-a ter\-kup}, \emph{commenting on an action that is so intense that it results in dying}; ``like crazy, \ldots{} to death''. 
\B{Sran, sran, oe\-ri pì\-tìk tsa\-tse\-nget a mì tal! Fkxa\-ke nì\-ftxan kum\-a ter\-kup.} Yeah, yeah, scratch that place on my back! It itches like crazy!~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIterkup}{\B{tì\-ter\-kup}}, death, dying.

% teswotìng 
\hi
\HT{teswotIng}{\B{te\cp{swo}tìng}} \I{[tE."{}swo.tIN]} \emph{vtr.} grant. 
\B{Fì\-ta\-ron\-yur a\-pxan te\-swo\-ti\-vìng äiet a\-ngay.} Grant this worthy warrior true vision.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Becoming_one_of_the_Na.27vi}{a\textsuperscript{c}}

% tete
\hi
\HT{tete}{\B{\cp{te}te}} \I{["tE.tE]} \emph{adj.} dull (\emph{of a blade}).
(\ding{214}~\HL{litx}{\B{litx}}, \ding{43}~\HL{fwem}{\B{fwem}}) 

% tewng
\hi
\HT{tewng}{\B{tewng}} \I{[tEwN]} \emph{n., text.} loincloth.

% tewti
\hi
\HT{tewti}{\B{\cp{tew}ti}} \I{["tEw.ti]} \emph{intj. expression of (pleasant) surprise}, ``Wow! Whoa!''.
\B{Tew\-ti, nga lu tsul\-fä\-tu i\-'en\-ä.} Wow, you are a master on the \emph{i'en}.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}

% teya / si  
\hi
\HT{teya}{\B{te\cp{ya}}} \I{[tE."ja]} \emph{adj.} full (\HL{ta}{\B{ta}}, of).
\B{Oe\-ri ta pe\-yä fa\-hew a\-ke\-wong on\-tu te\-ya lä\-ngu.} My nose is full of his alien smell.~\LINK{http://languagelog.ldc.upenn.edu/nll/?p=1977}{ll\slash a}
\B{Eo ay\-fo a fya\-'o la\-mu ay\-skxe\-ta te\-ya sì re\-nul\-ke.} The path ahead of them was full of rocks and irregular.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Ay\-nge\-yä ay\-srr tì\-flä\-ta te\-ya li\-vu ul\-te si\-va\-lew nì\-'o' nì\-'aw.} May your days be full of success and proceed only in a fun way.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/}{b} \\
\textsc{ii}. \B{te\cp{ya} si} \I{[tE."ja si]} \emph{vin.} (\emph{a.}) fill, make full. 
\B{Fì\-po\-stì mek\-tseng\-ur te\-ya si.} This post fills a gap (in our understanding of Na'vi syntax).~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}  
(\emph{b. idiom}) \B{oe\-ru te\-ya si}, ``s.t. fills me with satisfaction\slash joy''. 
\B{Sì ay\-zì\-sìt\-ä ka\-to | Sì ekong te'\-lan\-ä | Te'\-lan le\-Na'\-vi | Oe\-ru te\-ya si.} And the rhythm of the years | And the beats of the hearts | The hearts of the People | Fill me.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
\B{Fay\-lì\-'u a\-lor oe\-ru te\-ya si nì\-txan, ul\-te ke tswa\-ya' oel sat kaw\-krr.} These beautiful words fill me with joy and I'll never forget them.~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
\B{Tì\-tstun\-wi ay\-nge\-yä oe\-ru te\-ya so\-li.} Your kindness have filled me with joy.~\LINK{http://forum.learnnavi.org/community-calendar/karyu-pawls-birthday/msg312733/\#msg312733}{ln}
\B{Moe\-ru te\-ya si nì\-ngay.} We're greatly touched.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
(\ding{214}~\HL{mek}{\B{mek}})

% teylu
\hi
\HT{teylu}{\B{\cp{tey}lu}} \I{["tEj.lu]} \emph{n., fauna} beetle larva(e).   
\B{Tey\-lu kì\-mar lì\-yu a fì\-'u oe\-ru te\-ya si.} It fills me with joy that \emph{teylu} is about to be in season.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Oe new yi\-vom tey\-lut.} I want to eat \emph{teylu}.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/}{b}
\B{Fol fnan fu\-ta 'em tey\-lut a fì\-'u sìn pe\-yì?} How good are they at cooking \emph{teylu}?~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Fì\-tey\-lu\-ri ke nar\-mew Va'\-ru yi\-vom nì\-yll.} Va'ru didn't want to share this \emph{teylu} with the Omatikaya.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{teylupil}{\B{tey\-lu\-pil}}, \emph{teylu}\hyp{}face (\emph{childish insult}). 

% teylupil
\hi 
\HT{teylupil}{\B{\cp{tey}lupil}} \I{["tEj.lu.pil]} \emph{n.} \emph{teylu}\hyp{}face (\emph{a somewhat childish insult}). 
(\ding{43}~\HL{teylu}{\B{tey\-lu}}, \HL{pil}{\B{pil}})

% teynga 
\hi
\HT{teynga}{\B{\cp{tey}nga}} \I{["tEj.Na]} \emph{conj. to form indirect questions; shortened form of} \ding{43}~\HL{tI'eyng}{\B{tì\-'eyng a}}.   
\B{Tey\-nga lum\-pe fo ho\-lum ke lu law.} It's not clear why they left.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Tey\-nga pe\-su ni\-vew hi\-vum ke tsran\-ten, po\-ru pll\-txe san ru\-txe 'i\-vì\-'awn.} No matter who wants to leave, tell them to please stay.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
(\emph{a. with adp.s})
\B{Ay\-oe pe\-räng\-kxo te\-ri tey\-nga lum\-pe po ho\-lum.} We're chatting about why he left.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
\B{Tì\-flä la\-tem ìlä seyng\-a ftxey fkol sä\-nu\-met li\-vek fu\-ke.} Success depends on whether or not one follows instructions.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{teyngla}{\B{teyng\-la}}, \HL{teyngta}{\B{teyng\-ta}}, \HL{teyngra}{\B{teyng\-ra}}, \HL{teyngA}{\B{tey\-ngä}}, \HL{teyngria}{\B{teyng\-ri\-a}})

% teyngä
\hi
\HT{teyngA}{\B{\cp{tey}ngä}} \I{["tEj.N\ae]} \emph{conj. to form indirect questions; shortened form of} \ding{43}~\HL{tI'eyng}{\B{tì\-'ey\-ngä a}}.   
\B{Tì\-run\-ìl tey\-ngä lum\-pe po ho\-lum oe\-ti ke\-ftxo 'ey\-ko\-le\-fu.} The discovery of why he left saddened me.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{teynga}{\B{tey\-nga}}, \HL{teyngla}{\B{teyng\-la}}, \HL{teyngta}{\B{teyng\-ta}}, \HL{teyngra}{\B{teyng\-ra}}, \HL{teyngria}{\B{teyng\-ri\-a}})

% teyngla
\hi
\HT{teyngla}{\B{\cp{tey}ngla}} \I{["tEjN.la]} \emph{conj. to form indirect questions; shortened form of} \ding{43}~\HL{tI'eyng}{\B{tì\-'eyng\-ìl a}}.
\B{Teyng\-la lum\-pe po ho\-lum oe\-ti hey\-ko\-lang\-ham.} Why he left made me laugh.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{teynga}{\B{tey\-nga}}, \HL{teyngta}{\B{teyng\-ta}}, \HL{teyngra}{\B{teyng\-ra}}, \HL{teyngA}{\B{tey\-ngä}}, \HL{teyngria}{\B{teyng\-ri\-a}})

% teyngra
\hi
\HT{teyngra}{\B{\cp{teyng}ra}} \I{["tEjN.Ra]} \emph{conj. to form indirect questions; shortened form of} \ding{43}~\HL{tI'eyng}{\B{tì\-'eyng\-ur a}}.
\B{Ru\-txe law si\-vi teyng\-ra lum\-pe po ho\-lum.} Please clarify why he left.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{teynga}{\B{tey\-nga}}, \HL{teyngla}{\B{teyng\-la}}, \HL{teyngta}{\B{teyng\-ta}}, \HL{teyngA}{\B{tey\-ngä}}, \HL{teyngria}{\B{teyng\-ri\-a}})

% teyngria
\hi
\HT{teyngria}{\B{\cp{teyng}ria}} \I{["tEjN.Ri.a]} \emph{conj. to form indirect questions; shortened form of} \ding{43}~\HL{tI'eyng}{\B{tì\-'eyng\-ri a}}.   
\B{Teyng\-ria lum\-pe po ho\-lum oel ke tslam ke\-'ut.} I understand nothing about why he left.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{teynga}{\B{tey\-nga}}, \HL{teyngla}{\B{teyng\-la}}, \HL{teyngta}{\B{teyng\-ta}}, \HL{teyngra}{\B{teyng\-ra}}, \HL{teyngA}{\B{tey\-ngä}})

% teyngta
\hi
\HT{teyngta}{\B{\cp{tey}ngta}} \I{["tEjN.ta]} \emph{conj. to form indirect questions; shortened form of} \ding{43}~\HL{tI'eyng}{\B{tì\-'eyng\-it a}}.
\B{Vo\-lin oel teyng\-ta Ney\-ti\-ri kä pe\-seng\-ne.} I asked where Neytiri was going.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} 
\B{Ngal ke tslä\-ngam teyng\-ta fì\-tì\-ngä\-zìk\-ìri pe\-fyin\-ep\-'ang.} Unfortunately you don't understand how complex this problem is.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Ke o\-mä\-ngum oel teyng\-ta ftu pe\-seng zo\-la\-'u tsa\-pam\-rel\-vi alu t.} I'm afraid I don't know where that letter t came from.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/comment-page-1/\#comment-2612}{b} 
(\ding{43}~\HL{teynga}{\B{tey\-nga}}, \HL{teyngla}{\B{teyng\-la}}, \HL{teyngra}{\B{teyng\-ra}}, \HL{teyngA}{\B{tey\-ngä}}, \HL{teyngria}{\B{teyng\-ri\-a}})

% teyr
\hi
\HT{teyr}{\B{teyr}} \I{[tEjR]} \emph{adj.} white. 
\B{Ro\-lun ay\-oel tsrul\-mì hì\-'i\-a pxe\-lo\-it a\-teyr.} We found three little white eggs in the nest.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}
(\ding{214}~\HL{layon}{\B{la\-yon}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{teyrpin}{\B{teyr\-pin}}, the color white.

% teyrpin
\hi
\HT{teyrpin}{\B{\cp{teyr}pin}} \I{["tEjR.pin]} \emph{n.} the color \ding{43}~\HL{teyr}{\B{teyr}}.
(\ding{43}~\HL{'opin}{\B{'o\-pin}})

% -ti 
\hi
\HT{-ti}{\B{-ti}} \I{[.ti]} \emph{suf. to mark the patientive case of a n. ending in a vowel, pseudo\hyp{}vowel, diphthong, or consonant, preferred ending in} \textsc{rn}.
(\ding{43}~\HL{-it}{\B{-it}}, \HL{-t}{\B{-t}}) 

% ti'a
\hi 
\HT{ti'a}{\B{\cp{ti}'a}} \I{["ti.Pa]} \emph{adv.} surprisingly (\emph{for unexpectedly positive outcomes}). 
\B{Ra\-mu ke lu txur nìtxan, slä u\-van\-it yo\-lo\-ra', ti\-'a.} Ra\-mu isn't very strong, but surprisingly, he won the game.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} 
(\ding{214}~\HL{um'a}{\B{um\-'a}}) 

% tiam
\hi
\HT{tiam}{\B{\cp{ti}am}} \I{["ti.am]} \emph{vtr.} count (\HL{ta}{\B{ta}}, from; \HL{vay}{\B{vay}}, (up) to; \HL{raw}{\B{raw}}, (down) to).
\B{Ru\-txe ti\-vi\-am ay\-srok\-it tsa\-krr hol\-pxay\-ti pi\-veng oer.} Please count the beads and tell me the number.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
\B{Ti\-am (ta pxey) vay vol.} Count (from three) to eight.~\LINK{https://forum.learnnavi.org/language-updates/'count-to'-wa-and-sequential-genitive-(discussion)/}{ln}
\B{Ti\-am ta vo\-mrr raw pxey.} Count down from thirteen to three.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words\%ef\%bb\%bf/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{ketsuktiam}{\B{ke\-tsuk\-ti\-am}}, uncountable, infinite.

% til
\hi
\HT{til}{\B{til}} \I{[til]} \emph{n.} joint, hinge. 
(\ding{43}~\HL{il}{\B{il}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{kinamtil}{\B{ki\-nam\-til}}, knee.
\on{2}.~\HL{pxuntil}{\B{pxun\-til}}, elbow.
\on{3}.~\HL{ventil}{\B{ven\-til}}, ankle.
\on{4}.~\HL{tilI'u}{\B{ti\-lì\-'u}}, conjunction.
\on{5}.~\HL{zektil}{\B{zek\-til}}, knuckle.

% tilì'u
\hi 
\HT{tilI'u}{\B{\cp{ti}lì'u}} \I{["ti.lI.Pu]} \emph{n.} conjunction. 
(\ding{43}~\HL{til}{\B{til}}, \HL{lI'u}{\B{lì\-'u}})

% tinan 
\hi
\HT{tinan}{\B{\cp{ti}nan}} \I{["ti.nan]} \emph{n.} reading.   
\B{Tsmìm\-ìri wätx fol ti\-nan\-it nì\-ngay.} They're really bad at reading animal tracks.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
(\ding{43}~\HL{inan}{\B{inan}})

% Tipani
\hi
\HT{Tipani}{\B{\cp{Ti}pani}} \I{["ti.pa.ni]} \emph{n.} \emph{clan name, expert warriors and hunters}.

% tirea
\hi
\HT{tirea}{\B{ti\cp{re}a}} \I{[ti."RE.a]} \emph{n.} spirit (\emph{of a being}). 
\B{Nga\-ri hu Ey\-wa sa\-lew ti\-rea, tokx 'ì\-'awn, slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body stays behind to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake.27s_final_kill}{a\slash wiki}
\B{Slä am\-'a\-lu\-ke nge\-yä ti\-re\-al tsa\-tse\-nget ta\-ye\-i\-ok.} But without a doubt your spirit will be there.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/comment-page-1/\#comment-1569}{b}
\B{nrr a ti\-re\-a}, glowing spirit.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tireafya'o}{\B{tirea\-fya\-'o}}, spirit path. 
\on{2}.~\HL{tireaioang}{\B{tirea\-ioang}}, spirit animal. 
\on{3}.~\HL{tireapAngkxo}{\B{tirea\-päng\-kxo}}, commune.
\on{4}.~\HL{slantire}{\B{slan\-ti\-re}}, inspiration. 
\on{5}.~\HL{tiretu}{\B{ti\-re\-tu}}, shaman.
\on{6}.~\HL{nItrea}{\B{nìt\-re\-a}}, in spirit.

% tireafya'o
\hi
\HT{tireafya'o}{\B{ti\cp{re}afya'o}} \I{[ti."RE.a.""fja.Po]} \emph{n.} spirit path. 
\B{Ay\-wayl yìm ki\-fkey\-ä | 'Ì\-he\-yut a\-vo\-mrr | Sìn ti\-rea\-fya\-'o a\-vol | Na way\-te\-lem\-ä hìng.} The songs bind | the thirteen spirals of the solid world | To the eight spirit paths | Like the threads of a Songcord.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
(\ding{43}~\HL{tirea}{\B{ti\-rea}}, \HL{fya'o}{\B{fya\-'o}})

% tireaioang
\hi
\HT{tireaioang}{\B{ti\cp{re}aioang}} \I{[ti."RE.a.i.""{}o.aN]} \emph{n.} spirit animal.   
\B{Nge\-yä Ti\-re\-a\-i\-o\-ang zo\-la\-'u a krr, law la\-yu nga\-ru.} When your Spirit Animal comes, you will know.~\LINK{http://forum.learnnavi.org/language-updates/auxilary-verb-si-possessive-dative-krr/}{ln}
(\ding{43}~\HL{tirea}{\B{ti\-rea}}, \HL{ioang}{\B{io\-ang}})

% tireapängkxo
\hi
\HT{tireapAngkxo}{\B{ti\cp{re}apängkxo}} \I{[ti."RE.a.p$\cdot$\ae{}N.""k'$\cdot$o]} \emph{vin., inf.}\on{45} commune (\emph{about intimate contact}).   
\B{Ke lu kaw\-tu a nul\-ni\-vew oe po\-hu ti\-rea\-pi\-väng\-kxo äo Ut\-ral Ay\-mok\-ri\-yä.} There's nobody I'd rather commune with under the Tree of Voices.~\LINK{http://wiki.learnnavi.org/Corpus\#Lemondrop.com}{ld\slash wiki}
(\ding{43}~\HL{tirea}{\B{ti\-rea}}, \HL{pAngkxo}{\B{päng\-kxo}})

% tiretu
\hi 
\HT{tiretu}{\B{ti\cp{re}tu}} \I{[ti."RE.tu]} \emph{n.} shaman.
(\ding{43}~\HL{tirea}{\B{ti\-re\-a}}, \HL{tute1}{\B{tu\-te}})

% tì-
\hi
\HT{tI-}{\B{tì-}} \I{[tI.]} \emph{unprod. pref. to turn adjectives, verbs or nouns into abstract nouns.}
(\ding{43}~\HL{sA-}{\B{sä-}})

% tì- us
\hi
\B{tì-} + ‹\B{us}› \emph{prod. process to form the gerund of a verb}.
\B{Ay\-nge\-yä tì\-ftu\-si\-a 'o' li\-vu nì\-'aw!} May your studying be fun.~\LINK{http://naviteri.org/2012/10/sound-files-added-to-previous-post/}{b}
\B{Tì\-tu\-sa\-ron mo\-wan lu oer nì\-ngay.} Hunting really turns me on.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Oel kan\-'ìn tì\-'u\-sem\-it.} I specialize in cooking.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-296}{b}
\B{Ko\-ren A\-'aw\-ve Tì\-ru\-sey\-ä 'Aw\-si\-teng}, the First Rule of Living Together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}

% tì'al
\hi 
\HT{tI'al}{\B{tì\cp{'al}}} \I{[tI."Pal]} \emph{n.} wastefulness.
\B{Tì\-'al lu zop\-lo a tsa\-ri ke tsun txoa li\-vu nì\-ftue.} Wastlefulness is an offense that cannot easily be forgiven.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} 
(\ding{43}~\HL{'al}{\B{'al}})

% tì'awm
\hi
\HT{tI'awm}{\B{tì\cp{'awm}}} \I{[tI."Pawm]} \textsc{i}. \emph{n.} camping. \\
\textsc{ii}. \B{tì\cp{'awm} si} \I{[tI."Pawm si]} \emph{vin.} to camp. 
(\ding{43}~\HL{'awm}{\B{'awm}})

% tì'awnì'aw
\hi 
\HT{tI'awnI'aw}{\B{tì\cp{'aw}nì'aw}} \I{[tI."Paw.nI.Paw]} \emph{n.} uniqueness.  
(\ding{43}~\HL{le'awnI'aw}{\B{le\-'aw\-nì\-'aw}})

% tì'awpo
\hi
\HT{tI'awpo}{\B{tì\cp{'aw}po}} \I{[tI."Paw.po]} \emph{n.} individuality (\emph{negative connotation}); selfishness. 
(\ding{43}~\HL{'awpo}{\B{'aw\-po}})

% tì'älek
\hi 
\HT{tI'Alek}{\B{tì\cp{'ä}lek}} \I{[tI."P\ae.lEk\textcorner]} \emph{n.} determination. 
\B{Pe\-yä tì\-'ä\-lek oe\-ru ro\-lo\-'a nì\-txan.} His determination impressed me greatly.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{'Alek}{\B{'ä\-lek}})

% tì'efu
\hi
\HT{tI'efu}{\B{tì\cp{'e}fu}} \I{[tI."PE.fu]} \emph{n.} feeling.
\B{Fwa lawk ay\-sì\-'efu\-ti ay\-ey\-lan\-kip lu fpom\-ro\-nga'.} It's healthy among friends to discuss feelings.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\emph{a. idiom}) \B{tì\-'e\-fu\-mì oe\-yä}, ``in my opinion''. \B{New oe ti\-vìng ay\-ngar lì\-'ut a tì\-'e\-fu\-mì oe\-yä lu lor fra\-to mì lì'\-fya le\-Na'\-vi.} I want to give you the word that, in my opinion, is the most beautiful in the Na'vi language.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{tì\-'e\-fu\-wä oe\-yä}, ``contrary to my opinion''.~\LINK{https://forum.learnnavi.org/language-updates/'count-to'-wa-and-sequential-genitive-(discussion)/}{ln}
\B{Tì\-'efu\-wä oe\-yä, fpìl Pey\-ral\-ìl fu\-ta ke ze\-ne ay\-oeng ki\-vä.} I think we have to go, but Peyral doesn't.~\LINK{https://forum.learnnavi.org/language-updates/'count-to'-wa-and-sequential-genitive-(discussion)/}{ln}
(\ding{43}~\HL{'efu}{\B{'efu}})

% tì'em
\hi
\HT{tI'em}{\B{tì\cp{'em}}} \I{[tI."PEm]} \emph{n.} the art of cooking.
\B{So\-rewn\-ìl kan\-'ìn tì\-'em\-it nì\-txan ul\-te kxawm tsa\-txe\-le me\-nga\-ne za\-'a\-tsu nì\-'eng.} Sorewn is really into cooking and you two might share that in common.\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{'em}{\B{'em}})

% tì'eylan
\hi
\HT{tI'eylan}{\B{tì\cp{'ey}lan}} \I{[tI."PEj.lan]} \emph{n.} friendship.
\B{Nge\-yä tì\-eyk\-tan\-ìri, tì\-slan\-ìri sì tsran\-ten fra\-to a tì\-'ey\-lan\-ìri a ka ay\-zì\-sìt nì\-wotx, sei\-yi oe ira\-yo nì\-txan.} Thank you so much for your leadership, your support, and most importantly your friendship throughout the years.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{'eylan}{\B{'eylan}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tI'eylanga'}{\B{tì\-'ey\-la\-nga'}}, friendly (nfp.).

% tì'eylanga'
\hi 
\HT{tI'eylanga'}{\B{tì\cp{'ey}langa'}} \I{[tI."PEj.la.NaP]} \emph{adj., nfp.} friendly. 
\B{Nge\-yä ay\-lì\-'u\-ri a\-tì\-'ey\-la\-nga' sei\-yi ira\-yo.} Thank you for your friendly words.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b}
(\ding{43}~\HL{let'eylan}{\B{let\-'ey\-lan}}; \HL{tI'eylan}{\B{tì\-'ey\-lan}}) 

% tì'eyng
\hi
\HT{tI'eyng}{\B{tì\cp{'eyng}}} \I{[tI."PEjN]} \emph{n.} answer, response.   
\B{Fo\-ru 'upxa\-ret oel fpo\-le', slä vay set ke pa\-mä\-hä\-ngem kea tì\-'eyng.} I have sent them a message, but up to now no answer has arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Tì\-'eyng\-it oel to\-lel a krr, tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} When I receive an answer, I will then decide the best way to bring you the words and rules of Na'vi.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ay\-nge\-yä sì\-pawm\-ìri kop fma\-yi fì\-tse\-nge ti\-vìng sì\-'eyng\-it.} As for your questions, I will also try to give answers here.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\emph{a. to form indirect questions}) \B{Vo\-lin oel tì\-'eyng\-it a Ney\-ti\-ri kä pe\-seng\-ne.} I asked where Neytiri was going.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b} 
(\emph{b. contracted forms for indirect questions:}) \HL{teynga}{\B{tey\-nga}} \HL{teyngla}{\B{teyng\-la}}, \HL{teyngta}{\B{teyng\-ta}}, \HL{teyngra}{\B{teyng\-ra}}, \HL{teyngA}{\B{tey\-ngä}}, \HL{teyngria}{\B{teyng\-ri\-a}}
(\ding{43}~\HL{'eyng}{\B{'eyng}})

% tì'ewan
\hi 
\HT{tI'ewan}{\B{tì\cp{'e}wan}} \I{[tI."PE.wan]} \emph{n.} youth (\emph{time of life}). 
(\emph{a. proverb}) \B{Nì\-win 'ìp tì\-'e\-wan, nì\-win pä\-hem tì\-ko\-ak.} Youth vanishes quickly, old age arrives quickly.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}
(\ding{43}~\HL{'ewan}{\B{'e\-wan}}; \ding{214}~\HL{tIkoak}{\B{tì\-ko\-vak}})

% tì'i'a
\hi
\HT{tI'i'a}{\B{tì\cp{'i'}a}} \I{[tI."PiP.a]} \emph{n.} ending, conclusion. 
\B{Tì\-'i'\-a\-ri tsam\-ä ze\-ne Saw\-tu\-te vi\-ve\-lek ta\-lun tì\-txur Ey\-wa\-yä.} At the end of the war, the Sky People had to give up due to the power of Eywa.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{a. idiom, poetic}) \B{tì\-'i'\-a\-vay krr\-ä}, forever, 'til the end of time.~\LINK{http://forum.learnnavi.org/language-updates/karyuru-frakrr-irayo-si-ko!/}{ln}
\B{To\-ler\-kup tu\-te; pam\-tseo pe\-yä tì\-'i'\-a\-vay krr\-ä ra\-yey.} The man has died; his music will live forever.~\LINK{http://naviteri.org/2015/06/na-ayskxe-mi-telan-sad-news/}{b}
(\emph{b. to tell the full hour}) \B{tì\-'i'\-a trr\-pxì\-yä a\-mrr\-ve}, `five o'clock', \emph{usually abbreviated to} \B{trr\-pxì a\-mrr\-ve}, `five o'clock'  \emph{or even} \B{Lu mrr\-ve.} `It's five.'~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
(\ding{214}~\HL{sngA'itseng}{\B{sngä\-'i\-tseng}}, \HL{sngA'ikrr}{\B{sngä\-'i\-krr}} \ding{43}~\HL{'i'a}{\B{'i'a}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tI'iluke}{\B{tì\-'i\-lu\-ke}}, endless, never\hyp{}ending.

% tì'iluke
\hi 
\HT{tI'iluke}{\B{tì\cp{'i}luke}} \I{[tI."Pi.lu.kE]} \emph{adj.} endless, never\hyp{}ending. 
\B{Hum zì\-sìt a\-lal, pä\-hem pum a\-mip. Yo'\-ko\-fya a\-tì\-'i\-lu\-ke.} The old year leaves, the new one arrives. An endless cycle.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
(\ding{43}~\HL{tI'i'a}{\B{tì\-'i\-'a}}, \HL{luke}{\B{lu\-ke}}; \HL{txewluke}{\B{txew\-lu\-ke}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIt'iluke}{\B{nìt\-'i\-lu\-ke}}, never\hyp{}endingly, forever. 

% tì'ipu
\hi
\HT{tI'ipu}{\B{tì\cp{'i}pu}} \I{[tI."Pi.pu]} \emph{n.} the abstract concept of being humorous, that is, humor in general.
\B{Sra\-ke tsun nga ri\-vun fì\-tì\-fkey\-tok\-mì a tì\-'i\-put?} Can you find the humor in this situation?~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{'ipu}{\B{'ipu}}, \HL{sA'ipu}{\B{sä\-'ipu}})

% tì'o'
\hi
\HT{tI'o'}{\B{tì\cp{'o'}}} \I{[tI."PoP]} \emph{n.} fun, excitement.
\B{Mì zì\-sìt a\-mip lì\-ye\-vu ay\-nga\-ru nì\-wotx txan\-a fpom\-tokx, fpom\-ron, tì\-yawn, sì tì\-'o'.} May you have much physical health, mental health, love and fun in the new year.~\LINK{http://naviteri.org/2014/12/ayngaru-tsinga-yoratu-may-i-present-the-four-winners/}{b}
\B{Tì\-'o'\-ìri pe\-u su\-nu ngar fra\-to?} What is your favorite way to have fun?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{'o'}{\B{'o'}})

% tì'ong
\hi
\HT{tI'ong}{\B{tì\cp{'ong}}} \I{[tI."PoN]} \emph{n.} blooming, unfolding.
\B{tì\-'ong lì'\-fya\-yä le\-Na'\-vi}, unfolding\slash blossoming of the Na'vi language.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{'ong}{\B{'ong}}) 

% tì'ongokx
\hi 
\HT{tI'ongokx}{\B{tì\cp{'o}ngokx}} \I{[tI."Po.Nok']} \emph{n.} birth. 
\B{Tì\-'o\-ngokx\-ìri nge\-yä 'itan\-ä sey\-kxel sì nit\-ram!} Congratulations on the birth of your son!~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} 
(\ding{43}~\HL{'ongokx}{\B{'o\-ngokx}})

% tì'ul
\hi 
\HT{tI'ul}{\B{tì\cp{'ul}}} \I{[tI."{}ul]} \emph{or} \I{[."{}Ul]} \emph{n.} increase. 
\B{Som\-wew\-ä tì\-'ul a ka 'Rr\-ta sä\-peng\-hrr lu aw\-nga\-ru nì\-wotx.} Global warming (lit.: the increase in temperature across the Earth) is a warning to us all.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{'ul}{\B{'ul}})

% tìaho
\hi 
\HT{tIaho}{\B{tìa\cp{ho}}} \I{[tI.a."ho]} \emph{n.} prayer (\emph{in general, abstract idea}). 
(\ding{43}~\HL{aho}{\B{aho}})

% tìeyktan
\hi
\HT{tIeyktan}{\B{tì\cp{eyk}tan}} \I{[tI."Ejk\textcorner.tan]} \emph{n.} leadership.
\B{Ke fkey\-tok tì\-eyk\-tan a\-tì\-flä\-nga' lu\-ke sno\-tipx.} Successful leadership does not exist without self-control.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
\B{Oeyk tsa\-tì\-snaytx\-ä lu apxa sä\-nui tì\-eyk\-tan\-ä.} That loss was caused by a great failure of leadership.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Nge\-yä tì\-eyk\-tan\-ìri, tì\-slan\-ìri sì tsran\-ten fra\-to a tì\-'ey\-lan\-ìri a ka ay\-zì\-sìt nì\-wotx, sei\-yi oe ira\-yo nì\-txan.} Thank you so much for your leadership, your support, and most importantly your friendship throughout the years.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{eyktan}{\B{eyk\-tan}})

% tìfe'ul
\hi
\HT{tIfe'ul}{\B{tì\cp{fe}'ul}} \I{[tI."fE.Pul]} \emph{n.} worsening.
(\ding{214}~\HL{tItsan'ul}{\B{tì\-tsan\-'ul}}, \ding{43}~\HL{fe'ul}{\B{fe\-'ul}}, \HL{sAfe'ul}{\B{sä\-fe\-'ul}})

% tìfenge
\hi 
\HT{tIfenge}{\B{tì\cp{fe}nge}} \I{[tI."fE.NE]} \emph{n.} ridicule.
(\ding{43}~\HL{fenge}{\B{fe\-nge}})

% tìfnawe'
\hi 
\HT{tIfnawe'}{\B{tìfna\cp{we'}}} \I{[tI.fna."wEP]} \emph{n.} cowardice. 
\B{Nge\-yä tì\-fna\-we' ke\-muia so\-li fì\-soaia\-ru.} Your cowardice has dishonored this family.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
(\ding{43}~\HL{fnawe'}{\B{fna\-we'}})

% tìfkeytok
\hi
\HT{tIfkeytok}{\B{tì\cp{fkey}tok}} \I{[tI."fkEj.tok\textcorner]} \emph{n.} state, condition, situation.
\B{Tì\-fkey\-tok le\-fkrr le\-hrr\-ap lu nì\-txan.} The current situation is very dangerous.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Fì\-fne\-tì\-fkey\-tok\-ìl fngo' fu\-ta kem si\-vi fko pxi\-ye'\-rìn.} This kind of situation requires immediate action.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Nge\-yä tsay\-lì\-'ul tì\-fkey\-tok\-it fe\-'ey\-ko\-lul nì\-'aw.} Those words of yours have only made the situation worse.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Sra\-ke tsun nga ri\-vun fì\-tì\-fkey\-tok\-mì a tì\-'i\-put?} Can you find the humor in this situation?~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{a. conv. phrase}) \B{Nga\-ri\slash Nge\-yä tì\-fkey\-tok (lu) fya\-pe?} How are you doing?~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/comment-page-1/\#comment-646}{b}
\B{Tì\-fkey\-tok pe\-fya\slash fya\-pe?} `What's the existing situation?\slash How are things?'~\LINK{http://naviteri.org/2018/02/negative-questions-in-navi/\#comment-28201}{b}
(\ding{43}~\HL{fkeytok}{\B{fkey\-tok}}, \HL{-fkeyk}{\B{-fkeyk}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIfkeytongay}{\B{tì\-fkey\-to\-ngay}}, reality.

% tìfkeytongay
\hi 
\HT{tIfkeytongay}{\B{tìfkeyto\cp{ngay}}} \I{[tI.fkEj.to."Naj]} \emph{n.} reality. 
\B{Ay\-unil nge\-yä lu lor, slä fì\-txe\-le\-ri lu tì\-fkey\-to\-ngay ke\-teng.} Your dreams are beautiful, but the reality of this situation is different.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{tIfkeytok}{\B{tì\-fkey\-tok}}, \HL{ngay}{\B{ngay}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIfkeytongay}{\B{nì\-fkey\-to\-ngay}}, actually, as a matter of fact.

% tìflä
\hi
\HT{tIflA}{\B{tì\cp{flä}}} \I{[tI."fl\ae]} \emph{n.} success (in general).
\B{Tì\-flä la\-tem ìlä seyng\-a ftxey fkol sä\-nu\-met li\-vek fu\-ke.} Success depends on whether or not one follows instructions.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Tì\-flä\-ri lo\-lu po 'ä\-lek.} She was determined to succeed.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b}
\B{Ay\-nge\-yä ay\-srr tì\-flä\-ta te\-ya li\-vu ul\-te si\-va\-lew nì\-'o' nì\-'aw.} May your days be full of success and proceed only in a fun way.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/}{b}

(\ding{214}~\HL{tInui}{\B{tì\-nui}}, \ding{43}~\HL{flA}{\B{flä}}, \HL{sAflA}{\B{sä\-flä}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIflAnga'}{\B{tì\-flä\-nga'}}, successful.

% tìflänga'
\hi 
\HT{tIflAnga'}{\B{tì\cp{flä}nga'}} \I{[tI."fl\ae.NaP]} \emph{adj., nfp.} successful. 
\B{tì\-hawl a\-tì\-flä\-nga'}, a successful plan.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}  
(\ding{43}~\HL{tIflA}{\B{tì\-flä}})

% tìflel
\hi 
\HT{tIflel}{\B{tì\cp{flel}}} \I{[tI."flEl]} \emph{n.} trickery (\emph{abstract concept}). 
(\ding{43}~\HL{flel}{\B{flel}}, \HL{sAflel}{\B{sä\-flel}})

% tìflrr
\hi
\HT{tIflrr}{\B{tì\cp{flrr}}} \I{[tI."fl\s{r}]} \emph{n.} gentleness, tenderness.
\B{Hu\-fwa me\-fo le\-ru mun\-txa\-tu txan\-krr, mi lu mun\-snar ho\-na tì\-flrr a na pum me\-yaw\-ne\-tu\-ä a\-mip nì\-wotx.} Although the two of them have been mates a long time, they still have all the adorable tenderness of new sweethearts.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{flrr}{\B{flrr}})

% tìfmetok
\hi
\HT{tIfmetok}{\B{tì\cp{fme}tok}} \I{[tI."fmE.tok\textcorner]} \emph{n.} test. 
\B{Tsyìl Ik\-ni\-ma\-yat ul\-te tsa\-heyl si ik\-ran\-hu a fì\-'u lu tì\-fme\-tok a ze\-ne fra\-ta\-ron\-yu a\-'e\-wan em\-zi\-va\-'u.} Scaling Iknimaya and bonding with a banshee is a test that every young hunter must pass.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Sì\-fme\-tok\-it em\-zo\-la\-'u o\-hel.} I have passed the tests.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{wiki\slash a\textsuperscript{c}}
(\ding{43}~\HL{fmetok}{\B{fme\-tok}})

% tìfmi
\hi
\HT{tIfmi}{\B{tì\cp{fmi}}} \I{[tI."fmi]} \emph{n.} attempt.
\B{Fì\-tseo\-ri ke tsun kaw\-tu pi\-vä\-hem tì\-yo'\-ne; tsran\-ten tì\-pä\-hem\-ä tì\-fmi nì\-'aw.} In this art it's impossible to arrive at perfection; the only thing that matters is the attempt to arrive there.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{fmi}{\B{fmi}}, \HL{sAfmi}{\B{sä\-fmi}})

% tìfmong
\hi 
\HT{tIfmong}{\B{tì\cp{fmong}}} \I{[tI."fmoN]} \emph{n.} theft. 
(\ding{43}~\HL{fmong}{\B{fmong}})

% tìfnu
\hi
\HT{tIfnu}{\B{tì\cp{fnu}}} \I{[tI."fnu]} \emph{n.} quiet, silence. 
\B{Pam\-tse\-ol ngop ay\-re\-nut | Mì ron\-sem\-ä tì\-fnu}, Music creates patterns | In the silence of the mind.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Oel vo\-lin tì\-fnut.\slash Oe ätxä\-le so\-li tsnì li\-vu tì\-fnu.} I ask for silence.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/comment-page-1/\#comment-1060}{b}
(\ding{43}~\HL{fnu}{\B{fnu}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIfnunga'}{\B{tì\-fnu\-nga'}}, quiet.

% tìfnunga'
\hi
\HT{tIfnunga'}{\B{tì\cp{fnu}nga'}} \I{[tI."fnu.NaP]} \emph{adj.} quiet.
\B{New oe ri\-vun a\-sim tì\-fnu\-nga'\-a tseng\-it a tsa\-ro tsun syi\-vor tsi\-vu\-rokx fte spä\-pi\-veng.} I want to find a quiet place nearby where I can chill out and rest to get my head back on straight.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{tIfnu}{\B{tì\-fnu}}, \HL{-nga'}{\B{-nga'}})

% tìfpxamo
\hi 
\HT{tIfpxamo}{\B{tì\cp{fpxa}mo}} \I{[tI."fp'a.mo]} \emph{n.} horror. 
(\ding{43}~\HL{fpxamo}{\B{fpxa\-mo}})

% tìftang
\hi
\HT{tIftang}{\B{tì\cp{ftang}}} \I{[tI."ftaN]} \textsc{i}. \emph{n.} (\emph{a.}) stopping.
(\emph{b. linguistic}) glottal stop, the letter \B{'}, the sound \I{[P]}. 
\B{Ro sngä\-'i\-tseng tsa\-lì\-'u\-ä alu 'ey\-lan lu tì\-ftang.} At the beginning of the word \emph{'eylan} there's a glottal stop.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\
\textsc{ii}. \B{tì\cp{ftang} si} \I{[tI."ftaN si]} \emph{vin.} stop (s.o.\slash s.t.).   
\B{Ul\-te kaw\-tu ke tsun ay\-oer tì\-ftang si\-vi.} And no one can stop us.~\LINK{http://forum.learnnavi.org/language-updates/stop!/}{ln} 
\B{Oe po\-ru tì\-ftang so\-li.} I stopped him.~\LINK{http://forum.learnnavi.org/language-updates/stop!/}{ln}
(\ding{43}~\HL{ftang}{\B{ftang}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lukftang}{\B{luk\-ftang}}, constant, continual.

% tìftanglen
\hi 
\HT{tIftanglen}{\B{tìftang\cp{len}}} \I{[tI.ftaN."lEn]} \emph{n.} prevention. 
\B{Tì\-kan la\-'a\-yä a\-yll lu tì\-ftang\-len tì\-spxin\-ä.} The goal of social distancing is the prevention of disease.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b} 
(\ding{43}~\HL{ftanglen}{\B{ftang\-len}})

% tìftia
\hi
\HT{tIftia}{\B{tìfti\cp{a}}} \I{[tI.fti."a]} \emph{n.} study. 
(i.) \B{tì\-ftia ki\-fkey\-ä}, science [lit.: the study of the physical world]. 
(\ding{43}~\HL{ftia}{\B{ftia}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIftiatu}{\B{tì\-fti\-a\-tu}}, researcher, scientist.

% tìftiatu
\hi
\HT{tIftiatu}{\B{tìfti\cp{a}tu}} \I{[tI.fti."a.tu]} \emph{n.} researcher, scientist. 
(i.) \B{tì\-fti\-a\-tu ki\-fkey\-ä}, sientist. 
\B{Saw\-tu\-te\-ri sì\-fti\-a\-tu ki\-fkey\-ä var fmi\-vi Ey\-we\-veng\-it tsli\-vam, slä kaw\-krr ke fla\-yä.} The scientists among the Sky People keep trying to understand Pandora, but they will never succeed.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{tIftia}{\B{tì\-ftia}})

% tìftxavang
\hi
\HT{tIftxavang}{\B{tì\cp{ftxa}vang}} \I{[tI."ft'a.vaN]} \emph{n.} passion.
\B{Lu fra\-'u te\-ri tì\-ftxa\-vang.} Everything is about passion.~\LINK{http://forum.learnnavi.org/news-announcements/paul-frommer-at-grid-2010/}{ln}
(\ding{43}~\HL{ftxavang}{\B{ftxa\-vang}})

% tìftxey
\hi
\HT{tIftxey}{\B{tì\cp{ftxey}}} \I{[tI."ft'Ej]} \emph{n.} choice, option.
\B{Lu nge\-yä tì\-ftxey!} It's your choice!~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{Fì\-txe\-le\-ri tì\-nui ke lu tì\-ftxey.} In this matter, failure is not an option.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{ftxey}{\B{ftxey}})

% tìftxulì'u
\hi
\HT{tIftxulI'u}{\B{tìftxu\cp{lì}'u}} \I{[tI.ft'u."lI.Pu]} \emph{n.} speech\hyp{}making, public speaking.   
\B{Oe\-ri lu tì\-ftxu\-lì\-'u ngä\-zìk nì\-txan. Wätx nì\-'aw.} Public speaking is very difficult for me. I'm hopelessly bad at it.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{43}~\HL{ftxulI'u}{\B{ftxu\-lì\-'u}}, \HL{sAftxulI'u}{\B{sä\-ftxu\-lì\-'u}}) 

% tìfyawìntxu
\hi
\HT{tIfyawIntxu}{\B{tìfyawìn\cp{txu}}} \I{[tI.fja.wIn."t'u]} \emph{n.} guidance.
\B{Ul\-te o\-mum oel fu\-ta tì\-fya\-wìn\-txu\-ri oe\-yä pe\-rey ay\-nga nì\-wotx.} And I know you are all waiting for my guidance.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
(\ding{43}~\HL{fyawIntxu}{\B{fya\-wìn\-txu}}) 

% tìfyeyn
\hi
\HT{tIfyeyn}{\B{tì\cp{fyeyn}}} \I{[tI."fjEjn]} \emph{n.} ripeness, maturity, full fruition.
(\ding{43}~\HL{fyeyn}{\B{fyeyn}}) 

% tìhawl
\hi
\HT{tIhawl}{\B{tì\cp{hawl}}} \I{[tI."hawl]} \emph{n.} preparations, plan. 
\B{Tì\-hawl le\-sngä\-'i lu tì\-kang\-kem\-vi skxawng\-ä, slä pum alu fì\-'u yo' nì\-'aw.} The original plan was the work of an idiot, but this one is just perfect.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Hu\-fwa nge\-yä tì\-hawl\-ìri ke lu kea kxey\-ey, tsal\-su\-ngay oe\-ru ke ha' nì\-tam.} Although there’s nothing wrong with your plan, it just doesn't suit me.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Wo\-leyn Ìstawl yey\-fyat mì hll\-te fte oeyk\-ti\-vìng fra\-po\-ru tì\-hawl\-te\-ri sne\-yä.} Ìstaw drew a line on the ground to explain his plan to everyone.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{hawl}{\B{hawl}})

% tìhawnu / si
\hi
\HT{tIhawnu}{\B{tì\cp{haw}nu}} \I{[tI."haw.nu]} \textsc{i}. \emph{n.} protection.  \\
\textsc{ii}. \B{tì\cp{haw}nu si} \I{[tI."haw.nu si]} \emph{vin.} protect.
\B{Oma\-ti\-ka\-ya\-ru tì\-haw\-nu si\-vi.} Protect the people.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Eytukan.27s_Death}{wiki\slash a}
(\ding{43}~\HL{hawnu}{\B{haw\-nu}}) 

% tìhawnuwll
\hi
\HT{tIhawnuwll}{\B{tì\cp{haw}nuwll}} \I{[tI."haw.nu.w\s{l}]} \emph{n.} spartan, ``protection plant'', \emph{kuchenium polyphyllum}. 
(\ding{43}~\HL{tIhawnu}{\B{tì\-haw\-nu}}, \HL{'ewll}{\B{'ewll}})

% tìhan
\hi 
\HT{tIhan}{\B{tì\cp{han}}} \I{[tI."han]} \emph{n.} loss. 
\B{Maw tì\-han sa'\-nok\-ä, Txe\-wì a\-fpawng sar\-mä\-ngi zì\-sì\-to a\-pxay.} Sadly, after the loss of his mother, Txewì grieved for many years.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
\B{Nge\-yä tì\-han\-ìri ke\-ftxo.} Sorry for your loss.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/\#comment-30281}{b}
(\ding{43}~\HL{han}{\B{han}})

% tìhangham
\hi 
\HT{tIhangham}{\B{tì\cp{hang}ham}} \I{[tI."haN.ham]} \emph{n.} laughter. 
\B{Txa\-su\-nu oer fwa stawm nge\-yä tì\-hang\-ham\-it.} I love to hear your laughter.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}  
(\ding{43}~\HL{hangham}{\B{hang\-ham}})

% tìhipx
\hi 
\HT{tIhipx}{\B{tì\cp{hipx}}} \I{[tI."hip']} \emph{n.} control. 
(\ding{43}~\HL{hipx}{\B{hipx}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{snotipx}{\B{sno\-tipx}}, self\hyp{}control. 

% tìhìpey
\hi
\HT{tIhIpey}{\B{tì\cp{hì}pey}} \I{[tI."hI.pEj]} \emph{n.} hesitation.
\B{Tì\-hì\-pey tsun krro krro le\-sar li\-vu.} Hesitation can sometimes be useful.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
\B{Sar tsa\-lì\-'ut a fì\-'u\-ri lu oe\-ru tì\-hì\-pey nì\-'it.} I'm a bit hesitant about using that word.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{hIpey}{\B{hì\-pey}})

% tìhona
\hi
\HT{tIhona}{\B{tì\cp{ho}na}} \I{[tI."ho.na]} \emph{n.} cuteness, adorableness. 
\B{Pe\-yä 'itan\-ìri lu ho\-na nì\-txan a fì\-'u law lu fra\-por. Slä tì\-ho\-na nì\-'aw ke tam.} It's clear to everyone that his son is very cute. But cuteness alone isn't enough.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{hona}{\B{ho\-na}})

% tìhuslew
\hi 
\HT{tIhuslew}{\B{tì\cp{hu}slew}} \I{[tI."hu.slEw]} \emph{n.} accompaniment. 
(\ding{43}~\HL{huslew}{\B{hu\-slew}}) 

% tìk
\hi 
\HT{tIk}{\B{tìk}} \I{[tIk\textcorner]} \textsc{i}. \emph{adv.} immediately, without delay. 
\B{Tsa\-kem si tìk!} Do it immediately!~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} \\
\textsc{ii}. \emph{conj., indicating that a second action immediately follows a first}.
\B{Fì\-ioang ke tsun sli\-ve\-le; nem\-fa pay zup tìk spa\-kat.} This animal cannot swim; if it falls into the water, it immediately drowns.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\emph{a. idiom}) \B{Tse\-'a tìk yaw\-ne.} `Love at first sight.'~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\ding{43}~\HL{pxiye'rIn}{\B{pxi\-ye'\-rìn}})

% tìk'ìn
\hi
\HT{tIk'In}{\B{tìk\cp{'ìn}}} \I{[tIk\textcorner."PIn]} \emph{n.} free time, absence of being busy.  
\B{Tìk\-'ìn\-ìri kem\-pe si nga?} What do you do in your free time?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{'In}{\B{'ìn}}) 

% tìkoak
\hi 
\HT{tIkoak}{\B{tì\cp{ko}ak}} \I{[tI."ko.ak\textcorner]} \emph{n.} old age. 
(\emph{a. proverb}) \B{Nì\-win 'ìp tì\-'e\-wan, nì\-win pä\-hem tì\-ko\-ak.} Youth vanishes quickly, old age arrives quickly.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} 
(\ding{43}~\HL{koak}{\B{ko\-ak}}; \ding{214}~\HL{tI'ewan}{\B{tì\-'e\-wan}}) 

% tìktok
\hi 
\HT{tIktok}{\B{tìk\cp{tok}}} \I{[tIk\textcorner."tok\textcorner]} \emph{n.} absence. 
\B{Tìk\-tok\-it pe\-yä fkol tso\-le\-ri.} His absence was noted.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\emph{a. proverb}) \B{Tì\-tok slu tìk\-tok fa pam\-tsyìp a\-'aw.} Presence becomes absence by means of one little sound. [i.e. `Small things can make a big difference.']~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\ding{43}~\HL{ke}{\B{ke}}, \HL{tok}{\B{tok}}, \ding{214}~\HL{tItok}{\B{tì\-tok}}) 

% tìktseri
\hi
\HT{tIktseri}{\B{tìk\cp{tse}ri}} \I{[tIk\textcorner."\t{ts}E.Ri]} \emph{n.} unawareness, lack of notice.
(\emph{a. proverb}) \B{Tìk\-tse\-ri lu tì\-meyp.} Lack of awareness is (a form of) weakness.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{tseri}{\B{tse\-ri}})

% tìktstun
\hi 
\HT{tIktstun}{\B{tìk\cp{tstun}}} \I{[tIk\textcorner."\t{ts}tun]} \emph{or} \I{[."\t{ts}tUn]} \emph{n.} unkindness, emotional coldness.
(\ding{43}~\HL{ketstun}{\B{ke\-tstun}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIktstunga'}{\B{tìk\-tstu\-nga'}}, unkind, emotional coldness.

% tìktstunga'
\hi 
\HT{tIktstunga'}{\B{tìk\cp{tstu}nga'}} \I{[tIk\textcorner."\t{ts}tu.NaP]} \emph{adj., nfp.} unkind, emotional coldness.
(\ding{43}~\HL{tIktstun}{\B{tìk\-tstun}}, \HL{-nga'}{\B{-nga'}}; \HL{ketstun}{\B{ke\-tstun}}; \ding{214}~\HL{tItstunwinga'}{\B{tì\-tstun\-wi\-nga'}}) 

% tìkahena
\hi 
\HT{tIkahena}{\B{tìka\cp{he}na}} \I{[tI.ka."hE.na]} \emph{n.} transportation (\emph{abstract concept}).  
(\ding{43}~\HL{kahena}{\B{ka\-he\-na}}, \HL{sAkahena}{\B{sä\-ka\-he\-na}})

% tìkakmokri
\hi 
\HT{tIkakmokri}{\B{tìkak\cp{mok}ri}} \I{[tI.kak\textcorner."mok\textcorner.ri]} \emph{n.} muteness. 
(\ding{43}~\HL{kakmokri}{\B{kak\-mok\-ri}})

% tìkakpam
\hi
\HT{tIkakpam}{\B{tìkak\cp{pam}}} \I{[tI.kak\textcorner."pam]} \emph{n.} deafness. 
\B{Tì\-ta\-ron\-ìri lä\-ngu tì\-kak\-pam fe\-kum. Kin fkol fra\-inan\-fyat.} I'm sorry to say that deafness is a disadvantage for hunting. You need all your senses.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
\B{Po\-ri tì\-kak\-rel tì\-kak\-pam\-sì kum tsam\-ä lu.} His blindness and deafness are a result of the war.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{kakpam}{\B{kak\-pam}})

% tìkakrel
\hi
\HT{tIkakrel}{\B{tìkak\cp{rel}}} \I{[tI.kak\textcorner."REl]} \emph{n.} blindness. 
\B{Po\-ri tì\-kak\-rel tì\-kak\-pam\-sì kum tsam\-ä lu.} His blindness and deafness are a result of the war.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{kakrel}{\B{kak\-rel}})

% tìkakzir
\hi 
\HT{tIkakzir}{\B{tìkak\cp{zir}}} \I{[tI.kak\textcorner."zir]} \emph{n.} numbness (\emph{physical and emotional}). 
(\ding{43}~\HL{kakzir}{\B{kak\-zir}})

% tìkan
\hi
\HT{tIkan}{\B{tì\cp{kan}}} \I{[tI."kan]} \emph{n.} aim, goal, purpose, target. 
\B{Pì\-lok\-ä fì\-ha\-pxì\-yä tì\-kan lu law.} The aim of this part of the blog is clear.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Tì\-kan taw\-na\-tep!} (\emph{mil.}) Target lost!~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b\slash g}
\B{Lu pì\-lok\-ur pxe\-sì\-kan sì pxe\-fne\-'upxa\-re.} The blog has three aims and three kinds of messages.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Fu\-ria tsun fko nì\-'i'\-a tì\-kan\-it tsi\-ve\-'a, 'e\-fu oe nit\-ram nì\-txan nì\-'aw.} That one can finally see the aim, I feel very happy.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/comment-page-1/\#comment-2479}{b}
\B{Fì\-tì\-kan\-ìri ropx ke ha'.} For this purpose, a hole going right through the tree trunk isn't suitable.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{a. proverb}) \B{Sä\-fpìl a\-steng, tì\-kan a\-teng.} Great minds think alike [lit.: similar thought, same aim].~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\ding{43}~\HL{kan}{\B{kan}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIkangkem}{\B{tì\-kang\-kem}}, work.
\on{2}.~\HL{nItkan}{\B{nìt\-kan}}, intentionally. 
\on{3}.~\HL{tIkankxan}{\B{tì\-kan\-kxan}}, barrier to one's goals, source of frustration.
\on{4}.~\HL{mu'nitkan}{\B{mu'\-nit\-kan}}, effective.

% tìkankxan
\hi 
\HT{tIkankxan}{\B{tìkan\cp{kxan}}} \I{[tI.kan."k'an]} \emph{n.} barrier to one's goals, source of frustration. 
\B{Oe\-ri mì\-ftxe\-le, fwa pol oe\-ti ke slo\-lan lu tì\-kan\-kxan apxa.} In this regard, his not supporting me was a large barrier to (achieving) my goal.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
(\ding{43}~\HL{tIkan}{\B{tì\-kan}}, \HL{ekxan}{\B{e\-kxan}})

% tìkankxanga'
\hi 
\HT{tIkankxanga'}{\B{tìkan\cp{kxa}nga'}} \I{[tI.kan."k'a.NaP]} \emph{adj.} frustrating. 
\B{Fì\-tì\-fkey\-tok\-ìl a\-tì\-kan\-kxa\-nga' 'ey\-ke\-re\-fu oet lek\-ye\-'ung.} This frustrating situation is making me crazy.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
(\ding{43}~\HL{tIkankxan}{\B{tìkankxan}}, \HL{-nga'}{\B{-nga}}; \HL{lekxan}{\B{le\-kxan}})

% tìkanu
\hi
\HT{tIkanu}{\B{tì\cp{ka}nu}} \I{[tI."ka.nu]} \emph{n.} intelligence.
\B{Nge\-yä ke ftang a tì\-kang\-kem\-ìri sì tì\-ka\-nu\-ri sì fì\-lì'\-fya\-yä tì\-yawn\-ìri sei\-yi oe ira\-yo nì\-txan.} I thank you very much for your never\hyp{}ending work, intelligence and love for the language.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/comment-page-1/\#comment-1547}{b}
(\ding{43}~\HL{kanu}{\B{ka\-nu}})

% tìkangkem / si
\hi
\HT{tIkangkem}{\B{tì\cp{kang}kem}} \I{[tI."kaN.kEm]} \textsc{i}. \emph{n.} work (\emph{coll. shortened to} \ding{43}~\HL{kangkem}{\B{kang\-kem}}). 
\B{Fay\-sul\-fä\-tu\-ä tì\-kang\-kem ohe\-ru meuia lu\-yu nì\-ngay.} The work of these masters truly honors me.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Pe\-yä tì\-kang\-kem\-ìl mi vey\-krr\-ei\-yìn pot nì\-wotx.} His work is still completely overwhelming him (and I'm glad).~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Tì\-kang\-kem\-ìri var\-mrr\-ìn oe nì\-wotx.} I was completely swamped (overwhelmed) at work.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Poel tì\-kang\-kem\-it sngey\-ko\-lä\-'i.} She began the work.~\LINK{http://naviteri.org/2010/06/first-post/}{b}  \\
\textsc{ii}. \B{tì\cp{kang}kem si} \I{[tI."kaN.kEm si]} \emph{vin.} work.
\B{Slä nì\-aw\-no\-mum, ze\-ne oe 'aw\-si\-teng tì\-kang\-kem si\-vi fo\-hu.} But as you know, I must work together with them.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Tì\-kang\-kem si val!} Work hard!~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}
(\emph{with the topic}) work at\slash on s.t. \B{Fì\-tì\-kang\-kem\-vi\-ri le\-tsran\-ten ke new oe hu Ra\-lu tì\-kang\-kem si\-vi.} I don't want to work with Ralu on this important project.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{tIkan}{\B{tì\-kan}}, \HL{kem}{\B{kem}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIkangkemvi}{\B{tì\-kang\-kem\-vi}}, project.

% tìkangkemvi
\hi
\HT{tIkangkemvi}{\B{tì\cp{kang}kemvi}} \I{[tI."kaN.kEm.vi]} \emph{n.} project, piece of work.
\B{Hu\-fwa ro\-lun oel 'a\-'aw\-a kxey\-ey\-ti, fì\-tì\-kang\-kem\-vi lu txan\-tsan ka wotx.} Although I found a few errors, this piece of work is generally excellent.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Fwa oel fì\-tì\-kang\-kem\-vit sngey\-ki\-vä\-'i krr\-no\-lekx nì\-txan.} Me starting this project took a long time.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Nì\-'i'\-a po tsa\-tì\-kang\-kem\-vir ha\-sey so\-li.} She finally completed the project.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} 
\B{Fì\-tì\-kang\-kem\-vi\-ri le\-tsran\-ten ke new oe hu Ra\-lu tì\-kang\-kem si\-vi.} I don't want to work with Ralu on this important project.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Fì\-tì\-kang\-kem\-vi\-ri oel kin 'aw\-a trr\-ti nì\-'aw.} For this project I only need one day.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
(\ding{43}~\HL{tIkangkem}{\B{tì\-kang\-kem}})

% tìkara
\hi 
\HT{tIkara}{\B{tìka\cp{ra}}} \I{[tI.ka."Ra]} \emph{n.} resistance. 
(\emph{a. idiom}) \B{Tì\-ka\-ra lu ä\-txä\-le pa\-lu\-kan\-ur.} Resistance is futile [lit.: `reistance is a request to the thanator'].~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(\ding{43}~\HL{kara}{\B{ka\-ra}})

% tìkatìng
\hi 
\HT{tIkatIng}{\B{tì\cp{ka}tìng}} \I{[tI."ka.tIN]} \emph{n.} distribution.  
(\ding{43}~\HL{katIng}{\B{ka\-tìng}})

% tìkawng
\hi
\HT{tIkawng}{\B{tì\cp{kawng}}} \I{[tI."kawN]} \emph{n.} evil. 
\B{Eo ay\-oeng lu txan\-a tì\-kawng.} A great evil is upon us.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake_back_at_hometree}{wiki\slash a}
(\ding{43}~\HL{kawng}{\B{kawng}})

% tìkeftxo
\hi
\HT{tIkeftxo}{\B{tìke\cp{ftxo}}} \I{[tI.kE."ft'o]} \emph{n.} sadness.
(\ding{43}~\HL{keftxo}{\B{ke\-ftxo}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIkeftxonga'}{\B{tì\-ke\-ftxo\-nga'}}, sad (\emph{nfp.}).

% tìkeftxonga'
\hi
\HT{tIkeftxonga'}{\B{tìke\cp{ftxo}nga'}} \I{[tI.kE."ft'o.NaP]} \emph{adj., nfp.} sad.
\B{Txo tì\-ke\-ftxo\-nga'\-a fmawn\-it ke stil\-vawm ay\-ngal, ze\-ne oe pi\-veng san Unil\-tì\-ran\-tokx\-ä pam\-tseo\-ngop\-yu a\-yaw\-ne to\-ler\-kä\-ngup.} If you haven't heard the sad news, I have to inform that the dear music creator of \emph{Avatar} has died.~\LINK{http://naviteri.org/2015/06/na-ayskxe-mi-telan-sad-news/}{b}
(\ding{43}~\HL{tIkeftxo}{\B{tì\-ke\-ftxo}})

% tìkelpxìmrun
\hi 
\HT{tIkelpxImrun}{\B{tìkel\cp{pxìm}run}} \I{[tI.kEl."p'Im.Run]} \emph{or} \I{[.RUn]} \emph{n.} rarity, rareness. 
\B{Tì\-kel\-pxìm\-run\-it na\-fì\-'u\-a mok\-ri\-yä ke tsun fko wä\-ti\-ve.} The rarity of such a voice can't be disputed.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b}  
(\ding{43}~\HL{kelpxImrun}{\B{kel\-pxìm\-run}}; \HL{sAkelpxImrun}{\B{sä\-kel\-pxìm\-run}}) 

% tìkelu
\hi
\HT{tIkelu}{\B{tì\cp{ke}lu}} \I{[tI."kE.lu]} \emph{n.} lack. 
\B{Tì\-ke\-lu tì\-eyk\-tan\-ä a\-sìl\-tsan lä\-ngu mì olo' aw\-nge\-yä tì\-ngä\-zìk.} Unfortunately, the lack of good leadership in our clan is a problem.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tompakel}{\B{tom\-pa\-kel}}, drought.
\on{2}.~\HL{syuvekel}{\B{syu\-ve\-kel}}, famine.

% tìkemwiä
\hi 
\HT{tIkemwiA}{\B{tìkem\cp{wi}ä}} \I{[tI.kEm."wi.\ae]} \emph{n.} unfairness, injustice. 
(\ding{43}~\HL{kemwiA}{\B{kem\-wi\-ä}})

% tìkenong
\hi
\HT{tIkenong}{\B{tì\cp{ke}nong}} \I{[tI."kE.noN]} \emph{n.} example.   
\B{Fì\-po\-ti oel tspì\-yang, fte tì\-ke\-nong li\-ye\-vu ay\-la\-ru.} I will kill this one as a lesson to the others.~\LINK{http://forum.learnnavi.org/language-updates/capturing-avatar/}{ln\slash a\textsuperscript{c}}
\B{Su\-nu oer fay\-sì\-ke\-nong nge\-yä.} I like these examples of yours.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/comment-page-1/\#comment-7697}{b}
\B{Pol ngop fra\-krr sì\-ke\-nong\-it a hì\-no lu nì\-hawng.} He always creates excessively detailed examples.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Tsa\-tì\-ke\-nong\-ìri oel kxey\-eyt zo\-sley\-ko\-lu.} As for that example, I got the mistake fixed.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1903}{b}
(\ding{43}~\HL{kenong}{\B{ke\-nong}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{natkenong}{\B{nat\-ke\-nong}}, for example.

% tìketeng
\hi
\HT{tIketeng}{\B{tì\cp{ke}teng}} \I{[tI."kE.tEN]} \emph{n.} difference.
\B{Tì\-ke\-teng lu 'u\-pe?} What's the difference?~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/comment-page-1/\#comment-9347}{b}
\B{Ral\-ìri ke 'efu oel kea tì\-ke\-teng\-it.} I don't feel any difference concerning the meaning.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1903}{b}
(\ding{43}~\HL{keteng}{\B{ke\-teng}})

% tìkezin (si)
\hi 
\HT{tIkezin}{\B{tì\cp{ke}zin}} \I{[tI."kE.zin]} \textsc{i}. \emph{n.} (\emph{a.}) s.t. in an untangled state, ``untanglement''. (\emph{b.}) solution. 
\B{Tsa\-tì\-ngä\-zìk\-ìri tì\-ke\-zin lu fyin.} The solution to that problem is simple.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b}  
(\ding{43}~\HL{kezin}{\B{ke\-zin}}) \\
\textsc{ii}. \B{tì\cp{ke}zin si} \I{[tI."kE.zin si]} \emph{vin.} (\emph{a.}) untangle. (\emph{b.}) solve. 
\B{Sra\-ke tsun nga fì\-ing\-yen\-tsim\-ur tì\-ke\-zin si\-vi?} Can you solve this riddle?~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} 
\B{Ze\-ne aw\-nga tì\-ngä\-zìk\-ur sno\-ni\-vi\-yä tì\-ke\-zin si\-vi nì\-win.} We need to solve the hammock problem quickly.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/\#comment-33483}{b}
(\ding{43}~\HL{kezin}{\B{ke\-zin}})

% tìkin
\hi
\HT{tIkin}{\B{tì\cp{kin}}} \I{[tI."kin]} \emph{n.} need (\emph{can refer either to the general state or concept or to a specific instance}). 
\B{'On\-ìri tsko\-ä lu tì\-kin sä\-la\-tem\-ä a\-hì\-'i.} The form of the bow requires a small change.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\emph{a. with the dative to express}) (have the) need to (do s.t.) \B{Po\-ri aw\-nga\-ru lu tì\-kin a nu\-me nì\-'ul.} We need to learn more about him.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{wiki\slash a}  
(\emph{b. conv. phrase for responding to thanks}) \B{Kea tì\-kin.} ``You're welcome.'' [lit.: (there is) no need]
(\ding{43}~\HL{kin}{\B{kin}})

% tìkxal
\hi 
\HT{tIkxal}{\B{tì\cp{kxal}}} \I{[tI."k'al]} \emph{n.} strictness, discipline. 
(\emph{a. proverb}) \B{Tì\-kxal\-ìl tì\-kxäl\-it zey\-ko.} Discipline cures apathy.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}  
(\ding{43}~\HL{kxal}{\B{kxal}}) 

% tìkxäl 
\hi 
\HT{tIkxAl}{\B{tì\cp{kxäl}}} \I{[tI."k'\ae{}l]} \emph{n.} apathy. 
\B{Tì\-kxäl lu pe\-u? Lu tskxe\-pay a mì te'\-lan sì kxitx a mì el\-tu.} What is apathy? It's ice in the heart and a death in the brain.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} 
(\emph{a. proverb}) \B{Tì\-kxal\-ìl tì\-kxäl\-it zey\-ko.} Discipline cures apathy.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}  
(\ding{43}~\HL{kxAl}{\B{kxäl}}) 

% tìkxey / si
\hi
\HT{tIkxey}{\B{tì\cp{kxey}}} \I{[tI."k'Ej]} \textsc{i}. \emph{n.} incorrectness, mistakeness.
\B{Tì\-kxey zo\-slo\-lu.} The incorrectness has been fixed.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/comment-page-1/\#comment-2585}{b}
(\emph{a. idiom}) \B{Nga\-ru tì\-kxey.} You're wrong.
\B{Nì\-ngay lu ngar tì\-yawr, lu ngar tì\-kxey.} Truly, you are both right and wrong.~\LINK{http://naviteri.org/2010/07/vocabulary-update/comment-page-1/\#comment-95}{b} 
\B{Po ke tsun pi\-vll\-ngay san oe\-ru tì\-kxey.} He can't admit he's wrong.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} \\
\textsc{ii}. \B{tì\cp{kxey} si} \I{[tI."k'Ej si]} \emph{vin.} mess up, foul, do wrong, make a mistake.
\B{Krr\-o krr\-o fì\-tse\-nge oe tì\-kxey sa\-yi.} From time to time I will mess up here.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Ngay\-txoa, krro tì\-kxey si keng kar\-yu.} Sorry, sometimes even the teacher messes up.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{214}~\HL{tIyawr}{\B{tì\-yawr}}, \ding{43}~\HL{kxeyey}{\B{kxey\-ey}})

% tìkxìm
\hi
\HT{tIkxIm}{\B{tì\cp{kxìm}}} \I{[tI."k'Im]} \textsc{i}. \emph{n.} commanding, ordering, assigning tasks.
\B{Sìl\-tsan\-a eyk\-tan ze\-ne fni\-van tì\-kxìm\-ti.} A good leader must be skilled at assigning tasks.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b} \\
\textsc{ii}. \B{tì\cp{kxìm} si} \I{[tI."k'Im si]} \emph{vin.} be above s.o. in a hierarchy, be s.o.'s superior.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Po tì\-kxìm si oer.} He is above me (in some relevant hierarchy).\slash I am under him.\slash He has authority over me.\slash He is my boss.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Pe\-su tì\-kxìm si ngar?} Who's your boss?~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b} 
(\ding{43}~\HL{kxIm}{\B{kxìm}})

% tìkxuke
\hi
\HT{tIkxuke}{\B{tì\cp{kxu}ke}} \I{[tI."ku.kE]} \emph{n.} safety.
\B{Fmal tì\-kxu\-ket nge\-yä sì Mo\-'a\-ra\-yä.} Keep yourself and Mo\-'ara safe.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b}
(\ding{43}~\HL{kxuke}{\B{kxu\-ke}})

% tìla'um
\hi 
\HT{tIla'um}{\B{tì\cp{la}'um}} \I{[tI."la.Pum]} \emph{or} \I{[.PUm]} (\textsc{rn}: \B{tì\cp{la}ùm} \I{[tI."la.Um]}) \emph{n.} pretence. 
\B{Fu\-ria ke tsun tì\-kang\-kem si\-vi, pe\-yä sä\-spxin lu tì\-la\-'um nì\-'aw.} As for not being able to work, his illness is only a pretence.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b} 
(\ding{43}~\HL{la'um}{\B{la\-'um}})

% tìlätek
\hi 
\HT{tIlAtek}{\B{tì\cp{lä}tek}} \I{[tI."l\ae.tEk\textcorner]} \emph{n.} recognition. 
(\ding{43}~\HL{lAtek}{\B{lä\-tek}}) 

% tìlayl
\hi 
\HT{tIlayl}{\B{tì\cp{layl}}} \I{[tI."lajl]} \emph{n.} innocence.  
(\ding{43}~\HL{layl}{\B{layl}}; \ding{214}~\HL{tItokat}{\B{tì\-to\-kat}}) 

% tìlayro
\hi 
\HT{tIlayro}{\B{tì\cp{lay}ro}} \I{[tI."laj.ro]} \emph{n.} freedom. 
\B{Ay\-su\-tel nì\-wotx new tì\-lay\-rot.} All people want freedom.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} 
(\ding{43}~\HL{layro}{\B{lay\-ro}}) 

% tìlalim
\hi 
\HT{tIlalim}{\B{Tìla\cp{lim}}} \I{[tI.la."lim]} \emph{n., clan name} the Windtrader clan.~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b} 
\B{Way Tì\-la\-lim\-ä}, The Windtrader Song.~\LINK{https://naviteri.org/2026/02/way-tilalima-the-windtrader-song/}{b}

% tìlam
\hi
\HT{tIlam}{\B{tì\cp{lam}}} \I{[tI."lam]} \emph{n.} appearance. 
\B{Ul\-te kxawm ke lu tì\-fkey\-tok nì\-ftxan fe' na tì\-lam.} And maybe the situation is not as bad as it seems.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{lam}{\B{lam}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tatlam}{\B{tat\-lam}}, apparently.

% tìlang
\hi
\HT{tIlang}{\B{tì\cp{lang}}} \I{[tI."laN]} \emph{n.} exploration (general sense). 
\B{Sra\-ne, su\-nu Saw\-tu\-te\-ru tì\-lang, slä ke omum fol teyng\-ta kem\-pe ze\-ne si\-vi maw\-krr\-a 'uo\-ti ro\-lun.} Yes, the Skypeople love exploration, but they don't know what to do once they find something.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{lang}{\B{lang}}, \HL{sAlang}{\B{sä\-lang}})

% tìlapx
\hi 
\HT{tIlapx}{\B{tì\cp{lapx}}} \I{[tI."lap']} \emph{n.} regret. 
\B{Tsa\-tì\-pe\-'un\-ìri ke lu oe\-ru ke\-a tì\-lapx.} I have no reget(s) about that decision.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}  
(\ding{43}~\HL{lapx}{\B{lapx}})

% tìlatem
\hi
\HT{tIlatem}{\B{tì\cp{la}tem}} \I{[tI."la.tEm]} \emph{n.} change (abstract concept).
\B{Pxay\-a su\-te\-ri, tì\-la\-tem lu ngä\-zìk.} For many people, change is difficult.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{latem}{\B{la\-tem}}, \HL{sAlatem}{\B{sä\-la\-tem}})

% tìlen
\hi
\HT{tIlen}{\B{tì\cp{len}}} \I{[tI."lEn]} \emph{n.} event, happening. 
\B{tì\-len a\-fe'}, a bad event.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Maw\-krr\-a fko lie so\-li tì\-len\-ur a\-fpxa\-mo fì\-txan, tì\-rey ke lu teng kaw\-krr.} After experiencing such a terrible event, life is never the same.~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b} 
\B{… fu\-la tsun le\-fkrr\-a sì\-len\-it tswi\-va' hì\-krr (…) oe\-ti 'ey\-ke\-fu nit\-ram nì\-txan.} … that we can forget current events for a short time, makes me very happy.~\LINK{http://naviteri.org/2022/05/results-of-the-tttc/}{b}
(\ding{43}~\HL{len}{\B{len}})

% tìleym
\hi 
\HT{tIleym}{\B{tì\cp{leym}}} \I{[tI."lEjm]} \emph{n.} call. 
\B{Ey\-wal tì\-leym\-it aw\-nge\-yä sto\-lei\-awm!} Eywa has heard our call!~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{43}~\HL{leym}{\B{leym}})

% tìleymfe'
\hi 
\HT{tIleymfe'}{\B{tìleym\cp{fe'}}} \I{[tI.lEjm."fEP]} \emph{n.} complaining (\emph{abstract concept}). 
(\ding{43}~\HL{leymfe'}{\B{leym\-fe'}})

% tìleymkem
\hi 
\HT{tIleymkem}{\B{tìleym\cp{kem}}} \I{[tI.lEjm."kEm]} \emph{n.} protesting, protest (\emph{abstract concept}).
\B{Mì ftaw\-nem\-krr, sì\-leym\-kem a tì\-kem\-wiä\-wä maw kin\-trr a\-'aw to\-ler\-kup; set ke te\-rer\-kup ki 'e\-rul.} In the past, the protests against injustice have died after one week; now they are not dying but rather increase.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/\#comment-31259}{b}
(\ding{43}~\HL{leymkem}{\B{leym\-kem}})

% tìloho
\hi 
\HT{tIloho}{\B{tì\cp{lo}ho}} \I{[tI."lo.ho]} \emph{n.} surprise. 
\B{Epxang\-mì lu 'u\-pe?--Tì\-lo\-ho.} What's in the stone jar?--It's a surprise.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words}{b}
(\ding{43}~\HL{loho}{\B{lo\-ho}})

% tìlor
\hi
\HT{tIlor}{\B{tì\cp{lor}}} \I{[tI."loR]} \emph{n.} beauty.
\B{Nar\-mew oe fo\-ru na'\-rìng\-ä tì\-lor\-it wi\-vìn\-txu, slä ke tsä\-ngun fo tsli\-vam.} I wanted to show them the beauty of the forest, but sadly, they weren't able to understand.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{lor}{\B{lor}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{somtIlor}{\B{som\-tì\-lor}}, popsicle.

% tìm
\hi
\HT{tIm}{\B{tìm}}\textsuperscript{\on{1}} \I{[tIm]} \emph{adj.} low. 
\B{Ta\-lun\-a vo\-spxì\-o a\-mrr ke zo\-lup tom\-pa, lä\-ngu yì kil\-van\-ä tìm nì\-txan.} Because it hasn't rained for five months, the river level is very low.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{214}~\HL{kxayl}{\B{kxayl}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tImyI}{\B{tìm\-yì}}, low level.

% tìm
\hi
\B{tìm}\textsuperscript{\on{2}} : \HL{txIm}{\B{txìm}}. 

% tìmal
\hi
\HT{tImal}{\B{tì\cp{mal}}} \I{[tI."mal]} \emph{n.} trustworthiness.   
\B{Le\-kin lu tì\-txur, lu tì\-tstew. Slä le\-tsran\-ten fra\-to lu tì\-mal.} Strength and courage are necessary. But most important of all is trustworthiness.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{mal}{\B{mal}})

% tìme'em
\hi
\HT{tIme'em}{\B{tì\cp{me}'em}} \I{[tI."mE.PEm]} \emph{n.} harmony (general sense).
\B{Fì\-lì\-'u\-ä ral lu tì\-me\-'em sì tì\-ru\-sey mì hi\-fkey na Nawm\-a Sa'\-nok\-ä ha\-pxì.} The meaning of this word is ``harmony, living in the world as part of the Great Mother.''~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln} 
(\ding{43}~\HL{me'em}{\B{me\-'em}}, \HL{meoauniaea}{\B{me\-o\-a\-u\-ni\-a\-e\-a}})

% tìmeyp
\hi
\HT{tImeyp}{\B{tì\cp{meyp}}} \I{[tI."mEjp\textcorner]} \emph{n.} weakness.   
\B{Kun\-sìp\-ìri txan\-a tì\-meyp lu tsyal a mìn.} The gunship's main weakness is the rotor system.~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln\slash a\textsuperscript{c}}
(\emph{a. proverb}) \B{Tìk\-tse\-ri lu tì\-meyp.} Lack of awareness is (a form of) weakness.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{meyp}{\B{meyp}})

% tìmimu
\hi 
\HT{tImimu}{\B{tì\cp{mi}mu}} \I{[tI."mi.mu]} \emph{n.} development. 
(\ding{43}~\HL{mimu}{\B{mi\-mu}}) 

% tìmll'an
\hi
\HT{tImll'an}{\B{tìmll\cp{'an}}} \I{[tI.m\s{l}."Pan]} \emph{n.} acceptance. 
(\ding{214}~\HL{tItsyAr}{\B{tì\-tsyär}}, \ding{43}~\HL{mll'an}{\B{mll\-'an}})

% tìmok
\hi
\HT{tImok}{\B{tì\cp{mok}}} \I{[tI."mok\textcorner]} \emph{n.} suggestion, the abstract idea of suggesting.
\B{Fì\-txe\-le\-ri tì\-mok ke tam; ze\-ne fko fngi\-vo'.} In this matter, suggesting won't cut it; you need to demand.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
(\ding{43}~\HL{mok}{\B{mok}}, \HL{sAmok}{\B{sä\-mok}})

% tìmu'ni
\hi 
\HT{tImu'ni}{\B{tì\cp{mu'}ni}} \I{[tI."muP.ni]} \emph{or} \I{[."mUP.]} \emph{n.}  achievement, accomplishment.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{mu'ni}{\B{mu'\-ni}}) 

% tìmu'nitkan
\hi 
\HT{tImu'nitkan}{\B{tì\cp{mu'}nitkan}} \I{[tI."muP.nit\textcorner.kan]} \emph{or} \I{[."mUP.]} \emph{n.} effectiveness. 
(\ding{43}~\HL{mu'nitkan}{\B{mu'\-nit\-kan}}) 

% tìmuntxa
\hi
\HT{tImuntxa}{\B{tìmun\cp{txa}}} \I{[tI.mun."t'a]} \emph{or} \I{[.mUn.]} \emph{n.}  mating, Na'vi marriage. 
\B{Zì\-tsal\-trr\-ìri tì\-mun\-txa\-yä ay\-lrr\-tok!} Happy anniversary (of your marriage)!~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b} 
(\ding{43}~\HL{muntxa}{\B{mun\-txa}})

% tìmungwrr / si
\hi
\HT{tImungwrr}{\B{tìmung\cp{wrr}}} \I{[tI.muN."w\s{r}]} \emph{or} \I{[.mUN.]} (\textsc{rn}: \B{tìmùng\cp{wrr}} \I{[tI.mUN."w\s{r}]}) \textsc{i}. \emph{n.} exception.
\B{Fwa taw\-tu\-te slu Na'\-vi\-yä ha\-pxì lo\-lu tì\-mung\-wrr apxa.} It was a great exception for a human to become part of the People.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} \\
\textsc{ii}. \B{tìmung\cp{wrr} si} \I{[tI.muN."w\s{r} si]} \emph{vin.} make an exception.
\B{Ni\-nat tì\-mung\-wrr sìl\-mi fte Ra\-lu tsi\-vun ki\-vä hu tar\-po\-ngu.} Ninat just made an exception so that Ralu can go with the hunting party.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\ding{43}~\HL{mungwrr}{\B{mung\-wrr}})

% tìmweypey
\hi
\HT{tImweypey}{\B{tìm\cp{wey}pey}} \I{[tIm."wEj.pEj]} \emph{n.} patience.
\B{Tsa\-krr\-vay, ay\-nge\-yä tìm\-wey\-pey\-ri ira\-yo sei\-yi oe, ul\-te fì\-trr\-ä ftxo\-zä\-ri, sìl\-pey oe, ay\-nga\-ru prr\-te' li\-vu.} In the meantime, I thank you for your patience, and I hope you enjoy today's celebration.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
(\ding{43}~\HL{maweypey}{\B{ma\-wey\-pey}})

% tìmwiä
\hi
\HT{tImwiA}{\B{tìm\cp{wi}ä}} \I{[tIm."wi.\ae]} \emph{n.} fairness, justice.   
\B{tì\-ngay sì tìm\-wi\-ä}, truth and justice.~\LINK{http://forum.learnnavi.org/language-updates/tirey-lu-muia/}{ln}
\B{Oe\-ri ke lu tìm\-wiä, ha nga\-ti oel ngey\-ka\-syä\-'än nì\-teng.} For me there isn't justice, so I'll make you be miserable as well.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{muiA}{\B{mu\-i\-ä}})

% tìmyì
\hi
\HT{tImyI}{\B{\cp{tìm}yì}} \I{["tIm.jI]} \emph{n.} low level.
\B{Lì'\-fya\-ri tsa\-taw\-tu\-te pe\-yì? -- Tìm\-yì. Pol ke tslam stum ke\-'ut, omum lì\-'ut a\-vol nì\-'aw.} How is that Sky Person's Na'vi? -- Pretty bad. He understands almost nothing and only knows eight words.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{tIm}{\B{tìm}}, \HL{yI}{\B{yì}}, \HL{kxaylyI}{\B{kxayl\-yì}})

% tìn
\hi
\HT{tIn}{\B{tìn}} \I{[tIn]} \emph{n.} activity that keeps one busy. 
\B{Sìn A\-sok, Sìn Zu\-saw\-krr\-ä}, Recent and Upcoming Activities [lit.: recent activities, future's activities].~\LINK{http://naviteri.org/2013/04/sin-asok-sin-zusawkrra-recent-and-upcoming-activities/}{b}
(\ding{43}~\B{'ìn}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tInvi}{\B{tìn\-vi}}, task, errand, step.
\on{2}.~\HL{txintIn}{\B{txin\-tìn}}, primary role in society.

% tìnalsteng
\hi 
\HT{tInalsteng}{\B{tì\cp{nal}steng}} \I{[tI."nal.stEN]} \emph{n.} compassion, empathy. 
(\ding{43}~\HL{nalsteng}{\B{nal\-steng}}) \\
\textsc{Derivatives}: \ding{43}  
\on{1}.~\HL{tInalstenga'}{\B{tì\-nal\-ste\-nga'}}, compassionate (\emph{nfp.}).

% tìnalstenga'
\hi 
\HT{tInalstenga'}{\B{tì\cp{nal}stenga'}} \I{[tI."nal.stE.NaP]} \emph{adj., nfp.} compassionate. 
(\ding{43}~\HL{tInalsteng}{\B{tì\-nal\-steng}}, \HL{-nga'}{\B{-nga'}}; \HL{lenalsteng}{\B{le\-nal\-steng}})

% tìnatxu
\hi 
\HT{tInatxu}{\B{tìna\cp{txu}}} \I{[tI.na."t'u]} \emph{n.} disapproval. 
(\ding{43}~\HL{natxu}{\B{na\-txu}}, \ding{214}~\HL{tIsmaw}{\B{tì\-smaw}})

% tìnawri
\hi
\HT{tInawri}{\B{tì\cp{naw}ri}} \I{[tI."naw.ri]} \emph{n.} talent. 
\B{Tì\-ru\-sol\-ìri ke lu oe\-ru kea tì\-naw\-ri kaw\-'it.} I have absolutely no talent for singing.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 
(\ding{43}~\HL{nawri}{\B{naw\-ri}})

% tìnew
\hi
\HT{tInew}{\B{tì\cp{new}}} \I{[tI."nEw]} \emph{n.} desire (\emph{can refer either to the general state or concept or to a specific instance}).
\B{Tsam\-si\-yu\-ri lu tì\-yo\-ra'\-ä tì\-new le\-kin.} A warrior must have the desire for victory.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Lu oer tì\-new a tse\-'a txam\-pay\-it.} I have a desire to see the ocean.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Pxìm lu tì\-new le\-hawng kxu\-tu fpom\-ä.} Excessive desire is often the enemy of peace.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{new}{\B{new}})

% tìnitram
\hi 
\HT{tInitram}{\B{tìnit\cp{ram}}} \I{[tI.nit\textcorner."Ram]} \emph{n.} happiness (\emph{usually resulting from a particular situation or event}). 
\B{Lu oe\-ru tì\-nit\-ram.} I feel happy.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{nitram}{\B{nit\-ram}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tInitramnga'}{\B{tì\-nit\-ram\-nga'}}, happy (nfp.).

% tìnitramnga'
\hi 
\HT{tInitramnga'}{\B{tìnit\cp{ram}nga'}} \I{[tI.nit\textcorner."Ram.NaP]} \emph{or coll.} \I{[tI.nit\textcorner."Ra.NaP]} \emph{adj., nfp.} happy. 
\B{tì\-len a\-tì\-nit\-ram\-nga'}, happy event.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/\#comment-50634}{b} 
(\ding{43}~\HL{tInitram}{\B{tì\-nit\-ram}}, \HL{nitram}{\B{nit\-ram}})

% tìno
\hi
\HT{tIno}{\B{tì\cp{no}}} \I{[tI."no]} \emph{n.} thoroughness, attention to detail. 
\B{Lä\-ngu tì\-kang\-kem fe\-yä lu\-ke tì\-no.} Unfortunately there is no attention to detail in their work.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}

% tìnomum
\hi
\HT{tInomum}{\B{tì\cp{no}mum}} \I{[tI."no.mum]} \emph{or} \I{[.mUm]} (\textsc{rn}: \B{tì\cp{no}mùm} \I{[tI."no.mUm]}) \emph{n.} curiosity.
(\ding{43}~\HL{newomum}{\B{new\-o\-mum}})

% tìnui
\hi
\HT{tInui}{\B{tì\cp{nu}i}} \I{[tI."nu.i]} \emph{n.} failure.
\B{Fì\-txe\-le\-ri tì\-nui ke lu tì\-ftxey.} In this matter, failure is not an option.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{nui}{\B{nui}}, \HL{sAnui}{\B{sä\-nui}})

% tìnvi / si
\hi 
\HT{tInvi}{\B{\cp{tìn}vi}} \I{["tIn.vi]} \textsc{i}. \emph{n.} task, errand, step (in an instruction). \\
\textsc{ii}. \B{\cp{tìn}vi si} \I{["tIn.vi si]} \emph{vin.} perform a task, run an errand.
\B{Oer txoa li\-vu, ke tsun oe ki\-vä nga\-hu. Ze\-ne pxay\-a tìn\-vi si\-vi.} Sorry, I can't go with you. I have a lot of errands to run.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 
(\ding{43}~\HL{tIn}{\B{tìn}})

% tìng
\hi
\HT{tIng}{\B{tìng}} \I{[tIN]} \emph{vtr.} give (s.t. to s.o.).   
\B{Sìl\-pey oe, la\-yu oe\-ru ye'\-rìn sìl\-tsan\-a fmawn a tsun oe ay\-nga\-ru ti\-vìng.} I hope I will soon have good news to give you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Txep\-ìl tìng le\-fpom\-a sä\-nrr\-ti.} The fire gives a pleasant glow.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Zun ngal tsa\-fne\-syu\-vet tim\-vìng oer, zel li\-vu oe txur fì\-trr.} If you had given me that kind of food, I would be strong today.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
\B{Tì\-oeyk\-tìng\-it oel tì\-syìng mì upxa\-re a\-hay.} I intend to give an explanation in the next message.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/comment-page-1/\#comment-1267}{b}
(\emph{a. with} \B{rì'ìr}) reflect, give reflection. \B{Oe\-ri pay\-ìl tìng rì\-'ìr\-it key\-ä.} My face is reflected in the water.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\emph{b. with} \B{yoa} \emph{to convey trade}) \B{Oel to\-lìng nga\-ru tsngan\-it yoa fkxen.} I traded you meat for vegetables.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b} 
(\emph{c. special uses}) \\
(i.) \B{tìng ftxì} \I{[tIN ft'I]} \emph{vin.} (\texttt{+}\emph{control}) taste. (\ding{43}~\B{may'}, \B{ew\-ku}) \\
(ii.) \B{tìng \cp{mik}yun} \I{[tIm "mik\textcorner.jun]} \emph{vin.} (\texttt{+}\emph{control}) listen.
\B{Tìng mik\-yun fte tsli\-vam.} Listen to understand.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Oe\-ru tìng mik\-yun, ma Ra\-lu.} Listen to me, Ralu.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Ru\-txe ti\-vìng mik\-yun, ma fra\-po.} Your attention, please, everyone!~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Po pa\-mll\-txe a krr, fra\-po tar\-mìng mik\-yun nì\-pxi, ta\-lu\-na mok\-ri lu sä\-tsì\-syì\-tsyìp.} When he spoke, everyone listened intently, because his voice was a tiny whisper.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} (\ding{43}~\B{stawm}, \B{yu\-ne}) \\
(iii.) \B{tìng \cp{na}ri} \I{[tIn "na.Ri]} \emph{vin.} (\texttt{+}\emph{control}) look.
\B{Po\-ru tìng na\-ri.} Look at him.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#With_Direhorse}{wiki\slash a} 
\B{Ru\-txe ti\-vìng na\-ri ne\-kll.} Please, see below.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/comment-page-1/\#comment-847}{b} 
\B{Le\-hrr\-ap lu fwa evi\-tsyìp sle\-le mì hil\-van lu\-ke fwa fyeyn\-tu te\-rìng nari-} It's dangerous for tiny ones to swim in the river without an adult watching.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} (\ding{43}~\B{tse\-'a}, \B{nìn}) \\ 
(iv.) \B{tìng \cp{on}tu} \I{[tIN "{}on.tu]} \emph{vin.} (\texttt{+}\emph{control}) smell. (\ding{43}~\B{he\-fi}, \B{syam}) \\
(v.) \B{tìng tseng} \I{[tIN \t{ts}EN]} \emph{vin.} back down. \\ 
(vi.) \B{tìng lawr} \I{[tIN lawR]} \emph{vin.} sing wordlessly, give out a tune or melody. 
\B{'E\-van\-ìl alo a\-'aw\-ve nì\-'aw\-tu na'\-rìng\-it tar\-mok, ha to\-lìng lawr nì\-fwe\-fwi fte\-ke txo\-pu si\-vi.} The boy was alone in the forest for the first time, so he whistled a tune to calm his fears.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\ 
(vii.) \B{tìng zekwä} \I{[tIN "zEk\textcorner.w\ae]} \emph{vin.} (\texttt{+}\emph{control}) touch (intentionally). 
\B{Rä\-'ä tìng zek\-wä oer!} Don't touch me!~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Lu fì\-vul su\-syang nì\-txan. Txo fko ti\-vìng zek\-wä kxakx.} This branch is very fragile. If you touch it, it'll break.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 
(\ding{43}~\B{zìm}, \B{'am\-pi}) 
(viii.) \B{tìng syawn} \I{[tIN sjawn]} \emph{vin.} bless, give blessing. 
\B{Eywa ti\-vìng syawn ngar, ma 'i\-te.} May Ey\-wa bless you, my daughter.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\ 
(ix.) \B{tìng \cp{rä}zekx}  \I{[tIN "R\ae.zEk']} \emph{vin.} curse (s.o.), wishing s.o. ill.
\B{Oe nga\-ru tìng rä\-zekx, ma kal\-wey\-a\-veng!} I curse you, (OR: Damn you!), you son of a bitch.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
\B{Rä\-zekx ngar!} Damn you! [lit.: ``(I) curse you!'']~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/\#comment-54817}{b} \\ 
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{oeyktIng}{\B{oeyk\-tìng}}, explain.
\on{2}.~\HL{pAnutIng}{\B{pä\-nu\-tìng}}, promise. 
\on{3}.~\HL{stxenutIng}{\B{stxe\-nu\-tìng}}, offer. 
\on{4}.~\HL{teswotIng}{\B{te\-swo\-tìng}}, grant.  
\on{5}.~\HL{yomtIng}{\B{yom\-tìng}}, feed.

% tìnga'lini
\hi 
\HT{tInga'lini}{\B{tìnga'\cp{li}ni}} \I{[tI.NaP."li.ni]} \emph{n.} pregnancy (\emph{for animals}). 
(\ding{43}~\HL{nga'lini}{\B{nga'\-li\-ni}}, \HL{tInga'prrnen}{\B{tì\-nga'\-prr\-nen}}) 

% tìnga'prrnen
\hi 
\HT{tInga'prrnen}{\B{tìnga'\cp{prr}nen}} \I{[tI.NaP."p\s{r}.nEn]} \emph{n.} pregnancy (\emph{for people}). 
(\ding{43}~\HL{nga'prrnen}{\B{nga'\-prr\-nen}}, \HL{tInga'lini}{\B{tì\-nga'\-li\-ni}})

% tìngay
\hi
\HT{tIngay}{\B{tì\cp{ngay}}} \I{[tI."Naj]} \emph{n.} truth.   
\B{tì\-ngay sì tìm\-wi\-ä}, truth and justice.~\LINK{http://forum.learnnavi.org/language-updates/tirey-lu-muia/}{ln}
\B{Oe\-ri tì\-ngay\-ìl txe'\-lan\-it ti\-va\-kuk.} Let the truth strike my heart.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b} 
\B{Lo\-lu ka\-vuk, slä Tse\-nul tì\-ngay\-it ko\-lu\-lat.} There was treachery, but Tsenu revealed the truth.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{ngay}{\B{ngay}})

% tìngäzìk
\hi
\HT{tIngAzIk}{\B{tì\cp{ngä}zìk}} \I{[tI."N\ae.zIk\textcorner]} \emph{n.} difficulty, problem.
\B{Lu oe\-ru … ìì … tì\-ngä\-zìk a\-hì\-'i.} I have … um … a slight problem.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}
\B{'Ul tu\-te, 'ul tì\-ngä\-zìk.} The more people, the more problems.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Pol zo\-let oe\-yä sä\-fpìl\-it pxel pum a tì\-ngä\-zìk\-it ngop.} He treated my idea as though it created a problem.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Ngal ke tslä\-ngam teyng\-ta fì\-tì\-ngä\-zìk\-ìri pe\-fyin\-ep\-'ang.} Unfortunately you don't understand how complex this problem is.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} 
(\ding{43}~\HL{ngAzIk}{\B{ngä\-zìk}})

% tìngong
\hi
\HT{tIngong}{\B{tì\cp{ngong}}} \I{[tI."NoN]} \emph{n.} lethargy, laziness.
\B{Fwa Ìstaw\-hu 'aw\-si\-teng tì\-kang\-kem si ke su\-nu oer; tì\-ngong\-ìri ke lu kaw\-tu na po.} I don't like to work with Ìstaw; he's famous for his laziness.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{ngong}{\B{ngong}})

% tìngop
\hi
\HT{tIngop}{\B{tì\cp{ngop}}} \I{[tI."Nop\textcorner]} \emph{n.} creation.
\B{Ftxo\-zä\-ri fì\-tì\-ngop a\-lor lu swey\-a stxe\-li a to\-lel oel.} This beautiful creation is the best present I have received for (my birthday) celebration.~\LINK{http://forum.learnnavi.org/community-calendar/karyu-pawls-birthday/msg312733/\#msg312733}{ln}
(\ding{43}~\HL{ngop}{\B{ngop}}, \HL{sAngop}{\B{sä\-ngop}})

% tìoeyktìng
\hi
\HT{tIoeyktIng}{\B{tìo\cp{eyk}tìng}} \I{[tI.o."{}Ejk\textcorner.tIN]} \emph{or} \I{[tI."wEjk\textcorner.tIN]} \emph{n.} explanation. 
\B{Sìl\-pey oe tsnì fì\-tì\-oeyk\-tìng law li\-vu nga\-ru set.} I hope that this explanation is clear to you now.~\LINK{http://forum.learnnavi.org/language-updates/a-long-answer-to-a-quick-question/}{ln}
\B{Tì\-oeyk\-tìng za\-'u ye'\-rìn.} An explanation comes soon.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/comment-page-1/\#comment-520}{b}
\B{Ul\-te sìl\-pey oe, fay\-sì\-oeyk\-tìng li\-vu law nì\-wotx!} And I hope, all these explanations are clear.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Ru\-txe oe\-yä tì\-oeyk\-tìng\-it a\-saw\-nung ni\-vìn.} Please look at my added explanation.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/comment-page-1/\#comment-581}{b}
\B{Fì\-tì\-oeyk\-tìng\-ìri ira\-yo, ma tsmuk.} Thanks for this explanation, brother.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/comment-page-1/\#comment-10675}{b}
(\ding{43}~\HL{oeyktIng}{\B{oeyk\-tìng}})

% tìohakx
\hi
\HT{tIohakx}{\B{tìo\cp{hakx}}} \I{[tI.o."hak']} \emph{n.} hunger. 
(\ding{43}~\HL{ohakx}{\B{o\-hakx}})

% tìomum
\hi
\HT{tIomum}{\B{tì\cp{o}mum}} \I{[tI."{}o.mum]} \emph{or} \I{[.mUm]} (\textsc{rn}: \B{tì\cp{o}mùm} \I{[tI."{}o.mUm]}) \emph{n.} knowledge.
(\emph{idiom}) \B{tì\-o\-mum\-mì oe\-yä}, to my knowledge, as far as I know.
\B{Tì\-o\-mum\-mì oe\-yä, pol na'\-rìng\-it i\-nan nìl\-tsan.} As far as I know, he reads the forest well.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
(\ding{43}~\HL{omum}{\B{omum}}, \HL{sAomum}{\B{sä\-omum}}) 

% tìpalang
\hi 
\HT{tIpalang}{\B{tì\cp{pa}lang}} \I{[tI."pa.laN]} \emph{n.} (social) contact. 
\B{Po\-hu ke lu oe\-ru kea tì\-pa\-lang kaw\-'it.} I have no contact with him whatsoever.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} 
(\ding{43}~\HL{palang}{\B{pa\-lang}})

% tìpaw
\hi 
\HT{tIpaw}{\B{tì\cp{paw}}} \I{[tI."paw]} \emph{n.} growth. 
(\ding{43}~\HL{paw}{\B{paw}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zIskrrmipaw}{\B{zì\-skrr\-mi\-paw}}, spring (season).

% tìpawm
\hi
\HT{tIpawm}{\B{tì\cp{pawm}}} \I{[tI."pawm]} \emph{n.} question.   
\B{tì\-pawm a\-syen}, last question.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} 
\B{tì\-pawm a\-tor}, final question.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{Fra\-krr lu nge\-yä sì\-pawm ngä\-zìk slä\-kop le\-tsran\-ten.} Your questions are always difficult but also important.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Me\-sì\-pawm a\-sìl\-tsan nì\-txan nang!} Two very good questions!~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Fol oe\-yä tì\-pawm\-it ke sto\-lä\-ngawm.} Unfortunately they didn't hear my question.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Sä\-pll\-txel kar\-yu\-ä ke to\-la\-rä\-nge tì\-pawm\-it kaw\-'it.} The teacher's statement in no way pertained to the question. Drat!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Nge\-yä txan\-tsan\-a tì\-pawm\-ìri nga\-ru sei\-yi oe ira\-yo.} I thank you for your excellent question.~\LINK{http://forum.learnnavi.org/language-updates/another-email-from-frommer/}{ln}
(\ding{43}~\HL{pawm}{\B{pawm}})

% tìpähem
\hi
\HT{tIpAhem}{\B{tì\cp{pä}hem}} \I{[tI."p\ae.hEm]} \emph{n.} arrival.
\B{tì\-pä\-hem le\-ye'\-krr}, an early arrival.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Fì\-tseo\-ri ke tsun kaw\-tu pi\-vä\-hem tì\-yo'\-ne; tsran\-ten tì\-pä\-hem\-ä tì\-fmi nì\-'aw.} In this art it's impossible to arrive at perfection; the only thing that matters is the attempt to arrive there.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{pAhem}{\B{pä\-hem}})

% tìpängkxo
\hi
\HT{tIpAngkxo}{\B{tìpäng\cp{kxo}}} \I{[tI.p\ae{}N."k'o]} \emph{n.} conversation, discussion; dialog.  
\B{tì\-päng\-kxo le\-lì'\-fya}, language discussion.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Tsa\-ko\-ak\-te\-ri stawm\-tswo lu fe'. Po\-hu a tì\-päng\-kxo ngä\-zìk lu nì\-txan.} That old woman's hearing is poor. A conversation with her is very difficult.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Ke new oel fu\-ta fì\-tì\-päng\-kxot ay\-nge\-yä hi\-vul\-tstxem, slä tsun mi\-vä\-kxu hì\-krr nì\-'aw srak?.} I don't want to derail your chat, but can I interrupt for just a moment?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{pAngkxo}{\B{päng\-kxo}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIpAngkxotsyIp}{\B{tì\-päng\-kxo\-tsyìp}}, dialog.

% tìpängkxotsyìp
\hi
\HT{tIpAngkxotsyIp}{\B{tìpäng\cp{kxo}tsyìp}} \I{[tI.p\ae{}N."k'o.\t{ts}Ip\textcorner]} \emph{n.} little conversation, dialog.
\B{Sìl\-pey oe, ay\-nga\-ri fì\-tì\-päng\-kxo\-tsyìp el\-tur tì\-txen sil\-vi.} I hope, you have found this dialog interesting.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Ta\-fral ho\-lawl ay\-oel ay\-nga\-fpi 'a\-'aw\-a tì\-päng\-kxo\-tsyìp\-it.} That's why we've prepared several little dialogs for you.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\ding{43}~\HL{tIpAngkxo}{\B{tì\-päng\-kxo}})

% tìpe'ngay
\hi 
\HT{tIpe'ngay}{\B{tìpe'\cp{ngay}}} \I{[tI.pEP."Naj]} \emph{n.} conclusion (\emph{referring to a determination}). \\
\textsc{i}. \B{tì\-pe'\-ngayt wrr\-zä\-rìp} \emph{vtr.} infer.
\B{Ngey ay\-lì\-'u\-ftu wrr\-zä\-rìp oel tì\-pe'\-ngayt a lu ngar yew\-la.} From your words, I infer that you're disappointed.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} \\
\textsc{ii}. \B{tì\-pe'\-ngayt wrr\-zey\-kä\-rìp} \emph{vtr.} imply.
\B{Ngey ay\-lì\-'ul wrr\-zey\-kä\-rìp tì\-pe'\-ngayt a lu ngar yew\-la.} Your words imply that you're disappointed.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(\ding{43}~\HL{pe'ngay}{\B{pe'\-ngay}}) 

% tìpe'un
\hi
\HT{tIpe'un}{\B{tì\cp{pe}'un}} \I{[tI."pE.Pun]} (\textsc{rn}: \B{tì\cp{pe}ùn} \I{[tI."pE.Un]}) \emph{n.} decision. 
\B{Peyä tsa\-tì\-pe\-'un a swey\-lu txo wi\-vem ayoeng Oma\-ti\-ka\-ya\-wä lu fe'.} His decision to fight against the Omaticaya was a bad one.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Tsa\-tì\-pe\-'un\-ìri ke lu oe\-ru ke\-a tì\-lapx.} I have no reget(s) about that decision.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}
(\ding{43}~\HL{pe'un}{\B{pe\-'un}})

% tìprrte'
\hi
\HT{tIprrte'}{\B{tì\cp{prr}te'}} \I{[tI."p\s{r}.tEP]} \emph{n.} pleasure. 
(\emph{a. conv. phrase}) \B{Oe\-ru tì\-prr\-te'.} My pleasure.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/comment-page-1/\#comment-1851}{b}
(\ding{43}~\HL{prrte'}{\B{prr\-te'}})

% tìpsaw
\hi 
\HT{tIpsaw}{\B{tìp\cp{saw}}} \I{[tIp\textcorner."{}saw]} \emph{n.} clumsiness. 
\B{Po\-eyä tìp\-saw\-ìl txo\-pu sley\-ko\-lu ye\-rik\-it ha po hi\-fwo.} Her clumsiness scared the hexapede, so he escaped.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
(\ding{43}~\HL{pIsaw}{\B{pì\-saw}})

% tìpxul
\hi 
\HT{tIpxul}{\B{tì\cp{pxul}}} \I{[tI."p'ul]} \emph{or} \I{[.p'Ul]} \emph{n.} formidableness, imposingness.~\LINK{http://naviteri.org/2020/08/hiia-vur-a-teri-mefalukantsyip-a-little-story-about-two-cats/}{b} 
(\ding{43}~\HL{pxul}{\B{pxul}}) 

% tìra'un
\hi 
\HT{tIra'un}{\B{tì\cp{ra}'un}} \I{[tI."Ra.Pun]} \emph{or} \I{[.PUn]} \emph{n.} surrender, relinquishment. 
\B{Pe\-yä tì\-ra\-'un tì\-eyk\-tan\-ä le\-yew\-la lu nì\-txan.} His surrender of leadership is very disappointing.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{ra'un}{\B{ra\-'un}}) 

% tìralpeng
\hi
\HT{tIralpeng}{\B{tìral\cp{peng}}} \I{[tI.Ral."pEN]} \emph{n.} translation, interpretation.
\B{Spä\-ngaw oel fu\-ta fì\-'upxa\-re\-yä tì\-ral\-peng ke lu eyawr.} Unfortunately, I don't believe the translation of this message is correct.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\ding{43}~\HL{ralpeng}{\B{ral\-peng}})

% tìran
\hi
\HT{tIran}{\B{tì\cp{ran}}} \I{[tI."Ran]} \emph{vin.} (\emph{a.}) walk. 
\B{Ma 'ite, aw\-nge\-yä fya\-'o\-ri ze\-ne nga sä\-nume si\-vi po\-ru fte tsi\-vun pil\-vll\-txe sì ti\-vì\-ran nì\-ay\-oeng.} My daughter, you will teach him our way to speak and walk as we do.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{a\slash wiki}
\B{Sre fwa sìn tskxe\-pay tì\-ran, ze\-ne fko flì\-nutx\-it sti\-ve\-ftxaw.} It's necessary to check the thickness of the ice before walking on it.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Trr\-'ong\-ta Txon\-'ong\-vay po to\-lì\-ran.} He walked from dawn until dusk.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{b. with} \B{nì\-nu}) trip, ``walk failingly''.
\B{Kll\-te lu ekx\-txu. Na\-ri si txo\-ke\-fyaw tì\-ran nì\-nu.} The ground is rough. Be careful or you'll trip.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{c. of the weather}) \B{tì\-ran hu\-fwe}, it's breezy (but pleasantly so).~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b} 
\B{Hu\-fwe\-tsyìp le\-fpom tar\-mì\-ran.} A pleasant little breeze was blowing.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b} 
(\emph{d. rhyming expression}) \B{Tslikx, tì\-ran, tul; | Ftu yì ne yì tsan\-'ul.} \emph{when learning s.t. new, you have to proceed from step to step: baby steps first, then bigger ones} [lit.: `Crawl, walk, run; from level to level get better.']~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{uniltIranyu}{\B{u\-nil\-tì\-ran\-yu}}, dreamwalker. 
\on{2}.~\HL{uniltIrantokx}{\B{u\-nil\-tì\-ran\-tokx}}, avatar, dreamwalker body; (the movie) \emph{Avatar} (\on{2009}). 
\on{3}.~\HL{tIranpam}{\B{tì\-ran\-pam}}, footstep (\emph{sound}).

% tìranpam
\hi 
\HT{tIranpam}{\B{tì\cp{ran}pam}} \I{[tI."Ran.pam]} \emph{n.} footstep (\emph{sound}). 
\B{Oel stawm sì\-ran\-pam\-it! Le\-rok tu\-teo!} I hear footsteps! Someone is coming!~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words}{b} 
(\ding{43}~\HL{tIran}{\B{tì\-ran}}, \HL{pam}{\B{pam}})

% tìranyu
\hi
\HT{tIranyu}{\B{tì\cp{ran}yu}} \I{[tI."Ran.ju]} \emph{n.} walker, one who walks.
(\ding{43}~\HL{tIran}{\B{tì\-ran}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{uniltIranyu}{\B{u\-nil\-tì\-ran\-yu}}, dreamwalker.

% tìrawn
\hi
\HT{tIrawn}{\B{tì\cp{rawn}}} \I{[tI."Rawn]} \emph{n.} replacement, act of replacing.
\B{Po 'e\-fu ngeyn ul\-te kin tì\-rawn\-it nì\-txan.} He is tired and very much needs to be replaced.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{rawn}{\B{rawn}})

% tìrengop
\hi
\HT{tIrengop}{\B{tì\cp{re}ngop}} \I{[tI."RE.Nop\textcorner]} \emph{n.} (abstract) design.
\B{Tì\-re\-ngop\-ìri ioi\-yä lu sem\-pul pe\-yä tsul\-fä\-tu.} Her father is a master designer of ceremonial adornments.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{rengop}{\B{re\-ngop}}, \HL{sArengop}{\B{sä\-re\-ngop}})

% tìrey
\hi
\HT{tIrey}{\B{tì\cp{rey}}} \I{[tI."REj]} \emph{n.} life.  
\B{'Ìhe\-yu sì\-rey\-ä | Sì\-rey le\-Na'\-vi | Oe\-ru te\-ya si.} The spiral of the lives | the lives of the people | Fill me.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki} 
\B{mì oe\-yä le\-trr\-a tì\-rey}, in my daily life.~\LINK{http://forum.learnnavi.org/language-updates/frommerian-email/}{b} 
\B{Me\-nge\-yä tì\-rey\-ìl 'aw\-si\-teng me\-nga\-ru za\-mì\-ye\-vu\-nge txan\-a fpom\-it sì fpom\-tokx\-it.} May your life together bring you both much peace and health.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Oe\-ri Unil\-tì\-ran\-tokx\-ìl tì\-rey\-ti ley\-ko\-la\-te\-i\-em nì\-txan.} \emph{Avatar} has greatly changed my life for the better.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/msg350756/\#msg350756}{ln}
\B{Nawm\-a sa'\-nok tì\-haw\-nu si\-vi | Aw\-nge\-yä tì\-rey\-ur ki\-fkey\-ur\-sì.} May the Great Mother protect | Our life and world.~\LINK{http://forum.learnnavi.org/news-announcements/christmas-song-ninavi-on-youtube/}{ln}
\B{No\-lu\-me oe nì\-txan te\-ri tì\-tsun\-slu tì\-rey\-ä a hi\-fkey\-mì a\-la\-he.} I learned a lot about the possibility of life on other worlds.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
\B{Tsei\-un oe pi\-vll\-txe san oe\-ri lu tì\-rey sìl\-tsan nì\-ngay sìk.} I can happily say, `my life is truly good.'~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/\#comment-31216}{b}
(\ding{43}~\HL{rey}{\B{rey}})

% tìreyn
\hi
\HT{tIreyn}{\B{tì\cp{reyn}}} \I{[tI."REjn]} \emph{n.}, \textsc{lw} train.
\B{Su\-nu Tsyan\-ur tì\-reyn nì\-txan.} John likes trains a lot.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}

% tìrol
\hi
\HT{tIrol}{\B{tì\cp{rol}}} \I{[tI."Rol]} \emph{n.} song.
\B{Ul\-te lu yaw\-ne oer fì\-tì\-rol ay\-nge\-yä nì\-ngay!} And I truly love this song of yours!~\LINK{http://forum.learnnavi.org/community-calendar/karyu-pawls-birthday/msg312733/\#msg312733}{ln}
\B{Nge\-yä tì\-rol\-tsyìp\-ìri, oe\-yä ay\-syu\-lang set mì fay kll\-kxe\-rem.} As for your little song, my flowers are standing in water now.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/comment-page-1/\#comment-296}{b}
(\ding{43}~\HL{rol}{\B{rol}})

% tìronsrel
\hi
\HT{tIronsrel}{\B{tì\cp{ron}srel}} \I{[tI."Ron.sREl]} \emph{n.} imagination.   
\B{Lu po\-ru tì\-ron\-srel a\-txa\-na\-tan.} She has a vivid imagination.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
(\ding{43}~\HL{ronsrel}{\B{ron\-srel}})

% tìrun
\hi
\HT{tIrun}{\B{tì\cp{run}}} \I{[tI."Run]} \emph{or} \I{[."RUn]} \emph{n.} discovery.
\B{Tì\-run\-ìl tey\-ngä lum\-pe po ho\-lum oe\-ti ke\-ftxo 'ey\-ko\-le\-fu.} The discovery of why he left saddened me.~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{run}{\B{run}})

% tìsay
\hi 
\HT{tIsay}{\B{tì\cp{say}}} \I{[tI."{}saj]} \emph{n.} loyalty. 
\B{Tì\-'ey\-lan\-ìri tsran\-ten fra\-to tì\-say.} What matters most in friendship is loyalty.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
(\ding{43}~\HL{say}{\B{say}}) 

% tìsäfpìlyewn
\hi 
\HT{tIsAfpIlyewn}{\B{tìsä\cp{fpìl}yewn}} \I{[tI.s\ae."fpIl.jEwn]} \emph{or coll.}  \I{[\t{ts}\ae."fpIl.jEwn]} \emph{n.} communication. 
\B{Tìsäfpìlyewn ke lu aylì'u nì'aw.} Communication is far more than a matter of words.~\LINK{https://naviteri.org/2026/06/tskxekeng-a-mikyunfpi-pamrel-listening-exercise-text/}{b}
(\ding{43}~\HL{sAfpIlyewn}{\B{sä\-fpìl\-yewn}}) 

% tìska'a
\hi
\HT{tIska'a}{\B{tìska\cp{'a}}} \I{[tI.ska."Pa]} \emph{n.} destruction. 
\B{Tì\-ska\-'a\-ri Kel\-ut\-ral\-ä Na'\-vi wa\-ya\-lew pe\-fya?} With the destruction of Hometree, how will the Na’vi ever move on?~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
(\ding{43}~\HL{ska'a}{\B{ska\-'a}})

% tìsla'tsu
\hi
\HT{tIsla'tsu}{\B{tì\cp{sla'}tsu}} \I{[tI."{}slaP.\t{ts}u]} \emph{n.} description.
(\ding{43}~\HL{sla'tsu}{\B{sla'\-tsu}}) 

% tìslan
\hi
\HT{tIslan}{\B{tì\cp{slan}}} \I{[tI."{}slan]} \emph{n.} support.
\B{Nge\-yä tì\-eyk\-tan\-ìri, tì\-slan\-ìri sì tsran\-ten fra\-to a tì\-'ey\-lan\-ìri a ka ay\-zì\-sìt nì\-wotx, sei\-yi oe ira\-yo nì\-txan.} Thank you so much for your leadership, your support, and most importantly your friendship throughout the years.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{slan}{\B{slan}})

% tìsmaw
\hi 
\HT{tIsmaw}{\B{tì\cp{smaw}}} \I{[tI."smaw]} \emph{n.} approval. 
\B{Moe\-y\"a tì\-mun\-txa\-ri tì\-smaw nge\-y\"a oe\-ru te\-ya si.} Your approval of our marriage fills me (with joy).~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
(\ding{43}~\HL{smaw}{\B{smaw}}, \ding{214}~\HL{tInatxu}{\B{tì\-na\-txu}})

% tìsnaytx
\hi
\HT{tIsnaytx}{\B{tì\cp{snaytx}}} \I{[tI."{}snajt']} \emph{n.} loss (\emph{of a contest, game or fight}). 
\B{Oeyk tsa\-tì\-snaytx\-ä lu apxa sä\-nui tì\-eyk\-tan\-ä.} That loss was caused by a great failure of leadership.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{214}~\HL{tIyora'}{\B{tì\-yo\-ra'}}, \ding{43}~\HL{snaytx}{\B{snaytx}})

% tìso'ha
\hi
\HT{tIso'ha}{\B{tì\cp{so'}ha}} \I{[tI."{}soP.ha]} \emph{n.} enthusiasm; having a good attitude.
(\ding{43}~\HL{so'ha}{\B{so'\-ha}})

% tìsom
\hi
\HT{tIsom}{\B{tì\cp{som}}} \I{[tI."{}som]} \emph{n.} heat.
(\ding{214}~\HL{tIwew}{\B{tì\-wew}}, \ding{43}~\HL{som}{\B{som}})

% tìsop
\hi
\HT{tIsop}{\B{tì\cp{sop}}} \I{[tI."{}sop\textcorner]} \emph{n.} journey. 
\B{Nge\-yä fì\-tì\-sop\-ìri pe\-han\-tseng?} This journey of yours--what's its destination?~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} 
(\ding{43}~\HL{sop}{\B{sop}})

% tìspaw
\hi 
\HT{tIspaw}{\B{tì\cp{spaw}}} \I{[tI."{}spaw]} \emph{n.} belief (\emph{abstract concept}). 
\B{Tsran\-ten tì\-spaw, tsran\-ten nì\-'ul tì\-fkey\-to\-ngay.} Belief is important, but reality is more important.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
(\ding{43}~\HL{spaw}{\B{spaw}}; \HL{sAspaw}{\B{sä\-spaw}})

% tìspe'e
\hi
\HT{tIspe'e}{\B{tìspe\cp{'e}}} \I{[tI.spE."PE]} \emph{n.} capture.
(\ding{43}~\HL{spe'e}{\B{spe\-'e}})

% tìspxin
\hi
\HT{tIspxin}{\B{tì\cp{spxin}}} \I{[tI."{}sp'in]} \emph{n.} sickness, the state of being ill. 
(\ding{43}~\HL{spxin}{\B{spxin}}, \HL{sAspxin}{\B{sä\-spxin}})

% tìsraw / si / seyki
\hi
\HT{tIsraw}{\B{tì\cp{sraw}}} \I{[tI."{}sRaw]} \textsc{i}. \emph{n.} pain.
\B{Tì\-sraw le\-tokx sì tì\-ngu\-sä\-'än pxìm tä\-pa\-re fì\-tsap.} Physical pain and mental suffering are often interrelated.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 
\B{Tsam\-si\-yu ze\-ne tsi\-vun yi\-väkx sne\-yä tì\-sraw\-it.} Warriors must be able to ignore their own pain.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}  \\
\textsc{ii}. \B{tì\cp{sraw} si} \I{[tI."{}sRaw si]} \emph{vin.} hurt, be painful.
\B{Sra\-ne, skxir tì\-sraw si nì\-txan, slä ke nge\-rä\-'än oe kaw\-'it.} Yes, the wound is very painful, but I'm not in the least suffering emotionally.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Oe\-ri skxir a mì syokx tì\-sraw se\-ngi.} The wound on my hand hurts.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} \\
\textsc{iii}. \B{tì\cp{sraw} sey\cp{ki}} \I{[tI."{}sRaw sEj."k$\cdot$i]} \emph{vtr., inf.}\on{22} hurt (s.o.\slash s.t.).
\B{Tì\-'efu\-mì oe\-yä, nge\-yä fì\-sä\-ka\-nom lu le\-hrr\-ap ul\-te tsun nga\-ti tì\-sraw sey\-ki\-vi.} In my opinion, this acquisition of yours is dangerous and can hurt you.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Oe\-ri teng\-krr te\-rul mì na'\-rìng, tsun\-it a\-ski\-en tì\-sraw sey\-ko\-li.} While I was running in the forest, I hurt my right heel.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 
(\ding{43}~\HL{sraw}{\B{sraw}}, \HL{pung}{\B{pung}})

% tìsrese'a
\hi
\HT{tIsrese'a}{\B{tì\cp{sre}se'a}} \I{[tI."{}sRE.sE.Pa]} \emph{n.} prophecy. 
(\ding{43}~\HL{srese'a}{\B{sre\-se\-'a}})

% tìsteftxaw
\hi
\HT{tIsteftxaw}{\B{tìste\cp{ftxaw}}} \I{[tI.stE."ft'aw]} \emph{n.} examination.
\B{Tì\-ste\-ftxaw nge\-yä flo\-lä nìl\-tsan!} Your examination succeeded well!~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
(\ding{43}~\HL{steftxaw}{\B{ste\-ftxaw}})

% tìsti
\hi
\HT{tIsti}{\B{tì\cp{sti}}} \I{[tI."{}sti]} \emph{n.} anger. 
\B{Ney\-ti\-ri\-ri lu yì tì\-sti\-yä kxayl nì\-ngay.} Neytiri is really angry [lit.: Neytiri's level of anger is truly high].~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\ding{43}~\HL{sti}{\B{sti}})

% tìsung
\hi
\HT{tIsung}{\B{tì\cp{sung}}} \I{[tI."{}suN]} (\textsc{rn}: \B{tì\cp{sùng}} \I{[tI."sUN]}) \emph{n.} addition; post\hyp{}scriptum.
(\ding{43}~\HL{sung}{\B{sung}}, \HL{nIsung}{\B{nì\-sung}}) 

% tìswaran
\hi 
\HT{tIswaran}{\B{tì\cp{swa}ran}} \I{[tI."{s}wa.Ran]} \emph{n.} humility, humbleness.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{swaran}{\B{swa\-ran}})

% tìswäper
\hi 
\HT{tIswAper}{\B{tìswä\cp{per}}} \I{[tI.sw\ae."pER]} \emph{n.} self\hyp{}praise. 
(\ding{43}~\HL{sweri}{\B{swe\-ri}})

% tìsweri
\hi 
\HT{tIsweri}{\B{tì\cp{swe}ri}} \I{[tI."{}swE.Ri]} \emph{n.} praise (general sense).
(\ding{43}~\HL{sweri}{\B{swe\-ri}})

% tìsyìmawnun'i
\hi 
\HT{tIsyImawnun'i}{\B{tìsyìmawnun\cp{'i}}} \I{[tI.sjI.maw.nun."Pi]} \emph{or} \I{[.nUn.]} (\textsc{rn}: \B{tìsyìmawnùn\cp{'i}} \I{[tI.SI.maw.nUn."Pi]}) \emph{n.} independence, freedom from a pre\hyp{}existing relationship. (\emph{often shortened and spelled} \ding{43}~\HL{tsyImawnun'i}{\B{tsyì\-maw\-nun\-'i}})
(\ding{43}~\HL{syImawnun'i}{\B{syì\-maw\-nun\-'i}}) 

% tìsyor
\hi
\HT{tIsyor}{\B{tì\cp{syor}}} \I{[tI."{}sjoR]} \emph{n.} relaxation.
\B{Krr\-a fko ta\-ron ke lu kea skxom tì\-syor\-ä.} When one hunts there's no opportunity for relaxation.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{syor}{\B{syor}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIsyortsyIp}{\B{tì\-syor\-tsyìp}}, break, small rest or relaxation.

% tìsyortsyìp
\hi 
\HT{tIsyortsyIp}{\B{tì\cp{syor}tsyìp}} \I{[tI."{}sjoR.\t{ts}jIp\textcorner]} \emph{n.} break, small rest or relaxation. 
\B{Tì\-kang\-kem so\-li oe kawl nì\-txan, 'efu ngeyn, ul\-te kin oel tì\-syor\-tsyìp\-it.} I've worked hard, I'm tired, and I need a break.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
(\ding{43}~\HL{tIsyor}{\B{tì\-syor}}, \HL{-tsyIp}{\B{-tsyìp}})

% tìtaron
\hi
\HT{tItaron}{\B{tì\cp{ta}ron}} \I{[tI."ta.Ron]} \emph{n.} hunting (\emph{abstract}).
\B{Tsko swi\-zaw\-fa a tì\-ta\-ron\-ìri ze\-ne fko ni\-vu\-me.} You must learn to hunt with a bow and arrow.~\LINK{http://forum.learnnavi.org/language-updates/ordinals-nume/}{ln\slash a\textsuperscript{c}}
(\ding{43}~\HL{taron}{\B{ta\-ron}}, \HL{sAtaron}{\B{sä\-ta\-ron}})

% tìtengwotx
\hi 
\HT{tItengwotx}{\B{tì\cp{teng}wotx}} \I{[tI."tEN.wot']} \emph{n.} uniformity. 
\B{Naw\-num\-tseng\-ä ay\-nu\-me\-yu\-ri ley ay\-nga\-ru tì\-teng\-wotx, ay\-oe\-ru ke\-teng\-syon.} Regarding university students, you value uniformity, we value diversity.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b}
(\ding{43}~\HL{tengwotx}{\B{teng\-wotx}}) 

% tìterkup
\hi
\HT{tIterkup}{\B{tì\cp{ter}kup}} \I{[tI."tER.kup\textcorner]} (\textsc{rn}: \B{tì\cp{ter}kùp} \I{[tI."tER.kUp\textcorner]}) \emph{n.} death. 
(\ding{43}~\HL{terkup}{\B{ter\-kup}}, \HL{kxitx}{\B{kxitx}})

% tìtok
\hi 
\HT{tItok}{\B{tì\cp{tok}}} \I{[tI."tok\textcorner]} \emph{n.} presence. 
\B{Tì\-tok\-ìl nge\-yä ngop tì\-ngä\-zìk\-it fra\-po\-ru.} Your presence creates a difficulty for everyone.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\emph{a. proverb}) \B{Tì\-tok slu tìk\-tok fa pam\-tsyìp a\-'aw.} Presence becomes absence by means of one little sound. [i.e. `Small things can make a big difference.']~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\ding{43}~\HL{tok}{\B{tok}}, \ding{214}~\HL{tIktok}{\B{tìktok}}) 

% tìtokat
\hi 
\HT{tItokat}{\B{tì\cp{to}kat}} \I{[tI."to.kat\textcorner]} \emph{n.} guilt. 
(\ding{43}~\HL{tokat}{\B{to\-kat}}; \ding{214}~\HL{tIlayl}{\B{tì\-layl}}) 

% tìtunu
\hi 
\HT{tItunu}{\B{tì\cp{tu}nu}} \I{[tI."tu.nu]} \emph{n.} romance. 
\B{Aw\-nga zen\-ke ti\-vung fu\-ta fì\-tì\-tu\-nu vi\-var.} We must not allow this romance to continue.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} 
(\ding{43}~\HL{tunu}{\B{tu\-nu}})

% tìtxaw
\hi 
\HT{tItxaw}{\B{tì\cp{txaw}}} \I{[tI."t'aw]} \emph{n.} punishment. 
(\ding{43}~\HL{txaw}{\B{txaw}})  

% tìtxanew
\hi
\HT{tItxanew}{\B{tì\cp{txa}new}} \I{[tI."t'a.nEw]} \emph{n.} greed. 
(\ding{43}~\HL{txanew}{\B{txa\-new}})

% tìtxanro'a
\hi 
\HT{tItxanro'a}{\B{tìtxan\cp{ro}'a}} \I{[tI.t'an."Ro.Pa]} \emph{n.} fame, glory. 
\B{Tsa\-tu a new tì\-txan\-ro\-'at nì\-'aw ke sla\-yu eyk\-tan a\-txan\-tsan kaw\-krr.} The person who wants only fame will never become an excellent leader.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b} 
(\ding{43}~\HL{txanro'a}{\B{txan\-ro\-'a}})

% tìtxantslusam
\hi 
\HT{tItxantslusam}{\B{tì\cp{txan}tslusam}} \I{[tI."t'an.\t{ts}lu.sam]} \emph{n.} wisdom (\emph{seldom used}). 
(\ding{43}~\HL{txantslusam}{\B{txan\-tslu\-sam}}, \HL{hafyon}{\B{ha\-fyon}})

% tìtxap
\hi 
\HT{tItxap}{\B{tì\cp{txap}}} \I{[tI."t'ap\textcorner]} \emph{n.} pressure (\emph{physical}). 
(\ding{43}~\HL{txap}{\B{txap}})

% tìtxeym
\hi 
\HT{tItxeym}{\B{tì\cp{txeym}}} \I{[tI."t'Ejm]} \emph{uncountable n.} interest. 
\B{Fe\-yä vur\-ìri lo\-lu txan\-a tì\-txeym.} There was great interest in their story.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{txeym}{\B{txeym}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItxeymtsim}{\B{tì\-txeym\-tsim}}, interest (\emph{countable}).

% tìtxeymtsim
\hi 
\HT{tItxeymtsim}{\B{tì\cp{txeym}tsim}} \I{[tI."t'Ejm.\t{ts}im]} \emph{countable n.} interest. 
\B{Lu oe\-ru pxay\-a tì\-txeym\-tsim.} I have a lot of interests.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/\#comment-50634}{b} 
(\ding{43}~\HL{tItxeym}{\B{tì\-txey}}, \HL{tsim}{\B{tsim}}) 

% tìtxukxefu
\hi 
\HT{tItxukxefu}{\B{tìtxu\cp{kxe}fu}} \I{[tI.t'u."k'E.fu]} \emph{n.} care, concern. 
(\ding{43}~\HL{txukxefu}{\B{txu\-kxe\-fu}})  

% tìtxula
\hi
\HT{tItxula}{\B{tì\cp{txu}la}} \I{[tI."t'u.la]} (\textsc{rn}: \B{tì\cp{txù}la} \I{[tI."dU.la]}) \emph{n.} construction, constructed thing.
(\ding{43}~\HL{txula}{\B{txu\-la}}) 

% tìtxur
\hi
\HT{tItxur}{\B{tì\cp{txur}}} \I{[tI."t'uR]} \emph{n.} strength, power. 
\B{Ut\-ral\-ä A\-nawm | Ay\-ri\-na' lu ay\-oeng, | A pe\-yä tì\-txur mì hi\-nam aw\-nge\-yä.} We are all seeds | Of the Great Tree, | Whose strength is in our legs.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Tì\-'i'\-a\-ri tsam\-ä ze\-ne Saw\-tu\-te vi\-ve\-lek ta\-lun tì\-txur Ey\-wa\-yä.} At the end of the war, the Sky People had to give up due to the power of Eywa.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Tu\-te\-ri tì\-txur lu txew\-nga'.} There are limits to a person's strength.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Ey\-wa\-ru a\-ho, ma 'i\-tan, fte Nawm\-a Sa'\-nok\-ìl tì\-ye\-vìng ngar tì\-txur\-it.} Pray to Eywa, my son, that Great Mother will give you strength.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
(\ding{43}~\HL{txur}{\B{txur}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItxurnga'}{\B{tì\-txur\-nga'}}, powerful.

% tìtxurnga'
\hi
\HT{tItxurnga'}{\B{tì\cp{txur}nga'}} \I{[tI."t'uR.NaP]} \emph{adj., nfp.} powerful. 
\B{Po\-ri ay\-sam\-sä\-'o a\-tì\-txur\-nga' fra\-to mì hi\-fkey la\-yu mì syokx.} He will have the most powerful weapons of war in the world in his hands.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{tItxur}{\B{tì\-txur}})

% tìtxen / si
\hi
\HT{tItxen}{\B{tì\cp{txen}}} \textsc{i}. \emph{n.} awakeness, the state of waking. \\ 
\textsc{ii}. \B{tì\cp{txen} si} \I{[tI."t'En si]} \emph{vin.} to wake, to waken, wake up (\emph{with s.o.\slash s.t. in the dative}).
\B{Tì\-txen si, ma 'ite! Unil sar\-mi nga teng\-krr ze\-rawng. Lu fpom srak?} Wake up, daughter! You were dreaming and screaming. Are you okay?~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Sa'\-nok 'eveng\-ur tì\-txen si sre\-srr\-'ong.} Mother woke the child in the early morning.~\LINK{http://forum.learnnavi.org/language-updates/little-confirmation-titxen-si/}{ln}
(\emph{a. idiom}) \B{el\-tur tì\-txen si}, be interesting, intriguing [lit.: (s.t.) wakes the brain].   
\B{Tsa\-'u el\-tur tì\-txen si.} That's interesting.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{Tsa\-vur oe\-yä el\-tur tì\-txen sol\-i.} That story intrigued me.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{Tsa\-vur fe\-yä ay\-el\-tur tì\-txen so\-li.} They found that story interesting.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln} 
\B{Pxoe\-ri tsa\-vur el\-tur tì\-txen so\-li.} The three of us were interested in that story.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Lu poe tstun\-wi, ul\-te evi\-ru peng el\-tur tì\-txen si a ay\-vur\-it te\-ri 'Rr\-ta, alu tse\-nge a\-stxong nì\-ngay.} She is kind, and tells the kids interesting stories about Earth which is a truly strange place.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{txen}{\B{txen}})

% tìtsan'ul
\hi
\HT{tItsan'ul}{\B{tì\cp{tsan}'ul}} \I{[tI."\t{ts}an.Pul]} \emph{or} \I{[.PUl]} \emph{n.} improvement.
\B{Tskxe\-keng\-luke ke lu kea tì\-tsan\-'ul.} Without practice there is no improvement.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Tì\-tsan\-'ul\-ìri lu nga\-ru ay\-sä\-mok srak?} Do you have any suggestions for improvement?~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\ding{214}~\HL{tIfe'ul}{\B{tì\-fe\-'ul}}, \ding{43}~\HL{tsan'ul}{\B{tsan\-'ul}}, \HL{sAtsan'ul}{\B{sä\-tsan\-'ul}})

% tìtseri
\hi
\HT{tItseri}{\B{tì\cp{tse}ri}} \I{[tI."\t{ts}E.Ri]} \emph{n.} awareness, notice.
(\emph{a. idiom}) \B{to tì\cp{tse}ri} \I{[to tI."\t{ts}E.Ri]} than is apparent, than you are aware of. 
\B{Lu po\-e na nga nì\-'ul to tì\-tse\-ri.} She's more like you than you think (\emph{or:} than you know).~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
(\ding{43}~\HL{tseri}{\B{tse\-ri}})

% tìtslam
\hi
\HT{tItslam}{\B{tì\cp{tslam}}} \I{[tI."\t{ts}lam]} \emph{n.} understanding, intelligence.
\B{Slä tsal\-su\-ngay, txo tì\-tslam for, lu txa\-yo na'\-rìng\-to sìl\-tsan.} But even so, if they're smart they'll take open terrain over bush.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Sìl\-pey oe, nge\-yä tì\-tslam\-ur srung so\-li nì\-'it.} Hopefully, (it) has helped your understanding a bit.~\LINK{http://forum.learnnavi.org/language-updates/minor-confirmation-uvan-si/}{ln}
(\ding{43}~\HL{tslam}{\B{tslam}}) 

% tìtsleng
\hi 
\HT{tItsleng}{\B{tì\cp{tsleng}}} \I{[tI."\t{ts}lEN]} \emph{n.} falsehood, falsity. 
(\ding{43}~\HL{tsleng}{\B{tsleng}}; \HL{fleyul}{\B{fley\-ul}})

% tìtstew
\hi
\HT{tItstew}{\B{tì\cp{tstew}}} \I{[tI."\t{ts}tew]} \emph{n.} courage, bravery.
\B{Le\-kin lu tì\-txur, lu tì\-tstew. Slä le\-tsran\-ten fra\-to lu tì\-mal.} Strength and courage are necessary. But most important of all is trustworthiness.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{tstew}{\B{tstew}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItstewnga'}{\B{tì\-tstew\-nga'}}, couragous, brave.

% tìtstewnga'
\hi
\HT{tItstewnga'}{\B{tì\cp{tstew}nga'}} \I{[tI."\t{ts}tew.NaP]} \emph{adj.} (\emph{of things}) courageous, brave.
\B{Po pol\-txe san tì\-tstew\-nga'\-a stxe\-nu\-ri ira\-yo sei\-yi oe nga\-ru nì\-txan.} She said, ``I thank you very much for your courageous offer.''~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Kem a\-tì\-tstew\-nga' si tu\-te a\-tstew.} A brave person does brave deeds.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{tItstew}{\B{tì\-tstew}}, \HL{tstew}{\B{tstew}})

% tìtsukanom
\hi 
\HT{tItsukanom}{\B{tìtsu\cp{ka}nom}} \I{[tI.\t{ts}u."ka.nom]} (\textsc{rn}: \B{tìtsù\cp{ka}nom} \I{[tI.\t{ts}U."ka.nom]}) \emph{n.} availability. 
(\ding{43}~\HL{tsukanom}{\B{tsu\-ka\-nom}})

% tìtstunwi
\hi
\HT{tItstunwi}{\B{tì\cp{tstun}wi}} \I{[tI."\t{ts}tun.wi]} \emph{or} \I{[."\t{ts}tUn.]} \emph{n.} kindness.
\B{Tì\-tstun\-wi ay\-nge\-yä oe\-ru te\-ya so\-li.} Your kindness have filled me with joy.~\LINK{http://forum.learnnavi.org/community-calendar/karyu-pawls-birthday/msg312733/\#msg312733}{ln}
\B{Pxe\-fe\-yä tì\-tstun\-wi\-ri aw\-nga ira\-yo si\-vi ko!} Let's thank the three of them for their kindness.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\ding{43}~\HL{tstunwi}{\B{tstun\-wi}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItstunwinga'}{\B{tì\-tstun\-wi\-nga'}}, kind.

% tìtstunwinga'
\hi 
\HT{tItstunwinga'}{\B{tì\cp{tstun}winga'}} \I{[tI."\t{ts}tun.wi.NaP]} \emph{or} \I{[."\t{ts}tUn.]} \emph{adj., nfp.} kind. 
\B{ay\-lì\-'u a\-tì\-tstun\-wi\-nga'}, kind words.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
\B{Nge\-yä fay\-lì\-'u a\-tì\-tstun\-wi\-nga' oe\-ru meuia so\-li nì\-ngay.} These kind words of yours have honored me greatly.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
(\ding{43}~\HL{tItstunwi}{\B{tì\-tstun\-wi}}, \HL{-nga'}{\B{-nga'}}; \HL{tstunwi}{\B{tstun\-wi}} \ding{214}~\HL{tIktstunga'}{\B{tìk\-tstu\-nga'}})

% tìtsunslu
\hi
\HT{tItsunslu}{\B{tì\cp{tsun}slu}} \I{[tI."\t{ts}un.slu]} \emph{or} \I{["\t{ts}Un.]} (\textsc{rn}: \B{tì\cp{tsùn}slu} \I{[tI."\t{ts}Un.slu]}) \emph{n.} possibility.
\B{No\-lu\-me oe nì\-txan te\-ri tì\-tsun\-slu tì\-rey\-ä a hi\-fkey\-mì a\-la\-he.} I learned a lot about the possibility of life on other worlds.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{tsunslu}{\B{tsun\-slu}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{kxutslu}{\B{kxu\-tslu}}, risk.

% tìtsyär
\hi
\HT{tItsyAr}{\B{tì\cp{tsyär}}} \I{[tI."\t{ts}j\ae{}R]} \emph{n.} rejection. 
(\ding{214}~\HL{tImll'an}{\B{tì\-mll\-'an}}, \ding{43}~\HL{tsyAr}{\B{tsyär}})

% tìtsyul
\hi 
\HT{tItsyul}{\B{tì\cp{tsyul}}} \I{[tI."\t{ts}jul]} \emph{or} \I{[."\t{ts}jUl]} \emph{n.} beginning, start. 
\B{Sì\-tsyul nì\-wotx lu ngä\-zìk.} All beginnings are difficult.~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b}
(\ding{43}~\HL{tsyul}{\B{tsyul}})

% tìvawm
\hi
\HT{tIvawm}{\B{tì\cp{vawm}}} \I{[tI."vawm]} \emph{n.} darkness. 
\B{Lar\-mu tsa\-tsam\-si\-yu\-hu tì\-vawm a lu stxong ay\-oer.} That warrior carried with him a darkness unknown to us.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{vawm}{\B{vawm}})

% tìväng
\hi
\HT{tIvAng}{\B{tì\cp{väng}}} \I{[tI."v\ae{}N]} \emph{n.} thirst.
\B{Apxa tì\-väng\-ìl po\-ti stey\-ko\-li.} (His) great thirst made him angry.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{vAng}{\B{väng}})

% tìve'kì
\hi
\HT{tIve'kI}{\B{tìve'\cp{kì}}} \I{[tI.vEP."kI]} \emph{n.} hatred. 
(\ding{214}~\HL{tIyawn}{\B{tì\-yawn}}, \ding{43}~\HL{ve'kI}{\B{ve'\-kì}})

% tìvirä
\hi 
\HT{tIvirA}{\B{tìvi\cp{rä}}} \I{[tI.vi."r\ae]} \emph{n.} spread, proliferation. 
\B{Ke\-zem\-pll\-txe, ta\-lun tì\-vi\-rä fì\-sä\-spxin\-ä, lo\-la\-tä\-ngem ki\-fkey, lo\-la\-tä\-ngem tì\-rey.} Needless to say, because of the spread of this illness, the world and life changes.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} 
(\ding{43}~\HL{virA}{\B{virä}})

% tìwan
\hi
\HT{tIwan}{\B{tì\cp{wan}}} \I{[tI."wan]} \emph{n.} obfuscation, cover\hyp{}up.
(\ding{43}~\HL{wan}{\B{wan}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letwan}{\B{let\-wan}}, dodgy, sneaky

% tìwäsul
\hi
\HT{tIwAsul}{\B{tì\cp{wä}sul}} \I{[tI."w\ae.sul]} \emph{or} \I{[.sUl]}  \emph{n.} competition, the abstract idea of competition.
(\ding{43}~\HL{wAsul}{\B{wä\-sul}}, \HL{sAwAsul}{\B{sä\-wä\-sul}})

% tìwäte
\hi
\HT{tIwAte}{\B{tìwä\cp{te}}} \I{[tI.w\ae."tE]} \emph{n.} dispute, argument.
\B{Tì\-wä\-te\-ri ngal oe\-ti pe\-lun ke slan kaw\-krr?} Why don't you ever support me in an argument?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{wAte}{\B{wä\-te}}, \HL{sAwAte}{\B{sä\-wä\-te}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIwAtenga'}{\B{tì\-wä\-te\-nga'}}, controversial.

% tìwätenga'
\hi 
\HT{tIwAtenga'}{\B{tìwä\cp{te}nga'}} \I{[tI.w\ae."tE.NaP]} \emph{adj., nfp.} controversial. 
\B{Fì\-puk\-mì a ay\-sä\-fpìl lu tì\-wä\-te\-nga' nì\-txan.} The ideas in this book are very controversial.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{tIwAte}{\B{tì\-wä\-te}})

% tìwew
\hi
\HT{tIwew}{\B{tì\cp{wew}}} \I{[tI."wEw]} \emph{n.} coldness, cold.
(\ding{214}~\HL{tIsom}{\B{tì\-som}}, \ding{43}~\HL{wew}{\B{wew}})

% tìwesek
\hi 
\HT{tIwesek}{\B{tì\cp{we}sek}} \I{[tI."wE.sEk\textcorner]} \emph{n.} subtlety.
\B{Ke lo\-lu kea tì\-we\-sek mì sä\-ftxu\-lì\-'u a\-räp\-tum pe\-yä.} There was no subtlety in his coarse speech.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b} 
(\ding{43}~\HL{wesek}{\B{we\-sek}})

% tìwìngay
\hi
\HT{tIwIngay}{\B{tìwì\cp{ngay}}} \I{[tI.wI."Naj]} \emph{n.} proof, proving (\emph{abstract}).
\B{Txo new ngal fu\-ta su\-tel nge\-yä ay\-lì\-'u\-ti spi\-vaw, tsran\-ten tì\-wì\-ngay.} If you want people to believe you, proof is important.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{wIngay}{\B{wì\-ngay}}, \HL{sAwIngay}{\B{sä\-wì\-ngay}})

% tìyawn
\hi
\HT{tIyawn}{\B{tì\cp{yawn}}} \I{[tI."jawn]} \emph{n.} love.   
\B{Spaw Na'\-vil fu\-ta tì\-yawn Ey\-wa\-yä lu txew\-lu\-ke.} The Na'vi believe that Eywa's love is boundless.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
\B{Lì'\-fya\-ri le\-Na'\-vi oel 'e\-fu ay\-nge\-yä tì\-yawn\-it.} I feel your love for the Na'vi language.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Nge\-yä ke ftang a tì\-kang\-kem\-ìri sì tì\-ka\-nu\-ri sì fì\-lì'\-fya\-yä tì\-yawn\-ìri sei\-yi oe ira\-yo nì\-txan.} I thank you very much for your never\hyp{}ending work, intelligence and love for the language.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/comment-page-1/\#comment-1547}{b}
(\emph{a. idiom}) \B{tì\-yewn tì\-yawn\-ä} `an expression of love'~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{214}~\HL{tIve'kI}{\B{tì\-ve'\-kì}}, \ding{43}~\HL{yawne}{\B{yaw\-ne}})

% tìyawr
\hi
\HT{tIyawr}{\B{tì\cp{yawr}}} \I{[tI."jawR]} \emph{n.} correctness. 
(\emph{a. idiom}) \B{Nga\-ru tì\-yawr.} You're right. 
\B{Nga\-ru tì\-yawr nì\-wotx!} You are completely right.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/comment-page-1/\#comment-191}{b} 
\B{Nì\-ngay lu ngar tì\-yawr, lu ngar tì\-kxey.} Truly, you are both right and wrong.~\LINK{http://naviteri.org/2010/07/vocabulary-update/comment-page-1/\#comment-95}{b} 
\B{Oe\-ru tì\-yawr srak?} Am I right?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/comment-page-1/\#comment-1452}{b}
(\ding{214}~\HL{tIkxey}{\B{tì\-kxey}}, \ding{43}~\HL{eyawr}{\B{eyawr}})

% tìyäkx
\hi
\HT{tIyAkx}{\B{tì\cp{yäkx}}} \I{[tI."j\ae{}k']} \emph{n.} lack of notice; snubbing.
\B{Tì\-yäkx ke lu sru\-nga', ma tsmuk. Nga txo sti, oeyk\-tìng teyng\-ta pe\-lun.} Snubbing isn't helpful, brother. If you're angry, explain why.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\ding{43}~\HL{yAkx}{\B{yäkx}}, \HL{sAyAkx}{\B{sä\-yäkx}})

% tìyewn
\hi 
\HT{tIyewn}{\B{tì\cp{yewn}}} \I{[tI."jEwn]} \emph{n.} expression (\emph{of a thought or feeling}). 
(\emph{a. idiom}) \B{tì\-yewn tì\-yawn\-ä} `an expression of love'~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{yewn}{\B{yewn}}, \HL{lI'fyavi}{\B{lì'\-fya\-vi}})

% tìyo'
\hi
\HT{tIyo'}{\B{tì\cp{yo'}}} \I{[tI."joP]} \emph{n.} perfection.
\B{Fì\-tseo\-ri ke tsun kaw\-tu pi\-vä\-hem tì\-yo'\-ne; tsran\-ten tì\-pä\-hem\-ä tì\-fmi nì\-'aw.} In this art it's impossible to arrive at perfection; the only thing that matters is the attempt to arrive there.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{yo'}{\B{yo'}})

% tìyora'
\hi
\HT{tIyora'}{\B{tìyo\cp{ra'}}} \I{[tI.jo."RaP]} \emph{n.} victory, a win. 
\B{Ul\-te kxawm zì\-ye\-va\-'u nga\-ne tì\-yo\-ra' zì\-sìt\-ay!} And maybe victory might come to you next year!~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b}
\B{Tsam\-si\-yu\-ri lu tì\-yo\-ra'\-ä tì\-new le\-kin.} A warrior must have the desire for victory.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{214}~\HL{tIsnaytx}{\B{tì\-snaytx}}, \ding{43}~\HL{yora'}{\B{yo\-ra'}})

% tìzevakx
\hi 
\HT{tIzevakx}{\B{tì\cp{ze}vakx}} \I{[tI."zE.vak']} \emph{n.} cruelty. 
(\ding{43}~\HL{zevakx}{\B{zevakx}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIzevakxnga'}{\B{tì\-ze\-vakx\-nga'}}, cruel.

% tìzevakxnga'
\hi 
\HT{tIzevakxnga'}{\B{tì\cp{ze}vakxnga'}} \I{[tI."zE.vak'.NaP]} \emph{adj., nfp.} cruel. 
\B{ay\-lì\-'u a\-tì\-ze\-vakx\-nga'}, cruel words.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{zevakx}{\B{zevakx}}; \HL{tIzevakx}{\B{tì\-ze\-vakx}}, \HL{-nga'}{\B{-nga'}})

% tìzin (si)
\hi 
\HT{tIzin}{\B{tì\cp{zin}}} \I{[tI."zin]} \textsc{i}. \emph{n.} a tangle, tanglement; \emph{mass of s.t. twisted together}. 
\B{Fwa fpìl nì\-fya\-'o a\-ve\-nga' fì\-tì\-zin\-mì lu kel\-tsun.} It's impossible to think clearly (i.e., in an organized manner) in this muddled situation.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/\#comment-33483}{b}
(\ding{43}~\HL{zin}{\B{zin}}) \\
\textsc{ii}. \B{tì\cp{zin} si} \I{[tI."zin si]} \emph{vin.} tangle, tangle up.
\B{Na\-ri si fte\-ke ay\-tur\-tel\-ur tì\-zin si\-vi!} Be careful not to tangle the ropes.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} 

% to
\hi
\HT{to}{\B{to}} \I{[to]} \emph{part.} (\emph{comparative marker}) than.   
\B{Po to oe lu sìl\-tsan.} S\slash he is better than me.~\LINK{http://forum.learnnavi.org/language-updates/prrwll-moss-the-comparative-to-construct/}{ln}
\B{Nga nì\-aw\-no\-mum to oe\-tsyìp lu txur nì\-txan.} As everyone knows, you're a lot stronger than little old me.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Oe to nga tul nì\-win.} I run faster than you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Oel to ngal ye\-rik\-it ta\-ron nìl\-tsan.} I hunt yerik better than you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Oel pay\-ti nul\-new to swo\-at.} I prefer water to alcohol.~\LINK{http://forum.learnnavi.org/language-updates/prefer-x-to-y-24474/}{ln}
(\emph{can be attached as a suffix}) \B{Oe nga\-to yom nì\-'ul.} I eat more than you.~\LINK{http://forum.learnnavi.org/language-updates/prefer-x-to-y-24474/}{ln}
\B{Po\-to oe lu ko\-ak.} I am older than he.~\LINK{http://naviteri.org/2010/07/thoughts-on-ambiguity/}{b}
\B{Oe\-ri tsaw ke ley fì\-'u\-to.} I don't value that over this.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b}
(\emph{can refer to whole sentences}) \B{Ftu\-e lu fwa ta\-ron ngong\-a io\-ang\-it to fwa ta\-ron pum\-it a lu wa\-lak sì win.} It's easier to hunt lethargic animals than to hunt perky, speedy ones.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{a. idiom}) \B{to tì\cp{tse}ri} \I{[to tI."\t{ts}E.Ri]} than is apparent, than you are aware of. 
\B{Lu po\-e na nga nì\-'ul to tì\-tse\-ri.} She's more like you than you think (\emph{or:} than you know).~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b} 
(\ding{43}~\HL{frato}{\B{fra\-to}})

% Toitslan
% \hi
% \B{\cp{To}itslan} \I{["to.i.\t{ts}lan]} \emph{n.}, \textsc{lw} Germany.

% toitsye
% \hi
% \B{\cp{To}itsye} \I{"to.i.\t{ts}jE} \emph{adj.} German, German language.

% tok
\hi
\HT{tok}{\B{tok}} \I{[tok\textcorner]} \emph{vtr.} (\emph{a. local}) be at, occupy a space. 
\B{Nì\-'i'\-a tok oel nì\-mun fì\-tse\-nget!} Finally, I'm here again!~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Pep\-sul tok fì\-tse\-nget?} What three people are here?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
\B{Aw\-ngal tok kel\-kut.} We're home.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Tsìl\-pey\-ìl tok txe'\-lan\-it.} Hope lives within the heart.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Li pol te\-rok fì\-tseng\-it srak?} Is she already here?~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok.} Yesterday I was with Txewì in the forest.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
(\emph{b. can be omitted in casual speech when easily understood})
\B{Pol fì\-tse\-nget.} He's here.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
\B{Pol ke fì\-tse\-nget.} He's not here.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
\B{Po\-ru ke po\-leng oel ke\-'ut mung\-wrr fwa Ra\-lul ke tsa\-tse\-nget.} I told her nothing except that Ralu wasn't there.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}
(\emph{c. metaphorically with} \B{yì}) 
\B{Lì'\-fya\-ri pol tok pe\-yìt?} How good is her Na'vi?~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Fu\-ri\-a tä\-ftxu ngal tok yì\-pet? -- Tok yìt a\-ke\-sran.} How's your weaving? -- I'm so\hyp{}so.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\emph{d. idiom}) \B{Nga\-ri ke\-ftxo (fwa) ke tok.} ``We missed you.\slash Sorry you couldn't make it.\slash Too bad you couldn't be there.''~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{fkeytok}{\B{fkey\-tok}}, exist. 
\on{2}.~\HL{fewtusok}{\B{few\-tu\-sok}}, opposite.
\on{3}.~\HL{tItok}{\B{tì\-tok}}, presence.
\on{4}.~\HL{pxawtok}{\B{pxaw\-tok}}, surround.

% tokat
\hi 
\HT{tokat}{\B{\cp{to}kat}} \I{["to.kat\textcorner]} \emph{adj.} guilty. 
(\ding{214}~\HL{layl}{\B{layl}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tItokat}{\B{tì\-to\-kat}}, guilt. 
\on{2}.~\HL{nItokat}{\B{nì\-to\-kat}}, guiltily. 

% toktor
\hi
\HT{toktor}{\B{\cp{tok}tor}} \I{["tok\textcorner.toR]} \emph{n.}, \textsc{lw} doctor (\emph{academic degree}). 
\B{Lo\-lu tok\-tor Kì\-rey\-sì kar\-yu a\-sìl\-tsan.} Doctor Grace was a good teacher.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 

% tokx
\hi
\HT{tokx}{\B{tokx}} \I{[tok']} \emph{n.} body.
\B{Nga\-ri hu Ey\-wa sa\-lew ti\-rea, tokx 'ì\-'awn, slu Na'\-vi\-yä ha\-pxì.} Your spirit goes with Eywa, your body stays behind to become part of the People.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake.27s_final_kill}{a\slash wiki} 
\B{Fì\-po lu vrr\-tep a mì sokx a\-tsleng!} This one is a demon in a false body!~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Tsu.27tey_attacks_Jake}{wiki\slash a}
(\emph{a. idiom}) \B{tokx eo tokx}, face to face, in person.~\LINK{http://forum.learnnavi.org/language-updates/fmawno/}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letokx}{\B{le\-tokx}}, bodily, physical. 
\on{2}.~\HL{fpomtokx}{\B{fpom\-tokx}}, (physical) health. 
\on{3}.~\HL{uniltIrantokx}{\B{u\-nil\-tì\-ran\-tokx}}, avatar, ``dreamwalker body''. 
\on{4}.~\HL{yemstokx}{\B{yem\-stokx}}, put on (clothing). 

% toltem
\hi
\HT{toltem}{\B{tol\cp{tem}}} \I{[tol."tEm]} \emph{vtr.} shoot (s.o.\slash s.t.).
\B{Pll\-txe Ra\-lu san oe new ti\-vol\-tem ye\-rik\-it.} Ralu says he wants to shoot a hexapede.~\LINK{http://naviteri.org/2017/03/titusemteri-concerning-shooting/}{b}
(\ding{43}~\HL{tem}{\B{tem}}; \HL{tsweykayon}{\B{tsway\-on} (\emph{b. \ding{246}})})

% tompa
\hi
\HT{tompa}{\B{\cp{tom}pa}} \I{["tom.pa]} \emph{n.} rain.   
\B{Tom\-pa\-yä ka\-to | Tsaw\-ke\-yä ka\-to.} The rhythm of the rain | The rhythm of the sun.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
\B{Ye'\-rìn sngä\-'i tom\-pa.} Soon the rain starts.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg475008/\#msg475008}{ln}
(\emph{a. precipitation, with} \B{zup})
\B{Ze\-rup tom\-pa.} It's raining.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Nìk\-mar zup tom\-pa set.} Now is a good time for it to be raining.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{with} \B{'eko}) \B{Tom\-pa 'e\-ko nì\-hawng, ha ze\-ne aw\-nga kll\-wi\-vo.} The rain is too strong, so we must land.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
\textsc{Derivation}: \ding{43}
\on{1}.~\HL{tompameyp}{\B{tom\-pa\-meyp}}, drizzle. 
\on{2}.~\HL{tompawll}{\B{tom\-pa\-wll}}, geode. 
\on{3}.~\HL{txantompa}{\B{txan\-tom\-pa}}, rainstorm. 
\on{4}.~\HL{tomperwI}{\B{tom\-per\-wì}}, sleet. 
\on{5}.~\HL{tompIva}{\B{tom\-pì\-va}}, raindrop. 
\on{6}.~\HL{tompakel}{\B{tom\-pa\-kel}}, drought.
\on{7}.~\HL{tompatayng}{\B{tom\-pa\-tayng}}, rain thistle.

% tompakel
\hi
\HT{tompakel}{\B{\cp{tom}pakel}} \I{["tom.pa.kEl]} \emph{n.} drought.   
\B{Tom\-pa\-kel\-ta\-lun ze\-ne tu\-te Kä\-lì\-for\-ni\-a\-yä pay\-it si\-var nì\-nän.} Because of the drought, Californians have to use less water.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{tompa}{\B{tom\-pa}}, \HL{tIkelu}{\B{tì\-ke\-lu}})

% tompameyp
\hi
\HT{tompameyp}{\B{tompa\cp{meyp}}} \I{[tom.pa."mEjp\textcorner]} \emph{n.} drizzle.
\B{Fì\-re\-won tom\-pa\-meyp zar\-mup, slä set 'ì\-me\-ko nì\-txan nang.} It was drizzling this morning, but it's really started coming down now!~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
(\ding{43}~\HL{tompa}{\B{tom\-pa}}, \HL{meyp}{\B{meyp}})

% tompatayng
\hi 
\HT{tompatayng}{\B{\cp{tom}patayng}} \I{["tom.pa.tajN]} \emph{n.} \ding{96} rain thistle. 
(\ding{43}~\HL{tompa}{\B{tompa}}, \HL{tayng}{\B{tayng}})

% tompawll
\hi
\HT{tompawll}{\B{\cp{tom}pawll}} \I{["tom.pa.w\s{l}]} \emph{n.}, \ding{96} geode, ``rain plant'', \emph{tubineus azureus}.
(\ding{43}~\HL{tompa}{\B{tom\-pa}}, \HL{'ewll}{\B{'ewll}})

% tomperwì
\hi
\HT{tomperwI}{\B{\cp{tom}perwì}} \I{["tom.pER.wI]} \emph{n.} sleet.
(\ding{43}~\HL{tompa}{\B{tom\-pa}}, \HL{herwI}{\B{her\-wì}})

% tompìva
\hi
\HT{tompIva}{\B{\cp{tom}pìva}} \I{["tom.pI.va]} \emph{n.} raindrop.
\B{Oe\-ri ay\-som\-pì\-va sìn re\-'o var zi\-vup.} Raindrops keep falling on my head.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
(\ding{43}~\HL{tompa}{\B{tom\-pa}}, \HL{Ilva}{\B{ìl\-va}})

% tonowari
\hi
\HT{Tonowari}{\B{Tono\cp{wa}ri}} \I{[to.no."wa.ri]} \emph{n. male name}.

% tong
\hi 
\HT{tong}{\B{tong}} \I{[toN]} \emph{vtr.} put out, quench, exstinguish (\emph{of a fire}). 
\B{Maw\-krr\-a ngal txep\-it to\-long tsun hi\-vum.} After you've put out the fire you can leave.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
\B{Tong txep\-tsyìp\-it.} Extinguish the flame.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
\B{Lo\-nu\-sye tong txep\-tsyìp\-it.} Blow out the flame.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}

% tor
\hi
\HT{tor}{\B{tor}} \I{[toR]} \emph{adj.} last, ultimate, terminal, bring about finality.
\B{tì\-pawm a\-tor}, ultimate question.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} 
(\ding{43}~\HL{syen}{\B{syen}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{toruk}{\B{to\-ruk}}, great leonopteryx.

% toruk
\hi
\HT{toruk}{\B{\cp{to}ruk}} \I{["to.Ruk\textcorner]} \emph{or} \I{[.RUk\textcorner]} (\textsc{rn}: \B{\cp{to}rùk} \I{["to.RUk\textcorner]}) \emph{n.} (\emph{a. fauna}) great leonopteryx, ``last shadow'', \emph{leonopteryx rex}. 
\B{Tseyk tswa\-may\-on fa ik\-ran sre\-krr; ta\-fral fmo\-li fì\-kem si\-vi fa to\-ruk nì\-steng.} Jake flew with an ikran before; therefore he tried to do it with a \emph{toruk} in a similar fashion.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{ean na tsuk\-sìm to\-ruk\-ä}, blue like the chin of a great leonopteryx.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
(\emph{b. idiom})
\B{Ke tsun fko tspi\-vang to\-ruk\-it fa fwa pewn\-ti snew.} \emph{Proverbial expression for a method that will not work.} [lit.: You can't kill a great leonopteryx by constricting its throat]~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} 
(\emph{shortened to}) \B{pewn to\-ruk\-ä}. \B{Sra\-ke fpìl ay\-ngal fu\-ta aw\-nga tsun hu Saw\-tu\-te a tsam\-it 'i\-vaw\-nìm fa fwa päng\-kxo fo\-hu nì\-'aw, ma smuk? Pewn to\-ruk\-ä!} Do you think we can avoid a war with the Sky People solely by talking with them, my brothers? That won't work!~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/comment-page-1/\#comment-1755}{b} \\
(i.) \B{\cp{to}ruk \cp{mak}to} \I{["to.Ruk\textcorner{} "mak\textcorner.to]} \emph{n.} Rider of the Last Shadow (\emph{legendary hero of the Na'vi}).
\B{To\-ruk Mak\-to po\-lä\-hem a fì\-'u ro\-lo\-'a nì\-txan Oma\-ti\-ka\-ya\-ru.} The arrival of Toruk Makto made a great impression on the Omaticaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{tor}{\B{tor}}, \HL{uk}{\B{uk}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{torukspxam}{\B{to\-ruk\-spxam}}, octoshroom. 
\on{2}.~\HL{nawmtoruktek}{\B{nawm\-to\-ruk\-tek}}, Toruk Makto Totem.

% torukspxam
\hi
\HT{torukspxam}{\B{\cp{to}rukspxam}} \I{["to.Ruk\textcorner.sp'am]} \emph{or} \I{[.RUk\textcorner.]} (\textsc{rn}: \B{\cp{to}rùkspxam} \I{["to.RUk\textcorner.sp'am]}) \emph{n., funga} octoshroom, ``great leonopteryx fungus'', \emph{fungimonium giganteum}.
(\ding{43}~\HL{toruk}{\B{to\-ruk}}, \HL{spxam}{\B{spxam}})

% trr
\hi
\HT{trr}{\B{trr}} \I{[t\s{r}]} \emph{n.} day.   
\B{Tom\-pa\-yä ka\-to | Tsaw\-ke\-yä ka\-to | Trr\-ä sì txon\-ä.} The rhythm of the rain | The rhythm of the sun | Of day and night.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
\B{Tskot sngo\-lä\-'i po si\-var kam trr 'a\-'aw\-a.} He started to use the bow several days ago.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Fì\-trr lu trr\-pe\-ve?} What day is it today?~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Ay\-nge\-yä ay\-srr tì\-flä\-ta te\-ya li\-vu ul\-te si\-va\-lew nì\-'o' nì\-'aw.} May your days be full of success and proceed only in a fun way.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/}{b}
\B{Fì\-tì\-kang\-kem\-vi\-ri oel kin 'aw\-a trr\-ti nì\-'aw.} For this project I only need one day.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
(\emph{idiom}) \B{trr le\-fpom}, good day, good morning (cf. French \emph{bon jour}).
\B{Trr le\-fpom, ma Su\-te Ame\-ri\-ka\-yä.} Good morning, America.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} \\
(i.) \HT{trram}{\B{trr\cp{am}}} \I{[t\s{r}."{}am]} \emph{adv., n.} yesterday. 
\B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok, tso\-le\-'a syep\-tu\-tet a\-tsawl fra\-to mì sì\-rey.} Yesterday I was with Txewì in the forest and we saw the biggest Trapper I've ever seen.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
\B{Trr\-am\-ä ay\-u\-van\-ìri mak\-to Ak\-wey nì\-yew\-la ha snaytx.} Akwey rode disappointingly in yesterday's games so he lost.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
(ii.) \HT{trray}{\B{trr\cp{ay}}} \I{[t\s{r}."{}aj]} \emph{adv., n.} tomorrow. 
\B{Pam\-rel sa\-yi trr\-ay krr a skxom la\-tsu oe\-ru.} I will write tomorrow when I have the chance.~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln} 
\B{Trr\-ay\-ri li\-vu nì\-sìl\-pey ay\-lrr\-tok a\-txan ta Ey\-wa aw\-nga\-ru nì\-wotx.} As for tomorrow, hopefully much smiles from Eywa to you all.~\LINK{http://forum.learnnavi.org/language-updates/luke-kxu-without-harm/msg132308/\#msg132308}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{fratrr}{\B{fra\-trr}}, every day. 
\on{2}.~\HL{fItrr}{\B{fì\-trr}}, today. 
\on{3}.~\HL{letrr}{\B{le\-trr}}, daily. 
\on{4}.~\HL{nItrrtrr}{\B{nì\-trr\-trr}}, ordinarily. 
\on{5}.~\HL{trrtxon}{\B{trr\-txon}}, day and night cycle. 
\on{6}.~\HL{trr'ong}{\B{trr\-'ong}}, dawn, sunrise. 
\on{7}.~\HL{kxamtrr}{\B{kxam\-trr}}, midday, noon. 
\on{8}.~\HL{mrrtrr}{\B{mrr\-trr}}, (five\hyp{}day) work week. 
\on{9}.~\HL{kintrr}{\B{kin\-trr}}, seven\hyp{}day week. 
\on{10}.~\HL{muntrr}{\B{mun\-trr}}, weekend. 
\on{11}.~\HL{trr'awve}{\B{Trr\-'aw\-ve}}, Sunday. 
\on{12}.~\HL{trrmuve}{\B{Trr\-mu\-ve}}, Monday. 
\on{13}.~\HL{trrpxeyve}{\B{Trr\-pxey\-ve}}, Tuesday. 
\on{14}.~\HL{trrtsIve}{\B{Trr\-tsì\-ve}}, Wednesday. 
\on{15}.~\HL{trrmrrve}{\B{Trr\-mrr\-ve}}, Thursday. 
\on{16}.~\HL{trrpuve}{\B{Trr\-pu\-ve}}, Friday. 
\on{17}.~\HL{trrkive}{\B{Trr\-ki\-ve}}, Saturday. 
\on{18}.~\HL{zIsItsaltrr}{\B{zì\-sìt\-sal\-trr}}, (yearly) anniversary.
\on{19}.~\HL{trrpxI}{\B{trr\-pxì}}, part of day; hour.
\on{20}.~\HL{fleltrr}{\B{flel\-trr}}, April Fool's Day.

% trr'awve
\hi
\HT{trr'awve}{\B{Trr\cp{'aw}ve}} \I{[t\s{r}."Paw.vE]} \emph{n.} Sunday. 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{'awve}{\B{'aw\-ve}})

% trr'ong
\hi
\HT{trr'ong}{\B{trr\cp{'ong}}} \I{[t\s{r}."PoN]} \emph{n.}; \emph{adv.} daybreak, dawn.
\B{Trr\-'ong\-ta Txon\-'ong\-vay po to\-lì\-ran.} He walked from dawn until dusk.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{trr}{\B{trr}}, \HL{'ong}{\B{'ong}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sresrr'ong}{\B{sre\-srr\-'ong}}, before daybreak. 
\on{2}.~\HL{trr'ongmaw}{\B{trr\-'ong\-maw}}, after daybreak.  

% trr'ongmaw
\hi
\HT{trr'ongmaw}{\B{trr\cp{'ong}maw}} \I{[t\s{r}."PoN.maw]} \emph{n.}; \emph{adv.} after daybreak, after dawn.
(\ding{43}~\HL{trr'ong}{\B{trr\-'ong}}, \HL{maw}{\B{maw}})

% trrkive
\hi
\HT{trrkive}{\B{Trr\cp{ki}ve}} \I{[t\s{r}."ki.vE]} \emph{n.} Saturday. 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{kive}{\B{ki\-ve}})

% trrmuve
\hi
\HT{trrmuve}{\B{Trr\cp{mu}ve}} \I{[t\s{r}."mu.vE]} \emph{n.} Monday. 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{muve}{\B{mu\-ve}})

% trrmrrve
\hi
\HT{trrmrrve}{\B{Trr\cp{mrr}ve}} \I{[t\s{r}."m\s{r}.vE]} \emph{n.} Thursday. 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{mrrve}{\B{mrr\-ve}})

% trrpuve
\hi
\HT{trrpuve}{\B{Trr\cp{pu}ve}} \I{[t\s{r}."pu.vE]} \emph{n.} Friday. 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{puve}{\B{pu\-ve}})

% trrpxeyve
\hi
\HT{trrpxeyve}{\B{Trr\cp{pxey}ve}} \I{[t\s{r}."p'Ej.vE]} \emph{n.} Tuesday. 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{pxeyve}{\B{pxey\-ve}})

% trrpxì
\hi 
\HT{trrpxI}{\B{trr\cp{pxì}}} \I{[t\s{r}."p'I]} \textsc{i}. \emph{n.} (\emph{a. on Pandora}) part of day. 
(\emph{b. on Earth}) hour (\emph{shortened from:} \B{trr\-pxì Saw\-tu\-te\-yä} \emph{and a variant of:} \ding{43}~\HL{pxevotrrpxI}{\B{pxe\-vo\-trr\-pxì}})
\B{Po tì\-kang\-kem so\-li (ka) trr\-pxì\-o a\-mrr.} She has worked for five hours.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
(\emph{b.1 to tell the full hour}) \B{tì\-'i\-'a trr\-pxì\-yä a\-mrr\-ve}, `five o'clock' [lit.: end of the fifth hour], \emph{usually abbreviated to} \B{trr\-pxì a\-mrr\-ve}, `five o'clock' [lit.: fifth hour] \emph{or even} \B{Lu mrr\-ve.} `It's five.' [lit.: `it's the fifth']~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
(\emph{b.2 to tell minutes}) \B{trr\-pxì a\-mrr\-ve sì mawl} \emph{or coll.} \B{mrr\-ve sì mawl}, (it's) \on{5}:\on{30}.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
\B{trr\-pxì a\-mrr\-ve sì trr\-pxì\-vi a\-pxe\-vol} \emph{or coll.} \B{mrr\-ve sì pxe\-vol}, \on{5}:\on{24}.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
\B{trr\-pxì\-vi a\-vo\-mun sre srr\-pxì a\-mrr\-ve} \emph{or coll.} \B{vo\-mun sre mrr\-ve}, (it's) \on{4}:\on{50}, or `ten to five'.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} \\
\textsc{ii.} \B{trr\cp{pxì}pe} \I{[t\s{r}."p'I.pE]} \emph{or} \B{pe\-srr\-\cp{pxì}} \I{[pE.s\s{r}."p'I]} \emph{inter.} what part of the day? what hour? 
\B{Fo pä\-hem trr\-pxì\-pe?--Sre\-kam\-trr.} What part of the day will they arrive?--Before noon.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{hapxI}{\B{hapxì}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{trrpxIvi}{\B{trr\-pxì\-vi}}, minute.

% trrpxìvi
\hi 
\HT{trrpxIvi}{\B{trr\cp{pxì}vi}} \I{[t\s{r}."p'I.vi]} \emph{n.} minute (on Earth). 
\B{Oe pä\-hem maw\slash kay trr\-pxì\-vi a\-mrr.} I'll be there in five minutes.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} 
(\emph{a. to tell time}) \B{trr\-pxì a\-mrr\-ve sì mawl} \emph{or coll.} \B{mrr\-ve sì mawl}, (it's) \on{5}:\on{30}.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
\B{trr\-pxì a\-mrr\-ve sì trr\-pxì\-vi a\-pxe\-vol} \emph{or coll.} \B{mrr\-ve sì pxe\-vol}, (it's) \on{5}:\on{24}.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b}
\B{trr\-pxì\-vi a\-vo\-mun sre srr\-pxì a\-mrr\-ve} \emph{or coll.} \B{vo\-mun sre mrr\-ve}, (it's) \on{4}:\on{50}, or `ten to five'.~\LINK{https://naviteri.org/2024/05/krrteri-about-time/}{b} \\
(\ding{43}~\HL{trrpxI}{\B{trrpxì}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{trrpxIvitsyIp}{\B{trr\-pxì\-vi\-tsyìp}}, second. 

% trrpxìvitsyìp
\hi 
\HT{trrpxIvitsyIp}{\B{trr\cp{pxì}vitsyìp}} \I{[t\s{r}."p'I.vi.\t{ts}jIp\textcorner]} \emph{n.} second (on Earth). 
(\ding{43}~\HL{trrpxIvi}{\B{trr\-pxì\-vi}}) 

% trrtsìve
\hi
\HT{trrtsIve}{\B{Trr\cp{tsì}ve}} \I{[t\s{r}."\t{ts}I.vE]} \emph{n.} Wednesday. 
(\ding{43}~\HL{trr}{\B{trr}}, \HL{tsIve}{\B{tsì\-ve}})

% trrtxon
\hi
\HT{trrtxon}{\B{trr\cp{txon}}} \I{[t\s{r}."t'on]} \emph{n.} day night cycle.
\B{Oe tì\-kang\-kem si trr\-txon nì\-wotx, nga\-ru ke\-'u!} I work all day and all night, and you don't give a damn!~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
(\ding{43}~\HL{trr}{\B{trr}}, \HL{txon}{\B{txon}})

% -tu
\hi
\HT{-tu}{\B{-tu}} \I{[tu]} \emph{suf. non\hyp{}productive to mainly turn word classes other than v.s into an agent, e.g., of n.s:} \B{pam\-tseo\-tu}, musician; \emph{of adj.s:} \B{nawm\-tu}, great person; \emph{other word classes:} \B{wä\-tu}, opponent; \emph{etc.}
(\ding{43}~\HL{-yu}{\B{-yu}})

% Tuk
\hi
\HT{Tuk}{\B{Tuk}} \I{[tuk\textcorner]} \emph{or} \I{[tUk\textcorner]} \emph{n., female name} (\emph{short form of:} \ding{43}~\HL{Tuktirey}{\B{Tuk\-ti\-rey}}).

% Tuke
\hi
\HT{Tuke}{\B{\cp{Tu}ke}} \I{["tu.kE]} \emph{n., female name}.
\B{Tu\-ke\-ru pol\-txe Ak\-wey nam\-'a\-ke, omum fu\-ta ke tsun poe sti\-vo.} Akwey spoke to Tuke confidently, knowing that she couldn't refuse.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}

% tukru
\hi
\HT{tukru}{\B{tuk\cp{ru}}} \I{[tuk\textcorner."Ru]} \emph{or} \I{[tUk\textcorner.]} (\textsc{rn}: \B{tùk\cp{ru}} \I{[tUk\textcorner."Ru]}) \emph{n.} spear.
\B{Oe\-yä swi\-zaw nì\-ngay ti\-va\-kuk | Oe\-yä tuk\-rul txe'\-lan\-it ti\-va\-kuk.} Let my arrow strike true | Let my spear strike the heart.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Txe\-wìl tuk\-rut Lo\-ak\-ä nar\-mìn nì\-fmokx.} Txewì was eyeing Loak's spear enviously.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{tuk\-ru a txo\-lu\-la fkol ta rìn\slash tuk\-ru a ta rìn\slash tuk\-ru le\-rìn}, a spear made of wood.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}

% Tuktirey
\hi 
\HT{Tuktirey}{\B{Tuktirey}} \I{[tuk\textcorner.ti.REj]} \emph{n. female name} (\emph{short form:} \ding{43}~\HL{Tuk}{\B{Tuk}}).

% tul
\hi
\HT{tul}{\B{tul}} \I{[tul]} \emph{or} \I{[tUl]} \emph{vin.} (\emph{a.}) run. 
\B{Hum ne wrr\-pa! Tul!} Get outside! Run!~\LINK{http://forum.learnnavi.org/language-updates/txing-and-hum/}{ln\slash a\textsuperscript{c}} 
\B{Oe tul nì\-ftxan nì\-win na nga.} I run as fast as you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Oe to nga tul nì\-win.} I run faster than you.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Po tul nì\-win fra\-to.} She is the fastest.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Txo fko ti\-vul mì na'\-rìng nì\-hawm\-pam, stawm ay\-ioang.} If you run noisily in the forest, the animals will hear.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\emph{b. of the weather}) \B{Tul hu\-fwe nì\-win.} It's very windy.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
(\emph{c. rhyming expression}) \B{Tslikx, tì\-ran, tul; | Ftu yì ne yì tsan\-'ul.} \emph{when learning s.t. new, you have to proceed from step to step: baby steps first, then bigger ones} [lit.: `Crawl, walk, run; from level to level get better.']~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(i.) \B{sä\-wä\-sul a tul}, foot race.~\LINK{http://naviteri.org/2012/10/wina-sawasultsyip-ahii-a-quick-little-contest/}{b}

% tulkun
\hi
\HT{tulkun}{\B{\cp{tul}kun}} \I{["tul.kun]} \emph{or} \I{["tUl.kUn]} (\textsc{rn}: \B{\cp{tùl}kun} \I{["tUl.kun]}) \emph{n., fauna} Tulkun (\emph{sentient whale\hyp{}like Pandoran sea creature}). 

% tumpasuk
\hi 
\HT{tumpasuk}{\B{\cp{tum}pasuk}} \I{["tum.pa.suk\textcorner]} \emph{or} \I{["tUm.pa.sUk\textcorner]} (\textsc{rn}: \B{\cp{tum}pasùk} \I{["tum.pa.sUk\textcorner]}) \emph{n.} \ding{96} celia fruit tree, \emph{pampinus bacca acinum}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln} 

% tumpin
\hi
\HT{tumpin}{\B{\cp{tum}pin}} \I{["tum.pin]} \emph{or} \I{["tUm.]}  \emph{adj.} the color \ding{43}~\HL{tun}{\B{tun}}.
(\ding{43}~\HL{'opin}{\B{'o\-pin}})

% tun
\hi
\HT{tun}{\B{tun}} \I{[tun]} \emph{or} \I{[tUn]}  \emph{adj.} red, orange.
\B{Oe\-ri key tun slo\-lu nì\-txan.} I'm so embarrassed [lit.: my face has become very red].~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-694}{b}
\B{tun na eyk\-tan}, leader red (\emph{the reddish color that distinguishes the dress of Na'vi leaders}).~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tumpin}{\B{tum\-pin}}, the color red\slash orange. 
\on{2}.~\HL{reypaytun}{\B{rey\-pay\-tun}}, red. 
\on{3}.~\HL{txeptun}{\B{txep\-tun}}, orange. 

% tunu
\hi 
\HT{tunu}{\B{\cp{tu}nu}} \I{["tu.nu]} \emph{adj.} romantic. 
\B{Nga\-ri 'e\-fu oe tu\-nu.} I feel romantic towards you; I have romantic feelings for you.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b}  
\B{Po yaw\-ne lu oer, slä po\-ri ke 'e\-fu oe tu\-nu.} I love him, but I don't have romantic feelings for him.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItunu}{\B{tì\-tu\-nu}}, romance.
\on{2}.~\HL{tunutu}{\B{tu\-nu\-tu}}, object of desire; ``crush''.

% tunutu
\hi 
\HT{tunutu}{\B{\cp{tu}nutu}} \I{["tu.nu.tu]} \emph{n.} object of desire; ``crush''. 
(\ding{43}~\HL{tunu}{\B{tu\-nu}})

% tung
\hi
\HT{tung}{\B{tung}} \I{[tuN]} \emph{or} \I{[tUN]} (\textsc{rn}: \B{tùng} \I{[tUN]}) \emph{vtr.} allow, let, permit. 
\B{Ke tung fkol tì\-wu\-sem\-it fì\-tseng.} Fighting isn't allowed here.~\LINK{http://naviteri.org/2017/01/meliuteri-alu-tung-si-pllhrr-about-tung-and-pllhrr/}{b}
\B{Pol to\-lung oe\-ru fu\-ta ki\-vä.\slash Pol to\-lung fu\-ta oe ki\-vä.} He allowed me to go.~\LINK{http://naviteri.org/2017/01/meliuteri-alu-tung-si-pllhrr-about-tung-and-pllhrr/}{b}
\B{Tung oer fu\-ta kä!} Let me go!~\LINK{http://naviteri.org/2017/01/meliuteri-alu-tung-si-pllhrr-about-tung-and-pllhrr/}{b}
\B{Ke tung Na'\-vil fu\-ta tswìk kxe\-ner\-it Mo\-'a\-ra\-ka nì\-wotx.} The Na'vi have banned smoking in Mo\-'ara.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tungzup}{\B{tung\-zup}}, drop, let fall.

% tungzup
\hi
\HT{tungzup}{\B{tung\cp{zup}}} \I{[tuN."zup\textcorner]} \emph{or} \I{[tUN."zUp\textcorner]} (\textsc{rn}: \B{tùng\cp{zup}} \I{[tUN."zup\textcorner]}) \emph{vtr.} drop, let fall (\emph{accidentally, inadvertently}). 
\B{Ay\-oe sno\-laytx fì\-trr ta\-lun\-a oel rum\-it to\-lung\-zup.} We lost today because I dropped the ball.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{tung}{\B{tung}}, \HL{zup}{\B{zup}}, \HL{zeykup}{\B{zey\-kup}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIktungzup}{\B{nìk\-tung\-zup}}, (hold) carefully, firmly.

% tup
\hi
\HT{tup}{\B{tup}} \I{[tup\textcorner]} \emph{or} \I{[tUp\textcorner]} (\textsc{rn}: \B{tùp} \I{[tUp\textcorner]}) \emph{conj.} instead of, rather than. 
\B{Lam ngay oer, fo ka\-yä ìlä hil\-van tup na'\-rìng.} I bet they'll take the river rather than the forest.~\LINK{http://forum.learnnavi.org/language-updates/concerning-tup/}{ln\slash g}

% tupe
\hi
\HT{tupe}{\B{\cp{tu}pe}} \emph{or} \HL{pesu}{\B{pe\cp{su}}} \I{["tu.pE]} \emph{or} \I{[pE."{}su]} \emph{inter.} who? what person?
\B{Kal\-txì. Nge\-nga lu tu\-pe? -- Oe lu Txe\-wì.} Hello. Who are you? -- I'm Txewì.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Nge\-yä tsa\-fkxi\-let tu\-pel re\-ngo\-lop?} Who designed that necklace of yours?~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tsay\-sam\-si\-yu lu su\-pe? --  (Fo) lu Ka\-mun, Ra\-lu, Ìstaw, sì Ate\-yo.} Who are those warriors? [i.e. identify them] -- They're Ka\-mun, Ra\-lu, Ìstaw, and Ate\-yo.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
(\ding{43}~\HL{-pe}{\B{pe}}\texttt{+}, \HL{tute1}{\B{tu\-te}}) 

% tut
\hi
\HT{tut}{\B{tut}} \I{[tut\textcorner]} \emph{or} \I{[tUt\textcorner]} (\textsc{rn}: \B{tùt} \I{[tUt\textcorner]}) \emph{part. of continuation,} ``And what about \ldots{}?'' 
\B{Oe\-ru syaw fko Ney\-ti\-ri. Nga\-ru tut?} My name is Neytiri. And how about you?~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Nga\-ru lu fpom srak? -- Lu fpom. Nga\-ru tut?} Are you well? -- I'm well. And you?~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Oe zo\-la\-'u ftu Wa\-sying\-ton. Nga\-ri tut?} I'm from Washington. And you?~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Su\-nu nga\-ru fkxen srak? -- Ke\-zem\-pll\-txe, su\-nu oe\-ru nì\-txan. -- Vey tut? -- Ngay\-txoa, ke tsun oe yi\-vom vey\-ti.} Do you like veggie stuff? -- Of course, I like it a lot. -- How about meat? -- Sorry, I can't eat meat.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nItut}{\B{nì\-tut}}, continually.
\on{2}.~\HL{letut}{\B{le\-tut}}, constant, continual.

% tutampe
\hi
\HT{tutampe}{\B{tu\cp{tam}pe}} \emph{or} \HL{pestan}{\B{pe\cp{stan}}} \I{[tu."tam.pe]} \emph{inter.} who? what man?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
(\ding{43}~\HL{-pe}{\B{pe}}\texttt{+}, \HL{tutan}{\B{tu\-tan}}) 

% tutan
\hi
\HT{tutan}{\B{tu\cp{tan}}} \I{[tu."tan]} \emph{n.} male person, man. 
\B{Slä tsran\-ten fra\-to, tsa\-tu\-tan la\-yä\-ngu ye'\-rìn eyk\-tan.} But most important, that man will soon be leader.\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
\B{Ki\-fkey\-ri fì\-tu\-tan\-ìl a\-yay\-mak ke tslam stum ke\-'ut, ul\-te ke new ni\-vu\-me nì\-'ul.} This ignorant man understands almost nothing and doesn't want to learn more about the world.\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
\B{Fkol ftxo\-lä\-ngey na eyk\-tan tu\-tan\-ti a tì\-eyk\-tan\-ìri ke lu pxan kaw\-'it.} A man was chosen as leader who isn't worthy of leadership at all.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{tute1}{\B{tu\-te}}\textsuperscript{\on{1}})

% tute
\hi
\HT{tute1}{\B{\cp{tu}te}}\textsuperscript{\on{1}} \I{["tu.tE]} \emph{n.} person. 
\B{Ke tsun tu\-te a\-'aw tsa\-tskxe\-ti a\-ku'\-up kxi\-vel\-tek nì\-'aw\-tu.} One person alone can't lift that heavy rock.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Ke\-tsran tu\-tel 'i\-vem, tsa\-fne\-tsngan lu ftxì\-vä'.} That kind of meat is gross no matter who cooks it.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Sìl\-pey oe, tsa\-me\-su\-tel a mun\-txa so\-li nge\-yä ay\-lì\-'ut a\-lor tsì\-ye\-ve\-'a.} I hope, those two people who got married, will see your beautiful words.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/comment-page-1/\#comment-2619}{b}
\B{Oel tse\-'a 'a\-'aw\-a tu\-tet.} I see several people.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Ngal oe\-yä 'upxa\-ret ay\-su\-te\-ru fpo\-le'.} You sent my message to people.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki}
\B{Txam\-pxì\-ri su\-te\-yä, ke tsun fko ni\-vu\-me fte nì\-lì'\-fya a\-mip pi\-vll\-txe nìl\-tsan fa fwa tìng mik\-yun ho\-ren\-ur nì\-'aw.} For the majority of people, it's not possible to learn to speak a new language well simply by listening to rules.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Tu\-te\-ri tì\-txur lu txew\-nga'.} There are limits to a person's strength.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Ik\-ran\-ìri krr\-a hu tu\-te tsa\-heyl si, ftang li\-vu yrr.} When an \emph{ikran} bonds with a person, it ceases to be wild.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tì\-'e\-fu\-mì oe\-yä su\-te\-kip nì\-wotx, ftxey Na'\-vi ftxey Saw\-tu\-te, lu sìl\-tsan lu kawng.} In my opinion among all people, whether Na'vi or Skypeople, there are good and bad.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\ding{43}~\HL{-tu}{\B{-tu}}, \HL{-yu}{\B{-yu}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tuteo}{\B{tu\-teo}}, someone, somebody.
\on{2}.~\HL{tutan}{\B{tu\-tan}}, male person. 
\on{3}.~\HL{tute2}{\B{tu\-té}}, female person. 
\on{4}.~\HL{tawtute}{\B{taw\-tu\-te}}, human, skyperson.
\on{5}.~\HL{tutsena}{\B{tut\-sena}}, stretcher. 
\on{6}.~\HL{syeptute}{\B{syep\-tu\-te}}, hyneman.
\on{7}.~\HL{tiretu}{\B{ti\-re\-tu}}, shaman.
\on{8}.~\HL{nawmtu}{\B{nawm\-tu}}, noble person.
\on{9}.~\HL{tsantu}{\B{tsan\-tu}}, good person.
\on{10}.~\HL{kawngtu}{\B{kawng\-tu}}, bad person.
\on{11}.~\HL{matu}{\B{ma\-tu}}, ``excuse me, hey''.

% tuté
\hi
\HT{tute2}{\B{tu\cp{te}}}\textsuperscript{\on{2}} \emph{or} \B{tu\cp{té}} \I{[tu."tE]} \emph{n.} female (person), woman.
\B{Lu po lor\-a tu\-té a\-lor.} She's an extremely beautiful woman.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Fkxi\-le pewn\-ta tu\-té\-yä kur.} The bib necklace hangs from the woman's neck.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{tute1}{\B{tu\-te}}\textsuperscript{\on{1}})

% tuteo
\hi
\HT{tuteo}{\B{\cp{tu}teo}} \I{["tu.tE.o]} \emph{n., pn.} someone, somebody. 
\B{Txo tu\-te\-o ni\-vew hi\-vum, po\-ru pll\-txe san ru\-txe 'i\-vì\-'awn.} If someone wants to leave, tell them to please stay.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Lu su\-te\-o a lu 'e\-wan\-a ay\-nu\-me\-yu, ul\-te kop su\-te\-o a\-la\-he a tok fra\-kll\-pxìl\-tut ki\-fkey\-ä ul\-te ke lu 'e\-wan kaw\-'it.} There are some people who are young students, and also some other people in every territory of the world who are not young at all.~\LINK{http://forum.learnnavi.org/language-updates/ke-kawit/}{ln} 
(\ding{43}~\HL{tute1}{\B{tu\-te}}\textsuperscript{\on{1}}, \HL{-o}{\B{-o}})

% tutepe
\hi
\HT{tutepe}{\B{tu\cp{te}pe}} \emph{or} \HL{peste}{\B{pe\cp{ste}}} \I{[tu."tE.pE]} \emph{inter.} who? what woman?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
(\ding{43}~\HL{-pe}{\B{pe}}\texttt{+}, \HL{tute2}{\B{tu\-te}}\textsuperscript{\on{2}}) 

% tutsena
\hi
\HT{tutsena}{\B{\cp{tut}sena}} \I{["tut\textcorner.sE.na]} \emph{or} \I{[tUt\textcorner.]} \emph{n.} stretcher.
(\ding{43}~\HL{tute1}{\B{tu\-te}}\textsuperscript{\on{1}}, \HL{sAhena}{\B{sä\-he\-na}})

% tuvom
\hi
\HT{tuvom}{\B{tu\cp{vom}}} \I{[tu."vom]} \emph{adj.} greatest of all, exceedingly great.
\B{En\-tu lu tu\-vom\-a ta\-ron\-yu. Kaw\-tut na po ke tso\-le\-'a oel mì sì\-rey.} Entu is an incredible hunter. I've never seen anyone like him before.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Sa'\-sem\-ìri lu 'eveng\-ä kxitx yeng\-wal a\-tu\-vom.} For a parent, the greatest sorrow is the death of a child.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}

% tuvon
\hi
\HT{tuvon}{\B{\cp{tu}von}} \I{["tu.von]} \emph{vin.} lean.
\B{Ton\-ìri a\-la\-he, aw\-ngal yom wu\-tsot teng\-krr he\-reyn nì\-pxim, teng\-krr te\-ru\-von, ke tsran\-ten.} As for all other nights, it doesn't matter whether we eat dinner while sitting upright or leaning.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}

\end{multicols}