\pdfbookmark[0]{TX}{TX}
\begin{center}
\section*{TX \I{[t']} (\emph{TxeTx})}
\end{center}

\begin{multicols}{2}

% txa'
\hi
\HT{txa'}{\B{txa'}} \I{[t'aP]} \emph{adj.} hard. 
\B{Teng\-krr hu pa\-lu\-kan\-tsyìp uvan se\-ri zo\-lìm oel mì sa'\-leng a 'u\-ot a lu txa' sì ekx\-txu.} While playing with my cat I felt something hard and rough on his skin.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\ding{214}~\HL{hewne}{\B{hew\-ne}})

% txakrrfpìl
\hi 
\HT{txakrrfpIl}{\B{txa\cp{krr}fpìl}} \I{[t'a."k\s{r}.fp$\cdot$Il]} \emph{vtr., inf.}\on{33} consider, ponder. 
\B{Oel sä\-mok\-ti nge\-yä txa\-krr\-fpo\-lìl.} I have considered your suggestion.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{krr}{\B{krr}}, \HL{fpIl}{\B{fpìl}}) 

% txal
\hi
\HT{txal}{\B{txal}} \I{[t'al]} \emph{n.} back (of the body).
\B{Sran, sran, oe\-ri pì\-tìk tsa\-tse\-nget a mì tal! Fkxa\-ke nì\-ftxan kum\-a ter\-kup.} Yeah, yeah, scratch that place on my back! It itches like crazy!~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\emph{a. proverb}) \B{Txìm a\-'aw ke tsun hi\-veyn mì tal me\-fa'\-li\-yä.} You can't take both positions or sit on the fence; you need to decide. [lit.: One butt can't sit on the backs of two direhorses]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 

% txampam
\hi 
\HT{txampam}{\B{\cp{txam}pam}} \I{["t'am.pam]} \emph{n., fauna} soundblast colossus. 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{pa}{\B{pam}}) 

% txampay
\hi
\HT{txampay}{\B{txam\cp{pay}}} \I{[t'am."paj]} \emph{n.} sea, ocean. 
\B{'Rr\-ta\-mì a tam\-pay lu ho\-et.} The oceans on Earth are vast.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Lu oer tì\-new a tse\-'a txam\-pay\-it.} I have a desire to see the ocean.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Oe 'o\-long\-okx mì sray a txam\-pay\-ìri sim.} I was born in a town near the ocean.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Ne\-ni lew si fì\-txa\-yor vay txam\-pay nì\-wotx.} Sand covers this expanse all the way to the ocean.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{pay}{\B{pay}})

% txampxì
\hi
\HT{txampxI}{\B{txam\cp{pxì}}} \I{[t'am."p'I]} \emph{n.} majority, most, large part. 
\B{Hìm\-pxì Saw\-tu\-te\-yä lu tstun\-wi, slä fe\-yä txam\-pxì lä\-ngu kawng\-lan.} A minority of the Sky People are kind, but the majority are malicious.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Txam\-pxì\-ri su\-te\-yä, ke tsun fko ni\-vu\-me fte nì\-lì'\-fya a\-mip pi\-vll\-txe nìl\-tsan fa fwa tìng mik\-yun ho\-ren\-ur nì\-'aw.} For the majority of people, it's not possible to learn to speak a new language well simply by listening to rules.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Slä mì tam\-pxì krr\-ä, ke sar fkol kea hem\-lì\-'u\-vit fì\-lì\-'u\-mì.} But most of the time, one doesn't use infixes in this word.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/comment-page-1/\#comment-677}{b}
(\ding{214}~\HL{hImpxI}{\B{hìm\-pxì}}, \ding{43}~\HL{txan}{\B{txan}}, \HL{hapxI}{\B{ha\-pxì}})

% txan
\hi
\HT{txan}{\B{txan}} \I{[t'an]} \emph{adj.} (\emph{a. in quantity}) great, much. 
\B{Hu\-fwa tì\-sraw lu txan, tsun ay\-oe tsat li\-vewn.} Although the pain is great, we can endure it.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
\B{Eo ay\-oeng lu txan\-a tì\-kawng.} A great evil is upon us.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Jake_back_at_hometree}{wiki\slash a}
\B{Txan\-a srung so\-li po oer.} He helped me a lot.~\LINK{http://forum.learnnavi.org/language-updates/si-verbs-modification/}{ln}
\B{Txan\-a ira\-yo.} Much thanks.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b}
\B{Na\-ri si! Kll\-te\-sìn lu pay a\-txan. Fwi rä\-'ä!} Be careful! There’s a lot of water on the ground. Don't slip!~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Slär\-ìl nga' fwep\-it a\-txan.} The cave is very dusty.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} 
(\ding{214}~\HL{hIm}{\B{hìm}})
(\emph{b. of time}) long, much. \B{Yol\-a krr, txan\-a krr, ke tsran\-ten.} It doesn't matter how long it takes.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b\slash a\textsuperscript{c}}
\B{Krr a fol txepit tok lu txan nìtam srak?} Have they been in the fire long enough?~\LINK{http://forum.learnnavi.org/intermediate/translation-exercise-a-navajo-coyote-tale-in-navi/}{ln}
\B{Ke lu oer set krr a\-txan.} I don't have much time now.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/}{ln}
(\ding{214}~\HL{yol}{\B{yol}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{fItxan}{\B{fì\-txan}}, so, this much. 
\on{2}.~\HL{hImtxan}{\B{hìm\-txan}}, quantity.  
\on{3}.~\HL{nIftxan}{\B{nì\-ftxan}}, this much. 
\on{4}.~\HL{nItxan}{\B{nì\-txan}}, very.   
\on{5}.~\HL{txampay}{\B{txam\-pay}}, ocean, sea. 
\on{6}.~\HL{txampxI}{\B{txam\-pxì}}, majority. 
\on{7}.~\HL{txanatan}{\B{txa\-na\-tan}}, bright, vivid. 
\on{8}.~\HL{txanew}{\B{txa\-new}}, greedy. 
\on{9}.~\HL{txanfwerwI}{\B{txan\-fwer\-wì}}, blizzard. 
\on{10}.~\HL{txankrr}{\B{txan\-krr}}, for a long time. 
\on{11}.~\HL{txanro'a}{\B{txan\-ro\-'a}}, be famous. 
\on{12}.~\HL{txantompa}{\B{txan\-tom\-pa}}, rainstorm. 
\on{13}.~\HL{txantur}{\B{txan\-tur}}, powerful. 
\on{14}.~\HL{txantxew}{\B{txan\-txew}}, maximum. 
\on{15}.~\HL{txantsan}{\B{txan\-tsan}}, excellent. 
\on{16}.~\HL{txantslusam}{\B{txan\-tslu\-sam}}, wise. 
\on{17}.~\HL{txantstew}{\B{txan\-tstew}}, hero, heroine. 
\on{18}.~\HL{txanwawe}{\B{txan\-wa\-we}}, significant. 
\on{19}.~\HL{txasom}{\B{txa\-som}}, very hot. 
\on{20}.~\HL{txawew}{\B{txa\-wew}}, very cold. 
\on{21}.~\HL{txanlal}{\B{txan\-lal}}, ancient, very old. 
\on{22}.~\HL{txanlokxe}{\B{txan\-lo\-kxe}}, clan or tribal domain; country.
\on{23}.~\HL{txanso'hayu}{\B{txan\-so'\-ha\-yu}}, fan, enthusiast. 
\on{24}.~\HL{txankeltrrtrr}{\B{txan\-kel\-trr\-trr}}, extraordinary.
\on{25}.~\HL{txansngum}{\B{txan\-sngum}}, desperation.
\on{26}.~\HL{txantsawl}{\B{txan\-tsawl}}, huge, giant.
\on{27}.~\HL{txavA'}{\B{txa\-vä'}}, disgusting.
\on{28}.~\HL{txasunu}{\B{txa\-su\-nu}}, love greatly, enjoy tremendously.
\on{29}.~\HL{txanfwIngtu}{\B{txanfwìngtu}}, bastard, loser (\emph{insult})
\on{30}.~\HL{txantsawtsray}{\B{txan\-tsaw\-tsray}}, large city, metropolis.
\on{31}.~\HL{txantxewm}{\B{txan\-txewm}}, terrifying.
\on{32}.~\HL{txakrrfpIl}{\B{txa\-krr\-fpìl}}, consider, ponder.
\on{33}.~\HL{txampam}{\B{txam\-pam}}, soundblast colossus.

% txanatan
\hi
\HT{txanatan}{\B{\cp{txa}natan}} \I{["t'a.na.tan]} \emph{adj.} bright, vivid.
\B{Lu po\-ru tì\-ron\-srel a\-txa\-na\-tan.} She has a vivid imagination.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{atan}{\B{atan}})

% txanew
\hi
\HT{txanew}{\B{\cp{txa}new}} \I{["t'a.nEw]} \emph{adj.} greedy.
(\ding{43}~\HL{txan}{\B{txan}}, \HL{new}{\B{new}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItxanew}{\B{tì\-txa\-new}}, greed.

% txanfwerwì
\hi
\HT{txanfwerwI}{\B{txan\cp{fwer}wì}} \I{[t'an."fwER.wI]} \emph{n.} blizzard.
(\ding{43}~\HL{txan}{\B{txan}}, \HL{hufwe}{\B{hu\-fwe}}, \HL{herwI}{\B{her\-wì}})

% txanfwìngtu
\hi 
\HT{txanfwIngtu}{\B{txan\cp{fwìng}tu}} \I{[t'an."fwIN.tu]} \emph{n.} bastard, loser (\emph{insult}). 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{fwIng}{\B{fwìng}})

% txankeltrrtrr
\hi 
\HT{txankeltrrtrr}{\B{\cp{txan}kel\cp{trr}trr}} \I{["t'an.kEl."t\s{r}.t\s{r}]} \emph{adj.} extraordinary. 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{keltrrtrr}{\B{kel\-trr\-trr}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nItxankeltrrtrr}{\B{nì\-txan\-kel\-trr\-trr}}, extraordinarily.

% txankrr
\hi
\HT{txankrr}{\B{txan\cp{krr}}} \I{[t'an."k\s{r}]} \emph{adv.} for a long time.
\B{Txan\-krr ngal ke lawk pot kaw\-'it.} You haven't mentioned a thing about him in a long time.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Fì\-sä\-omum txan\-krr lo\-lu oer.} I had this information for a long time.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{krr}{\B{krr}})

% txanlal
\hi
\HT{txanlal}{\B{\cp{txan}lal}} \I{["t'an.lal]}  \emph{adj.} ancient, very old (\emph{not for people}).
\B{Po\-leng ay\-oe\-ru ko\-ak\-tel vur\-it a\-txan\-lal.} The old woman told us an ancient story.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{lal}{\B{lal}})

% txanlokxe
\hi
\HT{txanlokxe}{\B{txanlo\cp{kxe}}} \I{[t'an.lo."k'E]} \emph{n.}  (\emph{a. on Pandora}) clan or tribal domain. 
(\emph{b. on Earth}) country.
\B{Po yaw\-ne lu snor nì\-'aw; fpom txan\-lo\-kxe\-yä ke tsran\-ten.} He only loves himself; the well\hyp{}being of the country doesn't matter.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{olo'}{\B{o\-lo'}}, \HL{atxkxe}{\B{atx\-kxe}})

% txanro'a
\hi
\HT{txanro'a}{\B{txan\cp{ro}'a}} \I{[t'an."R$\cdot$o.P$\cdot$a]} \emph{vin., inf.}\on{23} be famous. 
\B{Fì\-'um\-tsat to\-lìng 'e\-veng\-ur a\-spxin zey\-ko\-yul a txan\-ro\-'a.} This medicine was given to the sick child by a famous healer.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
\B{Vay fwa zo\-la\-'u Tsyeyk\-Su\-li, To\-ruk Mak\-to alu pi\-za\-yu Ney\-ti\-ri\-yä txan\-rar\-mo\-'a fra\-to kip ay\-ha\-pxì\-tu Oma\-ti\-ka\-yaä.} Until Jake Sully arrived, Neytiri's ancestor was the most famous Toruk Makto among the Omatikaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{ro'a}{\B{ro\-'a}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{txanro'tu}{\B{txan\-ro'\-tu}}, famous person, celebrity.
\on{2}.\HL{tItxanro'a}{\B{tì\-txan\-ro\-'a}}, fame, glory.

% txanro'tu
\hi 
\HT{txanro'atu}{\B{txan\cp{ro'}tu}} \I{[t'an."RoP.tu]} \emph{n.} famous person, celebrity. 
\B{Krr\-a ul\-txa si nga tsa\-txan\-ro'\-tu\-hu, ze\-rok fu\-ta ze\-ne he\-nga si\-vi, ma 'e\-veng.} When you meet that famous person, remember that you have to be formally polite, child.~\LINK{http://naviteri.org/2022/02/lifyengteri-concerning-honorific-language/}{b} 
(\ding{43}~\HL{txanro'a}{\B{txan\-ro\-'a}}, \HL{tute1}{\B{tu\-te}})

% txansngum / si
\hi 
\HT{txansngum}{\B{txan\cp{sngum}}} \I{[t'an."{}sNum]} \emph{or} \I{[."{}sNUm]} \textsc{i}. \emph{n.} desperation, feeling of great worry.  \\
\textsc{ii}. \B{txan\cp{sngum} si} \I{[t'an."{}sNum si]} \emph{or} \I{[."{}sNum]} \emph{vin.} feel desperate.
\B{Ke lu syu\-ve ul\-te tu\-te a\-pxay txan\-sngum si.} There is no food, and many people are desperate.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{sngum}{\B{sngum}})

% txanso'hayu
\hi 
\HT{txanso'hayu}{\B{txan\cp{so'}hayu}} \I{[t'an."{}soP.ha.ju]} \emph{n.} fan, enthusiast; (\emph{coll. shortened to} \ding{43}~\HL{so'yu}{\B{so'\-yu}}). 
\B{Lu pxay\-a txan\-so'\-ha\-yu tsa\-rel\-ä a\-ru\-sikx alu Unil\-tì\-ran\-tokx ki\-fkey\-ka nì\-wotx.} There are many fans of \emph{Avatar} all over the world.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b} 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{so'ha}{\B{so'\-ha}})

% txantompa
\hi
\HT{txantompa}{\B{txan\cp{tom}pa}} \I{[t'an."tom.pa]} \emph{n.} rainstorm, heavy rain.
\B{Maw txan\-tom\-pa, pxay\-a rìk\-äo la\-mu tska\-lep pe\-yä, ha tsat ku\-lat ay\-oel.} After the rainstorm, his crossbow was under a lot of leaves, so we uncovered it.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{tompa}{\B{tom\-pa}})

% txantur
\hi
\HT{txantur}{\B{\cp{txan}tur}} \I{["t'an.tuR]} \emph{or} \I{[.tUR]} \emph{adj.} powerful.
\B{Tsa\-tu\-tan la\-yä\-ngu ye'\-rìn eyk\-tan a txan\-tur fra\-to 'Rr\-ta\-mì.} That man will soon be the most powerful leader on Earth.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
\B{tu\-té a\-txan\-tur}, a powerful woman.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{txur}{\B{txur}})

% txantxew
\hi
\HT{txantxew}{\B{txan\cp{txew}}} \I{[t'an."t'Ew]} \emph{n.} maximum. 
(\ding{214}~\HL{hImtxew}{\B{hìm\-txew}}, \ding{43}~\HL{txan}{\B{txan}}, \HL{txew}{\B{txew}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txantxewvay}{\B{txan\-txew\-vay}}, maximally.

% txantxewm
\hi 
\HT{txantxewm}{\B{\cp{txan}txewm}} \I{["t'an.t'Ewm]} \emph{adj.} terrifying.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{txewm}{\B{txewm}})

% txantxewvay
\hi
\HT{txantxewvay}{\B{txan\cp{txew}vay}} \I{[t'an."t'Ew.vaj]} \emph{adv.} (\emph{a.}) maximally. 
(\emph{b. used with adj.s and adv.s to mean}) as … as possible.
\B{Tì\-ran nì\-fnu txan\-txew\-vay fte\-ke ay\-ye\-rik\-ìl aw\-nga\-ti sti\-vawm.} Walk as quietly as possible so the hexapedes won't hear us.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b}
\B{Ay\-nga\-ri ftxo\-zä Hä\-lo\-win\-ä li\-vu 'o' sì snew\-sye txan\-txew\-vay.} May your Halloween be as fun and spooky as possible.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b}
(\ding{214}~\HL{hImtxewvay}{\B{hìm\-txew\-vay}}, \ding{43}~\HL{txantxew}{\B{txan\-txew}}, \HL{vay}{\B{vay}})

% txantsan
\hi
\HT{txantsan}{\B{\cp{txan}tsan}} \I{["t'an.\t{ts}an]} \emph{adj.} excellent. 
\B{Tsa\-lì\-'u\-ri alu yo\-ra'\-tu, nge\-yä sä\-pll\-txe\-vi lu txan\-tsan.} As for the word \emph{yo\-ra'\-tu} your comment is excellent.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/comment-page-1/\#comment-1267}{b}
\B{Nge\-yä txan\-tsan\-a tì\-pawm\-ìri nga\-ru sei\-yi oe ira\-yo.} I thank you for your excellent question.~\LINK{http://forum.learnnavi.org/language-updates/another-email-from-frommer/}{ln}
\B{Rel\-tseo\-tu a\-txan\-tsan lu nga!} You are an excellent artist.~\LINK{http://forum.learnnavi.org/language-updates/art-related-vocab/}{ln}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{sIltsan}{\B{sìl\-tsan}})

% txantsawl
\hi 
\HT{txantsawl}{\B{\cp{txan}tsawl}} \I{["t'an.\t{ts}awl]} \emph{adj.} giant, huge. 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{tsawl}{\B{tsawl}})

% txantsawtsray
\hi 
\HT{txantsawtsray}{\B{txan\cp{tsaw}tsray}} \I{[t'an."\t{ts}aw.\t{ts}Raj]} \emph{n.} large city, metropolis.
\B{Slä set tä\-ngok kop tan\-tsaw\-tsray\-ti ay\-oe\-yä 'uol a\-la\-he: tsak\-tap.} But now, something else is also at our large cities: violence.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{tsawtsray}{\B{tsaw\-tsray}} )

% txantslusam 
\hi
\HT{txantslusam}{\B{\cp{txan}tslusam}} \I{["t'an.\t{ts}lu.sam]} \emph{adj.} wise, much\hyp{}knowing.   
\B{Lu nga win sì txur | Lu nga txan\-tslu\-sam.} You are fast and strong | You are wise.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Lu fra\-eyk\-tan\-ur a\-sìl\-tsan txan\-tslu\-sam\-a ay\-mo\-war\-si\-yu.} Every good leader has wise advisors.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{tslam}{\B{tslam}}, \HL{lafyon}{\B{la\-fyon}}, \HL{hafyonga'}{\B{ha\-fyo\-nga'}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItxantslusam}{\B{tì\-txan\-tslu\-sam}}, wisdom.

% txantstew
\hi
\HT{txantstew}{\B{\cp{txan}tstew}} \I{["t'an.\t{ts}tEw]} \emph{n.} hero, heroine. 
\B{Txan\-tstew sä\-ro\-'a si, fna\-we'\-tu ke si.} A hero does great deeds, a coward does not.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{tstew}{\B{tstew}})

% txanwawe
\hi
\HT{txanwawe}{\B{txanwa\cp{we}}} \I{[t'an.wa."wE]} \emph{adj.} personally meaningful, significant. 
\B{Kì\-rey\-sì\-ri lu tsa\-puk\-ä ay\-vur a te\-ri 'Rr\-ta txan\-wa\-we nì\-ngay, slä oe\-ri, hu\-fwa el\-tur tì\-txen si, lu ral\-nga' nì\-'aw.} For Grace, the stories in that book relating to Earth are personally meaningful, but for me, although interesting, they’re simply instructive.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Na\-ri si!  Tsa\-tsko lu spu\-win ul\-te ke lu mi txur, slä oe\-ri txan\-wa\-we le\-iu. Lu stxe\-li a sem\-pul\-ta.} Careful! That bow is old and no longer strong, but it means a lot to me. It was a gift from my father.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{wawe}{\B{wa\-we}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txanwetseng}{\B{txan\-we\-tseng}}, personally significant or beloved place, \emph{heimat}.

% txanwetseng
\hi 
\HT{txanwetseng}{\B{txan\cp{we}tseng}} \I{[t'an."wE.\t{ts}EN]} \emph{n.} personally significant or beloved place, \emph{heimat}. 
(\ding{43}~\HL{txanwawe}{\B{txan\-wa\-we}}, \HL{tseng}{\B{tseng}})

% txap
\hi 
\HT{txap}{\B{txap}} \I{[t'ap\textcorner]} \emph{vtr.} press, press on, apply pressure to. 
\B{Txap skxir\-it fte\-ke rey\-pay wrr\-zi\-va\-'u.} Apply pressure to the wound so that the blood won't flow.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItxap}{\B{tì\-txap}}, pressure. 
\on{2}.~\HL{pxawtxap}{\B{pxaw\-txap}}, squeeze.

% txasom
\hi
\HT{txasom}{\B{\cp{txa}som}} \I{["t'a.som]} \emph{adj.} very hot. 
\B{Ya\-ri som\-wew\-pe (set)?---Ya lu txa\-som.} What's the temperature?---It's very hot.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
(\ding{214}~\HL{txawew}{\B{txa\-wew}}, \ding{43}~\HL{txan}{\B{txan}}, \HL{som}{\B{som}})

% txasunu
\hi 
\HT{txasunu}{\B{txa\cp{su}nu}} \I{[t'a."{}su.nu]} \emph{vin.} love greatly, enjoy tremendously. 
\B{Txa\-su\-nu oe\-ru tey\-lu!} I really love \emph{teylu}!~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
\B{Txa\-su\-nu nga\-ru oey way\-te\-lem, ke\-fyak?} You love my songcord, don't you?~\LINK{https://www.youtube.com/watch?v=13aPkyXDuEE}{yt\slash fop}
(\ding{43}~\HL{txan}{\B{txan}}, \HL{sunu}{\B{su\-nu}})

% txatx
\hi
\HT{txatx}{\B{txatx}} \I{["t'at']} \emph{n.} bubble.
\B{Yo\-sìn kil\-van\-ä lu tatx.} There are bubbles on the surface of the river.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}

% txavä'
\hi 
\HT{txavA'}{\B{txa\cp{vä'}}} \I{[t'a."v\ae{}P]} \emph{adj.} disgusting. 
\B{Lu tsa\-kem txa\-vä', ma tsmuk.} That's disgusting, bro.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{txan}{\B{txan}}, \HL{vA'}{\B{vä'}})

% txawew
\hi
\HT{txawew}{\B{\cp{txa}wew}} \I{["t'a.wEw]} \emph{adj.} very cold. 
\B{Ya\-ri som\-wew\-pe (set)?---Ya lu txa\-wew.} What's the temperature?---It's very cold.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Mì zì\-sì\-krr a\-txa\-wew slu ay\-ora leyr.} In the very cold season, the lakes freeze.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
(\ding{214}~\HL{txasom}{\B{txa\-som}}, \ding{43}~\HL{txan}{\B{txan}}, \HL{wew}{\B{wew}})

% txayo
\hi
\HT{txayo}{\B{\cp{txa}yo}} \I{["t'a.jo]} \emph{n.} field, plain, open terrain. 
\B{Oe lu Va'\-ru a ftu txa\-yo zo\-la\-'u.} I'm Va'ru from the plains.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Ne\-ni lew si fì\-txa\-yor vay txam\-pay nì\-wotx.} Sand covers this expanse all the way to the ocean.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Maw sä\-tsway\-on a\-yol ay\-oe kll\-po\-lä mì ta\-yo a lu ro\-fa kil\-van.} After a short flight we landed in a field beside the river.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{txay}{\B{txay}}, \HL{yo}{\B{yo}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{meitayo}{\B{mei\-ta\-yo}}, wetlands.

% txaw
\hi 
\HT{txaw}{\B{txaw}} \I{[t'aw]} \emph{vtr.} punish. 
\B{Sem\-pul\-ìl a\-sìl\-tsan sney eveng\-it ke txaw nì\-ka\-kan.} A good father doesn't punish his children roughly.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tItxaw}{\B{tì\-txaw}}, punishment.

% txawnulsrung
\hi 
\HT{txawnulsrung}{\B{txaw\cp{nul}srung}} \I{[t'aw."nul.sRuN]} \emph{or} \I{[."nUl.sRUN]} (\textsc{rn}: \B{txaw\cp{nùl}srùng} \I{[daw."nUl.sRUN]}) \emph{n.} machine. \\
(i.) \B{txaw\-\cp{nul}\-srung a yur}, washing machine.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b} \\
(ii.) \B{txaw\-\cp{nul}\-srung a tsway\-on}, airplane.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b} 
(\ding{43}~\HL{txula}{\B{txu\-la}}, \HL{sAsrung}{\B{sä\-srung}})

% txay
\hi
\HT{txay}{\B{txay}} \I{[t'aj]} \emph{vin.} be horizontal, lie flat. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nItxay}{\B{nì\-txay}}, horizontally. 
\on{2}.~\HL{klltxay}{\B{kll\-txay}}, lie on the ground. 
\on{3}.~\HL{txayo}{\B{txa\-yo}}, field, plain.

% txärem
\hi
\HT{txArem}{\B{\cp{txä}rem}} \I{[t'\ae.REm]} \emph{n.} bone.
\B{neyn na txä\-rem}, the light color of bone.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} 
\B{Tì\-kawng a su\-tel ngop var ri\-vey, tsìl\-tsan\-it pxìm kll\-yem fkol fe\-yä tä\-rem\-hu.} The evil that men do lives after them, The good is oft interred with their bones [\emph{Julius Caesar, Act III, Sc. 2}].~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{snatxArem}{\B{sna\-txä\-rem}}, skeleton. 
\on{2}.~\HL{txArpawk}{\B{txär\-pawk}}, Pa\-lu\-lu\-kan Bone Horn.

% txärpawk
\hi
\HT{txArpawk}{\B{\cp{txär}pawk}} \I{["t'\ae{}R.pawk\textcorner]} \emph{n.} Pa\-lu\-lu\-kan Bone Horn.
(\ding{43}~\HL{txArem}{\B{txärem}}, \HL{pawk}{\B{pawk}})

% txe'lan
\hi
\HT{txe'lan}{\B{txe'\cp{lan}}} \I{[t'EP."lan]} \emph{n.} heart. 
\B{Oe\-yä tuk\-rul txe'\-lan\-it ti\-va\-kuk | Oe\-ri tì\-ngay\-ìl txe'\-lan\-it ti\-va\-kuk | Oe\-yä txe'\-lan li\-vu ngay.} Let my spear strike the heart | Let the ruth strike my heart | Let my heart be true.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Ay\-nga\-ri teng\-krr ya wur sle\-ru nì\-'ul, sìl\-pey oe, li\-vu hel\-ku sang ul\-te te'\-lan le\-fpom.} As for you all, while it's getting cooler, I hope your homes may be warm and your hearts happy.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} 
\B{Ay\-lì\-'u na ay\-skxe mì te'\-lan.} The words (are) like stones in (my) heart.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{ln\slash a\textsuperscript{c}}
\B{Sì 'e\-kong te'\-lan\-ä.} And the beat of the hearts.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
(\emph{a. idiom}) \B{Nga\-ri txe'\-lan ma\-wey li\-vu.\slash Txe'\-lan ma\-wey.} ``Don't worry (about it).''~\LINK{http://forum.learnnavi.org/language-updates/fmawno/}{ln} 
(\emph{b. idiom}) \B{txe'\-lan\-ti wrr\-zä\-rìp} \emph{vtr.} to greatly move emotionally.
\B{Oe\-ri pe\-yä ay\-lì\-'ul txe'\-lan\-ti wrr\-zo\-lä\-rìp.} Her words moved me greatly.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{kawnglan}{\B{kawng\-lan}}, malicious, bad hearted. 
\on{2}.~\HL{txe'lankong}{\B{txe'\-lan\-kong}}, heartbeat.
\on{3}.~\HL{tstunlan}{\B{tstun\-lan}}, kind\hyp{}hearted.
\on{4}.~\HL{txeylan}{\B{txey\-lan}}, best friend.

% txe'lankong
\hi 
\HT{txe'lankong}{\B{txe'\cp{lan}kong}} \I{[t'E."lan.koN]} \emph{n.} heartbeat. 
(\ding{43}~\HL{txe'lan}{\B{txe'\-lan}}, \HL{'ekong}{\B{'ekong}})

% txekxumpay
\hi
\HT{txekxumpay}{\B{txe\cp{kxum}pay}} \I{[t'E."k'um.paj]} \emph{or} \I{[."k'Um.]} (\textsc{rn}: \B{txe\cp{kxùm}pay} \I{[dE."gUm.paj]}) \emph{n.} magma, lava.
\B{Txep\-ram pxor a krr, txan\-a txe\-kxum\-pay wrr\-za\-'u.} When a volcano erupts, a lot of lava comes out.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\ding{43}~\HL{txep}{\B{txep}}, \HL{kxumpay}{\B{kxum\-pay}})

% txele
\hi
\HT{txele}{\B{\cp{txe}le}} \I{["t'E.lE]} \emph{n.} matter (subject), topic.  
\B{Tsa\-hìk\-ur txe\-le lu.} This is a matter for the Tsahik.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Neytiri_talking_to_Tsu.27tey}{wiki\slash a}
\B{Lu aw\-ngar ay\-te\-le a\-pxay a te\-ri sa\-'u pi\-vll\-txe.} We have many matters to talk about.~\LINK{http://forum.learnnavi.org/language-updates/txantsana-pxesisung-kefyak/}{ln}
\B{So\-rewn\-ìl kan\-'ìn tì\-'em\-it nì\-txan ul\-te kxawm tsa\-txe\-le me\-nga\-ne za\-'a\-tsu nì\-'eng.} Sorewn is really into cooking and you two might share that in common.\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Nge\-yä te\-ri fay\-te\-le a ay\-sä\-nu\-me\-ri ngar ira\-yo sei\-yi ay\-oe nì\-wotx.} We all thank you for your teachings about these topics.~\LINK{http://forum.learnnavi.org/language-updates/combined-answers-1-\%28feb-16\%29/}{ln}
\B{Fì\-txe\-le\-ri tì\-mok ke tam; ze\-ne fko fngi\-vo'.} In this matter, suggesting won't cut it; you need to demand.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Fì\-txe\-le\-ri tì\-nui ke lu tì\-ftxey.} In this matter, failure is not an option.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{mIftxele}{\B{mì\-ftxe\-le}}, by the way. 
\on{2}.~\HL{mawftxele}{\B{maw\-ftxe\-le}}, belatedly, in hindsight.

% txen
\hi
\HT{txen}{\B{txen}} \I{[t'En]} \emph{adj.} awake.
\B{Nga lum\-pe lu txen nì\-mun, ma 'e\-vi?} Why are you awake again, my child?~\LINK{https://www.youtube.com/watch?v=13aPkyXDuEE}{yt\slash fop}
(\ding{214}~\HL{ngeyn}{\B{ngeyn}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItxen}{\B{tì\-txen}}, awakeness.

% txep
\hi
\HT{txep}{\B{txep}} \I{[t'Ep\textcorner]} \emph{n.} fire. 
\B{Txep a\-hì\-'i mì tep\-tseng par\-ma\-lon.} A little fire was burning in the fireplace.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b}
\B{Txep\-ìl tìng le\-fpom\-a sä\-nrr\-ti.} The fire gives a pleasant glow.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
\B{Txep\-ìri, yrr\-a rìn\-ti rä\-'ä sar, ma skxawng.} Don't use that green wood for a fire, you fool.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
\B{Maw\-krr\-a ngal txep\-it to\-long tsun hi\-vum.} After you've put out the fire you can leave.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txepIva}{\B{txe\-pì\-va}}, ash, cinder. 
\on{2}.~\HL{txeptseng}{\B{txep\-tseng}}, fire pit. 
\on{3}.~\HL{txepvi}{\B{txep\-vi}}, spark. 
\on{4}.~\HL{ylltxep}{\B{yll\-txep}}, communal fire. 
\on{5}.~\HL{txepram}{\B{txep\-ram}}, volcano. 
\on{6}.~\HL{txekxumpay}{\B{txe\-kxum\-pay}}, magma, lava.
\on{7}.~\HL{vrrtep}{\B{vrr\-tep}}, demon. 
\on{8}.~\HL{txeptun}{\B{txep\-tun}}, orange.
\on{9}.~\HL{txeptsyIp}{\B{txep\-tsyìp}}, flame.
\on{10}.~\HL{txeptsyal}{\B{txep\-tsyal}}, coronis.

% txepìva
\hi
\HT{txepIva}{\B{\cp{txe}pìva}} \I{["t'E.pI.va]} \emph{n.} ash, cinder.
(\ding{43}~\HL{txep}{\B{txep}}, \HL{Ilva}{\B{ìl\-va}})

% txepram
\hi
\HT{txepram}{\B{txep\cp{ram}}} \I{[t'Ep\textcorner."Ram]} \emph{n.} volcano.
\B{Txep\-ram pxor a krr, txan\-a txe\-kxum\-pay wrr\-za\-'u.} When a volcano erupts, a lot of lava comes out.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\ding{43}~\HL{txep}{\B{txep}}, \HL{ram}{\B{ram}})

% txeptun
\hi 
\HT{txeptun}{\B{\cp{txep}tun}} \I{["t'Ep\textcorner.tun]} \emph{or} \I{[.tUn]} \emph{adj.} orange. 
(\ding{43}~\HL{txep}{\B{txep}}, \HL{tun}{\B{tun}})

% txeptseng
\hi
\HT{txeptseng}{\B{\cp{txep}tseng}} \I{["t'Ep\textcorner.\t{ts}EN]} \emph{n.} place where a fire is burning or has burned.   
\B{Tsa\-txep\-tseng\-mì lä\-ngu ay\-ut\-ral a\-ke\-ru\-sey nì\-'aw.} Sadly, there are only dead trees where the fire has been.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{txep}{\B{txep}}, \HL{tseng}{\B{tse\-nge}})

% txeptsyal
\hi 
\HT{txeptsyal}{\B{\cp{txep}tsyal}} \I{["t'Ep\textcorner.\t{ts}jal]} \emph{n., fauna} coronis [lit.: `firewing']. 
(\ding{43}~\HL{txep}{\B{txep}}, \HL{tsyal}{\B{tsyal}})

% txeptsyìp 
\hi 
\HT{txeptsyIp}{\B{\cp{txep}tsyìp}} \I{["t'Ep'.\t{ts}jIp\textcorner]} \emph{n.} flame. 
\B{Tong txep\-tsyìp\-it.} Extinguish the flame.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
\B{Lo\-nu\-sye tong txep\-tsyìp\-it.} Blow out the flame.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
(\ding{43}~\HL{txep}{\B{txep}}) 

% txepvi
\hi
\HT{txepvi}{\B{\cp{txep}vi}} \I{["t'Ep\textcorner.vi]} \emph{n.} spark. 
(\ding{43}~\HL{txep}{\B{txep}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{txepvispxam}{\B{txep\-vi\-spxam}}, sparkle pod. 

% txepvispxam
\hi 
\HT{txepvispxam}{\B{\cp{txep}vispxam}} \I{["t'Ep\textcorner.vi.sp'am]} \emph{n., funga} sparkle pod. 
(\ding{43}~\HL{txepvi}{\B{txep\-vi}}, \HL{spxam}{\B{spxam}}) 

% txew
\hi
\HT{txew}{\B{txew}} \I{[t'Ew]} \emph{n.} edge, brink, limit, border, end.
\B{Sra\-ke pol la\-yok txew\-ti na'\-rìng\-ä?} Will he approach the edge of the forest?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Me\-tew\-it fì\-swek\-ä ey\-ki\-vil.} Pull the two ends of this bar together.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{Ik\-ran ya\-wo\-lo ftu txew 'awkx\-ä.} The banshee took to the air from the edge of a cliff.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}  
(\emph{a. idiom}) \B{Ke tsun aw\-nga pi\-vey nul\-krr--txew lok.} We can't wait any longer--time is almost up.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txewluke}{\B{txew\-lu\-ke}}, endless, boundless. 
\on{2}.~\HL{txantxew}{\B{txan\-txew}}, maximum. 
\on{3}.~\HL{hImtxew}{\B{hìm\-txew}}, minimum. 
\on{4}.~\HL{txewnga'}{\B{txew\-nga'}}, finite. 
\on{5}.~\HL{tawtxew}{\B{taw\-txew}}, horizon, skyline.

% txewì
\hi
\B{\cp{Txe}wì} \I{["t'E.wI]} \emph{n. male name.} 
\B{Oe\-ru syaw (fko) Txe\-wì.} My name is Txe\-wì.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Txe\-wìl 'o\-la\-ku fwep\-it ftum\-fa kel\-ku sne\-yä.} Txewì removed the dust from inside his house.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Su\-nu Txe\-wì\-ru tok\-tor Kì\-rey\-sì.} Txe\-wì likes doctor Grace.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Ni\-nat\-ìri tì\-ru\-sol Txe\-wì\-yä lu vä\-pam.} To Ni\-nat, Txe\-wì's singing is noise.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}

% txewk
\hi
\HT{txewk}{\B{txewk}} \I{[t'Ewk\textcorner]} \emph{n.} \ding{246} club (\emph{weapon}). 

% txewluke
\hi
\HT{txewluke}{\B{\cp{txew}luke}} \I{["t'Ew.lu.kE]} \emph{adj.} endless, boundless, without limit (\emph{for uncountable nouns}).
\B{Spaw Na'\-vil fu\-ta tì\-yawn Ey\-wa\-yä lu txew\-lu\-ke.} The Na'vi believe that Eywa's love is boundless.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b} 
(\ding{43}~\HL{ketsuktiam}{\B{ke\-tsuk\-ti\-am}})

% txewm
\hi
\HT{txewm}{\B{txewm}} \I{[t'Ewm]} \emph{adj.} scary, frightening.   
\B{Slä ma sa'\-nu, ik\-ran txewm lu! Oe txo\-pu si!} But Mommy, the banshee is scary! I'm afraid!~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txantxewm}{\B{txan\-txewm}}, terrifying.

% txewnga'
\hi
\HT{txenga'}{\B{\cp{txew}nga'}} \I{["t'Ew.NaP]} \emph{adj.} having a limit, not without bounds, finite.
\B{Tu\-te\-ri tì\-txur lu txew\-nga'.} There are limits to a person's strength.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{txew}{\B{txew}})

% txey
\hi
\HT{txey}{\B{txey}} \I{[t'Ej]} \emph{vin.} halt (\emph{usually used in military context}). 

% txeylan
\hi 
\HT{txeylan}{\B{\cp{txey}lan}} \I{["t'Ej.lan]} \emph{n.} best friend. 
(\ding{43}~\HL{txe'lan}{\B{txe'\-lan}}, \HL{'eylan}{\B{'ey\-lan}}) 

% txeym
\hi 
\HT{txeym}{\B{txeym}} \I{[t'Ejm]} \emph{adj.} interested (\emph{used with} \ding{43}~\HL{'efu}{\B{'e\-fu}}). 
\B{Tsa\-vur\-ìri oe 'e\-fu txeym.} I'm interested in that story.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tItxeym}{\B{tì\-txeym}}, interest.

% txi
\hi
\HT{txi}{\B{txi}} \I{[t'i]} \emph{n.} hurry, hurriedness, frenzy. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letxi}{\B{le\-txi}}, hurried, frenzied. 
\on{2}.~\HL{nItxi}{\B{nì\-txi}}, hurriedly, in a frenzied way. 
\on{3}.~\HL{letxiluke}{\B{le\-txi\-lu\-ke}}, unhurried. 
\on{4}.~\HL{nItxiluke}{\B{nì\-txi\-lu\-ke}}, unhurriedly, leisurely

% txikx
\hi 
\HT{txikx}{\B{txikx}} \I{[t'ik']} \emph{vtr.} chew. 

% Txilte
\hi
\B{\cp{Txil}te} \I{["t'il.tE]} \emph{n. female name}
\B{Txil\-te Ri\-ni\-sì tä\-pa\-re fì\-tsap nì\-soaia, slä tsal\-su\-ngay ke nì\-olo' ta\-krr\-a Ri\-ni mun\-txa slo\-lu.} Txilte and Rini are related by blood, but nevertheless not by clan since the time Rini got married.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}

% txim
\hi 
\HT{txim}{\B{txim}} \I{[t'im]} \emph{n.} spike, thorn of a plant. 
(\emph{a. proverbial expression}) \B{na txim a txìm\-mì}, s.t. extremely annoying [lit.: `like a thorn in the butt'].~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}  \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{naritxim}{\B{na\-ri\-txim}}, eyethorn.

% txin
\hi
\HT{txin}{\B{txin}} \I{[t'in]} \emph{adj.} main, primary. 
\B{Na'\-vi\-ri txin\-a sä\-'o tì\-tu\-sa\-ron\-ä lu tsko swi\-zaw.} For the Na'vi the bow and arrow is the main hunting tool.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Hu\-fwa lu oe fì\-trr kar\-yu a\-txin, le\-iu oer kop pxe\-srung\-si\-yu a fì\-lì'\-fya\-ri slo\-lu tsul\-fä\-tu.} Although I'm the main teacher today, I'm happy to say I also have three assistants who have become masters of the language.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txintIn}{\B{txin\-tìn}}, occupation. 
\on{2}.~\HL{txintseng}{\B{txin\-tseng}}, base of operations.
\on{3}.~\HL{txinfpIl}{\B{txin\-fpìl}}, main point.

% txinfpìl
\hi 
\HT{txinfpIl}{\B{\cp{txin}fpìl}} \I{["t'in.fpIl]} \emph{n.} main point (\emph{the primary idea or thesis statement of a presentation or argument}). 
\B{Oel nge\-yä txin\-fpìl\-it mi ke tslam.} I still don't understand your main point.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(\ding{43}~\HL{txin}{\B{txin}}, \HL{sAfpIl}{\B{sä\-fpìl}}; \HL{ngrrfpIl}{\B{ngrr\-fpìl}}) 

% txintìn
\hi
\HT{txintIn}{\B{\cp{txin}tìn}} \I{["t'in.tIn]} \emph{n.} occupation, primary role in society.   
\B{Nga\-ru lu pe\-fne\-txin\-tìn nì\-trr\-trr?} What is your primary role (in society)?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Oe\-yä txin\-tìn lu fwa stä'\-nì fay\-oang\-it.} My central societal occupation is to catch fish.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\ding{43}~\HL{txin}{\B{txin}}, \HL{tIn}{\B{tìn}})

% txintseng
\hi 
\HT{txintseng}{\B{\cp{txin}tseng}} \I{["t'in.\t{ts}EN]} \emph{n.} base of operations. 
(\ding{43}~\HL{txin}{\B{txin}}, \HL{tseng}{\B{tseng}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txintseng sawtuteyA}{\B{Txin\-tseng Saw\-tu\-te\-yä}}, Hell's Gate.

% Txintseng Sawtuteyä
\hi 
\HT{txintseng sawtuteyA}{\B{\cp{Txin}tseng \cp{Saw}tuteyä}} \I{["t'in.\t{ts}EN "{}saw.tu.tE.j\ae]} \emph{prop. n.} Hell's Gate (\emph{the main human base on Pandora}). 
(\ding{43}~\HL{txintseng}{\B{txin\-tseng}}, \HL{tawtute}{\B{taw\-tu\-te}})

% txìm
\hi
\HT{txIm}{\B{txìm}} \I{[t'Im]} \emph{n.} butt, ass, rear end (of the body). 
\B{Ean\-a txìm a\-tsawl.} Big, blue butt.~\LINK{http://www.pbs.org/wgbh/nova/blogs/secretlife/entertainment/paul-frommer/}{nova}
(\emph{a. proverb}) \B{Txìm a\-'aw ke tsun hi\-veyn mì tal me\-fa'\-li\-yä.} You can't take both positions or sit on the fence; you need to decide. [lit.: One butt can't sit on the backs of two direhorses]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 
(\emph{b. proverbial expression}) \B{na txim a txìm\-mì}, s.t. extremely annoying [lit.: `like a thorn in the butt'].~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txImya}{\B{txìm\-ya}}, fart.

% txìmya / si
\hi 
\HT{txImya}{\B{\cp{txìm}ya}} \I{["t'Im.ja]} \textsc{i}. \emph{n.} fart, gas emitted from the anus.~\LINK{https://forum.learnnavi.org/language-updates/new-words-fart-to-fart/}{ln} \\
\textsc{ii}. \B{txìm\-ya si} \I{["t'Im.ja si]} \emph{vin.} to fart.
(\ding{43}~\HL{txIm}{\B{txìm}}, \HL{ya}{\B{ya}})

% txìng
\hi
\HT{txIng}{\B{txìng}} \I{[t'IN]} \emph{vtr.} leave, abandon. 
\B{Txìng ay\-ioang\-it! Hum ne wrr\-pa! Tul!} Leave the animals! Get outside! Run!~\LINK{http://forum.learnnavi.org/language-updates/txing-and-hum/}{ln\slash a\textsuperscript{c}} 
\B{Swey\-lu txo ngal ke txi\-vìng sä\-fpìl\-it le\-tsim.} You shouldn't abandon your original idea.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}

% txll'u
\hi
\HT{txll'u}{\B{\cp{txll}'u}} \I{["t'\s{l}.Pu]} \emph{n.}, \ding{96} hookagourd, \emph{flaska ascendens}. 

% txo
\hi
\HT{txo}{\B{txo}} \I{[t'o]} \emph{conj.} if (\emph{conditional}), \emph{optionally together with} \B{tsa\-krr}, then, \emph{or the imperative}.
(\emph{a. factual conditions})
\B{Txo new nga ri\-vey, oe\-hu!} Come with me, if you want to live!~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#UGO_Movie_Blog_.28Twitter_Questions.29}{wiki}
\B{Tì\-tu\-sa\-ron\-ìri txo new fko sli\-vu tsul\-fä\-tu, ze\-ne smar\-to li\-vu wa\-lak.} If you want to become a master hunter, you have to be more active than your prey.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Txo new nga tsli\-vam, ze\-ne ki\-van\-fpìl.} If you want to understand, you have to concentrate.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\emph{b. hypothetical conditions}) 
\B{Pxan li\-vu txo nì\-'aw oe nga\-ri | Tsa\-krr nga Na'\-vi\-ru yom\-tì\-yìng.} Only if I am worthy of you | Will you feed the People.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Txo kxey\-ey\-ti ay\-ngal tsi\-ve\-'a, ru\-txe oe\-ru pi\-veng.} If you see a mistake, please tell me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Txo ke tsi\-ye\-vun oe tì\-ftang si\-vi for.}  If I should not succeed in stopping them.~\LINK{http://forum.learnnavi.org/language-updates/stop!/}{ln}
\B{Txo fkol ke fyi\-vel u\-ran\-it pay\-wä, ze\-ne fko sli\-ve\-le.} If one does not seal a boat against water, one must swim.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{c. with} \B{sweylu} \emph{and} ‹\B{iv}› \emph{for s.t. that hasn't yet happened}) should.
\B{Fì\-na\-er ftxì\-vä' lu nì\-hawng, ha swey\-lu txo ngal tsat kxi\-vukx nì\-win.} This drink tastes horrible, so you'd better swallow it quickly.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Swey\-lu txo nga ki\-vä.\slash Nga swey\-lu txo ki\-vä.} You should go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Swey\-lu txo nga ke ki\-vä.\slash Nga swey\-lu txo ke ki\-vä.} You shouldn't go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Swey\-lu txo ki\-vä.} I\slash you\slash she\slash one\slash etc. should go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
(\emph{d. proverb}) \B{Txo ke nì\-yo' tsa\-krr nì\-yol.} ``If you can't be flawless, at least be brief.''~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
(\ding{43}~\B{zun}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txokefyaw}{\B{txo\-ke\-fyaw}}, if not, or else, otherwise.
\on{2}.~\HL{mungwrrtxo}{\B{mung\-wrr\-txo}}, unless, except if.
\on{3}.~\HL{txotsafya}{\B{txo\-tsa\-fya}}, if that's the case.

% txoa
\hi
\HT{txoa}{\B{\cp{txo}a}} \I{["t'o.a]} \emph{n.} forgiveness. 
\B{Säm\-yam\-ìl po\-ru wa\-yìn\-txu fu\-ta nga\-ta lo\-lu li txo\-a.} A hug will show him that you've already forgiven him.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Tì\-'al lu zop\-lo a tsa\-ri ke tsun txoa li\-vu nì\-ftue.} Wastlefulness is an offense that cannot easily be forgiven.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} 
\B{Tu\-te a ke tsun txoa\-ti ti\-vìng, ke tsun tì\-yawn\-it ti\-vìng nì\-teng.} A person who cannot forgive, cannot love either.~\LINK{http://forum.learnnavi.org/navi-customs-and-culture/navi-proverbs/msg582576/\#msg582576}{ln}
(\emph{a. conv. phrase}) \B{Oe\-ru txoa li\-vu.} ``Forgive me.'' [lit.: may I have forgiveness] 
\B{Oe\-ru txoa li\-vu, ma 'ey\-lan. Rä\-'ä sti\-vi. Lu hì\-'i\-a sä\-'i\-pu nì\-'aw.} I'm sorry, friend. Don't be angry. It was just a small bit of humor.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{hItxoa}{\B{hì\-txo\-a}}, excuse me, pardon me. 
\on{2}.~\HL{ngaytxoa}{\B{ngay\-txo\-a}}, forgive me, sorry. 

% txokefyaw
\hi
\HT{txokefyaw}{\B{\cp{txo}kefyaw}} \I{["t'o.kE.fjaw]} \emph{conj.} if not, or else, otherwise. 
\B{Kll\-te lu ekx\-txu. Na\-ri si txo\-ke\-fyaw tì\-ran nì\-nu.} The ground is rough. Be careful or you'll trip.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Txo\-ke\-fyaw lì\-ye\-vu kxawm 'a\-'aw\-a tu\-te a pll\-txe san po tì\-kxey so\-li.} Otherwise, there may be several people who say, ``He messed up.''~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1951}{b}
(\ding{43}~\HL{txo}{\B{txo}}, \HL{fya'o}{\B{fya'o}}) 

% txolar
\hi
\HT{txolar}{\B{\cp{txo}lar}} \I{["t'o.laR]} \emph{n.}, \textsc{lw} dollar (\emph{currency}).
\B{Po\-el haw\-re'\-tsyìp\-it ko\-la\-nom yoa txo\-lar a\-me\-vol.} She bought a little cap for \$\on{16}.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}

% txon
\hi
\HT{txon}{\B{txon}} \I{[t'on]} \emph{n.}; \emph{adv.} night.   
\B{Tom\-pa\-yä ka\-to | Tsaw\-ke\-yä ka\-to | Trr\-ä sì txon\-ä.} The rhythm of the rain | The rhythm of the sun | Of day and night.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
\B{Fì\-txon na ton a\-la\-he nì\-wotx pe\-lun ke lu teng?} Why is this night not the same like all other nights?~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}
\B{Txon le\-fpom (li\-vu ngar).} (May you have a) Good night.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} \B{Ton\-ìri a\-la\-he, aw\-ngal yom fkxen\-ti le\-rìk nì\-wotx.} As for other nights, we eat leafy vegetable\hyp{}food.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn}  \\
(i.) \B{txo\cp{nam}} \I{[t'o."nam]} \emph{adv., n.} last night. \B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
(ii.) \B{txo\cp{nay}} \I{[t'o."naj]} \emph{adv., n.} tomorrow night. \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{trrtxon}{\B{trr\-txon}}, day night cycle.   
\on{2}.~\HL{txon'ong}{\B{txon\-'ong}}, nightfall.   
\on{3}.~\HL{txonkrr}{\B{txon\-krr}}, at night.   
\on{4}.~\HL{kxamtxon}{\B{kxam\-txon}}, midnight. 

% txon'ong
\hi
\HT{txon'ong}{\B{txon\cp{'ong}}} \I{[t'on."PoN]} \emph{n.}; \emph{adv.} nightfall, sunset, dusk. 
\B{Set sre\-fey oel fu\-ta tsam\-po\-ngu tä\-txaw maw txon\-'ong.} I'm currently expecting the war party back after nightfall.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Trr\-'ong\-ta Txon\-'ong\-vay po to\-lì\-ran.} He walked from dawn until dusk.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\
(i.) \B{txon'o\cp{ngam}} \I{[t'on.Po."Nam]} yesterday's nightfall\slash sunset. \\
(ii.) \B{txon'o\cp{ngay}} \I{[t'on.Po."Naj]} tomorrow's nightfall\slash sunset. \B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay.} Let's meet on the hill tomorrow at nightfall.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\B{txon}, \B{'ong}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sreton'ong}{\B{sre\-ton\-'ong}}, before nightfall. 
\on{2}.~\HL{txon'ongmaw}{\B{txon\-'ong\-maw}}, after nightfall.

% txon'ongmaw
\hi
\HT{txon'ongmaw}{\B{txon\cp{'ong}maw}} \I{[t'on."PoN.maw]} \emph{n.}; \emph{adv.} after nightfall\slash sunset\slash dusk. 
(\ding{43}~\HL{sreton'ong}{\B{sreton'ong}}, \HL{txon'ong}{\B{txon'ong}}, \HL{maw}{\B{maw}})

% txonkrr
\hi
\HT{txonkrr}{\B{\cp{txon}krr}} \I{["t'on.k\s{r}]} \emph{adv.} at night.
\B{Txon\-krr lu syu\-ra\-tan na'\-rìng\-ä Ey\-we\-veng\-ä lor nì\-txan.} At night, the bioluminescence of the Pandoran forest is very beautiful.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Na'\-rìng sngä\-'i ni\-vrr txon\-krr syu\-ra\-tan\-fa.} The forest begins to glow at night with bioluminescence.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{txon}{\B{txon}}, \HL{krr}{\B{krr}})

% txopu
\hi
\HT{txopu}{\B{\cp{txo}pu}} \I{["t'o.pu]} \textsc{i}. \emph{countable n.} fear.
\B{Txo\-pul pe\-yä vll fu\-ta kaw\-krr ke sla\-yu tsam\-si\-yu.} His fear indicates he'll never become a warrior.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
\B{Ru\-txe, ay\-nga\-ri fì\-kem fe\-yä 'e\-'a\-la to\-pur ka\-ngay sì\-yi nì\-'aw.} Please, this will only confirm their worst fears about you.~\LINK{http://forum.learnnavi.org/language-updates/two-words-in-use-eal-and-kangay-si/msg564284/\#msg564284}{ln} \\
\textsc{ii}.a \B{\cp{txo}pu si} \I{["t'o.pu si]} \emph{vin.} be afraid (\emph{with the topic} of). 
\B{Txo\-pu rä\-'ä si, lu ke\-tu\-wong\-o nì\-'aw.} Don't be afraid, it's just an alien.~\LINK{http://forum.learnnavi.org/language-updates/a-collection/}{ln\slash a}    
\B{Slä ma sa'\-nu, ik\-ran txewm lu! Oe txo\-pu si!} But Mommy, the banshee is scary! I'm afraid!~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{Ay\-ha\-pxì\-tu po\-ngu\-ä txo\-pu si nì\-nän ta\-krr\-a Va'\-rul pxe\-ku\-tut lä\-txayn.} The members of the group are less afraid since Va'\-ru defeated three of the enemy.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b} 
\B{Ul\-te zu\-saw\-krr\-ìri txo\-pu si tu\-te a\-pxay.} And many people are afraid of the future.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} \\
\textsc{ii}.b \B{\cp{txo}pu sley\cp{ku}} \I{["t'o.pu slEj."ku]} \emph{vtr.} frighten, scare, produce fear. \B{'Uol ik\-ran\-it txo\-pu sley\-ko\-la\-tsu, ta\-lun\-a po tsìk ya\-wo.} Something must have frightened the banshee, because it suddenly took to the air.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Nge\-yä ay\-lì\-'ul a\-kxap\-nga' txo\-pu ke sley\-ku oet kaw\-'it.} Your threatening words don't scare me one bit.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{txopunil}{\B{txo\-pu\-nil}}, nightmare.

% txopunil
\hi
\HT{txopunil}{\B{\cp{txo}punil}} \I{["t'o.pu.nil]} \emph{n.} nightmare.
(\ding{43}~\HL{txopu}{\B{txo\-pu}}, \HL{unil}{\B{u\-nil}})

% txotsafya
\hi 
\HT{txotsafya}{\B{\cp{txo}tsafya}} \I{["t'o.\t{ts}a.fja]} \emph{adv.} if that's the case, if that's so. 
\B{Ke su\-nu ngar tey\-lu srak? Txo\-tsa\-fya, tìng oer pum\-it nge\-yä!} You don’t like teylu? If that’s the case, give me yours!~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}  
(\ding{43}~\HL{txo}{\B{txo}}, \HL{tsafya}{\B{tsa\-fya}}, \HL{txokefyaw}{\B{txo\-ke\-fyaw}})

% txukx
\hi
\HT{txukx}{\B{txukx}} \I{[t'uk']} \emph{or} \I{[t'Uk']} \emph{adj.} (\emph{a. physical}) deep. 
(\emph{b. metaphorically for thoughts, ideas, analyses, etc.}) deep, thorough. 
\B{Ay\-sä\-fpìl a\-txukx.} Deep thoughts.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Tsun sä\-fpìl a\-txukx nì\-'aw tsa\-tì\-pawm\-it ti\-vìng, ke\-fyak?} Only deep thought can give that question, right?~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/comment-page-1/\#comment-225}{b}
(\ding{214}~\HL{reng}{\B{reng}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nItxukx}{\B{nì\-txukx}}, deeply.
\on{2}.~\HL{txukxefu}{\B{txu\-kxe\-fu}}, care, be concerned about.

% txukxefu
\hi 
\HT{txukxefu}{\B{txu\cp{kxe}fu}} \I{[t'u."k'$\cdot$E.f$\cdot$u]} \emph{vin., inf.}\on{23} care, be concerned about, have deep feelings for (with the topical or \ding{43}~\HL{teri}{\B{teri}}). 
\B{Nga\-ri po ke txu\-kxe\-fu kaw\-'it.} He doesn't care a bit about you.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}
\B{Fu\-ria te\-ri lì'\-fya aw\-nge\-yä nga txu\-kxe\-fu nì\-ftxan, sei\-yi i\-ra\-yo.} Thank you for caring so much about our language.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}  
(\ding{43}~\HL{txukx}{\B{txukx}}, \HL{'efu}{\B{'e\-fu}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tItxukxefu}{\B{tì\-txu\-kxe\-fu}}, care, concern. 

% txula
\hi
\HT{txula}{\B{\cp{txu}la}} \I{["t'u.la]} (\textsc{rn}: \B{\cp{txù}la} \I{["dU.la]}) \emph{vtr.} build, construct.
\B{Fu\-ri\-a txu\-la tsa\-lew\-ti lu srä sìl\-tsan to fngap.} For constructing that cover, woven cloth is better than metal.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Txo ngal tsat txi\-vu\-la, (tsa\-krr) fo za\-ya\-'u.} If you build that (then) they will come.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
\B{Maw fwa fkol Kel\-ut\-ral\-it sko\-la\-'a, lem\-rey\-a ha\-pxì\-tul tsa\-soai\-ä txo\-lu\-la mip\-a kel\-ku\-ti mì ta\-yo.} After Hometree was destroyed, the surviving members of that family built a new home on the plains.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Fì\-stxe\-lit fol txe\-ru\-la fpi olo'\-eyk\-tan.} They're constructing this gift for the chief.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{tuk\-ru a txo\-lu\-la fkol ta rìn}, a spear made of wood.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}
(\emph{a. with} \B{txep}) make a fire(place). \B{Frr\-nen hì\-'i lu nì\-txan a krr, apxa txep\-it txu\-la oel.} When my babies are very little, I build a big fire.~\LINK{http://forum.learnnavi.org/intermediate/translation-exercise-a-navajo-coyote-tale-in-navi/}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItxula}{\B{tì\-txu\-la}}, construction, constructed thing.
\on{2}.~\HL{txawnulsrung}{\B{txaw\-nul\-srung}}, machine.

% txum
\hi
\HT{txum}{\B{txum}} \I{[t'um]} \emph{or} \I{[t'Um]} (\textsc{rn}: \B{txùm} \I{[dUm]}) \emph{n.} poison. 
\B{Oe\-ri lu swoa fne\-txum.} I'm allergic to intoxicating beverage.~\LINK{http://forum.learnnavi.org/language-updates/new-words-from-the-2012-avatarmeet-navi-class/msg551479/\#msg551479}{b}  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txumnga'}{\B{txum\-nga'}}, poisonous. 
\on{2}.~\HL{txumpaywll}{\B{txum\-pay\-wll}}, scorpion thistle.  
\on{3}.~\HL{txumtsA'wll}{\B{txum\-tsä'\-wll}}, baja tickler.
\on{4}.~\HL{wAtum}{\B{wä\-tum}}, antidote.

% txumnga'
\hi
\HT{txumnga'}{\B{\cp{txum}nga'}} \I{["t'um.NaP]} \emph{or} \I{["t'Um.]} (\textsc{rn}: \B{\cp{txù}nga'} \I{["dU.NaP]}) \emph{adj.} poisonous.
(\ding{43}~\HL{txum}{\B{txum}}, \HL{-nga'}{\B{-nga'}})

% txumpaywll
\hi
\HT{txumpaywll}{\B{\cp{txum}paywll}} \I{["t'um.paj.w\s{l}]} \emph{or} \I{["t'Um.]} (\textsc{rn}: \B{\cp{txùm}paywll} \I{["dUm.paj.w\s{l}]}) \emph{n.}, \ding{96} scorpion thistle, ``poison water plant'', \emph{scorpioflora maxima}. 
(\ding{43}~\HL{txum}{\B{txum}}, \HL{pay}{\B{pay}}, \HL{'ewll}{\B{'ewll}})

% txumtsä'wll
\hi
\HT{txumtsA'wll}{\B{\cp{txum}tsä'wll}} \I{["t'um.\t{ts}\ae{}P.w\s{l}]} \emph{or} \I{["t'Um.]} (\textsc{rn}: \B{\cp{txùm}tsä'wll} \I{["dUm.\t{ts}E.w\s{l}]}) \emph{n.}, \ding{96} baja tickler, ``poison squirt plant'', \emph{flaska reclinata}. 
(\ding{43}~\HL{txum}{\B{txum}}, \HL{tsA'}{\B{tsä'}}, \HL{'ewll}{\B{'ewll}})

% txung
\hi 
\HT{txung}{\B{txung}} \I{[t'uN]} \emph{or} \I{[t'UN]} \emph{vtr.} disturb, disrupt, bother, affect negatively. 
\B{Fì\-sä\-spxin\-ìl ke txa\-yung aw\-ngey sì\-rey\-ti tì\-'i'\-a\-vay krr\-ä.} This illness will not disrupt our lifes forever.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b}
\B{Oey fpom\-it txung rä\-'ä!} Don't bother me!~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} 

% txur
\hi
\HT{txur}{\B{txur}} \I{[t'uR]} \emph{or} \I{[t'UR]} \emph{adj.} strong.   
\B{Lu nga win sì txur | Lu nga txan\-tslu\-sam.} You are fast and strong | You are wise.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
\B{Nga nì\-aw\-no\-mum to oe\-tsyìp lu txur nì\-txan.} As everyone knows, you're a lot stronger than little old me.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Krro lu 'Ìng\-lì\-sì txur nì\-hawng mì re\-'o.} Sometimes English is too strong in (my) head.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b}
\B{Tsa\-sä\-lä\-txayn Na'\-vi\-ru srung so\-li nì\-'aw fte sli\-vu txur nì\-'ul.} That defeat only helped the People become stronger.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Nga pam\-rel so\-li san sley\-ki\-vu ut\-ral\-ìl lì'\-fya\-yä le\-Na'\-vi pxay\-a ay\-vur\-it a\-txur sìk.} You wrote, ``May the Na'vi language tree produce many strong stories.''~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tItxur}{\B{tì\-txur}}, strength. 
\on{2}.~\HL{txantur}{\B{txan\-tur}}, powerful. 
\on{3}.~\HL{txurtel}{\B{txur\-tel}}, rope. 
\on{4}.~\HL{txurtu}{\B{txur\-tu}}, strongman\slash strongwoman, brawny person.
\on{5}.~\HL{txurtseng}{\B{txur\-tseng}}, fortification, fortress, bulwark.

% txurtel
\hi
\HT{txurtel}{\B{\cp{txur}tel}} \I{["t'uR.tEl]} \emph{or} \I{["t'UR.]} \emph{adj.} rope.
\B{Na\-ri si fte\-ke ay\-tur\-tel\-ur tì\-zin si\-vi!} Be careful not to tangle the ropes.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} 
\B{Txur\-tel\-mì fo fta so\-li fte\-ke ka tsyokx fwi\-vi.} They tied a knot in the rope so it wouldn't slip through their hands.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{txur}{\B{txur}}, \HL{telem}{\B{te\-lem}})

% txurtu
\hi 
\HT{txurtu}{\B{\cp{txur}tu}} \I{["t'uR.tu]} \emph{or} \I{["t'UR.]} \emph{n.} strongman\slash strongwoman, brawny person, \emph{a person of strong physique or character}.
(\ding{43}~\HL{txur}{\B{txur}}, \HL{-tu}{\B{-tu}}, \ding{43}~\HL{meyptu}{\B{meyptu}})

% txurtseng
\hi 
\HT{txurtseng}{\B{\cp{txur}tseng}} \I{["t'uR.\t{ts}EN]} \emph{or} \I{["t'UR.]} \emph{n.} fortification, fortress, bulwark. 
\B{Aw\-nga\-ri ke\-tsran tseng\-ne ki\-vä, fì\-soaia lu txur\-tseng.} Wherever we go, this family is our fortress.~\LINK{naviteri.org/2022/05/results-of-the-tttc/}{b} 
(\ding{43}~\HL{txur}{\B{txur}}, \HL{tseng}{\B{tse\-nge}}) 

\end{multicols}