\pdfbookmark[0]{TS}{TS}
\begin{center}
\section*{TS \I{[\t{ts}]} (\emph{Tsä})}
\end{center}

\begin{multicols}{2}

% tsa-
\hi 
\HT{tsa-}{\B{tsa-}} \I{[\t{ts}a.]} \textsc{i}. \emph{prod. pref. on n.s} that (there) (\emph{distal singular dem. pref.}). \B{vur}, (a) story, \B{tsa\-vur}, that story. (\ding{43}~\B{fì-}) \\
\textsc{ii}. \emph{shortened form of} \B{tsa'u\slash tsaw} \emph{with enclitic adp.s} \B{Po tsa\-ne kar\-mä a tseng\-it ke tsì\-me\-'a oel.} I didn't see where she was going. [lit.: I didn't see the place that she was going to (it)]~\LINK{http://forum.learnnavi.org/language-updates/combined-answers-1-\%28feb-16\%29/}{ln}
\B{Oel nge\-y\"a tì\-hawl\-it na\-txu ulte tsa\-w\"a wa\-syem.} I disapprove of your plan and will oppose (fight against) it.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b} 

% tsa'u
\hi
\HT{tsa'u}{\B{\cp{tsa}'u}} \I{["\t{ts}a.Pu]} (\emph{pl.} \B{(ay)sa'u}) \textsc{i}. \emph{n.} that thing, it. 
\B{Tsa\-'u el\-tur tì\-txen si.} That's interesting.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Tsa\-'u la\-tsu pe\-u?} What on earth is that?~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Ke lu oer am\-'a, tsa\-'u po\-lä\-hem a krr, su\-nu oe\-ru nì\-txan.} I have no doubt that when it arrives, I'm going to enjoy it very much.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{Ngal new a tsa\-'ut rä\-'ä wi\-vo, ma 'evi. Vi\-vin.} Don't reach for what you want, child. Ask for it.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Txo tsi\-ve\-'a ay\-ngal key\-eyt, ru\-txe oe\-ru pi\-veng fte tsi\-vun oe sa\-'ut ley\-ki\-va\-tem.} If you see mistakes, please tell me so that I can change them.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Tsa\-'u\-ri po pll\-ngay.} He admits it.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} \\
\textsc{ii}. \B{tsa'u} \emph{or} \B{tsaw} \I{[\t{ts}aw]} \emph{pn.} it, that. 
\B{Lä\-ngu fì\-seyn kel\-ho\-an nì\-ngay. Tsun oe hi\-veyn tsaw\-sìn 'a\-'aw\-a swaw\-tsyìp nì\-'aw.} This chair is really uncomfortable. I can only sit on it a few seconds.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
(\emph{shortened forms:} \B{tsal, tsat(i), tsar(u), tse\-yä, tsa\-ri})
\B{Tsat rä\-'ä yi\-vom!} Don't eat that.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Ke tsun oe fì\-'awkx\-it tsyi\-vìl. Ke lu tsa\-ru kea yì.} I can't scale this cliff. It doesn't have any footholds.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Fra\-'u\-ä ran ngä\-pop fa fra\-syon tse\-yä.} The \emph{ran} of each thing arises from the totality of its attributes.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Ye'\-rìn 'ì\-yi'\-a sä\-nu\-me a tsa\-ri kll\-fro' oe.} Soon the teaching that I am responsible for will end.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}

% tsay+
\hi 
\HT{tsay}{\B{tsay}}\texttt{+} \I{[\t{ts}aj.]} \emph{prod. pref. on n.s} those (there) (\emph{distal plural dem. pref.}). 
\B{Ke pol\-txe pol tsay\-lì\-'ut.} She didn't say that (i.e., those words).~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
(\ding{43}~\HL{tsa-}{\B{tsa-}}; \HL{fay}{\B{fay}\texttt{+}}) 

% tsafya
\hi
\HT{tsafya}{\B{\cp{tsa}fya}} \I{["\t{ts}a.fja]} \emph{adv.} that way, like that. 
\B{Sra\-ke tsun oe fay\-upxa\-ret tsli\-vam nì\-ftue? Tse … ze\-ne pi\-vll\-txe san pxìm tsa\-fya lu sìk.} Can I understand these messages easily? Well…  I must say ``Often that way''.~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln}
\B{Slä tsa\-fya pam\-rel rä\-'ä si!} But don't write like that.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1092}{b}
(\emph{a. idiom}) \B{fì\-fya tsa\-fya}, one way or another. \B{Slä fì\-fya tsa\-fya, nga\-ri ti\-re\-al tsa\-tse\-nget ta\-yok.} But one way or the other your spirit will be there.~\LINK{http://naviteri.org/2013/04/sin-asok-sin-zusawkrra-recent-and-upcoming-activities/comment-page-1/\#comment-2316}{b}

% tsa-hey
\hi
\HT{tsa-hey}{\B{tsa-hey}} \I{["\t{ts}a."hEj]} \emph{intj. expression of warning or frustration}.

% tsaheylu / tsaheyl si
\hi
\HT{tsaheylu}{\B{tsa\cp{hey}lu}} \I{[\t{ts}a."hEj.lu]} \textsc{i}. \emph{n.} bond, neural connection. 
\B{Sä\-tsway\-on a\-'aw\-ve tsa\-heyl\-ur ha\-sey si; ke tsun nga pi\-vey.} First flight seals the bond; you cannot wait.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{ii}. \B{tsa\cp{heyl} si} \I{[\t{ts}a."hEjl si]} \emph{vin.} bond (\B{hu}, with), establish a neural connection. 
\B{Ik\-ran\-ìri krr\-a hu tu\-te tsa\-heyl si, ftang li\-vu yrr.} When an \emph{ikran} bonds with a person, it ceases to be wild.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tsyìl Ik\-ni\-ma\-yat ul\-te tsa\-heyl si ik\-ran\-hu a fì\-'u lu tì\-fme\-tok a ze\-ne fra\-ta\-ron\-yu a\-'e\-wan em\-zi\-va\-'u.} Scaling Iknimaya and bonding with a banshee is a test that every young hunter must pass.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}

% tsahìk
\hi
\HT{tsahIk}{\B{\cp{tsa}hìk}} \I{["\t{ts}a.hIk]} \emph{n.} Tsa\-hik, high priestess (\emph{who interprets the will of Eywa}). 
\B{Mo\-'at \mbox{alu} Tsa\-hìk lu O\-ma\-ti\-ka\-ya\-ä le\-'awa ha\-pxì\-tu a \mbox{ioi} sä\-pi fa 'a\-re.} Mo'at, the Tsahik, is the only member of the Omatikaya who wears a poncho.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Sko Sa\-hìk ke tsun oe mì\-ftxe\-le tsngi\-vaw\-vìk, sko sa'\-nok tsun.} As Tsahik, I cannot weep over this matter, as mother I can.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Tsa\-hìk\-ur txe\-le lu.} This is a matter for the Tsahik.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Neytiri_talking_to_Tsu.27tey}{a\slash wiki}

% tsakem / tsakem si
\hi
\HT{tsakem}{\B{\cp{tsa}kem}} \I{["\t{ts}a.kEm]} \emph{or} \B{tsa\cp{kem}} \I{[\t{ts}a."kEm]} \textsc{i}. \emph{dem., pn.} that (action). 
\B{Tsa\-kem kop krr\-nekx.} That also takes time.~\LINK{http://forum.learnnavi.org/language-updates/just-a-few-interesting-little-words/}{ln} 
\B{Txo tsa\-kem li\-ven mì ftaw\-nem\-krr.} If that happens in the past.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1910}{b} \\ 
\textsc{ii}. \B{tsakem si} \I{["\t{ts}a.kEm si]} \emph{vin.} do that (action). 
\B{Tsa\-kem so\-li oe nga\-ru.} I did it to you.~\LINK{http://forum.learnnavi.org/language-updates/i-have-something-to-say/}{ln}
\B{Sìl\-pey oe tsnì tsì\-ye\-vun nì\-'i'\-a tsa\-kem si\-vi fì\-pì\-lok\-fa.} I hope that I can finally do that through this blog.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Zun ay\-oe li\-vu tsam\-si\-yu, zel tsa\-kem ke sim\-vi.} If we were warriors, we wouldn't have done that.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
(\emph{a. proverb}) \B{Kem\-ìri a nga\-ru prr\-te' ke lu, tsa\-kem rä\-'ä si\-vi ay\-la\-he\-ru.} As for any action that you don't like, don't act that way to others.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}
(\ding{43}~\HL{kem}{\B{kem}}, \HL{tsa-}{\B{tsa-}})

% tsakrr 
\hi
\HT{tsakrr}{\B{tsa\cp{krr}}} \I{[\t{ts}a."k\s{r}]} \textsc{i}. \emph{n.} that time. 
\B{Mun\-trr\-am oe ko\-lä\-teng hu Ra\-lu ul\-te su\-nu oer tsa\-krr.} I spent last weekend with Ralu and had a good time [lit.: that time was pleasant to me].~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} \\ 
\textsc{ii}. \emph{adv.} at that time, then.
\B{Tì\-'eyng\-it oel to\-lel a krr, tsa\-krr pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} When I receive an answer, I will then decide the best way to bring you the words and rules of Na'vi.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Oe\-yä ik\-ran sli\-vu nga, tsa\-krr oeng 'aw\-si\-teng mi\-vak\-to.} Be my ikran and let's ride together.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{ld}
\B{Tsa\-krr syay ay\-nge\-yä, syay o\-lo'\-ä oe\-yä la\-yu teng.} Then you will suffer the same fate as my clan.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
(\emph{a. optionally with} \B{txo}) \B{Pxan li\-vu txo nì\-'aw oe nga\-ri | Tsa\-krr nga Na'\-vi\-ru yom\-tì\-yìng.} Only if I am worthy of you | Will you feed the People.~\LINK{http://wiki.learnnavi.org/index.php/Hunting_Song}{wiki}
(\emph{b. proverb}) \B{Txo ke nì\-yo' tsa\-krr nì\-yol.} `If you can't be flawless, at least be brief.'~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{krr}{\B{krr}}, \HL{tsa-}{\B{tsa-}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsakrrvay}{\B{tsakrrvay}}, until then, in the meantime.

% tsakrrvay
\hi
\HT{tsakrrvay}{\B{tsa\cp{krr}vay}} \I{[\t{ts}a."k\s{r}.vaj]} (\emph{a. adv.}) until then, in the meantime.
\B{Tsa\-krr\-vay, ay\-nge\-yä tìm\-wey\-pey\-ri ira\-yo sei\-yi oe, ul\-te fì\-trr\-ä ftxo\-zä\-ri, sìl\-pey oe, ay\-nga\-ru prr\-te' li\-vu.} In the meantime, I thank you for your patience, and I hope you enjoy today's celebration.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
(\emph{b. conv.}) until then, see you then. \B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay. -- Ye\-io'! Tsa\-krr\-vay ko!} Let's meet on the hill tomorrow at nightfall. -- Perfect! See you then.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{43}~\HL{tsakrr}{\B{tsa\-krr}}, \HL{vay}{\B{vay}})

% tsaktap / si
\hi 
\HT{tsaktap}{\B{\cp{tsak}tap}} \I{["\t{ts}ak\textcorner.tap\textcorner]} \textsc{i}. \emph{n.} violence. 
\B{Slä set tä\-ngok kop tan\-tsaw\-tsray\-ti ay\-oe\-yä 'uol a\-la\-he: tsak\-tap.} But now, something else is also at our large cities: violence.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} \\
\textsc{ii}. \B{\cp{tsak}tap si} \I{["\t{ts}ak\textcorner.tap\textcorner{} si]} \emph{vin.} be violent, use violence.
\B{Tsak\-tap r\"a'\"a si kaw\-krr mung\-wrr\-txo ke li\-vu kea fya\-'o a\-la\-he.} Never use violence unless there is no other way. ~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letsaktap}{\B{le\-tsak\-tap}}, violent.

% tsal
\hi
\B{tsal} \I{[\t{ts}al]} : \HL{tsa'u}{\B{tsa'u}}

% tsala
\hi
\HT{tsala}{\B{\cp{tsa}la}}, \B{tsa\-\cp{'ul} a}, \emph{or} \B{a tsa\-\cp{'ul}} \I{["\t{ts}a.la]} \emph{conj.} that (\emph{to introduce an agentive subordinate clause}). 
(\ding{43}~\HL{fula}{\B{fu\-la}})

% tsalsungay
\hi
\HT{tsalsungay}{\B{tsalsu\cp{ngay}}} \I{[\t{ts}al.su."Naj]} \emph{adv.} nevertheless, even so.
\B{Slä tsal\-su\-ngay, txo tì\-tslam for, lu txa\-yo na'\-rìng\-to sìl\-tsan.} But even so, if they're smart they'll take open terrain over bush.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Ta\-fral kì\-syar a ay\-sä\-num\-vi sì ay\-lì'\-fya\-vi lu fyin. Tsal\-su\-ngay sìl\-pey oe, nu\-mul\-txa le\-sar lì\-ye\-vu.} That's why I'll teach lessons and expressions that are simple. Nevertheless, I hope the class will be useful.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b} 
\B{Hu\-fwa lu fil\-ur Va'\-ru\-ä fne\-fe'\-ran\-vi, tsal\-su\-ngay fpìl fu\-ta say\-rìp lu nì\-txan.} Although Va'ru's facial stripes are rather uneven, I still think he's very handsome.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 

% tsam
\hi
\HT{tsam}{\B{tsam}} \I{[\t{ts}am]} \emph{n.} war. 
\B{Hu Saw\-tu\-te a tsam lu ha\-sey.} The war with the Skypeople is over.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
\B{Sra\-ke fpìl ay\-ngal fu\-ta aw\-nga tsun hu Saw\-tu\-te a tsam\-it 'i\-vaw\-nìm fa fwa päng\-kxo fo\-hu nì\-'aw, ma smuk?} Do you think we can avoid a war with the Sky People solely by talking with them, my brothers?~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/comment-page-1/\#comment-1755}{b}
\B{Po\-ri tì\-kak\-rel tì\-kak\-pam\-sì kum tsam\-ä lu.} His blindness and deafness are a result of the war.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsamsiyu}{\B{tsam\-si\-yu}}, warrior. 
\on{2}.~\HL{tsampongu}{\B{tsam\-po\-ngu}}, war party. 
\on{3}.~\HL{tsamopin}{\B{tsa\-mo\-pin}}, war paint. 
\on{4}.~\HL{tsamkuk}{\B{tsam\-kuk}}, war drum.
\on{5}.~\HL{tsamsA'o}{\B{tsam\-sä\-'o}}, weapon of war.

% tsame-
\hi 
\HT{tsame}{\B{tsame}\texttt{+}} \I{[\t{ts}a.mE.]} \emph{prod. pref. on n.s} those two (there) (\emph{distal dual dem. pref.}). 
\B{Fpìl oel fu\-ta tsun ay\-nga tsli\-vam teyng\-ta tsa\-me\-stxo za\-'u ftu pe\-sim.} I think you understand where those two names come from.~\LINK{http://naviteri.org/2015/09/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise-2/}{b}
(\ding{43}~\HL{tsa-}{\B{tsa-}}; \HL{me}{\B{me}\texttt{+}}) 

% tsamkuk
\hi
\HT{tsamkuk}{\B{\cp{tsam}kuk}} \I{["\t{ts}am.kuk]} \emph{or} \I{[.kUk]} (\textsc{rn}: \B{\cp{tsam}kùk} \I{["\t{ts}am.kUk\textcorner]}) \emph{n.} war drum. 
(\ding{43}~\HL{tsam}{\B{tsam}}, \HL{takuk}{\B{ta\-kuk}})

% tsamopin
\hi
\HT{tsamopin}{\B{\cp{tsa}mopin}} \I{["\t{ts}a.mo.pin]} \emph{n.} warpaint.
(\ding{43}~\HL{tsam}{\B{tsam}}, \HL{'opin}{\B{'o\-pin}})

% tsampongu
\hi
\HT{tsampongu}{\B{\cp{tsam}pongu}} \I{["\t{ts}am.po.Nu]} \emph{n.} war party. 
\B{Set sre\-fey oel fu\-ta tsam\-po\-ngu tä\-txaw maw txon\-'ong.} I'm currently expecting the war party back after nightfall.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b} 
\B{Lu tsa\-tsam\-si\-yu le\-'aw\-a ha\-pxì\-tu tsam\-po\-ngu\-ä a mal lu moer.} That warrior is the only member of the war party that we both trust.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\ding{43}~\HL{tsam}{\B{tsam}}, \HL{pongu}{\B{po\-ngu}})

% tsamsä'o
\hi
\HT{tsamsA'o}{\B{\cp{tsam}sä'o}} \I{["\t{ts}am.s\ae.Po]} \emph{n.} weapon of war.
\B{Po\-ri ay\-sam\-sä\-'o a\-tì\-txur\-nga' fra\-to mì hi\-fkey la\-yu mì syokx.} He will have the most powerful weapons of war in the world in his hands.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{tsam}{\B{tsam}}, \HL{sA'o}{\B{sä'o}})

% tsamsiyu
\hi
\HT{tsamsiyu}{\B{\cp{tsam}siyu}} \I{["\t{ts}am.si.ju]} \emph{n.} warrior. 
\B{Oe lu tsam\-si\-yu.} I am a worrior.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
\B{Ke lu po tsam\-si\-yu kaw\-'it!} There's not a warrior's bone in his whole body!~\LINK{http://forum.learnnavi.org/language-updates/ke-kawit/msg174572/\#msg174572}{ln}
\B{Nì\-trr\-trr pxì\-mun\-'i sam\-si\-yul ay\-swi\-zaw\-it ku\-tu\-ä a\-law\-nä\-txayn sno\-kip nì\-'eng.} Warriors typically share the arrows of their defeated enemies among themselves.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Po\-ri wem\-tswo fra\-tsam\-si\-yur ro\-lo\-'a nì\-txan.} His ability to fight greatly impressed all the warriors.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Tsran\-ten fra\-to a syon tsam\-si\-yu\-ä lu tì\-tstew.} The most important characteristic of a warrior is bravery.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Tsam\-si\-yu\-ri lu tì\-yo\-ra'\-ä tì\-new le\-kin.} A warrior must have the desire for victory.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Lar\-mu tsa\-tsam\-si\-yu\-hu tì\-vawm a lu stxong ay\-oer.} That warrior carried with him a darkness unknown to us.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{tsam}{\B{tsam}})

% tsan'ul
\hi
\HT{tsan'ul}{\B{\cp{tsan}'ul}} \I{["\t{ts}an.P$\cdot$ul]} \emph{vin., inf.}\on{22} improve, get better.
\B{Saw\-tu\-te\-ri tì\-fkey\-tok ke tsan\-'o\-lul kaw\-'it.} The situation with the Skypeople hasn't improved one bit.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Lì'\-fya\-ri le\-Na'\-vi nga tsan\-'e\-re\-i\-ul fra\-trr.} I'm delighted that your Na'vi is improving every day.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Ru\-txe fì\-tì\-oeyk\-tìng\-it tsan\-'ey\-ki\-vul. Ke lu law kaw\-'it.} Please improve this explanation. It's not at all clear.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
(\emph{a. rhyming expression with stress change}) \B{Tslikx, tì\-ran, tul; | Ftu yì ne yì tsan\-\cp{'ul}.} \emph{when learning s.t. new, you have to proceed from step to step: baby steps first, then bigger ones} [lit.: `Crawl, walk, run; from level to level get better.']~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(\ding{214}~\B{fe\-'ul}, \ding{43}~\B{sìl\-tsan}, \B{'ul}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tItsan'ul}{\B{tì\-tsan\-'ul}}, improvement (abstract).
\on{2}.~\HL{sAtsan'ul}{\B{sä\-tsan\-'ul}}, improvement (specific).

% tsankum
\hi
\HT{tsankum}{\B{\cp{tsan}kum}} \I{["\t{ts}an.kum]} \emph{or} \I{[.kUm]}, \emph{coll. pronounced as} \I{["\t{ts}aN.kum]} \emph{n.} advantage, benefit, upside, gain.
\B{Tsa\-kem\-ìri tsan\-kum lu law.} The advantage of that action is clear.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} 
(\ding{43}~\HL{kum}{\B{kum}}, \ding{214}~\HL{fekum}{\B{fe\-kum}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsankumnga'}{\B{tsan\-kum\-nga'}}, adventageous.

% tsankumnga'
\hi
\HT{tsankumnga'}{\B{\cp{tsan}kumnga'}} \I{["\t{ts}an.kum.NaP]} \emph{or} \I{[.kUm.]}, \emph{coll. pronounced as} \I{["\t{ts}aN.kum.NaP]} \emph{adj.} advantageous.
(\ding{43}~\HL{tsankum}{\B{tsan\-kum}}, \ding{214}~\HL{fekumnga'}{\B{fe\-kum\-nga'}})

% tsantu
\hi 
\HT{tsantu}{\B{\cp{tsan}tu}} \I{["\t{ts}an.tu]} \emph{n.} good person, `good guy'. 
\B{Lal\-a tsa\-rel\-mì a\-ru\-sikx, yem\-stokx tsan\-tul haw\-re'\-ti a\-teyr, kawng\-tul pum\-it a\-la\-yon.} In that old movie, the good guy wears a white hat, the bad guy a black one.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 
(\ding{214}~\HL{kawngtu}{\B{kawng\-tu}}; \ding{43}~\HL{sIltsan}{\B{sìl\-tsan}}, \HL{tute1}{\B{tu\-te}})

% tsantxäl / si
\hi 
\HT{tsantxAl}{\B{tsan\cp{txäl}}} \I{[\t{ts}an."t'\ae{}l]} \textsc{i}. \emph{n.} invitation. 
\B{Nge\-yä fì\-tsan\-txäl\-ìri a\-tì\-tstun\-wi\-nga' ira\-yo, slä ke tsä\-ngun oe zi\-va\-'u.} Thank you for this kind invitation, but unfortunately I cannot come.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} \\
\textsc{ii}. \B{tsan\cp{txäl} si} \I{[\t{ts}an."t'\ae{}l si]} \emph{vin.} invite (\emph{with} \ding{43}~\HL{tsnI}{\B{tsnì}}).
\B{Po tsan\-txäl so\-li oer tsnì zi\-va\-'u kel\-ku\-ne.} She invited me to come to her home.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\ding{43}~\HL{sIlstan}{\B{sìl\-tsan}}, \HL{AtxAle}{\B{ä\-txä\-le}}) 

% tsang
\hi
\HT{tsang}{\B{tsang}} \I{[\t{ts}aN]} \emph{n.} a piercing. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{miktsang}{\B{mik\-tsang}}, earring.
\on{2}.~\HL{ontsang}{\B{on\-tsang}}, nose ring.

% tsap'alute / si 
\hi
\HT{tsap'alute}{\B{tsap\cp{'a}lute}} \I{[\t{ts}ap\textcorner."Pa.lu.tE]} \textsc{i}. (\emph{a. n.}) apology (for, \emph{with the topic}).
\B{Fì\-kxey\-ey\-ri tsap\-'a\-lu\-te. Ngi\-an la\-he\-a lì\-'u alu Pol ke lu kxey\-ey.} Apologies for this mistake. However the other word `pol' is not a mistake.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/comment-page-1/\#comment-520}{ln}
(\emph{b. intj.}) ``(my) apologies''. \B{Tsap\-'a\-lu\-te, mì upxa\-re a fpo\-le' oel ay\-nga\-ru trr\-am lu kxey\-ey.} My apologies, in the message I sent you yesterday is a mistake.~\LINK{http://forum.learnnavi.org/language-updates/another-email-from-frommer/msg68133/\#msg68133}{b} \\
\textsc{ii}. \B{tsap\cp{'a}lute si} \I{[\t{ts}ap\textcorner."Pa.lu.tE si]} \emph{vin.} apologize. 
\B{Fì\-skxawng\-ìri tsap\-'a\-lu\-te se\-ngi oe.} I apologize for this moron.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Po tsap\-'a\-lu\-te so\-li nìk\-mar.} He apologized at the right time.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Nga tsap\-'alu\-te so\-li srak? -- Soli.} Did you apologize? -- (Yes,) I have.~\LINK{http://forum.learnnavi.org/language-updates/auxilary-verb-si-possessive-dative-krr/}{ln}

% tsapo
\hi
\HT{tsapo}{\B{\cp{tsa}po}} \I{["\t{ts}a.po]} \emph{dem. pn.} that one.
\B{Tu\-te a ke\-ftxo fra\-to lu tsa\-po a tì\-yawn\-ur lie ke so\-li kaw\-krr.} The saddest person of all is the one who has never experienced love.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}
\B{Ri\-ni tsa\-po\-hu ho\-lum nì\-mal nì\-wotx.} Rini left with that guy without thinking twice about it.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
(\emph{a. in contrastive comparisons}) \B{Fì\-kar\-yu alu fì\-po lu tsul\-fä\-tu; tsa\-kar\-yu alu tsa\-po lu \mbox{skxawng}.} \emph{This} teacher is a master; \emph{that} teacher is a fool.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
(\ding{43}~\HL{po}{\B{po}}, \HL{tsa-}{\B{tsa-}})

% tsapxe-
\hi 
\HT{tsapxe}{\B{tsapxe}\texttt{+}} \I{[\t{ts}a.p'E.]} \emph{prod. pref. on n.s} those three (there) (\emph{distal trial dem. pref.}). 
\B{Tsa\-pxe\-sam\-si\-yu lu pe\-su\slash tu\-pe?} Who are those three warriors?~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
(\ding{43}~\HL{tsa-}{\B{tsa-}}; \HL{pxe}{\B{pxe}\texttt{+}}) 

% tsar / tsaru
\hi
\B{tsar} \I{["\t{ts}aR]} \emph{or} \B{\cp{tsa}ru} \I{["\t{ts}a.Ru]}~:~\HL{tsa'u}{\B{tsa'u}}.

% tsara
\hi
\HT{tsara}{\B{\cp{tsa}ra}}, \B{tsa\-\cp{'ur} a}, \emph{or} \B{a tsa\-\cp{{'ur}}} \I{["\t{ts}a.Ra]} \emph{conj.} that (\emph{to introduce a dative subordinate clause}). 
(\ding{43}~\HL{fura}{\B{fu\-ra}}) 

% tsari
\hi
\B{\cp{tsa}ri} \I{["\t{ts}a.Ri]}~:~\HL{tsa'u}{\B{tsa'u}}.

% tsaria
\hi
\HT{tsaria}{\B{\cp{tsa}ria}}, \B{tsa\-\cp{'u}ri a}, \emph{or} \B{a tsa\-\cp{{'u}ri}} \I{["\t{ts}a.Ri.a]} \emph{conj.} that (\emph{to introduce a topic subordinate clause}). 

% tsat / tsati
\hi
\B{tsat} \I{[\t{ts}at\textcorner]} \emph{or} \B{\cp{tsa}ti} \I{["\t{ts}a.ti]} : \HL{tsa'u}{\B{tsa'u}}.

% tsata
\hi
\HT{tsata}{\B{\cp{tsa}ta}}, \B{tsa\-\cp{'ut} a}, \emph{or} \B{a tsa\-\cp{'ut}} \I{["\t{ts}a.ta]} \emph{conj.} that (\emph{to introduce a patientive subordinate clause}). 

% tsatu
\hi
\HT{tsatu}{\B{\cp{tsa}tu}} \I{["\t{ts}a.tu]} \emph{pn.} that person. 
(\ding{43}~\HL{tute1}{\B{tu\-te}}, \HL{tsa-}{\B{tsa-}})

% tsatseng
\hi 
\HT{tsatseng}{\B{\cp{tsa}tseng}} \I{["\t{ts}a.\t{ts}EN]} \emph{or} \B{tsa\cp{tseng}} \I{[\t{ts}a."\t{ts}EN]} \textsc{i}. \emph{n.} there, that place. 
\B{Tsa\-tse\-nge le\-hrr\-ap lu fì\-txan kum\-a tsa\-ne ke kä aw\-nga kaw\-krr.} That place is so dangerous we never go there.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Slä am\-'a\-lu\-ke nge\-yä ti\-re\-al tsa\-tse\-nget ta\-ye\-i\-ok.} But without a doubt your spirit will be there.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/comment-page-1/\#comment-1569}{b}
\B{Fwam\-pop\-ä tem\-rey tsa\-tseng\-mì ngä\-zìk lu nì\-wotx.} It's very hard for a tapirus to survive there.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 
\B{Tsun oe nga\-hu tsa\-tse\-nge\-ne ki\-vä, slä nul\-new fu\-ta si\-vop oeng nì\-txi\-lu\-ke.} I can go there with you, but I prefer to travel leisurely.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\textsc{ii}. \emph{adv.} there. \B{Kil\-van\-ìri tsa\-tseng slo\-snep\-pe?} How wide is the river there?~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} 
(\ding{43}~\HL{tseng}{\B{tse\-nge}}, \HL{tsa-}{\B{tsa-}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{kAsatseng}{\B{kä\-sa\-tseng}}, out there.

% tsatsenge
\hi
\B{tsa\cp{tse}nge} \I{[\t{ts}a."\t{ts}E.NE]} : \HL{tseng}{\B{tsatseng}, \B{tsenge}} 

% tsaw
\hi
\B{tsaw} \I{[\t{ts}aw]} : \HL{tsa'u}{\B{tsa'u}}

% tsawa
\hi
\HT{tsawa}{\B{\cp{tsa}wa}}, \B{tsa\-\cp{'u} a}, \emph{or} \B{a tsa\-\cp{{'u}}} \I{["\t{ts}a.wa]} \emph{conj.} that (\emph{to introduce a subjective subordinate clause}). 

% tsawke
\hi
\HT{tsawke}{\B{\cp{tsaw}ke}} \I{["\t{ts}aw.kE]} \emph{n., astr.} sun. 
\B{Tom\-pa\-yä ka\-to | Tsaw\-ke\-yä ka\-to | Ka\-tot tä\-ftxu oel | Nì\-ean nì\-rim.} The rhythm of the rain | The rhythm of the sun | I weave the rhythm | In blue and yellow.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
\B{Fä\-za'u tsaw\-ke krr\-pe?} When will the sun come up?~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Tsaw\-ke fpxe\-rä\-kìm (nem\-fa taw).} The sun is rising.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsawkenay}{\B{tsaw\-ke\-nay}}, Alpha Centauri B. 
\on{2}.~\HL{tsawksyul}{\B{tsawk\-syul}}, sun lily.

% Tsawkenay
\hi 
\HT{tsawkenay}{\B{Tsawke\cp{nay}}} \I{[\t{ts}aw.kE."naj]} \emph{n., astr.} Alpha Centauri B. 
(\ding{43}~\HL{tsawke}{\B{tsaw\-ke}})

% tsawksyul
\hi 
\HT{tsawksyul}{\B{\cp{tsawk}syul}} \I{["\t{ts}awk\textcorner.sjul]} \emph{or} \I{[.sjUl]} \emph{n.} \ding{96} sun lily, \emph{stella lilliam}.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln} 
(\ding{43}~\HL{tsawke}{\B{tsaw\-ke}}, \HL{syulang}{\B{syu\-lang}})

% tsawl (tsawl slu)
\hi
\HT{tsawl}{\B{tsawl}} \I{[\t{ts}awl]} \textsc{i}. \emph{adj.} big (in size or stature); tall.
\B{Kel\-ku nge\-yä lu tsawl; pum oe\-yä lu hì\-'i.} Your house is large; mine is small.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Frìp, frìp, ma Evi\-tsyìp, To\-ruk\-ä tsawl\-a kxa!} The big mouth of the Great Leonopteryx bites, bites, children.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}
\B{Oel hu Txe\-wì trr\-am na'\-rìng\-it tar\-mok, tso\-le\-'a syep\-tu\-tet a\-tsawl fra\-to mì sì\-rey.} Yesterday I was with Txewì in the forest and we saw the biggest Trapper I've ever seen.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#New_York_Times}{wiki}
\B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
\B{Po oe\-to lu tsawl.} He's taller than I am.~\LINK{http://naviteri.org/2014/03/value-and-worth/}{b} \\
\textsc{ii}. \B{tsawl slu} \I{[\t{ts}awl slu]} \emph{vin.} grow. 
\B{Fì\-ut\-ral paw kil\-van\-lok nì\-'aw. Tsawl slu nì\-win nì\-txan.} This tree only grows (i.e., germinates, develops) near a river. It grows (i.e., gets big) very quickly.~\LINK{http://naviteri.org/2018/10/aysrr-ayvospxi-ayzisikrr-days-months-seasons/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsawlhI'}{\B{tsawl\-hì'}}, size. 
\on{2}.~\HL{tsawlultxa}{\B{tsawl\-ul\-txa}}, large gathering, conference. 
\on{3}.~\HL{tsawlapxangrr}{\B{tsawl\-apxa\-ngrr}}, unidelta tree. 
\on{4}.~\HL{tsawsngem}{\B{tsaw\-sngem}}, muscular. 
\on{5}.~\HL{txantsawl}{\B{txan\-tsawl}}, huge, giant.
\on{6}.~\HL{tsawtsray}{\B{tsaw\-tsray}}, small or medium\hyp{}sized city.
\on{7}.~\HL{flrrtsawl}{\B{flrr\-tsawl}}, sailfin goliath. 

% tsawlapx
\hi
\HT{tsawlapx}{\B{\cp{tsawl}apx}} \I{["\t{ts}awl.ap']} : \HL{tsawlapxangrr}{\B{tsawl\-apxa\-ngrr}}

% tsawlapxangrr
\hi
\HT{tsawlapxangrr}{\B{\cp{tsawl}apxangrr}} \I{["\t{ts}awl.a.p'a.N\s{r}]} \emph{n.}, \ding{96} uni\-del\-ta tree, ``tall large root'', \emph{magellum deltoids}. (\emph{short form} \ding{43}~\B{tsawl\-apx}) \LINK{http://forum.learnnavi.org/language-updates/ewll/}{ln}
(\ding{43}~\HL{tsawl}{\B{tsawl}}, \HL{apxa}{\B{a\-pxa}}, \HL{ngrr}{\B{ngrr}})

% tsawlhì'
\hi
\HT{tsawlhI'}{\B{tsawl\cp{hì'}}} \I{[\t{ts}awl."hIP]} \emph{n.} size. 
(\ding{43}~\HL{tsawl}{\B{tsawl}}, \HL{hI'i}{\B{hì\-'i}})

% tsawlultxa 
\hi
\HT{tsawlultxa}{\B{tsawlul\cp{txa}}} \I{[\t{ts}awl.ul."t'a]} \emph{or} \I{[.Ul.]} (\textsc{rn}: \B{tsawlùl\cp{txa}} \I{[\t{ts}awl.Ul."da]}) \emph{n.} large gathering, conference.   
\B{Krr\-ka tsawl\-ul\-txa vo\-spxì\-am.} During the great meeting last month.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}
\B{Tsawl\-ul\-txa\-maw a Ay\-lì\-'u}, Post\hyp{}meetup Words.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
\B{Tsa\-tsawl\-ul\-txa\-ri alu SETI\-kon a\-2ve, no\-lu\-me oe nì\-txan te\-ri tì\-tsun\-slu tì\-rey\-ä a hi\-fkey\-mì a\-la\-he.} At the \textsc{seti}con \textsc{ii} conference I learned a lot about the possibility of life on other worlds.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{tsawl}{\B{tsawl}}, \HL{ultxa}{\B{ul\-txa}})

% tsawn
\hi 
\HT{tsawn}{\B{tsawn}} \I{[\t{ts}awn]} \emph{vtr.} gather growing food from the forest; pick; (\emph{in agriculture}) harvest. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zIskrrtsawn}{\B{zì\-skrr\-tsawn}}, fall, autumn. 
\on{2}.~\HL{sAtsawn}{\B{sä\-tsawn}}, harvest.

% tsawng
\hi 
\HT{tsawng}{\B{tsawng}} \I{[\t{ts}awN]} \emph{vin.} shatter, break into pieces. 
\B{Ma sem\-pu, oey yom\-yo tso\-lawng!} Daddy, my plate broke!~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
\B{Ma En\-tu, ngal lum\-pe ngey tsmu\-ke\-yä yom\-yot tsey\-ko\-lawng?} En\-tu, why did you break your sister's plate?~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
\B{Kxew! Oey tsko tsìl\-mawng!} Damn! My bow just broke into pieces!~\LINK{https://naviteri.org/2025/11/kaltxi-nimun-hello-again/}{b}

% tsawsngem
\hi
\HT{tsawsngem}{\B{tsaw\cp{sngem}}} \I{[\t{ts}aw."sNEm]} \emph{adj.} muscular.
\B{Ak\-wey ke lu tsaw\-sngem kaw\-'it slä lu say\-rìp nì\-txan.} Akwey isn't at all muscular but he's very handsome.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{tsngem}{\B{tsngem}}, \HL{tsawl}{\B{tsawl}})

% tsawtsray
\hi
\HT{tsawtsray}{\B{\cp{tsaw}tsray}} \I{["\t{ts}aw.\t{ts}Raj]} \emph{n.} small or medium\hyp{}sized city.
(\ding{43}~\HL{tsawl}{\B{tsawl}}, \HL{tsray}{\B{tsray}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txantsawtsray}{\B{txan\-tsaw\-tsray}}, large city, metropolis.

% tsä'
\hi
\HT{tsA'}{\B{tsä'}} \I{[\t{ts}\ae{}P]} \emph{vin.} squirt. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txumtsA'wll}{\B{txum\-tsä'\-wll}}, baja tickler.

% tsäfpìlyewn
\hi 
\B{tsä\cp{fpìl}yewn} \I{[\t{ts}\ae."fpIl.jEwn]}~:~\HL{tIsAfpIlyewn}{\B{tì\-sä\-fpìl\-yewn}}

% tse
\hi
\HT{tse}{\B{tse}} \I{[\t{ts}E]} \emph{intj.} well … (\emph{to indicate resumption of conversation, or introduce a remark}). 
\B{Tse … za\-'u fra\-'u ne tu\-te lem\-wey\-pey, ke\-fyak?} Well, everything comes to the patient people, right?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}

% tse'a
\hi
\HT{tse'a}{\B{tse\cp{'a}}} \I{[\t{ts}E."Pa]} \emph{vtr.} (\texttt{-}\emph{control}) see (\emph{physical sense}).
\B{Pe\-ut tse\-'a ngal?} What do you see?~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Oel tse\-'a 'a\-'aw\-a tu\-tet.} I see several people.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Fpìr\-mìl oel fu\-ta ay\-nga na\-tsew tsi\-ve\-'a fì\-'ut.} I was just thinking that you might want to see this.~\LINK{http://forum.learnnavi.org/language-updates/the-evidential-at-last/msg106055/\#msg106055}{ln} 
\B{Spaw oel fu\-ta Mo\-'at\-ìl tso\-le\-'a Ney\-ti\-rit.} I believe Mo\-'at saw Ney\-ti\-ri.~\LINK{http://naviteri.org/2011/03/word-order-and-case-marking-with-modals/}{b}
\B{Ayfo tsä\-pa\-ye\-'a.} They will see themselves.~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln}
(\emph{a. idiom}) \B{Tse\-'a tìk yaw\-ne.} `Love at first sight.'~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}
(\ding{43}~\HL{nIn}{\B{nìn}}, \HL{tIng}{\B{tìng na\-ri}}; \HL{kame}{\B{ka\-me}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tse'atswo}{\B{tse\-'a\-tswo}}, sight, vision. 
\on{2}.~\HL{srese'a}{\B{sre\-se\-'a}}, predict. 
\on{3}.~\HL{uniltsa}{\B{unil\-tsa}}, dream of\slash about.
\on{4}.~\HL{sAtse'a}{\B{sä\-tse\-'a}}, sight.

% tse'atswo
\hi
\HT{tse'atswo}{\B{tse\cp{'a}tswo}} \I{[\t{ts}E."Pa.\t{ts}wo]} \emph{n.} (sense of) sight, vision.
(\ding{43}~\HL{tse'a}{\B{tse\-'a}}, \HL{tsu'o}{\B{tsu\-'o}})

% tsefta / si
\hi 
\HT{tsefta}{\B{\cp{tse}fta}} \I{["\t{ts}E.fta]} \textsc{i}. \emph{n.} vengeance, revenge. 
\B{Omum oel fu\-ta ngal pot ve'\-kì, slä tse\-fta ke lu tì\-'eyng a\-muiä.} I know you hate him, but vengeance is not a proper response.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} \\
\textsc{ii}. \B{\cp{tse}fta si} \I{["\t{ts}E.fta si]} \emph{vin.} take revenge.
\B{Tu\-tan\-ur a eyk\-ta\-nay\-ti tspo\-lang oe\-yä sem\-pul tse\-fta sa\-yi.} My father will take revenge on the man who killed the deputy.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} \\
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{tseftanga'}{\B{tse\-fta\-nga'}}, vengeful. 

% tseftanga'
\hi 
\HT{tseftanga'}{\B{\cp{tse}ftanga'}} \I{["\t{ts}E.fta.NaP]} \emph{adj.} vengeful. 
\B{tu\-te a\-tse\-fta\-nga'}, vengeful person.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} 
\B{ay\-lì\-'u a\-tse\-fta\-nga'}, vengeful words.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b} 
(\ding{43}~\HL{tsefta}{\B{tse\-fta}})

% tseltsul
\hi
\HT{tseltsul}{\B{\cp{tsel}tsul}} \I{["\t{ts}El.\t{ts}ul]} \emph{or} \I{[.\t{ts}Ul]} (\textsc{rn}: \B{\cp{tsel}tsùl} \I{["\t{ts}El.\t{ts}Ul]}) \emph{n.} whitewater rapids (\emph{countable, but only rarely}). 

% Tsenu 
\hi
\B{\cp{Tse}nu} \I{["\t{ts}E.nu]} \emph{n., female name}.
\B{Kal\-txì. Nga lu So\-rewn, ke\-fyak? Oer syaw fko Tse\-nu.} Hi. You're So\-rewn, right? I'm Tse\-nu.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{'En si oe, lor\-a tsa\-fkxi\-le\-ri a\-pxay\-opin so\-lar Tse\-nul srok\-it a\-vo\-zam.} I would guess that Tsenu used a thousand (lit. \on{512}) beads for that beautiful multi\hyp{}colored bib necklace.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b}
\B{Lu Tse\-nur yawn\-yew\-la a lam fwa Va'\-rul pot txì\-yìng.} Tse\-nu is broken hearted that Va'\-ru appears to be about to dump her.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Tse\-nur Lo\-ak ke hä\-nga'.} Loak isn't good for Tsenu.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% tsenga
\hi
\HT{tsenga}{\B{\cp{tse}nga}} \I{["\t{ts}E.Na]} \emph{conj.} where, place where.
\B{Tseng\-a 'aw\-steng\-yä\-pem fì\-me\-kem\-yo lu mek\-tseng.} Where these two walls come together there's a gap.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{tseng}{\B{tse\-nge}}, \HL{a}{\B{a}})

% tseng / tsenge
\hi
\HT{tseng}{\B{tseng}} \I{[\t{ts}EN]} \emph{or} \B{\cp{tse}nge}\I{["\t{ts}E.NE]} \textsc{i}. \emph{n.} place. 
\B{Lu poe tstun\-wi, ul\-te evi\-ru peng el\-tur tì\-txen si a ay\-vur\-it te\-ri 'Rr\-ta, alu tse\-nge a\-stxong nì\-ngay.} She is kind, and tells the kids interesting stories about Earth, a truly strange place.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}  \\
\textsc{ii}. \B{tìng tseng} \I{[tIN \t{ts}EN]} \emph{vin.} back down.  \\
\textsc{iii}. \B{\cp{tseng}pe} \I{["\t{ts}EN.pE]} \emph{or} \B{pe\cp{seng}} \I{[pE."{}sEN]} \emph{inter.} where? what place? 
\B{Nga z(ol)a\-'u ftu tseng\-pe\slash tseng\-pe\-ftu?} Where are you from?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{fItseng}{\B{fì\-tseng(e)}}, here. 
\on{2}.~\HL{tsatseng}{\B{tsa\-tseng(e)}}, there. 
\on{3}.~\HL{tsenga}{\B{tse\-nga}}, where, place where. 
\on{4}.~\HL{tsengo}{\B{tse\-ngo}}, somewhere. 
\on{5}.~\HL{fratseng}{\B{fra\-tseng}}, everywhere. 
\on{6}.~\HL{kawtseng}{\B{kaw\-tseng}}, nowhere.  
\on{7}.~\HL{klltseng}{\B{kll\-tseng}}, position. 
\on{8}.~\HL{numtseng}{\B{num\-tseng}}, school. 
\on{9}.~\HL{mektseng}{\B{mek\-tseng}}, gap, breach. 
\on{10}.~\HL{txeptseng}{\B{txep\-tseng}}, fire pit. 
\on{11}.~\HL{kxamtseng}{\B{kxam\-tseng}}, center. 
\on{12}.~\HL{zongtseng}{\B{zong\-tseng}}, safe place, refuge. 
\on{13}.~\HL{sngeltseng}{\B{sngel\-tseng}}, rubbish place, garbage dump. 
\on{14}.~\HL{sngA'itseng}{\B{sngä\-'i\-tseng}}, beginning, starting position. 
\on{15}.~\HL{fngA'tseng}{\B{fngä'\-tseng}}, restroom.
\on{16}.~\HL{txurtseng}{\B{txur\-tseng}}, fortification, fortress, bulwark.
\on{17}.~\HL{wantseng}{\B{wan\-tseng}}, (temporary) hiding place.

% tsengo
\hi
\HT{tsengo}{\B{\cp{tse}ngo}} \I{["\t{ts}E.No]} \emph{n.} somewhere.
\B{Sti\-wi rä\-'ä si, ma 'e\-veng! U\-van si mì se\-ngo a\-la\-he.} Don't be naughty, child! Play somewhere else.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
(\ding{43}~\HL{tseng}{\B{tse\-nge}}) 

% tseo
\hi
\HT{tseo}{\B{\cp{tse}o}} \I{["\t{ts}E.o]} \emph{n.} art.
\B{tseo tì\-ru\-sol\-ä}, art of singing.~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
\B{Fì\-tseo\-ri ke tsun kaw\-tu pi\-vä\-hem tì\-yo'\-ne; tsran\-ten tì\-pä\-hem\-ä tì\-fmi nì\-'aw.} In this art it's impossible to arrive at perfection; the only thing that matters is the attempt to arrive there.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tseotu}{\B{tseo\-tu}}, artist. 
\on{2}.~\HL{pamtseo}{\B{pam\-tseo}}, music. 
\on{3}.~\HL{reltseo}{\B{rel\-tseo}}, visual art.
\on{4}.~\HL{tseovi}{\B{tseo\-vi}}, work of art.

% tseotu
\hi
\HT{tseotu}{\B{\cp{tse}otu}} \I{["\t{ts}E.o.tu]} \emph{n.} artist (\emph{generic term}).
\B{Fu\-ri\-a lo\-lu oer tsa\-skxom a tseo\-tur mok\-ri\-yä srung si, nit\-ram 'efu oe nì\-txan.} I am very happy that I had the chance to help the voice artist.~\LINK{http://naviteri.org/2015/05/cirque-du-soleils-toruk-the-first-flight-teaser-video-online/comment-page-1/\#comment-19897}{b}
(\ding{43}~\HL{tseo}{\B{tseo}}, \HL{-tu}{\B{-tu}})

% tseovi
\hi 
\HT{tseovi}{\B{\cp{tse}ovi}} \I{["\t{ts}E.o.vi]} \emph{n.} work of art. 
\B{Fì\-fkxi\-le lu tseo\-vi a\-lor.} This necklace is a beautiful work of art.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{tseo}{\B{tseo}}) 

% tseri
\hi
\HT{tseri}{\B{\cp{tse}ri}} \I{["\t{ts}E.Ri]} \emph{vtr.} note, notice.
\B{Pey\-ral\-ä mik\-tsang\-it a\-mip ngal tso\-le\-ri srak?} Did you notice Pey\-ral's new earring?~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Oe\-ru txoa li\-vu. Ke tso\-le\-rä\-ngi oel fu\-ta nga\-ri kxe\-tse eo oe lu.} Forgive me. I didn't notice that your tail was in front of me.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{214}~\HL{yAkx}{\B{yäkx}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItseri}{\B{tì\-tse\-ri}}, awareness, notice. 
\on{2}.~\HL{tIktseri}{\B{tìk\-tse\-ri}}, unawareness, lack of notice. 
\on{3}.~\HL{sAtseri}{\B{sä\-tse\-ri}}, observation, something noticed.

% tsewtx
\hi
\HT{tsewtx}{\B{tsewtx}} \I{[\t{ts}Ewt']} \emph{adj.} dirty. 
(\ding{214}~\HL{laro}{\B{la\-ro}})

% tsiki
\hi 
\HT{tsiki}{\B{\cp{tsi}ki}} \I{["\t{ts}i.ki]} \emph{n., fauna} reef tick.~\LINK{https://forum.learnnavi.org/language-updates/expansion-to-flora-fauna-and-places-from-the-valley-of-mo'ara/}{ln}

% tsim
\hi
\HT{tsim}{\B{tsim}} \I{[\t{ts}im]} \emph{n.} source.
\B{Fpìl oel fu\-ta tsun ay\-nga tsli\-vam teyng\-ta tsa\-me\-stxo za\-'u ftu pe\-sim.} I think that you can understand which source those two names come from.~\LINK{http://naviteri.org/2015/09/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise-2/}{b} \\  
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{letsim}{\B{le\-tsim}}, original, unique. 
\on{2}.~\HL{nItsim}{\B{nì\-tsim}}, originally. 
\on{3}.~\HL{ingyentsim}{\B{ing\-yen\-tsim}}, mystery, riddle. 
\on{4}.~\HL{sngumtsim}{\B{sngum\-tsim}}, worrisome matter.   
\on{5}.~\HL{yayayrtsim}{\B{ya\-yayr\-tsim}}, s.t. confusing.
\on{6}.~\HL{paytsim}{\B{pay\-tsim}}, water fountain.
\on{7}.~\HL{lawnoltsim}{\B{law\-nol\-tsim}}, source of (great) joy.
\on{8}.~\HL{tItxeymsim}{\B{tìtxeymtsim}}, interest (\emph{countable}).

% tsin 
\hi 
\HT{tsin}{\B{tsin}} \I{[\t{ts}in]} \emph{n.} (\emph{a. body part}) nail. 
(\emph{b. of an animal}) claw. 
\B{Pa\-lu\-kan\-ìl oe\-yä po\-ti fa tsin tso\-lä\-ngupx.} Unfortunately my cat scratched her with its claw.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b} 

% Tsireya
\hi 
\HT{Tsireya}{\B{Tsi\cp{rey}a}} \I{[\t{ts}i."REj.a]} \emph{n. female name} (\emph{short form:} \ding{43}~\HL{Reya}{\B{Re\-ya}}).

% tsì-
\hi
\HT{tsI-}{\B{tsì-}} \I{[\t{ts}I.]} \emph{num. pref.} four (\emph{base form of} \ding{43}~\HL{tsIng}{\B{tsìng}} \emph{to form higher numbers or fractions}). 
\B{\cp{tsì}vol}, \on{32} (decimal), \on{40} (octal); 
\B{tsì\cp{pxì}}, one\hyp{}fourth, $\frac{1}{4}$; 
\B{tsì\-pxì a\-pxey}, three\hyp{}fourth, $\frac{3}{4}$.

% tsìk
\hi
\HT{tsIk}{\B{tsìk}} \I{[\t{ts}Ik\textcorner]} \emph{adv.} suddenly, without warning. 
\B{'Uol ik\-ran\-it txo\-pu sley\-ko\-la\-tsu, ta\-lun\-a po tsìk ya\-wo.} Something must have frightened the banshee, because it suddenly took to the air.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}

% tsìlpey
\hi
\HT{tsIlpey}{\B{tsìl\cp{pey}}} \I{[\t{ts}Il."pEj]} \emph{n.} hope (\emph{abstract idea}).
\B{Tsìl\-pey\-lu\-ke ke tsun kaw\-tu ri\-vey.} No one can live without hope.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tsìl\-pey\-ìl tok txe'\-lan\-it.} Hope lives within the heart.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{sIlpey}{\B{sìl\-pey}}, \HL{sAsIlpey}{\B{sä\-sìl\-pey}})

% tsìltsan
\hi 
\HT{tsIltsan}{\B{tsìl\cp{tsan}}} \I{[\t{ts}Il."\t{ts}an]} \emph{n} good (\emph{abstract concept}), goodness. 
\B{Tì\-kawng a su\-tel ngop var ri\-vey, tsìl\-tsan\-it pxìm kll\-yem fkol fe\-yä tä\-rem\-hu.} The evil that men do lives after them, The good is oft interred with their bones [\emph{Julius Caesar, Act III, Sc. 2}].~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}  
(\ding{43}~\HL{sIltsan}{\B{sìl\-tsan}})

% tsìng
\hi
\HT{tsIng}{\B{tsìng}} \I{[\t{ts}IN]} \emph{num., adj.} four. (\emph{base}: \ding{43}~\HL{tsI-}{\B{tsì-}}, \emph{rest}: \ding{43}~\HL{-sIng}{\B{\mbox{-sìng}}})
\B{Ay\-nga\-ru tsìng\-a yo\-ra'\-tut!} May I present the four winners!~\LINK{http://naviteri.org/2014/12/ayngaru-tsinga-yoratu-may-i-present-the-four-winners/}{b}
\B{Tsyokx\-ìri ke lu zek\-wä a\-tsìng ki a\-mrr.} The hand doesn't have four but five digits.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vospxItsIng}{\B{Vo\-spxì\-tsìng}}, April.

% tsìsyì 
\hi
\HT{tsIsyI}{\B{\cp{tsì}syì}} \I{["\t{ts}I.sjI]} \emph{vin.} whisper.   
\B{Ru\-txe pi\-vll\-txe nì\-wok nì\-'it, oel nga\-ti stum ke stä\-ngawm krr\-a nga fì\-fya tsye\-rì\-syì!} Please speak up a bit, I can barely hear you when you're whispering like this!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sAtsIsyI}{\B{sä\-tsì\-syì}}, whisper.
\on{2}.~\HL{nItsyIsyI}{\B{nì\-tsì\-syì}}, by whispering.

% tsìve
\hi
\HT{tsIve}{\B{\cp{tsì}ve}} \I{["\t{ts}I.vE]} \emph{adj., ord. num.} fourth.

% tsìvol
\hi
\HT{tsIvol}{\B{\cp{tsì}vol}} \I{["\t{ts}I.vol]} \emph{num., adj.} \on{32} (decimal), \on{40} (octal).

% tskaha
\hi
\B{\cp{Tska}ha} \I{["\t{ts}ka.ha]} \emph{n.} \emph{name, family name}.
\B{Ay\-nge\-nga\-ru ohe\-yä tsmuk\-it alu Ne\-wey te Tska\-ha So\-rewn\-'ite.} Allow me to introduce my sister, Ne\-wey te Tska\-ha So\-rewn\-'ite. (highly formal)~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}

% tskalep
\hi
\HT{tskalep}{\B{\cp{tska}lep}} \I{["\t{ts}ka.lEp\textcorner]} \emph{n.}, \ding{246} crossbow.
\B{Maw txan\-tom\-pa, pxay\-a rìk\-äo la\-mu tska\-lep pe\-yä, ha tsat ku\-lat ay\-oel.} After the rainstorm, his crossbow was under a lot of leaves, so we uncovered it.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Tska\-lep\-it oel to\-lìng oe\-yä tsmu\-kan\-ur alu Ìstaw.} I gave the crossbow to my brother Istaw.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}

% tskawr
\hi 
\HT{tskawr}{\B{tskawr}} \I{[\t{ts}kawR]} \emph{vin.} limp. 
\B{Oel tse\-ri fu\-ta nga tskawr. Sra\-ke ngal ve\-nut tì\-sraw sey\-ko\-li?} I see you're limping. Did you hurt your foot?~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 

% tski
\hi 
\HT{tski}{\B{tski}} \I{[\t{ts}ki]} \emph{n.} jaw. 

% tsko (tsko swizaw)
\hi
\HT{tsko}{\B{tsko}} \I{[\t{ts}ko]} \emph{n.}, \ding{246} bow (weapon). 
\B{Na\-ri si!  Tsa\-tsko lu spu\-win ul\-te ke lu mi txur, slä oe\-ri txan\-wa\-we le\-iu. Lu stxe\-li a sem\-pul\-ta.} Careful! That bow is old and no longer strong, but it means a lot to me. It was a gift from my father.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Tskot sngo\-lä\-'i po si\-var 'a\-'aw\-a trr\-kam.} He started to use the bow several days ago.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Fu\-ri\-a sne\-yä tsko\-ti ngop po kan\-fpìl.} He's concentrating on making his bow.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b} \\
(i.) \B{tsko swi\cp{zaw}} \I{[\t{ts}ko swi."zaw]} \emph{n.}, \ding{246} bow and arrow; archery. 
\B{Na'\-vi\-ri txin\-a sä\-'o tì\-tu\-sa\-ron\-ä lu tsko swi\-zaw.} For the Na'vi the bow and arrow is the main hunting tool.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Fu\-ri\-a fì\-txan fnä\-ngan Ul\-rey\-ìl tsko swi\-zaw\-it, lu oe\-ru fmokx.} I'm jealous of the fact that Ul\-rey is such an excellent archer.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Tsko swi\-zaw\-fa a tì\-ta\-ron\-ìri ze\-ne fko ni\-vu\-me.} You must learn to hunt with a bow and arrow.~\LINK{http://forum.learnnavi.org/language-updates/ordinals-nume/}{ln\slash a\textsuperscript{c}}
\B{Oe tskxe\-keng si sä\-su\-lìn\-ur alu tsko swi\-zaw.} I practice my hobby, which is archery.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}

% tskxaytsyìp
\hi
\HT{tskxaytsyIp}{\B{\cp{tskxay}tsyìp}} \I{["\t{ts}k'aj.\t{ts}jIp\textcorner]} \emph{n.} hail. 
(\ding{43}~\HL{tskxepay}{\B{tskxe\-pay}}, \HL{-tsyIp}{\B{-tsyìp}})

% tskxe
\hi
\HT{tskxe}{\B{tskxe}} \I{[\t{ts}k'E]} \emph{n.} rock, stone. 
\B{Ay\-skxe a mì sa\-sna\-'o ku'\-up lu nì\-txan.} The rocks in that pile are very heavy.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Lì\-'u\-ri alu tskxe pam\-rel fya\-pe?--Pam\-rel\-fya lu na Tsä, KxeKx, E.} How do you spell the word \emph{tskxe}?--It's spelled ts, kx, e.~\LINK{http://naviteri.org/2010/08/a-na\%E2\%80\%99vi-alphabet/}{b} 
\B{He\-tu\-wong\-ìl aw\-nge\-yä swo\-tut ska\-'a, fte kll\-ki\-vu\-lat ke\-ru\-sey\-a tskxet.} The aliens destroy our sacred place to dig up dead rock.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{Fì\-tskxe\-ri fa\-'o lu yey; ke lu ko\-um.} This rock has straight sides; it's not rounded.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{ngul na tskxe}, the drab color of stone.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
(\emph{a. idiom}) \B{Pay\-ìl a lipx tskxe\-ti ripx.} \emph{persistent effort can accomplish unexpected and amazing things} [lit.: `dripping water pierces a stone']~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letskxe}{\B{le\-tskxe}}, stony, of stone.
\on{2}.~\HL{tskxepay}{\B{tskxe\-pay}}, ice. 
\on{3}.~\HL{tskxevi}{\B{tskxe\-vi}}, pebble.
\on{4}.~\HL{lIngtskxe}{\B{lìng\-tskxe}}, unobtanium.
\on{5}.~\HL{tskxemauti}{\B{tskxe\-ma\-u\-ti}}, nut.

% tskxemauti
\hi 
\HT{tskxemauti}{\B{\cp{tskxe}mauti}} \I{["\t{ts}k'E.ma.u.ti]} \emph{n.} nut. 
(\ding{43}~\HL{tskxe}{\B{tskxe}}, \HL{mauti}{\B{mau\-ti}}) 

% tskxekeng / si 
\hi
\HT{tskxekeng}{\B{\cp{tskxe}keng}} \I{["\t{ts}k'E.kEN]} \textsc{i}. \emph{n.} training, exercise.
\B{Tskxe\-keng\-luke ke lu kea tì\-tsan\-'ul.} Without practice there is no improvement.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/}{b}
\B{Mll\-te oe nì\-wotx: fì\-fne\-tskxe\-keng le\-sar lu nì\-txan.} I agree completely: this kind of exercise is very useful.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/comment-page-1/\#comment-9348}{b} \\
\textsc{ii}. \B{\cp{tskxe}keng si} \I{["\t{ts}k'E.kEN si]} \emph{vin.} exercise, train.
\B{Oe tskxe\-keng si sä\-su\-lìn\-ur alu tsko swi\-zaw.} I practice my hobby, which is archery.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{'Ul tskxe\-keng si, 'ul fnan.} The more you practice, the better you'll get.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}

% tskxepay
\hi
\HT{tskxepay}{\B{\cp{tskxe}pay}} \I{["\t{ts}k'E.paj]} \emph{n.} ice. 
\B{Sre fwa sìn tskxe\-pay tì\-ran, ze\-ne fko flì\-nutx\-it sti\-ve\-ftxaw.} It's necessary to check the thickness of the ice before walking on it.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
(\ding{43}~\HL{tskxe}{\B{tskxe}}, \HL{pay}{\B{pay}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tskxaytsyIp}{\B{tskxay\-tsyìp}}, hail.

% tskxevi
\hi
\HT{tskxevi}{\B{\cp{tskxe}vi}} \I{["\t{ts}k'E.vi]} \emph{n.} pebble.
\B{'Eveng\-ìl skxe\-vit 'ey\-krr\-ko sko uvan.} Children roll pebbles as a game.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{tskxe}{\B{tskxe}})

% tslam
\hi
\HT{tslam}{\B{tslam}} \I{[\t{ts}lam]} \emph{vtr.} understand. 
\B{Maw\-krra Tsyeyk ftxo\-lu\-lì\-'u, tslam fra\-pol fu\-ta slu po O\-lo'\-eyk\-tan a\-mip.} After Jake spoke, everyone understood that he had become the new Clan Leader.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Sra\-ke tsun oe fay\-upxa\-ret tsli\-vam nì\-ftue?} Can I understand these messages easily?~\LINK{http://forum.learnnavi.org/language-updates/verb-for-write/msg175592/\#msg175592}{ln}
\B{Hì\-txoa, ke tslolam. Ru\-txe li\-veyn.} Sorry, I didn't get that. Could you repeat it, please?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Ngal ke tslä\-ngam teyng\-ta fì\-tì\-ngä\-zìk\-ìri pe\-fyin\-ep\-'ang.} Unfortunately you don't understand how complex this problem is.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} 
(\emph{a. conv.}) \B{tslo\cp{lam}} ``understood''. 
\B{Oe\-ri lu swoa fne\-txum.--Tslo\-lam.} I am allergic to alcohol.--Understood.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItslam}{\B{tì\-tslam}}, understanding, intelligence. 
\on{2}.~\HL{leytslam}{\B{ley\-tslam}}, appreciate.

% tsleng
\hi
\HT{tsleng}{\B{tsleng}} \I{[\t{ts}lEN]} \emph{adj.} false. 
\B{Fì\-po lu vrr\-tep a mì sokx a\-tsleng!} This one is a demon in a false body!~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Tsu.27tey_attacks_Jake}{wiki\slash a}
(\ding{214}~\HL{ngay}{\B{ngay}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItsleng}{\B{tì\-tsleng}}, falsehood, falsity. 

% tslikx
\hi 
\HT{tslikx}{\B{tslikx}} \I{[\t{ts}lik']} \emph{vin.} crawl. 
(\emph{a. rhyming expression}) \B{Tslikx, tì\-ran, tul; | Ftu yì ne yì tsan\-'ul.} \emph{when learning s.t. new, you have to proceed from step to step: baby steps first, then bigger ones} [lit.: `Crawl, walk, run; from level to level get better.']~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{tslikxyulatopin}{\B{tslikx\-yu la\-to\-pin}}, chamelion crawler.   
\on{2}.~\HL{tslikxyutsawlak}{\B{tslikx\-yu tsaw\-lak}}, scarab crawler.

% tslikxyu latopin
\hi 
\HT{tslikxyulatopin}{\B{\cp{tslikx}yu \cp{la}topin}} \I{["\t{ts}lik'.ju "la.to.pin]} \emph{n., fauna} chamelion crawler. 
(\ding{43}~\HL{tslikx}{\B{tslikx}}, \HL{latem}{\B{la\-tem}}, \HL{'opin}{\B{'opin}})

% tslikxyu tsawlak
\hi 
\HT{tslikxyutsawlak}{\B{\cp{tslikx}yu \cp{tsaw}lak}} \I{["\t{ts}lik'.ju "\t{ts}aw.lak\textcorner]} \emph{n., fauna} scarab crawler. 
(\ding{43}~\HL{tslikx}{\B{tslikx}}, \HL{tsawl}{\B{tsawl}}, \HL{lak}{\B{lak}})

% tsmi
\hi
\HT{tsmi}{\B{tsmi}} \I{[\t{ts}mi]} \emph{n.} nectar. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsmisnrr}{\B{tsmi\-snrr}}, nectar lantern, bladder lantern.

% tsmisnrr
\hi
\HT{tsmisnrr}{\B{\cp{tsmi}snrr}} \I{["\t{ts}mi.sn\s{r}]} \emph{n.} bladder lantern, nectar lantern. 
(\ding{43}~\HL{tsmi}{\B{tsmi}}, \HL{sAnrr}{\B{sä\-nrr}})

% tsmìm
\hi
\HT{tsmIm}{\B{tsmìm}} \I{[\t{ts}mIm]} \emph{n.} track (of animal or human origin).
\B{Äo fì\-ut\-ral lu tsmìm 'ang\-tsìk\-ä.} There's a hammerhead track under this tree.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Tsmìm\-ìri wätx fol ti\-nan\-it nì\-ngay.} They're really bad at reading animal tracks.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Ioang\-ìri, tu\-te\-ri, tsun fko si\-var lì\-'ut alu tsmìm.} One can use the word `track' for animals and people.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/comment-page-1/\#comment-1118}{b}

% tsmuk
\hi
\HT{tsmuk}{\B{tsmuk}} \I{[\t{ts}muk\textcorner]} \emph{or} \I{[\t{ts}mUk\textcorner]} \emph{n.} sibling. 
\B{Fya\-pe fko syaw ngar, ma tsmuk?--Oe\-ru syaw Tsyey\-son.} What are you called, brother (or sister)?--I'm called Jason.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Ay\-nga\-ru sei\-yi ira\-yo, ma smuk!} Thank you all, siblings!~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Ira\-yo, ma pxe\-smuk, ul\-te sey\-kxel sì nit\-ram.} Thank you, siblings, and congratulation!~\LINK{http://naviteri.org/2010/12/navi-writing-contest-the-first-place-winners/}{b}
\B{Ma fra\-po, ay\-nga\-ru oe\-yä tsmuk\-it alu Ne\-wey.} Everybody, please allow me to introduce (to you) my sister, Newey.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\ding{43}~\HL{tsmuktu}{\B{tsmuk\-tu}}, \HL{tsmukan}{\B{tsmu\-kan}}, \HL{tsmuke}{\B{tsmu\-ke}})

% tsmukan
\hi
\HT{tsmukan}{\B{\cp{tsmu}kan}} \I{["\t{ts}mu.kan]} \emph{n.} brother. 
\B{Tsa\-tu\-te a ngal tse\-'a lu tsmu\-kan oe\-yä.} That person you see is my brother.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/comment-page-1/\#comment-15463}{b}
\B{Oe\-yä tsmu\-kan\-ìl mip\-a tsko\-ti kì\-ma\-ne\-i\-om.} My brother just got a new bow, I'm happy to say.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Tska\-lep\-it oel to\-lìng oe\-yä tsmu\-kan\-ur alu Ìstaw.} I gave the crossbow to my brother Istaw.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\ding{43}~\HL{tsmuk}{\B{tsmuk}})

% tsmuke
\hi
\HT{tsmuke}{\B{\cp{tsmu}ke}} \I{["\t{ts}mu.kE]} \emph{n.} sister. 
\B{Ke lu po\-ru kea smu\-ke.} He doesn't have any sisters.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Tsa\-tsmu\-kel alu Ri\-ni mo\-lok fu\-ta oe ngar mu\-wä\-pi\-vìn\-txu.} Sister Rini over there suggested that I introduce myself.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Sa'\-nok\-ìl oe\-yä tsmu\-ket a\-mip no\-lokx trr\-am.} Mom gave birth to my new sister yesterday.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
(\ding{43}~\HL{tsmuk}{\B{tsmuk}})

% tsmuktu
\hi
\HT{tsmuktu}{\B{\cp{tsmuk}tu}} \I{["\t{ts}muk\textcorner.tu]} \emph{or} \I{["\t{ts}mUk\textcorner.]} \emph{n.} sibling.
\B{Rì\-'ìr rä\-'ä si\-vi tsmuk\-tur!} Don't imitate your sibling!~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
(\ding{43}~\HL{tsmuk}{\B{tsmuk}})

% tsnì
\hi
\HT{tsnI}{\B{tsnì}} \I{[\t{ts}nI]} \emph{conj.} that (\emph{to introduce subordinate clauses after the v.s} \ding{43}~\HL{sIlpey}{\B{sìl\-pey}}, \HL{AtxAle}{\B{ätxä\-le si}}, \HL{Azan}{\B{ä\-zan si}}, \HL{rangal}{\B{rang\-al}}, \HL{mowar}{\B{mo\-war si}}, \HL{leymkem}{\B{leym\-kem}}, \HL{leymfe'}{\B{leym\-fe'}}, \HL{fe'pey}{\B{fe'\-pey}}, \HL{la'um}{\B{la\-'um}}, \HL{pe'ngay}{\B{pe'\-ngay}}, \HL{tsantxAl}{\B{tsan\-txäl si}} \emph{and optionally} \HL{srefpIl}{\B{sre\-fpìl}}, \HL{srefey}{\B{sre\-fey}}). 
\B{Ä\-txä\-le si tsnì li\-vu o\-he\-ru U\-nil\-ta\-ron.} I respectfully request the Dream Hunt.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Behind_the_Scenes}{wiki} 
\B{Sìl\-pey oe tsnì fì\-tì\-oeyk\-tìng law li\-vu nga\-ru set.} I hope that this explanation is clear to you now.~\LINK{http://forum.learnnavi.org/language-updates/a-long-answer-to-a-quick-question/}{ln}
\B{Set sre\-fey oe tsnì tsam\-po\-ngu tä\-txaw maw txon\-'ong.} I'm currently expecting the war party back after nightfall.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Poe mo\-war so\-li poan\-ur tsnì hi\-vum.} She advised him to leave.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
\B{Sre\-fpìl Oma\-ti\-ka\-ya tsnì Tsyeyk kaw\-krr ke ta\-yä\-txaw maw ka\-vuk sne\-yä.} The Omaticaya assumed that Jake would never return after his treachery.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\emph{proverb}) \B{Ätxä\-le si pa\-lu\-lu\-kan\-ur tsnì smar\-it li\-vo\-nu.} \emph{a futile gesture, an attempt to achieve something that might be desirable but will clearly not happen} [lit.: Ask a thanator to release its prey.]~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}

% tsngal
\hi
\HT{tsgnal}{\B{tsngal}} \I{[\t{ts}Nal]} \emph{n.} cup. 
\B{Nge\-yä tsngal lum\-pe lu mek? Näk nì\-'ul ko!} Why is your cup empty? Drink up!~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{'Ä'! \mbox{Oel} tsngal\-it tì\-mung\-zup.} Whoops! I just dropped the cup.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tawtsngal}{\B{taw\-tsngal}}, panopyra.

% tsngan
\hi
\HT{tsngan}{\B{tsngan}} \I{[\t{ts}Nan]} \emph{n.} meat. 
\B{Ke\-tsran tu\-tel 'i\-vem, tsa\-fne\-tsngan lu ftxì\-vä'.} That kind of meat is gross no matter who cooks it.~\LINK{http://naviteri.org/2013/03/whoever-whatever-whenever/}{b}
\B{Fol tsngan\-it pxì\-mo\-lun\-'i nì\-'eng.} They shared the meat.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Fol tsngan\-it pxì\-mo\-lun\-'i nì\-yll.} They shared the meat with the entire clan.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Ke new oe mi\-vay' tsngan\-ti a 'o\-lem Ri\-nil.} I don't want to taste the meat that Rini cooked.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Fì\-tsngan\-ur a sno\-läm lä\-ngu fa\-hew a\-kxä\-näng.} This rotten meat has a putrid smell.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}

% tsngawpay
\hi
\HT{tsngawpay}{\B{\cp{tsngaw}pay}} \I{["\t{ts}Naw.paj]} \emph{n.} tears.
\B{Mì Na'\-rìng lu tsngaw\-pay.} There are tears in the forest.~\LINK{http://naviteri.org/2017/07/way-tiretua-the-shamans-song/}{b}
(\ding{43}~\HL{tsngawvIk}{\B{tsngaw\-vìk}}, \HL{pay}{\B{pay}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsngawpayvi}{\B{tsngaw\-pay\-vi}}, teardrop.

% tsngawpayvi
\hi
\HT{tsngawpayvi}{\B{\cp{tsngaw}payvi}} \I{["\t{ts}Naw.paj.vi]} \emph{n.} teardrop.
(\ding{43}~\HL{tsngawpay}{\B{tsngaw\-pay}})

% tsngawvìk
\hi
\HT{tsngawvIk}{\B{\cp{tsngaw}vìk}} \I{["\t{ts}Naw.vIk\textcorner]} \emph{vin.} cry, weep. 
\B{Sko Sa\-hìk ke tsun oe mì\-ftxe\-le tsngi\-vaw\-vìk, sko sa'\-nok tsun.} As Tsahik, I cannot weep over this matter, as mother I can.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Fpxa\-mo\-a fì\-kem\-ìl a\-fpxa\-mo ay\-oe\-ti tsngey\-ko\-law\-vìk, tsa\-krr ley\-ko\-leym\-kem.} This terrible, terrible action has made us cry, then protest.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}

% tsngem
\hi
\HT{tsngem}{\B{tsngem}} \I{[\t{ts}NEm]} \emph{n.} muscle. 
\B{Lu pa'\-lir sngem a\-txur.} A direhorse has strong muscles.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsawsngem}{\B{tsaw\-sngem}}, muscular.

% tson
\hi
\HT{tson}{\B{tson}} \I{[\t{ts}on]} \emph{n.} obligation, duty, imposed requirement, task.
\B{Lä\-ngu oe\-ru tson a fì\-fmawn\-it pi\-veng ngar.} I'm sorry that I'm obliged to tell you this news.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Za\-'u tsa\-tson ta Ey\-wa.} That obligation comes from Eywa.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Lu Ney\-ti\-rir ta Mo\-'at a tson a kar Tsyeyk\-ur ay\-fya\-'ot Na'\-vi\-yä.} Neytiri is under obligation by\slash from Mo'at to teach Jake the ways of the Na'vi.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b} 
\B{Po tì\-kang\-kem so\-li val nì\-txan fte tsa\-tson\-ur ha\-sey si\-vi.} She worked very hard to complete the task.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nItson}{\B{nì\-tson}}, dutifully, as an obligation. 
\on{2}.~\HL{tsonta}{\B{tson\-ta}}, to (\emph{with} \HL{kxIm}{\B{kxìm}})

% tsonta
\hi
\HT{tsonta}{\B{\cp{tson}ta}} \I{["\t{ts}on.ta]} \emph{conj.} to (with \ding{43}~\HL{kxIm}{\B{kxìm}}), (be obliged) to do a task; (\emph{short form of} \B{tson\-it a})
\B{Ay\-eveng\-ur kxo\-lìm sa'\-nok\-ìl tson\-ta pay\-it za\-mu\-nge.} The children were told by their mother to fetch water.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
(\ding{43}~\HL{tson}{\B{tson}})

% tsong 
\hi
\HT{tsong}{\B{tsong}} \I{[\t{ts}oN]} \emph{n.} valley.   
\B{Aw\-nga tsong\-ne ki\-vä fte sti\-var\-sìm tey\-lut.} Let's go to the valley to gather beetle larvae.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsongropx}{\B{tsong\-ropx}}, hole, cavity. 
\on{2}.~\HL{tsongtsyIp}{\B{tsong\-tsyìp}}, dimple.

% tsongropx
\hi
\HT{tsongropx}{\B{\cp{tsong}ropx}} \I{["\t{ts}oN.Rop']} \emph{n.} hole, cavity, excavation with a bottom (\emph{visible or presumed}).
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang. Fì\-tì\-kan\-ìri ropx ke ha'.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators. For this purpose, a hole going right through the tree trunk isn't suitable.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{tsong}{\B{tsong}}, \HL{ropx}{\B{ropx}})

% tsongtsyìp
\hi
\HT{tsongtsyIp}{\B{\cp{tsong}tsyìp}} \I{["\t{ts}oN.\t{ts}jIp\textcorner]} \emph{n.} (\emph{of the body}) dimple. 
\B{Prr\-nen lrr\-tok si a krr, srer me\-song\-tsyìp a\-ho\-na.} When the baby smiles, two adorable dimples appear.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{tsong}{\B{tsong}}, \HL{-tsyIp}{\B{-tsyìp}})

% tsopì 
\hi
\HT{tsopI}{\B{\cp{tso}pì}} \I{["\t{ts}o.pI]} \emph{n.} lung (\emph{of the body}). 

% tspang
\hi
\HT{tspang}{\B{tspang}} \I{[\t{ts}paN]} \emph{vtr.} kill. 
\B{Sna\-fpìl\-fya\-ri le\-Na'\-vi krra smar\-it fkol tspang, tsran\-ten nì\-txan fwa po ke ngä\-'än nì\-kel\-kin.} It's important in Na'vi philosophy that the prey not suffer unnecessarily when it's killed.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
\B{Tsa\-ik\-ran\-ìri ta\-lun\-a new nga\-ti tspi\-vang, law lu fwa mi ke lu zä\-fi.} Since that ikran wants to kill you, it's clear it's still not tame.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Fì\-po\-ti oel tspì\-yang, fte tì\-ke\-nong li\-ye\-vu ay\-la\-ru.} I will kill this one as a lesson to the others.~\LINK{http://forum.learnnavi.org/language-updates/capturing-avatar/}{ln\slash a\textsuperscript{c}}
\B{Tsa\-ye\-rik\-tsyìp\-ìl li 'an\-la sa'\-nok\-it a fkol tspì\-mang.} That little hexapede is already yearning for its mother that's just been killed.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\emph{a. idiom}) \B{Ke tsun fko tspi\-vang to\-ruk\-it fa fwa pewn\-ti snew.} \emph{Proverbial expression for a method that will not work.} [lit.: You can't kill a great leonopteryx by constricting its throat]~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tspangyu}{\B{tspang\-yu}}, killer.

% tspanyu
\hi 
\HT{tspangyu}{\B{\cp{tspang}yu}} \I{["\t{ts}paN.ju]} \emph{n.} killer. 
\B{Tsa\-tu\-tan\-ur a fkol tspo\-lang lo\-lu ta'\-leng a\-kll\-vawm; tspang\-yur lu pum a\-teyr.} That person that was killed had brown skin, the killer had white.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{tspang}{\B{tspang}}) 

% tspìng
\hi
\HT{tspIng}{\B{tspìng}} \I{[\t{ts}pIN]} \emph{n., fauna} austrapede. 

% tspu'
\hi 
\HT{tspu'}{\B{tspu'}} \I{[\t{ts}puP]} \emph{or} \I{[\t{ts}pUP]} (\emph{a. vin.}) extend (\emph{occupying a certain place, stretching to a certain point}). 
\B{Atx\-kxe ay\-oe\-yä tspu' nuä ay\-ram.} Our land extends beyond the mountains.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b}
(\emph{b. vtr.}) hold out (\emph{often to offer}).
\B{Pol tspo\-lu' oer u\-tu\-ma\-ut\-it tsa\-krr oel mu\-nge.} She held a banana fruit out to me and I took it.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b}

% tsranten
\hi
\HT{tsranten}{\B{\cp{tsran}ten}} \I{["\t{ts}Ran.tEn]} \emph{vin.} matter, be of importance\slash import. 
\B{Ton\-ìri a\-la\-he, aw\-ngal yom wu\-tsot teng\-krr he\-reyn nì\-pxim, teng\-krr te\-ru\-von, ke tsran\-ten.} As for all other nights, it doesn't matter whether we eat dinner while sitting upright or leaning.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
\B{Fì\-tseo\-ri ke tsun kaw\-tu pi\-vä\-hem tì\-yo'\-ne; tsran\-ten tì\-pä\-hem\-ä tì\-fmi nì\-'aw.} In this art it's impossible to arrive at perfection; the only thing that matters is the attempt to arrive there.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
\B{Nga oer tsran\-ten fra\-to.} You matter the most to me.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letsranten}{\B{le\-tsran\-ten}}, important. 
\on{2}.~\HL{ketsran}{\B{ke\-tsran}}, whatever, whoever, however.

% tsray
\hi
\HT{tsray}{\B{tsray}} \I{[\t{ts}Raj]} \emph{n.} village. 
\B{Mì sray apxa kel\-ku si so\-a\-i\-a\-hu.} (He) lives with his family in a large village.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsawtsray}{\B{tsaw\-tsray}}, small or medium\hyp{}sized city. 
\on{2}.~\HL{txantsawtsray}{\B{txan\-tsaw\-tsray}}, large city, metropolis.

% tsre'i
\hi
\HT{tsre'i}{\B{\cp{tsre}'i}} \I{["\t{ts}RE.Pi]} \emph{vtr.} throw (\B{ne}, to, towards; \HL{wA}{\B{wä}}, at). 
\B{Oel tsre\-'i tskxet ne nga.} I throw a stone towards you.~\LINK{http://forum.learnnavi.org/language-updates/to-throw-at-x/}{ln}
\B{Oel tsre\-'i tskxet nga\-ru.} I throw a stone to you.~\LINK{http://forum.learnnavi.org/language-updates/to-throw-at-x/}{ln}
\B{Oel tsre\-'i tskxet wä nga.} I throw a stone at you.~\LINK{http://forum.learnnavi.org/language-updates/to-throw-at-x/}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{kxapaytsre'}{\B{kxa\-pay\-tsre'}}, spit.

% tsrul
\hi
\HT{tsrul}{\B{tsrul}} \I{[\t{ts}Rul]} \emph{or} \I{[\t{ts}RUl]} \emph{n.} nest;  \emph{protected area serving as the home of Pandoran fauna}.
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Ro\-lun ay\-oel tsrul\-mì hì\-'i\-a pxe\-lo\-it a\-teyr.} We found three little white eggs in the nest.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yayotsrul}{\B{ya\-yo\-tsrul}}, bird's nest.

% tstal
\hi
\HT{tstal}{\B{tstal}} \I{[\t{ts}tal]} \emph{n.} knife. 
\B{El\-tu si! Tsa\-tstal a\-fwem lu litx nì\-txan.} Pay attention! That blunt knife is very sharp.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Oe\-yä tstal\-ìl tok pe\-se\-nget? -- Lu yo\-sìn.} Where's my knife? -- It's on the table.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b}
\B{Oel tstal\-it fyo\-lep fa aysre'.} I held the knife in my teeth.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tstalsena}{\B{tstal\-se\-na}}, knife sheath.

% tstalsena
\hi 
\HT{tstalsena}{\B{\cp{tstal}sena}} \I{["\t{ts}tal.sE.na]} \emph{n.} knife sheath.
\B{Pol tstal\-it wrr\-zo\-lä\-rìp tstal\-se\-na\-ftu.} He pulled the knife out of its sheath.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 
(\ding{43}~\HL{tstal}{\B{tstal}}, \HL{sAhena}{\B{sä\-he\-na}})

% tstew 
\hi
\HT{tstew}{\B{tstew}} \I{[\t{ts}Ew]} \emph{adj.} (\emph{of a person}) brave. 
\B{Tsam\-si\-yu a\-sìl\-tsan lu tstew slä\-kop swa\-ran.} A good warrior is courageous but also humble.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
\B{Kem a\-tì\-tstew\-nga' si tu\-te a\-tstew.} A brave person does brave deeds.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
\B{'Ur fki\-van tstew!} Do not show fear. [lit.: ``May your appearance come to the senses as brave!'']~\LINK{https://naviteri.org/2026/01/follow-up-to-txep-si-txeva/}{b\slash a\on{3}} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItstew}{\B{tì\-tstew}}, courage, bravery. 
\on{2}.~\HL{txantstew}{\B{txan\-tstew}}, hero. 
\on{3}.~\HL{nItstew}{\B{nì\-tstew}}, bravely, courageously.

% tstir
\hi 
\HT{tstir}{\B{tstir}} \I{[\t{ts}tiR]} \emph{n.} palm of a hand. 

% tstu
\hi
\HT{tstu}{\B{tstu}} \I{[\t{ts}tu]} \textsc{i}. \emph{adj.} (\emph{a.}) closed, shut.    (\emph{b. of the weather}) completely overcast, covered with clouds. \\
\textsc{ii}. \B{tstu si} \I{[\t{ts}tu si]} \emph{vin.} close. 
\B{Oe\-ri me\-na\-ri\-ru tstu si.} I close my eyes.~\LINK{http://forum.learnnavi.org/language-updates/open-close/}{ln} \\
\textsc{iii}. \B{tstu sä\cp{pi}} \I{[\t{ts}tu s\ae."pi]} \emph{vin.} close itself\slash by itself\slash on its own.
\B{Po\-ri me\-na\-ri tstu sä\-po\-li.} His eyes closed.~\LINK{http://forum.learnnavi.org/language-updates/open-close/}{ln}

% tstunkem / si
\hi
\HT{tstunkem}{\B{\cp{tstun}kem}} \I{["\t{ts}tun.kEm]} \emph{or} \I{["\t{ts}tUn.]} \emph{coll. often pronounced} \I{["\t{ts}tuN.]}  \textsc{i}. \emph{n.} favor, act of kindness.  
\B{Tung oer fu\-ta vin tstun\-kem\-it nga\-ta.} Let me ask you a favor.~\LINK{http://naviteri.org/2020/03/mipa-sawasultsyip-a-new-contest/}{b}
(\ding{43}~\HL{tstunwi}{\B{tstu\-nwi}}, \HL{kem}{\B{kem}}) \\
\textsc{ii}. \B{\cp{tstun}kem si} \I{["\t{ts}tun.kEm si]} \emph{or} \I{["\t{ts}tUn.]} \emph{vin.} do a favor.
\B{Tstun\-kem si oer ru\-txe.} Please do me a favor.~\LINK{http://naviteri.org/2020/03/mipa-sawasultsyip-a-new-contest/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tstunkemtsyIp}{\B{tstun\-kem\-tsyìp}}, little favor.

% tstunkemtsyìp
\hi
\HT{tstunkemtsyIp}{\B{\cp{tstun}kemtsyìp}} \I{["\t{ts}tun.kEm.\t{ts}jIp\textcorner]} \emph{or} \I{["\t{ts}tUn.]} \emph{coll. often pronounced} \I{["\t{ts}tuN.]} \emph{n.} a little favor, act of kindness. 
(\ding{43}~\HL{tstunkem}{\B{tstun\-kem}}, \HL{-tsyIp}{\B{-tsyìp}})

% tstunlan
\hi 
\HT{tstunlan}{\B{\cp{tstun}lan}} \I{["\t{ts}tun.lan]} \emph{or} \I{["\t{ts}tUn.]} \emph{adj.} kind\hyp{}hearted. 
(\ding{43}~\HL{tstunwi}{\B{tstun\-wi}}, \HL{txe'lan}{\B{txe'\-lan}}; \ding{214}~\HL{kawnglan}{\B{kawng\-lan}})

% tstunwi
\hi
\HT{tstunwi}{\B{\cp{tstun}wi}} \I{["\t{ts}tun.wi]} \emph{or} \I{["\t{ts}tUn.]} \emph{adj., ofp.} kind, thoughtful, considerate.  
\B{Le\-iu ay\-nga tstun\-wi nì\-txan nì\-wotx.} You are all very kind.~\LINK{http://forum.learnnavi.org/news-announcements/happy-birthday-to-karyu-pawl/msg556899/\#msg556899}{ln} 
\B{Lu nga tstun\-wi\-a tsmuk a\-tstun\-wi.} You are a truly kind sibling.~\LINK{http://naviteri.org/2013/03/tsanerul-feerul-getting-better-getting-worse/comment-page-1/\#comment-2136}{b}
(\emph{a. idiom}) \B{Tstun\-wi.} ``It's kind of you to say that.'' (\emph{in response to a compliment}).~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} 
(\ding{43}~\HL{tItstunwinga'}{\B{tì\-tstun\-wi\-nga'}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItstunwi}{\B{tì\-tstun\-wi}}, kindness. 
\on{2}.~\HL{tstunkem}{\B{tstun\-kem}}, favor, act of kindness.
\on{3}.~\HL{tstunlan}{\B{tstun\-lan}}, kind\hyp{}hearted.
\on{4}.~\HL{ketstun}{\B{ke\-tstun}}, unkind, emotionally cold.

% tstxa'a
\hi
\HT{tstxa'a}{\B{\cp{tstxa}'a}} \I{["\t{ts}t'a.Pa]} \emph{n.}, \ding{96} Canalyd, \emph{canalydium limacineum}.

% tstxo 
\hi
\HT{tstxo}{\B{tstxo}} \I{[\t{ts}t'o]} \emph{n.} name.
\B{Nga\-ri tstxo lu pe\-lì\-'u?} What is your name?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Tstxo oe\-yä lu X.} My name is X.~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln}
\B{Tstxo nge\-yä su\-nu oer nì\-ngay.} I truly like your name.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/comment-page-1/\#comment-9344}{b}
\B{Pol tstxo\-ti oe\-yä pol\-txe\-i\-e nì\-tsì\-syì.} He whispered my name, I'm happy to say.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tstxolI'u}{\B{tstxo\-lì\-'u}}, noun. 
\on{2}.~\HL{tstxolI'ukIngvi}{\B{tstxo\-lì\-'u\-kìng\-vi}}, noun phrase.

% tstxolì'u
\hi
\HT{tstxolI'u}{\B{\cp{tstxo}lì'u}} \I{["\t{ts}t'o.lI.Pu]} \emph{n.} noun. 
\B{Sra\-ne, lì\-'ul alu mì fra\-krr ley\-ka\-tem lì\-'ut a\-hay tsa\-fya--fwa li\-vu tstxo\-lì\-'u, li\-vu la\-he\-a fne\-lì\-'u ke tsran\-ten.} Yes, the word \emph{in} always changes the following word like that--whether it be a noun or another kind of word doesn't matter.~\LINK{http://forum.learnnavi.org/language-updates/fmawno/msg293595/\#msg293595}{ln}
(\ding{43}~\HL{tstxo}{\B{tstxo}}, \HL{lI'u}{\B{lì\-'u}})

% tstxolì'ukìngvi
\hi 
\HT{tstxolI'ukIngvi}{\B{tstxo\cp{lì}'ukìngvi}} \I{[\t{ts}t'o."lI.Pu.kIN.vi]} \emph{n.} noun phrase (\emph{a phrase whose pivotal or central word is a noun}). 
\B{Nì\-fkey\-to\-ngay, tsa\-pxe\-lì\-'u alu zì\-sìt a\-mip le\-fpom ke lu tstxo\-lì\-'u\-kìng\-vi.} In fact those three words \emph{zì\-sìt a\-mip le\-fpom} are not a noun phrase.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/\#comment-27964}{b} 
(\ding{43}~\HL{tstxo}{\B{tstxo}}, \HL{lI'ukIngvi}{\B{lì\-'u\-kìng\-vi}})

% tsu'o
\hi
\HT{tsu'o}{\B{\cp{tsu}'o}} \I{["\t{ts}u.Po]} \emph{n.} ability.   
\B{Tì\-ru\-sol\-ìri ke lu po\-ru kea tsu\-'o kaw\-'it.} As for singing, he has no ability whatsoever.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b} 
\B{Ze\-ne fko tsko swi\-zaw\-it si\-var nì\-trr\-trr fte\-ke fì\-tsu\-'o sni\-väm.} One must use a bow and arrow regularly to prevent this ability degrading over time.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{-tswo}{\B{-tswo}})

% Tsu'tey
\hi
\B{Tsu'\cp{tey}} \I{[\t{ts}uP."tEj]} \emph{n.} \emph{male name}.
\B{Ma Tsu'\-tey, kem\-pe si nga?} Tsu'\-tey, what are you doing?~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Neytiri_talking_to_Tsu.27tey}{wiki\slash a}
\B{Tsu'\-tey\-ìl to\-lìng oer mawl\-it smar\-ä.} Tsu'\-tey gave me a half of the prey.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Fo lu 'ey\-lan Tsu'\-tey\-ä.} They're Tsu'\-tey's friends.~\LINK{http://naviteri.org/2011/07/number-in-na\%E2\%80\%99vi/}{b}
\B{Tsu'\-tey\-ri lu fo\-ru nrr\-a a fì\-re\-won yo\-lo\-ra'.} They're proud of Tsu'\-tey for having won this morning.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}

% tsuk-
\hi
\HT{tsuk}{\B{tsuk-}} \I{[\t{ts}uk\textcorner.]} \emph{or} \I{[\t{ts}Uk\textcorner.]} (\textsc{rn}: \B{tsùk-} \I{[\t{ts}Uk\textcorner.]})\emph{prod. pref. on verbs to form receptive ability, i.e., s.t. is capable of receiving the action of the verb}, ``…-able'' .
\B{Fì\-i\-o\-ang lu tsuk\-yom.} This animal is edible\slash can be eaten.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}
\B{Tsuk\-yom\-a ioang lu le\-sar.} An edible animal is useful.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}
(\emph{a. more prod. than in English}) \B{Lu na'\-rìng tsuk\-ha\-haw.} One can sleep in the forest.\slash It's possible to sleep in the forest.\slash The forest is ``sleepable''.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}
\B{Fì\-tseng lu tsuk\-tsu\-rokx.} One can rest here.\slash It's possible to rest here.\slash This place is ``restable''.~\LINK{http://naviteri.org/2011/03/\%E2\%80\%9Creceptive-ability\%E2\%80\%9D-and-hesitation/}{b}
(\ding{214}~\HL{ketsuk-}{\B{ke\-tsuk-}}) 

% tsukanom
\hi 
\HT{tsukanom}{\B{tsu\cp{ka}nom}} \I{[\t{ts}u."ka.nom]} (\textsc{rn}: \B{tsù\cp{ka}nom} \I{[\t{ts}U."ka.nom]}) \emph{adj.} available, obtainable. 
\B{Tsay\-fa\-suk tsu\-ka\-nom lu krr\-ka fì\-zì\-sì\-krr nì\-'aw.} Those berries are available during this season only.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 
\B{Wä fì\-sä\-spxin a 'um\-tsa le\-iu set tsu\-ka\-nom.} Medicine against this disease is happily now available.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 
(\ding{43}~\HL{tsuk}{\B{tsuk-}}, \HL{kanom}{\B{ka\-nom}}; \ding{214}~\HL{ketsukanom}{\B{ke\-tsu\-ka\-nom}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItsukanom}{\B{tì\-tsu\-ka\-nom}}, availability.

% tsukmong
\hi 
\HT{tsukmong}{\B{tsuk\cp{mong}}} \I{[\t{ts}uk\textcorner."moN]} \emph{or} \I{[\t{ts}Uk\textcorner.]} (\textsc{rn}: \B{tsùk\cp{mong}} \I{[\t{ts}Uk\textcorner."moN]}) \emph{adj.} reliable, dependable. 
\B{tu\-te a\-tsuk\-mong}, dependable person.~\LINK{http://naviteri.org/2022/06/solalew-maul-zisita-half-the-year-is-over/}{b} 
\B{ay\-sì\-oeyk\-tìng a\-tsuk\-mong}, reliable explanations.~\LINK{http://naviteri.org/2022/06/solalew-maul-zisita-half-the-year-is-over/}{b} 
(\ding{43}~\HL{mong}{\B{mong}})

% tsuksìm
\hi
\HT{tsuksIm}{\B{\cp{tsuk}sìm}} \I{["\t{ts}uk\textcorner.sIm]} \emph{or} \I{[\t{ts}Uk\textcorner.]} \emph{n.} chin. 
\B{ean na tsuk\-sìm to\-ruk\-ä}, blue like the chin of a great leonopteryx.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}

% tsukx
\hi 
\HT{tsukx}{\B{tsukx}} \I{[\t{ts}uk']} \emph{or} \I{[\t{ts}Uk']} (\textsc{rn}: \B{tsùkx} \I{[\t{ts}Uk']}) \emph{vtr.} stab. 
\B{Ney\-ti\-ril nan\-tang\-it tso\-lukx fte pe\-yä tì\-sraw\-ti 'ey\-ki\-vi\-'a.} Neytiri stabbed the viperwolf to end its pain.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}

% tsulfä 
\hi
\HT{tsulfA}{\B{tsul\cp{fä}}} \I{[\t{ts}ul."f\ae]} \emph{or} \I{[\t{ts}Ul.]} (\textsc{rn}: \B{tsùl\cp{fä}} \I{[\t{ts}Ul."f\ae]}) \textsc{i}. \emph{n.} mastery.
\B{Pxe\-nge\-yä tsul\-fä lì'\-fya\-yä le\-Na'\-vi oe\-ru te\-ya si.} Your mastery of the Na'vi language fills me with joy.~\LINK{http://naviteri.org/2010/12/navi-writing-contest-the-first-place-winners/}{b}
(\emph{a. idiom}) \B{Nga\-ru tsul\-fä.} ``Thank you.'' [lit.: to you the mastery] (\emph{in response to a compliment, in a situation where the s.o. praising you is better at the activity you're being praised for than you are yourself}).~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} \\
\textsc{ii}. \B{tsul\cp{fä} si} \I{[\t{ts}ul."f\ae{} si]} (\textsc{rn}: \B{tsùl\cp{fä} si} \I{[\t{ts}Ul."f\ae{} si]}) \emph{vin.} master. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsulfAtu}{\B{tsul\-fä\-tu}}, master of an art.

% tsulfätu
\hi
\HT{tsulfAtu}{\B{tsul\cp{fä}tu}} \I{[\t{ts}ul."f\ae.tu]} \emph{or} \I{[\t{ts}Ul.]} (\textsc{rn}: \B{tsùl\cp{fä}tu} \I{[\t{ts}Ul."f\ae.tu]}) \emph{n.} master of an art, craft, or skill; expert.
\B{Tew\-ti, nga lu tsul\-fä\-tu i\-'en\-ä.} Wow, you are a master on the \emph{i'en}.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Fì\-me\-rel\-it 'o\-ngo\-lop aw\-nge\-yä tsul\-fä\-tul rel\-tseo\-ä alu Älìn.} The two portraits were created by our Master of Visual Arts, Alan.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b} 
\B{Fay\-sul\-fä\-tu\-ä tì\-kang\-kem ohe\-ru meuia lu\-yu nì\-ngay.} The work of these masters truly honors me.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
(\ding{43}~\HL{tsulfA}{\B{tsul\-fä}}, \HL{-tu}{\B{-tu}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsulfAtunay}{\B{tsul\-fä\-tu\-nay}}, near\hyp{}master.

% tsulfätunay
\hi
\HT{tsulfAtunay}{\B{tsulfätu\cp{nay}}} \I{[\t{ts}ul.f\ae.tu."naj]} \emph{or} \I{[\t{ts}Ul.]} (\textsc{rn}: \B{tsùlfätun\cp{ay}} \I{[\t{ts}Ul.fE.tu."naj]}) \emph{n.} near\hyp{}master, \emph{s.o. who is one step away from becoming a master}.
(\ding{43}~\HL{tsulfAtu}{\B{tsul\-fä\-tu}})

% tsun 1
\hi
\HT{tsun}{\B{tsun}}\textsuperscript{\on{1}} \I{[\t{ts}un]} \emph{or} \I{[\t{ts}Un]} (\textsc{rn}: \B{tsùn} \I{[\t{ts}Un]}) \emph{vin., modal} be able, can.
\B{Tsun oe nga\-hu nì\-Na'\-vi pi\-väng\-kxo a fì\-'u oe\-ru prr\-te' lu.} It's a pleasure to be able to chat with you in Na'vi.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki}
\B{Tsat ke tsun oe ron\-srel\-ngi\-vop.} I can't imagine that.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Txo tsi\-ve\-'a ay\-ngal key\-eyt, ru\-txe oe\-ru pi\-veng fte tsi\-vun oe sa\-'ut ley\-ki\-va\-tem.} If you see mistakes, please tell me so that I can change them.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Lo\-lu tok\-tor Kì\-rey\-sì kar\-yu a\-sìl\-tsan ul\-te po\-e tso\-lun nì\-Na'\-vi pi\-vll\-txe.} Doctor Grace was a good teacher and she could speak Na'vi.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 
(\emph{a. idiom}) \B{Tsun ti\-vam.} ``That's fine.'' [lit.: can suffice]~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Grace_and_Norm_in_the_Link_room}{wiki\slash a} 
\B{Nga tsun ni\-väk wew\-a pay\-it. Ul\-te ke ra\-you! -- Tse\-i\-un ti\-vam.} You can drink cold water. And not get drunk! -- That's just fine.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} 
\B{Tsun pe\-hem?} (\emph{short for} \B{Tsun fko pe\-hem si\-vi?}) ``What can one do?'' 
\B{Ri\-ni yaw\-ne lu oer, slä oe yaw\-ne ke lä\-ngu por kaw\-'it.---Tsun pe\-hem?} I'm in love with Rini, but she doesn't love me one bit.---What are you gonna do? That's life.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
(\emph{b. proverb}) \B{Ke tsun fko tspi\-vang to\-ruk\-it fa fwa pewn\-ti snew.} \emph{Proverbial expression for a method that will not work.} [lit.: You can't kill a great leonopteryx by constricting its throat]~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} 
\B{Txìm a\-'aw ke tsun hi\-veyn mì tal me\-fa'\-li\-yä.} You can't take both positions or sit on the fence; you need to decide. [lit.: One butt can't sit on the backs of two direhorses]~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsunslu}{\B{tsun\-slu}}, be possible. 

% tsun 2
\hi 
\HT{tsun2}{\B{tsun}}\textsuperscript{\on{2}} \I{[\t{ts}un]} \emph{or} \I{[\t{ts}Un]} \emph{n.} heel. 
\B{Oe\-ri teng\-krr te\-rul mì na'\-rìng, tsun\-it a\-ski\-en tì\-sraw sey\-ko\-li.} While I was running in the forest, I hurt my right heel.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}  

% tsunslu
\hi
\HT{tsunslu}{\B{\cp{tsun}slu}} \I{["\t{ts}un.slu]} \emph{or} \I{["\t{ts}Un.]} (\textsc{rn}: \B{\cp{tsùn}slu} \I{["\t{ts}Un.slu]}) \emph{vin.} may, be possible.
\B{Kehe, ke tsun\-slu fwa yem me\-syon\-lì\-'ut uo tstxo\-lì\-'u.} No, it's not possible to put two adjectives after a noun.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/\#comment-27964}{b} 
(\ding{43}~\HL{tsun}{\B{tsun}}, \HL{slu}{\B{slu}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItsunslu}{\B{tì\-tsun\-slu}}, possibility.
\on{2}.~\HL{letsunslu}{\B{le\-tsun\-slu}}, possible.

% tsup
\hi 
\HT{tsup}{\B{tsup}} \I{[\t{ts}up\textcorner]} \emph{or} \I{[\t{ts}Up\textcorner]} \emph{n.} chasm, ravine. 
(\emph{a. proverb}) \B{Rä\-'ä zup nem\-fa tsup.} \emph{don't fall into the trap that s.o. has laid for you} [lit.: ``don't fall into the chasm''].~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b}

% tsupx
\hi 
\HT{tsupx}{\B{tsupx}} \I{[\t{ts}up']} \emph{or} \I{[\t{ts}Up']} (\textsc{rn}: \B{tsùpx} \I{[\t{ts}Up']}) \emph{vtr.} scratch harmfully (\emph{as with a claw}). 
\B{Pa\-lu\-kan\-ìl oe\-yä po\-ti fa tsin tso\-lä\-ngupx.} Unfortunately my cat scratched her with its claw.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b} 
(\ding{43}~\HL{pItIk}{\B{pì\-tìk}}) 

% tsurak 
\hi
\HT{tsurak}{\B{\cp{tsu}rak}} \I{["\t{ts}u.Rak\textcorner]} \emph{n., fauna} Skimwing (\emph{temperamental Pandoran sea creature}). 

% tsurokx 
\hi
\HT{tsurokx}{\B{tsu\cp{rokx}}} \I{[\t{ts}u."Rok']} \emph{vin.} rest. 
\B{Sa'\-nur le\-ru haw\-tsyìp. Tsi\-vu\-rokx ko.} Mommy is taking a nap. Let her rest.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{New oe ri\-vun a\-sim tì\-fnu\-nga'\-a tseng\-it a tsa\-ro tsun syi\-vor tsi\-vu\-rokx fte spä\-pi\-veng.} I want to find a quiet place nearby where I can chill out and rest to get my head back on straight.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}

% tswa' 
\hi
\HT{tswa'}{\B{tswa'}} \I{[\t{ts}waP]} \emph{vtr.} forget.
\B{Tok pe\-se\-nget pam\-rel\-ìl? Tsat ngal tswo\-lä\-nga' nì\-lam. -- Nì\-fkey\-to\-ngay ke tswo\-la' kaw\-'it.} Where is the text? Apparently you've forgotten it. -- As a matter of fact, I haven't.~\LINK{http://naviteri.org/2014/06/teri-tsaliu-alu-nifkeytongay-about-nifkeytongay/}{b}
\B{Fay\-lì\-'u a\-lor oe\-ru te\-ya si nì\-txan, ul\-te ke tswa\-ya' oel sat kaw\-krr.} These beautiful words fill me with joy, and I will never forget them.~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
(\emph{a. idiom}) \B{Teng\-krr ftxo\-zä se\-rei\-yi aw\-nga, ke tswi\-va' ay\-lom\-tu\-ti ko!} A toast: To absent friends. [lit.: While we are celebrating, let us not forget those who we wish could also be here (but can't)]~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}

% tswayon  
\hi
\HT{tswayon}{\B{\cp{tsway}on}} \I{["\t{ts}waj.on]} (\emph{a. vin.}) fly (\HL{fa}{\B{fa}}, with\slash by means of). 
\B{Oe tsway\-on io Ay\-vit\-ra\-yä Ra\-mu\-nong.} I fly over the Tree of Souls.~\LINK{http://forum.learnnavi.org/language-updates/over-under-and-quickly/}{ln}
\B{Ri\-ti tswo\-lay\-on ftum\-fa slär.} The stingbat flew out of the cave.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
\B{Tseyk tswa\-may\-on fa ik\-ran sre\-krr; ta\-fral fmo\-li fì\-kem si\-vi fa to\-ruk nì\-steng.} Jake flew with an ikran before; therefore he tried to do it with a toruk in a similar fashion.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\HT{tsweykayon}{(\emph{b. \ding{246} as})} \B{swi\-zaw\-ti tswey\-kay\-on}, shoot an arrow.
\B{Swi\-zaw\-ti tswey\-kay\-on ne\-fä ne taw, tse\-nga zup ke lu law.} I shot an arrow into the air, | It fell to earth, I know not where.~\LINK{http://naviteri.org/2017/03/titusemteri-concerning-shooting/}{b} 
(\emph{c. idiom}) \B{Nge\-yä ik\-ran kaw\-krr ke tswi\-vay\-on nì\-'aw\-tu.} May your banshee never fly alone.~\LINK{http://forum.learnnavi.org/language-updates/idiomatic-expressions/}{ln}  \\
(i.) \B{txaw\-nul\-srung a tsway\-on}, airplane.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}
\on{1}.~\HL{sAtswayon}{\B{sätswayon}}, flight.

% tswal
\hi 
\HT{tswal}{\B{tswal}} \I{[\t{ts}wal]} \emph{n.} power (\emph{physical, psychological, emotional, or political}). \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{letswal}{\B{le\-tswal}}, powerful (\emph{ofp.}). 
\on{2}.~\HL{tswalnga'}{\B{tswal\-nga'}}, powerful (\emph{nfp.}).

% tswalnga'
\hi 
\HT{tswalnga'}{\B{\cp{tswal}nga'}} \I{["\t{ts}wal.NaP]} \emph{adj., nfp.} powerful. 
(\ding{43}~\HL{tswal}{\B{tswal}}, \HL{letswal}{\B{le\-tswal}})

% tswesya / si
\hi 
\HT{tswesya}{\B{\cp{tswe}sya}} \I{["\t{ts}wE.sja]} \textsc{i}. \emph{n.} current. \\
\textsc{ii}. \B{\cp{tswe}sya si} \I{["\t{ts}wE.sja si]} \emph{vin.} flow.
\B{Na\-ri si, ma 'itan. Kil\-van tswe\-sya si nì\-win nì\-txan.} Be careful, son. The river is flowing very swiftly.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}

% tswin
\hi
\HT{tswin}{\B{tswin}} \I{[\t{ts}win]} \emph{n.} queue, braid (\emph{neural queue of the Na'vi to enable them to make} \B{tsa\-hey\-lu}). 
\B{Nga\-ri tswin\-tsyìp se\-vin nì\-txan lu nang!} What a pretty little queue you have!~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\ding{43}~\HL{kuru}{\B{ku\-ru}})

% tswìk
\hi
\HT{tswIk}{\B{tswìk}} \I{[\t{ts}wIk\textcorner]} \emph{vtr.} suck. \\
(i.) \B{tswìk \cp{kxe}nerit} \I{[\t{ts}wIk\textcorner{} "k'E.nE.Rit\textcorner]} \emph{vin.} smoke (\emph{a cigarette})
\B{Ke tung Na'\-vil fu\-ta tswìk kxe\-ner\-it Mo\-'a\-ra\-ka nì\-wotx.} The Na'vi have banned smoking in Mo'ara.~\LINK{http://naviteri.org/2017/06/ayliu-a-ta-eywaeveng-kifkey-uniltirantokxa-words-from-disneys-pandora-the-world-of-avatar/}{b} \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{reyptswIk}{\B{reyp\-tswìk}}, wolf tick.

% -tswo
\hi
\HT{-tswo}{\B{-tswo}} \I{[.\t{ts}wo]} \emph{prod. suf. on verbs and non\hyp{}verbal elements of} \B{si}\hyp{}\emph{constructs to indicate the ability to perform the action of the verb.} 
\B{Po\-ri wem\-tswo fra\-tsam\-si\-yur ro\-lo\-'a nì\-txan.} His ability to fight greatly impressed all the warriors.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Kxa\-ri tstu\-tswo tsran\-ten krra ke lu kea sä\-fpìl le\-sar.} When one has no useful thoughts, the ability to close one's mouth is important.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
(\ding{43}~\HL{tsu'o}{\B{tsu\-'o}})

% tsyafe
\hi
\HT{tsyafe}{\B{\cp{tsya}fe}} \I{["\t{ts}ja.fE]} \emph{adj.} mild, moderate, comfortable.
\B{Ya\-ri som\-wew\-pe (set)? Ya lu tsya\-fe.} What's the temperature? It's nice.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} 

% tsyal
\hi
\HT{tsyal}{\B{tsyal}} \I{[\t{ts}jal]} \emph{n.} wing.
\B{me\-syal\-hu a ik\-ran}, an ikran with two wings.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#More_extracts_from_various_emails}{wiki}
\B{Kun\-sìp\-ìri txan\-a tì\-meyp lu tsyal a mìn.} The gunship's main weakness is the rotor system. [lit.: wing that turns]~\LINK{http://forum.learnnavi.org/language-updates/ftarpa-and-skienpa-added-to-dict/msg570362/\#msg570362}{ln\slash a\textsuperscript{c}} \\
\textsc{Derviations}: \ding{43} 
\on{1}.~\HL{lortsyal}{\B{lor\-tsyal}}, shimmyfly.
\on{2}.~\HL{txeptsyal}{\B{txep\-tsyal}}, coronis.
\on{3}.~\HL{slotsyal}{\B{slo\-tsyal}}, stormglider.

% tsyang
\hi 
\HT{tsyang}{\B{tsyang}} \I{[\t{ts}jaN]} \emph{n.} swarm (\emph{of insects and metaphorically of people, annoying and threatening}). 
\B{tsyang ay\-hì\-'ang\-ä}, a swarm of insects.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
\B{tsyang su\-te\-yä}, a swarm of people.~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b} 
(\ding{43}~\HL{sna}{\B{sna-}})

% tsyär
\hi
\HT{tsyAr}{\B{tsyär}} \I{[\t{ts}j\ae{}R]} \emph{vtr.} reject.
\B{Ngal pe\-lun fay\-stxe\-nut fra\-krr tsyär?} Why do you always reject these offers?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Ngal lum\-pe oe\-yä stxe\-nut tsyo\-lä\-ngär?} Why did you reject my offer?~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b} 
(\ding{214}~\HL{mll'an}{\B{mll\-'an}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItsyAr}{\B{tì\-tsyär}}, rejection.

% tsyey
\hi
\HT{tsyey}{\B{tsyey}} \I{[\t{ts}jEj]} \emph{n.} snack, light meal.
\B{Ke 'efu oe ohakx nì\-hawng; tam tsyey.} I'm not too hungry; a snack will do.~\LINK{http://naviteri.org/2015/07/aylifyavi-lereyfya-1-cultural-terms-1/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsyeytsyIp}{\B{tsyey\-tsyìp}}, tiny bite. 
\on{2}.~\HL{niktsyey}{\B{nik\-tsyey}}, food wrap.

% tsyeytsyìp
\hi
\HT{tsyeytsyIp}{\B{\cp{tsyey}tsyìp}} \I{["\t{ts}jEj.\t{ts}jIp\textcorner]} \emph{n.} tiny bite.
(\ding{43}~\HL{tsyey}{\B{tsyey}}, \HL{-tsyIp}{\B{-tsyìp}})

% tsyeym
\hi
\HT{tsyeym}{\B{tsyeym}} \I{[\t{ts}jEjm]} \emph{n.} treasure, \emph{s.t. rare and of great value}.
\B{Tsyan\-ìri sì oe\-ri lo\-le\-iu tsa\-kaym tsyeym a\-ngay.} There was a true treasure that evening for John and I.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}

% tsyìl
\hi
\HT{tsyIl}{\B{tsyìl}} \I{[\t{ts}jIl]} \emph{vtr.} climb, scale.   
\B{Tsyìl Ik\-ni\-ma\-yat ul\-te tsa\-heyl si ik\-ran\-hu a fì\-'u lu tì\-fme\-tok a ze\-ne fra\-ta\-ron\-yu a\-'e\-wan em\-zi\-va\-'u.} Scaling Iknimaya and bonding with a banshee is a test that every young hunter must pass.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Ke tsun oe fì\-'awkx\-it tsyi\-vìl.} I can't scale this cliff.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Sìn yì sngä\-'i\-yu\-ä, slä tsye\-rìl (hay\-a yì\-ne) nì\-'ul\-'ul.} They're at a beginner's level, but they're getting better and better.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}

% tsyìmawnun'i
\hi 
\HT{tsyImawnun'i}{\B{tsyìmawnun\cp{'i}}} \I{[\t{ts}jI.maw.nun."Pi]} \emph{or} \I{[.nUn.]} (\textsc{rn}: \B{tsyìmawnùn\cp{'i}} \I{[\t{tS}I.maw.nUn."Pi]}) \emph{n., shortened form of} \ding{43}~\HL{tIsyImawnun'i}{\B{tì\-syì\-maw\-nun\-'i}}, independence,  freedom from a pre\hyp{}existing relationship. 
\B{Ze\-ne fra\-'e\-veng tsyì\-maw\-nun\-'it a ta sa'\-sem nì\-'i'\-a mi\-vu'\-ni.} Every child must eventually achieve independence from his or her parents.~\LINK{http://naviteri.org/2023/07/trr-tsyimawnuniya-lefpom-happy-independence-day/}{b} 
\B{Trr Tsyì\-maw\-nun\-'i\-yä Le\-fpom!} Happy Independence Day!~\LINK{http://naviteri.org/2023/07/trr-tsyimawnuniya-lefpom-happy-independence-day/}{b} 
(\ding{43}~\HL{syImawnun'i}{\B{syì\-maw\-nun\-'i}}) 

% -tsyìp
\hi
\HT{-tsyIp}{\B{-tsyìp}} \I{[.\t{ts}jIp\textcorner]} \emph{prod. suf. to n.s and pronouns to form the diminutive form}.
(\emph{a. to form new lexical items that originated as small versions of a n.}) e.g. \ding{43} \HL{puktsyIp}{\B{puk\-tsyìp}}, booklet, pamphlet;  
\HL{utraltsyIp}{\B{ut\-ral\-tsyìp}}, bush; 
\HL{sAspxintsyIp}{\B{sä\-spxin\-tsyìp}}, minor ailment.
(\emph{b. to express affection or endearment})
\B{Za\-'u fì\-tseng, ma 'ite\-tsyìp.} Come here, little daughter.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Kam\-tsyìp\-ìl wu\-tsot ye\-rom.} Little Kamun is having dinner.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Nga\-tsyìp yaw\-ne lu oer.} I love you, little one.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Nga\-ri tswin\-tsyìp se\-vin nì\-txan lu nang!} What a pretty little queue you have!~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
(\emph{c. to express disparagement or insult})
\B{Nga nì\-aw\-no\-mum to oe\-tsyìp lu txur nì\-txan.} As everyone knows, you're a lot stronger than little old me.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Nga\-tsyìp\-ìl new pe\-ut ta oe?} What does little you want from me?~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}
\B{Fì\-ta\-ron\-yu\-tsyìp ke tsun ke\-'ut sti\-vä'\-nì.} This little hunter can't catch anything.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b}

% tsyo
\hi
\HT{tsyo}{\B{tsyo}} \I{[\t{ts}jo]} \emph{n.} flour. \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tsyosyu}{\B{tsyo\-syu}}, food made from flour. 
\on{2}.~\HL{tsyorina'wll}{\B{tsyo\-ri\-na'\-wll}}, Cycad.

% tsyokx
\hi
\HT{tsyokx}{\B{tsyokx}} \I{[\t{ts}jok']} \emph{n.} hand. 
\B{Oe\-ri nì\-'i'a tsyokx zo\-slo\-lu.} My hand is finally healed.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Oe\-ri skxir a mì syokx tì\-sraw se\-ngi.} The wound on my hand hurts.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Slä lu 'a\-'aw\-a tì\-ke\-teng, nat\-ke\-nong, tsyokx\-ìri ke lu zek\-wä a\-tsìng ki a\-mrr.} But there are several differences, for example, the hand doesn't have four but five digits.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Txur\-tel\-mì fo fta so\-li fte\-ke ka tsyokx fwi\-vi.} They tied a knot in the rope so it wouldn't slip through their hands.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{hawntsyokx}{\B{hawn\-tsyokx}}, glove. 
\on{2}.~\HL{taksyokx}{\B{tak\-syokx}}, clap hands.
\on{3}.~\HL{tsyokxtil}{\B{tsyokx\-til}}, wrist.
\on{4}.~\HL{rumtsyokx}{\B{rum\-tsyokx}}, fist.
\on{5}.~\HL{rIktsyokx}{\B{rìk\-tsyokx}}, flat hand.

% tsyokxtil
\hi 
\HT{tsyokxtil}{\B{\cp{tsyokx}til}} \I{["\t{ts}jok'.til]} \emph{usually} \I{["\t{ts}jok\textcorner.til]} \emph{n.} wrist. 
(\ding{43}~\HL{tsyokx}{\B{tsyokx}}, \HL{til}{\B{til}})

% tsyorina'wll
\hi
\HT{tsyorina'wll}{\B{tsyori\cp{na'}wll}} \I{[\t{ts}jo.ri."naP.w\s{l}]} \emph{n.}, \ding{96} Cycad, ``flour seed plant'', \emph{pseudocycas altissima}. 
(\ding{43}~\HL{tsyo}{\B{tsyo}}, \HL{rina'}{\B{ri\-na'}}, \HL{'ewll}{\B{'e\-wll}})

% tsyosyu
\hi
\HT{tsyosyu}{\B{\cp{tsyo}syu}} \I{["\t{ts}jo.sju]} \emph{n.} food made from flour.
(\ding{43}~\HL{tsyo}{\B{tsyo}}, \HL{syuve}{\B{syu\-ve}})

% tsyul
\hi 
\HT{tsyul}{\B{tsyul}} \I{[\t{ts}jul]} \emph{or} \I{[\t{ts}jUl]} \emph{vtr.} begin, start. 
\B{Pol tì\-kang\-kem\-it tsyo\-lul.} He began the work.
\B{Tì\-kang\-kem tsyä\-po\-lul.} The work began.~\LINK{http://naviteri.org/2018/08/fmawnti-stolawm-srak-have-you-heard-the-news/}{b}
(\ding{43}~\HL{sngA'i}{\B{sngä'i}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tItsyul}{\B{tì\-tsyul}}, beginning, start.

% Tsyungwen
\hi
\HT{tsyungwen}{\B{\cp{Tsyung}wen}} \I{["\t{ts}juN.wEn]} \emph{or} \I{[."\t{ts}jUN.]} (\textsc{rn}: \B{\cp{Tsyùng}wen} \I{["\t{tS}UN.wEn]}) \emph{n.} \textsc{lw} Mandarin Chinese.

\end{multicols}
