\pdfbookmark[0]{U}{U}
\begin{center}
\section*{U \I{[u\slash U]}}
\end{center}

\begin{multicols}{2}

% ue'
\hi 
\HT{ue'}{\B{u\cp{e'}}} \I{[u."{}EP]} (\emph{a. vin.}) vomit.
\B{Oey nan\-tang\-tsyìp o\-lu\-e' ta\-lu\-na yom nì\-hawng.} My dog vomited because it ate too much.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b}
(\emph{b. vtr.}) vomit up. 
\B{Prr\-nen\-ìl wu\-tsot o\-lu\-e'.} The baby vomited up its meal.~\LINK{http://naviteri.org/2016/06/mrrvola-lifyavi-amip-forty-new-expressions/}{b} 

% uk
\hi
\HT{uk}{\B{uk}} \I{[uk\textcorner]} \emph{or} \I{[Uk\textcorner]} (\textsc{rn}: \B{ùk} \I{[Uk\textcorner]}) \emph{n.} shadow.
\B{vawm na uk}, dark\hyp{}shadow color.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{toruk}{\B{to\-ruk}}, leonopteryx, the Last Shadow.
\on{2}.~\HL{ukyom}{\B{uk\-yom}}, eclipse.

% ukyom / si
\hi 
\HT{ukyom}{\B{\cp{uk}yom}} \I{["{}uk\textcorner.jom]} \emph{or} \I{["{}Uk\textcorner.]} (\textsc{rn}: \B{\cp{ùk}yom} \I{["{}Uk\textcorner.jom]}) \textsc{i}. \emph{n.} eclipse. \\ 
\textsc{ii}. \B{\cp{uk}yom si} \I{["{}uk\textcorner.jom si]} \emph{vin.} to eclipse s.t. (\emph{also metaphorically})
\B{Po\-ri tsa\-kxey\-ey a\-'aw fra\-kem\-ur a\-muiä uk\-yom so\-lä\-ngi.} Sadly, that one mistake eclipsed all of his good deeds.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b} 
(\ding{43}~\HL{uk}{\B{uk}}, \HL{yom}{\B{yom}}) 

% ukxo
\hi
\HT{ukxo}{\B{u\cp{kxo}}} \I{[u."k'o]} \emph{adj.} dry. 
\B{Ta\-kuk set tsaw\-ke. Sley\-ku fra\-'ut u\-kxo.} Now the sun bears down. (It) makes everything dry.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg475008/\#msg475008}{ln}
\B{Rìk a\-ukxo mì hu\-fwe\-tsyìp ka\-fi sar\-mi.} The dry leaf was sailing in the breeze.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b}
(\ding{214}~\HL{mei}{\B{mei}})

% ulte
\hi
\HT{ulte}{\B{\cp{ul}te}} \I{["ul.tE]} \emph{or} \I{["Ul.]} \emph{conj.} and (\emph{to connect sentences}) 
\B{Oel ay\-nga\-ti ka\-me\-ie, ma oe\-yä ey\-lan, ul\-te ay\-nga\-ru sei\-yi ira\-yo.} I \textsc{see} you, my friends, and I thank you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}  
\B{Kì\-ye\-va\-me ul\-te Ey\-wa nga\-hu.} \textsc{See} you again, and may Ey\-wa be with you.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
(\ding{43}~\HL{sI}{\B{sì}})

% ultxa / si
\hi
\HT{ultxa}{\B{ul\cp{txa}}} \I{[ul."t'a]} \emph{or} \I{[Ul.]} (\textsc{rn}: \B{ùl\cp{txa}} \I{[Ul."da]}) \textsc{i}. \emph{n.} meeting.
\B{ul\-txa a mì na'\-rìng}, the meeting in the forest.~\LINK{http://naviteri.org/2010/12/catching-up/}{b}
\B{Am\-'a\-lu\-ke fì\-ul\-txa 'o' la\-yu nì\-txan.} Without a doubt this meeting will be very exciting.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Pol mo\-lä\-kxu ul\-txa\-ti.} He interrupted the meeting.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{Sìl\-pey oe, tsun oeng ul\-txa si\-vi ro hay\-a ul\-txa!} I hope we can meet at the next meeting.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/comment-page-1/\#comment-844}{b} \\
\textsc{ii}. \B{ultxa si} \I{[ul."t'a si]} (\textsc{rn}: \B{ùl\cp{txa} si} \I{[Ul."da{} si]}) \emph{vin.} meet (intentionally with s.o., \B{hu}). 
\B{Tsa\-ria nga\-hu ye'\-rìn ul\-txa si nì\-mun, oe sre\-fe\-re\-i\-ey nì\-prr\-te'.} I'm looking forward to getting together with you again soon.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}
\B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay.} Let's meet on the hill tomorrow at nightfall.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Vol\-a ta\-ron\-yu mì na'\-rìng ul\-txa so\-li.} Eight hunters met (together) in the forest.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/comment-page-1/\#comment-1456}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{ultxatu}{\B{ul\-txa\-tu}}, meeting participant. 
\on{2}.~\HL{ultxarun}{\B{ul\-txa\-run}}, meet (by chance). 
\on{3}.~\HL{tsawlultxa}{\B{tsawl\-ul\-txa}}, great gathering, conference. 
\on{4}.~\HL{numultxa}{\B{nu\-mul\-txa}}, class (for instruction).

% ultxarun
\hi
\HT{ultxarun}{\B{ul\cp{txa}run}} \I{[ul."t'a.R$\cdot$un]} \emph{or} \I{[Ul."t'a.R$\cdot$Un]} (\textsc{rn}: \B{ùl\cp{txa}run} \I{[Ul."da.R$\cdot$un]}) \emph{vtr., inf.}\on{33} meet by chance, encounter. 
\B{Ul\-txa\-ro\-lun pol ye\-rik\-it.} He met a Hexapede.~\LINK{http://forum.learnnavi.org/intermediate/translation-exercise-a-navajo-coyote-tale-in-navi/}{ln} 
(\ding{43}~\HL{ultxa}{\B{ul\-txa}}, \HL{run}{\B{run}})

% ultxatu
\hi
\HT{ultxatu}{\B{ul\cp{txa}tu}} \I{[ul."t'a.tu]} \emph{or} \I{[Ul.]} (\textsc{rn}: \B{ùl\cp{txa}tu} \I{[Ul."da.tu]}) \emph{n.} meeting participant. 
\B{Kaym\-o zo\-la\-'u fray\-ul\-txa\-tu ne kel\-ku moe\-yä fte yi\-vom wu\-tsot, fti\-vi\-a nì\-'it lì'\-fya\-ti le\-Na'\-vi, ul\-te ki\-vä\-teng nì\-'o'.} All the participants came to our house one evening to have dinner, study a little Na'\-vi, hang out together, and have fun.~\LINK{http://naviteri.org/2014/09/stxeli-alor-the-text/}{b}
(\ding{43}~\HL{ultxa}{\B{ul\-txa}}, \HL{-tu}{\B{-tu}})

% um
\hi
\HT{um}{\B{um}} \I{[um]} \emph{or} \I{[Um]} (\textsc{rn}: \B{ùm} \I{[Um]}) \emph{adj.} loose.   
(\ding{214}~\HL{'ekxin}{\B{'ekxin}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{'ekxinum}{\B{'e\-kxi\-num}}, tightness, looseness.

% um'a
\hi 
\HT{um'a}{\B{\cp{um}'a}} \I{["{}um.Pa]} \emph{or} \I{["{}Um.Pa]} \emph{adv.} surprisingly (\emph{for unexpectedly negative outcomes}). 
\B{Pol tì\-kang\-kem\-it tsyo\-lul nì\-so'\-ha, slä tsa\-'ur ha\-sey ke so\-li, um\-'a.} He began the work enthusiastically, but surprisingly, he didn't finish it.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b}  
(\ding{214}~\HL{ti'a}{\B{ti\-'a}}) 

% unil
\hi
\HT{unil}{\B{\cp{u}nil}} \I{["u.nil]} \textsc{i}. \emph{n.} dream.
\B{Kap sì ay\-u\-nil say\-la\-he.} Threats, dreams, and other things.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} \\  
\textsc{ii}. \B{unil si} \I{["u.nil si]} \emph{vin.} dream.
\B{Tì\-txen si, ma 'ite! Unil sar\-mi nga teng\-krr ze\-rawng. Lu fpom srak?} Wake up, daughter! You were dreaming and screaming. Are you okay?~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\ding{43}~\HL{uniltsa}{\B{unil\-tsa}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{uniltaron}{\B{uniltaron}}, dreamhunt. 
\on{2}.~\HL{uniltIrantokx}{\B{u\-nil\-tì\-ran\-tokx}}, dreamwalker body, avatar; (the movie) \emph{Avatar}. 
\on{3}.~\HL{uniltIranyu}{\B{u\-nil\-tì\-ran\-yu}}, dreamwalker.
\on{4}.~\HL{uniltsa}{\B{u\-nil\-tsa}}, dream of\slash about. 
\on{5}.~\HL{txopunil}{\B{txo\-pu\-nil}}, nightmare.

% uniltaron
\hi
\HT{uniltaron}{\B{unil\cp{ta}ron}} \I{[""{}u.nil."ta.Ron]} \emph{n.} Dreamhunt. 
\B{Ätxä\-le si tsnì li\-vu ohe\-ru Unil\-ta\-ron.} \emph{or}
\B{Vu\-yin ohel Unil\-ta\-ron\-it.} I respectfully request the Dreamhunt.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
(\ding{43}~\HL{unil}{\B{unil}}, \HL{taron}{\B{ta\-ron}})

% uniltìrantokx
\hi
\HT{uniltIrantokx}{\B{uniltì\cp{ran}tokx}} \I{[""{}u.nil.tI."Ran.tok']} \emph{n.} (\emph{a.}) dream\-walk\-er body, avatar.  
\B{Me\-unil\-tì\-ran\-tokx Tok\-tor Kì\-rey\-sä sì Tsyeyk\-ä ke lu teng ki steng.} The avatars of Grace and Jake aren't the same, but they are similar.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Poe\-ri unil\-tì\-ran\-tokx\-it tar\-mok a krr lam stum nì\-ay\-fo.} When she is in her Avatar body, she appears almost like them.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
(\emph{b. the movie} Avatar (\on{2009})) \B{Oe\-ri Unil\-tì\-ran\-tokx\-ìl tì\-rey\-ti ley\-ko\-la\-te\-i\-em nì\-txan.} \emph{Avatar} has greatly changed my life for the better.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/msg350756/\#msg350756}{ln}
\B{Ze\-ne oe 'aw\-si\-teng tì\-kang\-kem si\-vi fo\-hu a Unil\-tì\-ran\-tokx\-it sì ki\-fkey\-ti Ey\-wa\-'e\-veng\-ä za\-mo\-lu\-nge aw\-ngar.} I must work together with those who have brought us ``Avatar'' and the world of Pandora.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Unil\-tì\-ran\-tokx\-ä pam\-tseo\-ngop\-yu a\-yaw\-ne to\-ler\-kä\-ngup.}The dear music creator of \emph{Avatar} has died.~\LINK{http://naviteri.org/2015/06/na-ayskxe-mi-telan-sad-news/}{b}
(\ding{43}~\HL{unil}{\B{unil}}, \HL{tIranyu}{\B{tì\-ran\-yu}}, \HL{tokx}{\B{tokx}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{uniltIrantokxolo'}{\B{u\-nil\-tì\-ran\-tokx\-olo'}}, \emph{Avatar} community.

% uniltìrantokxolo'
\hi
\HT{uniltIrantokxolo'}{\B{uniltì\cp{ran}tokxolo'}} \I{[""u.nil.tI."Ran.tok'.o.loP]} \emph{n.} \emph{Avatar} community.
\B{Krr\-ka tsawl\-ul\-txa Unil\-tì\-ran\-tokx\-olo'\-ä vo\-spxì\-am.} During the Avatar Community Meet\hyp{}up last month.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b}
(\ding{43}~\HL{uniltIrantokx}{\B{unil\-tì\-ran\-tokx}}, \HL{olo'}{\B{o\-lo'}})

% uniltìranyu
\hi
\HT{uniltIranyu}{\B{uniltì\cp{ran}yu}} \I{[""{}u.nil.tI."Ran.ju]} \emph{n.} dreamwalker. 
\B{Pol\-txe oe san ze\-ne kea unil\-tì\-ran\-yu ke zi\-va\-'u fì\-tseng.} I said no dreamwalker is to come here.~\LINK{http://wiki.learnnavi.org/index.php?title=Na\%27vi_from_Avatar_Movie\#Scene_in_Hometree}{wiki\slash a}
(\ding{43}~\HL{unil}{\B{unil}}, \HL{tIranyu}{\B{tì\-ran\-yu}})

% uniltsa
\hi
\HT{uniltsa}{\B{\cp{u}niltsa}} \I{["u.nil.\t{ts}$\cdot$a]} \emph{vtr., inf.}\on{33} dream of\slash about, dream (that).
\B{Nì\-trr\-trr oel unil\-tsa sa'\-nu\-ä tey\-lut.} I regularly dream of my mom's \emph{teylu}.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Unil\-tso\-la oel txon\-am fu\-ta tsway\-on Ney\-ti\-ri\-hu.} Last night I dreamed I was flying with Neytiri.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{43}~\HL{unil}{\B{unil}}, \HL{tse'a}{\B{tse\-'a}})

% unyor
\hi
\HT{unyor}{\B{un\cp{yor}}} \I{[un."joR]} \emph{adj.} sweetly aromatic, \emph{a flowery or aromatic woody sort of smell}. 

% uo
\hi
\HT{uo}{\B{\cp{u}o}} \I{["u.o]} \emph{adp.} behind. 
\B{Fnu, fnu, wä\-pan wä ku | Uo tsa\-ìpxa.} Quiet, quiet, hide yourself from harm | Behind that fern.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}
\B{Ut\-ral\-ti a tsa\-uo Lo\-ak wä\-par\-man pol vol fa kxe\-tse.} She used her tail to point out the tree Loak was hiding behind.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{214}~\HL{eo}{\B{eo}}) \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{uolI'uvi}{\B{uo\-lì\-'u\-vi}}, suffix. 
\on{2}.~\HL{ftuopa}{\B{ftuo\-pa}}, from behind.

% uolì'uvi
\hi 
\HT{uolI'uvi}{\B{\cp{u}olì'uvi}} \I{["{}u.o.lI.Pu.vi]} \emph{n.} suffix. 
\B{Lu tsa\-lì\-'ur alu fì\-haw\-re'\-ti me\-lì\-'u\-vi alu 'aw\-a eo\-lì\-'u\-vi sì 'aw\-a uo\-lì\-'u\-vi.} The word \emph{fì\-haw\-re'\-ti} has two affixes--one prefix and one suffix.~\LINK{http://naviteri.org/2017/10/more-language-for-talking-about-language/}{b}
(\ding{43}~\HL{uo}{\B{uo}}, \HL{lI'uvi}{\B{lì\-'u\-vi}})

% -ur
\hi
\HT{-ur}{\B{-ur}} \I{[.uR]} \emph{or} \I{[.UR]} \emph{suf. to mark the dative case of a n. ending in a consonant, the pseudo\hyp{}vowels ll and rr, or the diphthongs ey, ay, and aw}.
(\ding{43}~\HL{-r}{\B{-r}}, \HL{-ru}{\B{-ru}})

% uran
\hi
\HT{uran}{\B{\cp{u}ran}} \I{["u.Ran]} \emph{n.} boat. 
\B{Ay\-fo so\-lop ìlä hil\-van fa u\-ran.} They traveled along (up, down) the river by boat.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Txo fkol ke fyi\-vel u\-ran\-it pay\-wä, ze\-ne fko sli\-ve\-le.} If one does not seal a boat against water, one must swim.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{Derivations}: \ding{43} 
\on{1}.~\HL{kafiuran}{\B{ka\-fi\-uran}}, sailboat.

% ‹us›
\hi
‹\HT{us}{\B{us}}› \I{[u.s]} \emph{inf. in} ‹1› (to form the active participle) 
(\emph{a. always used attributively}) \B{Pa\-lu\-lu\-kan a\-tu\-sa\-ron lu le\-hrr\-ap.} A hunting thanator is dangerous.~\LINK{http://forum.learnnavi.org/language-updates/participles/}{ln}
\B{pay a\-zu\-sup}, falling water.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\emph{b. to productively form the gerund with} \HL{tI-}{\B{tì-}}) 
\B{Tì\-ru\-sol lu oe\-ru mo\-wan.} Singing is enjoyable to me.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Ay\-nge\-yä tì\-ftu\-si\-a 'o' li\-vu nì\-'aw!} May your studying be exciting only.~\LINK{http://naviteri.org/2012/10/sound-files-added-to-previous-post/}{b}
\B{Ran tì\-ru\-sol\-ä pe\-yä lu fyo\-le.} The \emph{ran} of her singing is sublime.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% utral 
\hi
\HT{utral}{\B{\cp{ut}ral}} \I{["ut\textcorner.Ral]} \emph{or} \I{["Ut\textcorner.]} \emph{n.} \ding{96} tree. 
\B{Ut\-ral a lu few pay\-fya a eo kel\-ku oe\-yä tsawl lu nì\-txan.} The tree on the other side of the stream in front of my house is very tall.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
\B{Nga pam\-rel so\-li san sley\-ki\-vu ut\-ral\-ìl lì'\-fya\-yä le\-Na'\-vi pxay\-a ay\-vur\-it a\-txur sìk.} You wrote, ``May the Na'vi language tree produce many strong stories.''~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
\B{Ngä\-zìk lu fwa var tsko\-ti fyi\-vep teng\-krr ut\-ral\-it tsye\-rìl.} It's hard to keep holding a bow while climbing a tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Ay\-ut\-ral\-ur lu frr\-ne\-syul}, the trees have buds.~\LINK{http://forum.learnnavi.org/news-announcements/christmas-song-ninavi-on-youtube/}{ln}
\B{Ut\-ral\-ä A\-nawm | Ay\-ri\-na' lu ay\-oeng.} We are all seeds | Of the Great Tree.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} 
\B{Tsa\-txon ay\-ut\-ral\-äo krr a pol oe\-ti no\-lìn nì\-wa\-we, olo\-me\-i\-um oel fa key\-rel fu\-ta lu yaw\-ne oe po\-ru.} That night beneath the trees when she looked at me meaningfully, I knew by the expression on her face that she loves me.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} \\
(i.) \B{\cp{Ut}ral Ay\cp{mok}riyä} \I{["ut\textcorner.Ral aj."mok\textcorner.Ri.j\ae]} \emph{n.} Tree of Voices.
\B{Ke lu kaw\-tu a nul\-ni\-vew oe po\-hu ti\-rea\-pi\-väng\-kxo äo Ut\-ral Ay\-mok\-ri\-yä.} There's nobody I'd rather commune with under the Tree of Voices.~\LINK{http://wiki.learnnavi.org/Corpus\#Lemondrop.com}{ld\slash wiki} \\  
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{utraltsyIp}{\B{ut\-ral\-tsyìp}}, bush. 
\on{2}.~\HL{kelutral}{\B{kel\-ut\-ral}}, Hometree. 
\on{3}.~\HL{tautral}{\B{tau\-tral}}, sky tree.  
\on{4}.~\HL{rumut}{\B{ru\-mut}}, puffball tree. 
\on{5}.~\HL{pxiut}{\B{pxi\-ut}}, razor palm. 
\on{6}.~\HL{vitrautral}{\B{vit\-ra\-ut\-ral}}, Tree of Souls. 
\on{7}.~\HL{utral aymokriyA}{\B{Ut\-ral Ay\-mok\-ri\-yä}}, Tree of Voices.
\on{8}.~\HL{vArumut}{\B{vä\-ru\-mut}}, vein pod. 
\on{9}.~\HL{vozampasukut}{\B{vo\-zam\-pa\-su\-kut}}, grinch tree.
\on{10}.~\HL{fyIpmaut}{\B{fyìp\-ma\-ut}}, squid fruit tree.
\on{11}.~\HL{koaktutral}{\B{ko\-ak\-tut\-ral}}, goblin thistle.
\on{12}.~\HL{lanutral}{\B{lanutral}}, dandetiger.
\on{13}.~\HL{rumaut}{\B{ru\-ma\-ut}}, cannonball fruit tree.

% Utral Aymokriyä
\hi
\HT{utral aymokriyA}{\B{\cp{Ut}ral Ay\cp{mok}riyä}} \emph{or} \B{\cp{Ut}raya \cp{Mok}ri} \I{["ut\textcorner.Ral aj."mok\textcorner.Ri.j\ae]} \emph{or} \I{["Ut\textcorner.]} \emph{n.} \ding{96} Tree of Voices (\emph{important spiritual site}).
\B{Ke lu kaw\-tu a nul\-ni\-vew oe po\-hu ti\-re\-a\-pi\-väng\-kxo äo Ut\-ral Ay\-mok\-ri\-yä.} There's nobody I'd rather commune with under the Tree of Voices.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{wiki}
(\ding{43}~\HL{utral}{\B{ut\-ral}}, \HL{mokri}{\B{mok\-ri}})

% utraltsyìp 
\hi
\HT{utraltsyIp}{\B{\cp{ut}raltsyìp}} \I{["ut\textcorner.Ral.\t{ts}jIp\textcorner]} \emph{or} \I{["Ut\textcorner.]} \emph{n.} \ding{96} bush. 
(\ding{43}~\HL{utral}{\B{ut\-ral}}, \HL{-tsyIp}{\B{-tsyìp}})

% Utraya Mokri
\hi
\HT{utraya mokri}{\B{\cp{Ut}raya \cp{Mok}ri}} \I{["ut\textcorner.Ra.ja "mok\textcorner.Ri]} : \HL{utral aymokriyA}{\B{Utral Aymokriyä}}

% utu
\hi
\HT{utu}{\B{\cp{u}tu}} \I{["u.tu]} \emph{n.} forest canopy.
(\emph{a. proverb}) \B{Kxìm utu\-ftu fna\-we'\-tu.} ``A coward commands from the canopy,'' i.e. \emph{a real leader will have `boots on the ground' and will help out, whereas a coward will only tell people (from afar) what to do}.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{utumauti}{\B{utu\-mau\-ti}}, banana fruit.

% utumauti
\hi
\HT{utumauti}{\B{\cp{u}tumauti}} \I{["u.tu.ma.u.ti]} \emph{n.} banana fruit.
(\ding{43}~\HL{utu}{\B{utu}}, \HL{mauti}{\B{ma\-u\-ti}})

% uturu
\hi 
\HT{uturu}{\B{u\cp{tu}ru}} \I{[u."tu.Ru]} (\textsc{rn}: \B{u\cp{tù}ru} \I{[u."tU.Ru]}) \emph{n.} sanctuary, place of refuge. 
\B{Vu\-yin o\-hel u\-tu\-rut.} I respectfully request sanctuary.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 
\B{Nga ke tsun wä\-pi\-van; nga\-ri ke lu ke\-a u\-tu\-ru kaw\-tseng.} You cannot hide; there is no sanctuary for you anywhere.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 

% uvan / si
\hi
\HT{uvan}{\B{u\cp{van}}} \I{[u."van]} \textsc{i}. \emph{n.} game.
\B{U\-van Vo\-mun\-a Pam\-rel\-vi\-yä.} The Ten Letter Game.~\LINK{http://www.pbs.org/wgbh/nova/blogs/secretlife/blogposts/learn-navi-with-avatars-paul-frommer/}{nova}
\B{Fì\-u\-van\-ìl oe\-ti syey\-kor nì\-txan.} I find this game very relaxing.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Yì\-mo\-ra' Tsu'\-tey\-ìl uvan\-it.} Tsu'tey just won the game.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Trr\-am\-ä ay\-u\-van\-ìri mak\-to Ak\-wey nì\-yew\-la ha snaytx.} Akwey rode disappointingly in yesterday's games so he lost.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Fra\-u\-van\-ìri lu yo\-ra'\-tu, lu snay\-tu.} For every game, there's a winner and a loser.~\LINK{http://naviteri.org/2011/12/a-note-on-the-word-\%E2\%80\%9Cyora\%E2\%80\%99tu\%E2\%80\%9D/}{b}
\B{Saw\-no'\-ha ioit ko\-la\-ne\-i\-om oel u\-van\-fa!} I got (won) the prized piece of jewelry in the game!~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
(i.) \B{u\cp{van} le\cp{tokx}} physical game\slash activity, sport.
\B{Ay\-u\-van le\-tokx 'o' lu nì\-txan.} Sports are great fun.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} \\
\textsc{ii}. \B{u\cp{van} si} \I{[u."van si]} \emph{vin.} play (a game).
\B{Nì\-'aw\-ve oeng yo\-lom wu\-tsot; maw\-krr uvan si.} First we had dinner; afterwards we played.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
\B{Teng\-krr hu pa\-lu\-kan\-tsyìp uvan se\-ri zo\-lìm oel mì sa'\-leng a 'u\-ot a lu txa' sì ekx\-txu.} While playing with my cat I felt something hard and rough on his skin.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b} 
\B{win\-a u\-van si}, play a quick game. \LINK{http://wiki.learnnavi.org/Canon/2010/UltxaAyharyu\%C3\%A4\#Refinements_of_si-construction_verbs}{wiki} \\
\textsc{Derivations}: \ding{43}
\on{1}.~\HL{lI'uvan}{\B{lì\-'u\-van}}, wordplay, pun. 
\on{2}.~\HL{pamuvan}{\B{pam\-u\-van}}, soundplay.

% ‹uy›
\hi
‹\HT{uy}{\B{uy}}› \I{[u.j]} \emph{inf. in} ‹2› (to indicate formal or ceremonial speech).   
\B{Fay\-sul\-fä\-tu\-ä tì\-kang\-kem ohe\-ru meuia lu\-yu nì\-ngay.} The work of these masters truly honors me.~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Ä\-txä\-le su\-yi o\-he pi\-vawm, pe\-o\-lo' lu\-yu pum nge\-nge\-yä?} May I ask what tribe you belong to?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} 
(\emph{a. in storytelling}) \emph{opening formula} \B{Ay\-fi\-za\-yu pll\-txu\-ye san trro la\-mu krr a …}, Once upon a time … [lit.: the ancestors say, once upon a time] 
\emph{closing formula} \B{Fì\-fya pll\-txu\-ye ay\-fi\-za\-yu.} Thus say the ancestors.~\LINK{http://forum.learnnavi.org/intermediate/translation-exercise-a-navajo-coyote-tale-in-navi/}{ln}
\\
\end{multicols}