\pdfbookmark[0]{V}{V}
\begin{center}
\section*{V \I{[v]} (\emph{Vä})}
\end{center}

\begin{multicols}{2}

% Va'ru
\hi
\B{\cp{Va'}ru} \I{["vaP.Ru]} \emph{n., male name}.
\B{Fì\-tey\-lu\-ri ke nar\-mew Va'\-ru yi\-vom nì\-yll.} Va'ru didn't want to share this \emph{teylu} with the Omatikaya.~\LINK{http://naviteri.org/2012/01/more-additions-to-the-lexicon/}{b}
\B{Ay\-ha\-pxì\-tu po\-ngu\-ä txo\-pu si nì\-nän ta\-krr\-a Va'\-rul pxe\-ku\-tut lä\-txayn.} The members of the group are less afraid since Va'ru defeated three of the enemy.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Hu\-fwa lu fil\-ur Va'\-ru\-ä fne\-fe'\-ran\-vi, tsal\-su\-ngay fpìl fu\-ta say\-rìp lu nì\-txan.} Although Va'ru's facial stripes are rather uneven, I still think he's very handsome.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% vakx
\hi 
\HT{vakx}{\B{vakx}} \I{[vak']} \emph{n., fauna} snake.  

% val
\hi 
\HT{val}{\B{val}} \I{[val]} \emph{adv.}, \textsc{rn} diligently, hard, with effort. 
\B{Mak\-to val!} Ride hard!~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b} 
\B{Po tì\-kang\-kem so\-li val nì\-txan fte tsa\-tson\-ur ha\-sey si\-vi.} She worked very hard to complete the task.~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}
\B{Tì\-kang\-kem si val!\slash Kang\-kem val!} Work hard!~\LINK{http://naviteri.org/2022/12/trr-anawm-polaheiem-the-great-day-has-arrived/}{b}
(\ding{43}~\HL{kawl}{\B{kawl}}) 

% var 
\hi
\HT{var}{\B{var}} \I{[vaR]} (\emph{a. vin.}) to persist in a state, to continue to perform an action.   
\B{Pol trr\-am ye\-rik\-it tar\-ma\-ron. -- Fì\-trr var.} He was hunting hexapede yesterday. -- He still is today.~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln}
\B{Aw\-nga ze\-ne vi\-var ìlä fì\-sa\-lew\-fya. Zen\-ke yak si\-vi.} We must continue in this direction. We must not go astray.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\emph{b. modal}) continue to, keep on. 
\B{Var ni\-vu\-me ko!} Let's keep learning!~\LINK{http://forum.learnnavi.org/language-updates/as-x-as-y-keep-on-keepin-on/}{ln}
\B{Oe\-ri ay\-som\-pì\-va sìn re\-'o var zi\-vup.} Raindrops keep falling on my head.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Ke lu kaw\-tu a nga\-to kem so\-li nì\-'ul fte tsi\-vun lì'\-fya le\-Na'\-vi vi\-var 'i\-vong.} There is nobody that does more than you so that the Na'vi language can continue to bloom.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/comment-page-1/\#comment-1547}{b}
(\emph{c. idiom}) \B{Pol sä\-fpìl\-it ve\-rar wi\-van.} He's keeping his idea a secret.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}

% Varang
\hi
\HT{Varang}{\B{\cp{Va}rang}} \I{["va.RaN]} \emph{n., female name}.

% vawm 
\hi
\HT{vawm}{\B{vawm}} \I{[vawm]} \emph{adj.} deep dark colors including browns.
\B{vawm na nik\-re}, the dark color of Na'\-vi hair.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
\B{vawm na uk}, dark\hyp{}shadow color.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b} 
\B{Tsaw\-ke slo\-lu vawm ta\-lun fwopx.} The sun became dark due to dust in the air.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIvawm}{\B{tì\-vawm}}, darkness. 
\on{2}.~\HL{vawmpin}{\B{vawm\-pin}}, the color \B{vawm}. 
\on{3}.~\HL{kllvawm}{\B{kll\-vawm}}, brown.

% vawmpin
\hi
\HT{vawmpin}{\B{\cp{vawm}pin}} \I{["vawm.pin]} \emph{n.} the color \ding{43}~\HL{vawm}{\B{vawm}}.
(\ding{43}~\HL{'opin}{\B{'o\-pin}})

% vawt 
\hi
\HT{vawt}{\B{vawt}} \I{[vawt\textcorner]} \emph{adj.} solid.
\B{Fì\-ut\-ral\-ìri ta\-ngek\-ä zir fkan vawt, hu\-fwa ke rey.} The trunk of the tree feels solid, although it's dead.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{214}~\HL{momek}{\B{mo\-mek}})

% vay
\hi
\HT{vay}{\B{vay}} \I{[vaj]} \emph{adp.} (\emph{a. local}) up to. 
\B{Ne\-ni lew si fì\-txa\-yor vay txam\-pay nì\-wotx.} Sand covers this expanse all the way to the ocean.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b}
(\emph{b. temporal}) until.
\B{Trr\-'ong\-ta Txon\-'ong\-vay po to\-lì\-ran.} He walked from dawn until dusk.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Vay fwa zo\-la\-'u Tsyeyk\-Su\-li, To\-ruk Mak\-to alu pi\-za\-yu Ney\-ti\-ri\-yä txan\-rar\-mo\-'a fra\-to kip ay\-ha\-pxì\-tu Oma\-ti\-ka\-yaä.} Until Jake Sully arrived, Neytiri's ancestor was the most famous Toruk Makto among the Omatikaya.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} \\
(\emph{c. with numbers}) (\emph{count up}) to. 
\B{Ti\-am (ta pxey) vay vol.} Count (from three) to eight.~\LINK{https://forum.learnnavi.org/language-updates/'count-to'-wa-and-sequential-genitive-(discussion)/}{ln} \\
(i.) \B{vay set}, up to now. \B{Fo\-ru 'upxa\-ret oel fpo\-le', slä vay set ke pa\-mä\-hä\-ngem kea tì\-'eyng.} I have sent them a message, but up to now no answer has arrived.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Lì'\-fya\-ri le\-Na'\-vi 'Rr\-ta\-mì, vay set 'al\-mong a fra\-'u ze\-ra\-'u ta ngrr\-po\-ngu.} Everything that has gone on with (blossomed regarding) Na'vi until now on Earth has come from a grassroots movement.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki} \\
(ii.) \B{tsa\cp{krr}vay}, \emph{conv.} until then, see you then. \B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay. -- Ye\-io'! Tsa\-krr\-vay ko!} Let's meet on the hill tomorrow at nightfall. -- Perfect! See you then.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\
(iii.) \B{txan\cp{txew}vay}, maximally, as … as possible. \B{Tì\-ran nì\-fnu txan\-txew\-vay fte\-ke ay\-ye\-rik\-ìl aw\-nga\-ti sti\-vawm.} Walk as quietly as possible so the hexapedes won't hear us.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} \\
(iv.) \B{ha\cp{ya}lovay}, until next time.~\LINK{http://naviteri.org/2010/07/diminutives-conversational-expressions/}{b} 
(\ding{214}~\HL{raw}{\B{raw}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vaykrr}{\B{vay\-krr}}, until (\emph{conj.}).

% vaykrr
\hi
\HT{vaykrr}{\B{vay\cp{krr}}} \I{[vaj."k\s{r}]} \emph{conj.} until.
(\ding{43}~\HL{vay}{\B{vay}}, \HL{krr}{\B{krr}})

% vä'
\hi
\HT{vA'}{\B{vä'}} \I{[v\ae{}P]} \emph{adj.} unpleasant to the senses. 
\B{Fay\-sä\-re\-ngop\-it a\-vä' oe\-ru rä\-'ä wìn\-txu nì\-mun, ru\-txe. Ke su\-nu oer keng nì\-'it.} Please don't show me these ugly designs again. I don't like them one bit.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{214}~\HL{lor}{\B{lor}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{ftxIvA'}{\B{ftxì\-vä'}}, bad\hyp{}tasting, disgusting. 
\on{2}.~\HL{onvA'}{\B{on\-vä'}}, bad\hyp{}smelling. 
\on{3}.~\HL{mikvA'}{\B{mikvä'}}, bad\hyp{}sounding. 
\on{4}.~\HL{narvA'}{\B{nar\-vä'}}, ugly, unsightly.  
\on{5}.~\HL{vApam}{\B{vä\-pam}}, noise, screech. 
\on{6}.~\HL{vAfewll}{\B{vä\-fe\-wll}}, Centipede. 
\on{7}.~\HL{vArumut}{\B{vä\-ru\-mut}}, vein pod.
\on{8}.~\HL{txavA'}{\B{txa\-vä'}}, disgusting.

% väfewll
\hi
\HT{vAfewll}{\B{\cp{vä}fewll}} \I{["v\ae.fE.w\s{l}]} \emph{n.}, \ding{96} Centipede, ``bad\hyp{}smell plant'', \emph{scolopendra virens}. 
(\ding{43}~\HL{vA'}{\B{vä'}}, \HL{fe'}{\B{fe'}}, \HL{'ewll}{\B{'e\-wll}})

% väng 
\hi
\HT{vAng}{\B{väng}} \I{[v\ae{}N]} \emph{adj.} thirsty.
\B{Nga 'e\-fu väng srak? -- Sran, 'e\-fu nì\-txan.} Are you thirsty? -- Yep, I'm parched.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Sre\-krr 'a\-me\-fu väng, set ye\-väng.} Before, I was thirsty; now my thirst has been quenched.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIvAng}{\B{tì\-väng}}, thirst. 
\on{2}.~\HL{yevAng}{\B{ye\-väng}}, satisfied, quenched.

% väpam
\hi
\HT{vApam}{\B{\cp{vä}pam}} \I{["v\ae.pam]} \emph{n.} noise (\emph{ugly or unpleasant sound}), screech.
\B{Ni\-nat\-ìri tì\-ru\-sol Txe\-wì\-yä lu vä\-pam.} To Ninat, Txewì's singing is noise.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\ding{43}~\HL{hawmpam}{\B{hawm\-pam}}, \HL{vA'}{\B{vä'}}, \HL{pam}{\B{pam}})

% värumut
\hi
\HT{vArumut}{\B{\cp{vä}rumut}} \I{["v\ae.Ru.mut\textcorner]} \emph{or} \I{[.mUt\textcorner]} \emph{n.}, \ding{96} vein pod, ``bad ball tree'', \emph{obesus fenestralis}. 
(\ding{43}~\HL{vA'}{\B{vä'}}, \HL{rum}{\B{rum}}, \HL{utral}{\B{ut\-ral}})

% -ve
\hi
\HT{-ve}{\B{-ve}} \I{[.vE]} \emph{suf. to form ordinal numbers}, e.g. \B{'aw\-ve}, first, \B{mu\-ve}, second, etc.
(\ding{43}~\HL{ve'o}{\B{ve\-'o}})

% ve'kì
\hi
\HT{ve'kI}{\B{ve'\cp{kì}}} \I{[vEP."kI]} \emph{vtr.} hate.
\B{Omum oel fu\-ta ngal pot ve'\-kì, slä tse\-fta ke lu tì\-'eyng a\-muiä.} I know you hate him, but vengeance is not a proper response.~\LINK{http://naviteri.org/2022/11/krr-ao-an-exciting-time/}{b}
\B{Txe\-wì\-yä sem\-pul\-ìri sngä\-'i\-krr Saw\-tu\-tet ve\-re'\-kì.} In the beginning, Txe\-wì's father hated the Skypeople.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIve'kI}{\B{tì\-ve'\-kì}}, hatred.

% ve'o
\hi
\HT{ve'o}{\B{\cp{ve}'o}} \I{["vEP.o]} \emph{n.} order (\emph{as opposed to disorder or chaos}), organization.   
\B{Maw\-krr\-a Saw\-tu\-te\-yä txam\-pxì ho\-lum, tä\-txaw Na'\-vi\-ne nì\-'i'\-a ve\-'o.} After most of the Sky People left, order finally returned to the People.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
(i.) \B{\cp{ve}'o ayru\cp{sey}ä}, ecosystem.~\LINK{https://naviteri.org/2026/04/tsivola-liu-amip-thirty-two-new-words/}{b}
(\ding{214}~\HL{keve'o}{\B{ke\-ve\-'o}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vezo}{\B{ve\-zo}}, be in order, be organzied.   
\on{2}.~\HL{vezeyko}{\B{ve\-zey\-ko}}, put in order, organize. 
\on{3}.~\HL{vefya}{\B{ve\-fya}}, system, process, procedure, approach. 
\on{4}.~\HL{velke}{\B{vel\-ke}}, chaotic, messy, disorganized, in shambles. 
\on{5}.~\HL{venga'}{\B{ve\-nga'}}, organized, `on top of things'. 
\on{6}.~\HL{keve'o}{\B{ke\-ve\-'o}}, chaos, disorder.
\on{7}.~\HL{velun}{\B{ve\-lun}}, logic.

% vefya
\hi
\HT{vefya}{\B{\cp{ve}fya}} \I{["vE.fja]} \emph{n.} system, process, procedure, approach.   
\B{Ney\-ti\-ril Tsyey\-kur wa\-mìn\-txu Oma\-ti\-ka\-ya\-ä ve\-fyat tì\-tu\-sa\-ron\-ä.} Ney\-ti\-ri showed Jake the Omatikaya's approach to hunting.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{ve'o}{\B{ve\-'o}}, \HL{fya'o}{\B{fya\-'o}})

% vekreng
\hi 
\HT{vekreng}{\B{\cp{vek}reng}} \I{["vEk\textcorner.REN]} \emph{n., fauna} cloaked panther. 

% velek
\hi
\HT{velek}{\B{\cp{ve}lek}} \I{["vE.lEk\textcorner]} \emph{vin.} give up, surrender, concede defeat.   
\B{Tì\-'i'\-a\-ri tsam\-ä ze\-ne Saw\-tu\-te vi\-ve\-lek ta\-lun tì\-txur Ey\-wa\-yä.} At the end of the war, the Sky People had to give up due to the power of Ey\-wa.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{ra'un}{\B{ra\-'un}})

% velke
\hi
\HT{velke}{\B{\cp{vel}ke}} \I{["vEl.kE]} \emph{adj.} chaotic, messy, disorganized, in shambles.   
\B{Eyk Ka\-mun a fra\-lo lä\-ngu tsa\-sä\-ta\-ron vel\-ke nì\-wotx. Ta\-ron\-yut yom smar\-ìl!} Every time Ka\-mun is in charge, the hunt is a mess. Everything goes wrong that can!~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{ve'o}{\B{ve'o}}, \HL{luke}{\B{lu\-ke}})

% velun
\hi 
\HT{velun}{\B{ve\cp{lun}}} \I{[vE."lun]} \emph{or} \I{[."lUn]} \emph{n.} logic. 
\B{Nge\-yä tì\-hawl\-ìri ke lä\-ngu kea ve\-lun.} I'm sorry to say there's no logic in your plan.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b}
(\ding{43}~\HL{ve'o}{\B{ve\-'o}}, \HL{lun}{\B{lun}})  

% ventil
\hi
\HT{ventil}{\B{\cp{ven}til}} \I{["vEn.til]} \emph{n.} ankle.
(\ding{43}~\HL{venu}{\B{ve\-nu}}, \HL{til}{\B{til}})

% venu
\hi
\HT{venu}{\B{\cp{ve}nu}} \I{["vE.nu]} \emph{n.} foot.
\B{Ri\-ni fmar\-mi hi\-vi\-fwo slä ve\-nu no\-lip äo tskxe.} Ri\-ni tried to escape but her foot got caught under a rock.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Oel tse\-ri fu\-ta nga tskawr. Sra\-ke ngal ve\-nut tì\-sraw sey\-ko\-li?} I see you're limping. Did you hurt your foot?~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{venzek}{\B{ven\-zek}}, toe. \\
\on{2}.~\HL{hawnven}{\B{hawn\-ven}}, shoe. \\
\on{3}.~\HL{ventil}{\B{ven\-til}}, ankle.

% venzek
\hi
\HT{venzek}{\B{\cp{ven}zek}} \I{["vEn.zEk\textcorner]} \emph{n.} toe. 
(\ding{43}~\HL{venu}{\B{ve\-nu}}, \HL{zekwA}{\B{zek\-wä}})

% venga'
\hi
\HT{venga'}{\B{\cp{ve}nga'}} \I{["vE.NaP]} \emph{adj.} organized, `on top of things'.
\B{Txo ni\-vew fko sä\-ro\-'a si\-vi, ze\-ne nì\-'aw\-ve ve\-nga' li\-vu.} If you want to accomplish great things, you first have to be organized.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Fwa fpìl nì\-fya\-'o a\-ve\-nga' fì\-tì\-zin\-mì lu kel\-tsun.} It's impossible to think clearly (i.e., in an organized manner) in this muddled situation.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/\#comment-33483}{b}
(\ding{43}~\HL{ve'o}{\B{ve'o}}, \HL{-nga'}{\B{-nga'}})

% vewng
\hi
\HT{vewng}{\B{vewng}} \I{[vEwN]} \emph{vtr.} look after, take care of.
\B{Oel vewng frr\-nen\-it.} I look after the infants.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Oel vewng ay\-sngel\-it.} I tend to the refuse.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Oel vewng fu\-ta ay\-e\-veng ni\-vu\-me te\-ri ay\-ewll na'\-rìng\-ä.} I see to it that the children learn about the forest plants.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Mong prr\-nen\-ìl fu\-ta sa'\-nul ve\-rewng nì\-tut.} The infant depends on her Mommy to look after her constantly.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}

% vey
\hi
\HT{vey}{\B{vey}} \I{[vEj]} \emph{n.} food of animal origin, meat; flesh.
\B{Ngay\-txoa, ke tsun oe yi\-vom vey\-ti.} Sorry, I can't eat meat.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yomvey}{\B{yom\-vey}}, dine on flesh, be carnivorous.

% vezeyko
\hi
\HT{vezeyko}{\B{vezey\cp{ko}}} \I{[vE.zEj."k$\cdot$o]} \emph{vtr., inf.}\on{33} put in order, organize.
\B{Nga\-ri sno\-mot krr\-pe ve\-zey\-ko, ma 'itan?} When are you going to organize your room, son?~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\ding{43}~\HL{vezo}{\B{ve\-zo}})

% vezo
\hi
\HT{vezo}{\B{ve\cp{zo}}} \I{[vE."z$\cdot$o]} \emph{vin., inf.}\on{22} be in order, be organized.
(\ding{43}~\HL{ve'o}{\B{ve\-'o}}, \HL{zo}{\B{zo}})

% -vi
\hi
\HT{-vi}{\B{-vi}} \I{[.vi]} \emph{unprod. suf. used loosely for the spawn of s.t. bigger.}

% vin 
\hi
\HT{vin}{\B{vin}} \I{[vin]} \emph{vtr.} (\emph{a.}) ask for, request.   
\B{Pol vo\-lin mip\-a tska\-lep\-it.} He asked for a new crossbow.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Vu\-yin ohel Unil\-ta\-ron\-it.} I respectfully request the Dreamhunt.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
\B{Ngal new a tsa\-'ut rä\-'ä wi\-vo, ma 'evi. Vi\-vin.} Don't reach for what you want, child. Ask for it.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{b. with indirect questions}) 
\B{Vo\-lin oel tì\-'eyng\-it a\slash teyng\-ta Ney\-ti\-ri kä pe\-seng\-ne.} I asked where Neytiri was going.~\LINK{http://naviteri.org/2011/08/reported-speech-reported-questions/}{b}
(\ding{43}~\HL{AtxAle}{\B{ä\-txä\-le si}})

% vingkap
\hi
\HT{vingkap}{\B{\cp{ving}kap}} \I{["viN.kap\textcorner]} \emph{vtr.} occur to one, strike one, pop into one's mind. 
\B{Vì\-ming\-kap oe\-ti fu\-la poe ke li ke pol\-txe san oe za\-sya\-'u.} It just occurred to me that she hasn't yet said she's coming.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIvingkap}{\B{nì\-ving\-kap}}, by the way, incidentally.

% virä
\hi
\HT{virA}{\B{vi\cp{rä}}} \I{[vi."R\ae{}]} \emph{vin.} spread, proliferate. 
\B{Fì\-tseng pay\-fya vi\-rä ka ngip a\-reng, ha tsun aw\-nga tsat si\-var sko em\-kä\-fya.} Here the stream spreads over a shallow area, so we can use it as a ford.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words!}{b} 
\B{Tsran\-ten nì\-txan fwa ftang\-len aw\-ngal fu\-ta fì\-sä\-spxin vi\-vi\-rä.} It's very important that we prevent this disease from spreading.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIvirA}{\B{tì\-vi\-rä}}, spread, proliferation.

% vitra
\hi
\HT{vitra}{\B{vit\cp{ra}}} \I{[vit\textcorner."Ra]} \emph{n.} soul.
(i.) \B{Ayvit\cp{ra}yä Ra\cp{mu}nong}, the Well of Souls (\emph{the closest connection to Eywa on Pandora, a point of extreme spiritual significance to the Na'vi}).
\B{Oe tsway\-on io Ay\-vit\-ra\-yä Ra\-mu\-nong.} I fly over the Tree of Souls.~\LINK{http://forum.learnnavi.org/language-updates/over-under-and-quickly/}{ln} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vitronvA'}{\B{vit\-ron\-vä'}}, asshole, dickhead.

% vitrautral
\hi
\HT{vitrautral}{\B{vit\cp{ra}utral}} \I{[vit\textcorner."Ra.ut\textcorner.Ral]} \emph{n.} Tree of Souls.
(\ding{43}~\HL{vitra}{\B{vit\-ra}}, \HL{utral}{\B{ut\-ral}})

% vitronvä'
\hi
\HT{vitronvA'}{\B{vitron\cp{vä'}}} \I{[vit\textcorner.Ron."v\ae{}P]} \emph{n., vulg.} butthole, asshole, dickhead (\emph{highly abusive and vulgar, and is never used in polite society}). \emph{Shortened to} \ding{43}~\HL{vonvA'}{\B{von\-vä'}}.
(\ding{43}~\HL{vitra}{\B{vit\-ra}}, \HL{onvA'}{\B{on\-vä'}})

% vll
\hi
\HT{vll}{\B{vll}} \I{[v\s{l}]} (\textsc{perf}, \B{vol})  (\emph{a. vin.}) indicate, point at.   
\B{Vll eyk\-yu ne\-fä fte po\-ngu fä\-ki\-vä.} The leader signals the party to ascend by pointing upwards.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{b. vtr.}) \B{Txo\-pul pe\-yä vll fu\-ta kaw\-krr ke sla\-yu tsam\-si\-yu.} His fear indicates he'll never become a warrior.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Ut\-ral\-ti a tsa\-uo Lo\-ak wä\-par\-man pol vol fa kxe\-tse.} She used her tail to point out the tree Loak was hiding behind.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{c. metaphorically}) \B{Txo\-pul pe\-yä vll fu\-ta kaw\-krr ke sla\-yu tsam\-si\-yu.} His fear indicates he'll never become a warrior.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sAvll}{\B{sä\-vll}}, sign, indication, signal.

% vo-
\hi
\HT{vo-}{\B{vo-}} \I{[vo.]} \emph{num. pref.} eight (\emph{base form of} \ding{43}~\HL{vol}{\B{vol}} \emph{to form higher numbers and fractions}).
\B{vo\-\cp{mun}}, \on{10} (decimal), \on{12} (octal);
\B{me\-vo\cp{mun}}, \on{18} (decimal), \on{22} (octal); 
\B{vo\cp{pxì}}, one\hyp{}eighth, $\frac{1}{8}$.

% vofu
\hi
\HT{vofu}{\B{vo\cp{fu}}} \I{[vo."fu]} \emph{num., adj.} fourteen, \on{14} (decimal), \on{16} (octal).  

% vohin
\hi
\HT{vohin}{\B{vo\cp{hin}}} \I{[vo."hin]} \emph{num., adj.} fifteen, \on{15} (decimal), \on{17} (octal).  

% voìk / si
\hi 
\HT{voIk}{\B{\cp{vo}ìk}} \I{["vo.Ik\textcorner]} \textsc{i}. \emph{n.} behavior, \emph{how one conducts oneself}. 
\B{Nga sìl\-mi a tsa\-kem ke lu vo\-ìk a\-mu\-iä!} What you just did was not proper behavior!~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}  \\
\textsc{ii}. \B{\cp{vo}ìk si} \I{["vo.Ik\textcorner{} si]} \emph{vin.} behave.
\B{Ney\-ti\-ril Tsyey\-kur oeyk\-to\-lìng teyng\-ta fya\-pe ze\-ne vo\-ìk si\-vi tsa\-tì\-fkey\-tok\-mì.} Neytiri explained to Jake how to behave in that situation.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} 

% vol
\hi
\HT{vol}{\B{vol}}\textsuperscript{\on{1}} \I{[vol]} \emph{num., adj.} eight (decimal), ten (octal). (\emph{base}: \ding{43}~\B{vo-})
\B{Re\-rol teng\-krr ke\-rä | Ìlä fya\-'o a\-vol | Ne kxam\-tseng.} We sing our way, down the eight paths, to the center.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Pol ke tslam stum ke\-'ut, o\-mum lì\-'ut a\-vol nì\-'aw.} He understands almost nothing and only knows eight words.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vospxIvol}{\B{Vo\-spxì\-vol}}, August.

\B{vol}\textsuperscript{\on{2}} : \HL{vll}{\B{vll}}

% volaw
\hi
\HT{volaw}{\B{vo\cp{law}}} \I{[vo."law]} \emph{num., adj.} nine (decimal), eleven (octal). 

% volve
\hi
\HT{volve}{\B{\cp{vol}ve}} \I{["vol.vE]} \emph{ord. num., adj.} eighth. 

% vomrr
\hi
\HT{vomrr}{\B{vo\cp{mrr}}} \I{[vo."m\s{r}]} \emph{num., adj.} thirteen, \on{13} (decimal), \on{15} (octal).
\B{Ay\-wayl yìm ki\-fkey\-ä | 'Ì\-he\-yut a\-vo\-mrr | Sìn ti\-rea\-fya\-'o a\-vol | Na way\-te\-lem\-ä hìng.} The songs bind | the thirteen spirals of the solid world | To the eight spirit paths | Like the threads of a Songcord.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}

% vomun
\hi
\HT{vomun}{\B{vo\cp{mun}}} \I{[vo."mun]} \emph{or} \I{[.mUn]} \emph{num., adj.} ten, \on{10} (decimal), \on{12} (octal).
\B{U\-van Vo\-mun\-a Pam\-rel\-vi\-yä.} The Ten Letter Game.~\LINK{http://www.pbs.org/wgbh/nova/blogs/secretlife/blogposts/learn-navi-with-avatars-paul-frommer/}{nova}
\B{Txe\-wì lu 'e\-veng\-an a so\-la\-lew zì\-sìt a\-vo\-mun.} Txewì is a ten\hyp{}year\hyp{}old boy.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b} 

% vonvä'
\hi 
\HT{vonvA'}{\B{von\cp{vä'}}} \I{[von."v\ae{}P]}, \emph{coll. pronounced as} \I{[v\~{o}."v\ae{}P]} \emph{n., vulg.} butthole, asshole, dickhead (\emph{highly abusive and vulgar, and is never used in polite society}). \emph{Shortened from} \HL{vitronvA'}{\B{vit\-ron\-vä'}}.
(\ding{43}~\HL{vitra}{\B{vit\-ra}}, \HL{onvA'}{\B{on\-vä'}})

% vopey
\hi
\HT{vopey}{\B{vo\cp{pey}}} \I{[vo."pEj]} \emph{num., adj.} eleven, \on{11} (decimal), \on{13} (octal).

% vopxì
\hi
\HT{vopxI}{\B{vo\cp{pxì}}} \I{[vo."p'I]} \emph{n., num.} one\hyp{}eighth, $\frac{1}{8}$.

% vosìng
\hi
\HT{vosIng}{\B{vo\cp{sìng}}} \I{[vo."sIN]} \emph{num., adj.} twelve, \on{12} (decimal), \on{14} (octal).

% vospxey
\hi 
\HT{vospxey}{\B{Vo\cp{spxey}}} \I{[vo."{}sp'Ej]} \emph{n.} March.
(\ding{43}~\HL{vosxpI}{\B{vospxì}}, \HL{pxey}{\B{pxey}})

% vospxì
\hi
\HT{vospxI}{\B{vo\cp{spxì}}} \I{[vo."{}sp'I]} \emph{n.} month (\emph{from} \B{vo\-sì\-pxì zì\-sìt\-ä}, a twelfth of the year).
\B{Mip\-a Vo\-spxì, Mip\-a Ay\-lì\-'u.} New Words for the New Month.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Ta\-lu\-na vo\-spxì\-o a\-mrr ke zo\-lup tom\-pa, lä\-ngu yì kil\-van\-ä tìm nì\-txan.} Because it hasn't rained for five months, the river level is very low.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Fì\-vo\-spxì\-yä Ay\-lì'\-fya\-vi A\-mip.} This Month's New Expressions.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}. \B{vospxì\cp{am}} \I{[vo.sp'I."am]} \emph{n., adv.} last month. \B{Krr\-ka tsawl\-ul\-txa vo\-spxì\-am.} During the great meeting last month.~\LINK{http://naviteri.org/2014/08/stxeli-alor-a-beautiful-gift/}{b} \\
\on{2}.	\B{vospxì\cp{ay}} \I{[vo.sp'I."aj]} \emph{n., adv.} next month. \B{Vo\-spxì\-ay\-vay!} Until next month.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b} 

% vospxì'aw
\hi 
\HT{vospxI'aw}{\B{Vospxì\cp{'aw}}} \I{[vo.sp'I."Paw]} \emph{also coll.} \I{[.spI.]} \emph{n.} January.  
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{'aw}{\B{'aw}})

% vospxìkin
\hi 
\HT{vospxIkin}{\B{Vospxì\cp{kin}}} \I{[vo.sp'I."kin]} \emph{also coll.} \I{[.spI.]} \emph{n.} July. 
\B{Sìl\-pey oe, ay\-nga\-ri nì\-wotx Vo\-spxì\-kin sngil\-vä\-'i nìl\-tsan.} I hope that as for your all July has started well.~\LINK{http://naviteri.org/2023/07/trr-tsyimawnuniya-lefpom-happy-independence-day/}{b} 
\B{Vo\-spì\-kin Le\-fpom!} Happy July!~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b}
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{kinA}{\B{ki\-nä}})

% vospxìmrr
\hi 
\HT{vospxImrr}{\B{Vospxì\cp{mrr}}} \I{[vo.sp'I."m\s{r}]} \emph{also coll.} \I{[.spI.]} \emph{n.} May. 
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{mrr}{\B{mrr}})

% vospxìmun
\hi 
\HT{vospxImun}{\B{Vospxì\cp{mun}}} \I{[vo.sp'I."mun]} \emph{or} \I{[."mUn]} \emph{also coll.} \I{[.spI.]} \emph{n.} February. 
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{mune}{\B{mu\-ne}})

% vospxìpuk
\hi 
\HT{vospxIpuk}{\B{Vospxì\cp{puk}}} \I{[vo.sp'I."puk\textcorner]} \emph{or} \I{[."pUk\textcorner]} \emph{also coll.} \I{[.spI.]} \emph{n.} June.  
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{pukap}{\B{pu\-kap}})

% vospxìtsìng
\hi 
\HT{vospxItsIng}{\B{Vospxì\cp{tsìng}}} \I{[vo.sp'I."\t{ts}IN]} \emph{also coll.} \I{[.spI.]} \emph{n.} April. 
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{tsIng}{\B{tsìng}})

% vospxìvol
\hi 
\HT{vospxIvol}{\B{Vospxì\cp{vol}}} \I{[vo.sp'I."vol]} \emph{also coll.} \I{[.spI.]} \emph{n.} August.  
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{vol}{\B{vol}})

% vospxìvolaw
\hi 
\HT{vospxIvolaw}{\B{Vospxìvo\cp{law}}} \I{[vo.sp'I.vo."law]} \emph{also coll.} \I{[.spI.]} \emph{n.} September. 
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{volaw}{\B{volaw}})

% vospxìvomun
\hi 
\HT{vospxIvomun}{\B{Vospxìvo\cp{mun}}} \I{[vo.sp'I.vo."mun]} \emph{or} \I{[."mUn]} \emph{also coll.} \I{[.spI.]} \emph{n.} October. 
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{vomun}{\B{vo\-mun}})

% vospxìvopey
\hi 
\HT{vospxIvopey}{\B{Vospxìvo\cp{pey}}} \I{[vo.sp'I.vo."pEj]} \emph{also coll.} \I{[.spI.]} \emph{n.} November. 
\B{Vo\-spxì\-vo\-pey\-ä ay\-lì\-'u a\-mip} November's new words.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b}
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{vopey}{\B{vo\-pey}})

% vospxìvosìng
\hi 
\HT{vospxIvosIng}{\B{Vospxìvo\cp{sìng}}} \I{[vo.sp'I.vo."{}sIN]} \emph{also coll.} \I{[.spI.]} \emph{n.} December. 
(\ding{43}~\HL{vospxI}{\B{vo\-spxì}}, \HL{vosIng}{\B{vo\-sìng}})

% vozam
\hi
\HT{vozam}{\B{vo\cp{zam}}} \I{[vo."zam]} \emph{num., adj.} \on{512} (decimal), \on{1.000} (octal).  
\B{'En si oe, lor\-a tsa\-fkxi\-le\-ri a\-pxay\-opin so\-lar Tse\-nul srok\-it a\-vo\-zam.} I would guess that Tsenu used a thousand (lit. \on{512}) beads for that beautiful multi\hyp{}colored bib necklace.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vozampasukut}{\B{vo\-zam\-pa\-su\-kut}}, grinch tree.

% vozampasukut
\hi 
\HT{vozampasukut}{\B{\cp{vo}zampasukut}} \I{["vo.zam.pa.suk\textcorner.ut\textcorner]} \emph{or} \I{[.sUk\textcorner.Ut\textcorner]} (\textsc{rn}: \B{\cp{vo}zampasùkut} \I{["vo.zam.pa.sU.kut\textcorner]}) \emph{n.}, \ding{96} grinch tree; thousand berry tree.
(\ding{43}~\HL{vozam}{\B{vo\-zam}}, \HL{pasuk}{\B{pa\-suk}}, \HL{utral}{\B{ut\-ral}})

% vrrìn
\hi
\HT{vrrIn}{\B{\cp{vrr}ìn}} \I{["v\s{r}.In]} \emph{vin.} be busy (\emph{negative sense}), \emph{be tired out and overwhelmed by an activity}.
\B{Tì\-kang\-kem\-ìri var\-mrr\-ìn oe nì\-wotx.} I was completely swamped (overwhelmed) at work.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Pe\-yä tì\-kang\-kem\-ìl mi vey\-krr\-ei\-yìn pot nì\-wotx.} His work is still completely overwhelming him (and I'm glad).~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
(\ding{43}~\HL{'In}{\B{'ìn}}, \ding{214}~\HL{sulIn}{\B{su\-lìn}})

% vrrtep
\hi
\HT{vrrtep}{\B{\cp{vrr}tep}} \I{["v\s{r}.tEp\textcorner]} \emph{n.} demon.   
\B{Txon\-am teng\-krr tar\-mì\-ran oe kxam\-lä na'\-rìng, sro\-ler eo ut\-ral a\-tsawl txewm\-a vrr\-tep.} Last night as I was walking through the forest, a frightening demon appeared in front of a big tree.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Saw\-tu\-te lu ay\-vrr\-tep nì\-wotx a sä\-fpìl\-it oel wä\-te.} I dispute the idea that the Skypeople are all demons.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Fay\-vrr\-tep fì\-tse\-nge lu kxa\-nì.} These demons are forbidden here.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Times_Online}{wiki\slash a}
(\ding{43}~\HL{txep}{\B{txep}})

% vul
\hi
\HT{vul}{\B{vul}} \I{[vul]} \emph{or} \I{[vUl]} \emph{n.} \ding{96} branch (of a tree). 
\B{Txo vul ivil nì\-hawng kxì\-ye\-vakx.} If a branch bends too much, it might break.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{Krro krro, flì\-a vul a\-ru\-sey to nutx\-a pum a\-ke\-ru\-sey lu txur.} A thin living branch is sometimes stronger than a thick dead one.~\LINK{http://naviteri.org/2011/11/\%E2\%80\%99a\%E2\%80\%99awa-ayli\%E2\%80\%99u-amip-ni\%E2\%80\%99aw-\%E2\%80\%94-a-few-new-words-only/}{b}
\B{'Ali\-'ät vul\-sìn key\-kur za\-'u fì\-tseng.} Hang the choker on the branch and come here.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
\B{A pe\-yä tì\-txur … Mì pun | Na ay\-vul a\-hu\-saw\-nu.} Whose strength is in our arms | As sheltering branches.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vultsyIp}{\B{vul\-tsyìp}}, stick. 
\on{2}.~\HL{pamrelvul}{\B{pam\-rel\-vul}}, pen, pencil. 
\on{3}.~\HL{relvul}{\B{rel\-vul}}, pen, pencil.

% vultsyìp
\hi 
\HT{vultsyIp}{\B{\cp{vul}tsyìp}} \I{["vul.\t{ts}jIp\textcorner]} \emph{or} \I{["vUl.]} \emph{n.} \ding{96} stick (\emph{of a tree, branch}). 
(\ding{43}~\HL{vul}{\B{vul}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{'opinvultsyIp}{\B{'opin\-vul\-tsyìp}}, crayon.

% vun
\hi 
\HT{vun}{\B{vun}} \I{[vun]} \emph{or} \I{[vUn]} \emph{vtr.} provide. 
\B{Sem\-pul\-ìl a\-sìl\-tsan vun syu\-vet soaia\-ru sne\-yä.} A good father provides food to his family.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}
\B{Ey\-wa va\-yun.} Ey\-wa will provide.~\LINK{http://naviteri.org/2020/05/aawa-liu-amip-si-vurway-alor-a-few-new-words-and-a-beautiful-poem/}{b}  

% vur
\hi
\HT{vur}{\B{vur}} \I{[vuR]} \emph{or} \I{[vUR]} \emph{n.} story. 
\B{Tsa\-vur oe\-yä el\-tur tì\-txen sol\-i.} That story intrigued me.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Nga pam\-rel so\-li san sley\-ki\-vu ut\-ral\-ìl lì'\-fya\-yä le\-Na'\-vi pxay\-a ay\-vur\-it a\-txur sìk.} You wrote, ``May the Na'vi language tree produce many strong stories.''~\LINK{http://forum.learnnavi.org/language-updates/life-death-and-gerunds/}{ln}
\B{Nga\-ri wa\-we fì\-vur\-ä lu 'u\-pe?} What is the significance of this story to you?~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vurvi}{\B{vur\-vi}}, summary. 
\on{2}.~\HL{'okvur}{\B{'ok\-vur}}, (recent) history. 
\on{3}.~\HL{hangvur}{\B{hang\-vur}}, joke. 
\on{4}.~\HL{vurway}{\B{vur\-way}}, story poem, narrative poem. 
\on{5}.~\HL{vurtu}{\B{vur\-tu}}, fictional character.

% vurtu
\hi 
\HT{vurtu}{\B{\cp{vur}tu}} \I{["vuR.tu]} \emph{or} \I{["vUR.]} \emph{n.} fictional character. 
\B{Txon Ki\-ho\-te lu vur\-tu nì\-'aw; ke fkey\-to\-lok kaw\-krr.} Don Quixote is a fictional character; he never existed.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{vur}{\B{vur}}) 

% vurway
\hi
\HT{vurway}{\B{\cp{vur}way}} \I{["vuR.waj]} \emph{or} \I{["vUR.]} \emph{n.} story poem, narrative poem. 
\B{Vur\-way a\-lor.} A beautiful narrative poem.~\LINK{http://naviteri.org/2017/10/vurway-alor-a-beautiful-narrative-poem/}{b}
\B{Fì\-vur\-wayt ivi\-nan nì\-'o' nì\-'aw!} Have fun reading this narrative poem!~\LINK{http://naviteri.org/2017/10/vurway-alor-a-beautiful-narrative-poem/}{b}

% vurvi
\hi
\HT{vurvi}{\B{vurvi}} \I{["vuR.vi]} \emph{or} \I{["vUR.]} \emph{n.} summary, synopsis. 
(\ding{43}~\HL{vur}{\B{vur}})  \\

\end{multicols}