\pdfbookmark[0]{W}{W}
\begin{center}
\section*{W \I{[w]} (\emph{Wä})}
\end{center}

\begin{multicols}{2}

% walak
\hi 
\HT{walak}{\B{\cp{wa}lak}} \I{["wa.lak\textcorner]} \emph{adj.} energetic, active.   
\B{Tì\-tu\-sa\-ron\-ìri txo new fko sli\-vu tsul\-fä\-tu, ze\-ne smar\-to li\-vu wa\-lak.} If you want to become a master hunter, you have to be more active than your prey.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Ftu\-e lu fwa ta\-ron ngong\-a io\-ang\-it to fwa ta\-ron pum\-it a lu wa\-lak sì win.} It's easier to hunt lethargic animals than to hunt perky, speedy ones.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
(\ding{214}~\HL{ngong}{\B{ngong}})

% walew
\hi 
\HT{walew}{\B{wa\cp{lew}}} \I{[wa."lEw]} \emph{vin.} get over, accept some fact, reconcile oneself, move on. 
\B{Fu\-ri\-a oe yaw\-ne ke lu Va'\-rur nul\-krr, ke tsä\-ngun oe wi\-va\-lew.} I can't get over the fact that Va'ru no longer loves me.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}
\B{Tì\-ska\-'a\-ri Kel\-ut\-ral\-ä Na'\-vi wa\-ya\-lew pe\-fya?} With the destruction of Hometree, how will the Na’vi ever move on?~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b}

% wan
\hi 
\HT{wan}{\B{wan}} \I{[wan]} \emph{vtr.} hide.   
\B{Pol sä\-'o\-ti wo\-lan äo ay\-rìk.} he hid his tool under the leaves.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
\B{Nì\-sngä\-'i fmawn\-it fo nar\-mew wi\-van, slä nì\-'i'\-a fra\-por lo\-lo\-nu.} They originally wanted to hide the news, but in the end they revealed it everyone.~\LINK{http://naviteri.org/2011/10/more-vocabulary-a-bit-of-grammar/}{b}
\B{Fnu, fnu, wä\-pan wä ku | Uo tsa\-ìpxa.} Quiet, quiet, hide yourself from harm | Behind that fern.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}
\B{Nga pe\-lun wä\-pe\-ran?} Why are you hiding?~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b}
\B{Ya\-mì tsun fko tsi\-ve\-'a lo\-ran\-it re\-nu\-ä kil\-van\-ä slä kll\-te\-sìn wä\-pan.} From the air you can see the grace of the river's form but from the ground it's hidden.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
(\emph{a. idiom}) \B{Pol sä\-fpìl\-it ve\-rar wi\-van.} He's keeping his idea a secret.~\LINK{http://naviteri.org/2011/02/new-vocabulary-ii\%E2\%80\%94part-1/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIwan}{\B{tì\-wan}}, obfuscation, cover\hyp{}up.   
\on{2}.~\HL{nIwan}{\B{nì\-wan}}, secretly; in hiding, by hiding.   
\on{3}.~\HL{letwan}{\B{let\-wan}}, dodgy, sneaky.
\on{4}.~\HL{wantseng}{\B{wan\-tseng}}, (temporary) hiding place.

% wantseng
\hi 
\HT{wantseng}{\B{\cp{wan}tseng}} \I{["wan.\t{ts}EN]} \emph{n.} (temporary) hiding place. 
\B{Tsa\-ye\-rik\-ìl mu'\-nit\-kan\-a wan\-tseng\-it ro\-lun; ke tsun fko pot tsi\-ve\-'a kaw\-'it.} That hexapede has found an effective hiding place. He can't be seen at all.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} 
(\ding{43}~\HL{wan}{\B{wan}}, \HL{tseng}{\B{tseng}})

% wapx
\hi 
\HT{wapx}{\B{wapx}} \I{[wap']} \emph{vin.} sink. 
\B{Ke omum teyng\-ta fì\-u\-ran a\-ku'\-up fu fwa\-yum fu wa\-yapx.} I don't know (or: It's not known) whether this heavy boat will float or sink.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b}

% wawe
\hi 
\HT{wawe}{\B{wa\cp{we}}} \I{[wa."wE]} \emph{n.} meaning, importance, significance.   
\B{Nga\-ri wa\-we fì\-vur\-ä lu 'u\-pe?} What is the significance of this story to you?~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
\B{Wa\-we\-ti ke tsun fko ral\-pi\-veng.} One can't explain (or: put into words) \emph{wawe}.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 
\B{Unil\-tì\-ran\-tokx\-ìri lì'\-fya\-yä le\-Na'\-vi wa\-wet fol tslam.} They understand the importance of the Na'vi language for \emph{Avatar}.~\LINK{http://naviteri.org/2013/08/fmawn-a-fkol-kilmulat-this-just-in/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txanwawe}{\B{txan\-wa\-we}}, personally meaningful, significant.   
\on{2}.~\HL{nIwawe}{\B{nì\-wa\-we}}, meaningfully, significantly.

% way / si
\hi 
\HT{way}{\B{way}} \I{[waj]} \textsc{i}. \emph{n.} song, poetry.
\B{Ay\-wayl yìm ki\-fkey\-ä | 'Ì\-he\-yut a\-vo\-mrr | Sìn ti\-rea\-fya\-'o a\-vol.} The songs bind | the thirteen spirals of the solid world | To the eight spirit paths.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Fì\-way\-ri hì\-no\-a re\-nut ngam\-pam\-ä ke tsä\-ngun oe tsl\-ivam.} I'm afraid I can't understand the intricate rhyme scheme of this poem.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
\B{Sìl\-pey oe, tsi\-vun oe nge\-yä ay\-wayt a\-mip i\-vi\-nan ye'\-rìn!} I hope, I can read your new songs soon.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/\#comment-33459}{b}
\B{Fì\-way\-ri hì\-no\-a re\-nut ngam\-pam\-ä ke tsä\-ngun oe tsli\-vam.} I'm afraid I can't understand the intricate rhyme scheme of this poem.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} 
(i.) \B{way a pll\-\cp{txe}}, spoken poem.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b} 
(ii.) \B{way a rol}, song.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b} 
(iii.) \B{\cp{way}ä ay\cp{lì}'u}, words of a poem, lyrics of a song.~\LINK{http://naviteri.org/2011/12/ulte-yora\%E2\%80\%99tu-leiu-and-the-winner-is/}{b} \\
\textsc{ii}. \B{\cp{way} si} \I{["waj si]} \emph{vin.} sing (singing of ancient, traditional song)
(\ding{43}~\HL{rol}{\B{rol}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{waytelem}{\B{way\-te\-lem}}, songcord. 
\on{2}.~\HL{taronway}{\B{ta\-ron\-way}}, hunt song. 
\on{4}.~\HL{vurway}{\B{vur\-way}}, story poem, narrative poem.
\on{5}.~\HL{haway}{\B{ha\-way}}, lullaby.

% waytelem
\hi 
\HT{waytelem}{\B{\cp{way}telem}} \I{["waj.tE.lEm]} \emph{n.} songcord. 
\B{Ay\-wayl yìm ki\-fkey\-ä | 'Ì\-he\-yut a\-vo\-mrr | Sìn ti\-rea\-fya\-'o a\-vol | Na way\-te\-lem\-ä hìng.} The songs bind | the thirteen spirals of the solid world | To the eight spirit paths | Like the threads of a Songcord.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp} 
\B{Trro la\-yu ngar lor\-a way\-te\-lem a\-ngim fte vur\-it nge\-yä pi\-veng.} One day you will have a long and beautiful songcord to tell your story.~\LINK{https://www.youtube.com/watch?v=13aPkyXDuEE}{yt\slash fop}
(\ding{43}~\HL{way}{\B{way}}, \HL{telem}{\B{te\-lem}})

% wä 
\hi 
\HT{wA}{\B{wä}} \I{[w\ae{}]} \emph{adp.}\texttt{+} (\emph{a.}) against, opposing. 
(\emph{a.1}) \B{fyel wä}, seal from\slash against. \B{Txo fkol ke fyi\-vel u\-ran\-it pay\-wä, ze\-ne fko sli\-ve\-le.} If one does not seal a boat against water, one must swim.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\emph{a.2}) \B{haw\-nu wä}, protect from. \B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang. Fì\-tì\-kan\-ìri ropx ke ha'.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators. For this purpose, a hole going right through the tree trunk isn't suitable.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\emph{a.3}) \B{kan wä}, aim s.t. against s.t.\slash s.o.
\B{Pol tsko\-ti ko\-lan wä ku\-tu.} He raised his bow against the enemy.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/comment-page-1/\#comment-22792}{g\slash b}
(\emph{a.4}) \B{kxap wä}, a threat to s.t.\slash s.o., threaten.
\B{Krr\-a kxap lar\-mu sì\-rey\-wä fe\-yä nì\-wotx.} When their lives were threatened.~\LINK{http://naviteri.org/2015/05/cirque-du-soleils-toruk-the-first-flight-teaser-video-online/}{b}
(\emph{a.5}) \B{tsre\-'i wä}, throw at. \B{Oel tsre\-'i tskxet wä nga.} I throw a stone at you.~\LINK{http://forum.learnnavi.org/language-updates/to-throw-at-x/}{ln}
(\emph{a.6}) \B{w(ä\-p)an wä}, hide (oneself) from. \B{Fnu, fnu, wä\-pan wä ku | Uo tsa\-ìpxa.} Quiet, quiet, hide yourself from harm | Behind that fern.~\LINK{http://forum.learnnavi.org/language-updates/learnings-from-the-siatlla-song-translations/msg473566/\#msg473566}{ln}
(\emph{a.7}) \B{wem wä}, fight against. \B{Peyä tsa\-tì\-pe\-'un a swey\-lu txo wi\-vem ayoeng Oma\-ti\-ka\-ya\-wä lu fe'.} His decision to fight against the Omaticaya was a bad one.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
(\emph{b.}) contrary to (what somebody said).
\B{Wä Fey\-ral, mun\-txa ke so\-li Ra\-lu sì Ne\-wey nì\-wan me\-srr\-am.} Contrary to what Peyral reported, Ralu and Newey were not secretly married the day before yesterday.~\LINK{https://forum.learnnavi.org/language-updates/'count-to'-wa-and-sequential-genitive-(discussion)/}{ln}
(\emph{b.1}) \B{tì\-'e\-fu\-wä oe\-yä}, ``contrary to my opinion''.
\B{Tì\-'efu\-wä oe\-yä, fpìl Pey\-ral\-ìl fu\-ta ke ze\-ne ay\-oeng ki\-vä.} I think we have to go, but Peyral doesn't.~\LINK{https://forum.learnnavi.org/language-updates/'count-to'-wa-and-sequential-genitive-(discussion)/}{ln} (\ding{214}~\HL{IlA}{\B{ìlä}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{wAsul}{\B{wä\-sul}}, compete.  
\on{2}.~\HL{wAte}{\B{wä\-te}}, argue, dispute. 
\on{3}.~\HL{wAtu}{\B{wä\-tu}}, opponent. 
\on{4}.~\HL{nIwA}{\B{nì\-wä}}, on the contrary, conversely. 
\on{5}.~\HL{wAralI'u}{\B{wä\-ra\-lì\-'u}}, antonym.
\on{6}.~\HL{wAtum}{\B{wä\-tum}}, antidote.

% wäralì'u
\hi 
\HT{wAralI'u}{\B{\cp{wä}ralì'u}} \I{["w\ae.Ra.lI.Pu]} \emph{n.} antonym. 
(\ding{43}~\HL{wA}{\B{wä}}, \HL{ral}{\B{ral}}, \HL{lI'u}{\B{lì\-'u}}; \ding{214}~\HL{tengralI'u}{\B{teng\-ra\-lì\-'u}}) 

% wäsul
\hi 
\HT{wAsul}{\B{\cp{wä}sul}} \I{["w\ae{}.s$\cdot$ul]} \emph{vin., inf.}\on{22} compete. 
\B{Oe new nga\-hu 'aw\-si\-teng tì\-kang\-kem si\-vi -- ke new fu\-ta wä\-si\-vul oeng.} I want to work together with you -- I don't want us to compete.~\LINK{http://naviteri.org/2012/10/wina-sawasultsyip-ahii-a-quick-little-contest/}{b} 
(\ding{43}~\HL{wA}{\B{wä}}, \HL{tul}{\B{tul}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIwAsul}{\B{tì\-wä\-sul}}, competition, competing.
\on{2}.~\HL{sAwAsul}{\B{sä\-wä\-sul}}, a competition.

% wäte
\hi 
\HT{wAte}{\B{wä\cp{te}}} \I{[w\ae{}."t$\cdot$E]} \emph{vtr., inf.}\on{22} argue, dispute.   
\B{Saw\-tu\-te lu ay\-vrr\-tep nì\-wotx a sä\-fpìl\-it oel wä\-te.} I dispute the idea that the Skypeople are all demons.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
(\ding{43}~\HL{wA}{\B{wä}}, \HL{plltxe}{\B{pll\-txe}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIwAte}{\B{tì\-wä\-te}}, dispute, argument. 
\on{2}.~\HL{sAwAte}{\B{sä\-wä\-te}}, point of contention, source of argument. 
\on{3}.~\HL{lewAte}{\B{le\-wä\-te}}, disagreeable, argumentative. 
\on{4}.~\HL{nIwAte}{\B{nì\-wä\-te}}, disagreeably, begrudgingly.

% wätu
\hi 
\HT{wAtu}{\B{\cp{wä}tu}} \I{["w\ae{}.tu]} \emph{n.} opponent.
\B{Oe pll\-ngay san mo\-lak\-to oe nì\-fe', ta\-fral sno\-laytx; wä\-tu lu oe\-to txur.} I acknowledge that I rode badly, so I lost; my opponent was stronger than I was.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} 
\B{Krr\-a wä\-tut tse\-'a, pe\-yä sä\-sìl\-pey a yo\-ra' 'o\-lìp.} When he saw his opponent, his hope of winning vanished.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{43}~\HL{wA}{\B{wä}}, \HL{tute1}{\B{tu\-te}}) 

% wätx
\hi 
\HT{wAtx}{\B{wätx}} \I{[w\ae{}t']} \emph{vtr.} be bad at. 
\B{Sra\-ke fnan ngal lì'\-fya\-ti le\-Na'\-vi? -- Ke\-he. Slä hu\-fwa oel (tsat) wätx nì\-'aw, tsal\-su\-ngay yaw\-ne lu oer nì\-wotx!} Are you good at Na'vi? -- No, I totally suck at it, but I still love it to death.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b}
\B{Tsmìm\-ìri wätx fol ti\-nan\-it nì\-ngay.} They're really bad at reading animal tracks.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b}
\B{Oe\-ri lu tì\-ftxu\-lì\-'u ngä\-zìk nì\-txan. Wätx nì\-'aw.} Public speaking is very difficult for me. I'm hopelessly bad at it.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
(\ding{214}~\HL{fnan}{\B{fnan}})

% wätum
\hi 
\HT{wAtum}{\B{\cp{wä}tum}} \I{["w\ae.tum]} \emph{or} \I{[.tUm]} (\textsc{rn}: \B{\cp{wä}tùm} \I{["w\ae.tUm]}) \emph{n.} antidote. 
\B{Pe\-yä ay\-lì\-'u\-ri a\-txum\-nga' wä\-tum\-ìl pe\-se\-nget?} Where is the antidote for his poisonous words?~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} 
(\ding{43}~\HL{wA}{\B{wä}}, \HL{txum}{\B{txum}}) 

% we'ay
\hi 
\HT{we'ay}{\B{\cp{we}'ay}} \I{["wE.Paj]} \emph{adj.} sour.

% wem
\hi 
\HT{wem}{\B{wem}} \I{[wEm]} \emph{vin.} fight (with\slash alongside, \HL{hu}{\B{hu}}; against, oppose, \HL{wA}{\B{wä}}).   
\B{Ru\-txe wi\-vem nì\-nän.} Please fight less.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Peyä tsa\-tì\-pe\-'un a swey\-lu txo wi\-vem ayoeng Oma\-ti\-ka\-ya\-wä lu fe'.} His decision to fight against the Omaticaya was a bad one.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b}
\B{Ay\-oe ke wa\-syem.} We don't intend to fight.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g} 
\B{Po\-ri wem\-tswo fra\-tsam\-si\-yur ro\-lo\-'a nì\-txan.} His ability to fight greatly impressed all the warriors.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Oel nge\-y\"a tì\-hawl\-it na\-txu ulte tsa\-w\"a wa\-syem.} I disapprove of your plan and will oppose (fight against) it.~\LINK{http://naviteri.org/2020/02/some-words-for-leap-year-day/}{b}  \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{wempongu}{\B{wem\-po\-ngu}}, squad, battle party.

% wempongu
\hi 
\HT{wempongu}{\B{\cp{wem}pongu}} \I{["wEm.po.Nu]} \emph{n.} squad, military clan, battle party. 
(\ding{43}~\HL{wem}{\B{wem}}, \HL{pongu}{\B{po\-ngu}})

% wew
\hi 
\HT{wew}{\B{wew}} \I{[wEw]} \emph{adj.} cold. 
\B{Ya\-ri som\-wew\-pe (set)? Ya lu wew.} What's the temperature? It's cold.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Nga tsun ni\-väk wew\-a pay\-it.} You can drink cold water.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b} 
(\ding{214}~\HL{}{\B{som}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIwew}{\B{tì\-wew}}, coldness, cold. 
\on{2}.~\HL{txawew}{\B{txa\-wew}}, very cold. 
\on{3}.~\HL{somwew}{\B{som\-wew}}, temperature.
\on{4}.~\HL{zIskrrwew}{\B{zì\-skrr\-wew}}, winter.

% weyn
\hi 
\HT{weyn}{\B{weyn}} \I{[wEjn]} \emph{vtr.} draw, illustrate. 
\B{Wo\-leyn Ìstawl yey\-fyat mì hll\-te fte oeyk\-ti\-vìng fra\-po\-ru tì\-hawl\-te\-ri sne\-yä.} Ìstaw drew a line on the ground to explain his plan to everyone.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Pol rel\-it wo\-leyn fa pin\-vul.} He drew a picture with a crayon.~\LINK{http://naviteri.org/2018/03/100a-liu-amip-64-new-words-part-1/\#comment-28355}{b}

% weopx
\hi 
\HT{weopx}{\B{we\cp{opx}}} \I{[wE."{}op']} \emph{n.} wave (of water). 
\B{Krr\-a hu\-fwe tul nì\-win, tsun fko tsi\-ve\-'a ay\-we\-opx\-it a sìn yo pay\-ä.} When there is strong wind, you can see waves on the water.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}  \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{leweopx}{\B{le\-we\-opx}}, wave\hyp{}like.

% win / si / säpi
\hi 
\HT{win}{\B{win}} \I{[win]} \textsc{i}. \emph{adj.} fast, quick, rapid.
\B{Lu nga win sì txur | Lu nga txan\-tslu\-sam.} You are fast and strong | You are wise.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b} 
\B{Po lu win fra\-to.} She is the fastest.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln}
\B{Oel tso\-le\-'a fwam\-pop\-it a\-win fra\-to.} I saw the fastest Tapirus.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln} 
\B{Win\-a sä\-wä\-sul\-tsyìp a\-hì\-'i.} A quick little contest.~\LINK{http://naviteri.org/2012/10/wina-sawasultsyip-ahii-a-quick-little-contest/}{b} 
\B{win\-a u\-van si}, play a quick game. \LINK{http://wiki.learnnavi.org/Canon/2010/UltxaAyharyu\%C3\%A4\#Refinements_of_si-construction_verbs}{wiki} \\
\textsc{ii}.a \B{win si} \I{[win si]} \emph{vin.} rush s.t., make s.t. fast. \\
\textsc{ii}.b \B{win sä\cp{pi}} \I{[win s\ae{}."pi]} \emph{vin.} hurry oneself. 
\B{Ke ze\-ne win sä\-pi\-vi. 'Ivong nìk\-'ong.} Take your time; don't rush. Slow is fine.~\LINK{http://naviteri.org/2010/09/quick-follow-up/}{b} \\  
\textsc{Derivatives}: \ding{43} \on{1}.~\HL{nIwin}{\B{nì\-win}}, fast, quickly, rapidly. 
\on{2}.~\HL{winzaw}{\B{win\-zaw}}, arrow deer.

% winzaw
\hi 
\HT{winzaw}{\B{\cp{win}zaw}} \I{["win.zaw]} \emph{n., fauna} arrow deer. 
(\ding{43}~\HL{win}{\B{win}}, \HL{swizaw}{\B{swi\-zaw}}) 

% wion
\hi 
\HT{wion}{\B{\cp{wi}on}} \I{["wi.on]} \emph{n.} reef. 
\B{Lì'\-fya Wi\-on\-ä}, Reef Na'vi, \emph{a dialect of the Na'vi language family} [lit.: `language of the reef'].~\LINK{http://naviteri.org/2023/01/reef-navi-part-1-phonetics-and-phonology/}{b}

% wip
\hi 
\HT{wip}{\B{wip}} \I{[wip\textcorner]} \emph{adj.} salty.

% wiya
\hi 
\HT{wiya}{\B{\cp{wi}ya}} \I{["wi.ja]} \emph{intj. expression of warning or frustration}.

% wìntxu
\hi 
\HT{wIntxu}{\B{wìn\cp{txu}}} \I{[wIn."t'u]} \emph{vtr.} show. 
\B{Fay\-sä\-re\-ngop\-it a\-vä' oe\-ru rä\-'ä wìn\-txu nì\-mun, ru\-txe. Ke su\-nu oer keng nì\-'it.} Please don't show me these ugly designs again. I don't like them one bit.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Nar\-mew oe fo\-ru na'\-rìng\-ä tì\-lor\-it wi\-vìn\-txu, slä ke tsä\-ngun fo tsli\-vam.} I wanted to show them the beauty of the forest, but sadly, they weren't able to understand.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Oel lum\-pe tsat so\-lung mì upxa\-re? Fte wi\-vìn\-txu fu\-ta le\-sar lu fì\-lì\-'u nì\-txan.} Why have I added it to the message? To show that this word is very useful.~\LINK{http://naviteri.org/2010/07/vocabulary-update/comment-page-1/\#comment-95}{b} 
\B{Tsa\-ta txe'\-lan\-it nge\-yä tok wo\-lìn\-txu ngal ay\-oe\-ru nì\-ngay.} You truly showed us where your heart is.~\LINK{http://masempul.org/2010/07/\%E2\%80\%99ite-polaheiem/}{sempul} 
\B{Ney\-ti\-ril Tsyey\-kur wa\-mìn\-txu Oma\-ti\-ka\-ya\-ä ve\-fyat tì\-tu\-sa\-ron\-ä.} Ney\-ti\-ri showed Jake the Omatikaya's approach to hunting.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sAwIntxu}{\B{sä\-wìn\-txu}}, show, showing.  
\on{2}.~\HL{muwIntxu}{\B{mu\-wìn\-txu}}, introduce, present.  
\on{3}.~\HL{fyawIntxu}{\B{fya\-wìn\-txu}}, guide. 
\on{4}.~\HL{wIngay}{\B{wì\-ngay}}, prove.

% wìngay
\hi 
\HT{wIngay}{\B{wì\cp{ngay}}} \I{[w$\cdot$I."Naj]} \emph{vtr., inf.}\on{11} prove.
\B{Fa fwa tsyìl kxem\-yot a\-kxayl fra\-to, pol ay\-oer wo\-lì\-ngay fu\-ta tsyìl\-tswo tsan\-'o\-lul.} By scaling the highest wall, he proved to us that his climbing ability had improved.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
(\ding{43}~\HL{wIntxu}{\B{wìn\-txu}}, \HL{ngay}{\B{ngay}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIwIngay}{\B{tì\-wì\-ngay}}, proving (abstract). 
\on{2}.~\HL{sAwIngay}{\B{sä\-wì\-ngay}}, proof.

% wo
\hi 
\HT{wo}{\B{wo}} \I{[wo]} \emph{vtr.} reach for.   
\B{Ngal new a tsa\-'ut rä\-'ä wi\-vo, ma 'evi. Vi\-vin.} Don't reach for what you want, child. Ask for it.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{kllwo}{\B{kll\-wo}}, alight, land, approach for land. 
\on{2}.~\HL{yawo}{\B{ya\-wo}}, take off, launch. 

% wok
\hi 
\HT{wok}{\B{wok}} \I{[wok\textcorner]} \emph{adj.} loud.
\B{Lu Saw\-tu\-te wok sì pì\-saw nì\-wotx, na prr\-nen.} The skypeople are all loud and clumsy, like a baby.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b} 
\B{Zo\-la\-'u ftu kä\-pxì num\-tseng\-vi\-yä sä\-mewn a\-wok.} From the back of the classroom came a loud sigh.~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} 
(\ding{214}~\HL{'ango}{\B{'ango}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIwok}{\B{nì\-wok}}, loudly. 
\on{2}.~\HL{wokau}{\B{wok\-au}}, pendulum drum.

% wokau
\hi 
\HT{wokau}{\B{wok\cp{a}u}} \I{[wok."a.u]} \emph{n.} pendulum drum. 
(\ding{43}~\HL{wok}{\B{wok}}, \HL{au}{\B{au}})

% wotx
\hi 
\HT{wotx}{\B{wotx}} \I{[wot']} \emph{n.} totality, whole. 
(\emph{a. idiom}) \B{ka wotx}, generally, for the most part. 
\B{Hu\-fwa ro\-lun oel 'a\-'aw\-a kxey\-ey\-ti, fì\-tì\-kang\-kem\-vi lu txan\-tsan ka wotx.} Although I found a few errors, this piece of work is generally excellent.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIwotx}{\B{nì\-wotx}}, all (of), completely.
\on{2}.~\HL{tengwotx}{\B{teng\-wotx}}, uniform, non\hyp{}diverse.

% wou
\hi 
\HT{wou}{\B{\cp{wo}u}} \I{["wo.u]} \emph{vin., slang} be amazing, be fascinating (\emph{slang}).   
\B{Wou Ney\-ti\-ri!} Neytiri is awesome!~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Oe\-ru wo\-eiu lì'\-fya le\-Na'\-vi.} I'm just fascinated by the Na'vi language.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Mì na'\-rìng Tsyeyk\-ìl lo\-re\-yu\-ti 'o\-lam\-pi a tsa\-swaw wou oer nì\-wotx!} That instant when Jake touched the \emph{Helicoradiam Spirale} in the forest I was totally amazed.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}
\B{Por war\-mou fra\-fne\-i\-o\-ang nì\-tut, slä nì\-pxi pxe\-syì\-rä\-fì.} He just couldn't get over all the animals, but was especially taken with the three giraffe.~\LINK{http://forum.learnnavi.org/language-updates/txelanit-hivawl/}{ln}

% wrrkä
\hi 
\HT{wrrkA}{\B{wrr\cp{kä}}} \I{[w\s{r}."k$\cdot$\ae]} \emph{vin., inf.}\on{22} go out, go outside. 
\B{Ke tsun wrr\-ki\-vä fte tì\-kang\-kem si\-vi.} (We) cannot go out to work.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} 
\B{A\-ha'\-ri wrr\-ko\-lä fte li\-mang si\-vu\-nu por.} Aha'ri is out enjoying the moot.~\LINK{https://forum.learnnavi.org/language-updates/limangmoot-from-afop-dialogue/}{ln\slash fop}
(\ding{43}~\HL{wrrpa}{\B{wrr\-pa}}, \HL{kA}{\B{kä}}; \ding{214}~\HL{wrrza'u}{\B{wrr\-za\-'u}})

% wrrpa
\hi 
\HT{wrrpa}{\B{\cp{wrr}pa}} \I{["w\s{r}.pa]} \emph{n., adv.} outside. 
\B{Txìng ay\-ioang\-it! Hum ne wrr\-pa! Tul!} Leave the animals! Get outside! Run!~\LINK{http://forum.learnnavi.org/language-updates/txing-and-hum/}{ln\slash a\textsuperscript{c}} 
\B{Ke za\-syup lì\-'O\-na ne kxu\-tu a mì\-fa fu a wrr\-pa.} The l'Ona will not perish to the enemy within or the enemy without.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{ln\slash g}
(\ding{214}~\HL{mIfa}{\B{mì\-fa}}, \ding{43}~\HL{pa'o}{\B{pa\-'o}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tawrrpa}{\B{ta\-wrr\-pa}}, external.

% wrrza'u
\hi 
\HT{wrrza'u}{\B{wrr\cp{za}'u}} \I{[w\s{r}."z$\cdot$a.P$\cdot$u]} \emph{vin., inf.}\on{23} come out, emerge.
\B{Txep\-ram pxor a krr, txan\-a txe\-kxum\-pay wrr\-za\-'u.} When a volcano erupts, a lot of lava comes out.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b} 
(\ding{43}~\HL{za'u}{\B{za\-'u}}; \ding{214}~\HL{wrrkA}{\B{wrr\-kä}})

% wrrzärìp
\hi 
\HT{wrrzArIp}{\B{wrr\cp{zä}rìp}} \I{[w\s{r}."z$\cdot$\ae.R$\cdot$Ip\textcorner]} \emph{vtr., inf.}\on{23} (\emph{a. physical}) pull out, extract. 
\B{Pol tstal\-it wrr\-zo\-lä\-rìp tstal\-se\-na\-ftu.} He pulled the knife out of its sheath.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 
(\emph{b. fig.}) (1.) \B{txe'\-lan\-ti wrr\-zä\-rìp} \emph{vtr.} to greatly move emotionally.
\B{Oe\-ri pe\-yä ay\-lì\-'ul txe'\-lan\-ti wrr\-zo\-lä\-rìp.} Her words moved me greatly.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(2.) \B{tì\-pe'\-ngayt wrr\-zä\-rìp} \emph{vtr.} infer.
\B{Ngey ay\-lì\-'u\-ftu wrr\-zä\-rìp oel tì\-pe'\-ngayt a lu ngar yew\-la.} From your words, I infer that you're disappointed.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b} 
(3.) \B{tì\-pe'\-ngayt wrr\-zey\-kä\-rìp} \emph{vtr.} imply.
\B{Ngey ay\-lì\-'ul wrr\-zey\-kä\-rìp tì\-pe'\-ngayt a lu ngar yew\-la.} Your words imply that you're disappointed.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(\ding{43}~\HL{wrrpa}{\B{wrr\-pa}}, \HL{za'ArIp}{\B{za\-'ä\-rìp}})

% wum
\hi 
\HT{wum}{\B{wum}} \I{[wum]} \emph{or} \I{[wUm]} (\textsc{rn}: \B{wùm} \I{[wUm]}) \emph{adv.} approximately, roughly.
\B{Oe\-ri so\-la\-lew wum zì\-sìt °a14 a krr, fo\-lrr\-fen spo\-not alo a\-'aw\-ve.} When I was about 12 years old I visited an island for the first time.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% wur
\hi 
\HT{wur}{\B{wur}} \I{[wuR]} \emph{or} \I{[wUR]} \emph{adj.} cool, a bit chilly. 
\B{Ya\-ri som\-wew\-pe (set)? Ya lu wur.} What's the temperature? It's chilly.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Ay\-nga\-ri teng\-krr ya wur sle\-ru nì\-'ul, sìl\-pey oe, li\-vu hel\-ku sang ul\-te te'\-lan le\-fpom.} As for you all, while it's getting cooler, I hope your homes may be warm and your hearts happy.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Pey\-ral\-ìl zet wur\-a wu\-tsot a\-'aw\-nem pxel sngel.} Peyral treats a cool cooked meal like garbage.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{214}~\HL{sang}{\B{sang}})

% wutso
\hi 
\HT{wutso}{\B{\cp{wu}tso}} \I{["wu.\t{ts}o]} \emph{n.} dinner, served meal.
\B{Nì\-fra\-krr fol 'o\-lem a wu\-tso ftxì\-vä' lu nì\-ngay.} As always, the dinner they cooked tasted really terrible.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{Nì\-'aw\-ve oeng yo\-lom wu\-tsot; maw\-krr uvan si.} First we had dinner; afterwards we played.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b} 
\B{Pey\-ral\-ìl zet wur\-a wu\-tsot a\-'aw\-nem pxel sngel.} Peyral treats a cool cooked meal like garbage.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% wuwuk
\hi 
\HT{wuwuk}{\B{\cp{wu}wuk}} \I{["wu.wuk\textcorner]} \emph{or} \I{[.wUk\textcorner]} (\textsc{rn}: \B{\cp{wù}wùk} \I{["wU.wUk\textcorner]}) \emph{n., fauna} lizard, \emph{any of a variety of lizard\hyp{}like creatures}. 
\\
\end{multicols}


