\pdfbookmark[0]{Y}{Y}
\begin{center}
\section*{Y \I{[j]} (\emph{Yä})}
\end{center}

\begin{multicols}{2}

% ya
\hi 
\HT{ya}{\B{ya}} \I{[ja]} \emph{n.} air.
\B{Ya\-ri som\-wew\-pe (set)? -- Ya lu txa\-wew\slash wew\slash wur\slash tsya\-fe\slash sang\slash som\slash txa\-som.} What's the temperature? -- It's very cold\slash cold\slash cool\slash mild\slash warm\slash hot\slash very hot.~\LINK{http://naviteri.org/2011/05/weather-part-2-and-a-bit-more-2/}{b}
\B{Ya\-mì tsun fko tsi\-ve\-'a lo\-ran\-it re\-nu\-ä kil\-van\-ä slä kll\-te\-sìn wä\-pan.} From the air you can see the grace of the river's form but from the ground it's hidden.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yafkeyk}{\B{ya\-fkeyk}}, weather.  
\on{2}.~\HL{yapay}{\B{ya\-pay}}, mist, fog, steam. 
\on{3}.~\HL{yawo}{\B{ya\-wo}}, take off, launch.  
\on{4}.~\HL{yayo}{\B{ya\-yo}}, bird. 
\on{5}.~\HL{yrrap}{\B{yrr\-ap}}, storm. 
\on{6}.~\HL{txImya}{\B{txìm\-ya}}, fart.

% -ya
\hi 
\HT{ya2}{\B{-ya}} \I{[.ja]} \emph{suf. vocative for collective nouns}.
\B{Ma\-wey, Na'\-vi\-ya, ma\-wey!} Be calm, people, be calm.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Neytiri_talking_to_Jake}{a\slash wiki}
(\ding{43}~\HL{ma}{\B{ma}})

% yafkeyk
\hi 
\HT{yafkeyk}{\B{\cp{ya}fkeyk}} \I{["ja.fkEjk\textcorner]} \emph{n.} weather. 
\B{Ya\-fkeyk (za\-'u) fya\-pe?} How's the weather?~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
\B{Ya\-fkeyk fya\-pe set? -- Taw lu pi\-ak.} What's the weather like now? -- The sky is clear.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Tsun fko ay\-on\-ti fì\-wopx\-ä ni\-vìn fte ya\-fkeyk\-it sre\-si\-ve\-'a.} The shapes of clouds can be used to predict the weather.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Ya\-fkeyk\-ìri päng\-kxo fra\-po.} Everyone talks about the weather.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
(\ding{43}~\HL{ya}{\B{ya}}, \HL{tIfkeytok}{\B{tì\-fkey\-tok}})

% yak / si
\hi 
\HT{yak}{\B{yak}} \I{[jak\textcorner]} \textsc{i}. \emph{n.} fork, branch, point of divergence.
\B{Hay\-a yak\-ro fti\-vang. Sa\-lew rä\-'ä.} Stop at the next fork. Do not proceed further.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} \\
\textsc{ii}. \B{yak si} \I{[jak\textcorner si]} \emph{vin.} diverge, change direction, go astray.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Ne 'o\-ra\-tsyìp po\-lä\-hem, yak si nì\-ftär.} When you arrive at the pond, turn to the left.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
\B{Aw\-nga ze\-ne vi\-var ìlä fì\-sa\-lew\-fya. Zen\-ke yak si\-vi.} We must continue in this direction. We must not go astray.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}

% yaney
\hi 
\HT{yaney}{\B{ya\cp{ney}}} \I{[ja."nEj]} \emph{n.} canoe. \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{spulyaney}{\B{spul\-ya\-ney}}, canoe paddle.

% yapay
\hi 
\HT{yapay}{\B{\cp{ya}pay}} \I{["ja.paj]} \emph{n.} mist, fog, steam.
\B{neyn na ya\-pay}, the light, nondescript color of mist or fog.~\LINK{http://naviteri.org/2010/08/mipa-ayopin-mipa-ayliu-new-colors-new-words/}{b}
(\ding{43}~\HL{ya}{\B{ya}}, \HL{pay}{\B{pay}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{leyapay}{\B{le\-ya\-pay}}, foggy, misty.

% yawne
\hi 
\HT{yawne}{\B{\cp{yaw}ne}} \I{["jaw.nE]} \textsc{i}. \emph{adj.} (\emph{a.}) beloved.  
\B{Ma oe\-yä ey\-lan a\-yaw\-ne a lì'\-fya\-olo'\-mì.} My beloved friends in the language community.~\LINK{http://forum.learnnavi.org/news-announcements/happy-birthday-to-karyu-pawl/msg556899/\#msg556899}{ln} 
(\emph{b. to express}) to love s.o.\slash s.t. \B{Nga yaw\-ne lu oer.} I love you.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Lemondrop.com}{ld\slash wiki}
\B{Ay\-lì'\-fya yaw\-ne le\-ru oer ta\-krr\-a 'e\-veng la\-mu.} I've loved languages since I was a child.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} 
\B{Po yaw\-ne lu por.} He\slash she loves him\slash her.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Po yaw\-ne lu snor.} He loves himself.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Me\-fo yaw\-ne lu (snor) fì\-tsap.} They (= those two) love each other.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b} 
(\emph{c. idiom}) \B{Tse\-'a tìk yaw\-ne.} `Love at first sight.'~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} \\
\textsc{ii}. \B{\cp{yaw}ne slu} \I{["jaw.nE slu]} \emph{vin.} become beloved, fall in love.
\B{Lu poe se\-vin nì\-ftxan kuma yaw\-ne slo\-lu oer.} She was so beautiful that I fell in love with her.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Ma mun\-txa\-tu, oeng yaw\-ne lu oe\-nga\-ru fì\-tsap, ke\-fyak?} We love each other, don't we, my spouse?~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIyawn}{\B{tì\-yawn}}, love. 
\on{2}.~\HL{yawnetu}{\B{yaw\-ne\-tu\slash yawn\-tu}}, loved one. 
\on{3}.~\HL{yawnyewla}{\B{yawn\-yew\-la}}, broken heart, broken\hyp{}heartedness. 

% yawnetu / yawntu
\hi 
\HT{yawnetu}{\B{\cp{yaw}netu}} \I{["jaw.nE.tu]} \emph{or} \B{\cp{yawn}tu} \I{["jawn.tu]}  \emph{n.} loved one. 
\B{Hu\-fwa me\-fo le\-ru mun\-txa\-tu txan\-krr, mi lu mun\-snar ho\-na tì\-flrr a na pum me\-yaw\-ne\-tu\-ä a\-mip nì\-wotx.} Although the two of them have been mates a long time, they still have all the adorable tenderness of new sweethearts.~\LINK{http://naviteri.org/2013/02/vospxi-ayol-posti-apup-short-post-for-a-short-month/}{b}
(\ding{43}~\HL{yawne}{\B{yaw\-ne}}, \HL{tute}{\B{tu\-te}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yawntutsyIp}{\B{yawn\-tu\-tsyìp}}, darling.

% yawntutsyìp
\hi 
\HT{yawntutsyIp}{\B{\cp{yawn}tutsyìp}} \I{["jawn.tu.\t{ts}jIp\textcorner]} \emph{n.} darling, little loved one (\emph{often reflects parental or familial love}). 
\B{Sem\-pu\-ti rä\-'ä srätx, ma yawn\-tu\-tsyìp. Tì\-kang\-kem se\-ri.} Don't bother daddy, little one. He's working.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}  
(\ding{43}~\HL{yawntu}{\B{yawn\-tu}})

% yawnyewla
\hi 
\HT{yawnyewla}{\B{yawn\cp{yew}la}} \I{[jawn."jEw.la]} \emph{n.} broken heart; broken\hyp{}heartedness.   
\B{Lu Tse\-nur yawn\-yew\-la a lam fwa Va'\-rul pot txì\-yìng.} Tse\-nu is broken hearted that Va'\-ru appears to be about to dump her.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
(\ding{43}~\HL{yawne}{\B{yaw\-ne}}, \HL{yewla}{\B{yew\-la}})

% yawr
\hi 
\HT{yawr}{\B{yawr}} \I{[jawR]} \emph{n.} feather. \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yawrwll}{\B{yawr\-wll}}, feather blade.

% yawrwll
\hi 
\HT{yawrwll}{\B{\cp{yawr}wll}} \I{["jawR.w\s{l}]} \emph{n.} \ding{96} feather blade [lit.: `feather plant']. 
(\ding{43}~\HL{yawr}{\B{yawr}}, \HL{'ewll}{\B{'e\-wll}}) 

% yayl
\hi 
\HT{yayl}{\B{yayl}} \I{[jajl]} \emph{n.} nonsense, gibberish. 
\B{Fay\-lì\-'u a pol\-txe nga lu yayl nì\-'aw.} What you said is nothing but nonsense.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\emph{a. vulg.}) \B{Yayl!} `Bullshit!'  \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{leyayl}{\B{le\-yayl}}, nonsensical.

% yawo
\hi 
\HT{yawo}{\B{ya\cp{wo}}} \I{[ja."w$\cdot$o]} \emph{vin., inf.}\on{22} take off, launch.   
\B{Fwa ya\-wo ftu kll\-te to fwa tsway\-on ftu 'awkx lu ngä\-zìk.} Taking off from the ground is harder than flying off a cliff.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{'Uol ik\-ran\-it txo\-pu sley\-ko\-la\-tsu, ta\-lun\-a po tsìk ya\-wo.} Something must have frightened the banshee, because it suddenly took to the air.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Ik\-ran ya\-wo\-lo ftu txew 'awkx\-ä.} The banshee took to the air from the edge of a cliff.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
(\ding{214}~\HL{kllwo}{\B{kll\-wo}}, \ding{43}~\HL{ya}{\B{ya}}, \HL{wo}{\B{wo}})

% yayayr 
\hi 
\HT{yayayr}{\B{ya\cp{yayr}}} \I{[ja."jajR]} \emph{n.} (\emph{a.}) confusion.   
\B{Fì\-ya\-yayr\-ìri oe\-ru txoa li\-vu.} Forgive this confusion.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1910}{b}
(\emph{b. with dative and} \B{lu}) to be confused. \B{Saw\-tu\-te\-yä hem\-ìri lu aw\-nga\-ru ya\-yayr.} The Skypeople's actions confuse us.~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yayayrnga'}{\B{ya\-yayr\-nga'}}, confusing. 
\on{2}.~\HL{yayayrtsim}{\B{ya\-yayr\-tsim}}, s.t. confusing.

% yayayrnga'
\hi 
\HT{yayayrnga'}{\B{ya\cp{yayr}nga'}} \I{[ja."jajR.NaP]} \emph{or coll.} \I{[ja."jaj.NaP]} (\emph{in rapid speech}) \emph{adj.} confusing.
\B{Tsa\-tì\-oeyk\-tìng a\-ya\-yayr\-nga' srung ke so\-li oer fte tsli\-vam teyng\-ta kem\-pe ze\-ne si\-vi.} That confusing explanation didn't help me understand what I have to do.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b} 
(\ding{43}~\HL{yayayr}{\B{ya\-yayr}}, \HL{-nga'}{\B{-nga'}})

% yayayrtsim
\hi 
\HT{yayayrtsim}{\B{ya\cp{yayr}tsim}} \I{[ja."jajR.\t{ts}im]} \emph{n.} s.t. confusing, source of confusion.
(\ding{43}~\HL{yayayr}{\B{ya\-yayr}}, \HL{tsim}{\B{tsim}})

% yaymak
\hi 
\HT{yaymak}{\B{\cp{yay}mak}} \I{["jaj.mak\textcorner]} \emph{adj.} foolish, ignorant.
\B{Ki\-fkey\-ri fì\-tu\-tan\-ìl a\-yay\-mak ke tslam stum ke\-'ut.} This foolish man understands almost nothing of the world.~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}

% yayo 
\hi 
\HT{yayo}{\B{\cp{ya}yo}} \I{["ja.jo]} \emph{n., fauna} bird. 
\B{Fì\-fne\-ya\-yol tsrul\-it txu\-la mì song\-ropx ut\-ral\-ä fte ay\-li\-nit hi\-vaw\-nu wä sar\-ni\-o\-ang. Fì\-tì\-kan\-ìri ropx ke ha'.} This kind of bird builds its nest in a tree cavity so as to protect its young from predators. For this purpose, a hole going right through the tree trunk isn't suitable.~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b} 
(\ding{43}~\HL{ya}{\B{ya}}, \HL{ioang}{\B{ioang}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yayotsrul}{\B{ya\-yo\-tsrul}}, bird's nest.

% yayotsrul
\hi 
\HT{yayotsrul}{\B{\cp{ya}yotsrul}} \I{["ja.jo.\t{ts}Rul]} \emph{or} \I{[.\t{ts}RUl]} \emph{n.} bird's nest. 
\B{Tsa\-ya\-yo\-tsrul\-it sweyn, ma 'i\-tan.} Don't disturb that bird's nest, son. ~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\ding{43}~\HL{yayo}{\B{ya\-yo}}, \HL{tsrul}{\B{tsrul}})

% -yä
\hi 
\HT{-yA}{\B{-yä}} \I{[.j\ae{}]} \emph{suf. to mark the genitive case of a n. ending in a vowel other than u and o}. 
(\ding{43}~\HL{-A}{\B{-ä}})

% yäkx
\hi 
\HT{yAkx}{\B{yäkx}} \I{[j\ae{}k']} \emph{vtr.} not notice; ignore, snub, overlook intentionally.
\B{Sra\-ke fo hang\-ham ta\-lu\-na nga snaytx? Fo\-ti yäkx.} They're laughing because you lost? Ignore them.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
\B{Tsam\-si\-yu ze\-ne tsi\-vun yi\-väkx sne\-yä tì\-sraw\-it.} Warriors must be able to ignore their own pain.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\ding{214}~\HL{tseri}{\B{tse\-ri}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIyAkx}{\B{tì\-yäkx}}, lack of notice; snubbing. 
\on{2}.~\HL{sAyAkx}{\B{sä\-yäkx}}, snub.

% yän
\hi 
\HT{yAn}{\B{yän}} \I{[j\ae{}n]} \emph{vtr.} fasten, tie down. 

% ye
\hi 
\HT{ye}{\B{ye}} \I{[jE]} \emph{adj.} satisfied, content; satiated, ``full'' (\emph{used with} \HL{'efu}{\B{'efu}}).   
\B{Tsa\-ri\-a fkol po\-le\-'un fu\-ta Lo\-ak slu ta\-ron\-yu, sem\-pul 'efu ye.} Father is content that it's been decided Loak will be a hunter.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Nge\-yä tì\-kang\-kem\-ìri 'e\-fe\-iu oe ye nì\-txan. Sey\-so\-nìl\-tsan!} I'm very satisfied with your work. Well done!~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Sra\-ke ye\-hakx? -- Ye. Tsun ti\-vam.} Did you get enough to eat? -- Yes. That'll be enough.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yehakx}{\B{ye\-hakx}}, satisfied from hunger by food. 
\on{2}.~\HL{yevAng}{\B{ye\-väng}}, satisfied from thirst by drink.

% ye'krr
\hi 
\HT{ye'krr}{\B{\cp{ye'}krr}} \I{["jEP.k\s{r}]} \emph{adv.} early.
\B{Kar\-yul fngo\-lo' fu\-ta ay\-nu\-me\-yu pi\-va\-te ye'\-krr.} The teacher required the students to arrive early.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{leye'krr}{\B{le\-ye'\-krr}}, early.

% ye'rìn
\hi 
\HT{ye'rIn}{\B{\cp{ye'}rìn}} \I{["jEP.RIn]} \emph{adv.} soon.   
\B{Sìl\-pey oe, la\-yu oe\-ru ye'\-rìn sìl\-tsan\-a fmawn.} I hope, I will soon have good news.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Tsa\-ria nga\-hu ye'\-rìn ul\-txa si nì\-mun, oe sre\-fe\-re\-i\-ey nì\-prr\-te'.} I'm looking forward to getting together with you again soon.~\LINK{http://naviteri.org/2011/02/new-vocabulary-part-2/}{b}

% yehaw
\hi 
\HT{yehaw}{\B{\cp{ye}haw}} \I{["jE.haw]} \emph{adj.} well\hyp{}rested, having had enough sleep. 
\B{Lam fra\-kem le\-tsun\-slu krr\-a fko 'e\-fu ye\-haw.} Everything seems possible when one is well\hyp{}rested.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} 
(\ding{43}~\HL{ye}{\B{ye}}, \HL{hahaw}{\B{ha\-haw}})

% yehakx
\hi 
\HT{yehakx}{\B{\cp{ye}hakx}} \I{["jE.hak']} \emph{adj.} satisfied from hunger by food, ``full stomach''.  
\B{Sra\-ke ye\-hakx? -- Stum.\slash Ye. Tsun ti\-vam.\slash Nì\-txan!\slash Nì\-haawwwng.} Did you get enough to eat? -- Almost.\slash Yes. That'll be enough.\slash Very! I'm quite full.\slash Oooh. I ate too much.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
(\ding{43}~\HL{ye}{\B{ye}}, \HL{ohakx}{\B{o\-hakx}})

% yem
\hi 
\HT{yem}{\B{yem}} \I{[jEm]} \emph{vtr.} put, place. 
\B{Ngey tsko\-ti yem tsa\-yì\-sìn tsa\-krr za\-'u fì\-tseng!} Put your bow on that ledge and come here!~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yemfpay}{\B{yem\-fpay}}, dipping. 
\on{2}.~\HL{yemstokx}{\B{yem\-stokx}}, put on (clothing). 
\on{3}.~\HL{kllyem}{\B{kll\-yem}}, bury. 
\on{4}.~\HL{'awstengyem}{\B{'aw\-steng\-yem}}, join together.

% yemfpay / si
\hi 
\HT{yemfpay}{\B{yem\cp{fpay}}} \I{[jEm."fpaj]} \textsc{i}. \emph{n.} dipping, immersion (into a liquid). \\
\textsc{ii}. \B{yem\cp{fpay} si} \I{[jEm."fpaj si]} \emph{vin.} dip (into a liquid). 
\B{Ton\-ìri a\-la\-he, aw\-nga ke yem\-fpay si keng 'aw\-lo; fì\-txon yem\-fpay si me\-lo.} As for other nights, we don't even dip once; this night (we) dip twice.~\LINK{http://whyisthisnight.com/addl-more.html\#na\%27vi}{tn} 
(\ding{43}~\HL{yem}{\B{yem}}, \HL{nemfa}{\B{nem\-fa}} \HL{pay}{\B{pay}})

% yemstokx
\hi
\HT{yemstokx}{\B{\cp{yem}stokx}} \I{["j$\cdot$Em.stok']} \emph{vtr., inf.}\on{11} put on (functional clothing), don; wear. (\emph{from} \B{yem sìn tokx}, ``put on the body'')
\B{Pen\-it yem\-stokx!} Get dressed!~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Su\-nu oer haw\-re' a ngal yo\-lem\-stokx.} I like the hat you're wearing.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Fì\-re\-won ngal lum\-pe kea haw\-re'\-it ke yo\-lem\-stokx?} Why didn't you put on a hat this morning?~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/}{b}
\B{Fì\-ra\-spu' 'e\-kxin lu nì\-hawng. Ke tsun oe yi\-vem\-stokx.} These leggings are too tight. I can't wear them.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\ding{43}~\HL{ioi si}{\B{ioi si}}, \ding{214}~\HL{'aku}{\B{'a\-ku}})

% yengwal
\hi 
\HT{yengwal}{\B{yeng\cp{wal}}} \I{[jEN."wal]} \emph{n.} sorrow
\B{Sa'\-sem\-ìri lu 'eveng\-ä kxitx yeng\-wal a\-tu\-vom.} For a parent, the greatest sorrow is the death of a child.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}

% yerik
\hi 
\HT{yerik}{\B{\cp{ye}rik}} \I{["jE.Rik\textcorner]} \emph{n., fauna} yerik, hexapede, \emph{sexcruscervus caeruleus}.  
\B{Ye\-rik lu swi\-rä a\-nim nì\-txan.} The hexapede is a very timid creature.~\LINK{http://naviteri.org/2011/09/miscellaneous-vocabulary/}{b}
\B{Tì\-ran nì\-fnu txan\-txew\-vay fte\-ke ay\-ye\-rik\-ìl aw\-nga\-ti sti\-vawm.} Walk as quietly as possible so the hexapedes won't hear us.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/}{b} 
\B{Pol ye\-rik\-it ta\-ron nìl\-tsan fra\-to.} He hunts yerik the best of all.~\LINK{http://forum.learnnavi.org/language-updates/confirmations-\%28comparisons-gerund\%29/}{ln} 
\B{Zawr ye\-rik\-ä lu 'a\-ngo.} The call of the hexapede is quiet.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
\B{Ra\-lu\-ri fkan na ye\-rik.} Ralu smells (looks\slash sounds) like a hexapede.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\emph{a. idiom}) \B{ye\-rik (a) mì yrr\-ap} \emph{extreme panic, anxiety or timidity}, [lit.: Hexapede in a storm].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% yeväng
\hi 
\HT{yevAng}{\B{\cp{ye}väng}} \I{["jE.v\ae{}N]} \emph{adj.} satisfied from thirst by drink, feeling quenched\slash slaked.
\B{Sre\-krr 'a\-me\-fu väng, set ye\-väng.} Before, I was thirsty; now my thirst has been quenched.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} 
(\ding{43}~\HL{ye}{\B{ye}}, \HL{vAng}{\B{väng}}; \ding{214}~\HL{vAng}{\B{väng}})

% yewla / si
\hi 
\HT{yewla}{\B{\cp{yew}la}} \I{["jEw.la]} \textsc{i}. \emph{n.} disappointment, emotional let\hyp{}down, failed expectation.   
(\emph{a. with the dative and} \HL{lu}{\B{lu}}) to be disappointed. 
\B{Oer lu txan\-a yew\-la a ke tsun nga oe\-hu ki\-vä\-teng me\-srr\-ay.} I'm very disappointed you can't hang out with me the day after tomorrow.
(\emph{b. conv. idiom}) \B{Yew\-la!} Bummer! That's a shame! What a shame!~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\
\textsc{ii}. \B{\cp{yew}la si} \I{["jEw.la si]} \emph{vin.} disappoint s.o.
\B{O\-mum oel fu\-ta sä\-nui pe\-yä nga\-ru yew\-la so\-li nì\-txan.} I know that his failure disappointed you greatly.~\LINK{http://naviteri.org/2023/12/mipa-zisit-ayliu-amip-new-year-new-words/}{b}  \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{leyewla}{\B{le\-yew\-la}}, disappointing. 
\on{2}.~\HL{nIyewla}{\B{nì\-yew\-la}}, in a disappointing fashion. 
\on{3}.~\HL{yawnyewla}{\B{yawn\-yew\-la}}, broken heart.

% yewn
\hi 
\HT{yewn}{\B{yewn}} \I{[jEwn]} \emph{vtr.} express, convey (\emph{a thought or feeling}). 
\B{Oe new oey sì\-'e\-fut yi\-vewn po\-e\-ru, slä ke tsä\-ngun.} I want to express my feelings to her, but, sadly, I can't.~\LINK{http://naviteri.org/2020/11/vospxivopeya-ayliu-amip-novembers-new-words/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIyewn}{\B{tì\-yewn}}, expression (\emph{of a thought or feeling}).
\on{2}.~\HL{sAfpIlyewn}{\B{sä\-fpìl\-yewn}}, communicate. 

% yey 
\hi 
\HT{yey}{\B{yey}} \I{[jEj]} \emph{adj.} straight.
\B{Fì\-tskxe\-ri fa\-'o lu yey; ke lu ko\-um.} This rock has straight sides; it's not rounded.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{214}~\HL{lIktap}{\B{lìk\-tap}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIyey}{\B{nì\-yey}}, directly, straight to the point; just. 
\on{2}.~\HL{yeyfya}{\B{yey\-fya}}, straight line.

% yeyfya
\hi 
\HT{yeyfya}{\B{\cp{yey}fya}} \I{["jEj.fja]} \emph{n.} straight line.
\B{Wo\-leyn Ìstawl yey\-fyat mì hll\-te fte oeyk\-ti\-vìng fra\-po\-ru tì\-hawl\-te\-ri sne\-yä.} Ìstaw drew a line on the ground to explain his plan to everyone.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b}
(\ding{43}~\HL{yey}{\B{yey}}, \HL{fya'o}{\B{fya'o}})  \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIyeyfya}{\B{nì\-yey\-fya}}, straight ahead, in a straight line.

% yì
\hi 
\HT{yI}{\B{yì}} \I{[jI]} \emph{n.} (\emph{a.}) shelf, ledge, level, step, rung, \emph{a small, flat area on which one can stand or place an object, such as a foot}.
\B{Ta\-lun\-a vo\-spxì\-o a\-mrr ke zo\-lup tom\-pa, lä\-ngu yì kil\-van\-ä tìm nì\-txan.} Because it hasn't rained for five months, the river level is very low.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Ngey tsko\-ti yem tsa\-yì\-sìn tsa\-krr za\-'u fì\-tseng!} Put your bow on that ledge and come here!~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
(\emph{b. metaphorically with} \HL{kllkxem}{\B{kll\-kxem}} \emph{to refer to the level of anything scalable: water level, temperature, talent, anger, etc.}) 
\B{Lì'\-fya\-ri po (kll\-kxem) sìn pe\-yì?} [\emph{coll.}: \I{[sIm.pE."jI]}]\slash \B{Lì'\-fyari po pe\-yì?} How good is her Na'vi?~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Sìn yì sngä\-'i\-yu\-ä, slä tsye\-rìl (hay\-a yì\-ne) nì\-'ul\-'ul.} They're at a beginner's level, but they're getting better and better.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b}
\B{Sìn yì a ke tsun kaw\-tu spi\-vaw, nì\-win fra\-to.} He runs at an incredible level, faster than anyone else.~\LINK{http://naviteri.org/2013/01/lifyari-po-peyi-how-good-is-her-navi/}{b} 
(\emph{c. rhyming expression}) \B{Tslikx, tì\-ran, tul; | Ftu yì ne yì tsan\-'ul.} \emph{when learning s.t. new, you have to proceed from step to step: baby steps first, then bigger ones} [lit.: `Crawl, walk, run; from level to level get better.']~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{snayI}{\B{sna\-yì}}, staircase, series of step\hyp{}like levels. 
\on{2}.~\HL{kxaylyI}{\B{kxayl\-yì}}, high level. 
\on{3}.~\HL{kxamyI}{\B{kxam\-yì}}, intermediate level. 
\on{4}.~\HL{tImyI}{\B{tìm\-yì}}, low level. 
\on{5}.~\HL{fyanyI}{\B{fyan\-yì}}, shelf.
\on{6}.~\HL{yIspul}{\B{yì\-spul}}, mermaid tail.
\on{7}.~\HL{heynyI}{\B{heyn\-yì}}, lap.
\on{8}.~\HL{swayI}{\B{swa\-yì}}, generation within a family.

% yìm
\hi 
\HT{yIm}{\B{yìm}} \I{[jIm]} \emph{vtr.} bind. 
\B{Ay\-wayl yìm ki\-fkey\-ä | 'Ì\-he\-yut a\-vo\-mrr | Sìn ti\-rea\-fya\-'o a\-vol | Na way\-te\-lem\-ä hìng.} The songs bind | the thirteen spirals of the solid world | To the eight spirit paths | Like the threads of a Songcord.~\LINK{http://www.pandorapedia.com/navi/music/ritual_music}{pp}
\B{Me\-fo\-ti yìm.} Tie them up.~\LINK{https://wiki.learnnavi.org/Na\%27vi_from_Avatar_Movie\#Attack_on_hometree}{a} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{sAyIm}{\B{sä\-yìm}}, tie, s.t. used for binding.

% yìspul
\hi 
\HT{yIspul}{\B{\cp{yì}spul}} \I{["jI.spul]} \emph{or} \I{[.spUl]} \emph{n.} \ding{96} mermaid tail. 
(\ding{43}~\HL{yI}{\B{yì}}, \HL{spule}{\B{spu\-le}})

% yll
\hi 
\HT{yll}{\B{yll}} \I{[j\s{l}]} \emph{adj.} communal. 
\B{Kxawm set 'u a\-ngä\-zìk fra\-to lu la\-'a a\-yll.} Maybe the most difficult thing now is social distance.~\LINK{http://naviteri.org/2020/03/teri-tifkeytok-lefkrr-about-the-present-situation/}{b} \\
(i.) \B{swoa\-sey a\-yll}, large social kava bowl.~\LINK{http://naviteri.org/2015/07/aylifyavi-lereyfya-1-cultural-terms-1/}{b} \\
(ii.) \B{ko\-ren a\-yll} law, societal rule.
\B{Pol aw\-nga\-ru ho\-ren\-it a\-yll ley\-kek.} He makes us obey the laws.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{nIyll}{\B{nì\-yll}}, communally, in a communal manner. 
\on{2}.~\HL{ylltxep}{\B{yll\-txep}}, communal fire.

% ylltxep
\hi 
\HT{ylltxep}{\B{\cp{yll}txep}} \I{["j\s{l}.t'Ep\textcorner]} \emph{n.} communal fire or fire pit.
\B{Nì\-trr\-trr yom Na'\-vil wu\-tsot 'aw\-si\-teng pxaw yll\-txep.} The Na'vi regularly eat dinner together around a communal fire.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}
(\ding{43}~\HL{yll}{\B{yll}}, \HL{txep}{\B{txep}}, \HL{txeptseng}{\B{txep\-tseng}}) 

% yo
\hi 
\HT{yo}{\B{yo}} \I{[jo]} \emph{n.} (\emph{a.}) surface.
\B{Yo\-sìn kil\-van\-ä lu tatx.} There are bubbles on the surface of the river.~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
\B{Krr\-a hu\-fwe tul nì\-win, tsun fko tsi\-ve\-'a ay\-we\-opx\-it a sìn yo pay\-ä.} When there is strong wind, you can see waves on the water.~\LINK{http://naviteri.org/2017/09/zisikrr-amip-ayliu-amip-new-words-for-the-new-season/}{b}
(\emph{b. coll. form of} \B{fyan\-yo}) table. 
\B{Yo\-ti fwal ru\-txe. Mei slo\-lu maw tom\-pa.} Please wipe the table. It's gotten wet after the rain.~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
\B{Oe\-yä tstal\-ìl tok pe\-se\-nget? -- Lu yo\-sìn.} Where's my knife? -- It's on the table.~\LINK{http://naviteri.org/2015/08/aylifyavi-lereyfya-2-cultural-terms-2-and-more/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{txayo}{\B{txa\-yo}}, field, plain. 
\on{2}.~\HL{kxemyo}{\B{kxem\-yo}}, wall.
\on{3}.~\HL{fyanyo}{\B{fyan\-yo}}, table. 
\on{4}.~\HL{yomyo}{\B{yom\-yo}}, plate (for food).
\on{5}.~\HL{fApyo}{\B{fäp\-yo}}, roof.

% yo'
\hi 
\HT{yo'}{\B{yo'}} \I{[joP]} \emph{vin.} be perfect, flawless.
\B{Tì\-hawl le\-sngä\-'i lu tì\-kang\-kem\-vi skxawng\-ä, slä pum alu fì\-'u yo' nì\-'aw.} The original plan was the work of an idiot, but this one is just perfect.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Fì\-stxel\-it fol txe\-ru\-la fpi olo'\-eyk\-tan. Ze\-ne yi\-vo' lu\-ke kxey\-ey\-o kaw\-'it.} They're constructing this gift for the chief. It must be perfect without a single flaw.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
\B{Ri\-ni\-ri nik\-re yä\-ngo' nì\-tut.} Rini's hair is always perfect.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\emph{b. conv. idiom}) \B{Ye\-io'}, perfect. 
\B{Ul\-txa si\-vi oeng sìn ram\-tsyìp txon\-'ong\-ay. -- Ye\-io'! Tsa\-krr\-vay ko!} Let's meet on the hill tomorrow at nightfall. -- Perfect! See you then.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIyo'}{\B{tì\-yo'}}, perfection. 
\on{2}.~\HL{nIyo'}{\B{nì\-yo'}}, pefectly, flawlessly. 
\on{3}.~\HL{yo'ko}{\B{yo'\-ko}}, circle. 

% yo'ko
\hi 
\HT{yo'ko}{\B{\cp{yo'}ko}} \I{["joP.ko]} \emph{n.} circle, \emph{a perfectly circular ring}. 
(\ding{43}~\HL{yo'}{\B{yo'}}, \HL{koum}{\B{ko\-um}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yo'kofya}{\B{yo'\-ko\-fya}}, cycle.

% yo'kofya
\hi
\HT{yo'kofya}{\B{\cp{yo'}kofya}} \I{["joP.ko.fja]} \emph{n.} cycle.
\B{Hum zì\-sìt a\-lal, pä\-hem pum a\-mip. Yo'\-ko\-fya a\-tì\-'i\-lu\-ke.} The old year leaves, the new one arrives. An endless cycle.~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b}
(\ding{43}~\HL{yo'ko}{\B{yo'\-ko}}, \HL{fya'o}{\B{fya\-'o}})

% yoa
\hi 
\HT{yoa}{\B{\cp{yo}a}} \I{["jo.a]} \emph{adp.} in exchange for (\emph{usually used with} \B{tìng}, \B{tel}, \B{ka\-nom}, \B{stxe\-nu\-tìng}, \B{mll\-'an})
\B{Oel to\-lìng nga\-ru tsngan\-it yoa fkxen.} I traded you meat for vegetables.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Fol ko\-la\-nom po\-ta ay\-srok\-it fay\-o\-ang\-yoa.} They bartered fish with him for beads.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Ta\-yel Tse\-nul pxe\-swi\-zaw\-ti yoa mun\-sna\-hawn\-ven.} Tsenu will receive three arrows in exchange for a pair of shoes.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Kä\-sro\-lìn oel nik\-ro\-it Pey\-ral\-ur yoa fwa po rol oer.} I loaned Peyral a hair ornament in exchange for her singing to me.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Fu\-ta nga\-ta tel pxen\-yoa srät, mll\-'e\-i\-an oel.} I'm happy to take cloth from you for finished garments.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Poel haw\-re'\-tsyìp\-it ko\-la\-nom yoa txo\-lar a\-me\-vol.} She bought a little cap for \$\on{16}.~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}
\B{Kì\-ma\-nom oel mip\-a el\-tut le\-fngap yoa ew\-ro.} I just bought a new computer (for an unspecified amount of euros).~\LINK{http://naviteri.org/2014/02/barter-and-exchange/}{b}

% yokx
\hi 
\HT{yokx}{\B{yokx}} \I{[jok']} \emph{n.} shield.

% yol
\hi 
\HT{yol}{\B{yol}} \I{[jol]} \emph{adj.} short (of time).
\B{Yol\-a krr, txan\-a krr, ke tsran\-ten.} It doesn't matter how long it takes.~\LINK{http://forum.learnnavi.org/language-updates/a-collection/}{ln\slash a\textsuperscript{c}}
\B{Sra\-ne, lu yol\-a lì'\-fya\-vi.} Yes, there is a short expression.~\LINK{http://naviteri.org/2012/10/snewsyea-ftxoza-halowina-a-spooky-halloween/comment-page-1/\#comment-1755}{b}
\B{Maw sä\-tsway\-on a\-yol ay\-oe kll\-po\-lä mì ta\-yo a lu ro\-fa kil\-van.} After a short flight we landed in a field beside the river.~\LINK{http://naviteri.org/2012/01/mipa-zisit-ayliu-amip-new-words-for-the-new-year/}{b}
(\ding{214}~\HL{nun}{\B{nun}}, \HL{txan}{\B{txan}}) 

% yom 
\hi 
\HT{yom}{\B{yom}} \I{[jom]} \emph{vtr.} eat.
\B{Yom tey\-lut keng oel!}  Even \emph{I} eat teylu!~\LINK{http://forum.learnnavi.org/language-updates/even-keng/}{ln}
\B{Oe new yi\-vom tey\-lut.} I want to eat teylu.~\LINK{http://wiki.learnnavi.org/index.php?title=Canon\#Extracts_from_various_emails}{wiki} 
\B{Oe ye\-rom set.} I'm eating now.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
\B{Nì\-'aw\-ve oeng yo\-lom wu\-tsot; maw\-krr uvan si.} First we had dinner; afterwards we played.~\LINK{http://naviteri.org/2011/08/new-vocabulary-clothing/comment-page-1/\#comment-917}{b}
\B{Oe nga\-to yom nì\-'ul.} I eat more than you.~\LINK{http://forum.learnnavi.org/language-updates/prefer-x-to-y-24474/}{ln}
(i.) \B{mo a yom}, dining room, dining area.
(\emph{a. proverb}) \B{Ta\-ron\-yut yom smar\-ìl.} \emph{A totally botched situation, chaos} [lit.: the prey eats the hunter].
\B{Eyk Ka\-mun a fra\-lo lä\-ngu tsa\-sä\-ta\-ron vel\-ke nì\-wotx. Ta\-ron\-yut yom smar\-ìl!} Every time Kamun is in charge, the hunt is a mess. Everything goes wrong that can!~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yomhI'ang}{\B{yom\-hì\-'ang}}, dakteron.  
\on{2}.~\HL{yomioang}{\B{yom\-io\-ang}}, chalice plant. 
\on{3}.~\HL{yomtIng}{\B{yom\-tìng}}, feed. 
\on{4}.~\HL{yomyo}{\B{yom\-yo}}, plate (for food).
\on{5}.~\HL{yomvey}{\B{yom\-vey}}, dine on flesh, be carnivorous.
\on{6}.~\HL{yomzeswa}{\B{yom\-ze\-swa}}, graze.
\on{7}.~\HL{ukyom}{\B{uk\-yom}}, eclipse.

% yomhì'ang
\hi 
\HT{yomhI'ang}{\B{yom\cp{hì'}ang}} \I{[jom."hIP.aN]} \emph{n.}, \ding{96} dakteron, `insectivore', \emph{pseudopenthes coralis}.
(\ding{43}~\HL{yom}{\B{yom}}, \HL{hI'ang}{\B{hì\-'ang}})

% yomioang
\hi 
\HT{yomioang}{\B{yomi\cp{o}ang}} \I{[jom.i"o.aN]} \emph{n.}, \ding{96} chalice plant, \emph{pseudocenia rosea}.
(\ding{43}~\HL{yom}{\B{yom}}, \HL{ioang}{\B{ioang}})

% yomtìng
\hi 
\HT{yomtIng}{\B{\cp{yom}tìng}} \I{["jom.t$\cdot$IN]} \emph{vin., inf.}\on{22} feed.   
\B{Sa'\-nok prr\-nen\-ur yom\-tìng fa okup sne\-yä.} A mother feeds an infant with her milk.~\LINK{http://naviteri.org/2013/06/looking-back-looking-forward/}{b}
\B{Ta\-ron\-yu a\-swey lu tsa\-tu\-te a ta\-ron fte yom\-ti\-vìng ay\-la\-ru.} The best hunter is the one who hunts to feeds others.~\LINK{http://forum.learnnavi.org/language-updates/clarification-on-the-use-of-awpo/}{ln}
\B{Ha ftxey | 'aw\-pot set ay\-ngal a lu ay\-nga\-kip | 'Aw\-pot a Na'\-vi\-ru yom\-tì\-yìng.} So choose | one among you | Who will feed the People.~\LINK{http://naviteri.org/2013/08/taronway-the-hunt-song/}{b}
(\ding{43}~\HL{yom}{\B{yom}}, \HL{tIng}{\B{tìng}})

% yomvey
\hi 
\HT{yomvey}{\B{yom\cp{vey}}} \I{[j$\cdot$om."vEj]} \emph{vin., inf.}\on{11} dine on flesh, be carnivorous. 
\B{Pa\-lu\-kan\-tsyìp yom\-vey nì\-wotx.} All cats are carnivorous.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{yom}{\B{yom}}, \HL{vey}{\B{vey}})

% yomyo
\hi 
\HT{yomyo}{\B{\cp{yom}yo}} \I{["jom.jo]} \emph{n.} plate (for food).
\B{Ma sem\-pu, oey yom\-yo tso\-lawng!} Daddy, my plate broke!~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
\B{Ma En\-tu, ngal lum\-pe ngey tsmu\-ke\-yä yom\-yot tsey\-ko\-lawng?} En\-tu, why did you break your sister's plate?~\LINK{http://naviteri.org/2020/04/aawa-u-amip-a-few-new-things/}{b}
(\ding{43}~\HL{yom}{\B{yom}}, \HL{yo}{\B{yo}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yomyolerIk}{\B{yom\-yo le\-rìk}}, leaf plate.

% yomyo lerìk
\hi 
\HT{yomyolerIk}{\B{\cp{yom}yo le\cp{rìk}}} \I{["jom.jo{} lE."rIk\textcorner]} \emph{n.} leaf plate; (\emph{coll.}: \HL{rIk}{\B{rìk}})
(\ding{43}~\HL{yomyo}{\B{yom\-yo}}, \HL{lerIk}{\B{le\-rìk}})

% yomzeswa
\hi 
\HT{yomzeswa}{\B{yom\cp{ze}swa}} \I{[j$\cdot$om."zE.swa]} \emph{vin., inf.}\on{11} graze. 
\B{Sna\-ye\-rik ye\-rom\-ze\-swa mì ta\-yo.} A herd of hexapedes are grazing in the field.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b} 
(\ding{43}~\HL{yom}{\B{yom}}, \HL{zeswa}{\B{ze\-swa}})

% yora'
\hi 
\HT{yora'}{\B{yo\cp{ra'}}} \I{[jo."RaP]} \emph{vtr.} win.
\B{Krr\-a wä\-tut tse\-'a, pe\-yä sä\-sìl\-pey a yo\-ra' 'o\-lìp.} When he saw his opponent, his hope of winning vanished.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tsu'\-tey\-ri lu fo\-ru nrr\-a a fì\-re\-won yo\-lo\-ra'.} They're proud of Tsu'\-tey for having won this morning.~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}  
\B{Am\-'a\-lu\-ke sna\-yaytx Saw\-tu\-te, ya\-yo\-ra' Na'\-vi.} Without a doubt the Sky People will lose and the People will win.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Yì\-mo\-ra' Tsu'\-tey\-ìl uvan\-it.} Tsu'\-tey just won the game.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b}
(\ding{214}~\HL{snaytx}{\B{snaytx}}) \\  
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{yora'tu}{\B{yo\-ra'\-tu}}, winner. 
\on{2}.~\HL{tIyora'}{\B{tì\-yo\-ra'}}, victory, a win.

% yora'tu
\hi 
\HT{yora'tu}{\B{yo\cp{ra'}tu}} \I{[jo."RaP.tu]} \emph{n.} winner. 
\B{Fra\-u\-van\-ìri lu yo\-ra'\-tu, lu snay\-tu.} For every game, there's a winner and a loser.~\LINK{http://naviteri.org/2011/12/a-note-on-the-word-\%E2\%80\%9Cyora\%E2\%80\%99tu\%E2\%80\%9D/}{b}
\B{Ay\-nga\-ru tsìng\-a yo\-ra'\-tut!} May I present the four winners!~\LINK{http://naviteri.org/2014/12/ayngaru-tsinga-yoratu-may-i-present-the-four-winners/}{b}
(\ding{214}~\HL{snaytu}{\B{snay\-tu}}, \ding{43}~\HL{yora'}{\B{yo\-ra'}})

% yoten
\hi 
\HT{yoten}{\B{\cp{yo}ten}} \I{["jo.tEn]} \emph{n., fauna} yoten. 

% yrr
\hi 
\HT{yrr}{\B{yrr}} \I{[j\s{r}]} \emph{adj.} wild, natural (\emph{s.t. in its original, unmodified, untampered\hyp{}with natural state}), (\emph{of food}) raw.
\B{Ik\-ran\-ìri krr\-a hu tu\-te tsa\-heyl si, ftang li\-vu yrr.} When an \emph{ikran} bonds with a person, it ceases to be wild.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Lu tsa\-fne\-pay\-oang ftxì\-lor fra\-to krr\-a lu yrr.} That kind of fish is most tasty when eaten as \emph{sashimi}.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Txep\-ìri, yrr\-a rìn\-ti rä\-'ä sar, ma skxawng.} Don't use that green wood for a fire, you fool.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{214}~\HL{zAfi}{\B{zä\-fi}})

% yrrap
\hi 
\HT{yrrap}{\B{\cp{yrr}ap}} \I{["j\s{r}.ap\textcorner]} \emph{n.} storm.   
\B{Fra\-po ne mì\-fa! Le\-rok yrr\-ap apxa!} Everybody inside! A big storm is approaching!~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b}
(\emph{a. idiom}) \B{ye\-rik (a) mì yrr\-ap} \emph{extreme panic, anxiety or timidity}, [lit.: Hexapede in a storm].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
(\ding{43}~\HL{ya}{\B{ya}}, \HL{hrrap}{\B{hrr\-ap}})

% -yu
\hi
\HT{-yu}{\B{-yu}} \I{[.ju]} \emph{suf. to turn verbs into an agent, e.g.:} \B{ta\-ron\-yu}, hunter; \B{ma\-wey\-pey\-yu}, patient person, \emph{etc.}
(\ding{43}~\HL{-tu}{\B{-tu}}) 

% yuey
\hi 
\HT{yuey}{\B{\cp{yu}ey}} \I{["ju.Ej]} \emph{adj.} beautiful (\emph{inner beauty, only for people}).
\B{Lu poe lor, lu yu\-ey nì\-teng.} She's beautiful on the outside \emph{and} the inside.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\ding{43}~\HL{lor}{\B{lor}})

% yune
\hi 
\HT{yune}{\B{\cp{yu}ne}} \I{["ju.nE]} \emph{vtr.} listen to (\texttt{+}\emph{control}). 
\B{Nga ze\-ne ay\-lì\-'ut kar\-yu\-ä yi\-vu\-ne, ma 'evi.} You must listen to your teacher, my son.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
\B{Yìl\-mu\-ne\-ie oel ay\-nge\-yä stä'\-nì\-pam\-it sok\-a tsa\-pos\-tì\-yä oe\-yä.} I was happy to listen to your recording of that recent blog post of mine.~\LINK{https://forum.learnnavi.org/projects/an-audio-message-for-pawl/msg670815/\#msg670815}{ln}
(\ding{43}~\HL{mikyun}{\B{tìng mik\-yun}}, \HL{stawm}{\B{stawm}})

% yur
\hi 
\HT{yur}{\B{yur}} \I{[juR]} \emph{or} \I{[jUR]} \emph{vtr.} wash, clean.  
\B{Ze\-ne nga yi\-vur pxìm fì\-skxir\-it.} You must clean the wound frequently.~\LINK{http://naviteri.org/2011/05/some-miscellaneous-vocabulary/}{b} 
\B{Oe yä\-pur.} I wash myself.~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln}
\B{Oe yä\-pe\-rur.} I'm washing myself.~\LINK{http://forum.learnnavi.org/language-updates/the-reflexive-\%28from-a-frommer-email!!!!\%29/}{ln} \\
\\
(i.) \B{txaw\-nul\-srung a yur}, washing machine.~\LINK{https://naviteri.org/2024/03/fleltrra-ayliu-words-for-april-fools-day/}{b}
\\
\end{multicols}