\pdfbookmark[0]{Z}{Z}
\begin{center}
\section*{Z \I{[z]} (\emph{Zä})}
\end{center}

\begin{multicols}{2}

% za'ärìp
\hi
\HT{za'ArIp}{\B{za\cp{'ä}rìp}} \I{[za."P$\cdot$\ae{}.R$\cdot$Ip\textcorner]} \emph{vtr., inf.}\on{23} pull (\emph{coll. also} \ding{43}~\HL{zArIp}{\B{zä\-rìp}}). 
(\ding{214}~\HL{kA'ArIp}{\B{kä\-'ä\-rìp}}, \ding{43}~\HL{'ArIp}{\B{'ärìp}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{wrrzArIp}{\B{wrr\-zä\-rìp}}, pull out, extract.

% za'u 
\hi
\HT{za'u}{\B{\cp{za}'u}} \I{["za.Pu]} \emph{vin.} come (\HL{ne}{\B{ne}}, to\slash towards (\emph{can be optionally omitted if the destination comes after the verb}); \HL{ftu}{\B{ftu}}, from (a place); \HL{ta}{\B{ta}}, from (\emph{fig.})).   
\B{Nga za\-'u fì\-tseng pxìm srak?} Do you come here often?~\LINK{http://www.lemondrop.com/2010/01/26/avatar-pick-up-lines-talk-dirty-in-Navi/}{ld}
\B{Ftxey nga za\-'u fu\-ke?} Are you coming (or not)?~\LINK{http://forum.learnnavi.org/language-updates/if-vs-whether-ftxey-fuke/msg156629/\#msg156629}{ln}
\B{Za\-'u kal\-txì si ko!} Come and say hello!~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-3/}{b} 
\B{Za\-'u fra\-'u ne tu\-te lem\-wey\-pey, ke\-fyak?} Everything comes to the patient people, right?~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-2/}{b}
\B{Hawng\-krr rä\-'ä zi\-va\-'u!} Don't come late!~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b} 
\B{Nga zo\-la\-'u ftu pe\-seng?} Where are you from?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Pxay\-a a\-to\-ki\-ri\-na' ne fì\-ke\-tu\-wong zì\-ma\-'u.} Many \emph{atokirina} just came to this alien.~\LINK{http://forum.learnnavi.org/language-updates/pseudo-canon-from-bd-live-content/}{a\textsuperscript{c}\slash ln}
(\emph{a. idiom}) \B{zo\-la\-'u nì\-prr\-te'}, ``welcome'' (\B{ne}, to) \LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Zo\-la\-'u nì\-prr\-te' ay\-nga\-ru nì\-wotx!} Welcome to you all!~\LINK{http://naviteri.org/2010/06/first-post/}{b}
\B{Kal\-txì, ma fra\-po, ul\-te zo\-la\-'u nì\-prr\-te' aw\-nge\-yä nu\-mul\-txa\-ne a te\-ri lì'\-fya le\-Na'\-vi.} Hello, everyone and welcome to our Na'\-vi language class.~\LINK{http://naviteri.org/2012/07/tskxekengtsyip-a-mikyunfpi-a-little-listening-exercise/}{b}
(\emph{b. idiom}) \B{za\-'u nì\-'eng ne}, share an interest in common with s.o. 
\B{Tì\-ru\-sol za\-'u ne fo nì\-'eng.} They share an interest in singing.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
\B{So\-rewn\-ìl kan\-'ìn tì\-'em\-it nì\-txan ul\-te kxawm tsa\-txe\-le me\-nga\-ne za\-'a\-tsu nì\-'eng.} Sorewn is really into cooking and you two might share that in common.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-1/}{b}
(\emph{c. of the weather}) \B{Ya\-fkeyk (za\-'u) fya\-pe?} How's the weather?~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b} 
(\emph{d. idiom}) \emph{with} \HL{ne}{\B{ne}}, \emph{an adv. and a sense organ to convey} `s.o. looks\slash sounds\slash tastes\slash smells good or bad'
\B{Nga za\-'u nìl\-tsan ne na\-ri.} You look good.~\LINK{https://naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b}
\B{Fì\-syu\-ve za\-'u nì\-fe' ne ftxì.} This food tastes bad.~\LINK{https://naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b} 
\B{Nga za\-'u nì\-fpxa\-mo ne mik\-yun. Mok\-ri\-ri kem\-pe lo\-len?} You sound terrible. What happened to your voice?~\LINK{https://naviteri.org/2025/07/ayu-leketengsyon-various-things/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{emza'u}{\B{emz\-a\-'u}}, pass, overcome.  
\on{2}.~\HL{fAza'u}{\B{fä\-za\-'u}}, come up, ascend. 
\on{3}.~\HL{kllza'u}{\B{kll\-za\-'u}}, come down, descend. 
\on{4}.~\HL{wrrza'u}{\B{wrr\-za\-'u}}, come out, emerge. 
\on{5}.~\HL{zIma'uyu}{\B{zì\-ma\-'u\-yu}}, newcomer. 
\on{6}.~\HL{zawprrte'}{\B{zaw\-prr\-te'}}, be enjoyable. 
\on{7}.~\HL{zusawkrr}{\B{zu\-saw\-krr}}, future.

% zakru
\hi 
\HT{zakru}{\B{\cp{zak}ru}} \I{["zak\textcorner.Ru]} \emph{n., fauna} zakru. 

% zal
\hi 
\HT{zal}{\B{zal}} \I{[zal]} \emph{n.} (natural) clay (\emph{usually for pottery}) or baked clay. 
\B{Fkol fì\-fne\-tsngal\-it txu\-la ta zal.} This kind of cup is made of clay.~\LINK{https://naviteri.org/2024/09/pxevola-liu-amip-two-dozen-new-words/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zAfizal}{\B{zä\-fi\-zal}}, fired clay, ceramic material.

% zam
\hi
\HT{zam}{\B{zam}} \I{[zam]} \emph{num., adj.} \on{64} (decimal), \on{100} (octal). (\emph{base}: \B{za-})
\B{Zam\-a Lì\-'u A\-mip.} \on{64} new words.~\LINK{http://naviteri.org/2018/04/100a-liu-amip-64-new-words-part-2/}{b}

% zam mì zam
\hi 
\HT{zammIzam}{\B{zam mì zam}} \I{[zam mI zam]} \emph{adv.} completely, one hundred percent. 
\B{Nga\-hu mll\-te oe zam mì zam.} I agree with you one hundred percent.~\LINK{http://naviteri.org/2021/12/zolau-niprrte-ma-3746-welcome-2022/}{b} 

% zamunge
\hi
\HT{zamunge}{\B{za\cp{mu}nge}} \I{[za."m$\cdot$u.N$\cdot$E]} (\textsc{rn}: \B{za\cp{mù}nge} \I{[za."m$\cdot$U.N$\cdot$E]}) \emph{vtr., inf.}\on{23} bring.
\B{Ay\-eveng\-ur kxo\-lìm sa'\-nok\-ìl tson\-ta pay\-it za\-mu\-nge.} The children were told by their mother to fetch water.~\LINK{http://naviteri.org/2014/10/tson-si-fpomron-obligation-and-mental-health/}{b}
\B{Pa\-ye\-'un swey\-a fya\-'ot a za\-mi\-vu\-nge oel ay\-ngar ay\-lì\-'ut ho\-ren\-ti\-sì lì'\-fya\-yä le\-Na'\-vi.} I will decide the best way to bring you the words and rules of Na'vi.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln}
\B{Ay\-fol za\-mo\-lu\-nge aw\-ngar ay\-ri\-na'\-it fa sä\-he\-na a\-pxa.} They brought us the seeds in a large container.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\ding{43}~\HL{munge}{\B{mu\-nge}})

% zapxì   
\hi
\HT{zapxI}{\B{za\cp{pxì}}} \I{[za."p'I]} \emph{n.} front (part or section). 
(\ding{43}~\HL{hapxI}{\B{ha\-pxì}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zare'}{\B{zare'}}, forehead, brow.

% zare'
\hi
\HT{zare'}{\B{za\cp{re'}}} \I{[za."REP]} \emph{n.} forehead, brow.  
(\ding{43}~\HL{zapxI}{\B{za\-pxì}}, \HL{re'o}{\B{re\-'o}})

% zasrìn 
\hi
\HT{zasrIn}{\B{za\cp{srìn}}} \I{[za."sR$\cdot$In]} \emph{vtr., inf.}\on{22} borrow. 
\B{New Va'\-ru tskot Ey\-tu\-kan\-ä za\-sri\-vìn.} Va'ru wants to borrow Eytukan's bow.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}
\B{Fra\-syu\-ra\-ti fkol za\-sro\-lìn nì\-'aw ul\-te trr\-o ze\-ne tey\-ki\-vä\-txaw.} All energy is only borrowed, and one day it will have to be given back.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b} 
(\ding{43}~\HL{srIn}{\B{srìn}})

% zazam 
\hi
\HT{zazam}{\B{\cp{za}zam}} \I{["za.zam]} \emph{num., adj.} \on{4096} (decimal), \on{10,000} (octal). 

% zawng
\hi
\HT{zawng}{\B{zawng}} \I{[zawN]} \emph{vin.} scream. 
\B{Tì\-txen si, ma 'ite! Unil sar\-mi nga teng\-krr ze\-rawng. Lu fpom srak?} Wake up, daughter! You were dreaming and screaming. Are you okay?~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}

% zawprrte'
\hi
\HT{zawprrte'}{\B{zaw\cp{prr}te'}} \I{[z$\cdot$aw."p\s{r}.tEP]} \emph{vin., inf.}\on{11} be enjoyable (\emph{deriving physical or emotional pleasure from s.t.}); \emph{used with} \ding{43}~\HL{ne}{\B{ne}}. 
\B{Tsa\-fne\-pam\-tseo ke zaw\-prr\-te' oe\-ne.} I don't enjoy that kind of music.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Sìl\-pey oe tsnì fpom li\-vu ay\-nga\-ru nì\-wotx ul\-te snge\-rä\-'i a zì\-sì\-krr a\-sang zi\-vaw\-prr\-te' ay\-nga\-ne.} I hope that you are all well and that you are enjoying the beginning warm season.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{zaw\-prr\-te' nì\-layl fko\-ne}, be pleasurable to one in an innocent way.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
\B{zaw\-prr\-te' nì\-to\-kat fko\-ne}, be pleasurable to one in a guilty way.~\LINK{http://naviteri.org/2023/09/aawa-ayliu-si-aylifyavi-amip-a-few-new-words-and-expressions/}{b}
(\ding{43}~\HL{za'u}{\B{za\-'u}}, \HL{nIprrte'}{\B{nì\-prr\-te'}})

% zawr
\hi
\HT{zawr}{\B{zawr}} \I{[zawR]} \emph{n.} animal call, an animal cry. 
\B{Zawr ye\-rik\-ä lu 'a\-ngo.} The call of the hexapede is quiet.~\LINK{http://naviteri.org/2014/11/twenty-before-the-holidays/}{b}
(\emph{a. idiom}) \B{zawr (a) mì na'\-rìng}, ``Old news.'' [lit.: an animal cry in the forest].~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}
\B{Sra\-ke ngal sto\-lawm fu\-ta Tse\-nu Ra\-lur mo\-wan lu nì\-txan? -- Zawr mì na'\-rìng, ma tsmuk! Tsat o\-mum oel kin\-trr\-o.} Have you heard that Ralu has the hots for Tsenu? -- Old news, brother! I've known it for a week.~\LINK{http://naviteri.org/2021/08/ulte-ayyoratu-leiu-and-the-winners-are-2/}{b}

% zawsena 
\hi
\HT{zawsena}{\B{\cp{zaw}sena}} \I{["zaw.sE.na]} \emph{n., coll.} quiver (\emph{attached to the ik\-ran's saddle}), \emph{shortened form of} \ding{43}~\HL{swizawsena}{\B{swi\-zaw\-se\-na}}.

% zäfi
\hi
\HT{zAfi}{\B{\cp{zä}fi}} \I{["z\ae.fi]} \emph{adj.} modified, interfered with, no longer in a natural state, tamed, (\emph{of food}) processed, cooked.
\B{Oel yom tsngan\-it a\-zä\-fi nì\-'aw.} I only eat cooked meat.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
\B{Tsa\-ik\-ran\-ìri ta\-lun\-a new nga\-ti tspi\-vang, law lu fwa mi ke lu zä\-fi.} Since that ikran wants to kill you, it's clear it's still not tame.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/}{b}
(\ding{214}~\HL{yrr}{\B{yrr}}) \\
\textsc{Derivatives}: \ding{43} 
\on{1}.~\HL{zAfizal}{\B{zä\-fi\-zal}}, fired clay, ceramic material.

% zäfizal
\hi 
\HT{zAfizal}{\B{\cp{zä}fizal}} \I{["z\ae.fi.zal]} \emph{n.} fired clay, ceramic material. 
\B{Fì\-tsngal\-it fkol txu\-la ta zä\-fi\-zal.} This is a ceramic cup.\slash This cup is made of fired clay\slash ceramic material.~\LINK{https://naviteri.org/2025/06/vospikin-lefpom-happy-july/}{b} 
(\ding{43}~\HL{zAfi}{\B{zä\-fi}}, \HL{zal}{\B{zal}}) 

% zärìp
\hi 
\HT{zArIp}{\B{\cp{zä}rìp}} \I{["z\ae.RIp\textcorner]} \emph{vtr.} (\emph{coll. form of}) pull. 
(\ding{43}~\HL{za'ArIp}{\B{za\-'ä\-rìp}})

% zeklor
\hi 
\HT{zeklor}{\B{zek\cp{lor}}} \I{[zEk\textcorner."loR]} \emph{adj.} pleasant to the touch. 
\B{Su\-nu oer fì\-fne\-srä ta\-lu\-na lu zek\-lor.} I like this kind of cloth because it feels good.~\LINK{https://naviteri.org/2024/06/ayhapxi-tokxa-si-ayliu-alahe-parts-of-the-body-and-more/}{b}
(\ding{43}~\HL{zekwA}{\B{zek\-wä}}, \HL{lor}{\B{lor}})

% zektil
\hi 
\HT{zektil}{\B{\cp{zek}til}} \I{["zEk\textcorner.til]} \emph{n.} knuckle. 
(\ding{43}~\HL{zekwA}{\B{zek\-wä}}, \HL{til}{\B{til}}) 

% zekwä 
\hi
\HT{zekwA}{\B{\cp{zek}wä}} \I{["zEk\textcorner.w\ae{}]} \emph{n.} finger. 
\B{Tsyokx\-ìri ke lu zek\-wä a\-tsìng ki a\-mrr.} The hand doesn't have four but five digits.~\LINK{http://naviteri.org/2010/07/ting-mikyun-fte-tslivam-listening-comprehension-1-3/}{b}
\B{Hol\-pxay ay\-zek\-wä fe\-yä.} The number of their fingers.~\LINK{http://forum.learnnavi.org/language-updates/talun-taweyk-tafral-lun-oeyk-holpxay-hol/}{ln} \\
(i.) \B{tìng zekwä}, \emph{vin.} (\texttt{+}\emph{control}) touch (intentionally). \B{Rä\-'ä tìng zek\-wä oer!} Don't touch me!~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b}
\B{Lu fì\-vul su\-syang nì\-txan. Txo fko ti\-vìng zek\-wä kxakx.} This branch is very fragile. If you touch it, it'll break.~\LINK{http://naviteri.org/2015/04/some-new-words-for-may-day/}{b} 
(\ding{43}~\HL{zIm}{\B{zìm}}, \HL{'ampi}{\B{'am\-pi}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{venzek}{\B{ven\-zek}}, toe. 
\on{2}.~\HL{pxayzekwA}{\B{pxay\-zek\-wä}}, spiny whips.
\on{3}.~\HL{zeklor}{\B{zek\-lor}}, pleasant to the touch.
\on{4}.~\HL{zektil}{\B{zek\-til}}, knuckle.

% zel
\hi
\HT{zel}{\B{zel}} \I{[zEl]} \emph{conj.} then (\emph{for counterfactuals})
\B{Zun tom\-pa zì\-ye\-vup trr\-ay, zel fo srì\-ye\-vew.\slash … zel fo srew.} If it rained tomorrow, they'd do a dance.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
\B{Zun ay\-oe li\-vu tsam\-si\-yu, zel tsa\-kem ke sim\-vi.} If we were warriors, we wouldn't have done that.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
\B{Zel swey\-lu zun nga kil\-vä.} You should have gone [lit.: `it would have been better if you had gone'].~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\emph{can be left out in casual conversation if both clauses have the same tempus}) \B{Swey\-lu zun nga kil\-vä.} You should have gone [lit.: `it would have been better if you had gone'].~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{zun}{\B{zun}})

% zene
\hi
\HT{zene}{\B{\cp{ze}ne}} \I{["zE.nE]} \emph{vin., modal} must. 
\B{Slä nì\-aw\-no\-mum, ze\-ne oe 'aw\-si\-teng tì\-kang\-kem si\-vi fo\-hu.} But as you know, I must work together with them.~\LINK{http://forum.learnnavi.org/news-announcements/a-response-from-paul-frommer!/}{ln} 
\B{Nga ze\-ne ki\-vä.} You must go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Ke ze\-ne nga ki\-vä.} It's not necessary\slash obligatory that you go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zenke}{\B{zen\-ke}}, must not. 
\on{2}.~\HL{kezemplltxe}{\B{ke\-zem\-pll\-txe}}, of course.
\on{3}.~\HL{nIzen}{\B{nì\-zen}}, necessarily.

% zenke
\hi
\HT{zenke}{\B{\cp{zen}ke}} \I{["z$\cdot$EN$\cdot$.kE]} \emph{vin., modal} must not. (!\emph{irregular infix distribution with second position infixes}! \B{\cp{ze}\-näng\-ke}, \B{\cp{ze}\-nei\-ke}, \B{\cp{ze}\-nu\-ye\-ke}, \B{\cp{ze}\-na\-tse\-ke} \I{["zE.n\ae{}N.kE, "zE.nE.i.kE, "zE.nu.jE.kE, "zE.na.\t{ts}E.kE]} \LINK{http://forum.learnnavi.org/language-updates/affective-obligation-\%28zenke\%29/}{ln})
\B{Nga zen\-ke ki\-vä.} You must not go.~\LINK{http://naviteri.org/2011/04/\%E2\%80\%99a\%E2\%80\%99awa-li\%E2\%80\%99fyavi-amip\%E2\%80\%94a-few-new-expressions/}{b}
\B{Hem nge\-yä zen\-ke fko\-ru li\-vu, tsay\-hem a nga\-ru prr\-te' ke lu.} Your actions must not be to others (as) those actions that you don't like.~\LINK{http://wiki.learnnavi.org/index.php?title=Corpus\#Good_Morning_America}{wiki}

% zeng
\hi 
\HT{zeng}{\B{zeng}} \I{[zEN]} \emph{n.} crest (\emph{on the head of a bird or animal}).

% zerok
\hi
\HT{zerok}{\B{\cp{ze}rok}} \I{["zE.Rok\textcorner]} \emph{vtr.} remember.
\B{Maw\-krr la\-ye\-iu oer krr nì\-'ul fte ngi\-vop ay\-lì\-'ut sì tsay\-fne\-sä\-num\-vit a tsun fra\-por srung si\-vi fte ni\-vu\-me sì zi\-ve\-rok nì\-swey.} After that I will have more time to create words and the kinds of lessons that can help everyone best learn and remember.~\LINK{http://forum.learnnavi.org/language-updates/trr-rrtaya/}{ln}
\B{Ze\-rok aw\-ngal nì\-wotx krr\-it a pol\-txe Tseyk san Eo ay\-oeng lu txan\-a tì\-kawng sìk.} We all remember the time that Jake said, `A great evil is upon us.'~\LINK{http://naviteri.org/2016/12/tengkrr-perahem-zisit-amip-as-the-new-year-arrives/}{b}
(\ding{43}~\HL{'ok}{\B{'ok}})

% zeswa
\hi
\HT{zeswa}{\B{\cp{ze}swa}} \I{["zE.swa]} \emph{n.} \ding{96} grass.
\B{Pa\-lu\-kan\-it tso\-le\-'a, ye\-rik lopx hi\-fwo kxam\-lä ze\-swa.} Spotting a thanator, the hexapede panicked and escaped through the grass.~\LINK{http://naviteri.org/2012/07/meetings-waterfalls-and-more/}{b} 
\B{Torukìl ke yom zeswat.} A toruk doesn't eat grass.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{lezeswa}{\B{le\-ze\-swa}}, grassy. 
\on{2}.~\HL{zeswavi}{\B{ze\-swa\-vi}}, blad of grass. 
\on{3}.~\HL{yomzeswa}{\B{yom\-ze\-swa}}, graze.

% zeswavi
\hi
\HT{zeswavi}{\B{\cp{ze}swavi}} \I{["zE.swa.vi]} \emph{n.} blade of grass.
(\ding{43}~\HL{zeswa}{\B{ze\-swa}}, \HL{-vi}{\B{-vi}})

% zet 
\hi
\HT{zet}{\B{zet}} \I{[zEt\textcorner]} \emph{vtr.} treat (\emph{emotionally}), display an attitude towards (\emph{with} \ding{43}~\HL{pxel}{\B{pxel}}).
\B{Pey\-ral\-ìl zet wur\-a wu\-tsot a\-'aw\-nem pxel sngel.} Peyral treats a cool cooked meal like garbage.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Pol zo\-let oe\-yä sä\-fpìl\-it pxel pum a tì\-ngä\-zìk\-it ngop.} He treated my idea as though it created a problem.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b} 
\B{Pol ze\-ret oe\-ti pxel tu\-te a ke inan pot.} He's treating me like I don't know him.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}
\B{Va'\-rul zä\-nget ik\-ran\-it sn\-eyä pxel ha\-pxì\-tu so\-a\-i\-ä.} Va'ru treats his ikran like a member of the family.~\LINK{http://naviteri.org/2012/10/mipa-vospxi-mipa-ayliu-new-words-for-the-new-month/}{b}

% zey
\hi
\HT{zey}{\B{zey}} \I{[zEj]} \emph{adj.} special, distinct.   
\B{Kel\-ut\-ral lu fne\-ut\-ral a\-zey.} Hometree is a special kind of tree.~\LINK{http://naviteri.org/2012/06/spring-vocabulary-part-3/}{b}

% zeyko
\hi
\HT{zeyko}{\B{zey\cp{ko}}} \I{[zEj."k$\cdot$o]} \emph{vtr., inf.}\on{22} heal, fix.
\B{Ey\-wal zey\-ki\-vo ngat nì\-win.} May Eywa heal you quickly.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{'Ä'! Ira\-yo! Zey\-ko\-leio.} Oops! Thanks! Gladly fixed!~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/comment-page-1/\#comment-7689}{b}
\B{Nì\-ra\-ngal lir\-vu ay\-oe\-ru eyk\-tan a tsi\-vun srung si\-vi, pum a li\-vu por ay\-lì\-'u a\-zu\-sey\-ko.} I wish to have a leader who can help, one who has healing words.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{zo}{\B{zo}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vezeyko}{\B{ve\-zey\-ko}}, put in order, organize. 
\on{2}.~\HL{zeykoyu}{\B{zey\-ko\-yu}}, healer.

% zeykoyu
\hi 
\HT{zeykoyu}{\B{zey\cp{ko}yu}} \I{[zEj."ko.ju]} \emph{n.} healer. 
\B{Fì\-'um\-tsat to\-lìng 'e\-veng\-ur a\-spxin zey\-ko\-yul a txan\-ro\-'a.} This medicine was given to the sick child by a famous healer.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}  
(\ding{43}~\HL{zeyko}{\B{zey\-ko}})

% zeykup
\hi
\HT{zeykup}{\B{zey\cp{kup}}} \I{[zEj."k$\cdot$up\textcorner]} \emph{or} \I{[."kUp\textcorner]} \emph{vtr., inf.}\on{22} (intentionally\slash deliberately) let fall.
\B{Nge\-yä tsko\-ti zey\-kup! Set!} Drop your bow! Now!~\LINK{http://naviteri.org/2013/04/wheres-the-bathroom-and-other-useful-things/}{b}
(\ding{43}~\HL{zup}{\B{zup}}, \HL{tungzup}{\B{tung\-zup}}) 

% zevakx
\hi 
\HT{zevakx}{\B{\cp{ze}vakx}} \I{["zE.vak']} \emph{adj., ofp.} cruel. 
\B{tu\-te a\-ze\-vakx}, cruel person.~\LINK{http://naviteri.org/2020/06/mi-tanlokxe-oeya-srr-afpxamo-terrible-days-in-my-country/}{b}
(\ding{43}~\HL{tIzevakxnga'}{\B{tì\-ze\-vakx\-nga'}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIzevakx}{\B{tì\-ze\-vakx}}, cruelty. 
\on{2}.~\HL{nIzevakx}{\B{nì\-ze\-vakx}}, cruelly.

% zin
\hi 
\HT{zin}{\B{zin}} \I{[zin]} \emph{adj.} tangled. 
\B{Lä\-ngu fay\-hìng zin nì\-wotx; ke tsun sat si\-var.} Unfortunately, these threads are all tangled up; they can't be used.~\LINK{http://naviteri.org/2021/04/mipa-ayliu-mipa-sioeykting-new-words-new-explanations/}{b} (\ding{214}~\HL{kezin}{\B{ke\-zin}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{tIzin}{\B{tì\-zin}}, tanglement. 
\on{2}.~\HL{kezin}{\B{ke\-zin}}, untangled.

% zir
\hi
\HT{zir}{\B{zir}} \I{[ziR]} \emph{n.} touch, feel, texture (\emph{sensation}).
\B{Fì\-ut\-ral\-ìri ta\-ngek\-ä zir fkan vawt, hu\-fwa ke rey.} The trunk of the tree feels solid, although it's dead.~\LINK{http://naviteri.org/2013/09/on-si-salewfya-shapes-and-directions/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{kakzir}{\B{kak\-zir}}, numb. 

% zize'
\hi
\HT{zize'}{\B{zi\cp{ze'}}} \I{[zi."zEP]} \emph{n., fauna} hellfire wasp.
\B{Oe\-ri heyr tì\-sraw sä\-ngi ta\-lu\-na zi\-ze'\-ìl oet sngo\-lap tsa\-tseng.} My chest hurts because a hellfire wasp stung me there.~\LINK{http://naviteri.org/2015/03/kap-si-ayunil-saylahe-threats-dreams-and-other-things/}{b}
(\emph{a. idiom}) \B{sre fwa sngap zi\-ze'}, ``as quickly as possible'' [lit: before the hellfire wasp stings].~\LINK{http://naviteri.org/2012/07/fivospxiya-aylifyavi-amip-this-months-new-expressions-pt-2/}{b}

% zìm 
\hi
\HT{zIm}{\B{zìm}} \I{[zIm]} \emph{vtr.} feel (\texttt{-}\emph{control}).  
\B{Teng\-krr hu pa\-lu\-kan\-tsyìp uvan se\-ri zo\-lìm oel mì sa'\-leng a 'u\-ot a lu txa' sì ekx\-txu.} While playing with my cat I felt something hard and rough on his skin.~\LINK{http://naviteri.org/2012/11/renu-ayinanfyaya-the-senses-paradigm/}{b}
(\ding{43}~\HL{'ampi}{\B{'am\-pi}}, \HL{zekwA}{\B{tìng zek\-wä}})

% zìma'uyu
\hi
\HT{zIma'uyu}{\B{zì\cp{ma}'uyu}} \I{[zI."ma.Pu.ju]} \emph{n.} newcomer, s.o. who has just arrived on the scene.   
\B{Lam oer fwa tsa\-zì\-ma\-'u\-yul ke fnan tì\-tu\-sa\-ron\-it.} It seems to me that that newcomer isn't any good at hunting.~\LINK{http://naviteri.org/2011/09/\%E2\%80\%9Cby-the-way-what-are-you-reading\%E2\%80\%9D/}{b} 
(\ding{43}~\HL{za'u}{\B{za\-'u}})

% zìmtswo
\hi
\HT{zImtswo}{\B{\cp{zìm}tswo}} \I{["zIm.\t{ts}wo]} \emph{n.} sense of touch.
(\ding{43}~\HL{zIm}{\B{zìm}}, \HL{-tswo}{\B{-tswo}})

% zìsìkrr
\hi
\HT{zIsIkrr}{\B{\cp{zì}sìkrr}} \I{["zI.sI.k\s{r}]} \emph{n.} season
\B{Fra\-zì\-sì\-krr pay kil\-van\-ä nän nì\-'ul\-'ul.} Every season the river dries up more and more.~\LINK{http://naviteri.org/2012/02/trr-asawnung-lefpom-happy-leap-day/}{b}
\B{Tsay\-fa\-suk tsu\-ka\-nom lu krr\-ka fì\-zì\-sì\-krr nì\-'aw.} Those berries are available during this season only.~\LINK{http://naviteri.org/2021/03/mipa-ayliu-si-aylifyavi-new-words-and-expressions/}{b}
(\ding{43}~\HL{zIsIt}{\B{zì\-sìt}}, \HL{krr}{\B{krr}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zIskrrsom}{\B{zì\-skrr\-som}}, summer. 
\on{2}.~\HL{zIskrrwew}{\B{zì\-skrr\-wew}}, winter.
\on{3}.~\HL{zIskrrmipaw}{\B{zì\-skrr\-mi\-paw}}, spring (season).
\on{4}.~\HL{zIskrrtsawn}{\B{zì\-skrr\-tsawn}}, fall, autumn.

% zìsìt 
\hi
\HT{zIsIt}{\B{\cp{zì}sìt}} \I{["zI.sIt\textcorner]} \emph{n.} year.   
\B{Mip\-a Zì\-sìt Le\-fpom, ma fra\-po!} Happy New Year, everyone!~\LINK{http://naviteri.org/2011/01/new-year-new-vocabulary/}{b}
\B{Sìl\-pey oe tsnì fì\-zì\-sìt\-ìl ay\-nga\-ru za\-mì\-ye\-vu\-nge txan\-a fpom\-it sì fpom\-tokx\-it.} I hope this year will bring you much happiness and health.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/}{b}
\B{Sì ay\-zì\-sìt\-ä ka\-to | Sì ekong te'\-lan\-ä.} And the rhyhtm of the years | And the beats of the hearts.~\LINK{http://wiki.learnnavi.org/index.php/Weaving_Song}{wiki}
\B{Zì\-sìt\-o a\-vol ke tsä\-po\-le\-'a fo fì\-tsap.} They haven't seen each other in eight years.~\LINK{http://naviteri.org/2011/12/one-more-for-2011/}{b}
\B{Nge\-yä tì\-eyk\-tan\-ìri, tì\-slan\-ìri sì tsran\-ten fra\-to a tì\-'ey\-lan\-ìri a ka ay\-zì\-sìt nì\-wotx, sei\-yi oe ira\-yo nì\-txan.} Thank you so much for your leadership, your support, and most importantly your friendship throughout the years.~\LINK{http://naviteri.org/2012/03/spring-vocabulary-part-1/}{b}
(\emph{a. used for stating age}) 
\B{Nga\-ri so\-la\-lew pol\-pxay\-a zì\-sìt?\slash Nga\-ri so\-la\-lew zì\-sìt a\-pol\-pxay?} How old are you?~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b}
\B{Oe\-ri so\-la\-lew (so\-lew) zì\-sìt a\-pxe\-vol.} I'm 24 years old.~\LINK{http://naviteri.org/2010/09/getting-to-know-you-part-2/}{b} \\
(i.) \B{zìsì\cp{tam}} \I{[zI.sI"tam]} \emph{adv.} last year. \\
(ii.) \B{zìsì\cp{tay}} \I{[zI.sI."taj]} \emph{adv.} next year. \B{Nì\-sìl\-pey zì\-sìt\-ay!} Hopefully next year.~\LINK{http://naviteri.org/2011/07/txantsana-ultxa-mi-siatll-great-meeting-in-seattle/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zIsIkrr}{\B{zì\-sì\-krr}}, season. 
\on{2}.~\HL{zIsItsaltrr}{\B{zì\-sìt\-sal\-trr}}, (yearly) anniversary.
\on{3}.~\HL{zIsItnun}{\B{zì\-sìt\-nun}}, leap year.

% zìsìtnun
\hi 
\HT{zIsItnun}{\B{zìsìt\cp{nun}}} \I{[zI.sIt\textcorner."nun]} \emph{or} \I{[."nUn]} (\textsc{rn}: \B{zìsìt\cp{nun}} \I{[zI.sIt\textcorner."nun]}) \emph{n.} leap year. 
(\ding{43}~\HL{zIsIt}{\B{zìsìt}}, \HL{nun}{\B{nun}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zIsItnuntrr}{\B{zì\-sìt\-nun\-trr}}, leap year day.

% zìsìtnuntrr
\hi 
\HT{zIsItnuntrr}{\B{zìsìt\cp{nun}trr}} \I{[zI.sIt\textcorner."nun.t\s{r}]} \emph{or} \I{[."nUn.]} (\textsc{rn}: \B{zìsìt\cp{nun}trr} \I{[zI.sIt\textcorner."nun.t\s{r}]}) \emph{n.} leap year day. 
\B{Zì\-sìt\-nun\-trr Le\-fpom.} Happy Leap Year Day!~\LINK{https://naviteri.org/2024/02/mipa-ayliu-si-aylifyavi-niul-more-new-words-and-expressions/}{b}
(\ding{43}~\HL{zIsItnun}{\B{zì\-sìt\-nun}}, \HL{trr}{\B{trr}})

% zìsìtsaltrr
\hi 
\HT{zIsItsaltrr}{\B{zìsìt\cp{sal}trr}} \I{[zI.sIt\textcorner."{}sal.t\s{r}]} \emph{n.} (yearly) anniversary; (\emph{coll. shortened to} \ding{43}~\HL{zItsaltrr}{\B{zìt\-sal\-trr}}). 
(\ding{43}~\HL{zIsIt}{\B{zì\-sìt}}, \HL{salew}{\B{sa\-lew}}, \HL{trr}{\B{trr}})

% zìskrrmipaw
\hi 
\HT{zIskrrmipaw}{\B{zìskrr\cp{mi}paw}} \I{[zI.sk\s{r}."mi.paw]} \emph{n.} spring (season of new growth). 
(\ding{43}~\HL{zIsIkrr}{\B{zì\-sì\-krr}}, \HL{mip}{\B{mip}}, \HL{paw}{\B{paw}}; \ding{214}~\HL{zIskrrtsawn}{\B{zì\-skrr\-tsawn}})

% zìskrrsom
\hi 
\HT{zIskrrsom}{\B{zìskrr\cp{som}}} \I{[zI.sk\s{r}."{}som]} \emph{n.} summer (\emph{hot season}). 
\B{Teng\-krr le\-rok zì\-skrr\-som\-ìl tì\-'i'\-at, sìl\-pey oe, ay\-nga\-ri te'\-lan li\-vu le\-fpom ul\-te tì\-rey zi\-vaw\-prr\-te'.} While the summer nears its end, I hope your hearts are well and you enjoy life.~\LINK{http://naviteri.org/2019/08/aawa-lifya-si-lifyavi-amip-a-few-new-words-and-expressions/}{b}
(\ding{43}~\HL{zIsIkrr}{\B{zì\-sì\-krr}}, \HL{som}{\B{som}}; \ding{214}~\HL{zIskrrwew}{\B{zì\-skrr\-wew}})

% zìskrrtsawn
\hi 
\HT{zIskrrtsawn}{\B{zìskrr\cp{tsawn}}} \I{[zI.sk\s{r}."\t{ts}awn]} \emph{n.} fall, autumn. 
\B{Sìl\-pey oe, ay\-nga\-ri zì\-skrr\-tsawn sir\-va\-lew pxay\-a lrr\-tok\-hu a ta Ey\-wa.} I hope that the fall season proceeds with many smiles from Eywa for you all.~\LINK{http://naviteri.org/2018/10/aysrr-ayvospxi-ayzisikrr-days-months-seasons/}{b}
(\ding{43}~\HL{zIsIkrr}{\B{zì\-sì\-krr}}, \HL{tsawn}{\B{tsawn}}; \ding{214}~\HL{zIskrrmipaw}{\B{zì\-skrr\-mi\-paw}})

% zìskrrwew
\hi 
\HT{zIskrrwew}{\B{zìskrr\cp{wew}}} \I{[zI.sk\s{r}."wEw]} \emph{n.} winter (\emph{cold season}).  
(\ding{43}~\HL{zIsIkrr}{\B{zì\-sì\-krr}}, \HL{wew}{\B{wew}}; \ding{214}~\HL{zIskrrsom}{\B{zì\-skrr\-som}})

% zìtsaltrr
\hi 
\HT{zItsaltrr}{\B{zìt\cp{sal}trr}} \I{[zIt\textcorner."{}sal.t\s{r}]} \emph{or} \I{[zI."\t{ts}al.t\s{r}]} \emph{n.} (yearly) anniversary; (\emph{coll. form of} \ding{43}~\HL{zIsItsaltrr}{\B{zì\-sìt\-sal\-trr}}). 
\B{Zì\-tsal\-trr\-ìri tì\-mun\-txa\-yä ay\-lrr\-tok!} Happy anniversary (of your marriage)!~\LINK{http://naviteri.org/2017/12/zisit-amip-lefpom-ma-eylan-happy-new-year-friends/}{b} 

% zo
\hi
\HT{zo}{\B{zo}} \I{[zo]} \emph{vin.} be well, be intact, be as it should be, work correctly or as nature intended.
\B{Nga zo srak?} Are you well? (\emph{e.g. have you recovered from your illness\slash fall, etc.?})~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
(\emph{a. conversational expressions}) \B{'Uo ke zo srak?\slash 'Uk zo srak?} `Is something wrong?'~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}
\B{'U\-pe ke zo?} `What’s the matter?\slash What's wrong?'~\LINK{http://naviteri.org/2020/07/vospxivol-lefpom-happy-august/}{b}  \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{vezo}{\B{ve\-zo}}, be in order, be organized. 
\on{2}.~\HL{zoslu}{\B{zo\-slu}}, heal, become well, get fixed. 
\on{3}.~\HL{frawzo}{\B{fraw\-zo}}, all is well. 
\on{4}.~\HL{zeyko}{\B{zey\-ko}}, heal, fix. 

% zong
\hi
\HT{zong}{\B{zong}} \I{[zoN]} \emph{vtr.} save, defend. 
% \B{To\-ruk nì\-'aw tsun srung si\-vi fte kel\-ut\-ral\-it zi\-vong.} Only the Last Shadow can help to save Hometree. 
\B{Te\-ya si oer fwa ngal paw\-nu\-nga ay\-ioang\-it zong.} It moves me that you save injured animals.~\LINK{http://naviteri.org/2021/09/aawa-ayliu-amip-a-few-new-words/}{b} 
(\emph{a. idiom}) \B{Mak\-to zong.} ``Take care. Travel safely.'' (\emph{coll. of} \B{Mi\-vak\-to nì\-zaw\-nong})~\LINK{http://forum.learnnavi.org/language-updates/makto-zong!/}{ln} 
(\emph{b. short form of} \HL{nIzawnong}{\B{nì\-zaw\-nong}}) \emph{to answer the question} \B{Mak\-to fya\-pe?} ``How are things? How's everything going? How are you doing?'' \B{Zong.} ``Well.''~\LINK{http://naviteri.org/2020/05/tipusawm-tiuseyng-si-okvur-a-eltur-titxen-si-asking-answering-and-an-interesting-story/}{b} 
(\ding{43}~\HL{tarep}{\B{ta\-rep}}, \HL{'avun}{\B{'a\-vun}}) \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zongtseng}{\B{zong\-tseng}}, safe place, refuge.
\on{2}.~\HL{nIzawnong}{\B{nì\-zaw\-nong}}, safely.

% zongtseng
\hi
\HT{zongtseng}{\B{\cp{zong}tseng}} \I{["zoN.\t{ts}EN]} \emph{n.} safe place, refuge. 
(\ding{43}~\HL{zong}{\B{zong}}, \HL{tseng}{\B{tse\-nge}})

% zoplo / si
\hi
\HT{zoplo}{\B{zop\cp{lo}}} \I{[zop\textcorner."lo]} \textsc{i}. \emph{n.} offense, insult. 
\B{Tì\-'al lu zop\-lo a tsa\-ri ke tsun txoa li\-vu nì\-ftue.} Wastefulness is an offense that cannot easily be forgiven.~\LINK{http://naviteri.org/2014/07/tsawlultxamaw-a-ayliu-post-meetup-words/}{b} \\
\textsc{ii}. \B{zop\cp{lo} si} \I{[zop\textcorner."lo s$\cdot$i]} \emph{vin.} offend, insult (s.o.)
\B{Pol\-txe oe san ze\-ne kea unil\-tì\-ran\-yu ke zi\-va\-'u fì\-tseng fte po kel\-kur zop\-lo si\-vi sìk.} I have said no dreamwalker will come here to offend our home.~\LINK{http://forum.learnnavi.org/language-updates/little-confirmation-on-teach-and-insult/}{ln}

% zoslu
\hi
\HT{zoslu}{\B{\cp{zo}slu}} \I{["zo.sl$\cdot$u]} \emph{vin., inf.}\on{22} heal, become well, get fixed.   
\B{Fu\-r\-i\-a 'll\-ngo nge\-yä zo\-slu, 'e\-fu oe nit\-ram nì\-'aw.} I'm just happy that your hip is getting better.~\LINK{http://naviteri.org/2013/07/pximaw-tsawlultxa-just-after-the-meet-up/comment-page-1/\#comment-2484}{b}
\B{Oe\-ri nì\-'i'a tsyokx zo\-slo\-lu.} My hand is finally healed.~\LINK{http://naviteri.org/2010/07/vocabulary-update/}{b}
\B{Tì\-kxey zo\-slo\-lu.} The incorrectness has been fixed.~\LINK{http://naviteri.org/2013/12/mipa-zisit-lefpom-happy-new-year/comment-page-1/\#comment-2585}{b}
\B{Tsa\-tì\-ke\-nong\-ìri oel kxey\-eyt zo\-sley\-ko\-lu.} As for that example, I got the mistake fixed.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1903}{b}
(\ding{43}~\HL{zo}{\B{zo}}, \HL{slu}{\B{slu}}, \HL{zeyko}{\B{zey\-ko}})

% zum
\hi
\HT{zum}{\B{zum}} \I{[zum]} \emph{or} \I{[zUm]} (\textsc{rn}: \B{zùm} \I{[zUm]}) \emph{n.} object, thing (\emph{physical or tangible}).
\B{Fay\-zum lu pe\-u?} What are these things?~\LINK{http://naviteri.org/2015/11/vomuna-liu-amip-ten-new-words/}{b}
(\ding{43}~\HL{'u}{\B{'u}})

% zun
\hi
\HT{zun}{\B{zun}} \I{[zun]} \emph{or} \I{[zUn]} \emph{conj.} if (\emph{for counterfactuals})
\B{Zun tom\-pa zì\-ye\-vup trr\-ay, zel fo srì\-ye\-vew.\slash … zel fo srew.} If it rained tomorrow, they'd do a dance.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
\B{Zun ay\-oe li\-vu tsam\-si\-yu, zel tsa\-kem ke sim\-vi.} If we were warriors, we wouldn't have done that.~\LINK{http://naviteri.org/2013/04/zun-zel-counterfactual-conditionals/}{b}
\emph{a. as counterfactual}) `should have'. 
\B{(Zel) swey\-lu zun nga kil\-vä.} You should have gone [lit.: `it would have been better if you had gone']~\LINK{https://naviteri.org/2024/12/odds-and-ends/}{b}
(\ding{43}~\HL{zel}{\B{zel}})

% zung
\hi 
\HT{zung}{\B{zung}} \I{[zuN]} \emph{or} \I{[zUN]} (\textsc{rn}: \B{zùng} \I{[zUN]}) \emph{vin.} crouch. 
\B{Zo\-lung ay\-oe ne\-kll fte\-ke ay\-ioang tsi\-vun ay\-oe\-ti tsi\-ve\-'a.} We crouched down so that the animals wouldn't be able to see us.~\LINK{http://naviteri.org/2019/06/50a-liu-amip-40-new-words/}{b}

% zup 
\hi
\HT{zup}{\B{zup}} \I{[zup\textcorner]} \emph{or} \I{[zUp\textcorner]} \emph{vin.} (\emph{a.}) fall. 
\B{Po pxun\-fa pxek tsa\-krr me\-fo zup ne\-kll!} He shoved the two of them and they fell down.~\LINK{http://naviteri.org/2014/05/mipa-ayliu-mipa-aysafpil-new-words-new-ideas/}{b}
(\emph{b. of the weather, precipitation}) \B{Zup tompa\slash her\-wì.} It's raining\slash snowing.~\LINK{http://naviteri.org/2012/10/audio-and-video-learning-materials-for-navi-101/}{b}
\B{Oe\-ri ay\-som\-pì\-va sìn re\-'o var zi\-vup.} Raindrops keep falling on my head.~\LINK{http://naviteri.org/2011/04/yafkeykiri-plltxe-frapo-everyone-talks-about-the-weather/}{b}
(\emph{c. with} \ding{43}~\HL{ne}{\B{ne}}) perish. 
\B{Ke za\-syup lì\-'O\-na ne kxu\-tu a mì\-fa fu a wrr\-pa.} The l'Ona will not perish to the enemy within or the enemy without.~\LINK{http://forum.learnnavi.org/language-updates/ltasygt-with-ke-and-nga/}{g\slash ln} 
(\emph{d. proverb}) \B{Rä\-'ä zup nem\-fa tsup.} \emph{don't fall into the trap that s.o. has laid for you} [lit.: ``don't fall into the chasm''].~\LINK{https://naviteri.org/2025/04/mevosinga-liu-amip-twenty-new-words/}{b} \\
\textsc{Derivatives}: \ding{43}
\on{1}.~\HL{zeykup}{\B{zey\-kup}}, let fall (intentionally). 
\on{2}.~\HL{tungzup}{\B{tung\-zup}}, drop, let fall (unintentionally).

% zusawkrr
\hi
\HT{zusawkrr}{\B{zu\cp{saw}krr}} \I{[zu."saw.k\s{r}]} \textsc{i}. \emph{n.} future time, future.
\B{Nì\-sung, txo tsa\-kem li\-ven mì zu\-saw\-krr, ze\-ne fko si\-var kem\-lì\-'u\-vit a\-lu ‹ìyev›.} Furthermore, if that happens in the future, one must use the infix ‹\B{ìyev}›.~\LINK{http://naviteri.org/2013/01/awvea-posti-zisita-amip-first-post-of-the-new-year/comment-page-1/\#comment-1910}{b}
\B{Sìn A\-sok, Sìn Zu\-saw\-krr\-ä}, Recent and Upcoming Activities [lit.: recent activities, future's activities].~\LINK{http://naviteri.org/2013/04/sin-asok-sin-zusawkrra-recent-and-upcoming-activities/}{b}  
\B{Olo'\-ìri po\-an zu\-saw\-krr\-ur 'on so\-li.} He shaped the tribe's future.~\LINK{http://naviteri.org/2018/07/ta-sulfatu-a-ayliu-niul-more-words-from-our-experts/}{b} \\
\textsc{ii}. \emph{adv.} in the future.
\B{'Aw\-a tì\-pawm\-ì\-ri 'i\-veyng oe set;  ay\-la\-ri zu\-saw\-krr 'ay\-eyng.} Let me answer one of the questions now; the others (I'll) answer in the future.~\LINK{http://forum.learnnavi.org/language-updates/email-from-frommer-re-numbers-\%28or-the-full-number-chart!\%29/}{ln}
(\ding{43}~\HL{za'u}{\B{za\-'u}}, \HL{krr}{\B{krr}})
\\
\end{multicols}
